### 2023-03-13 docs/manuscript/paper.tex
add tex files for paper

% mnras_guide.tex
%
% MNRAS LaTeX user guide
%
% v3.1 released 11 June 2020
%
% v3.0 released 22 May 2015
% (version numbers match those of mnras.cls)
%
% Copyright (C) Royal Astronomical Society 2015
% Authors:
% Keith T. Smith (Royal Astronomical Society)

% Change log
%
% v3.0   September 2013 - May 2015
%    First version: complete rewrite of the user guide
%    Basic structure taken from mnras_template.tex by the same author

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,usenatbib,useAMS]{mnras}

%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{multicol}        % Multi-column entries in tables
\usepackage{bm}		% Bold maths symbols, including upright Greek
\usepackage{pdflscape}	% Landscape pages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%%

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
\newcommand{\kms}{\,km\,s$^{-1}$} % kilometres per second
\newcommand{\bibtex}{\textsc{Bib}\!\TeX} % bibtex. Not quite the correct typesetting, but close enough

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}

% MNRAS is set in Times font. If you don't have this installed (most LaTeX
% installations will be fine) or prefer the old Computer Modern fonts, comment
% out the following line
\usepackage{newtxtext,newtxmath}
% Depending on your LaTeX fonts installation, you might get better results with one of these:
%\usepackage{mathptmx}
%\usepackage{txfonts}

%%%%%%%%%%%%%%%%%%% TITLE PAGE %%%%%%%%%%%%%%%%%%%

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman PTA]{State-space PTA}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[Kimpson]{Kimpson$^{1}$, O'Leary, Melatos, Evans, others, etc. %
\thanks{Contact e-mail: \href{mailto:mn@ras.ac.uk}{mn@ras.ac.uk}}%
\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
\\
% List of institutions
$^{1}$Royal Astronomical Society, Burlington House, Piccadilly, London W1J 0BQ, UK}

% These dates will be filled out by the publisher
\date{Last updated 2020 June 10; in original form 2013 September 5}

% Enter the current year, for the copyright statements etc.
\pubyear{2020}

% Don't change these lines
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% Abstract of the paper
\begin{abstract}
This is an abstract
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
editorials, notices -- miscellaneous
\end{keywords}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% BODY OF PAPER %%%%%%%%%%%%%%%%%%

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage

\section{Introduction}

\begin{itemize}
	\item Introduce PTAs generally.
	\item Types of astrophysical source to be detected with PTAs.
	\item Why we want to do parameter estimation
	\item Advantages of doing this with state-space approach
\end{itemize}

\section{Model}
The intrinsic pulsar frequency $f$ evolves as (See Vargas and Melatos)
\begin{equation}
	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\gamma$ a proporionality constant, and $\xi(t)$ a white noise process that satisfies,
\begin{equation}
	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
\end{equation}
Over the timescales that we are interested in, we can express the EM spindown simply as
\begin{equation}
	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
\end{equation}  
Completely, the frequency evolution is then
\begin{equation}
	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
	\label{eq:frequency_evolution_expanded}
\end{equation}
This intrinsic frequency can be related to a measured frequency as 
\begin{equation}
	f_M = f g(\bar{\theta},t) + N_M
\end{equation}
where $g(\theta,t)$ ("measurement function")is a function of some parameters, $\bar{\theta}$ and time $t$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
\begin{equation}
	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
\end{equation}
Explicitly, it can be shown that the measurement function is,
\begin{equation} \label{eq:final}
	g(\bar{\theta},t) = 1 - \frac{1}{2} \frac{h_{ij}(t) q(t)^i q(t)^j}{1 + \bar{n} \cdot \bar{q}(t)}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}(t)) d}]
\end{equation}
There are a few terms to unpack and define in this equation. $q(t)$ is the vector that connects the Earth and the pulsar, $n$ the vector that connects the Earth and the GW source, $\Omega$ the (constant) angular frequency of the GW and $d$ the Earth-pulsar distance. $h_{ij}(t)$ is the metric perturbation due to the GW $= g_{ij} - \eta_{ij}$. It is given by
\begin{equation}
h_{ij}(t) = H_{ij} e^{(i(\Omega (\bar{n} \cdot \bar{q} - t) + \Phi_0))}
\end{equation}
for phase offset (GW phase at Earth when $t=0$) $\Phi_0$. The amplitude tensor  $H_{ij}$ is
\begin{eqnarray}
	H_{ij} = h_+ e_{ij}^+(\bar{n},\psi) + h_{\times} e_{ij}^{\times}(\bar{n},\psi)
\end{eqnarray}
where $h_{+,\times}$ are the constant amplitudes of the gravitational plane wave, and $e_{ij}^{+, \times}(\bar{n}, \psi)$ are the polarisation tensors which are uniquely defined by the principal axes of the GW. $\psi$ is the polarisation angle of the GW. \newline

\noindent Going forward, for now we will take $q(t) = q$ i..e. the pulsar locations are constant with respect to the Earth. This may have already been "done" during the barycentreing when pulsar TOAs are generated, in which case $q$ is the vector from the SSB to the pulsar  \newline

\noindent Lets review and categorise all the parameters:
\begin{equation}
	\bar{\theta} =  \bar{\theta}_{\rm PSR} + \bar{\theta}_{\rm GW} + \bar{\theta}_{\rm noise}
\end{equation}
\begin{equation}
	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]
\end{equation}
\begin{equation}
	\bar{\theta}_{\rm GW} = [h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0]
\end{equation}
\begin{equation}
	\bar{\theta}_{\rm noise} = [\sigma, \Sigma]
\end{equation}
We can also express the measurement equation generally as,
\begin{equation} 
	g(\bar{\theta},t) = 1 -A \cos(\Omega(n\cdot q - t) + \Phi_0)
\end{equation}
where 
\begin{equation}
	A =  \frac{1}{2} \frac{H_{ij} q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
\end{equation}

\subsection{References}
\label{sec:ref_list}

\bibliographystyle{mnras}
\bibliography{example} % if your bibtex file is called example.bib

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Don't change these lines
\bsp	% typesetting comment
\label{lastpage}
\end{document}

% End of mnras_guide.tex

### 2023-03-13 src/kalman_filter.jl

@@ -32,7 +32,8 @@ function kalman_filter(observations::Matrix{NF},
     x_pulsar = observations[:,1] # guess that the intrinsic frequencies is the same as the measured frequency
     x_parameters = [parameters.h, parameters.ι, parameters.δ, parameters.α, parameters.ψ, parameters.ω, parameters.Φ0] 
 
-
+    println("The input x_parameters = ")
+    println(x_parameters)
     x = [x_pulsar; x_parameters] #concatenate to get the intial state
     #P = I(L) * σm*1e9 
     
@@ -90,8 +91,10 @@ function kalman_filter(observations::Matrix{NF},
         
          
          χ1 = F_function(χ_input,dt,parameters)
+        
+
          χ1 = [χ1 χ_remain]
-         
+
 
   
 
@@ -99,18 +102,25 @@ function kalman_filter(observations::Matrix{NF},
          x_predicted, P_xx,Δx= predict(χ1, ukf_weights)
          P_xx += Q
 
+         println("x predicted")
+         println(size(x_predicted), " ", size(P_xx))
+
          #Evolve the sigma vectors according to the measurement function
          #Note we are omitting Joe's extra step recalculating sigmas here.
          #Including this steps makes the performance worse -
          # after discussion we think the two methods are equivalent 
          #χ2 = measurement_function(χ,ti,dot_product,prefactor,ω,Φ0)
          χ2 = measurement_function(χ,ti,Npulsars,PTA.q,PTA.d)
+ 
          y_predicted, P_yy,Δy= predict(χ2, ukf_weights)
+         println("y_predicted")
+         println(size(y_predicted), " ", size(P_yy))
          
 
          P_yy += R
         
         #  #Measurement update 
+        println("update")
          x,P,l = update(Δx, Δy,ukf_weights,P_xx,P_yy,observation,x_predicted,y_predicted)
 
          println(x)
@@ -254,6 +264,7 @@ function update(Δx::Matrix{NF}, Δy::Matrix{NF},W::UKF_weights, P_xx::Matrix{NF
 
     x = x_predicted + transpose(K*innovation)
     P = P_xx - K*P_yy*transpose(K)
+   
     l = likelihood(P_yy,innovation)
 
 

### 2023-03-13 docs/manuscript/paper.tex
paper corrections to fundamental equations

@@
 $h_{ij}(t)$ is the metric perturbation due to the GW $= g_{ij} - \eta_{ij}$.
-It is given by
+It is given by the familiar plane-wave relation:
 \begin{equation}
-h_{ij}(t) = H_{ij} e^{(i(\Omega (\bar{n} \cdot \bar{q} - t) + \Phi_0))}
+	h_{ij}(t) = H_{ij} e^{(i(\bar{k} \cdot \bar{x} - \Omega t + \Phi_0))}
 \end{equation}
-for phase offset (GW phase at Earth when $t=0$) $\Phi_0$. The amplitude tensor  $H_{ij}$ is
+where $\bar{k}$ is the GW 3-wavevector $=\Omega \bar{n}$ and $x = -\bar{q} t$ (photon propagates from pulsar to Earth), and for phase offset (GW phase at Earth when $t=0$) $\Phi_0$.
+We can also write this as,
+\begin{equation}
+h_{ij}(t) = H_{ij} e^{(-i\Omega t (1 + \bar{n} \cdot \bar{q} ) + \Phi_0))}
+\end{equation}
+The amplitude tensor  $H_{ij}$ is
 \begin{eqnarray}
@@
 \begin{equation}
-	\bar{\theta} =  \bar{\theta}_{\rm PSR} + \bar{\theta}_{\rm GW} + \bar{\theta}_{\rm noise}
+	\bar{\theta} =  \bar{\theta}_{\rm PSR} \cup \bar{\theta}_{\rm GW} \cup \bar{\theta}_{\rm noise}
 \end{equation}
@@
 \begin{equation} 
-	g(\bar{\theta},t) = 1 -A \cos(\Omega(n\cdot q - t) + \Phi_0)
+	g(\bar{\theta},t) = 1 -A \cos(-\Omega t (1 + n\cdot q ) + \Phi_0)
 \end{equation}
@@
 \end{equation}
+
+\cite{Vargas}
+
+\section{PTA pulsars}
 

### 2023-03-13 src/GW.jl

@@ -5,7 +5,7 @@ struct gravitational_wave{NF<:AbstractFloat}
 
     m :: Vector{NF}  
     n :: Vector{NF}  
-    Ω :: Vector{NF}  
+    n̄ :: Vector{NF}  #this is a vector in the directin of the GW
 
     Hij :: Matrix{NF} 
 
@@ -19,7 +19,7 @@ end
 function gw_variables(NF,P) #P is either a SystemParameters object or a GuessedParameters object 
 
     m,n                 = principal_axes(π/2.0 - P.δ,P.α,P.ψ)    
-    Ω                   = cross(m,n)            
+    n̄                   = cross(m,n)            
     
     hp,hx               = h_amplitudes(P.h,P.ι)                                    
     e_plus              = [m[i]*m[j]-n[i]*n[j] for i=1:3,j=1:3]
@@ -27,7 +27,7 @@ function gw_variables(NF,P) #P is either a SystemParameters object or a GuessedP
     
     Hij                 = hp .* e_plus .+ hx * e_cross
     
-    return gravitational_wave{NF}(m,n,Ω,Hij,P.ω,P.Φ0)
+    return gravitational_wave{NF}(m,n̄,Ω,Hij,P.ω,P.Φ0)
 
 end 
 
@@ -57,29 +57,29 @@ end
 
 
 
-"""
-Given the location of the GW source (θ, ϕ) and the polarisation angle (ψ)
-determine the principal axes of propagation
-Vector method
-"""
-function principal_axes(θ::Vector{NF},ϕ::Vector{NF},ψ::Vector{NF}) where {NF<:AbstractFloat}
+# """
+# Given the location of the GW source (θ, ϕ) and the polarisation angle (ψ)
+# determine the principal axes of propagation
+# Vector method
+# """
+# function principal_axes(θ::Vector{NF},ϕ::Vector{NF},ψ::Vector{NF}) where {NF<:AbstractFloat}
 
-    println("you have valled the correct method")
-    m1 = sin.(ϕ).*cos.(ψ) .- sin.(ψ).*cos.(ϕ).*cos.(θ)
[…]

### 2023-03-13 src/pulsars.jl

@@ -4,9 +4,10 @@
 struct Pulsars{NF<:AbstractFloat}
 
     f0 :: Vector{NF} 
+    ḟ0 :: Vector{NF}
     d :: Vector{NF}
     γ :: Vector{NF}
-    n :: Vector{NF} 
+    #n :: Vector{NF} 
     q :: Matrix{NF} 
     t :: Vector{NF}
 
@@ -30,9 +31,13 @@ function setup_PTA(P::SystemParameters)
 
 
     f = pulsars[:,"F0"]
+    ḟ = pulsars[:,"F1"] 
+
+
     d = pulsars[:,"DIST"]*1e3*pc/c #this is in units of s^-1
-    γ = pulsars[:,"gamma"]
-    n = pulsars[:,"n"]
+    
+    γ = 1e-13 .* (pulsars[:,"gamma"] ./ pulsars[:,"gamma"])
+   # n = pulsars[:,"n"]
 
     δ = pulsars[:,"DECJD"]
     α = pulsars[:,"RAJD"]
@@ -45,7 +50,7 @@ function setup_PTA(P::SystemParameters)
     t = collect(0:step_seconds:end_seconds)
    
 
-    return Pulsars{P.NF}(f,d,γ,n,q,t,P.σp, P.σm,step_seconds) #convert to type NF 
+    return Pulsars{P.NF}(f,ḟ,d,γ,q,t,P.σp, P.σm,step_seconds) #convert to type NF 
 
 
 end 

### 2023-03-13 src/test.jl

@@ -1,0 +1,34 @@
+
+using DifferentialEquations, Noise, Plots 
+
+#Vector of parameters
+γ = [1e-13, 1e-14, 1e-15]
+a = [300,400,500]
+b = [-1e-15,2e-15,3e-15]
+
+#Initial conditions on y 
+y0 = [300,400,500]
+
+σ = 1e-20
+
+#Define the ODEs 
+
+f(du,u,p,t) = (du .= -γ.*u .+ γ.*(a .+ b*t) .+ b)
+g(du,u,p,t) = (du .= σ)
+
+#Specify the noise process to use 
+noise = WienerProcess(0., 0.) # WienerProcess(t0,W0) where t0 is the initial value of time and W0 the initial value of the process
+
+#Time over which to integrate 
+t = range(0.0, 3e8, length=500)
+tspan = (first(t),last(t))
+
+#Setup SDE problem and solve
+prob = SDEProblem(f,g,y0,tspan,tstops=t,noise=noise)
+solution = solve(prob,EM())
+
+
+i = 1
+solution_i = solution[i,:]
+plt = plot(t,solution_i)
+display(plt)

### 2023-03-14 docs/manuscript/paper.tex
save changes before checkout

@@
 \section{Model}
-The intrinsic pulsar frequency $f$ evolves as (See Vargas and Melatos)
+We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \cite{Vargas}.
+The frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
 \begin{equation}
@@
 The amplitude tensor  $H_{ij}$ is
-\begin{eqnarray}
+\begin{equation}
 	H_{ij} = h_+ e_{ij}^+(\bar{n},\psi) + h_{\times} e_{ij}^{\times}(\bar{n},\psi)
-\end{eqnarray}
+\end{equation}
 where $h_{+,\times}$ are the constant amplitudes of the gravitational plane wave, and $e_{ij}^{+, \times}(\bar{n}, \psi)$ are the polarisation tensors which are uniquely defined by the principal axes of the GW.
@@
 
-\noindent Lets review and categorise all the parameters:
+\noindent Bringing this together, and adopting a trigonometric form of the equations, we can express the measurement equation as
+
+\begin{equation} 
+	g(\bar{\theta},t) = 1 -A \cos(-\Omega t (1 + n\cdot q ) + \Phi_0)
+\end{equation}
+where 
+\begin{equation}
+	A =  \frac{1}{2} \frac{H_{ij} q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - \cos\left( {\Omega(1+\bar{n} \cdot \bar{q}) d}\right)]
+\end{equation}
+
+\subsection{Parameters of the model}
+\noindent Lets review and categorise all the parameters the the above model.
+We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar, the GW parameters and the noise parameters i.e.
 \begin{equation}
@@
 We can also express the measurement equation generally as,
-\begin{equation} 
-	g(\bar{\theta},t) = 1 -A \cos(-\Omega t (1 + n\cdot q ) + \Phi_0)
-\end{equation}
-where 
-\begin{equation}
-	A =  \frac{1}{2} \frac{H_{ij} q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
-\end{equation}
-
-\cite{Vargas}
 

### 2023-03-14 src/GW.jl

@@ -27,7 +27,7 @@ function gw_variables(NF,P) #P is either a SystemParameters object or a GuessedP
     
     Hij                 = hp .* e_plus .+ hx * e_cross
     
-    return gravitational_wave{NF}(m,n̄,Ω,Hij,P.ω,P.Φ0)
+    return gravitational_wave{NF}(m,n,n̄,Hij,P.ω,P.Φ0)
 
 end 
 
@@ -118,30 +118,68 @@ end
 
 
 
-function gw_prefactor(n̄:: Vector{NF},q::Matrix{NF},Hij::Matrix{NF},ω::NF, d::Vector{NF}) where {NF<:AbstractFloat}
+# function gw_prefactor(n̄:: Vector{NF},q::Matrix{NF},Hij::Matrix{NF},ω::NF, d::Vector{NF}) where {NF<:AbstractFloat}
 
-    dot_product  = [NF(1.0) .+ dot(n̄,q[i,:]) for i=1:size(q)[1]] 
-    hbar         = [sum([Hij[i,j]*q[k,i]*q[k,j] for i=1:3,j=1:3]) for k=1:size(q)[1]] # Size Npulsars. Is there a vectorised way to do this?
-    ratio        = hbar ./ dot_product
-    Hcoefficient = NF(1.0) .- cos.(ω.*d.*dot_product)
-    prefactor    = NF(0.5).*ratio.*Hcoefficient
+#     dot_product  = [NF(1.0) .+ dot(n̄,q[i,:]) for i=1:size(q)[1]] 
+#     hbar         = [sum([Hij[i,j]*q[k,i]*q[k,j] for i=1:3,j=1:3]) for k=1:size(q)[1]] # Size Npulsars. Is there a vectorised way to do this?
+#     ratio        = hbar ./ dot_product
+#     Hcoefficient = NF(1.0) .- cos.(ω.*d.*dot_product)
+#     prefactor    = NF(0.5).*ratio.*Hcoefficient
+
+#     return prefactor,dot_product
+
+# end 
+
+
+
+
+function gw_prefactor(Ω:: Vector{NF},q::Matrix{NF},Hij::Matrix{NF},ω::NF, d::Vector{NF}) where {NF<:AbstractFloat}
+
+
+
+    dot_product = [NF(1.0) .+ dot(Ω,q[i,:]) for i=1:size(q)[1]] 
[…]

### 2023-03-15 docs/manuscript/paper.tex
some linearisation work

@@
 There are a few terms to unpack and define in this equation.
-$q(t)$ is the vector that connects the Earth and the pulsar, $n$ the vector that connects the Earth and the GW source, $\Omega$ the (constant) angular frequency of the GW and $d$ the Earth-pulsar distance.
+$q(t)$ is the vector from the Earth to the pulsar, $n$ the propagation direction of the GW, $\Omega$ the (constant) angular frequency of the GW and $d$ the Earth-pulsar distance.
 $h_{ij}(t)$ is the metric perturbation due to the GW $= g_{ij} - \eta_{ij}$.
@@
 \section{PTA pulsars}
+In order to proceed and explore how well this state-space formulation works, we will need to specify a selection of pulsars to make up our PTA.
+We will take the 47 pulsars that make up the NANOGrav PTA.
+For each pulsar, we need to specify the complete set of $\bar{\theta}_{\rm PSR}$ as well as $\sigma$.
+The parameters $f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d$ are straightforward to set and can be read directly from  the current "present day" best estimates of the pulsar frequency, derivative and distance.
+
+\begin{figure}
+	\includegraphics[width=0.8\columnwidth]{images/pulsars}
+	\caption{Spatial distribution and distances of NANOGrav pulsars}
+	\label{fig:pulsar_distrib}
+\end{figure}
+
+\section{Methods}
+
+There are two potential methods we can take
+
+\begin{enumerate}
+	\item Let the states just be the frequencies $\bar{x} = (f)$.
+We have a linear equation for both the state evolution and the measurement matrix.
+Both of these depend on the parameters $\theta$.
+Use the likelihood output by a standard Kalman filter to run a nested sampler and try to recover $\theta$.
+	\item Let the states be the frequencies and all the unknown parameters $\bar{x} = (f, \bar{\theta}_{\rm GW},\bar{\theta}_{\rm PSR},\bar{\theta}_{\rm noise})$.
+Our state evolution is now linear but our measurement matrix is non-linear - a function of the states(parameters).
+Use a non-linear estimator such as a EKF/UKF to try to recover all of the states at once.
+\end{enumerate}
+
+\subsection{Method 1}
 

### 2023-03-15 src/GW.jl

@@ -31,6 +31,21 @@ function gw_variables(NF,P) #P is either a SystemParameters object or a GuessedP
 
 end 
 
+function gw_variables(h::NF,ι::NF,δ::NF,α::NF,ψ::NF) where {NF<:AbstractFloat} 
+
+    m,n                 = principal_axes(π/2.0 - δ,α,ψ)    
+    n̄                   = cross(m,n)            
+    
+    hp,hx               = h_amplitudes(h,ι)                                    
+    e_plus              = [m[i]*m[j]-n[i]*n[j] for i=1:3,j=1:3]
+    e_cross             = [m[i]*n[j]-n[i]*m[j] for i=1:3,j=1:3]
+    
+    Hij                 = hp .* e_plus .+ hx * e_cross
+    
+    return m,n,n̄,Hij
+
+end 
+
 
 
 """
@@ -55,40 +70,6 @@ function principal_axes(θ::NF,ϕ::NF,ψ::NF) where {NF<:AbstractFloat}
 
 end 
 
-
-
-# """
-# Given the location of the GW source (θ, ϕ) and the polarisation angle (ψ)
-# determine the principal axes of propagation
-# Vector method
-# """
-# function principal_axes(θ::Vector{NF},ϕ::Vector{NF},ψ::Vector{NF}) where {NF<:AbstractFloat}
-
-#     println("you have valled the correct method")
-#     m1 = sin.(ϕ).*cos.(ψ) .- sin.(ψ).*cos.(ϕ).*cos.(θ)
-#     m2 = -(cos.(ϕ).*cos.(ψ) .+ sin.(ψ).*sin.(ϕ).*cos.(θ))
-#     m3 = sin.(ψ).*sin.(θ)
-#     m = [m1 m2 m3]
-    
-
-#     n1 = -sin.(ϕ).*sin.(ψ) .- cos.(ψ).*cos.(ϕ).*cos.(θ)
-#     n2 = cos.(ϕ).*sin.(ψ) .- cos.(ψ).*sin.(ϕ).*cos.(θ)
-#     n3 = cos.(ψ).*sin.(θ)
-#     n = [n1 n2 n3]
-
-#     return m,n
-
-
-# end 
-
-
-
-
-
-
-
-
-
 """
 Given the strain h and the inclination ι, get the h+ and hx components
 """
@@ -102,81 +83,24 @@ function h_amplitudes(h::NF,ι::NF) where {NF<:AbstractFloat}
 
 end 
 
+"""
[…]

### 2023-03-15 src/StateSpacePTA.jl

@@ -8,7 +8,7 @@ using LinearAlgebra,DataFrames,CSV,DifferentialEquations, Noise,Plots,Statistics
 
 
 #Exports
-export UKF, plotter
+export KalmanFilter, plotter
 
 
 
@@ -20,7 +20,10 @@ include("GW.jl")
 include("guessed_parameters.jl")
 include("synthetic_data.jl")
 include("model.jl")
-include("kalman_filter.jl")
+#include("unscented_kalman_filter.jl")
+#include("extended_kalman_filter.jl")
+include("linear_kalman_filter.jl")
+
 include("plotting.jl")
 include("run.jl")
 

### 2023-03-15 src/linear_kalman_filter.jl

@@ -1,0 +1,129 @@
+
+
+
+"""
+Given some data recover the state and determine the likelihood
+"""
+function KF(observations::Matrix{NF},
+                     PTA::Pulsars,
+                     parameters::GuessedParameters,
+                     model::Symbol
+                      ) where {NF<:AbstractFloat}
+
+   
+
+    @info "Running the Kalman filter for the measurement model defined via: ", model
+
+    @unpack q,dt,t = PTA # PTA does have some parameters that ultimatley we might want to estimate . 
+
+    @unpack γ,σp,σm = parameters 
+
+    #Set the dimension of the state space 
+    Npulsars = size(observations)[1]
+    L = Npulsars 
+    
+    
+    N = size(observations)[2]     # number of timesteps
+    @info "Size of the state space is: ", L 
+    @info "Number of observations is : ", N
+
+
+    #Get the 0th order frequencies and reshape them 
+    #These are the PSR frequencies given by ANTF
+    #f0 = reshape(f0,(1,size(f0)[1])) #change f0 from a 1D vector to a 2D matrix
+
+    #Initialise x and P
+    x = observations[:,1] # guess that the intrinsic frequencies is the same as the measured frequency
+    #x_GW_parameters = [parameters.h, parameters.ι, parameters.δ, parameters.α, parameters.ψ, parameters.ω, parameters.Φ0] # guess the GW parameters 
+
+
+   # x = [x_pulsar_frequencies; x_GW_parameters] #concatenate to get the intial state
+    #P = I(L) * σm*1e9 
+    
+    tmp_sigma = NF(1e-3)
+    P = diagm(fill(tmp_sigma ,L)) #maybe we want the uncertainty in the frequencies and the uncertainty in the parameters to be different?
+    # #Initialise the weights for thexite UKF
+
+
+    #Calculate the time-independent Q-matrix
+    Q_function(γ,σp,dt)
+    
+
+
+    #Calculate measurement noise matrix
+    R = R_function(Npulsars,σm)
+
+    #Initialise an array to hold the results
+    x_results = zeros(NF,N,L)
+
+    #Initialise a likelihood variable
+    likelihood = NF(0.0)
+
+    #Set what measurement model to use
+    if model == :GW
+        measurement_function = H_function
+    elseif model == :null
+        measurement_function = null_function
+    else
[…]

### 2023-03-15 src/model.jl

@@ -4,8 +4,13 @@
 Transition function which takes the state of sigma points and advances
 by dt using a Euler step.
 """
-function F_function(χ::Matrix{NF},dt::NF,θ̂::GuessedParameters) where {NF<:AbstractFloat}
-    df = -θ̂.γ .*χ.^θ̂.n
+function F_function(χ::Matrix{NF},dt::NF) where {NF<:AbstractFloat}
+
+
+
+
+
+    df = -θ̂.γ #.* (χ 
   
     return χ .+ dt .* df 
 end 
@@ -15,117 +20,19 @@ end
 """
 Measurement function which takes the state and returns the measurement
 """
-function H_function(χ::Matrix{NF},t::NF,Npulsars::Int64,q::Matrix{NF},d::Vector{NF}) where {NF<:AbstractFloat}
+function H_function(parameters,t,q) where {NF<:AbstractFloat}
 
+    println("this is the H function")
 
-    frequencies = χ[:,1:Npulsars]           # sigma vector of frequencies
-    parameters = χ[:,Npulsars+1:Npulsars+7] # sigma vector of frequencies
-    J = size(χ)[1] #2L+1
-    
-    #Ω = parameters[:,6] #size 109 
+    @unpack h,ι,δ,α,ψ,ω,Φ0,d = parameters 
 
-    h = parameters[:,1]
-    ι = parameters[:,2]
-    δ = parameters[:,3]
-    α = parameters[:,4]
-    ψ = parameters[:,5]
-    ω = parameters[:,6]
-    Φ0= parameters[:,7]
+    m,n,n̄,Hij = gw_variables(h,ι, δ, α, ψ)
 
-    m,n                 = principal_axes(π/NF(2.0) .- δ,α,ψ) 
-    Ω                   = [cross(m[i,:],n[i,:]) for i=1:J]
+    prefactor,dot_product = gw_prefactor(n̄,q,Hij,ω,d)
 
-    hp,hx               = h_amplitudes(h,ι)                                    
+    GW_factor = gw_modulation(t, ω,Φ0,prefactor,dot_product)
 
-    
-
-    Hij_sigma = zeros(NF,3,3,J) #one 3x3 matrix for every sigma vector!
-    for i=1:J #for every sigma vector 
-        mi = m[i,:]
-        ni = n[i,:]
-
-        e_plus  = [mi[j]*mi[k]-ni[j]*ni[k] for j=1:3,k=1:3]
[…]

### 2023-03-15 src/run.jl

@@ -1,9 +1,9 @@
 """
 solution,model = orbit(NF,kwargs...)
 
-Runs RelativisticDynamics.jl with number format `NF` and any additional parameters in the keyword arguments
+Runs StateSpacePTA.jl with number format `NF` and any additional parameters in the keyword arguments
 `kwargs...`. Any unspecified parameters will use the default values as defined in `src/system_parameters.jl`."""
-function UKF(::Type{NF}=Float64;              # number format, use Float64 as default
+function KalmanFilter(::Type{NF}=Float64;              # number format, use Float64 as default
            kwargs...                        # all additional non-default parameters
            ) where {NF<:AbstractFloat}
 
@@ -23,7 +23,15 @@ function UKF(::Type{NF}=Float64;              # number format, use Float64 as de
     plotter(PTA.t,state,measurement,nothing,nothing,4)
 
     #θ̂ = guess_parameters(PTA,P)
-   # model_state_predictions,model_likelihood = kalman_filter(measurement,PTA,θ̂,:GW)
+
+    #println(θ̂)
+
+    #model_state_predictions,model_likelihood = KF(measurement,PTA,θ̂,:GW)
+
+    #model_state_predictions,model_likelihood = EKF(measurement,PTA,θ̂,:GW)
+
+    #model_state_predictions,model_likelihood = UKF(measurement,PTA,:GW)
+
     
     # null_state_predictions,null_likelihood = kalman_filter(measurement,PTA,θ̂,:null)
 

### 2023-03-16 docs/manuscript/paper.tex
explore BAT.jl

@@
 
+In the case where the states are just the intrinsic pulsar frequencies, we can write the ODEs in matrix form as
+\begin{equation}
+	d\bar{X} = \bar{A} \bar{X} dt + \bar{N}(t) dt + \bar{\Sigma} d\bar{B}(t) 
+\end{equation}
+where $\bar{A} = \text{diag}(\gamma_1, \gamma_2,...)$,  $\bar{X} = \text{diag}(f_1, f_2,...)$, $\bar{N} =\text{diag}(\gamma_1[a_1+b_1t] + b_1, ...)$
+and we have let $a = f_{EM}(0)$, $b = \dot{f}_{\rm EM}(0)$.
+
 \subsection{References}

### 2023-03-16 src/linear_kalman_filter.jl

@@ -14,39 +14,30 @@ function KF(observations::Matrix{NF},
 
     @info "Running the Kalman filter for the measurement model defined via: ", model
 
-    @unpack q,dt,t = PTA # PTA does have some parameters that ultimatley we might want to estimate . 
-
-    @unpack γ,σp,σm = parameters 
+    @unpack q,dt,t = PTA         #Get the parameters related to the PTA 
+    @unpack γ,σp,σm = parameters #Get some of the parameters that we want to infer 
 
     #Set the dimension of the state space 
     Npulsars = size(observations)[1]
     L = Npulsars 
-    
-    
     N = size(observations)[2]     # number of timesteps
     @info "Size of the state space is: ", L 
     @info "Number of observations is : ", N
 
 
-    #Get the 0th order frequencies and reshape them 
-    #These are the PSR frequencies given by ANTF
-    #f0 = reshape(f0,(1,size(f0)[1])) #change f0 from a 1D vector to a 2D matrix
 
     #Initialise x and P
     x = observations[:,1] # guess that the intrinsic frequencies is the same as the measured frequency
-    #x_GW_parameters = [parameters.h, parameters.ι, parameters.δ, parameters.α, parameters.ψ, parameters.ω, parameters.Φ0] # guess the GW parameters 
+    P = I(L) * σm*1e9 
+    #println(σm)
+    #P = I(L) * 1e-15
 
-
-   # x = [x_pulsar_frequencies; x_GW_parameters] #concatenate to get the intial state
-    #P = I(L) * σm*1e9 
-    
-    tmp_sigma = NF(1e-3)
-    P = diagm(fill(tmp_sigma ,L)) #maybe we want the uncertainty in the frequencies and the uncertainty in the parameters to be different?
-    # #Initialise the weights for thexite UKF
-
+    #tmp_sigma = NF(1e-3)
+    #P = diagm(fill(tmp_sigma ,L)) #maybe we want the uncertainty in the frequencies and the uncertainty in the parameters to be different?
+   
 
     #Calculate the time-independent Q-matrix
-    Q_function(γ,σp,dt)
+    Q = Q_function(γ,σp,dt)
     
 
 
@@ -71,25 +62,23 @@ function KF(observations::Matrix{NF},
 
     #Update step first, then iterate over remainders
[…]

### 2023-03-16 test/bat_test.jl

@@ -1,0 +1,52 @@
+using Random, LinearAlgebra, Statistics, Distributions, StatsBase
+using BAT, IntervalSets
+
+
+data = vcat(
+    rand(Normal(-1.0, 0.5), 500),
+    rand(Normal( 2.0, 0.5), 1000)
+)
+
+hist = append!(Histogram(-2:0.1:4), data)
+
+function fit_function(p::NamedTuple{(:a, :mu, :sigma)}, x::Real)
+    p.a[1] * pdf(Normal(p.mu[1], p.sigma), x) +
+    p.a[2] * pdf(Normal(p.mu[2], p.sigma), x)
+end
+
+
+true_par_values = (a = [500, 1000], mu = (-1.0, 2.0), sigma = 0.5)
+
+
+likelihood = let h = hist, f = fit_function
+    # Histogram counts for each bin as an array:
+    observed_counts = h.weights
+
+    # Histogram binning:
+    bin_edges = h.edges[1]
+    bin_edges_left = bin_edges[1:end-1]
+    bin_edges_right = bin_edges[2:end]
+    bin_widths = bin_edges_right - bin_edges_left
+    bin_centers = (bin_edges_right + bin_edges_left) / 2
+
+    params -> begin
+        # Log-likelihood for a single bin:
+        function bin_log_likelihood(i)
+            # Simple mid-point rule integration of fit function `f` over bin:
+            expected_counts = bin_widths[i] * f(params, bin_centers[i])
+            logpdf(Poisson(expected_counts), observed_counts[i])
+        end
+
+        # Sum log-likelihood over bins:
+        idxs = eachindex(observed_counts)
+        ll_value = bin_log_likelihood(idxs[1])
+        for i in idxs[2:end]
+            ll_value += bin_log_likelihood(i)
+        end
+
+        # Wrap `ll_value` in `LogDVal` so BAT knows it's a log density-value.
+        return LogDVal(ll_value)
+    end
+end
+
+likelihood(true_par_values)

### 2023-04-02 docs/manuscript/paper.tex
assorted runs

@@
 % If you need two or more lines of authors, add an extra line using \newauthor
-\author[Kimpson]{Kimpson$^{1}$, O'Leary, Melatos, Evans, others, etc. %
+\author[Kimpson]{Kimpson$^{1}$, Melatos,O'Leary, , Evans, others, etc. %
 \thanks{Contact e-mail: \href{mailto:mn@ras.ac.uk}{mn@ras.ac.uk}}%
@@
 
+The detection of high frequency ($\sim 1$Hz) gravitational waves (GWs) from coalescing BH binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
+Gravitational radiation from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2030$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper} especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
+Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
+The foremost technique for the detection of GWs in this nano-Hz regime is a via timing an ensemble of milliseconds pulsars; a pulsar timing array (PTA) \citep{2021hgwa.bookE...4V}.
+The presence  of a nano-Hz gravitational wave will influence the propagation of the pulsar radio beacon and consequently will imprint on the pulsar timing signal.
+By measuring the modulation of the received pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs.
+\newline
+
+\noindent Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}.
+These previously disparate efforts have now been joined in international collaboration, along with a number of newer PTAs, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}.
+The primary target of PTA observations is the inspiral of supermassive black hole binaries (SMBHBs) with masses $\sim 10^7 M_{\odot}$.
+These signals from SMBHBs can be broadly classified into either deterministic or stochastic.
+For the former, sufficiently bright and near binaries may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated.
+For the latter, the  incoherent superposition of multiple weaker SMBHBs sources leads to a stochastic background detectable at nano-Hz frequencies.
+Other potential sources for PTAs include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}, but the deterministic and stochastic GW signals from SMBHBs remain the primary targets.
+\newline
+
+\noindent The detection of loud, resolved sources with a PTA typically involves a parametrised model for the pulsar timing residuals induced by the modulation of the pulsar signal by a GW.
+One can then search for evidence that this model describes the data via the usual Bayesian likelihood techniques, and try to estimate the parameters of the model \citep[e.g.][]{Babak2016}.
+For detecting the stochastic background the approach is different; one measures the correlation in pulsar timing residuals between any two pair of pulsars.
+The presence of a GW induces a characteristic correlation function as a function of the angular separation between the pulsars; the Hellings-Downs curve \citep{Hellings}.
+For both classes of source, detection of GW signals in the timing residuals of a PTA is a challenging enterprise, and currently neither a stochastic background nor an individually resolved source has yet been detected \citep{2022MNRAS.510.4873A, 10.1093/nsr/nwx126}.
+\newline
+
+\noindent Motivated by the difficulties faced by the classic PTA analysis methods, in this work we present a novel approach to formulate PTA analysis and GW detection as a state-space problem.
+This approach enables the state-space evolution to be tracked optimally, and for a specific realisation of the pulsar process noise (i.e.
+the spin wandering of the pulsar) infer both the system parameters and the detectability of the GW signal.
+
+For this initial work we will focus exclusively on resolved, monochromatic GW sources. Discussion on
+
+This paper is organised as follows.
+
+In section 2 we will briefly outline the response 
+In section 3 we will then construct a state space form for determinsitic signals (3.1) outline the mehtods for expectaiton likelihood (3.2) and apply this state space kalma nform
+
+In section 4 we will extend to stochastic sources...? OR we will leave thus for a second paper...
+
+PTA sources can be braodly characterised into two main buckets: determinsitic and stocahstic. Derterminisitc sources are loud isngle
+
 \begin{itemize}
@@
 
-\section{Model}
-We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \cite{Vargas}.
-The frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
+\section{Evolution of the pulsar frequency}
+We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \citep{Vargas}.
+Within this model, $f$ evolves according to a combination of both deterministic torques (i.e.
+electromagnetic spin-down) and stochastic torques (i.e.
+`spin wandering', achromatic variations in the pulse TOA intrinsic to the star).
+Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
 \begin{equation}
@@
 \end{equation}
-where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\gamma$ a proporionality constant, and $\xi(t)$ a white noise process that satisfies,
+where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\gamma$ a proportionality constant, and $\xi(t)$ a white noise process that satisfies,
 \begin{equation}
@@
 \end{equation}
-Over the timescales that we are interested in, we can express the EM spindown simply as
+For PTA analysis, we are concerned with timescales on the order or years. Consequently, we can express the EM spindown straightforwardly as
 \begin{equation}
@@
 \end{equation}  
-Completely, the frequency evolution is then
+Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
 \begin{equation}
@@
 \end{equation}
+
+\section{Modulation of pulsar frequency due to a GW}
+In the presence of a GW, the $f(t)$ measured by an observer on Earth is different from that measured by an observer in the local NS reference frame.
+We want to determine how the GW influences the received frequency
+
+\subsection{Plane GW perturbation}
+We take a gravitational plane wave that perturbs a background Minkowski spacetime as
+\begin{equation}
+	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
+\end{equation}
+for spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
+We emphasise that $\Omega$ has no time dependence - we are concerned solely with monochromatic sources.
+Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{This point is important, since the phase offset is then the same between multiple pulsars.
+See also Melatos 2022 PT12}.
+The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
+\begin{align}
+	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
+\end{align}
+where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave. The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
+\begin{equation}
+	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
+\end{equation}
+\begin{equation}
+		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
+\end{equation}
+which are in turn specified via the location of the GW source on the sky (via $\theta$, $\phi$ coordinates) and the polarisation angle $\psi$
+
+\begin{align}
+	\vec{m} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
+	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
+	& +(\sin \psi \sin \theta) \hat{z} \\
+	\vec{n} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
+	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
+	& +(\cos \psi \sin \theta) \hat{z}
+\end{align}
+
+\subsection{Pulse frequency as a photon}
+\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$.
+Generally, the frequency of a photon with 4-momentum $p_{\mu}$ recorded by an observer with 4-velocity $u^{\mu}$ is
+\begin{equation}
+	\nu = p_{\alpha} u^{\alpha}
+\end{equation}
+\noindent We consider both our emitter and receiver to be stationary, such that  
+\begin{equation}
+	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
+\end{equation}
+\noindent Consequently the frequency can be identified with the temporal component of the covariant 4-momentum,
+\begin{equation}
+	\nu = p_t
+\end{equation}
+\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
+\begin{equation}
+	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
+\end{equation}
+\noindent where the overdot denotes a derivative w.r.t.
+$t$.
+Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.
+\footnote{\textbf{TK: See also e.g.
+Maggiore, Meltaos who takes the same approach...}} \newline
+
+\noindent To complete our expression, we now just need to determine $\dot{p}_t$ and integrate it.
+
+\subsection{Hamiltonian Mechanics}
+The Hamiltonian in covariant notation can be written as 
+\begin{equation}
+	H(x^{\mu}, p_{\mu}) = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
+\end{equation}
+\noindent which if we substitute in our expression for the perturbed metric is
+\begin{equation}
+	H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
+\end{equation}
+\noindent Now, Hamilton's equations are
+\begin{equation}
+	\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
+\end{equation}
+
+\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
+\begin{equation}
+	\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t))+\Phi_0}
+\end{equation}
+\footnote{This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset.}
+
+\noindent Therefore the derivative w.r.t coordinate time $t$ is,
+\begin{equation}
+	\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
+\end{equation}
+\noindent To make contact with Melatos 2022 it will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar \footnote{similarly $\bar{x}(t) = -\bar{q}t$} and $\omega$ is the \textit{constant} photon angular frequency.
+Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer (see Melatos 2022, PT16).
+\newline
+
+\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.  \newline
+
+\noindent Bringing this all together we can write $\dot{p}_t$ in a condensed form as,
+\begin{equation}
+	\dot{p}_t = A e^{i \gamma t + \Phi_0}
+\end{equation}
+where 
+\begin{equation}
+	\gamma = -\Omega (1 + \bar{n}\cdot \bar{q}) 
+\end{equation}
+\footnote{compare with Melatos 2022, PT11}
+\noindent and
+\begin{equation}
+	A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
+\end{equation}
+where the $\omega$ term in the numerator results from a $\omega^2$ terms that comes from the $p^i p^j$ term and a $\omega^{-1}$ term from $(p^t)^{-1}$.
+\subsection{Performing the integral}
+
+The frequency shift experienced by the observer relative to the source due to a GW is then
+\begin{align}
+	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = A \int_{t = \tau - d}^{t=\tau} e^{i \gamma t + \Phi_0}dt  \\
+	=\frac{-iA}{\gamma} e^{i \gamma \tau}e^{\Phi_0}[1 - e^{-i \gamma d}] 
+	\label{eq:concise}
+\end{align}
+
+\subsection{Explicit expression and comparison with Melatos 22}
+
+We can expand the concise expression of Eq. \ref{eq:concise} as,
+
+\begin{equation} \label{eq:final}
+	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{H_{ij}q^i q^j  }{(1 + \hat{n} \cdot \hat{q})}e^{-i \Omega \tau(1+\bar{n} \cdot \bar{q}) + \Phi_0} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+\end{equation}
+This expression is equivalent to Melatos 2022, PT13 where the dependence on time $t$ is made explicit. Using the Melatos notation where $h_{ij} = g_{ij} - \eta_{ij}$
+
+\begin{equation}
+	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{h_{ij} q^i q^j  }{(1 + \hat{n} \cdot \hat{q})} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+\end{equation}
+
+\newpage
+\section{State-space structure} \label{sec3}
+\noindent The general structure of a Kalman state-space problem is
+\begin{equation}
+	\dot{x} = f(x) + w
+\end{equation}
+where $x$ are the state variables, $f()$ a non-linear function of the states, and $w$ a stochastic zero-mean process.
+The process noise matrix $Q = E(w w^T)$.
+The states are related to the measurement $z$ via a non-linear measurement function $h()$
+\begin{equation}
+	z = h(x) + v
+\end{equation}
+where $v$ is a stochastic zero-mean process and measurement noise matrix $R = E(v v^T)$. \newline
+
+\noindent This structure maps onto the preceding equations nicely. Our hidden state is just the intrinsic pulsar frequency, which evolves as,
+\begin{eqnarray}
+	\dot{f}_P = -\gamma f_P^n + \xi(t)
+\end{eqnarray}
+where $n$ is a braking index (typical values $\sim 1-5$, we take a canonical value $=3$) \footnote{See A.
+Vargas talk re anomalous braking indices.
+P.S.
+is there an arXiv preprint?} $\gamma$ a proportionality constant, and $\xi(t)$ represents a stochastic white noise process.
+Equation \ref{eq:final} can be used to relate the intrinsic pulsar frequency to that measured by an observer on Earth:
+
+\begin{eqnarray}
+	f_M = f_p(1 - X) + N_M
+\end{eqnarray}
+where $X$ is the RHS of Eq.
+\ref{eq:final} and $N_M$ a Gaussian measurement noise.
+An example of the evolution of the state and the measurement frequencies for our PTA pulsars can be seen in Fig.
+\ref{fig:mean and std of nets}
+
+\section{Discretisation}
+
 This intrinsic frequency can be related to a measured frequency as 
@@
 
+\section{Kalman filtering}
+
 \section{Methods}

### 2023-04-02 py_src/bilby_wrapper.py

@@ -23,7 +23,7 @@ class BilbyLikelihood(bilby.Likelihood):
         return ll
             
 
-def BilbySampler(KalmanFilter,init_parameters,priors):
+def BilbySampler(KalmanFilter,init_parameters,priors,label,outdir):
    
     
     likelihood = BilbyLikelihood(KalmanFilter,init_parameters)
@@ -31,9 +31,9 @@ def BilbySampler(KalmanFilter,init_parameters,priors):
  
     # #Run the sampler
     print("RUN THE SAMPLER")
-    result = bilby.run_sampler(likelihood, priors, label = "JIT1_omega3",outdir=".",
+    result = bilby.run_sampler(likelihood, priors, label = label,outdir=outdir,
                             sampler ='dynesty',check_point_plot=False,
-                            sample='rwalk', walks=10, npoints=100,dlogz=1.0,
-                            npool=6,plot=True,resume=False)
+                            sample='rwalk', walks=10, npoints=400,dlogz=0.10,
+                            npool=10,plot=False,resume=True)
 
     return result

### 2023-04-02 py_src/main.py

@@ -39,7 +39,7 @@ if __name__=="__main__":
    
     #Bilby 
     init_parameters, priors = bilby_priors_dict(PTA)
-    BilbySampler(KF,init_parameters,priors)
+    BilbySampler(KF,init_parameters,priors,label="run_0304a",outdir="../data/nested_sampling/")
 
 
 

### 2023-04-02 py_src/system_parameters.py

@@ -12,8 +12,8 @@ def SystemParameters(NF=np.float64,    # the number format of the arguments
                      Ω=   5e-7,           # GW angular frequency
                      Φ0 = 0.20,        # GW phase offset at t=0
                      ψ =  2.50,         # GW polarisation angle
-                     ι =  0.0,          # GW source inclination
-                     δ =  0.0,          # GW source declination
+                     ι =  0.4,          # GW source inclination
+                     δ =  1.0,          # GW source declination
                      α =  1.0,          # GW source right ascension
                      h =  1e-2,         # GW strain
                      σp = 1e-16,       # process noise standard deviation

### 2023-04-06 docs/manuscript/paper.tex
small correction to control vector

@@
 The foremost technique for the detection of GWs in this nano-Hz regime is a via timing an ensemble of milliseconds pulsars; a pulsar timing array (PTA) \citep{2021hgwa.bookE...4V}.
-The presence  of a nano-Hz gravitational wave will influence the propagation of the pulsar radio beacon and consequently will imprint on the pulsar timing signal.
+The presence  of a nano-Hz gravitational wave will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the pulsar timing signal.
 By measuring the modulation of the received pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs.
@@
 These previously disparate efforts have now been joined in international collaboration, along with a number of newer PTAs, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}.
-The primary target of PTA observations is the inspiral of supermassive black hole binaries (SMBHBs) with masses $\sim 10^7 M_{\odot}$.
-These signals from SMBHBs can be broadly classified into either deterministic or stochastic.
+The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries (SMBHBs) with masses $\sim\mathcal{ 10^7 M_{\odot}}$.
+These GW signals from SMBHBs can be broadly classified into either deterministic or stochastic.
 For the former, sufficiently bright and near binaries may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated.
@@
 the spin wandering of the pulsar) infer both the system parameters and the detectability of the GW signal.
-
-For this initial work we will focus exclusively on resolved, monochromatic GW sources. Discussion on
+For this initial exploratory study we will focus exclusively on resolved, monochromatic GW sources.
 
@@
 
-In section 2 we will briefly outline the response 
-In section 3 we will then construct a state space form for determinsitic signals (3.1) outline the mehtods for expectaiton likelihood (3.2) and apply this state space kalma nform
+\footnote{1.
+Introduce PTAs generally.
+2.Types of astrophysical source to be detected with PTAs.
+3.
+Previous methods.
+4.
+Challenges/problems of previous methods.
+5.
+Our approach and its potential advantages.
+}
 
-In section 4 we will extend to stochastic sources...? OR we will leave thus for a second paper...
+\section{State-Space Model}
+We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state and the measurement state recorded by an observer.
+We will take our state to be the pulsar pulse frequency $f(t)$.
+We will now consider how the intrinsic pulse frequency evolves in time, completely separate from the influence of a GW, and then go on to derive the  influence of a GW perturbation on the frequency recorded by an observer at Earth.
 
-PTA sources can be braodly characterised into two main buckets: determinsitic and stocahstic. Derterminisitc sources are loud isngle
-
-\begin{itemize}
-	\item Introduce PTAs generally.
-	\item Types of astrophysical source to be detected with PTAs.
-	\item Why we want to do parameter estimation
-	\item Advantages of doing this with state-space approach
-\end{itemize}
-
-\section{Evolution of the pulsar frequency}
+\subsection{Evolution of the pulsar frequency}
 We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \citep{Vargas}.
@@
 `spin wandering', achromatic variations in the pulse TOA intrinsic to the star).
+The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process.
 Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
@@
 \end{equation}
-For PTA analysis, we are concerned with timescales on the order or years. Consequently, we can express the EM spindown straightforwardly as
+For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
 \begin{equation}
@@
 \end{equation}
+As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g.
+a model of the neutron star crust and superfluid components][]{Meyers2021}.
+However, for our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.
 
-\section{Modulation of pulsar frequency due to a GW}
+\subsection{Modulation of pulsar frequency due to a GW}
 In the presence of a GW, the $f(t)$ measured by an observer on Earth is different from that measured by an observer in the local NS reference frame.
@@
 
-\subsection{Plane GW perturbation}
+\subsubsection{Plane GW perturbation}
 We take a gravitational plane wave that perturbs a background Minkowski spacetime as
@@
 We emphasise that $\Omega$ has no time dependence - we are concerned solely with monochromatic sources.
-Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{This point is important, since the phase offset is then the same between multiple pulsars.
-See also Melatos 2022 PT12}.
+Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{\textbf{TK: This point is important, since the phase offset is then the same between multiple pulsars.
+See also Melatos 2022 PT12}}.
 The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
@@
 \begin{equation}
-	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
+	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b
 \end{equation}
@@
 which are in turn specified via the location of the GW source on the sky (via $\theta$, $\phi$ coordinates) and the polarisation angle $\psi$
-
 \begin{align}
@@
 
-\subsection{Pulse frequency as a photon}
-\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$.
-Generally, the frequency of a photon with 4-momentum $p_{\mu}$ recorded by an observer with 4-velocity $u^{\mu}$ is
+\subsubsection{Pulse frequency as a photon}
+\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$. Generally, the frequency of a photon recorded by an observer with 4-velocity $u^{\mu}$ is 
 \begin{equation}
@@
 \end{equation}
-\noindent Consequently the frequency can be identified with the temporal component of the covariant 4-momentum,
+\noindent Consequently the frequency can be directly identified with the temporal component of the covariant 4-momentum,
 \begin{equation}
-	\nu = p_t
+	f = p_t
 \end{equation}
@@
 
-\subsection{Hamiltonian Mechanics}
+\subsubsection{Hamiltonian Mechanics}
 The Hamiltonian in covariant notation can be written as 
@@
 \end{equation}
-\footnote{This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset.}
+\footnote{\textbf{TK: This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset}.}
 
@@
 \end{equation}
-\noindent To make contact with Melatos 2022 it will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar \footnote{similarly $\bar{x}(t) = -\bar{q}t$} and $\omega$ is the \textit{constant} photon angular frequency.
-Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer (see Melatos 2022, PT16).
+\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation.
+In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.
+It will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar and $\omega$ is the \textit{constant} photon angular frequency.
+Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer \footnote{see Melatos 2022, PT16}.
+We will consider  the pulsar locations to be constant with respect to the Earth.
+TK: \textbf{This may have already been "done" during the barycentreing when pulsar TOAs are generated, in which case $q$ is the vector from the SSB to the pulsar}.
+Similarly we can express the spatial coordinates as $\bar{x}(t) = -\bar{q}t$.
 \newline
-
-\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.  \newline
 
@@
 \end{equation}
-where the $\omega$ term in the numerator results from a $\omega^2$ terms that comes from the $p^i p^j$ term and a $\omega^{-1}$ term from $(p^t)^{-1}$.
-\subsection{Performing the integral}
+
+\subsubsection{Performing the integral}
 
@@
 \end{align}
+We can write this more explicitly as,
+\begin{align} \label{eq:final}
+	&p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = \\
+	&\frac{ \omega}{2} \frac{H_{ij}q^i q^j  }{(1 + \hat{n} \cdot \hat{q})}e^{-i \Omega \tau(1+\bar{n} \cdot \bar{q}) + \Phi_0} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+\end{align}
+or alternatively, with $h_{ij} = g_{ij} - \eta_{ij}$,
+\begin{equation}
+	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{h_{ij} q^i q^j  }{(1 + \hat{n} \cdot \hat{q})} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+	\label{eq:freq_shift}
+\end{equation}
 
-\subsection{Explicit expression and comparison with Melatos 22}
+\subsection{State-space structure} \label{sec3}
+We now have a model for the evolution of the pulsar frequency states and the measured frequency timeseries.
+We can express the intrinsic frequency evolution, Eq.
+\ref{eq:frequency_evolution_expanded}, in a alternative form as,
+\begin{equation}
+	df = A f dt + N(t) dt + \sigma dB(t)
+	\label{eq:state1}
+\end{equation}
+where $A = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process).
+This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by,
+\begin{equation}
+	f(t) = e^{At}f(0) + \int_0^t e^{A(t-t')} N(t') dt' + \int_0^t e^{A(t-t')} \sigma dB(t')
+\end{equation} 
+If we move from a solution in continuous time to discrete time then
+\begin{equation}
+	f(t_{i+1}) = F f(t_i) + T_i + \eta_i
+\end{equation}
+where
+\begin{equation}
+	F_i = e^{A (t_{i+1} - t_i)}
+\end{equation}
+\begin{equation}
+	T_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt'
+\end{equation}
+\begin{equation}
+	\eta_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} \sigma dt'
+\end{equation}
+Given the discrete solution $f(\bar{t})$ where $\bar{t} = (t_1, t_2, ...,t_N)$ the intrinsic frequency can be related to the measured frequency as 
+\begin{equation}
+	f_M = f g(\bar{\theta},t) + N_M
+\end{equation}
+where $g(\theta,t)$ ("measurement function")is a function of some parameters, $\bar{\theta}$ and time $t$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
+\begin{equation}
+	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
+\end{equation}
+The measurement function follows from Eq. \ref{eq:freq_shift}
+\begin{equation} \label{eq:final}
+	g(\bar{\theta},t) = 1 - \frac{1}{2} \frac{h_{ij}(t) q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+\end{equation}
+or in a more concise form where we have switched to a trigonometric form of the equations,
+\begin{equation} 
+	g(\bar{\theta},t) = 1 -A \cos(-\Omega t (1 + n\cdot q ) + \Phi_0)
+\end{equation}
+where 
+\begin{equation}
+	A =  \frac{1}{2} \frac{H_{ij} q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - \cos\left( {\Omega(1+\bar{n} \cdot \bar{q}) d}\right)]
+		\label{eq:state2}
+\end{equation}
 
-We can expand the concise expression of Eq. \ref{eq:concise} as,
+\section{Detection and parameter estimation}
+Given the proceeding state-space model we will now try to use this model to infer the unknown parameters of the system.
+Let's first review and categorise all the parameters of model.
+We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar, the GW parameters and the noise parameters i.e.
+\begin{equation}
+	\bar{\theta} =  \bar{\theta}_{\rm PSR} \cup \bar{\theta}_{\rm GW} \cup \bar{\theta}_{\rm noise}
+\end{equation}
+\begin{equation}
+	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]
+\end{equation}
+\begin{equation}
+	\bar{\theta}_{\rm GW} = [h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0]
+\end{equation}
+\begin{equation}
+	\bar{\theta}_{\rm noise} = [\sigma, \Sigma]
+\end{equation}
+Whilst the GW parameters are shared between measurements for each pulsar, the pulsar parameters are clearly not.
+For a PTA dataset of N pulsars we have $7 + 5N$ parameters to estimate.
+For now we will consider the measurement noise to be known, although in principle this too could be estimated.
+Note that whilst this is a large parameter space, in general the pulsar parameters are much better constrained than the GW parameters: for example we have rough estimates for the pulsar distances accurate to $\sim$ 10$\%$, but we have no prior information on the source location.
+\newline
 
-\begin{equation} \label{eq:final}
-	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{H_{ij}q^i q^j  }{(1 + \hat{n} \cdot \hat{q})}e^{-i \Omega \tau(1+\bar{n} \cdot \bar{q}) + \Phi_0} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+\noindent We will use a Kalman filter in order to obtain likelihood estimates of the data given a set of parameters.
+We will now review the Kalman filter and the associated likelihood in Section \label{sec:kalman_filter} before going on to discuss how to use nested sampling methods in conjunction with the filter for detection (model selection) and parameter estimation
+
+\subsection{Kalman Filtering}\label{sec:kalman_filter}
+The Kalman filter \citep{Kalman1} is a algorithmic technique for recovering a set of system state variables $\bar{x}$ given some noisy measurements $\bar{z}$.
+It is a common technique in signal processing that has also been applied more recently with great success in astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}.
+The linear Kalman filter operates on measurements that are related to states via a linear transformation
+\begin{equation}
+	z = Hx + v
 \end{equation}
-This expression is equivalent to Melatos 2022, PT13 where the dependence on time $t$ is made explicit. Using the Melatos notation where $h_{ij} = g_{ij} - \eta_{ij}$
+where $H$ is the measurement matrix and $v$ a Gaussian measurement noise. The underlying states are the solutions to the state-space equation 
+\begin{equation}
+	\dot{x} = Fx + Gu + w
+\end{equation}
+where $x$ are the state variables, $F$ the system dynamics matrix, $G u$ the control matrix/vector and $w$ a stochastic zero-mean process which .
+By comparison with the preceding equations Eqs.
+	\ref{eq:state1} - 	\ref{eq:state2} it is immediately obvious how our state space model maps onto the Kalman filter structure.
+Specifically, our states are just the $N$ intrinsic pulsar frequencies $\bar{x} = (f_1,f_2,...f_N)$ whilst our measurements are the $N$ measured pulse frequencies $\bar{z} = (f^{(M)}_1,f^{(M)}_2,...f^{(M)}_N)$.
+If we specialize to the case of constant time sampling between our observations, then for our formulation, the various components that make up the Kalman filter are as follows:
+\begin{equation}
+	F_i = F_{i+1} = e^{-\gamma \Delta t}
+\end{equation}
+\begin{align}
+	T_i &= \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt' \\
+	    & =
+\end{align}
+
+The Kalman filter includes the effect of process noise and measurement noise via the defines a measurement noise matrix $R = E[v v^T]$
+defines a process noise matrix $Q = E[w w^T]$.
 
 \begin{equation}
-	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{h_{ij} q^i q^j  }{(1 + \hat{n} \cdot \hat{q})} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+	\eta_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} \sigma dt'
 \end{equation}
 
-\newpage
-\section{State-space structure} \label{sec3}
+Whilst for our problem we only need concern ourselves with linear stuff
+
+%Extensions to non-linear state-transitions
+%[i.e. 𝑭𝑿 → 𝑓 (𝑿) in equation (11)], and non-linear measurement
+%equations [i.e. 𝑪𝑿 → 𝑐(𝑿) in equation (10)] can be achieved using a range of tools such as the extended Kalman filter (Jazwinski
+%1970), the unscented Kalman filter (Julier & Uhlmann 1997; Wan
+%& Merwe 2000; Julier & Uhlmann 2004), or particle filters (Del
+%Moral 1997).
+%The full set of Kalma
+%
+for a full review we refer the reader to X
+
+the state the states are related to a measurement via
+
+where this is that is this
+
+By comaprison with the state space equations we derived earlier (Eqs A to B) we can see that our continuous equations map ideally onto this structure.
+In particular, for our system our state $\bar{x}$ is simply the frequencies of each individual pulsar $= (f_1, f_2, f_3,...)$
+
+here is an example for an arbitary GW and arbitrary pulsar with good and bad parameters 
+\subsection{Nested Sampling}
+
+We can use NS for both param estimation and
+
+\section{Validation on synthetic data}
+
+\subsection{PTA pulsars}
+Throughout this work it will be necessary to select a group of pulsars to make up our PTA. We will take the 47 pulsars that make up the NANOGrav PTA \citep{2020ApJ...905L..34A}
+
+In order to proceed and explore how well this state-space formulation works, we will need to specify a selection of pulsars to make up our PTA.
+We will take the 47 pulsars that make up the NANOGrav PTA.
+For each pulsar, we need to specify the complete set of $\bar{\theta}_{\rm PSR}$ as well as $\sigma$.
+The parameters $f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d$ are straightforward to set and can be read directly from  the current "present day" best estimates of the pulsar frequency, derivative and distance.
+
+\begin{figure}
+	\includegraphics[width=0.8\columnwidth]{images/pulsars}
+	\caption{Spatial distribution and distances of NANOGrav pulsars}
+	\label{fig:pulsar_distrib}
+\end{figure}
+
+\subsection{Detection}
+
+\subsection{Parameter Estimation}
+
+\section{Kalman filtering}
+
+\section{Parameter estimation}
+
+\subsection{Nested sampling}
+
+\section{Detection}
+
 \noindent The general structure of a Kalman state-space problem is
@@
 This intrinsic frequency can be related to a measured frequency as
-\begin{equation}
-	f_M = f g(\bar{\theta},t) + N_M
-\end{equation}
-where $g(\theta,t)$ ("measurement function")is a function of some parameters, $\bar{\theta}$ and time $t$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
-\begin{equation}
-	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
-\end{equation}
-Explicitly, it can be shown that the measurement function is,
-\begin{equation} \label{eq:final}
-	g(\bar{\theta},t) = 1 - \frac{1}{2} \frac{h_{ij}(t) q(t)^i q(t)^j}{1 + \bar{n} \cdot \bar{q}(t)}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}(t)) d}]
-\end{equation}
+
 There are a few terms to unpack and define in this equation.
@@
 
-\noindent Going forward, for now we will take $q(t) = q$ i..e.
-the pulsar locations are constant with respect to the Earth.
-This may have already been "done" during the barycentreing when pulsar TOAs are generated, in which case $q$ is the vector from the SSB to the pulsar  \newline
+\noindent Going forward,  \newline
 
@@
 
-\begin{equation} 
-	g(\bar{\theta},t) = 1 -A \cos(-\Omega t (1 + n\cdot q ) + \Phi_0)
-\end{equation}
-where 
-\begin{equation}
-	A =  \frac{1}{2} \frac{H_{ij} q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - \cos\left( {\Omega(1+\bar{n} \cdot \bar{q}) d}\right)]
-\end{equation}
+\subsection{Parameters of the model}
+\noindent
 
-\subsection{Parameters of the model}
-\noindent Lets review and categorise all the parameters the the above model.
-We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar, the GW parameters and the noise parameters i.e.
-\begin{equation}
-	\bar{\theta} =  \bar{\theta}_{\rm PSR} \cup \bar{\theta}_{\rm GW} \cup \bar{\theta}_{\rm noise}
-\end{equation}
-\begin{equation}
-	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]
-\end{equation}
-\begin{equation}
-	\bar{\theta}_{\rm GW} = [h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0]
-\end{equation}
-\begin{equation}
-	\bar{\theta}_{\rm noise} = [\sigma, \Sigma]
-\end{equation}
 We can also express the measurement equation generally as,
 
-\section{PTA pulsars}
-In order to proceed and explore how well this state-space formulation works, we will need to specify a selection of pulsars to make up our PTA.
-We will take the 47 pulsars that make up the NANOGrav PTA.
-For each pulsar, we need to specify the complete set of $\bar{\theta}_{\rm PSR}$ as well as $\sigma$.
-The parameters $f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d$ are straightforward to set and can be read directly from  the current "present day" best estimates of the pulsar frequency, derivative and distance.
+Kalman filtering works as follows
 
-\begin{figure}
-	\includegraphics[width=0.8\columnwidth]{images/pulsars}
-	\caption{Spatial distribution and distances of NANOGrav pulsars}
-	\label{fig:pulsar_distrib}
-\end{figure}
-
-\section{Kalman filtering}
+you can see that it maps nicel onto our state separation
 

### 2023-04-06 py_src/kalman_filter.py

@@ -16,7 +16,7 @@ def log_likelihood(S,innovation):
     x = innovation / S 
     N = len(x)
     #slogdet = np.sum(np.log(S)) # Uses log rules and diagonality of covariance "matrix"
-    #return -0.5*(slogdet+innovation @ x + N*np.log(2*np.pi))
+    #eturn -0.5*(slogdet+innovation @ x + N*np.log(2*np.pi))
     return -0.5*(innovation @ x + N*np.log(2*np.pi))
 
 

### 2023-04-06 py_src/main.py

@@ -18,6 +18,8 @@ import bilby
 
 
 if __name__=="__main__":
+    import multiprocessing
+    multiprocessing.set_start_method("fork")
 
 
     P   = SystemParameters()       #define the system parameters as a class
@@ -39,7 +41,7 @@ if __name__=="__main__":
    
     #Bilby 
     init_parameters, priors = bilby_priors_dict(PTA,P)
-    BilbySampler(KF,init_parameters,priors,label="canonical_example2",outdir="../data/nested_sampling/")
+    BilbySampler(KF,init_parameters,priors,label="alpha_narrow_prior",outdir="../data/nested_sampling/")
 
 
 

### 2023-04-06 py_src/model.py

@@ -24,7 +24,11 @@ class LinearModel:
     def T_function(f0,fdot,gamma,t,dt):
 
        
-        fdot_time =  np.outer(t+dt,fdot) #This has shape(n times, n pulsars)
+        #fdot_time =  np.outer(t+dt,fdot) #This has shape(n times, n pulsars)
+        #value = f0 + fdot_time + fdot*dt - np.exp(-gamma*dt)*(f0+fdot_time)
+        
+        
+        fdot_time =  np.outer(t,fdot) #This has shape(n times, n pulsars)
         value = f0 + fdot_time + fdot*dt - np.exp(-gamma*dt)*(f0+fdot_time)
 
         return value

### 2023-04-06 docs/manuscript/paper.tex
EoD

@@
 %%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%%
-
+\usepackage{subcaption}
 % Please keep new commands to a minimum, and use \newcommand not \def to avoid
@@
 Specifically, our states are just the $N$ intrinsic pulsar frequencies $\bar{x} = (f_1,f_2,...f_N)$ whilst our measurements are the $N$ measured pulse frequencies $\bar{z} = (f^{(M)}_1,f^{(M)}_2,...f^{(M)}_N)$.
-If we specialize to the case of constant time sampling between our observations, then for our formulation, the various components that make up the Kalman filter are as follows:
+If we specialize to the case of constant time sampling between our observations, then for our formulation, the components that make up the Kalman filter are as follows:
 \begin{equation}
@@
 	T_i &= \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt' \\
-	    & =
+	    & = f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)  (\Delta t + t_i) - e^{-\gamma \Delta t} (f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)t_i)
 \end{align}
+\begin{equation}
+	H_i = 1 -A(\theta_{\rm GW}) \cos(-\Omega t_i (1 + n\cdot q ) + \Phi_0)
+\end{equation}
+and $A$ is a constant that is given by  Eq.
+\ref{eq:state2}.
+The Kalman filter includes the effect of process noise $w$ and the measurement noise $v$ via the definition of a process noise matrix $Q = E[w w^T]$ and a measurement noise matrix $R = E[v v^T]$, which have the discrete form,
+\begin{equation}
+	Q_i  \delta_{ij}= \langle \eta_i \eta_j^T \rangle = \frac{- \sigma^2}{2 \gamma} \left( e^{-2 \gamma \Delta t} -1\right)
+\end{equation}
+\begin{equation}
+	R_i = R_{i+1} = \Sigma^2
+\end{equation}
+For our formulation we are concerned only with the linear Kalman filter since the measurements are a linear function of the states and the state transitions are also linear.
+Extension to non-linear problems is straighforward using either an extended Kalman filter, the unscented Kalman filter or the particle filter.
+For a full review of the  Kalman filter equations which may e unfamilar to some, we refer the reader to the appendix X of \cite{Melatos2023}.
+\newline
 
-The Kalman filter includes the effect of process noise and measurement noise via the defines a measurement noise matrix $R = E[v v^T]$
-defines a process noise matrix $Q = E[w w^T]$.
+\noindent In order for the Kalman filter to function successfully, the parameters of the model, which appear in the various Kalman matrices, must be accurate (e.g.
+Fig \ref{fig:four figures}).
+The filter tracks the error in the
 
-\begin{equation}
-	\eta_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} \sigma dt'
-\end{equation}
+The Kalman filte rprovides an estiamte In this way,
 
-Whilst for our problem we only need concern ourselves with linear stuff
-
-%Extensions to non-linear state-transitions
-%[i.e. 𝑭𝑿 → 𝑓 (𝑿) in equation (11)], and non-linear measurement
-%equations [i.e. 𝑪𝑿 → 𝑐(𝑿) in equation (10)] can be achieved using a range of tools such as the extended Kalman filter (Jazwinski
-%1970), the unscented Kalman filter (Julier & Uhlmann 1997; Wan
-%& Merwe 2000; Julier & Uhlmann 2004), or particle filters (Del
-%Moral 1997).
-%The full set of Kalma
-%
-for a full review we refer the reader to X
-
-the state the states are related to a measurement via
-
-where this is that is this
-
-By comaprison with the state space equations we derived earlier (Eqs A to B) we can see that our continuous equations map ideally onto this structure.
-In particular, for our system our state $\bar{x}$ is simply the frequencies of each individual pulsar $= (f_1, f_2, f_3,...)$
+\begin{figure*}
+	%\centering % Not needed
+	\begin{subfigure}[b]{1\columnwidth}
+		\includegraphics[width=\textwidth]{images/KF1}
+		\caption{}
+		\label{fig:6MB_BFS}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{1\columnwidth}
+		\includegraphics[width=\textwidth]{images/KF2}
+		\caption{}
+		\label{fig:25MB_bfs}
+	\end{subfigure}
+	\label{fig:four figures}
+\end{figure*}
 

### 2023-04-06 py_src/plotting.py

@@ -31,7 +31,7 @@ def plot_statespace(t,states,measurements,psr_index):
 
 
 
-def plot_all(t,states,measurements,predictions,psr_index):
+def plot_all(t,states,measurements,predictions,psr_index,savefig=None):
 
     plt.style.use('science')
 
@@ -63,6 +63,9 @@ def plot_all(t,states,measurements,predictions,psr_index):
 
     plt.subplots_adjust(wspace=0.1, hspace=0.1)
     #plt.rcParams["font.family"] = "fantasy"
+
+    if savefig != None:
+        plt.savefig(f"../data/images/{savefig}.png", bbox_inches="tight",dpi=300)
    
     plt.show()
 

### 2023-04-11 docs/manuscript/paper.tex
update to paper

@@
 % If you need two or more lines of authors, add an extra line using \newauthor
-\author[Kimpson]{Kimpson$^{1}$, Melatos,O'Leary, , Evans, others, etc. %
+\author[Kimpson]{Kimpson$^{1}$, Melatos, O'Leary, Evans, others, etc. etc. %
 \thanks{Contact e-mail: \href{mailto:mn@ras.ac.uk}{mn@ras.ac.uk}}%
@@
 
-The detection of high frequency ($\sim 1$Hz) gravitational waves (GWs) from coalescing BH binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
+The detection of high frequency ($\sim1$ Hz) gravitational waves (GWs) from coalescing BH binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
 Gravitational radiation from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2030$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper} especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
@@
 
-\noindent Motivated by the difficulties faced by the classic PTA analysis methods, in this work we present a novel approach to formulate PTA analysis and GW detection as a state-space problem.
-This approach enables the state-space evolution to be tracked optimally, and for a specific realisation of the pulsar process noise (i.e.
-the spin wandering of the pulsar) infer both the system parameters and the detectability of the GW signal.
+\noindent The sensitivity of a PTA to a GW signal is heavily dependent on the total level of noise \citep{Wang2015}.
+In particular, the intrinsic pulsar timing noise - i.e.
+random variations in the pulse arrival time - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}.
+This pulsar timing noise has multiple potential theorized causes including microglitches \citep{Melatos2008}, glitch recovery \citep{Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Antonelli2023} and superfluid turbulence \citep{Melatos2014}.
+In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars which are known to be much more stable rotators with minimal timing noise.
+However,
+timing noise in MSPs could be a `latent' phenomenon \citep{Shannon2010}; as we increase the length of observation spans and the timing precision required to detect gravitational waves, timing noise in MSPs may become emergent and more apparent, analogous to that seen in younger pulsars.
+In addition to timing noise there are also secondary noise sources one must consider such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+\newline
+
+\noindent In the standard approach to PTA-GW analysis, timing noise is nuisance which must be accurately modelled and subtracted and fundamentally limits the sensitivity of the PTA detector to a GW signal and its ability to infer the source parameters.
+Motivated by these challenges faced by the classic PTA analysis methods, in this work we present a novel approach to formulate PTA analysis and GW detection as a state-space problem.
+This approach enables the pulsar state-space evolution to be tracked optimally, given a specific realisation of the pulsar process noise (i.e.
+the spin wandering of the pulsar), and determine both the presence of a GW signal in the pulsar data, and  infer the underlying source parameters.
 For this initial exploratory study we will focus exclusively on resolved, monochromatic GW sources.
+\newline
 
-This paper is organised as follows.
-
-\footnote{1.
-Introduce PTAs generally.
-2.Types of astrophysical source to be detected with PTAs.
-3.
-Previous methods.
-4.
-Challenges/problems of previous methods.
-5.
-Our approach and its potential advantages.
-}
+This paper is organised as follows:
 
@@
 \subsection{Kalman Filtering}\label{sec:kalman_filter}
-The Kalman filter \citep{Kalman1} is a algorithmic technique for recovering a set of system state variables $\bar{x}$ given some noisy measurements $\bar{z}$.
+The Kalman filter \citep{Kalman1} is a algorithmic technique for recovering a set of system state variables, $\bar{x}$, given some noisy measurements, $\bar{z}$.
 It is a common technique in signal processing that has also been applied more recently with great success in astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}.
@@
 where $x$ are the state variables, $F$ the system dynamics matrix, $G u$ the control matrix/vector and $w$ a stochastic zero-mean process which .
-By comparison with the preceding equations Eqs.
-	\ref{eq:state1} - 	\ref{eq:state2} it is immediately obvious how our state space model maps onto the Kalman filter structure.
+By comparison with the preceding equations, Eqs.
+	\ref{eq:state1} - 	\ref{eq:state2}, it is immediately obvious how our state space model maps onto the Kalman filter structure.
 Specifically, our states are just the $N$ intrinsic pulsar frequencies $\bar{x} = (f_1,f_2,...f_N)$ whilst our measurements are the $N$ measured pulse frequencies $\bar{z} = (f^{(M)}_1,f^{(M)}_2,...f^{(M)}_N)$.
-If we specialize to the case of constant time sampling between our observations, then for our formulation, the components that make up the Kalman filter are as follows:
+If we specialize to the case of constant time sampling between our observations,$\Delta t$, then for our formulation the components that make up the Kalman filter are as follows:
 \begin{equation}
@@
 For our formulation we are concerned only with the linear Kalman filter since the measurements are a linear function of the states and the state transitions are also linear.
-Extension to non-linear problems is straighforward using either an extended Kalman filter, the unscented Kalman filter or the particle filter.
-For a full review of the  Kalman filter equations which may e unfamilar to some, we refer the reader to the appendix X of \cite{Melatos2023}.
+Extension to non-linear problems is straightforward using either an extended Kalman filter, the unscented Kalman filter or the particle filter.
+For a full review of the  Kalman filter equations which may bee familiar to some, we refer the reader to the appendices of \cite{Melatos2023} and  \cite{Meyers2021}.
 \newline
 
-\noindent In order for the Kalman filter to function successfully, the parameters of the model, which appear in the various Kalman matrices, must be accurate (e.g.
+\noindent In order for the Kalman filter to function successfully, the parameters of the model, which appear in the various Kalman matrices, must be accurate.
+Erroneous or inaccurate parameters lead to inaccurate predictions of the underlying states (e.g.
 Fig \ref{fig:four figures}).
-The filter tracks the error in the
 
-The Kalman filte rprovides an estiamte In this way,
+The filter tracks the error in its predictions of the underlying states by projecting the state predictions back into measurement space $\hat{z}$.
+These measurement predictions can then be compared against the true observed measurements.
+The Gaussian log-likelihood as a function of the parameters is,
+\begin{eqnarray}
+	\log \mathcal{L} =  -\frac{1}{2} (N \log 2 \pi + \log \sigma^2 + \frac{y^2}{x})
+\end{eqnarray}
+where this is this is that....
+This likelihood provided by the Kalman filter can then be used for Bayesian inference based techniques of the underlying system parameters, which we will now explore in more detail.
 
@@
 	\end{subfigure}
+\caption{TK: caption needed}
 	\label{fig:four figures}
@@
 
-here is an example for an arbitary GW and arbitrary pulsar with good and bad parameters 
 \subsection{Nested Sampling}
+Nested Sampling \cite{Skilling} is an integration algorithm used for evaluating integrals of the Bayesian form
+\begin{eqnarray}
+	Z = \int L \pi (\theta) d \theta
+\end{eqnarray}
+where this is this is that...
+i.e.
+the model evidence.
+It is a highly advantageous approach over the usual MCMC-type approaches which are used solely for calculating the posterior but can be highly computationally expensive, does not return an evidence, and can struggle on multi-modal problems.
 
-We can use NS for both param estimation and
+Within the astrophysical community, it was first adopted in cosmology, CITE CITE but has become a widespread technique used wihin astronomy, astrophysics, particle physics and materials science CITE
 
-\section{Validation on synthetic data}
+Nested sampling genearlly works by algorithmically evolving a set of "live points" that seek to describe an underlying probability distribution.
+For a full review we refer the reader to original works by \cite{Skilling} and a recent review from a physics sciences perspective from \textbf{ASHTON}
+
+Multiple nested sampling libraries exist.
+For gravitational astrophysics it is common to use the dynesty sampler CITE, via the Bilby gravitational wave inference library and we continue to follow this precedent.
 
 \subsection{PTA pulsars}
-Throughout this work it will be necessary to select a group of pulsars to make up our PTA. We will take the 47 pulsars that make up the NANOGrav PTA \citep{2020ApJ...905L..34A}
+
+In order to proceed, it is now necessary to specify a PTA configuration.
+
+As discussed, multiple separate PTA detectors exist under the umbrella of the IPTA.
+Going forward we will take the 47 pulsars that make up the NANOGrav PTA \citep{2020ApJ...905L..34A}.
+NANOGrav is selected simply as a well-representative example of the typical pulsars that make up a PTA.
+Our results and formulation are not contingent on the choice of PTA, and naturally extend to other PTAs or PTAs with more pulsars.
+
+Within the state-space forumulation, the pulsar evolution is governed by a set of 5 parameters $\bar{\theta}_{\rm PSR}$ for each pulsar.
+The parameters $f_{\rm EM} (0) $,$ \dot{f}_{\rm EM} (0)$ and $d$ are well specified via existing pulsar datasets.
+We take the frequency and frequency derivative as returned from the pulsar datasets to simply be the values now at $t=0$.
+The pulsar distances are also known, though less well constrained.
+Going forward we take the distances retuned from the datasets as the true values of the pulsars that make up our synthetic PTA.
+
+The specification of $\gamma$ and $\sigma$ are more involved. $\gamma$ specifies an effective timescale of reversion to the mean
+
+\textbf{TK: discussion on how to select these two parameters c.f. Andres }
 
@@
 We will take the 47 pulsars that make up the NANOGrav PTA.
+
 For each pulsar, we need to specify the complete set of $\bar{\theta}_{\rm PSR}$ as well as $\sigma$.
@@
 
-\subsection{Detection}
+\subsection{Parameter estimation on synthetic data}
+We are now in a position to try to infer the parameters of our system
 
-\subsection{Parameter Estimation}
+\begin{figure*}
+	\includegraphics[width=0.8\textwidth]{images/corner_example}
+	\caption{Example corner plot}
+	\label{fig:corner}
+\end{figure*}
 
-\section{Kalman filtering}
-
-\section{Parameter estimation}
-
-\subsection{Nested sampling}
-
-\section{Detection}
+\subsection{Detection on synthetic data}
 
@@
 \ref{fig:mean and std of nets}
+
+Quality vs quantity: 
+This is important as it measn we dont need to do this for all pulsars, just a few. 
+%https://arxiv.org/abs/2211.03201
 
@@
 
+\section{Discussion}
+
+\section{Conclusion}
+
+PTAs are all about MSPs for small timing noise. Maybe we can get away with large timing noise, and use more pulsars?
+
+Radiomenter noise. In theory well known but this could also be estimated via this approach
+
 \subsection{References}

### 2023-04-11 py_src/plotting.py

@@ -71,7 +71,7 @@ def plot_all(t,states,measurements,predictions,psr_index,savefig=None):
 
 
 
-def plot_custom_corner(path,labels, injection_parameters,axes_scales):
+def plot_custom_corner(path,labels, injection_parameters,axes_scales,savefig):
 
 
     # Opening JSON file
@@ -99,6 +99,10 @@ def plot_custom_corner(path,labels, injection_parameters,axes_scales):
                         quantiles=[0.16, 0.84],
                         truths=injection_parameters,
                         axes_scales = axes_scales)
+    
+
+    if savefig != None:
+        plt.savefig(f"../data/images/{savefig}.png", bbox_inches="tight",dpi=300)
     
     # fig.set_figwidth(12)
     # fig.set_figheight(8)

### 2023-04-11 docs/manuscript/paper.tex
assorted updates

@@
 
-The detection of high frequency ($\sim1$ Hz) gravitational waves (GWs) from coalescing BH binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
-Gravitational radiation from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2030$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper} especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
+The detection of high frequency ($\sim1$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
+Gravitational radiation from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
 Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
@@
 \noindent Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}.
-These previously disparate efforts have now been joined in international collaboration, along with a number of newer PTAs, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}.
-The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries (SMBHBs) with masses $\sim\mathcal{ 10^7 M_{\odot}}$.
+These previously disparate efforts have now been joined in international collaboration, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}.
+The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries (SMBHBs) with masses $\sim 10^7 M_{\odot}$.
 These GW signals from SMBHBs can be broadly classified into either deterministic or stochastic.
-For the former, sufficiently bright and near binaries may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated.
-For the latter, the incoherent superposition of multiple weaker SMBHBs sources leads to a stochastic background detectable at nano-Hz frequencies.
+For the former, sufficiently bright and near binaries may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Zhu10}.
+For the latter, the incoherent superposition of multiple weaker SMBHBs sources leads to a stochastic background detectable at nano-Hz frequencies \citep{Sesana10}.
 Other potential sources for PTAs include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}, but the deterministic and stochastic GW signals from SMBHBs remain the primary targets.
@@
 The presence of a GW induces a characteristic correlation function as a function of the angular separation between the pulsars; the Hellings-Downs curve \citep{Hellings}.
-For both classes of source, detection of GW signals in the timing residuals of a PTA is a challenging enterprise, and currently neither a stochastic background nor an individually resolved source has yet been detected \citep{2022MNRAS.510.4873A, 10.1093/nsr/nwx126}.
+For both classes of source, the detection of GW signals in the timing residuals of a PTA is a challenging enterprise, and currently neither a stochastic background nor an individually resolved source has yet been detected \citep{2022MNRAS.510.4873A, 10.1093/nsr/nwx126}.
 \newline
 
-\noindent The sensitivity of a PTA to a GW signal is heavily dependent on the total level of noise \citep{Wang2015}.
+\noindent The sensitivity of a PTA to a GW signal is heavily dependent on the total level of noise in the array \citep{Wang2015}.
 In particular, the intrinsic pulsar timing noise - i.e.
@@
 This pulsar timing noise has multiple potential theorized causes including microglitches \citep{Melatos2008}, glitch recovery \citep{Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Antonelli2023} and superfluid turbulence \citep{Melatos2014}.
-In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars which are known to be much more stable rotators with minimal timing noise.
-However,
-timing noise in MSPs could be a `latent' phenomenon \citep{Shannon2010}; as we increase the length of observation spans and the timing precision required to detect gravitational waves, timing noise in MSPs may become emergent and more apparent, analogous to that seen in younger pulsars.
-In addition to timing noise there are also secondary noise sources one must consider such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise.
+However, timing noise in MSPs could be a `latent' phenomenon \citep{Shannon2010}; as we increase the length of observation spans and the measurement timing precision to the levels required to detect gravitational waves, the timing noise in MSPs may emerge and become much more apparent, visible in the same way that timing noise is seen in younger pulsars.
+In addition to timing noise there are also secondary noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
 \newline
 
-\noindent In the standard approach to PTA-GW analysis, timing noise is nuisance which must be accurately modelled and subtracted and fundamentally limits the sensitivity of the PTA detector to a GW signal and its ability to infer the source parameters.
+\noindent In the standard approach to PTA-GW analysis, timing noise is a nuisance which must be accurately modelled and subtracted.
+It fundamentally limits the sensitivity of the PTA detector to a GW signal, and the ability of an observer to infer the GW source parameters.
 Motivated by these challenges faced by the classic PTA analysis methods, in this work we present a novel approach to formulate PTA analysis and GW detection as a state-space problem.
@@
 For this initial exploratory study we will focus exclusively on resolved, monochromatic GW sources.
-\newline
+\textcolor{red}{TK: This motivation section is weak.
+Need to think about how to phrase this better.
+Be clear on the real challenges of the "old" approach and the advantages of the "new" one.} \newline
 
-This paper is organised as follows:
+\noindent This paper is organised as follows:  \textcolor{red}{TK: TBD} \newline
+
+\noindent We adopt the natural units, with $c = G = \hbar = 1$, and a
+$(-,+,+,+)$ metric signature. \newline
 
 \section{State-Space Model}
-We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state and the measurement state recorded by an observer.
-We will take our state to be the pulsar pulse frequency $f(t)$.
+We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, $\bar{x}$, and the measurement state recorded by an observer,$\bar{y}$.
+We will take our state variable to be the intrinsic pulsar pulse frequency $f(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar.
 We will now consider how the intrinsic pulse frequency evolves in time, completely separate from the influence of a GW, and then go on to derive the  influence of a GW perturbation on the frequency recorded by an observer at Earth.
+\textcolor{red}{Need some extra discussion here on why this choice of state/measurement space.
+Why not e.g.
+phase, TOA, voltage, etc.
+?}
 
 \subsection{Evolution of the pulsar frequency}
-We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \citep{Vargas}.
+We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \cite{Vargas}.
 Within this model, $f$ evolves according to a combination of both deterministic torques (i.e.
@@
 \end{equation}
-where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\gamma$ a proportionality constant, and $\xi(t)$ a white noise process that satisfies,
+where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation,$\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
 \begin{equation}
@@
 \end{equation}
-for spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
+for spatial coordinates $\bar{x}$, \textcolor{red}{(TK: We have a notation conflict between $x$ = state and $x$ = coordinates)} where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
 We emphasise that $\Omega$ has no time dependence - we are concerned solely with monochromatic sources.
@@
 
-\noindent The general structure of a Kalman state-space problem is
-\begin{equation}
-	\dot{x} = f(x) + w
-\end{equation}
-where $x$ are the state variables, $f()$ a non-linear function of the states, and $w$ a stochastic zero-mean process.
-The process noise matrix $Q = E(w w^T)$.
-The states are related to the measurement $z$ via a non-linear measurement function $h()$
-\begin{equation}
-	z = h(x) + v
-\end{equation}
-where $v$ is a stochastic zero-mean process and measurement noise matrix $R = E(v v^T)$. \newline
+We are interested in how detectable a GW is using PTA + state space method
 
-\noindent This structure maps onto the preceding equations nicely. Our hidden state is just the intrinsic pulsar frequency, which evolves as,
-\begin{eqnarray}
-	\dot{f}_P = -\gamma f_P^n + \xi(t)
-\end{eqnarray}
-where $n$ is a braking index (typical values $\sim 1-5$, we take a canonical value $=3$) \footnote{See A.
-Vargas talk re anomalous braking indices.
-P.S.
-is there an arXiv preprint?} $\gamma$ a proportionality constant, and $\xi(t)$ represents a stochastic white noise process.
-Equation \ref{eq:final} can be used to relate the intrinsic pulsar frequency to that measured by an observer on Earth:
+We can use our state-space tools  to try to solve the problem of GW detection with a PTA i.e. "Is there evidence of a GW in my data?.
 
-\begin{eqnarray}
-	f_M = f_p(1 - X) + N_M
-\end{eqnarray}
-where $X$ is the RHS of Eq.
-\ref{eq:final} and $N_M$ a Gaussian measurement noise.
-An example of the evolution of the state and the measurement frequencies for our PTA pulsars can be seen in Fig.
-\ref{fig:mean and std of nets}
+We can frame this as a model selection procedure where we have two models/hypotheses:
 
-Quality vs quantity: 
-This is important as it measn we dont need to do this for all pulsars, just a few. 
-%https://arxiv.org/abs/2211.03201
+* **Null Model** $M_0$. There is no GW in the data. In this case the measurement model of the Kalman filter simply returns the frequency states i.e. $g(t,\theta)= 0$
+* **Alternative model** $M_1$. There is a GW in the data. The measurement model uses the full expression for $g(t,\theta)$
 
-\section{Discretisation}
+In order to accept the alternative hypothesis $M_1$ over $M_0$ there are two approaches we could take:
 
-This intrinsic frequency can be related to a measured frequency as
+1.
+The first is a fully Bayesian search over all the parameters for each model, calculating the evidence for each model and then determining the Bayes ratio.
+This is perhaps the most consistent way, but it is obviously expensive and at this stage we are keen to explore how detectability varies with e.g.
+GW strain.
 
-There are a few terms to unpack and define in this equation.
-$q(t)$ is the vector from the Earth to the pulsar, $n$ the propagation direction of the GW, $\Omega$ the (constant) angular frequency of the GW and $d$ the Earth-pulsar distance.
-$h_{ij}(t)$ is the metric perturbation due to the GW $= g_{ij} - \eta_{ij}$.
-It is given by the familiar plane-wave relation:
-\begin{equation}
-	h_{ij}(t) = H_{ij} e^{(i(\bar{k} \cdot \bar{x} - \Omega t + \Phi_0))}
-\end{equation}
-where $\bar{k}$ is the GW 3-wavevector $=\Omega \bar{n}$ and $x = -\bar{q} t$ (photon propagates from pulsar to Earth), and for phase offset (GW phase at Earth when $t=0$) $\Phi_0$.
-We can also write this as,
-\begin{equation}
-h_{ij}(t) = H_{ij} e^{(-i\Omega t (1 + \bar{n} \cdot \bar{q} ) + \Phi_0))}
-\end{equation}
-The amplitude tensor  $H_{ij}$ is
-\begin{equation}
-	H_{ij} = h_+ e_{ij}^+(\bar{n},\psi) + h_{\times} e_{ij}^{\times}(\bar{n},\psi)
-\end{equation}
-where $h_{+,\times}$ are the constant amplitudes of the gravitational plane wave, and $e_{ij}^{+, \times}(\bar{n}, \psi)$ are the polarisation tensors which are uniquely defined by the principal axes of the GW.
-$\psi$ is the polarisation angle of the GW.
-\newline
+2.
+The second method is to recognize that $M_0$ and $M_1$ are hierarchically nested models and we can perform a likelihood ratio test.
+That is, given the maximum likelihood estimators $\hat{\theta}$ of the true parameters $\theta$, the likelihood of each model can be calculate.
+These likelihoods are just  point estimates of the Bayes factor numerator/denominators.
+They can then be compared via the likelihood ratio $\Lambda$.
 
-\noindent Going forward,  \newline
+Given the cheap cost we proceed with the second method.
 
-\noindent Bringing this together, and adopting a trigonometric form of the equations, we can express the measurement equation as
+Likelihood ratio test
+For the likelihood ratio test we do not perform any kind of maximum likelihood search over the parameters for each of the models.
+Instead we just artificially set the maximum likelihood estimators to be equal to the true parameters of the system i.e.
+$\hat{\theta} = \theta$.
+We assume that any maximum likelihood algorithm would converge to these parameters.
+This is obviously an oversimplification but will serve our purposes for now.
 
-\subsection{Parameters of the model}
-\noindent
+Interpreting the likelihood ratio $\Lambda$ also needs some consideration, since we have to account for the increased model complexity of $M_1$.
+Bayes factors penalise complexity by construction since one must integrate over a larger parameter space.
 
-We can also express the measurement equation generally as,
+There are many different ways to do this:
 
-Kalman filtering works as follows
+* AIC
+* BIC
+* Wilks Theorem
+* ...
 
-you can see that it maps nicel onto our state separation
+In practice we will use different criteria to check they agree.
 
-\section{Methods}
-
-There are two potential methods we can take
-
-\begin{enumerate}
-	\item Let the states just be the frequencies $\bar{x} = (f)$.
-We have a linear equation for both the state evolution and the measurement matrix.
-Both of these depend on the parameters $\theta$.
-Use the likelihood output by a standard Kalman filter to run a nested sampler and try to recover $\theta$.
-	\item Let the states be the frequencies and all the unknown parameters $\bar{x} = (f, \bar{\theta}_{\rm GW},\bar{\theta}_{\rm PSR},\bar{\theta}_{\rm noise})$.
-Our state evolution is now linear but our measurement matrix is non-linear - a function of the states(parameters).
-Use a non-linear estimator such as a EKF/UKF to try to recover all of the states at once.
-\end{enumerate}
-
-\subsection{Method 1}
-
-In the case where the states are just the intrinsic pulsar frequencies, we can write the ODEs in matrix form as
-\begin{equation}
-	d\bar{X} = \bar{A} \bar{X} dt + \bar{N}(t) dt + \bar{\Sigma} d\bar{B}(t) 
-\end{equation}
-where $\bar{A} = \text{diag}(\gamma_1, \gamma_2,...)$,  $\bar{X} = \text{diag}(f_1, f_2,...)$, $\bar{N} =\text{diag}(\gamma_1[a_1+b_1t] + b_1, ...)$
-and we have let $a = f_{EM}(0)$, $b = \dot{f}_{\rm EM}(0)$.
+\begin{figure}
+	%\centering % Not needed
+	\begin{subfigure}[b]{1\columnwidth}
+		\includegraphics[width=\textwidth]{images/snr_canonical}
+		\caption{}
+		\label{fig:6MB_BFS}
+	\end{subfigure}
+	\\
+	\begin{subfigure}[b]{1\columnwidth}
+		\includegraphics[width=\textwidth]{images/snr_canonical_1psr}
+		\caption{}
+		\label{fig:25MB_bfs}
+	\end{subfigure}
+	\caption{TK: caption needed}
+	\label{fig:four figures}
+\end{figure}
 
@@
 Radiomenter noise. In theory well known but this could also be estimated via this approach
+
+Quality vs quantity: 
+This is important as it measn we dont need to do this for all pulsars, just a few. 
+%https://arxiv.org/abs/2211.03201
 

### 2023-04-11 py_src/priors.py

@@ -151,8 +151,8 @@ def bilby_priors_dict(PTA,P):
 
 
     init_parameters["alpha_gw"] = None
-    priors["alpha_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'alpha_gw')
- #   priors["alpha_gw"] = P["alpha_gw"]
+    #priors["alpha_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'alpha_gw')
+    priors["alpha_gw"] = P["alpha_gw"]
 
 
     init_parameters["h"] = None

### 2023-04-12 docs/manuscript/paper.tex
more paper updates

@@
 \end{equation}
-For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
+for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
 \begin{equation}
@@
 As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g.
-a model of the neutron star crust and superfluid components][]{Meyers2021}.
-However, for our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.
+a model of the neutron star crust and superfluid, components][]{Meyers2021}.
+For our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.
 
 \subsection{Modulation of pulsar frequency due to a GW}
-In the presence of a GW, the $f(t)$ measured by an observer on Earth is different from that measured by an observer in the local NS reference frame.
-We want to determine how the GW influences the received frequency
+In the presence of a GW, the $f(t)$  measured by an observer in the local NS reference frame is different from that measured by an observer on Earth.
+We want to determine quantitatively determine this difference - how the GW influences the received pulse frequency.
 
@@
 \end{equation}
-for spatial coordinates $\bar{x}$, \textcolor{red}{(TK: We have a notation conflict between $x$ = state and $x$ = coordinates)} where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
+for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, \textcolor{red}{(TK: We have a notation conflict between $x$ = state and $x$ = coordinates)} where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
 We emphasise that $\Omega$ has no time dependence - we are concerned solely with monochromatic sources.
-Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{\textbf{TK: This point is important, since the phase offset is then the same between multiple pulsars.
+Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{\textcolor{red}{TK: This point is important, since the phase offset is then the same between multiple pulsars.
 See also Melatos 2022 PT12}}.
@@
 where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave. The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
-\begin{equation}
-	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b
-\end{equation}
-\begin{equation}
+\begin{align}
+	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
 		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
-\end{equation}
+\end{align}
 which are in turn specified via the location of the GW source on the sky (via $\theta$, $\phi$ coordinates) and the polarisation angle $\psi$
@@
 Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.
-\footnote{\textbf{TK: See also e.g.
+\footnote{\textcolor{red}{TK: See also e.g.
 Maggiore, Meltaos who takes the same approach...}} \newline
@@
 \end{equation}
-\footnote{\textbf{TK: This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset}.}
+\footnote{\textcolor{red}{TK: This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset}.}
 
@@
 It will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar and $\omega$ is the \textit{constant} photon angular frequency.
-Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer \footnote{see Melatos 2022, PT16}.
+Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer \footnote{ \textcolor{red}{TK: see Melatos 2022, PT16}}.
 We will consider  the pulsar locations to be constant with respect to the Earth.
-TK: \textbf{This may have already been "done" during the barycentreing when pulsar TOAs are generated, in which case $q$ is the vector from the SSB to the pulsar}.
-Similarly we can express the spatial coordinates as $\bar{x}(t) = -\bar{q}t$.
+In practice, the pulsar locations vary with respect to the Earth, but are constant with respect to the solar system barycentre (SSB).
+This barycentering correction is typically applied during the pulsar detection process.
+Our conclusions are generally unchanged by this choice - it is straightforward to transform $q$ if needed into a vector between the SSB and pulsar.
+The spatial coordinates along the photon path are related to the pulsar as $\bar{x}(t) = -\bar{q}t$.
 \newline
@@
 \end{equation}
-\footnote{compare with Melatos 2022, PT11}
+\footnote{\textcolor{red}{TK: compare with Melatos 2022, PT11}}
 \noindent and
@@
 \subsection{State-space structure} \label{sec3}
-We now have a model for the evolution of the pulsar frequency states and the measured frequency timeseries.
+We now have a model for the evolution of the intrinsic pulsar frequency states and the measured frequency timeseries.
 We can express the intrinsic frequency evolution, Eq.
@@
 \begin{equation}
-	df = A f dt + N(t) dt + \sigma dB(t)
+	df = \mathcal{A} f dt + N(t) dt + \sigma dB(t)
 	\label{eq:state1}
 \end{equation}
-where $A = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process).
-This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by,
+where $\mathcal{A} = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process).
+This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by \citep{gardiner2009stochastic},
 \begin{equation}
-	f(t) = e^{At}f(0) + \int_0^t e^{A(t-t')} N(t') dt' + \int_0^t e^{A(t-t')} \sigma dB(t')
+	f(t) = e^{\mathcal{A}t}f(0) + \int_0^t e^{\mathcal{A}(t-t')} N(t') dt' + \int_0^t e^{\mathcal{A}(t-t')} \sigma dB(t')
 \end{equation} 
@@
 \begin{equation}
-	f_M = f g(\bar{\theta},t) + N_M
+	f_M (\bar{t})= f g(\bar{\theta},\bar{t}) + N_M
 \end{equation}
-where $g(\theta,t)$ ("measurement function")is a function of some parameters, $\bar{\theta}$ and time $t$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
+where $g(\theta,t)$ ("measurement function") is a function of some constant parameters, $\bar{\theta}$, and discrete time $\bar{t}$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
 \begin{equation}
@@
 \end{equation}
-The measurement function follows from Eq. \ref{eq:freq_shift}
+for variance $\Sigma^2$. The measurement function follows from Eq. \ref{eq:freq_shift}
 \begin{equation} \label{eq:final}
@@
 \end{equation}
-or in a more concise form where we have switched to a trigonometric form of the equations,
+or in a more concise form where we have switched to a trigonometric form of the equations and made the time dependence explicit,
 \begin{equation} 
@@
 \section{Detection and parameter estimation}
-Given the proceeding state-space model we will now try to use this model to infer the unknown parameters of the system.
-Let's first review and categorise all the parameters of model.
+Let's  review and categorise all the parameters of above state-space model.
 We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar, the GW parameters and the noise parameters i.e.
@@
 \begin{equation}
-	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]
+	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]^{(N)}
 \end{equation}
@@
 \end{equation}
+\textcolor{red}{TK: $\sigma$ is currently a parameter shared between all pulsars in current code implementation.
+Probably need to update this since different pulsars have different process noise, but need to check identifiability and likelihood curves - only enters the Kalman filter only through the process noise matrix...}.
 Whilst the GW parameters are shared between measurements for each pulsar, the pulsar parameters are clearly not.
-For a PTA dataset of N pulsars we have $7 + 5N$ parameters to estimate.
+For a PTA dataset of $N$ pulsars we have $7 + 5N$ parameters to estimate.
 For now we will consider the measurement noise to be known, although in principle this too could be estimated.
@@
 
-\noindent We will use a Kalman filter in order to obtain likelihood estimates of the data given a set of parameters.
-We will now review the Kalman filter and the associated likelihood in Section \label{sec:kalman_filter} before going on to discuss how to use nested sampling methods in conjunction with the filter for detection (model selection) and parameter estimation
+\noindent We want be able to infer the values of the model parameters and look for evidence of a GW signal in a data stream.
+To this end, we will use a Kalman filter in to obtain likelihood estimates of the data given a set of parameters.
+We will then deploy the Kalman filter in conjunction with a nested sampling technique for optimisation of the parameters over the likelihood.
+We will now review the Kalman filter and the associated likelihood in Section \ref{sec:kalman_filter} before going on in Section \ref{sec:nested_sampling} to discuss how to use nested sampling methods together with the filter for detection (model selection) and parameter estimation.
+We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.
 
 \subsection{Kalman Filtering}\label{sec:kalman_filter}
+\begin{figure*}
+	%\centering % Not needed
+	\begin{subfigure}[b]{1\columnwidth}
+		\includegraphics[width=\textwidth]{images/KF1}
+		\caption{Using correct parameters}
+		\label{fig:6MB_BFS}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{1\columnwidth}
+		\includegraphics[width=\textwidth]{images/KF2}
+		\caption{Using incorrect parameters}
+		\label{fig:25MB_bfs}
+	\end{subfigure}
+	\caption{Application of a Kalman filter to recover the intrinsic frequency states (top panels) given a measured frequency signal (bottom panels) which has been modulated by the presence of a gravitational wave.
+In the case where the parameters of the filter are accurate (subfigure (a)) the underlying states are also recovered accurately.
+Conversely, when the parameters fed to the filter are inaccurate (subfigure (b)) the filter cannot recover the state.
+\textcolor{red}{TK: this needs to be just one stacked figure.}}
+	\label{fig:four figures}
+\end{figure*}
+
 The Kalman filter \citep{Kalman1} is a algorithmic technique for recovering a set of system state variables, $\bar{x}$, given some noisy measurements, $\bar{z}$.
@@
 \begin{equation}
-	z = Hx + v
+	\bar{z} = \bar{H} \bar{x} + \bar{v}
 \end{equation}
-where $H$ is the measurement matrix and $v$ a Gaussian measurement noise. The underlying states are the solutions to the state-space equation 
+where $\bar{H}$ is the measurement matrix and $\bar{v}$ a Gaussian measurement noise. The underlying states are the solutions to the state-space equation 
 \begin{equation}
-	\dot{x} = Fx + Gu + w
+	\dot{\bar{x}} = \bar{F} \bar{x} + \bar{G}\bar{u} + \bar{w}
 \end{equation}
-where $x$ are the state variables, $F$ the system dynamics matrix, $G u$ the control matrix/vector and $w$ a stochastic zero-mean process which .
+for the system dynamics matrix,$\bar{F}$,  control model $G$, control vector $\bar{u}$, and $w$ a stochastic zero-mean process.
 By comparison with the preceding equations, Eqs.
@@
 Specifically, our states are just the $N$ intrinsic pulsar frequencies $\bar{x} = (f_1,f_2,...f_N)$ whilst our measurements are the $N$ measured pulse frequencies $\bar{z} = (f^{(M)}_1,f^{(M)}_2,...f^{(M)}_N)$.
-If we specialize to the case of constant time sampling between our observations,$\Delta t$, then for our formulation the components that make up the Kalman filter are as follows:
+If we specialize to the case of constant time sampling between our observations, $\Delta t$, then for our formulation the components that make up the Kalman filter are as follows:
 \begin{equation}
 	F_i = F_{i+1} = e^{-\gamma \Delta t}
+	\label{eq:fmatrix}
 \end{equation}
@@
 \end{equation}
-and $A$ is a constant that is given by  Eq.
-\ref{eq:state2}.
-The Kalman filter includes the effect of process noise $w$ and the measurement noise $v$ via the definition of a process noise matrix $Q = E[w w^T]$ and a measurement noise matrix $R = E[v v^T]$, which have the discrete form,
+where the $i$ subscript labels the value at the $i$-th timestep, and $A$ is a constant that is given by  Eq. \ref{eq:state2}. \newline
+
+\noindent The Kalman filter includes the effect of process noise $\bar{w}$ and the measurement noise $\bar{v}$ via the definition of a process noise matrix $Q = E[w w^T]$ and a measurement noise matrix $R = E[v v^T]$, which have the discrete form,
 \begin{equation}
@@
 	R_i = R_{i+1} = \Sigma^2
+	\label{eq:rmatix}
 \end{equation}
+The above equations, Eqs.
+\ref{eq:fmatrix} - 	\ref{eq:rmatix}, apply to an operation on a single state.
+The extension to $N$ states is straightforward, since one just needs to construct a diagonal matrix for each of the Kalman components where each non-zero element corresponds to a separate pulsar frequency.
 For our formulation we are concerned only with the linear Kalman filter since the measurements are a linear function of the states and the state transitions are also linear.
-Extension to non-linear problems is straightforward using either an extended Kalman filter, the unscented Kalman filter or the particle filter.
-For a full review of the  Kalman filter equations which may bee familiar to some, we refer the reader to the appendices of \cite{Melatos2023} and  \cite{Meyers2021}.
+Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, the unscented Kalman filter \citep{882463van} or the particle filter \citep{Simon10}.
+For a full review of the  Kalman filter equations in an astrophysical context, which may be unfamiliar to some, we refer the reader to the appendices of \cite{Melatos2023} and  \cite{Meyers2021}.
 \newline
@@
 Fig \ref{fig:four figures}).
+The filter tracks the error in its predictions of the underlying states by projecting the state predictions back into measurement space, $\hat{x} \rightarrow \hat{z}$.
+These measurement predictions can then be compared against the true observed measurements i.e.
+$ \epsilon = \bar{z} - \hat{z}$.
+The Gaussian log-likelihood at timestep $i$, as a function of the parameters is,
+\begin{eqnarray}
+	\log \mathcal{L}_i =  -\frac{1}{2} (N \log 2 \pi + \log \det |S_i| + \epsilon_i^T S_i^{-1} \epsilon_i)
+\end{eqnarray}
+where $S_i$ is the covariance matrix of the error $\epsilon_i$. The net log-likelihood is simply the sum over all timesteps.
 
-The filter tracks the error in its predictions of the underlying states by projecting the state predictions back into measurement space $\hat{z}$.
-These measurement predictions can then be compared against the true observed measurements.
-The Gaussian log-likelihood as a function of the parameters is,
+\subsection{Nested Sampling}\label{sec:nested_sampling}
+Nested Sampling \citep{Skilling} is an integration algorithm used for evaluating integrals of the Bayesian evidence form,
 \begin{eqnarray}
-	\log \mathcal{L} =  -\frac{1}{2} (N \log 2 \pi + \log \sigma^2 + \frac{y^2}{x})
+	Z = \int \mathcal{L} \pi (\theta) d \theta
 \end{eqnarray}
-where this is this is that....
-This likelihood provided by the Kalman filter can then be used for Bayesian inference based techniques of the underlying system parameters, which we will now explore in more detail.
+where $\pi (\theta)$ is the prior which is a function of a set of unknown parameters $\theta$.
+The primary advantage of nested sampling is the ability to compute this evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches.
+Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}.
+For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
+Within this work we also use nested sampling for parameter estimation and model selection.
+\newline
 
-\begin{figure*}
-	%\centering % Not needed
-	\begin{subfigure}[b]{1\columnwidth}
-		\includegraphics[width=\textwidth]{images/KF1}
-		\caption{}
-		\label{fig:6MB_BFS}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{1\columnwidth}
-		\includegraphics[width=\textwidth]{images/KF2}
-		\caption{}
-		\label{fig:25MB_bfs}
-	\end{subfigure}
-\caption{TK: caption needed}
-	\label{fig:four figures}
-\end{figure*}
+\noindent Multiple nested sampling libraries exist.
+For gravitational astrophysics it is common to use the dynesty sampler CITE, via the Bilby gravitational wave inference library and we continue to follow this precedent.
+\textcolor{red}{TK: we may end up not even using Bilby/dynesty so leaving this section unfilled for now. Do we need a very short description of how nested sampling works?}
 
-\subsection{Nested Sampling}
-Nested Sampling \cite{Skilling} is an integration algorithm used for evaluating integrals of the Bayesian form
-\begin{eqnarray}
-	Z = \int L \pi (\theta) d \theta
-\end{eqnarray}
-where this is this is that...
-i.e.
-the model evidence.
-It is a highly advantageous approach over the usual MCMC-type approaches which are used solely for calculating the posterior but can be highly computationally expensive, does not return an evidence, and can struggle on multi-modal problems.
-
-Within the astrophysical community, it was first adopted in cosmology, CITE CITE but has become a widespread technique used wihin astronomy, astrophysics, particle physics and materials science CITE
-
-Nested sampling genearlly works by algorithmically evolving a set of "live points" that seek to describe an underlying probability distribution.
-For a full review we refer the reader to original works by \cite{Skilling} and a recent review from a physics sciences perspective from \textbf{ASHTON}
-
-Multiple nested sampling libraries exist.
-For gravitational astrophysics it is common to use the dynesty sampler CITE, via the Bilby gravitational wave inference library and we continue to follow this precedent.
-
-\subsection{PTA pulsars}
-
-In order to proceed, it is now necessary to specify a PTA configuration.
-
-As discussed, multiple separate PTA detectors exist under the umbrella of the IPTA.
-Going forward we will take the 47 pulsars that make up the NANOGrav PTA \citep{2020ApJ...905L..34A}.
-NANOGrav is selected simply as a well-representative example of the typical pulsars that make up a PTA.
-Our results and formulation are not contingent on the choice of PTA, and naturally extend to other PTAs or PTAs with more pulsars.
-
-Within the state-space forumulation, the pulsar evolution is governed by a set of 5 parameters $\bar{\theta}_{\rm PSR}$ for each pulsar.
-The parameters $f_{\rm EM} (0) $,$ \dot{f}_{\rm EM} (0)$ and $d$ are well specified via existing pulsar datasets.
-We take the frequency and frequency derivative as returned from the pulsar datasets to simply be the values now at $t=0$.
-The pulsar distances are also known, though less well constrained.
-Going forward we take the distances retuned from the datasets as the true values of the pulsars that make up our synthetic PTA.
-
-The specification of $\gamma$ and $\sigma$ are more involved. $\gamma$ specifies an effective timescale of reversion to the mean
-
-\textbf{TK: discussion on how to select these two parameters c.f. Andres }
-
-In order to proceed and explore how well this state-space formulation works, we will need to specify a selection of pulsars to make up our PTA.
-We will take the 47 pulsars that make up the NANOGrav PTA.
-
-For each pulsar, we need to specify the complete set of $\bar{\theta}_{\rm PSR}$ as well as $\sigma$.
-The parameters $f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d$ are straightforward to set and can be read directly from  the current "present day" best estimates of the pulsar frequency, derivative and distance.
-
+\subsection{PTA pulsars}\label{sec:pta_pulsars}
 \begin{figure}
@@
 \end{figure}
+\noindent With our Kalman filter and nested sampling techniques in hand, in order to proceed it is necessary to specify a PTA configuration.
+As discussed, multiple separate PTA detectors exist under the umbrella of the IPTA.
+Going forward we will take the 47 pulsars that make up the NANOGrav PTA \citep{2020ApJ...905L..34A}.
+NANOGrav is selected simply as a well-representative example of the typical pulsars that make up a PTA.
+Our results and formulation are not contingent on the choice of PTA, and naturally extend to other PTAs or PTAs with more pulsars.
+\newline
 
-\subsection{Parameter estimation on synthetic data}
-We are now in a position to try to infer the parameters of our system
+\noindent Within our state-space formulation, the pulsar evolution is governed by a set of 5 parameters, $\bar{\theta}_{\rm PSR}$ for each pulsar.
+The parameters $f_{\rm EM} (0) $,$ \dot{f}_{\rm EM} (0)$ and $d$ are well specified via existing pulsar datasets.
+We take the frequency and frequency derivative as returned from the pulsar datasets to simply be the values now at $t=0$.
+The pulsar distances are also known, though less well constrained.
+Going forward we take the distances retuned from the datasets as the true values of the pulsars that make up our synthetic PTA.
+The specification of $\gamma$ and $\sigma$ are more involved.
+$\gamma$ specifies an effective timescale of reversion to the mean \textcolor{red}{TK: Need some discussion and though on how to choose these two parameters to be astrophysically reasonable c.f.
+A.
+Vargas}
 
+\subsection{Parameter estimation on synthetic data}\label{sec:param_estimation}
+We are now in a position to try to infer the parameters of a GW system.
+We take our NANOGrav PTA and create a synthetic noisy dataset using the parameters described in Table \ref{tab:toy_example_parameters}.
+\begin{table}
+	\centering
+	\begin{tabular}{lc}
+		Parameter & Value  \\
+		\hline
+		$\omega$       & $5 \times 10^{-7}$  \\
+		$\alpha$          & $1.0$  \\
+		$\delta$              & $1.0$   \\
+		$\psi$              & $2.50$  \\
+		$\iota$             & $1.0$  \\
+		$\Phi_0$          & $0.20$  \\
+		$h$ & 1e-2 \\
+		$\sigma_{\rm p}$ & $10^{-13}$ \\
+		$\sigma_{\rm m}$ & $10^{-10}$ \\
+		$T_{obs}$ & 10 \text{ years} \\
+		$\Delta t$ & 7 \text{ days} \\
+		\hline
+	\end{tabular}
+	\caption{GW parameters used for generating synthetic data. There is nothing special about these parameters - they are just chosen arbitrarily.}
+	\label{tab:toy_example_parameters}
+\end{table}
+\noindent Given this synthetic data, we can try to use our KF + NS approach to recover some of the parameters.
+The results for inference on 5 parameters, $\omega.
+\Phi_0, \psi, \iota, h$ are shown in Fig \ref{fig:corner}.
+\textcolor{red}{TK: likelihood plots for each parameter could also be of interest to include here?
+Need to expand this section greatly for more astrophysical systems and more unknown parameters}
 \begin{figure*}
 	\includegraphics[width=0.8\textwidth]{images/corner_example}
-	\caption{Example corner plot}
+	\caption{Example corner plot. Aside: for this run the $\sigma_p$ value fed to the filter is different from that used to generate the data.}
 	\label{fig:corner}
@@
 
-\subsection{Detection on synthetic data}
+\subsection{Detection on synthetic data}\label{sec:detection}
 
-We are interested in how detectable a GW is using PTA + state space method
+In addition to estimating the parameters of the system, we are also interested in how detectable a GW is using PTA + state space method.
+We can use our state-space tools  to try to solve the problem of GW detection with a PTA i.e.
+\textit{"Is there evidence of a GW in my data?"}We can frame this as a model selection procedure where we have two models/hypotheses:
 
-We can use our state-space tools  to try to solve the problem of GW detection with a PTA i.e. "Is there evidence of a GW in my data?.
-
-We can frame this as a model selection procedure where we have two models/hypotheses:
-
-* **Null Model** $M_0$. There is no GW in the data. In this case the measurement model of the Kalman filter simply returns the frequency states i.e. $g(t,\theta)= 0$
-* **Alternative model** $M_1$. There is a GW in the data. The measurement model uses the full expression for $g(t,\theta)$
+\begin{itemize}
+\item Null Model, $M_0$. There is no GW in the data. In this case the measurement model of the Kalman filter simply returns the frequency states i.e. $g(t,\theta)= 0$
+\item Alternative model, $M_1$. There is a GW in the data. The measurement model uses the full expression for $g(t,\theta)$
+\end{itemize}
 
@@
 
-1.
-The first is a fully Bayesian search over all the parameters for each model, calculating the evidence for each model and then determining the Bayes ratio.
+\begin{enumerate}
+	\item The first is a fully Bayesian search over all the parameters for each model, calculating the evidence for each model and then determining the Bayes ratio.
 This is perhaps the most consistent way, but it is obviously expensive and at this stage we are keen to explore how detectability varies with e.g.
 GW strain.
-
-2.
-The second method is to recognize that $M_0$ and $M_1$ are hierarchically nested models and we can perform a likelihood ratio test.
+	\item The second method is to recognize that $M_0$ and $M_1$ are hierarchically nested models and we can perform a likelihood ratio test.
 That is, given the maximum likelihood estimators $\hat{\theta}$ of the true parameters $\theta$, the likelihood of each model can be calculate.
@@
 They can then be compared via the likelihood ratio $\Lambda$.
-
+\end{enumerate}
 Given the cheap cost we proceed with the second method.
-
-Likelihood ratio test
 For the likelihood ratio test we do not perform any kind of maximum likelihood search over the parameters for each of the models.
@@
 This is obviously an oversimplification but will serve our purposes for now.
+\newline
 
-Interpreting the likelihood ratio $\Lambda$ also needs some consideration, since we have to account for the increased model complexity of $M_1$.
+\noindent Interpreting the likelihood ratio $\Lambda$ also needs some consideration, since we have to account for the increased model complexity of $M_1$.
 Bayes factors penalise complexity by construction since one must integrate over a larger parameter space.
+There are many different ways to do this, AIC, BIC, Wilks Theorem...
+\newline
 
-There are many different ways to do this:
-
-* AIC
-* BIC
-* Wilks Theorem
-* ...
-
-In practice we will use different criteria to check they agree.
+\noindent
 
@@
 	\end{subfigure}
-	\caption{TK: caption needed}
-	\label{fig:four figures}
+	\caption{Likelihood ratio vs strain for \textit{(a)} full PTA and \textit{(b)} a single randomly chosen pulsar.
+Horizontal dashed line shows the minimum detectable cutoff, vertical dashed line is the corresponding GW strain.}
+	\label{fig:snr_detection}
 \end{figure}
@@
 
+Points to discuss:
+
+\begin{itemize}
+	\item Non constant time sampling? Different pulsars sampled at different times...
+\end{itemize}
+
 \section{Conclusion}
 
-PTAs are all about MSPs for small timing noise. Maybe we can get away with large timing noise, and use more pulsars?
+Questions
 
-Radiomenter noise. In theory well known but this could also be estimated via this approach
-
-Quality vs quantity: 
-This is important as it measn we dont need to do this for all pulsars, just a few. 
-%https://arxiv.org/abs/2211.03201
+\begin{itemize}
+	\item PTAs like MSPs for small timing noise. Can we get away with large timing noise, and use more pulsars? Some pulsars are more useful than others, e.g. arXiv 2211.03201
+	\item Can we also estimate radiometer noise? Is this useful?
+\end{itemize}
 

### 2023-04-17 docs/manuscript/paper.tex
MCMC expts

@@
 The results for inference on 5 parameters, $\omega.
-\Phi_0, \psi, \iota, h$ are shown in Fig \ref{fig:corner}.
+\Phi_0, \psi, \iota, h$ are shown in Fig \ref{fig:corner2}.
 \textcolor{red}{TK: likelihood plots for each parameter could also be of interest to include here?
 Need to expand this section greatly for more astrophysical systems and more unknown parameters}
+%\begin{figure*}
+%	\includegraphics[width=0.8\textwidth]{images/corner_example}
+%	\caption{Example corner plot. Aside: for this run the $\sigma_p$ value fed to the filter is different from that used to generate the data.}
+%	\label{fig:corner}
+%\end{figure*}
+
 \begin{figure*}
-	\includegraphics[width=0.8\textwidth]{images/corner_example}
+	\includegraphics[width=0.8\textwidth]{images/canonical_example}
 	\caption{Example corner plot. Aside: for this run the $\sigma_p$ value fed to the filter is different from that used to generate the data.}
-	\label{fig:corner}
+	\label{fig:corner2}
 \end{figure*}
@@
 \begin{figure}
-	%\centering % Not needed
-	\begin{subfigure}[b]{1\columnwidth}
-		\includegraphics[width=\textwidth]{images/snr_canonical}
-		\caption{}
-		\label{fig:6MB_BFS}
-	\end{subfigure}
-	\\
-	\begin{subfigure}[b]{1\columnwidth}
-		\includegraphics[width=\textwidth]{images/snr_canonical_1psr}
-		\caption{}
-		\label{fig:25MB_bfs}
-	\end{subfigure}
+	\includegraphics[width=\columnwidth]{images/snr_canonical}
 	\caption{Likelihood ratio vs strain for \textit{(a)} full PTA and \textit{(b)} a single randomly chosen pulsar.
-Horizontal dashed line shows the minimum detectable cutoff, vertical dashed line is the corresponding GW strain.}
-	\label{fig:snr_detection}
+Horizontal dashed line shows the minimum detectable cutoff, vertical dashed lines are the corresponding GW strains.}
+	\label{fig:SNR}
 \end{figure}
+
+\begin{figure*}
+	\includegraphics[width=0.8\textwidth]{images/snr_mollewide}
+	\caption{stuff}
+	\label{fig:mollewide}
+\end{figure*}
+
+%
+%
+%
+%\begin{figure}
+%	%\centering % Not needed
+%	\begin{subfigure}[b]{1\columnwidth}
+%		\includegraphics[width=\textwidth]{images/snr_canonical}
+%		\caption{}
+%		\label{fig:6MB_BFS}
+%	\end{subfigure}
+%	\\
+%	\begin{subfigure}[b]{1\columnwidth}
+%		\includegraphics[width=\textwidth]{images/snr_canonical_1psr}
+%		\caption{}
+%		\label{fig:25MB_bfs}
+%	\end{subfigure}
+%	\caption{}
+%	\label{fig:snr_detection}
+%\end{figure}
 

### 2023-04-17 py_src/bilby_wrapper.py

@@ -30,15 +30,23 @@ def BilbySampler(KalmanFilter,init_parameters,priors,label,outdir):
 
  
     # #Run the sampler
-    print("RUN THE SAMPLER")
+    print("RUN THE SAMPLER - MCMC")
 
-    result = bilby.run_sampler(likelihood, priors, 
-                             label = label,
-                             outdir=outdir,
-                            sampler ='dynesty',
-                            check_point_plot=False,
-                            npoints=100,dlogz=0.1,
-                            npool=10, plot=True,resume=False)
+
+
+    result = bilby.run_sampler(likelihood, priors, sampler="bilby_mcmc", nsamples=1000,label="MCMCtestrun") #https://lscsoft.docs.ligo.org/bilby/bilby-mcmc-guide.html
+
+
+
+
+    # result = bilby.run_sampler(likelihood, priors, 
+    #                          label = label,
+    #                          outdir=outdir,
+    #                         sampler ='dynesty',
+    #                         sample='rwalk',
+    #                         check_point_plot=False,
+    #                         npoints=100,dlogz=0.1,
+    #                         npool=10, plot=True,resume=False)
 
 
 

### 2023-04-17 py_src/kalman_filter.py

@@ -15,8 +15,10 @@ The log likelihood, designed for diagonal matrices where S is considered as a ve
 def log_likelihood(S,innovation):
     x = innovation / S 
     N = len(x)
-    #slogdet = np.sum(np.log(S)) # Uses log rules and diagonality of covariance "matrix"
+    slogdet = np.sum(np.log(S)) # Uses log rules and diagonality of covariance "matrix"
     #return -0.5*(slogdet+innovation @ x + N*np.log(2*np.pi))
+    #return np.sum(innovation**2)
+
 
 
 

### 2023-04-17 py_src/likelihood_surface.py

@@ -12,24 +12,29 @@ import matplotlib.cm as cm
 import matplotlib.pyplot as plt 
 import matplotlib.colors as mc
 import numpy as np 
-
-#container = np.load("../data/omega_delta_heatmap_data.npz")
-
-container = np.load("../data/psi_gw_phi0_gw_surface_data.npz")
+import sys 
 
 
+
+
+p1 = sys.argv[1]
+p2= sys.argv[2]
+true_value_1= float(sys.argv[3])
+true_value_2= float(sys.argv[4])
+
+
+
+container = np.load(f"../data/{p1}_{p2}_surface_data2.npz")
 
 
 data_dict = {name: container[name] for name in container}
 
-y = data_dict["phi0_gw"]
-x = data_dict["psi_gw"]
+y = data_dict[p1]
+x = data_dict[p2]
 z = np.log10(np.abs(data_dict["likelihood"]))
 z = data_dict["likelihood"]
 
 print(np.max(z))
-
-
 Y,X = np.meshgrid(y,x)
 Z = z
 
@@ -40,11 +45,11 @@ Z = z
 
 fig = plt.figure(figsize=(12, 8))
 ax = fig.add_subplot(111, projection='3d')
-#ax.plot_surface(X, Y, Z.T, cmap='viridis', alpha=0.6)
-ax.scatter(X, Y, Z.T, color='black', alpha=0.5, linewidths=1)
+ax.plot_surface(X, Y, Z.T, cmap='viridis', alpha=0.6)
+#ax.scatter(X, Y, Z.T, color='black', alpha=0.5, linewidths=1)
 #ax.set(xlabel='$omega$', ylabel='$delta$')
 #ax.set_zlabel('$f(y_1, y_2)$', labelpad=10)
 
-ax.scatter(0.20,2.50,0,c='r',s=100)
+ax.scatter(true_value_2,true_value_1,np.max(Z),c='r',s=100)
 
 plt.show()

### 2023-04-17 py_src/main_swarm.py

@@ -1,0 +1,98 @@
+
+
+
+
+
+from system_parameters import SystemParameters
+from pulsars import Pulsars
+from synthetic_data import SyntheticData
+#from plotting import plot_all
+from model import LinearModel
+from kalman_filter import KalmanFilter
+from bilby_wrapper import BilbySampler
+from priors import priors_dict,bilby_priors_dict
+from bilby_wrapper import BilbyLikelihood
+
+import numpy as np
+import bilby
+import pyswarms as ps
+
+if __name__=="__main__":
+    import multiprocessing
+    multiprocessing.set_start_method("fork")
+
+
+    P   = SystemParameters()       #define the system parameters as a class
+    PTA = Pulsars(P)               #setup the PTA
+    data = SyntheticData(PTA,P) #generate some synthetic data
+    
+
+    #Define the model 
+    model = LinearModel
+
+    #Initialise the Kalman filter
+    KF = KalmanFilter(model,data.f_measured,PTA)
+
+    # Run the KF once with the correct parameters.
+    # This allows JIT precompile
+    guessed_parameters = priors_dict(PTA,P)
+    model_likelihood = KF.likelihood(guessed_parameters)
+    print("Ideal likelihood = ", model_likelihood)
+
+    def helper(y):
+        #print("helper y = ", y)
+        guessed_parameters = priors_dict(PTA,P)
+
+        guessed_parameters["alpha_gw"] = y[0]
+        guessed_parameters["delta_gw"] = y[1]
+       
+        return -KF.likelihood(guessed_parameters)
+
+    def cost_function(x):
+
+            n_particles = x.shape[0]  # number of particles
+
+
+            dist = [helper(x[i,:]) for i in range(n_particles)]
+
+
+            return np.array(dist)
+
+            # print("This is the cost function")
+            # print("The x value inputs is:")
+            # print(len(x))
+            # print(x)
+            # l_returns = np.zeros_like(x)
+
+            # for i in range(len(x)):
+            #     print(i)
+  
+                
+            #     print("Attemping l returns")
+            #     l_returns[i] = KF.likelihood(guessed_parameters)
+
+
+            # print(l_returns)
+            # return l_returns
+
+
+
+
+    # Set-up hyperparameters
+    options = {'c1': 0.5, 'c2': 0.3, 'w':0.3}
+
+    # Create bounds
+    max_bound =  np.pi * np.ones(2)
[…]

### 2023-04-17 py_src/priors.py

@@ -131,8 +131,8 @@ def bilby_priors_dict(PTA,P):
 
 
     init_parameters["phi0_gw"] = None
-    #priors["phi0_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'phi0_gw')
-    priors["phi0_gw"] =P["phi0_gw"]
+    priors["phi0_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'phi0_gw')
+    #priors["phi0_gw"] =P["phi0_gw"]
 
     init_parameters["psi_gw"] = None
     #priors["psi_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'psi_gw',boundary="periodic")
@@ -156,10 +156,10 @@ def bilby_priors_dict(PTA,P):
 
 
     init_parameters["h"] = None
-    priors["h"] = bilby.core.prior.LogUniform(1e-4, 1e0, 'h')
+    #priors["h"] = bilby.core.prior.LogUniform(1e-4, 1e0, 'h')
     #priors["h"] = bilby.core.prior.LogUniform(1e-11, 1e-9, 'h')
 
-    #priors["h"] = P["h"]
+    priors["h"] = P["h"]
 
 
    

### 2023-04-17 src/blackbox.jl

@@ -1,0 +1,77 @@
+
+
+
+
+using BlackBoxOptim
+
+function black_box(::Type{NF}=Float64;              # number format, use Float64 as default
+kwargs...                        # all additional non-default parameters
+) where {NF<:AbstractFloat}
+
+
+
+    println("attempting a black box optimisation")
+
+    state,measurements,PTA,θ̂,P = setup()
+  
+        optimisation_wrapper = let measurements = measurements, PTA = PTA, P=P
+
+        params -> begin 
+
+
+            ω=params[1]
+            δ = params[2]
+
+
+            #println(ω, " ", δ)
+
+            P1 = SystemParameters(NF=Float64,ω_guess=ω,δ_guess=δ) 
+            θ̂1 = kalman_parameters(PTA,P1)
+
+
+            ll_value,x_results = KF(measurements,
+                                    PTA,
+                                    θ̂1,
+                                    )
+            return -ll_value
+        end
+        end
+
+
+    truths = [θ̂.ω;θ̂.δ]
+    x0 = [2e-7;0.30]
+    lower_limits = [2e-9;-π/2.0]
+    upper_limits = [1e-6;π/2.0]
+
+
+
+
+    res = bboptimize(optimisation_wrapper; SearchRange = [(4e-7,6e-7 ),(0.9,1.1)],Method = :de_rand_1_bin, MaxSteps=50000)
+
+
+
+    # nparticles = Int64(10*length(truths))
+    # opt = optimize(optimisation_wrapper, 
+    #                x0, 
+    #                ParticleSwarm(lower_limits,upper_limits,nparticles),
+    #                Optim.Options(show_trace=true, store_trace=true,iterations = Int64(1e4),show_every=100))
+
+
+
+#opt = ParticleSwarm(optimisation_wrapper, x0)
+
+
+    println("Opt completed")
+    # display(opt)
+    # output = Optim.minimizer(opt)
+    # # #
+
+    # for i in 1:length(output)
+    # println(i, " ", truths[i], " ", output[i])
+    # end
+
+
+
+
+
+end 

### 2023-04-17 src/linear_kalman_filter.jl

@@ -96,34 +96,5 @@ function log_likelihood(S::Vector{NF}, innovation::Vector{NF}) where {NF<:Abstra
 end 
 
 
-# function read_vector(x::Vector{NF}) where {NF<:AbstractFloat}
 
-#     # Vector of length x 
-#     # The last 9 elements are single parameters shared between every pulsar, including σp 
-#     # The first elements are f0,ḟ0,d,γ
-#     Npsr = Int64((length(x) - 9) / 4)
-
-
-#     f0 = x[1:Npsr]
-#     ḟ0 = x[Npsr+1:2*Npsr]
-#     d  = x[2*Npsr+1:3*Npsr]
-#     γ  = x[3*Npsr+1:4*Npsr]
-
-#     ω  = x[4*Npsr + 1]
-#     Φ0 = x[4*Npsr + 2]
-#     ψ  = x[4*Npsr + 3]
-#     ι  = x[4*Npsr + 4]
-#     δ  = x[4*Npsr + 5]
-#     α  = x[4*Npsr + 6]
-#     h  = x[4*Npsr + 7]
-#     σp = x[4*Npsr + 8]
-#     σm = x[4*Npsr + 9]
-
-
-
-#     #return ω, h,ι,δ,α,ψ,Φ0,σp,σm,f0,ḟ0,d,γ 
-#     return h,ι,δ,α,ψ,Φ0,σp,σm,f0,ḟ0,d,γ 
-
-
-# end 
 

### 2023-04-17 src/system_parameters.jl

@@ -19,8 +19,8 @@ The default values of the keywords define the default model setup.
     Φ0     ::Real = 0.20      
     ψ      ::Real = 2.5 
     cos_ι  ::Real = 0.921060994 #this is cos(0.4)
-    δ  ::Real = 0.40
-    α :: Real = 4.80
+    δ  ::Real = 1.0
+    α :: Real = 1.0
     h :: Real = 1e-2
 
     #Noise parameters

### 2023-04-20 docs/manuscript/paper.tex
assorted clean up

@@
 Bayes factors penalise complexity by construction since one must integrate over a larger parameter space.
-There are many different ways to do this, AIC, BIC, Wilks Theorem...
+There are many different ways to do this - fot now we will consider two: Akaike information criterion (AIC) and Wilks Theorem.
+For the latter, Wilks' Theorem which states that for a large number of samples \footnote{What counts as large?
+See \url{https://www.osti.gov/servlets/purl/1529145}} the distribution of the test statistic approaches the chi-squared distribution under the null hypothesis i.e.
+\begin{equation}
+	2 \log \Lambda \rightarrow \chi^2
+\end{equation}
+One can then compute $p$-values where the number of degrees of freedom is equal to the difference in the number of parameters of the two models; $M_1$ has 7 extra parameters over $M_0$ corresponding to the 7 parameters of $\bar{\theta}_{\rm GW}$.
+With 7 degrees of freedom and a target tolerance of 5(1) \% the test statistic is $\sim 14 (18.5)$.
 \newline
+
+An alternative approach is to use the AIC which is given by,
+\begin{equation}
+	\text{AIC} = 2k - 2\log\ \mathcal{L} 
+\end{equation}
+where $k$ is the number of degrees of freedom. The AIC can be computed for each model and the model with the minimum AIC is preferred. This can be straightforwardly mapped into a relative condition
+\begin{equation}
+	\log \mathcal{L}_1 - 	\log \mathcal{L}_0 > 7 
+\end{equation}
+Note this is the same as the Wilks case!
 
@@
 	\includegraphics[width=0.8\textwidth]{images/snr_mollewide}
-	\caption{stuff}
+	\caption{\textcolor{red}{TK: Just a placeholder. Likelihood ratio for a given $h$. Maybe more interesting, but more expensive, to get minimum detectable h?}}
 	\label{fig:mollewide}

### 2023-04-20 docs/old_notes/LectureNotes_Math571.tex
change location of old notes

\documentclass{tufte-handout} % A4 paper and 11pt font size
\usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage{mathpazo} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm, amssymb} % Math packages
\usepackage{pgf,tikz}
\usetikzlibrary{positioning,matrix,arrows}
\usepackage{float}
\usepackage{tikz-cd}
\usepackage{caption}
\usepackage{stmaryrd}
\usepackage{multicol}
\usepackage{booktabs}
\usepackage{verbatim}
\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
\usepackage{sectsty} % Allows customizing section commands
\allsectionsfont{\normalfont \bfseries} % Make all sections centered, the default font and small caps
\usepackage{enumerate}
\usepackage{pythonhighlight}
\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header
\allowdisplaybreaks

\usepackage{graphicx}
\usepackage{subcaption}

\usepackage{hyperref}
\hypersetup{  
	colorlinks=true,
	urlcolor=cyan,
}

\urlstyle{same}

% Turn on numbering for section and subsection headings
\setcounter{secnumdepth}{2}

\geometry{
	left=13mm, % left margin
	textwidth=130mm, % main text block
	marginparsep=8mm, % gutter between main text block and margin notes
	marginparwidth=55mm % width of margin notes
}
\fontsize{10}{20}\selectfont
%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\title{	
	\normalfont\normalsize 
	{Melbourne University - Summer 2023} \\ [0pt] % Your university, school and/or department name(s)
	\huge Notes on state-space formulation of PTA problem% The assignment title
}\author{T. Kimpson} % Your name
\date{\vspace{-5pt}\normalsize\today} % Today's date or a custom date

\begin{document}
\justifying 
\maketitle

\pagenumbering{gobble} %turn off page numbering
\tableofcontents

\section{Preamble}

These notes collect some ideas around how to formulate the PTA data analysis as a state-space problem. \newline

\noindent We will frequently reference previous works from A.Melatos, in particular Melatos 2018 (GR assignment UniMelb) and Melatos 2022 (private communication at UniMelb). \newline

\noindent Geometric units ($c=\hbar = G = 1$) are used throughout, with the usual convention of Roman indices ($i,j$) for spatial indices. Note that $i$ is used for both labelling tensor indices and as the imaginary number - I trust the difference to be obvious from the context.\newline

\noindent These are rough notes - likely full of typos and undefined terms! Hopefully they are still understandable. \newline

\section{Derivation of pulsar frequency modulation due to GW}
We want to know how the pulse frequency from a pulsar is influenced by a passing gravitational wave (GW). \newline

\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$  \footnote{See Melatos 2022 for nuances and discussion around this construction of "pulse train as photon"} \newline

\noindent We take a gravitational plane wave that perturbs a background Minkowski spacetime as \footnote{c.f. Melatos 2018, Equation 1}
\begin{equation}
g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
where the GW has angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$. Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{This point is important, since the phase offset is then the same between multiple pulsars. See also Melatos 2022 PT12}. The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
\begin{align}
	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
\end{align}
\noindent Note that we use a convention where the amplitudes $h_{+,\times}$ label the \textit{constant} amplitudes of the gravitational plane wave. All time variation in the strain is a result of the exponential term $e^{i\Omega t}$. This is in contrast to the notation used by some authors \footnote{see e.g. arXiv:1003.0677} where the amplitudes include the time variability $h_{+,\times} = h_{+,\times}(t)$. The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave. \newline

\subsection{Setting up the problem}
The frequency of a photon with 4-momentum $p_{\mu}$ recorded by an observer with 4-velocity $u^{\mu}$ is 
\begin{equation}
	\nu = p_{\alpha} u^{\alpha}
\end{equation}

\noindent We consider both our emitter and receiver to be stationary, such that  
\begin{equation}
u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}

\noindent Consequently the frequency can be identified with the temporal component of the covariant 4-momentum,
\begin{equation}
\nu = p_t
\end{equation}

\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,

\begin{equation}
	p_t(\tau)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=\tau} \dot{p}_t dt
	\end{equation}

\noindent where the overdot denotes a derivative w.r.t. $t$. Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $\tau = t_0 + d$ and consider the photon trajectory to be an unperturbed path. \footnote{See also e.g. Maggiore who takes the same approach...} \newline

\noindent To complete our expression, we now just need to determine $\dot{p}_t$ and integrate it.

\subsection{Hamiltonian Mechanics}

The Hamiltonian in covariant notation can be written as

\begin{equation}
H(x^{\mu}, p_{\mu}) = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
\end{equation}

\noindent which if we substitute in our expression for the perturbed metric is

\begin{equation}
H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}

\noindent Hamilton's equations are
\begin{equation}
\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
\end{equation}

\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
\begin{equation}
\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t))+\Phi_0}
\end{equation}
\footnote{This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset.}

\noindent Therefore the derivative w.r.t coordinate time $t$ is,
\begin{equation}
\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
\end{equation}
\noindent To make contact with Melatos 2022 it will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar \footnote{similarly $\bar{x}(t) = -\bar{q}t$} and $\omega$ is the \textit{constant} photon angular frequency. Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer (see Melatos 2022, PT16). \newline

\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.  \newline

\noindent Bringing this all together we can write $\dot{p}_t$ in a condensed form as,
\begin{equation}
\dot{p}_t = A e^{i \gamma t + \Phi_0}
\end{equation}
where 
\begin{equation}
\gamma = -\Omega (1 + \bar{n}\cdot \bar{q}) 
\end{equation}
\footnote{compare with Melatos 2022, PT11}
\noindent and
\begin{equation}
A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
\end{equation}
where the $\omega$ term in the numerator results from a $\omega^2$ terms that comes from the $p^i p^j$ term and a $\omega^{-1}$ term from $(p^t)^{-1}$.
\subsection{Performing the integral}

The frequency shift experienced by the observer relative to the source due to a GW is then
\begin{align}
p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = A \int_{t = \tau - d}^{t=\tau} e^{i \gamma t + \Phi_0}dt  \\
=\frac{-iA}{\gamma} e^{i \gamma \tau}e^{\Phi_0}[1 - e^{-i \gamma d}] 
\label{eq:concise}
\end{align}

\subsection{Explicit expression and comparison with Melatos 22}

We can expand the concise expression of Eq. \ref{eq:concise} as,

\begin{equation} \label{eq:final}
p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{H_{ij}q^i q^j  }{(1 + \hat{n} \cdot \hat{q})}e^{-i \Omega \tau(1+\bar{n} \cdot \bar{q}) + \Phi_0} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
\end{equation}
This expression is equivalent to Melatos 2022, PT13 where the dependence on time $t$ is made explicit. Using the Melatos notation where $h_{ij} = g_{ij} - \eta_{ij}$

\begin{equation}
	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{h_{ij} q^i q^j  }{(1 + \hat{n} \cdot \hat{q})} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
\end{equation}

\newpage
\section{State-space structure} \label{sec3}
\noindent The general structure of a Kalman state-space problem is
\begin{equation}
	\dot{x} = f(x) + w
\end{equation}
where $x$ are the state variables, $f()$ a non-linear function of the states, and $w$ a stochastic zero-mean process. The process noise matrix $Q = E(w w^T)$. The states are related to the measurement $z$ via a non-linear measurement function $h()$
\begin{equation}
	z = h(x) + v
\end{equation}
where $v$ is a stochastic zero-mean process and measurement noise matrix $R = E(v v^T)$. \newline

\noindent This structure maps onto the preceding equations nicely. Our hidden state is just the intrinsic pulsar frequency, which evolves as,
\begin{eqnarray}
	\dot{f}_P = -\gamma f_P^n + \xi(t)
\end{eqnarray}
where $n$ is a braking index (typical values $\sim 1-5$, we take a canonical value $=3$) \footnote{See A. Vargas talk re anomalous braking indices. P.S. is there an arXiv preprint?} $\gamma$ a proportionality constant, and $\xi(t)$ represents a stochastic white noise process. Equation \ref{eq:final} can be used to relate the intrinsic pulsar frequency to that measured by an observer on Earth:

\begin{eqnarray}
	f_M = f_p(1 - X) + N_M
\end{eqnarray}
where $X$ is the RHS of Eq. \ref{eq:final} and $N_M$ a Gaussian measurement noise. An example of the evolution of the state and the measurement frequencies for our PTA pulsars can be seen in Fig. \ref{fig:mean and std of nets}

\subsection{A note on magnitudes}

It is worth pausing to consider some typical values of our state/measurement variables, as well as the respective noises.

\begin{itemize}
	\item $f_p$ - PTAs typically deal with millisecond pulsars. We can take a standard value of $f_P = 100$ Hz.
	\item $\gamma, n$ - These two parameters describe the secular spin down. We can take $n=3$ as a standard value (spindown due to dipole emission). It follows that $\gamma$ is of the order $10^{-23}$ for typical PTA pulsars (e.g. PSR J$0023+0923$ has $f\sim 330$ Hz and $\dot{f} \sim 10^{-15} s^{-2}$)
	\item $\sigma_p$ - this is the parameter which quantifies the strength of  $\xi(t)$ i.e. the pulsar red noise (sometimes known as timing noise). For MSPs the red noise is typically small - we can set it to be some fraction of $\dot{f}$ e.g. if 10$\%$ then $\sigma_p \sim 10^{-16}$.
	\item $X$ - The magnitude of $X$ is set by the GW amplitude, which is expected to be $\lesssim 10^{-15}$ 
	\item $\sigma_m$ - this is the parameter which quantifies the strength of the Gaussian measurement noise i.e. the radiometer white noise. Typical measurement uncertainty in the pulsar TOAs is $\sim 10 - 100$ ns for our best MSPs. If we take an observation cadence of $\sim 1$ month then we can relate the uncertainty in the TOA to the uncertainty in the frequency as $\sigma_m / f \sim 10 \text{ns} / 1 \text{ month}  \sim 4 \times 10^{-15}$. \footnote{L.Dunn gave some good advice on how this works, but I am still a bit unsure whether this quite makes sense?}
\end{itemize}

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.475\textwidth}
		\centering
		\includegraphics[width=\textwidth]{images/fp1}
		\caption{State frequency, single pulsar}%

		\label{fig:mean and std of net14}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.475\textwidth}  
		\centering 
		\includegraphics[width=\textwidth]{images/fp_all}
		\caption{State frequency, multiple pulsars}
		\label{fig:mean and std of net24}
	\end{subfigure}
	\vskip\baselineskip
	\begin{subfigure}[b]{0.475\textwidth}   
		\centering 
		\includegraphics[width=\textwidth]{images/fm1}
		\caption{Measured frequency, single pulsar}  
		\label{fig:mean and std of net34}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.475\textwidth}   
		\centering 
		\includegraphics[width=\textwidth]{images/fm_all}
		\caption{Measured frequency, multiple pulsars}   
		\label{fig:mean and std of net44}
	\end{subfigure}
	\caption{Evolution of the state (top row) and measurement (bottom row) frequencies for pulsars influenced by an arbitrary GW source.}
	\label{fig:mean and std of nets}
\end{figure}

\newpage
\section{Constructing a PTA}\label{sec:PTA}

In order to proceed and explore how well this state-space formulation works, we will need to specify a selection of pulsars to make up our PTA. We will take the 47 pulsars that make up the NANOGrav PTA \footnote{Data obtained from the ANTF catalogue at \url{https://www.atnf.csiro.au/research/pulsar/psrcat/}} \newline

\noindent ANTF provides values for both $f$ and $\dot{f}$. For every pulsar we will assume that $n=3$ and calculate $\gamma$ accordingly. For example, PSR J1713+0747 has a spin frequency of 218 Hz and a derivative $-4 \times 10^{-16}$ s $^{-2}$. Taking $n=3$ gives $\gamma \sim 3.9 \times 10^{-23}$. \newline

\noindent Approximate pulsar distances are also provided. For now we will take these as known rather than additional parameters to be inferred. The median pulsar distance for NANOGrav pulsars is $\sim 1.3$ kpc. The distribution of NANOGrav pulsars on the sky is shown in Fig \ref{fig:pulsar_distrib}

\begin{figure}
\includegraphics[width=0.8\textwidth]{images/pulsars}
\caption{Spatial distribution and distances of NANOGrav pulsars}
\label{fig:pulsar_distrib}
\end{figure}

\newpage 
\section{Model selection using a UKF}
\noindent We can construct some noisy synthetic data using the equations and values of Section \ref{sec3} and then apply an unscented Kalman filter \footnote{See \href{https://groups.seas.harvard.edu/courses/cs281/papers/unscented.pdf}{Wan \& van der Merwe} for a very readable description of how the UKF works } to the measured frequency to try to recover the intrinsic state frequency. An example of this is shown below in Fig \ref{fig:model example} below for some large GW strain. We can see that for this example the filter can recover the state evolution well.

\begin{figure}
	\includegraphics[width=0.8\textwidth]{images/model}
	\caption{Recovery of the pulsar state frequency from the measurement frequency by a UKF.}
	\label{fig:model example}
\end{figure}
\noindent We can use this UKF to try to solve the problem of GW detection i.e. \textit{"Is there evidence of a GW in my data?}. We can frame this as a model selection procedure where we have two models/hypotheses:

\begin{itemize}
	\item \textbf{Null Model} $M_0$. There is no GW in the data. In this case the measurement model of the UKF simply returns the frequency states (i.e. $X= 0$)
	\item \textbf{Alternative model} $M_1$. There is a GW in the data. The measurement model uses the full expression for $X$
\end{itemize}
\begin{figure}
	\includegraphics[width=0.8\textwidth]{images/null}
	\caption{As Fig 3, but now for a the null model. The state is not well recovered.}
	\label{fig:null example}
\end{figure}
\noindent An example applying the null model to the same data as in Fig \ref{fig:model example} is shown in Fig \ref{fig:null example}. Evidently the null model does not capture the true underlying state well. \newline

\noindent In order to accept the alternative hypothesis $M_1$ over $M_0$ there are two approaches we could take:
\begin{itemize}
	\item The first is a fully Bayesian search over all the parameters for each model, calculating the evidence for each model and then determining the Bayes ratio. This is perhaps the most consistent way, but it is obviously expensive and at this stage we are keen to explore how detectability varies with e.g. GW strain.
	\item The second method is to recognize that $M_0$ and $M_1$ are hierarchically nested models and we can perform a likelihood ratio test. That is, given the maximum likelihood estimators $\hat{\theta}$ of the true parameters $\theta$, the likelihood of each model can be calculated and compared. These likelihoods are just  point estimates of the Bayes factor numerator/denominators.
\end{itemize}
Given the cheap cost we proceed with the second method.

\subsection{Likelihood ratio test}
For the likelihood ratio test we do not perform any kind of maximum likelihood search over the parameters for each of the models. Instead we just artificially set the maximum likelihood estimators to be equal to the true parameters of the system \footnote{i.e. $\hat{\theta} = \theta$}. We assume that any maximum likelihood algorithm would converge to these parameters \footnote{This is obviously an oversimplification but will serve our purposes for now}. \newline

\noindent Interpreting the likelihood ratio $\Lambda$ also needs some consideration, since we have to account for the increased model complexity of $M_1$ \footnote{Bayes factors penalise complexity by construction since one must integrate over a larger parameter space}. This can be accomplished via Wilks' Theorem which states that for a large nuber of samples \footnote{What counts as large? See \url{https://www.osti.gov/servlets/purl/1529145}} the distribution of the test statistic approaches the chi-squared distribution under the null hypothesis i.e. 
\begin{equation}
2 \log \Lambda \rightarrow \chi^2
\end{equation}
One can then compute $p$-values where the number of degrees of freedom is equal to the difference in the number of parameters of the two models; $M_1$ has 7 extra parameters over $M_0$

\begin{itemize}
	\item Parameters $M_0$: $\gamma$, $n$
	\item Parameters $M_1$: $\gamma$, $n$, $\omega$, $\theta$, $\phi$, $\psi$, $\iota$, $\Phi_0$,$h$
\end{itemize}
With 7 degrees of freedom and a target tolerance of 5(1) \% the test statistic is $\sim 14 (18.5)$. For the toy system exhibited in Figs. \ref{fig:model  example}, \ref{fig:null example}, the test statistic is $\gg 20$ - this makes intuitive sense since it is immediately obvious to the eye that the data is sinusoidal. But what about other systems?

\subsection{Detectability vs strain}

Lets take a system and explore how the detectability, as quantified by the test statistic varies with GW strain. For our PTA we use the NANOGrav pulsars described in Section \ref{sec:PTA}, with an observation length of 10 years and a weekly cadence. The GW source parameters are listed in Table \ref{tab:toy_example_parameters}. The results for this system are presented in Figure \ref{fig:SNR}, for a single noise realisation. We consider both the test statistic, $2 \log \Lambda$, and another parameter $\delta$ which is the median absolute error in the state estimates \footnote{In real-life we of course cannot compute predicted state - actual state, but it is useful here for exploring the performance of the UKF.} From the Figure we can see that the signal is a smooth function of the strain and is detectable up to $h \sim 10^{-13}$. Below this limit the two models cease to be hierarchical and the error in the state estimates is highly comparable. This is the regime where the noise is large with respect to the strain and depending on the specific realisation of the noise either the null or the alternative models can be preferred i.e. we are just measuring noise. A brief exploration of other noise realisations and different number of pulsars is presented in Fig. \ref{SNR2}.

 \begin{table}
	\centering
	\begin{tabular}{lc}
		Parameter & Value  \\
		\hline
		$\omega$       & $10^{-7}$  \\
		$\theta$          & $1.0$  \\
		$\phi$              & $0.0$   \\
		$\psi$              & $2.5$  \\
		$\iota$             & $0.0$  \\
		$\Phi_0$          & $0.2$  \\
		$\sigma_{\rm p}$ & $10^{-13}$ \\
		$\sigma_{\rm m}$ & $10^{-13}$ \\
		\hline
	\end{tabular}
	\caption{GW parameters used for generating synthetic data. There is nothing special about these parameters - they are just chosen arbitrarily.}
	\label{tab:toy_example_parameters}
\end{table}

\begin{figure}
	\includegraphics[width=0.8\textwidth]{images/SNR2}
	\caption{Variation of the test statistic (\textit{Top panel}) and the median state error (\textit{Bottom panel}) w.r.t GW strain. The grey horizontal line labels a test statistic of 14. Below $h \sim 10^{-12}$ the models cease to be consistently hierarchical - values with negative test statistics are set automatically $=1$ and marked with a red dot.}
	\label{fig:SNR}
\end{figure}

\begin{figure}
	\centering

	\begin{subfigure}[b]{0.475\textwidth}  
		\centering 
		\includegraphics[width=\textwidth]{images/SNR2_differentnoise}
		\caption{Full PTA, different noise realisation}
		\label{fig:SNR2a}
	\end{subfigure}
	\vskip\baselineskip
	\begin{subfigure}[b]{0.475\textwidth}   
		\centering 
		\includegraphics[width=\textwidth]{images/SNR2_frac02}
		\caption{Only using 20 \% of pulsars}  
		\label{fig:SNR2b}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.475\textwidth}   
		\centering 
		\includegraphics[width=\textwidth]{images/SNR2_n2}
		\caption{Only using 2 pulsars.}   
		\label{fig:SNR2c}
	\end{subfigure}
	\caption{As Figure 5, but for different noises and numbers of pulsars}
	\label{SNR2}
\end{figure}

 \newpage 
 \section{Caveats and questions}

 \begin{enumerate}

 	\item We should really do this using a ``proper" language and enforce consistent number formats. We are using the Python long double, but there is some uncertainty around how e.g. linear algebra operations in Python handle these formats - do they just get rounded to doubles?

 	\item We have assumed that all pulsars have $n=3$. What about some distribution?

 	\item We have only considered a single GW source at a particular location, polarisation, etc. What about others?

 	\item We have focused only on a monochromatic GW source. This is probably not a bad approximation, but maybe we can include some process noise on the phase evolution too?

 	 \item We assume all pulsars have observations taken at the same time

 	\item How reasonable are our noises? For instance, what is a reasonable measurement noise on the frequency evolution of a MSP?

 	 \item What about doing a full Bayesian search using e.g. Bilby to get a Bayes ratio?

 	 \item How good/bad an assumption is it that $\hat{\theta} \sim \theta$ for our likelihood ratio test?
 \end{enumerate}

\end{document}

### 2023-04-20 docs/measurement_eqn_deerivation/LectureNotes_Math571.tex
notes on measurement equation derivation

--- 2023-04-20 docs/old_notes/LectureNotes_Math571.tex
@@
 
-\noindent These are rough notes - likely full of typos and undefined terms! Hopefully they are still understandable. \newline
-
 \section{Derivation of pulsar frequency modulation due to GW}
@@
 where the GW has angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
-Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{This point is important, since the phase offset is then the same between multiple pulsars.
-See also Melatos 2022 PT12}.
-The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
+Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.} The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
 \begin{align}
@@
 \noindent Note that we use a convention where the amplitudes $h_{+,\times}$ label the \textit{constant} amplitudes of the gravitational plane wave.
-All time variation in the strain is a result of the exponential term $e^{i\Omega t}$.
+All time variation in the strain is a result of the exponential term.
 This is in contrast to the notation used by some authors \footnote{see e.g.
@@
 \begin{equation}
-\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t))+\Phi_0}
+\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)}
 \end{equation}
@@
 \end{equation}
-\noindent To make contact with Melatos 2022 it will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar \footnote{similarly $\bar{x}(t) = -\bar{q}t$} and $\omega$ is the \textit{constant} photon angular frequency.
+It will be useful to recognise that we can express the photon momentum in terms of the  unit vector between the Earth and pulsar $\bar{q}$
+\begin{equation}
+	p^{\mu} = \omega(1,-q^x,-q^y,-q^z)
+\end{equation}
+where $\omega$ is the \textit{constant} photon angular frequency.
 Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer (see Melatos 2022, PT16).
+Similarly, we can parameterize the spatial coordinates $\bar{x}$ as,
+\begin{equation}
+	\bar{x}(t) = - \bar{q} (t - \tau)
+\end{equation} 
+Note that when our integration variable $t = \tau$ (i.e.
+the photon is at Earth) then $\bar{x}(t=\tau) = 0$ which is what we expect given our chosen coordinate system.
+Also note that $\dot{p}_t$ is entirely a function of the GW perturbation.
+In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.
 \newline
 
-\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.  \newline
+\noindent Bringing this all together
 
-\noindent Bringing this all together we can write $\dot{p}_t$ in a condensed form as,
+\begin{align}
+	\dot{p}_t &= \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right) \\  
+	&= -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)} \left(\frac{1}{p^t}\right)
+	\\
+	&= \frac{-i\Omega \omega^2}{2} H_{ij}q^i q^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)} \left(\frac{1}{\omega}\right) \\
+	& = \frac{-i\Omega \omega}{2} H_{ij}q^i q^j  e^{i(\Omega(\bar{n}\cdot [- \bar{q} (t - \tau)] - t)+\Phi_0)} \\
+				& = \frac{-i\Omega \omega}{2} H_{ij}q^i q^j  e^{i(-\Omega t (1+ \bar{n} \cdot \bar{q}) + \Omega \tau  \bar{n} \cdot \bar{q} +\Phi_0)} \\
+						& =A e^{i (\gamma t + B)}
+\end{align}
+with 
 \begin{equation}
-\dot{p}_t = A e^{i \gamma t + \Phi_0}
+	A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
 \end{equation}
-where 
 \begin{equation}
@@
 \footnote{compare with Melatos 2022, PT11}
-\noindent and
+and 
 \begin{equation}
-A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
+	B = \Omega \tau \bar{n} \cdot \bar{q}  +\Phi_0
 \end{equation}
-where the $\omega$ term in the numerator results from a $\omega^2$ terms that comes from the $p^i p^j$ term and a $\omega^{-1}$ term from $(p^t)^{-1}$.
+
 \subsection{Performing the integral}
@@
 \begin{align}
-p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = A \int_{t = \tau - d}^{t=\tau} e^{i \gamma t + \Phi_0}dt  \\
-=\frac{-iA}{\gamma} e^{i \gamma \tau}e^{\Phi_0}[1 - e^{-i \gamma d}] 
+p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} &= A \int_{t = \tau - d}^{t=\tau} e^{i (\gamma t + B)}dt  \\
+&=\frac{-iA}{\gamma} e^{i B} \left(e^{i \gamma \tau} -e^{i \gamma (\tau - d)}\right) \\
+&=\frac{-iA}{\gamma} e^{i \Omega \tau \bar{n} \cdot \bar{q}} e^{i \Phi_0}e^{i \gamma \tau} \left(1 -e^{-i \gamma d}\right) \\
+&=\frac{-iA}{\gamma} e^{i \Omega \tau \bar{n} \cdot \bar{q}} e^{i \Phi_0}e^{i (-\Omega (1 + \bar{n}\cdot \bar{q}) ) \tau} \left(1 -e^{-i \gamma d}\right) \\
+&=\frac{-iA}{\gamma} e^{i \Phi_0}e^{-i \Omega \tau} \left(1 -e^{-i \gamma d}\right) \\
+&=\frac{-iA}{\gamma} e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i \gamma d}\right) 
 \label{eq:concise}
@@
 \subsection{Explicit expression and comparison with Melatos 22}
+We are now very close with making contact with the expression from Melatos 22. Let's be explicit and sub in our definitions for $A$ and $\gamma$ into \ref{eq:concise}
 
-We can expand the concise expression of Eq. \ref{eq:concise} as,
+\begin{align}
+p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} &=-i \frac{-\frac{i\Omega \omega}{2} H_{ij}q^i q^j }{-\Omega (1 + \bar{n}\cdot \bar{q}) } e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i (-\Omega (1 + \bar{n}\cdot \bar{q}) ) d}\right) \\
+&=\frac{\omega}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \\
+&=\frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right)
+\label{eq:final}
+\end{align}
+which is exactly the expression derived in Melatos 2022.
 
-\begin{equation} \label{eq:final}
-p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{H_{ij}q^i q^j  }{(1 + \hat{n} \cdot \hat{q})}e^{-i \Omega \tau(1+\bar{n} \cdot \bar{q}) + \Phi_0} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+\subsection{The $\Phi_0$ issue}
+There was previously some confusion over the presence of a single $\Phi_0$ parameter that was shared between multiple pulsars.
+Does this cause $\tau = 0$ to be a special time when all pulsars are in phase and then go on to evolve to be out of phase?
+This is just due to some sloppy transcription of equations in another document.
+Ignoring measurement noise, the measured photon frequency is related to the emitted photon frequency  as
+\begin{equation}
+	f_{\rm measured} = f_{\rm emitted} \left(1 - \frac{1}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \right)
 \end{equation}
-This expression is equivalent to Melatos 2022, PT13 where the dependence on time $t$ is made explicit. Using the Melatos notation where $h_{ij} = g_{ij} - \eta_{ij}$
-
-\begin{equation}
-	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{h_{ij} q^i q^j  }{(1 + \hat{n} \cdot \hat{q})} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
-\end{equation}
-
-\newpage
-\section{State-space structure} \label{sec3}
-\noindent The general structure of a Kalman state-space problem is
-\begin{equation}
-	\dot{x} = f(x) + w
-\end{equation}
-where $x$ are the state variables, $f()$ a non-linear function of the states, and $w$ a stochastic zero-mean process.
-The process noise matrix $Q = E(w w^T)$.
-The states are related to the measurement $z$ via a non-linear measurement function $h()$
-\begin{equation}
-	z = h(x) + v
-\end{equation}
-where $v$ is a stochastic zero-mean process and measurement noise matrix $R = E(v v^T)$. \newline
-
-\noindent This structure maps onto the preceding equations nicely. Our hidden state is just the intrinsic pulsar frequency, which evolves as,
-\begin{eqnarray}
-	\dot{f}_P = -\gamma f_P^n + \xi(t)
-\end{eqnarray}
-where $n$ is a braking index (typical values $\sim 1-5$, we take a canonical value $=3$) \footnote{See A.
-Vargas talk re anomalous braking indices.
-P.S.
-is there an arXiv preprint?} $\gamma$ a proportionality constant, and $\xi(t)$ represents a stochastic white noise process.
-Equation \ref{eq:final} can be used to relate the intrinsic pulsar frequency to that measured by an observer on Earth:
-
-\begin{eqnarray}
-	f_M = f_p(1 - X) + N_M
-\end{eqnarray}
-where $X$ is the RHS of Eq.
-\ref{eq:final} and $N_M$ a Gaussian measurement noise.
-An example of the evolution of the state and the measurement frequencies for our PTA pulsars can be seen in Fig.
-\ref{fig:mean and std of nets}
-
-\subsection{A note on magnitudes}
-
-It is worth pausing to consider some typical values of our state/measurement variables, as well as the respective noises.
-
-\begin{itemize}
-	\item $f_p$ - PTAs typically deal with millisecond pulsars. We can take a standard value of $f_P = 100$ Hz.
-	\item $\gamma, n$ - These two parameters describe the secular spin down.
-We can take $n=3$ as a standard value (spindown due to dipole emission).
-It follows that $\gamma$ is of the order $10^{-23}$ for typical PTA pulsars (e.g.
-PSR J$0023+0923$ has $f\sim 330$ Hz and $\dot{f} \sim 10^{-15} s^{-2}$)
-	\item $\sigma_p$ - this is the parameter which quantifies the strength of  $\xi(t)$ i.e.
-the pulsar red noise (sometimes known as timing noise).
-For MSPs the red noise is typically small - we can set it to be some fraction of $\dot{f}$ e.g.
-if 10$\%$ then $\sigma_p \sim 10^{-16}$.
-	\item $X$ - The magnitude of $X$ is set by the GW amplitude, which is expected to be $\lesssim 10^{-15}$ 
-	\item $\sigma_m$ - this is the parameter which quantifies the strength of the Gaussian measurement noise i.e.
-the radiometer white noise.
-Typical measurement uncertainty in the pulsar TOAs is $\sim 10 - 100$ ns for our best MSPs.
-If we take an observation cadence of $\sim 1$ month then we can relate the uncertainty in the TOA to the uncertainty in the frequency as $\sigma_m / f \sim 10 \text{ns} / 1 \text{ month}  \sim 4 \times 10^{-15}$.
-\footnote{L.Dunn gave some good advice on how this works, but I am still a bit unsure whether this quite makes sense?}
-\end{itemize}
-
-\begin{figure}
-	\centering
-	\begin{subfigure}[b]{0.475\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{images/fp1}
-		\caption{State frequency, single pulsar}%
-
-		\label{fig:mean and std of net14}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.475\textwidth}  
-		\centering 
-		\includegraphics[width=\textwidth]{images/fp_all}
-		\caption{State frequency, multiple pulsars}
-		\label{fig:mean and std of net24}
-	\end{subfigure}
-	\vskip\baselineskip
-	\begin{subfigure}[b]{0.475\textwidth}   
-		\centering 
-		\includegraphics[width=\textwidth]{images/fm1}
-		\caption{Measured frequency, single pulsar}  
-		\label{fig:mean and std of net34}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.475\textwidth}   
-		\centering 
-		\includegraphics[width=\textwidth]{images/fm_all}
-		\caption{Measured frequency, multiple pulsars}   
-		\label{fig:mean and std of net44}
-	\end{subfigure}
-	\caption{Evolution of the state (top row) and measurement (bottom row) frequencies for pulsars influenced by an arbitrary GW source.}
-	\label{fig:mean and std of nets}
-\end{figure}
-
-\newpage
-\section{Constructing a PTA}\label{sec:PTA}
-
-In order to proceed and explore how well this state-space formulation works, we will need to specify a selection of pulsars to make up our PTA.
-We will take the 47 pulsars that make up the NANOGrav PTA \footnote{Data obtained from the ANTF catalogue at \url{https://www.atnf.csiro.au/research/pulsar/psrcat/}} \newline
-
-\noindent ANTF provides values for both $f$ and $\dot{f}$.
-For every pulsar we will assume that $n=3$ and calculate $\gamma$ accordingly.
-For example, PSR J1713+0747 has a spin frequency of 218 Hz and a derivative $-4 \times 10^{-16}$ s $^{-2}$.
-Taking $n=3$ gives $\gamma \sim 3.9 \times 10^{-23}$.
-\newline
-
-\noindent Approximate pulsar distances are also provided.
-For now we will take these as known rather than additional parameters to be inferred.
-The median pulsar distance for NANOGrav pulsars is $\sim 1.3$ kpc.
-The distribution of NANOGrav pulsars on the sky is shown in Fig \ref{fig:pulsar_distrib}
-
-\begin{figure}
-\includegraphics[width=0.8\textwidth]{images/pulsars}
-\caption{Spatial distribution and distances of NANOGrav pulsars}
-\label{fig:pulsar_distrib}
-\end{figure}
-
-\newpage 
-\section{Model selection using a UKF}
-\noindent We can construct some noisy synthetic data using the equations and values of Section \ref{sec3} and then apply an unscented Kalman filter \footnote{See \href{https://groups.seas.harvard.edu/courses/cs281/papers/unscented.pdf}{Wan \& van der Merwe} for a very readable description of how the UKF works } to the measured frequency to try to recover the intrinsic state frequency.
-An example of this is shown below in Fig \ref{fig:model example} below for some large GW strain.
-We can see that for this example the filter can recover the state evolution well.
-
-\begin{figure}
-	\includegraphics[width=0.8\textwidth]{images/model}
-	\caption{Recovery of the pulsar state frequency from the measurement frequency by a UKF.}
-	\label{fig:model example}
-\end{figure}
-\noindent We can use this UKF to try to solve the problem of GW detection i.e.
-\textit{"Is there evidence of a GW in my data?}.
-We can frame this as a model selection procedure where we have two models/hypotheses:
-
-\begin{itemize}
-	\item \textbf{Null Model} $M_0$. There is no GW in the data. In this case the measurement model of the UKF simply returns the frequency states (i.e. $X= 0$)
-	\item \textbf{Alternative model} $M_1$. There is a GW in the data. The measurement model uses the full expression for $X$
-\end{itemize}
-\begin{figure}
-	\includegraphics[width=0.8\textwidth]{images/null}
-	\caption{As Fig 3, but now for a the null model. The state is not well recovered.}
-	\label{fig:null example}
-\end{figure}
-\noindent An example applying the null model to the same data as in Fig \ref{fig:model example} is shown in Fig \ref{fig:null example}.
-Evidently the null model does not capture the true underlying state well.
-\newline
-
-\noindent In order to accept the alternative hypothesis $M_1$ over $M_0$ there are two approaches we could take:
-\begin{itemize}
-	\item The first is a fully Bayesian search over all the parameters for each model, calculating the evidence for each model and then determining the Bayes ratio.
-This is perhaps the most consistent way, but it is obviously expensive and at this stage we are keen to explore how detectability varies with e.g.
-GW strain.
-	\item The second method is to recognize that $M_0$ and $M_1$ are hierarchically nested models and we can perform a likelihood ratio test.
-That is, given the maximum likelihood estimators $\hat{\theta}$ of the true parameters $\theta$, the likelihood of each model can be calculated and compared.
-These likelihoods are just  point estimates of the Bayes factor numerator/denominators.
-\end{itemize}
-Given the cheap cost we proceed with the second method.
-
-\subsection{Likelihood ratio test}
-For the likelihood ratio test we do not perform any kind of maximum likelihood search over the parameters for each of the models.
-Instead we just artificially set the maximum likelihood estimators to be equal to the true parameters of the system \footnote{i.e.
-$\hat{\theta} = \theta$}.
-We assume that any maximum likelihood algorithm would converge to these parameters \footnote{This is obviously an oversimplification but will serve our purposes for now}.
-\newline
-
-\noindent Interpreting the likelihood ratio $\Lambda$ also needs some consideration, since we have to account for the increased model complexity of $M_1$ \footnote{Bayes factors penalise complexity by construction since one must integrate over a larger parameter space}.
-This can be accomplished via Wilks' Theorem which states that for a large nuber of samples \footnote{What counts as large?
-See \url{https://www.osti.gov/servlets/purl/1529145}} the distribution of the test statistic approaches the chi-squared distribution under the null hypothesis i.e.
-\begin{equation}
-2 \log \Lambda \rightarrow \chi^2
-\end{equation}
-One can then compute $p$-values where the number of degrees of freedom is equal to the difference in the number of parameters of the two models; $M_1$ has 7 extra parameters over $M_0$
-
-\begin{itemize}
-	\item Parameters $M_0$: $\gamma$, $n$
-	\item Parameters $M_1$: $\gamma$, $n$, $\omega$, $\theta$, $\phi$, $\psi$, $\iota$, $\Phi_0$,$h$
-\end{itemize}
-With 7 degrees of freedom and a target tolerance of 5(1) \% the test statistic is $\sim 14 (18.5)$.
-For the toy system exhibited in Figs.
-\ref{fig:model  example}, \ref{fig:null example}, the test statistic is $\gg 20$ - this makes intuitive sense since it is immediately obvious to the eye that the data is sinusoidal.
-But what about other systems?
-
-\subsection{Detectability vs strain}
-
-Lets take a system and explore how the detectability, as quantified by the test statistic varies with GW strain.
-For our PTA we use the NANOGrav pulsars described in Section \ref{sec:PTA}, with an observation length of 10 years and a weekly cadence.
-The GW source parameters are listed in Table \ref{tab:toy_example_parameters}.
-The results for this system are presented in Figure \ref{fig:SNR}, for a single noise realisation.
-We consider both the test statistic, $2 \log \Lambda$, and another parameter $\delta$ which is the median absolute error in the state estimates \footnote{In real-life we of course cannot compute predicted state - actual state, but it is useful here for exploring the performance of the UKF.} From the Figure we can see that the signal is a smooth function of the strain and is detectable up to $h \sim 10^{-13}$.
-Below this limit the two models cease to be hierarchical and the error in the state estimates is highly comparable.
-This is the regime where the noise is large with respect to the strain and depending on the specific realisation of the noise either the null or the alternative models can be preferred i.e.
-we are just measuring noise.
-A brief exploration of other noise realisations and different number of pulsars is presented in Fig.
-\ref{SNR2}.
-
- \begin{table}
-	\centering
-	\begin{tabular}{lc}
-		Parameter & Value  \\
-		\hline
-		$\omega$       & $10^{-7}$  \\
-		$\theta$          & $1.0$  \\
-		$\phi$              & $0.0$   \\
-		$\psi$              & $2.5$  \\
-		$\iota$             & $0.0$  \\
-		$\Phi_0$          & $0.2$  \\
-		$\sigma_{\rm p}$ & $10^{-13}$ \\
-		$\sigma_{\rm m}$ & $10^{-13}$ \\
-		\hline
-	\end{tabular}
-	\caption{GW parameters used for generating synthetic data. There is nothing special about these parameters - they are just chosen arbitrarily.}
-	\label{tab:toy_example_parameters}
-\end{table}
-
-\begin{figure}
-	\includegraphics[width=0.8\textwidth]{images/SNR2}
-	\caption{Variation of the test statistic (\textit{Top panel}) and the median state error (\textit{Bottom panel}) w.r.t GW strain.
-The grey horizontal line labels a test statistic of 14.
-Below $h \sim 10^{-12}$ the models cease to be consistently hierarchical - values with negative test statistics are set automatically $=1$ and marked with a red dot.}
-	\label{fig:SNR}
-\end{figure}
-
-\begin{figure}
-	\centering
-
-	\begin{subfigure}[b]{0.475\textwidth}  
-		\centering 
-		\includegraphics[width=\textwidth]{images/SNR2_differentnoise}
-		\caption{Full PTA, different noise realisation}
-		\label{fig:SNR2a}
-	\end{subfigure}
-	\vskip\baselineskip
-	\begin{subfigure}[b]{0.475\textwidth}   
-		\centering 
-		\includegraphics[width=\textwidth]{images/SNR2_frac02}
-		\caption{Only using 20 \% of pulsars}  
-		\label{fig:SNR2b}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.475\textwidth}   
-		\centering 
-		\includegraphics[width=\textwidth]{images/SNR2_n2}
-		\caption{Only using 2 pulsars.}   
-		\label{fig:SNR2c}
-	\end{subfigure}
-	\caption{As Figure 5, but for different noises and numbers of pulsars}
-	\label{SNR2}
-\end{figure}
-
- \newpage 
- \section{Caveats and questions}
-
- \begin{enumerate}
-
- 	\item We should really do this using a ``proper" language and enforce consistent number formats.
-We are using the Python long double, but there is some uncertainty around how e.g.
-linear algebra operations in Python handle these formats - do they just get rounded to doubles?
-
- 	\item We have assumed that all pulsars have $n=3$. What about some distribution?
-
- 	\item We have only considered a single GW source at a particular location, polarisation, etc. What about others?
-
- 	\item We have focused only on a monochromatic GW source. This is probably not a bad approximation, but maybe we can include some process noise on the phase evolution too?
-
- 	 \item We assume all pulsars have observations taken at the same time
-
- 	\item How reasonable are our noises? For instance, what is a reasonable measurement noise on the frequency evolution of a MSP?
-
- 	 \item What about doing a full Bayesian search using e.g. Bilby to get a Bayes ratio?
-
- 	 \item How good/bad an assumption is it that $\hat{\theta} \sim \theta$ for our likelihood ratio test?
- \end{enumerate}
+Now the pulsar frequencies remain in phase and $\tau=0$ is no longer a special time.
+\footnote{This expression is slightly different from the (erroneous) one typed up in the manuscript!
+We have dropped a $1+ \bar{n} \cdot \bar{q}$ factor that was wrongly typed up in the other document}
 

### 2023-04-22 docs/measurement_eqn_deerivation/LectureNotes_Math571.tex
typos
misc updates

@@
 	\normalfont\normalsize 
-	{Melbourne University - Summer 2023} \\ [0pt] % Your university, school and/or department name(s)
+	{Melbourne University} \\ [0pt] % Your university, school and/or department name(s)
 	\huge Notes on state-space formulation of PTA problem% The assignment title
@@
 	f_{\rm measured} = f_{\rm emitted} \left(1 - \frac{1}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \right)
+	\label{eq:measureent}
 \end{equation}
@@
 
+ \subsection{Trigonometric form}
+
+ Lets express Eq. \ref{eq:measureent} as
+
+ \begin{equation}
+ 	f_{\rm measured} = f_{\rm emitted} \, g(\tau; \bar{\theta})
+ 	\label{eq:measureent}
+ \end{equation}
+where $\bar{\theta}$ is a set of parameters of the system.
+Lets write this in terms of the real part of the trig expansion.
+Let $a = -\Omega \tau +\Phi_0$ and $b =\Omega (1 + \bar{n}\cdot \bar{q})  d $
+ \begin{align}
+	g(\tau; \bar{\theta}) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } e^{ia} \left(1 -e^{i b}\right) \\
+	&= 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } [\cos(a) + i \sin(a)] \left(1 -[\cos(b) + i \sin(b)]\right) \\ 
+	&= 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left( \cos(a) + i \sin(a) - \cos(a) \cos(b) -i \cos(a) \sin(b) -i \sin(a) \cos(b) + \sin(a) \sin(b)\right) \\
+	& \stackrel{\rm Re}{=} 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left( \cos(a) - \cos(a) \cos(b) + \sin(a) \sin(b)\right) \\
+	& = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left( \cos(a) - \cos(a + b) \right) \\
+	& = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
+\end{align}
+
 \end{document}

### 2023-04-22 py_src/gravitational_waves.py

@@ -7,8 +7,16 @@ Given the GW parameters, the pulsar parameters and the time,
 compute the frequency correction factor.
 Returns an object of shape (n times, n pulsars)
 """
+
+
+
+
+
+
+
+#### EXP FORM
 @jit(nopython=True)
-def gw_prefactor_optimised(delta,alpha,psi,q,q_products,h,iota,omega,d,t,phi0):
+def gw_prefactor_optimised_exp(delta,alpha,psi,q,q_products,h,iota,omega,d,t,phi0):
         
 
         #GW direction axes
@@ -40,6 +48,98 @@ def gw_prefactor_optimised(delta,alpha,psi,q,q_products,h,iota,omega,d,t,phi0):
         GW_factor = 1.0 - 0.50 * (hij/dot_product) * (1.0 - np.exp(1j*omega*dot_product*d))
         return np.real(GW_factor) #This has shape(n times, n pulsars)
 
+
+#### OLD FORM
+@jit(nopython=True)
+def gw_prefactor_optimised_old(delta,alpha,psi,q,q_products,h,iota,omega,d,t,phi0):
+
+        print("alpha = ", alpha)
+
+        m,n                 = principal_axes(np.pi/2.0 - delta,alpha,psi)    
+        gw_direction        = np.cross(m,n)
+      
+        dot_product         = 1.0 + np.dot(q,gw_direction) #matmul might be a bit faster, but np.dot has JIT support
+
+
+
+
+        e_plus              = np.array([[m[i]*m[j]-n[i]*n[j] for i in range(3)] for j in range(3)]) #tensordot might be a bit faster, but list comprehension has JIT support
+        e_cross             = np.array([[m[i]*n[j]-n[i]*m[j] for i in range(3)] for j in range(3)])
+
+        hp,hx               = h_amplitudes(h,iota) 
+        Hij                 = hp * e_plus + hx * e_cross
+        Hij_flat            = Hij.flatten()
+
+        hbar                = np.dot(Hij_flat,q_products)
+
+
+
+
+
+        prefactor    = 0.5*(hbar / dot_product)*(1.0 - cos(omega*d*dot_product))
+
+
+        tensor = np.outer(t,dot_product) #This has shape(n times, n pulsars)
+        time_variation = cos(-omega*tensor + phi0)
+
+        GW_factor = 1.0 - prefactor * time_variation
+        return GW_factor #This has shape(n times, n pulsars)
+
+
+
+#### TRIG FORM
[…]

### 2023-04-22 py_src/plotting.py

@@ -151,8 +151,9 @@ def likelihoods_over_priors(parameters,priors,PTA,P,KF,sigma_p):
     plt.style.use('science')
     true_parameters = priors_dict(PTA,P)
     true_parameters["sigma_p"] = sigma_p
-    #true_parameters["alpha_gw"] = 2.0
+    #
     
+    print("likelihoods_over_priors")
 
     h,w = 20,12
     rows = 6
@@ -182,7 +183,8 @@ def likelihoods_over_priors(parameters,priors,PTA,P,KF,sigma_p):
         if key in logvalues:
            
             ax.set_xscale('log')
-            #ax.set_yscale('log')
+        
+        ax.set_yscale('log')
 
         #ax.set_yscale('log')
         i+=1

### 2023-05-03 docs/measurement_eqn_derivation/LectureNotes_Math571.tex
some updates to minutes

--- 2023-04-22 docs/measurement_eqn_deerivation/LectureNotes_Math571.tex
@@
 
+\section{Discretization of state equations}
+We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of Vargas \& Melatos.
+Within this model, $f$ evolves according to a combination of both deterministic torques (i.e.
+electromagnetic spin-down) and stochastic torques (i.e.
+`spin wandering', achromatic variations in the pulse TOA intrinsic to the star).
+The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process.
+Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
+\begin{equation}
+	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
+	\label{eq:frequency_evolution}
+\end{equation}
+where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation,$\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
+\begin{equation}
+	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
+\end{equation}
+for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
+\begin{equation}
+	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
+\end{equation}  
+Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
+\begin{equation}
+	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
+	\label{eq:frequency_evolution_expanded}
+\end{equation}
+We can express the intrinsic frequency evolution, Eq. \ref{eq:frequency_evolution_expanded}, in a alternative form as,
+\begin{equation}
+	df = \mathcal{A} f dt + N(t) dt + \sigma dB(t)
+	\label{eq:state1}
+\end{equation}
+where $\mathcal{A} = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process).
+This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by,
+\begin{equation}
+	f(t) = e^{\mathcal{A}t}f(0) + \int_0^t e^{\mathcal{A}(t-t')} N(t') dt' + \int_0^t e^{\mathcal{A}(t-t')} \sigma dB(t')
+\end{equation} 
+If we move from a solution in continuous time to discrete time then
+\begin{equation}
+	f(t_{i+1}) = F_i f(t_i) + T_i + \eta_i
+\end{equation}
+where
+\begin{equation}
+	F_i = e^{A (t_{i+1} - t_i)}
+\end{equation}
+\begin{equation}
+	T_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt'
+\end{equation}
+\begin{equation}
+	\eta_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} \sigma dt'
+\end{equation}
+\noindent  If we specialize to the case of constant time sampling between our observations, $\Delta t$, then for our formulation the components that make up the Kalman filter are as follows: 
+\begin{equation}
+	F_i = F_{i+1} = e^{-\gamma \Delta t}
+	\label{eq:fmatrix}
+\end{equation}
+\begin{align}
+	T_i &= \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt' \\
+	& = f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)  (\Delta t + t_i) - e^{-\gamma \Delta t} (f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)t_i)
+\end{align}
+whilst the discrete process noise matrix is
+\begin{equation}
+	Q_i  \delta_{ij}= \langle \eta_i \eta_j^T \rangle = \frac{- \sigma^2}{2 \gamma} \left( e^{-2 \gamma \Delta t} -1\right)
+\end{equation}
+At this point we can also specify our measurement matrix from Eq \ref{eq:measureent}
+\begin{equation}
+	H_i = g(t_i; \bar{\theta})
+\end{equation}
+and the measurement noise matrix 
+\begin{equation}
+	R_i = R_{i+1} = \Sigma^2
+	\label{eq:rmatix}
+\end{equation}
+
 \end{document}

### 2023-05-03 py_src/kalman_filter.py

@@ -17,10 +17,11 @@ The log likelihood, designed for diagonal matrices where S is considered as a ve
 def log_likelihood(S,innovation):
     x = innovation / S 
     N = len(x)
+    
     slogdet = np.sum(np.log(S)) # Uses log rules and diagonality of covariance "matrix"
     value = -0.5*(slogdet+innovation @ x + N*np.log(2*np.pi))
-    #return value
-    return -np.log(np.abs(value))
+    return value
+    #return -np.log(np.abs(value))
     #return -np.log(np.abs(innovation @ innovation))
     #return -np.sum(innovation**2)
 

### 2023-05-08 docs/minutes/slides/document.tex
canonical_psr_example_4

% !TeX TS-program = xelatex

\documentclass[10pt]{beamer}

\usetheme[progressbar=frametitle]{metropolis}
\usepackage{appendixnumberbeamer}
\usepackage{pgfplots}
\usepackage{xspace}

\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
%\setsansfont[BoldFont={Fira Sans}]{Fira Sans Light}
%\setmonofont{Fira Mono}
%\usepackage[sfdefault]{Fira Sans}

\title{}
\subtitle{}

\author{}

\date{EE/GW meeting, November 11, 2023}

\begin{document}

	\maketitle

	\begin{frame}{}

	\begin{itemize}
		\item Problem = overly narrow,biased posteriors 
	\end{itemize}

\end{frame}

\begin{frame}{}
	A match made in heaven: reduced precision + stochastic rounding \newline

	   \centering
		\begin{tikzpicture}
    \draw [-,>=stealth, color = black, thick] (-5, 0.0) -- (+5, 0.0); 
     \draw [densely dashed,>=stealth, color = black, thick] (0.0, -0.50) -- (0.0, +0.5); 
      \draw [-,>=stealth, color = black, thick] (-3, -0.50) -- (-3, +0.5); 
	\draw [-,>=stealth, color = black, thick] (+3, -0.50) -- (+3, +0.5); 
	\node[text=blue] at (3.2, -0.5) {$x_2$};
	\node[text=blue] at (-2.8, -0.5) {$x_1$};
	\node[text=blue] at (1.5, 1.2) {$x$};
	\draw [-stealth,color = blue, thick] (1.5, 1.01) -- (1.5, 0.0);

	\draw [-,>=stealth, color = red, thick] (-5, 1.0) -- (+5, 1.0);

	\end{tikzpicture}

\raggedright

\begin{itemize}
	\item Usual way: Round-to-nearest
	\item Alternatively: Round \alert{stochastically} 
\end{itemize}

Why bother? $\rightarrow \mathbb{E} \left(\mathcal{R}(x)\right) = x$

\end{frame}

\begin{frame}{}
\centering
\vspace{5mm}
%\includegraphics[width=0.75\textwidth,height=0.75\textwidth]{images/precip_ensemble_timeseries_decadal}

\raggedleft
\vspace{-1mm}
\tiny \url{arxiv.org/abs/2207.14598} \normalsize

\end{frame}

%	\begin{frame}{}
	%
	%Numerical weather and climate modelling 
	%
	%        \begin{itemize}
		%	        \item Reduced precision floats + stochastic rounding
		%	        \item Automatic Differentiation
		%        \end{itemize}
	%    
	%    \alert{Open call for collaboration!}
	%    
	%    
	%	\end{frame}

\end{document}

### 2023-05-08 py_src/plotting.py

@@ -128,13 +128,15 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
     selected_variables = []
     selected_injections = []
     selected_labels = []
+    selected_ranges = []
     for i in range(len(medians)):
         print(variables_to_plot[i], medians[i], variances[i])
 
-        if variances[i] > 1e-40:
+        if variances[i] > 1e-30:
             selected_variables.extend([variables_to_plot[i]])
             selected_injections.extend([injection_parameters[i]])
             selected_labels.extend([labels[i]])
+            #selected_ranges.extend([ranges[i]])
         else:
             print("Note! ", variables_to_plot[i], " has zero variance and will not be plotted")
    
@@ -147,6 +149,9 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
     plt.style.use('science')
 
     try: 
+
+        print("selected variabels are")
+        print (selected_variables)
         fig = corner.corner(y_post, 
                             color='C0',
                             show_titles=True,

### 2023-05-08 src/GW.jl

@@ -30,26 +30,30 @@ function gw_frequency_modulation_factor(δ::NF,α::NF,ψ::NF,h::NF,cos_ι::NF,ω
 
 
    
-
-
-    #
     m,n                  = principal_axes(NF(π/2.0) - δ,α,ψ)  
     gw_direction         = cross(m,n)
-    dot_product = NF(1.0) .+ q * gw_direction
+    dot_product          = NF(1.0) .+ q * gw_direction
 
     e_plus              = [m[i]*m[j]-n[i]*n[j] for i=1:3,j=1:3]
     e_cross             = [m[i]*n[j]-n[i]*m[j] for i=1:3,j=1:3]
     hp,hx               = h_amplitudes(h,cos_ι)   
     Hij                 = hp * e_plus + hx * e_cross
+
+
+    #Ampltiude terms 
     hbar         = [sum([Hij[i,j]*q[k,i]*q[k,j] for i=1:3,j=1:3]) for k=1:size(q)[1]] # Size Npulsars. Is there a vectorised way to do this?
-    prefactor    = NF(0.5).*(hbar ./ dot_product).*(NF(1.0) .- cos.(ω.*d.*dot_product))
-    tensor_product = t * dot_product' #this has shape(n times, n pulsars)
-    time_variation = cos.(-ω*tensor_product .+ Φ0)
-    GW_factor = NF(1.0) .- prefactor' .* time_variation
+    
+    
+    little_a = -ω*t .+ Φ0
+    little_b = ω.*dot_product.*d
+    
 
     
-    return GW_factor'
-    #@debug @assert size(dot_product)==length(q) # assert is not called unless 
+    blob = little_a .+ little_b'
+    trig_block = cos.(little_a) .- cos.(blob)
+    GW_factor = NF(1.0) .- NF(0.50).*(hbar ./ dot_product) .* trig_block' 
+
+    return GW_factor
 
     
 end 

### 2023-05-08 src/StateSpacePTA.jl

@@ -43,7 +43,6 @@ include("priors.jl")
 include("linear_kalman_filter.jl")
 include("particle_swarm.jl")
 include("particle_swarm_v2.jl")
-include("blackbox.jl")
 include("plotting.jl")
 include("run.jl")
 

### 2023-05-08 docs/minutes/slides/document.tex
try ultranest sampler

@@
 	\begin{itemize}
-		\item Problem = overly narrow,biased posteriors 
+		\item \alert{Problem = overly narrow,biased posteriors}. Even for ``easy" systems!
+	\end{itemize}
+
+\includegraphics[width=0.5\textwidth,height=0.5\textwidth]{images/example_system}
+
+$h = 10^{-2}, \sigma_m = 10^{-13}, \sigma_p = 0.0$
+
+\end{frame}
+
+	\begin{frame}{}
+
+	\begin{itemize}
+		\item Why? Very nasty likelihood curves
+	\end{itemize}
+
+	$$  p(\theta | d) \propto p(d | \theta) \pi({\theta})$$
+	$$ p(\theta | d) \propto \mathcal{L}(\theta) \pi({\theta})$$
+
+	For uniform prior, posterior shape $\sim$ likelihood shape.
+
+\end{frame}
+
+\begin{frame}{}
+
+	Example: $\mathcal{L}(\delta)$
+
+\includegraphics[width=0.7\textwidth,height=0.6\textwidth]{images/likelihood_delta_1}
+
+\end{frame}
+
+\begin{frame}{}
+
+	Example: $\mathcal{L}(\delta)$, zoomed 1
+
+	\includegraphics[width=0.7\textwidth,height=0.6\textwidth]{images/likelihood_delta_2}
+
+\end{frame}
+
+\begin{frame}{}
+
+	Example: $\mathcal{L}(\delta)$, zoomed 2
+
+	\includegraphics[width=0.7\textwidth,height=0.6\textwidth]{images/likelihood_delta_3}
+
+\end{frame}
+
+\begin{frame}{}
+
+	Example: $\mathcal{L}(\omega)$
+
+	\includegraphics[width=0.7\textwidth,height=0.6\textwidth]{images/likelihood_omega_1}
+
+\end{frame}
+
+\begin{frame}{}
+
+	Example: $\mathcal{L}(\omega)$, zoomed 1
+
+	\includegraphics[width=0.7\textwidth,height=0.6\textwidth]{images/likelihood_omega_2}
+
+\end{frame}
+
+\begin{frame}{}
+
+	Example: $\mathcal{L}(\omega)$, zoomed 2
+
+	\includegraphics[width=0.7\textwidth,height=0.6\textwidth]{images/likelihood_omega_3}
+
+\end{frame}
+
+\begin{frame}{}
+
+	All likelihood curves for $\bar{\theta}_{\rm GW}$:
+
+	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves}
+
+\end{frame}
+
+\begin{frame}{}
+
+		\begin{itemize}
+		\item \alert{Problem} = overly narrow,biased posteriors
+		\item Cause: noisy, likelihoods which are "locally Gaussian". Sampling gets stuck in a local optima and infers posteriors widths based on that optima. 
 	\end{itemize}
@@
 \begin{frame}{}
-	A match made in heaven: reduced precision + stochastic rounding \newline
+	Potential solutions:
 
-	   \centering
-		\begin{tikzpicture}
-    \draw [-,>=stealth, color = black, thick] (-5, 0.0) -- (+5, 0.0); 
-     \draw [densely dashed,>=stealth, color = black, thick] (0.0, -0.50) -- (0.0, +0.5); 
-      \draw [-,>=stealth, color = black, thick] (-3, -0.50) -- (-3, +0.5); 
-	\draw [-,>=stealth, color = black, thick] (+3, -0.50) -- (+3, +0.5); 
-	\node[text=blue] at (3.2, -0.5) {$x_2$};
-	\node[text=blue] at (-2.8, -0.5) {$x_1$};
-	\node[text=blue] at (1.5, 1.2) {$x$};
-	\draw [-stealth,color = blue, thick] (1.5, 1.01) -- (1.5, 0.0);
-
-	\draw [-,>=stealth, color = red, thick] (-5, 1.0) -- (+5, 1.0);
-
-	\end{tikzpicture}
-
-\raggedright
-
-\begin{itemize}
-	\item Usual way: Round-to-nearest
-	\item Alternatively: Round \alert{stochastically} 
-\end{itemize}
-
-Why bother? $\rightarrow \mathbb{E} \left(\mathcal{R}(x)\right) = x$
+	\begin{enumerate}
+		\item Don't be a Bayesian: Maximum likelihood / loss optimisation? 
+		\item Be a Bayesian: Drop PSR terms from model?
+		\item other...?
+	\end{enumerate}
 
@@
 
-\begin{frame}{}
-\centering
-\vspace{5mm}
-%\includegraphics[width=0.75\textwidth,height=0.75\textwidth]{images/precip_ensemble_timeseries_decadal}
+\begin{frame}{Option 2: Drop PSR terms from model?}
+	 \begin{equation}
+		f_{\rm measured} = f_{\rm emitted} \, g(\tau; \bar{\theta})
+		\label{eq:measureent}
+	\end{equation}
+	with
+	\begin{align}
+		g(\tau; \bar{\theta}) 
+		& = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
+	\end{align}
 
-\raggedleft
-\vspace{-1mm}
-\tiny \url{arxiv.org/abs/2207.14598} \normalsize
+Generate the fake data using the full expression for $g(\tau; \bar{\theta})$, but use $g'(\tau; \bar{\theta})$ in the Kalman measurement model
 
+		\begin{align}
+		g'(\tau; \bar{\theta}) 
+		& = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) \right]
+	\end{align}
+\end{frame}
+
+\begin{frame}{Option 2: Drop PSR terms from model?}
+
+		All likelihood curves for $\bar{\theta}_{\rm GW}$:
+		\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves_psr}
+
+\end{frame}
+
+\begin{frame}{Option 2: Drop PSR terms from model?}
+
+	All likelihood curves for $\bar{\theta}_{\rm GW}$:
+	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves_psr}
+
+\end{frame}
+
+\begin{frame}{Option 2: Drop PSR terms from model?}
+
+	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/canonical_mixed_result}
+
+	OK, but some biases...
 \end{frame}

### 2023-05-08 py_src/kalman_filter.py

@@ -21,7 +21,7 @@ def log_likelihood(S,innovation):
     slogdet = np.sum(np.log(S)) # Uses log rules and diagonality of covariance "matrix"
     value = -0.5*(slogdet+innovation @ x + N*np.log(2*np.pi))
     return value
-    #return np.sum(np.abs(innovation))
+    #return -np.log(np.sum(np.abs(innovation)))
     #return -np.log(innovation @ innovation)
 
 

### 2023-05-08 py_src/plotting.py

@@ -76,7 +76,7 @@ def plot_all(t,states,measurements,predictions_x,predictions_y,psr_index,savefig
     ax2.set_xlabel('t [years]', fontsize=fs)
     ax1.set_ylabel(r'$f_p$ [Hz]', fontsize=fs)
     ax2.set_ylabel(r'$f_M$ [Hz]', fontsize=fs)
-    ax2.set_ylabel(r'Residual [Hz]', fontsize=fs)
+    ax3.set_ylabel(r'Residual [Hz]', fontsize=fs)
     ax2.xaxis.set_tick_params(labelsize=fs-4)
     ax2.yaxis.set_tick_params(labelsize=fs-4)
     ax1.yaxis.set_tick_params(labelsize=fs-4)
@@ -194,7 +194,7 @@ def iterate_over_priors(variable, variable_range,true_parameters,KF):
 
 
 
-def likelihoods_over_priors(parameters,priors,PTA,P,KF):
+def likelihoods_over_priors(parameters,priors,PTA,P,KF,title,savefig):
 
 
 
@@ -225,7 +225,12 @@ def likelihoods_over_priors(parameters,priors,PTA,P,KF):
         ax = axes[i]
         ax.plot(prior,likelihood)
         ax.set_xlabel(key, fontsize = 16)
-        ax.axvline(value,linestyle='--', c='C2')
+        ax.axvline(value,linestyle='--', c='C2',label="truth")
+
+        idx = np.argmax(likelihood)
+
+
+        ax.axvline(prior[idx],linestyle='--', c='C3',label="maxima")
 
         if key in log_x_values:
             ax.set_xscale('log')
@@ -236,8 +241,14 @@ def likelihoods_over_priors(parameters,priors,PTA,P,KF):
     
 
     plt.subplots_adjust(hspace=0.5)
-    title = r"Likelihood Identifiability"
+   
     fig.suptitle(title, fontsize=20)
+    plt.legend()
+
+
+
+    if savefig != None:
+        plt.savefig(f"../data/images/{savefig}.png", bbox_inches="tight",dpi=300)
 
     plt.show()
 

### 2023-05-08 py_src/system_parameters.py

@@ -8,7 +8,7 @@ Function that returns a dict of parameters which define the system
 def SystemParameters(NF=np.float64,    # the number format of the arguments
                      T = 10,           # how long to integrate for in years
                      cadence=7,        # the interval between observations
-                     Ω=   5e-7,        # GW angular frequency
+                     Ω=   5.0e-7,        # GW angular frequency
                      Φ0 = 0.20,        # GW phase offset at t=0
                      ψ =  2.50,        # GW polarisation angle
                      ι = 1.0,          # GW source inclination

### 2023-05-08 src/linear_kalman_filter.jl

@@ -69,7 +69,8 @@ function update(x::Vector{NF},P::Vector{NF},observation::Vector{NF},R::NF,H::Vec
     Pnew = I_KH .* P .* I_KH .+ K .* R .* K #updated covariance
 
     #...and get the likelihood
-    l = log_likelihood(S,y)
+    #l = log_likelihood(S,y)
+    l = residuals(y)
     
     return xnew,Pnew,l
 
@@ -97,4 +98,10 @@ end
 
 
 
+function residuals(innovation::Vector{NF}) where {NF<:AbstractFloat}
+
+
+return -log(abs.(innovation⋅innovation))
+
+end 
 

### 2023-05-10 docs/minutes/slides/document.tex
fprior_earth_corrected prior
he-8 for earth model

@@
 
-\date{EE/GW meeting, November 11, 2023}
+\date{EE/GW meeting, May 11, 2023}
 
@@
 	\begin{itemize}
-		\item \alert{Problem = overly narrow,biased posteriors}. Even for ``easy" systems!
+		\item \alert{Problem = overly narrow,biased posteriors from nested sampling}. Even for ``easy" systems!
 	\end{itemize}
 
+Here is an easy system: $h = 10^{-2}, \sigma_m = 10^{-13}, \sigma_p = 0.0$ \newline 
+\centering
 \includegraphics[width=0.5\textwidth,height=0.5\textwidth]{images/example_system}
 
-$h = 10^{-2}, \sigma_m = 10^{-13}, \sigma_p = 0.0$
+\raggedright
+If the true declination $\delta = 1.0$, NS might give us a posterior of $\sim 0.95$ but variance v.v. small and posterior does not capture true value
 
@@
 		\item Cause: noisy, likelihoods which are "locally Gaussian". Sampling gets stuck in a local optima and infers posteriors widths based on that optima. 
+%		\item Require $n_{live} > \frac{\rm prior \, volume}{\rm mode \,  footprint} = \frac{X}{\sigma^2}$
+%		\begin{itemize}
+%			\item For $\delta$ require $n_{live} > \frac{2 \pi}{10^{-5}} \sim 10^5 $
+%			\item For $\omega$ require $n_{live} > \frac{10^4}{10^{-11}} \sim 10^7 $ 
+%			\item Currently $n_{live} = 500$ parallelized on 32 cores $\sim 700$ minutes
+%			\item Runtime scales $\mathcal{O}(n_{live}) \rightarrow 8000 $ hours for required $\delta$ sampling (just under 1 year!)
+%
+%		\end{itemize}
 	\end{itemize}
@@
 	\begin{enumerate}
-		\item Don't be a Bayesian: Maximum likelihood / loss optimisation? 
-		\item Be a Bayesian: Drop PSR terms from model?
+		\item Better settings for sampler? E.g. more live points (how many?), massively parallel, optimized likelihood evaluations,...
+		\item Maximum likelihood / loss optimisation? 
+		\item Drop PSR terms from model?
+		\item MCMC at higher temperatures (trades off evidence..., less efficient	)
 		\item other...?
@@
 
-\begin{frame}{Option 2: Drop PSR terms from model?}
+\begin{frame}{Drop PSR terms from model?}
 	 \begin{equation}
@@
 
-\begin{frame}{Option 2: Drop PSR terms from model?}
-
-		All likelihood curves for $\bar{\theta}_{\rm GW}$:
-		\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves_psr}
-
-\end{frame}
-
-\begin{frame}{Option 2: Drop PSR terms from model?}
+\begin{frame}{Drop PSR terms from model?}
 
@@
 
-	OK, but some biases...
+	OK, but some biases. c.f. location of the maxima on the previous slide
 \end{frame}

### 2023-05-10 py_src/main.py

@@ -29,7 +29,7 @@ if __name__=="__main__":
     multiprocessing.set_start_method("fork")
 
 
-    P   = SystemParameters(h=1e-2,σp=0.0,σm=1e-13,use_psr_terms_in_data=True,use_psr_terms_in_model=False,noise_model=False)       # define the system parameters as a class
+    P   = SystemParameters(h=1e-8,σp=0.0,σm=1e-13,use_psr_terms_in_data=True,use_psr_terms_in_model=False,noise_model=False)       # define the system parameters as a class
     PTA = Pulsars(P)                                       # setup the PTA
     data = SyntheticData(PTA,P)                            # generate some synthetic data
 

### 2023-05-15 docs/minutes/slides_11_05/document.tex
restart canoncial 12 and noise

--- 2023-05-10 docs/minutes/slides/document.tex
@@
 
+\begin{frame}{}
+
+	Lets proceed using this `Earth terms' model for now.
+
+   So far we have been considering inference of only the GW terms $\bar{\theta}_{\rm GW}$. It is important to be able to infer the pulsar terms $\bar{\theta}_{\rm PSR}$.
+
+   Lets try and include $f \, (t=0)$ and $\dot{f} \, (t=0)$ in our nested sampling.
+
+   We have a uniform prior of $\pm 1 \%$ about the true value for these two parameters
+
+\end{frame}
+
+\begin{frame}{Earth terms model $\bar{\theta}_{\rm GW}$ and $f \, (t=0)$, $\dot{f} \, (t=0)$}
+	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/fprior_earth_corrected_result}
+\end{frame}
+
 %	\begin{frame}{}

### 2023-05-15 docs/minutes/slides_19_05/document.tex
restart canoncial 12 and noise

--- 2023-05-15 docs/minutes/slides_11_05/document.tex
@@
 
-\date{EE/GW meeting, May 11, 2023}
+\date{EE/GW meeting, May 19, 2023}
 
@@
 
+		Last time: Problems at small $h$, nasty likelihood curves
+
+	All likelihood curves for $\bar{\theta}_{\rm GW}$:
+
+\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves}
+
+\end{frame}
+
+\begin{frame}{Drop PSR terms from model?}
+	\begin{equation}
+		f_{\rm measured} = f_{\rm emitted} \, g(\tau; \bar{\theta})
+		\label{eq:measureent}
+	\end{equation}
+	with
+	\begin{align}
+		g(\tau; \bar{\theta}) 
+		& = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
+	\end{align}
+
+	Generate the fake data using the full expression for $g(\tau; \bar{\theta})$, but use $g'(\tau; \bar{\theta})$ in the Kalman measurement model
+
+	\begin{align}
+		g'(\tau; \bar{\theta}) 
+		& = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) \right]
+	\end{align}
+\end{frame}
+
+\begin{frame}{Drop PSR terms from model?}
+
+	All likelihood curves for $\bar{\theta}_{\rm GW}$:
+	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves_psr}
+
+\end{frame}
+
+\begin{frame}
+	Full NS search for \alert{all parameters} \newline 
+	$\bar{\theta} = [\bar{\theta}_{\rm GW}, \, f, \, (t=0) \, \dot{f} \, (t=0) ]$ \newline 
+	with $h = 10^{-10}$, $\sigma_m = 10^{-11}$ \newline 
+	Note: pulsar distance is no longer part of model. $\gamma$ doesn't matter
+
+\end{frame}
+
+\begin{frame}
+		\includegraphics[width=0.9\textwidth,height=0.9\textwidth]{images/canonical_mixed_result}
+\end{frame}
+
+\begin{frame}
+	\includegraphics[width=0.9\textwidth,height=0.9\textwidth]{images/canonical_mixed_f_result}
+\end{frame}
+
+\begin{frame}
+	Next steps:
 	\begin{itemize}
-		\item \alert{Problem = overly narrow,biased posteriors from nested sampling}. Even for ``easy" systems!
+		\item Bayes factors vs. strain (currently running)
+		\item Process noise $\sigma_p$
+		\item Looser priors on $f$, $\dot{f}$? Do we need to?
+		\item Can we add PSR terms back in? Do we need to?
 	\end{itemize}
+\end{frame}
 
-Here is an easy system: $h = 10^{-2}, \sigma_m = 10^{-13}, \sigma_p = 0.0$ \newline 
-\centering
-\includegraphics[width=0.5\textwidth,height=0.5\textwidth]{images/example_system}
+\begin{frame}[standout]
+	Fin.
+\end{frame}
 
-\raggedright
-If the true declination $\delta = 1.0$, NS might give us a posterior of $\sim 0.95$ but variance v.v. small and posterior does not capture true value
+\appendix
 
+\begin{frame}[fragile]{Backup slides}
+	Sometimes, it is useful to add slides at the end of your presentation to
+	refer to during audience questions.
+
+	The best way to do this is to include the \verb|appendixnumberbeamer|
+	package in your preamble and call \verb|\appendix| before your backup slides.
+
+	\themename will automatically turn off slide numbering and progress bars for
+	slides in the appendix.
+\end{frame}
+
+\begin{frame}[fragile]{Where did $\sigma_m$ value come from?}
+The measurement noise is
+
+$$ \sigma_m = f \frac{\sigma_{\rm TOA}}{\rm cadence}$$
+
+So for a MSP ( $f \sim 100 Hz$) observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s:
+
+$$\sigma_m \sim 1.6 \times 10^{-10}$$
+
+The very best pulsars might have $\sigma_{\rm TOA} \sim 10$ ns:
+
+$$\sigma_m \sim 1.6 \times 10^{-12}$$
+\end{frame}
+
+\begin{frame}{$\gamma$ likelihood}
+	\includegraphics[width=0.8\textwidth,height=0.8\textwidth]{images/likelihood_gamma_1}
 \end{frame}
@@
 
-	All likelihood curves for $\bar{\theta}_{\rm GW}$:
-
-	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves}
-
-\end{frame}
-
-\begin{frame}{}
-
 		\begin{itemize}
@@
 
-\begin{frame}{Drop PSR terms from model?}
-	 \begin{equation}
-		f_{\rm measured} = f_{\rm emitted} \, g(\tau; \bar{\theta})
-		\label{eq:measureent}
-	\end{equation}
-	with
-	\begin{align}
-		g(\tau; \bar{\theta}) 
-		& = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
-	\end{align}
-
-Generate the fake data using the full expression for $g(\tau; \bar{\theta})$, but use $g'(\tau; \bar{\theta})$ in the Kalman measurement model
-
-		\begin{align}
-		g'(\tau; \bar{\theta}) 
-		& = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) \right]
-	\end{align}
-\end{frame}
-
-\begin{frame}{Drop PSR terms from model?}
-
-		All likelihood curves for $\bar{\theta}_{\rm GW}$:
-		\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves_psr}
-
-\end{frame}
-
 \begin{frame}{Option 2: Drop PSR terms from model?}
@@
 \begin{frame}{Earth terms model $\bar{\theta}_{\rm GW}$ and $f \, (t=0)$, $\dot{f} \, (t=0)$}
-	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/fprior_earth_corrected_result}
+	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/canonical_mixed_result}
 \end{frame}

### 2023-05-15 docs/minutes/slides_19_05/document.tex
test heterodyne run

@@
 \begin{frame}[standout]
-	Fin.
+	Backup slides
 \end{frame}
@@
 \appendix
-
-\begin{frame}[fragile]{Backup slides}
-	Sometimes, it is useful to add slides at the end of your presentation to
-	refer to during audience questions.
-
-	The best way to do this is to include the \verb|appendixnumberbeamer|
-	package in your preamble and call \verb|\appendix| before your backup slides.
-
-	\themename will automatically turn off slide numbering and progress bars for
-	slides in the appendix.
-\end{frame}
 
@@
 
-	\begin{frame}{}
+	\begin{frame}{What changed?}
+
+		AM: What changed such that now everything works?
 
 		\begin{itemize}
-		\item Why? Very nasty likelihood curves
+			\item Dropping PSR terms to smooth likelihoods
+			\item \textit{Increasing} measurement noise  
+			\item Don't need too many live points
+			\item Non-default Bilby sampler settings 
+			\begin{itemize}
+				\item e.g. docs recommend `sample=act-walk`or `sample=acceptance-walk`
+				\item More success with `sample=rwalk-dynesty` (not `sample=rwalk` which is the Bilby implementation)
 			\end{itemize}
-
-	$$  p(\theta | d) \propto p(d | \theta) \pi({\theta})$$
-	$$ p(\theta | d) \propto \mathcal{L}(\theta) \pi({\theta})$$
-
-	For uniform prior, posterior shape $\sim$ likelihood shape.
+		\item Be mindful of float errors - Python/Numpy/Bilby Nans get propagated, rather than error e.g. Fortran.
+		\end{itemize}
 

### 2023-05-18 docs/minutes/slides_19_05/document.tex
slides

@@
 \usepackage{xspace}
+
+\definecolor{links}{HTML}{2A1B81}
+
+\hypersetup{colorlinks,linkcolor=,urlcolor=links}
 
@@
 	Full NS search for \alert{all parameters} \newline 
-	$\bar{\theta} = [\bar{\theta}_{\rm GW}, \, f, \, (t=0) \, \dot{f} \, (t=0) ]$ \newline 
-	with $h = 10^{-10}$, $\sigma_m = 10^{-11}$ \newline 
+	$\bar{\theta} = [\bar{\theta}_{\rm GW}, \, f_0, \, \dot{f}_0]$ \newline 
+	with $h = [10^{-10}, 10^{-12}]$, $\sigma_m = 10^{-11}$ \newline 
 	Note: pulsar distance is no longer part of model. $\gamma$ doesn't matter
@@
 \begin{frame}
-		\includegraphics[width=0.9\textwidth,height=0.9\textwidth]{images/canonical_mixed_result}
+		\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical_mixed_result_10}
+		\only<1>{%
+			\tikz[overlay,remember picture]
+			\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-10}$,  $\bar{\theta}_{\rm GW}$};
+
+		}
+			\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=2.5cm]current page.center){Note some systematics in $\alpha$, $\psi$};
+
+	}
 \end{frame}
@@
 \begin{frame}
-	\includegraphics[width=0.9\textwidth,height=0.9\textwidth]{images/canonical_mixed_f_result}
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_result}
+	\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$,$\bar{\theta}_{\rm GW}$};
+
+	}
+	\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=2.5cm]current page.center){Note $h -\iota$ trade-off};
+
+	}
+
+\end{frame}
+
+\begin{frame}
+\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_f_result}
+\only<1>{%
+	\tikz[overlay,remember picture]
+	\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$, $f_0$, all (47) PSRs};
+
+}
+
+\end{frame}
+
+\begin{frame}
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_fdot_result}
+	\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$, $\dot{f}_0$, all (47) PSRs};
+
+	}
 \end{frame}
@@
 		\item Bayes factors vs. strain (currently running)
-		\item Process noise $\sigma_p$
-		\item Looser priors on $f$, $\dot{f}$? Do we need to?
+		\item Process noise $\sigma_p$ (next slides)
+		\item $[h, \iota] \rightarrow [h_{+}, h_{\times}]$  (discussed, easy, helpful?)
 		\item Can we add PSR terms back in? Do we need to?
+		\item Robustness for different $\bar{\theta}_{\rm GW}$ (overkill? canonical example sufficient?)
 	\end{itemize}
+\end{frame}
+
+\begin{frame}{Setting the process noise}
+
+Timing noise model of \href{https://arxiv.org/abs/1010.4794}{Shannon \& Cordes, 2010}
+\begin{eqnarray}
+ \ln \sigma_p^{\rm TOA} [\mu s] = \ln \alert{C} + \alert{\alpha} \ln \nu [s^{-1}] + \alert{\beta} \ln \dot{\nu} [10^{-15} s^{-2}] + \alert{\gamma} \ln T [years]
+\end{eqnarray}
+
+For MSP, $\ln C \sim -20$, $\alpha \sim 1$, $\beta\sim2$, $\gamma \sim 2.4$
+
+Taking $T = 1$ week, for our synthetic NANOGrav pulsars this puts:	 $$\sigma_p^{\rm TOA}  [s] \sim (10^{-19}, 10^{-17}, 10^{-13}) \, \, , \text{(min/median/max)}$$
+
+Relate to frequency:
+$$ \sigma_p^{\rm f} [Hz] = f \, \frac{\sigma_p^{\rm TOA}}{T}  \sim (10^{-23}, 10^{-21}, 10^{-16}) \, \, , \text{(min/median/max)}$$
+
+Could hit float epsilon issues here... may need to scale state evolution equations, similar to heterodyning the measured data...
+
+\end{frame}
+
+\begin{frame}{Rescaling to avoid float issues}
+	State evolution equation (Ornstein–Uhlenbeck)
+	\begin{eqnarray}
+		\frac{df}{dt}= -\gamma [f - f_{\rm EM}(t)] + \dot{f}_{\rm EM} + \xi(t; \sigma_p)
+	\end{eqnarray}
+i.e. 
+		\begin{eqnarray}
+		df = a(f,t) dt + \sigma_p dW
+	\end{eqnarray}
+
+Can solve numerically via e.g. Euler-Maruyama
+
+Partition interval $[0,T]$ into $N$ intervals, each width $\Delta t = T/N$ 
+\begin{eqnarray}
+	f_{n+1} = f_{n} + a(f_n, \tau_n) \Delta t + \sigma_p \Delta W_n
+\end{eqnarray}
+\begin{eqnarray}
+		\mathcal{O}(10^2) = \mathcal{O}(10^2) + \mathcal{O}(10^{-15}) \cdot \Delta t  + \mathcal{O}(10^{-20}) \Delta W_n
+\end{eqnarray}
+
+Practically: synthetic data (float64) generated with e.g. $\sigma_p = 10^{-19}$ = data with e.g. $\sigma_p = 10^{-20}$
+
+\end{frame}
+
+\begin{frame}{}
+	\includegraphics[width=0.90\textwidth,height=0.60\textwidth]{images/sigma_p_non_hetero}
+
+\end{frame}
+
+\begin{frame}{}
+	Let $f \, ' = f - f_{\rm EM}$
+	\begin{align}
+		\frac{df \, '}{dt}&= \frac{df}{dt} - \frac{df_{\rm EM}}{dt} \\
+		                    &= \frac{df}{dt} - \frac{df_{\rm EM}}{dt} \\
+		                    &= - \gamma f \, '  + \xi 
+	\end{align}
+	Now numerical solution looks like
+	\begin{eqnarray}
+				\mathcal{O}(10^{-12}) = \mathcal{O}(10^{-13}) \cdot \mathcal{O}(10^{-12}) \cdot \Delta t  + \mathcal{O}(10^{-20})
+	\end{eqnarray}
+	\begin{eqnarray}
+	\mathcal{O}(1) = \mathcal{O}(10^{-13}) \cdot \Delta t  + \mathcal{O}(10^{-8})
+\end{eqnarray}
+
+	Practically: synthetic data (float64) generated with e.g. $\sigma_p = 10^{-19}$ $\neq$ data with e.g. $\sigma_p = 10^{-20}$
+
+\end{frame}
+
+\begin{frame}{}
+	\includegraphics[width=0.90\textwidth,height=0.60\textwidth]{images/sigma_p_hetero}
+
+\end{frame}
+
+\begin{frame}{Heterodyne state and measurement?}
+
+	\begin{itemize}
+		\item Heterodyne the state: $f \, ' = f - f_{\rm EM}$ 
+		\item Heterodyne the measurement: $f \, '_{\rm M}  = f_{\rm M} - f^{\, *}_{\rm EM}$ 
+	\end{itemize}
+
+	$f_{\rm EM}$ is the exact, true solution parametrised by the unknowns $f_0, \dot{f}_0$
+
+	 $f^{\, *}_{\rm EM}$ is a guess based on the pulsar ephemeris.
+
+	 For our case where we have synthetic data we can set $f_{\rm EM} = f^{\, *}_{\rm EM}$, but obviously not true generally - we are trying to estimate $f_{\rm EM}$!
+
+	We can then write the measurement equation as
+	\begin{eqnarray}
+		f_{\rm M} = (1 - X)f + N_{\rm G} 
+	\end{eqnarray}
+\begin{eqnarray}
+	f \, '_{\rm M} = (1 - X)f \, ' - X f_{\rm EM} + \alert{(f_{\rm EM} - f^{\, *}_{\rm EM})} +  N_{\rm G}
+\end{eqnarray}
+
+	\begin{itemize}
+		\item Linear \checkmark (control vector has shifted from state equation to measurement equation)
+		\item Scales: $\mathcal{O}(h) = \mathcal{O}(10^{-12}) - \mathcal{O}(h)\cdot\mathcal{O}(10^2) + \alert{\mathcal{O}(0)} + \mathcal{O}(10^{-11})$ \checkmark
+	\end{itemize}
+
+\end{frame}
+
+\begin{frame}{}
+
+Insert corner plot here
+
 \end{frame}
@@
 
-		AM: What changed such that now everything works?
+		AM: \textit{"What changed such that now everything works?"}
 
@@
 			\item Don't need too many live points
+			\begin{itemize}
+				\item $n_{live} > $ number of parameters 
+				\item $n_{live} \times \frac{\rm mode \, volume}{\rm prior \, volume} > 1 $ i.e. at least 1 live point per mode
+				\item Runtime scales $\mathcal{O}(n_{live})$
+				\item Posterior/Evidence uncertainties scale as $\mathcal{O}(1 / \sqrt{n_{live}})$
+			\end{itemize}
+
 			\item Non-default Bilby sampler settings 
@@
 				\item More success with `sample=rwalk-dynesty` (not `sample=rwalk` which is the Bilby implementation)
+				\item bound = multi(dynesty), not the Bilby version 
 			\end{itemize}
-		\item Be mindful of float errors - Python/Numpy/Bilby Nans get propagated, rather than error e.g. Fortran.
+		\item Be mindful of float errors - Python/Numpy/Bilby NaNs get propagated, rather than error e.g. Fortran.
 		\end{itemize}
@@
 
-\begin{frame}{}
+\begin{frame}{More likelihood curves}
 
@@
 		\item Cause: noisy, likelihoods which are "locally Gaussian". Sampling gets stuck in a local optima and infers posteriors widths based on that optima.
-%		\item Require $n_{live} > \frac{\rm prior \, volume}{\rm mode \,  footprint} = \frac{X}{\sigma^2}$
-%		\begin{itemize}
-%			\item For $\delta$ require $n_{live} > \frac{2 \pi}{10^{-5}} \sim 10^5 $
-%			\item For $\omega$ require $n_{live} > \frac{10^4}{10^{-11}} \sim 10^7 $ 
-%			\item Currently $n_{live} = 500$ parallelized on 32 cores $\sim 700$ minutes
-%			\item Runtime scales $\mathcal{O}(n_{live}) \rightarrow 8000 $ hours for required $\delta$ sampling (just under 1 year!)
-%
+
 %		\end{itemize}
@@
 
-\begin{frame}{}
-	Potential solutions:
+\begin{frame}{Potential solutions to noisy likelihoods}
 
@@
 
-\end{frame}
-
-\begin{frame}{Option 2: Drop PSR terms from model?}
-
-	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/canonical_mixed_result}
-
-	OK, but some biases. c.f. location of the maxima on the previous slide
-\end{frame}
-
-\begin{frame}{}
-
-	Lets proceed using this `Earth terms' model for now.
-
-   So far we have been considering inference of only the GW terms $\bar{\theta}_{\rm GW}$. It is important to be able to infer the pulsar terms $\bar{\theta}_{\rm PSR}$.
-
-   Lets try and include $f \, (t=0)$ and $\dot{f} \, (t=0)$ in our nested sampling.
-
-   We have a uniform prior of $\pm 1 \%$ about the true value for these two parameters
-
-\end{frame}
-
-\begin{frame}{Earth terms model $\bar{\theta}_{\rm GW}$ and $f \, (t=0)$, $\dot{f} \, (t=0)$}
-	\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/canonical_mixed_result}
 \end{frame}

### 2023-05-22 docs/minutes/slides_25_05/document.tex
broaden sigma_p prior

--- 2023-05-18 docs/minutes/slides_19_05/document.tex
@@
 
-\date{EE/GW meeting, May 19, 2023}
+\date{EE/GW meeting, May 25, 2023}
 
@@
 
-		Last time: Problems at small $h$, nasty likelihood curves
+		Last time:
 
-	All likelihood curves for $\bar{\theta}_{\rm GW}$:
-
-\includegraphics[width=0.9\textwidth,height=0.7\textwidth]{images/all_likelihood_curves}
+\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_result}
 
 \end{frame}
+
+	\begin{frame}{}
+
+	Last time:
+
+	\begin{itemize}
+		\item Earth terms model $g'(\tau; \bar{\theta}) = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) \right]$
+		\item Zero process noise  $\sigma_p = 0$
+		\item No heterodyning
+		\item Single noise realisation
+    \end{itemize}
+
+\end{frame}
+
+	\begin{frame}{}
+
+	This time:
+
+	\begin{itemize}
+		\item Earth terms model 
+		\item \alert{Non - Zero process noise  $\sigma_p \neq 0$}
+		\item \alert{Heterodyning}
+		\item \alert{Multiple noise realisations}
+	\end{itemize}
+
+\end{frame}
+
+	\begin{frame}{}
+
+	Reminder of "heterodyned" equations
+
+	\begin{itemize}
+		\item Heterodyne the state: $f \, ' = f - f_{\rm EM}$ 
+		\item Heterodyne the measurement: $f \, '_{\rm M}  = f_{\rm M} - f^{\, *}_{\rm EM}$ 
+	\end{itemize}
+
+	Note the distinction between $ f_{\rm EM}$ and $f^{\, *}_{\rm EM}$
+
+	$f_{\rm EM}$ is the exact, true solution parametrised by the unknowns $f_0, \dot{f}_0$. This heterodyning is just a change of variables
+
+	$f^{\, *}_{\rm EM}$ is a guess of $f_{\rm EM}$ based on the pulsar ephemeris.
+It is just to scale the equations - it does not remove any degrees of freedom - we still are estimating $f_0, \dot{f}_0$ optimally!
+
+	\begin{equation}
+		\frac{df \, '}{dt} = - \gamma f \, '  + \xi 
+	\end{equation}
+
+	\begin{eqnarray}
+		f \, '_{\rm M} = (1 - X)f \, ' - X f_{\rm EM} + \alert{(f_{\rm EM} - f^{\, *}_{\rm EM})} +  N_{\rm G}
+	\end{eqnarray}
+
+	For our case where we have synthetic data we can set $f_{\rm EM} = f^{\, *}_{\rm EM}$, but obviously not true generally - we are trying to estimate $f_{\rm EM}$!
+
+\end{frame}
+
+\begin{frame}[standout]
+	Backup slides
+\end{frame}
+
+\appendix
 
@@
 
-\begin{frame}[standout]
-	Backup slides
-\end{frame}
-
-\appendix
-
 \begin{frame}[fragile]{Where did $\sigma_m$ value come from?}

### 2023-05-22 py_src/config_batch.py
corrected sigma_p prior

@@ -26,8 +26,8 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
     
     
 
-N = 10
-seeds = np.arange(1235,1235+N,1)
+N = 100
+seeds = np.arange(1235+10,1235+10+N,1)
 h = 1e-12 
 model = "earth"
 with open('batch.sh','w') as b: 

### 2023-05-30 docs/minutes/slides_25_05/document.tex
corrected sigma_p prior
loosen priors on f and fdot
bayes ratio at large scale
V5 Bayes run

@@
 		\item Single noise realisation
+		\item Single set of parameters realisation
     \end{itemize}
@@
 		\item Earth terms model 
-		\item \alert{Non - Zero process noise  $\sigma_p \neq 0$}
-		\item \alert{Heterodyning}
-		\item \alert{Multiple noise realisations}
+		\item \alert{Non - zero process noise  $\sigma_p \neq 0$} $\checkmark$
+		\item \alert{Heterodyning} $\checkmark$
+		\item \alert{Multiple noise realisations} $\checkmark$
+		\item Single set of parameters realisation
 	\end{itemize}
@@
 
-	$f_{\rm EM}$ is the exact, true solution parametrised by the unknowns $f_0, \dot{f}_0$. This heterodyning is just a change of variables
+	$f_{\rm EM}$ is the exact, true solution parametrised by the unknowns $f_0, \dot{f}_0$. This ``heterodyning" is just a change of variables
 
@@
 
-	For our case where we have synthetic data we can set $f_{\rm EM} = f^{\, *}_{\rm EM}$, but obviously not true generally - we are trying to estimate $f_{\rm EM}$!
+	For our case where we have synthetic data we can set $f_{\rm EM} = f^{\, *}_{\rm EM}$.
+In the general ``real-world" case we cannot - but this is just a scaling, not a  fundamental structure viz.
+d.o.f
+
+\end{frame}
+
+	\begin{frame}[label={first_slide}]
+
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/1237_broad_GW}
+		\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$, $\sigma_p = \mathcal{U}(10^{-21}, 10^{-19})$};
+	}
+	\only<1>{%
+	\tikz[overlay,remember picture]
+	\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.0cm]current page.center){Heterodyned equations, Full PTA};
+
+}
+	\only<1>{%
+	\tikz[overlay,remember picture]
+	\node[fill=white,text=black] at ([xshift=1.5cm,yshift=2.5cm]current page.center){$\bar{\theta}_{\rm GW}$, c.f. Slide \ref{first_slide_noise}};
+
+}
+
+\end{frame}
+
+	\begin{frame}{}
+
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/1237_broad_f}
+
+		\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){Slightly deceptive: very tight priors - to discuss};
+	}
+
+\end{frame}
+
+	\begin{frame}{}
+
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/1237_broad_fdot}
+
+	\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){Slightly deceptive: very tight priors - to discuss};
+	}
+
+\end{frame}
+
+	\begin{frame}{}
+
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/1237_broad_sigma_p}
+
+	\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){LogUniform prior};
+	}
+		\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.0cm]current page.center){Bad: not super accurate};
+
+	}
+		\only<1>{%
+	\tikz[overlay,remember picture]
+	\node[fill=white,text=black] at ([xshift=1.9cm,yshift=2.5cm]current page.center){Good: does not impact estimates of $\bar{\theta}_{\rm GW}$};
+
+}
+\end{frame}
+
+%\begin{frame}{}
+%	
+%	
+%	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/1237_broad_sigma_p}
+%	
+%	\only<1>{%
+%		\tikz[overlay,remember picture]
+%		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){LogUniform prior};
+%	}
+%\end{frame}
+
+\begin{frame}
+	But what about different realisations of a) measurement noise and b) process noise?
+
+\end{frame}
+
+	\begin{frame}[label={first_slide_noise}]
+
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/stacked_GW_plot}
+	\only<1>{%
+		\tikz[overlay,remember picture]
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){As Slide \ref{first_slide}, 5 noise realisations};
+	}
+
+\end{frame}
+
+	\begin{frame}
+
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/stacked_f_plot}
+
+\end{frame}
+
+	\begin{frame}
+
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/stacked_fdto_plot}
+
+\end{frame}
+
+	\begin{frame}
+
+	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/stacked_sgima_p_plot}
+
+\end{frame}
+
+\begin{frame}
+	How about 100 noise realisations? 
+\end{frame}
+
+	\begin{frame}
+
+	\includegraphics[width=0.32\textwidth,height=0.85\textwidth]{images/parameter_medians}
+	\includegraphics[width=0.32\textwidth,height=0.85\textwidth]{images/parameter_medians_f}
+	\includegraphics[width=0.32\textwidth,height=0.85\textwidth]{images/parameter_medians_fdot}
+
+\end{frame}
+
+	\begin{frame}	
+	\centering
+	\includegraphics[width=0.34\textwidth,height=0.85\textwidth]{images/parameter_medians_sigma_p}
+\end{frame}
+
+\begin{frame}
+	Case examples - iota, h 
+\end{frame}
+
+\begin{frame}{Bayes ratios}
+
+\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/BayesRatioPlot}
 

### 2023-05-30 py_src/plotting.py

@@ -94,8 +94,8 @@ def plot_all(t,states,measurements,predictions_x,predictions_y,psr_index,savefig
 
 
 
-def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges,axes_scales,savefig):
-
+def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges,axes_scales,savefig,logscale=False):
+    plt.style.use('science')
 
     # Opening JSON file
     f = open(path)
@@ -103,22 +103,11 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
     # returns JSON object as 
     # a dictionary
     data = json.load(f)
-
-
     print("The evidence is:", data["log_evidence"])
-
-    #Make it a dataframe. Nice for surfacing
-    df = pd.DataFrame(data["samples"]["content"]) # posterior
 
 
     #Make it a dataframe. Nice for surfacing
     df_posterior = pd.DataFrame(data["posterior"]["content"]) # posterior
-    df_samples = pd.DataFrame(data["samples"]["content"]) # posterior
-
-
-
-    #Now make it a numpy array
-    y_samp = df_samples.to_numpy() 
 
 
     print("Number of samples:")
@@ -128,57 +117,39 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
 
     medians = df_posterior[variables_to_plot].median()
     variances = df_posterior[variables_to_plot].var()
+    y_post = df_posterior[variables_to_plot].to_numpy()
 
-    selected_variables = []
-    selected_injections = []
-    selected_labels = []
-    selected_ranges = []
     for i in range(len(medians)):
-        print(injection_parameters[i],variables_to_plot[i], medians[i], variances[i])
-
-        if variances[i] > 1e-50: #is this necessary anymore?
-            selected_variables.extend([variables_to_plot[i]])
-            selected_injections.extend([injection_parameters[i]])
-            selected_labels.extend([labels[i]])
-            #selected_ranges.extend([ranges[i]])
-        else:
-            print("Note! ", variables_to_plot[i], " has zero variance and will not be plotted")
-   
+        print(labels[i], injection_parameters[i], medians[i], variances[i])
 
 
-    
-    y_post = df_posterior[selected_variables].to_numpy()
 
 
-    plt.style.use('science')
+    if logscale:
+        y_post = np.log10(y_post)
[…]

### 2023-06-13 docs/manuscript/paper.tex
document updates

@@
 
+This is how we usually search for single source PTAs
+
+estimate all unknown parameters optimally
+here is a difference approach
+
+Typical approaches for
+
 \noindent This paper is organised as follows:  \textcolor{red}{TK: TBD} \newline
@@
 
-\section{State-Space Model}
-We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, $\bar{x}$, and the measurement state recorded by an observer,$\bar{y}$.
-We will take our state variable to be the intrinsic pulsar pulse frequency $f(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar.
-We will now consider how the intrinsic pulse frequency evolves in time, completely separate from the influence of a GW, and then go on to derive the  influence of a GW perturbation on the frequency recorded by an observer at Earth.
-\textcolor{red}{Need some extra discussion here on why this choice of state/measurement space.
-Why not e.g.
-phase, TOA, voltage, etc.
-?}
+\section{State-Space Model}\label{sec:model}
+We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, and the measurement state recorded by an observer.
+For this work we will take our state variable to be the intrinsic pulsar pulse frequency $f_{\rm p}(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar.
+We will take our measurement variable to be the pulsar pulse frequency that is measured by an observer at Earth, $f_{\rm m}(t)$.
+In Sections \ref{sec:psr_frequency}, \ref{sec:psr_measured} we will describe our model for $f_{\rm p}(t)$ and  $f_{\rm m}(t)$ respectively.
+We will go on in Section \ref{sec:statespace} to bring together these two models into a cohesive, discretized state space model.
+\textcolor{red}{TK: Need some extra discussion here on why this choice of state/measurement space. Why not e.g. phase, TOA, voltage, etc. ?}
 
-\subsection{Evolution of the pulsar frequency}
-We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \cite{Vargas}.
-Within this model, $f$ evolves according to a combination of both deterministic torques (i.e.
+\subsection{Evolution of the pulsar frequency} \label{sec:psr_frequency}
+We will take as our model of the intrinsic pulsar frequency $f_{\rm p}(t)$ a variation of the phenomenological model of \cite{Vargas}.
+Within this model, $f_{\rm p}(t)$ evolves according to a combination of both deterministic torques (i.e.
 electromagnetic spin-down) and stochastic torques (i.e.
@@
 \begin{equation}
-	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
+	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p} - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
 	\label{eq:frequency_evolution}
 \end{equation}
-where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation,$\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
+where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
 \begin{equation}
@@
 \begin{equation}
-	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
+	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p}- f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
 	\label{eq:frequency_evolution_expanded}
@@
 
-\subsection{Modulation of pulsar frequency due to a GW}
-In the presence of a GW, the $f(t)$  measured by an observer in the local NS reference frame is different from that measured by an observer on Earth.
-We want to determine quantitatively determine this difference - how the GW influences the received pulse frequency.
+\subsection{Modulation of pulsar frequency due to a GW} \label{sec:psr_measured}
+In the presence of a GW, the pulse frequency measured by an observer in the local neutron star reference frame is different from that measured by an observer on Earth.
+We will now derive the influence of the GW on the received pulse frequency.
 
@@
 \end{equation}
-for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, \textcolor{red}{(TK: We have a notation conflict between $x$ = state and $x$ = coordinates)} where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
-We emphasise that $\Omega$ has no time dependence - we are concerned solely with monochromatic sources.
-Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{\textcolor{red}{TK: This point is important, since the phase offset is then the same between multiple pulsars.
-See also Melatos 2022 PT12}}.
+for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
+We emphasise that for this work $\Omega$ has no time dependence - we are concerned solely with monochromatic sources.
+We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the Earth.
 The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
@@
 \end{align}
-where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave. The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
+where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave which are given by
+\begin{align}
+	h_+ &= h(1 + \cos^2 \iota) \\
+		h_{\times} &= -2h\cos \iota
+\end{align}
+The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
 \begin{align}
@@
 \end{align}
-which are in turn specified via the location of the GW source on the sky (via $\theta$, $\phi$ coordinates) and the polarisation angle $\psi$
+which are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coodinate $\phi$) and the polarisation angle $\psi$
 \begin{align}
@@
 \begin{equation}
-	\nu = p_{\alpha} u^{\alpha}
+	f = p_{\alpha} u^{\alpha}
+	\label{eq:freq_temporal}
 \end{equation}
@@
 	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
+	\label{eq:integral}
 \end{equation}
@@
 Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.
-\footnote{\textcolor{red}{TK: See also e.g.
-Maggiore, Meltaos who takes the same approach...}} \newline
-
-\noindent To complete our expression, we now just need to determine $\dot{p}_t$ and integrate it.
 
@@
 \begin{equation}
-	H(x^{\mu}, p_{\mu}) = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
+	H = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
 \end{equation}
@@
 \begin{equation}
-	\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t))+\Phi_0}
+	\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)}
 \end{equation}
-\footnote{\textcolor{red}{TK: This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset}.}
-
-\noindent Therefore the derivative w.r.t coordinate time $t$ is,
+\noindent The derivative w.r.t coordinate time $t$ is
 \begin{equation}
@@
 \end{equation}
-\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation.
+\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic. 
+It will prove useful to recognise that 
+\begin{equation}
+	p^{\mu} = \omega(1,-q^x,-q^y,-q^z)
+\end{equation} 
+where $\bar{q}$ is the unit vector between the Earth and pulsar and $\omega$ is the \textit{constant} photon angular frequency.
+Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer.
+Similarly, we can parameterize the photon spatial coordinates $\bar{x}$ as,
+\begin{equation}
+	\bar{x}(t) = - \bar{q} (t - t_1)
+\end{equation} 
+Note that when our integration variable $t = t_1$ (i.e.
+the photon is at Earth) then $\bar{x}(t=t_1) = 0$ which is what we expect given our chosen coordinate system.
+Also note that $\dot{p}_t$ is entirely a function of the GW perturbation.
 In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.
-It will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar and $\omega$ is the \textit{constant} photon angular frequency.
-Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer \footnote{ \textcolor{red}{TK: see Melatos 2022, PT16}}.
-We will consider  the pulsar locations to be constant with respect to the Earth.
+We will treat the pulsar locations to be constant with respect to the Earth, i.e.
+$\bar{q}$ is not a function of time.
 In practice, the pulsar locations vary with respect to the Earth, but are constant with respect to the solar system barycentre (SSB).
@@
 Our conclusions are generally unchanged by this choice - it is straightforward to transform $q$ if needed into a vector between the SSB and pulsar.
-The spatial coordinates along the photon path are related to the pulsar as $\bar{x}(t) = -\bar{q}t$.
 \newline
@@
 \begin{equation}
-	\dot{p}_t = A e^{i \gamma t + \Phi_0}
+	\dot{p}_t = A e^{i B t + C}
 \end{equation}
-where 
-\begin{equation}
-	\gamma = -\Omega (1 + \bar{n}\cdot \bar{q}) 
-\end{equation}
-\footnote{\textcolor{red}{TK: compare with Melatos 2022, PT11}}
-\noindent and
+
+with 
 \begin{equation}
@@
 \end{equation}
-
-\subsubsection{Performing the integral}
-
-The frequency shift experienced by the observer relative to the source due to a GW is then
+\begin{equation}
+	B = -\Omega (1 + \bar{n}\cdot \bar{q}) 
+\end{equation}
+and 
+\begin{equation}
+	C = \Omega \tau \bar{n} \cdot \bar{q}  +\Phi_0
+\end{equation}
+From Eq \ref{eq:integral} the frequency shift experienced by the observer relative to the source due to a GW is then
 \begin{align}
-	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = A \int_{t = \tau - d}^{t=\tau} e^{i \gamma t + \Phi_0}dt  \\
-	=\frac{-iA}{\gamma} e^{i \gamma \tau}e^{\Phi_0}[1 - e^{-i \gamma d}] 
+	&p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = \\
+	 & A \int_{t = \tau - d}^{t=\tau} e^{i B t + C}dt  \\
+	& =\frac{-iA}{B} e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i B d}\right) \\
+	&= \frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right)
 	\label{eq:concise}
 \end{align}
-We can write this more explicitly as,
-\begin{align} \label{eq:final}
-	&p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = \\
-	&\frac{ \omega}{2} \frac{H_{ij}q^i q^j  }{(1 + \hat{n} \cdot \hat{q})}e^{-i \Omega \tau(1+\bar{n} \cdot \bar{q}) + \Phi_0} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
+for $h_{ij} = g_{ij} - \eta_{ij}$. Since via Equation  \ref{eq:freq_temporal} we can related the temporal component of the covariant momentum to the frequency, it follows that
+\begin{align}
+	&f(\tau)|_{\rm Earth} = f(\tau - d)|_{\rm source}  \left( 1- \frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \right)
+	\label{eq:main_eq}
 \end{align}
-or alternatively, with $h_{ij} = g_{ij} - \eta_{ij}$,
+where we have identified $\omega$ with $f(\tau - d)|_{\rm source}$.
+
+\section{Detection and parameter estimation}
+Let's review and categorise all the free parameters,$\bar{\theta}$, of above model.
+We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source
 \begin{equation}
-	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{h_{ij} q^i q^j  }{(1 + \hat{n} \cdot \hat{q})} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
-	\label{eq:freq_shift}
+	\bar{\theta} =  \bar{\theta}_{\rm PSR} \cup \bar{\theta}_{\rm GW}
 \end{equation}
+\begin{equation}
+	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]^{(N)}
+\end{equation}
+\begin{equation}
+	\bar{\theta}_{\rm GW} = [h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0]
+\end{equation}
+Whilst the GW parameters are shared between measurements for each pulsar, the pulsar parameters are clearly not.
+For a PTA dataset of $N$ pulsars we have $7 + 5N$ parameters to estimate.
+Note that whilst this is a large parameter space, in general the pulsar parameters are much better constrained than the GW parameters: for example we have rough estimates for the pulsar distances accurate to $\sim$ 10$\%$, but we have no prior information on the source location.
+\newline
 
-\subsection{State-space structure} \label{sec3}
-We now have a model for the evolution of the intrinsic pulsar frequency states and the measured frequency timeseries.
-We can express the intrinsic frequency evolution, Eq.
-\ref{eq:frequency_evolution_expanded}, in a alternative form as,
+In this section we present a new method to infer the model parameters and calculate the marginal likelihood (i.e.
+the model evidence).
+In Section \ref{sec:statespace} we will present a discretised version of the model of Section \ref{sec:model} which maps onto the discretely sampled observable $f(\tau)|_{\rm Earth}$.
+In Section \ref{sec:kalman_filter} we will outline how observations of this pulse frequency can be used to recover the underlying intrinsic pulsar state $f(\tau -d)|_{\rm source}$ using a Kalman Filter.
+In Section \ref{sec:nested_sampling} we will demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to recover the model parameters.
+We will now review the Kalman filter and the associated likelihood in Section \ref{sec:kalman_filter} before going on in Section \ref{sec:nested_sampling} to discuss how to use nested sampling methods together with the filter for detection (model selection) and parameter estimation.
+
+\subsection{Discretised model} \label{sec:statespace}
+We can express the intrinsic frequency evolution, Eq. \ref{eq:frequency_evolution_expanded}, in a alternative form as,
 \begin{equation}
@@
 \end{equation} 
-If we move from a solution in continuous time to discrete time then
+If we move from a solution in continuous time, $t$, to discrete time, $\bar{t} = (t_1, t_2, ...,t_K)$, then
 \begin{equation}
@@
 \begin{equation}
-	F_i = e^{A (t_{i+1} - t_i)}
+	F_i = e^{\mathcal{A} (t_{i+1} - t_i)}
 \end{equation}
 \begin{equation}
-	T_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt'
+	T_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} N(t') dt'
 \end{equation}
 \begin{equation}
-	\eta_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} \sigma dt'
+	\eta_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} \sigma dt'
 \end{equation}
-Given the discrete solution $f(\bar{t})$ where $\bar{t} = (t_1, t_2, ...,t_N)$ the intrinsic frequency can be related to the measured frequency as 
+The discrete solution $f(\bar{t})$ to the intrinsic frequency can be related to the discrete measured frequency via Eq. \ref{eq:main_eq} as,
 \begin{equation}
-	f_M (\bar{t})= f g(\bar{\theta},\bar{t}) + N_M
+	f_M (\bar{t})= f(\bar{t}) g(\bar{\theta},\bar{t}) + N_M
 \end{equation}
-where $g(\theta,t)$ ("measurement function") is a function of some constant parameters, $\bar{\theta}$, and discrete time $\bar{t}$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
+where $g(\theta,t)$ can be expressed in a trigonometric form as 
+\begin{equation}
+		X = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
+\end{equation} 
+whilst $N_M$ is a Gaussian measurement noise that satisfies 
 \begin{equation}
@@
 \end{equation}
-for variance $\Sigma^2$. The measurement function follows from Eq. \ref{eq:freq_shift}
-\begin{equation} \label{eq:final}
-	g(\bar{\theta},t) = 1 - \frac{1}{2} \frac{h_{ij}(t) q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
-\end{equation}
-or in a more concise form where we have switched to a trigonometric form of the equations and made the time dependence explicit,
-\begin{equation} 
-	g(\bar{\theta},t) = 1 -A \cos(-\Omega t (1 + n\cdot q ) + \Phi_0)
-\end{equation}
-where 
-\begin{equation}
-	A =  \frac{1}{2} \frac{H_{ij} q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - \cos\left( {\Omega(1+\bar{n} \cdot \bar{q}) d}\right)]
-		\label{eq:state2}
-\end{equation}
-
-\section{Detection and parameter estimation}
-Let's  review and categorise all the parameters of above state-space model.
-We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar, the GW parameters and the noise parameters i.e.
-\begin{equation}
-	\bar{\theta} =  \bar{\theta}_{\rm PSR} \cup \bar{\theta}_{\rm GW} \cup \bar{\theta}_{\rm noise}
-\end{equation}
-\begin{equation}
-	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]^{(N)}
-\end{equation}
-\begin{equation}
-	\bar{\theta}_{\rm GW} = [h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0]
-\end{equation}
-\begin{equation}
-	\bar{\theta}_{\rm noise} = [\sigma, \Sigma]
-\end{equation}
-\textcolor{red}{TK: $\sigma$ is currently a parameter shared between all pulsars in current code implementation.
-Probably need to update this since different pulsars have different process noise, but need to check identifiability and likelihood curves - only enters the Kalman filter only through the process noise matrix...}.
-Whilst the GW parameters are shared between measurements for each pulsar, the pulsar parameters are clearly not.
-For a PTA dataset of $N$ pulsars we have $7 + 5N$ parameters to estimate.
-For now we will consider the measurement noise to be known, although in principle this too could be estimated.
-Note that whilst this is a large parameter space, in general the pulsar parameters are much better constrained than the GW parameters: for example we have rough estimates for the pulsar distances accurate to $\sim$ 10$\%$, but we have no prior information on the source location.
-\newline
-
-\noindent We want be able to infer the values of the model parameters and look for evidence of a GW signal in a data stream.
-To this end, we will use a Kalman filter in to obtain likelihood estimates of the data given a set of parameters.
-We will then deploy the Kalman filter in conjunction with a nested sampling technique for optimisation of the parameters over the likelihood.
-We will now review the Kalman filter and the associated likelihood in Section \ref{sec:kalman_filter} before going on in Section \ref{sec:nested_sampling} to discuss how to use nested sampling methods together with the filter for detection (model selection) and parameter estimation.
-We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.
+for variance $\Sigma^2$.
 
@@
 
+\subsection{Practical considerations}
+
+heterodyeing
+
+pulsar terms 
+For now we will consider the measurement noise to be known, although in principle this too could be estimated.
+
+\section{Tests with synthetic data}
+
+We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.
+
 \subsection{PTA pulsars}\label{sec:pta_pulsars}
@@
 
-\subsection{Parameter estimation on synthetic data}\label{sec:param_estimation}
+\subsection{Parameter estimation}\label{sec:param_estimation}
 We are now in a position to try to infer the parameters of a GW system.
@@
 
-\subsection{Detection on synthetic data}\label{sec:detection}
+\subsection{Detection}\label{sec:detection}
 

### 2023-06-13 docs/manuscript/paper.tex
paper

@@
 \newline
+
+THe standard method of detecting single source psr is as follows
+
+drawback is does not account for noises, ooptimal estimators etc etc.
 
@@
 \end{equation}
+
+if we specialise to the case of constatn time sampling
+
 The discrete solution $f(\bar{t})$ to the intrinsic frequency can be related to the discrete measured frequency via Eq. \ref{eq:main_eq} as,

### 2023-06-15 docs/manuscript/paper.tex
rough first draft of intro

@@
 \author[Kimpson]{Kimpson$^{1}$, Melatos, O'Leary, Evans, others, etc. etc. %
-\thanks{Contact e-mail: \href{mailto:mn@ras.ac.uk}{mn@ras.ac.uk}}%
+\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
 \thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
@@
 % These dates will be filled out by the publisher
-\date{Last updated 2020 June 10; in original form 2013 September 5}
+%\date{Last updated 2020 June 10; in original form 2013 September 5}
+\date{Last updated \today}
 
 % Enter the current year, for the copyright statements etc.
-\pubyear{2020}
+\pubyear{2023}
 
@@
 \newpage
-
 \section{Introduction}
-
-The detection of high frequency ($\sim1$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
-Gravitational radiation from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
+The detection of high frequency ($\sim 100$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
+GWs from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
 Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
-The foremost technique for the detection of GWs in this nano-Hz regime is a via timing an ensemble of milliseconds pulsars; a pulsar timing array (PTA) \citep{2021hgwa.bookE...4V}.
-The presence  of a nano-Hz gravitational wave will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the pulsar timing signal.
-By measuring the modulation of the received pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs.
+The foremost technique for the detection of GWs in this nano-Hz regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}.
+The presence  of a nHz gravitational wave will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal.
+By measuring the modulation of the pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs.
+Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}.
+These previously disparate efforts have now been joined in international collaboration, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}.
 \newline
 
-\noindent Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}.
-These previously disparate efforts have now been joined in international collaboration, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}.
-The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries (SMBHBs) with masses $\sim 10^7 M_{\odot}$.
-These GW signals from SMBHBs can be broadly classified into either deterministic or stochastic.
-For the former, sufficiently bright and near binaries may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Zhu10}.
-For the latter, the incoherent superposition of multiple weaker SMBHBs sources leads to a stochastic background detectable at nano-Hz frequencies \citep{Sesana10}.
-Other potential sources for PTAs include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}, but the deterministic and stochastic GW signals from SMBHBs remain the primary targets.
+The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R}.
+Other potential sources for PTAs include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
+The incoherent superposition of multiple SMBHBs sources leads to a stochastic GW background detectable at nHz frequencies \citep{Sesana10}.
+Previous efforts have mainly focused on the detection of this stochastic background via measuring the correlation in the pulsar timing residuals between any two pair of pulsars; the presence of a GW induces a characteristic correlation as a function of the angular separation between the pulsars - the Hellings-Downs curve \citep{Hellings}.
+However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}.
 \newline
 
-\noindent The detection of loud, resolved sources with a PTA typically involves a parametrised model for the pulsar timing residuals induced by the modulation of the pulsar signal by a GW.
-One can then search for evidence that this model describes the data via the usual Bayesian likelihood techniques, and try to estimate the parameters of the model \citep[e.g.][]{Babak2016}.
-For detecting the stochastic background the approach is different; one measures the correlation in pulsar timing residuals between any two pair of pulsars.
-The presence of a GW induces a characteristic correlation function as a function of the angular separation between the pulsars; the Hellings-Downs curve \citep{Hellings}.
-For both classes of source, the detection of GW signals in the timing residuals of a PTA is a challenging enterprise, and currently neither a stochastic background nor an individually resolved source has yet been detected \citep{2022MNRAS.510.4873A, 10.1093/nsr/nwx126}.
-\newline
-
-\noindent The sensitivity of a PTA to a GW signal is heavily dependent on the total level of noise in the array \citep{Wang2015}.
-In particular, the intrinsic pulsar timing noise - i.e.
+SMBHBs which are sufficiently near and massive may be individually resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
+Moreover, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}.
+However, as with the stochastic background, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Zhu2014PPTA,Babak2016,Arzoumanian2023}.
+The intrinsic pulsar timing noise - i.e.
 random variations in the pulse arrival time - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}.
@@
 
-THe standard method of detecting single source psr is as follows
+In this work we present a novel approach to PTA data analysis for individual monochromatic SMBHB GW sources.
+This method can self-consistently account for the intrinsic stochastic processes inherent to the pulsar.
+We formulate PTA analysis as a state-space problem and demonstrate how to track the state-space evolution optimally by using a Kalman filter, determine the presence of a GW signal in the pulsar data and infer the underlying source parameters.
+\newline
 
-drawback is does not account for noises, ooptimal estimators etc etc.
+\noindent This paper is organised as follows.
+In Section \ref{sec:model} we present our state-space model for the pulsar pulse frequency which is subject the influence of a GW.
+In Section \ref{sec:detect} we discretise the model, develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence.
+In Section \ref{sec:testing} we test this new method on synthetic pulsar data.
+Throughout this work ee adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature.
+\newline
 
-\noindent In the standard approach to PTA-GW analysis, timing noise is a nuisance which must be accurately modelled and subtracted.
-It fundamentally limits the sensitivity of the PTA detector to a GW signal, and the ability of an observer to infer the GW source parameters.
-Motivated by these challenges faced by the classic PTA analysis methods, in this work we present a novel approach to formulate PTA analysis and GW detection as a state-space problem.
-This approach enables the pulsar state-space evolution to be tracked optimally, given a specific realisation of the pulsar process noise (i.e.
-the spin wandering of the pulsar), and determine both the presence of a GW signal in the pulsar data, and  infer the underlying source parameters.
-For this initial exploratory study we will focus exclusively on resolved, monochromatic GW sources.
-\textcolor{red}{TK: This motivation section is weak.
-Need to think about how to phrase this better.
-Be clear on the real challenges of the "old" approach and the advantages of the "new" one.} \newline
-
-This is how we usually search for single source PTAs
-
-estimate all unknown parameters optimally
-here is a difference approach
-
-Typical approaches for
-
-\noindent This paper is organised as follows:  \textcolor{red}{TK: TBD} \newline
-
-\noindent We adopt the natural units, with $c = G = \hbar = 1$, and a
-$(-,+,+,+)$ metric signature. \newline
-
-\section{State-Space Model}\label{sec:model}
+\section{State-Space Frequency Model}\label{sec:model}
 We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, and the measurement state recorded by an observer.
@@
 In Sections \ref{sec:psr_frequency}, \ref{sec:psr_measured} we will describe our model for $f_{\rm p}(t)$ and  $f_{\rm m}(t)$ respectively.
+
 We will go on in Section \ref{sec:statespace} to bring together these two models into a cohesive, discretized state space model. 
@@
 \end{align}
-which are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coodinate $\phi$) and the polarisation angle $\psi$
+which are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coordinate $\phi$) and the polarisation angle $\psi$
 \begin{align}
@@
 
-\section{Detection and parameter estimation}
+\section{Detection and parameter estimation} \label{sec:detect}
 Let's review and categorise all the free parameters,$\bar{\theta}$, of above model.
@@
 
-\section{Tests with synthetic data}
+\section{Tests with synthetic data} \label{sec:testing}
 

### 2023-06-15 py_src/kalman_filter.py

@@ -160,7 +160,7 @@ class KalmanFilter:
 
         #Initialise x and P
         x = self.x0 # guess that the intrinsic frequencies is the same as the measured frequency
-        P = np.ones(self.Npsr) * sigma_m * 1e3 #Guess that the uncertainty in the initial state is a few orders of magnitude greater than the measurement noise
+        P = np.ones(self.Npsr)* sigma_m * 1e3 #Guess that the uncertainty in the initial state is a few orders of magnitude greater than the measurement noise
 
 
         #Precompute the influence of the GW
@@ -199,6 +199,9 @@ class KalmanFilter:
             obs = self.observations[i,:]
             x_predict, P_predict   = predict(x,P,F,Q)
             x,P,l,ypred = update(x_predict,P_predict, obs,R,X_factor[i,:],f_EM[i,:])
+            
+           # if i > 250:
+                #print("barp")
             likelihood +=l
 
             x_results[i,:] = x

### 2023-06-15 py_src/pulsars.py

@@ -28,11 +28,30 @@ class Pulsars:
         self.f         = pulsars["F0"].to_numpy(dtype=NF)
         self.fdot      = pulsars["F1"] .to_numpy(dtype=NF)
         self.d         = pulsars["DIST"].to_numpy(dtype=NF)*1e3*pc/c #this is in units of s^-1
-        self.γ     = np.ones_like(self.f,dtype=NF) * 1e-13  #for every pulsar let γ be 1e-13
+        self.γ         = np.ones_like(self.f,dtype=NF) * 1e-13  #for every pulsar let γ be 1e-13
         self.δ         = pulsars["DECJD"].to_numpy()
         self.α         = pulsars["RAJD"].to_numpy()
-        self.q         = unit_vector(np.pi/2.0 -self.δ, self.α) #3 rows, N columns
 
+
+        if SystemParameters.orthogonal_pulsars:
+            print("Creating orthogonal pulsars")
+            #Overwrite true  RA/DEC
+            GW_direction = unit_vector(np.pi/2.0 -SystemParameters.δ, SystemParameters.α) 
+            print("GW_direction:", GW_direction)
+            for i in range(len(self.f)):
+                v = random_three_vector()
+                psr_directon = np.cross(GW_direction,v) # not a unit vector 
+                mag = np.sqrt(psr_directon.dot(psr_directon))
+                psr_directon = psr_directon/mag #now a unit vector 
+               
+                dec_i, ra_i = convert_vector_to_ra_dec(psr_directon)
+                self.δ[i] = dec_i 
+                self.α[i] = ra_i  
+                #print(np.dot(GW_direction,psr_directon))
+
+
+
+        self.q         = unit_vector(np.pi/2.0 -self.δ, self.α) #3 rows, N columns
         #Create a flattened q-vector for optimised calculations later
         self.q_products = np.zeros((len(self.f),9))
         k = 0
@@ -80,5 +99,32 @@ def unit_vector(theta,phi):
     return np.array([qx, qy, qz]).T
 
 
+def random_three_vector():
+    """
+    Generates a random 3D unit vector (direction) with a uniform spherical distribution
+    Algo from http://stackoverflow.com/questions/5408276/python-uniform-spherical-distribution
+    :return:
+    """
+    phi = np.random.uniform(0,np.pi*2)
[…]

### 2023-06-15 py_src/synthetic_data.py

@@ -45,7 +45,7 @@ class SyntheticData:
             GW_function = gw_earth_terms
             logging.info("You are using just the Earth terms in your synthetic data generation")
 
-
+        
         X_factor = GW_function(
                                         P.δ,
                                         P.α,

### 2023-06-15 py_src/system_parameters.py

@@ -28,8 +28,10 @@ class SystemParameters:
                  Npsr = 0,         # Number of pulsars to use in PTA. 0 = all
                  use_psr_terms_in_data=True, # when generating the synthetic data, include pulsar terms?
                  measurement_model='pulsar',# what do you want the KF measurement model to be? One of pulsar, earth,null
-                 seed = 1234, #this is the noise seed. It is used for sdeint and gaussian measurement noise
-                 σp_seed=1234): #this is the seed when we randomly generate simga_p parameter values
+                 seed = 1234,       # this is the noise seed. It is used for sdeint and gaussian measurement noise
+                 σp_seed=1234,      # this is the seed when we randomly generate simga_p parameter values
+                 orthogonal_pulsars=False #if True overwrite true RA/DEC of pulsars and create a ring of pulsars perpendicular to GW direction
+                 ): 
 
         logging.info("Welcome to the Kalman Filter Nested Sampler for PTA GW systems")
 
@@ -52,6 +54,7 @@ class SystemParameters:
         self.measurement_model = measurement_model
         self.seed = seed
         self.sigma_p_seed = σp_seed
+        self.orthogonal_pulsars =orthogonal_pulsars
 
         logging.info(f"Random seed is {self.seed}")
         if σp ==1.0:

### 2023-06-15 docs/canonical_PTA_approach/canonical_PTA_approach.tex

--- 2023-04-20 docs/measurement_eqn_deerivation/LectureNotes_Math571.tex
@@
 	\normalfont\normalsize 
-	{Melbourne University - Summer 2023} \\ [0pt] % Your university, school and/or department name(s)
-	\huge Notes on state-space formulation of PTA problem% The assignment title
+	{Melbourne University} \\ [0pt] % Your university, school and/or department name(s)
+	\huge Canonical method for single source detection with PTA% The assignment title
 }\author{T. Kimpson} % Your name
@@
 \section{Preamble}
+These notes summarise the "canonical" method by which the pulsar/PTA community detect single GW sources (i.e. a supermassive BH binary inspiral, rather than a stochastic background). \newline
 
-These notes collect some ideas around how to formulate the PTA data analysis as a state-space problem. \newline
+\noindent We will consider \href{https://arxiv.org/abs/1003.0677}{Sesana \& Vecchio 2010} (SV2010 hereafter) and \href{https://arxiv.org/abs/1204.4218}{Ellis et.
+al.
+2012} (Ellis2012 hereafter) as representative of these canonical methods.
+Other methods exist (e.g.
+\href{https://ui.adsabs.harvard.edu/abs/2016MNRAS.461.1317Z/abstract}{Zhu et al 2016}) but they are generally derivative of these two original works (e.g.
+Zhu 2016 uses the same base equations, but then works in the frequency domain and explores impact of PSR terms on SNR).
 
-\noindent We will frequently reference previous works from A.Melatos, in particular Melatos 2018 (GR assignment UniMelb) and Melatos 2022 (private communication at UniMelb). \newline
-
-\noindent Geometric units ($c=\hbar = G = 1$) are used throughout, with the usual convention of Roman indices ($i,j$) for spatial indices.
-Note that $i$ is used for both labelling tensor indices and as the imaginary number - I trust the difference to be obvious from the context.\newline
-
-\section{Derivation of pulsar frequency modulation due to GW}
-We want to know how the pulse frequency from a pulsar is influenced by a passing gravitational wave (GW). \newline
-
-\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$  \footnote{See Melatos 2022 for nuances and discussion around this construction of "pulse train as photon"} \newline
-
-\noindent We take a gravitational plane wave that perturbs a background Minkowski spacetime as \footnote{c.f. Melatos 2018, Equation 1}
-\begin{equation}
-g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
-\end{equation}
-where the GW has angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
-Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.} The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
-\begin{align}
-	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
-\end{align}
-\noindent Note that we use a convention where the amplitudes $h_{+,\times}$ label the \textit{constant} amplitudes of the gravitational plane wave.
-All time variation in the strain is a result of the exponential term.
-This is in contrast to the notation used by some authors \footnote{see e.g.
-arXiv:1003.0677} where the amplitudes include the time variability $h_{+,\times} = h_{+,\times}(t)$.
-The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave.
+\section{Fundamental Equations}
+Both SV2010 and Ellis2012 treat the PSR timing residual as the fundamental observable.
+That is, they assume the true PSR evolution is "known" - described by a pulsar timing model.
+Deviations between this known solution and the what they actually observe at the detector is characterised by a residual.
+The GW imprints onto these residuals.
 \newline
 
-\subsection{Setting up the problem}
-The frequency of a photon with 4-momentum $p_{\mu}$ recorded by an observer with 4-velocity $u^{\mu}$ is 
-\begin{equation}
-	\nu = p_{\alpha} u^{\alpha}
-\end{equation}
+\noindent The general approach is as follows:
 
-\noindent We consider both our emitter and receiver to be stationary, such that  
-\begin{equation}
-u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
-\end{equation}
+\begin{enumerate}
+	\item Take the data timeseries for a single pulsar as a residual corrupted by some noise
+	\begin{eqnarray}
+		d_{\alpha}(t) = r_{\alpha}(t) + n_{\alpha}(t)
+	\end{eqnarray}
 
-\noindent Consequently the frequency can be identified with the temporal component of the covariant 4-momentum,
-\begin{equation}
-\nu = p_t
-\end{equation}
+\item Use TEMPO and least-squares fitting to generate some post-fit residuals $\tilde{r}_{\alpha}(t)$ . i.e. update pulsar parameter estimations given this new data
 
-\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
+\item Define a signal model, $s$, for the timing residuals due to the influence of a GW. Define a Gaussian likelihood $\mathcal{L}(s|\tilde{r})$.
 
-\begin{equation}
-	p_t(\tau)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=\tau} \dot{p}_t dt
-	\end{equation}
+\item Drop the pulsar terms and derive an $F$-statistic by analytically maximising over the extrinsic parameters $\left(\frac{\mathcal{M}^{5/3}}{D}, \iota, \Phi_0, \psi \right)$ Note that for a single pulsar the maximisation is ill-poised.
+At least two pulsars are required.
 
-\noindent where the overdot denotes a derivative w.r.t.
-$t$.
-Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $\tau = t_0 + d$ and consider the photon trajectory to be an unperturbed path.
-\footnote{See also e.g.
-Maggiore who takes the same approach...} \newline
+\item Perform a maximum likelihood search given the $F$-statistic to search for the intrinsic parameters.
 
-\noindent To complete our expression, we now just need to determine $\dot{p}_t$ and integrate it.
-
-\subsection{Hamiltonian Mechanics}
-
-The Hamiltonian in covariant notation can be written as
-
-\begin{equation}
-H(x^{\mu}, p_{\mu}) = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
-\end{equation}
-
-\noindent which if we substitute in our expression for the perturbed metric is
-
-\begin{equation}
-H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
-\end{equation}
-
-\noindent Hamilton's equations are
-\begin{equation}
-\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
-\end{equation}
-
-\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
-\begin{equation}
-\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)}
-\end{equation}
-\footnote{This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset.}
-
-\noindent Therefore the derivative w.r.t coordinate time $t$ is,
-\begin{equation}
-\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
-\end{equation}
-It will be useful to recognise that we can express the photon momentum in terms of the  unit vector between the Earth and pulsar $\bar{q}$
-\begin{equation}
-	p^{\mu} = \omega(1,-q^x,-q^y,-q^z)
-\end{equation}
-where $\omega$ is the \textit{constant} photon angular frequency.
-Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer (see Melatos 2022, PT16).
-Similarly, we can parameterize the spatial coordinates $\bar{x}$ as,
-\begin{equation}
-	\bar{x}(t) = - \bar{q} (t - \tau)
-\end{equation} 
-Note that when our integration variable $t = \tau$ (i.e.
-the photon is at Earth) then $\bar{x}(t=\tau) = 0$ which is what we expect given our chosen coordinate system.
-Also note that $\dot{p}_t$ is entirely a function of the GW perturbation.
-In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.
-\newline
-
-\noindent Bringing this all together
-
-\begin{align}
-	\dot{p}_t &= \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right) \\  
-	&= -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)} \left(\frac{1}{p^t}\right)
-	\\
-	&= \frac{-i\Omega \omega^2}{2} H_{ij}q^i q^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)} \left(\frac{1}{\omega}\right) \\
-	& = \frac{-i\Omega \omega}{2} H_{ij}q^i q^j  e^{i(\Omega(\bar{n}\cdot [- \bar{q} (t - \tau)] - t)+\Phi_0)} \\
-				& = \frac{-i\Omega \omega}{2} H_{ij}q^i q^j  e^{i(-\Omega t (1+ \bar{n} \cdot \bar{q}) + \Omega \tau  \bar{n} \cdot \bar{q} +\Phi_0)} \\
-						& =A e^{i (\gamma t + B)}
-\end{align}
-with 
-\begin{equation}
-	A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
-\end{equation}
-\begin{equation}
-	\gamma = -\Omega (1 + \bar{n}\cdot \bar{q}) 
-\end{equation}
-\footnote{compare with Melatos 2022, PT11}
-and 
-\begin{equation}
-	B = \Omega \tau \bar{n} \cdot \bar{q}  +\Phi_0
-\end{equation}
-
-\subsection{Performing the integral}
-
-The frequency shift experienced by the observer relative to the source due to a GW is then
-\begin{align}
-p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} &= A \int_{t = \tau - d}^{t=\tau} e^{i (\gamma t + B)}dt  \\
-&=\frac{-iA}{\gamma} e^{i B} \left(e^{i \gamma \tau} -e^{i \gamma (\tau - d)}\right) \\
-&=\frac{-iA}{\gamma} e^{i \Omega \tau \bar{n} \cdot \bar{q}} e^{i \Phi_0}e^{i \gamma \tau} \left(1 -e^{-i \gamma d}\right) \\
-&=\frac{-iA}{\gamma} e^{i \Omega \tau \bar{n} \cdot \bar{q}} e^{i \Phi_0}e^{i (-\Omega (1 + \bar{n}\cdot \bar{q}) ) \tau} \left(1 -e^{-i \gamma d}\right) \\
-&=\frac{-iA}{\gamma} e^{i \Phi_0}e^{-i \Omega \tau} \left(1 -e^{-i \gamma d}\right) \\
-&=\frac{-iA}{\gamma} e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i \gamma d}\right) 
-\label{eq:concise}
-\end{align}
-
-\subsection{Explicit expression and comparison with Melatos 22}
-We are now very close with making contact with the expression from Melatos 22. Let's be explicit and sub in our definitions for $A$ and $\gamma$ into \ref{eq:concise}
-
-\begin{align}
-p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} &=-i \frac{-\frac{i\Omega \omega}{2} H_{ij}q^i q^j }{-\Omega (1 + \bar{n}\cdot \bar{q}) } e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i (-\Omega (1 + \bar{n}\cdot \bar{q}) ) d}\right) \\
-&=\frac{\omega}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \\
-&=\frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right)
-\label{eq:final}
-\end{align}
-which is exactly the expression derived in Melatos 2022.
-
-\subsection{The $\Phi_0$ issue}
-There was previously some confusion over the presence of a single $\Phi_0$ parameter that was shared between multiple pulsars.
-Does this cause $\tau = 0$ to be a special time when all pulsars are in phase and then go on to evolve to be out of phase?
-This is just due to some sloppy transcription of equations in another document.
-Ignoring measurement noise, the measured photon frequency is related to the emitted photon frequency  as
-\begin{equation}
-	f_{\rm measured} = f_{\rm emitted} \left(1 - \frac{1}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \right)
-\end{equation}
-Now the pulsar frequencies remain in phase and $\tau=0$ is no longer a special time.
-\footnote{This expression is slightly different from the (erroneous) one typed up in the manuscript!
-We have dropped a $1+ \bar{n} \cdot \bar{q}$ factor that was wrongly typed up in the other document}
+\end{enumerate}
 

### 2023-06-15 docs/minutes/slides_01_06/document.tex

--- 2023-05-30 docs/minutes/slides_25_05/document.tex
@@
 
-\date{EE/GW meeting, May 25, 2023}
+\date{EE/GW meeting, June 1, 2023}
 
@@
 
-	$f_{\rm EM}$ is the exact, true solution parametrised by the unknowns $f_0, \dot{f}_0$. This ``heterodyning" is just a change of variables
+	$f_{\rm EM}$ is the exact, true solution parametrised by the unknowns $f_0, \dot{f}_0$. This heterodyning is just a change of variables
 
@@
 \begin{frame}
-	But what about different realisations of a) measurement noise and b) process noise?
+	What about different realisations of:
+	\begin{itemize}
+		\item Measurement noise
+		\item Process noise
+	\end{itemize}
 
@@
 		\tikz[overlay,remember picture]
-		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){As Slide \ref{first_slide}, 5 noise realisations};
+		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){As Slide \ref{first_slide}, 5 noise realisations. (See also Slide \ref{iota_histogram})};
 	}
@@
 \begin{frame}
+	That was for 5 noise realisations - easy to show on corner plot.
+
 	How about 100 noise realisations? 
+\end{frame}
+
+\begin{frame} [label={iota_histogram}]
+	Case examples - $(\iota, h)$ c.f. Slide \ref{first_slide_noise}
+
+	\includegraphics[width=0.48\textwidth,height=0.5\textwidth]{images/single_histogram_iota_gw}
+	\includegraphics[width=0.48\textwidth,height=0.5\textwidth]{images/single_histogram_h}
+
+	(See also `notebooks/2.Multiple\_noise\_realisations.ipynb`)
 \end{frame}
@@
 
-\begin{frame}
-	Case examples - iota, h 
+\begin{frame}{Bayes ratios}
+
+\includegraphics[width=0.85\textwidth,height=0.65\textwidth]{images/BayesRatioPlot}
+
 \end{frame}
@@
 
-\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/BayesRatioPlot}
+	\includegraphics[width=0.85\textwidth,height=0.65\textwidth]{images/BayesRatioPlot_stacked}
 
@@
 
-\end{frame}
-
-\begin{frame}
-	Full NS search for \alert{all parameters} \newline 
-	$\bar{\theta} = [\bar{\theta}_{\rm GW}, \, f_0, \, \dot{f}_0]$ \newline 
-	with $h = [10^{-10}, 10^{-12}]$, $\sigma_m = 10^{-11}$ \newline 
-	Note: pulsar distance is no longer part of model. $\gamma$ doesn't matter
-
-\end{frame}
-
-\begin{frame}
-		\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical_mixed_result_10}
-		\only<1>{%
-			\tikz[overlay,remember picture]
-			\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-10}$,  $\bar{\theta}_{\rm GW}$};
-
-		}
-			\only<1>{%
-		\tikz[overlay,remember picture]
-		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=2.5cm]current page.center){Note some systematics in $\alpha$, $\psi$};
-
-	}
-\end{frame}
-
-\begin{frame}
-	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_result}
-	\only<1>{%
-		\tikz[overlay,remember picture]
-		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$,$\bar{\theta}_{\rm GW}$};
-
-	}
-	\only<1>{%
-		\tikz[overlay,remember picture]
-		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=2.5cm]current page.center){Note $h -\iota$ trade-off};
-
-	}
-
-\end{frame}
-
-\begin{frame}
-\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_f_result}
-\only<1>{%
-	\tikz[overlay,remember picture]
-	\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$, $f_0$, all (47) PSRs};
-
-}
-
-\end{frame}
-
-\begin{frame}
-	\includegraphics[width=0.85\textwidth,height=0.85\textwidth]{images/canonical12_fdot_result}
-	\only<1>{%
-		\tikz[overlay,remember picture]
-		\node[fill=white,text=black] at ([xshift=1.5cm,yshift=3.5cm]current page.center){$h = 10^{-12}$, $\dot{f}_0$, all (47) PSRs};
-
-	}
-\end{frame}
-
-\begin{frame}
-	Next steps:
-	\begin{itemize}
-		\item Bayes factors vs. strain (currently running)
-		\item Process noise $\sigma_p$ (next slides)
-		\item $[h, \iota] \rightarrow [h_{+}, h_{\times}]$  (discussed, easy, helpful?)
-		\item Can we add PSR terms back in? Do we need to?
-		\item Robustness for different $\bar{\theta}_{\rm GW}$ (overkill? canonical example sufficient?)
-	\end{itemize}
 \end{frame}
@@
 	\end{itemize}
-
-\end{frame}
-
-\begin{frame}{}
-
-Insert corner plot here
 
@@
 
-\begin{frame}{Potential solutions to noisy likelihoods}
-
-	\begin{enumerate}
-		\item Better settings for sampler? E.g. more live points (how many?), massively parallel, optimized likelihood evaluations,...
-		\item Maximum likelihood / loss optimisation? 
-		\item Drop PSR terms from model?
-		\item MCMC at higher temperatures (trades off evidence..., less efficient	)
-		\item other...?
-	\end{enumerate}
-
-\end{frame}
-
 %	\begin{frame}{}

### 2023-06-15 docs/manuscript/paper.tex
fix typos in intro

@@
 The foremost technique for the detection of GWs in this nano-Hz regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}.
-The presence  of a nHz gravitational wave will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal.
+The presence  of a nHz GW will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal.
 By measuring the modulation of the pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs.
 Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}.
-These previously disparate efforts have now been joined in international collaboration, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}.
+These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}.
 \newline
 
-The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R}.
-Other potential sources for PTAs include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
-The incoherent superposition of multiple SMBHBs sources leads to a stochastic GW background detectable at nHz frequencies \citep{Sesana10}.
+The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R}, in addition to other potential GW sources such as cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
+The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
 Previous efforts have mainly focused on the detection of this stochastic background via measuring the correlation in the pulsar timing residuals between any two pair of pulsars; the presence of a GW induces a characteristic correlation as a function of the angular separation between the pulsars - the Hellings-Downs curve \citep{Hellings}.
@@
 
-SMBHBs which are sufficiently near and massive may be individually resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
-Moreover, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}.
+SMBHBs which are sufficiently massive and nearby may be individually resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
+The stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}.
 However, as with the stochastic background, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Zhu2014PPTA,Babak2016,Arzoumanian2023}.
@@
 random variations in the pulse arrival time - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}.
-This pulsar timing noise has multiple potential theorized causes including microglitches \citep{Melatos2008}, glitch recovery \citep{Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Antonelli2023} and superfluid turbulence \citep{Melatos2014}.
+This pulsar timing noise has multiple theorized causes including microglitches \citep{Melatos2008}, glitch recovery \citep{Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Antonelli2023} and superfluid turbulence \citep{Melatos2014}.
 In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise.
-However, timing noise in MSPs could be a `latent' phenomenon \citep{Shannon2010}; as we increase the length of observation spans and the measurement timing precision to the levels required to detect gravitational waves, the timing noise in MSPs may emerge and become much more apparent, visible in the same way that timing noise is seen in younger pulsars.
-In addition to timing noise there are also secondary noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+However, timing noise in MSPs could be a `latent' phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}.
+In addition to achromatic timing noise there are also secondary, chromatic, noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
 \newline
 
-In this work we present a novel approach to PTA data analysis for individual monochromatic SMBHB GW sources.
+In this work we present a novel approach to PTA data analysis for individual, monochromatic, SMBHB GW sources.
 This method can self-consistently account for the intrinsic stochastic processes inherent to the pulsar.
-We formulate PTA analysis as a state-space problem and demonstrate how to track the state-space evolution optimally by using a Kalman filter, determine the presence of a GW signal in the pulsar data and infer the underlying source parameters.
+We formulate PTA analysis as a state-space problem and demonstrate how to track the state-space evolution optimally by using a Kalman filter.
+With optimal tracking of the intrinsic pulsar state, it is then possible to determine the presence of a GW signal in the pulsar data and infer the underlying source parameters.
 \newline
@@
 In Section \ref{sec:testing} we test this new method on synthetic pulsar data.
-Throughout this work ee adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature.
+Throughout this work we adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature.
 \newline

### 2023-06-16 docs/manuscript/paper.tex
response to Julian comments 1

@@
 \section{Introduction}
-The detection of high frequency ($\sim 100$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
-GWs from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
-Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
-The foremost technique for the detection of GWs in this nano-Hz regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}.
+The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R},  would result in the emission of nHz gravitational waves (GWs).
+Other potential GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
+The detection of these long-wavelength GWs has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
+The foremost technique for the detection of GWs in this  low-frequency regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}.
 The presence  of a nHz GW will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal.
@@
 Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}.
-These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}.
+These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, as MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}.
 \newline
 
-The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R}, in addition to other potential GW sources such as cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
 The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
-Previous efforts have mainly focused on the detection of this stochastic background via measuring the correlation in the pulsar timing residuals between any two pair of pulsars; the presence of a GW induces a characteristic correlation as a function of the angular separation between the pulsars - the Hellings-Downs curve \citep{Hellings}.
+Previous efforts have mainly focused on the detection of this stochastic background via measuring the correlation in the pulsar timing residuals between any two pair of pulsars; the presence of a GW induces a characteristic correlation as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
 However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}.
-\newline
-
 SMBHBs which are sufficiently massive and nearby may be individually resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
@@
 However, as with the stochastic background, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Zhu2014PPTA,Babak2016,Arzoumanian2023}.
-The intrinsic pulsar timing noise - i.e.
-random variations in the pulse arrival time - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}.
-This pulsar timing noise has multiple theorized causes including microglitches \citep{Melatos2008}, glitch recovery \citep{Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Antonelli2023} and superfluid turbulence \citep{Melatos2014}.
-In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise.
-However, timing noise in MSPs could be a `latent' phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}.
-In addition to achromatic timing noise there are also secondary, chromatic, noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
 \newline
 
-In this work we present a novel approach to PTA data analysis for individual, monochromatic, SMBHB GW sources.
-This method can self-consistently account for the intrinsic stochastic processes inherent to the pulsar.
-We formulate PTA analysis as a state-space problem and demonstrate how to track the state-space evolution optimally by using a Kalman filter.
-With optimal tracking of the intrinsic pulsar state, it is then possible to determine the presence of a GW signal in the pulsar data and infer the underlying source parameters.
+The intrinsic pulsar timing noise - i.e.
+random, unmodelled variations in the pulse arrival time due to irregularities in the rotation of the star - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}.
+This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J}, magnetospheric state switching \citep{magneto1,Lyne2010L} and superfluid turbulence \citep{Link10.111,Melatos2014}.
+In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise.
+However, timing noise in MSPs could be a ``latent" phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}.
+In addition to the temporally uncorrelated ``red" timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+\newline
+
+In this work we present a novel approach to PTA data analysis for individual, monochromatic, SMBHB GW sources that can self-consistently account for the intrinsic timing noise inherent to pulsar observations.
+We formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution by using a Kalman filter
+\citep{Kalman1,Meyers2021,Melatos2023}.
+With the tracking of the intrinsic pulsar state established, it is then possible to determine the presence of a GW signal in the pulsar data and infer the underlying source parameters.
 \newline
@@
 \newline
+
+%
+%The detection of high frequency ($\sim 100$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
+GWs from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
+
+%
+% Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), 
+%
 

### 2023-06-17 docs/manuscript/paper.tex
section 2 typos fix

@@
 \section{Introduction}
-The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R},  would result in the emission of nHz gravitational waves (GWs).
+The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R},  is predicted to result in the emission of nHz gravitational waves (GWs).
 Other potential GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
@@
 
-The intrinsic pulsar timing noise - i.e.
-random, unmodelled variations in the pulse arrival time due to irregularities in the rotation of the star - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}.
+The intrinsic pulsar timing noise -- i.e.
+random, unmodelled variations in the pulse arrival time due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}.
 This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J}, magnetospheric state switching \citep{magneto1,Lyne2010L} and superfluid turbulence \citep{Link10.111,Melatos2014}.
@@
 We will take our measurement variable to be the pulsar pulse frequency that is measured by an observer at Earth, $f_{\rm m}(t)$.
-In Sections \ref{sec:psr_frequency}, \ref{sec:psr_measured} we will describe our model for $f_{\rm p}(t)$ and  $f_{\rm m}(t)$ respectively.
-
-We will go on in Section \ref{sec:statespace} to bring together these two models into a cohesive, discretized state space model. 
-\textcolor{red}{TK: Need some extra discussion here on why this choice of state/measurement space. Why not e.g. phase, TOA, voltage, etc. ?}
-
 \subsection{Evolution of the pulsar frequency} \label{sec:psr_frequency}
@@
 electromagnetic spin-down) and stochastic torques (i.e.
-`spin wandering', achromatic variations in the pulse TOA intrinsic to the star).
+`spin wandering' or timing noise, achromatic variations in the pulse time of arrival intrinsic to the star).
 The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process.
@@
 \begin{equation}
-	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p} - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
+	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p} - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t) \ , 
 	\label{eq:frequency_evolution}
 \end{equation}
-where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
+where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies:
 \begin{equation}
-	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
+	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t') \ ,
 \end{equation}
-for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
+for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown as a simple linear relation:
 \begin{equation}
-	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
+	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t \ .
 \end{equation}  
@@
 As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g.
-a model of the neutron star crust and superfluid, components][]{Meyers2021}.
+a model of the neutron star crust and superfluid components][]{Meyers2021}.
 For our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.
@@
 \begin{equation}
-	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
+	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\hat{n} \cdot \bar{x} - t) + \Phi_0)} \ ,
 \end{equation}
-for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
-We emphasise that for this work $\Omega$ has no time dependence - we are concerned solely with monochromatic sources.
+for Minkowski metric $ \eta_{\mu \nu}$ with spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\hat{n}$-direction and has a phase offset of  $\Phi_0$.
+We emphasise that for this work $\Omega$ has no time dependence -- we are concerned solely with monochromatic sources.
 We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the Earth.
@@
 \begin{align}
-	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
+	H_{ij} = h_+ e_{ij}^+(\hat{n}) + h_{\times} e_{ij}^{\times}(\hat{n}) \ ,
 \end{align}
-where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave which are given by
+where $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave.
+The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes $\hat{k}$, $\hat{l}$ of the wave:
 \begin{align}
-	h_+ &= h(1 + \cos^2 \iota) \\
-		h_{\times} &= -2h\cos \iota
+	e_{i j}^{+}(\hat{n}) =\hat{k}_i \hat{k}_j-\hat{l}_i \hat{l}_j \ , \\
+		e_{i j}^{\times}(\hat{n}) =\hat{k}_i \hat{l}_j+\hat{l}_i \hat{k}_j \ .
 \end{align}
-The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
+The principal axes are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coordinate $\phi$) and the polarisation angle $\psi$
 \begin{align}
-	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
-		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
-\end{align}
-which are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coordinate $\phi$) and the polarisation angle $\psi$
-\begin{align}
-	\vec{m} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
+	\hat{k} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
 	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
 	& +(\sin \psi \sin \theta) \hat{z} \\
-	\vec{n} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
+	\hat{l} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
 	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
@@
 \end{align}
+where e.g. $\hat{x}$ is a unit vector in the direction of the $x$-axis of the reference frame. The direction of GW propagation is related to the principal axes as,
+\begin{equation}
+	\hat{n} = \hat{k} \times \hat{l} \ . 
+\end{equation}
 
 \subsubsection{Pulse frequency as a photon}
-\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$. Generally, the frequency of a photon recorded by an observer with 4-velocity $u^{\mu}$ is 
+In general radio pulses from a pulsar are transmitted as an amplitude-modulated, radio-frequency carrier wave, where the carrier wave (i.e.
+the photon) is at typical radio frequencies of $\sim$ GHz whilst the pulse frequency $f_p(t)$ is $\sim $kHz.
+We are interested in measuring how the pulse frequency is perturbed by the presence of a GW.
+We define the geometric object
 \begin{equation}
-	f = p_{\alpha} u^{\alpha}
+\vec{p} = \omega (1, -\bar{q})
+\end{equation} 
+where $\omega$ is the angular pulse frequency $(= 2 \pi f_p)$ and $\bar{q}$ is the local propagation direction of the radio photon.
+We can identify $\vec{p}$ as the 4-momentum of the radio photon.
+Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $u^{\mu}$ is given by the coordinate-independent expression:
+\begin{equation}
+	f = p_{\alpha} u^{\alpha} \ . 
 	\label{eq:freq_temporal}
 \end{equation}
-\noindent We consider both our emitter and receiver to be stationary, such that  
+We want to relate the pulse frequencies measured in two reference frames; one at the emitter, i.e.
+the pulsar, and one at the observer on Earth.
+For pulsar observations the kinematical Doppler shifts relative to the Solar System barycentre (SSB) are corrected in the generation of a timing solution so that both the pulsar and the Earth are at rest in the global Minkowski coordinates.
+After these kinematical corrections we can consider both our emitter and receiver to be stationary, such that
 \begin{equation}
@@
 \end{equation}
-\noindent Consequently the frequency can be directly identified with the temporal component of the covariant 4-momentum,
+\noindent Consequently the measured frequency in both frames can be directly identified with the temporal component of the covariant 4-momentum,
 \begin{equation}
@@
 \end{equation}
-\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
+\noindent The evolution of the pulse frequency as measured by the observer on Earth is then,
 \begin{equation}
-	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
+	p_t(\tau)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=\tau} \dot{p}_t dt
 	\label{eq:integral}
 \end{equation}
-\noindent where the overdot denotes a derivative w.r.t.
-$t$.
+\noindent where we use $\tau$ to denote the proper time recorded by the Earth observers clock, and the overdot denotes a derivative w.r.t.
+coordinate time along the geodesic $t$.
 Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.
@@
 \newline
+
+wave which are given by
+\begin{align}
+	h_+ &= h(1 + \cos^2 \iota) \\
+	h_{\times} &= -2h\cos \iota
+\end{align}
 

### 2023-07-04 docs/manuscript/paper.tex
implement AM corrections

--- 2023-06-16 docs/manuscript/paper.tex
@@
 % Keep the title short and informative.
-	\title[Kalman PTA]{State-space PTA}
+	\title[Kalman PTA]{Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array}
 
@@
 \begin{keywords}
-editorials, notices -- miscellaneous
+gravitational waves -- methods: data analysis -- pulsars: general
 \end{keywords}
@@
 \newpage
-\section{Introduction}
-The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R},  would result in the emission of nHz gravitational waves (GWs).
-Other potential GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
-The detection of these long-wavelength GWs has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
-The foremost technique for the detection of GWs in this  low-frequency regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}.
-The presence  of a nHz GW will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal.
-By measuring the modulation of the pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs.
-Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}.
-These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, as MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}.
+\section{Introduction}\label{sec:intro}
+The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022} is predicted to emit nHz gravitational waves (GWs).
+Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
+The detection of nHz GWs has necessitated the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
+The foremost method is timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}.
+A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth.
+By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs.
+Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}.
+These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}.
 \newline
 
+\textcolor{red}{TK: Content and references need to be updated in light of recent PTA results}.
 The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
-Previous efforts have mainly focused on the detection of this stochastic background via measuring the correlation in the pulsar timing residuals between any two pair of pulsars; the presence of a GW induces a characteristic correlation as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
+Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
 However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}.
-SMBHBs which are sufficiently massive and nearby may be individually resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
-The stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}.
-However, as with the stochastic background, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Zhu2014PPTA,Babak2016,Arzoumanian2023}.
 \newline
 
-The intrinsic pulsar timing noise - i.e.
-random, unmodelled variations in the pulse arrival time due to irregularities in the rotation of the star - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}.
-This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J}, magnetospheric state switching \citep{magneto1,Lyne2010L} and superfluid turbulence \citep{Link10.111,Melatos2014}.
-In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise.
-However, timing noise in MSPs could be a ``latent" phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}.
-In addition to the temporally uncorrelated ``red" timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
+Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}.
+Individual SMBHBs are continuous wave sources: they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals.
+Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs.
+However, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}.
 \newline
 
-In this work we present a novel approach to PTA data analysis for individual, monochromatic, SMBHB GW sources that can self-consistently account for the intrinsic timing noise inherent to pulsar observations.
-We formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution by using a Kalman filter
-\citep{Kalman1,Meyers2021,Melatos2023}.
-With the tracking of the intrinsic pulsar state established, it is then possible to determine the presence of a GW signal in the pulsar data and infer the underlying source parameters.
+Intrinsic pulsar timing noise -- i.e.
+random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}.
+This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
+In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are relatively stable rotators.
+However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
+In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red).
+In addition to the red timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
 \newline
 
+In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations.
+The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density.
+Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
+The new and existing approach are therefore complementary, but the new approach holds out the promise of somewhat higher sensitivity --- a promise which we aim to test as the key goal of this paper.
+In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
+We combine the Kalman tracking of the pulsars intrinsic rotational state with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood for model selection.\textcolor{red}{TK: a few extra sentences here on how we compare to existing approaches.
+Not quite sure yet how to do this.} \newline
+
 \noindent This paper is organised as follows.
-In Section \ref{sec:model} we present our state-space model for the pulsar pulse frequency which is subject the influence of a GW.
+In Section \ref{sec:model} we present the state-space model for the pulsar pulse frequency which is subject the influence of a GW.
 In Section \ref{sec:detect} we discretise the model, develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence.
-In Section \ref{sec:testing} we test this new method on synthetic pulsar data.
+\textcolor{red}{TK: this paper outline section will be completed once we have finalized the rest of the manuscript} In Section \ref{sec:testing} we test this new method on synthetic pulsar data.
 Throughout this work we adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature.
@@
 
-\section{State-Space Frequency Model}\label{sec:model}
-We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, and the measurement state recorded by an observer.
-For this work we will take our state variable to be the intrinsic pulsar pulse frequency $f_{\rm p}(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar.
-We will take our measurement variable to be the pulsar pulse frequency that is measured by an observer at Earth, $f_{\rm m}(t)$.
-In Sections \ref{sec:psr_frequency}, \ref{sec:psr_measured} we will describe our model for $f_{\rm p}(t)$ and  $f_{\rm m}(t)$ respectively.
+\section{State-Space Formulation}\label{sec:model}
+We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation.
+In this work we take the intrinsic state variable to be the pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local rest frame of the pulsar's centre of mass.
+A model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.
+We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.
+The measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_measured}.
+The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
+\subsection{Spin evolution} \label{sec:psr_frequency}
 
-We will go on in Section \ref{sec:statespace} to bring together these two models into a cohesive, discretized state space model. 
-\textcolor{red}{TK: Need some extra discussion here on why this choice of state/measurement space. Why not e.g. phase, TOA, voltage, etc. ?}
-
-\subsection{Evolution of the pulsar frequency} \label{sec:psr_frequency}
-We will take as our model of the intrinsic pulsar frequency $f_{\rm p}(t)$ a variation of the phenomenological model of \cite{Vargas}.
-Within this model, $f_{\rm p}(t)$ evolves according to a combination of both deterministic torques (i.e.
-electromagnetic spin-down) and stochastic torques (i.e.
-`spin wandering', achromatic variations in the pulse TOA intrinsic to the star).
-The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process.
-Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
+A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}.
+We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend.
+In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque.
+The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$.
+Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution can be approximated accurately by
 \begin{equation}
-	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p} - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
+ f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
+\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA.
+The stochastic torque is assumed to be a zero-mean white noise process.
+Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
+\begin{equation}
+	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)} +\xi^{(n)}(t) \ . 
 	\label{eq:frequency_evolution}
 \end{equation}
-where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
+In Eq \ref{eq:frequency_evolution} $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a proportionality constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
 \begin{equation}
-	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
+	\langle \xi^{(n)}(t) \rangle = 0 \ ,
 \end{equation}
-for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
 \begin{equation}
-	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
+	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle = [\sigma^{(n)}]^2 \delta(t - t') \ .
+	\label{eq:xieqn}
 \end{equation}
-Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
+$[\sigma^(n)]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Eq.
+\ref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
+\newline
+
+Equations \ref{eq:spinevol}--\ref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend [ref latest results paper from each of the three main PTAs].
+Equations \ref{eq:spinevol}--\ref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
+As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G}.
+An approach akin to Equations \ref{eq:spinevol}--\ref{eq:xieqn} has been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
+However, Equations \ref{eq:spinevol}--\ref{eq:xieqn}  involve significant idealizations, which must be recognized at the outset \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
+First, the white noise driver $\xi(t)$ in Equation \ref{eq:frequency_evolution} is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper.
+Second, the white spectrum assumed for $\xi(t)$ may or may not be suitable for millisecond pulsars in PTAs.
+It is challenging observationally to infer the spectrum of $\xi(t)$ from the observed spectrum of the phase residuals, because the inference is conditional on the (unknown) dynamical model governing $df_{\rm p}/dt$.
+For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless.
+Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004} which have not been ruled out by pulsar timing experiments to date.
+The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \ref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \ref{eq:frequency_evolution}.
+
+\subsection{Modulation of pulse TOAs by a GW} \label{sec:psr_measured}
+In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
+Indeed, the pulse TOAs are modulated harmonically at the GW frequency.
+We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section.
+The measurement equation is a key input into the Kalman filter in Section \ref{sec:kalman_filter}
+\subsubsection{Plane GW perturbation}
+We consider a gravitational plane wave from a single,distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
 \begin{equation}
-	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p}- f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
-	\label{eq:frequency_evolution_expanded}
+	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{[i(\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0)]} \ ,
 \end{equation}
-As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g.
-a model of the neutron star crust and superfluid, components][]{Meyers2021}.
-For our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.
+with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$.
+The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$.
+Throughout this paper we will work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB).
+We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB.
+In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic.
+Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves over decadal timescales as \citep[e.g.][]{Zhu10},
+\begin{equation}
+	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left(\frac{f_{\rm gw}(t=0)}{10^{-7} \mathrm{~Hz}}\right)^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right)
+	\label{eq:f_evolution}
+\end{equation}
+where $M_c$ is the chirp mass of the SMBHB, $f_{gw}(t=0)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ the length of the data timespan, which for PTAs is $\sim 10$ years.
+A source can be considered as monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution of $1/T_{\rm obs}$.
+From Eq.
+\ref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans.
+The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA with pulsars at 1.5 kpc, 78 \% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array (SKA2) this fraction drops to 52 \% \citep[Fig 7 of ][]{Rosado10.1093/mnras/stv1098}.
+We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
+\newline
 
-\subsection{Modulation of pulsar frequency due to a GW} \label{sec:psr_measured}
-In the presence of a GW, the pulse frequency measured by an observer in the local neutron star reference frame is different from that measured by an observer on Earth.
-We will now derive the influence of the GW on the received pulse frequency.
+The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$). The spatial part is
+\begin{align}
+	H_{ij} = h_+ e_{ij}^+(\boldsymbol{n}) + h_{\times} e_{ij}^{\times}(\boldsymbol{n}) \ ,
+\end{align}
+where $h_{+}$ and $h_{\times}$ are the respective polarisation amplitudes.
+The plus and cross polarisation tensors $e_{ij}^{+}$ and $e_{ij}^{\times}$ are uniquely defined by the principal axes of the wave, viz.\ the unit 3-vectors $\boldsymbol{k}$ and $\boldsymbol{l}$, according to
+\begin{align}
+	e_{i j}^{+}(\boldsymbol{n}) =k_i k_j-l_i l_j \ , \\
+		e_{i j}^{\times}(\boldsymbol{n}) =k_i l_j+l_i k_j \ .
+\end{align}
+The principal axes are in turn specified by the location of the GW source on the sky (via colatitude $\theta$ and azimuth $\phi$) and the polarisation angle $\psi$ according to
+\begin{align}
+	\boldsymbol{k}  = &(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
+	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}} \nonumber \\
+	& +(\sin \psi \sin \theta) \boldsymbol{\hat{z}} \\
+	\boldsymbol{l} = &(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
+	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}}\nonumber  \\
+	& +(\cos \psi \sin \theta) \boldsymbol{\hat{z}}
+\end{align}
+where e.g. $\boldsymbol{\hat{x}}$ is a unit vector in the direction of the $x$-axis. The direction of GW propagation is related to the principal axes by
+\begin{equation}
+	\boldsymbol{n} = \boldsymbol{k} \times \boldsymbol{l} \ . 
+\end{equation}
 
-\subsubsection{Plane GW perturbation}
-We take a gravitational plane wave that perturbs a background Minkowski spacetime as
+\subsubsection{Measurement equation}
+\textcolor{red}{TK: Does this subsection needs a better name?} \newline 
+\noindent \textcolor{red}{TK: this subsection has been overhauled per comments from AM.
+The structure is now: \newline 1.
+Radio pulses described by general geometric object $p$ \newline 2.
+Shift in $p$ due to a GW (no derivation, just definition of terms) \newline 3.Relate shift in $p$ to a shift in $f$ by clarifying that $u=(1,0,0,0)$ for PSR and observer.
+\newline 4.
+Give final measurement equation} \newline
+
+In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
+They are described by the geometric object $\vec{p}$ which we can identify  as the 4-momentum of the radio photon.
+The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e.
+$\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
 \begin{equation}
-	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
+ \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
+	\label{eq:momentum_shift}
 \end{equation}
-for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
-We emphasise that for this work $\Omega$ has no time dependence - we are concerned solely with monochromatic sources.
-We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the Earth.
-The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
+where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame (MCRF) of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
+We take the pulsar location $\boldsymbol{q}$ to be constant i.e.
+$\boldsymbol{q}$ is not a function of time.
+In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB.
+This barycentering correction is typically applied when generating TOAs using e.g.
+{\sc tempo2} and related timing software.
+Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\vec{u}$ is given by the coordinate-independent expression:
+\begin{equation}
+	f = p_{\alpha} u^{\alpha} \ . 
+	\label{eq:freq_temporal}
+\end{equation}
+Due to the kinematical corrections from the barycentering process, 
+\begin{equation}
+	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
+\end{equation}
+where the perturbations to $\vec{u}$ from the GW are of higher order and neglected.
+In this case the shift in the momentum from Equation \ref{eq:momentum_shift} can be related to a shift in the frequency as,
+\begin{equation}
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t)
+	\label{eq:measurement}
+\end{equation}
+where \textcolor{red}{TK: what is the best notation here to give $q^i$ an $(n)$ superscript?}
+\begin{equation}
+	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right)
+	\label{eq:g_func}
+\end{equation}
+It is will also prove instructive to express Eq. \ref{eq:g_func} in a trigonometric form as,
+or in a trigonometric form 
 \begin{align}
-	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
+	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
+	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right) \right]
+	\label{eq:g_func_trig}
 \end{align}
-where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave which are given by
+Equations \ref{eq:measurement}, \ref{eq:g_func_trig}, \ref{eq:g_func}  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.
+
+\section{Signal tracking and parameter estimation} \label{sec:detect}
+The set of static parameters $\bar{\theta}$ of the model in Section \ref{sec:model} cane be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e.
+\begin{equation}
+	\bar{\theta} =  \bar{\theta}_{\rm psr} \cup \bar{\theta}_{\rm gw}
+\end{equation}
+with
+\begin{equation}
+	\bar{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N}
+\end{equation}
+and
+\begin{equation}
+	\bar{\theta}_{\rm gw} = \left \{h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ . 
+\end{equation}
+We can reparameterize the two GW polarisation strains in terms of the GW amplitude $h_0$ and the system inclination $\iota$
 \begin{align}
@@
 \end{align}
-The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
-\begin{align}
-	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
-		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
-\end{align}
-which are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coordinate $\phi$) and the polarisation angle $\psi$
-\begin{align}
-	\vec{m} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
-	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
-	& +(\sin \psi \sin \theta) \hat{z} \\
-	\vec{n} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
-	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
-	& +(\cos \psi \sin \theta) \hat{z}
-\end{align}
-
-\subsubsection{Pulse frequency as a photon}
-\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$. Generally, the frequency of a photon recorded by an observer with 4-velocity $u^{\mu}$ is 
-\begin{equation}
-	f = p_{\alpha} u^{\alpha}
-	\label{eq:freq_temporal}
-\end{equation}
-\noindent We consider both our emitter and receiver to be stationary, such that  
-\begin{equation}
-	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
-\end{equation}
-\noindent Consequently the frequency can be directly identified with the temporal component of the covariant 4-momentum,
-\begin{equation}
-	f = p_t
-\end{equation}
-\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
-\begin{equation}
-	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
-	\label{eq:integral}
-\end{equation}
-\noindent where the overdot denotes a derivative w.r.t.
-$t$.
-Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.
-
-\subsubsection{Hamiltonian Mechanics}
-The Hamiltonian in covariant notation can be written as 
-\begin{equation}
-	H = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
-\end{equation}
-\noindent which if we substitute in our expression for the perturbed metric is
-\begin{equation}
-	H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
-\end{equation}
-\noindent Now, Hamilton's equations are
-\begin{equation}
-	\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
-\end{equation}
-
-\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
-\begin{equation}
-	\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)}
-\end{equation}
-\noindent The derivative w.r.t coordinate time $t$ is
-\begin{equation}
-	\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
-\end{equation}
-\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic. 
-It will prove useful to recognise that 
-\begin{equation}
-	p^{\mu} = \omega(1,-q^x,-q^y,-q^z)
-\end{equation} 
-where $\bar{q}$ is the unit vector between the Earth and pulsar and $\omega$ is the \textit{constant} photon angular frequency.
-Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer.
-Similarly, we can parameterize the photon spatial coordinates $\bar{x}$ as,
-\begin{equation}
-	\bar{x}(t) = - \bar{q} (t - t_1)
-\end{equation} 
-Note that when our integration variable $t = t_1$ (i.e.
-the photon is at Earth) then $\bar{x}(t=t_1) = 0$ which is what we expect given our chosen coordinate system.
-Also note that $\dot{p}_t$ is entirely a function of the GW perturbation.
-In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.
-We will treat the pulsar locations to be constant with respect to the Earth, i.e.
-$\bar{q}$ is not a function of time.
-In practice, the pulsar locations vary with respect to the Earth, but are constant with respect to the solar system barycentre (SSB).
-This barycentering correction is typically applied during the pulsar detection process.
-Our conclusions are generally unchanged by this choice - it is straightforward to transform $q$ if needed into a vector between the SSB and pulsar.
+where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$ and the observer line of sight, i.e.
+$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.
+Henceforth we use the parametrisation in terms of $h_0$ and $\iota$.
+For a PTA dataset containing $N$ pulsars we have $7 + 5N$ parameters to estimate.
+Typically the pulsar parameters are better constrained than the GW parameters: for example estimates of pulsar distances are accurate to $\sim$ 10$\%$, but we have no prior information on the source location \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.The variance of the measurement noise $\Sigma$ can be estimated approximately from the root-mean-square amplitude of phase residuals and we do not treat it as a parameter to be inferred in this work.
 \newline
 
-\noindent Bringing this all together we can write $\dot{p}_t$ in a condensed form as,
-\begin{equation}
-	\dot{p}_t = A e^{i B t + C}
-\end{equation}
+In this section we present a new method to infer $\bar{\theta}$ and calculate the marginal likelihood (i.e.
+the model evidence).
+In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency $f_{\rm m}^{(n)}(t)$ can be used to estimate the hidden state sequence $f_{\rm p}^{(n)}(t)$ using a Kalman filter.
+In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampler to perform Bayesian inference of the model parameters and calculate the model evidence.
+Model selection and the specification of the null model is described in Section \ref{sec:model_selection}.
 
-with 
-\begin{equation}
-	A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
-\end{equation}
-\begin{equation}
-	B = -\Omega (1 + \bar{n}\cdot \bar{q}) 
-\end{equation}
-and 
-\begin{equation}
-	C = \Omega \tau \bar{n} \cdot \bar{q}  +\Phi_0
-\end{equation}
-From Eq \ref{eq:integral} the frequency shift experienced by the observer relative to the source due to a GW is then
-\begin{align}
-	&p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = \\
-	 & A \int_{t = \tau - d}^{t=\tau} e^{i B t + C}dt  \\
-	& =\frac{-iA}{B} e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i B d}\right) \\
-	&= \frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right)
-	\label{eq:concise}
-\end{align}
-for $h_{ij} = g_{ij} - \eta_{ij}$. Since via Equation  \ref{eq:freq_temporal} we can related the temporal component of the covariant momentum to the frequency, it follows that
-\begin{align}
-	&f(\tau)|_{\rm Earth} = f(\tau - d)|_{\rm source}  \left( 1- \frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \right)
-	\label{eq:main_eq}
-\end{align}
-where we have identified $\omega$ with $f(\tau - d)|_{\rm source}$.
-
-\section{Detection and parameter estimation} \label{sec:detect}
-Let's review and categorise all the free parameters,$\bar{\theta}$, of above model.
-We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source
-\begin{equation}
-	\bar{\theta} =  \bar{\theta}_{\rm PSR} \cup \bar{\theta}_{\rm GW}
-\end{equation}
-\begin{equation}
-	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]^{(N)}
-\end{equation}
-\begin{equation}
-	\bar{\theta}_{\rm GW} = [h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0]
-\end{equation}
-Whilst the GW parameters are shared between measurements for each pulsar, the pulsar parameters are clearly not.
-For a PTA dataset of $N$ pulsars we have $7 + 5N$ parameters to estimate.
-Note that whilst this is a large parameter space, in general the pulsar parameters are much better constrained than the GW parameters: for example we have rough estimates for the pulsar distances accurate to $\sim$ 10$\%$, but we have no prior information on the source location.
-\newline
-
-In this section we present a new method to infer the model parameters and calculate the marginal likelihood (i.e.
-the model evidence).
-In Section \ref{sec:statespace} we will present a discretised version of the model of Section \ref{sec:model} which maps onto the discretely sampled observable $f(\tau)|_{\rm Earth}$.
-In Section \ref{sec:kalman_filter} we will outline how observations of this pulse frequency can be used to recover the underlying intrinsic pulsar state $f(\tau -d)|_{\rm source}$ using a Kalman Filter.
-In Section \ref{sec:nested_sampling} we will demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to recover the model parameters.
-We will now review the Kalman filter and the associated likelihood in Section \ref{sec:kalman_filter} before going on in Section \ref{sec:nested_sampling} to discuss how to use nested sampling methods together with the filter for detection (model selection) and parameter estimation.
-
-\subsection{Discretised model} \label{sec:statespace}
-We can express the intrinsic frequency evolution, Eq. \ref{eq:frequency_evolution_expanded}, in a alternative form as,
-\begin{equation}
-	df = \mathcal{A} f dt + N(t) dt + \sigma dB(t)
-	\label{eq:state1}
-\end{equation}
-where $\mathcal{A} = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process).
-This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by \citep{gardiner2009stochastic},
-\begin{equation}
-	f(t) = e^{\mathcal{A}t}f(0) + \int_0^t e^{\mathcal{A}(t-t')} N(t') dt' + \int_0^t e^{\mathcal{A}(t-t')} \sigma dB(t')
-\end{equation} 
-If we move from a solution in continuous time, $t$, to discrete time, $\bar{t} = (t_1, t_2, ...,t_K)$, then
-\begin{equation}
-	f(t_{i+1}) = F f(t_i) + T_i + \eta_i
-\end{equation}
-where
-\begin{equation}
-	F_i = e^{\mathcal{A} (t_{i+1} - t_i)}
-\end{equation}
-\begin{equation}
-	T_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} N(t') dt'
-\end{equation}
-\begin{equation}
-	\eta_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} \sigma dt'
-\end{equation}
-
-if we specialise to the case of constatn time sampling
-
-The discrete solution $f(\bar{t})$ to the intrinsic frequency can be related to the discrete measured frequency via Eq. \ref{eq:main_eq} as,
-\begin{equation}
-	f_M (\bar{t})= f(\bar{t}) g(\bar{\theta},\bar{t}) + N_M
-\end{equation}
-where $g(\theta,t)$ can be expressed in a trigonometric form as 
-\begin{equation}
-		X = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
-\end{equation} 
-whilst $N_M$ is a Gaussian measurement noise that satisfies 
-\begin{equation}
-	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
-\end{equation}
-for variance $\Sigma^2$.
-
-\subsection{Kalman Filtering}\label{sec:kalman_filter}
+\subsection{Kalman filter}\label{sec:kalman_filter}
 \begin{figure*}
@@
 
-The Kalman filter \citep{Kalman1} is a algorithmic technique for recovering a set of system state variables, $\bar{x}$, given some noisy measurements, $\bar{z}$.
-It is a common technique in signal processing that has also been applied more recently with great success in astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}.
-The linear Kalman filter operates on measurements that are related to states via a linear transformation
-\begin{equation}
-	\bar{z} = \bar{H} \bar{x} + \bar{v}
-\end{equation}
-where $\bar{H}$ is the measurement matrix and $\bar{v}$ a Gaussian measurement noise. The underlying states are the solutions to the state-space equation 
-\begin{equation}
-	\dot{\bar{x}} = \bar{F} \bar{x} + \bar{G}\bar{u} + \bar{w}
-\end{equation}
-for the system dynamics matrix,$\bar{F}$,  control model $G$, control vector $\bar{u}$, and $w$ a stochastic zero-mean process.
-By comparison with the preceding equations, Eqs.
-	\ref{eq:state1} - 	\ref{eq:state2}, it is immediately obvious how our state space model maps onto the Kalman filter structure.
-Specifically, our states are just the $N$ intrinsic pulsar frequencies $\bar{x} = (f_1,f_2,...f_N)$ whilst our measurements are the $N$ measured pulse frequencies $\bar{z} = (f^{(M)}_1,f^{(M)}_2,...f^{(M)}_N)$.
-If we specialize to the case of constant time sampling between our observations, $\Delta t$, then for our formulation the components that make up the Kalman filter are as follows:
-\begin{equation}
-	F_i = F_{i+1} = e^{-\gamma \Delta t}
-	\label{eq:fmatrix}
-\end{equation}
-\begin{align}
-	T_i &= \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt' \\
-	    & = f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)  (\Delta t + t_i) - e^{-\gamma \Delta t} (f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)t_i)
-\end{align}
-\begin{equation}
-	H_i = 1 -A(\theta_{\rm GW}) \cos(-\Omega t_i (1 + n\cdot q ) + \Phi_0)
-\end{equation}
-where the $i$ subscript labels the value at the $i$-th timestep, and $A$ is a constant that is given by  Eq. \ref{eq:state2}. \newline
-
-\noindent The Kalman filter includes the effect of process noise $\bar{w}$ and the measurement noise $\bar{v}$ via the definition of a process noise matrix $Q = E[w w^T]$ and a measurement noise matrix $R = E[v v^T]$, which have the discrete form,
-\begin{equation}
-	Q_i  \delta_{ij}= \langle \eta_i \eta_j^T \rangle = \frac{- \sigma^2}{2 \gamma} \left( e^{-2 \gamma \Delta t} -1\right)
-\end{equation}
-\begin{equation}
-	R_i = R_{i+1} = \Sigma^2
-	\label{eq:rmatix}
-\end{equation}
-The above equations, Eqs.
-\ref{eq:fmatrix} - 	\ref{eq:rmatix}, apply to an operation on a single state.
-The extension to $N$ states is straightforward, since one just needs to construct a diagonal matrix for each of the Kalman components where each non-zero element corresponds to a separate pulsar frequency.
-For our formulation we are concerned only with the linear Kalman filter since the measurements are a linear function of the states and the state transitions are also linear.
+The Kalman filter \citep{Kalman1} is an algorithm for recovering the most likely evolution of a set of system state variables, $\boldsymbol{X}$ given some noisy measurements $\boldsymbol{Y}$.
+It is a common technique in signal processing that has been applied successfully in neutron star astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}.
+In this work we use the linear Kalman filter which assumes a linear relation between $\boldsymbol{X}$ and $\boldsymbol{Y}$.
+Whilst the measurement equation, Eq.
+\ref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively.
 Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, the unscented Kalman filter \citep{882463van} or the particle filter \citep{Simon10}.
-For a full review of the  Kalman filter equations in an astrophysical context, which may be unfamiliar to some, we refer the reader to the appendices of \cite{Melatos2023} and  \cite{Meyers2021}.
 \newline
 
-\noindent In order for the Kalman filter to function successfully, the parameters of the model, which appear in the various Kalman matrices, must be accurate.
-Erroneous or inaccurate parameters lead to inaccurate predictions of the underlying states (e.g.
-Fig \ref{fig:four figures}).
-The filter tracks the error in its predictions of the underlying states by projecting the state predictions back into measurement space, $\hat{x} \rightarrow \hat{z}$.
-These measurement predictions can then be compared against the true observed measurements i.e.
-$ \epsilon = \bar{z} - \hat{z}$.
-The Gaussian log-likelihood at timestep $i$, as a function of the parameters is,
+The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}.
+At each discrete timestep $i = 1, ...
+, M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates $\boldsymbol{P}_i$.
+The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation.
+The measurement predictions can then be compared aganst the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation".
+The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$.
+The Gaussian log-likelihood can then be calculated at each timestep,
 \begin{eqnarray}
-	\log \mathcal{L}_i =  -\frac{1}{2} (N \log 2 \pi + \log \det |S_i| + \epsilon_i^T S_i^{-1} \epsilon_i)
+	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right )
 \end{eqnarray}
-where $S_i$ is the covariance matrix of the error $\epsilon_i$. The net log-likelihood is simply the sum over all timesteps.
+with the total log-likelihood the sum over all timesteps, i.e. 
+\begin{eqnarray}
+	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i  \label{eq:likelihood}
+\end{eqnarray}
+The likelihood $\mathcal{L}$, along $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of the static parameters of the model, $\bar{\theta}$.
+If the estimates of the system parameters, $\hat{\theta}$, that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised.
+This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the hidden state evolution with high fidelity.
+Conversely, if $\hat{\theta}$ is not close to $\bar{\theta}$ then the filter is unable to recover the state evolution.
+This is demonstrated in Figure \ref{fig:25MB_bfs} where the failure of the filter to track the hidden state is evident.
 
 \subsection{Nested Sampling}\label{sec:nested_sampling}
-Nested Sampling \citep{Skilling} is an integration algorithm used for evaluating integrals of the Bayesian evidence form,
-\begin{eqnarray}
-	Z = \int \mathcal{L} \pi (\theta) d \theta
-\end{eqnarray}
-where $\pi (\theta)$ is the prior which is a function of a set of unknown parameters $\theta$.
+
+We can use the likelihood returned by the Kalman filter, Eq \ref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\bar{\theta}$ by Bayes' Rule,
+\begin{equation}
+	p(\bar{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L} \cdot \pi(\theta)}{\mathcal{Z}}
+\end{equation}
+where $\pi(\theta)$ is the prior distribution on $\bar{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
+\begin{equation}
+	\mathcal{Z} = \int \mathcal{L} \pi (\theta) d \theta \label{eq:model_evidence}
+\end{equation}
+In order to estimate the posterior distribution and the model evidence we use nested sampling \citep[NS,][]{Skilling}.
+Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Eq.
+\ref{eq:model_evidence}, that also returns samples from the posterior $p(\bar{\theta} | \boldsymbol{Y})$.
 The primary advantage of nested sampling is the ability to compute this evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches.
@@
 For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
-Within this work we also use nested sampling for parameter estimation and model selection.
+Within this work we exclusively use nested sampling for Bayesian parameter estimation and model selection.
 \newline
@@
 For gravitational astrophysics it is common to use the dynesty sampler CITE, via the Bilby gravitational wave inference library and we continue to follow this precedent.
-\textcolor{red}{TK: we may end up not even using Bilby/dynesty so leaving this section unfilled for now. Do we need a very short description of how nested sampling works?}
+\textcolor{red}{TK: update text here from other version}
+
+\subsection{Model selection}\label{sec:model_selection}
+The evidence integral returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}$.
+This enables us to compare competing models via a Bayes factor,
+\begin{equation}
+	\beta = \frac{\mathcal{Z}(\mathcal{M}_1)}{\mathcal{Z}(\mathcal{M}_0)}
+\end{equation}
+where the subscript index identifies a particular model.
+Throughout this work we take $\mathcal{M}_1$ to be the complete model defined in Section \ref{sec:model}.
+$\mathcal{M}_0$ is our null model that assumes there is no GW in the data.
+This is equivalent to setting $g^{(n)}(t)$ from Equation \ref{eq:g_func}, \ref{eq:g_func_trig} $= 1$.
+The Bayes factors we quote in this paper therefore quantify whether the data shows evidence for a GW compared to no GW.
+
+\section{Tests with synthetic data} \label{sec:testing}
+
+We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.
 
@@
 For now we will consider the measurement noise to be known, although in principle this too could be estimated.
-
-\section{Tests with synthetic data} \label{sec:testing}
-
-We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.
 
@@
 
+\appendix
+
+\section{Kalman recursion equations} \label{sec:kalman}
+
+The linear Kalman filter operates on temporally discrete measurements which are related to unobservable system states via a linear transformation
+\begin{equation}
+	\bar{z} = \bar{H} \bar{x} + \bar{v}
+\end{equation}
+where $\bar{H}$ is the measurement matrix and $\bar{v}$ is a Gaussian measurement noise. The underlying states evolve according to are the state-space dynamical equation 
+\begin{equation}
+	\dot{\bar{x}} = \bar{F} \bar{x} + \bar{G}\bar{u} + \bar{w}
+\end{equation}
+for the system dynamics matrix,$\bar{F}$,  control model $G$, control vector $\bar{u}$, and $w$ a stochastic zero-mean process.
+By comparison with the preceding equations, Eqs.
+	\ref{eq:state1} - 	\ref{eq:state2}, it is immediately obvious how our state space model maps onto the Kalman filter structure.
+Specifically, our states are just the $N$ intrinsic pulsar frequencies $\bar{x} = (f_1,f_2,...f_N)$ whilst our measurements are the $N$ measured pulse frequencies $\bar{z} = (f^{(M)}_1,f^{(M)}_2,...f^{(M)}_N)$.
+If we specialize to the case of constant time sampling between our observations, $\Delta t$, then for our formulation the components that make up the Kalman filter are as follows:
+\begin{equation}
+	F_i = F_{i+1} = e^{-\gamma \Delta t}
+	\label{eq:fmatrix}
+\end{equation}
+\begin{align}
+	T_i &= \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt' \\
+	& = f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)  (\Delta t + t_i) - e^{-\gamma \Delta t} (f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)t_i)
+\end{align}
+\begin{equation}
+	H_i = 1 -A(\theta_{\rm GW}) \cos(-\Omega t_i (1 + n\cdot q ) + \Phi_0)
+\end{equation}
+where the $i$ subscript labels the value at the $i$-th timestep, and $A$ is a constant that is given by  Eq. \ref{eq:state2}. \newline
+
+\noindent The Kalman filter includes the effect of process noise $\bar{w}$ and the measurement noise $\bar{v}$ via the definition of a process noise matrix $Q = E[w w^T]$ and a measurement noise matrix $R = E[v v^T]$, which have the discrete form,
+\begin{equation}
+	Q_i  \delta_{ij}= \langle \eta_i \eta_j^T \rangle = \frac{- \sigma^2}{2 \gamma} \left( e^{-2 \gamma \Delta t} -1\right)
+\end{equation}
+\begin{equation}
+	R_i = R_{i+1} = \Sigma^2
+	\label{eq:rmatix}
+\end{equation}
+The above equations, Eqs.
+\ref{eq:fmatrix} - 	\ref{eq:rmatix}, apply to an operation on a single state.
+The extension to $N$ states is straightforward, since one just needs to construct a diagonal matrix for each of the Kalman components where each non-zero element corresponds to a separate pulsar frequency.
+
+In Section \ref{sec:statespace} we present a discretised version of the model of Section \ref{sec:model} which maps onto the discretely sampled observable $f(\tau)|_{\rm Earth}$.
+
+In order to make contact with discretely sampled data it is importnat to temporally discretise the model of Section \ref{sec:model}
+
+We can express the intrinsic frequency evolution, Eq. \ref{eq:frequency_evolution_expanded}, in a alternative form as,
+\textcolor{red}{TK all this needs updating}
+\begin{equation}
+	df = \gamma  f dt + N(t) dt + \sigma dB(t)
+	\label{eq:state1}
+\end{equation}
+where $\mathcal{A} = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process).
+This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by \citep{gardiner2009stochastic},
+\begin{equation}
+	f(t) = e^{\mathcal{A}t}f(0) + \int_0^t e^{\mathcal{A}(t-t')} N(t') dt' + \int_0^t e^{\mathcal{A}(t-t')} \sigma dB(t')
+\end{equation} 
+If we move from a solution in continuous time, $t$, to discrete time, \textcolor{red}{TK: AM reccomends deligint tabe sy,bol} $\bar{t} = (t_1, t_2, ...,t_K)$, then
+\begin{equation}
+	f(t_{i+1}) = F f(t_i) + T_i + \eta_i
+\end{equation}
+where
+\begin{equation}
+	F_i = e^{\mathcal{A} (t_{i+1} - t_i)}
+\end{equation}
+\begin{equation}
+	T_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} N(t') \, dt'
+\end{equation}
+\begin{equation}
+	\eta_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} \sigma \, dt'
+\end{equation}
+
+if we specialise to the case of constatn time sampling
+
+The discrete solution $f(\bar{t})$ to the intrinsic frequency can be related to the discrete measured frequency via Eq. \ref{eq:main_eq} as,
+\begin{equation}
+	f_M (\bar{t})= f(\bar{t}) g(\bar{\theta},\bar{t}) + N_M
+\end{equation}
+where $g(\theta,t)$ can be expressed in a trigonometric form as 
+\begin{equation}
+	X = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
+\end{equation} 
+whilst $N_M$ is a Gaussian measurement noise that satisfies 
+\begin{equation}
+	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
+\end{equation}
+for variance $\Sigma^2$.
+
 \subsection{References}

### 2023-07-04 docs/minutes/slides_28_06 /document.tex
implement AM corrections

% !TeX TS-program = xelatex

\documentclass[10pt]{beamer}

\usetheme[progressbar=frametitle]{metropolis}
\usepackage{appendixnumberbeamer}
\usepackage{pgfplots}
\usepackage{xspace}
\usepackage[normalem]{ulem}

\definecolor{links}{HTML}{2A1B81}

\hypersetup{colorlinks,linkcolor=,urlcolor=links}

\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
%\setsansfont[BoldFont={Fira Sans}]{Fira Sans Light}
%\setmonofont{Fira Mono}
%\usepackage[sfdefault]{Fira Sans}

\title{}
\subtitle{}

\author{}

\date{EE/GW meeting, June 29, 2023}

\begin{document}

	\maketitle

	\begin{frame}{}

		\begin{enumerate}
			\item IPTA conference
			\item Manuscript updates
			\item Identifiability analysis and frequency-space formulation from EE
		\end{enumerate}

\end{frame}

	\begin{frame}{IPTA conference}

	\begin{itemize}
		\item IPTA conference last week
		\item \sout{ Many results currently under embargo}
		\item \href{https://iopscience.iop.org/article/10.3847/2041-8213/acdac6}{The NANOGrav 15 yr Data Set: Evidence for a Gravitational-wave Background} 
		\item More papers to come c.f. single source. TK to sync with EE to discuss results and methods
	\end{itemize}

\end{frame}

		\begin{frame}{Manuscript updates}

		\begin{itemize}
			\item Plan for papers is as follows:
			\begin{enumerate}
				\item Single source. Earth terms + nested sampling (Bilby) 
				\item Single source. Pulsar terms, Earth terms + other likelihood inference methods (e.g. Bilby vs other nested sampling libs, MCMC, EM)
				\item Multiple sources (i.e. stochastic background, Hellings Downs)
			\end{enumerate}
			\item P1 manuscript currently being written up. Can be found at \href{https://github.com/tomkimpson/StateSpacePTA.jl/blob/main/docs/manuscript/paper.pdf}{github/StateSpacePTA.jl} 
			\item Majority of work for P1 is done - just a question of writing up
			\item Exception: comparison with existing approaches (e.g. \href{https://ui.adsabs.harvard.edu/abs/2014ApJ...794..141A/abstract}{NANOGrav 6 year}, \href{https://ui.adsabs.harvard.edu/abs/2019ApJ...880..116A/abstract}{NANOGrav 11 year}, \href{https://ui.adsabs.harvard.edu/abs/2023arXiv230103608A/abstract}{NANOGrav 12.5}, NANOGrav15). \alert{To discuss (next slide)}
			\item Completing P1 manuscript will be main focus over next few weeks, then back to a "research focus" re P2/P3
		\end{itemize}

	\end{frame}

			\begin{frame}{How to compare against existing methods?}

		Todo: comparison with existing approaches (e.g. \href{https://ui.adsabs.harvard.edu/abs/2014ApJ...794..141A/abstract}{NANOGrav 6 year}, \href{https://ui.adsabs.harvard.edu/abs/2019ApJ...880..116A/abstract}{NANOGrav 11 year}, \href{https://ui.adsabs.harvard.edu/abs/2023arXiv230103608A/abstract}{NANOGrav 12.5}, NANOGrav15).

		Existing methods:
		\begin{enumerate}
			\item Take some timing residuals ``observations" $\delta t$, (i.e. TOAs  - expected TOAs from timing model $M \epsilon$)
			\item Define a GW model for timing residuals $s(\theta_{\rm gw})$. 
			\item Define a noise model $n(\theta_{\rm n})$
			\item Define a likelihood $\mathcal{L}(\delta t | \epsilon,\theta_{\rm gw},\theta_{\rm n} )$
			\item Two approaches
			\begin{enumerate}
				\item Frequentist. Define an F-statistic from $\mathcal{L}$ (now disfavoured?)
				\item Bayesian. MCMC over parameters
			\end{enumerate}

		\end{enumerate}

 How to compare? Do we need to compare for P1?

	\end{frame}

			\begin{frame}{Identifiability and frequency-space}

		\begin{itemize}
			\item Rob + Bill presented frequency-space formulation of problem 
			\item Interesting/useful for identifiability analysis + potentially as an alternative inference/detection method
		\end{itemize}

	\end{frame}

	\appendix

%	
%	
%	\begin{frame}
%		
%		
%		
%		\begin{itemize}
%			\item CURN vs HD \url{https://arxiv.org/abs/2110.13184}
%			\item Model misspecification? \url{https://iopscience.iop.org/article/10.3847/2041-8213/ac17f4}
%		\end{itemize}
%	\end{frame}
%

\end{document}

### 2023-07-31 docs/manuscript/paper.tex
misc text updates

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{booktabs}
+\usepackage{multirow}
 
@@
 \end{equation}
-In Eq \ref{eq:frequency_evolution} $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a proportionality constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
+In Eq \ref{eq:frequency_evolution}, $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a proportionality constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
 \begin{equation}
@@
 where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame (MCRF) of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
-We take the pulsar location $\boldsymbol{q}$ to be constant i.e.
+We take the pulsar location to be constant i.e.
 $\boldsymbol{q}$ is not a function of time.
@@
 \section{Signal tracking and parameter estimation} \label{sec:detect}
-The set of static parameters $\bar{\theta}$ of the model in Section \ref{sec:model} cane be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e.
+The set of static parameters $\boldsymbol{\theta}$ of the model in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e.
 \begin{equation}
-	\bar{\theta} =  \bar{\theta}_{\rm psr} \cup \bar{\theta}_{\rm gw}
+	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \label{eq:params1}
 \end{equation}
@@
 \begin{equation}
-	\bar{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N}
+	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N}
 \end{equation}
@@
 \begin{equation}
-	\bar{\theta}_{\rm gw} = \left \{h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ . 
+	\boldsymbol{\theta}_{\rm gw} = \left \{h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
 \end{equation}
@@
 
-In this section we present a new method to infer $\bar{\theta}$ and calculate the marginal likelihood (i.e.
+In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e.
 the model evidence).
@@
 
+%The choice of prior distributions on the model parameters are described in
+
 \subsection{Kalman filter}\label{sec:kalman_filter}
@@
 	%\centering % Not needed
-	\begin{subfigure}[b]{1\columnwidth}
-		\includegraphics[width=\textwidth]{images/KF1}
-		\caption{Using correct parameters}
+	\begin{subfigure}[b]{1\textwidth}
+		\includegraphics[width=\textwidth]{images/Kalman_example_true_params}
+		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
 		\label{fig:6MB_BFS}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{1\columnwidth}
-		\includegraphics[width=\textwidth]{images/KF2}
-		\caption{Using incorrect parameters}
+	\end{subfigure} \newline
+
+	\begin{subfigure}[b]{1\textwidth}
+		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
+		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
 		\label{fig:25MB_bfs}
 	\end{subfigure}
-	\caption{Application of a Kalman filter to recover the intrinsic frequency states (top panels) given a measured frequency signal (bottom panels) which has been modulated by the presence of a gravitational wave.
-In the case where the parameters of the filter are accurate (subfigure (a)) the underlying states are also recovered accurately.
-Conversely, when the parameters fed to the filter are inaccurate (subfigure (b)) the filter cannot recover the state.
-\textcolor{red}{TK: this needs to be just one stacked figure.}}
+	\caption{Application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$.
+We here plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency.
+In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$.
+In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$.
+The top left panel shows the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line).
+The middle left panel shows the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta).
+The bottom left panel shows the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$.
+The right hand panel plots the distribution of the innovation over the total observation period.}
 	\label{fig:four figures}
@@
 , M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates $\boldsymbol{P}_i$.
-The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation.
+The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Eq.
+\ref{eq:measurement}d.
 The measurement predictions can then be compared aganst the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation".
@@
 If the estimates of the system parameters, $\hat{\theta}$, that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised.
-This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the hidden state evolution with high fidelity.
+This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the hidden state evolution with high fidelity - leavign only gaussian errrs
+
 Conversely, if $\hat{\theta}$ is not close to $\bar{\theta}$ then the filter is unable to recover the state evolution.
@@
 \end{equation}
-In order to estimate the posterior distribution and the model evidence we use nested sampling \citep[NS,][]{Skilling}.
+In order to estimate the posterior distribution and the model evidence we use nested sampling \citep[NS,][]{Skilling} throughout this work.
 Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Eq.
@@
 For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
-Within this work we exclusively use nested sampling for Bayesian parameter estimation and model selection.
-\newline
+For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}.
+Multiple nested sampling algorithms and computational libraries exist.
+\citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}.
+For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \cite{bilby.507.2037A} front-end library.
+We  follow this precedent and use \texttt{Bilby} for all Bayesian inference in this work.
 
-\noindent Multiple nested sampling libraries exist.
-For gravitational astrophysics it is common to use the dynesty sampler CITE, via the Bilby gravitational wave inference library and we continue to follow this precedent.
-\textcolor{red}{TK: update text here from other version}
+%from https://arxiv.org/pdf/2102.12478.pdf
+%Nested sampling performs these dual processes by evolving an
+%ensemble of 𝑛live live points by initially drawing them from the prior
+%distribution 𝜋, and then at each subsequent iteration: (a) discarding
+%the lowest likelihood live point and (b) replacing it by a live point
+%drawn from the prior 𝜋 subject to hard constraint that the new point is
+%at a higher likelihood than the discarded point. The set of discarded
+%dead points and likelihoods can be used to compute a statistical
+%estimate of the Bayesian evidence (1), and be weighted to form a set
+%of samples from the posterior distribution.
+%The nested sampling procedure therefore differs substantially from
+%traditional strategies such as Gibbs sampling, Metropolis-Hastings
+%or Hamiltonian Monte Carlo in that it is acts as a likelihood scanner
+%rather than a posterior sampler. It scans athermally, so in fact is
+%capable of producing samples at any temperature and computing the
+%full partition function in addition to the Bayesian evidence (Handley
+%2019b).
 
@@
 $\mathcal{M}_0$ is our null model that assumes there is no GW in the data.
-This is equivalent to setting $g^{(n)}(t)$ from Equation \ref{eq:g_func}, \ref{eq:g_func_trig} $= 1$.
+This is equivalent to setting $g^{(n)}(t)=1$ in Equation \ref{eq:g_func}, \ref{eq:g_func_trig}.
 The Bayes factors we quote in this paper therefore quantify whether the data shows evidence for a GW compared to no GW.
@@
 \section{Tests with synthetic data} \label{sec:testing}
-
 We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.
 
-\subsection{Practical considerations}
+\subsection{Synthetic data generation}
+
+\subsubsection{Constructing a synthetic PTA}\label{sec:synt_pta}
+
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
+	\caption{Spatial distribution of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
+The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc.
+The grey dashed line denotes the Galactic plane.}
+	\label{fig:pulsar_distrib}
+\end{figure}
+
+For this work we generate representative synthetic data as follows.
+We first specify a set of pulsars to make up our artificial PTA.
+We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}.
+For each pulsar we get the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$.
+The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Fig \ref{fig:pulsar_distrib}.
+The data is obtained via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$.
+We disused in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$.
+Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ \cite{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars.
+For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
+In order to set a physically reasonable value for $[\sigma^{(n)}]^2$ we use two complementary approaches.
+The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by
+\begin{eqnarray}
+	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right)+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \label{eq:sigmap}
+\end{eqnarray}
+where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$.
+Throughout this work we assume that all pulsars are observed with a weekly cadence i.e.
+$T = 1$ week.
+The frequency noise parameter can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as
+\begin{eqnarray}
+	\sigma^{(n)} = \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
+\end{eqnarray}
+For our synthetic NANOGrav PTA, the median $\sigma^{(n)} = 4.28 \times 10^{-21} $ s.
+Alternatively to this empirical best-fit model, we can also directly solve the state equation Eq.
+\ref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a phase solution
+\begin{eqnarray}
+	\phi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') 
+\end{eqnarray}
+We can then vary $\sigma^{(n)}$ and compare the synthetically generate phase residuals with the phase solution from the UTMOST pulsar timing program with the Mongolo Observatory Synthesis Telescope  \citep{UTMOST}.
+It is sufficient to simply compare the synthetic and real solutions visually, rather than e.g.
+defining some quantitative loss function (e.g.
+mean-squared error) between the two solutions and solving the optimisation problem of the value of $\sigma_{\rm TOA}^{(n)}$ which minimised this loss function; we emphasise that we are not overly concerned with getting the maximally accurate values for the noise in the pulsar frequency evolution, but just instead some reasonable, representative values.
+This approach of generating synthetic phase solutions and visually comparing with real solutions is also followed in \cite{Vargas}.
+In Figure \ref{fig:qualitative_compare} we demonstrate for a particular representative pulsar, PSR XXX, blah blah blah...
+\begin{figure}
+	\includegraphics[width=\columnwidth]{example-image-a}
+	\caption{Actual and synthetic phase evolution}
+	\label{fig:qualitative_compare}
+\end{figure}
+
+\subsubsection{Synthetic data generation}
+We generate $N$ synthetic noisy timeseries by integrating the state equations, Eq \ref{eq:spinevol}- \ref{eq:xieqn} numerically, mapping from state space to measurement space via the measurement equation, Eq.
+\ref{eq:measurement}, and finally adding to each timeseries some zero-mean Gaussian noise with covariance $\sigma_{\rm m}^2$.
+We take $\sigma_{\rm m}^2$ to be known \textit{a priori} rather than a parameter to be inferred.
+Throughout this work we fix $\sigma_{\rm m}^2 = 10^{-11}$.
+Equations \ref{eq:spinevol}- \ref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified by our construction of the synthetic PTA outlined in Section \ref{sec:synt_pta}.
+For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
+In practice, in order to avoid numerical precision issues that arise when solving Eq \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic evolution, equivalent to a change of variables:
+\begin{equation}
+	f_{\rm p}^{\prime (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)}
+\end{equation}  
+We similarly heterodyne the measurement variable 
+\begin{equation}
+	f_{\rm m}^{\prime (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{*(n)}
+\end{equation}
+where $ f_{\rm em}^{*(n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris.
+For synthetic data we can set $ f_{\rm em}^{*(n)} = f_{\rm em}^{(n)}$ but this is not true generally for real-world observations since the spin-down ephemeris is only known approximately, not exactly.
+The measurement equation Eq.
+\ref{eq:measurement} that relates the state and measurement variables is then updated as
+\begin{equation}
+	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}((t-d))(1-g^{(n)}(t))
+	\label{eq:measurement_cov}
+\end{equation}
+We emphasise that this change of variables is simply a convenient scaling to bring our numerical values into a reasonable dynamic range without having to use extended numerical floating point formats (e.g.
+long double, quadruple).
+It does not remove any degrees of freedom of the formulation.
+In particular both $f_{\rm em}^{(n)}(t_1)$
+and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain static parameters of the model, but now appear in the measurement equation rather than the state equation.
+
+\subsection{Representative example}\label{sec:rep_example}
+We initially characterise our method with a single example of a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
+The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are shown in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}.
+The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$ are as described in Section \ref{sec:synt_pta} and are also shown in Table \ref{tab:parameters_and_priors}.
+For the Bayesian methods used in this work, one must select reasonable priors for the complete set complete of static parameters, $\pi(\boldsymbol{\theta})$.
+For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
+The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires some additional discussion.
+The parameters which govern the deterministic evolution of the pulsar spin frequency, $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are well-determined by existing radio timing observations.
+We can identify $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar  barycentric rotation frequency and the time derivative of barycentric rotation frequency quoted by pulsar observation catalogues.
+For the 12.5 year NANOGrav pulsars the mean error in the barycentric rotation frequency is $\sim 1.1 \times 10^{-11}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.29 \times 10^{-19}$ s $^{-2}$.
+These mean errors are skewed by a minority of pulsars which are less well timed; the median error in  the barycentric rotation frequency is $\sim 7 \times 10^{-13}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.8 \times 10^{-20}$ s $^{-2}$.
+Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 1 \%$ either side of the true value.
+In this way we can test how well our method performs without requiring very precise measurements of the pulsar parameters to be made \textit{a priori}.
+Instead we consider the pulsar parameters similarly to the GW source parameters and estimate them optimally within a consistent framework.
+This counteracts the potential of a confusion problem c.f.
+the influence of timing noise and gravitational waves on the received pulsar signal.
+The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}.
+However, for this work whilst the pulsar distance is used when generating synthetic data, as discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d$.
+\textcolor{red}{TK: this next bit on gamma needs cleaning up}.
+For setting a prior on $\gamma^{(n)}$ since  $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$, $\gamma^{(n)}$ is in some sense ``unobservable" over these timescales and it is sufficient to set it at its true value i.e consider $\gamma$ to be known.
+We briefly explored a LogUniform prior of $\gamma$ over e.g.
+LogUniform($10^{-10}, 10^{-15}$) but the results are unchanged.
+In order to reduced the computational load we therefore set $\gamma$ to its true value.
+The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22}$.
+The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so the $\sigma^{(n)}$ inferred from Eq.
+\ref{eq:sigmap} \ref{eq:sigmap_f} $\sim 10^{-16}$.
+For the majority of pulsars we set an uninformative broad prior of $\sigma^{(n)}$ of LogUniform($10^{-19}, 10^{-23}$).
+For PSR J1939+2134 we use LogUniform($10^{-15}, 10^{-17}$).
+All static parameters and their priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
+\begin{table}
+	\centering
+	\resizebox{\columnwidth}{!}{%
+	\begin{tabular}{lccll}
+		\toprule
+		&Parameter & Injected Values & Units & Prior  \\
+		\hline
+		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
+	  & $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
+	  & $\delta$              & $1.0$  & radians & Cosine($-\pi/2, \pi/2$) \\
+	  & $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
+	  & $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
+	  & $h_0$            & $10^{-12}$ & - & LogUniform($10^{-9}$, $10^{-15}$) \\
+	  & $\iota$             & $1.0$ & radians & Sin($0, \pi$) \\ 
+		\hline
+		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ANTF}^{(n)}$ & Hz & Uniform(0.9$f_{\rm ANTF}^{n}$, $1.1 f_{\rm ANTF}^{n}$) \\
+		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ANTF}$ & s$^{-2}$ & Uniform(0.9$\dot{f}_{\rm ANTF}^{n}$, $1.1 \dot{f}_{\rm ANTF}^{n}$) \\
+		&  $d^{(n)}$       &$d_{\rm ANTF}^{(n)}$  & m & - \\
+		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & ? & LogUniform*($10^{-19}, 10^{-23}$) \\
+		& $\gamma$              & $10^{-13}$ & s$^{-1}$ & - \\
+		\bottomrule
+	\end{tabular}
+}
+	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter.
+The subscript ``ANTF" denotes values which have been obtained from the ANTF pulsar catalogue as described in Section \ref{sec:synt_pta}.
+The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Eq.
+\ref{eq:sigmap} \ref{eq:sigmap_f}.
+All pulsars use the prior on $\sigma^{(n)}$ as specificed in table, with the exception of PSR J1939+2134 which uses LogUniform($10^{-15}, 10^{-17}$)}
+	\label{tab:parameters_and_priors}
+\end{table}
+
+\subsubsection{Parameter estimation}\label{sec:parameter_estim}
+From Eq.
+\ref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms.
+The first term $\cos(-\Omega t + \Phi_0)$ depends only on the GW source parameters and is shared across all pulsars.
+The argument of the cosine corresponds to the GW phase at the observer on Earth.
+Conversely, the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$ varies between pulsars.
+The argument of the cosine corresponds to the GW phase at each the individual pulsar.
+The first and second terms are commonly referred to as the ``Earth term" and ``pulsar term" respectively.
+Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases and in standard PTA analysis are typically considered as a source of self noise and dropped from the analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
+In this work we follow the standard analysis approach and drop the pulsar terms from our model.
+We defer the inclusion of the pulsar terms in the measurement equation to a future work.
+We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
+from the Kalman filter model that feeds into the nested sampling algorithm.
+The synthetic data that we apply our method on does include the pulsar term in full.
+\newline
+
+First we consider a single noise realisation of the synthetic data for the representative example.
+We apply the Kalman filter using the earth terms only measurement equation in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters, which are shown in Figure X.
+
+\begin{figure*}
+	\includegraphics[width=0.8\textwidth]{images/1237_broad_GW}
+	\caption{some caption}
+	\label{fig:corner_plot_1}
+\end{figure*}
+
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/1237_broad_f}
+	\caption{some caption}
+	\label{fig:corner_plot_2}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/1237_broad_fdot}
+	\caption{some caption}
+	\label{fig:corner_plot_3}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/1237_broad_sigma_p}
+	\caption{some caption}
+	\label{fig:corner_plot_4}
+\end{figure}
+
+\subsubsection{Detection}
+
+\subsection{Practical considerations for inference}
 
@@
 
-pulsar terms 
-For now we will consider the measurement noise to be known, although in principle this too could be estimated.
+pulsar terms vs earth terms
+
+For this initial demonstrative work we use only the Earth terms
+
+loss of SNR, cite Zhu etc.
+
+We defer the inclusion of the pulsar terms for a future work
+
+reduces the dimensionality of the problem to $7 + 4N$ since the measurement equation eq X is no longer a function of the pulsar distance.
+
+we emphasise that whilst our alternative model $m$ disregards the pulsar terms, the synthetic data does have the pulsar terms
 
 \subsection{PTA pulsars}\label{sec:pta_pulsars}
-\begin{figure}
-	\includegraphics[width=0.8\columnwidth]{images/pulsars}
-	\caption{Spatial distribution and distances of NANOGrav pulsars}
-	\label{fig:pulsar_distrib}
-\end{figure}
-\noindent With our Kalman filter and nested sampling techniques in hand, in order to proceed it is necessary to specify a PTA configuration.
-As discussed, multiple separate PTA detectors exist under the umbrella of the IPTA.
-Going forward we will take the 47 pulsars that make up the NANOGrav PTA \citep{2020ApJ...905L..34A}.
-NANOGrav is selected simply as a well-representative example of the typical pulsars that make up a PTA.
-Our results and formulation are not contingent on the choice of PTA, and naturally extend to other PTAs or PTAs with more pulsars.
-\newline
+
+\noindent With our Kalman filter and nested sampling techniques in hand,
 
@@
 
+we consider the pulsars to have data over the entire
+
 \subsection{Parameter estimation}\label{sec:param_estimation}
-We are now in a position to try to infer the parameters of a GW system.
+
+%We are now in a position to try to infer the parameters of a GW system.
 We take our NANOGrav PTA and create a synthetic noisy dataset using the parameters described in Table \ref{tab:toy_example_parameters}.
-\begin{table}
-	\centering
-	\begin{tabular}{lc}
-		Parameter & Value  \\
-		\hline
-		$\omega$       & $5 \times 10^{-7}$  \\
-		$\alpha$          & $1.0$  \\
-		$\delta$              & $1.0$   \\
-		$\psi$              & $2.50$  \\
-		$\iota$             & $1.0$  \\
-		$\Phi_0$          & $0.20$  \\
-		$h$ & 1e-2 \\
-		$\sigma_{\rm p}$ & $10^{-13}$ \\
-		$\sigma_{\rm m}$ & $10^{-10}$ \\
-		$T_{obs}$ & 10 \text{ years} \\
-		$\Delta t$ & 7 \text{ days} \\
-		\hline
-	\end{tabular}
-	\caption{GW parameters used for generating synthetic data. There is nothing special about these parameters - they are just chosen arbitrarily.}
-	\label{tab:toy_example_parameters}
-\end{table}
+%\begin{table}
+%	\centering
+%	\begin{tabular}{lc}
+%		Parameter & Value  \\
+%		\hline
+%		$\omega$       & $5 \times 10^{-7}$  \\
+%		$\alpha$          & $1.0$  \\
+%		$\delta$              & $1.0$   \\
+%		$\psi$              & $2.50$  \\
+%		$\iota$             & $1.0$  \\
+%		$\Phi_0$          & $0.20$  \\
+%		$h$ & 1e-2 \\
+%		$\sigma_{\rm p}$ & $10^{-13}$ \\
+%		$\sigma_{\rm m}$ & $10^{-10}$ \\
+%		$T_{obs}$ & 10 \text{ years} \\
+%		$\Delta t$ & 7 \text{ days} \\
+%		\hline
+%	\end{tabular}
+%	\caption{GW parameters used for generating synthetic data. There is nothing special about these parameters - they are just chosen arbitrarily.}
+%	\label{tab:toy_example_parameters}
+%\end{table}
+%
+
 \noindent Given this synthetic data, we can try to use our KF + NS approach to recover some of the parameters.
@@
 \subsection{Detection}\label{sec:detection}
+
+spherical localisatin plot would be cool e.g.
+
+\url{https://github.com/stephengreen/lfi-gw}
 
@@
 \section{Conclusion}
+
+Our method is not contingent on the choice of PTA, and naturally extend to other PTAs or PTAs with more pulsars.
 

### 2023-08-01 docs/manuscript/paper.tex
added Vargas plot
updates to compare

@@
 By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs.
-Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}.
-These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}.
+Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA,][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}.
+These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep[CPTA,][]{Hobbs_2019}.
 \newline
 
-\textcolor{red}{TK: Content and references need to be updated in light of recent PTA results}.
 The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
 Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
-However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}.
-\newline
+After multiple non-detections \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A} consilient evidence for the GW background was presented by NANOGrav \citep{2023ApJ...951L...8A}, EPTA/InPTA \citep{2023arXiv230616214A}, PPTA \citep{2023ApJ...951L...6R} and the CPTA \citep{2023RAA....23g5024X} \newline
 
@@
 Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs.
-However, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}.
+However, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for a individual source at 4-5 nHz \citep{2023arXiv230616226A}.
 \newline
@@
 The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Eq.
-\ref{eq:measurement}d.
+\ref{eq:measurement}.
 The measurement predictions can then be compared aganst the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation".
@@
 \section{Tests with synthetic data} \label{sec:testing}
-We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.
 
-\subsection{Synthetic data generation}
+In this section we use synthetic data to test the ability of our method to detect the presence of a GW in noisy data and recover the static parameters $\boldsymbol{\theta}$.
+In Section \ref{sec:synth_data} we discuss how the synthetic data is generated, including the construction of an artificial PTA.
+In Section \ref{sec:rep_example} we generate synthetic data for a single representative example SMBHB system and apply our method to recover the system parameters and the Bayes factor.
+\textcolor{red}{TK: more here once this section is completed}.
+
+\subsection{Synthetic data generation} \label{sec:synth_data}
 
@@
 We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}.
+
 For each pulsar we get the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$.
@@
 \end{eqnarray}
-We can then vary $\sigma^{(n)}$ and compare the synthetically generate phase residuals with the phase solution from the UTMOST pulsar timing program with the Mongolo Observatory Synthesis Telescope  \citep{UTMOST}.
-It is sufficient to simply compare the synthetic and real solutions visually, rather than e.g.
-defining some quantitative loss function (e.g.
-mean-squared error) between the two solutions and solving the optimisation problem of the value of $\sigma_{\rm TOA}^{(n)}$ which minimised this loss function; we emphasise that we are not overly concerned with getting the maximally accurate values for the noise in the pulsar frequency evolution, but just instead some reasonable, representative values.
-This approach of generating synthetic phase solutions and visually comparing with real solutions is also followed in \cite{Vargas}.
-In Figure \ref{fig:qualitative_compare} we demonstrate for a particular representative pulsar, PSR XXX, blah blah blah...
+For a particular choice of $\gamma^{(n)}$ we can then vary $\sigma^{(n)}$ and compare the synthetically generated phase residuals with the empirical phase residuals.
+We obtain empirical phase residuals from the NANOGrav 12.5 year narrowband data set \cite{nanograv_narrowband_DR}.
+For the purposes of validating the $\sigma^{(n)}$ inferred from Eq \ref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real solutions visually.
+Whilst it would be possible to e.g.
+define some quantitative loss function (e.g.
+mean-squared error) between the two solutions and solve the optimisation problem to infer $\sigma^{(n)}$, for our purposes this is overkill and it is satisfactory to simply qualitatively compare the two solutions as a useful sanity check.
+We emphasise that we are not overly concerned with getting the maximally accurate values for $\sigma^{(n)}$, but just instead some reasonable values for constructing representative synthetic data.
+This approach of generating synthetic phase solutions and visually comparing with empirical solutions is also followed in \cite{Vargas}.
+In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for 3 NANOgrav pulsars: J0030+0451, J0340+4130 and J0613-0200.
+We can see that for all pulsars the synthetic and empirical residuals are qualitatively similar with the magnitude of the residuals being directly comparable for each pulsar.
    \begin{figure}
-	\includegraphics[width=\columnwidth]{example-image-a}
-	\caption{Actual and synthetic phase evolution}
+	\centering
+	\subfloat[PSR J0030+0451]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-a}}
+
+	\subfloat[PSR J0340+4130 ]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-b}}
+
+	\subfloat[PSR J0613-0200]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-c}}
+	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars.
+The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}.
+The synthetic residuals are generated by numerically solving Eq.
+\ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Eqs.
+\ref{eq:sigmap}, \ref{eq:sigmap_f}.
+For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
+\textcolor{red}{TK: these plots need updating, but need to discuss with Andres re use of baboo}}
 	\label{fig:qualitative_compare}
@@
 
+%
+%\textbf{phase solution from the UTMOST pulsar timing program with the Mongolo Observatory Synthesis Telescope  \citep{UTMOST}. I}
+%
+
 \subsubsection{Synthetic data generation}
-We generate $N$ synthetic noisy timeseries by integrating the state equations, Eq \ref{eq:spinevol}- \ref{eq:xieqn} numerically, mapping from state space to measurement space via the measurement equation, Eq.
+We generate $N$ synthetic noisy timeseries by integrating the state equations, Eq \ref{eq:spinevol}- \ref{eq:xieqn} numerically for the synthetic PTA described in Sectin \ref{sec:synt_pta}, mapping from state space to measurement space via the measurement equation, Eq.
 \ref{eq:measurement}, and finally adding to each timeseries some zero-mean Gaussian noise with covariance $\sigma_{\rm m}^2$.
-We take $\sigma_{\rm m}^2$ to be known \textit{a priori} rather than a parameter to be inferred.
-Throughout this work we fix $\sigma_{\rm m}^2 = 10^{-11}$.
 Equations \ref{eq:spinevol}- \ref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
@@
 For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
-In practice, in order to avoid numerical precision issues that arise when solving Eq \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic evolution, equivalent to a change of variables:
+We take $\sigma_{\rm m}^2$ to be known \textit{a priori} rather than a parameter to be inferred.
+The measurement noise can be approximately related to the uncertainty in the pulse TOA $\sigma_{\rm TOA}$ as
+\begin{equation}
+	\sigma_{\rm m} = f_{\rm p}^{(n)} \frac{\sigma_{\rm TOA}}{\text{cadence}} \ . 
+\end{equation}
+For a millisecond pulsar with $f_{\rm p}^{(n)}$ observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$ s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
+Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$.
+In practice, in order to avoid numerical precision issues that arise when solving Eq \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
 \begin{equation}
@@
 We defer the inclusion of the pulsar terms in the measurement equation to a future work.
+This choice reduces the dimensionality of the parameter estimation problem to $7 + 4N$ since the measurement equation Eq \ref{eq:measurement} is no longer a function of the pulsar distance.
 We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
 from the Kalman filter model that feeds into the nested sampling algorithm.
-The synthetic data that we apply our method on does include the pulsar term in full.
+The generated synthetic data that we use to test our method does include the pulsar term in full.
 \newline
 
-First we consider a single noise realisation of the synthetic data for the representative example.
-We apply the Kalman filter using the earth terms only measurement equation in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters, which are shown in Figure X.
+Initially we consider a single noise realisation of the synthetic data for the representative example.
+We apply the Kalman filter using the earth terms only measurement equation in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters, which are shown in Figure \ref{fig:corner_plot_1}.
+
+the gw parameters....
+
+the frequency parameters....
+
+the sigma p parameters....
+
+bias in parameters c.f. likelihood
 
@@
 
-\subsection{Practical considerations for inference}
+\subsection{Multiple noise realisations}
 
-heterodyeing
+The above discussion was just for a single realisaitn of the noise
 
-pulsar terms vs earth terms
+\subsubsection{Exploring a broader parameter space}
 
-For this initial demonstrative work we use only the Earth terms
+The above was for a particular SBHB system
 
-loss of SNR, cite Zhu etc.
+\section{Discussion}
 
-We defer the inclusion of the pulsar terms for a future work
+hellings downs
 
-reduces the dimensionality of the problem to $7 + 4N$ since the measurement equation eq X is no longer a function of the pulsar distance.
-
-we emphasise that whilst our alternative model $m$ disregards the pulsar terms, the synthetic data does have the pulsar terms
-
-\subsection{PTA pulsars}\label{sec:pta_pulsars}
-
-\noindent With our Kalman filter and nested sampling techniques in hand,
-
-\noindent Within our state-space formulation, the pulsar evolution is governed by a set of 5 parameters, $\bar{\theta}_{\rm PSR}$ for each pulsar.
-The parameters $f_{\rm EM} (0) $,$ \dot{f}_{\rm EM} (0)$ and $d$ are well specified via existing pulsar datasets.
-We take the frequency and frequency derivative as returned from the pulsar datasets to simply be the values now at $t=0$.
-The pulsar distances are also known, though less well constrained.
-Going forward we take the distances retuned from the datasets as the true values of the pulsars that make up our synthetic PTA.
-The specification of $\gamma$ and $\sigma$ are more involved.
-$\gamma$ specifies an effective timescale of reversion to the mean \textcolor{red}{TK: Need some discussion and though on how to choose these two parameters to be astrophysically reasonable c.f.
-A.
-Vargas}
-
-we consider the pulsars to have data over the entire
-
-\subsection{Parameter estimation}\label{sec:param_estimation}
-
-%We are now in a position to try to infer the parameters of a GW system.
-We take our NANOGrav PTA and create a synthetic noisy dataset using the parameters described in Table \ref{tab:toy_example_parameters}.
-%\begin{table}
-%	\centering
-%	\begin{tabular}{lc}
-%		Parameter & Value  \\
-%		\hline
-%		$\omega$       & $5 \times 10^{-7}$  \\
-%		$\alpha$          & $1.0$  \\
-%		$\delta$              & $1.0$   \\
-%		$\psi$              & $2.50$  \\
-%		$\iota$             & $1.0$  \\
-%		$\Phi_0$          & $0.20$  \\
-%		$h$ & 1e-2 \\
-%		$\sigma_{\rm p}$ & $10^{-13}$ \\
-%		$\sigma_{\rm m}$ & $10^{-10}$ \\
-%		$T_{obs}$ & 10 \text{ years} \\
-%		$\Delta t$ & 7 \text{ days} \\
-%		\hline
-%	\end{tabular}
-%	\caption{GW parameters used for generating synthetic data. There is nothing special about these parameters - they are just chosen arbitrarily.}
-%	\label{tab:toy_example_parameters}
-%\end{table}
-%
-
-\noindent Given this synthetic data, we can try to use our KF + NS approach to recover some of the parameters.
-The results for inference on 5 parameters, $\omega.
-\Phi_0, \psi, \iota, h$ are shown in Fig \ref{fig:corner2}.
-\textcolor{red}{TK: likelihood plots for each parameter could also be of interest to include here?
-Need to expand this section greatly for more astrophysical systems and more unknown parameters}
-%\begin{figure*}
-%	\includegraphics[width=0.8\textwidth]{images/corner_example}
-%	\caption{Example corner plot. Aside: for this run the $\sigma_p$ value fed to the filter is different from that used to generate the data.}
-%	\label{fig:corner}
-%\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.8\textwidth]{images/canonical_example}
-	\caption{Example corner plot. Aside: for this run the $\sigma_p$ value fed to the filter is different from that used to generate the data.}
-	\label{fig:corner2}
-\end{figure*}
-
-\subsection{Detection}\label{sec:detection}
+multiple sources
 
@@
 \url{https://github.com/stephengreen/lfi-gw}
-
-In addition to estimating the parameters of the system, we are also interested in how detectable a GW is using PTA + state space method.
-We can use our state-space tools  to try to solve the problem of GW detection with a PTA i.e.
-\textit{"Is there evidence of a GW in my data?"}We can frame this as a model selection procedure where we have two models/hypotheses:
-
-\begin{itemize}
-\item Null Model, $M_0$. There is no GW in the data. In this case the measurement model of the Kalman filter simply returns the frequency states i.e. $g(t,\theta)= 0$
-\item Alternative model, $M_1$. There is a GW in the data. The measurement model uses the full expression for $g(t,\theta)$
-\end{itemize}
-
-In order to accept the alternative hypothesis $M_1$ over $M_0$ there are two approaches we could take:
-
-\begin{enumerate}
-	\item The first is a fully Bayesian search over all the parameters for each model, calculating the evidence for each model and then determining the Bayes ratio.
-This is perhaps the most consistent way, but it is obviously expensive and at this stage we are keen to explore how detectability varies with e.g.
-GW strain.
-	\item The second method is to recognize that $M_0$ and $M_1$ are hierarchically nested models and we can perform a likelihood ratio test.
-That is, given the maximum likelihood estimators $\hat{\theta}$ of the true parameters $\theta$, the likelihood of each model can be calculate.
-These likelihoods are just  point estimates of the Bayes factor numerator/denominators.
-They can then be compared via the likelihood ratio $\Lambda$.
-\end{enumerate}
-Given the cheap cost we proceed with the second method.
-For the likelihood ratio test we do not perform any kind of maximum likelihood search over the parameters for each of the models.
-Instead we just artificially set the maximum likelihood estimators to be equal to the true parameters of the system i.e.
-$\hat{\theta} = \theta$.
-We assume that any maximum likelihood algorithm would converge to these parameters.
-This is obviously an oversimplification but will serve our purposes for now.
-\newline
-
-\noindent Interpreting the likelihood ratio $\Lambda$ also needs some consideration, since we have to account for the increased model complexity of $M_1$.
-Bayes factors penalise complexity by construction since one must integrate over a larger parameter space.
-There are many different ways to do this - fot now we will consider two: Akaike information criterion (AIC) and Wilks Theorem.
-For the latter, Wilks' Theorem which states that for a large number of samples \footnote{What counts as large?
-See \url{https://www.osti.gov/servlets/purl/1529145}} the distribution of the test statistic approaches the chi-squared distribution under the null hypothesis i.e.
-\begin{equation}
-	2 \log \Lambda \rightarrow \chi^2
-\end{equation}
-One can then compute $p$-values where the number of degrees of freedom is equal to the difference in the number of parameters of the two models; $M_1$ has 7 extra parameters over $M_0$ corresponding to the 7 parameters of $\bar{\theta}_{\rm GW}$.
-With 7 degrees of freedom and a target tolerance of 5(1) \% the test statistic is $\sim 14 (18.5)$.
-\newline
-
-An alternative approach is to use the AIC which is given by,
-\begin{equation}
-	\text{AIC} = 2k - 2\log\ \mathcal{L} 
-\end{equation}
-where $k$ is the number of degrees of freedom. The AIC can be computed for each model and the model with the minimum AIC is preferred. This can be straightforwardly mapped into a relative condition
-\begin{equation}
-	\log \mathcal{L}_1 - 	\log \mathcal{L}_0 > 7 
-\end{equation}
-Note this is the same as the Wilks case!
-
-\noindent
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/snr_canonical}
-	\caption{Likelihood ratio vs strain for \textit{(a)} full PTA and \textit{(b)} a single randomly chosen pulsar.
-Horizontal dashed line shows the minimum detectable cutoff, vertical dashed lines are the corresponding GW strains.}
-	\label{fig:SNR}
-\end{figure}
-
-\begin{figure*}
-	\includegraphics[width=0.8\textwidth]{images/snr_mollewide}
-	\caption{\textcolor{red}{TK: Just a placeholder. Likelihood ratio for a given $h$. Maybe more interesting, but more expensive, to get minimum detectable h?}}
-	\label{fig:mollewide}
-\end{figure*}
-
-%
-%
-%
-%\begin{figure}
-%	%\centering % Not needed
-%	\begin{subfigure}[b]{1\columnwidth}
-%		\includegraphics[width=\textwidth]{images/snr_canonical}
-%		\caption{}
-%		\label{fig:6MB_BFS}
-%	\end{subfigure}
-%	\\
-%	\begin{subfigure}[b]{1\columnwidth}
-%		\includegraphics[width=\textwidth]{images/snr_canonical_1psr}
-%		\caption{}
-%		\label{fig:25MB_bfs}
-%	\end{subfigure}
-%	\caption{}
-%	\label{fig:snr_detection}
-%\end{figure}
-
-\section{Discussion}
 
@@
 
+ \newpage 
 \appendix

### 2023-08-02 docs/manuscript/paper.tex
manuscript parameter sec

@@
 \begin{align}
-	h_+ &= h(1 + \cos^2 \iota) \\
-	h_{\times} &= -2h\cos \iota
+	h_+ &= h(1 + \cos^2 \iota) 	\label{eq:hphx} \\
+	h_{\times} &= -2h\cos \iota 	\label{eq:hphx2}
 \end{align}
@@
 For PSR J1939+2134 we use LogUniform($10^{-15}, 10^{-17}$).
+Since we are not setting priors on $\gamma$ or $d$ this reduces our dimensionality to $7 + 3N$.
+We will use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space.
 All static parameters and their priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
@@
 Initially we consider a single noise realisation of the synthetic data for the representative example.
-We apply the Kalman filter using the earth terms only measurement equation in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters, which are shown in Figure \ref{fig:corner_plot_1}.
-
-the gw parameters....
-
-the frequency parameters....
-
-the sigma p parameters....
-
-bias in parameters c.f. likelihood
-
+We apply the Kalman filter using the earth terms only measurement equation in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters, which are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:corner_plot_1} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ in Figure \ref{fig:example}.
+It is evident that for this representative example we are able to accurately estimate all the static parameters of the system.
+Of particular interest are the estimates of $\boldsymbol{\theta}_{\rm gw}$.
+Whilst some parameters are recovered identically (e.g.
+$\omega, \Phi_0, \delta, \iota$) for others the posterior median is shifted away from the injected value (e.g.
+$\psi, \alpha$).
+For a single realisation of the noise it is not clear if this is a systematic effect, i.e.
+a bias in the parameter estimate, or else just a random variation.
+We will explore this further in Section \ref{sec:multiple_noise} and show that this is a bias that results from dropping the Earth terms, similar to the bias reported in \cite{Zhupulsarterms}.
+Regarding $\boldsymbol{\theta}_{\rm psr, reduced}$, the frequency parameters $f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1)$ are generally well recovered for all pulsars, as are the noise amplitude parameters $\sigma^{(n)}$.
 \begin{figure*}
 	\includegraphics[width=0.8\textwidth]{images/1237_broad_GW}
-	\caption{some caption}
+	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for a single realisation of the system noise.
+The vertical orange lines indicate the true injected values, c.f.
+Table \ref{tab:parameters_and_priors}.
+The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+We are able to accurately estimate each parameter of interest.}
 	\label{fig:corner_plot_1}
 \end{figure*}
+\begin{figure*}%
+	\centering
+	\subfloat[\centering label 1]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_f} }}%
+	\qquad
+	\subfloat[\centering label 2]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_fdot} }}%
+	\qquad
+	\subfloat[\centering label 2]{{\includegraphics[width=0.3\textwidth]{images/1237_broad_sigma_p} }}%
 
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/1237_broad_f}
-	\caption{some caption}
-	\label{fig:corner_plot_2}
-\end{figure}
+	\caption{As Fig \ref{fig:corner_plot_1} but for the $N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$. We show only a selection of the first 5 pulsars rather than the full PTA.}
+	\label{fig:example}%
+\end{figure*}
 
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/1237_broad_fdot}
-	\caption{some caption}
-	\label{fig:corner_plot_3}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/1237_broad_sigma_p}
-	\caption{some caption}
-	\label{fig:corner_plot_4}
-\end{figure}
+%
+%
+%
+%
+%\begin{figure}
+%	\includegraphics[width=\columnwidth]{images/1237_broad_f}
+%	\caption{some caption}
+%	\label{fig:corner_plot_2}
+%\end{figure}
+%
+%\begin{figure}
+%	\includegraphics[width=\columnwidth]{images/1237_broad_fdot}
+%	\caption{some caption}
+%	\label{fig:corner_plot_3}
+%\end{figure}
+%
+%
+%
+%\begin{figure}
+%	\includegraphics[width=\columnwidth]{images/1237_broad_sigma_p}
+%	\caption{some caption}
+%	\label{fig:corner_plot_4}
+%\end{figure}
 
@@
 
-\subsection{Multiple noise realisations}
+\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
 
 The above discussion was just for a single realisaitn of the noise
+
+iota is also shifted.
+
+due to the degeneracy this impacts h
+
+The estimates on $h$ and $\iota$ are generally less accurate than the other parameters of $\boldsymbol{\theta}_{\rm gw}$.
+
+From Eq \ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+Whilst both parameters are generally identifiable the algebraic relation between the two makes it difficult for our algorithm to estimate both simultaneously.
+For example, from Eq \ref{eq:hphx2} it is not clear if a larger $h_{\times}$ is due to a larger strain $h$ or a different inclination $\iota$.
+This degeneracy is broken by the relation for $h_{+}$ which does separate $h$ and $\iota$ but the algebraic relation of how these parameters get mapped into the measurement equation tends for our algorithm to trade off one parameter against the other.
+If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy.
+This is demonstrated in Fig X which is analogous to the representative exxample, but now we have fixed $h$ at its true value (i.e.
+reduced the prior to a delta function).
+WE can see that the method can now infer $\iota$ which much greater accuracy.
 

### 2023-08-04 docs/manuscript/paper.tex
paper updates

@@
 \usepackage{multirow}
+
+\usepackage[section]{placeins} %ensures figures go in their section e.g https://tex.stackexchange.com/questions/279/how-do-i-ensure-that-figures-appear-in-the-section-theyre-associated-with
 
@@
 In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
-We combine the Kalman tracking of the pulsars intrinsic rotational state with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood for model selection.\textcolor{red}{TK: a few extra sentences here on how we compare to existing approaches.
-Not quite sure yet how to do this.} \newline
+We combine the Kalman tracking of the pulsars intrinsic rotational state with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood for model selection.
+\newline
 
@@
 \end{equation}
-$[\sigma^(n)]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+$[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
 It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Eq.
@@
 The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$.
-Throughout this paper we will work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB).
+Throughout this paper we work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB).
 We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB.
@@
 \begin{equation}
-	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right)
+	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right) \ .
 	\label{eq:g_func}
@@
 	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
-	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right) \right]
+	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right) \right] \ .
 	\label{eq:g_func_trig}
@@
 \section{Signal tracking and parameter estimation} \label{sec:detect}
-The set of static parameters $\boldsymbol{\theta}$ of the model in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e.
+The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e.
 \begin{equation}
-	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \label{eq:params1}
+	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
 \end{equation}
@@
 \begin{equation}
-	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N}
+	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
 \end{equation}
@@
 \begin{equation}
-	\boldsymbol{\theta}_{\rm gw} = \left \{h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
+	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
 \end{equation}
-We can reparameterize the two GW polarisation strains in terms of the GW amplitude $h_0$ and the system inclination $\iota$
+In Eq \ref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
 \begin{align}
-	h_+ &= h(1 + \cos^2 \iota) 	\label{eq:hphx} \\
-	h_{\times} &= -2h\cos \iota 	\label{eq:hphx2}
+	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
+	h_{\times} &= -2h\cos \iota 	\label{eq:hphx2} \ , 
 \end{align}
-where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$ and the observer line of sight, i.e.
+where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e.
 $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.
-Henceforth we use the parametrisation in terms of $h_0$ and $\iota$.
-For a PTA dataset containing $N$ pulsars we have $7 + 5N$ parameters to estimate.
-Typically the pulsar parameters are better constrained than the GW parameters: for example estimates of pulsar distances are accurate to $\sim$ 10$\%$, but we have no prior information on the source location \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.The variance of the measurement noise $\Sigma$ can be estimated approximately from the root-mean-square amplitude of phase residuals and we do not treat it as a parameter to be inferred in this work.
+We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
+For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate.
+Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.
 \newline
@@
 the model evidence).
-In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency $f_{\rm m}^{(n)}(t)$ can be used to estimate the hidden state sequence $f_{\rm p}^{(n)}(t)$ using a Kalman filter.
-In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampler to perform Bayesian inference of the model parameters and calculate the model evidence.
+In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter.
+In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference of the model parameters and calculate the model evidence.
 Model selection and the specification of the null model is described in Section \ref{sec:model_selection}.
@@
 	\end{subfigure}
-	\caption{Application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$.
+	\caption{Example application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$.
 We here plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency.
-In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$.
-In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$.
-The top left panel shows the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line).
-The middle left panel shows the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta).
-The bottom left panel shows the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$.
-The right hand panel plots the distribution of the innovation over the total observation period.}
+In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ and the state is estimated accurately.
+In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$ and the state is not estimated accurately.
+The top left panels show the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line).
+The middle left panels show the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta).
+The bottom left panels show the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$.
+The right hand panels plot the distribution of the innovation over the total observation period.}
 	\label{fig:four figures}
@@
 
-The Kalman filter \citep{Kalman1} is an algorithm for recovering the most likely evolution of a set of system state variables, $\boldsymbol{X}$ given some noisy measurements $\boldsymbol{Y}$.
+The Kalman filter \citep{Kalman1} is an algorithm for recovering the most likely evolution of a set of system state variables, $\boldsymbol{X}$, given some noisy measurements, $\boldsymbol{Y}$.
 It is a common technique in signal processing that has been applied successfully in neutron star astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}.
@@
 At each discrete timestep $i = 1, ...
-, M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates $\boldsymbol{P}_i$.
+, M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$.
 The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Eq.
@@
 The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$.
-The Gaussian log-likelihood can then be calculated at each timestep,
+The Gaussian log-likelihood can then be calculated at each timestep:
 \begin{eqnarray}
-	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right )
+	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ ,
 \end{eqnarray}
-with the total log-likelihood the sum over all timesteps, i.e. 
+with the total log-likelihood simply the sum over all timesteps, i.e. 
 \begin{eqnarray}
-	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i  \label{eq:likelihood}
+	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i \ . \label{eq:likelihood}
 \end{eqnarray}
-The likelihood $\mathcal{L}$, along $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of the static parameters of the model, $\bar{\theta}$.
-If the estimates of the system parameters, $\hat{\theta}$, that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised.
-This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the hidden state evolution with high fidelity - leavign only gaussian errrs
-
-Conversely, if $\hat{\theta}$ is not close to $\bar{\theta}$ then the filter is unable to recover the state evolution.
-This is demonstrated in Figure \ref{fig:25MB_bfs} where the failure of the filter to track the hidden state is evident.
+For fixed data $\boldsymbol{Y}$, $\mathcal{L}$ is a function of the static parameters of the model, i.e.
+$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})$.
+Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are also functions of $\boldsymbol{\theta}$.
+If the estimates of the system parameters, $\boldsymbol{\hat{\theta}}$ , that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised.
+This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
+The residuals correspond to random noise and are normally distributed.
+Conversely, if $\boldsymbol{\hat{\theta}}$ is not close to the true parameters then the filter is unable to recover the state evolution.
+This is demonstrated in Figure \ref{fig:25MB_bfs} where the Kalman filter was run with $\boldsymbol{\hat{\theta}}$ slightly perturbed away from the true values.
+In this case the filter cannot track the state variable accurately and the residuals are no longer Gaussian.
 
 \subsection{Nested Sampling}\label{sec:nested_sampling}
-
-We can use the likelihood returned by the Kalman filter, Eq \ref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\bar{\theta}$ by Bayes' Rule,
+We can use the likelihood returned by the Kalman filter, Eq \ref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
 \begin{equation}
-	p(\bar{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L} \cdot \pi(\theta)}{\mathcal{Z}}
+	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}}
 \end{equation}
-where $\pi(\theta)$ is the prior distribution on $\bar{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
+where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
 \begin{equation}
-	\mathcal{Z} = \int \mathcal{L} \pi (\theta) d \theta \label{eq:model_evidence}
+	\mathcal{Z} = \int \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta}) d \boldsymbol{\theta} \ . \label{eq:model_evidence}
 \end{equation}
-In order to estimate the posterior distribution and the model evidence we use nested sampling \citep[NS,][]{Skilling} throughout this work.
+In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work.
 Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Eq.
-\ref{eq:model_evidence}, that also returns samples from the posterior $p(\bar{\theta} | \boldsymbol{Y})$.
-The primary advantage of nested sampling is the ability to compute this evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches.
+\ref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
+The primary advantage of nested sampling is the ability to compute the evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches.
 Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}.
@@
 For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \cite{bilby.507.2037A} front-end library.
-We  follow this precedent and use \texttt{Bilby} for all Bayesian inference in this work.
+We  follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work.
+The primary tunable parameter in nested sampling is $n_{\rm live}$, where a greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}(1/\sqrt{n_{\rm live}})$.
+However the computational runtime scales as $\mathcal{O}(n_{\rm live})$ and so one must make a trade-off between uncertainty and runtime.
+\cite{Ashton2022} offer a rule-of-thumb where the minimum number of live points should be greater than the number of static parameters.
+The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$.
+We take $n_{\rm live} = 500$ for all results presented in this work.
 
-%from https://arxiv.org/pdf/2102.12478.pdf
-%Nested sampling performs these dual processes by evolving an
-%ensemble of 𝑛live live points by initially drawing them from the prior
-%distribution 𝜋, and then at each subsequent iteration: (a) discarding
-%the lowest likelihood live point and (b) replacing it by a live point
-%drawn from the prior 𝜋 subject to hard constraint that the new point is
-%at a higher likelihood than the discarded point. The set of discarded
-%dead points and likelihoods can be used to compute a statistical
-%estimate of the Bayesian evidence (1), and be weighted to form a set
-%of samples from the posterior distribution.
-%The nested sampling procedure therefore differs substantially from
-%traditional strategies such as Gibbs sampling, Metropolis-Hastings
-%or Hamiltonian Monte Carlo in that it is acts as a likelihood scanner
-%rather than a posterior sampler. It scans athermally, so in fact is
-%capable of producing samples at any temperature and computing the
-%full partition function in addition to the Bayesian evidence (Handley
-%2019b).
+%See https://arxiv.org/pdf/2102.12478.pdf for a good NS explanation
 
 \subsection{Model selection}\label{sec:model_selection}
-The evidence integral returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}$.
+The evidence integral,$\mathcal{Z}$, returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}_i$.
 This enables us to compare competing models via a Bayes factor,
 \begin{equation}
-	\beta = \frac{\mathcal{Z}(\mathcal{M}_1)}{\mathcal{Z}(\mathcal{M}_0)}
+	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ .
 \end{equation}
-where the subscript index identifies a particular model.
 Throughout this work we take $\mathcal{M}_1$ to be the complete model defined in Section \ref{sec:model}.
@@
 This is equivalent to setting $g^{(n)}(t)=1$ in Equation \ref{eq:g_func}, \ref{eq:g_func_trig}.
-The Bayes factors we quote in this paper therefore quantify whether the data shows evidence for a GW compared to no GW.
+The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
 
 \section{Tests with synthetic data} \label{sec:testing}
-
-In this section we use synthetic data to test the ability of our method to detect the presence of a GW in noisy data and recover the static parameters $\boldsymbol{\theta}$.
+In this section we use synthetic data to test the ability of our method to detect the presence of a GW in noisy data and recover the static parameters, $\boldsymbol{\theta}$.
 In Section \ref{sec:synth_data} we discuss how the synthetic data is generated, including the construction of an artificial PTA.
-In Section \ref{sec:rep_example} we generate synthetic data for a single representative example SMBHB system and apply our method to recover the system parameters and the Bayes factor.
-\textcolor{red}{TK: more here once this section is completed}.
+In Section \ref{sec:rep_example} we generate synthetic data for a single representative example SMBHB system and apply our method to estimate $\boldsymbol{\theta}$ and calculate the statistical evidence for a GW in the data via the Bayes factor.
+In this section we also explore how the detectability of the system changes for different magnitudes of $h_0$.
+In Section \ref{sec:parameter_space} we move beyond a single example system and explore a broader parameter space...
 
@@
 We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}.
-
-For each pulsar we get the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$.
+For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$.
 The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Fig \ref{fig:pulsar_distrib}.
-The data is obtained via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+The data is acquired via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
 The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$.
-We disused in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$.
-Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ \cite{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars.
+We discussed in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$.
+Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ \citep{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars.
 For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
 In order to set a physically reasonable value for $[\sigma^{(n)}]^2$ we use two complementary approaches.
-The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by
-\begin{eqnarray}
-	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right)+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \label{eq:sigmap}
-\end{eqnarray}
+The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by:
+\begin{align}
+	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)&= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
+	&+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
+\end{align}
 where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$.
@@
 \begin{eqnarray}
-	\sigma^{(n)} = \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
+	\sigma^{(n)} \sim \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
 \end{eqnarray}
-For our synthetic NANOGrav PTA, the median $\sigma^{(n)} = 4.28 \times 10^{-21} $ s.
-Alternatively to this empirical best-fit model, we can also directly solve the state equation Eq.
-\ref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a phase solution
+For our synthetic NANOGrav PTA, the median $\sigma^{(n)}$ calculated in this way is $ = 4.28 \times 10^{-21} $ s. \newline
+
+As a sanity check, we can also calculate $\sigma^{(n)}$ using a complementary numerical approach to the empirical best-fit model outlined above.
+We can directly solve the state equation Eq.
+\ref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a synthetic phase solution,
 \begin{eqnarray}
-	\phi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') 
+	\phi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
 \end{eqnarray}
-For a particular choice of $\gamma^{(n)}$ we can then vary $\sigma^{(n)}$ and compare the synthetically generated phase residuals with the empirical phase residuals.
+and generate TOAs and phase residuals.
+We calibrate the noise amplitude $\sigma^{(n)}$  to generate phase residuals that qualitatively resemble empirical phase residuals measured from real pulsars.
 We obtain empirical phase residuals from the NANOGrav 12.5 year narrowband data set \cite{nanograv_narrowband_DR}.
-For the purposes of validating the $\sigma^{(n)}$ inferred from Eq \ref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real solutions visually.
-Whilst it would be possible to e.g.
+For the purposes of validating the $\sigma^{(n)}$ inferred from Eq \ref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually.
+Whilst it would also be possible to e.g.
 define some quantitative loss function (e.g.
-mean-squared error) between the two solutions and solve the optimisation problem to infer $\sigma^{(n)}$, for our purposes this is overkill and it is satisfactory to simply qualitatively compare the two solutions as a useful sanity check.
-We emphasise that we are not overly concerned with getting the maximally accurate values for $\sigma^{(n)}$, but just instead some reasonable values for constructing representative synthetic data.
+mean-squared error) between the two solutions and solve the optimisation problem to infer $\sigma^{(n)}$, for our purposes this is excessive.
+It is satisfactory to simply qualitatively compare the two solutions as a useful sanity check.
+We emphasise that we are not overly concerned with calculating the maximally accurate values for $\sigma^{(n)}$, but instead some reasonable values for constructing representative synthetic data.
 This approach of generating synthetic phase solutions and visually comparing with empirical solutions is also followed in \cite{Vargas}.
@@
 We can see that for all pulsars the synthetic and empirical residuals are qualitatively similar with the magnitude of the residuals being directly comparable for each pulsar.
-   \begin{figure}
+%   \begin{figure}
+%	\centering
+%	\subfloat[PSR J0030+0451]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-a}}
+%	
+%	\subfloat[PSR J0340+4130 ]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-b}}
+%	
+%	\subfloat[PSR J0613-0200]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-c}}
+%
+%
+%\end{figure}
+
+\begin{figure*}%
 	\centering
-	\subfloat[PSR J0030+0451]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-a}}
+	\subfloat[\centering $f_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/example_residuals_plot} }}%
+	\qquad
+	\subfloat[\centering $\dot{f}_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/example_residuals_plot} }}%
+	\qquad
+	\subfloat[\centering $\sigma^{(n)}$]{{\includegraphics[width=0.3\textwidth]{images/example_residuals_plot} }}%
 
-	\subfloat[PSR J0340+4130 ]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-b}}
-
-	\subfloat[PSR J0613-0200]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-c}}
 	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars.
@@
 For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
-\textcolor{red}{TK: these plots need updating, but need to discuss with Andres re use of baboo}}
+\textcolor{red}{TK: these plots need updating, but need to discuss with Andres re use of baboo.
+Maybe just one is sufficient just as an example.}}
 	\label{fig:qualitative_compare}
-\end{figure}
-
-%
-%\textbf{phase solution from the UTMOST pulsar timing program with the Mongolo Observatory Synthesis Telescope  \citep{UTMOST}. I}
-%
+\end{figure*}
 
 \subsubsection{Synthetic data generation}
-We generate $N$ synthetic noisy timeseries by integrating the state equations, Eq \ref{eq:spinevol}- \ref{eq:xieqn} numerically for the synthetic PTA described in Sectin \ref{sec:synt_pta}, mapping from state space to measurement space via the measurement equation, Eq.
-\ref{eq:measurement}, and finally adding to each timeseries some zero-mean Gaussian noise with covariance $\sigma_{\rm m}^2$.
+We generate $N$ synthetic noisy timeseries - one for each pulsar - as follows:
+
+\begin{enumerate}
+	\item Integrate the state equations, Eq \ref{eq:spinevol}- \ref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
+	\item Map from state space to measurement space via the measurement equation, Eq. \ref{eq:measurement}
+	\item Add zero-mean Gaussian measurement noise to each timeseries, i.e.
+$f_{\rm m} ^{(n)} + N_{\rm f}^{(n)}$ where 	$\langle N_{\rm f}(t) N_{\rm f}(t') \rangle = \sigma_{\rm m}^2 \delta(t - t')$ for covariance $\sigma_{\rm m}^2$.
+\end{enumerate}
 Equations \ref{eq:spinevol}- \ref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
@@
 For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
-We take $\sigma_{\rm m}^2$ to be known \textit{a priori} rather than a parameter to be inferred.
-The measurement noise can be approximately related to the uncertainty in the pulse TOA $\sigma_{\rm TOA}$ as
+The measurement noise can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
 \begin{equation}
-	\sigma_{\rm m} = f_{\rm p}^{(n)} \frac{\sigma_{\rm TOA}}{\text{cadence}} \ . 
+	\sigma_{\rm m} \sim f_{\rm p}^{(n)} \frac{\sigma_{\rm TOA}}{\text{cadence}} \ . 
 \end{equation}
-For a millisecond pulsar with $f_{\rm p}^{(n)}$ observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$ s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
-Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$ s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
+Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$ and it to be known \textit{a priori} rather than a parameter to be inferred.
+Note that $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar; whilst for our construction every pulsar is sampled at the same time and so would experience the same realisation of the noise, in actual pulsar astronomy different pulsars will be sampled at different times and so subject to different realisations of the measurement noise.
+\newline
+
 In practice, in order to avoid numerical precision issues that arise when solving Eq \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
@@
 We initially characterise our method with a single example of a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are shown in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}.
+The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected arbitrarily, subject to being astrophysically reasonable and representative.
+The injected $\boldsymbol{\theta}_{\rm gw}$ are summarised in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}.
 The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$ are as described in Section \ref{sec:synt_pta} and are also shown in Table \ref{tab:parameters_and_priors}.
-For the Bayesian methods used in this work, one must select reasonable priors for the complete set complete of static parameters, $\pi(\boldsymbol{\theta})$.
+
+\subsubsection{Prior distributions}\label{sec:priors}
+
+For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters.
 For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
 The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires some additional discussion.
+\newline
+
 The parameters which govern the deterministic evolution of the pulsar spin frequency, $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are well-determined by existing radio timing observations.
@@
 Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 1 \%$ either side of the true value.
-In this way we can test how well our method performs without requiring very precise measurements of the pulsar parameters to be made \textit{a priori}.
+In this way we can test how well our method performs without requiring exceptionally precise and accurate measurements of the pulsar parameters to be made \textit{a priori}.
 Instead we consider the pulsar parameters similarly to the GW source parameters and estimate them optimally within a consistent framework.
-This counteracts the potential of a confusion problem c.f.
-the influence of timing noise and gravitational waves on the received pulsar signal.
+\newline
+
 The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}.
-However, for this work whilst the pulsar distance is used when generating synthetic data, as discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d$.
-\textcolor{red}{TK: this next bit on gamma needs cleaning up}.
-For setting a prior on $\gamma^{(n)}$ since  $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$, $\gamma^{(n)}$ is in some sense ``unobservable" over these timescales and it is sufficient to set it at its true value i.e consider $\gamma$ to be known.
-We briefly explored a LogUniform prior of $\gamma$ over e.g.
-LogUniform($10^{-10}, 10^{-15}$) but the results are unchanged.
-In order to reduced the computational load we therefore set $\gamma$ to its true value.
-The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22}$.
-The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so the $\sigma^{(n)}$ inferred from Eq.
-\ref{eq:sigmap} \ref{eq:sigmap_f} $\sim 10^{-16}$.
-For the majority of pulsars we set an uninformative broad prior of $\sigma^{(n)}$ of LogUniform($10^{-19}, 10^{-23}$).
-For PSR J1939+2134 we use LogUniform($10^{-15}, 10^{-17}$).
-Since we are not setting priors on $\gamma$ or $d$ this reduces our dimensionality to $7 + 3N$.
-We will use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space.
-All static parameters and their priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
+However for this work, whilst the pulsar distance is used when generating synthetic data, as discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d$.
+Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
+That is, for $T_{\rm obs}$ =10 years the solution of Eq \ref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ is satisfied.
+It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value.
+We briefly explored setting an uninformative prior on $\gamma$ over e.g.
+LogUniform($10^{-10}, 10^{-15}$) as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g.
+set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
+The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22}$ as calculated from Eqs.
+\ref{eq:sigmap}, \ref{eq:sigmap_f}.
+For these pulsars we set an uninformative broad prior of LogUniform($10^{-19}, 10^{-23}$).
+The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16}$.
+For PSR J1939+2134 we set the prior at LogUniform($10^{-15}, 10^{-17}$).
+\newline
+
+Since we are not setting priors on $\gamma$ or $d$ this reduces the dimensionality of the parameter space to $7 + 3N$.
+We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space.
+Explicitly,
+\begin{equation}
+	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
+\end{equation}
+All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
 \begin{table}
@@
 	\resizebox{\columnwidth}{!}{%
+	\renewcommand{\arraystretch}{1.5} % Default value: 1
 	\begin{tabular}{lccll}
@@
 		\hline
-		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
+		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
 	  & $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
@@
 		\hline
-		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ANTF}^{(n)}$ & Hz & Uniform(0.9$f_{\rm ANTF}^{n}$, $1.1 f_{\rm ANTF}^{n}$) \\
-		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ANTF}$ & s$^{-2}$ & Uniform(0.9$\dot{f}_{\rm ANTF}^{n}$, $1.1 \dot{f}_{\rm ANTF}^{n}$) \\
-		&  $d^{(n)}$       &$d_{\rm ANTF}^{(n)}$  & m & - \\
+		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform(0.9$f_{\rm ATNF}^{n}$, $1.1 f_{\rm ATNF}^{n}$) \\
+		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform(0.9$\dot{f}_{\rm ATNF}^{n}$, $1.1 \dot{f}_{\rm ATNF}^{n}$) \\
+		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
 		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & ? & LogUniform*($10^{-19}, 10^{-23}$) \\
@@
 	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter.
-The subscript ``ANTF" denotes values which have been obtained from the ANTF pulsar catalogue as described in Section \ref{sec:synt_pta}.
+The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
 The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Eq.
-\ref{eq:sigmap} \ref{eq:sigmap_f}.
-All pulsars use the prior on $\sigma^{(n)}$ as specificed in table, with the exception of PSR J1939+2134 which uses LogUniform($10^{-15}, 10^{-17}$)}
+\ref{eq:sigmap}, \ref{eq:sigmap_f} from \protect \cite{Shannon2010}.
+All pulsars use the prior on $\sigma^{(n)}$ as specified in the Table, with the exception of PSR J1939+2134 which uses LogUniform($10^{-15}, 10^{-17}$).}
 	\label{tab:parameters_and_priors}
@@
 \subsubsection{Parameter estimation}\label{sec:parameter_estim}
+
 From Eq.
 \ref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms.
-The first term $\cos(-\Omega t + \Phi_0)$ depends only on the GW source parameters and is shared across all pulsars.
+The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
 The argument of the cosine corresponds to the GW phase at the observer on Earth.
-Conversely, the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$ varies between pulsars.
+Conversely the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
 The argument of the cosine corresponds to the GW phase at each the individual pulsar.
@@
 In this work we follow the standard analysis approach and drop the pulsar terms from our model.
+Explicitly the measurement equation used in the Kalman filter is
+\begin{equation}
+		f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) \ , 
+		\label{eq:measuremen_earth}
+	\end{equation}
+	with
+	\begin{equation}
+		g^{(n)}_{\rm Earth}(t) = 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \cos(-\Omega t +\Phi_0)  \ .
+		\label{eq:g_func_trig_earth}
+	\end{equation}
 	We defer the inclusion of the pulsar terms in the measurement equation to a future work.
-This choice reduces the dimensionality of the parameter estimation problem to $7 + 4N$ since the measurement equation Eq \ref{eq:measurement} is no longer a function of the pulsar distance.
+As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation Eq \ref{eq:measurement} is no longer a function of the pulsar distance.
 We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
@@
 Initially we consider a single noise realisation of the synthetic data for the representative example.
-We apply the Kalman filter using the earth terms only measurement equation in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters, which are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:corner_plot_1} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ in Figure \ref{fig:example}.
+We apply the Kalman filter using the Earth terms only measurement equation, Eq.
+\ref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
+The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:corner_plot_1} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ in Figure \ref{fig:example}.
 It is evident that for this representative example we are able to accurately estimate all the static parameters of the system.
-Of particular interest are the estimates of $\boldsymbol{\theta}_{\rm gw}$.
-Whilst some parameters are recovered identically (e.g.
-$\omega, \Phi_0, \delta, \iota$) for others the posterior median is shifted away from the injected value (e.g.
-$\psi, \alpha$).
-For a single realisation of the noise it is not clear if this is a systematic effect, i.e.
+Regarding the estimates of $\boldsymbol{\theta}_{\rm gw}$, whilst all the parameters are generally recovered accurately, for some the posterior median is almost identically equal to the injected value  (e.g.
+$\Omega, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the  injected value (e.g.
+$\psi, \alpha$), whilst still being well within the distribution uncertainty.
+For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
 a bias in the parameter estimate, or else just a random variation.
@@
 Regarding $\boldsymbol{\theta}_{\rm psr, reduced}$, the frequency parameters $f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1)$ are generally well recovered for all pulsars, as are the noise amplitude parameters $\sigma^{(n)}$.
+
 \begin{figure*}
-	\includegraphics[width=0.8\textwidth]{images/1237_broad_GW}
+	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/1237_broad_GW}
 	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for a single realisation of the system noise.
@@
 \end{figure*}
+
 \begin{figure*}%
 	\centering
-	\subfloat[\centering label 1]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_f} }}%
+	\subfloat[\centering $f_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_f} }}%
 	\qquad
-	\subfloat[\centering label 2]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_fdot} }}%
+	\subfloat[\centering $\dot{f}_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_fdot} }}%
 	\qquad
-	\subfloat[\centering label 2]{{\includegraphics[width=0.3\textwidth]{images/1237_broad_sigma_p} }}%
+	\subfloat[\centering $\sigma^{(n)}$]{{\includegraphics[width=0.3\textwidth]{images/1237_broad_sigma_p} }}%
 
-	\caption{As Fig \ref{fig:corner_plot_1} but for the $N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$. We show only a selection of the first 5 pulsars rather than the full PTA.}
+	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$. We show only a selection of the first 5 pulsars rather than the full PTA.}
 	\label{fig:example}%
@@
 
+We frame the detection problem in terms of model selection as described in Section \ref{sec:model_selection}
+
+The variance of the measurement noise $\Sigma$ can be estimated approximately from the root-mean-square amplitude of phase residuals and we do not treat it as a parameter to be inferred in this work.
+
 \subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
@@
 
-\subsubsection{Exploring a broader parameter space}
+\subsection{Exploring a broader parameter space} \label{sec:parameter_space}
 

### 2023-08-07 docs/manuscript/paper.tex
canonical example with a fixed h prior

@@
 \begin{equation}
-	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ .
+	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
 \end{equation}
@@
 
-\begin{figure*}%
-	\centering
-	\subfloat[\centering $f_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/example_residuals_plot} }}%
-	\qquad
-	\subfloat[\centering $\dot{f}_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/example_residuals_plot} }}%
-	\qquad
-	\subfloat[\centering $\sigma^{(n)}$]{{\includegraphics[width=0.3\textwidth]{images/example_residuals_plot} }}%
-
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/example_residuals_plot}
 	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars.
@@
 	\label{fig:qualitative_compare}
-\end{figure*}
+\end{figure}
+
+%
+%\begin{figure*}%
+%	\centering
+%	\subfloat[\centering $f_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/example_residuals_plot} }}%
+%	\qquad
+%	\subfloat[\centering $\dot{f}_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/example_residuals_plot} }}%
+%	\qquad
+%	\subfloat[\centering $\sigma^{(n)}$]{{\includegraphics[width=0.3\textwidth]{images/example_residuals_plot} }}%
+%	
+%	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars.
+The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}.
+The synthetic residuals are generated by numerically solving Eq.
+\ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Eqs.
+\ref{eq:sigmap}, \ref{eq:sigmap_f}.
+For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
+\textcolor{red}{TK: these plots need updating, but need to discuss with Andres re use of baboo.
+Maybe just one is sufficient just as an example.}}
+%	\label{fig:qualitative_compare}
+%\end{figure*}
 
@@
 \subsubsection{Detection}
+We frame the detection problem in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
 
-We frame the detection problem in terms of model selection as described in Section \ref{sec:model_selection}
+The Bayes factor $\beta$ as defined in Eq. \ref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative example system, where we now vary the magnitude of the GW strain, $h_0$.
 
-The variance of the measurement noise $\Sigma$ can be estimated approximately from the root-mean-square amplitude of phase residuals and we do not treat it as a parameter to be inferred in this work.
+The noise is an identical realisation for each value of $h_0$
+
+We take $\beta = 10$ as a tolerance cutoff above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
+
+For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3 \times 10^{-15}$
+
+The small perturbations in the general log-linear $\beta - h_0$ trend arise due to noise in the nested sampling method.
+For these points the sampler did not converge as well for $\mathcal{M}_1$ and so the evidence is lower than would be expected.
+This is purely a noise artefact: running the nested sampler again and recreating the $\beta - h_0$ leads to either no perturbations, or perturbations at different points in the curve.
+
+Below $\beta = 10$ this noise in the sampler starts to dominate and the hierarchical structure of the models $\mathcal{M}_0 , \mathcal{M}_1$ fails.
+That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds
+
+and the $\beta - h_0$ curve becomes much less smooth
+
+This is simply stating that at these small strains the system is completely noise dominated and the two models $\mathcal{M}_1, \mathcal{M}_0$ cannot be separated.
+
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/PaperBayesRatioPlotn1000_stacked}
+	\caption{}
+	\label{fig:bayes}
+\end{figure}
 
@@
 
+\subsection{Bias by dropping pulsar terms}
+
+some example likelihood curves.
+
+why are some affected but not others?
+
 \section{Discussion}
@@
 \url{https://github.com/stephengreen/lfi-gw}
+
+comparison with alternative methods
 

### 2023-08-07 py_src/priors.py

@@ -197,7 +197,8 @@ def set_prior_on_measurement_parameters(init_parameters,priors,measurement_model
 
         init_parameters["h"] = None
         #priors["h"] = bilby.core.prior.LogUniform(h/100.0, h*10.0, 'h')
-        priors["h"] = bilby.core.prior.LogUniform(1e-15, 1e-12, 'h')
+        #priors["h"] = bilby.core.prior.LogUniform(1e-15, 1e-12, 'h')
+        priors["h"] = h #experiment fixing $h$ to compare performance in $\iota$
 
     return init_parameters,priors 
 

### 2023-08-07 docs/manuscript/paper.tex
create PP plots, N=200

@@
 \subsubsection{Detection}
-We frame the detection problem in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-
-The Bayes factor $\beta$ as defined in Eq. \ref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative example system, where we now vary the magnitude of the GW strain, $h_0$.
-
-The noise is an identical realisation for each value of $h_0$
-
-We take $\beta = 10$ as a tolerance cutoff above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
-
-For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3 \times 10^{-15}$
-
+We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
+The Bayes factor, $\beta$, as defined in Eq.
+\ref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
+The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
+We can see that there is a general log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
+We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
+For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3 \times 10^{-15}$.
 The small perturbations in the general log-linear $\beta - h_0$ trend arise due to noise in the nested sampling method.
-For these points the sampler did not converge as well for $\mathcal{M}_1$ and so the evidence is lower than would be expected.
-This is purely a noise artefact: running the nested sampler again and recreating the $\beta - h_0$ leads to either no perturbations, or perturbations at different points in the curve.
-
-Below $\beta = 10$ this noise in the sampler starts to dominate and the hierarchical structure of the models $\mathcal{M}_0 , \mathcal{M}_1$ fails.
-That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds
-
-and the $\beta - h_0$ curve becomes much less smooth
-
-This is simply stating that at these small strains the system is completely noise dominated and the two models $\mathcal{M}_1, \mathcal{M}_0$ cannot be separated.
-
+For these points the sampler converged sub-optimally for $\mathcal{M}_1$ and so the evidence $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)$ is lower than would be expected.
+This is purely a noise artefact; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these perturbations appear at different values of $h_0$, or else are not present at all.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler also dampens these perturbations and smooths the $\beta - h_0$ curve, since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
+Below $\beta = 10$ this noise in the sampler starts to dominate and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
+That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+This is simply due to the fact that at these small strains the system is completely noise dominated and the two models $\mathcal{M}_1, \mathcal{M}_0$ cannot be separated.
 \begin{figure}
 	\includegraphics[width=\columnwidth]{images/PaperBayesRatioPlotn1000_stacked}
-	\caption{}
+	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M_1}$ (GW present in data) and $M_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
+The horizontal grey dashed line labels the detection tolerance cut-off of $\beta = 10$.
+The minimal detectable strain, below which $\beta < 10$, is $\sim 3 \times 10^{-15}$.}
 	\label{fig:bayes}
@@
 
+can we also get a sky map using the KF? 
 \subsection{Bias by dropping pulsar terms}

### 2023-08-07 py_src/batch.sh

@@ -1,202 +1,200 @@
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1e-15_model_earth_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1e-15_model_null_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.071519305237605e-15_model_earth_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.071519305237605e-15_model_null_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.1481536214968841e-15_model_earth_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.1481536214968841e-15_model_null_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.2302687708123812e-15_model_earth_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.2302687708123812e-15_model_null_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.3182567385564047e-15_model_earth_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.3182567385564047e-15_model_null_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.4125375446227554e-15_model_earth_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.4125375446227554e-15_model_null_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.5135612484362072e-15_model_earth_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.5135612484362072e-15_model_null_seed_1237.sh & 
-sbatch slurm_jobs/slurm_paper_bayes_ratios_n1000_h_1.6218100973589331e-15_model_earth_seed_1237.sh & 
[…]

### 2023-08-07 py_src/main_pp.py

@@ -1,0 +1,85 @@
+
+
+
+
+
+from system_parameters import SystemParameters
+from pulsars import Pulsars
+from synthetic_data import SyntheticData
+from model import LinearModel
+from kalman_filter import KalmanFilter
+from bilby_wrapper import BilbySampler
+from priors import priors_dict,bilby_priors_dict
+from bilby_wrapper import BilbyLikelihood
+
+import numpy as np
+import bilby
+import sys
+
+import logging 
+#logging.basicConfig()
+#logging.getLogger(name="KalmanGW").setLevel(logging.INFO)
+
+
+arg_name = sys.argv[1]  #reference name
+h        =  float(sys.argv[2]) #strain
+measurement_model =  sys.argv[3] #whether to use the H0 or H1 model
+seed = int(sys.argv[4]) #the seeding
+
+
+#Extra params for pp plot
+omega =  float(sys.argv[5]) 
+phi_0 =  float(sys.argv[6]) 
+psi   =  float(sys.argv[7]) 
+alpha =  float(sys.argv[8]) 
+delta =  float(sys.argv[9]) 
+
+
+from numba import jit, config
+
+
+if __name__=="__main__":
+    logger = logging.getLogger().setLevel(logging.INFO)
+   
+    import multiprocessing
+    multiprocessing.set_start_method("fork")
+
+
+    #Setup the system
+    P   = SystemParameters(h=h,σp=None,σm=1e-11,use_psr_terms_in_data=True,measurement_model=measurement_model,seed=seed,
+                            Ω=omega,Φ0=phi_0,ψ=psi,α=alpha,δ=delta) # define the system parameters as a dict. Todo: make this a class
+
+    PTA = Pulsars(P)                                       # setup the PTA
+    data = SyntheticData(PTA,P)                            # generate some synthetic data
+
+    #Define the model 
+    model = LinearModel(P)
+
+    #Initialise the Kalman filter
+    KF = KalmanFilter(model,data.f_measured,PTA)
+
+    #Run the KF once with the correct parameters.
+    #This allows JIT precompile
+    optimal_parameters = priors_dict(PTA,P)
+    model_likelihood, state_predictions,measurement_predictions = KF.likelihood(optimal_parameters)
+    logging.info(f"Ideal likelihood given optimal parameters = {model_likelihood}")
+    
+    #Bilby
+    init_parameters, priors = bilby_priors_dict(PTA,P)
+   
+
[…]

### 2023-08-07 py_src/priors.py
multiple noise realisations, shifted prior

@@ -196,8 +196,8 @@ def set_prior_on_measurement_parameters(init_parameters,priors,measurement_model
 
 
         init_parameters["h"] = None
-        #priors["h"] = bilby.core.prior.LogUniform(h/100.0, h*10.0, 'h')
-        priors["h"] = bilby.core.prior.LogUniform(1e-15, 1e-12, 'h')
+        priors["h"] = bilby.core.prior.LogUniform(h/100.0, h*10.0, 'h')
+        #priors["h"] = bilby.core.prior.LogUniform(1e-15, 1e-12, 'h')
         #priors["h"] = h #experiment fixing $h$ to compare performance in $\iota$
 
     return init_parameters,priors 

### 2023-08-08 py_src/bilby_wrapper.py
rerun the canonical example, with a large dlogz

@@ -34,7 +34,7 @@ def BilbySampler(KalmanFilter,init_parameters,priors,label,outdir):
 			                  sample='rwalk_dynesty',
                               check_point_plot=False,
                               npoints=1000,
-                              dlogz=1e-6,
+                              dlogz=10.0,
                               npool=32,
 			                  plot=False,resume=False)
 

### 2023-08-08 py_src/pulsars.py

@@ -23,7 +23,7 @@ class Pulsars:
         if SystemParameters.Npsr != 0:
             pulsars = pulsars.sample(SystemParameters.Npsr,random_state=SystemParameters.seed) #can also use  pulsars.head(N) to sample  
 
-        print("gamma cha4nged!")
+        
         #Extract the parameters
         self.f         = pulsars["F0"].to_numpy(dtype=NF)
         self.fdot      = pulsars["F1"] .to_numpy(dtype=NF)

### 2023-08-09 docs/manuscript/paper.tex
some manuscript updates
multiple noise realisations, shifted prior
rerun the canonical example, with a large dlogz
fixed h, multiple noise realisations, 1 core

@@
 	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
-	h_{\times} &= -2h\cos \iota 	\label{eq:hphx2} \ , 
+	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ , 
 \end{align}
@@
 For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
-\textcolor{red}{TK: these plots need updating, but need to discuss with Andres re use of baboo.
-Maybe just one is sufficient just as an example.}}
+\textcolor{red}{TK: potential sinusoidal feature?
+Need to discuss with Andres re use of baboo.}}
 	\label{fig:qualitative_compare}
@@
 \ref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
-The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:corner_plot_1} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ in Figure \ref{fig:example}.
+The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:example_psr_params} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ in Figure \ref{fig:example}.
 It is evident that for this representative example we are able to accurately estimate all the static parameters of the system.
@@
 
+need some discussion here since htey are not perfecty recovered...
+
 \begin{figure*}
-	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/1237_broad_GW}
+	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/representative_example_v2_GW}
 	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for a single realisation of the system noise.
@@
 	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$. We show only a selection of the first 5 pulsars rather than the full PTA.}
-	\label{fig:example}%
+	\label{fig:example_psr_params}%
 \end{figure*}
@@
 
-\subsubsection{Detection}
+\subsubsection{Detection} \label{sec:detection}
 We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
@@
 	\includegraphics[width=\columnwidth]{images/PaperBayesRatioPlotn1000_stacked}
-	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M_1}$ (GW present in data) and $M_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
+	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
 The horizontal grey dashed line labels the detection tolerance cut-off of $\beta = 10$.
@@
 
-The above discussion was just for a single realisaitn of the noise
+The above discussion was just for a single realisation of the noise processes.
+It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method.
+
+To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.
+
+For each realisation of the data we can then independently estimate the static parameters $\boldsymbol{\theta}$.
+
+The estimated values of $\boldsymbol{\theta}_{\rm gw}$ for 5 realisations of the noise are shown in Fig \ref{fig:corner_plot_2} which is analogous to Fig \ref{fig:corner_plot_1} but now shows multiple realisations of the noise.
+
+We can see that the results agree well with the case of a single noise realisation.
+The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different realisations, agreeing with the injected values with high fidelity.
+
+The variance in the estimates of $\iota$ and $h_0$ is much greater than for the other parameters.
+
+This is due to the nested sampler effectively trading off accuracy in one parameter against the other; from Eq \ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+For example, from Eq \ref{eq:hphx2} it is not clear if a larger $h_{\times}$ is due to a larger strain $h$ or a different inclination $\iota$.
+This degeneracy is broken by the relation for $h_{+}$ which does separate $h$ and $\iota$
+
+Whilst both parameters are then generally identifiable, the specific algebraic relation between the two and how these parameters get mapped into the measurement equation makes it difficult for our algorithm to estimate both simultaneously, instead typically trading off one parameter against the other.
+
+If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy.
+This is demonstrated in Fig X which is analogous to the representative exxample, but now we have fixed $h$ at its true value (i.e.
+reduced the prior to a delta function).
+WE can see that the method can now infer $\iota$ which much greater accuracy.
+
+Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.
+Fig \ref{fig:example_psr_params}) show a consistency between the single and multiple noise realisations though for brevity we do not show the results here \textcolor{red}{TK: is this fair?
+Don't want to clog up paper with extra figures that don't offer any real insight.}
+
+In the case where h is fixed...
+
+whilst the estimates on  are much more varied.
+
+In order to visualise the 1000 realisations, it is convenient to consider the distribution in the median of the posterior that is inferred for each static parameter, for each realisation of the noise.
+The resulting distributions are presented in Fig \ref{fig:median_distriubutins}
+
+The bias that was evident for some of the parameters
+
+  \textcolor{red}{with the estimate on iota}
+
+are entirely consistent with the single
+
+, which is analogous to Fig Y but with multiple  now shown.
+
+\begin{figure*}
+	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/stacked_GW_plot_1000}
+	\caption{}
+	\label{fig:corner_plot_2}
+\end{figure*}
+
+\begin{figure*}
+	\setkeys{Gin}{width=\linewidth}   
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics{images/distribution_omega_gw}
+		\caption{$\Omega$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics{images/distribution_phi0_gw}
+		\caption{$\Phi_0$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics{images/distribution_psi_gw}
+		\caption{$\psi$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics{images/distribution_iota_gw}
+		\caption{$\iota$}
+	\end{subfigure}
+
+	\medskip
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics{images/distribution_delta_gw}
+		\caption{$\delta$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics{images/distribution_alpha_gw}
+		\caption{$\alpha$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics{images/distribution_h}
+		\caption{$h$}
+	\end{subfigure}
+
+	\caption{some caption \textcolor{red}{TK: these plots are ugly. Is there a better way to show this?}}\label{fig:median_distriubutins}
+\end{figure*}
+
+to generate 1000 different datasets
+
+the above example was not simply a particualry
 
@@
 The estimates on $h$ and $\iota$ are generally less accurate than the other parameters of $\boldsymbol{\theta}_{\rm gw}$.
-
-From Eq \ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
-Whilst both parameters are generally identifiable the algebraic relation between the two makes it difficult for our algorithm to estimate both simultaneously.
-For example, from Eq \ref{eq:hphx2} it is not clear if a larger $h_{\times}$ is due to a larger strain $h$ or a different inclination $\iota$.
-This degeneracy is broken by the relation for $h_{+}$ which does separate $h$ and $\iota$ but the algebraic relation of how these parameters get mapped into the measurement equation tends for our algorithm to trade off one parameter against the other.
-If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy.
-This is demonstrated in Fig X which is analogous to the representative exxample, but now we have fixed $h$ at its true value (i.e.
-reduced the prior to a delta function).
-WE can see that the method can now infer $\iota$ which much greater accuracy.
 
@@
 
+this could also go in "Multiple noise realisations"
+
 \section{Discussion}
+
+single pulsar performance.
 
@@
 
+more stringent priors on f and fdot 
 Points to discuss:
@@
 	\item Non constant time sampling? Different pulsars sampled at different times...
+	\item pulsar terms obv
 \end{itemize}

### 2023-08-09 py_src/plotting.py

@@ -107,17 +107,37 @@ def plot_all(t,states,measurements,measurements_clean,predictions_x,predictions_
 def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges,axes_scales,savefig,logscale=False,title=None):
     plt.style.use('science')
 
-    # Opening JSON file
-    f = open(path)
-    
-    # returns JSON object as 
-    # a dictionary
-    data = json.load(f)
-    print("The evidence is:", data["log_evidence"])
-    f.close()
 
-    #Make it a dataframe. Nice for surfacing
-    df_posterior = pd.DataFrame(data["posterior"]["content"]) # posterior
+
+
+    print(path.split('.')[-1])
+    if path.split('.')[-1] == 'json':
+
+        # Opening JSON file
+        f = open(path)
+        
+        # returns JSON object as 
+        # a dictionary
+        data = json.load(f)
+        print("The evidence is:", data["log_evidence"])
+        f.close()
+
+        #Make it a dataframe. Nice for surfacing
+        df_posterior = pd.DataFrame(data["posterior"]["content"]) # posterior
+
+
+
+    else: #is a parquet gzip
+        df_posterior = pd.read_parquet(path) 
+
+
+
+    #Make omega into nHz
+    df_posterior["omega_gw"] = df_posterior["omega_gw"]*1e9
+    df_posterior["h"] = df_posterior["h"]*1e12
+
+    display(df_posterior)
+
 
 
     print("Number of samples:")
@@ -143,47 +163,55 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
     import warnings
     warnings.filterwarnings("error")
 
-    try:
-        print("running with increased label size")
-        fs = 20
-        fig = corner.corner(y_post, 
-                            color='C0',
-                            show_titles=True,
-                            smooth=True,smooth1d=True,
-                            truth_color='C2',
-                            quantiles=[0.16, 0.84],
-                            truths = injection_parameters,
-                            range=ranges,
-                            labels = labels,
-                            label_kwargs=dict(fontsize=fs),
-                            axes_scales = axes_scales)
-                
-
-        #Pretty-ify
-        for ax in fig.axes:
-
-            if ax.lines: #is anything plotted on this axis?
-                
-                ax.yaxis.set_major_locator(plt.MaxNLocator(4))
-                ax.yaxis.set_tick_params(labelsize=fs-6)
-                ax.xaxis.set_tick_params(labelsize=fs-6)
-
-            
+    #try:
+    
[…]

### 2023-08-09 py_src/priors.py

@@ -196,9 +196,9 @@ def set_prior_on_measurement_parameters(init_parameters,priors,measurement_model
 
 
         init_parameters["h"] = None
-        priors["h"] = bilby.core.prior.LogUniform(h/100.0, h*10.0, 'h')
+        #priors["h"] = bilby.core.prior.LogUniform(h/100.0, h*10.0, 'h')
         #priors["h"] = bilby.core.prior.LogUniform(1e-15, 1e-12, 'h')
-        #priors["h"] = h #experiment fixing $h$ to compare performance in $\iota$
+        priors["h"] = h #experiment fixing $h$ to compare performance in $\iota$
 
     return init_parameters,priors 
 

### 2023-08-10 docs/manuscript/paper.tex
example likelihood surfaces

@@
 \subsubsection{Measurement equation}
-\textcolor{red}{TK: Does this subsection needs a better name?} \newline 
-\noindent \textcolor{red}{TK: this subsection has been overhauled per comments from AM.
-The structure is now: \newline 1.
-Radio pulses described by general geometric object $p$ \newline 2.
-Shift in $p$ due to a GW (no derivation, just definition of terms) \newline 3.Relate shift in $p$ to a shift in $f$ by clarifying that $u=(1,0,0,0)$ for PSR and observer.
-\newline 4.
-Give final measurement equation} \newline
-
 In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
@@
 \newline
+
+In practice looser, leave them loose to subject method to more stringent test i.e. can we still estimate GW parameters without highly accurate t0 frequencies.
 
@@
 \subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
-
-The above discussion was just for a single realisation of the noise processes.
-It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method.
-
-To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.
-
-For each realisation of the data we can then independently estimate the static parameters $\boldsymbol{\theta}$.
-
-The estimated values of $\boldsymbol{\theta}_{\rm gw}$ for 5 realisations of the noise are shown in Fig \ref{fig:corner_plot_2} which is analogous to Fig \ref{fig:corner_plot_1} but now shows multiple realisations of the noise.
-
-We can see that the results agree well with the case of a single noise realisation.
-The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different realisations, agreeing with the injected values with high fidelity.
-
-The variance in the estimates of $\iota$ and $h_0$ is much greater than for the other parameters.
-
-This is due to the nested sampler effectively trading off accuracy in one parameter against the other; from Eq \ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
-For example, from Eq \ref{eq:hphx2} it is not clear if a larger $h_{\times}$ is due to a larger strain $h$ or a different inclination $\iota$.
-This degeneracy is broken by the relation for $h_{+}$ which does separate $h$ and $\iota$
-
-Whilst both parameters are then generally identifiable, the specific algebraic relation between the two and how these parameters get mapped into the measurement equation makes it difficult for our algorithm to estimate both simultaneously, instead typically trading off one parameter against the other.
-
-If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy.
-This is demonstrated in Fig X which is analogous to the representative exxample, but now we have fixed $h$ at its true value (i.e.
-reduced the prior to a delta function).
-WE can see that the method can now infer $\iota$ which much greater accuracy.
-
-Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.
-Fig \ref{fig:example_psr_params}) show a consistency between the single and multiple noise realisations though for brevity we do not show the results here \textcolor{red}{TK: is this fair?
-Don't want to clog up paper with extra figures that don't offer any real insight.}
-
-In the case where h is fixed...
-
-whilst the estimates on  are much more varied.
-
-In order to visualise the 1000 realisations, it is convenient to consider the distribution in the median of the posterior that is inferred for each static parameter, for each realisation of the noise.
-The resulting distributions are presented in Fig \ref{fig:median_distriubutins}
-
-The bias that was evident for some of the parameters
-
-  \textcolor{red}{with the estimate on iota}
-
-are entirely consistent with the single
-
-, which is analogous to Fig Y but with multiple  now shown.
-
 \begin{figure*}
@@
 \end{figure*}
-
+The above discussion was just for a single realisation of the noise processes.
+It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method.
+To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.
+For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
+In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for 9 realisations of the noise.
+This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}.
+We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
+We can see from Figure \ref{fig:corner_plot_2} that the results for multiple noise realisations agree well with the results that we saw in  Figure \ref{fig:corner_plot_1} for a single noise realisation.
+The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations i.e.
+the variance between the different marginalized posteriors is generally low.
+The estimates generally agree with the injected values with high fidelity.
+Conversely,for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise.
+\newline
+
+ The large degree of variance in the estimates of $\iota$ and $h_0$ is due to the nested sampling algorithm effectively trading off accuracy in one parameter against the accuracy in other.
+From Eqs.
+\ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+For example, from Eq \ref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
+This degeneracy is broken generally by the relation for $h_{+}$, Eq \ref{eq:hphx}.
+Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of reasonable values (\textcolor{red}{TK: Figure here would be useful}).
+Consequently it is difficult for our algorithm to select a particular point in this plateau.
+If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau surface and the variance in the estimates between different noise realisations decreases.
+This is demonstrated in Figure \ref{fig:iota}
+
+As we will see in Section \ref{sec:bias}
+
+ larger strain or
+
+ from
+
+ If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy. This is demonstrated in 
+for the inferred posteriors of $\iota$ for the same 5 noise realisations of Fig \ref{fig:corner_plot_2}, but now taking $h$ as a known parameter fixed at its true value.
+
+\begin{figure*}
+
+		\begin{subfigure}[b]{0.48\textwidth}
+		\includegraphics[width=\textwidth]{images/stacked_GW_plot_iota_h}
+		\caption{}
+		\label{fig:example1_iota}
+	\end{subfigure} 
+	\hfill
+	\begin{subfigure}[b]{0.48\textwidth}
+		\includegraphics[width=\textwidth]{images/stacked_GW_plot_iota_h_fixed}
+		\caption{}
+		\label{fig:example2_iota}
+	\end{subfigure}
+
+	\caption{Figure showing distribution of $\iota$ for a known/fixed $h$}
+	\label{fig:iota}
+\end{figure*}
+We can see... \newline
+
+Note that whilst the variance between the different noise realisations has decreased, the discrepancy between the estimated value and the true recovered value has increased.
+
+Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.
+Fig \ref{fig:example_psr_params}) show a consistency across multiple noise realisations, though for brevity we do not show the results here \textcolor{red}{TK: is this fair?
+Don't want to clog up paper with extra figures that don't offer any real insight.} \newline
+
+For each realisation of the noise we infer a marginalised 1-D posterior for each parameter.
+Each posterior has a median value.
+We plot the distribution of these medians over 1000 noise realisations for each parameter of $\boldsymbol{\theta}_{\rm gw}$ in Figure \ref{fig:median_distriubutins}.
+For each of these distributions over the noise realisations we can also calculate the median value (i.e.
+the median of the medians) and compare it with the true injection value.
+This is shown by the red and orange dashed lines respectively in the Figure.
 \begin{figure*}
 	\setkeys{Gin}{width=\linewidth}
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics{images/distribution_omega_gw}
+
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
 		\caption{$\Omega$}
@@
 	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics{images/distribution_phi0_gw}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
 		\caption{$\Phi_0$}
@@
 	\hfill	
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics{images/distribution_psi_gw}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
 		\caption{$\psi$}
 	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics{images/distribution_iota_gw}
+	\medskip
+
+		\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
 		\caption{$\iota$}
 	\end{subfigure}
-
-	\medskip
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics{images/distribution_delta_gw}
+	\hfill
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
 		\caption{$\delta$}
@@
 	\hfill	
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics{images/distribution_alpha_gw}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
 		\caption{$\alpha$}
 	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
+	\medskip
+
+	\medskip
+	\begin{subfigure}[b]{0.33\textwidth}
 		\includegraphics{images/distribution_h}
@@
 \end{figure*}
+Analogous to the results we saw in Figure \ref{fig:corner_plot_2}, the distributions of the median values of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are very narrow with a generally small discrepancy between the inferred value and the injected value.
+The parameters $\psi$ and $\alpha$ do exhibit a small bias of magnitude $\sim 0.1$ and $\sim 0.04$ respectively which we will discuss in Section \ref{sec:bias}.
 
@@
 
-The above was for a particular SBHB system
-
-can we also get a sky map using the KF? 
-\subsection{Bias by dropping pulsar terms}
+\textcolor{red}{TK: TBD}
+
+\subsection{Bias by dropping pulsar terms}\label{sec:bias}
 
@@
 
-single pulsar performance.
-
-hellings downs
-
-multiple sources
-
-spherical localisatin plot would be cool e.g.
-
-\url{https://github.com/stephengreen/lfi-gw}
-
-comparison with alternative methods
-
-more stringent priors on f and fdot 
-Points to discuss:
+Some potential further points for discussion
 
 \begin{itemize}
-	\item Non constant time sampling? Different pulsars sampled at different times...
-	\item pulsar terms obv
-\end{itemize}
-
-\section{Conclusion}
-
-Our method is not contingent on the choice of PTA, and naturally extend to other PTAs or PTAs with more pulsars.
-
-Questions
-
-\begin{itemize}
+	\item What is the performance of a single pulsar?
+	\item How does this map to Hellings-downs?
+	\item How does this map to having more than one source? Or a chrmomatic source?
+	\item Non constant time sampling?
+	\item Any further discussion on pulsar terms?
 	\item PTAs like MSPs for small timing noise. Can we get away with large timing noise, and use more pulsars? Some pulsars are more useful than others, e.g. arXiv 2211.03201
@@
 
+\section{Conclusion}
+
+\textcolor{red}{TK: TBD}
+
  \newpage 
@@
 \section{Kalman recursion equations} \label{sec:kalman}
+
+\textcolor{red}{TK: This section needs cleaning up}
 

### 2023-08-10 notebooks/plot_likelihood_surfaces.py

@@ -1,0 +1,57 @@
+
+
+import numpy as np
+import matplotlib.pyplot as plt  
+
+
+
+
+
+import matplotlib.pyplot as plt 
+fig = plt.figure(figsize=(6,6))
+ax = fig.add_subplot(111, projection='3d')
+
+path1 = 'tmp_likelihood_surface.npz'
+path2 = 'tmp_likelihood_surface_earth.npz'
+
+def load_and_plot(path,ax):
+
+    data = np.load(path)
+
+
+    iota_values = data['x']
+    h_values = np.log10(data['y'])
+    surface_pulsar = data['z']
+    iota_idx, h_idx = np.unravel_index(surface_pulsar.argmax(), surface_pulsar.shape)
+
+
+
+    xc = iota_values[iota_idx]
+    yc = h_values[h_idx]
+    zc = surface_pulsar[iota_idx,h_idx]
+
+
+    X,Y = np.meshgrid(h_values,iota_values)
+
+
+   
+    lx = len(iota_values)
+    ly = len(h_values)
+    z = np.reshape(surface_pulsar, (lx, ly))
+   
+    # Plot a 3D surface
+    ax.plot_surface(X, Y, z,alpha=0.5)
+
+
+    print(path, yc,xc,zc)
+    ax.scatter(yc,xc,zc, s=20)
+ 
+
+
+
+
+load_and_plot(path1,ax)
+load_and_plot(path2,ax)
+
+#ax.scatter(1e-12,1.0,-586119.6803021401,c='r', s=20)
+plt.show()

### 2023-08-14 docs/manuscript/paper.tex
added likelihood curves

@@
 \thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
-\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
+%\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
 \\
@@
 Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
-After multiple non-detections \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A} consilient evidence for the GW background was presented by NANOGrav \citep{2023ApJ...951L...8A}, EPTA/InPTA \citep{2023arXiv230616214A}, PPTA \citep{2023ApJ...951L...6R} and the CPTA \citep{2023RAA....23g5024X} \newline
+After multiple non-detections \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A} consilient evidence for the GW background was presented by NANOGrav \citep{2023ApJ...951L...8A}, EPTA/InPTA \citep{2023arXiv230616214A}, PPTA \citep{2023ApJ...951L...6R} and the CPTA \citep{2023RAA....23g5024X}.
+\newline
 
@@
 Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
-The new and existing approach are therefore complementary, but the new approach holds out the promise of somewhat higher sensitivity --- a promise which we aim to test as the key goal of this paper.
+The new and existing approach are therefore complementary, but the new approach holds out the promise of somewhat higher sensitivity.
 In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
@@
 
+% --- a promise which we aim to test as the key goal of this paper. 
 \noindent This paper is organised as follows.
 In Section \ref{sec:model} we present the state-space model for the pulsar pulse frequency which is subject the influence of a GW.
-In Section \ref{sec:detect} we discretise the model, develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence.
-\textcolor{red}{TK: this paper outline section will be completed once we have finalized the rest of the manuscript} In Section \ref{sec:testing} we test this new method on synthetic pulsar data.
+In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence.
+In Section \ref{sec:testing} we test this new method on synthetic pulsar data.
 Throughout this work we adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature.
@@
 
-Equations \ref{eq:spinevol}--\ref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend [ref latest results paper from each of the three main PTAs].
+Equations \ref{eq:spinevol}--\ref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend \citep{NANOgrav2023} \citep{EPTA2023} \citep{Zic2023arXiv230616230Z}.
 Equations \ref{eq:spinevol}--\ref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
@@
 \begin{equation}
-	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left(\frac{f_{\rm gw}(t=0)}{10^{-7} \mathrm{~Hz}}\right)^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right)
+	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left(\frac{f_{\rm gw}(t=0)}{10^{-7} \mathrm{~Hz}}\right)^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
 	\label{eq:f_evolution}
@@
 \end{align}
-Equations \ref{eq:measurement}, \ref{eq:g_func_trig}, \ref{eq:g_func}  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.
+Equations \ref{eq:measurement},\ref{eq:g_func} ,\ref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.
 
@@
 In Section \ref{sec:rep_example} we generate synthetic data for a single representative example SMBHB system and apply our method to estimate $\boldsymbol{\theta}$ and calculate the statistical evidence for a GW in the data via the Bayes factor.
-In this section we also explore how the detectability of the system changes for different magnitudes of $h_0$.
-In Section \ref{sec:parameter_space} we move beyond a single example system and explore a broader parameter space...
+In this section we also explore how the detectability of the system changes for different strain magnitudes, $h_0$ and the performance of the method for different realisations of the system noise.
+In Section \ref{sec:parameter_space} we move beyond a single example system and test our method across a broader parameter space.
+In Section \ref{sec:bias} we discuss biases that are evident in some of the parameter estimates.
 
@@
 We calibrate the noise amplitude $\sigma^{(n)}$  to generate phase residuals that qualitatively resemble empirical phase residuals measured from real pulsars.
-We obtain empirical phase residuals from the NANOGrav 12.5 year narrowband data set \cite{nanograv_narrowband_DR}.
+We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} from the  Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}.
 For the purposes of validating the $\sigma^{(n)}$ inferred from Eq \ref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually.
@@
 This approach of generating synthetic phase solutions and visually comparing with empirical solutions is also followed in \cite{Vargas}.
-In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for 3 NANOgrav pulsars: J0030+0451, J0340+4130 and J0613-0200.
-We can see that for all pulsars the synthetic and empirical residuals are qualitatively similar with the magnitude of the residuals being directly comparable for each pulsar.
-%   \begin{figure}
-%	\centering
-%	\subfloat[PSR J0030+0451]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-a}}
-%	
-%	\subfloat[PSR J0340+4130 ]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-b}}
-%	
-%	\subfloat[PSR J0613-0200]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-c}}
-%
-%
-%\end{figure}
-
+In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for the NANOGrav pulsar J0030+0451.
+We can see that the synthetic and empirical residuals are qualitatively similar.
 \begin{figure}
@@
 \end{figure}
-
-%
-%\begin{figure*}%
-%	\centering
-%	\subfloat[\centering $f_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/example_residuals_plot} }}%
-%	\qquad
-%	\subfloat[\centering $\dot{f}_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/example_residuals_plot} }}%
-%	\qquad
-%	\subfloat[\centering $\sigma^{(n)}$]{{\includegraphics[width=0.3\textwidth]{images/example_residuals_plot} }}%
-%	
-%	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars.
-The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}.
-The synthetic residuals are generated by numerically solving Eq.
-\ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Eqs.
-\ref{eq:sigmap}, \ref{eq:sigmap_f}.
-For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
-\textcolor{red}{TK: these plots need updating, but need to discuss with Andres re use of baboo.
-Maybe just one is sufficient just as an example.}}
-%	\label{fig:qualitative_compare}
-%\end{figure*}
-
 \subsubsection{Synthetic data generation}
@@
 	\item Add zero-mean Gaussian measurement noise to each timeseries, i.e.
-$f_{\rm m} ^{(n)} + N_{\rm f}^{(n)}$ where 	$\langle N_{\rm f}(t) N_{\rm f}(t') \rangle = \sigma_{\rm m}^2 \delta(t - t')$ for covariance $\sigma_{\rm m}^2$.
+$f_{\rm m} ^{(n)} + N_{\rm f}^{(n)}$ where 	$\langle N_{\rm f}(t) N_{\rm f}(t') \rangle = \sigma_{\rm m}^2 \delta(t - t')$ for measurement noise covariance $\sigma_{\rm m}^2$.
 \end{enumerate}
@@
 \end{equation}
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$ s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
-Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$ and it to be known \textit{a priori} rather than a parameter to be inferred.
-Note that $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar; whilst for our construction every pulsar is sampled at the same time and so would experience the same realisation of the noise, in actual pulsar astronomy different pulsars will be sampled at different times and so subject to different realisations of the measurement noise.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
+Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$ and take it to be known \textit{a priori} rather than a parameter to be inferred.
+Note that $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar; whilst for our construction every pulsar is sampled at the same time and so would experience the same realisation of the noise, in actual pulsar astronomy different pulsars will be sampled at different times and so subject to different noise realisations.
 \newline
@@
 \begin{equation}
-	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}((t-d))(1-g^{(n)}(t))
+	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right]
 	\label{eq:measurement_cov}
@@
 Instead we consider the pulsar parameters similarly to the GW source parameters and estimate them optimally within a consistent framework.
-\newline
-
-In practice looser, leave them loose to subject method to more stringent test i.e. can we still estimate GW parameters without highly accurate t0 frequencies.
+Leaving the priors looser in this way allows us to subject our method to a more stringent test and confirm that we still estimate GW parameters without highly accurate initial estimates of $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$.
+\newline
 
@@
 \begin{equation}
-	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
+	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ .
 \end{equation}
@@
 		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
-		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & ? & LogUniform*($10^{-19}, 10^{-23}$) \\
+		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform*($10^{-19}, 10^{-23}$) \\
 		& $\gamma$              & $10^{-13}$ & s$^{-1}$ & - \\
@@
 
-Initially we consider a single noise realisation of the synthetic data for the representative example.
+Initially we consider a single noise realisation of the synthetic data for the representative example system described in Table \ref{tab:parameters_and_priors}.
 We apply the Kalman filter using the Earth terms only measurement equation, Eq.
 \ref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
-The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:example_psr_params} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ in Figure \ref{fig:example}.
-It is evident that for this representative example we are able to accurately estimate all the static parameters of the system.
+The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:example_psr_params} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$.
+It is evident that for this representative example we are able to accurately estimate all the static parameters of the system using our method.
 Regarding the estimates of $\boldsymbol{\theta}_{\rm gw}$, whilst all the parameters are generally recovered accurately, for some the posterior median is almost identically equal to the injected value  (e.g.
@@
 Regarding $\boldsymbol{\theta}_{\rm psr, reduced}$, the frequency parameters $f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1)$ are generally well recovered for all pulsars, as are the noise amplitude parameters $\sigma^{(n)}$.
-
-need some discussion here since htey are not perfecty recovered...
-
+\textcolor{red}{TK: Could do with some extra discussion here re $\boldsymbol{\theta}_{\rm psr, reduced}$ since they are not perfectly recovered, but the important thing is that the GW parameters and the GW parameters are the interesting bit...}
 \begin{figure*}
@@
 
-	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$. We show only a selection of the first 5 pulsars rather than the full PTA.}
+	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$.
+We show only a selection of the first 5 pulsars rather than the full PTA.
+\textcolor{red}{TK: these figures are a little small to resolve nicely}}
 	\label{fig:example_psr_params}%
 \end{figure*}
-
-%
-%
-%
-%
-%\begin{figure}
-%	\includegraphics[width=\columnwidth]{images/1237_broad_f}
-%	\caption{some caption}
-%	\label{fig:corner_plot_2}
-%\end{figure}
-%
-%\begin{figure}
-%	\includegraphics[width=\columnwidth]{images/1237_broad_fdot}
-%	\caption{some caption}
-%	\label{fig:corner_plot_3}
-%\end{figure}
-%
-%
-%
-%\begin{figure}
-%	\includegraphics[width=\columnwidth]{images/1237_broad_sigma_p}
-%	\caption{some caption}
-%	\label{fig:corner_plot_4}
-%\end{figure}
 
@@
 This is simply due to the fact that at these small strains the system is completely noise dominated and the two models $\mathcal{M}_1, \mathcal{M}_0$ cannot be separated.
+\textcolor{red}{TK: Rather than discussing the bumps, maybe just run it again a few times and remove the bumps.}
 \begin{figure}
@@
 The horizontal grey dashed line labels the detection tolerance cut-off of $\beta = 10$.
-The minimal detectable strain, below which $\beta < 10$, is $\sim 3 \times 10^{-15}$.}
+The minimal detectable strain, below which $\beta < 10$, is $\sim 3 \times 10^{-15}$.
+\textcolor{red}{TK: probably need to regenerate this figure to get rid of the "bumps" which are just artefacts from a poor sampler convergence at that point}}
 	\label{fig:bayes}
@@
 	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/stacked_GW_plot_1000}
-	\caption{}
+	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
+The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
+The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance.}
 	\label{fig:corner_plot_2}
+\end{figure*}
+\begin{figure*}
+	\centering
+	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
+	\caption{1D marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$.
+The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column).
+The thick orange line labels the true injected value.
+Fixing $h$ reduces the variance between different noise realisations but results in a clear bias between the true and inferred values of $\iota$.}
+	\label{fig:iota}
+\end{figure*}
+\begin{figure*}
+	\setkeys{Gin}{width=\linewidth}
+
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
+		\caption{$\Omega$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
+		\caption{$\Phi_0$}
+	\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
+		\caption{$\psi$}
+	\end{subfigure}
+	\medskip
+
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
+		\caption{$\iota$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
+		\caption{$\delta$}
+	\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
+		\caption{$\alpha$}
+	\end{subfigure}
+	\medskip
+
+	\medskip
+	\begin{subfigure}[b]{0.33\textwidth}
+		\includegraphics{images/distribution_h}
+		\caption{$h$}
+	\end{subfigure}
+
+	\caption{Distribution in the median values of the 1D marginalized posteriors inferred by nested sampling for each static parameters of $\boldsymbol{\theta}_{\rm gw}$ over 1000 noise realisations.
+The orange dashed line labels the true injection value and the red dashed line labels the median of the distribution.
+\textcolor{red}{TK: these plots are ugly.
+Is there a better way to show this?}}\label{fig:median_distriubutins}
 \end{figure*}
@@
 The estimates generally agree with the injected values with high fidelity.
-Conversely,for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise.
-\newline
+Conversely, for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise.
+Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.
+Fig \ref{fig:example_psr_params}) show a consistency across multiple noise realisations, though for brevity we do not show the results here \textcolor{red}{TK: is this fair?
+Don't want to clog up paper with extra figures that don't offer any additional insight...} \newline
 
@@
 From Eqs.
-\ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+\ref{eq:hphx}, \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
 For example, from Eq \ref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
 This degeneracy is broken generally by the relation for $h_{+}$, Eq \ref{eq:hphx}.
-Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of reasonable values (\textcolor{red}{TK: Figure here would be useful}).
+Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of physically reasonable values (\textcolor{red}{TK: Could include a figure here showing this, but don't want to overclutter}).
 Consequently it is difficult for our algorithm to select a particular point in this plateau.
-If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau surface and the variance in the estimates between different noise realisations decreases.
-This is demonstrated in Figure \ref{fig:iota}
-
-As we will see in Section \ref{sec:bias}
-
- larger strain or
-
- from
-
- If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy. This is demonstrated in 
-for the inferred posteriors of $\iota$ for the same 5 noise realisations of Fig \ref{fig:corner_plot_2}, but now taking $h$ as a known parameter fixed at its true value.
-
-\begin{figure*}
-
-		\begin{subfigure}[b]{0.48\textwidth}
-		\includegraphics[width=\textwidth]{images/stacked_GW_plot_iota_h}
-		\caption{}
-		\label{fig:example1_iota}
-	\end{subfigure} 
-	\hfill
-	\begin{subfigure}[b]{0.48\textwidth}
-		\includegraphics[width=\textwidth]{images/stacked_GW_plot_iota_h_fixed}
-		\caption{}
-		\label{fig:example2_iota}
-	\end{subfigure}
-
-	\caption{Figure showing distribution of $\iota$ for a known/fixed $h$}
-	\label{fig:iota}
-\end{figure*}
-We can see... \newline
-
-Note that whilst the variance between the different noise realisations has decreased, the discrepancy between the estimated value and the true recovered value has increased.
-
-Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.
-Fig \ref{fig:example_psr_params}) show a consistency across multiple noise realisations, though for brevity we do not show the results here \textcolor{red}{TK: is this fair?
-Don't want to clog up paper with extra figures that don't offer any real insight.} \newline
+If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the variance in the estimates between different noise realisations decreases.
+This is demonstrated in Figure \ref{fig:iota} where we plot the marginalized 1D posterior of $\iota$ for 9 realisations of the noise in both the original case where $h_0$ is a free parameter (i.e.
+Figure \ref{fig:corner_plot_2}) and in the case where we hold $h_0$ fixed at its injection value, $h_0 = 10^{-12}$.
+We can see that whilst the variance between the individual posteriors is large in the original case, once we fix $h_0$ the variance decreases dramatically and the sampling algorithm repeatedly converges consistent estimates of $\iota$.
+There is a notable bias in Figure \ref{fig:iota} in the estimates for $\iota$ inferred for the fixed $h_0$ case relative to the true injected value; whilst the variance has decreased, the mean accuracy of the estimates has also decreased.
+This bias is a result of dropping the pulsar terms from our measurement equation and is discussed in Section \ref{sec:bias}.
+\newline
 
 For each realisation of the noise we infer a marginalised 1-D posterior for each parameter.
-Each posterior has a median value.
-We plot the distribution of these medians over 1000 noise realisations for each parameter of $\boldsymbol{\theta}_{\rm gw}$ in Figure \ref{fig:median_distriubutins}.
-For each of these distributions over the noise realisations we can also calculate the median value (i.e.
+In turn, each posterior distribution has a median value.
+We plot the distribution of these medians over the 1000 noise realisations for each parameter of $\boldsymbol{\theta}_{\rm gw}$ in Figure \ref{fig:median_distriubutins}.
+For each of these distributions we can also calculate the median value (i.e.
 the median of the medians) and compare it with the true injection value.
 This is shown by the red and orange dashed lines respectively in the Figure.
+Analogous to the results we saw in Figure \ref{fig:corner_plot_2}, the distributions of the median values of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are very narrow, with a generally small discrepancy between the inferred value and the injected value.
+Conversely, the distributions for $\iota$ and $h_0$ are much broader and exhibit a bias of $\sim 0.18$ radians and $0.17 \times 10^{-12}$ respectively.
+This agrees with the results we discussed for the 9 noise realisations in Figure \ref{fig:corner_plot_2} where a bias in $\iota$ and $h_0$ was also observed.
+In addition to  $\iota$ and $h_0$  exhibiting a strong bias, the parameters $\psi$ and $\alpha$ also exhibit a small bias of magnitude $\sim 0.1$ and $\sim 0.04$ radians respectively.
+This bias is again a result of dropping the pulsar terms from the measurement equation and will be discussed in Section \ref{sec:bias}.
+
+\subsection{Exploring a broader parameter space} \label{sec:parameter_space}
+
+\textcolor{red}{TK: TBD.
+Maybe lets just change the sky position - it it too expensive to generate a sky map of minimum detectable strain.
+Given the biases I don't know if a PP plot quite makes sense at this stage...}
+
+\subsection{Bias by dropping pulsar terms}\label{sec:bias}
+\textcolor{red}{TK: TBD. Will show some likelihood plots and how the likelihood optima does not coincide with the injection value when the pulsar term is dropped.}
+
 \begin{figure*}
@@
 	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
+		\includegraphics[width=\textwidth]{example-image-a}
 		\caption{$\Omega$}
@@
 	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
+		\includegraphics[width=\textwidth]{example-image-a}
 		\caption{$\Phi_0$}
@@
 	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
+		\includegraphics[width=\textwidth]{example-image-a}
 		\caption{$\psi$}
@@
    \medskip
-
-		\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
+	\begin{subfigure}[b]{0.45\textwidth}
+		\includegraphics[width=\textwidth]{example-image-a}
 		\caption{$\iota$}
@@
 	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
+	\begin{subfigure}[b]{0.45\textwidth}
+		\includegraphics[width=\textwidth]{example-image-a}
 		\caption{$\delta$}
 	\end{subfigure}
+	\medskip
+	\begin{subfigure}[b]{0.45\textwidth}
+		\includegraphics[width=\textwidth]{example-image-a}
+		\caption{$\alpha$}
+	\end{subfigure}
 	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
-		\caption{$\alpha$}
-	\end{subfigure}
-	\medskip
-
-	\medskip
-	\begin{subfigure}[b]{0.33\textwidth}
-		\includegraphics{images/distribution_h}
+	\begin{subfigure}[b]{0.45\textwidth}
+		\includegraphics{example-image-a}
 		\caption{$h$}
@@
 
-	\caption{some caption \textcolor{red}{TK: these plots are ugly. Is there a better way to show this?}}\label{fig:median_distriubutins}
+	\caption{}\label{fig:likelihood_curves}
 \end{figure*}
-Analogous to the results we saw in Figure \ref{fig:corner_plot_2}, the distributions of the median values of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are very narrow with a generally small discrepancy between the inferred value and the injected value.
-The parameters $\psi$ and $\alpha$ do exhibit a small bias of magnitude $\sim 0.1$ and $\sim 0.04$ respectively which we will discuss in Section \ref{sec:bias}.
-
-to generate 1000 different datasets
-
-the above example was not simply a particualry
-
-iota is also shifted.
-
-due to the degeneracy this impacts h
-
-The estimates on $h$ and $\iota$ are generally less accurate than the other parameters of $\boldsymbol{\theta}_{\rm gw}$.
-
-\subsection{Exploring a broader parameter space} \label{sec:parameter_space}
-
-\textcolor{red}{TK: TBD}
-
-\subsection{Bias by dropping pulsar terms}\label{sec:bias}
-
-some example likelihood curves.
-
-why are some affected but not others?
-
-this could also go in "Multiple noise realisations"
+
+%\begin{figure*}
+%	\centering
+%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}} 
+%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}} 
+%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}}
+%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}}
+%	\caption{(a) blah (b) blah (c) blah (d) blah}
+%	\label{fig:foobar}
+%\end{figure*}
 

### 2023-08-15 docs/manuscript/paper.tex
generate a skymap of bayes factors

@@
 \subsection{Bias by dropping pulsar terms}\label{sec:bias}
-\textcolor{red}{TK: TBD. Will show some likelihood plots and how the likelihood optima does not coincide with the injection value when the pulsar term is dropped.}
+Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (i.e.
+the median of the 1D marginalised posterior) is biased away from the true injection value.
+This effect is most severe for $\iota$ and $h_0$ (e.g.
+Figures \ref{fig:iota}, \ref{fig:median_distriubutins}) but also present to a lesser extent in other static parameters such as $\alpha$ and $\psi$.
+In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
+
+In Figure \ref{fig:likelihood_curves_earth} we plot the variation in the log-likelihood returned by the Kalman filter, Eq.
+\ref{eq:likelihood}, as we vary a single static parameter of $\boldsymbol{\theta}_{\rm gw}$, holding all other parameters constant at their true injected value.
+For instance, in Figure \ref{fig:omega_likelihood} we pass different values of $\Omega$ in the range $10^{-9}$ to $10^{-5}$ Hz into the Kalman filter, whilst holding the remaining parameters constant, and plot the retuned log-likelihood as a function of $\Omega$.
+The dashed orange line labels the true injection value whilst the dashed red line labels the location of the maximum log-likelihood.
+Finding the location of this maximum across all parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
+
+The key observation from Figure \ref{fig:likelihood_curves_earth} is that for many parameters the location of the likelihood optima and the location of the injected value are not the same, that is, the orange and red dashed lines are not coincident.
+
+This feature shows up as a bias in
+
+Taking a specific example, in Figure \ref{fig:iota_likelihood}
+Consequently
+
+Indeed this is a 1D representation of a multidimensinal problem and the exact shape of the likelihood curves depends on the
+
+Whilst the earth term likelihoo dcurves are generally smooth...
+
+If the effect is present in our analysis, also present in other analysis?
 
@@
 
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{example-image-a}
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_omega_earth}
 		\caption{$\Omega$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{example-image-a}
+		\label{fig:omega_likelihood}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_phi0_earth}
 		\caption{$\Phi_0$}
@@
 	\hfill	
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{example-image-a}
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_psi_earth}
 		\caption{$\psi$}
 	\end{subfigure}
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_iota_earth}
+		\caption{$\iota$}
+			\label{fig:iota_likelihood}
+	\end{subfigure}
 	\medskip
-	\begin{subfigure}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth]{example-image-a}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_delta_earth}
+		\caption{$\delta$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_alpha_earth}
+		\caption{$\alpha$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics{images/likelihood_h_earth}
+		\caption{$h$}
+	\end{subfigure}
+
+	\caption{}\label{fig:likelihood_curves_earth}
+\end{figure*}
+
+\begin{figure*}
+	\setkeys{Gin}{width=\linewidth}
+
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_omega_psr}
+		\caption{$\Omega$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_phi0_psr}
+		\caption{$\Phi_0$}
+	\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_psi_psr}
+		\caption{$\psi$}
+	\end{subfigure}
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_iota_psr}
 		\caption{$\iota$}
 	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth]{example-image-a}
+	\medskip
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_delta_psr}
 		\caption{$\delta$}
 	\end{subfigure}
-	\medskip
-	\begin{subfigure}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth]{example-image-a}
+	\hfill
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/likelihood_alpha_psr}
 		\caption{$\alpha$}
@@
 	\hfill
-	\begin{subfigure}[b]{0.45\textwidth}
-		\includegraphics{example-image-a}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics{images/likelihood_h_psr}
 		\caption{$h$}
@@
 
-	\caption{}\label{fig:likelihood_curves}
+	\caption{As Figure \ref{fig:likelihood_curves_earth} but now including the pulsar terms in the measurement equation.}\label{fig:likelihood_curves_psr}
 \end{figure*}

### 2023-08-15 py_src/batch.sh

@@ -1,200 +1,200 @@
-sbatch slurm_jobs/slurm_pp_plot_h_1e-12_model_earth_seed_1237_omega_3.238135383654817e-07_phi0_0.29064304869624064_psi_2.233533122609889_delta_0.576622103841384_alpha_2.8958273726369277.sh & 
-sbatch slurm_jobs/slurm_pp_plot_h_1e-12_model_earth_seed_1237_omega_4.175795004126788e-09_phi0_1.5244686525279478_psi_2.7584342099821373_delta_1.5334517538974524_alpha_2.0097595520876816.sh & 
-sbatch slurm_jobs/slurm_pp_plot_h_1e-12_model_earth_seed_1237_omega_3.244743719418169e-06_phi0_0.777693967993837_psi_0.7925015155291241_delta_1.3560844514447792_alpha_0.5290918000050544.sh & 
-sbatch slurm_jobs/slurm_pp_plot_h_1e-12_model_earth_seed_1237_omega_4.047059861771545e-07_phi0_0.4674671442124611_psi_2.8555279045246422_delta_0.47575415785695435_alpha_2.0517401400946036.sh & 
-sbatch slurm_jobs/slurm_pp_plot_h_1e-12_model_earth_seed_1237_omega_9.182369185646294e-08_phi0_0.9790417108514293_psi_2.8966457544826434_delta_1.3628314419241472_alpha_0.021859570913558307.sh & 
-sbatch slurm_jobs/slurm_pp_plot_h_1e-12_model_earth_seed_1237_omega_9.006757742139388e-09_phi0_1.132859862091618_psi_1.969321027083716_delta_1.367122743049036_alpha_1.4062992757161272.sh & 
-sbatch slurm_jobs/slurm_pp_plot_h_1e-12_model_earth_seed_1237_omega_4.932658186144806e-06_phi0_1.2309740586772084_psi_1.8873086260749967_delta_0.5996020135574363_alpha_2.147341131621517.sh & 
-sbatch slurm_jobs/slurm_pp_plot_h_1e-12_model_earth_seed_1237_omega_6.01712259589448e-09_phi0_0.3867075623942348_psi_1.7828074433800987_delta_0.7860397869262657_alpha_1.7694668905089181.sh & 
[…]

### 2023-08-15 docs/manuscript/paper.tex
some paper comments

@@
 \end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA.
-The stochastic torque is assumed to be a zero-mean white noise process.
+The stochastic torque is assumed to be a zero-mean, white noise process.
 Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
@@
 \end{equation}
-In Eq \ref{eq:frequency_evolution}, $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a proportionality constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
+In Eq \ref{eq:frequency_evolution}, $f_{\rm em}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
 \begin{equation}
@@
 
-Equations \ref{eq:spinevol}--\ref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend \citep{NANOgrav2023} \citep{EPTA2023} \citep{Zic2023arXiv230616230Z}.
+Equations \ref{eq:spinevol}--\ref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
 Equations \ref{eq:spinevol}--\ref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
@@
 A source can be considered as monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution of $1/T_{\rm obs}$.
-From Eq.
-\ref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans.
-The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA with pulsars at 1.5 kpc, 78 \% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array (SKA2) this fraction drops to 52 \% \citep[Fig 7 of ][]{Rosado10.1093/mnras/stv1098}.
+From Equation \ref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans.
+The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA with pulsars at 1.5 kpc, 78 \% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52 \%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
 We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
@@
 	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}} \nonumber \\
-	& +(\sin \psi \sin \theta) \boldsymbol{\hat{z}} \\
+	& +(\sin \psi \sin \theta) \boldsymbol{\hat{z}} \ , \\
 	\boldsymbol{l} = &(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
 	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}}\nonumber  \\
-	& +(\cos \psi \sin \theta) \boldsymbol{\hat{z}}
+	& +(\cos \psi \sin \theta) \boldsymbol{\hat{z}} \ ,
 \end{align}
@@
 {\sc tempo2} and related timing software.
-Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\vec{u}$ is given by the coordinate-independent expression:
+Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression:
 \begin{equation}
@@
 \begin{equation}
-	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
-\end{equation}
-where the perturbations to $\vec{u}$ from the GW are of higher order and neglected.
+	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0) \ ,
+\end{equation}
+where the perturbations to $\boldsymbol{u}$ from the GW are of higher order and neglected.
 In this case the shift in the momentum from Equation \ref{eq:momentum_shift} can be related to a shift in the frequency as,
 \begin{equation}
-	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t)
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t) \ ,
 	\label{eq:measurement}
@@
 \end{equation}
-It is will also prove instructive to express Eq. \ref{eq:g_func} in a trigonometric form as,
+It is will also prove instructive to express Equation \ref{eq:g_func} in a trigonometric form as,
 or in a trigonometric form 
@@
 \end{align}
-Equations \ref{eq:measurement},\ref{eq:g_func} ,\ref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.
+Equations \ref{eq:measurement} -- \ref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.
 
@@
 \end{equation}
-In Eq \ref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
+In Equation \ref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
 \begin{align}
@@
 In this work we use the linear Kalman filter which assumes a linear relation between $\boldsymbol{X}$ and $\boldsymbol{Y}$.
-Whilst the measurement equation, Eq.
-\ref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively.
+Whilst the measurement equation, Equation \ref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively.
 Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, the unscented Kalman filter \citep{882463van} or the particle filter \citep{Simon10}.
@@
 , M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$.
-The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Eq.
-\ref{eq:measurement}.
-The measurement predictions can then be compared aganst the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation".
+The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Equation \ref{eq:measurement}.
+The measurement predictions can then be compared against the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation".
 The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$.
@@
 If the estimates of the system parameters, $\boldsymbol{\hat{\theta}}$ , that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised.
-This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
+This is illustrated in Figure \ref{fig:6MB_BFS}; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
 The residuals correspond to random noise and are normally distributed.
@@
 \subsection{Nested Sampling}\label{sec:nested_sampling}
-We can use the likelihood returned by the Kalman filter, Eq \ref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
+We can use the likelihood returned by the Kalman filter, Equation \ref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
 \begin{equation}
@@
 In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work.
-Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Eq.
-\ref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \ref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
 It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
@@
 As a sanity check, we can also calculate $\sigma^{(n)}$ using a complementary numerical approach to the empirical best-fit model outlined above.
-We can directly solve the state equation Eq.
-\ref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a synthetic phase solution,
+We can directly solve the state equation Equation \ref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a synthetic phase solution,
 \begin{eqnarray}
@@
 We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} from the  Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}.
-For the purposes of validating the $\sigma^{(n)}$ inferred from Eq \ref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually.
+For the purposes of validating the $\sigma^{(n)}$ inferred from Equation \ref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually.
 Whilst it would also be possible to e.g.
@@
 The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}.
-The synthetic residuals are generated by numerically solving Eq.
-\ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Eqs.
+The synthetic residuals are generated by numerically solving Equation \ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations.
 \ref{eq:sigmap}, \ref{eq:sigmap_f}.
@@
 \begin{enumerate}
-	\item Integrate the state equations, Eq \ref{eq:spinevol}- \ref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
-	\item Map from state space to measurement space via the measurement equation, Eq. \ref{eq:measurement}
+	\item Integrate the state equations, Equations \ref{eq:spinevol}- \ref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
+	\item Map from state space to measurement space via the measurement equation, Equation \ref{eq:measurement}
 	\item Add zero-mean Gaussian measurement noise to each timeseries, i.e.
@@
 
-In practice, in order to avoid numerical precision issues that arise when solving Eq \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
+In practice, in order to avoid numerical precision issues that arise when solving Equations \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
 \begin{equation}
@@
 For synthetic data we can set $ f_{\rm em}^{*(n)} = f_{\rm em}^{(n)}$ but this is not true generally for real-world observations since the spin-down ephemeris is only known approximately, not exactly.
-The measurement equation Eq.
-\ref{eq:measurement} that relates the state and measurement variables is then updated as
+The measurement equation Equation \ref{eq:measurement} that relates the state and measurement variables is then updated as
 \begin{equation}
@@
 Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
-That is, for $T_{\rm obs}$ =10 years the solution of Eq \ref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ is satisfied.
+That is, for $T_{\rm obs}$ =10 years the solution of Equation \ref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ is satisfied.
 It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value.
@@
 set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
-The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22}$ as calculated from Eqs.
-\ref{eq:sigmap}, \ref{eq:sigmap_f}.
+The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22}$ as calculated from Equations \ref{eq:sigmap}, \ref{eq:sigmap_f}.
 For these pulsars we set an uninformative broad prior of LogUniform($10^{-19}, 10^{-23}$).
@@
 The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Eq.
-\ref{eq:sigmap}, \ref{eq:sigmap_f} from \protect \cite{Shannon2010}.
+The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \ref{eq:sigmap}, \ref{eq:sigmap_f} from \protect \cite{Shannon2010}.
 All pulsars use the prior on $\sigma^{(n)}$ as specified in the Table, with the exception of PSR J1939+2134 which uses LogUniform($10^{-15}, 10^{-17}$).}
@@
 \subsubsection{Parameter estimation}\label{sec:parameter_estim}
-
-From Eq.
-\ref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms.
+From Equation \ref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms.
 The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
@@
 	We defer the inclusion of the pulsar terms in the measurement equation to a future work.
-As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation Eq \ref{eq:measurement} is no longer a function of the pulsar distance.
+As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation Equation \ref{eq:measurement} is no longer a function of the pulsar distance.
 We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
@@
 Initially we consider a single noise realisation of the synthetic data for the representative example system described in Table \ref{tab:parameters_and_priors}.
-We apply the Kalman filter using the Earth terms only measurement equation, Eq.
-\ref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
+We apply the Kalman filter using the Earth terms only measurement equation, Equation \ref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
 The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:example_psr_params} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$.
@@
 We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-The Bayes factor, $\beta$, as defined in Eq.
-\ref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
+The Bayes factor, $\beta$, as defined in Equation \ref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
 The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
@@
 The large degree of variance in the estimates of $\iota$ and $h_0$ is due to the nested sampling algorithm effectively trading off accuracy in one parameter against the accuracy in other.
-From Eqs.
-\ref{eq:hphx}, \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
-For example, from Eq \ref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
-This degeneracy is broken generally by the relation for $h_{+}$, Eq \ref{eq:hphx}.
+From Equations \ref{eq:hphx}, \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+For example, from Equation \ref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
+This degeneracy is broken generally by the relation for $h_{+}$, Equation \ref{eq:hphx}.
 Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of physically reasonable values (\textcolor{red}{TK: Could include a figure here showing this, but don't want to overclutter}).
@@
 In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
-
-In Figure \ref{fig:likelihood_curves_earth} we plot the variation in the log-likelihood returned by the Kalman filter, Eq.
-\ref{eq:likelihood}, as we vary a single static parameter of $\boldsymbol{\theta}_{\rm gw}$, holding all other parameters constant at their true injected value.
+\newline
+
+In Figure \ref{fig:likelihood_curves_earth} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \ref{eq:likelihood}, as we vary a single static parameter of $\boldsymbol{\theta}_{\rm gw}$, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
 For instance, in Figure \ref{fig:omega_likelihood} we pass different values of $\Omega$ in the range $10^{-9}$ to $10^{-5}$ Hz into the Kalman filter, whilst holding the remaining parameters constant, and plot the retuned log-likelihood as a function of $\Omega$.
@@
 Finding the location of this maximum across all parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
-
-The key observation from Figure \ref{fig:likelihood_curves_earth} is that for many parameters the location of the likelihood optima and the location of the injected value are not the same, that is, the orange and red dashed lines are not coincident.
-
-This feature shows up as a bias in
-
-Taking a specific example, in Figure \ref{fig:iota_likelihood}
-Consequently
-
-Indeed this is a 1D representation of a multidimensinal problem and the exact shape of the likelihood curves depends on the
-
-Whilst the earth term likelihoo dcurves are generally smooth...
-
-If the effect is present in our analysis, also present in other analysis?
+The key observation from Figure \ref{fig:likelihood_curves_earth} is that for many parameters the location of the likelihood optima and the location of the injected value are not the same i.e.
+the orange and red dashed lines are not coincident.
+Consequently Bayesian likelihood inference techniques such as nested sampling settle in optima that do not necessarily align with the true injection value.
+This is the origin of the bias in the 1D marginalised posteriors that we observe.
+Taking a specific example, in Figure \ref{fig:iota_likelihood} we can see that the optima of the likelihood curve lies around $\iota \sim 0.7$ radians, whereas the injected value was $\iota = 1.0$ radians, This difference corresponds to the bias observed in the right-hand panel of Figure \ref{fig:iota} where the median of the marginalised posterior for $\iota$ was $\sim 0.65$ for an injection value of $\iota = 1.0$ radians.
+The discrepancy between $\iota \sim 0.7$ in the log-likelihood curve and $\iota \sim 0.65$ in the nested sampling inference is a result of the likelihood curves being a 1D representation of a multidimensinal problem and the exact shape of the likelihood curves depends on the values of all the other static parameters.
+That is to say, the likelihood curve for $\iota$ will have a different shape with a different optima if we set e.g.
+$\psi$ equal to its likelihood optima value rather than its true injection value.
+
+We can compare the likelihood curves of Figure \ref{fig:likelihood_curves_earth} generated using the earth -terms only measurement equation, Equations \ref{eq:measuremen_earth} \ref{eq:g_func_trig_earth}, with the likelihood curves derived using the measurement equation which includes the pulsar terms, Equation \ref{eq:measurement}.
+Whilst we have not used the nested sampling method together with the pulsar terms measurement equation in this  work, we can run the Kalman filter using Equation \ref{eq:measurement} and consider how the resulting likelihood varies for different parameter values.
+The results are shown in Figure \ref{fig:likelihood_curves_psr}.
+There are two interesting features in the pulsar-terms likelihood curves.
+The first is that the optima of the likelihood curves now align with the true injected values for all parameters.
+This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.The second is that the log-likelihood curves for $\Omega, \delta$ and $\alpha$ are much more noisy than in the Earth-terms case.
+Whilst clear optima exist above the noise, and whilst local to the true value the curve becomes much more smooth, this noise presents an additional challenge to typical likelihood inference techniques which perform best on smooth likelihoods with a single clear global optima.
+It affects these parameters in particular since Equation \ref{eq:g_func_trig} contains a phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, which is essentially a combination of these 3 parameters ($ \boldsymbol{n}$ is defined by $\delta$ and $\alpha$).
+When calculating a likelihood we are asking "How likely is this data, given these parameters".
+The inclusion of the pulsar terms results in effectively 2 continuous waves with different frequencies and phase.
+Perturbing e.g.
+$\alpha$ influences wave 1 in some non-liner way and wave 2 in some different non linear way.
+Whether this new set of parameters makes the data more likely depends on the particular combination of those two waves.
+\textcolor{red}{TK: this explanation of why these parameters is clunky because it is not properly clear to myself.
+Think in terms of $g(\theta)$ and $g(\hat{\theta})$}.
+
+We have shown in this section how the dropping of the pulsar terms results in a bias in the inferred parameters, not just in the sky localisation parameters  \citep[e.g.][]{Zhupulsarterms,Chen2022} but additionally $\psi, h_0$ and $\iota$.
+We have only considered a single example system to explore the effect of the bias.
+Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for each of the parameters, it is present and will need to be accounted for.
+This bias is not particular to our method but a shared feature across likelihood based methods that do not include the pulsar terms in the inference model.
+The inclusion of the pulsar terms in the methods presented in this work will be a key goal of a future work, which will alleviate these bias problems.
 

### 2023-08-16 docs/manuscript/paper.tex
Bayes ratio, 1000 points, 2 noise realisations

@@
 
-Equations \ref{eq:spinevol}--\ref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
-Equations \ref{eq:spinevol}--\ref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
+Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
+Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
 As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G}.
-An approach akin to Equations \ref{eq:spinevol}--\ref{eq:xieqn} has been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
-However, Equations \ref{eq:spinevol}--\ref{eq:xieqn}  involve significant idealizations, which must be recognized at the outset \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
-First, the white noise driver $\xi(t)$ in Equation \ref{eq:frequency_evolution} is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper.
+An approach akin to Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} has been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
+However, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}  involve significant idealizations, which must be recognized at the outset \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
+First, the white noise driver $\xi(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper.
 Second, the white spectrum assumed for $\xi(t)$ may or may not be suitable for millisecond pulsars in PTAs.
@@
 Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004} which have not been ruled out by pulsar timing experiments to date.
-The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \ref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \ref{eq:frequency_evolution}.
-
-\subsection{Modulation of pulse TOAs by a GW} \label{sec:psr_measured}
-In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
+The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \eqref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \eqref{eq:frequency_evolution}.
+
+\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
+In the presence of a GW, the pulsar frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
 Indeed, the pulse TOAs are modulated harmonically at the GW frequency.
 We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section.
-The measurement equation is a key input into the Kalman filter in Section \ref{sec:kalman_filter}
+The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
 \subsubsection{Plane GW perturbation}
@@
 A source can be considered as monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution of $1/T_{\rm obs}$.
-From Equation \ref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans.
+From Equation \eqref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans.
 The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA with pulsars at 1.5 kpc, 78 \% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52 \%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
@@
 \end{equation}
-where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame (MCRF) of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
+where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
 We take the pulsar location to be constant i.e.
-$\boldsymbol{q}$ is not a function of time.
+neither $\boldsymbol{q}$ nor $d$ is not a function of time.
 In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB.
@@
 where the perturbations to $\boldsymbol{u}$ from the GW are of higher order and neglected.
-In this case the shift in the momentum from Equation \ref{eq:momentum_shift} can be related to a shift in the frequency as,
+In this case the shift in the momentum from Equation \eqref{eq:momentum_shift} can be related to a shift in the frequency as,
 \begin{equation}
@@
 \end{equation}
-It is will also prove instructive to express Equation \ref{eq:g_func} in a trigonometric form as,
+It is will also prove instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
 or in a trigonometric form 
@@
 \end{align}
-Equations \ref{eq:measurement} -- \ref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.
+Equations \eqref{eq:measurement} -- \eqref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.
 
@@
 \end{equation}
-In Equation \ref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
+In Equation \eqref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
 \begin{align}
@@
 	\caption{Example application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$.
-We here plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency.
+We plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency.
 In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ and the state is estimated accurately.
@@
 The bottom left panels show the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$.
-The right hand panels plot the distribution of the innovation over the total observation period.}
+The right hand panels plot the distribution of the innovation over the total observation period.
+In (a) the greed/blue and red/magenta lines are completely overlaid.}
 	\label{fig:four figures}
@@
 In this work we use the linear Kalman filter which assumes a linear relation between $\boldsymbol{X}$ and $\boldsymbol{Y}$.
-Whilst the measurement equation, Equation \ref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively.
+Whilst the measurement equation, Equation \eqref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively.
 Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, the unscented Kalman filter \citep{882463van} or the particle filter \citep{Simon10}.
@@
 , M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$.
-The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Equation \ref{eq:measurement}.
+The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Equation \eqref{eq:measurement}.
 The measurement predictions can then be compared against the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation".
@@
 \subsection{Nested Sampling}\label{sec:nested_sampling}
-We can use the likelihood returned by the Kalman filter, Equation \ref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
+We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
 \begin{equation}
@@
 \begin{equation}
-	\mathcal{Z} = \int \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta}) d \boldsymbol{\theta} \ . \label{eq:model_evidence}
+	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
 \end{equation}
 In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work.
-Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \ref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
 It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
@@
 \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}.
-For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \cite{bilby.507.2037A} front-end library.
+For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library.
 We  follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work.
@@
 \subsection{Model selection}\label{sec:model_selection}
-The evidence integral,$\mathcal{Z}$, returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}_i$.
+The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}_i$.
 This enables us to compare competing models via a Bayes factor,
@@
 $\mathcal{M}_0$ is our null model that assumes there is no GW in the data.
-This is equivalent to setting $g^{(n)}(t)=1$ in Equation \ref{eq:g_func}, \ref{eq:g_func_trig}.
+This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func}, \eqref{eq:g_func_trig}.
 The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
@@
 As a sanity check, we can also calculate $\sigma^{(n)}$ using a complementary numerical approach to the empirical best-fit model outlined above.
-We can directly solve the state equation Equation \ref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a synthetic phase solution,
+We can directly solve the state equation Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a synthetic phase solution,
 \begin{eqnarray}
@@
 We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} from the  Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}.
-For the purposes of validating the $\sigma^{(n)}$ inferred from Equation \ref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually.
+For the purposes of validating the $\sigma^{(n)}$ inferred from Equation \eqref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually.
 Whilst it would also be possible to e.g.
@@
 The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}.
-The synthetic residuals are generated by numerically solving Equation \ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations.
-\ref{eq:sigmap}, \ref{eq:sigmap_f}.
+The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations.
+\eqref{eq:sigmap}, \eqref{eq:sigmap_f}.
 For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
@@
 \begin{enumerate}
-	\item Integrate the state equations, Equations \ref{eq:spinevol}- \ref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
-	\item Map from state space to measurement space via the measurement equation, Equation \ref{eq:measurement}
+	\item Integrate the state equations, Equations \eqref{eq:spinevol}- \eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
+	\item Map from state space to measurement space via the measurement equation, Equation \eqref{eq:measurement}
 	\item Add zero-mean Gaussian measurement noise to each timeseries, i.e.
@@
 \end{enumerate}
-Equations \ref{eq:spinevol}- \ref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
-The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified by our construction of the synthetic PTA outlined in Section \ref{sec:synt_pta}.
+Equations \eqref{eq:spinevol}- \eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified by our construction of the synthetic PTA outlined in Section \eqref{sec:synt_pta}.
 For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
@@
 
-In practice, in order to avoid numerical precision issues that arise when solving Equations \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
-\begin{equation}
-	f_{\rm p}^{\prime (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)}
+In practice, in order to avoid numerical precision issues that arise when solving Equations \eqref{eq:spinevol}- \eqref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
+\begin{equation}
+	f_{\rm p}^{\prime (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ .
 \end{equation}  
@@
 \begin{equation}
-	f_{\rm m}^{\prime (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{*(n)}
+	f_{\rm m}^{\prime (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{*(n)} \ ,
 \end{equation}
 where $ f_{\rm em}^{*(n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris.
-For synthetic data we can set $ f_{\rm em}^{*(n)} = f_{\rm em}^{(n)}$ but this is not true generally for real-world observations since the spin-down ephemeris is only known approximately, not exactly.
-The measurement equation Equation \ref{eq:measurement} that relates the state and measurement variables is then updated as
-\begin{equation}
-	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right]
+For synthetic data we can set $ f_{\rm em}^{*(n)} = f_{\rm em}^{(n)}$ but this is not true generally for real-world observations since the spin-down ephemeris is only known approximately.
+The measurement equation Equation \eqref{eq:measurement} that relates the state and measurement variables is then updated as
+\begin{equation}
+	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
 	\label{eq:measurement_cov}
@@
 These mean errors are skewed by a minority of pulsars which are less well timed; the median error in  the barycentric rotation frequency is $\sim 7 \times 10^{-13}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.8 \times 10^{-20}$ s $^{-2}$.
-Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 1 \%$ either side of the true value.
+Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 10 \%$ either side of the true value.
 In this way we can test how well our method performs without requiring exceptionally precise and accurate measurements of the pulsar parameters to be made \textit{a priori}.
 Instead we consider the pulsar parameters similarly to the GW source parameters and estimate them optimally within a consistent framework.
-Leaving the priors looser in this way allows us to subject our method to a more stringent test and confirm that we still estimate GW parameters without highly accurate initial estimates of $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$.
+Adopting looser priors in this way allows us to subject our method to a more stringent test and confirm that we still estimate GW parameters without highly accurate initial estimates of $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$.
 \newline
@@
 Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
-That is, for $T_{\rm obs}$ =10 years the solution of Equation \ref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ is satisfied.
+That is, for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ is satisfied.
 It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value.
 We briefly explored setting an uninformative prior on $\gamma$ over e.g.
-LogUniform($10^{-10}, 10^{-15}$) as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g.
+LogUniform($10^{-15}, 10^{-10}$) as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g.
 set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
-The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22}$ as calculated from Equations \ref{eq:sigmap}, \ref{eq:sigmap_f}.
-For these pulsars we set an uninformative broad prior of LogUniform($10^{-19}, 10^{-23}$).
+The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
+For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$).
 The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16}$.
@@
 	  & $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
-	  & $h_0$            & $10^{-12}$ & - & LogUniform($10^{-9}$, $10^{-15}$) \\
+	  & $h_0$            & $10^{-12}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
 	  & $\iota$             & $1.0$ & radians & Sin($0, \pi$) \\ 
@@
 		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
-		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform*($10^{-19}, 10^{-23}$) \\
+		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform*($10^{-23}, 10^{-19}$) \\
 		& $\gamma$              & $10^{-13}$ & s$^{-1}$ & - \\
@@
 The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \ref{eq:sigmap}, \ref{eq:sigmap_f} from \protect \cite{Shannon2010}.
+The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.
 All pulsars use the prior on $\sigma^{(n)}$ as specified in the Table, with the exception of PSR J1939+2134 which uses LogUniform($10^{-15}, 10^{-17}$).}
@@
 \subsubsection{Parameter estimation}\label{sec:parameter_estim}
-From Equation \ref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms.
+From Equation \eqref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms.
 The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
@@
 	We defer the inclusion of the pulsar terms in the measurement equation to a future work.
-As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation Equation \ref{eq:measurement} is no longer a function of the pulsar distance.
+As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation Equation \eqref{eq:measurement} is no longer a function of the pulsar distance.
 We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
@@
 Initially we consider a single noise realisation of the synthetic data for the representative example system described in Table \ref{tab:parameters_and_priors}.
-We apply the Kalman filter using the Earth terms only measurement equation, Equation \ref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
+We apply the Kalman filter using the Earth terms only measurement equation, Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
+\newline
+
  The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:example_psr_params} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$.
+The histograms shown in the Figures are the 1D posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
+The 2-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
 It is evident that for this representative example we are able to accurately estimate all the static parameters of the system using our method.
@@
 We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-The Bayes factor, $\beta$, as defined in Equation \ref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
 The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
@@
 We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
-We can see from Figure \ref{fig:corner_plot_2} that the results for multiple noise realisations agree well with the results that we saw in  Figure \ref{fig:corner_plot_1} for a single noise realisation.
+We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1} for a single noise realisation.
 The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations i.e.
@@
 Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.
-Fig \ref{fig:example_psr_params}) show a consistency across multiple noise realisations, though for brevity we do not show the results here \textcolor{red}{TK: is this fair?
-Don't want to clog up paper with extra figures that don't offer any additional insight...} \newline
+Fig \ref{fig:example_psr_params}) show a consistency across multiple noise realisations, though for brevity we do not show the results here.
+\newline
 
@@
 From Equations \ref{eq:hphx}, \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
-For example, from Equation \ref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
-This degeneracy is broken generally by the relation for $h_{+}$, Equation \ref{eq:hphx}.
+For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
+This degeneracy is broken generally by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
 Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of physically reasonable values (\textcolor{red}{TK: Could include a figure here showing this, but don't want to overclutter}).
@@
 
-In Figure \ref{fig:likelihood_curves_earth} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \ref{eq:likelihood}, as we vary a single static parameter of $\boldsymbol{\theta}_{\rm gw}$, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
+In Figure \ref{fig:likelihood_curves_earth} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, as we vary a single static parameter of $\boldsymbol{\theta}_{\rm gw}$, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
 For instance, in Figure \ref{fig:omega_likelihood} we pass different values of $\Omega$ in the range $10^{-9}$ to $10^{-5}$ Hz into the Kalman filter, whilst holding the remaining parameters constant, and plot the retuned log-likelihood as a function of $\Omega$.
@@
 
-We can compare the likelihood curves of Figure \ref{fig:likelihood_curves_earth} generated using the earth -terms only measurement equation, Equations \ref{eq:measuremen_earth} \ref{eq:g_func_trig_earth}, with the likelihood curves derived using the measurement equation which includes the pulsar terms, Equation \ref{eq:measurement}.
-Whilst we have not used the nested sampling method together with the pulsar terms measurement equation in this  work, we can run the Kalman filter using Equation \ref{eq:measurement} and consider how the resulting likelihood varies for different parameter values.
+We can compare the likelihood curves of Figure \ref{fig:likelihood_curves_earth} generated using the earth -terms only measurement equation, Equations \eqref{eq:measuremen_earth} \eqref{eq:g_func_trig_earth}, with the likelihood curves derived using the measurement equation which includes the pulsar terms, Equation \eqref{eq:measurement}.
+Whilst we have not used the nested sampling method together with the pulsar terms measurement equation in this  work, we can run the Kalman filter using Equation \eqref{eq:measurement} and consider how the resulting likelihood varies for different parameter values.
 The results are shown in Figure \ref{fig:likelihood_curves_psr}.
@@
 Whilst clear optima exist above the noise, and whilst local to the true value the curve becomes much more smooth, this noise presents an additional challenge to typical likelihood inference techniques which perform best on smooth likelihoods with a single clear global optima.
-It affects these parameters in particular since Equation \ref{eq:g_func_trig} contains a phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, which is essentially a combination of these 3 parameters ($ \boldsymbol{n}$ is defined by $\delta$ and $\alpha$).
+It affects these parameters in particular since Equation \eqref{eq:g_func_trig} contains a phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, which is essentially a combination of these 3 parameters ($ \boldsymbol{n}$ is defined by $\delta$ and $\alpha$).
 When calculating a likelihood we are asking "How likely is this data, given these parameters".

### 2023-08-16 py_src/priors.py

@@ -188,16 +188,18 @@ def set_prior_on_measurement_parameters(init_parameters,priors,measurement_model
 
 
         init_parameters["delta_gw"] = None
-        priors["delta_gw"] = bilby.core.prior.Uniform(-np.pi/2.0, np.pi/2, 'delta_gw')
+        #priors["delta_gw"] = bilby.core.prior.Uniform(-np.pi/2.0, np.pi/2, 'delta_gw')
+        priors["delta_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2, 'delta_gw')
 
 
         init_parameters["alpha_gw"] = None
-        priors["alpha_gw"] = bilby.core.prior.Uniform(0.0, 2*np.pi, 'alpha_gw')
+        #priors["alpha_gw"] = bilby.core.prior.Uniform(0.0, 2*np.pi, 'alpha_gw')
+        priors["alpha_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'alpha_gw')
 
 
         init_parameters["h"] = None
-        priors["h"] = bilby.core.prior.LogUniform(h/100.0, h*10.0, 'h')
-        #priors["h"] = bilby.core.prior.LogUniform(1e-15, 1e-12, 'h')
+        #priors["h"] = bilby.core.prior.LogUniform(h/100.0, h*10.0, 'h')
+        priors["h"] = bilby.core.prior.LogUniform(1e-15, 1e-12, 'h')
         #priors["h"] = h #experiment fixing $h$ to compare performance in $\iota$
 
     return init_parameters,priors 

### 2023-08-18 docs/manuscript/paper.tex
run a single example with dlogz=0.1

@@
 % If you need two or more lines of authors, add an extra line using \newauthor
-\author[Kimpson]{Kimpson$^{1}$, Melatos, O'Leary, Evans, others, etc. etc. %
+\author[Kimpson]{Kimpson$^{1}$, Melatos, O'Leary, Evans, others, etc. etc.%
 \thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
@@
 \begin{abstract}
-This is an abstract
+	\textcolor{red}{TK: will obviously drop the bullet points, but useful for writing a concise abstract.}
+
+	\begin{itemize}
+		\item Continuous nHz gravitational waves from individual supermassive black hole binaries are a primary target for Pulsar Timing Array (PTA) observations.
+		\item We present a new and complementary state-space approach to PTA data analysis.
+		\item We show how to combine a Kalman filter and nested sampling to estimate te
+
+	\end{itemize}
+
+ using the combination of a Kalman filter and .
+
+that is able to self consistenyl track the intrinsic timingnoise in PTA pulsars and disentangle it from the GW-induced TOA modulat
+
+We illustrate how our method is able to recover the full posterior distributions of the N model parameters
+
+We combine the to...
+
+The method is tested...
+
+bias observed....
+
 \end{abstract}
@@
 \begin{equation}
-	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{[i(\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0)]} \ ,
+	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
 \end{equation}
@@
 We take the pulsar location to be constant i.e.
-neither $\boldsymbol{q}$ nor $d$ is not a function of time.
+neither $\boldsymbol{q}$ nor $d$ is a function of time.
 In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB.
 This barycentering correction is typically applied when generating TOAs using e.g.
-{\sc tempo2} and related timing software.
+{\sc tempo2} \citep{tempo2} and related timing software.
 Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression:
@@
 \end{equation}
-where \textcolor{red}{TK: what is the best notation here to give $q^i$ an $(n)$ superscript?}
-\begin{equation}
-	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right) \ .
+where
+\begin{equation}
+	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i, \,(n)} q^{j, \,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right) \ .
 	\label{eq:g_func}
 \end{equation}
-It is will also prove instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
-or in a trigonometric form 
+and $q^{i, \,(n)}$ labels the $i$-th coordinate component of the of the $n$-th pulsars position vector $\boldsymbol{q}^{(n)}$.
+It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
 \begin{align}
-	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
-	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right) \right] \ .
+	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^{i, \,(n)} q^{j, \,(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
+	& \times \Big\{ \cos\left[-\Omega t +\Phi_0\right] \nonumber \\
+	&- \cos \left[-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right] \Big\} \ .
 	\label{eq:g_func_trig}
 \end{align}
-Equations \eqref{eq:measurement} -- \eqref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.
+Equations \eqref{eq:measurement} -- \eqref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the frequency measured by an observer on Earth.
+
+\textbf{MAYB ESUPRESS N ? rm \,}
+
+\textbf{Introduce puslar terms dropping here insttea?}
 
@@
 The right hand panels plot the distribution of the innovation over the total observation period.
-In (a) the greed/blue and red/magenta lines are completely overlaid.}
+In (a) the greed/blue and red/magenta lines are completely overlaid.
+\textcolor{red}{TK: See Fig \ref{fig:kalmanexampleboth} in Appendix for an alternative version of this figure.
+Which do we prefer?}}
 	\label{fig:four figures}
@@
 
-The Kalman filter \citep{Kalman1} is an algorithm for recovering the most likely evolution of a set of system state variables, $\boldsymbol{X}$, given some noisy measurements, $\boldsymbol{Y}$.
+The Kalman filter \citep{Kalman1} is a Gaussian-Markov model used to algorithmically recover the most likely set of unknown system state variables, $\boldsymbol{X}$, given some noisy measurements, $\boldsymbol{Y}$.
 It is a common technique in signal processing that has been applied successfully in neutron star astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}.
@@
 The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}.
-At each discrete timestep $i = 1, ...
-, M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$.
+At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$.
 The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Equation \eqref{eq:measurement}.
@@
 Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are also functions of $\boldsymbol{\theta}$.
-If the estimates of the system parameters, $\boldsymbol{\hat{\theta}}$ , that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised.
+If the estimates of the system parameters, $\boldsymbol{\hat{\theta}}$ , that we pass to the Kalman filter are close to the true underlying parameters then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are	 minimized and $\mathcal{L}$ is maximised.
 This is illustrated in Figure \ref{fig:6MB_BFS}; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
@@
 We take $n_{\rm live} = 500$ for all results presented in this work.
+\newline
 
@@
 \end{equation}
-Throughout this work we take $\mathcal{M}_1$ to be the complete model defined in Section \ref{sec:model}.
-$\mathcal{M}_0$ is our null model that assumes there is no GW in the data.
+Throughout this work we take $\mathcal{M}_1$ to be the state space model that includes the presence of a GW $\mathcal{M}_0$ is our null model that assumes there is no GW in the data.
 This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func}, \eqref{eq:g_func_trig}.
 The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
+
+\subsection{Summary of workflow}
+To summarise the above, the order of operations in our method for estimating the posteriors on the static parameters and calculating the marginal likelihood for the inference model is as follows:
+
+\begin{enumerate}[wide = 0pt]
+	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\pi(\boldsymbol{\theta})_{\rm sample}$
+	\item Pass $\pi(\boldsymbol{\theta})_{\rm sample}$ to the Kalman filter
+	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
+	\item Return $\log \mathcal{L}$ to the nested sampler
+	\item Nested sampler updates $\boldsymbol{\theta}$
+	\item Repeat
+\end{enumerate}
+In Section \ref{sec:testing} we will apply this workflow on a synthetic data set.
 
@@
 
-For this work we generate representative synthetic data as follows.
+For this work we generate synthetic data that is astrophysically representative as follows.
 We first specify a set of pulsars to make up our artificial PTA.
@@
 The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}.
-The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations.
-\eqref{eq:sigmap}, \eqref{eq:sigmap_f}.
+The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f}.
 For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
@@
 We initially characterise our method with a single example of a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected arbitrarily, subject to being astrophysically reasonable and representative.
+The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected to be astrophysically reasonable and representative.
 The injected $\boldsymbol{\theta}_{\rm gw}$ are summarised in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}.
@@
 \subsubsection{Prior distributions}\label{sec:priors}
+
+\textbf{CLARIFY re uninformative c.f loguniform}
+
+\textbf{sin/cos priors realy uniform over cos(x)}
+
+\textbf{referecne NANOgrav puslars and n index c.f. TAble ANTF}
+
+\textbf{why did you choose NANOGrav you dick}
+
+\textbf{priors on f, fdot, shrink to e.g. factor x ANTF error}
 
@@
 We can identify $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar  barycentric rotation frequency and the time derivative of barycentric rotation frequency quoted by pulsar observation catalogues.
-For the 12.5 year NANOGrav pulsars the mean error in the barycentric rotation frequency is $\sim 1.1 \times 10^{-11}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.29 \times 10^{-19}$ s $^{-2}$.
-These mean errors are skewed by a minority of pulsars which are less well timed; the median error in  the barycentric rotation frequency is $\sim 7 \times 10^{-13}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.8 \times 10^{-20}$ s $^{-2}$.
+For the 12.5 year NANOGrav pulsars the median error in  the barycentric rotation frequency is $\sim 7 \times 10^{-13}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.8 \times 10^{-20}$ s $^{-2}$.
 Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 10 \%$ either side of the true value.
@@
 Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
-That is, for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ is satisfied.
+That is, for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
 It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value.
@@
 The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16}$.
-For PSR J1939+2134 we set the prior at LogUniform($10^{-15}, 10^{-17}$).
-\newline
+For our artificial dataset we set $\sigma = 10^{21}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a constrained prior
 
@@
 	  & $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
-	  & $\delta$              & $1.0$  & radians & Cosine($-\pi/2, \pi/2$) \\
+	  & $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
 	  & $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
@@
 	  & $h_0$            & $10^{-12}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
-	  & $\iota$             & $1.0$ & radians & Sin($0, \pi$) \\ 
+	  & $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
 		\hline
@@
 		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
-		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform*($10^{-23}, 10^{-19}$) \\
+		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
 		& $\gamma$              & $10^{-13}$ & s$^{-1}$ & - \\
@@
 The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.
-All pulsars use the prior on $\sigma^{(n)}$ as specified in the Table, with the exception of PSR J1939+2134 which uses LogUniform($10^{-15}, 10^{-17}$).}
+The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.}
 	\label{tab:parameters_and_priors}
@@
 	\begin{equation}
-		g^{(n)}_{\rm Earth}(t) = 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \cos(-\Omega t +\Phi_0)  \ .
+		g^{(n)}_{\rm Earth}(t) = 1 - \frac{1}{2} \frac{ H_{ij}q^{i, \, (n)} q^{j, \, (n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \cos(-\Omega t +\Phi_0)  \ .
 		\label{eq:g_func_trig_earth}
@@
 The generated synthetic data that we use to test our method does include the pulsar term in full.
-\newline
-
 Initially we consider a single noise realisation of the synthetic data for the representative example system described in Table \ref{tab:parameters_and_priors}.
@@
 
- The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:example_psr_params} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$.
-The histograms shown in the Figures are the 1D posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
+ The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$.
+The histograms shown in the Figure are the 1D-posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
 The 2-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
-It is evident that for this representative example we are able to accurately estimate all the static parameters of the system using our method.
-Regarding the estimates of $\boldsymbol{\theta}_{\rm gw}$, whilst all the parameters are generally recovered accurately, for some the posterior median is almost identically equal to the injected value  (e.g.
+It is evident that for this representative example we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for the majority of the parameters.
+The clear exception is $\iota$ which exhibits a notable bias, and lies outside the with the 90\% credible interval, with the posterior median shifted from the injected value by $\sim 0.3$ radians.
+For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
+a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation.
+We will explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.
+If we consider the median values of the 1D-posteriors we can see that for some parameters the median is almost identically equal to the injected value  (e.g.
 $\Omega, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the injected value (e.g.
-$\psi, \alpha$), whilst still being well within the distribution uncertainty.
-For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
-a bias in the parameter estimate, or else just a random variation.
-We will explore this further in Section \ref{sec:multiple_noise} and show that this is a bias that results from dropping the Earth terms, similar to the bias reported in \cite{Zhupulsarterms}.
-Regarding $\boldsymbol{\theta}_{\rm psr, reduced}$, the frequency parameters $f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1)$ are generally well recovered for all pulsars, as are the noise amplitude parameters $\sigma^{(n)}$.
+$\psi, \alpha$), whilst still being contained within the credible interval.
+This potential bias will also be investigated alongside the bias in $\iota$ in Section \ref{sec:multiple_noise} and Section \ref{sec:bias}.
+Similar results can be derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
+We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.
+
+% and in Figure \ref{fig:example_psr_params} for . 
+% 
+% Regarding $\boldsymbol{\theta}_{\rm psr, reduced}$, the frequency parameters $f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1)$ are generally well recovered for all pulsars, as are the noise amplitude parameters $\sigma^{(n)}$.
 \textcolor{red}{TK: Could do with some extra discussion here re $\boldsymbol{\theta}_{\rm psr, reduced}$ since they are not perfectly recovered, but the important thing is that the GW parameters and the GW parameters are the interesting bit...}
+
 \begin{figure*}
@@
 The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-We are able to accurately estimate each parameter of interest.}
+We are able to accurately estimate each parameter of interest.
+$\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^12} = h_0 \times 10^{12}$.}
 	\label{fig:corner_plot_1}
@@
 
-\begin{figure*}%
-	\centering
-	\subfloat[\centering $f_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_f} }}%
-	\qquad
-	\subfloat[\centering $\dot{f}_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_fdot} }}%
-	\qquad
-	\subfloat[\centering $\sigma^{(n)}$]{{\includegraphics[width=0.3\textwidth]{images/1237_broad_sigma_p} }}%
-
-	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$.
+%\begin{figure*}%
+%	\centering
+%	\subfloat[\centering $f_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_f} }}%
+%	\qquad
+%	\subfloat[\centering $\dot{f}_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_fdot} }}%
+%	\qquad
+%	\subfloat[\centering $\sigma^{(n)}$]{{\includegraphics[width=0.3\textwidth]{images/1237_broad_sigma_p} }}%
+%	
+%	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$.
 We show only a selection of the first 5 pulsars rather than the full PTA.
 \textcolor{red}{TK: these figures are a little small to resolve nicely}}
-	\label{fig:example_psr_params}%
-\end{figure*}
+%	\label{fig:example_psr_params}%
+%\end{figure*}
 
@@
 We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
+$\mathcal{M}_1$ is the Earth-terms only model, i.e.
+Equation \eqref{eq:measuremen_earth}.
 The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
@@
 For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3 \times 10^{-15}$.
+This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
+$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
+This will be explored in more detail in Section \ref{sec:parameter_space}.
 The small perturbations in the general log-linear $\beta - h_0$ trend arise due to noise in the nested sampling method.
@@
 \end{figure*}
-The above discussion was just for a single realisation of the noise processes.
+The above discussion was for a single realisation of the noise processes.
 It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method.
@@
 We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
+As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here.
+\newline
+
 We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1} for a single noise realisation.
-The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations i.e.
-the variance between the different marginalized posteriors is generally low.
-The estimates generally agree with the injected values with high fidelity.
-Conversely, for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise.
-Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.
-Fig \ref{fig:example_psr_params}) show a consistency across multiple noise realisations, though for brevity we do not show the results here.
-\newline
+The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations with the injected value contained in the 90 \% credible interval for each of the marginalized posteriors.
+The variance between each of the individual 1D histograms is also generally low.
+
+Conversely for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise.
+Typically, for a given noise realisation, we do not contain the injected value within the 90\% credible interval for both $\iota$ and $h_0$ simultaneously.
 
@@
 This degeneracy is broken generally by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
-Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of physically reasonable values (\textcolor{red}{TK: Could include a figure here showing this, but don't want to overclutter}).
+Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of physically reasonable values.
 Consequently it is difficult for our algorithm to select a particular point in this plateau.
@@
 	\item Can we also estimate radiometer noise? Is this useful?
+	\item Lets preempt reviewer and make it explicit WHY we are not using TOAs
 \end{itemize}
@@
 
+\section{Alternative version of Fig 1}
+
+\begin{figure*}
+	\centering
+	\includegraphics[width=\textwidth]{images/Kalman_example_both}
+	\caption{}
+	\label{fig:kalmanexampleboth}
+\end{figure*}
+
 \subsection{References}

### 2023-08-18 notebooks/plot_likelihood_surfaces.py

@@ -51,12 +51,12 @@ def load_and_plot(path,ax):
     z = np.reshape(surface_pulsar, (lx, ly))
    
     # Plot a 3D surface
-    #ax.plot_surface(X, Y, z,alpha=0.5)
+    ax.plot_surface(X, Y, z,alpha=0.5)
 
-    CS = ax.pcolormesh(X, Y, z) #clim=(0.0, 5e5)
+    #CS = ax.pcolormesh(X, Y, z) #clim=(0.0, 5e5)
     #plt.contour(X, Y, Z, 4, colors='k')
-    plt.colorbar(CS)
-    ax.set_xscale('log')
+    #plt.colorbar(CS)
+   # ax.set_xscale('log')
 
 
 
@@ -67,8 +67,8 @@ def load_and_plot(path,ax):
 
 
 #load_and_plot('tmp_likelihood_surface2_earth.npz',ax)
-load_and_plot(path1,ax)
-#load_and_plot(path2,ax)
+#load_and_plot(path1,ax)
+load_and_plot(path2,ax)
 
 #ax.scatter(1e-12,1.0,-586119.6803021401,c='r', s=20)
 plt.show()

### 2023-08-18 py_src/bilby_wrapper.py

@@ -34,7 +34,8 @@ def BilbySampler(KalmanFilter,init_parameters,priors,label,outdir):
 			                  sample='rwalk_dynesty',
                               check_point_plot=False,
                               npoints=1000,
-                              dlogz=1e-6,
+                              #dlogz=1e-6,
+                              dlogz=0.1,
                               npool=1,
 			                  plot=False,resume=False)
 

### 2023-08-18 py_src/plotting.py

@@ -170,7 +170,7 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
     fig = corner.corner(y_post, 
                         color='C0',
                         show_titles=True,
-                        smooth=True,smooth1d=True,
+                        smooth=True,smooth1d=False,
                         truth_color='C2',
                         quantiles=[0.16, 0.84],
                         truths = injection_parameters,

### 2023-08-20 docs/manuscript/paper.tex
run with tighter prior on f

@@
 % If you need two or more lines of authors, add an extra line using \newauthor
-\author[Kimpson]{Kimpson$^{1}$, Melatos, O'Leary, Evans, others, etc. etc.%
+\author[Kimpson]{Kimpson$^{1}$, Melatos$^{2}$, O'Leary, Evans$^{3}$, others, etc. etc.%
 \thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
@@
 % List of institutions
-$^{1}$Royal Astronomical Society, Burlington House, Piccadilly, London W1J 0BQ, UK}
+$^{1}$School of Physics, University of Melbourne, Parkville, VIC 3010, Australia \\
+$^{2}$OzGrav, University of Melbourne, Parkville, VIC 3010, Australia \\
+$^{3}$Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia }
 
@@
 \begin{abstract}	
-	\textcolor{red}{TK: will obviously drop the bullet points, but useful for writing a concise abstract.}
-
 	\begin{itemize}
@@
 		\item We present a new and complementary state-space approach to PTA data analysis.
-		\item We show how to combine a Kalman filter and nested sampling to estimate te
-
+		\item We show how to combine a Kalman filter and nested sampling for Bayesian inference of the model parameters and model selection.
+		\item The method is tested on synthetic, astrophysically representative data and shown to be able to estimate the full posterior distributions of all parameters of the model, as well calculating the marginalized likelihood for model selection.
+		\item The bias introduced by dropping the `pulsar term' from the state-space model is analysed and discussed. 
 	\end{itemize}
-
- using the combination of a Kalman filter and .
-
-that is able to self consistenyl track the intrinsic timingnoise in PTA pulsars and disentangle it from the GW-induced TOA modulat
-
-We illustrate how our method is able to recover the full posterior distributions of the N model parameters
-
-We combine the to...
-
-The method is tested...
-
-bias observed....
 
@@
 \begin{equation}
-	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i, \,(n)} q^{j, \,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right) \ .
+	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i,(n)} q^{j,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right) \ .
 	\label{eq:g_func}
 \end{equation}
-and $q^{i, \,(n)}$ labels the $i$-th coordinate component of the of the $n$-th pulsars position vector $\boldsymbol{q}^{(n)}$.
+and $q^{i,(n)}$ labels the $i$-th coordinate component of the of the $n$-th pulsars position vector $\boldsymbol{q}^{(n)}$.
 It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
 \begin{align}
-	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^{i, \,(n)} q^{j, \,(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
+	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^{i,(n)} q^{j,(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
 	& \times \Big\{ \cos\left[-\Omega t +\Phi_0\right] \nonumber \\
@@
 Equations \eqref{eq:measurement} -- \eqref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the frequency measured by an observer on Earth.
-
-\textbf{MAYB ESUPRESS N ? rm \,}
-
-\textbf{Introduce puslar terms dropping here insttea?}
 
@@
 
-\begin{enumerate}[wide = 0pt]
+\begin{enumerate}
 	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\pi(\boldsymbol{\theta})_{\rm sample}$
@@
 We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}.
+The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA and the method naturally extends to other PTAs of the IPTA.
 For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$.
@@
 	\begin{equation}
-		g^{(n)}_{\rm Earth}(t) = 1 - \frac{1}{2} \frac{ H_{ij}q^{i, \, (n)} q^{j, \, (n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \cos(-\Omega t +\Phi_0)  \ .
+		g^{(n)}_{\rm Earth}(t) = 1 - \frac{1}{2} \frac{ H_{ij}q^{i,(n)} q^{j,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \cos(-\Omega t +\Phi_0)  \ .
 		\label{eq:g_func_trig_earth}

### 2023-08-20 py_src/bilby_wrapper.py

@@ -34,8 +34,8 @@ def BilbySampler(KalmanFilter,init_parameters,priors,label,outdir):
 			                  sample='rwalk_dynesty',
                               check_point_plot=False,
                               npoints=1000,
-                              #dlogz=1e-6,
-                              dlogz=0.1,
+                              dlogz=1e-6,
+                              #dlogz=0.1,
                               npool=1,
 			                  plot=False,resume=False)
 

### 2023-08-20 py_src/config_batch_skymap.py

@@ -27,7 +27,7 @@ def create_slurm_job(arg_name,h,measurement_model,seed,omega,phi0,psi,delta,alph
     
     
 
-Nsamples = 10
+Nsamples = 100
 seed = 1237 
 h = 1e-12 
 
@@ -39,7 +39,7 @@ psi_gw = 2.50
 
 
 #eps prevents railing against prior edge
-eps =0.01
+eps =0.05
 deltas = np.linspace(-np.pi/2.0+eps,np.pi/2.0 - eps,Nsamples)
 alphas = np.linspace(0+eps,2*np.pi-eps,Nsamples)
 
@@ -52,7 +52,7 @@ with open('batch.sh','w') as b:
             for a in alphas:
 
    
-                arg_name = f"skymap_h_{h}_model_{m}_seed_{seed}_omega_{omega_gw}_phi0_{phi0_gw}_psi_{psi_gw}_delta_{d}_alpha_{a}"
+                arg_name = f"skymap100_h_{h}_model_{m}_seed_{seed}_omega_{omega_gw}_phi0_{phi0_gw}_psi_{psi_gw}_delta_{d}_alpha_{a}"
                 print(arg_name)
                 create_slurm_job(arg_name,h,m,seed,omega_gw,phi0_gw,psi_gw,d,a)
                 b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")

### 2023-08-20 py_src/priors.py

@@ -109,7 +109,7 @@ def set_prior_on_state_parameters(init_parameters,priors,f,fdot,σp,γ,d):
 
 
 
-    init_parameters,priors = add_to_bibly_priors_dict_uniform(f,"f0",init_parameters,priors,tol=0.01)      #uniform
+    init_parameters,priors = add_to_bibly_priors_dict_uniform(f,"f0",init_parameters,priors,tol=1e-10)      #uniform
     init_parameters,priors = add_to_bibly_priors_dict_uniform(fdot,"fdot",init_parameters,priors,tol=0.01) #uniform
 
     #If we set the true process noise to zero, then don't bother searching over this parameter

### 2023-08-21 docs/manuscript/paper.tex
100 noise realisations, h5e-15, narrow f prior

@@
 
-\section{Signal tracking and parameter estimation} \label{sec:detect}
+\section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
 The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e.
@@
 \begin{enumerate}
-	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\pi(\boldsymbol{\theta})_{\rm sample}$
-	\item Pass $\pi(\boldsymbol{\theta})_{\rm sample}$ to the Kalman filter
+	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\boldsymbol{\theta}_{\rm sample}$
+	\item Pass $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
 	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
@@
 \subsubsection{Prior distributions}\label{sec:priors}
-
-\textbf{CLARIFY re uninformative c.f loguniform}
-
-\textbf{sin/cos priors realy uniform over cos(x)}
-
-\textbf{referecne NANOgrav puslars and n index c.f. TAble ANTF}
-
-\textbf{why did you choose NANOGrav you dick}
-
-\textbf{priors on f, fdot, shrink to e.g. factor x ANTF error}
-
 For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters.
@@
 The variance between each of the individual 1D histograms is also generally low.
-
 Conversely for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise.
 Typically, for a given noise realisation, we do not contain the injected value within the 90\% credible interval for both $\iota$ and $h_0$ simultaneously.
+\newline
 
@@
 
-%\begin{figure*}
-%	\centering
-%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}} 
-%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}} 
-%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}}
-%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}}
-%	\caption{(a) blah (b) blah (c) blah (d) blah}
-%	\label{fig:foobar}
-%\end{figure*}
-
 \section{Discussion}
-
-Some potential further points for discussion
-
-\begin{itemize}
-	\item What is the performance of a single pulsar?
-	\item How does this map to Hellings-downs?
-	\item How does this map to having more than one source? Or a chrmomatic source?
-	\item Non constant time sampling?
-	\item Any further discussion on pulsar terms?
-	\item PTAs like MSPs for small timing noise. Can we get away with large timing noise, and use more pulsars? Some pulsars are more useful than others, e.g. arXiv 2211.03201
-	\item Can we also estimate radiometer noise? Is this useful?
-	\item Lets preempt reviewer and make it explicit WHY we are not using TOAs
-\end{itemize}
+We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
+
+The natural first extension of this work is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
+
+Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred 1D marginalized posteriors for the model parameters.
+
+Biases in synthetic datasets from dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.
+Alternatively, if one wishes to continue to use just the Earth terms it would be highly desirable to
+
+We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up NANOGrav.
+Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA.
+Indeed, due to the increased computational demands as pulsars are added to the arrays it may be advantageous to optimally select a subset of the total number of available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.
+
+Throughout this work we have assumed that all pulsars are observed at the same time, with a constant time sampling.
+In practice this assumption does not hold and pulsars will be observed with different cadences at different time intervals.
+Applying this method to non-constant time sampling is a key further extension.
+
+We have also assumed that there is only a single, quasi-monochromatic GW source that influences the pulsar frequencies.However, it may be possible to resolve multiple continuous GW source simultaneously.
+Our method naturally extends to multiple GW sources; Equation \ref{eq:measurement} is modified to become the linear superposition of multiple GWs.
+The stochastic background itself is the incoherent summation of many individual GW sources.
+It therefore stands to reason that we could extend this method to also search for a stochastic GW background without requiring cross-correlating the pulsar residuals to uncover the Hellings-Downs curve.
+
+\textcolor{red}{TK: Lets pre-empt reviewer and make it explicit WHY we are not using TOAs}
+We have formulated the state space model in terms of frequencies, where
 

### 2023-08-21 py_src/config_batch.py

@@ -26,32 +26,32 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
     
     
 
-# N = 10
-# seeds = np.arange(1235+10,1235+10+N,1)
-# h = 1e-12 
-# model = "earth"
-# with open('batch.sh','w') as b: 
+N = 100
+seeds = np.arange(1235+10,1235+10+N,1)
+h = 5e-15 
+model = "earth"
+with open('batch.sh','w') as b: 
 
-#     for s in seeds:
-#         arg_name = f"canonical_model_earth_batch_shifted_prior_fixed_h{s}"
-#         create_slurm_job(arg_name,h,model,s)
-#         b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+    for s in seeds:
+        arg_name = f"small_h_model_earth_batch_{s}"
+        create_slurm_job(arg_name,h,model,s)
+        b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
        
 
-h_range = np.logspace(-15,-12,101)
-noise_models = ["earth", "null"]
-seeds = [1237,1245]
+# h_range = np.logspace(-15,-12,101)
+# noise_models = ["earth", "null"]
+# seeds = [1237,1245]
 
-with open('batch.sh','w') as b:
+# with open('batch.sh','w') as b:
     
-    for s in seeds:
-        for h in h_range:
-            for n in noise_models:
-                 arg_name = f"paper_bayes_ratios_n1000_V2_h_{h}_model_{n}_seed_{s}"
-                 print(arg_name)
-                 create_slurm_job(arg_name,h,n,s)
+#     for s in seeds:
+#         for h in h_range:
+#             for n in noise_models:
+#                  arg_name = f"paper_bayes_ratios_n1000_V2_h_{h}_model_{n}_seed_{s}"
+#                  print(arg_name)
+#                  create_slurm_job(arg_name,h,n,s)
 
-                 b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+#                  b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
 
     
 

### 2023-08-21 docs/manuscript/paper.tex
bayes, 2000 points

@@
 For the 12.5 year NANOGrav pulsars the median error in  the barycentric rotation frequency is $\sim 7 \times 10^{-13}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.8 \times 10^{-20}$ s $^{-2}$.
-Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 10 \%$ either side of the true value.
+Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$, $\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 10^3 \epsilon^{(n)}_{a}$ either side of the true value, where $\epsilon^{(n)}_{a}$ is the error for that pulsar quoted in the Australia Telescope National Facility (ATNF) catalogue and the subscript $a = \{ f, \dot{f} \}$ identifies the error as being in $f_{\rm em}^{(n)} (t_1)$, or $\dot{f}_{\rm em}^{(n)} (t_1)$ respectively.
 In this way we can test how well our method performs without requiring exceptionally precise and accurate measurements of the pulsar parameters to be made \textit{a priori}.
@@
 The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16}$.
-For our artificial dataset we set $\sigma = 10^{21}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a constrained prior
+For our artificial dataset we set $\sigma = 10^{21}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a constrained prior.
+\newline
 
@@
 		\hline
-		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform(0.9$f_{\rm ATNF}^{n}$, $1.1 f_{\rm ATNF}^{n}$) \\
-		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform(0.9$\dot{f}_{\rm ATNF}^{n}$, $1.1 \dot{f}_{\rm ATNF}^{n}$) \\
+		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left( f_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{f} \right)$ \\
+		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left( \dot{f}_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{\dot{f}} \right)$ \\
 		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
@@
 We are able to accurately estimate each parameter of interest.
-$\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^12} = h_0 \times 10^{12}$.}
+$\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
 	\label{fig:corner_plot_1}
@@
 This will be explored in more detail in Section \ref{sec:parameter_space}.
+
 The small perturbations in the general log-linear $\beta - h_0$ trend arise due to noise in the nested sampling method.
@@
 Increasing the number of live points $n_{\rm live}$ used by the nested sampler also dampens these perturbations and smooths the $\beta - h_0$ curve, since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
+
 Below $\beta = 10$ this noise in the sampler starts to dominate and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
@@
 This is simply due to the fact that at these small strains the system is completely noise dominated and the two models $\mathcal{M}_1, \mathcal{M}_0$ cannot be separated.
-\textcolor{red}{TK: Rather than discussing the bumps, maybe just run it again a few times and remove the bumps.}
 \begin{figure}
@@
 \end{figure*}
-\begin{figure*}
-	\setkeys{Gin}{width=\linewidth}
-
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
-		\caption{$\Omega$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
-		\caption{$\Phi_0$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
-		\caption{$\psi$}
-	\end{subfigure}
-	\medskip
-
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
-		\caption{$\iota$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
-		\caption{$\delta$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
-		\caption{$\alpha$}
-	\end{subfigure}
-	\medskip
-
-	\medskip
-	\begin{subfigure}[b]{0.33\textwidth}
-		\includegraphics{images/distribution_h}
-		\caption{$h$}
-	\end{subfigure}
-
-	\caption{Distribution in the median values of the 1D marginalized posteriors inferred by nested sampling for each static parameters of $\boldsymbol{\theta}_{\rm gw}$ over 1000 noise realisations.
+%\begin{figure*}
+%	\setkeys{Gin}{width=\linewidth}   
+%	
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
+%		\caption{$\Omega$}
+%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
+%		\caption{$\Phi_0$}
+%	\end{subfigure}
+%	\hfill	
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
+%		\caption{$\psi$}
+%	\end{subfigure}
+%	\medskip
+%	
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
+%		\caption{$\iota$}
+%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
+%		\caption{$\delta$}
+%	\end{subfigure}
+%	\hfill	
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
+%		\caption{$\alpha$}
+%	\end{subfigure}
+%	\medskip
+%	
+%	
+%	\medskip
+%	\begin{subfigure}[b]{0.33\textwidth}
+%		\includegraphics{images/distribution_h}
+%		\caption{$h$}
+%	\end{subfigure}
+%	
+%	\caption{Distribution in the median values of the 1D marginalized posteriors inferred by nested sampling for each static parameters of $\boldsymbol{\theta}_{\rm gw}$ over 1000 noise realisations.
 The orange dashed line labels the true injection value and the red dashed line labels the median of the distribution.
@@
 Is there a better way to show this?}}\label{fig:median_distriubutins}
-\end{figure*}
+%\end{figure*}
+
 The above discussion was for a single realisation of the noise processes.
@@
 For each realisation of the noise we infer a marginalised 1-D posterior for each parameter.
+
+High SNR makes the posteriors effectively point estimates. Therefore maybe credible intervals dont make as much sense
+
+%
+%.
 In turn, each posterior distribution has a median value.
@@
 
+%%%%%%%%%%%%%%%%Scratch space
+
+%Taking a tangible example, for pulsar J0023+0923 ATNF returns a frequency $f_{\rm em}^{(n)} (t_1) \sim 327.8$ Hz with an error $\epsilon \sim 4 \times 10^{13}$ Hz.
+	The prior on $f_{\rm em}^{(n)} (t_1) \sim 327.8$ for this pulsar is then Uniform$(327.8 - 4 \times 10^{10},327.8 + 4 \times 10^{10} )$
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-08-21 docs/manuscript/paper.tex
commit before hplus branch

@@
 \begin{align}
-	H_{ij} = h_+ e_{ij}^+(\boldsymbol{n}) + h_{\times} e_{ij}^{\times}(\boldsymbol{n}) \ ,
+	H_{ij} = h_+ e_{ij}^+(\boldsymbol{n}) + h_{\times} e_{ij}^{\times}(\boldsymbol{n}) \ , \label{eq:hij}
 \end{align}
@@
 
-We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1} for a single noise realisation.
-The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations with the injected value contained in the 90 \% credible interval for each of the marginalized posteriors.
-The variance between each of the individual 1D histograms is also generally low.
-Conversely for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise.
-Typically, for a given noise realisation, we do not contain the injected value within the 90\% credible interval for both $\iota$ and $h_0$ simultaneously.
-\newline
-
-The large degree of variance in the estimates of $\iota$ and $h_0$ is due to the nested sampling algorithm effectively trading off accuracy in one parameter against the accuracy in other.
+We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1}.
+Considering the 1D marginalized posteriors, the inferred probability distributions of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations.
+Moreover, the variance of each of the individual 1D posteriors is generally low i.e.
+each of individual the distributions are narrow compared to the prior space.
+That is to say, for each different noise realisation the method infers a consistent maximum \textit{a posteriori} probability (MAP) estimate.
+Conversely for the parameters $\iota$ and $h_0$ the inferred 1D posteriors are not consistent across different noise realisations and the variance of an individual posterior is generally larger than for the other parameters.
+That is to say, we infer very different MAP estimates of these parameters depending on the particular realisation of the noise.
+\newline
+
+The large degree of variance in the 1D posteriors of $\iota$ and $h_0$ is due to the nested sampling algorithm effectively trading off accuracy in one parameter against the accuracy in other.
 From Equations \ref{eq:hphx}, \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
 For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
+
+The situation can be further elucidated if we consider instead the alternative parametrisation in terms of $h_{+}$ and $h_{\times}$.
+From Equation \ref{eq:hij} the amplitude of the plane GW perturbation at a particular sky location $\boldsymbol{n}$ is a linear combination of $h_{+}$ and $h_{\times}$.
+
+he log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of physically reasonable values.
+
 This degeneracy is broken generally by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
-Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of physically reasonable values.
+Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since t
+
 Consequently it is difficult for our algorithm to select a particular point in this plateau.
@@
 For each realisation of the noise we infer a marginalised 1-D posterior for each parameter.
+
+Since we are in the high SHR regime (c.f. Fgiure X for h = 1012) the posteriors are narrow high confidence in narrow range of values.
+
+In this case we are justified in talking abotu the mode of posteriors
 

### 2023-08-22 docs/manuscript/paper.tex
run pp plot for fixed small strain

@@
 \begin{figure*}
-	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/representative_example_v2_GW}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
 	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for a single realisation of the system noise.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{images/PaperBayesRatioPlotn1000_stacked}
+	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot}
 	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
@@
 \begin{figure*}
-	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/stacked_GW_plot_1000}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
 	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
@@
 \end{figure*}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_surface}
+	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation.
+The red point labels the location of the likelihood maxima.
+The likelihood surface has been normalised with respect to the absolute value of the maxima.
+A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
+	\label{fig:likelihood_surface}
+\end{figure}
 \begin{figure*}
@@
 	\caption{1D marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$.
-The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column).
-The thick orange line labels the true injected value.
-Fixing $h$ reduces the variance between different noise realisations but results in a clear bias between the true and inferred values of $\iota$.}
+The thick vertical orange line labels the true injected value.
+The dashed vertical lines denote the median of the corresponding coloured posterior.
+Fixing $h$ reduces the variance between different noise realisations but results in a clear bias between the true and inferred values of $\iota$.
+The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column), but with the $(0.16,0.84)$ quantiles replaced with the $0.5$ quantile.}
 	\label{fig:iota}
 \end{figure*}
-%\begin{figure*}
-%	\setkeys{Gin}{width=\linewidth}   
-%	
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
-%		\caption{$\Omega$}
-%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
-%		\caption{$\Phi_0$}
-%	\end{subfigure}
-%	\hfill	
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
-%		\caption{$\psi$}
-%	\end{subfigure}
-%	\medskip
-%	
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
-%		\caption{$\iota$}
-%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
-%		\caption{$\delta$}
-%	\end{subfigure}
-%	\hfill	
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
-%		\caption{$\alpha$}
-%	\end{subfigure}
-%	\medskip
-%	
-%	
-%	\medskip
-%	\begin{subfigure}[b]{0.33\textwidth}
-%		\includegraphics{images/distribution_h}
-%		\caption{$h$}
-%	\end{subfigure}
-%	
-%	\caption{Distribution in the median values of the 1D marginalized posteriors inferred by nested sampling for each static parameters of $\boldsymbol{\theta}_{\rm gw}$ over 1000 noise realisations.
-The orange dashed line labels the true injection value and the red dashed line labels the median of the distribution.
-\textcolor{red}{TK: these plots are ugly.
-Is there a better way to show this?}}\label{fig:median_distriubutins}
-%\end{figure*}
 
@@
 It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method.
+In this Section we confirm that we are able to infer comparable results for different noise realisations.
 To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.
@@
 
-The large degree of variance in the 1D posteriors of $\iota$ and $h_0$ is due to the nested sampling algorithm effectively trading off accuracy in one parameter against the accuracy in other.
-From Equations \ref{eq:hphx}, \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+The large degree of variance in the 1D posteriors of $\iota$ and $h_0$ is due to these parameters being only weakly-identifiable.
+Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
+Sources of non-identifiability are generally categorised as either ``structural" - arising from the structure of the model - or ``practical" - arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
+Regarding $\iota$ and $h_0$ we are concerned with structural identifiability.
+Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \ref{eq:hphx}, \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
 For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
-
-The situation can be further elucidated if we consider instead the alternative parametrisation in terms of $h_{+}$ and $h_{\times}$.
-From Equation \ref{eq:hij} the amplitude of the plane GW perturbation at a particular sky location $\boldsymbol{n}$ is a linear combination of $h_{+}$ and $h_{\times}$.
-
-he log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of physically reasonable values.
-
-This degeneracy is broken generally by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
-Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since t
-
-Consequently it is difficult for our algorithm to select a particular point in this plateau.
+This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
+However, it can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
+That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a non-zero gradient close to the maximum likelihood solution.
+This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
+Whilst a single, unique maxima exists (labelled by the red point in the figure), along ridge in the parameter space the likelihood values are all very similar.
+Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
 If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the variance in the estimates between different noise realisations decreases.
-This is demonstrated in Figure \ref{fig:iota} where we plot the marginalized 1D posterior of $\iota$ for 9 realisations of the noise in both the original case where $h_0$ is a free parameter (i.e.
-Figure \ref{fig:corner_plot_2}) and in the case where we hold $h_0$ fixed at its injection value, $h_0 = 10^{-12}$.
+This is demonstrated in Figure \ref{fig:iota} where we plot the marginalized 1D posterior of $\iota$ for 9 realisations of the noise in both the original case where $h_0$ is a free parameter (equivalent to the 1D posterior for $\iota$ in Figure \ref{fig:corner_plot_2}) and in the case where we hold $h_0$ fixed at its injection value, $h_0 = 10^{-12}$.
 We can see that whilst the variance between the individual posteriors is large in the original case, once we fix $h_0$ the variance decreases dramatically and the sampling algorithm repeatedly converges consistent estimates of $\iota$.
 There is a notable bias in Figure \ref{fig:iota} in the estimates for $\iota$ inferred for the fixed $h_0$ case relative to the true injected value; whilst the variance has decreased, the mean accuracy of the estimates has also decreased.
+An equivalent bias in $h_0$ is observed if we fix $\iota$.
 This bias is a result of dropping the pulsar terms from our measurement equation and is discussed in Section \ref{sec:bias}.
@@
 
+We now consider the complete set of 1000 noise realisations, rather than the subset of 9 realisations in the preceding discussion.
 For each realisation of the noise we infer a marginalised 1-D posterior for each parameter.
-
-Since we are in the high SHR regime (c.f. Fgiure X for h = 1012) the posteriors are narrow high confidence in narrow range of values.
-
-In this case we are justified in talking abotu the mode of posteriors
-
-High SNR makes the posteriors effectively point estimates. Therefore maybe credible intervals dont make as much sense
-
-%
-%.
-In turn, each posterior distribution has a median value.
-We plot the distribution of these medians over the 1000 noise realisations for each parameter of $\boldsymbol{\theta}_{\rm gw}$ in Figure \ref{fig:median_distriubutins}.
-For each of these distributions we can also calculate the median value (i.e.
-the median of the medians) and compare it with the true injection value.
-This is shown by the red and orange dashed lines respectively in the Figure.
-Analogous to the results we saw in Figure \ref{fig:corner_plot_2}, the distributions of the median values of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are very narrow, with a generally small discrepancy between the inferred value and the injected value.
-Conversely, the distributions for $\iota$ and $h_0$ are much broader and exhibit a bias of $\sim 0.18$ radians and $0.17 \times 10^{-12}$ respectively.
-This agrees with the results we discussed for the 9 noise realisations in Figure \ref{fig:corner_plot_2} where a bias in $\iota$ and $h_0$ was also observed.
-In addition to  $\iota$ and $h_0$  exhibiting a strong bias, the parameters $\psi$ and $\alpha$ also exhibit a small bias of magnitude $\sim 0.1$ and $\sim 0.04$ radians respectively.
-This bias is again a result of dropping the pulsar terms from the measurement equation and will be discussed in Section \ref{sec:bias}.
+For this injected strain ($h_0 = 10^{12}$) the SNR is large (c.f.
+Figure \ref{fig:bayes}) and consequently the posteriors are typically narrow compared to the prior space and well described by their median value.
+To compare each of the posteriors across different noise realisations, we consider the normalised standard deviation in the set of 1000 median values.
+For $\Omega, \Phi_0, \psi, \delta, \alpha$ the normalised standard deviation is $\sim 4 \times 10^{-5}, 0.03,0.004,0.03,0.02$ respectively.
+We can therefore see that the method is able to reliably converge to a consistent MAP estimate.
+Conversely, as we saw before, the MAP estimates for $\iota$ and $h$ are not consistent across different noise realisations, with normalised standard deviation of $0.3$ and $0.1$ respectively.Whilst the estimates are consistent across different noise realisations, for some parameters a small bias is observed between the MAP estimate and the injected value.
+This was evident in Figure \ref{fig:iota} for $\iota$ (and equivalently $h_0$s) which exhibits a strong bias of $\sim 0.35$ radians, but over the 1000 realisations biases are also observed in $\psi$ ($\sim 0.1$ radians) and $\alpha$ ($\sim 0.04$ radians).
+These biases are a result of dropping the pulsar terms from the measurement equation and will be discussed in Section \ref{sec:bias}
 
@@
 
-\begin{figure*}
-	\setkeys{Gin}{width=\linewidth}
-
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_omega_earth}
-		\caption{$\Omega$}
-		\label{fig:omega_likelihood}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_phi0_earth}
-		\caption{$\Phi_0$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_psi_earth}
-		\caption{$\psi$}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_iota_earth}
-		\caption{$\iota$}
-			\label{fig:iota_likelihood}
-	\end{subfigure}
-	\medskip
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_delta_earth}
-		\caption{$\delta$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_alpha_earth}
-		\caption{$\alpha$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics{images/likelihood_h_earth}
-		\caption{$h$}
-	\end{subfigure}
-
-	\caption{}\label{fig:likelihood_curves_earth}
-\end{figure*}
-
-\begin{figure*}
-	\setkeys{Gin}{width=\linewidth}
-
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_omega_psr}
-		\caption{$\Omega$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_phi0_psr}
-		\caption{$\Phi_0$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_psi_psr}
-		\caption{$\psi$}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_iota_psr}
-		\caption{$\iota$}
-	\end{subfigure}
-	\medskip
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_delta_psr}
-		\caption{$\delta$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/likelihood_alpha_psr}
-		\caption{$\alpha$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics{images/likelihood_h_psr}
-		\caption{$h$}
-	\end{subfigure}
-
-	\caption{As Figure \ref{fig:likelihood_curves_earth} but now including the pulsar terms in the measurement equation.}\label{fig:likelihood_curves_psr}
-\end{figure*}
+%\begin{figure*}
+%	\setkeys{Gin}{width=\linewidth}   
+%	
+%	\begin{subfigure}[b]{0.22\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_omega_earth}
+%		\caption{$\Omega$}
+%		\label{fig:omega_likelihood}
+%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.22\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_phi0_earth}
+%		\caption{$\Phi_0$}
+%	\end{subfigure}
+%	\hfill	
+%	\begin{subfigure}[b]{0.22\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_psi_earth}
+%		\caption{$\psi$}
+%	\end{subfigure}
+%	\begin{subfigure}[b]{0.22\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_iota_earth}
+%		\caption{$\iota$}
+%			\label{fig:iota_likelihood}
+%	\end{subfigure}
+%	\medskip
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_delta_earth}
+%		\caption{$\delta$}
+%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_alpha_earth}
+%		\caption{$\alpha$}
+%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics{images/likelihood_h_earth}
+%		\caption{$h$}
+%	\end{subfigure}
+%	
+%	\caption{}\label{fig:likelihood_curves_earth}
+%\end{figure*}
+%
+%
+%
+%
+%\begin{figure*}
+%	\setkeys{Gin}{width=\linewidth}   
+%	
+%	\begin{subfigure}[b]{0.22\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_omega_psr}
+%		\caption{$\Omega$}
+%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.22\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_phi0_psr}
+%		\caption{$\Phi_0$}
+%	\end{subfigure}
+%	\hfill	
+%	\begin{subfigure}[b]{0.22\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_psi_psr}
+%		\caption{$\psi$}
+%	\end{subfigure}
+%	\begin{subfigure}[b]{0.22\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_iota_psr}
+%		\caption{$\iota$}
+%	\end{subfigure}
+%	\medskip
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_delta_psr}
+%		\caption{$\delta$}
+%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics[width=\textwidth]{images/likelihood_alpha_psr}
+%		\caption{$\alpha$}
+%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.3\textwidth}
+%		\includegraphics{images/likelihood_h_psr}
+%		\caption{$h$}
+%	\end{subfigure}
+%	
+%	\caption{As Figure \ref{fig:likelihood_curves_earth} but now including the pulsar terms in the measurement equation.}\label{fig:likelihood_curves_psr}
+%\end{figure*}
+%
 
@@
 
+\section{Appendix on identifiability}\label{appendix_identifiability}
+
+This paper is useful: \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
 \section{Alternative version of Fig 1}
@@
 
+%Alternatively, one can consider instead the alternative parametrisation in terms of $h_{+}$ and $h_{\times}$. 
+%
+%from Equation \ref{eq:hij} the amplitude of the plane GW perturbation at a particular sky location $\boldsymbol{n}$ is a linear combination of $h_{+}$ and $h_{\times}$. 
+%
+%
+%
+%Considering a single pulsar where the only unknown parameters of the model are $h_{+}$ and $h_{\times}$, the system is clearly under-determined - we have one equation with two unknowns - and the parameters are not identifiable
+%
+%
+
+%\begin{figure*}
+%	\setkeys{Gin}{width=\linewidth}   
+%	
+%	\begin{subfigure}[b]{0.3\textwidth}
+	%		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
+	%		\caption{$\Omega$}
+	%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.3\textwidth}
+	%		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
+	%		\caption{$\Phi_0$}
+	%	\end{subfigure}
+%	\hfill	
+%	\begin{subfigure}[b]{0.3\textwidth}
+	%		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
+	%		\caption{$\psi$}
+	%	\end{subfigure}
+%	\medskip
+%	
+%	\begin{subfigure}[b]{0.3\textwidth}
+	%		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
+	%		\caption{$\iota$}
+	%	\end{subfigure}
+%	\hfill
+%	\begin{subfigure}[b]{0.3\textwidth}
+	%		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
+	%		\caption{$\delta$}
+	%	\end{subfigure}
+%	\hfill	
+%	\begin{subfigure}[b]{0.3\textwidth}
+	%		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
+	%		\caption{$\alpha$}
+	%	\end{subfigure}
+%	\medskip
+%	
+%	
+%	\medskip
+%	\begin{subfigure}[b]{0.33\textwidth}
+	%		\includegraphics{images/distribution_h}
+	%		\caption{$h$}
+	%	\end{subfigure}
+%	
+%	\caption{Distribution in the median values of the 1D marginalized posteriors inferred by nested sampling for each static parameters of $\boldsymbol{\theta}_{\rm gw}$ over 1000 noise realisations.
+The orange dashed line labels the true injection value and the red dashed line labels the median of the distribution.
+\textcolor{red}{TK: these plots are ugly.
+Is there a better way to show this?}}\label{fig:median_distriubutins}
+%\end{figure*}
+
+%
+%.
+In turn, each posterior distribution has a median value.
+We plot the distribution of these medians over the 1000 noise realisations for each parameter of $\boldsymbol{\theta}_{\rm gw}$ in Figure \ref{fig:median_distriubutins}.
+For each of these distributions we can also calculate the median value (i.e.
+the median of the medians) and compare it with the true injection value.
+This is shown by the red and orange dashed lines respectively in the Figure.
+Analogous to the results we saw in Figure \ref{fig:corner_plot_2}, the distributions of the median values of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are very narrow, with a generally small discrepancy between the inferred value and the injected value.
+Conversely, the distributions for $\iota$ and $h_0$ are much broader and exhibit a bias of $\sim 0.18$ radians and $0.17 \times 10^{-12}$ respectively.
+This agrees with the results we discussed for the 9 noise realisations in Figure \ref{fig:corner_plot_2} where a bias in $\iota$ and $h_0$ was also observed.
+In addition to  $\iota$ and $h_0$  exhibiting a strong bias, the parameters $\psi$ and $\alpha$ also exhibit a small bias of magnitude $\sim 0.1$ and $\sim 0.04$ radians respectively.
+This bias is again a result of dropping the pulsar terms from the measurement equation and will be discussed in .
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-08-22 py_src/config_batch_pp.py

@@ -29,7 +29,7 @@ def create_slurm_job(arg_name,h,measurement_model,seed,omega,phi0,psi,delta,alph
 
 N = 200
 seed = 1237 
-h = 1e-12 
+h = 5e-15 
 model = "earth"
         
 #Create a bilby prior object to sample from

### 2023-08-24 docs/manuscript/paper.tex
pre-meeting commit

@@
 \begin{abstract}	
-	\begin{itemize}
-		\item Continuous nHz gravitational waves from individual supermassive black hole binaries are a primary target for Pulsar Timing Array (PTA) observations.
-		\item We present a new and complementary state-space approach to PTA data analysis.
-		\item We show how to combine a Kalman filter and nested sampling for Bayesian inference of the model parameters and model selection.
-		\item The method is tested on synthetic, astrophysically representative data and shown to be able to estimate the full posterior distributions of all parameters of the model, as well calculating the marginalized likelihood for model selection.
-		\item The bias introduced by dropping the `pulsar term' from the state-space model is analysed and discussed. 
-	\end{itemize}
-
+Continuous nHz gravitational waves from individual supermassive black hole binaries are detectable with Pulsar Timing Array observations.
+A novel search strategy is developed wherein intrinsic pulsar timing variations are tracked simultaneously with the modulation induced by a putative gravitational wave.
+A two-step procedure is applied within a state-space framework, where the model parameters are estimated with a Kalman filter, which provides a likelihood for nested sampling.
+This enables self-consistent inference of model parameters and calculation of the Bayesian evidence.
+It is shown via an astrophysically representative software injection campaign that the procedure can distinguish a gravitational wave from pure noise down to a strain of $h_0 \sim 4 \times 10^{15}$, given an inclination angle $\iota = 1.0$ rad.
+Full posterior distributions of model parameters are recovered.
+There is a bias in the estimates of some parameters, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
 \end{abstract}
@@
 \end{enumerate}
-In Section \ref{sec:testing} we will apply this workflow on a synthetic data set.
+In Section \ref{sec:testing} we apply this workflow on a synthetic data set.
 
@@
 	\includegraphics[width=\columnwidth]{images/example_residuals_plot}
-	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars.
-The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}.
+	\caption{Actual (left panels) and synthetic (right panels) phase residuals for NANOGrav millisecond pulsar J0030+0451.
+The actual residuals are obtained via UTMOST pulsar timing program \citep{UTMOST}.
 The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f}.
 For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
-\textcolor{red}{TK: potential sinusoidal feature?
-Need to discuss with Andres re use of baboo.}}
+}
 	\label{fig:qualitative_compare}
@@
 a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation.
-We will explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.
+We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.
 If we consider the median values of the 1D-posteriors we can see that for some parameters the median is almost identically equal to the injected value  (e.g.
@@
 $\mathcal{M}_1$ is the Earth-terms only model, i.e.
-Equation \eqref{eq:measuremen_earth}.
+the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
 The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
 The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
-We can see that there is a general log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
+We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
 We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
-For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3 \times 10^{-15}$.
+For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
 This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
@@
 This will be explored in more detail in Section \ref{sec:parameter_space}.
-
-The small perturbations in the general log-linear $\beta - h_0$ trend arise due to noise in the nested sampling method.
-For these points the sampler converged sub-optimally for $\mathcal{M}_1$ and so the evidence $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)$ is lower than would be expected.
-This is purely a noise artefact; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these perturbations appear at different values of $h_0$, or else are not present at all.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler also dampens these perturbations and smooths the $\beta - h_0$ curve, since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
-
-Below $\beta = 10$ this noise in the sampler starts to dominate and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
+\newline
+
+The data gaps that are evident in the curve for some values of $h_0$ below $\beta =10$ are due to the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
+For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
 That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-This is simply due to the fact that at these small strains the system is completely noise dominated and the two models $\mathcal{M}_1, \mathcal{M}_0$ cannot be separated.
+There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these gaps appear at different values of $h_0$, for $\beta < 10$.
+For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
+For the results presented here we set $n_{\rm live} =2000$.
 \begin{figure}
-	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot}
+	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
 	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
 The horizontal grey dashed line labels the detection tolerance cut-off of $\beta = 10$.
-The minimal detectable strain, below which $\beta < 10$, is $\sim 3 \times 10^{-15}$.
-\textcolor{red}{TK: probably need to regenerate this figure to get rid of the "bumps" which are just artefacts from a poor sampler convergence at that point}}
+The minimal detectable strain, below which $\beta < 10$, is $\sim 3.9 \times 10^{-15}$.
+Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$}
 	\label{fig:bayes}
@@
 To compare each of the posteriors across different noise realisations, we consider the normalised standard deviation in the set of 1000 median values.
-For $\Omega, \Phi_0, \psi, \delta, \alpha$ the normalised standard deviation is $\sim 4 \times 10^{-5}, 0.03,0.004,0.03,0.02$ respectively.
+For $\{\Omega, \Phi_0, \psi, \delta, \alpha \}$ the normalised standard deviation is $\sim \{4 \times 10^{-5}, 0.03,0.004,0.03,0.02 \}$ respectively.
 We can therefore see that the method is able to reliably converge to a consistent MAP estimate.
@@
 
-\textcolor{red}{TK: TBD.
-Maybe lets just change the sky position - it it too expensive to generate a sky map of minimum detectable strain.
-Given the biases I don't know if a PP plot quite makes sense at this stage...}
+\textcolor{red}{TK: TBD}
 
 \subsection{Bias by dropping pulsar terms}\label{sec:bias}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/iota_likelihood_surface}
+	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, orange line) and the pulsar terms model (Equation \ref{eq:measurement}, blue line), with respect to the inclination $\iota$, over the prior space.
+All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
+The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model.
+For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$).
+For the pulsar-terms model the location of the maxima and injected value do coincide.}
+	\label{fig:likelihood_surface_iota}
+\end{figure}
 Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (i.e.
 the median of the 1D marginalised posterior) is biased away from the true injection value.
-This effect is most severe for $\iota$ and $h_0$ (e.g.
-Figures \ref{fig:iota}, \ref{fig:median_distriubutins}) but also present to a lesser extent in other static parameters such as $\alpha$ and $\psi$.
+This effect is most severe for $\iota$ and $h_0$ but also present to a lesser extent in other static parameters such as $\alpha$ and $\psi$, and additionally to an even small extent in $\delta$ and $\Phi_0$.
 In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
@@
 
-In Figure \ref{fig:likelihood_curves_earth} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, as we vary a single static parameter of $\boldsymbol{\theta}_{\rm gw}$, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
-For instance, in Figure \ref{fig:omega_likelihood} we pass different values of $\Omega$ in the range $10^{-9}$ to $10^{-5}$ Hz into the Kalman filter, whilst holding the remaining parameters constant, and plot the retuned log-likelihood as a function of $\Omega$.
-The dashed orange line labels the true injection value whilst the dashed red line labels the location of the maximum log-likelihood.
-Finding the location of this maximum across all parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
-The key observation from Figure \ref{fig:likelihood_curves_earth} is that for many parameters the location of the likelihood optima and the location of the injected value are not the same i.e.
-the orange and red dashed lines are not coincident.
-Consequently Bayesian likelihood inference techniques such as nested sampling settle in optima that do not necessarily align with the true injection value.
+In Figure \ref{fig:likelihood_surface_iota} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, as we vary $\iota$ across the prior space, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
+The solid orange line is the likelihood curve with just the Earth-terms included (i.e.
+using Equation \ref{eq:measuremen_earth} in the Kalman filter) and the solid blue line is the likelihood curve with the additional pulsar terms included (i.e.
+using Equation \ref{eq:measurement}).
+Whilst we have not used the nested sampling method together with the pulsar terms measurement equation in this work, we are able to run the Kalman filter using the pulsar terms and consider how the resulting likelihood varies for different parameter values.
+The dashed lines show the location of the likelihood maxima for each of the two likelihood solutions; finding the location of the maxima across all static parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
+\newline
+
+The key observation from Figure \ref{fig:likelihood_surface_iota} is that for the Earth-terms only solution the location of the likelihood optima and the location of the injected value are not the same.
+Consequently Bayesian likelihood inference techniques settle in optima that do not necessarily align with the true injection value.
 This is the origin of the bias in the 1D marginalised posteriors that we observe.
-Taking a specific example, in Figure \ref{fig:iota_likelihood} we can see that the optima of the likelihood curve lies around $\iota \sim 0.7$ radians, whereas the injected value was $\iota = 1.0$ radians, This difference corresponds to the bias observed in the right-hand panel of Figure \ref{fig:iota} where the median of the marginalised posterior for $\iota$ was $\sim 0.65$ for an injection value of $\iota = 1.0$ radians.
-The discrepancy between $\iota \sim 0.7$ in the log-likelihood curve and $\iota \sim 0.65$ in the nested sampling inference is a result of the likelihood curves being a 1D representation of a multidimensinal problem and the exact shape of the likelihood curves depends on the values of all the other static parameters.
-That is to say, the likelihood curve for $\iota$ will have a different shape with a different optima if we set e.g.
-$\psi$ equal to its likelihood optima value rather than its true injection value.
-
-We can compare the likelihood curves of Figure \ref{fig:likelihood_curves_earth} generated using the earth -terms only measurement equation, Equations \eqref{eq:measuremen_earth} \eqref{eq:g_func_trig_earth}, with the likelihood curves derived using the measurement equation which includes the pulsar terms, Equation \eqref{eq:measurement}.
-Whilst we have not used the nested sampling method together with the pulsar terms measurement equation in this  work, we can run the Kalman filter using Equation \eqref{eq:measurement} and consider how the resulting likelihood varies for different parameter values.
-The results are shown in Figure \ref{fig:likelihood_curves_psr}.
-There are two interesting features in the pulsar-terms likelihood curves.
-The first is that the optima of the likelihood curves now align with the true injected values for all parameters.
-This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.The second is that the log-likelihood curves for $\Omega, \delta$ and $\alpha$ are much more noisy than in the Earth-terms case.
-Whilst clear optima exist above the noise, and whilst local to the true value the curve becomes much more smooth, this noise presents an additional challenge to typical likelihood inference techniques which perform best on smooth likelihoods with a single clear global optima.
-It affects these parameters in particular since Equation \eqref{eq:g_func_trig} contains a phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, which is essentially a combination of these 3 parameters ($ \boldsymbol{n}$ is defined by $\delta$ and $\alpha$).
-When calculating a likelihood we are asking "How likely is this data, given these parameters".
-The inclusion of the pulsar terms results in effectively 2 continuous waves with different frequencies and phase.
-Perturbing e.g.
-$\alpha$ influences wave 1 in some non-liner way and wave 2 in some different non linear way.
-Whether this new set of parameters makes the data more likely depends on the particular combination of those two waves.
-\textcolor{red}{TK: this explanation of why these parameters is clunky because it is not properly clear to myself.
-Think in terms of $g(\theta)$ and $g(\hat{\theta})$}.
+For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ radians, whereas the injected value was $\iota = 1.0$ radians, This difference is comparable to the bias observed in the right-hand panel of Figure \ref{fig:iota} where the median of the marginalised posterior for $\iota$ was $\sim 0.65$ radians.
+The discrepancy between $\iota \sim 0.7$ in the log-likelihood curve and $\iota \sim 0.65$ in the nested sampling inference is a result of the likelihood curves being a 1D representation of a multidimensional problem.
+The exact shape of the likelihood curve naturally depends on the values of all of the other parameters that we hold constant.
+In the nested sampling case the estimate for $\psi$ is sightly biased at $\psi \sim 2.60$ radians compared to the true injected value of $\psi = 2.50$ radians that is used for the results of Figure \ref{fig:likelihood_surface_iota}.
+\newline
+
+In contrast to the Earth-terms solution, for the solution which includes the pulsar terms the optima of the likelihood curve now aligns with the true injected value.
+This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.
+Similar results are observed for the other parameters which exhibit a bias in the nested sampling inference.
+\newline
+
+Why does dropping the pulsar terms bias these parameters in particular, whilst leaving others unaffected?
+For instance, why is the bias induced in $\iota$ so strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
+When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is effective modulated by the interference of two equal-amplitude, phase shifted waves (c.f.
+Equation \ref{eq:measurement}).
+These additional modulations from the phase shifted wave influence the amplitude of the resulting wave interference wave, but do not affect its frequency.
+When we model this data using the Kalman filter without the pulsar terms, the additional amplitude modulations due to the extra phase shifted wave must be described by the single wave (i.e.
+Earth terms only) model.
+As a result the single wave model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude of the single wave.
+Consequently the parameters which are most affected are those which determine the amplitude of the GW.
+Terms which influence the amplitude strongly such as $h_0$ and $\iota$ are strongly affected and exhibit a strong bias.
+Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
+The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
+
+For loud sources with high SNR the uncertainty in the 1D posterior is less than the bias and so these biases corrupt the inferred MAP estimate of the parameter.
+
+For quieter sources
+
+For most parameters the bias is less than the uncertainty
+
+For loud sources
+
+for less loud sources the unertainty dominates the errors
 
@@
 We have only considered a single example system to explore the effect of the bias.
-Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for each of the parameters, it is present and will need to be accounted for.
+Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for each of the parameters,\textbf{biases will be different for different GW systems}  it is present and will need to be accounted for.
 This bias is not particular to our method but a shared feature across likelihood based methods that do not include the pulsar terms in the inference model.
@@
 
-%\begin{figure*}
-%	\setkeys{Gin}{width=\linewidth}   
-%	
-%	\begin{subfigure}[b]{0.22\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_omega_earth}
-%		\caption{$\Omega$}
-%		\label{fig:omega_likelihood}
-%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.22\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_phi0_earth}
-%		\caption{$\Phi_0$}
-%	\end{subfigure}
-%	\hfill	
-%	\begin{subfigure}[b]{0.22\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_psi_earth}
-%		\caption{$\psi$}
-%	\end{subfigure}
-%	\begin{subfigure}[b]{0.22\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_iota_earth}
-%		\caption{$\iota$}
-%			\label{fig:iota_likelihood}
-%	\end{subfigure}
-%	\medskip
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_delta_earth}
-%		\caption{$\delta$}
-%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_alpha_earth}
-%		\caption{$\alpha$}
-%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics{images/likelihood_h_earth}
-%		\caption{$h$}
-%	\end{subfigure}
-%	
-%	\caption{}\label{fig:likelihood_curves_earth}
-%\end{figure*}
-%
-%
-%
-%
-%\begin{figure*}
-%	\setkeys{Gin}{width=\linewidth}   
-%	
-%	\begin{subfigure}[b]{0.22\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_omega_psr}
-%		\caption{$\Omega$}
-%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.22\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_phi0_psr}
-%		\caption{$\Phi_0$}
-%	\end{subfigure}
-%	\hfill	
-%	\begin{subfigure}[b]{0.22\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_psi_psr}
-%		\caption{$\psi$}
-%	\end{subfigure}
-%	\begin{subfigure}[b]{0.22\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_iota_psr}
-%		\caption{$\iota$}
-%	\end{subfigure}
-%	\medskip
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_delta_psr}
-%		\caption{$\delta$}
-%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics[width=\textwidth]{images/likelihood_alpha_psr}
-%		\caption{$\alpha$}
-%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.3\textwidth}
-%		\includegraphics{images/likelihood_h_psr}
-%		\caption{$h$}
-%	\end{subfigure}
-%	
-%	\caption{As Figure \ref{fig:likelihood_curves_earth} but now including the pulsar terms in the measurement equation.}\label{fig:likelihood_curves_psr}
-%\end{figure*}
-%
+not just present in our formulation but in all works. whetehr this is important depends.... real astrophysica sources
 
@@
 	\includegraphics[width=\textwidth]{images/Kalman_example_both}
-	\caption{}
+	\caption{\textcolor{red}{TK: For consideration. Do we prefer this figure to Figure \ref{fig:four figures}}?}
 	\label{fig:kalmanexampleboth}
@@
 %
+%
+
+%
+%
+%
+%
+%For instance, in Figure \ref{fig:omega_likelihood} we pass different values of $\Omega$ in the range $10^{-9}$ to $10^{-5}$ Hz into the Kalman filter, whilst holding the remaining parameters constant, and plot the retuned log-likelihood as a function of $\Omega$.
 %
@@
 %
+%The second is that the log-likelihood curves for $\Omega, \delta$ and $\alpha$ are much more noisy than in the Earth-terms case.
+Whilst clear optima exist above the noise, and whilst local to the true value the curve becomes much more smooth, this noise presents an additional challenge to typical likelihood inference techniques which perform best on smooth likelihoods with a single clear global optima.
+It affects these parameters in particular since Equation \eqref{eq:g_func_trig} contains a phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, which is essentially a combination of these 3 parameters ($ \boldsymbol{n}$ is defined by $\delta$ and $\alpha$).
+When calculating a likelihood we are asking "How likely is this data, given these parameters".
+The inclusion of the pulsar terms results in effectively 2 continuous waves with different frequencies and phase.
+Perturbing e.g.
+$\alpha$ influences wave 1 in some non-liner way and wave 2 in some different non linear way.
+Whether this new set of parameters makes the data more likely depends on the particular combination of those two waves.
+\textcolor{red}{TK: this explanation of why these parameters is clunky because it is not properly clear to myself.
+Think in terms of $g(\theta)$ and $g(\hat{\theta})$}.
+
+%
 %.

### 2023-08-24 notebooks/plot_likelihood_surfaces_3D.py

@@ -1,0 +1,79 @@
+
+
+import numpy as np
+import matplotlib.pyplot as plt  
+
+
+
+
+from matplotlib import ticker, cm
+import matplotlib.pyplot as plt 
+
+import scienceplots
+
+
+
+
+import matplotlib
+
+plt.style.use('science')
+fig = plt.figure(figsize=(6,6))
+ax = fig.add_subplot(111, projection='3d')
+
+
+
+
+path_to_earth_model = 'likelihood_surface_h_iota_mm_earth.npz'
+path_to_earth_model = 'likelihood_surface_h_iota_mm_earth_broader.npz'
+path_to_earth_model = 'likelihood_surface_h_iota_mm_earth_coarse.npz'
+
+
+def load_and_plot(path,ax):
+
+
+    #Load the data
+    data = np.load(path)
+    iota_values = data['x']
+    h_values = np.log10(data['y'])
+    surface_pulsar = data['z'] 
+    surface_pulsar = surface_pulsar / np.abs(np.max(surface_pulsar)) #Normalize
+
+
+    double_log = False #Do you want to log the likelihood again?
+    if double_log:
+        surface_pulsar = np.log10(np.abs(surface_pulsar))
+        #Extract location of minima
+        iota_idx, h_idx = np.unravel_index(surface_pulsar.argmin(), surface_pulsar.shape)
+    else:
+        iota_idx, h_idx = np.unravel_index(surface_pulsar.argmax(), surface_pulsar.shape)
+
+    
+    
+    xc = iota_values[iota_idx]
+    yc = h_values[h_idx]
+    zc = surface_pulsar[iota_idx,h_idx]
+
+
+    #Cast to 2D mesh
+    X,Y = np.meshgrid(h_values,iota_values)
+    lx = len(iota_values)
+    ly = len(h_values)
+    z = np.reshape(surface_pulsar, (lx, ly))
+
+
+
+    #Plot colormap
+    ax.plot_surface(X, Y, z,alpha=0.5)
+   
+
+    #Config
+  
+    ax.scatter(yc,xc,zc, s=20,c='C3')
+    fs = 20
+    ax.set_xlabel(r'$h_0$', fontsize=fs)
+    ax.set_ylabel(r'$\iota$', fontsize=fs)
+    ax.xaxis.set_tick_params(labelsize=fs-4)
+    ax.yaxis.set_tick_params(labelsize=fs-4)
+    
+load_and_plot(path_to_earth_model,ax)
+plt.show()

### 2023-08-25 docs/manuscript/paper.tex
pp plots at a larger h value
PP plot for just omega

@@
 The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
-\subsubsection{Plane GW perturbation}
+\subsubsection{Plane GW perturbation}\label{sec:plane_gw}
 We consider a gravitational plane wave from a single,distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
@@
 The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$.
-We take $n_{\rm live} = 500$ for all results presented in this work.
+Unless stated otherwise we take $n_{\rm live} = 500$ for all results presented in this work.
 \newline
@@
 The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16}$.
-For our artificial dataset we set $\sigma = 10^{21}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a constrained prior.
+For our artificial dataset we set $\sigma = 10^{-21}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a constrained prior.
 \newline
@@
 
-\textcolor{red}{TK: TBD}
+We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
 
@@
 	\centering
-	\includegraphics[width=\columnwidth]{images/iota_likelihood_surface}
+	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
 	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, orange line) and the pulsar terms model (Equation \ref{eq:measurement}, blue line), with respect to the inclination $\iota$, over the prior space.
@@
 Equation \ref{eq:measurement}).
-These additional modulations from the phase shifted wave influence the amplitude of the resulting wave interference wave, but do not affect its frequency.
-When we model this data using the Kalman filter without the pulsar terms, the additional amplitude modulations due to the extra phase shifted wave must be described by the single wave (i.e.
-Earth terms only) model.
-As a result the single wave model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude of the single wave.
+Explicitly, we have the ``Earth-terms wave" with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
+The net interference wave that modulates $f_{\rm p}$ is
+\begin{eqnarray}
+	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right)
+\end{eqnarray}
+The additional modulations from $y_{\rm pulsar}$ influence the amplitude of the resulting interference wave, but do not affect its frequency.
+When we model this data using the Kalman filter without the pulsar terms, the additional amplitude modulations from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
+As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude $A$ of the single wave.
 Consequently the parameters which are most affected are those which determine the amplitude of the GW.
@@
 Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
-The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
-
-For loud sources with high SNR the uncertainty in the 1D posterior is less than the bias and so these biases corrupt the inferred MAP estimate of the parameter.
-
-For quieter sources
-
-For most parameters the bias is less than the uncertainty
-
-For loud sources
-
-for less loud sources the unertainty dominates the errors
-
-We have shown in this section how the dropping of the pulsar terms results in a bias in the inferred parameters, not just in the sky localisation parameters  \citep[e.g.][]{Zhupulsarterms,Chen2022} but additionally $\psi, h_0$ and $\iota$.
-We have only considered a single example system to explore the effect of the bias.
-Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for each of the parameters,\textbf{biases will be different for different GW systems}  it is present and will need to be accounted for.
+The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
+\newline
+
+In order to ascertain how important the bias is for PTA continuous wave searches requires a thorough exploration of the astrophysical parameter space which we do not carry out in this work.
+The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
+If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the 1D marginalised posterior may dominate and the bias may not be important for searches.
+Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties then the bias may come to systematically affect the inferred parameters.
+\newline
+
+We have shown in this section how the dropping of the pulsar terms results in a bias in the inferred parameters, not just in the sky localisation parameters  \citep[e.g.][]{Zhupulsarterms,Chen2022} but additionally all parameters which influence the wave amplitude.
+Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for our example system, it is present and may need to be accounted for for analysis in other parts of the parameter space.
 This bias is not particular to our method but a shared feature across likelihood based methods that do not include the pulsar terms in the inference model.
-The inclusion of the pulsar terms in the methods presented in this work will be a key goal of a future work, which will alleviate these bias problems.
-
-not just present in our formulation but in all works. whetehr this is important depends.... real astrophysica sources
+The inclusion of the pulsar terms in the methods presented in this work in order to alleviate the bias will be a key goal of a future work.
 
@@
 We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
-
-The natural first extension of this work is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
-
+There are a number of potential extensions of this work which we now discuss.
+\newline
+
+The natural first extension is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
 Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred 1D marginalized posteriors for the model parameters.
-
-Biases in synthetic datasets from dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.
-Alternatively, if one wishes to continue to use just the Earth terms it would be highly desirable to
-
-We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up NANOGrav.
+Biases in sky localisation for synthetic datasets as a result of dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.
+Alternatively, if the method continues to be used with solely the Earth terms then it would be highly desirable to make systematic, quantitative estimates of the incurred biases in the model parameters across an astrophysically representative parameter space.
+\newline
+
+We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up NANOGrav (Section \ref{sec:synt_pta}).
 Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA.
-Indeed, due to the increased computational demands as pulsars are added to the arrays it may be advantageous to optimally select a subset of the total number of available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.
+Indeed, since adding pulsars to PTAs increases the computational demands of the data analysis it may be advantageous to optimally select a subset of the total available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.
+\newline
 
 Throughout this work we have assumed that all pulsars are observed at the same time, with a constant time sampling.
-In practice this assumption does not hold and pulsars will be observed with different cadences at different time intervals.
-Applying this method to non-constant time sampling is a key further extension.
-
-We have also assumed that there is only a single, quasi-monochromatic GW source that influences the pulsar frequencies.However, it may be possible to resolve multiple continuous GW source simultaneously.
-Our method naturally extends to multiple GW sources; Equation \ref{eq:measurement} is modified to become the linear superposition of multiple GWs.
+In practice this assumption does not hold and different pulsars will be observed with different cadences at different times.
+Extending this method to non-constant time sampling is straightforward and the performance of the method should be evaluated in this regime.
+\newline
+
+We have taken our GW source to be monochromatic, i.e.
+$\Omega$ is a fixed parameter with no time dependence.
+Whilst this is an astronomically well-justified assumption (see discussion in Section \ref{sec:plane_gw}), and appropriate for this initial methods work, it would also be of interest to extend the state-space construction to enable the GW frequency to evolve in time.
+The GW source is monochromatic over the decadal PTA observation period, but the timescale set by the Earth-pulsar distance is comparable to the SMBHB evolution timescale \citep{Sesana2010}.
+For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \ref{eq:f_evolution}) the GW source frequency evolution manifests as an incoherent source of noise in the pulsar terms, whilst for  $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms can affect the phase coherency of the Earth terms \citep{Perrodin2018}.
+Careful consideration of the evolution of the GW frequency will be important as we look to extend our method to include the pulsar terms in the inference model.
+\newline
+
+We have also assumed that there is only a single GW source that influences the received pulsar frequencies.
+However, it may be possible to resolve multiple continuous GW sources simultaneously \citep{PhysRevD.85.044034}.
+Our method naturally extends to detecting multiple GW sources simultaneously; Equation \ref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
+Evaluating the performance of the method under these conditions would be a further interesting pursuit.
 The stochastic background itself is the incoherent summation of many individual GW sources.
-It therefore stands to reason that we could extend this method to also search for a stochastic GW background without requiring cross-correlating the pulsar residuals to uncover the Hellings-Downs curve.
-
-\textcolor{red}{TK: Lets pre-empt reviewer and make it explicit WHY we are not using TOAs}
-We have formulated the state space model in terms of frequencies, where
-
+It therefore stands to reason that it may be possible to extend this method to also search for a stochastic GW background, providing a complementary approach to detect the stochastic background without requiring a cross-correlation of the pulsar residuals to uncover the Hellings-Downs curve.
+\newline
+
+\textcolor{red}{TK: Lets also pre-empt reviewer and make it explicit why we are using frequencies rather than TOAs. Need to chat with group to get this clear in my head.}
+
+%https://arxiv.org/pdf/2107.03047.pdf
 \section{Conclusion}
 
-\textcolor{red}{TK: TBD}
+In this paper we demonstrate a new method for the detection and parameter estimation of GWs from monochromatic SMBHBs.
+We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method rather than subtracting the ensemble-averaged statistics.
+That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW induced modulations.
+This enables the inference of GW parameters conditioned on the specific observed realisation of the noisy data.
+
+We introduce the Kalman filter in order to track the intrinsic state evolution of the pulsar and combine this with a Bayesian sampling framework to estimate the posterior distributions of each parameter of the system, as well as the associated evidence (marginalised likelihood) of the model.
+
+We test our method on synthetic data. We focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
+
+The parameters of the model are accurately recovered and the minimal detectable strain is estimated via comparing the GW model with a null, noise-only model to describe the data.
+
+The method is then further tested over 1000 noise realisations.
+Consistent MAP estimates of the parameters across all noise realisations are obtained for the majority of parameters, whilst the structural weak-identifiability of $\iota$ and $h_0$ results in a larger variance in the MAP estimates between individual noise realisations.
+
+\textcolor{red}{Validation tests with 200 randomly sampled parameter vectors result in reliable posterior distributions
+that accurately contain the injected parameter vectors, which indicates that the percent errors cited above are consistent with statistical
+fluctuations. This test confirms that the method can be applied to a
+wide variety of pulsars without hand-tuning.}
+
+Biases are observed in t...
 

### 2023-08-25 docs/manuscript/paper.tex
now do the same, but searching over alpha

@@
 
+It is also important to test the method in different regions of the parameter space.
+
+To this end we ...
+
 \subsection{Bias by dropping pulsar terms}\label{sec:bias}
@@
 \section{Conclusion}
-
-In this paper we demonstrate a new method for the detection and parameter estimation of GWs from monochromatic SMBHBs.
-We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method rather than subtracting the ensemble-averaged statistics.
+In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
+We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
 That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW induced modulations.
 This enables the inference of GW parameters conditioned on the specific observed realisation of the noisy data.
-
 We introduce the Kalman filter in order to track the intrinsic state evolution of the pulsar and combine this with a Bayesian sampling framework to estimate the posterior distributions of each parameter of the system, as well as the associated evidence (marginalised likelihood) of the model.
-
-We test our method on synthetic data. We focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
-
-The parameters of the model are accurately recovered and the minimal detectable strain is estimated via comparing the GW model with a null, noise-only model to describe the data.
-
+We test our method on synthetic data.
+We initially focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
+The parameters of the model are accurately recovered and the minimal detectable strain is estimated to be $h_0 \sim 4 \times 10^{-15}$.
 The method is then further tested over 1000 noise realisations.
-Consistent MAP estimates of the parameters across all noise realisations are obtained for the majority of parameters, whilst the structural weak-identifiability of $\iota$ and $h_0$ results in a larger variance in the MAP estimates between individual noise realisations.
-
+Consistent MAP estimates of the parameters across all noise realisations are obtained for the majority of parameters, whilst the structural weak-identifiability of $\iota$ and $h_0$ results in a larger variance in the MAP estimates between individual noise realisations for these parameters.
 \textcolor{red}{Validation tests with 200 randomly sampled parameter vectors result in reliable posterior distributions
@@
 wide variety of pulsars without hand-tuning.}
-
-Biases are observed in t...
+As a result of dropping the pulsar terms from the inference model, biases are observed in some of the model parameters.
+The extension of this method to include the pulsar term is a key challenge for a future work.
 
@@
 \section{Kalman recursion equations} \label{sec:kalman}
-
-\textcolor{red}{TK: This section needs cleaning up}
-
-The linear Kalman filter operates on temporally discrete measurements which are related to unobservable system states via a linear transformation
-\begin{equation}
-	\bar{z} = \bar{H} \bar{x} + \bar{v}
-\end{equation}
-where $\bar{H}$ is the measurement matrix and $\bar{v}$ is a Gaussian measurement noise. The underlying states evolve according to are the state-space dynamical equation 
-\begin{equation}
-	\dot{\bar{x}} = \bar{F} \bar{x} + \bar{G}\bar{u} + \bar{w}
-\end{equation}
+The linear Kalman filter operate on temporally discrete, noisy measurements $\boldsymbol{Y}_k$ which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$ via a linear transformation:
+\begin{equation}
+	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \label{eq:kalman1}
+\end{equation}
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
+
+The Kalman filter takes the underlying states to evolve according to
+\begin{equation}
+	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \label{eq:kalman2}
+\end{equation}
+
+In order to deploy the Kalman filter we must map our state-space model described in X onto the Kalman filter structure of Equatons
+
+The according the dynamical equation
+\begin{equation}
+	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{N} dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
+\end{equation}
+where $\boldsymbol{A}$ is the state matrix, $\boldsymbol{N}$ the input or control matrix and $d\boldsymbol{B}(t)$ denotes a Wiener process with covariance matrix $\boldsymbol{\Sigma}$.
+
+The general solution to Equation \ref{eq:kalmn2}. an Ornstein-Uhlenbeck process
+
+By comparing Equation \ref{eq:kalman1} with Equation \ref{eq:measurement} and Equation \ref{eq:kalmn2} with Equation \ref{eq:frequency_evolution} it is immediately clear how the
+
 for the system dynamics matrix,$\bar{F}$,  control model $G$, control vector $\bar{u}$, and $w$ a stochastic zero-mean process.

### 2023-08-25 py_src/config_batch_pp.py
generate a different set of alphas

@@ -62,7 +62,7 @@ with open('batch.sh','w') as b:
 
        
    
-        arg_name = f"OMEGA_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
+        arg_name = f"ALPHA_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
         print(arg_name)
         create_slurm_job(arg_name,h,model,seed,p['omega_gw'],p['phi0_gw'],p['psi_gw'],p['delta_gw'],p['alpha_gw'])
         b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")

### 2023-08-28 docs/manuscript/paper.tex
generate a different set of alphas
complete manuscript

@@
 % If you need two or more lines of authors, add an extra line using \newauthor
-\author[Kimpson]{Kimpson$^{1}$, Melatos$^{2}$, O'Leary, Evans$^{3}$, others, etc. etc.%
+\author[Kimpson]{Kimpson$^{1}$, Melatos$^{2}$, O'Leary, Evans$^{3}$, Carlin others, etc. etc.%
 \thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
@@
 \begin{abstract}	
-Continuous nHz gravitational waves from individual supermassive black hole binaries are detectable with Pulsar Timing Array observations.
+Continuous nHz gravitational waves from individual supermassive black hole binaries may be detectable with Pulsar Timing Array observations.
 A novel search strategy is developed wherein intrinsic pulsar timing variations are tracked simultaneously with the modulation induced by a putative gravitational wave.
@@
 \subsection{Spin evolution} \label{sec:psr_frequency}
-
 A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}.
@@
 \section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
-The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e.
+The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar, and parameters of the GW source, i.e.
 \begin{equation}
@@
 \begin{equation}
-	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
+	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
 \end{equation}
@@
 Model selection and the specification of the null model is described in Section \ref{sec:model_selection}.
-
-%The choice of prior distributions on the model parameters are described in
+A complete summary of the method is presented in Section \ref{sec:methodsummary}.
 
@@
 \begin{equation}
-	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}}
+	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
 \end{equation}
@@
 The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$.
-Unless stated otherwise we take $n_{\rm live} = 500$ for all results presented in this work.
+Unless stated otherwise we take $n_{\rm live} = 1000$ for all results presented in this work.
 \newline
@@
 
-\subsection{Summary of workflow}
+\subsection{Summary of workflow}\label{sec:methodsummary}
 To summarise the above, the order of operations in our method for estimating the posteriors on the static parameters and calculating the marginal likelihood for the inference model is as follows:
@@
 	\item Return $\log \mathcal{L}$ to the nested sampler
-	\item Nested sampler updates $\boldsymbol{\theta}$
+	\item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$
 	\item Repeat
@@
 For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$.
-The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Fig \ref{fig:pulsar_distrib}.
+The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Figure \ref{fig:pulsar_distrib}.
 The data is acquired via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
@@
 $T = 1$ week.
-The frequency noise parameter can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as
+The frequency process noise parameter $\sigma^{(n)}$ can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as
 \begin{eqnarray}
@@
 \begin{eqnarray}
-	\phi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
+	\xi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
 \end{eqnarray}
@@
 Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$ and take it to be known \textit{a priori} rather than a parameter to be inferred.
-Note that $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar; whilst for our construction every pulsar is sampled at the same time and so would experience the same realisation of the noise, in actual pulsar astronomy different pulsars will be sampled at different times and so subject to different noise realisations.
+Note that $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar; whilst for our construction every pulsar is sampled at the same time and so would experience the same realisation of the noise, in actual pulsar astronomy different pulsars will be sampled at different times by different radio interferometers and so subject to different noise realisations.
 \newline
@@
 The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16}$.
-For our artificial dataset we set $\sigma = 10^{-21}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a constrained prior.
+For our artificial dataset we set $\sigma = 10^{-21}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
 \newline
@@
 Since we are not setting priors on $\gamma$ or $d$ this reduces the dimensionality of the parameter space to $7 + 3N$.
-We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space.
+We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space, c.f.
+Equation \ref{eq:psrparams}.
 Explicitly,
@@
 The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.}
+The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.
+The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
 	\label{tab:parameters_and_priors}
@@
 	We defer the inclusion of the pulsar terms in the measurement equation to a future work.
-As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation Equation \eqref{eq:measurement} is no longer a function of the pulsar distance.
+As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
 We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
@@
 
-% and in Figure \ref{fig:example_psr_params} for . 
-% 
-% Regarding $\boldsymbol{\theta}_{\rm psr, reduced}$, the frequency parameters $f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1)$ are generally well recovered for all pulsars, as are the noise amplitude parameters $\sigma^{(n)}$.
-\textcolor{red}{TK: Could do with some extra discussion here re $\boldsymbol{\theta}_{\rm psr, reduced}$ since they are not perfectly recovered, but the important thing is that the GW parameters and the GW parameters are the interesting bit...}
-
 \begin{figure*}
 	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
-	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for a single realisation of the system noise.
-The vertical orange lines indicate the true injected values, c.f.
-Table \ref{tab:parameters_and_priors}.
+	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The vertical orange lines indicate the true injected values.
 The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
 We are able to accurately estimate each parameter of interest.
-$\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
+We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
 	\label{fig:corner_plot_1}
 \end{figure*}
-
-%\begin{figure*}%
-%	\centering
-%	\subfloat[\centering $f_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_f} }}%
-%	\qquad
-%	\subfloat[\centering $\dot{f}_{\rm em}^{(n)}(t_1)$]{{\includegraphics[width=0.3 \textwidth]{images/1237_broad_fdot} }}%
-%	\qquad
-%	\subfloat[\centering $\sigma^{(n)}$]{{\includegraphics[width=0.3\textwidth]{images/1237_broad_sigma_p} }}%
-%	
-%	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$.
-We show only a selection of the first 5 pulsars rather than the full PTA.
-\textcolor{red}{TK: these figures are a little small to resolve nicely}}
-%	\label{fig:example_psr_params}%
-%\end{figure*}
 
@@
 the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
-The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
 The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
@@
 $T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
-This will be explored in more detail in Section \ref{sec:parameter_space}.
+\ref{sec:parameter_space}.
 \newline
@@
 The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
-The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance.}
+The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f.
+Figure \ref{fig:likelihood_surface}.}
 	\label{fig:corner_plot_2}
@@
 	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
-	\caption{1D marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$.
+	\caption{1D marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel).
 The thick vertical orange line labels the true injected value.
@@
 \end{figure*}
-
 The above discussion was for a single realisation of the noise processes.
@@
 This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
-Whilst a single, unique maxima exists (labelled by the red point in the figure), along ridge in the parameter space the likelihood values are all very similar.
+Whilst a single, unique maxima exists (labelled by the red point in the figure), along the ridge in the parameter space the likelihood values are all very similar.
 Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
@@
 We can see that whilst the variance between the individual posteriors is large in the original case, once we fix $h_0$ the variance decreases dramatically and the sampling algorithm repeatedly converges consistent estimates of $\iota$.
-There is a notable bias in Figure \ref{fig:iota} in the estimates for $\iota$ inferred for the fixed $h_0$ case relative to the true injected value; whilst the variance has decreased, the mean accuracy of the estimates has also decreased.
+For the fixed $h_0$ case, evident in Figure \ref{fig:iota} is a strong bias of $\sim 0.35$ radians in the estimates for $\iota$ relative to the true injected value.
+Whilst the variance in the estimates across the noise realisations has decreased, the mean accuracy of the estimates has also decreased.
 An equivalent bias in $h_0$ is observed if we fix $\iota$.
@@
 
-We now consider the complete set of 1000 noise realisations, rather than the subset of 9 realisations in the preceding discussion.
+We now consider the complete set of $10^3$ noise realisations, rather than the subset of 9 realisations in the preceding discussion.
 For each realisation of the noise we infer a marginalised 1-D posterior for each parameter.
-For this injected strain ($h_0 = 10^{12}$) the SNR is large (c.f.
+For this injected strain ($h_0 = 10^{-12}$) the signal-to-noise ratio (SNR) is large (c.f.
 Figure \ref{fig:bayes}) and consequently the posteriors are typically narrow compared to the prior space and well described by their median value.
-To compare each of the posteriors across different noise realisations, we consider the normalised standard deviation in the set of 1000 median values.
-For $\{\Omega, \Phi_0, \psi, \delta, \alpha \}$ the normalised standard deviation is $\sim \{4 \times 10^{-5}, 0.03,0.004,0.03,0.02 \}$ respectively.
+To compare each of the posteriors across different noise realisations, we consider the normalised standard deviation in the set of $10^3$ median values i.e.
+\begin{equation}
+	\rho \left( \boldsymbol{\theta}\right)= \frac{1}{\mu(\boldsymbol{\theta})} \sqrt{ \frac{\sum_{i=1}^{i=10^3}\left[x_i(\boldsymbol{\theta}) - \mu(\boldsymbol{\theta}) \right]}{10^3}} 
+\end{equation} 
+where $\mu(\boldsymbol{\theta})$ is the mean value across the distribution of $10^3$ samples for parameter $\boldsymbol{\theta}$, and $x_i$ the value of the $i$-th sample.
+For the parameters $\{\Omega, \Phi_0, \psi, \delta, \alpha \}$ we have $\rho \sim \{4 \times 10^{-5}, 0.03,0.004,0.03,0.02 \}$ respectively.
 We can therefore see that the method is able to reliably converge to a consistent MAP estimate.
-Conversely, as we saw before, the MAP estimates for $\iota$ and $h$ are not consistent across different noise realisations, with normalised standard deviation of $0.3$ and $0.1$ respectively.Whilst the estimates are consistent across different noise realisations, for some parameters a small bias is observed between the MAP estimate and the injected value.
-This was evident in Figure \ref{fig:iota} for $\iota$ (and equivalently $h_0$s) which exhibits a strong bias of $\sim 0.35$ radians, but over the 1000 realisations biases are also observed in $\psi$ ($\sim 0.1$ radians) and $\alpha$ ($\sim 0.04$ radians).
+Conversely, as we saw before, the MAP estimates for $\iota$ and $h$ are not consistent across different noise realisations, with $\rho \sim 0.3$ and $\rho \sim 0.1$ respectively.
+This is due to the weak-identifiability issue that we have discussed.
+\newline
+
+We saw previously strong biases in $\iota$ and $h_0$ when considering just 9 noise realisations.
+These biases persist with a similar magnitude when we consider the full set of $10^3$ noise realisations.
+Moreover, smaller biases are also observed in $\psi$ ($\sim 0.1$ radians) and $\alpha$ ($\sim 0.04$ radians) as well as smaller still biases in $\delta$ and $\Phi_0$.
 These biases are a result of dropping the pulsar terms from the measurement equation and will be discussed in Section \ref{sec:bias}
-
 \subsection{Exploring a broader parameter space} \label{sec:parameter_space}
-
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/pp_plot_new}
+	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}.
+Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$.
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad in order to maintain an approximately constant SNR.
+The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals.
+Well estimated posteriors should fall along the diagonal $y=x$.
+Many parameters show evidence of being over-constrained due to a modelling bias.
+}
+	\label{fig:parameter_space}
+\end{figure}
 We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
-
 It is also important to test the method in different regions of the parameter space.
-
-To this end we ...
+To this end we consider 200 simulated injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining 5 parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space.
+For each simulated injection we can then attempt to recover the posterior distributions for each parameter.
+To summarise the results across the parameter space we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+A PP plot describes the fraction of the total number of injected parameters which are included within a given confidence interval of the estimated posterior with respect to the confidence interval itself.
+In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage confidence interval.
+The results are shown in Figure \ref{fig:parameter_space}.
+The shaded grey contoured regions label the first 3 $\sigma$ significance levels i.e.
+how much deviation from the $y=x$ diagonal one might expect statistically given 200 simulations.
+We can see that it only $\Omega$ is well-behaved and falls within the shaded region.
+The other parameters deviate from the $y=x$ diagonal.
+The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$.
+The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value confidence intervals than would be expected statistically, and vice versa.
+This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally highly confident i.e.
+narrow, but are systematically biased away from the true injection values.
+This manifests in the PP plot as posteriors which are insufficiently narrow to capture the injected value within the appropriate confidence interval.
+An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.
 
@@
 	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
-	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, orange line) and the pulsar terms model (Equation \ref{eq:measurement}, blue line), with respect to the inclination $\iota$, over the prior space.
+	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and the Earth terms model for Earth terms only synthetic data (green line), with respect to the inclination $\iota$, over the prior space.
 All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
@@
 For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$).
-For the pulsar-terms model the location of the maxima and injected value do coincide.}
+For the pulsar-terms model the location of the maxima and injected value do coincide.
+10 noise realisations are plotted but the solutions are completely overlaid.}
 	\label{fig:likelihood_surface_iota}
@@
 the median of the 1D marginalised posterior) is biased away from the true injection value.
-This effect is most severe for $\iota$ and $h_0$ but also present to a lesser extent in other static parameters such as $\alpha$ and $\psi$, and additionally to an even small extent in $\delta$ and $\Phi_0$.
+This effect is most severe for $\iota$ and $h_0$ but also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
 In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
@@
 In Figure \ref{fig:likelihood_surface_iota} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, as we vary $\iota$ across the prior space, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
-The solid orange line is the likelihood curve with just the Earth-terms included (i.e.
-using Equation \ref{eq:measuremen_earth} in the Kalman filter) and the solid blue line is the likelihood curve with the additional pulsar terms included (i.e.
-using Equation \ref{eq:measurement}).
-Whilst we have not used the nested sampling method together with the pulsar terms measurement equation in this work, we are able to run the Kalman filter using the pulsar terms and consider how the resulting likelihood varies for different parameter values.
-The dashed lines show the location of the likelihood maxima for each of the two likelihood solutions; finding the location of the maxima across all static parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
+We do this for 10 separate realisations of the system noise.
+We consider 3 separate solutions: \textit{(i)} The solid blue line is the likelihood curve with just the Earth-terms included (i.e.
+using Equation \ref{eq:measuremen_earth} in the Kalman filter), \textit{(ii)} the solid orange line is the likelihood curve with the additional pulsar terms included (i.e.
+using Equation \ref{eq:measurement}), \textit{(iii)} the solid green line is the solution where we drop the pulsar terms from both the synthetic data and the measurement equation.
+Whilst we have not used the nested sampling method with either solutions \textit{(ii)} or \textit{(iii)} in this work, we are able to run the Kalman filter either using the pulsar terms in the model or else dropping them from the synthetic data consider how the resulting likelihood varies for different parameter values.
+The dashed lines show the location of the likelihood maxima for each of the likelihood solutions; finding the location of the maxima across all static parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
 \newline
@@
 In contrast to the Earth-terms solution, for the solution which includes the pulsar terms the optima of the likelihood curve now aligns with the true injected value.
+The corollary is also true; if we drop the pulsar terms from the synthetic data and use the Earth terms equation for inference the optima is also correctly located.
 This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.
@@
 \begin{eqnarray}
-	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right)
+	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ .
 \end{eqnarray}
@@
 When we model this data using the Kalman filter without the pulsar terms, the additional amplitude modulations from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
-As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude $A$ of the single wave.
+As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude of the single wave.
 Consequently the parameters which are most affected are those which determine the amplitude of the GW.
@@
 The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
+This hierarchy of biases for each parameter is also what we observe in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
 \newline
@@
 We have also assumed that there is only a single GW source that influences the received pulsar frequencies.
-However, it may be possible to resolve multiple continuous GW sources simultaneously \citep{PhysRevD.85.044034}.
+However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
 Our method naturally extends to detecting multiple GW sources simultaneously; Equation \ref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
@@
 Consistent MAP estimates of the parameters across all noise realisations are obtained for the majority of parameters, whilst the structural weak-identifiability of $\iota$ and $h_0$ results in a larger variance in the MAP estimates between individual noise realisations for these parameters.
-\textcolor{red}{Validation tests with 200 randomly sampled parameter vectors result in reliable posterior distributions
-that accurately contain the injected parameter vectors, which indicates that the percent errors cited above are consistent with statistical
-fluctuations. This test confirms that the method can be applied to a
-wide variety of pulsars without hand-tuning.}
-As a result of dropping the pulsar terms from the inference model, biases are observed in some of the model parameters.
+Exploration of a broader parameter space via 200 randomly sampled parameter vectors highlights biases in some parameters as a result of dropping the pulsar terms from the inference model.
+The cause of these biases is discussed.
 The extension of this method to include the pulsar term is a key challenge for a future work.
 
-\newpage 
 \appendix
@@
 where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
-
 The Kalman filter takes the underlying states to evolve according to
@@
 \end{equation}
-
-In order to deploy the Kalman filter we must map our state-space model described in X onto the Kalman filter structure of Equatons
-
-The according the dynamical equation
-\begin{equation}
-	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{N} dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
-\end{equation}
-where $\boldsymbol{A}$ is the state matrix, $\boldsymbol{N}$ the input or control matrix and $d\boldsymbol{B}(t)$ denotes a Wiener process with covariance matrix $\boldsymbol{\Sigma}$.
-
-The general solution to Equation \ref{eq:kalmn2}. an Ornstein-Uhlenbeck process
-
-By comparing Equation \ref{eq:kalman1} with Equation \ref{eq:measurement} and Equation \ref{eq:kalmn2} with Equation \ref{eq:frequency_evolution} it is immediately clear how the
-
-for the system dynamics matrix,$\bar{F}$,  control model $G$, control vector $\bar{u}$, and $w$ a stochastic zero-mean process.
-By comparison with the preceding equations, Eqs.
-	\ref{eq:state1} - 	\ref{eq:state2}, it is immediately obvious how our state space model maps onto the Kalman filter structure.
-Specifically, our states are just the $N$ intrinsic pulsar frequencies $\bar{x} = (f_1,f_2,...f_N)$ whilst our measurements are the $N$ measured pulse frequencies $\bar{z} = (f^{(M)}_1,f^{(M)}_2,...f^{(M)}_N)$.
-If we specialize to the case of constant time sampling between our observations, $\Delta t$, then for our formulation the components that make up the Kalman filter are as follows:
-\begin{equation}
-	F_i = F_{i+1} = e^{-\gamma \Delta t}
-	\label{eq:fmatrix}
-\end{equation}
-\begin{align}
-	T_i &= \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt' \\
-	& = f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)  (\Delta t + t_i) - e^{-\gamma \Delta t} (f_{\rm EM}(0) + \dot{f}_{\rm EM}(0)t_i)
-\end{align}
-\begin{equation}
-	H_i = 1 -A(\theta_{\rm GW}) \cos(-\Omega t_i (1 + n\cdot q ) + \Phi_0)
-\end{equation}
-where the $i$ subscript labels the value at the $i$-th timestep, and $A$ is a constant that is given by  Eq. \ref{eq:state2}. \newline
-
-\noindent The Kalman filter includes the effect of process noise $\bar{w}$ and the measurement noise $\bar{v}$ via the definition of a process noise matrix $Q = E[w w^T]$ and a measurement noise matrix $R = E[v v^T]$, which have the discrete form,
-\begin{equation}
-	Q_i  \delta_{ij}= \langle \eta_i \eta_j^T \rangle = \frac{- \sigma^2}{2 \gamma} \left( e^{-2 \gamma \Delta t} -1\right)
-\end{equation}
-\begin{equation}
-	R_i = R_{i+1} = \Sigma^2
-	\label{eq:rmatix}
-\end{equation}
-The above equations, Eqs.
-\ref{eq:fmatrix} - 	\ref{eq:rmatix}, apply to an operation on a single state.
-The extension to $N$ states is straightforward, since one just needs to construct a diagonal matrix for each of the Kalman components where each non-zero element corresponds to a separate pulsar frequency.
-
-In Section \ref{sec:statespace} we present a discretised version of the model of Section \ref{sec:model} which maps onto the discretely sampled observable $f(\tau)|_{\rm Earth}$.
-
-In order to make contact with discretely sampled data it is importnat to temporally discretise the model of Section \ref{sec:model}
-
-We can express the intrinsic frequency evolution, Eq. \ref{eq:frequency_evolution_expanded}, in a alternative form as,
-\textcolor{red}{TK all this needs updating}
-\begin{equation}
-	df = \gamma  f dt + N(t) dt + \sigma dB(t)
-	\label{eq:state1}
-\end{equation}
-where $\mathcal{A} = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process).
-This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by \citep{gardiner2009stochastic},
-\begin{equation}
-	f(t) = e^{\mathcal{A}t}f(0) + \int_0^t e^{\mathcal{A}(t-t')} N(t') dt' + \int_0^t e^{\mathcal{A}(t-t')} \sigma dB(t')
-\end{equation} 
-If we move from a solution in continuous time, $t$, to discrete time, \textcolor{red}{TK: AM reccomends deligint tabe sy,bol} $\bar{t} = (t_1, t_2, ...,t_K)$, then
-\begin{equation}
-	f(t_{i+1}) = F f(t_i) + T_i + \eta_i
+where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the input or control matrix.
+and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$.
+\newline
+
+The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage where the state at the current time-step is predicted based on the state at the previous time-step, and an ``update" stage where the predicted state estimate is refined based on the measurement at that time-step.
+The predict stage updates the estimate of the state $\hat{\boldsymbol{X}}_{k|k-1}$ and the associated covariance $\hat{\boldsymbol{P}}_{k|k-1}$where the subscript notation ${k|k-1}$ denotes an estimate at discrete step $k$, given measurements from step $k-1$ and earlier.
+The predict step proceeds as,
+\begin{equation}
+\hat{\boldsymbol{X}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k
+\end{equation}
+\begin{equation}
+	\hat{\boldsymbol{P}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k 
+\end{equation}
+The update step then uses the measurement $\boldsymbol{Y}_k$ to update $\hat{\boldsymbol{X}}_{k|k}$  and $\hat{\boldsymbol{P}}_{k|k}$ as,
+\begin{equation}
+	\boldsymbol{\epsilon}_{k} = \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1}
+\end{equation} 
+\begin{equation}
+	\boldsymbol{S}_k = \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k
+\end{equation}
+\begin{equation}
+	\boldsymbol{K}_k = \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \label{eq:kalman gain}
+\end{equation}
+\begin{equation}
+	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}
+\end{equation}
+\begin{equation}
+		\hat{\boldsymbol{P}}_{k|k} = \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1}
+\end{equation}
+Equation \ref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which yields the minimal mean square error in the measurement post fit residual, $\boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k}$.
+\newline
+
+We can map our state-space model described in Section \ref{sec:model} onto the Kalman filter structure of Equations  
+\ref{eq:kalman1}, \ref{eq:kalman2} as follows.
+We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e $\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right)$.
+Similarly  $\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)$.
+The states evolve according the continuous dynamical equation
+\begin{equation}
+	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
+\end{equation}
+where $\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right)$ is the state matrix, $\boldsymbol{C} = \left(C^{(1)}, C^{(2)}, ..., C^{(N)}\right)$ the input or control matrix where $C^{(i)} =\gamma^{(i)} \left(f_{\rm em}^{(i)} (0) + \dot{f}_{\rm em}(0)^{(i)} t \right) +  \dot{f}_{\rm em}(0)^{(i)}$ and $d\boldsymbol{B}(t)$ denotes a Wiener process with covariance matrix $\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)$.
+Equation \ref{eq:kalmn2} is a Langevin equation (equivalently an Ornstein-Uhlenbeck process) which has a general solution given by \citep{gardiner2009stochastic},
+\begin{equation}
+	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \label{eq:gardenier}
+\end{equation} 
+From Equation \ref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ as Equation \ref{eq:kalman2} where
+\begin{equation}
+\boldsymbol{F}_k = e^{\boldsymbol{A}\left( t_{k+1} - t_k \right)} 
+\end{equation}
+\begin{equation}
+	\boldsymbol{G}_k = \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' 
+\end{equation}
+\begin{equation}
+	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t')
+	\end{equation}
+For our system, the state, control and process noise covariance matrices then take the form:
+\begin{equation}
+	\boldsymbol{F}_k = 	\text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right)
+\end{equation}
+\begin{equation}
+\boldsymbol{G}_k	= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right)
 \end{equation}
@@
 \begin{equation}
-	F_i = e^{\mathcal{A} (t_{i+1} - t_i)}
-\end{equation}
-\begin{equation}
-	T_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} N(t') \, dt'
-\end{equation}
-\begin{equation}
-	\eta_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} \sigma \, dt'
-\end{equation}
-
-if we specialise to the case of constatn time sampling
-
-The discrete solution $f(\bar{t})$ to the intrinsic frequency can be related to the discrete measured frequency via Eq. \ref{eq:main_eq} as,
-\begin{equation}
-	f_M (\bar{t})= f(\bar{t}) g(\bar{\theta},\bar{t}) + N_M
-\end{equation}
-where $g(\theta,t)$ can be expressed in a trigonometric form as 
-\begin{equation}
-	X = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
-\end{equation} 
-whilst $N_M$ is a Gaussian measurement noise that satisfies 
-\begin{equation}
-	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
-\end{equation}
-for variance $\Sigma^2$.
+	G_k^{(n)} =    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] - e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
+\end{equation}
+and
+\begin{equation}
+	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1), Q^{(2)}},...,Q^{(N)}\right) 
+\end{equation}
+for 
+\begin{equation}
+	Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right)
+\end{equation}
+where $\Delta t$ is the constant time sampling interval $=t_{k+1} -t_{k}$. \newline
+
+The remaining component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
+These are defined straightforwardly from Equation \ref{eq:measurement}.
+Specifically,
+$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \ref{eq:g_func}.
+$\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
 
 \section{Appendix on identifiability}\label{appendix_identifiability}
+
+\textcolor{red}{TK TBD}
 

### 2023-08-30 docs/manuscript/paper.tex
1000 noise realisations for h=8e-15

@@
 This enables self-consistent inference of model parameters and calculation of the Bayesian evidence.
-It is shown via an astrophysically representative software injection campaign that the procedure can distinguish a gravitational wave from pure noise down to a strain of $h_0 \sim 4 \times 10^{15}$, given an inclination angle $\iota = 1.0$ rad.
+It is shown via an astrophysically representative software injection campaign that the procedure can distinguish a gravitational wave from pure noise down to a strain of $h_0 \sim 4 \times 10^{-15}$, given an inclination angle $\iota = 1.0$ rad.
 Full posterior distributions of model parameters are recovered.
@@
 \noindent This paper is organised as follows.
-In Section \ref{sec:model} we present the state-space model for the pulsar pulse frequency which is subject the influence of a GW.
+In Section \ref{sec:model} we present the state-space model for the pulsar frequency which is subject to the influence of a GW.
 In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence.
@@
 The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$.
-Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution can be approximated accurately by
+Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution $f_{\rm em}^{(n)}(t)$ can be approximated accurately by
 \begin{equation}
@@
 \end{equation}
-In Eq \ref{eq:frequency_evolution}, $f_{\rm em}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
-\begin{equation}
-	\langle \xi^{(n)}(t) \rangle = 0 \ ,
-\end{equation}
-\begin{equation}
-	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle = [\sigma^{(n)}]^2 \delta(t - t') \ .
-	\label{eq:xieqn}
-\end{equation}
+In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm em}$ is the spin derivative, $\gamma^{(n)}$ a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
+\begin{align}
+	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
+	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle &= [\sigma^{(n)}]^2 \delta(t - t') \ .	\label{eq:xieqn}
+\end{align}
 $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
-It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Eq.
-\ref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
+It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
 \newline
@@
 In the presence of a GW, the pulsar frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
-Indeed, the pulse TOAs are modulated harmonically at the GW frequency.
+Specifically, the pulsar frequency is modulated harmonically at the GW frequency.
 We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section.
@@
 \end{equation}
+where $\delta$ is the declination of the GW source and $\alpha$ the right ascension.
 In Equation \eqref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
@@
 \subsection{Kalman filter}\label{sec:kalman_filter}
+
+%
+%\begin{figure*}
+%	\centering
+%	
+%	\subfloat[a]{\includegraphics[width=0.5\textwidth]{images/Kalman_example_both}} 
+%	\subfloat[b]{\includegraphics[width=0.3\textwidth]{images/Kalman_example_both}}
+%	\caption{\textcolor{red}{TK: For consideration. Do we prefer this figure to Figure \ref{fig:four figures}}?}
+%	\label{fig:kalmanexampleboth}
+%\end{figure*}
+%
+
 \begin{figure*}
 	%\centering % Not needed
-	\begin{subfigure}[b]{1\textwidth}
-		\includegraphics[width=\textwidth]{images/Kalman_example_true_params}
+	\begin{subfigure}[b]{0.49\textwidth}
+		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
 		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
 		\label{fig:6MB_BFS}
-	\end{subfigure} \newline
-
-	\begin{subfigure}[b]{1\textwidth}
+	\end{subfigure}  
+    \hfill
+	\begin{subfigure}[b]{0.49\textwidth}
 		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
@@
 In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$ and the state is not estimated accurately.
-The top left panels show the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line).
-The middle left panels show the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta).
-The bottom left panels show the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$.
-The right hand panels plot the distribution of the innovation over the total observation period.
-In (a) the greed/blue and red/magenta lines are completely overlaid.
-\textcolor{red}{TK: See Fig \ref{fig:kalmanexampleboth} in Appendix for an alternative version of this figure.
-Which do we prefer?}}
+The top panels show the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line).
+The middle panels show the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta).
+The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$.
+In (a) the green/blue and red/magenta lines are completely overlaid.}
 	\label{fig:four figures}
@@
 Throughout this work we take $\mathcal{M}_1$ to be the state space model that includes the presence of a GW $\mathcal{M}_0$ is our null model that assumes there is no GW in the data.
-This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func}, \eqref{eq:g_func_trig}.
+This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
 The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
@@
 \subsection{Summary of workflow}\label{sec:methodsummary}
-To summarise the above, the order of operations in our method for estimating the posteriors on the static parameters and calculating the marginal likelihood for the inference model is as follows:
-
+To summarise the above, the order of operations in our method for a single iteration (i.e. for a single live point) is as follows:
 \begin{enumerate}
 	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\boldsymbol{\theta}_{\rm sample}$
-	\item Pass $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
+    \item Pass $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
 	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
@@
     \item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$
-	\item Repeat
 \end{enumerate}
@@
 In Section \ref{sec:rep_example} we generate synthetic data for a single representative example SMBHB system and apply our method to estimate $\boldsymbol{\theta}$ and calculate the statistical evidence for a GW in the data via the Bayes factor.
-In this section we also explore how the detectability of the system changes for different strain magnitudes, $h_0$ and the performance of the method for different realisations of the system noise.
+In this section we also explore how the detectability of the system changes for different strain magnitudes, $h_0$ and the performance of the method for different realisations of the pulsar process noise and the measurement noise.
 In Section \ref{sec:parameter_space} we move beyond a single example system and test our method across a broader parameter space.
@@
 \end{figure}
-
 For this work we generate synthetic data that is astrophysically representative as follows.
@@
 We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}.
-The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA and the method naturally extends to other PTAs of the IPTA.
-For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$.
+The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA and the method naturally extends to any other PTA.
+For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{(n)}$), the distance, the barycentric rotation frequency (which we identify with $f_{\rm em}^{(n)} (t_1)$), and the time derivative of the the barycentric rotation frequency (which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$).
 The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Figure \ref{fig:pulsar_distrib}.
-The data is acquired via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+The data is acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
 The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$.
 We discussed in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$.
-Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ \citep{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars.
+Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars.
 For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
@@
 where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$.
-Throughout this work we assume that all pulsars are observed with a weekly cadence i.e.
-$T = 1$ week.
+Throughout this work we assume that all pulsars are observed with a weekly cadence.
+In order to relate the empirical TOA noise model, Equation \ref{eq:sigmap}, to our Ornstein-Uhlenbeck frequency model, Equation \ref{eq:spinevol}, we set first $T = 1$ week in Equation \ref{eq:sigmap} to obtain an estimate of the TOA noise over a weekly time interval, i.e.
+the size of our numerical integration timestep.
 The frequency process noise parameter $\sigma^{(n)}$ can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as
@@
 \end{eqnarray}
-For our synthetic NANOGrav PTA, the median $\sigma^{(n)}$ calculated in this way is $ = 4.28 \times 10^{-21} $ s. \newline
+For our synthetic NANOGrav PTA, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s. \newline
 
@@
 The actual residuals are obtained via UTMOST pulsar timing program \citep{UTMOST}.
-The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f}.
+The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
 For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
@@
 \subsubsection{Synthetic data generation}
-We generate $N$ synthetic noisy timeseries - one for each pulsar - as follows:
+We generate $N$ synthetic noisy timeseries, one for each pulsar, as follows:
 
 \begin{enumerate}
-	\item Integrate the state equations, Equations \eqref{eq:spinevol}- \eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
+	\item Integrate the state equations, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
 	\item Map from state space to measurement space via the measurement equation, Equation \eqref{eq:measurement}
 	\item Add zero-mean Gaussian measurement noise to each timeseries, i.e.
-$f_{\rm m} ^{(n)} + N_{\rm f}^{(n)}$ where 	$\langle N_{\rm f}(t) N_{\rm f}(t') \rangle = \sigma_{\rm m}^2 \delta(t - t')$ for measurement noise covariance $\sigma_{\rm m}^2$.
+$f_{\rm m} ^{(n)} + \varepsilon^{(n)}$ where 	$\langle \varepsilon(t) \varepsilon(t') \rangle = \sigma_{\rm m}^2 \delta(t - t')$ for measurement noise covariance $\sigma_{\rm m}^2$.
 \end{enumerate}
-Equations \eqref{eq:spinevol}- \eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
 The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified by our construction of the synthetic PTA outlined in Section \eqref{sec:synt_pta}.
@@
 \begin{equation}
-	\sigma_{\rm m} \sim f_{\rm p}^{(n)} \frac{\sigma_{\rm TOA}}{\text{cadence}} \ . 
+	\sigma_{\rm m} \sim f_{\rm p}^{(n)} \frac{\sigma_{\rm TOA}}{\text{T}} \ . 
 \end{equation}
 For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
-Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$ and take it to be known \textit{a priori} rather than a parameter to be inferred.
+Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it to be known \textit{a priori} rather than a parameter to be inferred.
 Note that $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar; whilst for our construction every pulsar is sampled at the same time and so would experience the same realisation of the noise, in actual pulsar astronomy different pulsars will be sampled at different times by different radio interferometers and so subject to different noise realisations.
@@
 
-In practice, in order to avoid numerical precision issues that arise when solving Equations \eqref{eq:spinevol}- \eqref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
+In practice, in order to avoid numerical precision issues that arise when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
 \begin{equation}
@@
 
-The parameters which govern the deterministic evolution of the pulsar spin frequency, $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ are well-determined by existing radio timing observations.
-We can identify $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar  barycentric rotation frequency and the time derivative of barycentric rotation frequency quoted by pulsar observation catalogues.
+The parameters which govern the deterministic evolution of the pulsar spin frequency, $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ are well-determined by existing radio timing observations.
+We can identify $f_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the time derivative of barycentric rotation frequency quoted by pulsar observation catalogues.
 For the 12.5 year NANOGrav pulsars the median error in  the barycentric rotation frequency is $\sim 7 \times 10^{-13}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.8 \times 10^{-20}$ s $^{-2}$.
-Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$, $\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 10^3 \epsilon^{(n)}_{a}$ either side of the true value, where $\epsilon^{(n)}_{a}$ is the error for that pulsar quoted in the Australia Telescope National Facility (ATNF) catalogue and the subscript $a = \{ f, \dot{f} \}$ identifies the error as being in $f_{\rm em}^{(n)} (t_1)$, or $\dot{f}_{\rm em}^{(n)} (t_1)$ respectively.
+Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$, $\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 10^3 \epsilon^{(n)}_{a}$ either side of the true value, where $\epsilon^{(n)}_{a}$ is the error for that pulsar quoted in the ATNF catalogue and the subscript $a = \{ f, \dot{f} \}$ identifies the error as being in $f_{\rm em}^{(n)} (t_1)$, or $\dot{f}_{\rm em}^{(n)} (t_1)$ respectively.
 In this way we can test how well our method performs without requiring exceptionally precise and accurate measurements of the pulsar parameters to be made \textit{a priori}.
 Instead we consider the pulsar parameters similarly to the GW source parameters and estimate them optimally within a consistent framework.
-Adopting looser priors in this way allows us to subject our method to a more stringent test and confirm that we still estimate GW parameters without highly accurate initial estimates of $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$.
-\newline
-
-The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$,$\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}.
-However for this work, whilst the pulsar distance is used when generating synthetic data, as discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d$.
-Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
+Adopting looser priors in this way allows us to subject our method to a more stringent test and confirm that we still accurately estimate GW parameters without highly accurate initial estimates of $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$.
+\newline
+
+The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}.
+However for this work, whilst the pulsar distance is used when generating synthetic data, as discussed in Section \ref{sec:parameter_estim}, the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$.
+Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
 That is, for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
 It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value.
-We briefly explored setting an uninformative prior on $\gamma$ over e.g.
-LogUniform($10^{-15}, 10^{-10}$) as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g.
+We briefly explored setting an uninformative prior on $\gamma^{(n)}$ over e.g.
+LogUniform($10^{-15}, 10^{-10}$) s$^{-1}$ as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g.
 set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
-The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
-For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$).
-The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm p}^{(n)}$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16}$.
-For our artificial dataset we set $\sigma = 10^{-21}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
-\newline
-
-Since we are not setting priors on $\gamma$ or $d$ this reduces the dimensionality of the parameter space to $7 + 3N$.
+The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{s}^{-5/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
+For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-5/2}$).
+The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-5/2}$.
+For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-5/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
+\newline
+
+Since we are not setting priors on $\gamma^{(n)}$ or $d^{(n)}$ this reduces the dimensionality of the parameter space to $7 + 3N$.
 We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space, c.f.
@@
 All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
-\begin{table}
+\begin{table*}
 	\centering
-	\resizebox{\columnwidth}{!}{%
-	\renewcommand{\arraystretch}{1.5} % Default value: 1
+%	\resizebox{\textwidth}{!}{%
+%	\renewcommand{\arraystretch}{1.0} % Default value: 1
 	\begin{tabular}{lccll}
@@
 		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
-		& $\gamma$              & $10^{-13}$ & s$^{-1}$ & - \\
+		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
 		\bottomrule
 	\end{tabular}
-}
 	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter.
 The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.
+The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.
 The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
 	\label{tab:parameters_and_priors}
-\end{table}
+\end{table*}
 
@@
 
- The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$.
+ The results are shown in Figure \ref{fig:corner_plot_1} for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$.
 The histograms shown in the Figure are the 1D-posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
@@
 	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
-The horizontal grey dashed line labels the detection tolerance cut-off of $\beta = 10$.
+The horizontal grey dashed line labels the detection threshold of $\beta = 10$.
 The minimal detectable strain, below which $\beta < 10$, is $\sim 3.9 \times 10^{-15}$.
-Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$}
+Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$.}
 	\label{fig:bayes}
@@
 It is also important to test the method in different regions of the parameter space.
-To this end we consider 200 simulated injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining 5 parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
+To this end we consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
 We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space.
@@
 
-\subsection{References}
+\section{Wasserstein distance}
+In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD).
+The WD intuitively defines a distance between two distributions as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$.
+The $p$-Wasserstein distance between two distributions $\mu, \nu$ is
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p}
+\end{eqnarray}
+where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function.
+It is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
+\begin{eqnarray}
+	\boldsymbol{E} [W_1(\mu_n,\mu)] \asymp n^{-\frac{1}{d}}
+\end{eqnarray}
+To sidestep this issue, we marginalize onto the distributions spanned by the individual gridpoints and take the WD between these 1-dimensional distributions.
+Such an approach is also adopted in \cite{Paxton2021,Vissio2020}.
+We also take $p=1$ and $c(x,y) = |x-y|$ such that the WD values quoted in this work are given by
+\begin{eqnarray}
+	W_1(\mu,\nu)=  \inf_{\gamma \in \Gamma(\mu, \nu)} \int |x-y| d \gamma (x,y)
+\end{eqnarray}
+From the perspective of climate modelling, the WD has a nice physical interpretation given by the Monge-Kantorovich duality:
+\begin{eqnarray}
+	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
+\end{eqnarray}
+i.e.
+the difference in the expected value of some random variable between two distributions is no bigger than the WD value itself.
+More explicitly, if the WD for surface temperature between two distributions is 1K, then this tells us that the difference in the expected temperature between the distributions is no greater than 1K.
+For a full review of the Wasserstein distance, including favourable properties and comparison with other metrics we refer the reader to the Appendix of Pax21.
+
+\section{References}
 \label{sec:ref_list}

### 2023-08-30 py_src/plotting.py

@@ -170,7 +170,7 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
     fig = corner.corner(y_post, 
                         color='C0',
                         show_titles=True,
-                        smooth=True,smooth1d=False,
+                        smooth=True,smooth1d=True,
                         truth_color='C2',
                         quantiles=[0.16, 0.84],
                         truths = injection_parameters,

### 2023-08-30 docs/manuscript/paper.tex
correct priors

@@
  The results are shown in Figure \ref{fig:corner_plot_1} for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$.
-The histograms shown in the Figure are the 1D-posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
-The 2-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
+The histograms shown in the Figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
+The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
 It is evident that for this representative example we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for the majority of the parameters.
@@
 We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.
-If we consider the median values of the 1D-posteriors we can see that for some parameters the median is almost identically equal to the injected value  (e.g.
+If we consider the median values of the one-dimensional posteriors we can see that for some parameters the median is almost identically equal to the injected value  (e.g.
 $\Omega, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the injected value (e.g.
@@
 This potential bias will also be investigated alongside the bias in $\iota$ in Section \ref{sec:multiple_noise} and Section \ref{sec:bias}.
-Similar results can be derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
+Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
 We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.
@@
 $T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
-\ref{sec:parameter_space}.
-\newline
-
-The data gaps that are evident in the curve for some values of $h_0$ below $\beta =10$ are due to the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
+\newline
+
+There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10{-15}$.
+The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
+The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
 For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
 That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these gaps appear at different values of $h_0$, for $\beta < 10$.
+There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
 For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
@@
 	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
-	\caption{1D marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel).
+	\caption{One-dimensional marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel).
 The thick vertical orange line labels the true injected value.
@@
 The above discussion was for a single realisation of the noise processes.
-It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method.
+It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} and \ref{sec:detection} is not particularly advantageous or convenient for our method.
 In this Section we confirm that we are able to infer comparable results for different noise realisations.
@@
 For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
-In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for 9 realisations of the noise.
+In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise.
 This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}.
@@
 We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1}.
-Considering the 1D marginalized posteriors, the inferred probability distributions of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations.
-Moreover, the variance of each of the individual 1D posteriors is generally low i.e.
-each of individual the distributions are narrow compared to the prior space.
-That is to say, for each different noise realisation the method infers a consistent maximum \textit{a posteriori} probability (MAP) estimate.
-Conversely for the parameters $\iota$ and $h_0$ the inferred 1D posteriors are not consistent across different noise realisations and the variance of an individual posterior is generally larger than for the other parameters.
+Considering the one-dimensional marginalized posteriors, the inferred probability distributions of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ broadly overlap across the different noise realisations.
+Moreover, the variance of each of the individual one-dimensional posteriors is generally low i.e.
+each of the individual distributions are narrow compared to the prior space.
+For each different noise realisation the method infers a consistent maximum \textit{a posteriori} probability (MAP) estimate.
+Conversely for the parameters $\iota$ and $h_0$ the inferred one-dimensional posteriors do not broadly overlap are not consistent across different noise realisations and the variance of an individual posterior is generally larger than for the other parameters.
 That is to say, we infer very different MAP estimates of these parameters depending on the particular realisation of the noise.
@@
 
-The large degree of variance in the 1D posteriors of $\iota$ and $h_0$ is due to these parameters being only weakly-identifiable.
+The posteriors for $\iota$ and $h_0$ do not overlap for the different noise realisations due to these parameters being only weakly-identifiable.
 Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
-Sources of non-identifiability are generally categorised as either ``structural" - arising from the structure of the model - or ``practical" - arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
-Regarding $\iota$ and $h_0$ we are concerned with structural identifiability.
-Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \ref{eq:hphx}, \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+Sources of non-identifiability are generally categorised as either ``structural" (arising from the structure of the model) or ``practical" ( arising from insufficient data or measurement errors) \citep{GUILLAUME2019418}.
+Regarding $\iota$ and $h_0$, we are concerned with structural identifiability.
+Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx}, \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
 For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
@@
 Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
-If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the variance in the estimates between different noise realisations decreases.
-This is demonstrated in Figure \ref{fig:iota} where we plot the marginalized 1D posterior of $\iota$ for 9 realisations of the noise in both the original case where $h_0$ is a free parameter (equivalent to the 1D posterior for $\iota$ in Figure \ref{fig:corner_plot_2}) and in the case where we hold $h_0$ fixed at its injection value, $h_0 = 10^{-12}$.
+If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the the posteriors for the different noise realisations overlap.
+This is demonstrated in Figure \ref{fig:iota} where we plot the marginalized one-dimensional posterior of $\iota$ for 9 realisations of the noise in both the original case where $h_0$ is a free parameter (equivalent to the one-dimensional posterior for $\iota$ in Figure \ref{fig:corner_plot_2}) and in the case where we hold $h_0$ fixed at its injection value, $h_0 = 10^{-12}$.
 We can see that whilst the variance between the individual posteriors is large in the original case, once we fix $h_0$ the variance decreases dramatically and the sampling algorithm repeatedly converges consistent estimates of $\iota$.
@@
 Figure \ref{fig:bayes}) and consequently the posteriors are typically narrow compared to the prior space and well described by their median value.
-To compare each of the posteriors across different noise realisations, we consider the normalised standard deviation in the set of $10^3$ median values i.e.
+To compare each of the posteriors across different noise realisations,
+
+\textcolor{red}{Wasserstein intro and discussion here}
+
+we consider the normalised standard deviation in the set of $10^3$ median values i.e. 
 \begin{equation}
@@
 We can therefore see that the method is able to reliably converge to a consistent MAP estimate.
-Conversely, as we saw before, the MAP estimates for $\iota$ and $h$ are not consistent across different noise realisations, with $\rho \sim 0.3$ and $\rho \sim 0.1$ respectively.
+Conversely, as we saw before, the MAP estimates for $\iota$ and $h_0$ are not consistent across different noise realisations, with $\rho \sim 0.3$ and $\rho \sim 0.1$ respectively.
 This is due to the weak-identifiability issue that we have discussed.
@@
 To summarise the results across the parameter space we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
-A PP plot describes the fraction of the total number of injected parameters which are included within a given confidence interval of the estimated posterior with respect to the confidence interval itself.
-In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage confidence interval.
+A PP plot describes the fraction of the total number of injected parameters which are included within a given credible interval of the estimated posterior with respect to the credible interval itself.
+In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage credible interval.
 The results are shown in Figure \ref{fig:parameter_space}.
@@
 The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$.
-The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value confidence intervals than would be expected statistically, and vice versa.
+The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and vice versa.
 This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally highly confident i.e.
 narrow, but are systematically biased away from the true injection values.
-This manifests in the PP plot as posteriors which are insufficiently narrow to capture the injected value within the appropriate confidence interval.
+This manifests in the PP plot as posteriors which are insufficiently narrow to capture the injected value within the appropriate credible interval.
 An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.
@@
 Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (i.e.
-the median of the 1D marginalised posterior) is biased away from the true injection value.
+the median of the one-dimensional marginalised posterior) is biased away from the true injection value.
 This effect is most severe for $\iota$ and $h_0$ but also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
@@
 Consequently Bayesian likelihood inference techniques settle in optima that do not necessarily align with the true injection value.
-This is the origin of the bias in the 1D marginalised posteriors that we observe.
+This is the origin of the bias in the one-dimensional marginalised posteriors that we observe.
 For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ radians, whereas the injected value was $\iota = 1.0$ radians, This difference is comparable to the bias observed in the right-hand panel of Figure \ref{fig:iota} where the median of the marginalised posterior for $\iota$ was $\sim 0.65$ radians.
-The discrepancy between $\iota \sim 0.7$ in the log-likelihood curve and $\iota \sim 0.65$ in the nested sampling inference is a result of the likelihood curves being a 1D representation of a multidimensional problem.
+The discrepancy between $\iota \sim 0.7$ in the log-likelihood curve and $\iota \sim 0.65$ in the nested sampling inference is a result of the likelihood curves being a one-dimensional representation of a multidimensional problem.
 The exact shape of the likelihood curve naturally depends on the values of all of the other parameters that we hold constant.
@@
 Equation \ref{eq:measurement}).
-Explicitly, we have the ``Earth-terms wave" with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
+Explicitly, we have the ``Earth-terms wave with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
 The net interference wave that modulates $f_{\rm p}$ is
@@
 The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
-If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the 1D marginalised posterior may dominate and the bias may not be important for searches.
+If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate and the bias may not be important for searches.
 Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties then the bias may come to systematically affect the inferred parameters.
@@
 The natural first extension is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
-Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred 1D marginalized posteriors for the model parameters.
+Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred one-dimensional marginalized posteriors for the model parameters.
 Biases in sky localisation for synthetic datasets as a result of dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.

### 2023-08-30 py_src/priors.py

@@ -174,25 +174,25 @@ def set_prior_on_measurement_parameters(init_parameters,priors,measurement_model
         
         #Add all the GW quantities
         init_parameters["omega_gw"] = None
-        #priors["omega_gw"] = bilby.core.prior.LogUniform(1e-8, 1e-5, 'omega_gw')
-        priors["omega_gw"] = P.Ω
+        priors["omega_gw"] = bilby.core.prior.LogUniform(1e-8, 1e-5, 'omega_gw')
+        #priors["omega_gw"] = P.Ω
 
         init_parameters["phi0_gw"] = None
-        #priors["phi0_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'phi0_gw')
-        priors["phi0_gw"] = P.Φ0
+        priors["phi0_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'phi0_gw')
+        #priors["phi0_gw"] = P.Φ0
 
         init_parameters["psi_gw"] = None
-        #priors["psi_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'psi_gw')
-        priors["psi_gw"] = P.ψ
+        priors["psi_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'psi_gw')
+        #priors["psi_gw"] = P.ψ
 
         init_parameters["iota_gw"] = None
-        #priors["iota_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'iota_gw')
-        priors["iota_gw"] = P.ι
+        priors["iota_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'iota_gw')
+        #priors["iota_gw"] = P.ι
 
         init_parameters["delta_gw"] = None
         #priors["delta_gw"] = bilby.core.prior.Uniform(-np.pi/2.0, np.pi/2, 'delta_gw')
-        #priors["delta_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2, 'delta_gw')
-        priors["delta_gw"] = P.δ
+        priors["delta_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2, 'delta_gw')
+        #priors["delta_gw"] = P.δ
 
         init_parameters["alpha_gw"] = None
         #priors["alpha_gw"] = bilby.core.prior.Uniform(0.0, 2*np.pi, 'alpha_gw')
[…]

### 2023-08-30 docs/manuscript/paper.tex
commit before updating posteriori plots

@@
 \subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
+%\begin{figure*}
+%	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
+%	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
+The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
+The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f.
+Figure \ref{fig:likelihood_surface}.}
+%	\label{fig:corner_plot_2}
+%\end{figure*}
+
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
+	%\centering % Not needed
+	\begin{subfigure}[b]{0.49\textwidth}
+		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
+		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
+		\label{fig:6MB_BFS}
+	\end{subfigure}  
+	\hfill
+	\begin{subfigure}[b]{0.49\textwidth}
+		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
+		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
+		\label{fig:25MB_bfs}
+	\end{subfigure}
 	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
@@
 Figure \ref{fig:likelihood_surface}.}
-	\label{fig:corner_plot_2}
+	\label{fig:four figures2}
 \end{figure*}
+
 \begin{figure}
@@
 The results are shown in Figure \ref{fig:parameter_space}.
-The shaded grey contoured regions label the first 3 $\sigma$ significance levels i.e.
-how much deviation from the $y=x$ diagonal one might expect statistically given 200 simulations.
-We can see that it only $\Omega$ is well-behaved and falls within the shaded region.
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections."
+We can see that only $\Omega$ falls within the $3\sigma$ shaded region.
 The other parameters deviate from the $y=x$ diagonal.
 The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$.
-The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and vice versa.
-This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally highly confident i.e.
+The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and there are more injections contained within the low value credible intervals.
+This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally highly confident, i.e.
 narrow, but are systematically biased away from the true injection values.
-This manifests in the PP plot as posteriors which are insufficiently narrow to capture the injected value within the appropriate credible interval.
+This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval.
 An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.
@@
 \end{figure}
-Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (i.e.
+Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g.
 the median of the one-dimensional marginalised posterior) is biased away from the true injection value.
@@
 We do this for 10 separate realisations of the system noise.
-We consider 3 separate solutions: \textit{(i)} The solid blue line is the likelihood curve with just the Earth-terms included (i.e.
+We consider three separate situations: \textit{(i)} The solid blue line is the likelihood curve with just the Earth-terms included (i.e.
 using Equation \ref{eq:measuremen_earth} in the Kalman filter), \textit{(ii)} the solid orange line is the likelihood curve with the additional pulsar terms included (i.e.
@@
 The GW source is monochromatic over the decadal PTA observation period, but the timescale set by the Earth-pulsar distance is comparable to the SMBHB evolution timescale \citep{Sesana2010}.
-For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \ref{eq:f_evolution}) the GW source frequency evolution manifests as an incoherent source of noise in the pulsar terms, whilst for  $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms can affect the phase coherency of the Earth terms \citep{Perrodin2018}.
+For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) the GW source frequency evolution manifests as an incoherent source of noise in the pulsar terms, whilst for  $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms can affect the phase coherency of the Earth terms \citep{Perrodin2018}.
 Careful consideration of the evolution of the GW frequency will be important as we look to extend our method to include the pulsar terms in the inference model.
@@
 However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
-Our method naturally extends to detecting multiple GW sources simultaneously; Equation \ref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
-Evaluating the performance of the method under these conditions would be a further interesting pursuit.
+Our method naturally extends to detecting multiple GW sources simultaneously; Equation \eqref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
 The stochastic background itself is the incoherent summation of many individual GW sources.
@@
 We initially focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
-The parameters of the model are accurately recovered and the minimal detectable strain is estimated to be $h_0 \sim 4 \times 10^{-15}$.
+The parameters of the model are accurately recovered and the minimal detectable strain is estimated to be $h_0 \sim 4 \times 10^{-15}$ for $\iota=1.0$ rad.
 The method is then further tested over 1000 noise realisations.

### 2023-08-30 notebooks/plot_likelihood_surfaces_2D.py

@@ -25,8 +25,9 @@ ax = fig.add_subplot(111)
 
 path_to_earth_model = 'likelihood_surface_h_iota_mm_earth.npz'
 path_to_earth_model = 'likelihood_surface_h_iota_mm_earth_broader.npz'
-path_to_earth_model = 'likelihood_surface_h_iota_mm_earth_coarse.npz'
-
+path_to_earth_model = 'likelihood_surface_h_iota_mm_earth_coarse.npz' #this is the one currently in publication
+path_to_earth_model = 'likelihood_surface_h_iota_mm_earth_coarse_small_h.npz'
+#path_to_earth_model = 'likelihood_surface_h_iota_mm_earth_coarse_new.npz' 
 
 def load_and_plot(path,ax):
 
@@ -54,8 +55,11 @@ def load_and_plot(path,ax):
 
 
     #Plot colormap
-    CS = ax.pcolormesh(X, Y, z,clim=(2.0*np.max(surface_pulsar), np.max(surface_pulsar)),shading='gouraud',cmap='viridis')
+    #CS = ax.pcolormesh(X, Y, z,clim=(2.0*np.max(surface_pulsar), np.max(surface_pulsar)),shading='gouraud',cmap='viridis')
     #CS = ax.pcolormesh(X, Y, z,shading='gouraud',cmap='viridis')
+    eps=1e-5
+    CS = ax.pcolormesh(X, Y, z,clim=(1.0-eps, 1.0),shading='gouraud',cmap='viridis')
+
 
     clb = plt.colorbar(CS)
 
@@ -74,7 +78,7 @@ def load_and_plot(path,ax):
     savefig = 'likelihood_surface'
     ax.yaxis.set_major_locator(plt.MaxNLocator(5))
     plt.setp(ax.get_yticklabels()[0], visible=False)   #no 0th label to prevent overlap  
-    plt.savefig(f"../data/images/{savefig}.png", bbox_inches="tight",dpi=300)
+    #plt.savefig(f"../data/images/{savefig}.png", bbox_inches="tight",dpi=300)
 
 
 

### 2023-08-30 py_src/plotting.py

@@ -134,7 +134,8 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
 
     #Make omega into nHz
     df_posterior["omega_gw"] = df_posterior["omega_gw"]*1e9
-    df_posterior["h"] = df_posterior["h"]*1e15
+    #df_posterior["h"] = df_posterior["h"]*1e15
+    df_posterior["h"] = df_posterior["h"]*1e12
 
     display(df_posterior)
 
@@ -174,7 +175,7 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
                         show_titles=True,
                         smooth=True,smooth1d=True,
                         truth_color='C2',
-                        quantiles=[0.16, 0.84],
+                        quantiles=[0.16, 0.84], #[0.16, 0.84]
                         truths = injection_parameters,
                         range=ranges,
                         labels = labels,

### 2023-08-30 docs/manuscript/paper.tex
bayes for multiple noise realisations

@@
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
 	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
@@
 
-\subsubsection{Detection} \label{sec:detection}
-We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-$\mathcal{M}_1$ is the Earth-terms only model, i.e.
-the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
-The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
-The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
-We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
-We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
-For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
-This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
-$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
-\newline
-
-There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10{-15}$.
-The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
-The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
-For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
-That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
-For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
-For the results presented here we set $n_{\rm live} =2000$.
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
-	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
-The horizontal grey dashed line labels the detection threshold of $\beta = 10$.
-The minimal detectable strain, below which $\beta < 10$, is $\sim 3.9 \times 10^{-15}$.
-Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$.}
-	\label{fig:bayes}
-\end{figure}
+%\begin{figure*}
+%	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
+%	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The vertical orange lines indicate the true injected values.
+The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+We are able to accurately estimate each parameter of interest.
+We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
+%	\label{fig:corner_plot_1}
+%\end{figure*}
 
 \subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
+	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
+The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
+The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f.
+Figure \ref{fig:likelihood_surface}.}
+	\label{fig:corner_plot_2}
+\end{figure*}
+
 %\begin{figure*}
-%	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
+%	%\centering % Not needed
+%	\begin{subfigure}[b]{0.49\textwidth}
+%		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
+%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
+%		\label{fig:6MB_BFS}
+%	\end{subfigure}  
+%	\hfill
+%	\begin{subfigure}[b]{0.49\textwidth}
+%		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
+%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
+%		\label{fig:25MB_bfs}
+%	\end{subfigure}
 %	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
@@
 Figure \ref{fig:likelihood_surface}.}
-%	\label{fig:corner_plot_2}
+%	\label{fig:four figures2}
 %\end{figure*}
-
-\begin{figure*}
-	%\centering % Not needed
-	\begin{subfigure}[b]{0.49\textwidth}
-		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
-		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
-		\label{fig:6MB_BFS}
-	\end{subfigure}  
-	\hfill
-	\begin{subfigure}[b]{0.49\textwidth}
-		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
-		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
-		\label{fig:25MB_bfs}
-	\end{subfigure}
-	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
-The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
-The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f.
-Figure \ref{fig:likelihood_surface}.}
-	\label{fig:four figures2}
-\end{figure*}
 
@@
 \end{figure}
-\begin{figure*}
-	\centering
-	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
-	\caption{One-dimensional marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel).
+%\begin{figure*}
+%	\centering
+%	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
+%	\caption{One-dimensional marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel).
 The thick vertical orange line labels the true injected value.
@@
 The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column), but with the $(0.16,0.84)$ quantiles replaced with the $0.5$ quantile.}
-	\label{fig:iota}
-\end{figure*}
+%	\label{fig:iota}
+%\end{figure*}
 The above discussion was for a single realisation of the noise processes.
@@
 
-We now consider the complete set of $10^3$ noise realisations, rather than the subset of 9 realisations in the preceding discussion.
-For each realisation of the noise we infer a marginalised 1-D posterior for each parameter.
+We now consider the complete set of $10^3$ noise realisations, rather than the subset of ten realisations in the preceding discussion.
+For each realisation of the noise we infer a marginalised one-dimensional posterior for each parameter.
+We want to compare these posteriors for different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
+
+To this end we use the the Wasserstein distance \citep[WD,][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive of similarity between probability distributions.
+
+A full review of the Wasserstein distance which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}
+
+In order to quantitatively define `similar'  we use . The WD is a metric which quantifies the distance between two probability distributions 
+as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$.
+The $p$-Wasserstein distance between two distributions $\mu, \nu$ is
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p}
+\end{eqnarray}
+where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function.
+
+In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD). The WD intuitively defines a distance between s
+
+We also take $p=1$ and $c(x,y) = |x-y|$ such that the WD values quoted in this work are given by
+\begin{eqnarray}
+	W_1(\mu,\nu)=  \inf_{\gamma \in \Gamma(\mu, \nu)} \int |x-y| d \gamma (x,y)
+\end{eqnarray}
+From the perspective of climate modelling, the WD has a nice physical interpretation given by the Monge-Kantorovich duality:
+\begin{eqnarray}
+	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
+\end{eqnarray}
+i.e.
+the difference in the expected value of some random variable between two distributions is no bigger than the WD value itself.
+More explicitly, if the WD for surface temperature between two distributions is 1K, then this tells us that the difference in the expected temperature between the distributions is no greater than 1K.
+For a full review of the Wasserstein distance, including favourable properties and comparison with other metrics we refer the reader to the Appendix of Pax21.
+
 For this injected strain ($h_0 = 10^{-12}$) the signal-to-noise ratio (SNR) is large (c.f.
@@
 These biases are a result of dropping the pulsar terms from the measurement equation and will be discussed in Section \ref{sec:bias}
+
+\subsubsection{Detection} \label{sec:detection}
+We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
+$\mathcal{M}_1$ is the Earth-terms only model, i.e.
+the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
+The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
+We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
+We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
+For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
+This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
+$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
+\newline
+
+There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10{-15}$.
+The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
+The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
+For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
+That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
+For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
+For the results presented here we set $n_{\rm live} =2000$.
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
+	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
+The horizontal grey dashed line labels the detection threshold of $\beta = 10$.
+The minimal detectable strain, below which $\beta < 10$, is $\sim 3.9 \times 10^{-15}$.
+Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$.}
+	\label{fig:bayes}
+\end{figure}
+
 \subsection{Exploring a broader parameter space} \label{sec:parameter_space}
@@
 
-\section{Wasserstein distance}
+\section{Wasserstein distance}\label{sec:wasserstein}
+
+It is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
+\begin{eqnarray}
+	\boldsymbol{E} [W_1(\mu_n,\mu)] \asymp n^{-\frac{1}{d}}
+\end{eqnarray}
+To sidestep this issue, we marginalize onto the distributions spanned by the individual gridpoints and take the WD between these 1-dimensional distributions.
+Such an approach is also adopted in \cite{Paxton2021,Vissio2020}.
+
 In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD).

### 2023-08-30 docs/manuscript/paper.tex
plot updates for smaller strains

@@
 	  & $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
-	  & $h_0$            & $10^{-12}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
+	  & $h_0$            & $5 \times 10^{-15}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
 	  & $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
@@
 The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
-It is evident that for this representative example we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for the majority of the parameters.
+It is evident that for this representative example with strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
+
+ If we consider the median values of the one-dimensional posteriors we can see that for some parameters the median is almost identically equal to the injected value (e.g. $\Omega)
+
  The clear exception is $\iota$ which exhibits a notable bias, and lies outside the with the 90\% credible interval, with the posterior median shifted from the injected value by $\sim 0.3$ radians.
+
  For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
@@
 We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.
-If we consider the median values of the one-dimensional posteriors we can see that for some parameters the median is almost identically equal to the injected value  (e.g.
-$\Omega, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the injected value (e.g.
-$\psi, \alpha$), whilst still being contained within the credible interval.
+
+  $, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the injected value (e.g. $\psi, \alpha$), whilst still being contained within the credible interval.
+
   This potential bias will also be investigated alongside the bias in $\iota$ in Section \ref{sec:multiple_noise} and Section \ref{sec:bias}.
+
  Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
@@
 We are able to accurately estimate each parameter of interest.
-We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
+We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
+Note that the $x$-axis scaling of the one-dimensional posteriors does not span the total prior space.}
 	\label{fig:corner_plot_1}
@@
 \subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
+%\begin{figure*}
+%	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
+%	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
+The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
+The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f.
+Figure \ref{fig:likelihood_surface}.}
+%	\label{fig:corner_plot_2}
+%\end{figure*}
+
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
 	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
@@
 
-To this end we use the the Wasserstein distance \citep[WD,][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive of similarity between probability distributions.
-
-A full review of the Wasserstein distance which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}
-
-In order to quantitatively define `similar'  we use . The WD is a metric which quantifies the distance between two probability distributions 
-as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$.
-The $p$-Wasserstein distance between two distributions $\mu, \nu$ is
+To this end we use the the Wasserstein distance \citep[WD,][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
+The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
+Throughout this work we use the first moment of the Wasserstein distance which for one-dimensional distributions $\mu$ and  $\nu$ is defined
 \begin{eqnarray}
-	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p}
+	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y)
 \end{eqnarray}
-where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function.
-
-In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD). The WD intuitively defines a distance between s
-
-We also take $p=1$ and $c(x,y) = |x-y|$ such that the WD values quoted in this work are given by
-\begin{eqnarray}
-	W_1(\mu,\nu)=  \inf_{\gamma \in \Gamma(\mu, \nu)} \int |x-y| d \gamma (x,y)
-\end{eqnarray}
-From the perspective of climate modelling, the WD has a nice physical interpretation given by the Monge-Kantorovich duality:
+where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.
+The WD has an additional useful physical interpretation given by the Monge-Kantorovich duality:
 \begin{eqnarray}
@@
 \end{eqnarray}
-i.e.
-the difference in the expected value of some random variable between two distributions is no bigger than the WD value itself.
-More explicitly, if the WD for surface temperature between two distributions is 1K, then this tells us that the difference in the expected temperature between the distributions is no greater than 1K.
-For a full review of the Wasserstein distance, including favourable properties and comparison with other metrics we refer the reader to the Appendix of Pax21.
+where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
+We can see that the WD bounds the difference in the expected values.
+For example, if we compare two posteriors for $\iota$ calculated for two different noise realisations, a WD = 1.0 rad tells us that the difference in the expected value of $\iota$ from the two distributions is less than 1 rad.
+Beyond optimal transport theory, the WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
+A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein} \textcolor{red}{TK: is this too much info on Wasserstein?
+Maybe just say "its a metric between probability distributions" and move the rest to appendix so as not to interrupt main astrophysics arguments.}
+
+For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate the WD between every pair of posteriors across our $10^3$ noise realisations.
 
@@
 To compare each of the posteriors across different noise realisations,
-
-\textcolor{red}{Wasserstein intro and discussion here}
 
@@
 \end{figure}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
+	\caption{} 
+	\label{fig:bias_for_large_h}
+\end{figure*}
+
 Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g.
@@
 \newline
+
+for low signal bias is covered by noise
 
@@
 \appendix
-
+\newpage
 \section{Kalman recursion equations} \label{sec:kalman}
@@
 This paper is useful: \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
-\section{Alternative version of Fig 1}
-
-\begin{figure*}
-	\centering
-	\includegraphics[width=\textwidth]{images/Kalman_example_both}
-	\caption{\textcolor{red}{TK: For consideration. Do we prefer this figure to Figure \ref{fig:four figures}}?}
-	\label{fig:kalmanexampleboth}
-\end{figure*}
 
@@
 
-\section{References}
-\label{sec:ref_list}
+%\section{References}
+%\label{sec:ref_list}
 

### 2023-08-30 py_src/plotting.py

@@ -134,8 +134,8 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
 
     #Make omega into nHz
     df_posterior["omega_gw"] = df_posterior["omega_gw"]*1e9
-    df_posterior["h"] = df_posterior["h"]*1e15
-    #df_posterior["h"] = df_posterior["h"]*1e12
+    #df_posterior["h"] = df_posterior["h"]*1e15
+    df_posterior["h"] = df_posterior["h"]*1e12
 
     display(df_posterior)
 

### 2023-08-30 docs/manuscript/paper.tex
PP plot at canonical low h

@@
 
- If we consider the median values of the one-dimensional posteriors we can see that for some parameters the median is almost identically equal to the injected value (e.g. $\Omega)
-
- The clear exception is $\iota$ which exhibits a notable bias, and lies outside the with the 90\% credible interval, with the posterior median shifted from the injected value by $\sim 0.3$ radians.
-
+ For some of the parameters the estimated posterior is approximately symmetric about the true injected value (e.g.
+$\Omega$).
+In this case the posterior median and the injected value are approximately coincident.
+For other parameters (e.g.
+$\iota$) the distribution is not symmetric about the injected value and the posterior median and the injected value are not coincident.
+The median value of the posterior for $\iota$ is shifted by $\sim 0.35$ rad relative to the injected value, although the injected value does still remain inside the 90\% credible interval.
+Similar effects are seen, albeit with a smaller shift, in other parameters such as $\delta$ and $\alpha$.
 For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
@@
 We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.
-
-  $, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the injected value (e.g. $\psi, \alpha$), whilst still being contained within the credible interval.
-
-  This potential bias will also be investigated alongside the bias in $\iota$ in Section \ref{sec:multiple_noise} and Section \ref{sec:bias}.
-
 Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
@@
 The above discussion was for a single realisation of the noise processes.
-It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} and \ref{sec:detection} is not particularly advantageous or convenient for our method.
+It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} is not particularly advantageous or convenient for our method.
 In this Section we confirm that we are able to infer comparable results for different noise realisations.
@@
 
-To this end we use the the Wasserstein distance \citep[WD,][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
+To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
 The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.

### 2023-08-31 docs/manuscript/paper.tex
Bayes 2000, 10 noise realisations

@@
 \subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
-%\begin{figure*}
-%	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
-%	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
-The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
-The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f.
-Figure \ref{fig:likelihood_surface}.}
-%	\label{fig:corner_plot_2}
-%\end{figure*}
-
 \begin{figure*}
@@
 \end{figure*}
-
-%\begin{figure*}
-%	%\centering % Not needed
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
-%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
-%		\label{fig:6MB_BFS}
-%	\end{subfigure}  
-%	\hfill
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
-%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
-%		\label{fig:25MB_bfs}
-%	\end{subfigure}
-%	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
-The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
-The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f.
-Figure \ref{fig:likelihood_surface}.}
-%	\label{fig:four figures2}
-%\end{figure*}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_surface}
-	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation.
-The red point labels the location of the likelihood maxima.
-The likelihood surface has been normalised with respect to the absolute value of the maxima.
-A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
-	\label{fig:likelihood_surface}
-\end{figure}
-%\begin{figure*}
-%	\centering
-%	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
-%	\caption{One-dimensional marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel).
-The thick vertical orange line labels the true injected value.
-The dashed vertical lines denote the median of the corresponding coloured posterior.
-Fixing $h$ reduces the variance between different noise realisations but results in a clear bias between the true and inferred values of $\iota$.
-The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column), but with the $(0.16,0.84)$ quantiles replaced with the $0.5$ quantile.}
-%	\label{fig:iota}
-%\end{figure*}
 The above discussion was for a single realisation of the noise processes.
@@
 We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
-As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here.
+As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr, reduced}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here.
 \newline
@@
 We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1}.
-Considering the one-dimensional marginalized posteriors, the inferred probability distributions of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ broadly overlap across the different noise realisations.
-Moreover, the variance of each of the individual one-dimensional posteriors is generally low i.e.
-each of the individual distributions are narrow compared to the prior space.
-For each different noise realisation the method infers a consistent maximum \textit{a posteriori} probability (MAP) estimate.
-Conversely for the parameters $\iota$ and $h_0$ the inferred one-dimensional posteriors do not broadly overlap are not consistent across different noise realisations and the variance of an individual posterior is generally larger than for the other parameters.
-That is to say, we infer very different MAP estimates of these parameters depending on the particular realisation of the noise.
-\newline
-
-The posteriors for $\iota$ and $h_0$ do not overlap for the different noise realisations due to these parameters being only weakly-identifiable.
-Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
-Sources of non-identifiability are generally categorised as either ``structural" (arising from the structure of the model) or ``practical" ( arising from insufficient data or measurement errors) \citep{GUILLAUME2019418}.
-Regarding $\iota$ and $h_0$, we are concerned with structural identifiability.
-Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx}, \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
-For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
-This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
-However, it can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
-That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a non-zero gradient close to the maximum likelihood solution.
-This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
-Whilst a single, unique maxima exists (labelled by the red point in the figure), along the ridge in the parameter space the likelihood values are all very similar.
-Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
-If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the the posteriors for the different noise realisations overlap.
-This is demonstrated in Figure \ref{fig:iota} where we plot the marginalized one-dimensional posterior of $\iota$ for 9 realisations of the noise in both the original case where $h_0$ is a free parameter (equivalent to the one-dimensional posterior for $\iota$ in Figure \ref{fig:corner_plot_2}) and in the case where we hold $h_0$ fixed at its injection value, $h_0 = 10^{-12}$.
-We can see that whilst the variance between the individual posteriors is large in the original case, once we fix $h_0$ the variance decreases dramatically and the sampling algorithm repeatedly converges consistent estimates of $\iota$.
-For the fixed $h_0$ case, evident in Figure \ref{fig:iota} is a strong bias of $\sim 0.35$ radians in the estimates for $\iota$ relative to the true injected value.
-Whilst the variance in the estimates across the noise realisations has decreased, the mean accuracy of the estimates has also decreased.
-An equivalent bias in $h_0$ is observed if we fix $\iota$.
-This bias is a result of dropping the pulsar terms from our measurement equation and is discussed in Section \ref{sec:bias}.
-\newline
-
-We now consider the complete set of $10^3$ noise realisations, rather than the subset of ten realisations in the preceding discussion.
-For each realisation of the noise we infer a marginalised one-dimensional posterior for each parameter.
+Considering the one-dimensional marginalized posteriors, the inferred probability distributions of the parameters broadly overlap across the different noise realisations.
+For the vast majority of parameters and noise realisations the injected value is contained within the 90 \% credible interval.
+\newline
+
+We now consider the complete set of $10^3$ noise realisations, rather than the subset of nine realisations in the preceding discussion.
+For each realisation of the noise, for each parameter, we infer a marginalised one-dimensional posterior.
 We want to compare these posteriors for different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
-
 To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
+The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
+A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}.
 The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
-Throughout this work we use the first moment of the Wasserstein distance which for one-dimensional distributions $\mu$ and  $\nu$ is defined
+Throughout this work we use the first moment of the Wasserstein distance $W_1(\mu,\nu)$ which is defined between one-dimensional probability distributions $\mu$ and $\nu$ as
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.
+where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.  \newline
+
+For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
+To form a single metric we then take the median value for each parameter across these pairwise calculations.
+The results are shown in Table \ref{tab:Wasserstein}.
+\begin{table}
+	\centering
+		\begin{tabular}{ccll}
+			\toprule
+			Parameter & Injected Values & Units & WD  \\
+			\hline
+			 $\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
+			 $\Phi_0$          & $0.20$ & rad & $0.13$ \\
+			 $\psi$              & $2.50$ & rad & $0.16$ \\
+			 $\iota$             & $1.0$ & rad & $0.14$ \\ 
+			 $\delta$              & $1.0$  & rad & $0.09$ \\
+			 $\alpha$          & $1.0$  & rad & $0.17$\\
+			 $h_0$            & $5 \times 10^{-15}$ & - & $8 \times 10^{-16}$ \\
+			\bottomrule
+		\end{tabular}
+		\caption{}
+		\label{tab:Wasserstein}
+	\end{table}
 The WD has an additional useful physical interpretation given by the Monge-Kantorovich duality:
@@
 We can see that the WD bounds the difference in the expected values.
-For example, if we compare two posteriors for $\iota$ calculated for two different noise realisations, a WD = 1.0 rad tells us that the difference in the expected value of $\iota$ from the two distributions is less than 1 rad.
-Beyond optimal transport theory, the WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
-A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein} \textcolor{red}{TK: is this too much info on Wasserstein?
-Maybe just say "its a metric between probability distributions" and move the rest to appendix so as not to interrupt main astrophysics arguments.}
-
-For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate the WD between every pair of posteriors across our $10^3$ noise realisations.
+For example, in Table \ref{tab:Wasserstein} we report a median $W_1 = 0.14$ rad for $\iota$.
+This tells us that the across all noise realisations the difference in the expected value of $\iota$ is less than 1 rad.
+We can see that the bounds from the WD are tight for all parameters; as a percentage of the width of the prior space, the difference in the expected values for the distributions of each parameter are $\lesssim 2 \%$.
+
+See also Figure \ref{fig:pairwise_wasserstein}
+
+\begin{figure*}
+	\setkeys{Gin}{width=\linewidth}
+
+	\begin{subfigure}[b]{0.22\textwidth}
+			\includegraphics[width=\textwidth]{images/WD_0}
+			\caption{$\Omega$}
+		\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+			\includegraphics[width=\textwidth]{images/WD_1}
+			\caption{$\Phi_0$}
+		\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.22\textwidth}
+			\includegraphics[width=\textwidth]{images/WD_2}
+			\caption{$\psi$}
+		\end{subfigure}
+    \hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+			\includegraphics[width=\textwidth]{images/WD_3}
+			\caption{$\iota$}
+		\end{subfigure}
+	\medskip
+	\begin{subfigure}[b]{0.3\textwidth}
+			\includegraphics[width=\textwidth]{images/WD_4}
+			\caption{$\delta$}
+		\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.3\textwidth}
+			\includegraphics[width=\textwidth]{images/WD_5}
+			\caption{$\alpha$}
+		\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.3\textwidth}
+			\includegraphics[width=\textwidth]{images/WD_6}
+			\caption{$h$}
+		\end{subfigure}
+	\caption{Wassersteins}\label{fig:pairwise_wasserstein}
+\end{figure*}
+
+if we compare two posteriors for $\iota$ calculated for two different noise realisations, a WD = 1.0 rad tells us that the difference in the expected value of from the two distributions is less than 1 rad.
+\newline
+
+Considering now the
+
+Wd tells us only how well different distributions agreee. It doesnt tell use about the accuray of the posteriors themselves. c.f. bias
+
+As in the singe noise case, there are tentative signs in a bias
+
+Considering now the
+
+$\Omega, \Phi_0, \psi, \delta$ and $\alpha$
+
+%Moreover, the variance of each of the individual one-dimensional posteriors is generally low i.e.
+each of the individual distributions are narrow compared to the prior space.
+For each different noise realisation the method infers a consistent maximum \textit{a posteriori} probability (MAP) estimate.
+Conversely for the parameters $\iota$ and $h_0$ the inferred one-dimensional posteriors do not broadly overlap are not consistent across different noise realisations and the variance of an individual posterior is generally larger than for the other parameters.
+That is to say, we infer very different MAP estimates of these parameters depending on the particular realisation of the noise.
+\newline
+
+We now consider the complete set of
 
@@
 
-we consider the normalised standard deviation in the set of $10^3$ median values i.e. 
-\begin{equation}
-	\rho \left( \boldsymbol{\theta}\right)= \frac{1}{\mu(\boldsymbol{\theta})} \sqrt{ \frac{\sum_{i=1}^{i=10^3}\left[x_i(\boldsymbol{\theta}) - \mu(\boldsymbol{\theta}) \right]}{10^3}} 
-\end{equation} 
-where $\mu(\boldsymbol{\theta})$ is the mean value across the distribution of $10^3$ samples for parameter $\boldsymbol{\theta}$, and $x_i$ the value of the $i$-th sample.
+%
+%
+%we consider the normalised standard deviation in the set of $10^3$ median values i.e. 
+%\begin{equation}
+%	\rho \left( \boldsymbol{\theta}\right)= \frac{1}{\mu(\boldsymbol{\theta})} \sqrt{ \frac{\sum_{i=1}^{i=10^3}\left[x_i(\boldsymbol{\theta}) - \mu(\boldsymbol{\theta}) \right]}{10^3}} 
+%\end{equation} 
+%where $\mu(\boldsymbol{\theta})$ is the mean value across the distribution of $10^3$ samples for parameter $\boldsymbol{\theta}$, and $x_i$ the value of the $i$-th sample.
 For the parameters $\{\Omega, \Phi_0, \psi, \delta, \alpha \}$ we have $\rho \sim \{4 \times 10^{-5}, 0.03,0.004,0.03,0.02 \}$ respectively.
@@
 \newline
+%
+%
+%
+%
+%
 
@@
 Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$.
-The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad in order to maintain an approximately constant SNR.
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-14}$ and 1.0 rad in order to maintain an approximately constant SNR.
 The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals.
@@
 An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.
+
+\subsection{Identifiability and bias}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
+	\caption{}
+	\label{}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_surface}
+	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation.
+The red point labels the location of the likelihood maxima.
+The likelihood surface has been normalised with respect to the absolute value of the maxima.
+A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
+	\label{fig:likelihood_surface}
+\end{figure}
+
+The posteriors for $\iota$ and $h_0$ do not overlap for the different noise realisations due to these parameters being only weakly-identifiable.
+Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
+Sources of non-identifiability are generally categorised as either ``structural" (arising from the structure of the model) or ``practical" ( arising from insufficient data or measurement errors) \citep{GUILLAUME2019418}.
+Regarding $\iota$ and $h_0$, we are concerned with structural identifiability.
+Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx}, \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
+This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
+However, it can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
+That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a non-zero gradient close to the maximum likelihood solution.
+This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
+Whilst a single, unique maxima exists (labelled by the red point in the figure), along the ridge in the parameter space the likelihood values are all very similar.
+Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
+If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the the posteriors for the different noise realisations overlap.
+This is demonstrated in Figure \ref{fig:iota} where we plot the marginalized one-dimensional posterior of $\iota$ for 9 realisations of the noise in both the original case where $h_0$ is a free parameter (equivalent to the one-dimensional posterior for $\iota$ in Figure \ref{fig:corner_plot_2}) and in the case where we hold $h_0$ fixed at its injection value, $h_0 = 10^{-12}$.
+We can see that whilst the variance between the individual posteriors is large in the original case, once we fix $h_0$ the variance decreases dramatically and the sampling algorithm repeatedly converges consistent estimates of $\iota$.
+For the fixed $h_0$ case, evident in Figure \ref{fig:iota} is a strong bias of $\sim 0.35$ radians in the estimates for $\iota$ relative to the true injected value.
+Whilst the variance in the estimates across the noise realisations has decreased, the mean accuracy of the estimates has also decreased.
+An equivalent bias in $h_0$ is observed if we fix $\iota$.
+This bias is a result of dropping the pulsar terms from our measurement equation and is discussed in Section \ref{sec:bias}.
+\newline
 
@@
 
-%\begin{figure*}
-%	\setkeys{Gin}{width=\linewidth}   
-%	
-%	\begin{subfigure}[b]{0.3\textwidth}
-	%		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
-	%		\caption{$\Omega$}
-	%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.3\textwidth}
-	%		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
-	%		\caption{$\Phi_0$}
-	%	\end{subfigure}
-%	\hfill	
-%	\begin{subfigure}[b]{0.3\textwidth}
-	%		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
-	%		\caption{$\psi$}
-	%	\end{subfigure}
-%	\medskip
-%	
-%	\begin{subfigure}[b]{0.3\textwidth}
-	%		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
-	%		\caption{$\iota$}
-	%	\end{subfigure}
-%	\hfill
-%	\begin{subfigure}[b]{0.3\textwidth}
-	%		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
-	%		\caption{$\delta$}
-	%	\end{subfigure}
-%	\hfill	
-%	\begin{subfigure}[b]{0.3\textwidth}
-	%		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
-	%		\caption{$\alpha$}
-	%	\end{subfigure}
-%	\medskip
-%	
-%	
-%	\medskip
-%	\begin{subfigure}[b]{0.33\textwidth}
-	%		\includegraphics{images/distribution_h}
-	%		\caption{$h$}
-	%	\end{subfigure}
-%	
-%	\caption{Distribution in the median values of the 1D marginalized posteriors inferred by nested sampling for each static parameters of $\boldsymbol{\theta}_{\rm gw}$ over 1000 noise realisations.
-The orange dashed line labels the true injection value and the red dashed line labels the median of the distribution.
-\textcolor{red}{TK: these plots are ugly.
-Is there a better way to show this?}}\label{fig:median_distriubutins}
-%\end{figure*}
-
 %

### 2023-08-31 docs/manuscript/paper.tex
1000 noise runs at 5e-15 h

@@
 \end{figure*}
-The above discussion was for a single realisation of the noise processes.
-It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} is not particularly advantageous or convenient for our method.
-In this Section we confirm that we are able to infer comparable results for different noise realisations.
-To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.
-For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
-In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise.
-This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}.
-We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
-As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr, reduced}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here.
-\newline
-
-We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1}.
-Considering the one-dimensional marginalized posteriors, the inferred probability distributions of the parameters broadly overlap across the different noise realisations.
-For the vast majority of parameters and noise realisations the injected value is contained within the 90 \% credible interval.
-\newline
-
-We now consider the complete set of $10^3$ noise realisations, rather than the subset of nine realisations in the preceding discussion.
-For each realisation of the noise, for each parameter, we infer a marginalised one-dimensional posterior.
-We want to compare these posteriors for different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
-To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
-The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
-A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}.
-The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
-Throughout this work we use the first moment of the Wasserstein distance $W_1(\mu,\nu)$ which is defined between one-dimensional probability distributions $\mu$ and $\nu$ as
-\begin{eqnarray}
-	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y)
-\end{eqnarray}
-where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.  \newline
-
-For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
-To form a single metric we then take the median value for each parameter across these pairwise calculations.
-The results are shown in Table \ref{tab:Wasserstein}.
 \begin{table}
@@
 		\toprule
-			Parameter & Injected Values & Units & WD  \\
+		Parameter & Injected Values & Units & $W_{1, \rm median}$  \\
 		\hline
@@
 	\end{tabular}
-		\caption{}
+	\caption{Median Wasserstein distance $W_1$ for each parameter calculated across every pair of the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}}
 	\label{tab:Wasserstein}
 \end{table}
-The WD has an additional useful physical interpretation given by the Monge-Kantorovich duality:
-\begin{eqnarray}
-	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
-\end{eqnarray}
-where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
-We can see that the WD bounds the difference in the expected values.
-For example, in Table \ref{tab:Wasserstein} we report a median $W_1 = 0.14$ rad for $\iota$.
-This tells us that the across all noise realisations the difference in the expected value of $\iota$ is less than 1 rad.
-We can see that the bounds from the WD are tight for all parameters; as a percentage of the width of the prior space, the difference in the expected values for the distributions of each parameter are $\lesssim 2 \%$.
-
-See also Figure \ref{fig:pairwise_wasserstein}
-
 \begin{figure*}
@@
 		\includegraphics[width=\textwidth]{images/WD_2}
-			\caption{$\psi$}
+		\caption{$\psi$} \label{figWD_psi}
 	\end{subfigure}
@@
 	\end{subfigure}
-	\caption{Wassersteins}\label{fig:pairwise_wasserstein}
+	\caption{First moment of the Wasserstein distance $W_1$ between each pair of one-dimensional posteriors estimated for each parameter, for each noise realisation.
+$W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
+$W_1$ is generally small across all parameters and all posteriors, with some correlated greater values for $\alpha$ and $\psi$.}\label{fig:pairwise_wasserstein}
 \end{figure*}
 
+The above discussion was for a single realisation of the noise processes.
+It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} is not particularly advantageous or convenient for our method.
+In this Section we confirm that we are able to infer comparable results for different noise realisations.
+To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.
+For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
+In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise.
+This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}.
+We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
+As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr, reduced}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here.
+\newline
+
+We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1}.
+Considering the one-dimensional marginalized posteriors, the inferred probability distributions of the parameters broadly overlap across the different noise realisations.
+For the vast majority of parameters and noise realisations the injected value is contained within the 90 \% credible interval.
+\newline
+
+We now consider the complete set of $10^3$ noise realisations, rather than the subset of nine realisations in the preceding discussion.
+For each realisation of the noise, for each parameter, we infer a marginalised one-dimensional posterior.
+We want to compare these posteriors for different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
+To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
+The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
+A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}.
+The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
+Throughout this work we use the first moment of the Wasserstein distance $W_1(\mu,\nu)$ which is defined between one-dimensional probability distributions $\mu$ and $\nu$ as
+\begin{eqnarray}
+	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y)
+\end{eqnarray}
+where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.  \newline
+
+For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
+The results are presented in Figure \ref{fig:pairwise_wasserstein} which shows the $W_1$ value for each pair of posteriors.
+The WD has a useful physical interpretation given by the Monge-Kantorovich duality:
+\begin{eqnarray}
+	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
+\end{eqnarray}
+where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
+We can see that the $W_1$ bounds the difference in the expected values.
+That is to say, if $W_1 =0.5$ rad for two one-dimensional posterior distributions of e.g.
+$\iota$ this means that the difference in the expected values from these distributions is no greater than $0.5$.
+
+From Figure \ref{fig:pairwise_wasserstein} it is evident that generally $W_1$ is small across all pairs of posteriors.
+The median values across all pairs for each parameter are quoted in Table \ref{tab:Wasserstein}.
+
+As a percentage of the width of the prior space (i.e.
+$=2 \pi$ for $\alpha$), the median $W_1$ for each parameter is $\lesssim 2 \%$.
+Generally speaking, the method converges to consistent one-dimensional posteriors for each parameter across different noise realisations.
+
+There are a small fraction of the total number of noise realisations
+which have a large distance from all the other posteriors of all the other noise realisations. These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
+
+These correspond exclusively to posteriors which have converged poorly and not found the global maximum.
+For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi} corresponds to a posterior which has converged poorly, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
+This can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
+There is a strong correlation between the $W_1$ inferred for $\alpha$ and $\psi$ (Pearson correlation coefficient $=0.96$)
+
+indicating that poor estimats of the one parameter that have converged poorly
+
+ which result in posteriors with larger values of $W_1$.
+
+  indicating that the posterior inferred for this noise realisation has a large distance
+
+Some parameters
+
+We can see that the posteriors are bro
+
+To form a single metric we then take the median value for each parameter across these pairwise calculations. The results are shown in.
+
+See also Figure \ref{fig:pairwise_wasserstein}
+
 if we compare two posteriors for $\iota$ calculated for two different noise realisations, a WD = 1.0 rad tells us that the difference in the expected value of from the two distributions is less than 1 rad.
@@
 
-Wd tells us only how well different distributions agreee. It doesnt tell use about the accuray of the posteriors themselves. c.f. bias
-
+\textcolor{red}{We could also have a PP plot here}
+
+WD tells us only how well different distributions agreee. It doesnt tell use about the accuray of the posteriors themselves. c.f. bias
+
+\textcolor{red}{Intro biases here}
 As in the singe noise case, there are tentative signs in a bias 
-
-Considering now the
-
-$\Omega, \Phi_0, \psi, \delta$ and $\alpha$
-
-%Moreover, the variance of each of the individual one-dimensional posteriors is generally low i.e.
-each of the individual distributions are narrow compared to the prior space.
-For each different noise realisation the method infers a consistent maximum \textit{a posteriori} probability (MAP) estimate.
-Conversely for the parameters $\iota$ and $h_0$ the inferred one-dimensional posteriors do not broadly overlap are not consistent across different noise realisations and the variance of an individual posterior is generally larger than for the other parameters.
-That is to say, we infer very different MAP estimates of these parameters depending on the particular realisation of the noise.
-\newline
-
-We now consider the complete set of
-
-For this injected strain ($h_0 = 10^{-12}$) the signal-to-noise ratio (SNR) is large (c.f.
-Figure \ref{fig:bayes}) and consequently the posteriors are typically narrow compared to the prior space and well described by their median value.
-To compare each of the posteriors across different noise realisations,
-
-%
-%
-%we consider the normalised standard deviation in the set of $10^3$ median values i.e. 
-%\begin{equation}
-%	\rho \left( \boldsymbol{\theta}\right)= \frac{1}{\mu(\boldsymbol{\theta})} \sqrt{ \frac{\sum_{i=1}^{i=10^3}\left[x_i(\boldsymbol{\theta}) - \mu(\boldsymbol{\theta}) \right]}{10^3}} 
-%\end{equation} 
-%where $\mu(\boldsymbol{\theta})$ is the mean value across the distribution of $10^3$ samples for parameter $\boldsymbol{\theta}$, and $x_i$ the value of the $i$-th sample.
-For the parameters $\{\Omega, \Phi_0, \psi, \delta, \alpha \}$ we have $\rho \sim \{4 \times 10^{-5}, 0.03,0.004,0.03,0.02 \}$ respectively.
-We can therefore see that the method is able to reliably converge to a consistent MAP estimate.
-Conversely, as we saw before, the MAP estimates for $\iota$ and $h_0$ are not consistent across different noise realisations, with $\rho \sim 0.3$ and $\rho \sim 0.1$ respectively.
-This is due to the weak-identifiability issue that we have discussed.
-\newline
-%
-%
-%
-%
-%
-
 We saw previously strong biases in $\iota$ and $h_0$ when considering just 9 noise realisations.
@@
 \subsection{Identifiability and bias}
-
+There are two underlying issues in the preceding tests that we now explore in more detail.
+The first relates to parameter identifiability, i.e.
+can we uniquely identify the parameters of the model given the data and the model structure?
+The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
+In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ for our representative example system (Table \ref{tab:parameters_and_priors})
+
+\subsubsection{Identifiability}
 \begin{figure}
@@
 	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
-	\caption{}
-	\label{}
+	\caption{Posterior distribution over for the GW source parameters $\iota$ and $h_0$ for the representative system described in Table \ref{tab:parameters_and_priors}, in the high SNR regime, over nine noise realisations.
+Despite being in the high SNR regime, the posteriors do not overlap as much as expected \textit{a priori}, and much less than the other parameters of $\boldsymbol{\theta}_{\rm gw}$.}
+	\label{fig:just_iota_h}
 \end{figure}
-
 \begin{figure}
@@
 \end{figure}
-
-The posteriors for $\iota$ and $h_0$ do not overlap for the different noise realisations due to these parameters being only weakly-identifiable.
+In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations.
+In the case of high-SNR we might expect that our estimates would be increasingly accurate over the low-SNR case, and that the method converges robustly to similar posteriors.
+Whilst this is generally true for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ and the posteriors across noise realisations are largely indistinguishable, for $\iota$ and $h_0$ this is not the case.
+Instead the inferred one-dimensional posteriors do not broadly overlap are not consistent across different noise realisations.
+Specifically the median WD across $10^3$ noise realisations is $W_1 = 0.24$ rad and $\iota$ and $W_1 = 1.6 \times 10^{-13}$ for $h_0$.
+For $\iota$ this WD is greater than the value inferred in the low SNR case (0.14, Table \ref{tab:parameters_and_priors}, whilst the WD value for $h_0$ is also greater as a proportion of the scale set by the strain amplitude.
+This is in contrast to the median WD inferred for the other parameters; e.g.
+for$\Phi_0$, $W_1 = 0.0045$ rad.
+\newline
+
+This issue of inferring very different posteriors depending on the particular realisation of the noise is due to these parameters being only weakly-identifiable.
 Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
@@
 Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
-If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the the posteriors for the different noise realisations overlap.
-This is demonstrated in Figure \ref{fig:iota} where we plot the marginalized one-dimensional posterior of $\iota$ for 9 realisations of the noise in both the original case where $h_0$ is a free parameter (equivalent to the one-dimensional posterior for $\iota$ in Figure \ref{fig:corner_plot_2}) and in the case where we hold $h_0$ fixed at its injection value, $h_0 = 10^{-12}$.
-We can see that whilst the variance between the individual posteriors is large in the original case, once we fix $h_0$ the variance decreases dramatically and the sampling algorithm repeatedly converges consistent estimates of $\iota$.
-For the fixed $h_0$ case, evident in Figure \ref{fig:iota} is a strong bias of $\sim 0.35$ radians in the estimates for $\iota$ relative to the true injected value.
-Whilst the variance in the estimates across the noise realisations has decreased, the mean accuracy of the estimates has also decreased.
-An equivalent bias in $h_0$ is observed if we fix $\iota$.
-This bias is a result of dropping the pulsar terms from our measurement equation and is discussed in Section \ref{sec:bias}.
-\newline
-
-\subsection{Bias by dropping pulsar terms}\label{sec:bias}
+If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably.
+However, the resulting maximum a posteriori probability estimate for $\iota$ is strongly biased away from the true value by $\sim 0.35$ rad.
+This bias will be discussed further in Section \ref{sec:bias}.
+In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" the likelihood ridge, leading to a more concave likelihood surface.
+This enables consistent posteriors to be estimated for $\iota$ and $h$.
+
+\subsubsection{Bias by dropping pulsar terms}\label{sec:bias}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
+	\caption{As Figure \ref{fig:corner_plot_1} but for the low-SNR system with $h_0 = 10^{-12}$.
+For the majority of parameters the injection value does not fall within the 90 \% credible interval and is not visible over these scales.
+The estimated posteriors are biased away from the injected value due to the dropping of the pulsar terms.}
+	\label{fig:bias_for_large_h}
+\end{figure*}
+
 \begin{figure}
@@
 \end{figure}
-
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
-	\caption{} 
-	\label{fig:bias_for_large_h}
-\end{figure*}
-
 Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g.
 the median of the one-dimensional marginalised posterior) is biased away from the true injection value.
-This effect is most severe for $\iota$ and $h_0$ but also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
+In Figure \ref{fig:bias_for_large_h} we show the resulting posteriors for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but now for the high-SNR system with $h_0 = 10^{-12}$.
+We can see that for all parameters, with the exception of $\Omega$ and $h_0$, the injected value lies outside 90 \% credible interval, to the extent that it is not visible on the scales over which we have plotted the posteriors.
+This effect is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
+Whilst $h_0$ does not exhibit any noticeable bias for this particular noise realisation, in general across multiple noise realisations it does exhibit a correlated bias with $\iota$.
 In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
 \newline
-
-for low signal bias is covered by noise
 
@@
 This is the origin of the bias in the one-dimensional marginalised posteriors that we observe.
-For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ radians, whereas the injected value was $\iota = 1.0$ radians, This difference is comparable to the bias observed in the right-hand panel of Figure \ref{fig:iota} where the median of the marginalised posterior for $\iota$ was $\sim 0.65$ radians.
+For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ radians, whereas the injected value was $\iota = 1.0$ radians.
+This optima is comparable to the median of the posterior inferred for $\iota$ in Figure \ref{fig:bias_for_large_h}.
+\newline
+
+%This difference is comparable to the bias observed in the right-hand panel of Figure \ref{fig:iota} where the median of the marginalised posterior for $\iota$ was $\sim 0.65$ radians.
 The discrepancy between $\iota \sim 0.7$ in the log-likelihood curve and $\iota \sim 0.65$ in the nested sampling inference is a result of the likelihood curves being a one-dimensional representation of a multidimensional problem.
@@
 
-In contrast to the Earth-terms solution, for the solution which includes the pulsar terms the optima of the likelihood curve now aligns with the true injected value.
-The corollary is also true; if we drop the pulsar terms from the synthetic data and use the Earth terms equation for inference the optima is also correctly located.
+In contrast to the Earth-terms solution, for the pulsar-terms solution the optima of the likelihood curve now aligns with the true injected value.
+The corollary is also true; if we drop the pulsar terms from the synthetic data and use the Earth terms equation for inference then the optima is also correctly located.
 This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.
@@
 Equation \ref{eq:measurement}).
-Explicitly, we have the ``Earth-terms wave with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
+Explicitly, we have the ``Earth-terms wave"	 with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
 The net interference wave that modulates $f_{\rm p}$ is
@@
 If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate and the bias may not be important for searches.
-Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties then the bias may come to systematically affect the inferred parameters.
-\newline
+
+This is what we observe for our synthetic data (e.g.
+Figure \ref{fig:corner_plot_2}); the decreased SNR manifests as a broader marginalised posterior which generally overwhelms the shift due to the bias.
+
+\textcolor{red}{However, PP plot shows...}
+
+Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may come to systematically affect the inferred parameters. \newline
 
@@
 
-\section{Appendix on identifiability}\label{appendix_identifiability}
+\section{Identifiability}\label{appendix_identifiability}
 

### 2023-09-02 docs/manuscript/paper.tex
Bayes, 2000 points, 10 realisations

@@
 The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}.
-However for this work, whilst the pulsar distance is used when generating synthetic data, as discussed in Section \ref{sec:parameter_estim}, the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$.
+However for this work, whilst the pulsar distance is used when generating synthetic data, as discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$.
 Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
@@
 It is evident that for this representative example with strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
+\newline
 
@@
 We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
-Note that the $x$-axis scaling of the one-dimensional posteriors does not span the total prior space.}
+Note that the $x$-axis scaling does not span the total prior space.}
 	\label{fig:corner_plot_1}
@@
 	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
-	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes.
+	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes.
+The injection value is generally contained within the 90 \% credible interval for the vast of majority posteriors.
 The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
-The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f.
-Figure \ref{fig:likelihood_surface}.}
+A slight skew-bias is apparent in some of the parameters, e.g.
+$\iota$.
+Correlations can be seen between $\Omega - \Phi_0$, $\psi-\alpha$ and $\iota - h_0$.}
 	\label{fig:corner_plot_2}
@@
 	\end{tabular}
-	\caption{Median Wasserstein distance $W_1$ for each parameter calculated across every pair of the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}}
+	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
+$W_{1, \rm median}$  sets the bound on the difference in expectation value between the one dimensional posteriors, and is generally small with respect to the width of the prior space.}
 	\label{tab:Wasserstein}
@@
 	\end{subfigure}
-	\caption{First moment of the Wasserstein distance $W_1$ between each pair of one-dimensional posteriors estimated for each parameter, for each noise realisation.
+	\caption{For each parameter of $\boldsymbol{\theta}_{\rm gw}$ the first moment of the Wasserstein distance, $W_1$, calculated between each pair of one-dimensional posteriors for our representative system across $10^3$ realisations.
 $W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
-$W_1$ is generally small across all parameters and all posteriors, with some correlated greater values for $\alpha$ and $\psi$.}\label{fig:pairwise_wasserstein}
+$W_1$ is generally small across all parameters and all posteriors, but some posteriors converge poorly leading to large values of $W_1$.
+The $W_1$ values inferred for $\psi$ and $\alpha$ are strongly correlated with correlation coefficient = 0.96.}\label{fig:pairwise_wasserstein}
 \end{figure*}
-
-The above discussion was for a single realisation of the noise processes.
-It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} is not particularly advantageous or convenient for our method.
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_surface_alpha_psi}
+	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\alpha-\psi$ parameter space for a single noise realisation.
+The red point labels the location of the maxima, and the magenta point labels the location of the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data.
+The likelihood has been normalised with respect to the absolute value of the maxima.
+A multi-modal structure is clearly visible and for this noise realisation the nested sampler has settled in the local optima.
+This can be solved generally by increasing the number of live points $n_{\rm live}$}
+	\label{fig:likelihood_surface_alpha_psi}
+\end{figure}
+The above example was for a single realisation of the noise processes.
+It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous or convenient for our method.
 In this Section we confirm that we are able to infer comparable results for different noise realisations.
-To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.
+To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate $1000$ \footnote{\tiny\textcolor{red}{TK: For the results plotted here we currently just do 100 noise realisations.
+1000 is a hangover when we were using a larger strain ($h=10^{-12}$) as our canonical example.
+The $10^3$ noise realisations are currently running and plots will be updated once completed, but I don't expect any meaningful changes \textcolor{red}{since for $h=10^{-12}$ the results/conclusions of 100 realisations are the same as 1000 realisations)}}\normalsize} realisations of the process noise and measurement noise.
 For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
@@
 This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}.
-We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
+We plot only nine realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
 As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr, reduced}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here.
@@
 For the vast majority of parameters and noise realisations the injected value is contained within the 90 \% credible interval.
-\newline
+As in the singe noise case, there are tentative signs of bias for some of the parameters, with the one-dimensional posteriors not symmetric about the injected value.
+For example, the maximum a posteriori probability estimates of $\iota$ across the noise realisations all appear left-skewed, consistently underestimating the injected value.
+Given the uncertainty in the posteriors it is difficult to draw strong conclusions on a potential bias and this question will be discussed in detail in Section \ref{sec:bias} \newline
+
+%
+%We saw previously strong biases in $\iota$ and $h_0$ when considering just 9 noise realisations.
+These biases persist with a similar magnitude when we consider the full set of $10^3$ noise realisations.
+Moreover, smaller biases are also observed in $\psi$ ($\sim 0.1$ radians) and $\alpha$ ($\sim 0.04$ radians) as well as smaller still biases in $\delta$ and $\Phi_0$.
+%
+%
+%
+%These biases are a result of dropping the pulsar terms from the measurement equation and will be discussed in Section
 
@@
 For each realisation of the noise, for each parameter, we infer a marginalised one-dimensional posterior.
-We want to compare these posteriors for different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
+Our goal is to compare these posteriors between different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
 To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
-The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
+\footnote{\textcolor{red}{\tiny TK: How much of the following is ok, how much to move to the appendix?
+I don't want to get bogged down in WD and distract from astro arguments but we use $W_1$ as metric so I feel like it should be defined in the main body of the text, and the bounding result is important...} \normalsize} The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
 A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}.
@@
 The results are presented in Figure \ref{fig:pairwise_wasserstein} which shows the $W_1$ value for each pair of posteriors.
-The WD has a useful physical interpretation given by the Monge-Kantorovich duality:
+The first moment of the WD has a useful physical interpretation given by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
 \begin{eqnarray}
@@
 We can see that the $W_1$ bounds the difference in the expected values.
-That is to say, if $W_1 =0.5$ rad for two one-dimensional posterior distributions of e.g.
-$\iota$ this means that the difference in the expected values from these distributions is no greater than $0.5$.
-
+That is to say, if $W_1 =0.5$ rad between two one-dimensional posterior distributions of e.g.
+$\iota$ this means that the difference in the expected values from these distributions is no greater than $0.5$ rad.
 From Figure \ref{fig:pairwise_wasserstein} it is evident that generally $W_1$ is small across all pairs of posteriors.
 The median values across all pairs for each parameter are quoted in Table \ref{tab:Wasserstein}.
-
-As a percentage of the width of the prior space (i.e.
-$=2 \pi$ for $\alpha$), the median $W_1$ for each parameter is $\lesssim 2 \%$.
+As a percentage of the width of the prior space (e.g.
+for $\alpha$ the width of the prior $=2 \pi$), the median $W_1$ for each parameter is $\lesssim 3 \%$.
 Generally speaking, the method converges to consistent one-dimensional posteriors for each parameter across different noise realisations.
+\newline
 
 There are a small fraction of the total number of noise realisations
-which have a large distance from all the other posteriors of all the other noise realisations. These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
-
-These correspond exclusively to posteriors which have converged poorly and not found the global maximum.
-For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi} corresponds to a posterior which has converged poorly, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
-This can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
-There is a strong correlation between the $W_1$ inferred for $\alpha$ and $\psi$ (Pearson correlation coefficient $=0.96$)
-
-indicating that poor estimats of the one parameter that have converged poorly
-
- which result in posteriors with larger values of $W_1$.
-
-  indicating that the posterior inferred for this noise realisation has a large distance
-
-Some parameters
-
-We can see that the posteriors are bro
-
-To form a single metric we then take the median value for each parameter across these pairwise calculations. The results are shown in.
-
-See also Figure \ref{fig:pairwise_wasserstein}
-
-if we compare two posteriors for $\iota$ calculated for two different noise realisations, a WD = 1.0 rad tells us that the difference in the expected value of from the two distributions is less than 1 rad.
-\newline
-
-Considering now the
-
-\textcolor{red}{We could also have a PP plot here}
-
-WD tells us only how well different distributions agreee. It doesnt tell use about the accuray of the posteriors themselves. c.f. bias
-
-\textcolor{red}{Intro biases here}
-As in the singe noise case, there are tentative signs in a bias 
-We saw previously strong biases in $\iota$ and $h_0$ when considering just 9 noise realisations.
-These biases persist with a similar magnitude when we consider the full set of $10^3$ noise realisations.
-Moreover, smaller biases are also observed in $\psi$ ($\sim 0.1$ radians) and $\alpha$ ($\sim 0.04$ radians) as well as smaller still biases in $\delta$ and $\Phi_0$.
-These biases are a result of dropping the pulsar terms from the measurement equation and will be discussed in Section \ref{sec:bias}
+which have a large distance from all the other posteriors of all the other noise realisations.
+These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
+These correspond exclusively to posteriors which have converged poorly and not found the likelihood global maximum.
+For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the Figure, corresponds to a posterior which has converged poorly, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
+The $W_1$ between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ radians.
+This problem can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
+We also observe a a strong correlation (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
+This is due to multimodal likelihood surface for $\psi-\alpha$, shown in Figure \ref{fig:likelihood_surface_alpha_psi}.
+We can see that the likelihood surface has a global optima at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local optima that corresponds to the location of the poorly converged posterior discussed previously with $\psi = 1.01$ rad and a large $W_1$.
 
@@
 Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
-For the results presented here we set $n_{\rm live} =2000$.
+For the results presented in this subsection we set $n_{\rm live} =2000$.
 \begin{figure}
@@
 Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$.
-The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-14}$ and 1.0 rad in order to maintain an approximately constant SNR.
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad in order to maintain an approximately constant SNR.
 The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals.
@@
 The results are shown in Figure \ref{fig:parameter_space}.
-The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections."
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections.
 We can see that only $\Omega$ falls within the $3\sigma$ shaded region.
@@
 The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
-In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ for our representative example system (Table \ref{tab:parameters_and_priors})
+In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f.
+Figure \ref{fig:bayes}) for our representative example system (Table \ref{tab:parameters_and_priors}).
+\footnote{\tiny \textcolor{red}{TK: This discussion on bias and identifiability is perhaps less relevant than it was when our representative example had a 'high' strain ($10^{-12}$).
+However it IS still a feature/bug of our approach and I think it is worth discussing in the main body rather than relegating to an appendix.
+Opinions welcome...} \normalsize}
 
@@
 	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
-	\caption{Posterior distribution over for the GW source parameters $\iota$ and $h_0$ for the representative system described in Table \ref{tab:parameters_and_priors}, in the high SNR regime, over nine noise realisations.
-Despite being in the high SNR regime, the posteriors do not overlap as much as expected \textit{a priori}, and much less than the other parameters of $\boldsymbol{\theta}_{\rm gw}$.}
+	\caption{Posterior distributions over nine noise realisations for the GW source parameters $\iota$ and $h_0$ for the representative system described in Table \ref{tab:parameters_and_priors}, in the high SNR regime ($h_0 = 10^{-12}$).
+Unlike the other parameters of  $\boldsymbol{\theta}_{\rm gw}$, the one-dimensional posteriors for $\iota$ and $h_0$ do not strongly overlap.
+A correlation between  $\iota$ and $h_0$ is evident.}
 	\label{fig:just_iota_h}
@@
 In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations.
-In the case of high-SNR we might expect that our estimates would be increasingly accurate over the low-SNR case, and that the method converges robustly to similar posteriors.
-Whilst this is generally true for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ and the posteriors across noise realisations are largely indistinguishable, for $\iota$ and $h_0$ this is not the case.
+In the case of high-SNR we might expect \textit{a priori} that our estimates would be increasingly accurate over the low-SNR case, and that the method would converge robustly to similar posteriors irrespective of the noise realisation (i.e.
+small $W_1$).
+This is generally true for the majority of the component parameters of $\boldsymbol{\theta}_{\rm gw}$; the estimated posteriors across noise realisations are largely indistinguishable.
+However for $\iota$ and $h_0$ this is not the case.
 Instead the inferred one-dimensional posteriors do not broadly overlap are not consistent across different noise realisations.
-Specifically the median WD across $10^3$ noise realisations is $W_1 = 0.24$ rad and $\iota$ and $W_1 = 1.6 \times 10^{-13}$ for $h_0$.
-For $\iota$ this WD is greater than the value inferred in the low SNR case (0.14, Table \ref{tab:parameters_and_priors}, whilst the WD value for $h_0$ is also greater as a proportion of the scale set by the strain amplitude.
-This is in contrast to the median WD inferred for the other parameters; e.g.
-for$\Phi_0$, $W_1 = 0.0045$ rad.
+Specifically across $10^3$ noise realisations the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
+For $\iota$ this $W_{1,\rm median}$ is greater than the value inferred in the low SNR case (0.14, Table \ref{tab:Wasserstein}, whilst for $h_0$ $W_{1,\rm median}$ is also greater as a proportion of the scale set by the strain amplitude.
+This is in contrast to $W_{1,\rm median}$ inferred for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ in the high-SNR regime; e.g.
+for $\Phi_0$, $W_1 = 4.5 \times 10^{-3}$ rad.
 \newline
@@
 Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
-If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably.
-However, the resulting maximum a posteriori probability estimate for $\iota$ is strongly biased away from the true value by $\sim 0.35$ rad.
-This bias will be discussed further in Section \ref{sec:bias}.
-In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" the likelihood ridge, leading to a more concave likelihood surface.
+If during the inference we fix one of the parameters at its true value (i.e.
+set a delta function prior), taking a slice through the likelihood surface it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably.
+However if we fix e.g $h_0$ at its true value then the one dimensional posteriors for $\iota$ are all strongly biased away from the injected value; the maximum a posteriori probability estimates underestimate the injected value by $\sim 0.34$ rad.
+This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be discussed further in Section \ref{sec:bias}.
+In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" \footnote{\tiny \textcolor{red}{TK: "washes out" may be too colloquial / unspecific, but I think it is intuitive...}\normalsize}the likelihood ridge, leading to a more concave likelihood surface.
 This enables consistent posteriors to be estimated for $\iota$ and $h$.
@@
 	\caption{As Figure \ref{fig:corner_plot_1} but for the low-SNR system with $h_0 = 10^{-12}$.
-For the majority of parameters the injection value does not fall within the 90 \% credible interval and is not visible over these scales.
-The estimated posteriors are biased away from the injected value due to the dropping of the pulsar terms.}
+For the majority of parameters the injection value does not fall within the 90\% credible interval and is not visible over these scales.
+The estimated posteriors are biased away from the injected value due to the dropping of the pulsar terms.
+Note that the scale of the $x$-axes are generally highly constrained compared to Figure \ref{fig:corner_plot_1}.}
 	\label{fig:bias_for_large_h}
@@
 In Figure \ref{fig:bias_for_large_h} we show the resulting posteriors for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but now for the high-SNR system with $h_0 = 10^{-12}$.
-We can see that for all parameters, with the exception of $\Omega$ and $h_0$, the injected value lies outside 90 \% credible interval, to the extent that it is not visible on the scales over which we have plotted the posteriors.
+We can see that for all parameters, with the exception of $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it is not visible on the scales over which we have plotted the posteriors.
 This effect is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
@@
 If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate and the bias may not be important for searches.
-
 This is what we observe for our synthetic data (e.g.
 Figure \ref{fig:corner_plot_2}); the decreased SNR manifests as a broader marginalised posterior which generally overwhelms the shift due to the bias.
-
-\textcolor{red}{However, PP plot shows...}
-
-Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may come to systematically affect the inferred parameters. \newline
+Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may come to systematically affect the inferred parameters.
+\newline
 
@@
 The method is then further tested over 1000 noise realisations.
-Consistent MAP estimates of the parameters across all noise realisations are obtained for the majority of parameters, whilst the structural weak-identifiability of $\iota$ and $h_0$ results in a larger variance in the MAP estimates between individual noise realisations for these parameters.
+Consistent posterior estimates of the parameters are obtained for the majority of noise realisations, with the median Wasserstein distance across all noise realisations $W_{1, \rm median}$ generally small compared to the prior space ($\lesssim 3 \%$).
+For a minority of noise realisations the nested sampler converges poorly and settles into a local likelihood optima for the posteriors estimates of $\psi$ and $\alpha$.
+This problem can be solved in general by increasing the number of live points $n_{\rm live}$ used by the sampler.
 Exploration of a broader parameter space via 200 randomly sampled parameter vectors highlights biases in some parameters as a result of dropping the pulsar terms from the inference model.
-The cause of these biases is discussed.
+The cause of these biases is examined for the case of high SNR, along with the structural weak-identifiability between $\iota$ and $h_0$.
 The extension of this method to include the pulsar term is a key challenge for a future work.
@@
 \section{Identifiability}\label{appendix_identifiability}
-
-\textcolor{red}{TK TBD}
-
-This paper is useful: \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
+\textcolor{red}{TK TBD. An example calculation for a simple system using the methods of e.g. } \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
 
 \section{Wasserstein distance}\label{sec:wasserstein}
-
-It is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
+The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
+It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$.
+For this reason it is sometimes known as the "Earth mover's distance".
+\newline
+
+The $p$-th order Wasserstein distance between two distributions is
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p} \, 
+\end{eqnarray}
+where $p \in [1,\infty)$ and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$ (i.e.
+the set of couplings of $\mu$ and $\nu$).
+\newline
+
+For finite, empirical, discrete samples drawn from underlying probability distributions defined on $\mathbb{R}^d$ it is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
 \begin{eqnarray}
@@
 \end{eqnarray}
+In order to circumvent this issue in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors
+
+For empirical distributins the
+
+Pracical definitin for emiral distributions.
+
+First order and units of WD
+
+Curse of dimenionality for emipiral distributions
+
+Advantages
+
+symmetric
+
+triangle inequality
+
+non parameteric
+
+intuitive
+
+takes account of geometry
+
+comparison with KL
+
 To sidestep this issue, we marginalize onto the distributions spanned by the individual gridpoints and take the WD between these 1-dimensional distributions.
@@
 The WD intuitively defines a distance between two distributions as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$.
-The $p$-Wasserstein distance between two distributions $\mu, \nu$ is
-\begin{eqnarray}
-	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p}
-\end{eqnarray}
 where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function.

### 2023-09-02 notebooks/plot_likelihood_surfaces_2D_alpha_psi.py

@@ -1,0 +1,94 @@
+
+
+import numpy as np
+import matplotlib.pyplot as plt  
+
+
+
+
+from matplotlib import ticker, cm
+import matplotlib.pyplot as plt 
+
+import scienceplots
+
+
+
+
+import matplotlib
+
+plt.style.use('science')
+fig = plt.figure(figsize=(6,6))
+ax = fig.add_subplot(111)
+
+
+
+path_to_earth_model = 'likelihood_surface_alpha_psi_mm_earth_coarse.npz'
+
+
+def load_and_plot(path,ax):
+
+
+    #Load the data
+    data = np.load(path)
+    alpha_values = data['x']
+    psi_values = data['y']
+    surface_pulsar = data['z'] 
+    surface_pulsar = surface_pulsar / np.abs(np.max(surface_pulsar)) #Normalize
+
+    #Extract location of maxima
+    alpha_idx, psi_idx = np.unravel_index(surface_pulsar.argmax(), surface_pulsar.shape)
+    xc = alpha_values[alpha_idx]
+    yc = psi_values[psi_idx]
+    zc = surface_pulsar[alpha_idx,psi_idx]
+
+    print('Inferred maxima = ', xc,yc,zc)
+
+
+    #Cast to 2D mesh
+    X,Y = np.meshgrid(psi_values,alpha_values)
+    lx = len(alpha_values)
+    ly = len(psi_values)
+    z = np.reshape(surface_pulsar, (lx, ly))
+
+
+
+    #Plot colormap
+    #CS = ax.pcolormesh(X, Y, z,clim=(2.0*np.max(surface_pulsar), np.max(surface_pulsar)),shading='gouraud',cmap='viridis')
+    CS = ax.pcolormesh(X, Y, z,shading='gouraud',cmap='viridis')
+    eps=2e-5
+    #CS = ax.pcolormesh(X, Y, z,clim=(1.0-eps, 1.0),shading='gouraud',cmap='viridis')
+
+
+    clb = plt.colorbar(CS)
+
+
+
+    #Config
+    ax.scatter(1.00,2.50, s=20,c='C3') #Location of true maxima
+    ax.scatter(2.70,1.01, s=20,c='C4') #Location of settled maxima
+    #ax.set_xscale('log')
+    #ax.scatter(yc,xc, s=20,c='C3')
+    #ax.scatter(1.01,2.70, s=20,c='C4')
+
+
+    
+
+
[…]

### 2023-09-02 py_src/config_batch.py

@@ -39,8 +39,8 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
        
 
 h_range = np.logspace(-15,-13,101)
-#noise_models = ["earth", "null"]
-noise_models = ["earth"]
+noise_models = ["earth", "null"]
+#noise_models = ["earth"]
 
 seeds = np.arange(1245,1245+10)
 
@@ -49,7 +49,7 @@ with open('batch.sh','w') as b:
     for s in seeds:
         for h in h_range:
             for n in noise_models:
-                 arg_name = f"NMULTIpaper_bayes_ratios_n2000_V4_h_{h}_model_{n}_seed_{s}"
+                 arg_name = f"JMULTIpaper_bayes_ratios_n2000_V4_h_{h}_model_{n}_seed_{s}"
                  print(arg_name)
                  create_slurm_job(arg_name,h,n,s)
 

### 2023-09-14 docs/manuscript/paper.tex
response to AM comments

@@
 
+\usepackage{enumitem}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % If you need two or more lines of authors, add an extra line using \newauthor
-\author[Kimpson]{Kimpson$^{1}$, Melatos$^{2}$, O'Leary, Evans$^{3}$, Carlin others, etc. etc.%
-\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
+\author[Kimpson]{Tom Kimpson$^{1,2}$\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}, Andrew Melatos$^{1,2}$, Joseph O'Leary$^{1,2}$, Julian B.
+Carlin$^{1,2}$, Robin J.
+Evans$^{3}$, \newauthor William Moran$^{3}$, Tong Cheunchitra$^{1,2}$, Wenhao Dong$^{1,2}$, Liam Dunn$^{1,2}$, Julian Greentree$^{3}$, Nicholas J.
+O'Neill$^{1,2}$, \newauthor Sofia Suvorova$^{3}$, Kok Hong Thong$^{1,2}$, Andrés F.
+Vargas$^{1,2}$%
 %\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
@@
 \begin{abstract}	
-Continuous nHz gravitational waves from individual supermassive black hole binaries may be detectable with Pulsar Timing Array observations.
-A novel search strategy is developed wherein intrinsic pulsar timing variations are tracked simultaneously with the modulation induced by a putative gravitational wave.
-A two-step procedure is applied within a state-space framework, where the model parameters are estimated with a Kalman filter, which provides a likelihood for nested sampling.
-This enables self-consistent inference of model parameters and calculation of the Bayesian evidence.
-It is shown via an astrophysically representative software injection campaign that the procedure can distinguish a gravitational wave from pure noise down to a strain of $h_0 \sim 4 \times 10^{-15}$, given an inclination angle $\iota = 1.0$ rad.
+Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing array observations.
+A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
+A two-step procedure is applied within a state-space framework, such that the model parameters are estimated with a Kalman filter, which then provides a likelihood for nested sampling.
+The procedure enables self-consistent inference of model parameters and calculation of the Bayesian evidence, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional analysis methods.
+It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$.
 Full posterior distributions of model parameters are recovered.
-There is a bias in the estimates of some parameters, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
+There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
+Smaller biases $\lesssim 1 \%$ are also observed in some of the other parameters of the system.
 \end{abstract}
@@
 Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
-The detection of nHz GWs has necessitated the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
-The foremost method is timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}.
+The detection of nHz GWs has inspired the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
+The foremost method is timing an ensemble of pulsars, i.e.
+a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}.
 A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth.
@@
 
-Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
+Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the early stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
 Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}.
-Individual SMBHBs are continuous wave sources: they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals.
+Individual SMBHBs are continuous wave sources; they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals.
 Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs.
-However, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for a individual source at 4-5 nHz \citep{2023arXiv230616226A}.
+However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}.
 \newline
@@
 This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
-In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are relatively stable rotators.
+In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators.
 However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
 In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red).
-In addition to the red timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
 \newline
@@
 Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
-The new and existing approach are therefore complementary, but the new approach holds out the promise of somewhat higher sensitivity.
+The new and existing approaches are therefore complementary.
 In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
-We combine the Kalman tracking of the pulsars intrinsic rotational state with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood for model selection.
+We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood (i.e.
+the model evidence) for model selection.
 \newline
@@
 \noindent This paper is organised as follows.
-In Section \ref{sec:model} we present the state-space model for the pulsar frequency which is subject to the influence of a GW.
-In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence.
-In Section \ref{sec:testing} we test this new method on synthetic pulsar data.
-Throughout this work we adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature.
-\newline
-
-%
-%The detection of high frequency ($\sim 100$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
-GWs from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
-
-%
-% Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), 
-%
+In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW.
+In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence.
+In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method.
+We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative example GW system.
+In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space.
+Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.
+The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}.
+It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization which is deferred to future work.
+Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$.
+\newline
 
@@
 We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation.
-In this work we take the intrinsic state variable to be the pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local rest frame of the pulsar's centre of mass.
-A model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.
+In this work we take the intrinsic state variable to be the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local,freely-falling rest frame of the pulsar's centre of mass.
+A phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.
 We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.
@@
 The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
+The subtle problem of generalizing the measurement variable to pulse TOAs is postponed to future work, as noted in Section \ref{sec:intro}.
 \subsection{Spin evolution} \label{sec:psr_frequency}
@@
 In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque.
-The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$.
+The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
 Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution $f_{\rm em}^{(n)}(t)$ can be approximated accurately by
@@
 \begin{equation}
-	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)} +\xi^{(n)}(t) \ . 
+	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
 	\label{eq:frequency_evolution}
 \end{equation}
-In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm em}$ is the spin derivative, $\gamma^{(n)}$ a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
+In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
 \begin{align}
 	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle &= [\sigma^{(n)}]^2 \delta(t - t') \ .	\label{eq:xieqn}
+	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
 \end{align}
-$[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
-It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
+In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise.
+Combined with the mean reversion it gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+It is important to note that white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
 \newline
@@
 Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
-As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G}.
+As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}.
 An approach akin to Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} has been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
@@
 For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless.
-Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004} which have not been ruled out by pulsar timing experiments to date.
+Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, which have not been ruled out by pulsar timing experiments to date.
 The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \eqref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \eqref{eq:frequency_evolution}.
@@
 \subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
-In the presence of a GW, the pulsar frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
-Specifically, the pulsar frequency is modulated harmonically at the GW frequency.
+In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
+Specifically, the pulse frequency is modulated harmonically at the GW frequency.
 We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section.
@@
 \subsubsection{Plane GW perturbation}\label{sec:plane_gw}
-We consider a gravitational plane wave from a single,distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
+We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
 \begin{equation}
@@
 In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic.
-Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves over decadal timescales as \citep[e.g.][]{Zhu10},
-\begin{equation}
-	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left(\frac{f_{\rm gw}(t=0)}{10^{-7} \mathrm{~Hz}}\right)^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
+Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Zhu10},
+\begin{equation}
+	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
 	\label{eq:f_evolution}
 \end{equation}
-where $M_c$ is the chirp mass of the SMBHB, $f_{gw}(t=0)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ the length of the data timespan, which for PTAs is $\sim 10$ years.
-A source can be considered as monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution of $1/T_{\rm obs}$.
-From Equation \eqref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans.
-The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA with pulsars at 1.5 kpc, 78 \% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52 \%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
-We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
+where $M_c$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
+A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
+From Equation \eqref{eq:f_evolution} we can see that only those binaries which are massive or at high frequency experience significant frequency evolution over typical PTA timespans.
+The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
+We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
 \newline
@@
 \end{align}
-The principal axes are in turn specified by the location of the GW source on the sky (via colatitude $\theta$ and azimuth $\phi$) and the polarisation angle $\psi$ according to
+The principal axes are in turn specified by the location of the GW source on the sky (colatitude $\theta$, azimuth $\phi$) and the polarisation angle $\psi$ according to
 \begin{align}
@@
 In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
-They are described by the geometric object $\vec{p}$ which we can identify  as the 4-momentum of the radio photon.
-The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e.
-$\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
-\begin{equation}
- \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
+They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio photon.
+The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e.
+$\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $.
+One obtains \citep[e.g.][]{Maggiore}
+\begin{equation}
+ \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
 	\label{eq:momentum_shift}
 \end{equation}
-where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
+where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer.
+We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
 We take the pulsar location to be constant i.e.
-neither $\boldsymbol{q}$ nor $d$ is a function of time.
+neither $\boldsymbol{q}$ nor $d$ are functions of time.
 In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB.
-This barycentering correction is typically applied when generating TOAs using e.g.
-{\sc tempo2} \citep{tempo2} and related timing software.
-Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression:
-\begin{equation}
-	f = p_{\alpha} u^{\alpha} \ . 
-	\label{eq:freq_temporal}
-\end{equation}
-Due to the kinematical corrections from the barycentering process, 
-\begin{equation}
-	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0) \ ,
-\end{equation}
-where the perturbations to $\boldsymbol{u}$ from the GW are of higher order and neglected.
-In this case the shift in the momentum from Equation \eqref{eq:momentum_shift} can be related to a shift in the frequency as,
-\begin{equation}
-	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t) \ ,
+The barycentering correction is typically applied when generating TOAs, e.g.
+with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
+Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$.
+After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames,
+\begin{equation}
+	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
+\end{equation}
+Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
+\begin{equation}
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d^{(n)}) g^{(n)}(t) +  \varepsilon^{(n)}\ .
 	\label{eq:measurement}
 \end{equation}
-where
-\begin{equation}
-	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i,(n)} q^{j,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right) \ .
+where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies: 
+\begin{align}
+	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
+	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'}(t - t') \ .	\label{eq:vareps}
+\end{align}
+where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
+\begin{equation}
+	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
 	\label{eq:g_func}
 \end{equation}
-and $q^{i,(n)}$ labels the $i$-th coordinate component of the of the $n$-th pulsars position vector $\boldsymbol{q}^{(n)}$.
+and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$.
 It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
 \begin{align}
-	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^{i,(n)} q^{j,(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
-	& \times \Big\{ \cos\left[-\Omega t +\Phi_0\right] \nonumber \\
-	&- \cos \left[-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right] \Big\} \ .
+	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
+	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
+	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
 	\label{eq:g_func_trig}
 \end{align}
-Equations \eqref{eq:measurement} -- \eqref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the frequency measured by an observer on Earth.
+Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.
 
 \section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
-The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar, and parameters of the GW source, i.e.
+The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters controlling the intrinsic frequency evolution of the pulsars in the array and the GW source, i.e.
 \begin{equation}
@@
 \begin{equation}
-	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
+	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
 \end{equation}
@@
 \begin{equation}
-	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
-\end{equation}
-where $\delta$ is the declination of the GW source and $\alpha$ the right ascension.
-In Equation \eqref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
+	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
+\end{equation}
+where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively \footnote{$\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
+$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.}.
+In Equation \eqref{eq:params3} we reparameterize the two GW polarisation amplitudes, $h_{+}$ and $h_{\times}$, in terms of $\iota$ and the characteristic wave strain $h_0$ through
 \begin{align}
 	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
-	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ , 
+	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ .
 \end{align}
-where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e.
-$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.
-We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
-For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate.
-Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.
+We use the parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
+A PTA containing $N$ pulsars comprises $7 + 5N$ parameters to estimate.
+Typically the pulsar parameters are constrained better \textit{a priori} by electromagnetic observations than the GW parameters.
+For example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information about $\delta$ and $\alpha$.
 \newline
@@
 In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter.
-In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference of the model parameters and calculate the model evidence.
-Model selection and the specification of the null model is described in Section \ref{sec:model_selection}.
-A complete summary of the method is presented in Section \ref{sec:methodsummary}.
-
-\subsection{Kalman filter}\label{sec:kalman_filter}
-
-%
+In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference.
+Model selection and the specification of the null model are described in Section \ref{sec:model_selection}.
+A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
+The method complements traditional PTA analyses because it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations.
+
+\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
 %\begin{figure*}
-%	\centering
-%	
-%	\subfloat[a]{\includegraphics[width=0.5\textwidth]{images/Kalman_example_both}} 
-%	\subfloat[b]{\includegraphics[width=0.3\textwidth]{images/Kalman_example_both}}
-%	\caption{\textcolor{red}{TK: For consideration. Do we prefer this figure to Figure \ref{fig:four figures}}?}
-%	\label{fig:kalmanexampleboth}
+%	%\centering % Not needed
+%	\begin{subfigure}[b]{0.49\textwidth}
+%		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
+%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
+%		\label{fig:6MB_BFS}
+%	\end{subfigure}  
+%    \hfill
+%	\begin{subfigure}[b]{0.49\textwidth}
+%		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
+%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
+%		\label{fig:25MB_bfs}
+%	\end{subfigure}
+%	\caption{
+%		}
+%	\label{fig:four figures}
 %\end{figure*}
-%
 
 \begin{figure*}
-	%\centering % Not needed
-	\begin{subfigure}[b]{0.49\textwidth}
-		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
-		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
-		\label{fig:6MB_BFS}
-	\end{subfigure}  
-    \hfill
-	\begin{subfigure}[b]{0.49\textwidth}
-		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
-		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
-		\label{fig:25MB_bfs}
-	\end{subfigure}
-	\caption{Example application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$.
-We plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency.
-In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ and the state is estimated accurately.
-In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$ and the state is not estimated accurately.
-The top panels show the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line).
-The middle panels show the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta).
-The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$.
-In (a) the green/blue and red/magenta lines are completely overlaid.}
-	\label{fig:four figures}
+	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
+	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed systems state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are identically correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column).
+The top panels show the true pulsar state $f_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$  (green curve).
+We plot the subtracted state to better illustrate the stochastic wandering of the pulsar frequency.
+In the left hand column, the blue/green solutions are completely overlaid.
+The middle panels show the true measured frequency $f_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve).
+Again we plot the subtracted state, and in the left hand column the red/magenta solutions are completely overlaid.
+The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{1}(t) - \hat{f}_{\rm m}^{1}(t)$.
+The results are shown for a single pulsar.
+In the right hand column $\Omega$ has been perturbed from its true value by 20\%}
+	\label{fig:kalman_example}
 \end{figure*}
 
-The Kalman filter \citep{Kalman1} is a Gaussian-Markov model used to algorithmically recover the most likely set of unknown system state variables, $\boldsymbol{X}$, given some noisy measurements, $\boldsymbol{Y}$.
-It is a common technique in signal processing that has been applied successfully in neutron star astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}.
-In this work we use the linear Kalman filter which assumes a linear relation between $\boldsymbol{X}$ and $\boldsymbol{Y}$.
-Whilst the measurement equation, Equation \eqref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively.
-Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, the unscented Kalman filter \citep{882463van} or the particle filter \citep{Simon10}.
-\newline
-
-The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}.
-At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$.
-The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Equation \eqref{eq:measurement}.
-The measurement predictions can then be compared against the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation".
-The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$.
-The Gaussian log-likelihood can then be calculated at each timestep:
+The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$.
+It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
+In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
+Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
+Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
+Hence inferring the static parameters is a non-linear exercise, to be tackled separately after the linear Kalman filter operates on the data for a fixed set of static parameters.
+Inference of the static parameters in Equations \eqref{eq:measurement} and \eqref{eq:g_func} by nested sampling is discussed in Section \ref{sec:nested_sampling}.
+\newline
+
+Implementation of the linear Kalman filter for the PTA state-space model in Section 3.2, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
+At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.
+The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by converting ${\boldsymbol{\hat X}}_i$ into predicted measurements ${\boldsymbol{\hat Y}}_i$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func} and comparing with the actual, noisy measurements ${\boldsymbol{\hat Y}}_i$.
+This defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation.
+The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$.
+The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}.
+For a fixed set of static parameters, the Kalman filter returns an estimate of the the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error.
+We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline
+
+The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ can  then be calculated at each timestep from the Kalman filter output according to
 \begin{eqnarray}
-	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ ,
+	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ .
 \end{eqnarray}
-with the total log-likelihood simply the sum over all timesteps, i.e. 
+The total log-likelihood for the entire sequence is
 \begin{eqnarray}
-	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i \ . \label{eq:likelihood}
+	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
 \end{eqnarray}
-For fixed data $\boldsymbol{Y}$, $\mathcal{L}$ is a function of the static parameters of the model, i.e.
-$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})$.
-Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are also functions of $\boldsymbol{\theta}$.
-If the estimates of the system parameters, $\boldsymbol{\hat{\theta}}$ , that we pass to the Kalman filter are close to the true underlying parameters then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are	 minimized and $\mathcal{L}$ is maximised.
-This is illustrated in Figure \ref{fig:6MB_BFS}; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
+Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
+$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$.
+Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$.
+If $\boldsymbol{\hat{\theta}}$ is close to the true underlying parameters $\boldsymbol{\theta}$, then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are minimized and $\mathcal{L}$ is maximised.
+This is illustrated with synthetic data in Figure \ref{fig:kalman_example}.
+In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations (1)--(4), (16), and (17) for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$.
+The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve).
+The predicted state $\hat{f}_{\rm p}^{(1)}(t)$ is converted  into a predicted measurement $\hat{f}_{\rm m}^{(1)}(t)$ (middle panel, magenta curve), which again overlaps almost perfectly with the true measurement.
+The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel).
+By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters to the Kalman filter, where $\Omega$ has been perturbed from its true value by $20 \%$.
+In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates.
+It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel).
+
+%n Figure ; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
 The residuals correspond to random noise and are normally distributed.
@@
 
-\subsection{Nested Sampling}\label{sec:nested_sampling}
+\subsection{Nested sampling}\label{sec:nested_sampling}
 We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
@@
 \end{equation}
-In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work.
-Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
-It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
-The primary advantage of nested sampling is the ability to compute the evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches.
-Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}.
-For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
+ We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper.
+Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, which also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
+The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies.
+Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}.
+For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
 For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}.
@@
 \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}.
-For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library.
+In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library.
 We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work.
-The primary tunable parameter in nested sampling is $n_{\rm live}$, where a greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}(1/\sqrt{n_{\rm live}})$.
-However the computational runtime scales as $\mathcal{O}(n_{\rm live})$ and so one must make a trade-off between uncertainty and runtime.
-\cite{Ashton2022} offer a rule-of-thumb where the minimum number of live points should be greater than the number of static parameters.
-The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$.
+\newline
+
+The primary tunable parameter in nested sampling is $n_{\rm live}$.
+A greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
+However the computational runtime scales as $\mathcal{O}(n_{\rm live})$.
+\cite{Ashton2022} offered a rule-of-thumb trade-off, where the minimum number of live points should be greater than the number of static parameters.
+Informal empirical tests conducted as part of this paper support the trade-off suggested by \cite{Ashton2022}; we find typically that the true ${\boldsymbol{\theta}}$ is contained within the 90\% credible interval of the derived 1D marginalised posteriors of ${\boldsymbol{\hat{\theta}}}$ for $n_{\rm live} > 7 + 5N$ with $N \leq 50$.
 Unless stated otherwise we take $n_{\rm live} = 1000$ for all results presented in this work.
@@
 \subsection{Model selection}\label{sec:model_selection}
-The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}_i$.
-This enables us to compare competing models via a Bayes factor,
+The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
 \begin{equation}
@@
 \end{equation}
-Throughout this work we take $\mathcal{M}_1$ to be the state space model that includes the presence of a GW $\mathcal{M}_0$ is our null model that assumes there is no GW in the data.
+Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
+$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
 This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
@@
 \subsection{Summary of workflow}\label{sec:methodsummary}
-To summarise the above, the order of operations in our method for a single iteration (i.e. for a single live point) is as follows:
-\begin{enumerate}
-	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\boldsymbol{\theta}_{\rm sample}$
-    \item Pass $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
-	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
-	\item Return $\log \mathcal{L}$ to the nested sampler
-    \item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$
+For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
+\begin{enumerate}[leftmargin=2em]
+	\item Specify a PTA composed of $N$ pulsars 
+	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
+	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$. 
+	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
+	 \item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
+	 \item For each live point:
+\begin{enumerate}[leftmargin=2em]
+	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
+	\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
 \end{enumerate}
-In Section \ref{sec:testing} we apply this workflow on a synthetic data set.
-
-\section{Tests with synthetic data} \label{sec:testing}
-In this section we use synthetic data to test the ability of our method to detect the presence of a GW in noisy data and recover the static parameters, $\boldsymbol{\theta}$.
-In Section \ref{sec:synth_data} we discuss how the synthetic data is generated, including the construction of an artificial PTA.
-In Section \ref{sec:rep_example} we generate synthetic data for a single representative example SMBHB system and apply our method to estimate $\boldsymbol{\theta}$ and calculate the statistical evidence for a GW in the data via the Bayes factor.
-In this section we also explore how the detectability of the system changes for different strain magnitudes, $h_0$ and the performance of the method for different realisations of the pulsar process noise and the measurement noise.
-In Section \ref{sec:parameter_space} we move beyond a single example system and test our method across a broader parameter space.
-In Section \ref{sec:bias} we discuss biases that are evident in some of the parameter estimates.
-
-\subsection{Synthetic data generation} \label{sec:synth_data}
-
-\subsubsection{Constructing a synthetic PTA}\label{sec:synt_pta}
-
+	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
+	\item Constrained sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood $\log \mathcal{L}_{\rm new}$ > $\log \mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (a)--(b).
+	\item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ 
+	\item Repeat steps (vii) - (ix) until convergence criteria are satisfied.
+\end{enumerate}
+In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$.
+The resulting $\mathcal{Z}$ values can then be compared.
+We remind the reader that the above workflow differs from a realistic PTA analysis in on important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
+The generalization to TOAs is subtle and will be tackled in a forthcoming paper.
+
+\section{Validation procedure with synthetic data} \label{sec:testing}
+In this section we outline how synthetic data is generated in order to validate the analysis scheme in Section \ref{sec:detect}.
+In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
+In Section \ref{sec:gendata} we demonstrate how to solve the equations of Section \ref{sec:model} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.
+
+\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
 \begin{figure}
@@
 \end{figure}
-For this work we generate synthetic data that is astrophysically representative as follows.
-We first specify a set of pulsars to make up our artificial PTA.
-We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}.
-The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA and the method naturally extends to any other PTA.
-For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{(n)}$), the distance, the barycentric rotation frequency (which we identify with $f_{\rm em}^{(n)} (t_1)$), and the time derivative of the the barycentric rotation frequency (which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$).
-The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Figure \ref{fig:pulsar_distrib}.
-The data is acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
-The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$.
-We discussed in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$.
-Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars.
-For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
-In order to set a physically reasonable value for $[\sigma^{(n)}]^2$ we use two complementary approaches.
-The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by:
+We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
+The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
+For each pulsar we adopt the fiducial NANOGrav values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
+A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility.
+The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}.
+The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist.
+The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
+In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
+We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$.
+The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
 \begin{align}
-	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)&= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
-	&+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
+	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
+	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
 \end{align}
-where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$.
-Throughout this work we assume that all pulsars are observed with a weekly cadence.
-In order to relate the empirical TOA noise model, Equation \ref{eq:sigmap}, to our Ornstein-Uhlenbeck frequency model, Equation \ref{eq:spinevol}, we set first $T = 1$ week in Equation \ref{eq:sigmap} to obtain an estimate of the TOA noise over a weekly time interval, i.e.
-the size of our numerical integration timestep.
-The frequency process noise parameter $\sigma^{(n)}$ can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as
+where $T_{\rm cad}$ is the cadence of the timing observations and MSPs have $\ln \alpha_1 \approx -20 $, $\alpha_2 \approx 1$, $\alpha_3 \approx 2$ and $\alpha_4 \approx 2.4$.
+Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
+In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
 \begin{eqnarray}
-	\sigma^{(n)} \sim \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
+	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-1} \ . \label{eq:sigmap_f}
 \end{eqnarray}
-For our synthetic NANOGrav PTA, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s. \newline
-
-As a sanity check, we can also calculate $\sigma^{(n)}$ using a complementary numerical approach to the empirical best-fit model outlined above.
-We can directly solve the state equation Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a synthetic phase solution,
+For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s$^{-5/2}$, $\min [ \sigma^{(n)} ] = 1.3 \times 10^{-23}$s$^{-5/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.0 \times 10^{-16}$ s$^{-5/2}$ for PSR J1939+2134 \newline
+
+As a cross-check, we calculate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
 \begin{eqnarray}
-	\xi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
+	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
 \end{eqnarray}
-and generate TOAs and phase residuals.
-We calibrate the noise amplitude $\sigma^{(n)}$  to generate phase residuals that qualitatively resemble empirical phase residuals measured from real pulsars.
-We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} from the  Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}.
-For the purposes of validating the $\sigma^{(n)}$ inferred from Equation \eqref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually.
-Whilst it would also be possible to e.g.
-define some quantitative loss function (e.g.
-mean-squared error) between the two solutions and solve the optimisation problem to infer $\sigma^{(n)}$, for our purposes this is excessive.
-It is satisfactory to simply qualitatively compare the two solutions as a useful sanity check.
-We emphasise that we are not overly concerned with calculating the maximally accurate values for $\sigma^{(n)}$, but instead some reasonable values for constructing representative synthetic data.
-This approach of generating synthetic phase solutions and visually comparing with empirical solutions is also followed in \cite{Vargas}.
-In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for the NANOGrav pulsar J0030+0451.
-We can see that the synthetic and empirical residuals are qualitatively similar.
+and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach.
+We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} conducted with the Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}.
+A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$.
+In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$.
+In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J0030+0451, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}.
+We see that the synthetic and empirical residuals are qualitatively similar.
+Comparable behaviour is observed for other pulsars in our array.
 \begin{figure}
 	\includegraphics[width=\columnwidth]{images/example_residuals_plot}
-	\caption{Actual (left panels) and synthetic (right panels) phase residuals for NANOGrav millisecond pulsar J0030+0451.
-The actual residuals are obtained via UTMOST pulsar timing program \citep{UTMOST}.
+	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J0030+0451.
+The actual residuals are sourced from the UTMOST pulsar timing program \citep{UTMOST}.
 The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
-For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
-}
+The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}.
+Following \citet{Vargas} we set the TOA uncertainty to be a constant $10^{-4}$ s.
+The blue and orange residuals are qualitatively similar by inspection.
+The other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} are tuned to produce comparable results.}
 	\label{fig:qualitative_compare}
 \end{figure}
-\subsubsection{Synthetic data generation}
-We generate $N$ synthetic noisy timeseries, one for each pulsar, as follows:
-
-\begin{enumerate}
-	\item Integrate the state equations, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
-	\item Map from state space to measurement space via the measurement equation, Equation \eqref{eq:measurement}
-	\item Add zero-mean Gaussian measurement noise to each timeseries, i.e.
-$f_{\rm m} ^{(n)} + \varepsilon^{(n)}$ where 	$\langle \varepsilon(t) \varepsilon(t') \rangle = \sigma_{\rm m}^2 \delta(t - t')$ for measurement noise covariance $\sigma_{\rm m}^2$.
+\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
+We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
+\begin{enumerate}[leftmargin=2em]
+	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
+	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}
 \end{enumerate}
 Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
-The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified by our construction of the synthetic PTA outlined in Section \eqref{sec:synt_pta}.
+The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \eqref{sec:synt_pta}.
 For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
-The measurement noise can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
-\begin{equation}
-	\sigma_{\rm m} \sim f_{\rm p}^{(n)} \frac{\sigma_{\rm TOA}}{\text{T}} \ . 
-\end{equation}
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
-Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it to be known \textit{a priori} rather than a parameter to be inferred.
-Note that $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar; whilst for our construction every pulsar is sampled at the same time and so would experience the same realisation of the noise, in actual pulsar astronomy different pulsars will be sampled at different times by different radio interferometers and so subject to different noise realisations.
-\newline
-
-In practice, in order to avoid numerical precision issues that arise when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
-\begin{equation}
-	f_{\rm p}^{\prime (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ .
-\end{equation}  
-We similarly heterodyne the measurement variable 
-\begin{equation}
-	f_{\rm m}^{\prime (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{*(n)} \ ,
-\end{equation}
-where $ f_{\rm em}^{*(n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris.
-For synthetic data we can set $ f_{\rm em}^{*(n)} = f_{\rm em}^{(n)}$ but this is not true generally for real-world observations since the spin-down ephemeris is only known approximately.
-The measurement equation Equation \eqref{eq:measurement} that relates the state and measurement variables is then updated as
-\begin{equation}
-	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
+The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
+\begin{equation}
+	\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
+\end{equation}
+Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$.
+The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
+Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
+Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar.
+\newline
+
+Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } \sigma^{(n)} \Delta W_t \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta W_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
+To fix the problem, we subtract the deterministic frequency evolution and track the new variable, equivalent to a change of variables:
+\begin{equation}
+	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ .
+\end{equation}  
+We similarly modify the measurement variable to be
+\begin{equation}
+	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ ,
+\end{equation}
+where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris which is measured to high accuracy in practice by {\sc tempo2}.
+For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately.
+Equation \eqref{eq:measurement} is then updated to read
+\begin{equation}
+	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
 	\label{eq:measurement_cov}
 \end{equation}
-We emphasise that this change of variables is simply a convenient scaling to bring our numerical values into a reasonable dynamic range without having to use extended numerical floating point formats (e.g.
+We emphasise that this change of variables is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g.
 long double, quadruple).
-It does not remove any degrees of freedom of the formulation.
-In particular both $f_{\rm em}^{(n)}(t_1)$
-and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain static parameters of the model, but now appear in the measurement equation rather than the state equation.
-
-\subsection{Representative example}\label{sec:rep_example}
-We initially characterise our method with a single example of a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
+It does not remove any degrees of freedom nor does it involve an approximation.
+In particular $f_{\rm em}^{(n)}(t_1)$
+and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \ref{eq:spinevol}--\eqref{eq:xieqn}.
+
+\section{Representative example: single SMBHB source}\label{sec:rep_example}
+In this Section we apply the analysis scheme of Section \ref{sec:detect} and the validation procedure of Section \ref{sec:testing} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
+We remind the reader that the analysis in this paper targets a single SMBHB source.
+The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
+In Section \ref{sec:priors} we outline the priors chosen on $\boldsymbol{\theta}$.
+In Section \ref{sec:parameter_estim} we apply the workflow of Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
+We go on in Section \ref{sec:multiple_noise} to extend this parameter estimation procedure to multiple realisations of the noise processes.
+In Section \ref{sec:detection} we calculate the the detectability of the source as a function of $h_0$.
+\newline
+
 The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected to be astrophysically reasonable and representative.
@@
 
-\subsubsection{Prior distributions}\label{sec:priors}
+\subsection{Prior distributions}\label{sec:priors}
 For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters.
@@
 Instead we consider the pulsar parameters similarly to the GW source parameters and estimate them optimally within a consistent framework.
-Adopting looser priors in this way allows us to subject our method to a more stringent test and confirm that we still accurately estimate GW parameters without highly accurate initial estimates of $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$.
+Adopting looser priors in this way allows us to subject our method to a more stringent test and confirm that we still accurately estimate GW parameters without highly accurate and precise initial estimates of $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$.
 \newline
@@
 The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}.
-However for this work, whilst the pulsar distance is used when generating synthetic data, as discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$.
+However for this work, whilst the pulsar distance is used when generating synthetic data, as is discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$.
 Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
@@
 set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
-The majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{s}^{-5/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
+Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-5/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
 For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-5/2}$).
@@
 
-\subsubsection{Parameter estimation}\label{sec:parameter_estim}
-From Equation \eqref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms.
+\subsection{Parameter estimation}\label{sec:parameter_estim}
+In this section we derive the posterior probability distributions for the static parameters.
+However, it is first necessary to discuss a modification to the general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}.
+It can be seen that the equation generally separates into two cosine terms.
 The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
@@
 	\begin{equation}
-		g^{(n)}_{\rm Earth}(t) = 1 - \frac{1}{2} \frac{ H_{ij}q^{i,(n)} q^{j,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \cos(-\Omega t +\Phi_0)  \ .
+		g^{(n)}_{\rm Earth}(t) = 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \cos(-\Omega t +\Phi_0)  \ .
 		\label{eq:g_func_trig_earth}
 	\end{equation}
-	We defer the inclusion of the pulsar terms in the measurement equation to a future work.
+	We defer the inclusion of the pulsar terms to a future work.
 As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
@@
 The generated synthetic data that we use to test our method does include the pulsar term in full.
+\newline
+
 	Initially we consider a single noise realisation of the synthetic data for the representative example system described in Table \ref{tab:parameters_and_priors}.
-We apply the Kalman filter using the Earth terms only measurement equation, Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
-\newline
-
+We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
 The results are shown in Figure \ref{fig:corner_plot_1} for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$.
-The histograms shown in the Figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
+The histograms shown in the figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
 The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
-It is evident that for this representative example with strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
+It is evident that for this representative example with characteristic strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
 \newline
@@
 a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation.
-We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.
+We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
 Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
@@
 The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
 We are able to accurately estimate each parameter of interest.
-We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
+We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
 Note that the $x$-axis scaling does not span the total prior space.}
@@
 
-\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
+\subsection{Multiple noise realisations} \label{sec:multiple_noise}
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
 	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes.
+The subtitles of the marginalized 1D posteriors refer to values for the median posterior across the noise realisations.
 The injection value is generally contained within the 90 \% credible interval for the vast of majority posteriors.
@@
 	\end{subfigure}
-	\caption{For each parameter of $\boldsymbol{\theta}_{\rm gw}$ the first moment of the Wasserstein distance, $W_1$, calculated between each pair of one-dimensional posteriors for our representative system across $10^3$ realisations.
-$W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
+	\caption{The first moment of the Wasserstein distance, $W_1$, calculated between each pair of one-dimensional posteriors for our representative system across $10^3$ realisations, for each parameter of $\boldsymbol{\theta}_{\rm gw}$  $W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
 $W_1$ is generally small across all parameters and all posteriors, but some posteriors converge poorly leading to large values of $W_1$.
-The $W_1$ values inferred for $\psi$ and $\alpha$ are strongly correlated with correlation coefficient = 0.96.}\label{fig:pairwise_wasserstein}
+The $W_1$ values inferred for $\psi$ and $\alpha$ are strongly correlated with correlation coefficient = 0.96.
+An example noise realisation with particularly large pairwise $W_1$ values is labelled with an asterisk in (c) and is discussed in the text.} \label{fig:pairwise_wasserstein}
 \end{figure*}
@@
 	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\alpha-\psi$ parameter space for a single noise realisation.
-The red point labels the location of the maxima, and the magenta point labels the location of the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data.
+The red point labels the location of the maxima, and the magenta point labels the the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data.
 The likelihood has been normalised with respect to the absolute value of the maxima.
 A multi-modal structure is clearly visible and for this noise realisation the nested sampler has settled in the local optima.
-This can be solved generally by increasing the number of live points $n_{\rm live}$}
+This can be solved generally by increasing the number of live points $n_{\rm live}$.}
 	\label{fig:likelihood_surface_alpha_psi}
@@
 1000 is a hangover when we were using a larger strain ($h=10^{-12}$) as our canonical example.
-The $10^3$ noise realisations are currently running and plots will be updated once completed, but I don't expect any meaningful changes \textcolor{red}{since for $h=10^{-12}$ the results/conclusions of 100 realisations are the same as 1000 realisations)}}\normalsize} realisations of the process noise and measurement noise.
+The $10^3$ noise realisations are currently running and plots will be updated once completed, but I don't expect any meaningful changes \textcolor{red}{; for $h=10^{-12}$ the results/conclusions of 100 realisations are the same as 1000 realisations)}}\normalsize} realisations of the process noise and measurement noise.
 For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
@@
 Considering the one-dimensional marginalized posteriors, the inferred probability distributions of the parameters broadly overlap across the different noise realisations.
-For the vast majority of parameters and noise realisations the injected value is contained within the 90 \% credible interval.
-As in the singe noise case, there are tentative signs of bias for some of the parameters, with the one-dimensional posteriors not symmetric about the injected value.
+For the vast majority of parameters and noise realisations the injected value is contained within the 90\% credible interval.
+As in the single noise case, there are tentative signs of bias for some of the parameters, with the one-dimensional posteriors not symmetric about the injected value.
 For example, the maximum a posteriori probability estimates of $\iota$ across the noise realisations all appear left-skewed, consistently underestimating the injected value.
 Given the uncertainty in the posteriors it is difficult to draw strong conclusions on a potential bias and this question will be discussed in detail in Section \ref{sec:bias} \newline
-
-%
-%We saw previously strong biases in $\iota$ and $h_0$ when considering just 9 noise realisations.
-These biases persist with a similar magnitude when we consider the full set of $10^3$ noise realisations.
-Moreover, smaller biases are also observed in $\psi$ ($\sim 0.1$ radians) and $\alpha$ ($\sim 0.04$ radians) as well as smaller still biases in $\delta$ and $\Phi_0$.
-%
-%
-%
-%These biases are a result of dropping the pulsar terms from the measurement equation and will be discussed in Section
 
@@
 The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
-Throughout this work we use the first moment of the Wasserstein distance $W_1(\mu,\nu)$ which is defined between one-dimensional probability distributions $\mu$ and $\nu$ as
+Throughout this work we use the first moment of the Wasserstein distance, $W_1(\mu,\nu)$, which is defined between one-dimensional probability distributions $\mu$ and $\nu$ as
 \begin{eqnarray}
-	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y)
+	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y) \ ,
 \end{eqnarray}
-where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.  \newline
-
-For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
-The results are presented in Figure \ref{fig:pairwise_wasserstein} which shows the $W_1$ value for each pair of posteriors.
-The first moment of the WD has a useful physical interpretation given by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
+where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.
+The first moment of the WD has a convenient physical interpretation given by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
 \begin{eqnarray}
-	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
+	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
 \end{eqnarray}
 where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
-We can see that the $W_1$ bounds the difference in the expected values.
-That is to say, if $W_1 =0.5$ rad between two one-dimensional posterior distributions of e.g.
-$\iota$ this means that the difference in the expected values from these distributions is no greater than $0.5$ rad.
-From Figure \ref{fig:pairwise_wasserstein} it is evident that generally $W_1$ is small across all pairs of posteriors.
-The median values across all pairs for each parameter are quoted in Table \ref{tab:Wasserstein}.
+We can see that $W_1$ bounds the difference in the expected values.
+Taking a concrete example, suppose that we infer two one-dimensional posterior distributions for $\iota$, for two different realisations of the noise, and calculate $W_1 =0.5$ rad between them.
+This means that the difference in the expected value of two random variables drawn from the two posteriors is no greater than $0.5$ rad.
+\newline
+
+For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
+The results are presented in Figure \ref{fig:pairwise_wasserstein}.
+The figure shows a lower triangular heat-map where each point denotes the $W_1$ value between a pair of posteriors.
+Small values of $W_1$ are magenta moving to red for large values of $W_1$.
+Note that the heat-map colour scaling is not the same for all subplots of the figure.
+It is evident that $W_1$ is generally small across all pairs of posteriors.
+We define $W_{1, \rm median}$ as the median value of $W_1$ across all $10^3 \choose 2$ pairs of noise realisations.
+The values of $W_{1, \rm median}$ for each parameter are quoted in Table \ref{tab:Wasserstein}.
 As a percentage of the width of the prior space (e.g.
-for $\alpha$ the width of the prior $=2 \pi$), the median $W_1$ for each parameter is $\lesssim 3 \%$.
+for $\alpha$ the width of the prior is $2 \pi$), the median $W_1$ for each parameter is $\lesssim 3 \%$.
 Generally speaking, the method converges to consistent one-dimensional posteriors for each parameter across different noise realisations.
@@
 
-There are a small fraction of the total number of noise realisations
-which have a large distance from all the other posteriors of all the other noise realisations.
+There are a small fraction of the total number of noise realisations which have a large Wasserstein distance from all the other posteriors of all the other noise realisations.
 These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
 These correspond exclusively to posteriors which have converged poorly and not found the likelihood global maximum.
-For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the Figure, corresponds to a posterior which has converged poorly, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
-The $W_1$ between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ radians.
-This problem can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
-We also observe a a strong correlation (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
+For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the figure, corresponds to a posterior for $\psi$ which has converged sub-optimally, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
+The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ radians.
+We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
 This is due to multimodal likelihood surface for $\psi-\alpha$, shown in Figure \ref{fig:likelihood_surface_alpha_psi}.
-We can see that the likelihood surface has a global optima at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local optima that corresponds to the location of the poorly converged posterior discussed previously with $\psi = 1.01$ rad and a large $W_1$.
-
-\subsubsection{Detection} \label{sec:detection}
+We can see that the likelihood surface has a global optima at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local optima that corresponds to the location of the example poorly converged posterior discussed previously, with $\psi = 1.01$ rad.
+This problem of poor global convergence can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
+
+\subsection{Detection} \label{sec:detection}
 We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
@@
 
-There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10{-15}$.
+There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
 The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
@@
 
-\subsection{Exploring a broader parameter space} \label{sec:parameter_space}
+\section{Exploring a broader parameter space} \label{sec:parameter_space}
 \begin{figure}
@@
 Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$.
-The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad in order to maintain an approximately constant SNR.
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
 The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals.
 Well estimated posteriors should fall along the diagonal $y=x$.
-Many parameters show evidence of being over-constrained due to a modelling bias.
+$\Omega$ is generally well-estimated but many other parameters show evidence of being over-constrained due to a modelling bias.
 }
@@
 We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
-It is also important to test the method in different regions of the parameter space.
-To this end we consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
+In this section we test the method in different regions of the parameter space, varying the source parameters through astrophysically relevant ranges.
+\newline
+
+We consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
 We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space.
@@
 The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and there are more injections contained within the low value credible intervals.
-This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally highly confident, i.e.
-narrow, but are systematically biased away from the true injection values.
+This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally confident, but are systematically biased away from the true injection values.
 This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval.
@@
 
-\subsection{Identifiability and bias}
+\section{Identifiability and bias} \label{sec:bias_and_identifiability}
 There are two underlying issues in the preceding tests that we now explore in more detail.
@@
 can we uniquely identify the parameters of the model given the data and the model structure?
+This question is explored in Section \ref{sec:identif}.
 The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
+This is the focus of Section \ref{sec:bias}.
 In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f.
@@
 
-\subsubsection{Identifiability}
+\subsection{Identifiability}\label{sec:identif}
 \begin{figure}
@@
 	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
-	\caption{Posterior distributions over nine noise realisations for the GW source parameters $\iota$ and $h_0$ for the representative system described in Table \ref{tab:parameters_and_priors}, in the high SNR regime ($h_0 = 10^{-12}$).
+	\caption{Posterior probability distributions for the representative system described in Table \ref{tab:parameters_and_priors} over nine noise realisations for the GW source parameters $\iota$ and $h_0$ in the high SNR regime ($h_0 = 10^{-12}$).
 Unlike the other parameters of  $\boldsymbol{\theta}_{\rm gw}$, the one-dimensional posteriors for $\iota$ and $h_0$ do not strongly overlap.
@@
 In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations. 
+The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but just for $\iota$ and $h_0$ rather than all members of $\boldsymbol{\theta}_{\rm gw}$.
 In the case of high-SNR we might expect \textit{a priori} that our estimates would be increasingly accurate over the low-SNR case, and that the method would converge robustly to similar posteriors irrespective of the noise realisation (i.e.
-small $W_1$).
+$W_1 \to 0$).
 This is generally true for the majority of the component parameters of $\boldsymbol{\theta}_{\rm gw}$; the estimated posteriors across noise realisations are largely indistinguishable.
 However for $\iota$ and $h_0$ this is not the case.
-Instead the inferred one-dimensional posteriors do not broadly overlap are not consistent across different noise realisations.
+Instead the inferred one-dimensional posteriors do not broadly overlap and are not consistent across different noise realisations.
 Specifically across $10^3$ noise realisations the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
-For $\iota$ this $W_{1,\rm median}$ is greater than the value inferred in the low SNR case (0.14, Table \ref{tab:Wasserstein}, whilst for $h_0$ $W_{1,\rm median}$ is also greater as a proportion of the scale set by the strain amplitude.
-This is in contrast to $W_{1,\rm median}$ inferred for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ in the high-SNR regime; e.g.
-for $\Phi_0$, $W_1 = 4.5 \times 10^{-3}$ rad.
+For $\iota$, $W_{1,\rm median}$ is greater than the value inferred in the low SNR case ($=0.14$, Table \ref{tab:Wasserstein}), whilst for $h_0$ $W_{1,\rm median}$ is also greater as a proportion of the scale set by the strain amplitude.
+This is in contrast to $W_{1,\rm median}$ inferred for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ in the high-SNR regime where as SNR increases, $W_1$ decreases; e.g.
+for $\Phi_0$, $W_{1, \rm median} = 4.5 \times 10^{-3}$ rad.
 \newline
@@
 Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
-Sources of non-identifiability are generally categorised as either ``structural" (arising from the structure of the model) or ``practical" ( arising from insufficient data or measurement errors) \citep{GUILLAUME2019418}.
+Sources of non-identifiability are generally categorised as either ``structural", arising from the structure of the model, or ``practical", arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
 Regarding $\iota$ and $h_0$, we are concerned with structural identifiability.
-Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx}, \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx} and \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
 For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
 This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
-However, it can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
-That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a non-zero gradient close to the maximum likelihood solution.
+It can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
+That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a small, non-zero gradient close to the maximum likelihood solution.
 This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
@@
 If during the inference we fix one of the parameters at its true value (i.e.
-set a delta function prior), taking a slice through the likelihood surface it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably.
-However if we fix e.g $h_0$ at its true value then the one dimensional posteriors for $\iota$ are all strongly biased away from the injected value; the maximum a posteriori probability estimates underestimate the injected value by $\sim 0.34$ rad.
-This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be discussed further in Section \ref{sec:bias}.
+set a delta function prior), taking a slice through the likelihood surface it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably, i.e.
+$W_1 \to 0$.
 In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" \footnote{\tiny \textcolor{red}{TK: "washes out" may be too colloquial / unspecific, but I think it is intuitive...}\normalsize}the likelihood ridge, leading to a more concave likelihood surface.
 This enables consistent posteriors to be estimated for $\iota$ and $h$.
-
-\subsubsection{Bias by dropping pulsar terms}\label{sec:bias}
+In the case where we fix e.g $h_0$ at its true value to infer indistinguishable posteriors for $\iota$, all of the posteriors for $\iota$ are all strongly biased away from the injected value; the maximum a posteriori probability estimates underestimate the injected value by $\sim 0.34$ rad.
+This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be now be discussed further in Section \ref{sec:bias}.
+\subsection{Bias by dropping pulsar terms}\label{sec:bias}
 
@@
 	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
-	\caption{As Figure \ref{fig:corner_plot_1} but for the low-SNR system with $h_0 = 10^{-12}$.
+	\caption{As Figure \ref{fig:corner_plot_1} but for the high-SNR system with $h_0 = 10^{-12}$.
 For the majority of parameters the injection value does not fall within the 90\% credible interval and is not visible over these scales.
@@
 All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
-The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model.
+The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model; the green/orange dashed lines are overlaid on this scale.
 For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$).
 For the pulsar-terms model the location of the maxima and injected value do coincide.
-10 noise realisations are plotted but the solutions are completely overlaid.}
+10 noise realisations are plotted for each situation but the solutions are completely overlaid on this scale.}
 	\label{fig:likelihood_surface_iota}
@@
 This effect is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
+Similar results are obtained for other noise realisations.
 Whilst $h_0$ does not exhibit any noticeable bias for this particular noise realisation, in general across multiple noise realisations it does exhibit a correlated bias with $\iota$.
@@
 using Equation \ref{eq:measurement}), \textit{(iii)} the solid green line is the solution where we drop the pulsar terms from both the synthetic data and the measurement equation.
-Whilst we have not used the nested sampling method with either solutions \textit{(ii)} or \textit{(iii)} in this work, we are able to run the Kalman filter either using the pulsar terms in the model or else dropping them from the synthetic data consider how the resulting likelihood varies for different parameter values.
+We have not used nested sampling for situations \textit{(ii)} or \textit{(iii)} in this work.
+However, we are able to run the Kalman filter either using the pulsar terms in the model or else dropping them from the synthetic data consider how the resulting likelihood varies for different parameter values.
 The dashed lines show the location of the likelihood maxima for each of the likelihood solutions; finding the location of the maxima across all static parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
@@
 This optima is comparable to the median of the posterior inferred for $\iota$ in Figure \ref{fig:bias_for_large_h}.
-\newline
-
-%This difference is comparable to the bias observed in the right-hand panel of Figure \ref{fig:iota} where the median of the marginalised posterior for $\iota$ was $\sim 0.65$ radians.
-The discrepancy between $\iota \sim 0.7$ in the log-likelihood curve and $\iota \sim 0.65$ in the nested sampling inference is a result of the likelihood curves being a one-dimensional representation of a multidimensional problem.
-The exact shape of the likelihood curve naturally depends on the values of all of the other parameters that we hold constant.
-In the nested sampling case the estimate for $\psi$ is sightly biased at $\psi \sim 2.60$ radians compared to the true injected value of $\psi = 2.50$ radians that is used for the results of Figure \ref{fig:likelihood_surface_iota}.
 \newline
@@
 Consequently the parameters which are most affected are those which determine the amplitude of the GW.
-Terms which influence the amplitude strongly such as $h_0$ and $\iota$ are strongly affected and exhibit a strong bias.
+Terms which influence the amplitude strongly, such as $h_0$ and $\iota$, are strongly affected and exhibit a strong bias.
 Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
-The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
+The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
+Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
 This hierarchy of biases for each parameter is also what we observe in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
@@
 
-\section{Discussion}
+\section{Discussion}\label{sec:discussion}
 We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
@@
 
-We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up NANOGrav (Section \ref{sec:synt_pta}).
+We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
 Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA.
@@
 %https://arxiv.org/pdf/2107.03047.pdf
-\section{Conclusion}
+\section{Conclusion}\label{sec:conclusion}
 In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
@@
 \appendix
-\newpage
+
 \section{Kalman recursion equations} \label{sec:kalman}
@@
 \begin{equation}
-	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}
+	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
 \end{equation}
@@
 \end{equation}
-Equation \ref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which yields the minimal mean square error in the measurement post fit residual, $\boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k}$.
+Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which yields the minimal mean square error in the measurement post fit residual, $\boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k}$.
 \newline
@@
 We can map our state-space model described in Section \ref{sec:model} onto the Kalman filter structure of Equations  
-\ref{eq:kalman1}, \ref{eq:kalman2} as follows.
-We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e $\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right)$.
-Similarly  $\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)$.
+\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
+\begin{equation}
+	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
+\end{equation}
+Similarly,  
+\begin{equation}
+\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
+\end{equation}
 The states evolve according the continuous dynamical equation
@@
 \end{equation} 
-From Equation \ref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ as Equation \ref{eq:kalman2} where
+From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ as Equation \eqref{eq:kalman2} where
 \begin{equation}
@@
 The remaining component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
-These are defined straightforwardly from Equation \ref{eq:measurement}.
+These are defined straightforwardly from Equation \eqref{eq:measurement}.
 Specifically,
-$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \ref{eq:g_func}.
+$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
 $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
 
+\section{Fiducial pulsar values}\label{appendix_fiducial}
+\textcolor{red}{TK TBD} 
 \section{Identifiability}\label{appendix_identifiability}
@@
 For this reason it is sometimes known as the "Earth mover's distance".
-\newline
-
 The $p$-th order Wasserstein distance between two distributions is
 \begin{eqnarray}
-	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p} \, 
+	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
 \end{eqnarray}
-where $p \in [1,\infty)$ and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$ (i.e.
-the set of couplings of $\mu$ and $\nu$).
-\newline
-
-For finite, empirical, discrete samples drawn from underlying probability distributions defined on $\mathbb{R}^d$ it is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
+where $p \in [1,\infty)$ and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
+$\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
+\newline
+
+The cost function can be freely chosen for the nature of the problem, but is commonly taken to be the absolute value function, $c(x,y) = |x-y|$, and we adopt the same cost function in this work.
+Given this choice of $c(x,y)$, for an empirical distribution $\hat{\mu}_n$ composed of $n$ i.i.d samples drawn from the underlying probability distribution $\mu$, the solution to the discrete form of Equation \eqref{eq:wasserstein} can be computed generally by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
+However, for $\mu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
 \begin{eqnarray}
-	\boldsymbol{E} [W_1(\mu_n,\mu)] \asymp n^{-\frac{1}{d}}
+	\boldsymbol{E} [W_1(\hat{\mu}_n,\mu)] \asymp n^{-\frac{1}{d}} \ .
 \end{eqnarray}
-In order to circumvent this issue in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors
-
-For empirical distributins the
-
-Pracical definitin for emiral distributions.
-
-First order and units of WD
-
-Curse of dimenionality for emipiral distributions
-
-Advantages
-
-symmetric
-
-triangle inequality
-
-non parameteric
-
-intuitive
-
-takes account of geometry
-
-comparison with KL
-
-To sidestep this issue, we marginalize onto the distributions spanned by the individual gridpoints and take the WD between these 1-dimensional distributions.
-Such an approach is also adopted in \cite{Paxton2021,Vissio2020}.
-
-In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD).
-The WD intuitively defines a distance between two distributions as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$.
-where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function.
-It is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
+In order to circumvent this issue, in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors, i.e.
+setting $d=1$.
+In this case Equation \eqref{eq:wasserstein} has a convenient, easily computed, closed form:
 \begin{eqnarray}
-	\boldsymbol{E} [W_1(\mu_n,\mu)] \asymp n^{-\frac{1}{d}}
+	W_p(\mu,\nu)= \left(\int_0^1  F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) dz \right)^{1/p} \, \label{eq:wasserstein}
 \end{eqnarray}
-To sidestep this issue, we marginalize onto the distributions spanned by the individual gridpoints and take the WD between these 1-dimensional distributions.
-Such an approach is also adopted in \cite{Paxton2021,Vissio2020}.
-We also take $p=1$ and $c(x,y) = |x-y|$ such that the WD values quoted in this work are given by
-\begin{eqnarray}
-	W_1(\mu,\nu)=  \inf_{\gamma \in \Gamma(\mu, \nu)} \int |x-y| d \gamma (x,y)
-\end{eqnarray}
-From the perspective of climate modelling, the WD has a nice physical interpretation given by the Monge-Kantorovich duality:
-\begin{eqnarray}
-	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
-\end{eqnarray}
-i.e.
-the difference in the expected value of some random variable between two distributions is no bigger than the WD value itself.
-More explicitly, if the WD for surface temperature between two distributions is 1K, then this tells us that the difference in the expected temperature between the distributions is no greater than 1K.
-For a full review of the Wasserstein distance, including favourable properties and comparison with other metrics we refer the reader to the Appendix of Pax21.
+where $F_{\mu}(z)$ is the cumulative distribution function of $\mu$. \newline
+
+The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages.
+Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another.
+It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g.
+radians then we can interpret this as the cost to transform between the two distributions is 1 radian.
+Further interpretability is provided by the Monge-Kantarovih duality, Equation \eqref{eq:WDdefn}; the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian.
+Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g.
+Gaussianity.
+Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces.
+Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality.
+It also respects the geometry of the underlying metric space over which the distributions are defined.
 

### 2023-09-18 docs/manuscript/paper.tex
some paper corrections, ldunn

@@
 with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
+Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.
 Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$.
@@
 	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
-	\caption{Spatial distribution of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
+	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
 The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc.
@@
 \end{eqnarray}
-For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s$^{-5/2}$, $\min [ \sigma^{(n)} ] = 1.3 \times 10^{-23}$s$^{-5/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.0 \times 10^{-16}$ s$^{-5/2}$ for PSR J1939+2134 \newline
-
-As a cross-check, we calculate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
+For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.3 \times 10^{-23}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.0 \times 10^{-16}$ s$^{-3/2}$ for PSR J1939+2134 \newline
+
+As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
 \begin{eqnarray}
@@
 Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$.
-The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
 Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
@@
 set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
-Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-5/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
-For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-5/2}$).
-The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-5/2}$.
-For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-5/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
+Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-3/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
+For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-3/2}$).
+The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-3/2}$.
+For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-3/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
 \newline
@@
 		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
-		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
+		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
 		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
@@
 \section{Kalman recursion equations} \label{sec:kalman}
-The linear Kalman filter operate on temporally discrete, noisy measurements $\boldsymbol{Y}_k$ which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$ via a linear transformation:
-\begin{equation}
-	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \label{eq:kalman1}
+The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
+\begin{equation}
+	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
 \end{equation}
@@
 \begin{equation}
-	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \label{eq:kalman2}
-\end{equation}
-where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the input or control matrix.
-and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$.
-\newline
-
-The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage where the state at the current time-step is predicted based on the state at the previous time-step, and an ``update" stage where the predicted state estimate is refined based on the measurement at that time-step.
+	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
+\end{equation}
+where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix.
+$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$.
+\newline
+
+The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage where the state at timestep $k+1$ is predicted based on the state at timestep $k$, and an ``update" stage where the predicted state estimate is refined based on the measurement at that time-step.
 The predict stage updates the estimate of the state $\hat{\boldsymbol{X}}_{k|k-1}$ and the associated covariance $\hat{\boldsymbol{P}}_{k|k-1}$where the subscript notation ${k|k-1}$ denotes an estimate at discrete step $k$, given measurements from step $k-1$ and earlier.
@@
 
-We can map our state-space model described in Section \ref{sec:model} onto the Kalman filter structure of Equations  
+We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
 \eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 

### 2023-09-18 docs/manuscript/paper.tex
misc paper updates

@@
 In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
-They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio photon.
+They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio pulse.
 The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{images/example_residuals_plot}
+	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
 	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J0030+0451.
@@
 \end{equation}
-where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance, and the subscript $k$ labels the time-step.
 The Kalman filter takes the underlying states to evolve according to
@@
 where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix.
-$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$.
-\newline
-
-The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage where the state at timestep $k+1$ is predicted based on the state at timestep $k$, and an ``update" stage where the predicted state estimate is refined based on the measurement at that time-step.
-The predict stage updates the estimate of the state $\hat{\boldsymbol{X}}_{k|k-1}$ and the associated covariance $\hat{\boldsymbol{P}}_{k|k-1}$where the subscript notation ${k|k-1}$ denotes an estimate at discrete step $k$, given measurements from step $k-1$ and earlier.
-The predict step proceeds as,
+$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$.
+\newline
+
+The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage.
+The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
+Specifically, the predict step proceeds as,
 \begin{equation}
@@
 \end{equation}
-The update step then uses the measurement $\boldsymbol{Y}_k$ to update $\hat{\boldsymbol{X}}_{k|k}$  and $\hat{\boldsymbol{P}}_{k|k}$ as,
+Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
 \begin{equation}
@@
 \end{equation}
-Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which yields the minimal mean square error in the measurement post fit residual, $\boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k}$.
+Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
+$\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$.
+For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
 \newline
@@
 \end{equation}
-The states evolve according the continuous dynamical equation
+The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
 \begin{equation}
@@
 \end{equation}
-where $\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right)$ is the state matrix, $\boldsymbol{C} = \left(C^{(1)}, C^{(2)}, ..., C^{(N)}\right)$ the input or control matrix where $C^{(i)} =\gamma^{(i)} \left(f_{\rm em}^{(i)} (0) + \dot{f}_{\rm em}(0)^{(i)} t \right) +  \dot{f}_{\rm em}(0)^{(i)}$ and $d\boldsymbol{B}(t)$ denotes a Wiener process with covariance matrix $\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)$.
-Equation \ref{eq:kalmn2} is a Langevin equation (equivalently an Ornstein-Uhlenbeck process) which has a general solution given by \citep{gardiner2009stochastic},
+where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix
+\begin{equation}
+\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
+\end{equation}
+and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
+\begin{equation}
+ C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
+\end{equation}
+The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
+\begin{equation}
+\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)
+\end{equation}
+\newline
+
+The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
 \begin{equation}
@@
 \end{equation} 
-From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ as Equation \eqref{eq:kalman2} where
+From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
 \begin{equation}
@@
 \end{equation}
-where
-\begin{equation}
-	G_k^{(n)} =    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] - e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
-\end{equation}
+with
+\begin{align}
+	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
+	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
+\end{align}
 and
 \begin{equation}
-	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1), Q^{(2)}},...,Q^{(N)}\right) 
+	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) 
 \end{equation}
@@
 
-The remaining component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
+The remaining component matrices of the Kalman filter that we have not yet specified are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
 These are defined straightforwardly from Equation \eqref{eq:measurement}.
@@
 $\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
-$\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
+The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
 

### 2023-09-19 docs/manuscript/paper.tex
liam comments included

@@
 The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
-For each pulsar we adopt the fiducial NANOGrav values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
+For each pulsar we adopt the fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
 A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility.
@@
 \end{align}
-where $T_{\rm cad}$ is the cadence of the timing observations and MSPs have $\ln \alpha_1 \approx -20 $, $\alpha_2 \approx 1$, $\alpha_3 \approx 2$ and $\alpha_4 \approx 2.4$.
+where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits.
+The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value.
 Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
@@
 and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach.
-We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} conducted with the Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}.
+We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
 A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$.
 In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$.
-In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J0030+0451, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}.
+In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}.
 We see that the synthetic and empirical residuals are qualitatively similar.
-Comparable behaviour is observed for other pulsars in our array.
+For this pulsar the synthetic residuals were generated using $\sigma^{(n)} = 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the value inferred from the noise model of Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, using the best fit parameters, $\sigma^{(n)} = 2.0 \times 10^{-16}$ s$^{-3/2}$.
+This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars.
+This discrepancy between the two estimates of $\sigma^{(n)}$ is particular to PSR J1939+2134; for all other pulsars we find the two estimates of $\sigma^{(n)}$ are comparable to within a few orders of magnitude.
 \begin{figure}
 	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
-	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J0030+0451.
-The actual residuals are sourced from the UTMOST pulsar timing program \citep{UTMOST}.
-The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
+	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J1939+2134.
+The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
+The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$.
 The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}.
-Following \citet{Vargas} we set the TOA uncertainty to be a constant $10^{-4}$ s.
+We set the TOA uncertainty to be a constant $0.1 \mu$s.
 The blue and orange residuals are qualitatively similar by inspection.
-The other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} are tuned to produce comparable results.}
+Similar results are obtained for the other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} by tuning the value of $\sigma^{(n)}$.}
 	\label{fig:qualitative_compare}
@@
 To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
-\footnote{\textcolor{red}{\tiny TK: How much of the following is ok, how much to move to the appendix?
-I don't want to get bogged down in WD and distract from astro arguments but we use $W_1$ as metric so I feel like it should be defined in the main body of the text, and the bounding result is important...} \normalsize} The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
+\footnote{\textcolor{red}{\tiny TK: How much of the following Wasserstein stuff is ok here, how much to move to the appendix?
+I don't want to get bogged down in WD and distract from astro arguments, but we use $W_1$ as metric so I feel like it should be defined in the main body of the text, and the bounding result is important...} \normalsize} The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
 A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}.
@@
 \section{Fiducial pulsar values}\label{appendix_fiducial}
-\textcolor{red}{TK TBD} 
+\textcolor{red}{TK: Do we really need this?
+The argument is for reproducibility, but we have a nice github repo that holds all the data we actually used, jupyter notebooks that show how the calculations are done.
+Perhaps it is sufficient to point the reader to the repo...
+}
 \section{Identifiability}\label{appendix_identifiability}

### 2023-09-20 docs/identifiability/identifiability.tex
start identifiability notes

--- 2023-04-20 docs/measurement_eqn_deerivation/LectureNotes_Math571.tex
@@
 	\normalfont\normalsize 
-	{Melbourne University - Summer 2023} \\ [0pt] % Your university, school and/or department name(s)
-	\huge Notes on state-space formulation of PTA problem% The assignment title
+	{Melbourne University} \\ [0pt] % Your university, school and/or department name(s)
+	\huge Notes on identifiability of PTA problem% The assignment title
 }\author{T. Kimpson} % Your name
@@
 
-These notes collect some ideas around how to formulate the PTA data analysis as a state-space problem. \newline
+These notes collect some ideas around how to check if parameters of our PTA-GW state space model are identifiable. \newline
 
-\noindent We will frequently reference previous works from A.Melatos, in particular Melatos 2018 (GR assignment UniMelb) and Melatos 2022 (private communication at UniMelb). \newline
+\noindent We will deal with just 2 parameters $\iota$ and $h_0$.
 
-\noindent Geometric units ($c=\hbar = G = 1$) are used throughout, with the usual convention of Roman indices ($i,j$) for spatial indices.
-Note that $i$ is used for both labelling tensor indices and as the imaginary number - I trust the difference to be obvious from the context.\newline
+\section{General identifiability method}\label{sec:intro}
 
-\section{Derivation of pulsar frequency modulation due to GW}
-We want to know how the pulse frequency from a pulsar is influenced by a passing gravitational wave (GW). \newline
+How do we tell if parameters are identifiable? Lets use the approach from Karlsson 2012 \footnote{\url{https://www.sciencedirect.com/science/article/pii/S1474667015380745}}
 
-\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$  \footnote{See Melatos 2022 for nuances and discussion around this construction of "pulse train as photon"} \newline
+\noindent Take a state space model
+\begin{equation}
+	\dot{x}(t) = f(x(t), u(t), \theta)
+\end{equation}
+\begin{equation}
+	y(t) = g(x(t), u(t), \theta)
+\end{equation}
+where state $x$ has dimension $n$, $\theta$ has dimension $d$ and $y$ has dimension $p$. \newline
 
-\noindent We take a gravitational plane wave that perturbs a background Minkowski spacetime as \footnote{c.f. Melatos 2018, Equation 1}
+\noindent Now define an extended Lie derivative operator along the vector field $f$
 \begin{equation}
-g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
+	\mathcal{L}_f = \sum_{i=1}^{n} f_i \frac{\partial}{\partial x_i} + \sum_{i=0}^{i = \infty} u^{(i+1)} \frac{\partial}{\partial u^{(i)}}
 \end{equation}
-where the GW has angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$.
-Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.} The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
-\begin{align}
-	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
-\end{align}
-\noindent Note that we use a convention where the amplitudes $h_{+,\times}$ label the \textit{constant} amplitudes of the gravitational plane wave.
-All time variation in the strain is a result of the exponential term.
-This is in contrast to the notation used by some authors \footnote{see e.g.
-arXiv:1003.0677} where the amplitudes include the time variability $h_{+,\times} = h_{+,\times}(t)$.
-The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave.
-\newline
+\footnote{Is the vector field $f$ the same as the state space function $f$?
+\url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005153} suggests that is IS.
+We will proceed assuming that they are the same thing.}.
+We use the notation $\mathcal{L}_f^{(k)}$ to refer to the Lie operator applied $k$ times.
+We also define a variable
+\begin{equation}
+	\alpha^{(k)} = \mathcal{L}_f^{(k)} g(x,y,\theta)
+\end{equation}
+Note that Karlsson 2012 refer to this as $y^{(j)}(0)$. \newline
 
-\subsection{Setting up the problem}
-The frequency of a photon with 4-momentum $p_{\mu}$ recorded by an observer with 4-velocity $u^{\mu}$ is 
+\noindent We now define a vector $\mathcal{Y}$ which contains $\alpha^{(k)}$, $k=0,...,n+d-1$. We construct a Jacobian
 \begin{equation}
-	\nu = p_{\alpha} u^{\alpha}
+	J = \frac{\mathcal{Y}(x,\theta)}{\partial(x, \theta)}
 \end{equation}
+\footnote{See Equation 6 in \url{https://www.sciencedirect.com/science/article/pii/S0025556412000922} for a nice visual example of what this Jacobian looks like.} \newline 
+\textbf{Rank test for structural identifiability: } if $J$ is full rank \href{https://www.cds.caltech.edu/~murray/amwiki/index.php/FAQ:_What_does_it_mean_for_a_non-square_matrix_to_be_full_rank%3F#:~:text=A%20square%20matrix%20is%20full,in%20number)%20are%20linearly%20dependent.}{(see here)} then the parameters are identifiable.
 
-\noindent We consider both our emitter and receiver to be stationary, such that  
+\section{Relation to PTA-GW model}
+Our model is
 \begin{equation}
-u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
+	\dot{f}_{\rm p} = - \gamma f_{\rm p} + u(t)
 \end{equation}
+where $u(t) = \gamma f_{\rm EM}(t) + \dot{f}_{\rm EM} (t)$. The measurement equation is
+\begin{equation}
+	f_{\rm m}(t) = \left[1 - H(t,\theta) \right] f_{\rm p}
+\end{equation}
+By comparison with Section \ref{sec:intro}:
 
-\noindent Consequently the frequency can be identified with the temporal component of the covariant 4-momentum,
-\begin{equation}
-\nu = p_t
-\end{equation}
+\begin{itemize}
+	\item $x = f_{\rm p}$
+	\item $y = f_{\rm m}$
+	\item function $f(x) = -\gamma x + u(t)$
+	\item function $g(x) = \left[1 - H(t,\theta) \right] x$
+	\item $p = m = N_{\rm psr}$
+	\item $d = 2$ (We are just dealing with $h_0$ and $\iota$)
+\end{itemize}
 
-\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
+\section{Calculating $\alpha$}
 
-\begin{equation}
-	p_t(\tau)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=\tau} \dot{p}_t dt
-	\end{equation}
-
-\noindent where the overdot denotes a derivative w.r.t.
-$t$.
-Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $\tau = t_0 + d$ and consider the photon trajectory to be an unperturbed path.
-\footnote{See also e.g.
-Maggiore who takes the same approach...} \newline
-
-\noindent To complete our expression, we now just need to determine $\dot{p}_t$ and integrate it.
-
-\subsection{Hamiltonian Mechanics}
-
-The Hamiltonian in covariant notation can be written as
-
-\begin{equation}
-H(x^{\mu}, p_{\mu}) = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
-\end{equation}
-
-\noindent which if we substitute in our expression for the perturbed metric is
-
-\begin{equation}
-H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
-\end{equation}
-
-\noindent Hamilton's equations are
-\begin{equation}
-\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
-\end{equation}
-
-\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
-\begin{equation}
-\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)}
-\end{equation}
-\footnote{This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset.}
-
-\noindent Therefore the derivative w.r.t coordinate time $t$ is,
-\begin{equation}
-\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
-\end{equation}
-It will be useful to recognise that we can express the photon momentum in terms of the  unit vector between the Earth and pulsar $\bar{q}$
-\begin{equation}
-	p^{\mu} = \omega(1,-q^x,-q^y,-q^z)
-\end{equation}
-where $\omega$ is the \textit{constant} photon angular frequency.
-Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer (see Melatos 2022, PT16).
-Similarly, we can parameterize the spatial coordinates $\bar{x}$ as,
-\begin{equation}
-	\bar{x}(t) = - \bar{q} (t - \tau)
-\end{equation} 
-Note that when our integration variable $t = \tau$ (i.e.
-the photon is at Earth) then $\bar{x}(t=\tau) = 0$ which is what we expect given our chosen coordinate system.
-Also note that $\dot{p}_t$ is entirely a function of the GW perturbation.
-In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.
-\newline
-
-\noindent Bringing this all together
+Lets get some values for $\alpha^{(k)}$ and see if there is a pattern.
 
 \begin{align}
-	\dot{p}_t &= \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right) \\  
-	&= -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)} \left(\frac{1}{p^t}\right)
-	\\
-	&= \frac{-i\Omega \omega^2}{2} H_{ij}q^i q^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)} \left(\frac{1}{\omega}\right) \\
-	& = \frac{-i\Omega \omega}{2} H_{ij}q^i q^j  e^{i(\Omega(\bar{n}\cdot [- \bar{q} (t - \tau)] - t)+\Phi_0)} \\
-				& = \frac{-i\Omega \omega}{2} H_{ij}q^i q^j  e^{i(-\Omega t (1+ \bar{n} \cdot \bar{q}) + \Omega \tau  \bar{n} \cdot \bar{q} +\Phi_0)} \\
-						& =A e^{i (\gamma t + B)}
-\end{align}
-with 
-\begin{equation}
-	A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
-\end{equation}
-\begin{equation}
-	\gamma = -\Omega (1 + \bar{n}\cdot \bar{q}) 
-\end{equation}
-\footnote{compare with Melatos 2022, PT11}
-and 
-\begin{equation}
-	B = \Omega \tau \bar{n} \cdot \bar{q}  +\Phi_0
-\end{equation}
-
-\subsection{Performing the integral}
-
-The frequency shift experienced by the observer relative to the source due to a GW is then
-\begin{align}
-p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} &= A \int_{t = \tau - d}^{t=\tau} e^{i (\gamma t + B)}dt  \\
-&=\frac{-iA}{\gamma} e^{i B} \left(e^{i \gamma \tau} -e^{i \gamma (\tau - d)}\right) \\
-&=\frac{-iA}{\gamma} e^{i \Omega \tau \bar{n} \cdot \bar{q}} e^{i \Phi_0}e^{i \gamma \tau} \left(1 -e^{-i \gamma d}\right) \\
-&=\frac{-iA}{\gamma} e^{i \Omega \tau \bar{n} \cdot \bar{q}} e^{i \Phi_0}e^{i (-\Omega (1 + \bar{n}\cdot \bar{q}) ) \tau} \left(1 -e^{-i \gamma d}\right) \\
-&=\frac{-iA}{\gamma} e^{i \Phi_0}e^{-i \Omega \tau} \left(1 -e^{-i \gamma d}\right) \\
-&=\frac{-iA}{\gamma} e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i \gamma d}\right) 
-\label{eq:concise}
+	\alpha^{(1)} =& \mathcal{L}_f^{(1)} g(x,y,\theta) \\
+	                     =& \mathcal{L}_f \left[1 - H(t,\theta) \right] x \\
+	                     =&  \sum_{i=1}^{n} f_i \frac{\partial}{\partial x_i}  \left \{ \left[1 - H(t,\theta) \right] x\right \}+ \sum_{i=0}^{i = \infty} u^{(i+1)} \frac{\partial}{\partial u^{(i)}} \left \{ \left[1 - H(t,\theta) \right] x\right \}
 \end{align}
 
-\subsection{Explicit expression and comparison with Melatos 22}
-We are now very close with making contact with the expression from Melatos 22. Let's be explicit and sub in our definitions for $A$ and $\gamma$ into \ref{eq:concise}
-
-\begin{align}
-p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} &=-i \frac{-\frac{i\Omega \omega}{2} H_{ij}q^i q^j }{-\Omega (1 + \bar{n}\cdot \bar{q}) } e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i (-\Omega (1 + \bar{n}\cdot \bar{q}) ) d}\right) \\
-&=\frac{\omega}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \\
-&=\frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right)
-\label{eq:final}
-\end{align}
-which is exactly the expression derived in Melatos 2022.
-
-\subsection{The $\Phi_0$ issue}
-There was previously some confusion over the presence of a single $\Phi_0$ parameter that was shared between multiple pulsars.
-Does this cause $\tau = 0$ to be a special time when all pulsars are in phase and then go on to evolve to be out of phase?
-This is just due to some sloppy transcription of equations in another document.
-Ignoring measurement noise, the measured photon frequency is related to the emitted photon frequency  as
-\begin{equation}
-	f_{\rm measured} = f_{\rm emitted} \left(1 - \frac{1}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \right)
-\end{equation}
-Now the pulsar frequencies remain in phase and $\tau=0$ is no longer a special time.
-\footnote{This expression is slightly different from the (erroneous) one typed up in the manuscript!
-We have dropped a $1+ \bar{n} \cdot \bar{q}$ factor that was wrongly typed up in the other document}
-
 \end{document}

### 2023-09-20 docs/manuscript/internal_comments_3/difference_file/paper.tex
start identifiability notes

--- 2023-09-19 docs/manuscript/paper.tex
@@
 In this Section we confirm that we are able to infer comparable results for different noise realisations.
-To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate $1000$ \footnote{\tiny\textcolor{red}{TK: For the results plotted here we currently just do 100 noise realisations.
-1000 is a hangover when we were using a larger strain ($h=10^{-12}$) as our canonical example.
-The $10^3$ noise realisations are currently running and plots will be updated once completed, but I don't expect any meaningful changes \textcolor{red}{; for $h=10^{-12}$ the results/conclusions of 100 realisations are the same as 1000 realisations)}}\normalsize} realisations of the process noise and measurement noise.
+To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise and measurement noise.
 For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.

### 2023-09-26 docs/manuscript/paper.tex
extra small changes
do a check with smaller process noise

@@
 \begin{eqnarray}
-	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-1} \ . \label{eq:sigmap_f}
+	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
 \end{eqnarray}
-For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.3 \times 10^{-23}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.0 \times 10^{-16}$ s$^{-3/2}$ for PSR J1939+2134 \newline
+For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline
 
@@
 We see that the synthetic and empirical residuals are qualitatively similar.
-For this pulsar the synthetic residuals were generated using $\sigma^{(n)} = 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the value inferred from the noise model of Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, using the best fit parameters, $\sigma^{(n)} = 2.0 \times 10^{-16}$ s$^{-3/2}$.
+For this pulsar the synthetic residuals were generated using $\sigma^{(n)} = 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the value inferred from the noise model of Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, using the best fit parameters, $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$.
 This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars.
+If we set $\ln \alpha_1 = -38$ in Equation \eqref{eq:sigmap}, close to its lower limit, then the estimated $\sigma^{(n)}$ for  PSR J1939+2134 is $\sigma^{(n)} =3.90 \times 10^{-27}$ which is in much better agreement with the value inferred from the residuals comparison method.
 This discrepancy between the two estimates of $\sigma^{(n)}$ is particular to PSR J1939+2134; for all other pulsars we find the two estimates of $\sigma^{(n)}$ are comparable to within a few orders of magnitude.

### 2023-09-26 manuscript/manuscript2/paper.tex
some initial structure

--- 2023-09-26 docs/manuscript/paper.tex
@@
 % Keep the title short and informative.
-	\title[Kalman PTA]{Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array}
+	\title[Kalman PTA]{Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms}
 
@@
 \begin{abstract}	
-Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing array observations.
-A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-A two-step procedure is applied within a state-space framework, such that the model parameters are estimated with a Kalman filter, which then provides a likelihood for nested sampling.
-The procedure enables self-consistent inference of model parameters and calculation of the Bayesian evidence, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional analysis methods.
-It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$.
-Full posterior distributions of model parameters are recovered.
-There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
-Smaller biases $\lesssim 1 \%$ are also observed in some of the other parameters of the system.
+	The detection of continuous nanohertz gravitational waves from individual supermassive black hole binaries can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
+Such an approach complements traditional analysis methods by enabling self-consistent inference of model parameters and calculation of the Bayesian evidence, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise.
+Previous analyses have worked exclusively with the `Earth term', leading to biases in the parameter estimation.
+Here it is shown how to extend previous analyses to include the so-called `pulsar terms'.
+It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms increases the parameter accuracy relative to the Earth terms approach by X and the sensitivity by Y.
 \end{abstract}
@@
 \section{State-Space Formulation}\label{sec:model}
-We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation.
-In this work we take the intrinsic state variable to be the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local,freely-falling rest frame of the pulsar's centre of mass.
+
+In this Section we briefly review the state-space formulation of PTA analysis presented in K1.
+
+The intrinsic state variable is the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local,freely-falling rest frame of the pulsar's centre of mass.
+The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
+$f_{\rm p}^{(n)}(t)$ evolves according to a stochastic differential equation and is related to the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$ via a measurement equation.
+
 A phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.
-We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.
 The measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_measured}.
-The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
-The subtle problem of generalizing the measurement variable to pulse TOAs is postponed to future work, as noted in Section \ref{sec:intro}.
+
 \subsection{Spin evolution} \label{sec:psr_frequency}
-A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}.
-We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend.
+
+In our model
+
+Equation X is inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend.
+At present, a predictive, first-principles theory of timing noise does not exist and there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}.
+
 In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque.
@@
 
+As discussed in K1
 Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
@@
 \newline
+
+In K1
+
+The workflow presented in K1, hich we also follow here proceeds as follows
 
@@
 This discrepancy between the two estimates of $\sigma^{(n)}$ is particular to PSR J1939+2134; for all other pulsars we find the two estimates of $\sigma^{(n)}$ are comparable to within a few orders of magnitude.
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
-	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J1939+2134.
-The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
-The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$.
-The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}.
-We set the TOA uncertainty to be a constant $0.1 \mu$s.
-The blue and orange residuals are qualitatively similar by inspection.
-Similar results are obtained for the other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} by tuning the value of $\sigma^{(n)}$.}
-	\label{fig:qualitative_compare}
-\end{figure}
+
 \subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
@@
 
-\subsection{Multiple noise realisations} \label{sec:multiple_noise}
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
-	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes.
-The subtitles of the marginalized 1D posteriors refer to values for the median posterior across the noise realisations.
-The injection value is generally contained within the 90 \% credible interval for the vast of majority posteriors.
-The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
-A slight skew-bias is apparent in some of the parameters, e.g.
-$\iota$.
-Correlations can be seen between $\Omega - \Phi_0$, $\psi-\alpha$ and $\iota - h_0$.}
-	\label{fig:corner_plot_2}
-\end{figure*}
-\begin{table}
-	\centering
-	\begin{tabular}{ccll}
-		\toprule
-		Parameter & Injected Values & Units & $W_{1, \rm median}$  \\
-		\hline
-		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
-		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
-		$\psi$              & $2.50$ & rad & $0.16$ \\
-		$\iota$             & $1.0$ & rad & $0.14$ \\ 
-		$\delta$              & $1.0$  & rad & $0.09$ \\
-		$\alpha$          & $1.0$  & rad & $0.17$\\
-		$h_0$            & $5 \times 10^{-15}$ & - & $8 \times 10^{-16}$ \\
-		\bottomrule
-	\end{tabular}
-	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
-$W_{1, \rm median}$  sets the bound on the difference in expectation value between the one dimensional posteriors, and is generally small with respect to the width of the prior space.}
-	\label{tab:Wasserstein}
-\end{table}
-\begin{figure*}
-	\setkeys{Gin}{width=\linewidth}
-
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_0}
-		\caption{$\Omega$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_1}
-		\caption{$\Phi_0$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_2}
-		\caption{$\psi$} \label{figWD_psi}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_3}
-		\caption{$\iota$}
-	\end{subfigure}
-	\medskip
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_4}
-		\caption{$\delta$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_5}
-		\caption{$\alpha$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_6}
-		\caption{$h$}
-	\end{subfigure}
-	\caption{The first moment of the Wasserstein distance, $W_1$, calculated between each pair of one-dimensional posteriors for our representative system across $10^3$ realisations, for each parameter of $\boldsymbol{\theta}_{\rm gw}$  $W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
-$W_1$ is generally small across all parameters and all posteriors, but some posteriors converge poorly leading to large values of $W_1$.
-The $W_1$ values inferred for $\psi$ and $\alpha$ are strongly correlated with correlation coefficient = 0.96.
-An example noise realisation with particularly large pairwise $W_1$ values is labelled with an asterisk in (c) and is discussed in the text.} \label{fig:pairwise_wasserstein}
-\end{figure*}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_surface_alpha_psi}
-	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\alpha-\psi$ parameter space for a single noise realisation.
-The red point labels the location of the maxima, and the magenta point labels the the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data.
-The likelihood has been normalised with respect to the absolute value of the maxima.
-A multi-modal structure is clearly visible and for this noise realisation the nested sampler has settled in the local optima.
-This can be solved generally by increasing the number of live points $n_{\rm live}$.}
-	\label{fig:likelihood_surface_alpha_psi}
-\end{figure}
-The above example was for a single realisation of the noise processes.
-It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous or convenient for our method.
-In this Section we confirm that we are able to infer comparable results for different noise realisations.
-To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise and measurement noise.
-For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
-In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise.
-This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}.
-We plot only nine realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
-As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr, reduced}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here.
-\newline
-
-We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1}.
-Considering the one-dimensional marginalized posteriors, the inferred probability distributions of the parameters broadly overlap across the different noise realisations.
-For the vast majority of parameters and noise realisations the injected value is contained within the 90\% credible interval.
-As in the single noise case, there are tentative signs of bias for some of the parameters, with the one-dimensional posteriors not symmetric about the injected value.
-For example, the maximum a posteriori probability estimates of $\iota$ across the noise realisations all appear left-skewed, consistently underestimating the injected value.
-Given the uncertainty in the posteriors it is difficult to draw strong conclusions on a potential bias and this question will be discussed in detail in Section \ref{sec:bias} \newline
-
-We now consider the complete set of $10^3$ noise realisations, rather than the subset of nine realisations in the preceding discussion.
-For each realisation of the noise, for each parameter, we infer a marginalised one-dimensional posterior.
-Our goal is to compare these posteriors between different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
-To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
-\footnote{\textcolor{red}{\tiny TK: How much of the following Wasserstein stuff is ok here, how much to move to the appendix?
-I don't want to get bogged down in WD and distract from astro arguments, but we use $W_1$ as metric so I feel like it should be defined in the main body of the text, and the bounding result is important...} \normalsize} The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
-A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}.
-The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
-Throughout this work we use the first moment of the Wasserstein distance, $W_1(\mu,\nu)$, which is defined between one-dimensional probability distributions $\mu$ and $\nu$ as
-\begin{eqnarray}
-	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y) \ ,
-\end{eqnarray}
-where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.
-The first moment of the WD has a convenient physical interpretation given by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
-\begin{eqnarray}
-	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
-\end{eqnarray}
-where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
-We can see that $W_1$ bounds the difference in the expected values.
-Taking a concrete example, suppose that we infer two one-dimensional posterior distributions for $\iota$, for two different realisations of the noise, and calculate $W_1 =0.5$ rad between them.
-This means that the difference in the expected value of two random variables drawn from the two posteriors is no greater than $0.5$ rad.
-\newline
-
 For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.

### 2023-09-26 manuscript/manuscript_1/paper.tex
some initial structure

--- 2023-09-18 docs/manuscript/paper.tex
@@
 
-\usepackage{enumitem}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % If you need two or more lines of authors, add an extra line using \newauthor
-\author[Kimpson]{Tom Kimpson$^{1,2}$\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}, Andrew Melatos$^{1,2}$, Joseph O'Leary$^{1,2}$, Julian B.
-Carlin$^{1,2}$, Robin J.
-Evans$^{3}$, \newauthor William Moran$^{3}$, Tong Cheunchitra$^{1,2}$, Wenhao Dong$^{1,2}$, Liam Dunn$^{1,2}$, Julian Greentree$^{3}$, Nicholas J.
-O'Neill$^{1,2}$, \newauthor Sofia Suvorova$^{3}$, Kok Hong Thong$^{1,2}$, Andrés F.
-Vargas$^{1,2}$%
+\author[Kimpson]{Kimpson$^{1}$, Melatos$^{2}$, O'Leary, Evans$^{3}$, Carlin others, etc. etc.%
+\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
 %\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
@@
 \begin{abstract}	
-Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing array observations.
-A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-A two-step procedure is applied within a state-space framework, such that the model parameters are estimated with a Kalman filter, which then provides a likelihood for nested sampling.
-The procedure enables self-consistent inference of model parameters and calculation of the Bayesian evidence, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional analysis methods.
-It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$.
+Continuous nHz gravitational waves from individual supermassive black hole binaries may be detectable with Pulsar Timing Array observations.
+A novel search strategy is developed wherein intrinsic pulsar timing variations are tracked simultaneously with the modulation induced by a putative gravitational wave.
+A two-step procedure is applied within a state-space framework, where the model parameters are estimated with a Kalman filter, which provides a likelihood for nested sampling.
+This enables self-consistent inference of model parameters and calculation of the Bayesian evidence.
+It is shown via an astrophysically representative software injection campaign that the procedure can distinguish a gravitational wave from pure noise down to a strain of $h_0 \sim 4 \times 10^{-15}$, given an inclination angle $\iota = 1.0$ rad.
 Full posterior distributions of model parameters are recovered.
-There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
-Smaller biases $\lesssim 1 \%$ are also observed in some of the other parameters of the system.
+There is a bias in the estimates of some parameters, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
 \end{abstract}
@@
 Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
-The detection of nHz GWs has inspired the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
-The foremost method is timing an ensemble of pulsars, i.e.
-a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}.
+The detection of nHz GWs has necessitated the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
+The foremost method is timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}.
 A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth.
@@
 
-Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the early stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
+Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
 Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}.
@@
 Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs.
-However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}.
+However, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for a individual source at 4-5 nHz \citep{2023arXiv230616226A}.
 \newline
@@
 This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
-In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators.
+In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are relatively stable rotators.
 However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
 In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red).
-In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+In addition to the red timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
 \newline
@@
 Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
-The new and existing approaches are therefore complementary.
+The new and existing approach are therefore complementary, but the new approach holds out the promise of somewhat higher sensitivity.
 In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
-We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood (i.e.
-the model evidence) for model selection.
+We combine the Kalman tracking of the pulsars intrinsic rotational state with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood for model selection.
 \newline
@@
 \noindent This paper is organised as follows.
-In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW.
-In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence.
-In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method.
-We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative example GW system.
-In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space.
-Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.
-The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}.
-It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization which is deferred to future work.
-Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$.
-\newline
+In Section \ref{sec:model} we present the state-space model for the pulsar frequency which is subject to the influence of a GW.
+In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence.
+In Section \ref{sec:testing} we test this new method on synthetic pulsar data.
+Throughout this work we adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature.
+\newline
+
+%
+%The detection of high frequency ($\sim 100$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}.
+GWs from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.
+
+%
+% Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), 
+%
 
@@
 We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation.
-In this work we take the intrinsic state variable to be the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local,freely-falling rest frame of the pulsar's centre of mass.
-A phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.
+In this work we take the intrinsic state variable to be the pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local rest frame of the pulsar's centre of mass.
+A model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.
 We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.
@@
 The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
-The subtle problem of generalizing the measurement variable to pulse TOAs is postponed to future work, as noted in Section \ref{sec:intro}.
 \subsection{Spin evolution} \label{sec:psr_frequency}
@@
 In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque.
-The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
+The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$.
 Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution $f_{\rm em}^{(n)}(t)$ can be approximated accurately by
@@
 \begin{equation}
-	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
+	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)} +\xi^{(n)}(t) \ . 
 	\label{eq:frequency_evolution}
 \end{equation}
-In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
+In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm em}$ is the spin derivative, $\gamma^{(n)}$ a damping constant whose reciprocal specifies the mean-reversion timescale, and a white noise stochastic process $\xi^{(n)}(t)$ which satisfies:
 \begin{align}
 	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
+	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle &= [\sigma^{(n)}]^2 \delta(t - t') \ .	\label{eq:xieqn}
 \end{align}
-In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise.
-Combined with the mean reversion it gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
-It is important to note that white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
+$[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
 \newline
@@
 Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
-As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}.
+As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G}.
 An approach akin to Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} has been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
@@
 For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless.
-Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, which have not been ruled out by pulsar timing experiments to date.
+Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004} which have not been ruled out by pulsar timing experiments to date.
 The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \eqref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \eqref{eq:frequency_evolution}.
@@
 \subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
-In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
-Specifically, the pulse frequency is modulated harmonically at the GW frequency.
+In the presence of a GW, the pulsar frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
+Specifically, the pulsar frequency is modulated harmonically at the GW frequency.
 We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section.
@@
 \subsubsection{Plane GW perturbation}\label{sec:plane_gw}
-We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
+We consider a gravitational plane wave from a single,distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
 \begin{equation}
@@
 In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic.
-Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Zhu10},
-\begin{equation}
-	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
+Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves over decadal timescales as \citep[e.g.][]{Zhu10},
+\begin{equation}
+	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left(\frac{f_{\rm gw}(t=0)}{10^{-7} \mathrm{~Hz}}\right)^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
 	\label{eq:f_evolution}
 \end{equation}
-where $M_c$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
-A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
-From Equation \eqref{eq:f_evolution} we can see that only those binaries which are massive or at high frequency experience significant frequency evolution over typical PTA timespans.
+where $M_c$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=0)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ the length of the data timespan, which for PTAs is $\sim 10$ years.
+A source can be considered as monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution of $1/T_{\rm obs}$.
+From Equation \eqref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans.
 The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
-We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
+We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
 \newline
@@
 \end{align}
-The principal axes are in turn specified by the location of the GW source on the sky (colatitude $\theta$, azimuth $\phi$) and the polarisation angle $\psi$ according to
+The principal axes are in turn specified by the location of the GW source on the sky (via colatitude $\theta$ and azimuth $\phi$) and the polarisation angle $\psi$ according to
 \begin{align}
@@
 In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
-They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio pulse.
-The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e.
-$\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $.
-One obtains \citep[e.g.][]{Maggiore}
-\begin{equation}
- \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
+They are described by the geometric object $\boldsymbol{p}$ which we can identify  as the 4-momentum of the radio photon.
+The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e.
+$\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
+\begin{equation}
+ \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
 	\label{eq:momentum_shift}
 \end{equation}
-where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer.
-We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
+where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
 We take the pulsar location to be constant i.e.
-neither $\boldsymbol{q}$ nor $d$ are functions of time.
+neither $\boldsymbol{q}$ nor $d$ is a function of time.
 In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB.
-The barycentering correction is typically applied when generating TOAs, e.g.
-with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
-Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.
-Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$.
-After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames,
-\begin{equation}
-	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
-\end{equation}
-Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
-\begin{equation}
-	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d^{(n)}) g^{(n)}(t) +  \varepsilon^{(n)}\ .
+This barycentering correction is typically applied when generating TOAs using e.g.
+{\sc tempo2} \citep{tempo2} and related timing software.
+Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression:
+\begin{equation}
+	f = p_{\alpha} u^{\alpha} \ . 
+	\label{eq:freq_temporal}
+\end{equation}
+Due to the kinematical corrections from the barycentering process, 
+\begin{equation}
+	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0) \ ,
+\end{equation}
+where the perturbations to $\boldsymbol{u}$ from the GW are of higher order and neglected.
+In this case the shift in the momentum from Equation \eqref{eq:momentum_shift} can be related to a shift in the frequency as,
+\begin{equation}
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t) \ ,
 	\label{eq:measurement}
 \end{equation}
-where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies: 
-\begin{align}
-	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'}(t - t') \ .	\label{eq:vareps}
-\end{align}
-where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
-\begin{equation}
-	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
+where
+\begin{equation}
+	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i,(n)} q^{j,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right) \ .
 	\label{eq:g_func}
 \end{equation}
-and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$.
+and $q^{i,(n)}$ labels the $i$-th coordinate component of the of the $n$-th pulsars position vector $\boldsymbol{q}^{(n)}$.
 It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
 \begin{align}
-	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
-	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
-	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
+	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^{i,(n)} q^{j,(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
+	& \times \Big\{ \cos\left[-\Omega t +\Phi_0\right] \nonumber \\
+	&- \cos \left[-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right] \Big\} \ .
 	\label{eq:g_func_trig}
 \end{align}
-Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.
+Equations \eqref{eq:measurement} -- \eqref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the frequency measured by an observer on Earth.
 
 \section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
-The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters controlling the intrinsic frequency evolution of the pulsars in the array and the GW source, i.e.
+The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar, and parameters of the GW source, i.e.
 \begin{equation}
@@
 \begin{equation}
-	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
+	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
 \end{equation}
@@
 \end{equation}
-where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively \footnote{$\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
-$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.}.
-In Equation \eqref{eq:params3} we reparameterize the two GW polarisation amplitudes, $h_{+}$ and $h_{\times}$, in terms of $\iota$ and the characteristic wave strain $h_0$ through
+where $\delta$ is the declination of the GW source and $\alpha$ the right ascension.
+In Equation \eqref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
 \begin{align}
 	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
-	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ .
+	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ , 
 \end{align}
-We use the parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
-A PTA containing $N$ pulsars comprises $7 + 5N$ parameters to estimate.
-Typically the pulsar parameters are constrained better \textit{a priori} by electromagnetic observations than the GW parameters.
-For example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information about $\delta$ and $\alpha$.
+where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e.
+$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.
+We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
+For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate.
+Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.
 \newline
@@
 In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter.
-In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference.
-Model selection and the specification of the null model are described in Section \ref{sec:model_selection}.
-A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
-The method complements traditional PTA analyses because it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations.
-
-\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
+In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference of the model parameters and calculate the model evidence.
+Model selection and the specification of the null model is described in Section \ref{sec:model_selection}.
+A complete summary of the method is presented in Section \ref{sec:methodsummary}.
+
+\subsection{Kalman filter}\label{sec:kalman_filter}
+
+%
 %\begin{figure*}
-%	%\centering % Not needed
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
-%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
-%		\label{fig:6MB_BFS}
-%	\end{subfigure}  
-%    \hfill
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
-%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
-%		\label{fig:25MB_bfs}
-%	\end{subfigure}
-%	\caption{
-%		}
-%	\label{fig:four figures}
+%	\centering
+%	
+%	\subfloat[a]{\includegraphics[width=0.5\textwidth]{images/Kalman_example_both}} 
+%	\subfloat[b]{\includegraphics[width=0.3\textwidth]{images/Kalman_example_both}}
+%	\caption{\textcolor{red}{TK: For consideration. Do we prefer this figure to Figure \ref{fig:four figures}}?}
+%	\label{fig:kalmanexampleboth}
 %\end{figure*}
+%
 
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
-	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed systems state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are identically correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column).
-The top panels show the true pulsar state $f_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$  (green curve).
-We plot the subtracted state to better illustrate the stochastic wandering of the pulsar frequency.
-In the left hand column, the blue/green solutions are completely overlaid.
-The middle panels show the true measured frequency $f_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve).
-Again we plot the subtracted state, and in the left hand column the red/magenta solutions are completely overlaid.
-The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{1}(t) - \hat{f}_{\rm m}^{1}(t)$.
-The results are shown for a single pulsar.
-In the right hand column $\Omega$ has been perturbed from its true value by 20\%}
-	\label{fig:kalman_example}
+	%\centering % Not needed
+	\begin{subfigure}[b]{0.49\textwidth}
+		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
+		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
+		\label{fig:6MB_BFS}
+	\end{subfigure}  
+    \hfill
+	\begin{subfigure}[b]{0.49\textwidth}
+		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
+		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
+		\label{fig:25MB_bfs}
+	\end{subfigure}
+	\caption{Example application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$.
+We plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency.
+In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ and the state is estimated accurately.
+In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$ and the state is not estimated accurately.
+The top panels show the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line).
+The middle panels show the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta).
+The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$.
+In (a) the green/blue and red/magenta lines are completely overlaid.}
+	\label{fig:four figures}
 \end{figure*}
 
-The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$.
-It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
-In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
-Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
-Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
-Hence inferring the static parameters is a non-linear exercise, to be tackled separately after the linear Kalman filter operates on the data for a fixed set of static parameters.
-Inference of the static parameters in Equations \eqref{eq:measurement} and \eqref{eq:g_func} by nested sampling is discussed in Section \ref{sec:nested_sampling}.
-\newline
-
-Implementation of the linear Kalman filter for the PTA state-space model in Section 3.2, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
-At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.
-The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by converting ${\boldsymbol{\hat X}}_i$ into predicted measurements ${\boldsymbol{\hat Y}}_i$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func} and comparing with the actual, noisy measurements ${\boldsymbol{\hat Y}}_i$.
-This defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation.
-The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$.
-The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}.
-For a fixed set of static parameters, the Kalman filter returns an estimate of the the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error.
-We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline
-
-The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ can  then be calculated at each timestep from the Kalman filter output according to
+The Kalman filter \citep{Kalman1} is a Gaussian-Markov model used to algorithmically recover the most likely set of unknown system state variables, $\boldsymbol{X}$, given some noisy measurements, $\boldsymbol{Y}$.
+It is a common technique in signal processing that has been applied successfully in neutron star astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}.
+In this work we use the linear Kalman filter which assumes a linear relation between $\boldsymbol{X}$ and $\boldsymbol{Y}$.
+Whilst the measurement equation, Equation \eqref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively.
+Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, the unscented Kalman filter \citep{882463van} or the particle filter \citep{Simon10}.
+\newline
+
+The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}.
+At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$.
+The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Equation \eqref{eq:measurement}.
+The measurement predictions can then be compared against the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation".
+The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$.
+The Gaussian log-likelihood can then be calculated at each timestep:
 \begin{eqnarray}
-	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ .
+	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ ,
 \end{eqnarray}
-The total log-likelihood for the entire sequence is
+with the total log-likelihood simply the sum over all timesteps, i.e. 
 \begin{eqnarray}
-	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
+	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i \ . \label{eq:likelihood}
 \end{eqnarray}
-Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
-$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$.
-Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$.
-If $\boldsymbol{\hat{\theta}}$ is close to the true underlying parameters $\boldsymbol{\theta}$, then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are minimized and $\mathcal{L}$ is maximised.
-This is illustrated with synthetic data in Figure \ref{fig:kalman_example}.
-In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations (1)--(4), (16), and (17) for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$.
-The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve).
-The predicted state $\hat{f}_{\rm p}^{(1)}(t)$ is converted  into a predicted measurement $\hat{f}_{\rm m}^{(1)}(t)$ (middle panel, magenta curve), which again overlaps almost perfectly with the true measurement.
-The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel).
-By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters to the Kalman filter, where $\Omega$ has been perturbed from its true value by $20 \%$.
-In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates.
-It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel).
-
-%n Figure ; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
+For fixed data $\boldsymbol{Y}$, $\mathcal{L}$ is a function of the static parameters of the model, i.e.
+$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})$.
+Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are also functions of $\boldsymbol{\theta}$.
+If the estimates of the system parameters, $\boldsymbol{\hat{\theta}}$ , that we pass to the Kalman filter are close to the true underlying parameters then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are	 minimized and $\mathcal{L}$ is maximised.
+This is illustrated in Figure \ref{fig:6MB_BFS}; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
 The residuals correspond to random noise and are normally distributed.
@@
 
-\subsection{Nested sampling}\label{sec:nested_sampling}
+\subsection{Nested Sampling}\label{sec:nested_sampling}
 We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
@@
 \end{equation}
- We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper.
-Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, which also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
-It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
-The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies.
-Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}.
-For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
+In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work.
+Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
+The primary advantage of nested sampling is the ability to compute the evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches.
+Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}.
+For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
 For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}.
@@
 \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}.
-In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library.
+For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library.
 We  follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work.
-\newline
-
-The primary tunable parameter in nested sampling is $n_{\rm live}$.
-A greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
-However the computational runtime scales as $\mathcal{O}(n_{\rm live})$.
-\cite{Ashton2022} offered a rule-of-thumb trade-off, where the minimum number of live points should be greater than the number of static parameters.
-Informal empirical tests conducted as part of this paper support the trade-off suggested by \cite{Ashton2022}; we find typically that the true ${\boldsymbol{\theta}}$ is contained within the 90\% credible interval of the derived 1D marginalised posteriors of ${\boldsymbol{\hat{\theta}}}$ for $n_{\rm live} > 7 + 5N$ with $N \leq 50$.
+The primary tunable parameter in nested sampling is $n_{\rm live}$, where a greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}(1/\sqrt{n_{\rm live}})$.
+However the computational runtime scales as $\mathcal{O}(n_{\rm live})$ and so one must make a trade-off between uncertainty and runtime.
+\cite{Ashton2022} offer a rule-of-thumb where the minimum number of live points should be greater than the number of static parameters.
+The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$.
 Unless stated otherwise we take $n_{\rm live} = 1000$ for all results presented in this work.
@@
 \subsection{Model selection}\label{sec:model_selection}
-The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
+The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}_i$.
+This enables us to compare competing models via a Bayes factor,
 \begin{equation}
@@
 \end{equation}
-Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
-$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
+Throughout this work we take $\mathcal{M}_1$ to be the state space model that includes the presence of a GW.
+$\mathcal{M}_0$ is our null model that assumes there is no GW in the data.
 This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
@@
 \subsection{Summary of workflow}\label{sec:methodsummary}
-For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
-\begin{enumerate}[leftmargin=2em]
-	\item Specify a PTA composed of $N$ pulsars 
-	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
-	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$. 
-	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
-	 \item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
-	 \item For each live point:
-\begin{enumerate}[leftmargin=2em]
-	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
-	\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
+To summarise the above, the order of operations in our method for a single iteration (i.e. for a single live point) is as follows:
+\begin{enumerate}
+	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\boldsymbol{\theta}_{\rm sample}$
+    \item Pass $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
+	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
+	\item Return $\log \mathcal{L}$ to the nested sampler
+    \item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$
 \end{enumerate}
-	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
-	\item Constrained sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood $\log \mathcal{L}_{\rm new}$ > $\log \mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (a)--(b).
-	\item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ 
-	\item Repeat steps (vii) - (ix) until convergence criteria are satisfied.
-\end{enumerate}
-In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$.
-The resulting $\mathcal{Z}$ values can then be compared.
-We remind the reader that the above workflow differs from a realistic PTA analysis in on important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
-The generalization to TOAs is subtle and will be tackled in a forthcoming paper.
-
-\section{Validation procedure with synthetic data} \label{sec:testing}
-In this section we outline how synthetic data is generated in order to validate the analysis scheme in Section \ref{sec:detect}.
-In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
-In Section \ref{sec:gendata} we demonstrate how to solve the equations of Section \ref{sec:model} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.
-
-\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
+In Section \ref{sec:testing} we apply this workflow on a synthetic data set.
+
+\section{Tests with synthetic data} \label{sec:testing}
+In this section we use synthetic data to test the ability of our method to detect the presence of a GW in noisy data and recover the static parameters, $\boldsymbol{\theta}$.
+In Section \ref{sec:synth_data} we discuss how the synthetic data is generated, including the construction of an artificial PTA.
+In Section \ref{sec:rep_example} we generate synthetic data for a single representative example SMBHB system and apply our method to estimate $\boldsymbol{\theta}$ and calculate the statistical evidence for a GW in the data via the Bayes factor.
+In this section we also explore how the detectability of the system changes for different strain magnitudes, $h_0$ and the performance of the method for different realisations of the pulsar process noise and the measurement noise.
+In Section \ref{sec:parameter_space} we move beyond a single example system and test our method across a broader parameter space.
+In Section \ref{sec:bias_and_identifiability} we discuss biases that are evident in some of the parameter estimates.
+
+\subsection{Synthetic data generation} \label{sec:synth_data}
+
+\subsubsection{Constructing a synthetic PTA}\label{sec:synt_pta}
+
 \begin{figure}
 	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
-	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
+	\caption{Spatial distribution of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
 The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc.
@@
 \end{figure}
-We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
-The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
-For each pulsar we adopt the fiducial NANOGrav values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
-A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility.
-The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}.
-The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
-The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist.
-The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
-In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
-We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$.
-The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
+For this work we generate synthetic data that is astrophysically representative as follows.
+We first specify a set of pulsars to make up our artificial PTA.
+We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}.
+The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA and the method naturally extends to any other PTA.
+For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{(n)}$), the distance, the barycentric rotation frequency (which we identify with $f_{\rm em}^{(n)} (t_1)$), and the time derivative of the the barycentric rotation frequency (which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$).
+The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Figure \ref{fig:pulsar_distrib}.
+The data is acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$.
+We discussed in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$.
+Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars.
+For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
+In order to set a physically reasonable value for $[\sigma^{(n)}]^2$ we use two complementary approaches.
+The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by:
 \begin{align}
-	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
-	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
+	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)&= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
+	&+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
 \end{align}
-where $T_{\rm cad}$ is the cadence of the timing observations and MSPs have $\ln \alpha_1 \approx -20 $, $\alpha_2 \approx 1$, $\alpha_3 \approx 2$ and $\alpha_4 \approx 2.4$.
-Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
-In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
+where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$.
+Throughout this work we assume that all pulsars are observed with a weekly cadence.
+In order to relate the empirical TOA noise model, Equation \eqref{eq:sigmap}, to our Ornstein-Uhlenbeck frequency model, Equation \eqref{eq:spinevol}, we set first $T = 1$ week in Equation \ref{eq:sigmap} to obtain an estimate of the TOA noise over a weekly time interval, i.e.
+the size of our numerical integration timestep.
+The frequency process noise parameter $\sigma^{(n)}$ can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as
 \begin{eqnarray}
-	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-1} \ . \label{eq:sigmap_f}
+	\sigma^{(n)} \sim \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
 \end{eqnarray}
-For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.3 \times 10^{-23}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.0 \times 10^{-16}$ s$^{-3/2}$ for PSR J1939+2134 \newline
-
-As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
+For our synthetic NANOGrav PTA, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s. \newline
+
+As a sanity check, we can also calculate $\sigma^{(n)}$ using a complementary numerical approach to the empirical best-fit model outlined above.
+We can directly solve the state equation Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a synthetic phase solution,
 \begin{eqnarray}
-	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
+	\xi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
 \end{eqnarray}
-and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach.
-We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} conducted with the Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}.
-A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$.
-In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$.
-In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J0030+0451, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}.
-We see that the synthetic and empirical residuals are qualitatively similar.
-Comparable behaviour is observed for other pulsars in our array.
+and generate TOAs and phase residuals.
+We calibrate the noise amplitude $\sigma^{(n)}$  to generate phase residuals that qualitatively resemble empirical phase residuals measured from real pulsars.
+We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} from the  Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}.
+For the purposes of validating the $\sigma^{(n)}$ inferred from Equation \eqref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually.
+Whilst it would also be possible to e.g.
+define some quantitative loss function (e.g.
+mean-squared error) between the two solutions and solve the optimisation problem to infer $\sigma^{(n)}$, for our purposes this is excessive.
+It is satisfactory to simply qualitatively compare the two solutions as a useful sanity check.
+We emphasise that we are not overly concerned with calculating the maximally accurate values for $\sigma^{(n)}$, but instead some reasonable values for constructing representative synthetic data.
+This approach of generating synthetic phase solutions and visually comparing with empirical solutions is also followed in \cite{Vargas}.
+In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for the NANOGrav pulsar J0030+0451.
+We can see that the synthetic and empirical residuals are qualitatively similar, and comparable behaviour is observed for other pulsars in our array.
 \begin{figure}
-	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
-	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J0030+0451.
-The actual residuals are sourced from the UTMOST pulsar timing program \citep{UTMOST}.
+	\includegraphics[width=\columnwidth]{images/example_residuals_plot}
+	\caption{Actual (left panels) and synthetic (right panels) phase residuals for NANOGrav millisecond pulsar J0030+0451.
+The actual residuals are obtained via UTMOST pulsar timing program \citep{UTMOST}.
 The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
-The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}.
-Following \citet{Vargas} we set the TOA uncertainty to be a constant $10^{-4}$ s.
-The blue and orange residuals are qualitatively similar by inspection.
-The other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} are tuned to produce comparable results.}
+For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals.
+}
 	\label{fig:qualitative_compare}
 \end{figure}
-\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
-We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
-\begin{enumerate}[leftmargin=2em]
-	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
-	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}
+\subsubsection{Synthetic data generation}
+We generate $N$ synthetic noisy timeseries, one for each pulsar, as follows:
+
+\begin{enumerate}
+	\item Integrate the state equations, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}.
+	\item Map from state space to measurement space via the measurement equation, Equation \eqref{eq:measurement}
+	\item Add zero-mean Gaussian measurement noise to each timeseries, i.e.
+$f_{\rm m} ^{(n)} + \varepsilon^{(n)}$ where 	$\langle \varepsilon(t) \varepsilon(t') \rangle = \sigma_{\rm m}^2 \delta(t - t')$ for measurement noise covariance $\sigma_{\rm m}^2$.
 \end{enumerate}
 Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
-The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \eqref{sec:synt_pta}.
+The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified by our construction of the synthetic PTA outlined in Section \eqref{sec:synt_pta}.
 For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
-The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
-\begin{equation}
-	\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
-\end{equation}
-Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
-The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
-Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
-Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar.
-\newline
-
-Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } \sigma^{(n)} \Delta W_t \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta W_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
-To fix the problem, we subtract the deterministic frequency evolution and track the new variable, equivalent to a change of variables:
-\begin{equation}
-	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ .
-\end{equation}  
-We similarly modify the measurement variable to be
-\begin{equation}
-	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ ,
-\end{equation}
-where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris which is measured to high accuracy in practice by {\sc tempo2}.
-For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately.
-Equation \eqref{eq:measurement} is then updated to read
-\begin{equation}
-	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
+The measurement noise can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
+\begin{equation}
+	\sigma_{\rm m} \sim f_{\rm p}^{(n)} \frac{\sigma_{\rm TOA}}{\text{T}} \ . 
+\end{equation}
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$.
+Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it to be known \textit{a priori} rather than a parameter to be inferred.
+Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise; whilst for our construction every pulsar is sampled at the same time and so would experience the same realisation of the noise, in actual pulsar astronomy different pulsars will be sampled at different times by different radio interferometers and so subject to different noise realisations.
+\newline
+
+In practice, in order to avoid numerical precision issues that arise when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
+\begin{equation}
+	f_{\rm p}^{\prime (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ .
+\end{equation}  
+We similarly heterodyne the measurement variable 
+\begin{equation}
+	f_{\rm m}^{\prime (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{*(n)} \ ,
+\end{equation}
+where $ f_{\rm em}^{*(n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris.
+For synthetic data we can set $ f_{\rm em}^{*(n)} = f_{\rm em}^{(n)}$ but this is not true generally for real-world observations since the spin-down ephemeris is only known approximately.
+The measurement equation Equation \eqref{eq:measurement} that relates the state and measurement variables is then updated as
+\begin{equation}
+	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
 	\label{eq:measurement_cov}
 \end{equation}
-We emphasise that this change of variables is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g.
+We emphasise that this change of variables is simply a convenient scaling to bring our numerical values into a reasonable dynamic range without having to use extended numerical floating point formats (e.g.
 long double, quadruple).
-It does not remove any degrees of freedom nor does it involve an approximation.
-In particular $f_{\rm em}^{(n)}(t_1)$
-and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \ref{eq:spinevol}--\eqref{eq:xieqn}.
-
-\section{Representative example: single SMBHB source}\label{sec:rep_example}
-In this Section we apply the analysis scheme of Section \ref{sec:detect} and the validation procedure of Section \ref{sec:testing} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-We remind the reader that the analysis in this paper targets a single SMBHB source.
-The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-In Section \ref{sec:priors} we outline the priors chosen on $\boldsymbol{\theta}$.
-In Section \ref{sec:parameter_estim} we apply the workflow of Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
-We go on in Section \ref{sec:multiple_noise} to extend this parameter estimation procedure to multiple realisations of the noise processes.
-In Section \ref{sec:detection} we calculate the the detectability of the source as a function of $h_0$.
-\newline
-
+It does not remove any degrees of freedom of the formulation.
+In particular both $f_{\rm em}^{(n)}(t_1)$
+and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain static parameters of the model, but now appear in the measurement equation rather than the state equation.
+
+\subsection{Representative example}\label{sec:rep_example}
+We initially characterise our method with a single example of a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
 The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected to be astrophysically reasonable and representative.
@@
 
-\subsection{Prior distributions}\label{sec:priors}
+\subsubsection{Prior distributions}\label{sec:priors}
 For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters.
@@
 set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
-Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-3/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
-For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-3/2}$).
-The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-3/2}$.
-For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-3/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
+Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-5/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
+For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-5/2}$).
+The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-5/2}$.
+For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-5/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
 \newline
@@
 		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
-		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
+		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
 		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
@@
 
-\subsection{Parameter estimation}\label{sec:parameter_estim}
-In this section we derive the posterior probability distributions for the static parameters.
-However, it is first necessary to discuss a modification to the general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}.
-It can be seen that the equation generally separates into two cosine terms.
+\subsubsection{Parameter estimation}\label{sec:parameter_estim}
+From Equation \eqref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms.
 The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
@@
 	\begin{equation}
-		g^{(n)}_{\rm Earth}(t) = 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \cos(-\Omega t +\Phi_0)  \ .
+		g^{(n)}_{\rm Earth}(t) = 1 - \frac{1}{2} \frac{ H_{ij}q^{i,(n)} q^{j,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \cos(-\Omega t +\Phi_0)  \ .
 		\label{eq:g_func_trig_earth}
 	\end{equation}
-	We defer the inclusion of the pulsar terms to a future work.
+	We defer the inclusion of the pulsar terms in the measurement equation to a future work.
 As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
@@
 The generated synthetic data that we use to test our method does include the pulsar term in full.
-\newline
-
 Initially we consider a single noise realisation of the synthetic data for the representative example system described in Table \ref{tab:parameters_and_priors}.
-We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
+We apply the Kalman filter using the Earth terms only measurement equation, Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
+\newline
+
  The results are shown in Figure \ref{fig:corner_plot_1} for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$.
-The histograms shown in the figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
+The histograms shown in the Figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
 The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
-It is evident that for this representative example with characteristic strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
+It is evident that for this representative example with strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
 \newline
@@
 a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation.
-We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
+We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.
 Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
@@
 The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
 We are able to accurately estimate each parameter of interest.
@@
 
-\subsection{Multiple noise realisations} \label{sec:multiple_noise}
+\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
 	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes.
-The subtitles of the marginalized 1D posteriors refer to values for the median posterior across the noise realisations.
 The injection value is generally contained within the 90 \% credible interval for the vast of majority posteriors.
@@
 1000 is a hangover when we were using a larger strain ($h=10^{-12}$) as our canonical example.
-The $10^3$ noise realisations are currently running and plots will be updated once completed, but I don't expect any meaningful changes \textcolor{red}{; for $h=10^{-12}$ the results/conclusions of 100 realisations are the same as 1000 realisations)}}\normalsize} realisations of the process noise and measurement noise.
+The $10^3$ noise realisations are currently running and plots will be updated once completed, but I don't expect any meaningful changes \textcolor{red}{;for $h=10^{-12}$ the results/conclusions of 100 realisations are the same as 1000 realisations)}}\normalsize} realisations of the process noise and measurement noise.
 For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
@@
 \end{eqnarray}
-where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.
-The first moment of the WD has a convenient physical interpretation given by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
+where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.  \newline
+
+For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
+The results are presented in Figure \ref{fig:pairwise_wasserstein} which shows the $W_1$ value for each pair of posteriors.
+The first moment of the WD has a useful physical interpretation given by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
 \begin{eqnarray}
@@
 We can see that $W_1$ bounds the difference in the expected values.
-Taking a concrete example, suppose that we infer two one-dimensional posterior distributions for $\iota$, for two different realisations of the noise, and calculate $W_1 =0.5$ rad between them.
-This means that the difference in the expected value of two random variables drawn from the two posteriors is no greater than $0.5$ rad.
-\newline
-
-For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
-The results are presented in Figure \ref{fig:pairwise_wasserstein}.
-The figure shows a lower triangular heat-map where each point denotes the $W_1$ value between a pair of posteriors.
-Small values of $W_1$ are magenta moving to red for large values of $W_1$.
-Note that the heat-map colour scaling is not the same for all subplots of the figure.
-It is evident that $W_1$ is generally small across all pairs of posteriors.
+That is to say, if $W_1 =0.5$ rad between two one-dimensional posterior distributions estimated for e.g.
+$\iota$, this means that the difference in the expected values in random variables drawn from these distributions is no greater than $0.5$ rad.
+From Figure \ref{fig:pairwise_wasserstein} it is evident that $W_1$ is generally small across all pairs of posteriors.
 We define $W_{1, \rm median}$ as the median value of $W_1$ across all $10^3 \choose 2$ pairs of noise realisations.
@@
 For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the figure, corresponds to a posterior for $\psi$ which has converged sub-optimally, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
-The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ radians.
+The $W_1$ between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ radians.
 We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
@@
 
-\subsection{Detection} \label{sec:detection}
+\subsubsection{Detection} \label{sec:detection}
 We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
@@
 
-There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
+There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10{-15}$.
 The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
@@
 
-\section{Exploring a broader parameter space} \label{sec:parameter_space}
+\subsection{Exploring a broader parameter space} \label{sec:parameter_space}
 \begin{figure}
@@
 We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
-In this section we test the method in different regions of the parameter space, varying the source parameters through astrophysically relevant ranges.
-\newline
-
-We consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
+It is also important to test the method in different regions of the parameter space.
+To this end we consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
 We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space.
@@
 
-\section{Identifiability and bias} \label{sec:bias_and_identifiability}
+\subsection{Identifiability and bias} \label{sec:bias_and_identifiability}
 There are two underlying issues in the preceding tests that we now explore in more detail.
@@
 can we uniquely identify the parameters of the model given the data and the model structure?
-This question is explored in Section \ref{sec:identif}.
 The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
-This is the focus of Section \ref{sec:bias}.
 In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f.
@@
 
-\subsection{Identifiability}\label{sec:identif}
+\subsubsection{Identifiability}
 \begin{figure}
@@
 In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations.
-The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but just for $\iota$ and $h_0$ rather than all members of $\boldsymbol{\theta}_{\rm gw}$.
 In the case of high-SNR we might expect \textit{a priori} that our estimates would be increasingly accurate over the low-SNR case, and that the method would converge robustly to similar posteriors irrespective of the noise realisation (i.e.
@@
 This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be now be discussed further in Section \ref{sec:bias}.
-\subsection{Bias by dropping pulsar terms}\label{sec:bias}
+\subsubsection{Bias by dropping pulsar terms}\label{sec:bias}
 
@@
 
-\section{Discussion}\label{sec:discussion}
+\section{Discussion}
 We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
@@
 %https://arxiv.org/pdf/2107.03047.pdf
-\section{Conclusion}\label{sec:conclusion}
+\section{Conclusion}
 In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
@@
 \section{Kalman recursion equations} \label{sec:kalman}
-The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
-\begin{equation}
-	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
-\end{equation}
-where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance, and the subscript $k$ labels the time-step.
+The linear Kalman filter operate on temporally discrete, noisy measurements $\boldsymbol{Y}_k$ which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$ via a linear transformation:
+\begin{equation}
+	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \label{eq:kalman1}
+\end{equation}
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
 The Kalman filter takes the underlying states to evolve according to
 \begin{equation}
-	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
-\end{equation}
-where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix.
-$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$.
-\newline
-
-The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage.
-The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
-Specifically, the predict step proceeds as,
+	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \label{eq:kalman2}
+\end{equation}
+where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the input or control matrix.
+and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$.
+\newline
+
+The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage where the state at the current time-step is predicted based on the state at the previous time-step, and an ``update" stage where the predicted state estimate is refined based on the measurement at that time-step.
+The predict stage updates the estimate of the state $\hat{\boldsymbol{X}}_{k|k-1}$ and the associated covariance $\hat{\boldsymbol{P}}_{k|k-1}$where the subscript notation ${k|k-1}$ denotes an estimate at discrete step $k$, given measurements from step $k-1$ and earlier.
+The predict step proceeds as,
 \begin{equation}
@@
 \end{equation}
-Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
+The update step then uses the measurement $\boldsymbol{Y}_k$ to update $\hat{\boldsymbol{X}}_{k|k}$  and $\hat{\boldsymbol{P}}_{k|k}$ as,
 \begin{equation}
@@
 \begin{equation}
-	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
+	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}
 \end{equation}
@@
 \end{equation}
-Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
-$\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$.
-For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
-\newline
-
-We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
-\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
-\begin{equation}
-	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
-\end{equation}
-Similarly,  
-\begin{equation}
-\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
-\end{equation}
-The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
+Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which yields the minimal mean square error in the measurement post fit residual, $\boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k}$.
+\newline
+
+We can map our state-space model described in Section \ref{sec:model} onto the Kalman filter structure of Equations  
+\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows.
+We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e $\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right)$.
+Similarly  $\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)$.
+The states evolve according the continuous dynamical equation
 \begin{equation}
@@
 \end{equation}
-where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix
-\begin{equation}
-\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
-\end{equation}
-and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
-\begin{equation}
- C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
-\end{equation}
-The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
-\begin{equation}
-\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)
-\end{equation}
-\newline
-
-The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
+where $\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right)$ is the state matrix, $\boldsymbol{C} = \left(C^{(1)}, C^{(2)}, ..., C^{(N)}\right)$ the input or control matrix where $C^{(i)} =\gamma^{(i)} \left(f_{\rm em}^{(i)} (0) + \dot{f}_{\rm em}(0)^{(i)} t \right) +  \dot{f}_{\rm em}(0)^{(i)}$ and $d\boldsymbol{B}(t)$ denotes a Wiener process with covariance matrix $\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)$.
+Equation \ref{eq:kalmn2} is a Langevin equation (equivalently an Ornstein-Uhlenbeck process) which has a general solution given by \citep{gardiner2009stochastic},
 \begin{equation}
@@
 \end{equation} 
-From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
+From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ as Equation \eqref{eq:kalman2} where
 \begin{equation}
@@
 \end{equation}
-with
-\begin{align}
-	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
-	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
-\end{align}
+where
+\begin{equation}
+	G_k^{(n)} =    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] - e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
+\end{equation}
 and
 \begin{equation}
-	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) 
+	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1), Q^{(2)}},...,Q^{(N)}\right) 
 \end{equation}
@@
 
-The remaining component matrices of the Kalman filter that we have not yet specified are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
+The remaining component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
 These are defined straightforwardly from Equation \eqref{eq:measurement}.
@@
 $\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
-The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
-
-\section{Fiducial pulsar values}\label{appendix_fiducial}
-\textcolor{red}{TK TBD} 
+$\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
+
 \section{Identifiability}\label{appendix_identifiability}

### 2023-10-06 manuscript/manuscript_2/paper.tex
introduction to P2

--- 2023-09-26 manuscript/manuscript2/paper.tex
@@
 \section{Introduction}\label{sec:intro}
-The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022} is predicted to emit nHz gravitational waves (GWs).
-Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
-The detection of nHz GWs has inspired the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
-The foremost method is timing an ensemble of pulsars, i.e.
-a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}.
-A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth.
-By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs.
-Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA,][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}.
-These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep[CPTA,][]{Hobbs_2019}.
-\newline
-
-The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
-Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
-After multiple non-detections \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A} consilient evidence for the GW background was presented by NANOGrav \citep{2023ApJ...951L...8A}, EPTA/InPTA \citep{2023arXiv230616214A}, PPTA \citep{2023ApJ...951L...6R} and the CPTA \citep{2023RAA....23g5024X}.
-\newline
-
-Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the early stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
-Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}.
-Individual SMBHBs are continuous wave sources; they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals.
-Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs.
-However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}.
-\newline
-
-Intrinsic pulsar timing noise -- i.e.
-random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}.
-This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
+Nanohertz gravitational waves (GWs) can be produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}.
+A nHz GW influences the times of arrival (TOAs) of radio pulses from pulsars measured at the Earth.
+A pulsar timing array \citep[PTA][]{Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars.
+In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers.
+\newline
+
+ PTAs search for two types of SMBHB GW signals.
+The first is a stochastic GW background due to the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
+Consilient evidence for the GW background has been presented by multiple separate PTAs \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X}, by measuring the cross-correlation between pairs of pulsars in the pulsar timing residuals as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
+The second is resolved GWs from sufficiently close and massive individual SMBHBs \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
+Individual SMBHBs are sources of continuous GWs.
+This results in persistent, quasi-monochromatic modulations in the pulsar timing residuals which can be modelled.
+Therefore, rather than cross-correlating pulsar pairs, individual SMBHBs are detected by either frequentist matched filtering \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or Bayesian inference \citep{Ellis2016,Arzoumanian2020A}.
+Observational efforts to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, although some inconclusive evidence at low significance has been presented \citep{2023arXiv230616226A}.
+\newline
+
+ Irregularities in the intrinsic rotation of the pulsar result in random, unmodelled, red-spectrum TOA fluctuations known as `timing noise'.
+Timing noise limits the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}.
+Additional white noise sources such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021} must also be considered.
+\cite{K1} (K23 hereafter) presented an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations.
+Specifically, K23 formulate PTA analysis as a state-space problem and use a Kalman filter \citep{Kalman1} to track the pulsars' intrinsic rotational state in conjunction with a Bayesian nested sampler \citep{Skilling, Ashton2022} for parameter estimation and model selection.
+\newline
+
+ It is common in standard PTA analysis to drop the so called "pulsar term" from the modelling \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+Unlike the "Earth term", the pulsar term is not phase coherent and cannot be summed across the array to increase the total signal-to-noise.
+Instead it is considered as a source of self noise.
+However, dropping the pulsar terms results in a reduced detection probability ($\sim 5 \%$) bias in the parameter estimates \citep{Zhupulsarterms,Chen2022}.
+The method presented in K23 also drops the pulsar term.
+In this paper we remove this limitation and present the state-space approach to PTA data analysis including the pulsar term.
+\newline
+
+ This paper is organised as follows.
+In Section \ref{sec:model} we briefly review the state-space formulation of PTA data analysis introduced in \cite{K1}.
+We do not review the methods for state tracking, parameter inference and model selection and refer the reader to K23.
+In Section \ref{sec:pulsar_term} we introduce how to include the pulsar term in the state-space model.
+In Section \ref{sec:rep_example1} we test the updated method on the synthetic data for a single representative example GW system, and compare the results to the original Earth-terms model.
+In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space.
+Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.
+Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$.
+\newline
+
+% Could include the below block if we need more fodder
+
+%This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
 In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators.
 However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
-In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red).
-In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
-\newline
-
-In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations.
-The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density.
-Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
-The new and existing approaches are therefore complementary.
-In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
-We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood (i.e.
-the model evidence) for model selection.
-\newline
-
-% --- a promise which we aim to test as the key goal of this paper. 
-\noindent This paper is organised as follows.
-In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW.
-In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence.
-In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method.
-We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative example GW system.
-In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space.
-Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.
-The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}.
-It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization which is deferred to future work.
-Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$.
-\newline
 
 \section{State-Space Formulation}\label{sec:model}
-
-In this Section we briefly review the state-space formulation of PTA analysis presented in K1.
+In this section we briefly review the state-space formulation of PTA analysis presented in \citep{K1}.
 
@@
 
+\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
+
+A naive approach is challenging - look at this noisy likelihood curve
+
+A reparamterisation is good
+
+\section{Representative example: single SMBHB source and Comparison with Earth terms}\label{sec:rep_example1}
+
 \section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
@@
 %\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
-	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed systems state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are identically correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column).
-The top panels show the true pulsar state $f_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$  (green curve).
-We plot the subtracted state to better illustrate the stochastic wandering of the pulsar frequency.
-In the left hand column, the blue/green solutions are completely overlaid.
-The middle panels show the true measured frequency $f_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve).
-Again we plot the subtracted state, and in the left hand column the red/magenta solutions are completely overlaid.
-The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{1}(t) - \hat{f}_{\rm m}^{1}(t)$.
-The results are shown for a single pulsar.
-In the right hand column $\Omega$ has been perturbed from its true value by 20\%}
-	\label{fig:kalman_example}
-\end{figure*}
 
@@
 \subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
-	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
-The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc.
-The grey dashed line denotes the Galactic plane.}
-	\label{fig:pulsar_distrib}
-\end{figure}
+
 We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
+	\includegraphics[width=\textwidth, height =\textwidth ]{example-image-a}
 	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{images/pp_plot_new}
+	\includegraphics[width=\columnwidth]{example-image-a}
 	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}.
@@
 
-\subsection{Identifiability}\label{sec:identif}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
-	\caption{Posterior probability distributions for the representative system described in Table \ref{tab:parameters_and_priors} over nine noise realisations for the GW source parameters $\iota$ and $h_0$ in the high SNR regime ($h_0 = 10^{-12}$).
-Unlike the other parameters of  $\boldsymbol{\theta}_{\rm gw}$, the one-dimensional posteriors for $\iota$ and $h_0$ do not strongly overlap.
-A correlation between  $\iota$ and $h_0$ is evident.}
-	\label{fig:just_iota_h}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_surface}
-	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation.
-The red point labels the location of the likelihood maxima.
-The likelihood surface has been normalised with respect to the absolute value of the maxima.
-A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
-	\label{fig:likelihood_surface}
-\end{figure}
-In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations. 
-The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but just for $\iota$ and $h_0$ rather than all members of $\boldsymbol{\theta}_{\rm gw}$.
-In the case of high-SNR we might expect \textit{a priori} that our estimates would be increasingly accurate over the low-SNR case, and that the method would converge robustly to similar posteriors irrespective of the noise realisation (i.e.
-$W_1 \to 0$).
-This is generally true for the majority of the component parameters of $\boldsymbol{\theta}_{\rm gw}$; the estimated posteriors across noise realisations are largely indistinguishable.
-However for $\iota$ and $h_0$ this is not the case.
-Instead the inferred one-dimensional posteriors do not broadly overlap and are not consistent across different noise realisations.
-Specifically across $10^3$ noise realisations the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
-For $\iota$, $W_{1,\rm median}$ is greater than the value inferred in the low SNR case ($=0.14$, Table \ref{tab:Wasserstein}), whilst for $h_0$ $W_{1,\rm median}$ is also greater as a proportion of the scale set by the strain amplitude.
-This is in contrast to $W_{1,\rm median}$ inferred for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ in the high-SNR regime where as SNR increases, $W_1$ decreases; e.g.
-for $\Phi_0$, $W_{1, \rm median} = 4.5 \times 10^{-3}$ rad.
-\newline
-
-This issue of inferring very different posteriors depending on the particular realisation of the noise is due to these parameters being only weakly-identifiable.
-Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
-Sources of non-identifiability are generally categorised as either ``structural", arising from the structure of the model, or ``practical", arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
-Regarding $\iota$ and $h_0$, we are concerned with structural identifiability.
-Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx} and \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
-For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
-This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
-It can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
-That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a small, non-zero gradient close to the maximum likelihood solution.
-This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
-Whilst a single, unique maxima exists (labelled by the red point in the figure), along the ridge in the parameter space the likelihood values are all very similar.
-Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
-If during the inference we fix one of the parameters at its true value (i.e.
-set a delta function prior), taking a slice through the likelihood surface it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably, i.e.
-$W_1 \to 0$.
-In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" \footnote{\tiny \textcolor{red}{TK: "washes out" may be too colloquial / unspecific, but I think it is intuitive...}\normalsize}the likelihood ridge, leading to a more concave likelihood surface.
-This enables consistent posteriors to be estimated for $\iota$ and $h$.
-In the case where we fix e.g $h_0$ at its true value to infer indistinguishable posteriors for $\iota$, all of the posteriors for $\iota$ are all strongly biased away from the injected value; the maximum a posteriori probability estimates underestimate the injected value by $\sim 0.34$ rad.
-This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be now be discussed further in Section \ref{sec:bias}.
 \subsection{Bias by dropping pulsar terms}\label{sec:bias}
 
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
-	\caption{As Figure \ref{fig:corner_plot_1} but for the high-SNR system with $h_0 = 10^{-12}$.
-For the majority of parameters the injection value does not fall within the 90\% credible interval and is not visible over these scales.
-The estimated posteriors are biased away from the injected value due to the dropping of the pulsar terms.
-Note that the scale of the $x$-axes are generally highly constrained compared to Figure \ref{fig:corner_plot_1}.}
-	\label{fig:bias_for_large_h}
-\end{figure*}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
-	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and the Earth terms model for Earth terms only synthetic data (green line), with respect to the inclination $\iota$, over the prior space.
-All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
-The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model; the green/orange dashed lines are overlaid on this scale.
-For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$).
-For the pulsar-terms model the location of the maxima and injected value do coincide.
-10 noise realisations are plotted for each situation but the solutions are completely overlaid on this scale.}
-	\label{fig:likelihood_surface_iota}
-\end{figure}
 Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g.

### 2023-10-09 manuscript/manuscript_2/paper.tex
correct Bayes run with variable rates

@@
 
- It is common in standard PTA analysis to drop the so called "pulsar term" from the modelling \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+ It is common in standard PTA analysis to drop the so called ``pulsar term" from the modelling \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
 Unlike the "Earth term", the pulsar term is not phase coherent and cannot be summed across the array to increase the total signal-to-noise.
@@
 In Section \ref{sec:model} we briefly review the state-space formulation of PTA data analysis introduced in \cite{K1}.
-We do not review the methods for state tracking, parameter inference and model selection and refer the reader to K23.
+We do not review the methods for state tracking, parameter inference and model selection and refer the reader to K23 for details.
 In Section \ref{sec:pulsar_term} we introduce how to include the pulsar term in the state-space model.
@@
 \section{State-Space Formulation}\label{sec:model}
-In this section we briefly review the state-space formulation of PTA analysis presented in \citep{K1}.
-
-The intrinsic state variable is the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local,freely-falling rest frame of the pulsar's centre of mass.
-The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
-$f_{\rm p}^{(n)}(t)$ evolves according to a stochastic differential equation and is related to the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$ via a measurement equation.
-
-A phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.
-The measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_measured}.
+In this section we briefly review the state-space formulation of PTA analysis presented in \citep{K1}. \newline
+
+There are $N$ pulsars in the array.
+The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$.
+The intrinsic spin frequency of the $n$-th pulsar as measured in the local, freely-falling rest frame of the pulsar's centre of mass, $f_{\rm p}^{(n)}(t)$, evolves according to a stochastic differential equation.
+The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation.
+In Section \ref{sec:psr_frequency} we review the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$.
+In Section \ref{sec:psr_measured} we review the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$.
+In Section \ref{sec:ss_params} we summarise the parameters of the model.
 
 \subsection{Spin evolution} \label{sec:psr_frequency}
-
-In our model
-
-Equation X is inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend.
-At present, a predictive, first-principles theory of timing noise does not exist and there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}.
-
-In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque.
-The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
-Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution $f_{\rm em}^{(n)}(t)$ can be approximated accurately by
-\begin{equation}
- f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
-\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA.
-The stochastic torque is assumed to be a zero-mean, white noise process.
-Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
-\begin{equation}
-	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
+In the model $f_{\rm p}^{(n)}(t)$ evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
+\begin{equation}
+	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ ,
 	\label{eq:frequency_evolution}
 \end{equation}
-In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
+where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
 \begin{align}
@@
 \end{align}
-In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise.
-Combined with the mean reversion it gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
-It is important to note that white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
-\newline
-
-As discussed in K1
-Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
-Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
+In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
+PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and can be approximated accurately by
+\begin{equation}
+	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
+\end{equation} 
+where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline
+
+Equation \ref{eq:frequency_evolution} is inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
+Such a phenomenological model is necessary since a predictive, first-principles theory of timing noise does not currently exist c.f.
+the multiple theorized mechanisms of timing noise, see the discussion in Section 1 of K23.
+Similar approaches, analogous to Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
+However, Equation \eqref{eq:frequency_evolution} does contain certain modelling omissions and idealisations which much be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
+These include: exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}, the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs and the formal interpretation of $d^2 f_{\rm p} / dt^2$.
+
+%
+%However, as discussed in K23, First, Equation \ref{eq:frequency_evolution} excludes particular phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}.
+Second, the white spectrum assumed for $\xi(t)$ may or may not be suitable for millisecond pulsars in PTAs.
+It is challenging observationally to infer the spectrum of $\xi(t)$ from the observed spectrum of the phase residuals, because the inference is conditional on the (unknown) dynamical model governing $df_{\rm p}/dt$.
+For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless.
+White noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
+Third, $\xi(t)$ is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper.
+Finally, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, which have not been ruled out by pulsar timing experiments to date.
+The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \eqref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \eqref{eq:frequency_evolution}.
+
+%This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
+In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators.
+However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
+%
+%
+
+%
+%Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
 As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}.
-An approach akin to Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} has been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
-However, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}  involve significant idealizations, which must be recognized at the outset \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
+%
+%
+%
+%However, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}  involve significant idealizations, which must be recognized at the outset \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
 First, the white noise driver $\xi(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper.
@@
 \subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
-In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
-Specifically, the pulse frequency is modulated harmonically at the GW frequency.
-We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section.
-The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
-\subsubsection{Plane GW perturbation}\label{sec:plane_gw}
-We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
-\begin{equation}
-	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
-\end{equation}
-with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$.
-The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$.
-Throughout this paper we work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB).
-We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB.
-In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic.
-Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Zhu10},
-\begin{equation}
-	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
-	\label{eq:f_evolution}
-\end{equation}
-where $M_c$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
-A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
-From Equation \eqref{eq:f_evolution} we can see that only those binaries which are massive or at high frequency experience significant frequency evolution over typical PTA timespans.
-The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
-We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
-\newline
-
-The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$). The spatial part is
-\begin{align}
-	H_{ij} = h_+ e_{ij}^+(\boldsymbol{n}) + h_{\times} e_{ij}^{\times}(\boldsymbol{n}) \ , \label{eq:hij}
-\end{align}
-where $h_{+}$ and $h_{\times}$ are the respective polarisation amplitudes.
-The plus and cross polarisation tensors $e_{ij}^{+}$ and $e_{ij}^{\times}$ are uniquely defined by the principal axes of the wave, viz.\ the unit 3-vectors $\boldsymbol{k}$ and $\boldsymbol{l}$, according to
-\begin{align}
-	e_{i j}^{+}(\boldsymbol{n}) =k_i k_j-l_i l_j \ , \\
-		e_{i j}^{\times}(\boldsymbol{n}) =k_i l_j+l_i k_j \ .
-\end{align}
-The principal axes are in turn specified by the location of the GW source on the sky (colatitude $\theta$, azimuth $\phi$) and the polarisation angle $\psi$ according to
-\begin{align}
-	\boldsymbol{k}  = &(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
-	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}} \nonumber \\
-	& +(\sin \psi \sin \theta) \boldsymbol{\hat{z}} \ , \\
-	\boldsymbol{l} = &(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
-	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}}\nonumber  \\
-	& +(\cos \psi \sin \theta) \boldsymbol{\hat{z}} \ ,
-\end{align}
-where e.g. $\boldsymbol{\hat{x}}$ is a unit vector in the direction of the $x$-axis. The direction of GW propagation is related to the principal axes by
-\begin{equation}
-	\boldsymbol{n} = \boldsymbol{k} \times \boldsymbol{l} \ . 
-\end{equation}
-
-\subsubsection{Measurement equation}
-In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
-They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio pulse.
-The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e.
-$\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $.
-One obtains \citep[e.g.][]{Maggiore}
-\begin{equation}
- \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
-	\label{eq:momentum_shift}
-\end{equation}
-where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer.
-We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
-We take the pulsar location to be constant i.e.
-neither $\boldsymbol{q}$ nor $d$ are functions of time.
-In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB.
-The barycentering correction is typically applied when generating TOAs, e.g.
-with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
-Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.
-Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$.
-After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames,
-\begin{equation}
-	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
-\end{equation}
-Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
+In the presence of a GW the intrinsic pulsar spin frequency for the  $n$-th pulsar as is related to the radio pulse frequency measured by an observer on Earth as
 \begin{equation}
@@
 where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
-\begin{equation}
-	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
-	\label{eq:g_func}
-\end{equation}
-and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$.
-It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
 \begin{align}
@@
 \end{align}
-Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.
+where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\boldsymbol{n}$ the propagation direction of the GW, $\Omega$ is the constant angular frequency of the GW, $H_{ij}$ is the spatial part of the GW amplitude tensor and $\Phi_0$ the phase offset of the GW.
+\newline
+
+Equation \eqref{eq:g_func_trig} contains a number of simplifying assumptions:
+\begin{enumerate}
+	\item The GW is monochromatic plane wave.
+	\item The pulsar locations are constant with respect to the observer 
+\end{enumerate}
+Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Zhu10},
+\begin{equation}
+	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
+	\label{eq:f_evolution}
+\end{equation}
+where $M_c$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
+A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
+From Equation \eqref{eq:f_evolution} we can see that only those binaries which are massive or at high frequency experience significant frequency evolution over typical PTA timespans.
+The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
+We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
+Regarding point (ii) whilst in practice the pulsar locations vary with respect to the Earth, pulsar TOAs are defined relative to the Solar System barycentre (SSB).The barycentering correction is typically applied when generating TOAs, e.g.
+with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
+Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.
+
+%The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
+%\subsubsection{Plane GW perturbation}\label{sec:plane_gw}
+%We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
+%\begin{equation}
+%	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
+%\end{equation}
+%with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$.
+The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$.
+Throughout this paper we work with
+%
+%
+% We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic.
+
+%The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$). The spatial part is
+%\begin{align}
+%	H_{ij} = h_+ e_{ij}^+(\boldsymbol{n}) + h_{\times} e_{ij}^{\times}(\boldsymbol{n}) \ , \label{eq:hij}
+%\end{align}
+%where $h_{+}$ and $h_{\times}$ are the respective polarisation amplitudes.
+The plus and cross polarisation tensors $e_{ij}^{+}$ and $e_{ij}^{\times}$ are uniquely defined by the principal axes of the wave, viz.\ the unit 3-vectors $\boldsymbol{k}$ and $\boldsymbol{l}$, according to
+%\begin{align}
+%	e_{i j}^{+}(\boldsymbol{n}) =k_i k_j-l_i l_j \ , \\
+%		e_{i j}^{\times}(\boldsymbol{n}) =k_i l_j+l_i k_j \ .
+%\end{align}
+%The principal axes are in turn specified by the location of the GW source on the sky (colatitude $\theta$, azimuth $\phi$) and the polarisation angle $\psi$ according to
+%\begin{align}
+%	\boldsymbol{k}  = &(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
+%	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}} \nonumber \\
+%	& +(\sin \psi \sin \theta) \boldsymbol{\hat{z}} \ , \\
+%	\boldsymbol{l} = &(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
+%	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}}\nonumber  \\
+%	& +(\cos \psi \sin \theta) \boldsymbol{\hat{z}} \ ,
+%\end{align}
+%where e.g. $\boldsymbol{\hat{x}}$ is a unit vector in the direction of the $x$-axis. The direction of GW propagation is related to the principal axes by
+%\begin{equation}
+%	\boldsymbol{n} = \boldsymbol{k} \times \boldsymbol{l} \ . 
+%\end{equation}
+
+%\subsubsection{Measurement equation}
+%In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
+They are described by the geometric object $\boldsymbol{p}$ which we identify sas the 4-momentum of the radio pulse.
+The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e.
+$\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $.
+One obtains \citep[e.g.][]{Maggiore}
+%\begin{equation}
+% \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
+%	\label{eq:momentum_shift}
+%\end{equation}
+%where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer.
+We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
+We take the pulsar location to be constant i.e.
+neither $\boldsymbol{q}$ nor $d$ are functions of time.
+%
+%
+%
+%\begin{equation}
+%	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
+%\end{equation}
+%
+%Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.
+
+\subsection{Parameters of the model}\label{sec:ss_params}
+The set of static parameters which determine the intrinsic frequency evolution of the pulsars in the array outlined in Section \ref{sec:psr_frequency} are:
+\begin{equation}
+	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
+\end{equation}
+The parameters which describe the GW source are:
+\begin{equation}
+	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \psi, \delta, \alpha, , \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
+\end{equation}
+where $h_0$ is the characteristic wave strain, $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
+$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$, $\psi$ the polarization angle of the GW and $\delta,\alpha$ the declination and right ascension of the GW source respectively.
+The complete set of static parameters is denoted	$\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw}$.
+There are $7 + 5N$ parameters to estimate, although the pulsar parameters are constrained better \textit{a priori} by electromagnetic observations than the GW parameters.
+For example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information about $\delta$ and $\alpha$.
+\newline
+
+%
+%The model  and Section \ref{sec:psr_measured} depends on a set of static parameters $\boldsymbol{\theta}$.
+%
+%
+%
+%
+%
+%The model outlined in Section \ref{sec:psr_frequency} and Section \ref{sec:psr_measured} depends on a set of static parameters $\boldsymbol{\theta}$.
+%
+%
+%
+%The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters controlling t and the GW source, i.e. 
+%\begin{equation}
+% \ , \label{eq:params1}
+%\end{equation}
+%with
+%
+%and
+%
+%where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively.
+In Equation \eqref{eq:params3} we reparameterize the two GW polarisation amplitudes, $h_{+}$ and $h_{\times}$, in terms of $\iota$ and the characteristic wave strain $h_0$ through
+%\begin{align}
+%	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
+%	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ .
+%\end{align}
+%We use the parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
 
@@
 
+In this section we do the following...
+
+Measurement equation is parameterised as,
+
 A naive approach is challenging - look at this noisy likelihood curve
@@
 
+In this section we derive the posterior probability distributions for the static parameters.
+However, it is first necessary to discuss a modification to the general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}.
+It can be seen that the equation generally separates into two cosine terms.
+The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
+The argument of the cosine corresponds to the GW phase at the observer on Earth.
+Conversely the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
+The argument of the cosine corresponds to the GW phase at each the individual pulsar.
+The first and second terms are commonly referred to as the ``Earth term" and ``pulsar term" respectively.
+Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases and in standard PTA analysis are typically considered as a source of self noise and dropped from the analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
+In this work we follow the standard analysis approach and drop the pulsar terms from our model.
+Explicitly the measurement equation used in the Kalman filter is
+\begin{equation}
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) \ , 
+	\label{eq:measuremen_earth}
+\end{equation}
+with
+\begin{equation}
+	g^{(n)}_{\rm Earth}(t) = 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \cos(-\Omega t +\Phi_0)  \ .
+	\label{eq:g_func_trig_earth}
+\end{equation}
+We defer the inclusion of the pulsar terms to a future work.
+As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
+We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
+from the Kalman filter model that feeds into the nested sampling algorithm.
+The generated synthetic data that we use to test our method does include the pulsar term in full.
+\newline
+
 \section{Representative example: single SMBHB source and Comparison with Earth terms}\label{sec:rep_example1}
@@
 \section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
-The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters controlling the intrinsic frequency evolution of the pulsars in the array and the GW source, i.e.
-\begin{equation}
-	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
-\end{equation}
-with
-\begin{equation}
-	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
-\end{equation}
-and
-\begin{equation}
-	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
-\end{equation}
-where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively \footnote{$\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
-$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.}.
-In Equation \eqref{eq:params3} we reparameterize the two GW polarisation amplitudes, $h_{+}$ and $h_{\times}$, in terms of $\iota$ and the characteristic wave strain $h_0$ through
-\begin{align}
-	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
-	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ .
-\end{align}
-We use the parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
-A PTA containing $N$ pulsars comprises $7 + 5N$ parameters to estimate.
-Typically the pulsar parameters are constrained better \textit{a priori} by electromagnetic observations than the GW parameters.
-For example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information about $\delta$ and $\alpha$.
-\newline
 
@@
 \subsection{Parameter estimation}\label{sec:parameter_estim}
-In this section we derive the posterior probability distributions for the static parameters.
-However, it is first necessary to discuss a modification to the general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}.
-It can be seen that the equation generally separates into two cosine terms.
-The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
-The argument of the cosine corresponds to the GW phase at the observer on Earth.
-Conversely the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
-The argument of the cosine corresponds to the GW phase at each the individual pulsar.
-The first and second terms are commonly referred to as the ``Earth term" and ``pulsar term" respectively.
-Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases and in standard PTA analysis are typically considered as a source of self noise and dropped from the analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
-In this work we follow the standard analysis approach and drop the pulsar terms from our model.
-Explicitly the measurement equation used in the Kalman filter is
-\begin{equation}
-		f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) \ , 
-		\label{eq:measuremen_earth}
-	\end{equation}
-	with
-	\begin{equation}
-		g^{(n)}_{\rm Earth}(t) = 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \cos(-\Omega t +\Phi_0)  \ .
-		\label{eq:g_func_trig_earth}
-	\end{equation}
-	We defer the inclusion of the pulsar terms to a future work.
-As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
-We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
-from the Kalman filter model that feeds into the nested sampling algorithm.
-The generated synthetic data that we use to test our method does include the pulsar term in full.
-\newline
 

### 2023-10-09 py_src/configs/config_batch_bayes.py

@@ -32,7 +32,7 @@ with open('batch.sh','w') as b:
     
     for h in h_range:
         for n in noise_models:
-            arg_name = f"beta_canonical_bayes_h_{h}_model_{n}"
+            arg_name = f"gamma_canonical_bayes_h_{h}_model_{n}"
             print(arg_name)
             create_slurm_job(arg_name,h,n,s)
 

### 2023-10-10 manuscript/manuscript_2/paper.tex
PP plot, beta

@@
 \section{State-Space Formulation}\label{sec:model}
-In this section we briefly review the state-space formulation of PTA analysis presented in \citep{K1}. \newline
+In this section we briefly review the state-space formulation of PTA analysis presented in K23. \newline
 
@@
 \section{Inclusion of the pulsar term}\label{sec:pulsar_term}
-
-In this section we do the following...
-
-Measurement equation is parameterised as,
-
-A naive approach is challenging - look at this noisy likelihood curve
-
-A reparamterisation is good
-
-In this section we derive the posterior probability distributions for the static parameters.
-However, it is first necessary to discuss a modification to the general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}.
-It can be seen that the equation generally separates into two cosine terms.
-The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
+In this section we review the modification to Equation \eqref{eq:g_func_trig} used in K23, the issues that arise when using the full form of Equation \eqref{eq:g_func_trig}  for likelihood-based inference, and a new parametrisation that alleviates these issues.
+\newline
+
+\subsection{Earth term}
+Equation \eqref{eq:g_func_trig} is the sum of two equal amplitude, phase shifted waves.
+The first wave, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
 The argument of the cosine corresponds to the GW phase at the observer on Earth.
-Conversely the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
+Conversely the second wave, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
 The argument of the cosine corresponds to the GW phase at each the individual pulsar.
@@
 Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases and in standard PTA analysis are typically considered as a source of self noise and dropped from the analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
-In this work we follow the standard analysis approach and drop the pulsar terms from our model.
-Explicitly the measurement equation used in the Kalman filter is
+K23 also followed this example and dropped the pulsar terms from the inference model.
+Explicitly the measurement equation used is
 \begin{equation}
@@
 \end{equation}
-We defer the inclusion of the pulsar terms to a future work.
-As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
-We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
-from the Kalman filter model that feeds into the nested sampling algorithm.
-The generated synthetic data that we use to test our method does include the pulsar term in full.
-\newline
+
+\subsection{Challenges of pulsar term}
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
+	\caption{abc}
+	\label{fig:likelihood_comparison}
+\end{figure*}
+
+K23 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection.
+That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by
+Bayes' Rule,
+\begin{equation}
+	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
+\end{equation}
+where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
+\begin{equation}
+	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
+\end{equation}
+and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}.
+Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
+Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e.
+$\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$.
+We will use the notation $\mathcal{L}(\Theta)$ to refer to the likelihood returned by the Kalman filter for parameter $\Theta$, where $\Theta \in \boldsymbol{\theta}$ and all other members of $\boldsymbol{\theta}$ are held constant at their true injected values.
+\newline
+
+To elucidate the complications introduced by including the pulsar term, we generate some noisy synthetic data $\boldsymbol{Y}$ using the methods outlined in K23.
+$\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
+The choice of $\boldsymbol{\theta}_{\rm psr}$ corresponds to the 47 pulsars in the 12.5 year NANOGrav data set  \citep{2020ApJ...905L..34A}.
+We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig} on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad.
+We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve).
+The dashed grey line indicates the injected value of $\alpha$.
+The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-3} \le \alpha \le 1.0 +10^{-3}$ rad.
+There are two important features in the figure.
+The first is that on the scale of radians the curve of $\mathcal{L}(\alpha)$ is very noisy.
+The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist.
+
+This means that whilst it is theoretically possible to use likelihood-based methods for
+
+practically the noise in the curve
+
+Why noise is bad
+
+Similar are observed for $\Omega$ and $\delta$, the other members of $\theta_{\rm gw}$ are smooth and concave.
+
+$\mathcal{L}(\Theta)$ is noisy
+
+a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$.
+It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
+
+$p(\boldsymbol{\theta} | \boldsymbol{Y})$ is calculated via nested sampling
+
+%Given  Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
+$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})
+%
+%
+
+\subsection{New parameterisation}
+
+To alleviate the probelsm of seciton X,
+
+We define a new parameter 
+\begin{equation}
+\chi^{(n)} = \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} 
+\end{equation}
+Such that Equation \eqref{eq:g_func_trig} becomes
+\begin{align}
+	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
+	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
+	&- \cos \left \{-\Omega t +\Phi_0 + \chi^{(n)} \right \} \Big ] \ .
+	\label{eq:g_func_trig_chi}
+\end{align}
+and the static pulsar parameters to be inferred become
+\begin{equation}
+	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
+\end{equation}
+whilst $\boldsymbol{\theta}_{\rm gw}$  remain unchanged.
+
+In figure Xa we plot
+
+Clearly there is a relation between abc. This will nbe becked and shown ti no be a problm in Section Y
 
@@
 
-The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$.
-It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
+The Kalman filter \citep{Kalman1} is
+
  In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
@@
 \subsection{Nested sampling}\label{sec:nested_sampling}
-We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
-\begin{equation}
-	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
-\end{equation}
-where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
-\begin{equation}
-	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
-\end{equation}
+We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of
+
  We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper.

### 2023-10-10 py_src/kalman_filter.py

@@ -159,7 +159,7 @@ class KalmanFilter:
         #print("running the KF with omega = ", omega_gw)
         #print('...and chi = ', chi)
 
-    
+        
         #Precompute transition/Q/R Kalman matrices
         #F,Q,R are time-independent functions of the parameters
         F = F_function(gamma,self.dt)

### 2023-10-10 py_src/main.py

@@ -8,22 +8,9 @@ h        =  float(sys.argv[2])   # strain
 measurement_model =  sys.argv[3] # whether to use the H0(null) or H1(earth/pulsar) model
 seed = int(sys.argv[4])          # the seeding
 
-
-
-
-#Extra params for pp plot
-#omega =  float(sys.argv[5]) 
-#phi_0 =  float(sys.argv[6]) 
-#psi   =  float(sys.argv[7]) 
-#alpha =  float(sys.argv[8]) 
-#delta =  float(sys.argv[9]) 
-
 if __name__=="__main__":
     bilby_inference_run(arg_name,h,measurement_model,seed)
 
-
-
-    #pp_plot_run(arg_name,h,measurement_model,seed,omega,phi_0,psi,alpha,delta)
 
 
 

### 2023-10-11 manuscript/manuscript_2/paper.tex
PP plots, beta

@@
 
-\subsection{Challenges of pulsar term}
+\subsection{Challenges of pulsar term}\label{sec:pulsar_terms_challenge}
 \begin{figure*}
@@
 $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$.
-We will use the notation $\mathcal{L}(\Theta)$ to refer to the likelihood returned by the Kalman filter for parameter $\Theta$, where $\Theta \in \boldsymbol{\theta}$ and all other members of $\boldsymbol{\theta}$ are held constant at their true injected values.
+We will use the notation ONE DIMENSIONAL $\mathcal{L}(\Theta)$ to refer to the likelihood returned by the Kalman filter for parameter $\Theta$, where $\Theta \in \boldsymbol{\theta}$ and all other members of $\boldsymbol{\theta}$ are held constant at their true injected values.
 \newline
@@
 There are two important features in the figure.
-The first is that on the scale of radians the curve of $\mathcal{L}(\alpha)$ is very noisy.
-The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist.
-
-This means that whilst it is theoretically possible to use likelihood-based methods for
-
-practically the noise in the curve
-
-Why noise is bad
-
-Similar are observed for $\Omega$ and $\delta$, the other members of $\theta_{\rm gw}$ are smooth and concave.
-
-$\mathcal{L}(\Theta)$ is noisy
-
-a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$.
-It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
-
-$p(\boldsymbol{\theta} | \boldsymbol{Y})$ is calculated via nested sampling
+
+The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
+The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.
+
+This means that whilst it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$, in practice the computation is intractable since the noise in the likelihood curve inhibits sampling algorithms to reliably converge to a definitive posterior.
+
+Naturally $\mathcal{L}(\alpha)$ is only one dimensional. The situation is further esacerbated when we move to $N +D $ dimensions.
+
+However, the majority of parameters are well-behaved and exhibit smooth concanve 1D
+
+It is just omega, delta and alpha
+
+Why these parameters?
+
+These are the parameters which correspond to the pulsar term
+
+go through N cycles
+
+Not all parameters are noisy
+
+Whilst noisy likelihood inference is possible...
 
@@
 
-\subsection{New parameterisation}
-
-To alleviate the probelsm of seciton X,
+\subsection{New parametrisation}
+In this section we introduce a new parametrisation of the model described in Section \ref{sec:model} in order to alleviate the problems outlined in Section \ref{sec:pulsar_terms_challenge}. \newline
 

### 2023-10-11 py_src/bilby_wrapper.py

@@ -27,6 +27,7 @@ def BilbySampler(KalmanFilter,init_parameters,priors,label,outdir,npoints):
                               outdir=outdir,
                               sampler ='dynesty',
                               sample='rwalk_dynesty',
+                              bound='single', # https://dynesty.readthedocs.io/en/latest/faq.html
                               check_point_plot=False,
                               npoints=npoints,
                               dlogz=1e-1,

### 2023-10-12 manuscript/manuscript_2/paper.tex
PP plot for h = 1e-14

@@
 Here it is shown how to extend previous analyses to include the so-called `pulsar terms'.
-It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms increases the parameter accuracy relative to the Earth terms approach by X and the sensitivity by Y.
+It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy relative to the Earth terms approach by X and the sensitivity by Y.
 \end{abstract}
@@
 Specifically, K23 formulate PTA analysis as a state-space problem and use a Kalman filter \citep{Kalman1} to track the pulsars' intrinsic rotational state in conjunction with a Bayesian nested sampler \citep{Skilling, Ashton2022} for parameter estimation and model selection.
-\newline
+\textbf{VALIDATION TESTS ON SYNTHETIC DATA...
+}newline
 
@@
 The argument of the cosine corresponds to the GW phase at the observer on Earth.
-Conversely the second wave, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
+Conversely the second wave, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
 The argument of the cosine corresponds to the GW phase at each the individual pulsar.
@@
 	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
-	\caption{abc}
+	\caption{Logarithm of the one-dimensional likelihood $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
+All other static parameters of $\boldsymbol{\theta}$ are held constant at their true injected values.
+In the left-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}(\alpha)$, i.e.
+inclusive of the pulsar terms.
+In the right-hand panel the Kalman filter uses Equation \ref{eq:g_func_trig_earth}, i.e.
+just the Earth terms.
+The vertical dashed lines indicate the true injected value of $\alpha$ ($=1.0$ rad) used to generate $\boldsymbol{Y}$.
+The inset plot covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
+In the left hand panel $\mathcal{L}(\alpha)$ is very noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
+Conversely, in the right hand panel $\mathcal{L}(\alpha)$ is globally smooth.
+}
 	\label{fig:likelihood_comparison}
@@
 $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$.
-We will use the notation ONE DIMENSIONAL $\mathcal{L}(\Theta)$ to refer to the likelihood returned by the Kalman filter for parameter $\Theta$, where $\Theta \in \boldsymbol{\theta}$ and all other members of $\boldsymbol{\theta}$ are held constant at their true injected values.
+We will use the notation $\mathcal{L}(\Theta)$ to refer to the likelihood returned by the Kalman filter for parameter $\Theta$, where $\Theta \in \boldsymbol{\theta}$ and all other members of $\boldsymbol{\theta}$ are held constant at their true injected values.
 \newline
@@
 We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve).
-The dashed grey line indicates the injected value of $\alpha$.
-The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-3} \le \alpha \le 1.0 +10^{-3}$ rad.
-There are two important features in the figure.
-
+The dashed grey line indicates the injected value of $\alpha$ (= 1.0 rad).
+The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
+\newline
+
+There are two important features to discuss in the left-hand panel of Figure \ref{fig:likelihood_comparison}.
 The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
+Similar noisy curves are obtained for $\mathcal{L}(\Omega)$ and $\mathcal{L}(\delta)$.
+For all other static parameters, the $\mathcal{L}(\Theta)$ curves are smooth and globally concave.
 The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.
-
 This means that whilst it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$, in practice the computation is intractable since the noise in the likelihood curve inhibits sampling algorithms to reliably converge to a definitive posterior.
-
-Naturally $\mathcal{L}(\alpha)$ is only one dimensional. The situation is further esacerbated when we move to $N +D $ dimensions.
-
-However, the majority of parameters are well-behaved and exhibit smooth concanve 1D
-
-It is just omega, delta and alpha
-
-Why these parameters?
-
-These are the parameters which correspond to the pulsar term
-
-go through N cycles
-
-Not all parameters are noisy
+Moreover, $\mathcal{L}(\alpha)$ is only a one-dimensional function.
+The challenge for sampling algorithms is further exacerbated when we consider the three-dimensional parameter space in $\alpha - \delta - \Omega$, or the full $7 + 5N$ dimensions of Section \ref{sec:ss_params}.
+\newline
+
+Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\Theta)$ curves whereas other static parameters do not?
+These parameters appear in the pulsar term as a product in the pulsar-dependent phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, where we remind the reader that $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
+Over the typical pulsar distances $d^{(n)} \sim 1$ kpc, a 10 nHz GW can accumulate $10^3$ cycles.
+This creates a confusion problem when taking the cosine of  $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ since we do not know how many cycles the wave has gone through, modulo $2 \pi$.
+\textcolor{red}{TK: this explanation is rubbish!
+How can we say this better?}.
+Conversely, if we drop the pulsar term,  the $\mathcal{L}(\alpha)$ curve (as well as $\mathcal{L}(\Omega)$ , $\mathcal{L}(\delta)$   )
+This is shown by the orange curve in the right-hand panel of Figure \ref{fig:likelihood_comparison} where we plot $\mathcal{L}(\alpha)$, analogous to the left-hand panel, but now use Equation \ref{eq:g_func_trig_earth} in our Kalman filter (i.e.
+just the Earth terms) rather than the full Equation \ref{eq:g_func_trig}.
+In this case there is no confusion regarding the number of cycles the wave has gone through over a distance $d^{(n)}$ and the curve is smooth.
+
+\subsection{New parametrisation}
+In order to include the pulsar terms in an inference model and apply the methods of K23 it is necessary to be able to handle the noisy likelihood curves presented in Section \ref{sec:pulsar_terms_challenge}.
+
+Whilst computational methods for approximate inference from noisy likelihoods do exists \citep[e.g.][]{doi:10.1137/21M1410853,2021arXiv210403942J,2019arXiv190501252J}
 
@@
 
-%Given  Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
-$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})
-%
-%
-
-\subsection{New parametrisation}
 In this section we introduce a new parametrisation of the model described in Section \ref{sec:model} in order to alleviate the problems outlined in Section \ref{sec:pulsar_terms_challenge}. \newline

### 2023-10-12 py_src/configs/config_batch_pp.py

@@ -29,7 +29,7 @@ def create_slurm_job(arg_name,h,measurement_model,seed,omega,phi0,psi,delta,alph
 
 N = 200
 seed = 1250 
-h = 5e-15 
+h = 1e-14 
 model = "pulsar"
         
 #Create a bilby prior object to sample from
@@ -57,7 +57,7 @@ with open('batch.sh','w') as b:
     for i in range(N):
         p = priors.sample()
 
-        arg_name = f"beta_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
+        arg_name = f"gamma_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
         print(arg_name)
         create_slurm_job(arg_name,h,model,seed,p['omega_gw'],p['phi0_gw'],p['psi_gw'],p['delta_gw'],p['alpha_gw'])
         b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")

### 2023-10-12 manuscript/manuscript_2/paper.tex
misc changes

@@
 In the left hand panel $\mathcal{L}(\alpha)$ is very noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
+This noise hampers the ability of sampling algorithms to converge to a posterior distribution.
 Conversely, in the right hand panel $\mathcal{L}(\alpha)$ is globally smooth.
@@
 In order to include the pulsar terms in an inference model and apply the methods of K23 it is necessary to be able to handle the noisy likelihood curves presented in Section \ref{sec:pulsar_terms_challenge}.
-
-Whilst computational methods for approximate inference from noisy likelihoods do exists \citep[e.g.][]{doi:10.1137/21M1410853,2021arXiv210403942J,2019arXiv190501252J}
-
-Whilst noisy likelihood inference is possible...
-
-In this section we introduce a new parametrisation of the model described in Section \ref{sec:model} in order to alleviate the problems outlined in Section \ref{sec:pulsar_terms_challenge}. \newline
+In this section we introduce a new parametrisation of the model described in Section \ref{sec:model} in order to alleviate the noisy likelihoods.
+\newline
 
@@
 \begin{equation}
-	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
-\end{equation}
-whilst $\boldsymbol{\theta}_{\rm gw}$  remain unchanged.
-
-In figure Xa we plot
-
-Clearly there is a relation between abc. This will nbe becked and shown ti no be a problm in Section Y
+	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
+\end{equation}
+whilst $\boldsymbol{\theta}_{\rm gw}$  remain unchanged. \newline
+
+This parametrisation in terms of $\chi^{(n)}$ treats the pulsar dependent phase correction as an explicit parameter to be inferred for every pulsar.
+The advantage of this approach is that the likelihood curves $\mathcal{L}(\alpha),\mathcal{L}(\delta),\mathcal{L}(\Omega)$ become smooth and resemble the right-hand panel of Figure \ref{fig:likelihood_comparison}.
+This enables sampling algorithms to efficiently find the likelihood maxima and estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+The new parametrisation also does not change the dimension of the parameter space, since we have traded $d^{(n)}$ for $\chi^{(n)}$.
 
@@
 
-\section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
-
-In K1
-
-The workflow presented in K1, hich we also follow here proceeds as follows
-
-In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e.
-the model evidence).
-In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter.
-In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference.
-Model selection and the specification of the null model are described in Section \ref{sec:model_selection}.
-A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
-The method complements traditional PTA analyses because it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations.
-
-\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
-%\begin{figure*}
-%	%\centering % Not needed
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
-%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
-%		\label{fig:6MB_BFS}
-%	\end{subfigure}  
-%    \hfill
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
-%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
-%		\label{fig:25MB_bfs}
-%	\end{subfigure}
-%	\caption{
-%		}
-%	\label{fig:four figures}
-%\end{figure*}
-
-The Kalman filter \citep{Kalman1} is
-
- In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
-Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
-Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
-Hence inferring the static parameters is a non-linear exercise, to be tackled separately after the linear Kalman filter operates on the data for a fixed set of static parameters.
-Inference of the static parameters in Equations \eqref{eq:measurement} and \eqref{eq:g_func} by nested sampling is discussed in Section \ref{sec:nested_sampling}.
-\newline
-
-Implementation of the linear Kalman filter for the PTA state-space model in Section 3.2, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
-At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.
-The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by converting ${\boldsymbol{\hat X}}_i$ into predicted measurements ${\boldsymbol{\hat Y}}_i$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func} and comparing with the actual, noisy measurements ${\boldsymbol{\hat Y}}_i$.
-This defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation.
-The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$.
-The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}.
-For a fixed set of static parameters, the Kalman filter returns an estimate of the the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error.
-We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline
-
-The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ can  then be calculated at each timestep from the Kalman filter output according to
-\begin{eqnarray}
-	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ .
-\end{eqnarray}
-The total log-likelihood for the entire sequence is
-\begin{eqnarray}
-	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
-\end{eqnarray}
-Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
-$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$.
-Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$.
-If $\boldsymbol{\hat{\theta}}$ is close to the true underlying parameters $\boldsymbol{\theta}$, then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are minimized and $\mathcal{L}$ is maximised.
-This is illustrated with synthetic data in Figure \ref{fig:kalman_example}.
-In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations (1)--(4), (16), and (17) for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$.
-The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve).
-The predicted state $\hat{f}_{\rm p}^{(1)}(t)$ is converted  into a predicted measurement $\hat{f}_{\rm m}^{(1)}(t)$ (middle panel, magenta curve), which again overlaps almost perfectly with the true measurement.
-The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel).
-By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters to the Kalman filter, where $\Omega$ has been perturbed from its true value by $20 \%$.
-In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates.
-It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel).
-
-%n Figure ; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
-The residuals correspond to random noise and are normally distributed.
-Conversely, if $\boldsymbol{\hat{\theta}}$ is not close to the true parameters then the filter is unable to recover the state evolution.
-This is demonstrated in Figure \ref{fig:25MB_bfs} where the Kalman filter was run with $\boldsymbol{\hat{\theta}}$ slightly perturbed away from the true values.
-In this case the filter cannot track the state variable accurately and the residuals are no longer Gaussian.
-
-\subsection{Nested sampling}\label{sec:nested_sampling}
-We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of
-
- We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper.
-Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, which also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
-It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
-The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies.
-Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}.
-For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
-For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}.
-Multiple nested sampling algorithms and computational libraries exist.
-\citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}.
-In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library.
-We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work.
-\newline
-
-The primary tunable parameter in nested sampling is $n_{\rm live}$.
-A greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
-However the computational runtime scales as $\mathcal{O}(n_{\rm live})$.
-\cite{Ashton2022} offered a rule-of-thumb trade-off, where the minimum number of live points should be greater than the number of static parameters.
-Informal empirical tests conducted as part of this paper support the trade-off suggested by \cite{Ashton2022}; we find typically that the true ${\boldsymbol{\theta}}$ is contained within the 90\% credible interval of the derived 1D marginalised posteriors of ${\boldsymbol{\hat{\theta}}}$ for $n_{\rm live} > 7 + 5N$ with $N \leq 50$.
-Unless stated otherwise we take $n_{\rm live} = 1000$ for all results presented in this work.
-\newline
-
-%See https://arxiv.org/pdf/2102.12478.pdf for a good NS explanation
-
-\subsection{Model selection}\label{sec:model_selection}
-The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
-\begin{equation}
-	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
-\end{equation}
-Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
-$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
-This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
-The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
-
-\subsection{Summary of workflow}\label{sec:methodsummary}
-For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
-\begin{enumerate}[leftmargin=2em]
-	\item Specify a PTA composed of $N$ pulsars 
-	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
-	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$. 
-	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
-	 \item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
-	 \item For each live point:
-\begin{enumerate}[leftmargin=2em]
-	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
-	\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
-\end{enumerate}
-	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
-	\item Constrained sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood $\log \mathcal{L}_{\rm new}$ > $\log \mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (a)--(b).
-	\item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ 
-	\item Repeat steps (vii) - (ix) until convergence criteria are satisfied.
-\end{enumerate}
-In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$.
-The resulting $\mathcal{Z}$ values can then be compared.
-We remind the reader that the above workflow differs from a realistic PTA analysis in on important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
-The generalization to TOAs is subtle and will be tackled in a forthcoming paper.
-
-\section{Validation procedure with synthetic data} \label{sec:testing}
-In this section we outline how synthetic data is generated in order to validate the analysis scheme in Section \ref{sec:detect}.
-In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
-In Section \ref{sec:gendata} we demonstrate how to solve the equations of Section \ref{sec:model} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.
-
-\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
-
-We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
-The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
-For each pulsar we adopt the fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
-A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility.
-The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}.
-The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
-The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist.
-The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
-In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
-We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$.
-The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
-\begin{align}
-	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
-	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
-\end{align}
-where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits.
-The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value.
-Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
-In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
-\begin{eqnarray}
-	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
-\end{eqnarray}
-For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline
-
-As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
-\begin{eqnarray}
-	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
-\end{eqnarray}
-and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach.
-We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
-A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$.
-In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$.
-In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}.
-We see that the synthetic and empirical residuals are qualitatively similar.
-For this pulsar the synthetic residuals were generated using $\sigma^{(n)} = 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the value inferred from the noise model of Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, using the best fit parameters, $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$.
-This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars.
-If we set $\ln \alpha_1 = -38$ in Equation \eqref{eq:sigmap}, close to its lower limit, then the estimated $\sigma^{(n)}$ for  PSR J1939+2134 is $\sigma^{(n)} =3.90 \times 10^{-27}$ which is in much better agreement with the value inferred from the residuals comparison method.
-This discrepancy between the two estimates of $\sigma^{(n)}$ is particular to PSR J1939+2134; for all other pulsars we find the two estimates of $\sigma^{(n)}$ are comparable to within a few orders of magnitude.
-
-\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
-We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
-\begin{enumerate}[leftmargin=2em]
-	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
-	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}
-\end{enumerate}
-Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
-The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \eqref{sec:synt_pta}.
-For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
-The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
-\begin{equation}
-	\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
-\end{equation}
-Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
-The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
-Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
-Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar.
-\newline
-
-Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } \sigma^{(n)} \Delta W_t \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta W_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
-To fix the problem, we subtract the deterministic frequency evolution and track the new variable, equivalent to a change of variables:
-\begin{equation}
-	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ .
-\end{equation}  
-We similarly modify the measurement variable to be
-\begin{equation}
-	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ ,
-\end{equation}
-where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris which is measured to high accuracy in practice by {\sc tempo2}.
-For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately.
-Equation \eqref{eq:measurement} is then updated to read
-\begin{equation}
-	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
-	\label{eq:measurement_cov}
-\end{equation}
-We emphasise that this change of variables is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g.
-long double, quadruple).
-It does not remove any degrees of freedom nor does it involve an approximation.
-In particular $f_{\rm em}^{(n)}(t_1)$
-and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \ref{eq:spinevol}--\eqref{eq:xieqn}.
-
-\section{Representative example: single SMBHB source}\label{sec:rep_example}
-In this Section we apply the analysis scheme of Section \ref{sec:detect} and the validation procedure of Section \ref{sec:testing} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-We remind the reader that the analysis in this paper targets a single SMBHB source.
-The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-In Section \ref{sec:priors} we outline the priors chosen on $\boldsymbol{\theta}$.
-In Section \ref{sec:parameter_estim} we apply the workflow of Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
-We go on in Section \ref{sec:multiple_noise} to extend this parameter estimation procedure to multiple realisations of the noise processes.
-In Section \ref{sec:detection} we calculate the the detectability of the source as a function of $h_0$.
-\newline
-
-The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected to be astrophysically reasonable and representative.
-The injected $\boldsymbol{\theta}_{\rm gw}$ are summarised in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}.
-The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$ are as described in Section \ref{sec:synt_pta} and are also shown in Table \ref{tab:parameters_and_priors}.
-
-\subsection{Prior distributions}\label{sec:priors}
-For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters.
-For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
-The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires some additional discussion.
-\newline
-
-The parameters which govern the deterministic evolution of the pulsar spin frequency, $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ are well-determined by existing radio timing observations.
-We can identify $f_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the time derivative of barycentric rotation frequency quoted by pulsar observation catalogues.
-For the 12.5 year NANOGrav pulsars the median error in  the barycentric rotation frequency is $\sim 7 \times 10^{-13}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.8 \times 10^{-20}$ s $^{-2}$.
-Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$, $\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 10^3 \epsilon^{(n)}_{a}$ either side of the true value, where $\epsilon^{(n)}_{a}$ is the error for that pulsar quoted in the ATNF catalogue and the subscript $a = \{ f, \dot{f} \}$ identifies the error as being in $f_{\rm em}^{(n)} (t_1)$, or $\dot{f}_{\rm em}^{(n)} (t_1)$ respectively.
-In this way we can test how well our method performs without requiring exceptionally precise and accurate measurements of the pulsar parameters to be made \textit{a priori}.
-Instead we consider the pulsar parameters similarly to the GW source parameters and estimate them optimally within a consistent framework.
-Adopting looser priors in this way allows us to subject our method to a more stringent test and confirm that we still accurately estimate GW parameters without highly accurate and precise initial estimates of $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$.
-\newline
-
-The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}.
-However for this work, whilst the pulsar distance is used when generating synthetic data, as is discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$.
-Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
-That is, for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
-It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value.
-We briefly explored setting an uninformative prior on $\gamma^{(n)}$ over e.g.
-LogUniform($10^{-15}, 10^{-10}$) s$^{-1}$ as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g.
-set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
-Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-3/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
-For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-3/2}$).
-The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-3/2}$.
-For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-3/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
-\newline
-
-Since we are not setting priors on $\gamma^{(n)}$ or $d^{(n)}$ this reduces the dimensionality of the parameter space to $7 + 3N$.
-We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space, c.f.
-Equation \ref{eq:psrparams}.
-Explicitly,
-\begin{equation}
-	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ .
-\end{equation}
-All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
+In this section we apply the analysis scheme of K23 and the new model parametrisation of Section \ref{sec:pulsar_term} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
+In Section \ref{sec:synthetic_data} we briefly review how synthetic data is generated in order to validate the analysis scheme.
+In Section \ref{sec:param_estimation} we apply the analysis scheme in conjunction with the model of Section \ref{sec:pulsar_term} in order to estimate $\boldsymbol{\theta}$ and compare the results with those using Earth-terms model, Equation \eqref{eq:g_func_trig_earth}.
+In Section \ref{sec:detection_new} we calculate the detectability of the GW source as a function of $h_0$ for both the pulsar terms and the Earth terms model.
+\newline
+
+This section considers only a single realisation of the system (i.e.
+process and measurement) noise and a particular, arbitrary set of injection parameters $\boldsymbol{\theta}$.
+Extension to multiple noise realisations and broader parameter spaces is considered in Sections X and Y respectively.
+
+\subsection{Generating synthetic data}\label{sec:synthetic_data}
 \begin{table*}
@@
 
-\subsection{Parameter estimation}\label{sec:parameter_estim}
-
-	Initially we consider a single noise realisation of the synthetic data for the representative example system described in Table \ref{tab:parameters_and_priors}.
-We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
-The results are shown in Figure \ref{fig:corner_plot_1} for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$.
-The histograms shown in the figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
+Following K23, we consider a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
+The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
+Fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, are used with the latter two quantities evaluated at the Solar System barycenter.
+These values are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+No direct measurements exist for $\gamma^{(n)}$, though the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
+In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
+Similarly, no direct measurements exist for $\sigma^{(n)}$.
+Following K23, we use
+the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} in order to  set physically reasonable values.
+Within this model , the standard deviation in the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, is
+\begin{align}
+	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
+	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
+\end{align}
+where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits.
+The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value.
+Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
+In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
+\begin{eqnarray}
+	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
+\end{eqnarray}
+For the synthetic 12.5 year NANOGrav dataset, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline
+
+This completely specifies the static injection parameters for $\theta_{\rm psr}$.
+The static injection GW source parameters $\boldsymbol{\theta}_{\rm gw}$ are selected arbitrarily, subject to being astrophysically reasonable and representative.
+We consider a "high-SNR" source with $h_0 = 10^{-12}$ and a "low-SNR" source with $h_0 =5 \times 10^{-15}$.
+The complete set of injected parameters $\boldsymbol{\theta}$ is summarised in Table \ref{tab:parameters_and_priors} \newline
+
+Given the injection parameters $\boldsymbol{\theta}$, we generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
+\begin{enumerate}[leftmargin=2em]
+	\item Integrate Equation \eqref{eq:frequency_evolution} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
+	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func_trig}
+\end{enumerate}
+We consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
+The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
+\begin{equation}
+	\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
+\end{equation}
+Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
+Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
+Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar.
+\newline
+
+\subsection{Parameter estimation}\label{sec:param_estimation}
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_highSNR_compare}
+	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The vertical orange lines indicate the true injected values.
+The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
+We are able to accurately estimate each parameter of interest.
+We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
+Note that the $x$-axis scaling does not span the total prior space.}
+	\label{fig:corner_plot_1}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
+	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The vertical orange lines indicate the true injected values.
+The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
+We are able to accurately estimate each parameter of interest.
+We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
+Note that the $x$-axis scaling does not span the total prior space.}
+	\label{fig:corner_plot_2}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=0.45\textwidth, height =0.45\textwidth ]{images/corner_highSNR_chi} 	
+	\includegraphics[width=0.45\textwidth, height =0.45\textwidth ]{images/corner_lowSNR_chi}
+	\caption{hellow}
+	\label{fig:corner_plot_3}
+\end{figure*}
+
+Initially we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$ for the representative example system described in Table \ref{tab:parameters_and_priors}.
+We consider a "high-SNR" system with $h_0 = 10^{-12}$ and a ``low-SNR" system with $h_0 = 5 \times 10^{-15}$.
+We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ for both the high/low SNR systems.
+Priors on $\boldsymbol{\theta}$ are selected as follows.
+For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
+For $\pi(\boldsymbol{\theta}_{\rm psr})$ we use the same priors discussed in K23.
+The addition here is the prior on $\chi^{(n)}$ which we take to be $\text{Uniform}(0,2 \pi)$.
+	All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}
+
+The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_1} for the high-SNR system and in Figure \ref{fig:corner_plot_2} for the low-SNR system.
+
+The histograms shown in the figures are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the solid orange line indicates the true injection value.
 The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
+The green curves are the results derived using the Earth-terms measurement equation, Equation \ref{eq:g_func_trig_earth}.
+The blue curves are the results derived using the full measurement equation, including the pulsar terms, Equation \eqref{eq:g_func_trig}
+
+For the high-SNR case, Figure \ref{fig:corner_plot_1} the one-dimensional posteriors for the Earth-terms solution are evidently biased, as was observed in K23.
+Particular biases are observed in $\Phi_0, \psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
+
+The inclusion of the pulsar terms corrects for this bias. The green posteriors
+
+Moreover, the injection values are contained witinin the 2-sigma equivalent contours for every pair of parameters. This is not true in the Earth-terms case.
+
+For the low-SNR case
+
+les important for PE in low SNR
+
+but is important generally for detectio nas we will show.
+
+Takeaways
+
+We are able to estimate hte static parameters
+
+the inclusion of the pulsar terms general improves the accuracy
+
+the chi terms are also recoverable
+
 It is evident that for this representative example with characteristic strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
@@
 
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{example-image-a}
-	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
-The vertical orange lines indicate the true injected values.
-The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
-We are able to accurately estimate each parameter of interest.
-We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
-Note that the $x$-axis scaling does not span the total prior space.}
-	\label{fig:corner_plot_1}
-\end{figure*}
+\subsection{Detection}\label{sec:detection_new}
+
+\section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
+
+In K1
+
+The workflow presented in K1, hich we also follow here proceeds as follows
+
+In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e.
+the model evidence).
+In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter.
+In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference.
+Model selection and the specification of the null model are described in Section \ref{sec:model_selection}.
+A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
+The method complements traditional PTA analyses because it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations.
+
+\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
+%\begin{figure*}
+%	%\centering % Not needed
+%	\begin{subfigure}[b]{0.49\textwidth}
+%		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
+%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
+%		\label{fig:6MB_BFS}
+%	\end{subfigure}  
+%    \hfill
+%	\begin{subfigure}[b]{0.49\textwidth}
+%		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
+%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
+%		\label{fig:25MB_bfs}
+%	\end{subfigure}
+%	\caption{
+%		}
+%	\label{fig:four figures}
+%\end{figure*}
+
+The Kalman filter \citep{Kalman1} is
+
+ In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
+Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
+Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
+Hence inferring the static parameters is a non-linear exercise, to be tackled separately after the linear Kalman filter operates on the data for a fixed set of static parameters.
+Inference of the static parameters in Equations \eqref{eq:measurement} and \eqref{eq:g_func} by nested sampling is discussed in Section \ref{sec:nested_sampling}.
+\newline
+
+Implementation of the linear Kalman filter for the PTA state-space model in Section 3.2, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
+At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.
+The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by converting ${\boldsymbol{\hat X}}_i$ into predicted measurements ${\boldsymbol{\hat Y}}_i$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func} and comparing with the actual, noisy measurements ${\boldsymbol{\hat Y}}_i$.
+This defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation.
+The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$.
+The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}.
+For a fixed set of static parameters, the Kalman filter returns an estimate of the the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error.
+We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline
+
+The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ can  then be calculated at each timestep from the Kalman filter output according to
+\begin{eqnarray}
+	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ .
+\end{eqnarray}
+The total log-likelihood for the entire sequence is
+\begin{eqnarray}
+	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
+\end{eqnarray}
+Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
+$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$.
+Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$.
+If $\boldsymbol{\hat{\theta}}$ is close to the true underlying parameters $\boldsymbol{\theta}$, then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are minimized and $\mathcal{L}$ is maximised.
+This is illustrated with synthetic data in Figure \ref{fig:kalman_example}.
+In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations (1)--(4), (16), and (17) for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$.
+The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve).
+The predicted state $\hat{f}_{\rm p}^{(1)}(t)$ is converted  into a predicted measurement $\hat{f}_{\rm m}^{(1)}(t)$ (middle panel, magenta curve), which again overlaps almost perfectly with the true measurement.
+The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel).
+By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters to the Kalman filter, where $\Omega$ has been perturbed from its true value by $20 \%$.
+In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates.
+It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel).
+
+%n Figure ; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
+The residuals correspond to random noise and are normally distributed.
+Conversely, if $\boldsymbol{\hat{\theta}}$ is not close to the true parameters then the filter is unable to recover the state evolution.
+This is demonstrated in Figure \ref{fig:25MB_bfs} where the Kalman filter was run with $\boldsymbol{\hat{\theta}}$ slightly perturbed away from the true values.
+In this case the filter cannot track the state variable accurately and the residuals are no longer Gaussian.
+
+\subsection{Nested sampling}\label{sec:nested_sampling}
+We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of
+
+ We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper.
+Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, which also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
+The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies.
+Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}.
+For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
+For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}.
+Multiple nested sampling algorithms and computational libraries exist.
+\citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}.
+In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library.
+We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work.
+\newline
+
+The primary tunable parameter in nested sampling is $n_{\rm live}$.
+A greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
+However the computational runtime scales as $\mathcal{O}(n_{\rm live})$.
+\cite{Ashton2022} offered a rule-of-thumb trade-off, where the minimum number of live points should be greater than the number of static parameters.
+Informal empirical tests conducted as part of this paper support the trade-off suggested by \cite{Ashton2022}; we find typically that the true ${\boldsymbol{\theta}}$ is contained within the 90\% credible interval of the derived 1D marginalised posteriors of ${\boldsymbol{\hat{\theta}}}$ for $n_{\rm live} > 7 + 5N$ with $N \leq 50$.
+Unless stated otherwise we take $n_{\rm live} = 1000$ for all results presented in this work.
+\newline
+
+%See https://arxiv.org/pdf/2102.12478.pdf for a good NS explanation
+
+\subsection{Model selection}\label{sec:model_selection}
+The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
+\begin{equation}
+	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
+\end{equation}
+Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
+$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
+This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
+The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
+
+\subsection{Summary of workflow}\label{sec:methodsummary}
+For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
+\begin{enumerate}[leftmargin=2em]
+	\item Specify a PTA composed of $N$ pulsars 
+	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
+	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$. 
+	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
+	 \item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
+	 \item For each live point:
+\begin{enumerate}[leftmargin=2em]
+	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
+	\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
+\end{enumerate}
+	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
+	\item Constrained sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood $\log \mathcal{L}_{\rm new}$ > $\log \mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (a)--(b).
+	\item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ 
+	\item Repeat steps (vii) - (ix) until convergence criteria are satisfied.
+\end{enumerate}
+In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$.
+The resulting $\mathcal{Z}$ values can then be compared.
+We remind the reader that the above workflow differs from a realistic PTA analysis in on important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
+The generalization to TOAs is subtle and will be tackled in a forthcoming paper.
+
+\section{Validation procedure with synthetic data} \label{sec:testing}
+In this section we outline how synthetic data is generated in order to validate the analysis scheme in Section \ref{sec:detect}.
+In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
+In Section \ref{sec:gendata} we demonstrate how to solve the equations of Section \ref{sec:model} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.
+
+\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
+
+\section{Representative example: single SMBHB source}\label{sec:rep_example}
+In this Section we apply the analysis scheme of Section \ref{sec:detect} and the validation procedure of Section \ref{sec:testing} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
+We remind the reader that the analysis in this paper targets a single SMBHB source.
+The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
+In Section \ref{sec:priors} we outline the priors chosen on $\boldsymbol{\theta}$.
+In Section \ref{sec:parameter_estim} we apply the workflow of Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
+We go on in Section \ref{sec:multiple_noise} to extend this parameter estimation procedure to multiple realisations of the noise processes.
+In Section \ref{sec:detection} we calculate the the detectability of the source as a function of $h_0$.
+\newline
+
+\subsection{Prior distributions}\label{sec:priors}
 
@@
 \section{Discussion}\label{sec:discussion}
+
+Whilst computational methods for approximate inference from noisy likelihoods do exist \citep[e.g.][]{doi:10.1137/21M1410853,2021arXiv210403942J,2019arXiv190501252J},
+
 We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.

### 2023-10-12 py_src/plotting.py

@@ -251,6 +251,8 @@ def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,label
 
         y_post,injection_parameters_idx,ranges_idx= _extract_posterior_results(f,variables_to_plot,injection_parameters_idx,ranges_idx,scalings=scalings)
 
+        errors = get_posterior_accuracy(y_post,injection_parameters_idx,labels)
+
         k = i 
         if k ==2:
             k = k+1 #convoluted way of skipping C2 color. Surely a better way exists
@@ -366,6 +368,21 @@ def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,label
         plt.savefig(f"../data/images/{savefig}.png", bbox_inches="tight",dpi=300)
 
 
+
+
+
+def get_posterior_accuracy(posterior,injection,labels):
+
+    print("The error in the 1D posteriors is as follows:")
+    for i in range(posterior.shape[-1]):
+        y = posterior[:,i]
+        inj = injection[i]
+        error = np.mean(np.abs(inj - y) / inj)
+        print(labels[i], error)
+    print('*****************************')
+
+
+    return 1
 
 
 

### 2023-10-18 docs/manuscript/paper.tex
extra sentence for into of sections
all comments up to start of section 4

@@
 \begin{abstract}	
-Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing array observations.
+Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing arrays.
 A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-A two-step procedure is applied within a state-space framework, such that the model parameters are estimated with a Kalman filter, which then provides a likelihood for nested sampling.
-The procedure enables self-consistent inference of model parameters and calculation of the Bayesian evidence, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional analysis methods.
+A two-step procedure is applied within a state-space framework, such that the modulation is tracked with a Kalman filter, which then provides a likelihood for nested sampling.
+The procedure enables the static parameters in the problem, such as the sky position of the source, to be inferred without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional parameter estimation methods.
+It also returns the Bayes factor relating a model with a single-source gravitational wave to one without, complementing traditional detection methods.
 It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$.
-Full posterior distributions of model parameters are recovered.
-There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
-Smaller biases $\lesssim 1 \%$ are also observed in some of the other parameters of the system.
+Full posterior distributions of model parameters are recovered and tested for accuracy.
+There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms'.
+Smaller biases $\lesssim 1 \%$ are also observed in some other static parameters.
 \end{abstract}
@@
 Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}.
-The detection of nHz GWs has inspired the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines.
+The detection of nHz GWs has inspired the development of new observational methods, since it is impractical to engineer terrestrial interferometric detectors with sufficiently long baselines.
 The foremost method is timing an ensemble of pulsars, i.e.
 a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}.
-A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth.
+A nHz GW influences the trajectory and frequency of individual radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth.
 By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs.
@@
 Individual SMBHBs are continuous wave sources; they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals.
-Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs.
-However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}.
+Consequently, they are detected more efficiently by either a frequentist matched filter, e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA}, or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A}, rather than by cross-correlating pulsar pairs.
+However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}.
+Inconclusive evidence at low significance was presented recently by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}.
 \newline
@@
 random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}.
-This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
+This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Myers2021MNRAS.502.3113M,Meyers2021,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
 In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators.
@@
 The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density.
-Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
+Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations summarized by the power spectral density, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
 The new and existing approaches are therefore complementary.
 In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
-We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood (i.e.
+We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the static GW parameters and calculate the marginal likelihood (i.e.
 the model evidence) for model selection.
@@
 
-% --- a promise which we aim to test as the key goal of this paper. 
-\noindent This paper is organised as follows.
+This paper is organised as follows.
 In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW.
@@
 In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method.
-We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative example GW system.
+We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source.
 In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space.
+In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates.
 Discussion on  future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.
@@
 We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation.
-In this work we take the intrinsic state variable to be the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local,freely-falling rest frame of the pulsar's centre of mass.
+In this work we take the intrinsic state variable to be the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local, freely-falling rest frame of the pulsar's centre of mass.
 A phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.
@@
 A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}.
-We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend.
+We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend.
 In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque.
-The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
+The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ for the sake of definiteness \citep{1969ApJ...157..869G}.
 Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution $f_{\rm em}^{(n)}(t)$ can be approximated accurately by
@@
 	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
+	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'} \delta (t - t') \ .	\label{eq:xieqn}
 \end{align}
@@
 Combined with the mean reversion it gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
-It is important to note that white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
+It is important to note that white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma^{(n)}$ in Equation \eqref{eq:frequency_evolution}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
 \newline
@@
 In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth.
-Specifically, the pulse frequency is modulated harmonically at the GW frequency.
+Specifically, the pulse frequency at the Earth is modulated harmonically at the GW frequency.
 We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section.
-The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
+The measurement equation is a key input into the Kalman filter in Section \ref{sec:kalman_filter}
 \subsubsection{Plane GW perturbation}\label{sec:plane_gw}
@@
 \end{equation}
-with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$.
+with nearly Lorentz spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$ \citep{schutz2022}.
 The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$.
@@
 We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB.
+\newline
+
 In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic.
-Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Zhu10},
-\begin{equation}
-	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
+Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) changes during the inspiral by \citep[e.g.][]{Zhu10}
+\begin{equation}
+	\Delta f_{\rm gw} = 3.94 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
 	\label{eq:f_evolution}
 \end{equation}
-where $M_c$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
+where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
 A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
-From Equation \eqref{eq:f_evolution} we can see that only those binaries which are massive or at high frequency experience significant frequency evolution over typical PTA timespans.
+Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^9 M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 10^{-7} \mathrm{~Hz}]^{11/3} (T_{\rm obs}^2/10 \mathrm{yr}) \gtrsim 8 \mathrm{yr}$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$.
 The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
@@
 In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
-They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio pulse.
-The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e.
+They are described by the geometric object $\boldsymbol{p}$, which we identify as the momentum 4-vector of the radio pulse.
+The presence of a GW induces a shift in the temporal component of the associated momentum one-form, while the photon travels from the emitter to the observer, i.e.
 $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $.
@@
 \end{equation}
-where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer.
-We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
+where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer at rest in the nearly Lorentz coordinates $(t,{\boldsymbol{x}})$.
+In Equation \eqref{eq:momentum_shift} $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar, $d$ is the distance to the pulsar and we write $h_{ij} = g_{ij} - \eta_{ij}$.
 We take the pulsar location to be constant i.e.
@@
 with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
-Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.
+Some pulsars, including some PTA pulsars, do have non-negligible proper motions of order $10^2$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.\newline
+
 Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$.
-After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames,
-\begin{equation}
-	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
-\end{equation}
+After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames.
 Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
 \begin{equation}
-	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d^{(n)}) g^{(n)}(t) +  \varepsilon^{(n)}\ .
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}\ ,
 	\label{eq:measurement}
 \end{equation}
-where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies: 
+where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
 \begin{align}
 	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'}(t - t') \ .	\label{eq:vareps}
+	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
 \end{align}
-where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
+where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
 \begin{equation}
@@
 \end{equation}
-where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively \footnote{$\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
+where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively \footnote{$\iota$ is the angle between the unit normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
 $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.}.
@@
 A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
-The method complements traditional PTA analyses because it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations.
+The method complements traditional PTA analyses because (i) it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations; and (ii) it infers ${\boldsymbol{\theta}}$ conditional on the specific, time-ordered, random realization of the noise corresponding to the observed sequence of TOAs rather than on ensemble-averaged quantities like the power spectral density.
 
@@
 	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
-	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed systems state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are identically correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column).
-The top panels show the true pulsar state $f_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$  (green curve).
-We plot the subtracted state to better illustrate the stochastic wandering of the pulsar frequency.
-In the left hand column, the blue/green solutions are completely overlaid.
-The middle panels show the true measured frequency $f_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve).
-Again we plot the subtracted state, and in the left hand column the red/magenta solutions are completely overlaid.
-The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{1}(t) - \hat{f}_{\rm m}^{1}(t)$.
+	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column).
+The top panels show the true pulsar state $f_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$  (green curve).
+We subtract $f_{\rm em}^{(1)}(t)$ to better illustrate the stochastic wandering of the pulsar frequency.
+In the left column, the blue/green solutions overlap almost perfectly; in the right column, they do not.
+The middle panels show the true measured frequency $f_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve).
+Again we subtract $f_{\rm em}^{(1)}(t)$.
+In the left column the red/magenta solutions overlap almost perfectly; in the right column, they do not.
+The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{(1)}(t) - \hat{f}_{\rm m}^{(1)}(t)$.
 The results are shown for a single pulsar.
-In the right hand column $\Omega$ has been perturbed from its true value by 20\%}
+In the right column $\Omega$ is displaced from its true value by 20\%.}
 	\label{fig:kalman_example}
@@
 The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}.
-For a fixed set of static parameters, the Kalman filter returns an estimate of the the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error.
+For a fixed set of static parameters, the Kalman filter returns an estimate of the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error.
 We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline
@@
 This is illustrated with synthetic data in Figure \ref{fig:kalman_example}.
-In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations (1)--(4), (16), and (17) for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$.
+In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations \eqref{eq:spinevol}-- \eqref{eq:xieqn}, \eqref{eq:measurement}, and \eqref{eq:g_func} for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$.
 The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve).
@@
 The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel).
-By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters to the Kalman filter, where $\Omega$ has been perturbed from its true value by $20 \%$.
+By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$) to the Kalman filter, where $\Omega$ is displaced from its true value by $20 \%$.
 In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates.
@@
 \begin{equation}
-	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
+	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
 \end{equation}
@@
  We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper.
-Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, which also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}.
+It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
 It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
@@
 The primary tunable parameter in nested sampling is $n_{\rm live}$.
-A greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
+More live points assists with large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
 However the computational runtime scales as $\mathcal{O}(n_{\rm live})$.
 \cite{Ashton2022} offered a rule-of-thumb trade-off, where the minimum number of live points should be greater than the number of static parameters.
-Informal empirical tests conducted as part of this paper support the trade-off suggested by \cite{Ashton2022}; we find typically that the true ${\boldsymbol{\theta}}$ is contained within the 90\% credible interval of the derived 1D marginalised posteriors of ${\boldsymbol{\hat{\theta}}}$ for $n_{\rm live} > 7 + 5N$ with $N \leq 50$.
+Informal empirical tests conducted as part of this paper support the trade-off suggested by \cite{Ashton2022}; we find typically that the true ${\boldsymbol{\theta}}$ is contained within the 90\% credible interval of the one-dimensional marginalised posteriors of ${\boldsymbol{\hat{\theta}}}$ for $n_{\rm live} > 7 + 5N$ with $N \leq 50$.
 Unless stated otherwise we take $n_{\rm live} = 1000$ for all results presented in this work.
@@
 Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
-$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
-This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
-The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
+$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data, and is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
+The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW signal compared to there being no GW signal present.
 
@@
 	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
-	\item Constrained sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood $\log \mathcal{L}_{\rm new}$ > $\log \mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (a)--(b).
-	\item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ 
+	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a) -- (vi)(b).
+	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
 	\item Repeat steps (vii) - (ix) until convergence criteria are satisfied.
@@
 The resulting $\mathcal{Z}$ values can then be compared.
-We remind the reader that the above workflow differs from a realistic PTA analysis in on important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
+We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
 The generalization to TOAs is subtle and will be tackled in a forthcoming paper.
@@
 	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
-The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc.
-The grey dashed line denotes the Galactic plane.}
+The pulsar distances relative to the observer are also indicated, with a distance $\leq 2 \, {\rm kpc}$ for $38$ pulsars.
+The grey dashed curve denotes the Galactic plane.}
 	\label{fig:pulsar_distrib}
@@
 We go on in Section \ref{sec:multiple_noise} to extend this parameter estimation procedure to multiple realisations of the noise processes.
-In Section \ref{sec:detection} we calculate the the detectability of the source as a function of $h_0$.
+In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$.
 \newline
@@
 We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
-Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
+Similar results are derived for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
 We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.
@@
 	\end{tabular}
-	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
+	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
 $W_{1, \rm median}$  sets the bound on the difference in expectation value between the one dimensional posteriors, and is generally small with respect to the width of the prior space.}
@@
 	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\alpha-\psi$ parameter space for a single noise realisation.
-The red point labels the location of the maxima, and the magenta point labels the the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data.
+The red point labels the location of the maxima, and the magenta point labels the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data.
 The likelihood has been normalised with respect to the absolute value of the maxima.
@@
 Our goal is to compare these posteriors between different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
-To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
+To this end we use the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
 \footnote{\textcolor{red}{\tiny TK: How much of the following Wasserstein stuff is ok here, how much to move to the appendix?

### 2023-10-19 docs/manuscript/paper.tex
snapshot of more comments

@@
 \begin{equation}
-	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}\ ,
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
 	\label{eq:measurement}
@@
 
-\section{Validation procedure with synthetic data} \label{sec:testing}
-In this section we outline how synthetic data is generated in order to validate the analysis scheme in Section \ref{sec:detect}.
+\section{Validation with synthetic data} \label{sec:testing}
+In this section we outline how synthetic data are generated in order to validate the analysis scheme in Section \ref{sec:detect}.
+Synthetic data enable validation to occur systematically and under controlled conditions.
 In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
-In Section \ref{sec:gendata} we demonstrate how to solve the equations of Section \ref{sec:model} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.
+In Section \ref{sec:gendata} we demonstrate how to solve Equations \eqref{eq:spinevol} -- \eqref{eq:frequency_evolution}, and \eqref{eq:measurement} -- \eqref{eq:g_func} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.
 
@@
 The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
-For each pulsar we adopt the fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
+\newline
+
+For each pulsar we adopt fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
 A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility.
@@
 The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+\newline
+
 The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist.
-The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
+The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency}, and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
 In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
@@
 \end{align}
-where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits.
-The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value.
-Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
-In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
+where $T_{\rm cad}$ is the cadence of the timing observations.
+For MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$ ($\pm 2\sigma$ confidence limits).
+The uncertainties are broad; for the purpose of generating astrophysically representative synthetic data we straightforwardly adopt the central values in this paper.
+Throughout this work we assume for simplicity that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
+In order to relate Equation \eqref{eq:sigmap} to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
 \begin{eqnarray}
@@
 \end{eqnarray}
-For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline
+For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, with $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline
 
@@
 We see that the synthetic and empirical residuals are qualitatively similar.
-For this pulsar the synthetic residuals were generated using $\sigma^{(n)} = 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the value inferred from the noise model of Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, using the best fit parameters, $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$.
-This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars.
+For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ for PSR J1939$+$2134.
+
+%This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars.
 If we set $\ln \alpha_1 = -38$ in Equation \eqref{eq:sigmap}, close to its lower limit, then the estimated $\sigma^{(n)}$ for  PSR J1939+2134 is $\sigma^{(n)} =3.90 \times 10^{-27}$ which is in much better agreement with the value inferred from the residuals comparison method.
-This discrepancy between the two estimates of $\sigma^{(n)}$ is particular to PSR J1939+2134; for all other pulsars we find the two estimates of $\sigma^{(n)}$ are comparable to within a few orders of magnitude.
+
 \begin{figure}
@@
 The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}.
-We set the TOA uncertainty to be a constant $0.1 \mu$s.
+We set the TOA uncertainty to be a constant, $0.1 \mu$s.
 The blue and orange residuals are qualitatively similar by inspection.
@@
 	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
-	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}
+	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}.
 \end{enumerate}
 Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
-The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \eqref{sec:synt_pta}.
+The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \ref{sec:synt_pta}.
 For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
-The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
-\begin{equation}
-	\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
-\end{equation}
-Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
-The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
-Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
-Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar.
-\newline
-
-Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } \sigma^{(n)} \Delta W_t \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta W_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
-To fix the problem, we subtract the deterministic frequency evolution and track the new variable, equivalent to a change of variables:
-\begin{equation}
-	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ .
-\end{equation}  
-We similarly modify the measurement variable to be
-\begin{equation}
-	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ ,
-\end{equation}
-where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris which is measured to high accuracy in practice by {\sc tempo2}.
+\newline
+
+The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, through
+\begin{equation}
+	\sigma_{\rm m} \approx \sigma_{\rm TOA} f_{\rm p}^{(n)}  \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
+\end{equation}
+Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct.
+Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s, Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz.
+Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred for the sake of simplicity; this assumption can be relaxed easily when analysing real data.
+Whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}(t)$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar.
+\newline
+
+Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} \Delta B_t | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta B_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
+To fix the problem, we subtract the deterministic frequency evolution and track the new variable
+\begin{equation}
+	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ , \label{eq:hetero1}
+\end{equation}  
+equivalent to a change of variables. We similarly modify the measurement variable to be
+\begin{equation}
+	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ , \label{eq:hetero2}
+\end{equation}
+where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice.
 For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately.
@@
 \end{equation}
-We emphasise that this change of variables is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g.
+We emphasise that the change of variables in Equations \eqref{eq:hetero1} and \eqref{eq:hetero2} is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g.
 long double, quadruple).
@@
 In particular $f_{\rm em}^{(n)}(t_1)$
-and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \ref{eq:spinevol}--\eqref{eq:xieqn}.
-
-\section{Representative example: single SMBHB source}\label{sec:rep_example}
-In this Section we apply the analysis scheme of Section \ref{sec:detect} and the validation procedure of Section \ref{sec:testing} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-We remind the reader that the analysis in this paper targets a single SMBHB source.
+and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.
+
+\section{Representative end-to-end analysis} \label{sec:rep_example}
+In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
 The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-In Section \ref{sec:priors} we outline the priors chosen on $\boldsymbol{\theta}$.
-In Section \ref{sec:parameter_estim} we apply the workflow of Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
-We go on in Section \ref{sec:multiple_noise} to extend this parameter estimation procedure to multiple realisations of the noise processes.
+The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space}--\ref{sec:discussion}.
+To assist with reproducibility, we present intermediate outputs from each step.
+In Section \ref{sec:priors} we define the priors chosen on $\boldsymbol{\theta}$.
+
+In Section \ref{sec:earth_psr_terms} we do this...
+
+\textcolor{red}{TK to do}
+
+In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
+In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates.
 In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$.
@@
 
-The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected to be astrophysically reasonable and representative.
-The injected $\boldsymbol{\theta}_{\rm gw}$ are summarised in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}.
-The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$ are as described in Section \ref{sec:synt_pta} and are also shown in Table \ref{tab:parameters_and_priors}.
-
-\subsection{Prior distributions}\label{sec:priors}
-For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters.
+The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
+The injected elements of $\boldsymbol{\theta}_{\rm gw}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
+The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$, described in Section \ref{sec:synt_pta}, are also recorded in Table \ref{tab:parameters_and_priors}.
+
+\subsection{Prior distribution}\label{sec:priors}
+The first step is to select a reasonable Bayesian prior, $\pi(\boldsymbol{\theta})$, for the static parameters.
 For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
-The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires some additional discussion.
-\newline
-
-The parameters which govern the deterministic evolution of the pulsar spin frequency, $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ are well-determined by existing radio timing observations.
-We can identify $f_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the time derivative of barycentric rotation frequency quoted by pulsar observation catalogues.
-For the 12.5 year NANOGrav pulsars the median error in  the barycentric rotation frequency is $\sim 7 \times 10^{-13}$ Hz, and in the time derivative of barycentric rotation frequency is $\sim 1.8 \times 10^{-20}$ s $^{-2}$.
-Whilst it is then clear that $f_{\rm em}^{(n)} (t_1)$, $\dot{f}_{\rm em}^{(n)} (t_1)$ are typically measured with very high precisions, for this work we adopt a broader uniform prior, $\pm 10^3 \epsilon^{(n)}_{a}$ either side of the true value, where $\epsilon^{(n)}_{a}$ is the error for that pulsar quoted in the ATNF catalogue and the subscript $a = \{ f, \dot{f} \}$ identifies the error as being in $f_{\rm em}^{(n)} (t_1)$, or $\dot{f}_{\rm em}^{(n)} (t_1)$ respectively.
-In this way we can test how well our method performs without requiring exceptionally precise and accurate measurements of the pulsar parameters to be made \textit{a priori}.
-Instead we consider the pulsar parameters similarly to the GW source parameters and estimate them optimally within a consistent framework.
-Adopting looser priors in this way allows us to subject our method to a more stringent test and confirm that we still accurately estimate GW parameters without highly accurate and precise initial estimates of $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$.
-\newline
-
-The pulsar distances $d^{(n)}$ are typically not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with uncertainties typically on the order of $\sim 10 \%$ \citep{Arzoumanian2018ApJS..235...37A,Yao2017}.
-However for this work, whilst the pulsar distance is used when generating synthetic data, as is discussed in Section \ref{sec:parameter_estim} the pulsar distance is not used for our inference model and so we do not need to set a prior on $d^{(n)}$.
-Similarly we do not set a prior on $\gamma^{(n)}$; since $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ this means that $\gamma^{(n)}$ is effectively ``unobservable" over the decadal timescales that we are interested in.
-That is, for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is effectively independent of the choice of $\gamma^{(n)}$ as long as the condition $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
-It is therefore sufficient to consider $\gamma^{(n)}$ to be known \textit{a priori} and set it at its true injected value.
-We briefly explored setting an uninformative prior on $\gamma^{(n)}$ over e.g.
-LogUniform($10^{-15}, 10^{-10}$) s$^{-1}$ as well as setting $\gamma^{(n)}$ at some fixed value away from the true injected value (e.g.
-set $\gamma^{(n)} = 10^{-14} \text{s}^{-1}$ rather than $10^{-13} \text{s}^{-1}$) but the results are unchanged.
-Regarding $\sigma^{(n)}$, the majority of pulsars in our synthetic PTA have $\sigma^{(n)} \sim 10^{-20} - 10^{-22} \text{ s}^{-3/2}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
-For these pulsars we set an uninformative broad prior of LogUniform($10^{-23}, 10^{-19}$ $\text{s}^{-3/2}$).
-The single exception is PSR J1939+2134 which has a particularly large $\dot{f}_{\rm em}^{(n)} (t_1)$ compared to the other pulsars in the array and so $\sigma \sim 10^{-16} \text{s}^{-3/2}$.
-For our artificial dataset we set $\sigma = 10^{-21} \text{s}^{-3/2}$ for this pulsar in order to bring the process noise for all pulsars into a consistent range that can be described by a single constrained prior.
-\newline
-
-Since we are not setting priors on $\gamma^{(n)}$ or $d^{(n)}$ this reduces the dimensionality of the parameter space to $7 + 3N$.
-We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space, c.f.
+The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires additional discussion.
+\newline
+
+The parameters that govern the deterministic evolution of $f_{\rm p}^{(n)}(t)$, namely $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, are well-determined by radio timing observations.
+We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues \citep{Manchester2005}.
+For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} \%$ and $\pm 2.31 \times 10^{-3} \%$ respectively.
+In this paper we adopt uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database \citep{Manchester2005}.
+Adopting more relaxed priors tests the method stringently.
+The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are relaxed or narrow.
+\newline
+
+The pulsar distances $d^{(n)}$ are not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with typical uncertainties $\sim 10 \%$ \citep{Yao2017,Arzoumanian2018ApJS..235...37A}.
+In this paper, however, we omit the term involving $d^{(n)}$ from Equation \eqref{eq:g_func}, as justified in Section \ref{sec:parameter_estim}, so there is no need to set a prior on $d^{(n)}$.
+Furthermore we do not set a prior on $\gamma^{(n)}$, because we have $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and $\gamma^{(n)}$ is effectively ``unobservable" over a decade; that is,for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is approximately independent of $\gamma^{(n)}$, as long as $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
+It is therefore sufficient for validation purposes to carry $\gamma^{(n)}$ through the analysis at its injected value.
+\footnote{We check this empirically in two independent ways for an informal selection of test cases: (i) we set an uninformative prior $\pi\left [\gamma^{(n)} / 1 \, {\rm s^{-1}} \right] \sim$ LogUniform($10^{-15}, 10^{-10}$) , and (ii) we deliberately displace $\gamma^{(n)}$ in the inference analysis from its true, injected value in the synthetic data, e.g.
+$\gamma^{(n)} = 10^{-14} \text{ s}^{-1}$ versus $10^{-13} \text{ s}^{-1}$, respectively.
+The results in (i) and (ii) are the same as those reported in Section \ref{sec:rep_example}.} \newline
+
+Most pulsars in the synthetic PTA have $10^{-22} \leq \sigma^{(n)} / (1 \, {\rm s^{-3/2}}) \leq 10^{-20}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
+For these pulsars we set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$).
+The sole exception is PSR J1939+2134 which has $\sigma^{(n)} \sim 10^{-16} \text{s}^{-3/2}$.
+For validation purposes, we artificially set $\sigma^{(n)} = 10^{-20} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within the prior.
+This is done purely for testing convenience; it is straightforward to expand the prior when analysing real, astronomical data.
+\newline
+
+By not setting priors on $\gamma^{(n)}$ and $d^{(n)}$ we reduce the number of parameters to $7 + 3N$.
+We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter set, c.f.
 Equation \ref{eq:psrparams}.
-Explicitly,
-\begin{equation}
-	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ .
-\end{equation}
-All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
+Explicitly, we write
+\begin{equation}
+	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ . \label{eq:psrreduced}
+\end{equation}
+The injected static parameters and their corresponding priors for the representative analysis in Section \ref{sec:rep_example} are summarised in Table \ref{tab:parameters_and_priors}.
 \begin{table*}
@@
 		\toprule
-		&Parameter & Injected Values & Units & Prior  \\
+		&Parameter & Injected values & Units & Prior  \\
 		\hline
 		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
-	  & $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
-	  & $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
-	  & $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
-	  & $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
-	  & $h_0$            & $5 \times 10^{-15}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
-	  & $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
+	  & $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
+	  & $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
+	  & $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
+	  & $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
+	  & $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
+	  & $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
 		\hline
-		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left( f_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{f} \right)$ \\
-		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left( \dot{f}_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{\dot{f}} \right)$ \\
-		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
+		\vspace{1mm}\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
+		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
+		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
 		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
-		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
+		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
 		\bottomrule
 	\end{tabular}
-	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter.
-The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.
-The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
+	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the prior used for Bayesian inference on each parameter (rightmost column).
+The top and bottom halves of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ respectively.The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
+The subscript ``sc" indicates that the injected value has been calculated using Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} in \protect \cite{Shannon2010}.
+The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
+No priors are set for $d^{(n)}$ and $\gamma^{(n)}$ since those parameters do not enter the inference model,  as discussed in Section \ref{sec:priors}, c.f.
+Equation \eqref{eq:psrreduced}.
+}
 	\label{tab:parameters_and_priors}
@@
 
-\subsection{Parameter estimation}\label{sec:parameter_estim}
-In this section we derive the posterior probability distributions for the static parameters.
-However, it is first necessary to discuss a modification to the general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}.
-It can be seen that the equation generally separates into two cosine terms.
+\subsection{Earth and pulsar terms}\label{sec:earth_psr_terms}
+The general measurement equation used in the inference model, Equation \eqref{eq:g_func_trig}, separates into two cosine terms.
 The first term, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
-The argument of the cosine corresponds to the GW phase at the observer on Earth.
-Conversely the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
-The argument of the cosine corresponds to the GW phase at each the individual pulsar.
+The argument of the cosine corresponds to the GW phase at the Earth.
+The second term, $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d \right \}$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and varies between pulsars.
+The argument of the cosine corresponds to the GW phase at each individual pulsar.
 The first and second terms are commonly referred to as the ``Earth term" and ``pulsar term" respectively.
-Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases and in standard PTA analysis are typically considered as a source of self noise and dropped from the analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
-In this work we follow the standard analysis approach and drop the pulsar terms from our model.
-Explicitly the measurement equation used in the Kalman filter is
+Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases.
+They are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
+In this work we follow the standard approach and drop the pulsar terms from the inference model (but not the source model).
+Explicitly the measurement equation used in the Kalman filter reduces to
 \begin{equation}
@@
 	\end{equation}
-	We defer the inclusion of the pulsar terms to a future work.
+	We defer the inclusion of the pulsar terms to a future paper.
 As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
-We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e.
+We stress that the pulsar terms are dropped only when doing Bayesian inference, i.e.
 from the Kalman filter model that feeds into the nested sampling algorithm.
-The generated synthetic data that we use to test our method does include the pulsar term in full.
-\newline
-
-	Initially we consider a single noise realisation of the synthetic data for the representative example system described in Table \ref{tab:parameters_and_priors}.
-We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters.
-The results are shown in Figure \ref{fig:corner_plot_1} for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$.
-The histograms shown in the figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value.
-The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
-It is evident that for this representative example with characteristic strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
-\newline
-
- For some of the parameters the estimated posterior is approximately symmetric about the true injected value (e.g.
-$\Omega$).
-In this case the posterior median and the injected value are approximately coincident.
+The synthetic data include the pulsar term in full.
+\newline
+
+	\subsection{Posterior distribution}\label{sec:parameter_estim}
+	In this section, we calculate the posterior probability distribution for the static parameters $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$ and compare it to the known, injected values.
+The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior), and (ii) to give a preliminary sense of its accuracy.
+Initially we consider a single noise realisation when generating the synthetic data for the representative example in Table \ref{tab:parameters_and_priors}.
+We apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$.
+\newline
+
+Figure \ref{fig:corner_plot_1} displays results for the seven parameters in  $\boldsymbol{\theta}_{\rm gw}$ in the form of a traditional corner plot.The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
+The dashed vertical blue lines mark the 0.16 and 0.84 quantiles; the solid orange line marks the known injected value.
+The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces.
+All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values.
+There is scant evidence of railing against the prior bounds.
+Hence it is evident that for this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, that the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
+The injected values are contained within the 90\% credible interval for all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ .
+\newline
+
+The posterior is approximately symmetric about the known injected value for some parameters such as $\Omega$, where the posterior median and the injected value are coincide approximately.
 For other parameters (e.g.
-$\iota$) the distribution is not symmetric about the injected value and the posterior median and the injected value are not coincident.
-The median value of the posterior for $\iota$ is shifted by $\sim 0.35$ rad relative to the injected value, although the injected value does still remain inside the 90\% credible interval.
+$\iota$) the distribution is not symmetric about the injected value and the posterior median and the injected value do not coincide.
+The median value of the posterior for $\iota$ is shifted by $\approx 0.35$ rad relative to the injected value, although the injected value does still remain inside the 90\% credible interval.
 Similar effects are seen, albeit with a smaller shift, in other parameters such as $\delta$ and $\alpha$.
 For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
-a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation.
-We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
-Similar results are derived for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
-We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.
+a bias or a random outcome specific to this particular noise realisation.
+We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} using more noise realizations and show that indeed there is a bias from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
+\newline
+
+Similar results are obtained for the 3$N$ parameters in $\boldsymbol{\theta}_{\rm psr, reduced}$.
+Again, the parameters are recovered unambiguously, in the sense that the nested sampler converges smoothly to a unimodal joint posterior near the known, injected parameter values, with all $3N$ injected parameters lying within the 90 \% credible interval of the $3N$ one-dimensional, marginalized posteriors.
+We do not display the resulting corner plot because it is too big ($3N = 141$), and because inferring ${\boldsymbol{\theta}}_{\rm gw}$ is the focus of this paper and contemporary PTA analyses more generally.
 
@@
 
-\subsection{Multiple noise realisations} \label{sec:multiple_noise}
+\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
 \begin{figure*}
@@
 \end{figure}
-The above example was for a single realisation of the noise processes.
-It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous or convenient for our method.
-In this Section we confirm that we are able to infer comparable results for different noise realisations.
-To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise and measurement noise.
-For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$.
-In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise.
-This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}.
-We plot only nine realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure.
-As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr, reduced}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here.
-\newline
-
-We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1}.
-Considering the one-dimensional marginalized posteriors, the inferred probability distributions of the parameters broadly overlap across the different noise realisations.
-For the vast majority of parameters and noise realisations the injected value is contained within the 90\% credible interval.
+The results in Section \ref{sec:parameter_estim} are for a single realisation of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
+It is important to confirm that the analysis scheme returns accurate answers for arbitrary noise realisations and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous by accident.
+It is also important to quantify, albeit approximately, the natural random dispersion in the one-dimensional posterior medians from one noise realization to the next, as the dispersion is a practical measure of the accuracy of the parameter estimation scheme, when it is applied to real astronomical data, where the true parameter values and specific noise realisation are unknown.
+\newline
+
+To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
+For each realisation we independently estimate the static parameters, $\boldsymbol{\theta}$.
+In Figure \ref{fig:corner_plot_2} we plot the estimates of $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary realisations.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, which stems from one realisation.
+We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
+As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm gw}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
+\newline
+
+Figure \ref{fig:corner_plot_2} confirms two main points (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim 10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
+
+ Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval.
 As in the single noise case, there are tentative signs of bias for some of the parameters, with the one-dimensional posteriors not symmetric about the injected value.
@@
 To this end we use the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
-\footnote{\textcolor{red}{\tiny TK: How much of the following Wasserstein stuff is ok here, how much to move to the appendix?
-I don't want to get bogged down in WD and distract from astro arguments, but we use $W_1$ as metric so I feel like it should be defined in the main body of the text, and the bounding result is important...} \normalsize} The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
+The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
 A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}.
@@
 For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the figure, corresponds to a posterior for $\psi$ which has converged sub-optimally, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
-The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ radians.
+The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ rad.
 We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
@@
 This is the origin of the bias in the one-dimensional marginalised posteriors that we observe.
-For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ radians, whereas the injected value was $\iota = 1.0$ radians.
+For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ rad, whereas the injected value was $\iota = 1.0$ rad.
 This optima is comparable to the median of the posterior inferred for $\iota$ in Figure \ref{fig:bias_for_large_h}.

### 2023-10-23 docs/manuscript/internal_comments_4/difference_file/new.tex
returned package to AM

--- 2023-10-19 docs/manuscript/paper.tex
@@
 
-\usepackage[section]{placeins} %ensures figures go in their section e.g https://tex.stackexchange.com/questions/279/how-do-i-ensure-that-figures-appear-in-the-section-theyre-associated-with
+%
+%\usepackage[section]{placeins} %ensures figures go in their section e.g https://tex.stackexchange.com/questions/279/how-do-i-ensure-that-figures-appear-in-the-section-theyre-associated-with
 
@@
 Similar results are obtained for the 3$N$ parameters in $\boldsymbol{\theta}_{\rm psr, reduced}$.
-Again, the parameters are recovered unambiguously, in the sense that the nested sampler converges smoothly to a unimodal joint posterior near the known, injected parameter values, with all $3N$ injected parameters lying within the 90 \% credible interval of the $3N$ one-dimensional, marginalized posteriors.
+Again, the parameters are recovered unambiguously, in the sense that the nested sampler converges smoothly to a unimodal joint posterior near the known, injected parameter values, with all $3N$ injected parameters lying within the 90\% credible interval of the $3N$ one-dimensional, marginalized posteriors.
 We do not display the resulting corner plot because it is too big ($3N = 141$), and because inferring ${\boldsymbol{\theta}}_{\rm gw}$ is the focus of this paper and contemporary PTA analyses more generally.
@@
 	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
-	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
-The vertical orange lines indicate the true injected values.
-The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
-We are able to accurately estimate each parameter of interest.
+	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The horizontal and vertical orange lines indicate the true injected values.
+The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters.
+The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter.
+The supertitles of the marginalized one-dimensional histograms specify the posterior median and the 0.16 and 0.84 quantiles.
 We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
-Note that the $x$-axis scaling does not span the total prior space.}
+The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+The horizontal axes span a subset of the prior domain for all seven parameters.}
 	\label{fig:corner_plot_1}
@@
 	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
-	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes.
-The subtitles of the marginalized 1D posteriors refer to values for the median posterior across the noise realisations.
-The injection value is generally contained within the 90 \% credible interval for the vast of majority posteriors.
-The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
+	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely.
+The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation.
+The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations.
+There is an appreciable dispersion among the peaks of the one-dimensional posteriors is appreciable, with coefficient of variation $\sim10 \%$ across the 63 combinations.
 A slight skew-bias is apparent in some of the parameters, e.g.
 $\iota$.
-Correlations can be seen between $\Omega - \Phi_0$, $\psi-\alpha$ and $\iota - h_0$.}
+Weak correlations can be seen between $\Omega - \Phi_0$, $\psi-\alpha$ and $\iota - h_0$.}
 	\label{fig:corner_plot_2}
@@
 	\end{tabular}
-	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
-$W_{1, \rm median}$  sets the bound on the difference in expectation value between the one dimensional posteriors, and is generally small with respect to the width of the prior space.}
+	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
+$W_{1, \rm median}$ is generally smaller than the prior domain.}
 	\label{tab:Wasserstein}
 \end{table}
+
+The results in Section \ref{sec:parameter_estim} are for a single realisation of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
+It is important to confirm that the analysis scheme returns accurate answers for arbitrary noise realisations and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous by accident.
+It is also important to quantify, albeit approximately, the natural random dispersion in the one-dimensional posterior medians from one noise realization to the next, as the dispersion is a practical measure of the accuracy of the parameter estimation scheme, when it is applied to real astronomical data, where the true parameter values and specific noise realisation are unknown.
+\newline
+
+To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
+For each realisation we independently estimate the static parameters, $\boldsymbol{\theta}$.
+In Figure \ref{fig:corner_plot_2} we plot the estimates of $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary realisations.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, which stems from one realisation.
+We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
+As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm gw}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
+\newline
+
+Figure \ref{fig:corner_plot_2} confirms two main points (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
+Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 out of the 63 (i.e.
+$92 \%$) possible combinations of the seven parameters and nine realizations.
+Figure \ref{fig:corner_plot_2}, like Figure \ref{fig:corner_plot_1}, displays tentative signs of bias, where the one-dimensional posteriors are not symmetric about the injected value.
+For example, the maximum a posteriori probability estimates of $\iota$ appear left-skewed, consistently underestimating the injected value in all nine realisations.
+A similar but messier trend is seen in $\delta$.
+However, the width of the posteriors is comparable to the putative bias, so it is difficult to draw strong conclusions.
+Bias is discussed in detail in Section \ref{sec:bias}.
+There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.
+\newline
+
+We now consider the complete set of $10^3$ noise realisations, going beyond than the subset of nine realisations in the preceding discussion.
+We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$?
+There is no unique way to answer this question.
+In this paper we appeal to the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory, which defines an intuitive notion of similarity between probability distributions.
+The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}; a short review is presented in Appendix \ref{sec:wasserstein}.
+The WD measures the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
+In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}.
+Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$" with respect to the PDFs $\mu$ and $\nu$.
+Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad.
+Then we can infer $| \langle \iota \rangle_\mu - \langle \iota \rangle_\nu | \leq 0.5 \, {\rm rad}$.
+\newline
+
+Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$.
+As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
+These values are an approximate measure of the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
+Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.14 \times 10^{-4}$ \% for $h_0$ to 2.69\% for $\psi$.
+These values serve as confirmation that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for a wide variety of noise realizations.
+Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
+
+\subsection{Detectability versus $h_0$} \label{sec:detection}
+We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}.
+In Equatin \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-terms-only model, i.e.
+the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
+To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
+\footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, simply adds uninformative scatter to the trend in Figure \ref{fig:bayes}.} \newline
+
+We see in Figure \ref{fig:bayes} an approximate quadratic relationship $\beta \propto h_0^2$ exists.
+The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 4 \times 10^{-15}$.
+Of course, the minimal detectable strain is particular to the system in Table \ref{tab:parameters_and_priors}.
+It is influenced in general by e.g.
+$T_{\rm obs}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$, as discussed in Section \ref{sec:parameter_space}.
+Adjusting $\sigma_{\rm m}$ linearly scales the effective signal-to-noise ratio $h_0 / \sigma_{\rm m}$, such that an approximate quadratic relationship $\beta \propto h_0^2 / \sigma_{\rm m}^2$ also exists.
+\newline
+
+The odds ratio $\beta$ drops off for $h_0 \lesssim 4 \times10^{-15}$.
+This happens because the two competing models becoming increasingly indistinguishable once the measurement noise dominates the GW signal.
+Moreover, the points in Figure \ref{fig:bayes} become sparser for $h_0 \lesssim 4 \times 10^{-15}$.
+This happens because we find $\beta <0$ for the missing, intermediate points.
+This is a noise artefact of the nested sampler.When the sampler converges sub-optimally, the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails, and $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+There is nothing special about the particular missing points; if one recreates the $\beta(h_0)$ curve by rerunning the nested sampler with another random seed, a different set of points are missing.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$, because the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(n_{\rm live}^{-1/2})$ (see Section \ref{sec:nested_sampling}).
+In this section we set $n_{\rm live} =2000$.
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
+	\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW amplitudes, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
+The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
+The minimum detectable strain for $\beta < 10$, is $ 4 \times 10^{-15}$.
+Missing points for $\beta\leq 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
+	\label{fig:bayes}
+\end{figure}
+
+\section{Exploring a broader parameter space} \label{sec:parameter_space}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/pp_plot_new}
+	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}.
+Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$.
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
+The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals.
+Well estimated posteriors should fall along the diagonal $y=x$.
+$\Omega$ is generally well-estimated but many other parameters show evidence of being over-constrained due to a modelling bias.
+}
+	\label{fig:parameter_space}
+\end{figure}
+We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
+In this section we test the method in different regions of the parameter space, varying the source parameters through astrophysically relevant ranges.
+\newline
+
+We consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space.
+For each simulated injection we can then attempt to recover the posterior distributions for each parameter.
+To summarise the results across the parameter space we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+A PP plot describes the fraction of the total number of injected parameters which are included within a given credible interval of the estimated posterior with respect to the credible interval itself.
+In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage credible interval.
+The results are shown in Figure \ref{fig:parameter_space}.
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections.
+We can see that only $\Omega$ falls within the $3\sigma$ shaded region.
+The other parameters deviate from the $y=x$ diagonal.
+The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$.
+The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and there are more injections contained within the low value credible intervals.
+This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally confident, but are systematically biased away from the true injection values.
+This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval.
+An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.
+
+\section{Identifiability and bias} \label{sec:bias_and_identifiability}
+There are two underlying issues in the preceding tests that we now explore in more detail.
+The first relates to parameter identifiability, i.e.
+can we uniquely identify the parameters of the model given the data and the model structure?
+This question is explored in Section \ref{sec:identif}.
+The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
+This is the focus of Section \ref{sec:bias}.
+In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f.
+Figure \ref{fig:bayes}) for our representative example system (Table \ref{tab:parameters_and_priors}).
+
+\subsection{Identifiability}\label{sec:identif}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
+	\caption{Posterior probability distributions for the representative system described in Table \ref{tab:parameters_and_priors} over nine noise realisations for the GW source parameters $\iota$ and $h_0$ in the high SNR regime ($h_0 = 10^{-12}$).
+Unlike the other parameters of  $\boldsymbol{\theta}_{\rm gw}$, the one-dimensional posteriors for $\iota$ and $h_0$ do not strongly overlap.
+A correlation between  $\iota$ and $h_0$ is evident.}
+	\label{fig:just_iota_h}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_surface}
+	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation.
+The red point labels the location of the likelihood maxima.
+The likelihood surface has been normalised with respect to the absolute value of the maxima.
+A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
+	\label{fig:likelihood_surface}
+\end{figure}
+In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations. 
+The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but just for $\iota$ and $h_0$ rather than all members of $\boldsymbol{\theta}_{\rm gw}$.
+In the case of high-SNR we might expect \textit{a priori} that our estimates would be increasingly accurate over the low-SNR case, and that the method would converge robustly to similar posteriors irrespective of the noise realisation (i.e.
+$W_1 \to 0$).
+This is generally true for the majority of the component parameters of $\boldsymbol{\theta}_{\rm gw}$; the estimated posteriors across noise realisations are largely indistinguishable.
+However for $\iota$ and $h_0$ this is not the case.
+Instead the inferred one-dimensional posteriors do not broadly overlap and are not consistent across different noise realisations.
+Specifically across $10^3$ noise realisations the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
+For $\iota$, $W_{1,\rm median}$ is greater than the value inferred in the low SNR case ($=0.14$, Table \ref{tab:Wasserstein}), whilst for $h_0$ $W_{1,\rm median}$ is also greater as a proportion of the scale set by the strain amplitude.
+This is in contrast to $W_{1,\rm median}$ inferred for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ in the high-SNR regime where as SNR increases, $W_1$ decreases; e.g.
+for $\Phi_0$, $W_{1, \rm median} = 4.5 \times 10^{-3}$ rad.
+\newline
+
+This issue of inferring very different posteriors depending on the particular realisation of the noise is due to these parameters being only weakly-identifiable.
+Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
+Sources of non-identifiability are generally categorised as either ``structural", arising from the structure of the model, or ``practical", arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
+Regarding $\iota$ and $h_0$, we are concerned with structural identifiability.
+Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx} and \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
+For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
+This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
+It can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
+That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a small, non-zero gradient close to the maximum likelihood solution.
+This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
+Whilst a single, unique maxima exists (labelled by the red point in the figure), along the ridge in the parameter space the likelihood values are all very similar.
+Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
+If during the inference we fix one of the parameters at its true value (i.e.
+set a delta function prior), taking a slice through the likelihood surface it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably, i.e.
+$W_1 \to 0$.
+In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" \footnote{\tiny \textcolor{red}{TK: "washes out" may be too colloquial / unspecific, but I think it is intuitive...}\normalsize}the likelihood ridge, leading to a more concave likelihood surface.
+This enables consistent posteriors to be estimated for $\iota$ and $h$.
+In the case where we fix e.g $h_0$ at its true value to infer indistinguishable posteriors for $\iota$, all of the posteriors for $\iota$ are all strongly biased away from the injected value; the maximum a posteriori probability estimates underestimate the injected value by $\sim 0.34$ rad.
+This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be now be discussed further in Section \ref{sec:bias}.
+\subsection{Bias by dropping pulsar terms}\label{sec:bias}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
+	\caption{As Figure \ref{fig:corner_plot_1} but for the high-SNR system with $h_0 = 10^{-12}$.
+For the majority of parameters the injection value does not fall within the 90\% credible interval and is not visible over these scales.
+The estimated posteriors are biased away from the injected value due to the dropping of the pulsar terms.
+Note that the scale of the $x$-axes are generally highly constrained compared to Figure \ref{fig:corner_plot_1}.}
+	\label{fig:bias_for_large_h}
+\end{figure*}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
+	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and the Earth terms model for Earth terms only synthetic data (green line), with respect to the inclination $\iota$, over the prior space.
+All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
+The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model; the green/orange dashed lines are overlaid on this scale.
+For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$).
+For the pulsar-terms model the location of the maxima and injected value do coincide.
+10 noise realisations are plotted for each situation but the solutions are completely overlaid on this scale.}
+	\label{fig:likelihood_surface_iota}
+\end{figure}
+Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g.
+the median of the one-dimensional marginalised posterior) is biased away from the true injection value.
+In Figure \ref{fig:bias_for_large_h} we show the resulting posteriors for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but now for the high-SNR system with $h_0 = 10^{-12}$.
+We can see that for all parameters, with the exception of $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it is not visible on the scales over which we have plotted the posteriors.
+This effect is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
+Similar results are obtained for other noise realisations.
+Whilst $h_0$ does not exhibit any noticeable bias for this particular noise realisation, in general across multiple noise realisations it does exhibit a correlated bias with $\iota$.
+In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
+\newline
+
+In Figure \ref{fig:likelihood_surface_iota} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, as we vary $\iota$ across the prior space, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
+We do this for 10 separate realisations of the system noise.
+We consider three separate situations: \textit{(i)} The solid blue line is the likelihood curve with just the Earth-terms included (i.e.
+using Equation \ref{eq:measuremen_earth} in the Kalman filter), \textit{(ii)} the solid orange line is the likelihood curve with the additional pulsar terms included (i.e.
+using Equation \ref{eq:measurement}), \textit{(iii)} the solid green line is the solution where we drop the pulsar terms from both the synthetic data and the measurement equation.
+We have not used nested sampling for situations \textit{(ii)} or \textit{(iii)} in this work.
+However, we are able to run the Kalman filter either using the pulsar terms in the model or else dropping them from the synthetic data consider how the resulting likelihood varies for different parameter values.
+The dashed lines show the location of the likelihood maxima for each of the likelihood solutions; finding the location of the maxima across all static parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
+\newline
+
+The key observation from Figure \ref{fig:likelihood_surface_iota} is that for the Earth-terms only solution the location of the likelihood optima and the location of the injected value are not the same.
+Consequently Bayesian likelihood inference techniques settle in optima that do not necessarily align with the true injection value.
+This is the origin of the bias in the one-dimensional marginalised posteriors that we observe.
+For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ rad, whereas the injected value was $\iota = 1.0$ rad.
+This optima is comparable to the median of the posterior inferred for $\iota$ in Figure \ref{fig:bias_for_large_h}.
+\newline
+
+In contrast to the Earth-terms solution, for the pulsar-terms solution the optima of the likelihood curve now aligns with the true injected value.
+The corollary is also true; if we drop the pulsar terms from the synthetic data and use the Earth terms equation for inference then the optima is also correctly located.
+This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.
+Similar results are observed for the other parameters which exhibit a bias in the nested sampling inference.
+\newline
+
+Why does dropping the pulsar terms bias these parameters in particular, whilst leaving others unaffected?
+For instance, why is the bias induced in $\iota$ so strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
+When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is effective modulated by the interference of two equal-amplitude, phase shifted waves (c.f.
+Equation \ref{eq:measurement}).
+Explicitly, we have the ``Earth-terms wave"	 with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
+The net interference wave that modulates $f_{\rm p}$ is
+\begin{eqnarray}
+	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ .
+\end{eqnarray}
+The additional modulations from $y_{\rm pulsar}$ influence the amplitude of the resulting interference wave, but do not affect its frequency.
+When we model this data using the Kalman filter without the pulsar terms, the additional amplitude modulations from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
+As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude of the single wave.
+Consequently the parameters which are most affected are those which determine the amplitude of the GW.
+Terms which influence the amplitude strongly, such as $h_0$ and $\iota$, are strongly affected and exhibit a strong bias.
+Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
+The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
+Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
+This hierarchy of biases for each parameter is also what we observe in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
+\newline
+
+In order to ascertain how important the bias is for PTA continuous wave searches requires a thorough exploration of the astrophysical parameter space which we do not carry out in this work.
+The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
+If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate and the bias may not be important for searches.
+This is what we observe for our synthetic data (e.g.
+Figure \ref{fig:corner_plot_2}); the decreased SNR manifests as a broader marginalised posterior which generally overwhelms the shift due to the bias.
+Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may come to systematically affect the inferred parameters.
+\newline
+
+We have shown in this section how the dropping of the pulsar terms results in a bias in the inferred parameters, not just in the sky localisation parameters  \citep[e.g.][]{Zhupulsarterms,Chen2022} but additionally all parameters which influence the wave amplitude.
+Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for our example system, it is present and may need to be accounted for for analysis in other parts of the parameter space.
+This bias is not particular to our method but a shared feature across likelihood based methods that do not include the pulsar terms in the inference model.
+The inclusion of the pulsar terms in the methods presented in this work in order to alleviate the bias will be a key goal of a future work.
+
+\section{Discussion}\label{sec:discussion}
+We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
+There are a number of potential extensions of this work which we now discuss.
+\newline
+
+The natural first extension is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
+Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred one-dimensional marginalized posteriors for the model parameters.
+Biases in sky localisation for synthetic datasets as a result of dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.
+Alternatively, if the method continues to be used with solely the Earth terms then it would be highly desirable to make systematic, quantitative estimates of the incurred biases in the model parameters across an astrophysically representative parameter space.
+\newline
+
+We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
+Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA.
+Indeed, since adding pulsars to PTAs increases the computational demands of the data analysis it may be advantageous to optimally select a subset of the total available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.
+\newline
+
+Throughout this work we have assumed that all pulsars are observed at the same time, with a constant time sampling.
+In practice this assumption does not hold and different pulsars will be observed with different cadences at different times.
+Extending this method to non-constant time sampling is straightforward and the performance of the method should be evaluated in this regime.
+\newline
+
+We have taken our GW source to be monochromatic, i.e.
+$\Omega$ is a fixed parameter with no time dependence.
+Whilst this is an astronomically well-justified assumption (see discussion in Section \ref{sec:plane_gw}), and appropriate for this initial methods work, it would also be of interest to extend the state-space construction to enable the GW frequency to evolve in time.
+The GW source is monochromatic over the decadal PTA observation period, but the timescale set by the Earth-pulsar distance is comparable to the SMBHB evolution timescale \citep{Sesana2010}.
+For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) the GW source frequency evolution manifests as an incoherent source of noise in the pulsar terms, whilst for  $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms can affect the phase coherency of the Earth terms \citep{Perrodin2018}.
+Careful consideration of the evolution of the GW frequency will be important as we look to extend our method to include the pulsar terms in the inference model.
+\newline
+
+We have also assumed that there is only a single GW source that influences the received pulsar frequencies.
+However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
+Our method naturally extends to detecting multiple GW sources simultaneously; Equation \eqref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
+The stochastic background itself is the incoherent summation of many individual GW sources.
+It therefore stands to reason that it may be possible to extend this method to also search for a stochastic GW background, providing a complementary approach to detect the stochastic background without requiring a cross-correlation of the pulsar residuals to uncover the Hellings-Downs curve.
+\newline
+
+\textcolor{red}{TK: Lets also pre-empt reviewer and make it explicit why we are using frequencies rather than TOAs. Need to chat with group to get this clear in my head.}
+
+%https://arxiv.org/pdf/2107.03047.pdf
+\section{Conclusion}\label{sec:conclusion}
+In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
+We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
+That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW induced modulations.
+This enables the inference of GW parameters conditioned on the specific observed realisation of the noisy data.
+We introduce the Kalman filter in order to track the intrinsic state evolution of the pulsar and combine this with a Bayesian sampling framework to estimate the posterior distributions of each parameter of the system, as well as the associated evidence (marginalised likelihood) of the model.
+We test our method on synthetic data.
+We initially focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
+The parameters of the model are accurately recovered and the minimal detectable strain is estimated to be $h_0 \sim 4 \times 10^{-15}$ for $\iota=1.0$ rad.
+The method is then further tested over 1000 noise realisations.
+Consistent posterior estimates of the parameters are obtained for the majority of noise realisations, with the median Wasserstein distance across all noise realisations $W_{1, \rm median}$ generally small compared to the prior space ($\lesssim 3 \%$).
+For a minority of noise realisations the nested sampler converges poorly and settles into a local likelihood optima for the posteriors estimates of $\psi$ and $\alpha$.
+This problem can be solved in general by increasing the number of live points $n_{\rm live}$ used by the sampler.
+Exploration of a broader parameter space via 200 randomly sampled parameter vectors highlights biases in some parameters as a result of dropping the pulsar terms from the inference model.
+The cause of these biases is examined for the case of high SNR, along with the structural weak-identifiability between $\iota$ and $h_0$.
+The extension of this method to include the pulsar term is a key challenge for a future work.
+
+\appendix
+
+\section{Kalman recursion equations} \label{sec:kalman}
+The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
+\begin{equation}
+	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
+\end{equation}
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance, and the subscript $k$ labels the time-step.
+The Kalman filter takes the underlying states to evolve according to
+\begin{equation}
+	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
+\end{equation}
+where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix.
+$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$.
+\newline
+
+The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage.
+The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
+Specifically, the predict step proceeds as,
+\begin{equation}
+	\hat{\boldsymbol{X}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k
+\end{equation}
+\begin{equation}
+	\hat{\boldsymbol{P}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k 
+\end{equation}
+Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
+\begin{equation}
+	\boldsymbol{\epsilon}_{k} = \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1}
+\end{equation} 
+\begin{equation}
+	\boldsymbol{S}_k = \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k
+\end{equation}
+\begin{equation}
+	\boldsymbol{K}_k = \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \label{eq:kalman gain}
+\end{equation}
+\begin{equation}
+	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
+\end{equation}
+\begin{equation}
+		\hat{\boldsymbol{P}}_{k|k} = \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1}
+\end{equation}
+Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
+$\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$.
+For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
+\newline
+
+We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
+\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
+\begin{equation}
+	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
+\end{equation}
+Similarly,  
+\begin{equation}
+\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
+\end{equation}
+The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
+\begin{equation}
+	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
+\end{equation}
+where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix
+\begin{equation}
+\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
+\end{equation}
+and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
+\begin{equation}
+ C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
+\end{equation}
+The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
+\begin{equation}
+\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)
+\end{equation}
+\newline
+
+The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
+\begin{equation}
+	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \label{eq:gardenier}
+\end{equation} 
+From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
+\begin{equation}
+\boldsymbol{F}_k = e^{\boldsymbol{A}\left( t_{k+1} - t_k \right)} 
+\end{equation}
+\begin{equation}
+	\boldsymbol{G}_k = \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' 
+\end{equation}
+\begin{equation}
+	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t')
+	\end{equation}
+For our system, the state, control and process noise covariance matrices then take the form:
+\begin{equation}
+	\boldsymbol{F}_k = 	\text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right)
+\end{equation}
+\begin{equation}
+\boldsymbol{G}_k	= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right)
+\end{equation}
+with
+\begin{align}
+	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
+	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
+\end{align}
+and
+\begin{equation}
+	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) 
+\end{equation}
+for 
+\begin{equation}
+	Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right)
+\end{equation}
+where $\Delta t$ is the constant time sampling interval $=t_{k+1} -t_{k}$. \newline
+
+The remaining component matrices of the Kalman filter that we have not yet specified are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
+These are defined straightforwardly from Equation \eqref{eq:measurement}.
+Specifically,
+$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
+The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
+
+\section{Fiducial pulsar values}\label{appendix_fiducial}
+\textcolor{red}{TK: Do we really need this?
+The argument is for reproducibility, but we have a nice github repo that holds all the data we actually used, jupyter notebooks that show how the calculations are done.
+Perhaps it is sufficient to point the reader to the repo...
+}
+\section{Identifiability}\label{appendix_identifiability}
+\textcolor{red}{TK TBD. An example calculation for a simple system using the methods of e.g. } \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
+
+\section{Wasserstein distance}\label{sec:wasserstein}
+The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
+It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$.
+For this reason it is sometimes known as the "Earth mover's distance".
+The $p$-th order Wasserstein distance between two distributions is
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \lambda (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
+\end{eqnarray}
+where $p \in [1,\infty)$, $\lambda(x,y)$ is the transport plan, and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
+$\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
+
+The cost function can be freely chosen for the nature of the problem, but is commonly taken to be the absolute value function, $c(x,y) = |x-y|$, and we adopt the same cost function in this work.
+Given this choice of $c(x,y)$, for an empirical distribution $\hat{\mu}_n$ composed of $n$ i.i.d samples drawn from the underlying probability distribution $\mu$, the solution to the discrete form of Equation \eqref{eq:wasserstein} can be computed generally by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
+However, for $\mu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
+\begin{eqnarray}
+	\boldsymbol{E} [W_1(\hat{\mu}_n,\mu)] \asymp n^{-\frac{1}{d}} \ .
+\end{eqnarray}
+In order to circumvent this issue, in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors, i.e.
+setting $d=1$.
+In this case Equation \eqref{eq:wasserstein} has a convenient, easily computed, closed form:
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left(\int_0^1  F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) dz \right)^{1/p} \, \label{eq:wasserstein}
+\end{eqnarray}
+where $F_{\mu}(z)$ is the cumulative distribution function of $\mu$. \newline
+
+The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages.
+Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another.
+It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g.
+radians then we can interpret this as the cost to transform between the two distributions is 1 radian.
+Further interpretability is provided by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
+\begin{eqnarray}
+	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
+\end{eqnarray}
+where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
+That is, if $W_1 =1$ then the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian.
+Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g.
+Gaussianity.
+Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces.
+Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality.
+It also respects the geometry of the underlying metric space over which the distributions are defined.
+\newline
+
+For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
+The results are presented in Figure \ref{fig:pairwise_wasserstein}.
+The figure shows a lower triangular heat-map where each point denotes the $W_1$ value between a pair of posteriors.
+Small values of $W_1$ are magenta moving to red for large values of $W_1$.
+Note that the heat-map colour scaling is not the same for all subplots of the figure.
+It is evident that $W_1$ is generally small across all pairs of posteriors.
+We define $W_{1, \rm median}$ as the median value of $W_1$ across all $10^3 \choose 2$ pairs of noise realisations.
+The values of $W_{1, \rm median}$ for each parameter are quoted in Table \ref{tab:Wasserstein}.
+As a percentage of the width of the prior space (e.g.
+for $\alpha$ the width of the prior is $2 \pi$), the median $W_1$ for each parameter is $\lesssim 3 \%$.
+Generally speaking, the method converges to consistent one-dimensional posteriors for each parameter across different noise realisations.
+\newline
+
+There are a small fraction of the total number of noise realisations which have a large Wasserstein distance from all the other posteriors of all the other noise realisations.
+These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
+These correspond exclusively to posteriors which have converged poorly and not found the likelihood global maximum.
+For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the figure, corresponds to a posterior for $\psi$ which has converged sub-optimally, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
+The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ rad.
+We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
+This is due to multimodal likelihood surface for $\psi-\alpha$, shown in Figure \ref{fig:likelihood_surface_alpha_psi}.
+We can see that the likelihood surface has a global optima at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local optima that corresponds to the location of the example poorly converged posterior discussed previously, with $\psi = 1.01$ rad.
+This problem of poor global convergence can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
+
+%%%%%%%%%%%%%%%
+
 \begin{figure*}
@@
 \end{figure}
-The results in Section \ref{sec:parameter_estim} are for a single realisation of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
-It is important to confirm that the analysis scheme returns accurate answers for arbitrary noise realisations and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous by accident.
-It is also important to quantify, albeit approximately, the natural random dispersion in the one-dimensional posterior medians from one noise realization to the next, as the dispersion is a practical measure of the accuracy of the parameter estimation scheme, when it is applied to real astronomical data, where the true parameter values and specific noise realisation are unknown.
-\newline
-
-To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
-For each realisation we independently estimate the static parameters, $\boldsymbol{\theta}$.
-In Figure \ref{fig:corner_plot_2} we plot the estimates of $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary realisations.
-The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, which stems from one realisation.
-We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
-As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm gw}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
-\newline
-
-Figure \ref{fig:corner_plot_2} confirms two main points (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim 10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
-
- Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval.
-As in the single noise case, there are tentative signs of bias for some of the parameters, with the one-dimensional posteriors not symmetric about the injected value.
-For example, the maximum a posteriori probability estimates of $\iota$ across the noise realisations all appear left-skewed, consistently underestimating the injected value.
-Given the uncertainty in the posteriors it is difficult to draw strong conclusions on a potential bias and this question will be discussed in detail in Section \ref{sec:bias} \newline
-
-We now consider the complete set of $10^3$ noise realisations, rather than the subset of nine realisations in the preceding discussion.
-For each realisation of the noise, for each parameter, we infer a marginalised one-dimensional posterior.
-Our goal is to compare these posteriors between different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.
-To this end we use the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions.
-The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}.
-A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}.
-The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
-Throughout this work we use the first moment of the Wasserstein distance, $W_1(\mu,\nu)$, which is defined between one-dimensional probability distributions $\mu$ and $\nu$ as
-\begin{eqnarray}
-	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y) \ ,
-\end{eqnarray}
-where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function.
-The first moment of the WD has a convenient physical interpretation given by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
-\begin{eqnarray}
-	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
-\end{eqnarray}
-where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
-We can see that $W_1$ bounds the difference in the expected values.
-Taking a concrete example, suppose that we infer two one-dimensional posterior distributions for $\iota$, for two different realisations of the noise, and calculate $W_1 =0.5$ rad between them.
-This means that the difference in the expected value of two random variables drawn from the two posteriors is no greater than $0.5$ rad.
-\newline
-
-For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
-The results are presented in Figure \ref{fig:pairwise_wasserstein}.
-The figure shows a lower triangular heat-map where each point denotes the $W_1$ value between a pair of posteriors.
-Small values of $W_1$ are magenta moving to red for large values of $W_1$.
-Note that the heat-map colour scaling is not the same for all subplots of the figure.
-It is evident that $W_1$ is generally small across all pairs of posteriors.
-We define $W_{1, \rm median}$ as the median value of $W_1$ across all $10^3 \choose 2$ pairs of noise realisations.
-The values of $W_{1, \rm median}$ for each parameter are quoted in Table \ref{tab:Wasserstein}.
-As a percentage of the width of the prior space (e.g.
-for $\alpha$ the width of the prior is $2 \pi$), the median $W_1$ for each parameter is $\lesssim 3 \%$.
-Generally speaking, the method converges to consistent one-dimensional posteriors for each parameter across different noise realisations.
-\newline
-
-There are a small fraction of the total number of noise realisations which have a large Wasserstein distance from all the other posteriors of all the other noise realisations.
-These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
-These correspond exclusively to posteriors which have converged poorly and not found the likelihood global maximum.
-For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the figure, corresponds to a posterior for $\psi$ which has converged sub-optimally, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
-The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ rad.
-We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
-This is due to multimodal likelihood surface for $\psi-\alpha$, shown in Figure \ref{fig:likelihood_surface_alpha_psi}.
-We can see that the likelihood surface has a global optima at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local optima that corresponds to the location of the example poorly converged posterior discussed previously, with $\psi = 1.01$ rad.
-This problem of poor global convergence can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
-
-\subsection{Detection} \label{sec:detection}
-We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-$\mathcal{M}_1$ is the Earth-terms only model, i.e.
-the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
-The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
-The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
-We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
-We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
-For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
-This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
-$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
-\newline
-
-There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
-The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
-The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
-For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
-That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
-For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
-For the results presented in this subsection we set $n_{\rm live} =2000$.
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
-	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
-The horizontal grey dashed line labels the detection threshold of $\beta = 10$.
-The minimal detectable strain, below which $\beta < 10$, is $\sim 3.9 \times 10^{-15}$.
-Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$.}
-	\label{fig:bayes}
-\end{figure}
-
-\section{Exploring a broader parameter space} \label{sec:parameter_space}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/pp_plot_new}
-	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}.
-Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$.
-The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
-The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals.
-Well estimated posteriors should fall along the diagonal $y=x$.
-$\Omega$ is generally well-estimated but many other parameters show evidence of being over-constrained due to a modelling bias.
-}
-	\label{fig:parameter_space}
-\end{figure}
-We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
-In this section we test the method in different regions of the parameter space, varying the source parameters through astrophysically relevant ranges.
-\newline
-
-We consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
-We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space.
-For each simulated injection we can then attempt to recover the posterior distributions for each parameter.
-To summarise the results across the parameter space we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
-A PP plot describes the fraction of the total number of injected parameters which are included within a given credible interval of the estimated posterior with respect to the credible interval itself.
-In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage credible interval.
-The results are shown in Figure \ref{fig:parameter_space}.
-The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections.
-We can see that only $\Omega$ falls within the $3\sigma$ shaded region.
-The other parameters deviate from the $y=x$ diagonal.
-The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$.
-The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and there are more injections contained within the low value credible intervals.
-This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally confident, but are systematically biased away from the true injection values.
-This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval.
-An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.
-
-\section{Identifiability and bias} \label{sec:bias_and_identifiability}
-There are two underlying issues in the preceding tests that we now explore in more detail.
-The first relates to parameter identifiability, i.e.
-can we uniquely identify the parameters of the model given the data and the model structure?
-This question is explored in Section \ref{sec:identif}.
-The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
-This is the focus of Section \ref{sec:bias}.
-In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f.
-Figure \ref{fig:bayes}) for our representative example system (Table \ref{tab:parameters_and_priors}).
-\footnote{\tiny \textcolor{red}{TK: This discussion on bias and identifiability is perhaps less relevant than it was when our representative example had a 'high' strain ($10^{-12}$).
-However it IS still a feature/bug of our approach and I think it is worth discussing in the main body rather than relegating to an appendix.
-Opinions welcome...} \normalsize}
-
-\subsection{Identifiability}\label{sec:identif}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
-	\caption{Posterior probability distributions for the representative system described in Table \ref{tab:parameters_and_priors} over nine noise realisations for the GW source parameters $\iota$ and $h_0$ in the high SNR regime ($h_0 = 10^{-12}$).
-Unlike the other parameters of  $\boldsymbol{\theta}_{\rm gw}$, the one-dimensional posteriors for $\iota$ and $h_0$ do not strongly overlap.
-A correlation between  $\iota$ and $h_0$ is evident.}
-	\label{fig:just_iota_h}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_surface}
-	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation.
-The red point labels the location of the likelihood maxima.
-The likelihood surface has been normalised with respect to the absolute value of the maxima.
-A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
-	\label{fig:likelihood_surface}
-\end{figure}
-In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations. 
-The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but just for $\iota$ and $h_0$ rather than all members of $\boldsymbol{\theta}_{\rm gw}$.
-In the case of high-SNR we might expect \textit{a priori} that our estimates would be increasingly accurate over the low-SNR case, and that the method would converge robustly to similar posteriors irrespective of the noise realisation (i.e.
-$W_1 \to 0$).
-This is generally true for the majority of the component parameters of $\boldsymbol{\theta}_{\rm gw}$; the estimated posteriors across noise realisations are largely indistinguishable.
-However for $\iota$ and $h_0$ this is not the case.
-Instead the inferred one-dimensional posteriors do not broadly overlap and are not consistent across different noise realisations.
-Specifically across $10^3$ noise realisations the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
-For $\iota$, $W_{1,\rm median}$ is greater than the value inferred in the low SNR case ($=0.14$, Table \ref{tab:Wasserstein}), whilst for $h_0$ $W_{1,\rm median}$ is also greater as a proportion of the scale set by the strain amplitude.
-This is in contrast to $W_{1,\rm median}$ inferred for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ in the high-SNR regime where as SNR increases, $W_1$ decreases; e.g.
-for $\Phi_0$, $W_{1, \rm median} = 4.5 \times 10^{-3}$ rad.
-\newline
-
-This issue of inferring very different posteriors depending on the particular realisation of the noise is due to these parameters being only weakly-identifiable.
-Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
-Sources of non-identifiability are generally categorised as either ``structural", arising from the structure of the model, or ``practical", arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
-Regarding $\iota$ and $h_0$, we are concerned with structural identifiability.
-Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx} and \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
-For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
-This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
-It can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
-That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a small, non-zero gradient close to the maximum likelihood solution.
-This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
-Whilst a single, unique maxima exists (labelled by the red point in the figure), along the ridge in the parameter space the likelihood values are all very similar.
-Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
-If during the inference we fix one of the parameters at its true value (i.e.
-set a delta function prior), taking a slice through the likelihood surface it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably, i.e.
-$W_1 \to 0$.
-In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" \footnote{\tiny \textcolor{red}{TK: "washes out" may be too colloquial / unspecific, but I think it is intuitive...}\normalsize}the likelihood ridge, leading to a more concave likelihood surface.
-This enables consistent posteriors to be estimated for $\iota$ and $h$.
-In the case where we fix e.g $h_0$ at its true value to infer indistinguishable posteriors for $\iota$, all of the posteriors for $\iota$ are all strongly biased away from the injected value; the maximum a posteriori probability estimates underestimate the injected value by $\sim 0.34$ rad.
-This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be now be discussed further in Section \ref{sec:bias}.
-\subsection{Bias by dropping pulsar terms}\label{sec:bias}
-
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
-	\caption{As Figure \ref{fig:corner_plot_1} but for the high-SNR system with $h_0 = 10^{-12}$.
-For the majority of parameters the injection value does not fall within the 90\% credible interval and is not visible over these scales.
-The estimated posteriors are biased away from the injected value due to the dropping of the pulsar terms.
-Note that the scale of the $x$-axes are generally highly constrained compared to Figure \ref{fig:corner_plot_1}.}
-	\label{fig:bias_for_large_h}
-\end{figure*}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
-	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and the Earth terms model for Earth terms only synthetic data (green line), with respect to the inclination $\iota$, over the prior space.
-All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
-The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model; the green/orange dashed lines are overlaid on this scale.
-For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$).
-For the pulsar-terms model the location of the maxima and injected value do coincide.
-10 noise realisations are plotted for each situation but the solutions are completely overlaid on this scale.}
-	\label{fig:likelihood_surface_iota}
-\end{figure}
-Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g.
-the median of the one-dimensional marginalised posterior) is biased away from the true injection value.
-In Figure \ref{fig:bias_for_large_h} we show the resulting posteriors for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but now for the high-SNR system with $h_0 = 10^{-12}$.
-We can see that for all parameters, with the exception of $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it is not visible on the scales over which we have plotted the posteriors.
-This effect is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
-Similar results are obtained for other noise realisations.
-Whilst $h_0$ does not exhibit any noticeable bias for this particular noise realisation, in general across multiple noise realisations it does exhibit a correlated bias with $\iota$.
-In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
-\newline
-
-In Figure \ref{fig:likelihood_surface_iota} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, as we vary $\iota$ across the prior space, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
-We do this for 10 separate realisations of the system noise.
-We consider three separate situations: \textit{(i)} The solid blue line is the likelihood curve with just the Earth-terms included (i.e.
-using Equation \ref{eq:measuremen_earth} in the Kalman filter), \textit{(ii)} the solid orange line is the likelihood curve with the additional pulsar terms included (i.e.
-using Equation \ref{eq:measurement}), \textit{(iii)} the solid green line is the solution where we drop the pulsar terms from both the synthetic data and the measurement equation.
-We have not used nested sampling for situations \textit{(ii)} or \textit{(iii)} in this work.
-However, we are able to run the Kalman filter either using the pulsar terms in the model or else dropping them from the synthetic data consider how the resulting likelihood varies for different parameter values.
-The dashed lines show the location of the likelihood maxima for each of the likelihood solutions; finding the location of the maxima across all static parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
-\newline
-
-The key observation from Figure \ref{fig:likelihood_surface_iota} is that for the Earth-terms only solution the location of the likelihood optima and the location of the injected value are not the same.
-Consequently Bayesian likelihood inference techniques settle in optima that do not necessarily align with the true injection value.
-This is the origin of the bias in the one-dimensional marginalised posteriors that we observe.
-For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ rad, whereas the injected value was $\iota = 1.0$ rad.
-This optima is comparable to the median of the posterior inferred for $\iota$ in Figure \ref{fig:bias_for_large_h}.
-\newline
-
-In contrast to the Earth-terms solution, for the pulsar-terms solution the optima of the likelihood curve now aligns with the true injected value.
-The corollary is also true; if we drop the pulsar terms from the synthetic data and use the Earth terms equation for inference then the optima is also correctly located.
-This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.
-Similar results are observed for the other parameters which exhibit a bias in the nested sampling inference.
-\newline
-
-Why does dropping the pulsar terms bias these parameters in particular, whilst leaving others unaffected?
-For instance, why is the bias induced in $\iota$ so strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
-When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is effective modulated by the interference of two equal-amplitude, phase shifted waves (c.f.
-Equation \ref{eq:measurement}).
-Explicitly, we have the ``Earth-terms wave"	 with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
-The net interference wave that modulates $f_{\rm p}$ is
-\begin{eqnarray}
-	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ .
-\end{eqnarray}
-The additional modulations from $y_{\rm pulsar}$ influence the amplitude of the resulting interference wave, but do not affect its frequency.
-When we model this data using the Kalman filter without the pulsar terms, the additional amplitude modulations from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
-As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude of the single wave.
-Consequently the parameters which are most affected are those which determine the amplitude of the GW.
-Terms which influence the amplitude strongly, such as $h_0$ and $\iota$, are strongly affected and exhibit a strong bias.
-Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
-The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
-Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
-This hierarchy of biases for each parameter is also what we observe in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
-\newline
-
-In order to ascertain how important the bias is for PTA continuous wave searches requires a thorough exploration of the astrophysical parameter space which we do not carry out in this work.
-The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
-If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate and the bias may not be important for searches.
-This is what we observe for our synthetic data (e.g.
-Figure \ref{fig:corner_plot_2}); the decreased SNR manifests as a broader marginalised posterior which generally overwhelms the shift due to the bias.
-Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may come to systematically affect the inferred parameters.
-\newline
-
-We have shown in this section how the dropping of the pulsar terms results in a bias in the inferred parameters, not just in the sky localisation parameters  \citep[e.g.][]{Zhupulsarterms,Chen2022} but additionally all parameters which influence the wave amplitude.
-Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for our example system, it is present and may need to be accounted for for analysis in other parts of the parameter space.
-This bias is not particular to our method but a shared feature across likelihood based methods that do not include the pulsar terms in the inference model.
-The inclusion of the pulsar terms in the methods presented in this work in order to alleviate the bias will be a key goal of a future work.
-
-\section{Discussion}\label{sec:discussion}
-We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
-There are a number of potential extensions of this work which we now discuss.
-\newline
-
-The natural first extension is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
-Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred one-dimensional marginalized posteriors for the model parameters.
-Biases in sky localisation for synthetic datasets as a result of dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.
-Alternatively, if the method continues to be used with solely the Earth terms then it would be highly desirable to make systematic, quantitative estimates of the incurred biases in the model parameters across an astrophysically representative parameter space.
-\newline
-
-We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
-Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA.
-Indeed, since adding pulsars to PTAs increases the computational demands of the data analysis it may be advantageous to optimally select a subset of the total available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.
-\newline
-
-Throughout this work we have assumed that all pulsars are observed at the same time, with a constant time sampling.
-In practice this assumption does not hold and different pulsars will be observed with different cadences at different times.
-Extending this method to non-constant time sampling is straightforward and the performance of the method should be evaluated in this regime.
-\newline
-
-We have taken our GW source to be monochromatic, i.e.
-$\Omega$ is a fixed parameter with no time dependence.
-Whilst this is an astronomically well-justified assumption (see discussion in Section \ref{sec:plane_gw}), and appropriate for this initial methods work, it would also be of interest to extend the state-space construction to enable the GW frequency to evolve in time.
-The GW source is monochromatic over the decadal PTA observation period, but the timescale set by the Earth-pulsar distance is comparable to the SMBHB evolution timescale \citep{Sesana2010}.
-For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) the GW source frequency evolution manifests as an incoherent source of noise in the pulsar terms, whilst for  $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms can affect the phase coherency of the Earth terms \citep{Perrodin2018}.
-Careful consideration of the evolution of the GW frequency will be important as we look to extend our method to include the pulsar terms in the inference model.
-\newline
-
-We have also assumed that there is only a single GW source that influences the received pulsar frequencies.
-However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
-Our method naturally extends to detecting multiple GW sources simultaneously; Equation \eqref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
-The stochastic background itself is the incoherent summation of many individual GW sources.
-It therefore stands to reason that it may be possible to extend this method to also search for a stochastic GW background, providing a complementary approach to detect the stochastic background without requiring a cross-correlation of the pulsar residuals to uncover the Hellings-Downs curve.
-\newline
-
-\textcolor{red}{TK: Lets also pre-empt reviewer and make it explicit why we are using frequencies rather than TOAs. Need to chat with group to get this clear in my head.}
-
-%https://arxiv.org/pdf/2107.03047.pdf
-\section{Conclusion}\label{sec:conclusion}
-In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
-We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
-That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW induced modulations.
-This enables the inference of GW parameters conditioned on the specific observed realisation of the noisy data.
-We introduce the Kalman filter in order to track the intrinsic state evolution of the pulsar and combine this with a Bayesian sampling framework to estimate the posterior distributions of each parameter of the system, as well as the associated evidence (marginalised likelihood) of the model.
-We test our method on synthetic data.
-We initially focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
-The parameters of the model are accurately recovered and the minimal detectable strain is estimated to be $h_0 \sim 4 \times 10^{-15}$ for $\iota=1.0$ rad.
-The method is then further tested over 1000 noise realisations.
-Consistent posterior estimates of the parameters are obtained for the majority of noise realisations, with the median Wasserstein distance across all noise realisations $W_{1, \rm median}$ generally small compared to the prior space ($\lesssim 3 \%$).
-For a minority of noise realisations the nested sampler converges poorly and settles into a local likelihood optima for the posteriors estimates of $\psi$ and $\alpha$.
-This problem can be solved in general by increasing the number of live points $n_{\rm live}$ used by the sampler.
-Exploration of a broader parameter space via 200 randomly sampled parameter vectors highlights biases in some parameters as a result of dropping the pulsar terms from the inference model.
-The cause of these biases is examined for the case of high SNR, along with the structural weak-identifiability between $\iota$ and $h_0$.
-The extension of this method to include the pulsar term is a key challenge for a future work.
-
-\appendix
-
-\section{Kalman recursion equations} \label{sec:kalman}
-The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
-\begin{equation}
-	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
-\end{equation}
-where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance, and the subscript $k$ labels the time-step.
-The Kalman filter takes the underlying states to evolve according to
-\begin{equation}
-	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
-\end{equation}
-where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix.
-$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$.
-\newline
-
-The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage.
-The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
-Specifically, the predict step proceeds as,
-\begin{equation}
-	\hat{\boldsymbol{X}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k
-\end{equation}
-\begin{equation}
-	\hat{\boldsymbol{P}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k 
-\end{equation}
-Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
-\begin{equation}
-	\boldsymbol{\epsilon}_{k} = \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1}
-\end{equation} 
-\begin{equation}
-	\boldsymbol{S}_k = \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k
-\end{equation}
-\begin{equation}
-	\boldsymbol{K}_k = \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \label{eq:kalman gain}
-\end{equation}
-\begin{equation}
-	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
-\end{equation}
-\begin{equation}
-		\hat{\boldsymbol{P}}_{k|k} = \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1}
-\end{equation}
-Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
-$\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$.
-For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
-\newline
-
-We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
-\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
-\begin{equation}
-	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
-\end{equation}
-Similarly,  
-\begin{equation}
-\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
-\end{equation}
-The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
-\begin{equation}
-	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
-\end{equation}
-where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix
-\begin{equation}
-\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
-\end{equation}
-and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
-\begin{equation}
- C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
-\end{equation}
-The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
-\begin{equation}
-\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)
-\end{equation}
-\newline
-
-The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
-\begin{equation}
-	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \label{eq:gardenier}
-\end{equation} 
-From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
-\begin{equation}
-\boldsymbol{F}_k = e^{\boldsymbol{A}\left( t_{k+1} - t_k \right)} 
-\end{equation}
-\begin{equation}
-	\boldsymbol{G}_k = \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' 
-\end{equation}
-\begin{equation}
-	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t')
-	\end{equation}
-For our system, the state, control and process noise covariance matrices then take the form:
-\begin{equation}
-	\boldsymbol{F}_k = 	\text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right)
-\end{equation}
-\begin{equation}
-\boldsymbol{G}_k	= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right)
-\end{equation}
-with
-\begin{align}
-	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
-	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
-\end{align}
-and
-\begin{equation}
-	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) 
-\end{equation}
-for 
-\begin{equation}
-	Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right)
-\end{equation}
-where $\Delta t$ is the constant time sampling interval $=t_{k+1} -t_{k}$. \newline
-
-The remaining component matrices of the Kalman filter that we have not yet specified are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
-These are defined straightforwardly from Equation \eqref{eq:measurement}.
-Specifically,
-$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
-The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
-
-\section{Fiducial pulsar values}\label{appendix_fiducial}
-\textcolor{red}{TK: Do we really need this?
-The argument is for reproducibility, but we have a nice github repo that holds all the data we actually used, jupyter notebooks that show how the calculations are done.
-Perhaps it is sufficient to point the reader to the repo...
-}
-\section{Identifiability}\label{appendix_identifiability}
-\textcolor{red}{TK TBD. An example calculation for a simple system using the methods of e.g. } \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
-
-\section{Wasserstein distance}\label{sec:wasserstein}
-The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
-It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$.
-For this reason it is sometimes known as the "Earth mover's distance".
-The $p$-th order Wasserstein distance between two distributions is
-\begin{eqnarray}
-	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
-\end{eqnarray}
-where $p \in [1,\infty)$ and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
-$\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
-\newline
-
-The cost function can be freely chosen for the nature of the problem, but is commonly taken to be the absolute value function, $c(x,y) = |x-y|$, and we adopt the same cost function in this work.
-Given this choice of $c(x,y)$, for an empirical distribution $\hat{\mu}_n$ composed of $n$ i.i.d samples drawn from the underlying probability distribution $\mu$, the solution to the discrete form of Equation \eqref{eq:wasserstein} can be computed generally by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
-However, for $\mu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
-\begin{eqnarray}
-	\boldsymbol{E} [W_1(\hat{\mu}_n,\mu)] \asymp n^{-\frac{1}{d}} \ .
-\end{eqnarray}
-In order to circumvent this issue, in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors, i.e.
-setting $d=1$.
-In this case Equation \eqref{eq:wasserstein} has a convenient, easily computed, closed form:
-\begin{eqnarray}
-	W_p(\mu,\nu)= \left(\int_0^1  F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) dz \right)^{1/p} \, \label{eq:wasserstein}
-\end{eqnarray}
-where $F_{\mu}(z)$ is the cumulative distribution function of $\mu$. \newline
-
-The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages.
-Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another.
-It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g.
-radians then we can interpret this as the cost to transform between the two distributions is 1 radian.
-Further interpretability is provided by the Monge-Kantarovih duality, Equation \eqref{eq:WDdefn}; the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian.
-Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g.
-Gaussianity.
-Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces.
-Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality.
-It also respects the geometry of the underlying metric space over which the distributions are defined.
 

### 2023-10-23 py_src/plotting.py

@@ -104,10 +104,8 @@ def plot_all(t,states,measurements,measurements_clean,predictions_x,predictions_
 
 
 
-def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges,axes_scales,savefig,logscale=False,title=None):
+def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges,axes_scales,savefig,logscale=False,title=None,quantiles=[0.16, 0.84],smooth=True,smooth1d=True):
     plt.style.use('science')
-
-
 
 
     print(path.split('.')[-1])
@@ -175,7 +173,7 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
                         show_titles=True,
                         smooth=True,smooth1d=True,
                         truth_color='C2',
-                        quantiles=[0.16, 0.84], #[0.16, 0.84]
+                        quantiles=quantiles, #[0.16, 0.84] #
                         truths = injection_parameters,
                         range=ranges,
                         labels = labels,

### 2023-10-27 docs/manuscript/paper.tex
try some new PP runs

@@
 % Keep the title short and informative.
-	\title[Kalman PTA]{Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array}
+	\title[Kalman PTA]{Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array}
 
@@
 Full posterior distributions of model parameters are recovered and tested for accuracy.
-There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms'.
-Smaller biases $\lesssim 1 \%$ are also observed in some other static parameters.
+There is a bias of $\sim 0.3$ rad in the marginalised one-dimensional posterior for the orbital inclination $\iota$, introduced by dropping the so-called `pulsar terms'.
+Smaller biases $\lesssim 1 \%$ are also observed in other static parameters.
 \end{abstract}
@@
 
-The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
+The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
 Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
@@
 However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}.
-Inconclusive evidence at low significance was presented recently by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}.
+Inconclusive evidence at low significance was presented recently by the EPTA for an individual source at 4--5 nHz \citep{2023arXiv230616226A}.
 \newline
@@
 However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
-In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red).
+In modern Bayesian PTA searches, the power spectral density of the red intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the noise induced by a stochastic GW background (whose spectrum is also red).
 In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
@@
 
-This paper is organised as follows.
+The paper is organised as follows.
 In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW.
 In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence.
-In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method.
+In Section \ref{sec:testing} we describe how we create synthetic validation data with which to test the method.
 We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source.
-In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space.
+In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant domain of SMBHB source parameters.
 In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates.
-Discussion on  future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.
+Conclusions are drawn in Section \ref{sec:discussion}.
 The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}.
@@
 The stochastic torque is assumed to be a zero-mean, white noise process.
-Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
+Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
 \begin{equation}
@@
 \end{equation}
-In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
+In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
 \begin{align}
@@
 The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$.
-Throughout this paper we work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB).
-We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB.
-\newline
-
-In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic.
-Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) changes during the inspiral by \citep[e.g.][]{Zhu10}
-\begin{equation}
-	\Delta f_{\rm gw} = 3.94 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
+Throughout this paper we work with pulsar TOAs defined relative to the Solar System barycentre (SSB).
+We are free to choose coordinates such that $\Phi_0$ is the GW phase at $t=0$ at the SSB.
+\newline
+
+In this paper the source is approximated as monochromatic.
+In reality the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) increases during the inspiral by an amount \citep[e.g.][]{Sesana2010}
+\begin{equation}
+	\Delta f_{\rm gw} = 0.05 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^{8.5} M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{50 \mathrm{~nHz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
 	\label{eq:f_evolution}
 \end{equation}
+%
+%\begin{equation}
+%	\Delta f_{\rm gw} = 3.94 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
+%	\label{eq:f_evolution}
+%\end{equation}
 where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
 A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
-Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^9 M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 10^{-7} \mathrm{~Hz}]^{11/3} (T_{\rm obs}^2/10 \mathrm{yr}) \gtrsim 8 \mathrm{yr}$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$.
+Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \gtrsim 20$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$.
 The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
@@
 and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$.
-It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
+It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as
 \begin{align}
@@
 In the left column the red/magenta solutions overlap almost perfectly; in the right column, they do not.
-The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{(1)}(t) - \hat{f}_{\rm m}^{(1)}(t)$.
-The results are shown for a single pulsar.
-In the right column $\Omega$ is displaced from its true value by 20\%.}
+The bottom panels show the residual or innovation $\epsilon(t) =f_{\rm m}^{(1)}(t) - \hat{f}_{\rm m}^{(1)}(t)$.
+In the right column $\Omega$ is displaced from its true value by 20\%.
+Results are shown for a single pulsar.
+}
 	\label{fig:kalman_example}
@@
 It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel).
-
-%n Figure ; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
-The residuals correspond to random noise and are normally distributed.
-Conversely, if $\boldsymbol{\hat{\theta}}$ is not close to the true parameters then the filter is unable to recover the state evolution.
-This is demonstrated in Figure \ref{fig:25MB_bfs} where the Kalman filter was run with $\boldsymbol{\hat{\theta}}$ slightly perturbed away from the true values.
-In this case the filter cannot track the state variable accurately and the residuals are no longer Gaussian.
+In Section \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$.
 
@@
 For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}.
-For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}.
+For reviews of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}.
 Multiple nested sampling algorithms and computational libraries exist.
@@
 	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
-	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$. 
+	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$
 	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
@@
 \begin{enumerate}[leftmargin=2em]
-	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
+	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
 	\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
@@
 	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
-	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a) -- (vi)(b).
+	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a)--(vi)(b).
 	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
-	\item Repeat steps (vii) - (ix) until convergence criteria are satisfied.
+	\item Repeat steps (vii)--(ix) until convergence criteria are satisfied.
 \end{enumerate}
@@
 In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
-In Section \ref{sec:gendata} we demonstrate how to solve Equations \eqref{eq:spinevol} -- \eqref{eq:frequency_evolution}, and \eqref{eq:measurement} -- \eqref{eq:g_func} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.
+In Section \ref{sec:gendata} we demonstrate how to solve Equations \eqref{eq:spinevol}--\eqref{eq:frequency_evolution}, and \eqref{eq:measurement}--\eqref{eq:g_func} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) $.
 
@@
 	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
-The pulsar distances relative to the observer are also indicated, with a distance $\leq 2 \, {\rm kpc}$ for $38$ pulsars.
+The pulsar distances relative to the observer are also indicated, with distance $\leq 2 \, {\rm kpc}$ for $38$ pulsars.
 The grey dashed curve denotes the Galactic plane.}
@@
 We see that the synthetic and empirical residuals are qualitatively similar.
-For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ for PSR J1939$+$2134.
+For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is much less than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ for PSR J1939$+$2134.
 
@@
 The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
-The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$.
+The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$.
 The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}.
-We set the TOA uncertainty to be a constant, $0.1 \mu$s.
+We set the TOA uncertainty to be a constant, viz.
+$0.1 \mu$s.
 The blue and orange residuals are qualitatively similar by inspection.
@@
 Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s, Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 10^2$ Hz observed with weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s, Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
 The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz.
@@
 
-Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} \Delta B_t | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta B_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
+Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} dB(t) | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta B_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
 To fix the problem, we subtract the deterministic frequency evolution and track the new variable
@@
 \end{equation}
-where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice.
+where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic evolution based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice.
 For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately.
@@
 In particular $f_{\rm em}^{(n)}(t_1)$
-and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.
+and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, i.e.
+Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.
 
@@
 In Section \ref{sec:priors} we define the priors chosen on $\boldsymbol{\theta}$.
-
-In Section \ref{sec:earth_psr_terms} we do this...
-
-\textcolor{red}{TK to do}
-
+In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter.
 In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
@@
 The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
-The injected elements of $\boldsymbol{\theta}_{\rm gw}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
-The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$, described in Section \ref{sec:synt_pta}, are also recorded in Table \ref{tab:parameters_and_priors}.
+The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
 
@@
 For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
-The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires additional discussion.
+The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires additional discussion below.
 \newline
@@
 We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues \citep{Manchester2005}.
-For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} \%$ and $\pm 2.31 \times 10^{-3} \%$ respectively.
+For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} $ per cent and $\pm 2.31 \times 10^{-3}$ per cent respectively.
 In this paper we adopt uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database \citep{Manchester2005}.
-Adopting more relaxed priors tests the method stringently.
-The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are relaxed or narrow.
-\newline
-
-The pulsar distances $d^{(n)}$ are not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with typical uncertainties $\sim 10 \%$ \citep{Yao2017,Arzoumanian2018ApJS..235...37A}.
+Wider-than-necessary priors, such as those above, test the method more stringently than narrow priors.
+The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are wide or narrow.
+\newline
+
+The pulsar distances $d^{(n)}$ are less constrained than $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with typical uncertainties $\sim 10 \%$ \citep{Yao2017,Arzoumanian2018ApJS..235...37A}.
 In this paper, however, we omit the term involving $d^{(n)}$ from Equation \eqref{eq:g_func}, as justified in Section \ref{sec:parameter_estim}, so there is no need to set a prior on $d^{(n)}$.
-Furthermore we do not set a prior on $\gamma^{(n)}$, because we have $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and $\gamma^{(n)}$ is effectively ``unobservable" over a decade; that is,for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is approximately independent of $\gamma^{(n)}$, as long as $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
+Furthermore we do not set a prior on $\gamma^{(n)}$, because we have $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and $\gamma^{(n)}$ is effectively ``unobservable'' over a decade; that is, for $T_{\rm obs}=10$ years the solution of Equation \eqref{eq:spinevol} is approximately independent of $\gamma^{(n)}$, as long as $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
 It is therefore sufficient for validation purposes to carry $\gamma^{(n)}$ through the analysis at its injected value.
-\footnote{We check this empirically in two independent ways for an informal selection of test cases: (i) we set an uninformative prior $\pi\left [\gamma^{(n)} / 1 \, {\rm s^{-1}} \right] \sim$ LogUniform($10^{-15}, 10^{-10}$) , and (ii) we deliberately displace $\gamma^{(n)}$ in the inference analysis from its true, injected value in the synthetic data, e.g.
+\footnote{We check this empirically in two independent ways for an informal selection of test cases: (i) we set an uninformative prior $\pi\left [\gamma^{(n)} / 1 \, {\rm s^{-1}} \right] \sim$ LogUniform($10^{-15}, 10^{-10}$); and (ii) we deliberately displace $\gamma^{(n)}$ in the inference analysis from its true, injected value in the synthetic data, e.g.
 $\gamma^{(n)} = 10^{-14} \text{ s}^{-1}$ versus $10^{-13} \text{ s}^{-1}$, respectively.
-The results in (i) and (ii) are the same as those reported in Section \ref{sec:rep_example}.} \newline
+The results from (i) and (ii) are found to be the same as those reported in Section \ref{sec:rep_example}.} \newline
 
@@
 The sole exception is PSR J1939+2134 which has $\sigma^{(n)} \sim 10^{-16} \text{s}^{-3/2}$.
-For validation purposes, we artificially set $\sigma^{(n)} = 10^{-20} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within the prior.
+For validation purposes, we artificially set $\sigma^{(n)} = 10^{-20} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within the aforementioned log-uniform prior.
 This is done purely for testing convenience; it is straightforward to expand the prior when analysing real, astronomical data.
@@
 We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter set, c.f.
-Equation \ref{eq:psrparams}.
+Equation \eqref{eq:psrparams}.
 Explicitly, we write
@@
 		\toprule
-		&Parameter & Injected values & Units & Prior  \\
+		&Parameter & Injected value & Units & Prior  \\
 		\hline
@@
 		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
-		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
-		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
+		\vspace{1mm} &  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
+		\vspace{1mm} & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
 		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
@@
 	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the prior used for Bayesian inference on each parameter (rightmost column).
-The top and bottom halves of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ respectively.The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``sc" indicates that the injected value has been calculated using Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} in \protect \cite{Shannon2010}.
-The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
+The top and bottom halves of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ respectively.The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
+The subscript ``SC'' indicates that the injected value has been calculated using Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} in \protect \cite{Shannon2010}.
+The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
 No priors are set for $d^{(n)}$ and $\gamma^{(n)}$ since those parameters do not enter the inference model,  as discussed in Section \ref{sec:priors}, c.f.
@@
 The argument of the cosine corresponds to the GW phase at each individual pulsar.
-The first and second terms are commonly referred to as the ``Earth term" and ``pulsar term" respectively.
+The first and second terms are commonly referred to as the ``Earth term'' and ``pulsar term'' respectively.
 Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases.
-They are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
-In this work we follow the standard approach and drop the pulsar terms from the inference model (but not the source model).
+They are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the inferred sky position \citep{Zhupulsarterms,Chen2022}.
+In this work we follow the standard approach and drop the pulsar terms from the inference calculation (but not the source model).
 Explicitly the measurement equation used in the Kalman filter reduces to
@@
 	\end{equation}
-	We defer the inclusion of the pulsar terms to a future paper.
-As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
+	We defer the inclusion of the pulsar terms in the inference calculation to a future paper.
+As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem because the measurement equation, Equation \eqref{eq:g_func_trig_earth}, is no longer a function of the pulsar distance.
 We stress that the pulsar terms are dropped only when doing Bayesian inference, i.e.
@@
 
-Figure \ref{fig:corner_plot_1} displays results for the seven parameters in  $\boldsymbol{\theta}_{\rm gw}$ in the form of a traditional corner plot.The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
+Figure \ref{fig:corner_plot_1} displays results for the seven parameters in  $\boldsymbol{\theta}_{\rm gw}$ in the form of a traditional corner plot.
+The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
 The dashed vertical blue lines mark the 0.16 and 0.84 quantiles; the solid orange line marks the known injected value.
@@
 There is scant evidence of railing against the prior bounds.
-Hence it is evident that for this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, that the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
+Hence it is evident for this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, that the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
 The injected values are contained within the 90\% credible interval for all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ .
@@
 
-The posterior is approximately symmetric about the known injected value for some parameters such as $\Omega$, where the posterior median and the injected value are coincide approximately.
+The posterior is approximately symmetric about the known injected value for some parameters such as $\Omega$, where the posterior median and the injected value coincide approximately.
 For other parameters (e.g.
@@
 For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
-a bias or a random outcome specific to this particular noise realisation.
-We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} using more noise realizations and show that indeed there is a bias from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
+a bias, or a random outcome specific to this particular noise realisation.
+We explore this further in Sections \ref{sec:multiple_noise} and \ref{sec:bias} using more noise realizations and show that indeed there is a systematic bias from dropping the pulsar terms, similar to that reported by \cite{Zhupulsarterms}.
 \newline
@@
 	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
-	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes.
-The subtitles of the marginalized 1D posteriors refer to values for the median posterior across the noise realisations.
-The injection value is generally contained within the 90 \% credible interval for the vast of majority posteriors.
-The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters.
+	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely.
+The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation.
+The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations.
+There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation $\sim10 \%$ across the 63 combinations.
 A slight skew-bias is apparent in some of the parameters, e.g.
 $\iota$.
-Correlations can be seen between $\Omega - \Phi_0$, $\psi-\alpha$ and $\iota - h_0$.}
+Weak correlations can be seen between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.}
 	\label{fig:corner_plot_2}
@@
 		\toprule
-		Parameter & Injected Values & Units & $W_{1, \rm median}$  \\
+		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
 		\hline
@@
 	\end{tabular}
-	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
-$W_{1, \rm median}$  sets the bound on the difference in expectation value between the one dimensional posteriors, and is generally small with respect to the width of the prior space.}
+	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
+$W_{1, \rm median}$ is generally smaller than the prior domain.}
 	\label{tab:Wasserstein}
 \end{table}
+
+The results in Section \ref{sec:parameter_estim} are obtained for a single realisation of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
+It is important to confirm that the analysis scheme returns accurate answers for arbitrary noise realisations and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous by accident.
+It is also important to quantify, albeit approximately, the natural random dispersion in the one-dimensional posterior medians from one noise realization to the next, as the dispersion is a practical measure of the accuracy of the parameter estimation scheme, when it is applied to real astronomical data, where the true parameter values and specific noise realisation are unknown.
+\newline
+
+To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
+For each realisation we independently estimate the static parameters, $\boldsymbol{\theta}$.
+In Figure \ref{fig:corner_plot_2} we plot the estimates of $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary realisations.
+The seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$ are recovered unambiguously.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, which stems from one realisation.
+We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
+As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm psr}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
+\newline
+
+Figure \ref{fig:corner_plot_2} confirms two main points: (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
+Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 (i.e.
+$92 \%$) out of the 63 possible combinations of the seven parameters and nine realizations.
+Figure \ref{fig:corner_plot_2}, like Figure \ref{fig:corner_plot_1}, displays tentative signs of bias, where the one-dimensional posteriors are not symmetric about the injected value.
+For example, the maximum a posteriori probability estimates of $\iota$ appear left-skewed, consistently underestimating the injected value in all nine realisations.
+A similar but messier trend is seen in $\delta$.
+However, the width of the posteriors is comparable to the putative bias, so it is difficult to draw strong conclusions.
+Biases are discussed in detail in Section \ref{sec:bias_and_identifiability}.
+There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.
+\newline
+
+We now consider the complete set of $10^3$ noise realisations, going beyond the subset of nine realisations in the preceding discussion.
+We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$?
+There is no unique way to answer this question.
+In this paper we appeal to the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory, which defines an intuitive notion of similarity between probability distributions.
+The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}; a short review is presented in Appendix \ref{sec:wasserstein}.
+The WD measures the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
+In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}.
+Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$ with respect to the PDFs $\mu$ and $\nu$.
+Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad.
+Then we can infer $| \langle \iota \rangle_\mu - \langle \iota \rangle_\nu | \leq 0.5 \, {\rm rad}$.
+\newline
+
+Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$.
+As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
+These values are an approximate measure of the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
+Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$.
+These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for a wide variety of noise realizations.
+Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
+
+\subsection{Detectability versus $h_0$} \label{sec:detection}
+We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}.
+In Equation \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-terms-only model with a GW present, i.e.
+the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
+To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
+\footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.} \newline
+
+We see in Figure \ref{fig:bayes} an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$.
+The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 4 \times 10^{-15}$.
+Of course, the minimum detectable strain is particular to the system in Table \ref{tab:parameters_and_priors}.
+It is influenced in general by $T_{\rm obs}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$, as discussed in Section \ref{sec:parameter_space}.
+Adjusting $\sigma_{\rm m}$ we find that an approximate quadratic relationship $\beta \propto h_0^2 / \sigma_{\rm m}^2$ also exists, where $h_0 / \sigma_{\rm m}$ is the effective signal-to-noise ratio.
+\newline
+
+The odds ratio $\beta$ drops off for $h_0 \lesssim 4 \times10^{-15}$.
+This happens because the two competing models becoming increasingly indistinguishable once the measurement noise dominates the GW signal.
+Moreover, the points in Figure \ref{fig:bayes} become sparser for $h_0 \lesssim 4 \times 10^{-15}$.
+This happens because we obtain $\beta <0$ for the missing, intermediate points.
+This is a noise artefact of the nested sampler.
+When the sampler converges sub-optimally, the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails, and $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+There is nothing special about the particular missing points; if one recreates the $\beta(h_0)$ curve by rerunning the nested sampler with another random seed, a different set of points are missing.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$, because the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(n_{\rm live}^{-1/2})$ (see Section \ref{sec:nested_sampling}).
+In this section we set $n_{\rm live} =2000$.
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
+	\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strains, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
+The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
+The minimum detectable strain, for $\beta < 10$, equals $4 \times 10^{-15}$.
+Missing points for $\beta \lesssim 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
+	\label{fig:bayes}
+\end{figure}
+
+\section{SMBHB source parameters} \label{sec:parameter_space}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/pp_plot_new}
+	\caption{The fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
+PP plot).
+The injections are 200 simulated GW sources for a subset of five parameters in $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions in Table \ref{tab:parameters_and_priors}.
+Each coloured curve corresponds to a different parameter (see legend).
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
+The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals given our statistical sample.
+For parameters with well estimated posteriors, the PP plot curve should fall along the diagonal of unit slope.
+$\Omega$ is generally well-estimated (i.e.
+it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e.
+the curves lie above the unit diagonal for low credible intervals, below the unit diagonal for high credible intervals).
+This is due to a modelling bias.}
+	\label{fig:parameter_space}
+\end{figure}
+Section \ref{sec:rep_example} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}.
+In this section we test the method for various systems, varying the SMBHB source parameters through astrophysically relevant ranges.
+\newline
+
+We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
+For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
+To summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself.
+In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
+\newline
+
+ Figure \ref{fig:parameter_space} displays the results of the numerical experiment described in the previous paragraph.
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections.
+We see that only $\Omega$ falls wholly within the $3\sigma$ shaded region.
+The PP curves for the other parameters deviate from the diagonal of unit slope.
+The deviation is more pronounced for $\alpha$ and $\delta$ and less for $\psi$ and $\Phi_0$.
+The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within higher-value credible intervals than would be expected statistically, and there are more injections contained within lower-value credible intervals.
+
+ This stems from the bias noted in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally narrow, but are systematically biased away from the true injection values.
+
+ Consequently the
+
+ and the PP plot follows a characterstic S-shape, diagnostic of an under constrained posterior
+
+ Consequently at lower-value credible intervals
+
+ This translates in the PP plot to posteriors which are too narrow to contain the injected value within the credible interval. The origin of the bias is discussed in Section \ref{sec:bias}.
+
+ This stems from the bias noted in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally confident, but are systematically biased away from the true injection values.
+This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval.
+The origin of the bias is discussed in Section \ref{sec:bias}.
+
+\section{Systematic biases} \label{sec:bias_and_identifiability}
+ Two systematic aspects of the tests in Sections \ref{sec:rep_example} and \ref{sec:parameter_space} merit further study.
+The first relates to parameter identifiability, i.e.
+can we uniquely identify the parameters of the model given the data and the model structure?
+This question is explored in Section \ref{sec:identif}.
+The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
+It is the focus of Section \ref{sec:bias}.
+In order to elucidate these matters without confusion from the measurement noise, we switch to the high-SNR regime and set $h_0 = 10^{-12}$ (c.f.
+Figure \ref{fig:bayes}) for the representative system in Table \ref{tab:parameters_and_priors}.
+
+\subsection{Identifiability}\label{sec:identif}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
+	\caption{Posterior probability distributions for the representative system described in Table \ref{tab:parameters_and_priors} over nine noise realisations for the GW source parameters $\iota$ and $h_0$ in the high SNR regime ($h_0 = 10^{-12}$).
+Unlike the other parameters of  $\boldsymbol{\theta}_{\rm gw}$, the one-dimensional posteriors for $\iota$ and $h_0$ do not strongly overlap.
+A correlation between  $\iota$ and $h_0$ is evident.}
+	\label{fig:just_iota_h}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_surface}
+	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota - h_0$ plane for a single noise realisation.
+The red dot marks the maximum of $\log {\cal L}$.
+The colour scale is normalised with respect to the absolute value of the maximum.
+An S-shaped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
+	\label{fig:likelihood_surface}
+\end{figure}
+In Figure \ref{fig:just_iota_h} we plot the one-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations. 
+The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$.
+One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, which is related to the bias discussed in Section 7.2; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high.
+One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$.
+Specifically, across $10^3$ noise realisations, the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
+Both $W_{1,{\rm median}}$ values are higher (as a fraction of the true, injected value) than in the lower-SNR tests conducted in Sections \ref{sec:rep_example} and \ref{sec:parameter_space}.
+In contrast, $W_{1,\rm median}$ for the other five parameters in $\boldsymbol{\theta}_{\rm gw}$ decreases, as the SNR increases.
+\newline
+
+We infer different $h_0$ and $\iota$ posteriors for different noise realizations, because $h_0$ and $\iota$ are weakly identifiable.
+Identifiability is a formal concept from the theory of signal processing, which is applied routinely to engineering problems \citep{e5be7c83a0d24500826f6e1b414d1733}.
+It refers to whether it is theoretically possible to infer unique and accurate parameter values, given the measured data and the model structure .
+Sources of non-identifiability are generally categorised as either ``structural'' , arising from the structure of the model (e.g.
+its invertibility), or ``practical'', arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
+Regarding $\iota$ and $h_0$, the source is structural.
+Equations \eqref{eq:hphx} and \eqref{eq:hphx2} contain a weak degeneracy between $\iota$ and $h_0$.
+For example, the cross polarisation strain $h_{\times}$ involves a trade-off between $\iota$ and $h_0$, which is broken by knowing $h_+$ via Equation \eqref{eq:hphx}.
+However, the measured strain is a linear combination of $h_\times$ and $h_+$; the two polarizations are not measured independently, which makes it challenging to separate $\iota$ and $h_0$.
+In fact $\iota$ and $h_0$ are structurally identifiable --- but only weakly.
+That is, the $\iota$-$h_0$ likelihood contours plateau to a ridge with a small, non-zero gradient close to the maximum likelihood solution.
+The likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
+Whilst a single, unique maximum does exists (labelled by the red dot in the figure), it lies along an S-shaped ridge where the likelihood values are all similar.
+Consequently it is difficult for the nested sampling algorithm to settle on the red dot; it trades off accuracy in one parameter against the other, depending on the particular realisation of the noise.
+By way of verification, if we fix one parameter at its true value (i.e.
+set a delta function prior) during the inference analysis, the posteriors for the other parameter overlap indistinguishably  for different noise realisations, i.e.
+$W_1 \to 0$.
+
+\subsection{Dropping the pulsar terms}\label{sec:pulsar_terms_drop}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
+	\caption{Same as Figure \ref{fig:corner_plot_1} but for a high-SNR system with $h_0 = 10^{-12}$.
+For five out of seven parameters the true, injected value does not fall within the 90\% credible interval and falls outside the plotted domain.
+The estimated posteriors are biased away from the injected value due to dropping the pulsar terms.
+Note that the plotted domain is narrower than in Figure \ref{fig:corner_plot_1}.}
+	\label{fig:bias_for_large_h}
+\end{figure*}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
+	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter model (green line).
+All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
+The coloured dashed vertical lines mark the maximum for the corresponding model; the green and orange dashed lines overlap at this resolution.
+For the Earth-terms model the likelihood maximum ($\iota \approx 0.7$) does not coincide with the injected value ($\iota = 1.0$).
+For the pulsar-terms model the maximum and injected value do coincide.
+10 noise realisations are plotted for each situation but overlap at this resolution; they are indistinguishable to the eye.}
+	\label{fig:likelihood_surface_iota}
+\end{figure}
+
+Sections \ref{sec:rep_example} and \ref{sec:parameter_space} present preliminary evidence that the inferred value of ${\boldsymbol{\theta}}_{\rm gw}$ (e.g.
+the medians of the one-dimensional marginalised posteriors) is biased away from the true injected value.
+Figure \ref{fig:bias_for_large_h} displays a corner plot for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but for high SNR $(h_0 = 10^{-12})$.
+Except for $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it falls outside the plotted domain.
+The deviation is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$.
+Similar results are obtained for other noise realisations.
+Whilst $h_0$ does not exhibit any noticeable bias for the particular noise realisation generating Figure \ref{fig:bias_for_large_h}, in general it does exhibit a correlated bias with $\iota$ across multiple noise realisations, as noted in the bottom left panel of Figure \ref{fig:just_iota_h}.
+In this section we demonstrate how the bias stems from dropping the pulsar terms described in Section \ref{sec:parameter_estim}.
+\newline
+
+In Figure \ref{fig:likelihood_surface_iota} we plot the log-likelihood returned by the Kalman filter, viz.
+Equation \eqref{eq:likelihood}, as a function of $\iota$ (i.e.
+$\log \mathcal{L}(\iota)$) across the prior domain, holding all other parameters constant at the true injected value of the representative system in Table \ref{tab:parameters_and_priors}.
+We generate 10 $\log \mathcal{L}(\iota)$ solutions corresponding to 10 separate realisations of the system noise.
+The solid blue curves only includes the Earth terms (i.e.
+using Equation \eqref{eq:measuremen_earth} in the Kalman filter).
+The solid orange curves include the Earth and pulsar terms (i.e.
+using Equation \eqref{eq:measurement}).
+The solid green curve corresponds to a control experiment, where we drop the pulsar terms from both the synthetic data and the measurement equation.
+The vertical dashed lines mark the maxima, which nested sampling seeks.
+\newline
+
+The key observation from Figure \ref{fig:likelihood_surface_iota} is that the location of the maximum ($\iota \approx 0.7 \, {\rm rad}$) does not coincide with the true, injected value ($\iota = 1.0 \, {\rm rad}$), when only the Earth terms are included.
+This is the origin of the bias in the one-dimensional marginalised posteriors.
+In contrast, the maximum aligns with the true, injected value, when the pulsar terms are included.
+As a cross-check, agreement is also achieved, if we drop the pulsar terms from the synthetic data (which is not the situation in reality, of course) and use the Earth terms only for inference.
+Similar results are obtained for other parameters, e.g.
+$\delta$.
+A discussion on why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches is presented in Appendix \ref{sec:bias_appendix}.
+\newline
+
+We show in this section that dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just in the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}.
+The bias is not particular to our method; it is shared by likelihood based methods that do not include the pulsar terms in the inference model.
+Generalizing the Kalman tracker and nested sampler to correct for the bias will be a key goal of a future paper.
+
+\section{Discussion}\label{sec:discussion}
+We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
+There are a number of potential extensions of this work which we now discuss.
+\newline
+
+The natural first extension is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
+Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred one-dimensional marginalized posteriors for the model parameters.
+Biases in sky localisation for synthetic datasets as a result of dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.
+Alternatively, if the method continues to be used with solely the Earth terms then it would be highly desirable to make systematic, quantitative estimates of the incurred biases in the model parameters across an astrophysically representative parameter space.
+\newline
+
+We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
+Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA.
+Indeed, since adding pulsars to PTAs increases the computational demands of the data analysis it may be advantageous to optimally select a subset of the total available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.
+\newline
+
+Throughout this work we have assumed that all pulsars are observed at the same time, with a constant time sampling.
+In practice this assumption does not hold and different pulsars will be observed with different cadences at different times.
+Extending this method to non-constant time sampling is straightforward and the performance of the method should be evaluated in this regime.
+\newline
+
+We have taken our GW source to be monochromatic, i.e.
+$\Omega$ is a fixed parameter with no time dependence.
+Whilst this is an astronomically well-justified assumption (see discussion in Section \ref{sec:plane_gw}), and appropriate for this initial methods work, it would also be of interest to extend the state-space construction to enable the GW frequency to evolve in time.
+The GW source is monochromatic over the decadal PTA observation period, but the timescale set by the Earth-pulsar distance is comparable to the SMBHB evolution timescale \citep{Sesana2010}.
+For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) the GW source frequency evolution manifests as an incoherent source of noise in the pulsar terms, whilst for  $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms can affect the phase coherency of the Earth terms \citep{Perrodin2018}.
+Careful consideration of the evolution of the GW frequency will be important as we look to extend our method to include the pulsar terms in the inference model.
+\newline
+
+We have also assumed that there is only a single GW source that influences the received pulsar frequencies.
+However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
+Our method naturally extends to detecting multiple GW sources simultaneously; Equation \eqref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
+The stochastic background itself is the incoherent summation of many individual GW sources.
+It therefore stands to reason that it may be possible to extend this method to also search for a stochastic GW background, providing a complementary approach to detect the stochastic background without requiring a cross-correlation of the pulsar residuals to uncover the Hellings-Downs curve.
+\newline
+
+To conclude, in this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
+We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
+That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW induced modulations.
+This enables the inference of GW parameters conditioned on the specific observed realisation of the noisy data.
+We introduce the Kalman filter in order to track the intrinsic state evolution of the pulsar and combine this with a Bayesian sampling framework to estimate the posterior distributions of each parameter of the system, as well as the associated evidence (marginalised likelihood) of the model.
+We test our method on synthetic data.
+We initially focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
+The parameters of the model are accurately recovered and the minimal detectable strain is estimated to be $h_0 \sim 4 \times 10^{-15}$ for $\iota=1.0$ rad.
+The method is then further tested over 1000 noise realisations.
+Consistent posterior estimates of the parameters are obtained for the majority of noise realisations, with the median Wasserstein distance across all noise realisations $W_{1, \rm median}$ generally small compared to the prior space ($\lesssim 3 \%$).
+For a minority of noise realisations the nested sampler converges poorly and settles into a local likelihood optima for the posteriors estimates of $\psi$ and $\alpha$.
+This problem can be solved in general by increasing the number of live points $n_{\rm live}$ used by the sampler.
+Exploration of a broader parameter space via 200 randomly sampled parameter vectors highlights biases in some parameters as a result of dropping the pulsar terms from the inference model.
+The cause of these biases is examined for the case of high SNR, along with the structural weak-identifiability between $\iota$ and $h_0$.
+The extension of this method to include the pulsar term is a key challenge for a future work.
+
+\appendix
+
+\section{Kalman recursion equations} \label{sec:kalman}
+The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
+\begin{equation}
+	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
+\end{equation}
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance, and the subscript $k$ labels the time-step.
+The Kalman filter takes the underlying states to evolve according to
+\begin{equation}
+	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
+\end{equation}
+where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix.
+$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$.
+\newline
+
+The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage.
+The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
+Specifically, the predict step proceeds as,
+\begin{equation}
+	\hat{\boldsymbol{X}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k
+\end{equation}
+\begin{equation}
+	\hat{\boldsymbol{P}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k 
+\end{equation}
+Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
+\begin{equation}
+	\boldsymbol{\epsilon}_{k} = \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1}
+\end{equation} 
+\begin{equation}
+	\boldsymbol{S}_k = \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k
+\end{equation}
+\begin{equation}
+	\boldsymbol{K}_k = \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \label{eq:kalman gain}
+\end{equation}
+\begin{equation}
+	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
+\end{equation}
+\begin{equation}
+		\hat{\boldsymbol{P}}_{k|k} = \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1}
+\end{equation}
+Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
+$\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$.
+For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
+\newline
+
+We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
+\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
+\begin{equation}
+	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
+\end{equation}
+Similarly,  
+\begin{equation}
+\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
+\end{equation}
+The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
+\begin{equation}
+	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
+\end{equation}
+where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix
+\begin{equation}
+\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
+\end{equation}
+and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
+\begin{equation}
+ C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
+\end{equation}
+The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
+\begin{equation}
+\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)
+\end{equation}
+\newline
+
+The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
+\begin{equation}
+	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \label{eq:gardenier}
+\end{equation} 
+From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
+\begin{equation}
+\boldsymbol{F}_k = e^{\boldsymbol{A}\left( t_{k+1} - t_k \right)} 
+\end{equation}
+\begin{equation}
+	\boldsymbol{G}_k = \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' 
+\end{equation}
+\begin{equation}
+	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t')
+	\end{equation}
+For our system, the state, control and process noise covariance matrices then take the form:
+\begin{equation}
+	\boldsymbol{F}_k = 	\text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right)
+\end{equation}
+\begin{equation}
+\boldsymbol{G}_k	= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right)
+\end{equation}
+with
+\begin{align}
+	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
+	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
+\end{align}
+and
+\begin{equation}
+	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) 
+\end{equation}
+for 
+\begin{equation}
+	Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right)
+\end{equation}
+where $\Delta t$ is the constant time sampling interval $=t_{k+1} -t_{k}$. \newline
+
+The remaining component matrices of the Kalman filter that we have not yet specified are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
+These are defined straightforwardly from Equation \eqref{eq:measurement}.
+Specifically,
+$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
+The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
+
+\section{Fiducial pulsar values}\label{appendix_fiducial}
+\textcolor{red}{TK: Do we really need this?
+The argument is for reproducibility, but we have a nice github repo that holds all the data we actually used, jupyter notebooks that show how the calculations are done.
+Perhaps it is sufficient to point the reader to the repo...
+}
+
+\section{Wasserstein distance}\label{sec:wasserstein}
+The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
+It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$.
+For this reason it is sometimes known as the "Earth mover's distance".
+The $p$-th order Wasserstein distance between two distributions is
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \lambda (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
+\end{eqnarray}
+where $p \in [1,\infty)$, $\lambda(x,y)$ is the transport plan, and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
+$\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
+
+The cost function can be freely chosen for the nature of the problem, but is commonly taken to be the absolute value function, $c(x,y) = |x-y|$, and we adopt the same cost function in this work.
+Given this choice of $c(x,y)$, for an empirical distribution $\hat{\mu}_n$ composed of $n$ i.i.d samples drawn from the underlying probability distribution $\mu$, the solution to the discrete form of Equation \eqref{eq:wasserstein} can be computed generally by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
+However, for $\mu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
+\begin{eqnarray}
+	\boldsymbol{E} [W_1(\hat{\mu}_n,\mu)] \asymp n^{-\frac{1}{d}} \ .
+\end{eqnarray}
+In order to circumvent this issue, in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors, i.e.
+setting $d=1$.
+In this case Equation \eqref{eq:wasserstein} has a convenient, easily computed, closed form:
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left(\int_0^1  F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) dz \right)^{1/p} \, \label{eq:wasserstein}
+\end{eqnarray}
+where $F_{\mu}(z)$ is the cumulative distribution function of $\mu$. \newline
+
+The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages.
+Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another.
+It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g.
+radians then we can interpret this as the cost to transform between the two distributions is 1 radian.
+Further interpretability is provided by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
+\begin{eqnarray}
+	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
+\end{eqnarray}
+where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
+That is, if $W_1 =1$ then the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian.
+Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g.
+Gaussianity.
+Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces.
+Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality.
+It also respects the geometry of the underlying metric space over which the distributions are defined.
+\newline
+
+For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
+The results are presented in Figure \ref{fig:pairwise_wasserstein}.
+The figure shows a lower triangular heat-map where each point denotes the $W_1$ value between a pair of posteriors.
+Small values of $W_1$ are magenta moving to red for large values of $W_1$.
+Note that the heat-map colour scaling is not the same for all subplots of the figure.
+It is evident that $W_1$ is generally small across all pairs of posteriors.
+We define $W_{1, \rm median}$ as the median value of $W_1$ across all $10^3 \choose 2$ pairs of noise realisations.
+The values of $W_{1, \rm median}$ for each parameter are quoted in Table \ref{tab:Wasserstein}.
+As a percentage of the width of the prior space (e.g.
+for $\alpha$ the width of the prior is $2 \pi$), the median $W_1$ for each parameter is $\lesssim 3 \%$.
+Generally speaking, the method converges to consistent one-dimensional posteriors for each parameter across different noise realisations.
+\newline
+
+There are a small fraction of the total number of noise realisations which have a large Wasserstein distance from all the other posteriors of all the other noise realisations.
+These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
+These correspond exclusively to posteriors which have converged poorly and not found the likelihood global maximum.
+For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the figure, corresponds to a posterior for $\psi$ which has converged sub-optimally, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
+The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ rad.
+We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
+This is due to multimodal likelihood surface for $\psi-\alpha$, shown in Figure \ref{fig:likelihood_surface_alpha_psi}.
+We can see that the likelihood surface has a global optima at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local optima that corresponds to the location of the example poorly converged posterior discussed previously, with $\psi = 1.01$ rad.
+This problem of poor global convergence can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
+
+\section{Bias from dropping the pulsar terms}\label{sec:bias_appendix}
+Why does dropping the pulsar terms bias the above parameters in particular, whilst leaving others nearly unaffected?
+For instance, why is the bias in $\iota$ so strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
+When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is effective modulated by the interference of two equal-amplitude, phase shifted waves (c.f.
+Equation \eqref{eq:measurement}).
+Explicitly we have an
+``Earth-terms wave'' with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$, and the ``pulsar-terms wave'' with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
+These two terms are related to the measurement equation, Equation \eqref{eq:g_func_trig} as,
+\begin{eqnarray}
+	1 - g^{(n)}(t) =  y_{\rm Earth}  - y_{\rm pulsar} \ .
+\end{eqnarray}
+The net interference wave that modulates $f_{\rm p}$ is 
+\begin{eqnarray}
+	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ .
+\end{eqnarray}
+The inclusion of  $y_{\rm pulsar}$ influences the amplitude of the resulting interference wave by a factor $ 2 \sin (\kappa/2)$, but does but do not affect its frequency.
+When we model this data using the Kalman filter without the pulsar terms, the additional factor from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
+As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude.
+Terms which influence the amplitude strongly, such as $h_0$ and $\iota$, are strongly affected and exhibit a strong bias.
+Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
+The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
+Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
+The same hierarchy of biases is seen also in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
+\newline
+
+In order to ascertain how important the bias is for PTA continuous wave searches requires a thorough exploration of the astrophysical parameter space which we do not carry out in this work.
+The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
+Quantifying the bias in preparation for a practical PTA analysis with real data lies outside the scope of this introductory paper.
+If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate.
+This is what we observe for our synthetic data (e.g.
+Figure \ref{fig:corner_plot_2}); the marginalised posterior broadens as the SNR decreases, until it overwhelms the shift due to the bias.
+Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may systematically affect the inferred parameters.
+\newline
+
+%%%%%%%%%%%%%%%
+
 \begin{figure*}

### 2023-10-27 py_src/config_batch_pp.py

@@ -58,7 +58,7 @@ with open('batch.sh','w') as b:
     for i in range(N):
         p = priors.sample()
 
-        arg_name = f"FINAL_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
+        arg_name = f"OCT_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
         print(arg_name)
         create_slurm_job(arg_name,h,model,seed,p['omega_gw'],p['phi0_gw'],p['psi_gw'],p['delta_gw'],p['alpha_gw'])
         b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")

### 2023-10-29 docs/manuscript/paper.tex
add pp plot script

@@
 
- This stems from the bias noted in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally narrow, but are systematically biased away from the true injection values.
-
- Consequently the
+ This stems from the bias noted in Section \ref{sec:multiple_noise};
+
+\textbf{ NEW DATA INCOMING. See OCT pp}
+
+ the posteriors inferred for these parameters are generally narrow, but are systematically biased away from the true injection values.
+
+ Consequently at higher-value credible intervals the posterior is too narrow to
 
@@
 
- This translates in the PP plot to posteriors which are too narrow to contain the injected value within the credible interval. The origin of the bias is discussed in Section \ref{sec:bias}.
+ This translates in the PP plot to posteriors which are too narrow to contain the injected value within the credible interval.
+
+ The origin of the bias is discussed in Section \ref{sec:bias}.
 

### 2023-10-30 docs/manuscript/paper.tex
package returned to AM

--- 2023-10-27 docs/manuscript/paper.tex
@@
 a bias, or a random outcome specific to this particular noise realisation.
-We explore this further in Sections \ref{sec:multiple_noise} and \ref{sec:bias} using more noise realizations and show that indeed there is a systematic bias from dropping the pulsar terms, similar to that reported by \cite{Zhupulsarterms}.
+We explore this further in Sections \ref{sec:multiple_noise} and \ref{sec:pulsar_terms_drop} using more noise realizations and show that indeed there is a systematic bias from dropping the pulsar terms, similar to that reported by \cite{Zhupulsarterms}.
 \newline
@@
 	\includegraphics[width=\columnwidth]{images/pp_plot_new}
-	\caption{The fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
+	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
 PP plot).
-The injections are 200 simulated GW sources for a subset of five parameters in $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions in Table \ref{tab:parameters_and_priors}.
+The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}.
 Each coloured curve corresponds to a different parameter (see legend).
 The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
-The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals given our statistical sample.
-For parameters with well estimated posteriors, the PP plot curve should fall along the diagonal of unit slope.
+The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals.
+For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope.
 $\Omega$ is generally well-estimated (i.e.
 it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e.
-the curves lie above the unit diagonal for low credible intervals, below the unit diagonal for high credible intervals).
-This is due to a modelling bias.}
+the curves lie above the unit diagonal for low credible intervals, and below the unit diagonal for high credible intervals).
+This is due to a modelling bias whose origin is discussed in Sections \ref{sec:parameter_estim} and \ref{sec:pulsar_terms_drop}.}
 	\label{fig:parameter_space}
@@
 The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within higher-value credible intervals than would be expected statistically, and there are more injections contained within lower-value credible intervals.
-
- This stems from the bias noted in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally narrow, but are systematically biased away from the true injection values.
-
- Consequently the
-
- and the PP plot follows a characterstic S-shape, diagnostic of an under constrained posterior
-
- Consequently at lower-value credible intervals
-
- This translates in the PP plot to posteriors which are too narrow to contain the injected value within the credible interval. The origin of the bias is discussed in Section \ref{sec:bias}.
-
- This stems from the bias noted in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally confident, but are systematically biased away from the true injection values.
-This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval.
-The origin of the bias is discussed in Section \ref{sec:bias}.
+This stems from the bias noted in Section \ref{sec:multiple_noise}.
+The origin of the bias is discussed in Section \ref{sec:pulsar_terms_drop}.
 
@@
 The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
-It is the focus of Section \ref{sec:bias}.
+It is the focus of Section \ref{sec:pulsar_terms_drop}.
 In order to elucidate these matters without confusion from the measurement noise, we switch to the high-SNR regime and set $h_0 = 10^{-12}$ (c.f.
@@
 	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
-	\caption{Posterior probability distributions for the representative system described in Table \ref{tab:parameters_and_priors} over nine noise realisations for the GW source parameters $\iota$ and $h_0$ in the high SNR regime ($h_0 = 10^{-12}$).
-Unlike the other parameters of  $\boldsymbol{\theta}_{\rm gw}$, the one-dimensional posteriors for $\iota$ and $h_0$ do not strongly overlap.
-A correlation between  $\iota$ and $h_0$ is evident.}
+	\caption{Corner plot of the joint posterior distribution of the GW source parameters $h_0$ and $\iota$ for the representative system in Table 1 in the high-SNR regime ($h_0 = 10^{-12}$).
+Nine coloured curves are plotted for nine random noise realizations.
+The posteriors overlap less than for the other five parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+A correlation between $h_0$ and $\iota$ is evident in the bottom left panel.}
 	\label{fig:just_iota_h}
@@
 	\includegraphics[width=\columnwidth]{images/likelihood_surface}
-	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota - h_0$ plane for a single noise realisation.
+	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$- $h_0$ plane for a single noise realisation.
 The red dot marks the maximum of $\log {\cal L}$.
@@
 \end{figure}
-In Figure \ref{fig:just_iota_h} we plot the one-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations. 
+In Figure \ref{fig:just_iota_h} we plot the one- and two-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations.
 The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$.
-One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, which is related to the bias discussed in Section 7.2; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high.
+One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section 7.2; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high.
 One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$.
@@
 Identifiability is a formal concept from the theory of signal processing, which is applied routinely to engineering problems \citep{e5be7c83a0d24500826f6e1b414d1733}.
-It refers to whether it is theoretically possible to infer unique and accurate parameter values, given the measured data and the model structure .
+It refers to whether it is theoretically possible to infer unique and accurate parameter values, given the measured data and the model structure.
 Sources of non-identifiability are generally categorised as either ``structural'' , arising from the structure of the model (e.g.
@@
 The likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
-Whilst a single, unique maximum does exists (labelled by the red dot in the figure), it lies along an S-shaped ridge where the likelihood values are all similar.
+Whilst a single, unique maximum does exist (labelled by the red dot in the figure), it lies along an S-shaped ridge where the likelihood values are all similar.
 Consequently it is difficult for the nested sampling algorithm to settle on the red dot; it trades off accuracy in one parameter against the other, depending on the particular realisation of the noise.
@@
 	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
-	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter model (green line).
+	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth terms in the Kalman filter (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms in the Kalman filter (Equation \ref{eq:measurement}, orange line), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter (green line).
 All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
@@
 Figure \ref{fig:bias_for_large_h} displays a corner plot for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but for high SNR $(h_0 = 10^{-12})$.
-Except for $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it falls outside the plotted domain.
+Except for the elements $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it falls outside the plotted domain.
 The deviation is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$.
@@
 $\log \mathcal{L}(\iota)$) across the prior domain, holding all other parameters constant at the true injected value of the representative system in Table \ref{tab:parameters_and_priors}.
-We generate 10 $\log \mathcal{L}(\iota)$ solutions corresponding to 10 separate realisations of the system noise.
-The solid blue curves only includes the Earth terms (i.e.
+We generate 10 plots of $\log \mathcal{L}(\iota)$ corresponding to 10 separate realisations of the system noise.
+The solid blue curves only includes the Earth terms in the measurement equation (i.e.
 using Equation \eqref{eq:measuremen_earth} in the Kalman filter).
-The solid orange curves include the Earth and pulsar terms (i.e.
+The solid orange curves include the Earth and pulsar terms in the measurement equation (i.e.
 using Equation \eqref{eq:measurement}).
 The solid green curve corresponds to a control experiment, where we drop the pulsar terms from both the synthetic data and the measurement equation.
-The vertical dashed lines mark the maxima, which nested sampling seeks.
+The vertical dashed lines mark the corresponding maxima, which nested sampling seeks.
 \newline
@@
 The key observation from Figure \ref{fig:likelihood_surface_iota} is that the location of the maximum ($\iota \approx 0.7 \, {\rm rad}$) does not coincide with the true, injected value ($\iota = 1.0 \, {\rm rad}$), when only the Earth terms are included.
-This is the origin of the bias in the one-dimensional marginalised posteriors.
+This is the origin of the bias in the one-dimensional marginalised posteriors in Figure \ref{fig:bias_for_large_h}.
 In contrast, the maximum aligns with the true, injected value, when the pulsar terms are included.
@@
 $\delta$.
-A discussion on why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches is presented in Appendix \ref{sec:bias_appendix}.
-\newline
-
-We show in this section that dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just in the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}.
-The bias is not particular to our method; it is shared by likelihood based methods that do not include the pulsar terms in the inference model.
-Generalizing the Kalman tracker and nested sampler to correct for the bias will be a key goal of a future paper.
-
-\section{Discussion}\label{sec:discussion}
-We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
-There are a number of potential extensions of this work which we now discuss.
-\newline
-
-The natural first extension is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
-Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred one-dimensional marginalized posteriors for the model parameters.
-Biases in sky localisation for synthetic datasets as a result of dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.
-Alternatively, if the method continues to be used with solely the Earth terms then it would be highly desirable to make systematic, quantitative estimates of the incurred biases in the model parameters across an astrophysically representative parameter space.
-\newline
-
-We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
-Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA.
-Indeed, since adding pulsars to PTAs increases the computational demands of the data analysis it may be advantageous to optimally select a subset of the total available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.
-\newline
-
-Throughout this work we have assumed that all pulsars are observed at the same time, with a constant time sampling.
-In practice this assumption does not hold and different pulsars will be observed with different cadences at different times.
-Extending this method to non-constant time sampling is straightforward and the performance of the method should be evaluated in this regime.
-\newline
-
-We have taken our GW source to be monochromatic, i.e.
-$\Omega$ is a fixed parameter with no time dependence.
-Whilst this is an astronomically well-justified assumption (see discussion in Section \ref{sec:plane_gw}), and appropriate for this initial methods work, it would also be of interest to extend the state-space construction to enable the GW frequency to evolve in time.
-The GW source is monochromatic over the decadal PTA observation period, but the timescale set by the Earth-pulsar distance is comparable to the SMBHB evolution timescale \citep{Sesana2010}.
-For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) the GW source frequency evolution manifests as an incoherent source of noise in the pulsar terms, whilst for  $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms can affect the phase coherency of the Earth terms \citep{Perrodin2018}.
-Careful consideration of the evolution of the GW frequency will be important as we look to extend our method to include the pulsar terms in the inference model.
-\newline
-
-We have also assumed that there is only a single GW source that influences the received pulsar frequencies.
+A discussion on why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches, are presented in Appendix \ref{sec:bias_appendix}.
+Dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}.
+The bias is not particular to our method; it is shared by all likelihood-based methods that do not include the pulsar terms in the inference model.
+Generalizing the Kalman tracker and nested sampler to correct for the bias will be a key goal of a forthcoming paper.
+
+\section{Conclusion}\label{sec:discussion}
+
+In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs in PTA data.
+The new method is complementary to traditional approaches.
+We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
+That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data.
+We implement a Kalman filter in order to track the intrinsic rotational state evolution of the pulsar and combine it with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model.
+\newline
+
+ We test the new method on synthetic data and find that it detects injected signals successfully and estimates their static parameters accurately with relatively low computational cost.
+We initially focus on a single, astrophysically representative, SMBHB GW source observed synthetically by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$.
+The minimum detectable strain is estimated to be $\min(h_0) \approx 4 \times 10^{-15}$ for $\iota=1.0$ rad.
+We then repeat the parameter estimation exercise for 1000 noise realisations to compute the natural dispersion (``cosmic variance'') in the recovered values and hence quantify the statistical accuracy of the method in a real astrophysical application, when the true parameter values are unknown.
+Consistent posteriors are obtained for most realisations, with the median Wasserstein distance limited to $\lesssim 3\%$ of the width of the prior domain (i.e.\ negligible railing).
+The median Wasserstein distance as a fraction of the injected value, an approximate measure of the natural dispersion, ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
+\newline
+
+Exploration of a broader SMBHB parameter domain at fixed $\iota$ and $h_0$ via 200 randomly sampled parameter vectors suggests at the presence of a bias in the estimates of the static parameters $\delta$ and $\alpha$, (and to a lesser extent in $\psi$ and $\Phi_0$).
+The bias is examined for a specific SMBHB GW source in the limit of high-SNR.
+It is greatest for $\iota$ at $\sim 0.3$ rad, but smaller biases of $ \lessapprox 0.1$ rad are also observed in $\Phi_0, \psi, \delta$ and $\alpha$.
+The bias is shown to result from dropping the pulsar terms from the measurement equation in the Kalman filter.
+The method also exhibits a trade-off between $h_0$ and $\iota$, which follows formally from the weak identifiability of $h_0$ and $\iota$.
+\newline
+
+We emphasize that the Kalman tracker and nested sampler in this paper do not supplant traditional PTA analysis approaches; they complement traditional approaches and are most powerful when used in tandem.
+Relatedly, it is misleading to ask whether the Kalman tracker and nested sampler are more or less sensitive than traditional approaches for two reasons.
+First, one must still generalize the Kalman tracker and nested sampler to ingest TOAs directly, as happens traditionally, instead of ingesting a frequency time series, as in this paper.
+Generalizing the algorithm is a subtle task and is postponed to a forthcoming paper.
+Second, the Kalman tracker and nested sampler are conditional on a noise model that is related but different to the noise model in traditional approaches.
+The analysis in this paper assumes a specific, time-ordered realization of a mean-reverting Ornstein-Uhlenbeck process satisfying Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law PSD, whose amplitude and exponent are adjustable.
+Hence the sensitivities cannot be compared directly.
+Clarifying the similarities and differences between various approaches promises to be a fruitful avenue of future work.
+It is also the subject of attention in audio-band GW data analysis involving hidden Markov models applied to data from terrestrial long-baseline interferometers \citep{PhysRevD.102.023006,PhysRevD.105.022002,Abbott_2022SCO,2022PhRvD.106f2002A}.
+\newline
+
+ The approach in this paper can be extended in at least five ways, enumerated as (i)--(v) below. \newline
+
+(i) A natural first extension is to retain the pulsar terms in Equation \eqref{eq:measurement}.
+Alternatively, if the method continues to be used with solely the Earth terms, in line with standard practice in published PTA analyses, it would be desirable to evaluate systematically the incurred biases in the model parameters across an astrophysically representative parameter domain, supplementing the results obtained by other authors \citep{Zhupulsarterms,Chen2022}.\newline
+
+(ii) In this paper, we consider one specific configuration of the synthetic PTA, namely the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
+It is interesting to compare the performance of the method using different pulsar configurations, e.g.
+those in the PPTA and EPTA.
+Adding pulsars to PTAs increases the computational cost so it may be advantageous to select a subset of pulsars by exploiting formal optimization techniques from electrical engineering\citep{2023MNRAS.518.1802S}, although the Kalman filter and nested sampler are already cheaper computationally than some other methods.
+\newline
+
+(iii) In practice different pulsars in a PTA are be observed with different cadences at different times.
+Extending the Kalman filter to non-uniform time sampling is straightforward \citep{zarchan2000fundamentals}.
+\newline
+
+(iv) The assumption of a monochromatic source is well-justified astrophysically in various regimes (see Section \ref{sec:plane_gw}) and is an appropriate starting point for this introductory paper.
+Nevertheless, SMBHBs are not strictly monochromatic.
+It is interesting to extend the state-space framework such that $f_{\rm gw}$ evolves in time.
+For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) evolution adds noise incoherently to the pulsar terms, whilst for $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms induce phase shifts that affect the overall phase coherence \citep{Sesana2010,Perrodin2018}.
+Careful consideration of the evolution of $f_{\rm gw}$ will be needed when including the pulsar terms in the inference model, as in point (i).
+\newline
+
+(v) Finally, we assume in this paper that there is only one GW source.
 However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
-Our method naturally extends to detecting multiple GW sources simultaneously; Equation \eqref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
-The stochastic background itself is the incoherent summation of many individual GW sources.
-It therefore stands to reason that it may be possible to extend this method to also search for a stochastic GW background, providing a complementary approach to detect the stochastic background without requiring a cross-correlation of the pulsar residuals to uncover the Hellings-Downs curve.
-\newline
-
-To conclude, in this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
-We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
-That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW induced modulations.
-This enables the inference of GW parameters conditioned on the specific observed realisation of the noisy data.
-We introduce the Kalman filter in order to track the intrinsic state evolution of the pulsar and combine this with a Bayesian sampling framework to estimate the posterior distributions of each parameter of the system, as well as the associated evidence (marginalised likelihood) of the model.
-We test our method on synthetic data.
-We initially focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
-The parameters of the model are accurately recovered and the minimal detectable strain is estimated to be $h_0 \sim 4 \times 10^{-15}$ for $\iota=1.0$ rad.
-The method is then further tested over 1000 noise realisations.
-Consistent posterior estimates of the parameters are obtained for the majority of noise realisations, with the median Wasserstein distance across all noise realisations $W_{1, \rm median}$ generally small compared to the prior space ($\lesssim 3 \%$).
-For a minority of noise realisations the nested sampler converges poorly and settles into a local likelihood optima for the posteriors estimates of $\psi$ and $\alpha$.
-This problem can be solved in general by increasing the number of live points $n_{\rm live}$ used by the sampler.
-Exploration of a broader parameter space via 200 randomly sampled parameter vectors highlights biases in some parameters as a result of dropping the pulsar terms from the inference model.
-The cause of these biases is examined for the case of high SNR, along with the structural weak-identifiability between $\iota$ and $h_0$.
-The extension of this method to include the pulsar term is a key challenge for a future work.
+The Kalman filter extends naturally to multiple sources; one can modify Equation \eqref{eq:measurement} easily to accommodate a linear superposition of GWs.
+Taking the logic further, the stochastic background itself is an incoherent sum of many individual GW sources.
+As long as a way can be found to summarize economically the many static parameters associated with the background sources, it should be possible for a Kalman filter and nested sampler to operate together to detect the stochastic background by generalizing the model selection procedure in Sections \ref{sec:model_selection} and \ref{sec:detection}.
+Summarizing the parameter set economically, while respecting the mathematical structure of the Kalman filter, is a subtle challenge, which we postpone to a forthcoming paper.
+If successful, it will complement the traditional approach of cross-correlating pulsar residuals to uncover the Hellings-Downs curve \citep{Hellings,2023ApJ...951L...8A}.
+\newline
+
+\section*{Acknowledgements}
+This research was supported by the Australian Research Council Centre of Excellence for Gravitational Wave Discovery (OzGrav), grant number CE170100004.
+A.
+The numerical calculations were performed on the OzSTAR supercomputer facility at Swinburne University of Technology.
+The OzSTAR program receives funding in part from the Astronomy National Collaborative Research Infrastructure Strategy (NCRIS) allocation provided by the Australian Government.
+
+%%%% REFERENCES
+\bibliographystyle{mnras}
+\bibliography{example} % if your bibtex file is called example.bib
 
@@
 
-\section{Kalman recursion equations} \label{sec:kalman}
-The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
+\section{Kalman filtering} \label{sec:kalman}
+In this appendix, we describe the Kalman filter algorithm used in this paper.
+The general recursion relations for the discrete-time Kalman filter applied to an arbitrary linear dynamical system.
+are outlined in Section \ref{sec_kalman_general}.
+The mapping onto our specific continuous-time state-space model described in Section \ref{sec:model} is given in Section \ref{sec_kalman_specific}.
+
+\subsection{Recursion equations}\label{sec_kalman_general}
+The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation
 \begin{equation}
@@
 \end{equation}
-where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance, and the subscript $k$ labels the time-step.
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
 The Kalman filter takes the underlying states to evolve according to
@@
 \end{equation}
-where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix.
-$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$.
-\newline
+where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ the control matrix.
+$\boldsymbol{u}_k$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline
 
@@
 Specifically, the predict step proceeds as,
-\begin{equation}
-	\hat{\boldsymbol{X}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k
-\end{equation}
-\begin{equation}
-	\hat{\boldsymbol{P}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k 
-\end{equation}
+\begin{align}
+	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
+		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ .
+\end{align}
 Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
-\begin{equation}
-	\boldsymbol{\epsilon}_{k} = \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1}
-\end{equation} 
-\begin{equation}
-	\boldsymbol{S}_k = \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k
-\end{equation}
-\begin{equation}
-	\boldsymbol{K}_k = \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \label{eq:kalman gain}
-\end{equation}
-\begin{equation}
-	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
-\end{equation}
-\begin{equation}
-		\hat{\boldsymbol{P}}_{k|k} = \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1}
-\end{equation}
+\begin{align}
+	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \\
+	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \\
+	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
+		\hat{\boldsymbol{X}}_{k|k} &=\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \ , \label{eq:kalmangainupdate} \\
+			\hat{\boldsymbol{P}}_{k|k} &= \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1} \ .
+\end{align}
 Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
@@
 
-We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
-\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
-\begin{equation}
-	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
-\end{equation}
-Similarly,  
-\begin{equation}
-\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
+To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$,$\boldsymbol{Y}_k$,$\boldsymbol{F}_k$,$\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$,$\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$.
+In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.
+
+\subsection{Full state space representation}\label{sec_kalman_specific}
+We apply the Kalman machinery of Section \ref{sec_kalman_general} onto the state-space model described in Section \ref{sec:model} as follows. \newline
+
+We identify $\boldsymbol{X}(t)$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
+\begin{equation}
+	\boldsymbol{X}(t) = \left(f_{\rm p}^{(1)}(t), f_{\rm p}^{(2)}(t), ..., f_{\rm p}^{(N)}(t)\right) \ .
+\end{equation}
+Analogously,  
+\begin{equation}
+	\boldsymbol{Y}(t) = \left(f_{\rm m}^{(1)}(t), f_{\rm m}^{(2)}(t), ..., f_{\rm m}^{(N)}(t) \right) \ .
 \end{equation}
@@
 \end{equation}
-where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix
+where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix,
 \begin{equation}
@@
 \begin{equation}
-\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)
-\end{equation}
-\newline
+	\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right) \ .
+\end{equation}
 
@@
 \begin{equation}
-	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \label{eq:gardenier}
-\end{equation} 
-From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
-\begin{equation}
-\boldsymbol{F}_k = e^{\boldsymbol{A}\left( t_{k+1} - t_k \right)} 
-\end{equation}
-\begin{equation}
-	\boldsymbol{G}_k = \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' 
-\end{equation}
-\begin{equation}
-	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t')
-	\end{equation}
-For our system, the state, control and process noise covariance matrices then take the form:
-\begin{equation}
-	\boldsymbol{F}_k = 	\text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right)
-\end{equation}
-\begin{equation}
-\boldsymbol{G}_k	= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right)
-\end{equation}
-with
+	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ .
+\label{eq:gardenier}
+\end{equation} 
+From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
+\begin{align}
+	\boldsymbol{F}_k &= e^{\boldsymbol{A}\left(\Delta t \right)} \  \\
+	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
+\end{align}
+\begin{align}
+	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
+	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ' ,
+\end{align}
+\begin{equation}
+	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \label{eq:appendix_noise}
+\end{equation}
+where $ \Delta t = t_{k+1} - t_k$ and 
 \begin{align}
 	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
-	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
+	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ .
 \end{align}
-and
-\begin{equation}
-	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) 
+From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
+\begin{equation}
+	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) \ ,
 \end{equation}
@@
 \begin{equation}
-	Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right)
-\end{equation}
-where $\Delta t$ is the constant time sampling interval $=t_{k+1} -t_{k}$. \newline
-
-The remaining component matrices of the Kalman filter that we have not yet specified are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
-These are defined straightforwardly from Equation \eqref{eq:measurement}.
+Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right) \ .
+\end{equation}
+
+The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
+These are defined straightforwardly from Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
 Specifically,
@@
 \section{Fiducial pulsar values}\label{appendix_fiducial}
-\textcolor{red}{TK: Do we really need this?
+\textcolor{red}{TK: AM suggests a table of the pulsars used with there fiducial values.
+Do we really need this?
 The argument is for reproducibility, but we have a nice github repo that holds all the data we actually used, jupyter notebooks that show how the calculations are done.
 Perhaps it is sufficient to point the reader to the repo...
-}
+MNRAS encourages a "data availability" section before Acknowledgements that points readers to code/data/github repository}
 
 \section{Wasserstein distance}\label{sec:wasserstein}
+In this section we describe the Wasserstein distance used in Section \ref{sec:multiple_noise}.
+In Section \ref{sec:appendix_overview_WD} we give a brief overview of the Wasserstein distance generally.
+In Section \ref{sec:appendix_WD_results} we present for reproducibility the Wasserstein distance calculated between every pair of posteriors for each static parameter in $\boldsymbol{\theta}_{\rm gw}$ across the $10^3$ noise realisations of Section  \ref{sec:multiple_noise}.
+This corresponds to the mean values reported in Table \ref{tab:Wasserstein}.
+
+\subsection{Overview of the Wasserstein distance}\label{sec:appendix_overview_WD}
+
 The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
@@
 
-For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
-The results are presented in Figure \ref{fig:pairwise_wasserstein}.
-The figure shows a lower triangular heat-map where each point denotes the $W_1$ value between a pair of posteriors.
-Small values of $W_1$ are magenta moving to red for large values of $W_1$.
+\subsection{Wasserstein distance for static parameters} \label{sec:appendix_WD_results}
+
+Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+In Figure \ref{fig:pairwise_wasserstein} we present the WD values for a subset ($=10^2$) of those noise realisations.
+We plot a subset of realisations so as not to overcrowd the figure and enable certain features in the figure to be resolved.
+\newline
+
+The figure is composed of seven subplots, one for each element or ${\boldsymbol{\theta}}_{\rm gw}$.
+Each individual subplot is a lower triangular heat-map, where each point denotes the $W_1$ value between a pair of one-dimensional posteriors for that element.
+Smaller values of $W_1$ are magenta moving to through red to yellow for larger values of $W_1$.
 Note that the heat-map colour scaling is not the same for all subplots of the figure.
-It is evident that $W_1$ is generally small across all pairs of posteriors.
-We define $W_{1, \rm median}$ as the median value of $W_1$ across all $10^3 \choose 2$ pairs of noise realisations.
-The values of $W_{1, \rm median}$ for each parameter are quoted in Table \ref{tab:Wasserstein}.
-As a percentage of the width of the prior space (e.g.
-for $\alpha$ the width of the prior is $2 \pi$), the median $W_1$ for each parameter is $\lesssim 3 \%$.
-Generally speaking, the method converges to consistent one-dimensional posteriors for each parameter across different noise realisations.
-\newline
-
-There are a small fraction of the total number of noise realisations which have a large Wasserstein distance from all the other posteriors of all the other noise realisations.
+\newline
+
+As discussed in Section \ref{sec:multiple_noise} $W_1$ is generally small across all pairs of posteriors.
+The median $W_1$ across all $10^3 \choose 2$ pairs of noise realisations is reported in Table \ref{tab:Wasserstein}.
+However for some static parameters there are a small fraction ($\lesssim 5 \%$) of posterior distributions which have a notably larger $W_1$ value from all the other posteriors of all the other noise realisations.
 These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
@@
 
-%\section{References}
-%\label{sec:ref_list}
-
-\bibliographystyle{mnras}
-\bibliography{example} % if your bibtex file is called example.bib
-
 %%%%%%%%%%%%%%%%Scratch space

### 2023-11-03 notebooks/plot_likelihood_surfaces.py
major clean up c.f. reproducibility

@@ -1,103 +1,0 @@
-
-
-import numpy as np
-import matplotlib.pyplot as plt  
-
-
-
-
-
-import matplotlib.pyplot as plt 
-fig = plt.figure(figsize=(6,6))
-ax = fig.add_subplot(111, projection='3d')
-#ax = fig.add_subplot(111)
-
-
-
-
-
-
-
-
-path_to_earth_model = 'likelihood_surface_h_iota_mm_earth.npz'
-path_to_earth_model = 'likelihood_surface_h_iota_mm_earth_coarse.npz'
-
-
-# path1 = 'tmp_likelihood_surface.npz'
-# path2 = 'tmp_likelihood_surface_earth.npz'
-
-def load_and_plot(path,ax):
-
-    data = np.load(path)
-
-
-    iota_values = data['x']
-    h_values = data['y']
-    surface_pulsar = data['z']
-    iota_idx, h_idx = np.unravel_index(surface_pulsar.argmax(), surface_pulsar.shape)
-
-
-
-
-
-
-
-    xc = iota_values[iota_idx]
-    yc = h_values[h_idx]
-    zc = surface_pulsar[iota_idx,h_idx]
-
-
-    X,Y = np.meshgrid(h_values,iota_values)
-
-
-   
-    lx = len(iota_values)
-    ly = len(h_values)
-    z = np.reshape(surface_pulsar, (lx, ly))
-   
-    # Plot a 3D surface
-    ax.plot_surface(X, Y, z,alpha=0.5)
-
-    #CS = ax.pcolormesh(X, Y, z) #clim=(0.0, 5e5)
-    #plt.contour(X, Y, Z, 4, colors='k')
-    #plt.colorbar(CS)
-   # ax.set_xscale('log')
-
-
-
-    print(path, yc,xc,zc)
-    ax.scatter(yc,xc,zc, s=20)
-    ax.scatter(1e-12,1.0,zc, s=20)
-
-
-    # https://stackoverflow.com/questions/54495612/numpy-get-index-of-row-with-second-largest-value
-    row,col = np.unravel_index(np.argsort(surface_pulsar.ravel()),surface_pulsar.shape)
-    row,col = row[::-1],col[::-1]
-    
-    for k in np.arange(10):
-        idx=k
-        row_i,col_i = row[idx], col[idx]
-
-        xc = iota_values[row_i]
-        yc = h_values[col_i]
[…]

### 2023-11-04 docs/manuscript/paper.tex
major clean up c.f. reproducibility
initial fixes

@@
 Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing arrays.
-A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-A two-step procedure is applied within a state-space framework, such that the modulation is tracked with a Kalman filter, which then provides a likelihood for nested sampling.
-The procedure enables the static parameters in the problem, such as the sky position of the source, to be inferred without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional parameter estimation methods.
-It also returns the Bayes factor relating a model with a single-source gravitational wave to one without, complementing traditional detection methods.
+A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single gravitational wave source in the pulse times of arrival.
+A two-step inference procedure is applied within a state-space framework, such that the modulation is tracked with a Kalman filter, which then provides a likelihood for nested sampling.
+The procedure estimates the static parameters in the problem, such as the sky position of the source, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional parameter estimation methods.
+It also returns the Bayes factor relating a model with a single gravitational wave source to one without, complementing traditional detection methods.
 It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$.
@@
 \subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
-%\begin{figure*}
-%	%\centering % Not needed
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
-%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
-%		\label{fig:6MB_BFS}
-%	\end{subfigure}  
-%    \hfill
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
-%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
-%		\label{fig:25MB_bfs}
-%	\end{subfigure}
-%	\caption{
-%		}
-%	\label{fig:four figures}
-%\end{figure*}
 
@@
 We see that the synthetic and empirical residuals are qualitatively similar.
-For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is much less than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ for PSR J1939$+$2134.
+For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is much less than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ s$^{-3/2}$ for PSR J1939$+$2134.
 
@@
 The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space}--\ref{sec:discussion}.
+The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space} and \ref{sec:bias_and_identifiability}.
 To assist with reproducibility, we present intermediate outputs from each step.
-In Section \ref{sec:priors} we define the priors chosen on $\boldsymbol{\theta}$.
+In Section \ref{sec:priors} we define the priors on $\boldsymbol{\theta}$.
 In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter.
@@
 		\hline
-		\vspace{1mm}\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
+%		\vspace{1mm}\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
+		 & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
 		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
-		\vspace{1mm} &  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
+		\vspace{1mm} $\boldsymbol{\theta}_{\rm psr}$ &  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
 		\vspace{1mm} & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
@@
 from the Kalman filter model that feeds into the nested sampling algorithm.
-The synthetic data include the pulsar term in full.
+The synthetic data include the pulsar terms in full.
 \newline
@@
 There is scant evidence of railing against the prior bounds.
-Hence it is evident for this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, that the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
+For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
 The injected values are contained within the 90\% credible interval for all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ .
@@
 		$\alpha$          & $1.0$  & rad & $0.17$\\
-		$h_0$            & $5 \times 10^{-15}$ & - & $8 \times 10^{-16}$ \\
+		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
 		\bottomrule

### 2023-11-04 docs/manuscript/paper.tex
all correctins up to appendix

@@
 The supertitles of the marginalized one-dimensional histograms specify the posterior median and the 0.16 and 0.84 quantiles.
-We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
+We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$.
 The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
 	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely.
@@
 As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
-These values are an approximate measure of the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
+These values quantify approximately the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
 Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$.
-These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for a wide variety of noise realizations.
+These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for $10^3$ noise realizations.
 Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
@@
 We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}.
-In Equation \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-terms-only model with a GW present, i.e.
+In Equation \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-term-only model with a GW present, i.e.
 the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}.
@@
 
-We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
 We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
@@
 	\includegraphics[width=\columnwidth]{images/likelihood_surface}
-	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$- $h_0$ plane for a single noise realisation.
+	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$-$h_0$ plane for a single noise realisation.
 The red dot marks the maximum of $\log {\cal L}$.
 The colour scale is normalised with respect to the absolute value of the maximum.
-An S-shaped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
+An S-shaped, flat-topped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
 	\label{fig:likelihood_surface}
@@
 Equations \eqref{eq:hphx} and \eqref{eq:hphx2} contain a weak degeneracy between $\iota$ and $h_0$.
-For example, the cross polarisation strain $h_{\times}$ involves a trade-off between $\iota$ and $h_0$, which is broken by knowing $h_+$ via Equation \eqref{eq:hphx}.
+For example, the cross polarisation strain $h_{\times}$ involves a trade-off between $\iota$ and $h_0$, which is resolved by knowing $h_+$ via Equation \eqref{eq:hphx}.
 However, the measured strain is a linear combination of $h_\times$ and $h_+$; the two polarizations are not measured independently, which makes it challenging to separate $\iota$ and $h_0$.
 In fact $\iota$ and $h_0$ are structurally identifiable --- but only weakly.
-That is, the $\iota$-$h_0$ likelihood contours plateau to a ridge with a small, non-zero gradient close to the maximum likelihood solution.
+That is, the $\iota$-$h_0$ likelihood contours plateau to a flat-topped ridge with a small, non-zero gradient close to the maximum likelihood solution.
 The likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
@@
 	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
-	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth terms in the Kalman filter (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms in the Kalman filter (Equation \ref{eq:measurement}, orange line), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter (green line).
+	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth term in the Kalman filter (Equation \ref{eq:measuremen_earth}, blue curve), the pulsar terms in the Kalman filter (Equation \ref{eq:measurement}, orange curve), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter (green curve).
 All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
 The coloured dashed vertical lines mark the maximum for the corresponding model; the green and orange dashed lines overlap at this resolution.
-For the Earth-terms model the likelihood maximum ($\iota \approx 0.7$) does not coincide with the injected value ($\iota = 1.0$).
+For the Earth-term model the likelihood maximum ($\iota \approx 0.7$ rad) does not coincide with the injected value ($\iota = 1.0$ rad).
 For the pulsar-terms model the maximum and injected value do coincide.
@@
 Figure \ref{fig:bias_for_large_h} displays a corner plot for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but for high SNR $(h_0 = 10^{-12})$.
-Except for the elements $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it falls outside the plotted domain.
-The deviation is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$.
+Except for the elements $\Omega$ and $h_0$, the injected values lie outside the 90\% credible interval, and indeed fall outside the plotted domain.
+The deviation is most severe for $\iota$ with a bias of $\approx 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$.
 Similar results are obtained for other noise realisations.
@@
 We generate 10 plots of $\log \mathcal{L}(\iota)$ corresponding to 10 separate realisations of the system noise.
-The solid blue curves only includes the Earth terms in the measurement equation (i.e.
+The solid blue curves only includes the Earth term in the measurement equation (i.e.
 using Equation \eqref{eq:measuremen_earth} in the Kalman filter).
@@
 In contrast, the maximum aligns with the true, injected value, when the pulsar terms are included.
-As a cross-check, agreement is also achieved, if we drop the pulsar terms from the synthetic data (which is not the situation in reality, of course) and use the Earth terms only for inference.
+As a cross-check, agreement is also achieved, if we drop the pulsar terms from the synthetic data (which is not the situation in reality, of course) and use the Earth term only for inference.
 Similar results are obtained for other parameters, e.g.
 $\delta$.
-A discussion on why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches, are presented in Appendix \ref{sec:bias_appendix}.
+A discussion about why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches, are presented in Appendix \ref{sec:bias_appendix}.
 Dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}.
@@
 The new method is complementary to traditional approaches.
-We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
+We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than fitting for the ensemble-averaged power spectral density of the noise.
 That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data.
-We implement a Kalman filter in order to track the intrinsic rotational state evolution of the pulsar and combine it with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model.
+We implement a Kalman filter in order to track the intrinsic rotational state evolution of the pulsar and combine it with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model with and without a GW.
 \newline
@@
 We then repeat the parameter estimation exercise for 1000 noise realisations to compute the natural dispersion (``cosmic variance'') in the recovered values and hence quantify the statistical accuracy of the method in a real astrophysical application, when the true parameter values are unknown.
-Consistent posteriors are obtained for most realisations, with the median Wasserstein distance limited to $\lesssim 3\%$ of the width of the prior domain (i.e.\ negligible railing).
-The median Wasserstein distance as a fraction of the injected value, an approximate measure of the natural dispersion, ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
-\newline
-
-Exploration of a broader SMBHB parameter domain at fixed $\iota$ and $h_0$ via 200 randomly sampled parameter vectors suggests at the presence of a bias in the estimates of the static parameters $\delta$ and $\alpha$, (and to a lesser extent in $\psi$ and $\Phi_0$).
+Consistent posteriors are obtained for most realisations.
+The median WD is limited to $\lesssim 3\%$ of the width of the prior domain (i.e.\ negligible railing).
+The median WD divided by the injected value, an approximate measure of the natural dispersion, ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
+\newline
+
+Exploration of a broader SMBHB parameter domain at fixed $\iota$ and $h_0$ via 200 randomly sampled parameter vectors reveals a bias in the estimates of the static parameters $\delta$ and $\alpha$, and to a lesser extent $\psi$ and $\Phi_0$.
 The bias is examined for a specific SMBHB GW source in the limit of high-SNR.
-It is greatest for $\iota$ at $\sim 0.3$ rad, but smaller biases of $ \lessapprox 0.1$ rad are also observed in $\Phi_0, \psi, \delta$ and $\alpha$.
-The bias is shown to result from dropping the pulsar terms from the measurement equation in the Kalman filter.
+It is greatest for $\iota$, amounting to $\approx 0.3$ rad.
+Smaller biases of $\lesssim 0.1$ rad are also observed in $\Phi_0, \psi, \delta$ and $\alpha$.
+The bias is shown to result from dropping the pulsar terms from the measurement equation in the Kalman filter, consistent with the work of previous authors \citep{Zhupulsarterms,Chen2022}.
 The method also exhibits a trade-off between $h_0$ and $\iota$, which follows formally from the weak identifiability of $h_0$ and $\iota$.
@@
 Second, the Kalman tracker and nested sampler are conditional on a noise model that is related but different to the noise model in traditional approaches.
-The analysis in this paper assumes a specific, time-ordered realization of a mean-reverting Ornstein-Uhlenbeck process satisfying Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law PSD, whose amplitude and exponent are adjustable.
+The analysis in this paper assumes a specific, time-ordered realization of a mean-reverting Ornstein-Uhlenbeck process satisfying Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law power spectral density, whose amplitude and exponent are adjustable.
 Hence the sensitivities cannot be compared directly.
 Clarifying the similarities and differences between various approaches promises to be a fruitful avenue of future work.
-It is also the subject of attention in audio-band GW data analysis involving hidden Markov models applied to data from terrestrial long-baseline interferometers \citep{PhysRevD.102.023006,PhysRevD.105.022002,Abbott_2022SCO,2022PhRvD.106f2002A}.
+It is also a subject of attention in audio-band GW data analysis involving hidden Markov models applied to data from terrestrial long-baseline interferometers \citep{PhysRevD.102.023006,PhysRevD.105.022002,Abbott_2022SCO,2022PhRvD.106f2002A}.
 \newline
@@
 (i) A natural first extension is to retain the pulsar terms in Equation \eqref{eq:measurement}.
-Alternatively, if the method continues to be used with solely the Earth terms, in line with standard practice in published PTA analyses, it would be desirable to evaluate systematically the incurred biases in the model parameters across an astrophysically representative parameter domain, supplementing the results obtained by other authors \citep{Zhupulsarterms,Chen2022}.\newline
+Alternatively, if the method continues to be used with solely the Earth term, in line with standard practice in published PTA analyses, it would be desirable to evaluate systematically the incurred biases in the model parameters across an astrophysically representative parameter domain, supplementing the results obtained by other authors \citep{Zhupulsarterms,Chen2022}.\newline
 
@@
 those in the PPTA and EPTA.
-Adding pulsars to PTAs increases the computational cost so it may be advantageous to select a subset of pulsars by exploiting formal optimization techniques from electrical engineering\citep{2023MNRAS.518.1802S}, although the Kalman filter and nested sampler are already cheaper computationally than some other methods.
-\newline
-
-(iii) In practice different pulsars in a PTA are be observed with different cadences at different times.
+Adding pulsars to PTAs increases the computational cost so it may be advantageous to select a subset of pulsars by exploiting formal optimization techniques from electrical engineering \citep{2023MNRAS.518.1802S}, although the Kalman filter and nested sampler are already cheaper computationally than some other methods.
+\newline
+
+(iii) In practice different PTA pulsars are observed with different cadences at different times.
 Extending the Kalman filter to non-uniform time sampling is straightforward \citep{zarchan2000fundamentals}.
@@
 The Kalman filter extends naturally to multiple sources; one can modify Equation \eqref{eq:measurement} easily to accommodate a linear superposition of GWs.
-Taking the logic further, the stochastic background itself is an incoherent sum of many individual GW sources.
+Taking the logic further, the stochastic background itself is arguably an incoherent sum of many individual GW sources.
 As long as a way can be found to summarize economically the many static parameters associated with the background sources, it should be possible for a Kalman filter and nested sampler to operate together to detect the stochastic background by generalizing the model selection procedure in Sections \ref{sec:model_selection} and \ref{sec:detection}.
@@
 This research was supported by the Australian Research Council Centre of Excellence for Gravitational Wave Discovery (OzGrav), grant number CE170100004.
-A.
 The numerical calculations were performed on the OzSTAR supercomputer facility at Swinburne University of Technology.

### 2023-11-05 docs/manuscript/paper.tex
all comments addressed

@@
 
-For each pulsar we adopt fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
+For each pulsar we adopt fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
 A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility.
@@
 
-\section{Kalman filtering} \label{sec:kalman}
+\section{Kalman filter} \label{sec:kalman}
 In this appendix, we describe the Kalman filter algorithm used in this paper.
-The general recursion relations for the discrete-time Kalman filter applied to an arbitrary linear dynamical system.
-are outlined in Section \ref{sec_kalman_general}.
-The mapping onto our specific continuous-time state-space model described in Section \ref{sec:model} is given in Section \ref{sec_kalman_specific}.
+General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}.
+The mapping onto the specific continuous-time state-space model in Section \ref{sec:model} is written down in Section \ref{sec_kalman_specific}.
 
@@
 where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
-The Kalman filter takes the underlying states to evolve according to
+The Kalman filter evolves the underlying states according to
 \begin{equation}
@@
 \end{equation}
-where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ the control matrix.
-$\boldsymbol{u}_k$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline
-
-The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage.
-The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
-Specifically, the predict step proceeds as,
+where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix.
+$\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline
+
+The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage.
+The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
+Specifically, the predict step proceeds as
 \begin{align}
 	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
-		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ .
+		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ ,
 \end{align}
-Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
+where $\hat{\boldsymbol{P}}_{k|k-1}$ is the covariance of the prediction.
+Note that the predict stage is independent of the measurements.
+The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
 \begin{align}
@@
 Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
-$\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$.
+$\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$.
 For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
@@
 
-To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$,$\boldsymbol{Y}_k$,$\boldsymbol{F}_k$,$\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$,$\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$.
+To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$.
 In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.
 
-\subsection{Full state space representation}\label{sec_kalman_specific}
-We apply the Kalman machinery of Section \ref{sec_kalman_general} onto the state-space model described in Section \ref{sec:model} as follows. \newline
-
-We identify $\boldsymbol{X}(t)$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
+\subsection{State space representation of a PTA analysis}\label{sec_kalman_specific}
+We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:model} as follows. \newline
+
+We identify $\boldsymbol{X}(t)$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e. 
 \begin{equation}
@@
 \end{equation}
-Analogously,  
+Analogously,  we package the measured pulsar frequencies as
 \begin{equation}
@@
 \end{equation}
-The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
-\begin{equation}
-	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
-\end{equation}
-where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix,
-\begin{equation}
-	\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
-\end{equation}
-and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
-\begin{equation}
-	C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
-\end{equation}
-The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
+The states evolve according to the continuous stochastic differential equation (c.f. Equation \eqref{eq:frequency_evolution})
+\begin{equation}
+	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \ , \label{eq:kalmn2}
+\end{equation}
+where $\boldsymbol{A}$ is a diagonal $N \times N$ matrix,
+\begin{equation}
+	\boldsymbol{A} = \text{diag} \left(-\gamma^{(1)}, -\gamma^{(2)}, ..., -\gamma^{(N)}\right) \ ,
+\end{equation}
+and $\boldsymbol{C}(t)$ is a time-dependent vector with $n$-th component
+\begin{equation}
+	C^{(n)} =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
+\end{equation}
+The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, with
 \begin{equation}
@@
 
-The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
+In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic},
 \begin{equation}
@@
 \end{equation} 
-From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
+From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2}, with
 \begin{align}
-	\boldsymbol{F}_k &= e^{\boldsymbol{A}\left(\Delta t \right)} \  \\
+	\boldsymbol{F}_k &= e^{\boldsymbol{A} \Delta t } \  \\
 	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
@@
 \begin{align}
-	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
-	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ' ,
+	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
+	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ,
 \end{align}
 \begin{equation}
-	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \label{eq:appendix_noise}
-\end{equation}
-where $ \Delta t = t_{k+1} - t_k$ and 
+	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \ ,  \label{eq:appendix_noise}
+\end{equation}
 \begin{align}
-	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
-	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ .
+	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left(\Delta t + t_k \right) \nonumber \\ 
+	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ ,
 \end{align}
-From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
+and $\Delta t = t_{k+1} - t_k$. From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
 \begin{equation}
@@
 \end{equation}
-for 
-\begin{equation}
-Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right) \ .
+with 
+\begin{equation}
+Q^{(n)} = \frac{[\sigma^{n}]^2}{2 \gamma^{(n)}} \left[ 1 - e^{-2 \gamma^{(n)} \Delta t}\right] \ .
 \end{equation}
@@
 $\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
-The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
-
-\section{Fiducial pulsar values}\label{appendix_fiducial}
-\textcolor{red}{TK: AM suggests a table of the pulsars used with there fiducial values.
-Do we really need this?
-The argument is for reproducibility, but we have a nice github repo that holds all the data we actually used, jupyter notebooks that show how the calculations are done.
-Perhaps it is sufficient to point the reader to the repo...
-MNRAS encourages a "data availability" section before Acknowledgements that points readers to code/data/github repository}
-
-\section{Wasserstein distance}\label{sec:wasserstein}
-In this section we describe the Wasserstein distance used in Section \ref{sec:multiple_noise}.
-In Section \ref{sec:appendix_overview_WD} we give a brief overview of the Wasserstein distance generally.
-In Section \ref{sec:appendix_WD_results} we present for reproducibility the Wasserstein distance calculated between every pair of posteriors for each static parameter in $\boldsymbol{\theta}_{\rm gw}$ across the $10^3$ noise realisations of Section  \ref{sec:multiple_noise}.
-This corresponds to the mean values reported in Table \ref{tab:Wasserstein}.
-
-\subsection{Overview of the Wasserstein distance}\label{sec:appendix_overview_WD}
-
-The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
+The measurement covariance satisfies $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$ for all $k$.
+
+\section{Synthetic PTA specifications}\label{appendix_fiducial}
+The synthetic PTA deployed for testing in this paper is constructed to mimic the $N=47$ pulsars from the 12.5-year NANOGrav data release, whose sky positions are plotted in Figure \ref{fig:pulsar_distrib}.
+The construction recipe is described in full in Section \ref{sec:synt_pta}.
+In this appendix, we record a complete list of the pulsar parameters in Table \ref{tab:pulsar_values}, to assist the interested reader in implementing the analysis scheme and reproducing the test results in Sections \ref{sec:rep_example}--\ref{sec:bias_and_identifiability}.
+
+\begin{table*}
+\begin{tabular}{lccccc}
+	\toprule
+	JNAME &    $f_{\rm em} (t_1)$ (Hz) &  $\dot{f}^{(n)}_{\rm em}(t_1)\times 10^{16}$ (s$^{-2}$) &  $\alpha$ (deg) &  $\delta$ (deg) &  $d$ (kpc) \\
+	\midrule
+	J0023+0923 & 327.8 &      -12.3 &   5.8 &    9.4 &   1.8 \\
+	J0030+0451 & 205.5 &       -4.3 &   7.6 &    4.9 &   0.3 \\
+	J0340+4130 & 303.1 &       -6.5 &  55.1 &   41.5 &   1.6 \\
+	J0613-0200 & 326.6 &      -10.2 &  93.4 &   -2.0 &   0.9 \\
+	J0636+5128 & 348.6 &       -4.2 &  99.0 &   51.5 &   0.7 \\
+	J0645+5158 & 112.9 &       -0.6 & 101.5 &   52.0 &   1.2 \\
+	J0740+6620 & 346.5 &      -14.6 & 115.2 &   66.3 &   1.1 \\
+	J0931-1902 & 215.6 &       -1.7 & 142.8 &  -19.0 &   3.7 \\
+	J1012+5307 & 190.3 &       -6.2 & 153.1 &   53.1 &   0.7 \\
+	J1024-0719 & 193.7 &       -7.0 & 156.2 &   -7.3 &   1.2 \\
+	J1125+7819 & 238.0 &       -3.9 & 171.5 &   78.3 &   0.9 \\
+	J1453+1902 & 172.6 &       -3.5 & 223.4 &   19.0 &   1.3 \\
+	J1455-3330 & 125.2 &       -3.8 & 223.9 &  -33.5 &   0.7 \\
+	J1600-3053 & 277.9 &       -7.3 & 240.2 &  -30.9 &   1.9 \\
+	J1614-2230 & 317.4 &       -9.7 & 243.7 &  -22.5 &   0.7 \\
+	J1640+2224 & 316.1 &       -2.8 & 250.1 &   22.4 &   1.5 \\
+	J1643-1224 & 216.4 &       -8.6 & 250.9 &  -12.4 &   0.7 \\
+	J1713+0747 & 218.8 &       -4.1 & 258.5 &    7.8 &   1.3 \\
+	J1738+0333 & 170.9 &       -7.0 & 264.7 &    3.6 &   1.5 \\
+	J1741+1351 & 266.9 &      -21.5 & 265.4 &   13.9 &   1.7 \\
+	J1744-1134 & 245.4 &       -5.4 & 266.1 &  -11.6 &   0.4 \\
+	J1747-4036 & 607.7 &      -48.5 & 267.0 &  -40.6 &   7.1 \\
+	J1832-0836 & 367.8 &      -11.2 & 278.1 &   -8.6 &   2.1 \\
+	J1853+1303 & 244.4 &       -5.2 & 283.5 &   13.1 &   2.1 \\
+	J1857+0943 & 186.5 &       -6.2 & 284.4 &    9.7 &   1.2 \\
+	J1903+0327 & 465.1 &      -40.7 & 285.8 &    3.5 &   7.0 \\
+	J1909-3744 & 339.3 &      -16.1 & 287.4 &  -37.7 &   1.1 \\
+	J1910+1256 & 200.7 &       -3.9 & 287.5 &   12.9 &   1.5 \\
+	J1911+1347 & 216.2 &       -7.9 & 288.0 &   13.8 &   1.4 \\
+	J1918-0642 & 130.8 &       -4.4 & 289.7 &   -6.7 &   1.1 \\
+	J1923+2515 & 264.0 &       -6.7 & 290.8 &   25.3 &   1.2 \\
+	J1939+2134 & 641.9 &     -433.1 & 294.9 &   21.6 &   3.5 \\
+	J1944+0907 & 192.9 &       -6.4 & 296.0 &    9.1 &   1.2 \\
+	J1946+3417 & 315.4 &       -3.1 & 296.6 &   34.3 &   6.9 \\
+	J1955+2908 & 163.0 &       -7.9 & 298.9 &   29.1 &   6.3 \\
+	J2010-1323 & 191.5 &       -1.8 & 302.7 &  -13.4 &   2.4 \\
+	J2017+0603 & 345.3 &       -9.5 & 304.3 &    6.1 &   1.4 \\
+	J2033+1734 & 168.1 &       -3.1 & 308.4 &   17.6 &   1.7 \\
+	J2043+1711 & 420.2 &       -9.3 & 310.8 &   17.2 &   1.4 \\
+	J2145-0750 &  62.3 &       -1.2 & 326.5 &   -7.8 &   0.7 \\
+	J2214+3000 & 320.6 &      -15.1 & 333.7 &   30.0 &   0.6 \\
+	J2229+2643 & 335.8 &       -1.7 & 337.5 &   26.7 &   1.8 \\
+	J2234+0611 & 279.6 &       -9.4 & 338.6 &    6.2 &   1.0 \\
+	J2234+0944 & 275.7 &      -15.3 & 338.7 &    9.7 &   1.6 \\
+	J2302+4442 & 192.6 &       -5.1 & 345.7 &   44.7 &   0.9 \\
+	J2317+1439 & 290.3 &       -2.0 & 349.3 &   14.7 &   1.7 \\
+	J2322+2057 & 208.0 &       -4.2 & 350.6 &   21.0 &   1.0 \\
+	\bottomrule
+\end{tabular}
+\caption{List of the fiducial pulsar parameters for the 47 pulsars used to construct the synthetic PTA in Section \ref{sec:synt_pta}.
+The right ascension and declination of an individual pulsar in J2000 coordinates are labelled by $\alpha$ and $\delta$ respectively.
+Parameter values are obtained from the ATNF pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.}
+\label{tab:pulsar_values}
+\end{table*}
+
+\section{Dispersion of ${\theta}_{\rm gw}$ estimates}\label{sec:wasserstein}
+Every random realization of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$ leads to a different ${\boldsymbol{\theta}}_{\rm gw}$ posterior, when the synthetic data are analysed according to the procedure in Section \ref{sec:methodsummary}.
+This cosmic variance is unavoidable.
+It is a key source of uncertainty, when the analysis is performed on real astronomical data, where the specific random realizations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$ are unknown.
+To quantify the cosmic variance, one needs a metric to describe the distance between two posteriors (i.e.\ how similar they are).
+Many valid metrics of this sort exist, including those related to the Kolmogorov-Smirnoff test \citep{corder2014nonparametric}.
+In this paper, we use the WD \citep{Wasserstein,Villani2009} which is popular in machine learning \citep[e.g.][]{2017arXiv170107875A} and other domains.
+In Section \ref{sec:appendix_overview_WD} we define the WD and summarize its main properties.
+In Section \ref{sec:appendix_WD_results} we present for reproducibility the WD calculated between every pair of posteriors for each static parameter in $\boldsymbol{\theta}_{\rm gw}$ across the $10^3$ noise realisations of Section  \ref{sec:multiple_noise}.
+The results are summarised via the median values reported in Table \ref{tab:Wasserstein}.
+
+\subsection{Overview of the WD}\label{sec:appendix_overview_WD}
+The WD is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
 It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$.
-For this reason it is sometimes known as the "Earth mover's distance".
+For this reason it is sometimes known as the ``Earth mover's distance''.
 The $p$-th order Wasserstein distance between two distributions is
 \begin{eqnarray}
-	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \lambda (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
+	W_p(\mu,\nu)= \left[ \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \lambda (x,y)\right]^{1/p} \label{eq:wasserstein} \ ,
 \end{eqnarray}
-where $p \in [1,\infty)$, $\lambda(x,y)$ is the transport plan, and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
+for $p \in [1,\infty)$, where $\lambda(x,y)$ is the transport plan, and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
 $\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
-
-The cost function can be freely chosen for the nature of the problem, but is commonly taken to be the absolute value function, $c(x,y) = |x-y|$, and we adopt the same cost function in this work.
-Given this choice of $c(x,y)$, for an empirical distribution $\hat{\mu}_n$ composed of $n$ i.i.d samples drawn from the underlying probability distribution $\mu$, the solution to the discrete form of Equation \eqref{eq:wasserstein} can be computed generally by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
-However, for $\mu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
+\newline
+
+The cost function can be freely chosen to suit the nature of the problem.
+Often, as in this paper, it is taken to be the absolute value function, $c(x,y) = |x-y|$.
+In general, $W_p(\mu,\nu)$ with $c(x,y)=|x-y|$ can be computed from $n$ samples by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
+However, for $\mu$ and $\nu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that $E[W_p(\mu,\nu)]$ converges slowly $\propto n^{-1/d}$.
+In this paper, we calculate the WD exclusively between the one-dimensional marginalized posteriors, setting $d=1$ and obtaining
 \begin{eqnarray}
-	\boldsymbol{E} [W_1(\hat{\mu}_n,\mu)] \asymp n^{-\frac{1}{d}} \ .
+	W_p(\mu,\nu)= \left\{\int_0^1 dz \left[ F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) \right] \right\}^{1/p} \ , \label{eq:wasserstein}
 \end{eqnarray}
-In order to circumvent this issue, in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors, i.e.
-setting $d=1$.
-In this case Equation \eqref{eq:wasserstein} has a convenient, easily computed, closed form:
-\begin{eqnarray}
-	W_p(\mu,\nu)= \left(\int_0^1  F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) dz \right)^{1/p} \, \label{eq:wasserstein}
-\end{eqnarray}
-where $F_{\mu}(z)$ is the cumulative distribution function of $\mu$. \newline
-
-The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages.
-Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another.
-It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g.
-radians then we can interpret this as the cost to transform between the two distributions is 1 radian.
-Further interpretability is provided by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
+where $F_{\mu}^{-1}(z)$ is the inverse cumulative distribution function of $\mu$. \newline
+
+The WD holds certain advantages over valid alternatives like the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test \citep{gelman2013bayesian,corder2014nonparametric}.
+sIt is intuitive, being the minimum cost required to transform one distribution into another.
+It satisfies the Monge-Kantorovich duality \citep{villani2003topics,Villani2009},
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
-That is, if $W_1 =1$ then the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian.
-Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g.
-Gaussianity.
-Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces.
-Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality.
-It also respects the geometry of the underlying metric space over which the distributions are defined.
-\newline
-
-\subsection{Wasserstein distance for static parameters} \label{sec:appendix_WD_results}
-
-Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
-In Figure \ref{fig:pairwise_wasserstein} we present the WD values for a subset ($=10^2$) of those noise realisations.
-We plot a subset of realisations so as not to overcrowd the figure and enable certain features in the figure to be resolved.
-\newline
-
-The figure is composed of seven subplots, one for each element or ${\boldsymbol{\theta}}_{\rm gw}$.
-Each individual subplot is a lower triangular heat-map, where each point denotes the $W_1$ value between a pair of one-dimensional posteriors for that element.
-Smaller values of $W_1$ are magenta moving to through red to yellow for larger values of $W_1$.
-Note that the heat-map colour scaling is not the same for all subplots of the figure.
-\newline
-
-As discussed in Section \ref{sec:multiple_noise} $W_1$ is generally small across all pairs of posteriors.
-The median $W_1$ across all $10^3 \choose 2$ pairs of noise realisations is reported in Table \ref{tab:Wasserstein}.
-However for some static parameters there are a small fraction ($\lesssim 5 \%$) of posterior distributions which have a notably larger $W_1$ value from all the other posteriors of all the other noise realisations.
-These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
-These correspond exclusively to posteriors which have converged poorly and not found the likelihood global maximum.
-For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the figure, corresponds to a posterior for $\psi$ which has converged sub-optimally, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
-The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ rad.
-We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
-This is due to multimodal likelihood surface for $\psi-\alpha$, shown in Figure \ref{fig:likelihood_surface_alpha_psi}.
-We can see that the likelihood surface has a global optima at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local optima that corresponds to the location of the example poorly converged posterior discussed previously, with $\psi = 1.01$ rad.
-This problem of poor global convergence can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
+where $X_{\mu}$ and $Y_{\nu}$ are random variates drawn from the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
+It also satisfies desirable properties of a measure of distance, such as symmetry and the triangle inequality.
+For $p=1$, the WD inherits the units of the underlying distributions.
+Finally, the WD is versatile; any two distributions can be compared, irrespective of whether they are continuous, discrete, or singular.
+\newline
+
+\subsection{WD for $\theta_{\rm gw}$} \label{sec:appendix_WD_results}
+Table \ref{tab:Wasserstein} summarizes the median WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realisations analysed in Section \ref{sec:multiple_noise}, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+To assist with reproducibility, we present in Figure \ref{fig:pairwise_wasserstein} we present the WD values for a subset (numbering $10^2$) of the realisations.
+We plot a subset of realisations to avoid overcrowding the figure.
+Figure \ref{fig:pairwise_wasserstein} contains seven subplots, one for each element or ${\boldsymbol{\theta}}_{\rm gw}$.
+Each subplot is a lower triangular heat-map, where each point denotes the $W_1$ value between a pair of one-dimensional posteriors for that element.
+Lower values of $W_1$ are magenta; higher values of $W_1$ are yellow.
+Note that the heat-map colour scale is not the same in every subplot.
+\newline
+
+ Figure \ref{fig:pairwise_wasserstein} confirms that $W_1$ is generally small across all pairs of posteriors, consistent with Table \ref{tab:Wasserstein}.
+However for some static parameters a small fraction ($\lesssim 5 \%$) of posteriors differ significantly from all the other posteriors.
+These correspond to the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
+They are posteriors that converge poorly and miss the likelihood global maximum.
+For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk, corresponds to a posterior for $\psi$ which converges to $\psi = 1.01 \pm 0.06$ rad, far from the true injected value of $2.50$ rad.
+The $W_1$ value between this posterior and all the others is $\approx 1.7$ rad.
+We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations that corrupt the inference of $\psi$ also affect $\alpha$, due to the multimodal likelihood surface for $\psi-\alpha$, shown in Figure \ref{fig:likelihood_surface_alpha_psi}.
+The likelihood surface has a global maximum at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local maxima at the location of the example poorly converged posterior discussed previously, with $\psi = 1.01$ rad.
+Poor global convergence can be solved in principle by increasing $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima, raising the computational cost.
 
 \section{Bias from dropping the pulsar terms}\label{sec:bias_appendix}
-Why does dropping the pulsar terms bias the above parameters in particular, whilst leaving others nearly unaffected?
-For instance, why is the bias in $\iota$ so strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
-When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is effective modulated by the interference of two equal-amplitude, phase shifted waves (c.f.
+Why does dropping the pulsar terms in Equation \ref{eq:g_func_trig} bias some components of $\boldsymbol{\theta}_{\rm gw}$, whilst leaving others nearly unaffected?
+For instance, why is the bias in $\iota$ strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
+In this appendix, we examine the above questions in the context of the Kalman analysis in this paper, complementing work on similar questions in the context of traditional PTA analyses \citep{Zhupulsarterms,Chen2022}.
+\newline
+
+ When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is modulated by the interference of two equal-amplitude, phase-shifted oscillations (c.f.
 Equation \eqref{eq:measurement}).
-Explicitly we have an
-``Earth-terms wave'' with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$, and the ``pulsar-terms wave'' with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
-These two terms are related to the measurement equation, Equation \eqref{eq:g_func_trig} as,
-\begin{eqnarray}
-	1 - g^{(n)}(t) =  y_{\rm Earth}  - y_{\rm pulsar} \ .
-\end{eqnarray}
-The net interference wave that modulates $f_{\rm p}$ is 
-\begin{eqnarray}
-	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ .
-\end{eqnarray}
-The inclusion of  $y_{\rm pulsar}$ influences the amplitude of the resulting interference wave by a factor $ 2 \sin (\kappa/2)$, but does but do not affect its frequency.
-When we model this data using the Kalman filter without the pulsar terms, the additional factor from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
-As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude.
-Terms which influence the amplitude strongly, such as $h_0$ and $\iota$, are strongly affected and exhibit a strong bias.
+Specifically we have an
+``Earth-term oscillation'' of form $y_{\rm Earth} = A \cos \xi$, with amplitude $A$ and phase $\xi$, and a ``pulsar-terms oscillation'' of form $y_{\rm pulsar} = A \cos (\xi + \kappa)$, with phase offset $\kappa$.
+The two terms are related to the measurement equation, Equation \eqref{eq:g_func_trig}, through
+\begin{align}
+	1 - g^{(n)}(t) &=  y_{\rm Earth}  - y_{\rm pulsar} \  \nonumber \\
+		 &= 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ . \label{eq:psr_interference}
+\end{align}
+The interference in Equation \ref{eq:psr_interference} multiplies the amplitude by $ 2 \sin (\kappa/2)$, but leaves the frequency unchanged.
+When we model the data using the Kalman filter without the pulsar terms, $y_{\rm Earth}$ must try to compensate for the modified amplitude on the right-hand side of Equation \ref{eq:psr_interference} by adjusting the parameters which govern the amplitude, such as $h_0$ and $\iota$.
 Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
-The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
-Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
-The same hierarchy of biases is seen also in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
-\newline
-
-In order to ascertain how important the bias is for PTA continuous wave searches requires a thorough exploration of the astrophysical parameter space which we do not carry out in this work.
-The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
-Quantifying the bias in preparation for a practical PTA analysis with real data lies outside the scope of this introductory paper.
+The wave amplitude is independent of $\Omega$, and so $\Omega$ exhibits no bias.
+A shift in phase is covariant with the amplitude, so $\Phi_0$ also exhibits a small bias.
+The same hierarchy of biases is seen also in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are over-constrained whilst $\Omega$ is well-estimated.
+\newline
+
+Quantifying the bias for practical, astronomical PTA continuous wave searches requires a thorough exploration of the SMBHB parameter space which falls outside the scope of this introductory paper.
+The bias also depends on the PTA configuration.
 If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate.
-This is what we observe for our synthetic data (e.g.
-Figure \ref{fig:corner_plot_2}); the marginalised posterior broadens as the SNR decreases, until it overwhelms the shift due to the bias.
-Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may systematically affect the inferred parameters.
+This is what we observe for the synthetic data, for example in Figure \ref{fig:corner_plot_2}; the marginalised posterior broadens as the SNR decreases, until it overwhelms the shift due to the bias.
+Conversely if the bias is sufficiently strong, or the source is sufficiently loud, then the bias may dominate.
 \newline
@@
 	\end{subfigure}
-	\caption{The first moment of the Wasserstein distance, $W_1$, calculated between each pair of one-dimensional posteriors for our representative system across $10^3$ realisations, for each parameter of $\boldsymbol{\theta}_{\rm gw}$  $W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
-$W_1$ is generally small across all parameters and all posteriors, but some posteriors converge poorly leading to large values of $W_1$.
-The $W_1$ values inferred for $\psi$ and $\alpha$ are strongly correlated with correlation coefficient = 0.96.
-An example noise realisation with particularly large pairwise $W_1$ values is labelled with an asterisk in (c) and is discussed in the text.} \label{fig:pairwise_wasserstein}
+	\caption{First moment of the WD, $W_1$ (colour scale), calculated between each pair of one-dimensional posteriors for the representative system in Table \ref{tab:parameters_and_priors} across $10^3$ realisations as discussed in Section \ref{sec:appendix_WD_results}, for each component of $\boldsymbol{\theta}_{\rm gw}$.
+$W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
+$W_1$ is generally low across all parameters and posteriors, but some posteriors converge poorly and return high values of $W_1$.
+The $W_1$ values inferred for $\psi$ and $\alpha$ are correlated with Pearson correlation coefficient = 0.96.
+An example of a noise realisation with particularly high $W_1$ is labelled with an asterisk in (c) and is discussed in the text.} \label{fig:pairwise_wasserstein}
 \end{figure*}
@@
 	\includegraphics[width=\columnwidth]{images/likelihood_surface_alpha_psi}
-	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\alpha-\psi$ parameter space for a single noise realisation.
-The red point labels the location of the maxima, and the magenta point labels the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data.
-The likelihood has been normalised with respect to the absolute value of the maxima.
-A multi-modal structure is clearly visible and for this noise realisation the nested sampler has settled in the local optima.
-This can be solved generally by increasing the number of live points $n_{\rm live}$.}
+	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\alpha$-$\psi$ plane for a single noise realisation.
+The red point labels the maximum, and the magenta point labels the median values of the inferred one-dimensional posteriors for $\alpha$ and $\psi$.
+The likelihood is normalised with respect to the absolute value of the maximum; note the compressed colour scale.
+A multi-modal structure is clearly visible for this noise realisation.
+The nested sampler finds a local maximum (magenta point) instead of the global maximum (red point)}
 	\label{fig:likelihood_surface_alpha_psi}

### 2023-11-06 docs/manuscript/paper.tex
version submitted to mnras

@@
  f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
-\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA.
+\end{equation} where an overdot denotes a derivative with respect to $t$, and $t_1$ labels the time of the first TOA.
 The stochastic torque is assumed to be a zero-mean, white noise process.
@@
 
-Implementation of the linear Kalman filter for the PTA state-space model in Section 3.2, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
+Implementation of the linear Kalman filter for the PTA state-space model in Section \ref{sec:model}, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
 At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.
@@
 
-Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} dB(t) | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta B_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
+Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} dB(t) | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $dB(t)$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
 To fix the problem, we subtract the deterministic frequency evolution and track the new variable
@@
 	\centering
-	\includegraphics[width=\columnwidth]{images/pp_plot_new}
+	\includegraphics[width=\columnwidth]{images/pp_plot_new1}
 	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
@@
 The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$.
-One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section 7.2; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high.
+One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section \ref{sec:pulsar_terms_drop}; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high.
 One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$.

### 2023-11-13 manuscript/manuscript_2_old/paper.tex
movign some bits around

--- 2023-10-12 manuscript/manuscript_2/paper.tex
@@
 
+\begin{table}
+	\centering
+	%	\resizebox{\textwidth}{!}{%
+		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+		\begin{tabular}{lcclc}
+			\toprule
+			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm pulsar}(\theta) -\Delta_{\rm earth}(\theta) $   \\
+			\hline
+			\multirow{7}{2mm}{High SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
+			& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
+			& $\psi$ &0.0386& 0.000254& -0.0384 \\
+			& $\iota$ & 0.374 &0.126 &-0.248 \\
+           & $\delta$ & 0.00223 &0.000242& -0.00198 \\
+            &$\alpha$ &0.0397 &0.000582 &-0.0391 \\
+            &$h_0$ & 0.168 &0.0846 &-0.0839 \\
+			\hline
+			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
+& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
+& $\psi$ &0.0386& 0.000254& -0.0384 \\
+& $\iota$ & 0.374 &0.126 &-0.248 \\
+& $\delta$ & 0.00223 &0.000242& -0.00198 \\
+&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
+&$h_0$ & 0.168 &0.0846 &-0.0839 \\
+
+			\bottomrule
+		\end{tabular}
+		\caption{dfdfddd}
+		\label{tab:posterior_errors}
+	\end{table}
+
+%\begin{table}
+%	\centering
+%	%	\resizebox{\textwidth}{!}{%
+%		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+%		\begin{tabular}{cccc}
+%			\toprule
+%			$\theta$ & \multicolumn{2}{c}{RMSE$(\theta)$} & $\Delta$ \\
+%			& Earth & Pulsar & \\
+%			\hline
+%			$\Omega$ & 1 & 2 & 3 \\ 
+%			
+%			\bottomrule
+%		\end{tabular}
+%		\caption{}
+%		\label{tab:posterior_errors}
+%	\end{table}
+
+%
+%\begin{table}
+%	\centering
+%	%	\resizebox{\textwidth}{!}{%
+%		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+%		\begin{tabular}{lccll}
+%			\toprule
+%			$\theta$& \multirow{2}{*}{RMSE$(\theta)$} & Injected Values & Units & Prior  \\
+%			& 
+%			\hline
+%			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
+%			& $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
+%			& $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
+%			& $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
+%
+%
+%			\bottomrule
+%		\end{tabular}
+%		\caption{}
+%		\label{tab:posterior_errors}
+%	\end{table}
+
 Initially we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$ for the representative example system described in Table \ref{tab:parameters_and_priors}.
 We consider a "high-SNR" system with $h_0 = 10^{-12}$ and a ``low-SNR" system with $h_0 = 5 \times 10^{-15}$.
-We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ for both the high/low SNR systems.
 Priors on $\boldsymbol{\theta}$ are selected as follows.
@@
 For $\pi(\boldsymbol{\theta}_{\rm psr})$ we use the same priors discussed in K23.
-The addition here is the prior on $\chi^{(n)}$ which we take to be $\text{Uniform}(0,2 \pi)$.
-	All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}
-
+Additionally, we set a prior on $\chi^{(n)}$ as $\text{Uniform}(0,2 \pi)$.
+All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
+\newline
+
+We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ for both the high/low SNR systems.
 The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_1} for the high-SNR system and in Figure \ref{fig:corner_plot_2} for the low-SNR system.
-
 The histograms shown in the figures are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the solid orange line indicates the true injection value.
@@
 The green curves are the results derived using the Earth-terms measurement equation, Equation \ref{eq:g_func_trig_earth}.
-The blue curves are the results derived using the full measurement equation, including the pulsar terms, Equation \eqref{eq:g_func_trig}
-
-For the high-SNR case, Figure \ref{fig:corner_plot_1} the one-dimensional posteriors for the Earth-terms solution are evidently biased, as was observed in K23.
-Particular biases are observed in $\Phi_0, \psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
-
-The inclusion of the pulsar terms corrects for this bias. The green posteriors
-
-Moreover, the injection values are contained witinin the 2-sigma equivalent contours for every pair of parameters. This is not true in the Earth-terms case.
-
-For the low-SNR case
-
-les important for PE in low SNR
-
-but is important generally for detectio nas we will show.
-
-Takeaways
-
-We are able to estimate hte static parameters
-
-the inclusion of the pulsar terms general improves the accuracy
-
-the chi terms are also recoverable
-
-It is evident that for this representative example with characteristic strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
-\newline
-
-For some of the parameters the estimated posterior is approximately symmetric about the true injected value (e.g.
-$\Omega$).
-In this case the posterior median and the injected value are approximately coincident.
-For other parameters (e.g.
-$\iota$) the distribution is not symmetric about the injected value and the posterior median and the injected value are not coincident.
-The median value of the posterior for $\iota$ is shifted by $\sim 0.35$ rad relative to the injected value, although the injected value does still remain inside the 90\% credible interval.
-Similar effects are seen, albeit with a smaller shift, in other parameters such as $\delta$ and $\alpha$.
-For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
-a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation.
-We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
-Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
-We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.
+The blue curves are the results derived using the full measurement equation, including the pulsar terms, Equation \eqref{eq:g_func_trig}.
+\newline
+
+For the high-SNR case, Figure \ref{fig:corner_plot_1}, the one-dimensional posteriors for the Earth-terms solution are evidently biased, as was observed in K23.
+Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
+For example, for $\iota$, whilst the injected value is contained within the 90 \% credible interval, the median of the posterior of the Earth-terms solution is shifted from the injected value by 0.35 radians.
+The inclusion of the pulsar terms corrects for this bias.
+The green one-dimensional posteriors exhibit no bias and a generally symmetric about the injected value.
+Moreover, for every pair of parameter the injection values are contained within the 2-sigma equivalent contours.
+This is not true in the Earth-terms case.
+More quantitatively, we define a relative error metric to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
+\begin{equation}
+	\Delta_{\rm M}(\Theta) = \frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}} 
+\end{equation}
+where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ the total number of samples, $\hat{\theta}$ the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimates using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+$\Delta_{\rm M}(\theta)$ then represents the average relative error in the one dimensional posterior.
+The values for each parameter in Figure \ref{fig:corner_plot_1} are summarised it Table \ref{tab:posterior_errors}, along with the relative difference between the two models, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$.
+For every parameter except $\Omega$, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is negative, indicating that the parameter estimates are more accurate using the pulsar terms.
+$\Omega$ is recovered with high accuracy in both cases, and the relative error is small, $\Delta_{\rm M}(\Omega) \sim 10^{-5}$.
+The improvement in $\iota$ is largest, with $\Delta_{\rm pulsar}(\iota) - \Delta_{\rm earth}(\iota) = 0.25$.
+\newline
+
+For the low-SNR case,  Figure \ref{fig:corner_plot_2}, there is less of a notable improvement in the parameter estimates.
+Qualitatively, the green and blue one-dimensional posteriors are generally overlaid.
+The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest.
+The relative errors 	$\Delta_{\rm M}(\Theta)$ for the low SNR case are also reported in Table \ref{tab:posterior_errors}.
+We see that....
+\textcolor{red}{TK: numbers to be inserted here}.
+We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case.
+This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates.
+Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:detection_new} that they are important for increasing the detection probability for a given signal strength.
+\newline
+
+In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
+In both the high and low SNR cases $\chi^{(n)}$ is unambiguously recovered.
+This is shown in Figure \ref{fig:corner_plot_3} where we plot the results for five of the $\chi^{(n)}$ values for both high SNR (left-hand panel) and low SNR (right hand panel).
+The estimates are naturally more constrained in the high SNR case, but in both cases the injection is contained within the 90 \% credible interval.
+Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
+
+In summary, the main results of this section are:
+\begin{itemize}
+	\item We are able to estimate all the static parameters of the system using the full pulsar-terms model, for this representative system with characteristic strain $h_0 = 5 \times 10^{-15}$.
+	\item The inclusion of pulsar terms improves the parameter estimation accuracy, especially when the GW signal is loud with respect to the noise.	
+\end{itemize}
 
 \subsection{Detection}\label{sec:detection_new}
+
+\begin{figure}
+	\includegraphics[width=\columnwidth, height = \columnwidth ]{images/CanonicalBayesPlot2000} 	
+	\caption{hellow}
+	\label{fig:bayes1}
+\end{figure}
+
+The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
+\begin{equation}
+	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
+\end{equation}
+Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
+$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
+This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
+The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
+
+We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
+$\mathcal{M}_1$ is the Earth-terms only model, i.e.
+the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
+The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
+We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
+We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
+For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
+This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
+$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
+\newline
+
+There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
+The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
+The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
+For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
+That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
+For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
+For the results presented in this subsection we set $n_{\rm live} =2000$.
+
+\section{Exploring a broader parameter space}
+
+In this section we focus on pulsar terms exclusively and no longer compare with the Earth terms
 
@@
 \subsection{Model selection}\label{sec:model_selection}
-The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
-\begin{equation}
-	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
-\end{equation}
-Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
-$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
-This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
-The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
 
@@
 \subsection{Detection} \label{sec:detection}
-We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-$\mathcal{M}_1$ is the Earth-terms only model, i.e.
-the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
-The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
-The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
-We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
-We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
-For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
-This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
-$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
-\newline
-
-There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
-The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
-The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
-For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
-That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
-For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
-For the results presented in this subsection we set $n_{\rm live} =2000$.
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
-	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
+
+%\begin{figure}
+%	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
+%	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
 The horizontal grey dashed line labels the detection threshold of $\beta = 10$.
@@
 Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$.}
-	\label{fig:bayes}
-\end{figure}
+%	\label{fig:bayes}
+%\end{figure}
 

### 2023-11-13 manuscript/manuscript_2/paper.tex
assorted manuscript updates

@@
 % Basic setup. Most papers should leave these options alone.
+%\documentclass[fleqn,usenatbib,useAMS,onecolumn]{mnras}
 \documentclass[fleqn,usenatbib,useAMS]{mnras}
@@
 % Keep the title short and informative.
-	\title[Kalman PTA]{Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms}
+	\title[Kalman PTA]{Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array: inclusion of pulsar terms}
 
@@
 \begin{abstract}
-	The detection of continuous nanohertz gravitational waves from individual supermassive black hole binaries can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-Such an approach complements traditional analysis methods by enabling self-consistent inference of model parameters and calculation of the Bayesian evidence, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise.
-Previous analyses have worked exclusively with the `Earth term', leading to biases in the parameter estimation.
-Here it is shown how to extend previous analyses to include the so-called `pulsar terms'.
-It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy relative to the Earth terms approach by X and the sensitivity by Y.
+
+	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.
+
+	 The gravitational wave signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
+
+	 	 The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
+
+We show how to extend previous state-space analyses to include the pulsar terms.
+
+It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy relative to the Earth terms approach by X and the sensitivity by Y
+
 \end{abstract}
@@
 \section{Introduction}\label{sec:intro}
-Nanohertz gravitational waves (GWs) can be produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}.
-A nHz GW influences the times of arrival (TOAs) of radio pulses from pulsars measured at the Earth.
-A pulsar timing array \citep[PTA][]{Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars.
+Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}.
+A nHz GW influences the time of arrival (TOA) at the Earth of radio pulses from pulsars.
+A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars.
 In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers.
-\newline
-
- PTAs search for two types of SMBHB GW signals.
-The first is a stochastic GW background due to the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
-Consilient evidence for the GW background has been presented by multiple separate PTAs \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X}, by measuring the cross-correlation between pairs of pulsars in the pulsar timing residuals as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}.
-The second is resolved GWs from sufficiently close and massive individual SMBHBs \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}.
-Individual SMBHBs are sources of continuous GWs.
-This results in persistent, quasi-monochromatic modulations in the pulsar timing residuals which can be modelled.
-Therefore, rather than cross-correlating pulsar pairs, individual SMBHBs are detected by either frequentist matched filtering \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or Bayesian inference \citep{Ellis2016,Arzoumanian2020A}.
-Observational efforts to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, although some inconclusive evidence at low significance has been presented \citep{2023arXiv230616226A}.
-\newline
-
- Irregularities in the intrinsic rotation of the pulsar result in random, unmodelled, red-spectrum TOA fluctuations known as `timing noise'.
-Timing noise limits the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}.
-Additional white noise sources such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021} must also be considered.
-\cite{K1} (K23 hereafter) presented an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations.
-Specifically, K23 formulate PTA analysis as a state-space problem and use a Kalman filter \citep{Kalman1} to track the pulsars' intrinsic rotational state in conjunction with a Bayesian nested sampler \citep{Skilling, Ashton2022} for parameter estimation and model selection.
-\textbf{VALIDATION TESTS ON SYNTHETIC DATA...
-}newline
-
- It is common in standard PTA analysis to drop the so called ``pulsar term" from the modelling \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-Unlike the "Earth term", the pulsar term is not phase coherent and cannot be summed across the array to increase the total signal-to-noise.
-Instead it is considered as a source of self noise.
-However, dropping the pulsar terms results in a reduced detection probability ($\sim 5 \%$) bias in the parameter estimates \citep{Zhupulsarterms,Chen2022}.
-The method presented in K23 also drops the pulsar term.
-In this paper we remove this limitation and present the state-space approach to PTA data analysis including the pulsar term.
-\newline
-
- This paper is organised as follows.
-In Section \ref{sec:model} we briefly review the state-space formulation of PTA data analysis introduced in \cite{K1}.
-We do not review the methods for state tracking, parameter inference and model selection and refer the reader to K23 for details.
-In Section \ref{sec:pulsar_term} we introduce how to include the pulsar term in the state-space model.
-In Section \ref{sec:rep_example1} we test the updated method on the synthetic data for a single representative example GW system, and compare the results to the original Earth-terms model.
-In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space.
-Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.
-Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$.
-\newline
-
-% Could include the below block if we need more fodder
-
-%This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
+Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
+SMBHBs which are sufficiently close and massive may be individually resolvable, though conclusive evidence is currently lacking \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
+\newline
+
+The modulation in the pulsar TOAs due to the presence of a GW has two contributions: an ``Earth term" and a ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
+Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases.
+Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+However, dropping the pulsar terms leads to biases in the inferred parameters and reduced detection probabilities \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
+\newline
+
+\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations.
+The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term.
+In this paper we undertake two key tasks.
+Firstly we extend the methods presented in K24 to include the pulsar term.
+We show how to reparametrise the model in terms of an additional pulsars-specific phase term.
+We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}.
+Secondly we quantify how the inclusion of the pulsar term affects the accuracy with which the static GW parameters can be estimated, and the minimal detectable GW strain.
+
+This paper is organised as follows. In Section \ref{sec:2} we briefly review the state-space formulation of PTA data analysis introduced in \cite{KimpsonPTA}.
+
+Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline
+
+%In Section 3 we introduce how to include the pulsar term in the state-space model.
+In Section 4 we test the updated method on the synthetic data for a single representative example GW system, and compare the results to the original Earth-terms model.
+In Section 9 we extend the tests to cover an astrophysically relevant parameter space.
+Discussion on future extensions and conclusions are made in Sections 11 and 12 respectively.
+Throughout the paper we adopt the natural units, with 𝑐 = 𝐺 = ħ = 1, and metric signature (−, +, +, +).
+
+%
+%
+%Intrinsic pulsar timing noise -- i.e.
+random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}.
+This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Myers2021MNRAS.502.3113M,Meyers2021,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
 In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators.
 However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
-
-\section{State-Space Formulation}\label{sec:model}
-In this section we briefly review the state-space formulation of PTA analysis presented in K23. \newline
+In modern Bayesian PTA searches, the power spectral density of the red intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the noise induced by a stochastic GW background (whose spectrum is also red).
+In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+\newline
+
+%In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations.
+The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density.
+Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations summarized by the power spectral density, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
+The new and existing approaches are therefore complementary.
+In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
+We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the static GW parameters and calculate the marginal likelihood (i.e.
+the model evidence) for model selection.
+\newline
+
+%The paper is organised as follows.
+In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW.
+In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence.
+In Section \ref{sec:testing} we describe how we create synthetic validation data with which to test the method.
+We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source.
+In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant domain of SMBHB source parameters.
+In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates.
+Conclusions are drawn in Section \ref{sec:discussion}.
+
+\section{State-Space Formulation}\label{sec:2}
+In this section we briefly review the state-space formulation of PTA analysis presented in K24. \newline
 
@@
 The intrinsic spin frequency of the $n$-th pulsar as measured in the local, freely-falling rest frame of the pulsar's centre of mass, $f_{\rm p}^{(n)}(t)$, evolves according to a stochastic differential equation.
-The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation.
-In Section \ref{sec:psr_frequency} we review the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$.
-In Section \ref{sec:psr_measured} we review the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$.
+The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW.
+In Section \ref{sec:psr_frequency} we outline the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$.
+In Section \ref{sec:psr_measured} we outline the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$.
 In Section \ref{sec:ss_params} we summarise the parameters of the model.
+The model is formulated in terms of pulse frequency time series as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
+It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data.
+We defer this subtle generalization to future work.
 
 \subsection{Spin evolution} \label{sec:psr_frequency}
-In the model $f_{\rm p}^{(n)}(t)$ evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
-\begin{equation}
-	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ ,
+The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
+\begin{equation}
+	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
 	\label{eq:frequency_evolution}
@@
 
-Equation \ref{eq:frequency_evolution} is inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
-Such a phenomenological model is necessary since a predictive, first-principles theory of timing noise does not currently exist c.f.
-the multiple theorized mechanisms of timing noise, see the discussion in Section 1 of K23.
-Similar approaches, analogous to Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
-However, Equation \eqref{eq:frequency_evolution} does contain certain modelling omissions and idealisations which much be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
-These include: exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}, the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs and the formal interpretation of $d^2 f_{\rm p} / dt^2$.
-
-%
-%However, as discussed in K23, First, Equation \ref{eq:frequency_evolution} excludes particular phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}.
-Second, the white spectrum assumed for $\xi(t)$ may or may not be suitable for millisecond pulsars in PTAs.
-It is challenging observationally to infer the spectrum of $\xi(t)$ from the observed spectrum of the phase residuals, because the inference is conditional on the (unknown) dynamical model governing $df_{\rm p}/dt$.
-For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless.
-White noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments.
-Third, $\xi(t)$ is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper.
-Finally, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, which have not been ruled out by pulsar timing experiments to date.
-The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \eqref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \eqref{eq:frequency_evolution}.
-
-%This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
-In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators.
-However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
-%
-%
-
-%
-%Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are not derived from first principles by applying a microphysical theory.
-As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}.
-%
-%
-%
-%However, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}  involve significant idealizations, which must be recognized at the outset \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
-First, the white noise driver $\xi(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper.
-Second, the white spectrum assumed for $\xi(t)$ may or may not be suitable for millisecond pulsars in PTAs.
-It is challenging observationally to infer the spectrum of $\xi(t)$ from the observed spectrum of the phase residuals, because the inference is conditional on the (unknown) dynamical model governing $df_{\rm p}/dt$.
-For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless.
-Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, which have not been ruled out by pulsar timing experiments to date.
-The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \eqref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \eqref{eq:frequency_evolution}.
+Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} are inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
+Such a phenomenological model is necessary as a predictive, first-principles theory of timing noise does not currently exist c.f.
+the multiple theorized mechanisms of timing noise (see discussion in Section 1 of \cite{KimpsonPTA}).
+Similar approaches, analogous to Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
+However, Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} do contain certain modelling omissions and idealisations which much be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
+These include the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and the formal interpretation of $d^2 f_{\rm p} / dt^2$ as $\xi(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.
 
 \subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
-In the presence of a GW the intrinsic pulsar spin frequency for the  $n$-th pulsar as is related to the radio pulse frequency measured by an observer on Earth as
-\begin{equation}
-	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d^{(n)}) g^{(n)}(t) +  \varepsilon^{(n)}\ .
+In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a a non-linear measurement equation:
+\begin{equation}
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
 	\label{eq:measurement}
 \end{equation}
-where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies: 
+where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
 \begin{align}
 	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'}(t - t') \ .	\label{eq:vareps}
+	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
 \end{align}
-where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
+where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
 \begin{align}
@@
 \end{align}
-where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\boldsymbol{n}$ the propagation direction of the GW, $\Omega$ is the constant angular frequency of the GW, $H_{ij}$ is the spatial part of the GW amplitude tensor and $\Phi_0$ the phase offset of the GW.
-\newline
-
-Equation \eqref{eq:g_func_trig} contains a number of simplifying assumptions:
+where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor and $\Phi_0$ the phase offset of the GW with respect to some reference time.
+\newline
+
+Equation \eqref{eq:g_func_trig} contains two simplifying assumptions which must be recognised:  
 \begin{enumerate}
-	\item The GW is monochromatic plane wave.
+	\item The GW is a monochromatic plane wave.
 	\item The pulsar locations are constant with respect to the observer 
 \end{enumerate}
-Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Zhu10},
-\begin{equation}
-	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
+Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Sesana2010},
+\begin{equation}
+	\Delta f_{\rm gw} = 0.05 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^{8.5} M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{50 \mathrm{~nHz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
 	\label{eq:f_evolution}
 \end{equation}
-where $M_c$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
+where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
 A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
-From Equation \eqref{eq:f_evolution} we can see that only those binaries which are massive or at high frequency experience significant frequency evolution over typical PTA timespans.
+Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \gtrsim 20$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$.
 The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
 We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
+\newline
+
 Regarding point (ii) whilst in practice the pulsar locations vary with respect to the Earth, pulsar TOAs are defined relative to the Solar System barycentre (SSB).The barycentering correction is typically applied when generating TOAs, e.g.
 with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
-Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.
-
-%The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
-%\subsubsection{Plane GW perturbation}\label{sec:plane_gw}
-%We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
-%\begin{equation}
-%	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
-%\end{equation}
-%with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$.
-The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$.
-Throughout this paper we work with
-%
-%
-% We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic.
-
-%The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$). The spatial part is
-%\begin{align}
-%	H_{ij} = h_+ e_{ij}^+(\boldsymbol{n}) + h_{\times} e_{ij}^{\times}(\boldsymbol{n}) \ , \label{eq:hij}
-%\end{align}
-%where $h_{+}$ and $h_{\times}$ are the respective polarisation amplitudes.
-The plus and cross polarisation tensors $e_{ij}^{+}$ and $e_{ij}^{\times}$ are uniquely defined by the principal axes of the wave, viz.\ the unit 3-vectors $\boldsymbol{k}$ and $\boldsymbol{l}$, according to
-%\begin{align}
-%	e_{i j}^{+}(\boldsymbol{n}) =k_i k_j-l_i l_j \ , \\
-%		e_{i j}^{\times}(\boldsymbol{n}) =k_i l_j+l_i k_j \ .
-%\end{align}
-%The principal axes are in turn specified by the location of the GW source on the sky (colatitude $\theta$, azimuth $\phi$) and the polarisation angle $\psi$ according to
-%\begin{align}
-%	\boldsymbol{k}  = &(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
-%	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}} \nonumber \\
-%	& +(\sin \psi \sin \theta) \boldsymbol{\hat{z}} \ , \\
-%	\boldsymbol{l} = &(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
-%	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}}\nonumber  \\
-%	& +(\cos \psi \sin \theta) \boldsymbol{\hat{z}} \ ,
-%\end{align}
-%where e.g. $\boldsymbol{\hat{x}}$ is a unit vector in the direction of the $x$-axis. The direction of GW propagation is related to the principal axes by
-%\begin{equation}
-%	\boldsymbol{n} = \boldsymbol{k} \times \boldsymbol{l} \ . 
-%\end{equation}
-
-%\subsubsection{Measurement equation}
-%In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave.
-They are described by the geometric object $\boldsymbol{p}$ which we identify sas the 4-momentum of the radio pulse.
-The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e.
-$\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $.
-One obtains \citep[e.g.][]{Maggiore}
-%\begin{equation}
-% \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
-%	\label{eq:momentum_shift}
-%\end{equation}
-%where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer.
-We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar.
-We take the pulsar location to be constant i.e.
-neither $\boldsymbol{q}$ nor $d$ are functions of time.
-%
-%
-%
-%\begin{equation}
-%	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
-%\end{equation}
-%
-%Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.
-
-\subsection{Parameters of the model}\label{sec:ss_params}
-The set of static parameters which determine the intrinsic frequency evolution of the pulsars in the array outlined in Section \ref{sec:psr_frequency} are:
+Some pulsars, including some PTA pulsars, do have non-negligible proper motions of order $10^2$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.
+
+\subsection{Static parameters of the model}\label{sec:ss_params}
+For the model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured}, the set of static parameters which are specific to the pulsars in the array are
 \begin{equation}
@@
 \end{equation}
-The parameters which describe the GW source are:
-\begin{equation}
-	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \psi, \delta, \alpha, , \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
-\end{equation}
-where $h_0$ is the characteristic wave strain, $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
-$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$, $\psi$ the polarization angle of the GW and $\delta,\alpha$ the declination and right ascension of the GW source respectively.
-The complete set of static parameters is denoted	$\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw}$.
-There are $7 + 5N$ parameters to estimate, although the pulsar parameters are constrained better \textit{a priori} by electromagnetic observations than the GW parameters.
-For example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information about $\delta$ and $\alpha$.
-\newline
-
-%
-%The model  and Section \ref{sec:psr_measured} depends on a set of static parameters $\boldsymbol{\theta}$.
-%
-%
-%
-%
-%
-%The model outlined in Section \ref{sec:psr_frequency} and Section \ref{sec:psr_measured} depends on a set of static parameters $\boldsymbol{\theta}$.
-%
-%
-%
-%The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters controlling t and the GW source, i.e. 
-%\begin{equation}
-% \ , \label{eq:params1}
-%\end{equation}
-%with
-%
-%and
-%
-%where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively.
-In Equation \eqref{eq:params3} we reparameterize the two GW polarisation amplitudes, $h_{+}$ and $h_{\times}$, in terms of $\iota$ and the characteristic wave strain $h_0$ through
-%\begin{align}
-%	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
-%	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ .
-%\end{align}
-%We use the parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
+The parameters which are specific to the GW source are 
+\begin{equation}
+	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \psi, \delta, \alpha, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
+\end{equation}
+where $h_0$ is the characteristic wave strain, $\iota$ the orbital inclination, $\psi$ the polarisation angle of the wave, $\delta$ the declination of the source and $\alpha$ its right ascension.
+These parameters enter the model through Equation \eqref{eq:g_func_trig} where $H_{ij} = H_{ij}(h_0, \iota, \psi, \delta, \alpha)$ and $\boldsymbol{n}=\boldsymbol{n}(\delta,\alpha)$.
+The complete set of $7 + 5N$ static parameters is denoted $\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw}$.
+While we have no prior information about the values of the members of  	$\boldsymbol{\theta}_{\rm gw}$, the members of $\boldsymbol{\theta}_{\rm psr}$
+do have constraints from electromagnetic observations; for example
+for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
 
 \section{Inclusion of the pulsar term}\label{sec:pulsar_term}
-In this section we review the modification to Equation \eqref{eq:g_func_trig} used in K23, the issues that arise when using the full form of Equation \eqref{eq:g_func_trig}  for likelihood-based inference, and a new parametrisation that alleviates these issues.
-\newline
-
-\subsection{Earth term}
-Equation \eqref{eq:g_func_trig} is the sum of two equal amplitude, phase shifted waves.
-The first wave, $\cos(-\Omega t + \Phi_0)$, depends only on the GW source parameters and is shared across all pulsars.
-The argument of the cosine corresponds to the GW phase at the observer on Earth.
-Conversely the second wave, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right)$, also depends on $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
-The argument of the cosine corresponds to the GW phase at each the individual pulsar.
-The first and second terms are commonly referred to as the ``Earth term" and ``pulsar term" respectively.
-Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases and in standard PTA analysis are typically considered as a source of self noise and dropped from the analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
-K23 also followed this example and dropped the pulsar terms from the inference model.
-Explicitly the measurement equation used is
+K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection.
+That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
+\begin{equation}
+	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
+\end{equation}
+where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
+\begin{equation}
+	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
+\end{equation}
+and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}.
+Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
+We refer the reader to Appendix X for a review of Kalman filtering and nested sampling in the context of PTA searches for GWs.
+\newline
+
+The inference model used by the Kalman filter in K24 drops the pulsar term, consistent with other PTA analyses.
+In this section we show how to remove this limitation and include the pulsar term in the state-space inference model passed to the Kalman filter.
+In Section \ref{sec:earth_term} we review how the pulsar term is dropped from typical PTA analyses and the modification to Equation \eqref{eq:g_func_trig} used in K24.
+In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar term for likelihood-based inference methods like nested sampling.
+This new parametrisation is validated with synthetic data in Section \ref{sec:testing}.
+
+\subsection{Earth term}\label{sec:earth_term}
+The influence of the GW on the pulsar radio pulses is completely described by Equation \eqref{eq:g_func_trig}.
+The GW signal is composed of an ``Earth term", corresponding to the $\cos(-\Omega t + \Phi_0)$ component, and a ``pulsar term", corresponding to the $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \}$ component.
+The Earth term describes the GW phase at the observer on Earth, while the pulsar term describes the GW phase at each the individual pulsar.
+The Earth term depends only on the GW source parameters and so is shared across all pulsars.
+In contrast the pulsar term is a function of $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
+\newline
+
+Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases which cannot be coherently summed and are instead generally considered as a source of self noise.
+The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties much greater than the typical target GW wavelength.
+It is therefore not possible to set the phase of the pulsar term.
+Consequently the pulsar terms is typically dropped from in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space, making the parameter estimation computationally cheaper.
+Within the state-space model described in Section \ref{sec:2}, dropping the pulsar terms equates to using a modified measurement equation
 \begin{equation}
@@
 \end{equation}
-
-\subsection{Challenges of pulsar term}\label{sec:pulsar_terms_challenge}
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
-	\caption{Logarithm of the one-dimensional likelihood $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
-All other static parameters of $\boldsymbol{\theta}$ are held constant at their true injected values.
-In the left-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}(\alpha)$, i.e.
-inclusive of the pulsar terms.
-In the right-hand panel the Kalman filter uses Equation \ref{eq:g_func_trig_earth}, i.e.
-just the Earth terms.
-The vertical dashed lines indicate the true injected value of $\alpha$ ($=1.0$ rad) used to generate $\boldsymbol{Y}$.
-The inset plot covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
-In the left hand panel $\mathcal{L}(\alpha)$ is very noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
-This noise hampers the ability of sampling algorithms to converge to a posterior distribution.
-Conversely, in the right hand panel $\mathcal{L}(\alpha)$ is globally smooth.
-}
-	\label{fig:likelihood_comparison}
-\end{figure*}
-
-K23 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection.
-That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by
-Bayes' Rule,
-\begin{equation}
-	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
-\end{equation}
-where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
-\begin{equation}
-	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
-\end{equation}
-and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}.
-Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
-Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e.
-$\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$.
-We will use the notation $\mathcal{L}(\Theta)$ to refer to the likelihood returned by the Kalman filter for parameter $\Theta$, where $\Theta \in \boldsymbol{\theta}$ and all other members of $\boldsymbol{\theta}$ are held constant at their true injected values.
-\newline
-
-To elucidate the complications introduced by including the pulsar term, we generate some noisy synthetic data $\boldsymbol{Y}$ using the methods outlined in K23.
-$\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
-The choice of $\boldsymbol{\theta}_{\rm psr}$ corresponds to the 47 pulsars in the 12.5 year NANOGrav data set  \citep{2020ApJ...905L..34A}.
-We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig} on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad.
-We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve).
-The dashed grey line indicates the injected value of $\alpha$ (= 1.0 rad).
-The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
-\newline
-
-There are two important features to discuss in the left-hand panel of Figure \ref{fig:likelihood_comparison}.
-The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
-Similar noisy curves are obtained for $\mathcal{L}(\Omega)$ and $\mathcal{L}(\delta)$.
-For all other static parameters, the $\mathcal{L}(\Theta)$ curves are smooth and globally concave.
-The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.
-This means that whilst it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$, in practice the computation is intractable since the noise in the likelihood curve inhibits sampling algorithms to reliably converge to a definitive posterior.
-Moreover, $\mathcal{L}(\alpha)$ is only a one-dimensional function.
-The challenge for sampling algorithms is further exacerbated when we consider the three-dimensional parameter space in $\alpha - \delta - \Omega$, or the full $7 + 5N$ dimensions of Section \ref{sec:ss_params}.
-\newline
-
-Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\Theta)$ curves whereas other static parameters do not?
-These parameters appear in the pulsar term as a product in the pulsar-dependent phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, where we remind the reader that $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
-Over the typical pulsar distances $d^{(n)} \sim 1$ kpc, a 10 nHz GW can accumulate $10^3$ cycles.
-This creates a confusion problem when taking the cosine of  $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ since we do not know how many cycles the wave has gone through, modulo $2 \pi$.
-\textcolor{red}{TK: this explanation is rubbish!
-How can we say this better?}.
-Conversely, if we drop the pulsar term,  the $\mathcal{L}(\alpha)$ curve (as well as $\mathcal{L}(\Omega)$ , $\mathcal{L}(\delta)$   )
-This is shown by the orange curve in the right-hand panel of Figure \ref{fig:likelihood_comparison} where we plot $\mathcal{L}(\alpha)$, analogous to the left-hand panel, but now use Equation \ref{eq:g_func_trig_earth} in our Kalman filter (i.e.
-just the Earth terms) rather than the full Equation \ref{eq:g_func_trig}.
-In this case there is no confusion regarding the number of cycles the wave has gone through over a distance $d^{(n)}$ and the curve is smooth.
-
-\subsection{New parametrisation}
-In order to include the pulsar terms in an inference model and apply the methods of K23 it is necessary to be able to handle the noisy likelihood curves presented in Section \ref{sec:pulsar_terms_challenge}.
-In this section we introduce a new parametrisation of the model described in Section \ref{sec:model} in order to alleviate the noisy likelihoods.
-\newline
-
+However, dropping the pulsar term leads to biases in the parameter estimation, in particular the source localisation, and a reduced detection probability \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
+It is therefore important to remove this limitation and include the pulsar term in PTA data analysis pipelines.
+
+\subsection{Reparametrisation of the pulsar term} \label{sec:pulsar_term2}
 We define a new parameter 
@@
 \end{equation}
-whilst $\boldsymbol{\theta}_{\rm gw}$  remain unchanged. \newline
-
-This parametrisation in terms of $\chi^{(n)}$ treats the pulsar dependent phase correction as an explicit parameter to be inferred for every pulsar.
-The advantage of this approach is that the likelihood curves $\mathcal{L}(\alpha),\mathcal{L}(\delta),\mathcal{L}(\Omega)$ become smooth and resemble the right-hand panel of Figure \ref{fig:likelihood_comparison}.
-This enables sampling algorithms to efficiently find the likelihood maxima and estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
-The new parametrisation also does not change the dimension of the parameter space, since we have traded $d^{(n)}$ for $\chi^{(n)}$.
+whilst $\boldsymbol{\theta}_{\rm gw}$  remain unchanged. We define $\boldsymbol{\theta'} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta'}_{\rm gw}$. \newline
+
+The reparametrisation in terms of $\chi^{(n)}$ treats the pulsar-dependent phase correction $\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ in the argument of the cosine of the pulsar term as an explicit parameter to be inferred for each pulsar.
+The advantage of this approach, compared to trying to infer $d^{(n)}$ directly, is that the resulting likelihood surfaces are smoother enables sampling algorithms to efficiently find the likelihood maxima and estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+This point is discussed in more detail in Appendix YY.
+The new parametrisation also does not increase the dimension of the parameter space, because we have traded $d^{(n)}$ for $\chi^{(n)}$.
 
@@
 
+	\begin{table}
+		\centering
+		%	\resizebox{\textwidth}{!}{%
+			%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+			\begin{tabular}{lcclc}
+				\toprule
+				&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm pulsar}(\theta) -\Delta_{\rm earth}(\theta) $   \\
+				\hline
+				\multirow{7}{2mm}{High SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
+				& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
+				& $\psi$ &0.0386& 0.000254& -0.0384 \\
+				& $\iota$ & 0.374 &0.126 &-0.248 \\
+				& $\delta$ & 0.00223 &0.000242& -0.00198 \\
+				&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
+				&$h_0$ & 0.168 &0.0846 &-0.0839 \\
+				\hline
+				\multirow{7}{2mm}{Low SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
+				& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
+				& $\psi$ &0.0386& 0.000254& -0.0384 \\
+				& $\iota$ & 0.374 &0.126 &-0.248 \\
+				& $\delta$ & 0.00223 &0.000242& -0.00198 \\
+				&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
+				&$h_0$ & 0.168 &0.0846 &-0.0839 \\
+
+				\bottomrule
+			\end{tabular}
+			\caption{dfdfddd}
+			\label{tab:posterior_errors}
+		\end{table}
+
+		%\begin{table}
+		%	\centering
+		%	%	\resizebox{\textwidth}{!}{%
+			%		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+			%		\begin{tabular}{cccc}
+				%			\toprule
+				%			$\theta$ & \multicolumn{2}{c}{RMSE$(\theta)$} & $\Delta$ \\
+				%			& Earth & Pulsar & \\
+				%			\hline
+				%			$\Omega$ & 1 & 2 & 3 \\ 
+				%			
+				%			\bottomrule
+				%		\end{tabular}
+			%		\caption{}
+			%		\label{tab:posterior_errors}
+			%	\end{table}
+
+		%
+		%\begin{table}
+		%	\centering
+		%	%	\resizebox{\textwidth}{!}{%
+			%		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+			%		\begin{tabular}{lccll}
+				%			\toprule
+				%			$\theta$& \multirow{2}{*}{RMSE$(\theta)$} & Injected Values & Units & Prior  \\
+				%			& 
+				%			\hline
+				%			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
+				%			& $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
+				%			& $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
+				%			& $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
+				%
+				%
+				%			\bottomrule
+				%		\end{tabular}
+			%		\caption{}
+			%		\label{tab:posterior_errors}
+			%	\end{table}
+
 		Initially we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$ for the representative example system described in Table \ref{tab:parameters_and_priors}.
 We consider a "high-SNR" system with $h_0 = 10^{-12}$ and a ``low-SNR" system with $h_0 = 5 \times 10^{-15}$.
-We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ for both the high/low SNR systems.
 Priors on $\boldsymbol{\theta}$ are selected as follows.
@@
 For $\pi(\boldsymbol{\theta}_{\rm psr})$ we use the same priors discussed in K23.
-The addition here is the prior on $\chi^{(n)}$ which we take to be $\text{Uniform}(0,2 \pi)$.
-	All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}
-
+Additionally, we set a prior on $\chi^{(n)}$ as $\text{Uniform}(0,2 \pi)$.
+All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
+\newline
+
+		We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ for both the high/low SNR systems.
 The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_1} for the high-SNR system and in Figure \ref{fig:corner_plot_2} for the low-SNR system.
-
 The histograms shown in the figures are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the solid orange line indicates the true injection value.
@@
 The green curves are the results derived using the Earth-terms measurement equation, Equation \ref{eq:g_func_trig_earth}.
-The blue curves are the results derived using the full measurement equation, including the pulsar terms, Equation \eqref{eq:g_func_trig}
-
-For the high-SNR case, Figure \ref{fig:corner_plot_1} the one-dimensional posteriors for the Earth-terms solution are evidently biased, as was observed in K23.
-Particular biases are observed in $\Phi_0, \psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
-
-The inclusion of the pulsar terms corrects for this bias. The green posteriors
-
-Moreover, the injection values are contained witinin the 2-sigma equivalent contours for every pair of parameters. This is not true in the Earth-terms case.
-
-For the low-SNR case
-
-les important for PE in low SNR
-
-but is important generally for detectio nas we will show.
-
-Takeaways
-
-We are able to estimate hte static parameters
-
-the inclusion of the pulsar terms general improves the accuracy
-
-the chi terms are also recoverable
-
-It is evident that for this representative example with characteristic strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.
-\newline
-
-For some of the parameters the estimated posterior is approximately symmetric about the true injected value (e.g.
-$\Omega$).
-In this case the posterior median and the injected value are approximately coincident.
+The blue curves are the results derived using the full measurement equation, including the pulsar terms, Equation \eqref{eq:g_func_trig}.
+\newline
+
+		For the high-SNR case, Figure \ref{fig:corner_plot_1}, the one-dimensional posteriors for the Earth-terms solution are evidently biased, as was observed in K23.
+Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
+For example, for $\iota$, whilst the injected value is contained within the 90 \% credible interval, the median of the posterior of the Earth-terms solution is shifted from the injected value by 0.35 radians.
+The inclusion of the pulsar terms corrects for this bias.
+The green one-dimensional posteriors exhibit no bias and a generally symmetric about the injected value.
+Moreover, for every pair of parameter the injection values are contained within the 2-sigma equivalent contours.
+This is not true in the Earth-terms case.
+More quantitatively, we define a relative error metric to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
+		\begin{equation}
+			\Delta_{\rm M}(\Theta) = \frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}} 
+		\end{equation}
+		where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ the total number of samples, $\hat{\theta}$ the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimates using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+$\Delta_{\rm M}(\theta)$ then represents the average relative error in the one dimensional posterior.
+The values for each parameter in Figure \ref{fig:corner_plot_1} are summarised it Table \ref{tab:posterior_errors}, along with the relative difference between the two models, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$.
+For every parameter except $\Omega$, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is negative, indicating that the parameter estimates are more accurate using the pulsar terms.
+$\Omega$ is recovered with high accuracy in both cases, and the relative error is small, $\Delta_{\rm M}(\Omega) \sim 10^{-5}$.
+The improvement in $\iota$ is largest, with $\Delta_{\rm pulsar}(\iota) - \Delta_{\rm earth}(\iota) = 0.25$.
+\newline
+
+		For the low-SNR case,  Figure \ref{fig:corner_plot_2}, there is less of a notable improvement in the parameter estimates.
+Qualitatively, the green and blue one-dimensional posteriors are generally overlaid.
+The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest.
+The relative errors 	$\Delta_{\rm M}(\Theta)$ for the low SNR case are also reported in Table \ref{tab:posterior_errors}.
+We see that....
+\textcolor{red}{TK: numbers to be inserted here}.
+We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case.
+This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates.
+Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:detection_new} that they are important for increasing the detection probability for a given signal strength.
+\newline
+
+		In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
+In both the high and low SNR cases $\chi^{(n)}$ is unambiguously recovered.
+		This is shown in Figure \ref{fig:corner_plot_3} where we plot the results for five of the $\chi^{(n)}$ values for both high SNR (left-hand panel) and low SNR (right hand panel).
+The estimates are naturally more constrained in the high SNR case, but in both cases the injection is contained within the 90 \% credible interval.
+Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
+
+		In summary, the main results of this section are:
+		\begin{itemize}
+			\item We are able to estimate all the static parameters of the system using the full pulsar-terms model, for this representative system with characteristic strain $h_0 = 5 \times 10^{-15}$.
+			\item The inclusion of pulsar terms improves the parameter estimation accuracy, especially when the GW signal is loud with respect to the noise.	
+		\end{itemize}
+
+		\subsection{Detection}\label{sec:detection_new}
+
+		\begin{figure}
+			\includegraphics[width=\columnwidth, height = \columnwidth ]{images/CanonicalBayesPlot2000} 	
+			\caption{hellow}
+			\label{fig:bayes1}
+		\end{figure}
+
+		The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
+		\begin{equation}
+			\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
+		\end{equation}
+		Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
+$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
+This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
+The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
+
+		We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
+$\mathcal{M}_1$ is the Earth-terms only model, i.e.
+the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
+The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
+We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
+We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
+For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
+This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
+$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
+\newline
+
+		There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
+The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
+The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
+For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
+That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
+For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
+For the results presented in this subsection we set $n_{\rm live} =2000$.
+
+\section{Validation with synthetic data} \label{sec:testing}
+In this section we outline how synthetic data are generated in order to validate the analysis scheme in Section \ref{sec:detect}.
+Synthetic data enable validation to occur systematically and under controlled conditions.
+In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
+In Section \ref{sec:gendata} we demonstrate how to solve Equations \eqref{eq:spinevol}--\eqref{eq:frequency_evolution}, and \eqref{eq:measurement}--\eqref{eq:g_func} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) $.
+
+\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
+	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
+The pulsar distances relative to the observer are also indicated, with distance $\leq 2 \, {\rm kpc}$ for $38$ pulsars.
+The grey dashed curve denotes the Galactic plane.}
+	\label{fig:pulsar_distrib}
+\end{figure}
+We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
+The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
+\newline
+
+For each pulsar we adopt fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
+A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility.
+The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}.
+The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+\newline
+
+The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist.
+The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency}, and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
+In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
+We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$.
+The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
+\begin{align}
+	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
+	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
+\end{align}
+where $T_{\rm cad}$ is the cadence of the timing observations.
+For MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$ ($\pm 2\sigma$ confidence limits).
+The uncertainties are broad; for the purpose of generating astrophysically representative synthetic data we straightforwardly adopt the central values in this paper.
+Throughout this work we assume for simplicity that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
+In order to relate Equation \eqref{eq:sigmap} to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
+\begin{eqnarray}
+	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
+\end{eqnarray}
+For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, with $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline
+
+As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
+\begin{eqnarray}
+	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
+\end{eqnarray}
+and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach.
+We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
+A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$.
+In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$.
+In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}.
+We see that the synthetic and empirical residuals are qualitatively similar.
+For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is much less than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ s$^{-3/2}$ for PSR J1939$+$2134.
+
+%This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars.
+If we set $\ln \alpha_1 = -38$ in Equation \eqref{eq:sigmap}, close to its lower limit, then the estimated $\sigma^{(n)}$ for  PSR J1939+2134 is $\sigma^{(n)} =3.90 \times 10^{-27}$ which is in much better agreement with the value inferred from the residuals comparison method.
+
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
+	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J1939+2134.
+The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
+The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$.
+The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}.
+We set the TOA uncertainty to be a constant, viz.
+$0.1 \mu$s.
+The blue and orange residuals are qualitatively similar by inspection.
+Similar results are obtained for the other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} by tuning the value of $\sigma^{(n)}$.}
+	\label{fig:qualitative_compare}
+\end{figure}
+\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
+We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
+\begin{enumerate}[leftmargin=2em]
+	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
+	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}.
+\end{enumerate}
+Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \ref{sec:synt_pta}.
+For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
+\newline
+
+The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, through
+\begin{equation}
+	\sigma_{\rm m} \approx \sigma_{\rm TOA} f_{\rm p}^{(n)}  \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
+\end{equation}
+Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct.
+Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 10^2$ Hz observed with weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s, Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz.
+Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred for the sake of simplicity; this assumption can be relaxed easily when analysing real data.
+Whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}(t)$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar.
+\newline
+
+Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} dB(t) | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $dB(t)$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
+To fix the problem, we subtract the deterministic frequency evolution and track the new variable
+\begin{equation}
+	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ , \label{eq:hetero1}
+\end{equation}  
+equivalent to a change of variables. We similarly modify the measurement variable to be
+\begin{equation}
+	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ , \label{eq:hetero2}
+\end{equation}
+where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic evolution based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice.
+For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately.
+Equation \eqref{eq:measurement} is then updated to read
+\begin{equation}
+	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
+	\label{eq:measurement_cov}
+\end{equation}
+We emphasise that the change of variables in Equations \eqref{eq:hetero1} and \eqref{eq:hetero2} is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g.
+long double, quadruple).
+It does not remove any degrees of freedom nor does it involve an approximation.
+In particular $f_{\rm em}^{(n)}(t_1)$
+and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, i.e.
+Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.
+
+\section{Representative end-to-end analysis} \label{sec:rep_example}
+In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
+The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
+The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space} and \ref{sec:bias_and_identifiability}.
+To assist with reproducibility, we present intermediate outputs from each step.
+In Section \ref{sec:priors} we define the priors on $\boldsymbol{\theta}$.
+In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter.
+In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
+In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates.
+In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$.
+\newline
+
+The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
+The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
+
+\subsection{Prior distribution}\label{sec:priors}
+The first step is to select a reasonable Bayesian prior, $\pi(\boldsymbol{\theta})$, for the static parameters.
+For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
+The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires additional discussion below.
+\newline
+
+The parameters that govern the deterministic evolution of $f_{\rm p}^{(n)}(t)$, namely $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, are well-determined by radio timing observations.
+We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues \citep{Manchester2005}.
+For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} $ per cent and $\pm 2.31 \times 10^{-3}$ per cent respectively.
+In this paper we adopt uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database \citep{Manchester2005}.
+Wider-than-necessary priors, such as those above, test the method more stringently than narrow priors.
+The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are wide or narrow.
+\newline
+
+The pulsar distances $d^{(n)}$ are less constrained than $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with typical uncertainties $\sim 10 \%$ \citep{Yao2017,Arzoumanian2018ApJS..235...37A}.
+In this paper, however, we omit the term involving $d^{(n)}$ from Equation \eqref{eq:g_func}, as justified in Section \ref{sec:parameter_estim}, so there is no need to set a prior on $d^{(n)}$.
+Furthermore we do not set a prior on $\gamma^{(n)}$, because we have $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and $\gamma^{(n)}$ is effectively ``unobservable'' over a decade; that is, for $T_{\rm obs}=10$ years the solution of Equation \eqref{eq:spinevol} is approximately independent of $\gamma^{(n)}$, as long as $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
+It is therefore sufficient for validation purposes to carry $\gamma^{(n)}$ through the analysis at its injected value.
+\footnote{We check this empirically in two independent ways for an informal selection of test cases: (i) we set an uninformative prior $\pi\left [\gamma^{(n)} / 1 \, {\rm s^{-1}} \right] \sim$ LogUniform($10^{-15}, 10^{-10}$); and (ii) we deliberately displace $\gamma^{(n)}$ in the inference analysis from its true, injected value in the synthetic data, e.g.
+$\gamma^{(n)} = 10^{-14} \text{ s}^{-1}$ versus $10^{-13} \text{ s}^{-1}$, respectively.
+The results from (i) and (ii) are found to be the same as those reported in Section \ref{sec:rep_example}.} \newline
+
+Most pulsars in the synthetic PTA have $10^{-22} \leq \sigma^{(n)} / (1 \, {\rm s^{-3/2}}) \leq 10^{-20}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
+For these pulsars we set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$).
+The sole exception is PSR J1939+2134 which has $\sigma^{(n)} \sim 10^{-16} \text{s}^{-3/2}$.
+For validation purposes, we artificially set $\sigma^{(n)} = 10^{-20} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within the aforementioned log-uniform prior.
+This is done purely for testing convenience; it is straightforward to expand the prior when analysing real, astronomical data.
+\newline
+
+By not setting priors on $\gamma^{(n)}$ and $d^{(n)}$ we reduce the number of parameters to $7 + 3N$.
+We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter set, c.f.
+Equation \eqref{eq:psrparams}.
+Explicitly, we write
+\begin{equation}
+	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ . \label{eq:psrreduced}
+\end{equation}
+The injected static parameters and their corresponding priors for the representative analysis in Section \ref{sec:rep_example} are summarised in Table \ref{tab:parameters_and_priors}.
+\begin{table*}
+	\centering
+%	\resizebox{\textwidth}{!}{%
+%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+	\begin{tabular}{lccll}
+		\toprule
+		&Parameter & Injected value & Units & Prior  \\
+		\hline
+		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
+	  & $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
+	  & $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
+	  & $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
+	  & $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
+	  & $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
+	  & $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
+		\hline
+%		\vspace{1mm}\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
+		 & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
+		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
+		\vspace{1mm} $\boldsymbol{\theta}_{\rm psr}$ &  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
+		\vspace{1mm} & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
+		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
+		\bottomrule
+	\end{tabular}
+	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the prior used for Bayesian inference on each parameter (rightmost column).
+The top and bottom halves of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ respectively.The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
+The subscript ``SC'' indicates that the injected value has been calculated using Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} in \protect \cite{Shannon2010}.
+The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
+No priors are set for $d^{(n)}$ and $\gamma^{(n)}$ since those parameters do not enter the inference model,  as discussed in Section \ref{sec:priors}, c.f.
+Equation \eqref{eq:psrreduced}.
+}
+	\label{tab:parameters_and_priors}
+\end{table*}
+
+\subsection{Earth and pulsar terms}\label{sec:earth_psr_terms}
+
+	\subsection{Posterior distribution}\label{sec:parameter_estim}
+	In this section, we calculate the posterior probability distribution for the static parameters $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$ and compare it to the known, injected values.
+The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior), and (ii) to give a preliminary sense of its accuracy.
+Initially we consider a single noise realisation when generating the synthetic data for the representative example in Table \ref{tab:parameters_and_priors}.
+We apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$.
+\newline
+
+Figure \ref{fig:corner_plot_1} displays results for the seven parameters in  $\boldsymbol{\theta}_{\rm gw}$ in the form of a traditional corner plot.
+The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
+The dashed vertical blue lines mark the 0.16 and 0.84 quantiles; the solid orange line marks the known injected value.
+The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces.
+All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values.
+There is scant evidence of railing against the prior bounds.
+For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
+The injected values are contained within the 90\% credible interval for all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ .
+\newline
+
+The posterior is approximately symmetric about the known injected value for some parameters such as $\Omega$, where the posterior median and the injected value coincide approximately.
 For other parameters (e.g.
-$\iota$) the distribution is not symmetric about the injected value and the posterior median and the injected value are not coincident.
-The median value of the posterior for $\iota$ is shifted by $\sim 0.35$ rad relative to the injected value, although the injected value does still remain inside the 90\% credible interval.
+$\iota$) the distribution is not symmetric about the injected value and the posterior median and the injected value do not coincide.
+The median value of the posterior for $\iota$ is shifted by $\approx 0.35$ rad relative to the injected value, although the injected value does still remain inside the 90\% credible interval.
 Similar effects are seen, albeit with a smaller shift, in other parameters such as $\delta$ and $\alpha$.
 For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
-a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation.
-We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
-Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
-We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.
-
-\subsection{Detection}\label{sec:detection_new}
-
+a bias, or a random outcome specific to this particular noise realisation.
+We explore this further in Sections \ref{sec:multiple_noise} and \ref{sec:pulsar_terms_drop} using more noise realizations and show that indeed there is a systematic bias from dropping the pulsar terms, similar to that reported by \cite{Zhupulsarterms}.
+\newline
+
+Similar results are obtained for the 3$N$ parameters in $\boldsymbol{\theta}_{\rm psr, reduced}$.
+Again, the parameters are recovered unambiguously, in the sense that the nested sampler converges smoothly to a unimodal joint posterior near the known, injected parameter values, with all $3N$ injected parameters lying within the 90\% credible interval of the $3N$ one-dimensional, marginalized posteriors.
+We do not display the resulting corner plot because it is too big ($3N = 141$), and because inferring ${\boldsymbol{\theta}}_{\rm gw}$ is the focus of this paper and contemporary PTA analyses more generally.
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
+	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The horizontal and vertical orange lines indicate the true injected values.
+The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters.
+The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter.
+The supertitles of the marginalized one-dimensional histograms specify the posterior median and the 0.16 and 0.84 quantiles.
+We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$.
+The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+The horizontal axes span a subset of the prior domain for all seven parameters.}
+	\label{fig:corner_plot_1}
+\end{figure*}
+
+%\begin{figure*}
+%	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
+%	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The vertical orange lines indicate the true injected values.
+The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+We are able to accurately estimate each parameter of interest.
+We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
+%	\label{fig:corner_plot_1}
+%\end{figure*}
+
+\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
+	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely.
+The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation.
+The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations.
+There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation $\sim10 \%$ across the 63 combinations.
+A slight skew-bias is apparent in some of the parameters, e.g.
+$\iota$.
+Weak correlations can be seen between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.}
+	\label{fig:corner_plot_2}
+\end{figure*}
+\begin{table}
+	\centering
+	\begin{tabular}{ccll}
+		\toprule
+		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
+		\hline
+		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
+		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
+		$\psi$              & $2.50$ & rad & $0.16$ \\
+		$\iota$             & $1.0$ & rad & $0.14$ \\ 
+		$\delta$              & $1.0$  & rad & $0.09$ \\
+		$\alpha$          & $1.0$  & rad & $0.17$\\
+		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
+		\bottomrule
+	\end{tabular}
+	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
+$W_{1, \rm median}$ is generally smaller than the prior domain.}
+	\label{tab:Wasserstein}
+\end{table}
+
+The results in Section \ref{sec:parameter_estim} are obtained for a single realisation of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
+It is important to confirm that the analysis scheme returns accurate answers for arbitrary noise realisations and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous by accident.
+It is also important to quantify, albeit approximately, the natural random dispersion in the one-dimensional posterior medians from one noise realization to the next, as the dispersion is a practical measure of the accuracy of the parameter estimation scheme, when it is applied to real astronomical data, where the true parameter values and specific noise realisation are unknown.
+\newline
+
+To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
+For each realisation we independently estimate the static parameters, $\boldsymbol{\theta}$.
+In Figure \ref{fig:corner_plot_2} we plot the estimates of $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary realisations.
+The seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$ are recovered unambiguously.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, which stems from one realisation.
+We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
+As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm psr}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
+\newline
+
+Figure \ref{fig:corner_plot_2} confirms two main points: (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
+Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 (i.e.
+$92 \%$) out of the 63 possible combinations of the seven parameters and nine realizations.
+Figure \ref{fig:corner_plot_2}, like Figure \ref{fig:corner_plot_1}, displays tentative signs of bias, where the one-dimensional posteriors are not symmetric about the injected value.
+For example, the maximum a posteriori probability estimates of $\iota$ appear left-skewed, consistently underestimating the injected value in all nine realisations.
+A similar but messier trend is seen in $\delta$.
+However, the width of the posteriors is comparable to the putative bias, so it is difficult to draw strong conclusions.
+Biases are discussed in detail in Section \ref{sec:bias_and_identifiability}.
+There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.
+\newline
+
+We now consider the complete set of $10^3$ noise realisations, going beyond the subset of nine realisations in the preceding discussion.
+We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$?
+There is no unique way to answer this question.
+In this paper we appeal to the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory, which defines an intuitive notion of similarity between probability distributions.
+The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}; a short review is presented in Appendix \ref{sec:wasserstein}.
+The WD measures the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
+In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}.
+Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$ with respect to the PDFs $\mu$ and $\nu$.
+Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad.
+Then we can infer $| \langle \iota \rangle_\mu - \langle \iota \rangle_\nu | \leq 0.5 \, {\rm rad}$.
+\newline
+
+Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$.
+As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
+These values quantify approximately the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
+Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$.
+These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for $10^3$ noise realizations.
+Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
+
+\subsection{Detectability versus $h_0$} \label{sec:detection}
+We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}.
+In Equation \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-term-only model with a GW present, i.e.
+the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
+To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
+\footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.} \newline
+
+We see in Figure \ref{fig:bayes} an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$.
+The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 4 \times 10^{-15}$.
+Of course, the minimum detectable strain is particular to the system in Table \ref{tab:parameters_and_priors}.
+It is influenced in general by $T_{\rm obs}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$, as discussed in Section \ref{sec:parameter_space}.
+Adjusting $\sigma_{\rm m}$ we find that an approximate quadratic relationship $\beta \propto h_0^2 / \sigma_{\rm m}^2$ also exists, where $h_0 / \sigma_{\rm m}$ is the effective signal-to-noise ratio.
+\newline
+
+The odds ratio $\beta$ drops off for $h_0 \lesssim 4 \times10^{-15}$.
+This happens because the two competing models becoming increasingly indistinguishable once the measurement noise dominates the GW signal.
+Moreover, the points in Figure \ref{fig:bayes} become sparser for $h_0 \lesssim 4 \times 10^{-15}$.
+This happens because we obtain $\beta <0$ for the missing, intermediate points.
+This is a noise artefact of the nested sampler.
+When the sampler converges sub-optimally, the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails, and $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+There is nothing special about the particular missing points; if one recreates the $\beta(h_0)$ curve by rerunning the nested sampler with another random seed, a different set of points are missing.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$, because the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(n_{\rm live}^{-1/2})$ (see Section \ref{sec:nested_sampling}).
+In this section we set $n_{\rm live} =2000$.
+\begin{figure}
+	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
+	\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strains, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
+The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
+The minimum detectable strain, for $\beta < 10$, equals $4 \times 10^{-15}$.
+Missing points for $\beta \lesssim 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
+	\label{fig:bayes}
+\end{figure}
+
+\section{SMBHB source parameters} \label{sec:parameter_space}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/pp_plot_new1}
+	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
+PP plot).
+The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}.
+Each coloured curve corresponds to a different parameter (see legend).
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
+The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals.
+For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope.
+$\Omega$ is generally well-estimated (i.e.
+it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e.
+the curves lie above the unit diagonal for low credible intervals, and below the unit diagonal for high credible intervals).
+This is due to a modelling bias whose origin is discussed in Sections \ref{sec:parameter_estim} and \ref{sec:pulsar_terms_drop}.}
+	\label{fig:parameter_space}
+\end{figure}
+Section \ref{sec:rep_example} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}.
+In this section we test the method for various systems, varying the SMBHB source parameters through astrophysically relevant ranges.
+\newline
+
+We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
+For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
+To summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself.
+In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
+\newline
+
+ Figure \ref{fig:parameter_space} displays the results of the numerical experiment described in the previous paragraph.
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections.
+We see that only $\Omega$ falls wholly within the $3\sigma$ shaded region.
+The PP curves for the other parameters deviate from the diagonal of unit slope.
+The deviation is more pronounced for $\alpha$ and $\delta$ and less for $\psi$ and $\Phi_0$.
+The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within higher-value credible intervals than would be expected statistically, and there are more injections contained within lower-value credible intervals.
+This stems from the bias noted in Section \ref{sec:multiple_noise}.
+The origin of the bias is discussed in Section \ref{sec:pulsar_terms_drop}.
+
+\section{Systematic biases} \label{sec:bias_and_identifiability}
+ Two systematic aspects of the tests in Sections \ref{sec:rep_example} and \ref{sec:parameter_space} merit further study.
+The first relates to parameter identifiability, i.e.
+can we uniquely identify the parameters of the model given the data and the model structure?
+This question is explored in Section \ref{sec:identif}.
+The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
+It is the focus of Section \ref{sec:pulsar_terms_drop}.
+In order to elucidate these matters without confusion from the measurement noise, we switch to the high-SNR regime and set $h_0 = 10^{-12}$ (c.f.
+Figure \ref{fig:bayes}) for the representative system in Table \ref{tab:parameters_and_priors}.
+
+\subsection{Identifiability}\label{sec:identif}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
+	\caption{Corner plot of the joint posterior distribution of the GW source parameters $h_0$ and $\iota$ for the representative system in Table 1 in the high-SNR regime ($h_0 = 10^{-12}$).
+Nine coloured curves are plotted for nine random noise realizations.
+The posteriors overlap less than for the other five parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+A correlation between $h_0$ and $\iota$ is evident in the bottom left panel.}
+	\label{fig:just_iota_h}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_surface}
+	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$-$h_0$ plane for a single noise realisation.
+The red dot marks the maximum of $\log {\cal L}$.
+The colour scale is normalised with respect to the absolute value of the maximum.
+An S-shaped, flat-topped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
+	\label{fig:likelihood_surface}
+\end{figure}
+In Figure \ref{fig:just_iota_h} we plot the one- and two-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations.
+The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$.
+One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section \ref{sec:pulsar_terms_drop}; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high.
+One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$.
+Specifically, across $10^3$ noise realisations, the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
+Both $W_{1,{\rm median}}$ values are higher (as a fraction of the true, injected value) than in the lower-SNR tests conducted in Sections \ref{sec:rep_example} and \ref{sec:parameter_space}.
+In contrast, $W_{1,\rm median}$ for the other five parameters in $\boldsymbol{\theta}_{\rm gw}$ decreases, as the SNR increases.
+\newline
+
+We infer different $h_0$ and $\iota$ posteriors for different noise realizations, because $h_0$ and $\iota$ are weakly identifiable.
+Identifiability is a formal concept from the theory of signal processing, which is applied routinely to engineering problems \citep{e5be7c83a0d24500826f6e1b414d1733}.
+It refers to whether it is theoretically possible to infer unique and accurate parameter values, given the measured data and the model structure.
+Sources of non-identifiability are generally categorised as either ``structural'' , arising from the structure of the model (e.g.
+its invertibility), or ``practical'', arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
+Regarding $\iota$ and $h_0$, the source is structural.
+Equations \eqref{eq:hphx} and \eqref{eq:hphx2} contain a weak degeneracy between $\iota$ and $h_0$.
+For example, the cross polarisation strain $h_{\times}$ involves a trade-off between $\iota$ and $h_0$, which is resolved by knowing $h_+$ via Equation \eqref{eq:hphx}.
+However, the measured strain is a linear combination of $h_\times$ and $h_+$; the two polarizations are not measured independently, which makes it challenging to separate $\iota$ and $h_0$.
+In fact $\iota$ and $h_0$ are structurally identifiable --- but only weakly.
+That is, the $\iota$-$h_0$ likelihood contours plateau to a flat-topped ridge with a small, non-zero gradient close to the maximum likelihood solution.
+The likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
+Whilst a single, unique maximum does exist (labelled by the red dot in the figure), it lies along an S-shaped ridge where the likelihood values are all similar.
+Consequently it is difficult for the nested sampling algorithm to settle on the red dot; it trades off accuracy in one parameter against the other, depending on the particular realisation of the noise.
+By way of verification, if we fix one parameter at its true value (i.e.
+set a delta function prior) during the inference analysis, the posteriors for the other parameter overlap indistinguishably  for different noise realisations, i.e.
+$W_1 \to 0$.
+
+\subsection{Dropping the pulsar terms}\label{sec:pulsar_terms_drop}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
+	\caption{Same as Figure \ref{fig:corner_plot_1} but for a high-SNR system with $h_0 = 10^{-12}$.
+For five out of seven parameters the true, injected value does not fall within the 90\% credible interval and falls outside the plotted domain.
+The estimated posteriors are biased away from the injected value due to dropping the pulsar terms.
+Note that the plotted domain is narrower than in Figure \ref{fig:corner_plot_1}.}
+	\label{fig:bias_for_large_h}
+\end{figure*}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
+	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth term in the Kalman filter (Equation \ref{eq:measuremen_earth}, blue curve), the pulsar terms in the Kalman filter (Equation \ref{eq:measurement}, orange curve), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter (green curve).
+All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
+The coloured dashed vertical lines mark the maximum for the corresponding model; the green and orange dashed lines overlap at this resolution.
+For the Earth-term model the likelihood maximum ($\iota \approx 0.7$ rad) does not coincide with the injected value ($\iota = 1.0$ rad).
+For the pulsar-terms model the maximum and injected value do coincide.
+10 noise realisations are plotted for each situation but overlap at this resolution; they are indistinguishable to the eye.}
+	\label{fig:likelihood_surface_iota}
+\end{figure}
+
+Sections \ref{sec:rep_example} and \ref{sec:parameter_space} present preliminary evidence that the inferred value of ${\boldsymbol{\theta}}_{\rm gw}$ (e.g.
+the medians of the one-dimensional marginalised posteriors) is biased away from the true injected value.
+Figure \ref{fig:bias_for_large_h} displays a corner plot for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but for high SNR $(h_0 = 10^{-12})$.
+Except for the elements $\Omega$ and $h_0$, the injected values lie outside the 90\% credible interval, and indeed fall outside the plotted domain.
+The deviation is most severe for $\iota$ with a bias of $\approx 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$.
+Similar results are obtained for other noise realisations.
+Whilst $h_0$ does not exhibit any noticeable bias for the particular noise realisation generating Figure \ref{fig:bias_for_large_h}, in general it does exhibit a correlated bias with $\iota$ across multiple noise realisations, as noted in the bottom left panel of Figure \ref{fig:just_iota_h}.
+In this section we demonstrate how the bias stems from dropping the pulsar terms described in Section \ref{sec:parameter_estim}.
+\newline
+
+In Figure \ref{fig:likelihood_surface_iota} we plot the log-likelihood returned by the Kalman filter, viz.
+Equation \eqref{eq:likelihood}, as a function of $\iota$ (i.e.
+$\log \mathcal{L}(\iota)$) across the prior domain, holding all other parameters constant at the true injected value of the representative system in Table \ref{tab:parameters_and_priors}.
+We generate 10 plots of $\log \mathcal{L}(\iota)$ corresponding to 10 separate realisations of the system noise.
+The solid blue curves only includes the Earth term in the measurement equation (i.e.
+using Equation \eqref{eq:measuremen_earth} in the Kalman filter).
+The solid orange curves include the Earth and pulsar terms in the measurement equation (i.e.
+using Equation \eqref{eq:measurement}).
+The solid green curve corresponds to a control experiment, where we drop the pulsar terms from both the synthetic data and the measurement equation.
+The vertical dashed lines mark the corresponding maxima, which nested sampling seeks.
+\newline
+
+The key observation from Figure \ref{fig:likelihood_surface_iota} is that the location of the maximum ($\iota \approx 0.7 \, {\rm rad}$) does not coincide with the true, injected value ($\iota = 1.0 \, {\rm rad}$), when only the Earth terms are included.
+This is the origin of the bias in the one-dimensional marginalised posteriors in Figure \ref{fig:bias_for_large_h}.
+In contrast, the maximum aligns with the true, injected value, when the pulsar terms are included.
+As a cross-check, agreement is also achieved, if we drop the pulsar terms from the synthetic data (which is not the situation in reality, of course) and use the Earth term only for inference.
+Similar results are obtained for other parameters, e.g.
+$\delta$.
+A discussion about why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches, are presented in Appendix \ref{sec:bias_appendix}.
+Dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}.
+The bias is not particular to our method; it is shared by all likelihood-based methods that do not include the pulsar terms in the inference model.
+Generalizing the Kalman tracker and nested sampler to correct for the bias will be a key goal of a forthcoming paper.
+
+\section{Conclusion}\label{sec:discussion}
+
+In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs in PTA data.
+The new method is complementary to traditional approaches.
+We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than fitting for the ensemble-averaged power spectral density of the noise.
+That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data.
+We implement a Kalman filter in order to track the intrinsic rotational state evolution of the pulsar and combine it with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model with and without a GW.
+\newline
+
+ We test the new method on synthetic data and find that it detects injected signals successfully and estimates their static parameters accurately with relatively low computational cost.
+We initially focus on a single, astrophysically representative, SMBHB GW source observed synthetically by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$.
+The minimum detectable strain is estimated to be $\min(h_0) \approx 4 \times 10^{-15}$ for $\iota=1.0$ rad.
+We then repeat the parameter estimation exercise for 1000 noise realisations to compute the natural dispersion (``cosmic variance'') in the recovered values and hence quantify the statistical accuracy of the method in a real astrophysical application, when the true parameter values are unknown.
+Consistent posteriors are obtained for most realisations.
+The median WD is limited to $\lesssim 3\%$ of the width of the prior domain (i.e.\ negligible railing).
+The median WD divided by the injected value, an approximate measure of the natural dispersion, ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
+\newline
+
+Exploration of a broader SMBHB parameter domain at fixed $\iota$ and $h_0$ via 200 randomly sampled parameter vectors reveals a bias in the estimates of the static parameters $\delta$ and $\alpha$, and to a lesser extent $\psi$ and $\Phi_0$.
+The bias is examined for a specific SMBHB GW source in the limit of high-SNR.
+It is greatest for $\iota$, amounting to $\approx 0.3$ rad.
+Smaller biases of $\lesssim 0.1$ rad are also observed in $\Phi_0, \psi, \delta$ and $\alpha$.
+The bias is shown to result from dropping the pulsar terms from the measurement equation in the Kalman filter, consistent with the work of previous authors \citep{Zhupulsarterms,Chen2022}.
+The method also exhibits a trade-off between $h_0$ and $\iota$, which follows formally from the weak identifiability of $h_0$ and $\iota$.
+\newline
+
+We emphasize that the Kalman tracker and nested sampler in this paper do not supplant traditional PTA analysis approaches; they complement traditional approaches and are most powerful when used in tandem.
+Relatedly, it is misleading to ask whether the Kalman tracker and nested sampler are more or less sensitive than traditional approaches for two reasons.
+First, one must still generalize the Kalman tracker and nested sampler to ingest TOAs directly, as happens traditionally, instead of ingesting a frequency time series, as in this paper.
+Generalizing the algorithm is a subtle task and is postponed to a forthcoming paper.
+Second, the Kalman tracker and nested sampler are conditional on a noise model that is related but different to the noise model in traditional approaches.
+The analysis in this paper assumes a specific, time-ordered realization of a mean-reverting Ornstein-Uhlenbeck process satisfying Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law power spectral density, whose amplitude and exponent are adjustable.
+Hence the sensitivities cannot be compared directly.
+Clarifying the similarities and differences between various approaches promises to be a fruitful avenue of future work.
+It is also a subject of attention in audio-band GW data analysis involving hidden Markov models applied to data from terrestrial long-baseline interferometers \citep{PhysRevD.102.023006,PhysRevD.105.022002,Abbott_2022SCO,2022PhRvD.106f2002A}.
+\newline
+
+ The approach in this paper can be extended in at least five ways, enumerated as (i)--(v) below. \newline
+
+(i) A natural first extension is to retain the pulsar terms in Equation \eqref{eq:measurement}.
+Alternatively, if the method continues to be used with solely the Earth term, in line with standard practice in published PTA analyses, it would be desirable to evaluate systematically the incurred biases in the model parameters across an astrophysically representative parameter domain, supplementing the results obtained by other authors \citep{Zhupulsarterms,Chen2022}.\newline
+
+(ii) In this paper, we consider one specific configuration of the synthetic PTA, namely the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
+It is interesting to compare the performance of the method using different pulsar configurations, e.g.
+those in the PPTA and EPTA.
+Adding pulsars to PTAs increases the computational cost so it may be advantageous to select a subset of pulsars by exploiting formal optimization techniques from electrical engineering \citep{2023MNRAS.518.1802S}, although the Kalman filter and nested sampler are already cheaper computationally than some other methods.
+\newline
+
+(iii) In practice different PTA pulsars are observed with different cadences at different times.
+Extending the Kalman filter to non-uniform time sampling is straightforward \citep{zarchan2000fundamentals}.
+\newline
+
+(iv) The assumption of a monochromatic source is well-justified astrophysically in various regimes (see Section \ref{sec:plane_gw}) and is an appropriate starting point for this introductory paper.
+Nevertheless, SMBHBs are not strictly monochromatic.
+It is interesting to extend the state-space framework such that $f_{\rm gw}$ evolves in time.
+For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) evolution adds noise incoherently to the pulsar terms, whilst for $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms induce phase shifts that affect the overall phase coherence \citep{Sesana2010,Perrodin2018}.
+Careful consideration of the evolution of $f_{\rm gw}$ will be needed when including the pulsar terms in the inference model, as in point (i).
+\newline
+
+(v) Finally, we assume in this paper that there is only one GW source.
+However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
+The Kalman filter extends naturally to multiple sources; one can modify Equation \eqref{eq:measurement} easily to accommodate a linear superposition of GWs.
+Taking the logic further, the stochastic background itself is arguably an incoherent sum of many individual GW sources.
+As long as a way can be found to summarize economically the many static parameters associated with the background sources, it should be possible for a Kalman filter and nested sampler to operate together to detect the stochastic background by generalizing the model selection procedure in Sections \ref{sec:model_selection} and \ref{sec:detection}.
+Summarizing the parameter set economically, while respecting the mathematical structure of the Kalman filter, is a subtle challenge, which we postpone to a forthcoming paper.
+If successful, it will complement the traditional approach of cross-correlating pulsar residuals to uncover the Hellings-Downs curve \citep{Hellings,2023ApJ...951L...8A}.
+\newline
+
+\section*{Acknowledgements}
+This research was supported by the Australian Research Council Centre of Excellence for Gravitational Wave Discovery (OzGrav), grant number CE170100004.
+The numerical calculations were performed on the OzSTAR supercomputer facility at Swinburne University of Technology.
+The OzSTAR program receives funding in part from the Astronomy National Collaborative Research Infrastructure Strategy (NCRIS) allocation provided by the Australian Government.
+
+\section*{Data Availability}
+No new data were generated or analysed in support of this research.
+
+%%%% REFERENCES
+\bibliographystyle{mnras}
+\bibliography{example} % if your bibtex file is called example.bib
+
+\appendix
+
+\section{Kalman filter} \label{sec:kalman}
+In this appendix, we describe the Kalman filter algorithm used in this paper.
+General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}.
+The mapping onto the specific continuous-time state-space model in Section \ref{sec:model} is written down in Section \ref{sec_kalman_specific}.
+
+\subsection{Recursion equations}\label{sec_kalman_general}
+The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation
+\begin{equation}
+	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
+\end{equation}
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
+The Kalman filter evolves the underlying states according to
+\begin{equation}
+	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
+\end{equation}
+where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix.
+$\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline
+
+The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage.
+The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
+Specifically, the predict step proceeds as
+\begin{align}
+	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
+		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ ,
+\end{align}
+where $\hat{\boldsymbol{P}}_{k|k-1}$ is the covariance of the prediction.
+Note that the predict stage is independent of the measurements.
+The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
+\begin{align}
+	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \\
+	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \\
+	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
+		\hat{\boldsymbol{X}}_{k|k} &=\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \ , \label{eq:kalmangainupdate} \\
+			\hat{\boldsymbol{P}}_{k|k} &= \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1} \ .
+\end{align}
+Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
+$\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$.
+For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
+\newline
+
+To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$.
+In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.
+
+\subsection{State space representation of a PTA analysis}\label{sec_kalman_specific}
+We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:model} as follows. \newline
+
+We identify $\boldsymbol{X}(t)$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e. 
+\begin{equation}
+	\boldsymbol{X}(t) = \left(f_{\rm p}^{(1)}(t), f_{\rm p}^{(2)}(t), ..., f_{\rm p}^{(N)}(t)\right) \ .
+\end{equation}
+Analogously,  we package the measured pulsar frequencies as
+\begin{equation}
+	\boldsymbol{Y}(t) = \left(f_{\rm m}^{(1)}(t), f_{\rm m}^{(2)}(t), ..., f_{\rm m}^{(N)}(t) \right) \ .
+\end{equation}
+The states evolve according to the continuous stochastic differential equation (c.f. Equation \eqref{eq:frequency_evolution})
+\begin{equation}
+	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \ , \label{eq:kalmn2}
+\end{equation}
+where $\boldsymbol{A}$ is a diagonal $N \times N$ matrix,
+\begin{equation}
+	\boldsymbol{A} = \text{diag} \left(-\gamma^{(1)}, -\gamma^{(2)}, ..., -\gamma^{(N)}\right) \ ,
+\end{equation}
+and $\boldsymbol{C}(t)$ is a time-dependent vector with $n$-th component
+\begin{equation}
+	C^{(n)} =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
+\end{equation}
+The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, with
+\begin{equation}
+	\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right) \ .
+\end{equation}
+
+In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic},
+\begin{equation}
+	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ .
+\label{eq:gardenier}
+\end{equation} 
+From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2}, with
+\begin{align}
+	\boldsymbol{F}_k &= e^{\boldsymbol{A} \Delta t } \  \\
+	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
+\end{align}
+\begin{align}
+	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
+	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ,
+\end{align}
+\begin{equation}
+	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \ ,  \label{eq:appendix_noise}
+\end{equation}
+\begin{align}
+	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left(\Delta t + t_k \right) \nonumber \\ 
+	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ ,
+\end{align}
+and $\Delta t = t_{k+1} - t_k$. From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
+\begin{equation}
+	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) \ ,
+\end{equation}
+with 
+\begin{equation}
+Q^{(n)} = \frac{[\sigma^{n}]^2}{2 \gamma^{(n)}} \left[ 1 - e^{-2 \gamma^{(n)} \Delta t}\right] \ .
+\end{equation}
+
+The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
+These are defined straightforwardly from Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
+Specifically,
+$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
+The measurement covariance satisfies $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$ for all $k$.
+
+\section{Synthetic PTA specifications}\label{appendix_fiducial}
+The synthetic PTA deployed for testing in this paper is constructed to mimic the $N=47$ pulsars from the 12.5-year NANOGrav data release, whose sky positions are plotted in Figure \ref{fig:pulsar_distrib}.
+The construction recipe is described in full in Section \ref{sec:synt_pta}.
+In this appendix, we record a complete list of the pulsar parameters in Table \ref{tab:pulsar_values}, to assist the interested reader in implementing the analysis scheme and reproducing the test results in Sections \ref{sec:rep_example}--\ref{sec:bias_and_identifiability}.
+
+\begin{table*}
+\begin{tabular}{lccccc}
+	\toprule
+	JNAME &    $f_{\rm em} (t_1)$ (Hz) &  $\dot{f}^{(n)}_{\rm em}(t_1)\times 10^{16}$ (s$^{-2}$) &  $\alpha$ (deg) &  $\delta$ (deg) &  $d$ (kpc) \\
+	\midrule
+	J0023+0923 & 327.8 &      -12.3 &   5.8 &    9.4 &   1.8 \\
+	J0030+0451 & 205.5 &       -4.3 &   7.6 &    4.9 &   0.3 \\
+	J0340+4130 & 303.1 &       -6.5 &  55.1 &   41.5 &   1.6 \\
+	J0613-0200 & 326.6 &      -10.2 &  93.4 &   -2.0 &   0.9 \\
+	J0636+5128 & 348.6 &       -4.2 &  99.0 &   51.5 &   0.7 \\
+	J0645+5158 & 112.9 &       -0.6 & 101.5 &   52.0 &   1.2 \\
+	J0740+6620 & 346.5 &      -14.6 & 115.2 &   66.3 &   1.1 \\
+	J0931-1902 & 215.6 &       -1.7 & 142.8 &  -19.0 &   3.7 \\
+	J1012+5307 & 190.3 &       -6.2 & 153.1 &   53.1 &   0.7 \\
+	J1024-0719 & 193.7 &       -7.0 & 156.2 &   -7.3 &   1.2 \\
+	J1125+7819 & 238.0 &       -3.9 & 171.5 &   78.3 &   0.9 \\
+	J1453+1902 & 172.6 &       -3.5 & 223.4 &   19.0 &   1.3 \\
+	J1455-3330 & 125.2 &       -3.8 & 223.9 &  -33.5 &   0.7 \\
+	J1600-3053 & 277.9 &       -7.3 & 240.2 &  -30.9 &   1.9 \\
+	J1614-2230 & 317.4 &       -9.7 & 243.7 &  -22.5 &   0.7 \\
+	J1640+2224 & 316.1 &       -2.8 & 250.1 &   22.4 &   1.5 \\
+	J1643-1224 & 216.4 &       -8.6 & 250.9 &  -12.4 &   0.7 \\
+	J1713+0747 & 218.8 &       -4.1 & 258.5 &    7.8 &   1.3 \\
+	J1738+0333 & 170.9 &       -7.0 & 264.7 &    3.6 &   1.5 \\
+	J1741+1351 & 266.9 &      -21.5 & 265.4 &   13.9 &   1.7 \\
+	J1744-1134 & 245.4 &       -5.4 & 266.1 &  -11.6 &   0.4 \\
+	J1747-4036 & 607.7 &      -48.5 & 267.0 &  -40.6 &   7.1 \\
+	J1832-0836 & 367.8 &      -11.2 & 278.1 &   -8.6 &   2.1 \\
+	J1853+1303 & 244.4 &       -5.2 & 283.5 &   13.1 &   2.1 \\
+	J1857+0943 & 186.5 &       -6.2 & 284.4 &    9.7 &   1.2 \\
+	J1903+0327 & 465.1 &      -40.7 & 285.8 &    3.5 &   7.0 \\
+	J1909-3744 & 339.3 &      -16.1 & 287.4 &  -37.7 &   1.1 \\
+	J1910+1256 & 200.7 &       -3.9 & 287.5 &   12.9 &   1.5 \\
+	J1911+1347 & 216.2 &       -7.9 & 288.0 &   13.8 &   1.4 \\
+	J1918-0642 & 130.8 &       -4.4 & 289.7 &   -6.7 &   1.1 \\
+	J1923+2515 & 264.0 &       -6.7 & 290.8 &   25.3 &   1.2 \\
+	J1939+2134 & 641.9 &     -433.1 & 294.9 &   21.6 &   3.5 \\
+	J1944+0907 & 192.9 &       -6.4 & 296.0 &    9.1 &   1.2 \\
+	J1946+3417 & 315.4 &       -3.1 & 296.6 &   34.3 &   6.9 \\
+	J1955+2908 & 163.0 &       -7.9 & 298.9 &   29.1 &   6.3 \\
+	J2010-1323 & 191.5 &       -1.8 & 302.7 &  -13.4 &   2.4 \\
+	J2017+0603 & 345.3 &       -9.5 & 304.3 &    6.1 &   1.4 \\
+	J2033+1734 & 168.1 &       -3.1 & 308.4 &   17.6 &   1.7 \\
+	J2043+1711 & 420.2 &       -9.3 & 310.8 &   17.2 &   1.4 \\
+	J2145-0750 &  62.3 &       -1.2 & 326.5 &   -7.8 &   0.7 \\
+	J2214+3000 & 320.6 &      -15.1 & 333.7 &   30.0 &   0.6 \\
+	J2229+2643 & 335.8 &       -1.7 & 337.5 &   26.7 &   1.8 \\
+	J2234+0611 & 279.6 &       -9.4 & 338.6 &    6.2 &   1.0 \\
+	J2234+0944 & 275.7 &      -15.3 & 338.7 &    9.7 &   1.6 \\
+	J2302+4442 & 192.6 &       -5.1 & 345.7 &   44.7 &   0.9 \\
+	J2317+1439 & 290.3 &       -2.0 & 349.3 &   14.7 &   1.7 \\
+	J2322+2057 & 208.0 &       -4.2 & 350.6 &   21.0 &   1.0 \\
+	\bottomrule
+\end{tabular}
+\caption{List of the fiducial pulsar parameters for the 47 pulsars used to construct the synthetic PTA in Section \ref{sec:synt_pta}.
+The right ascension and declination of an individual pulsar in J2000 coordinates are labelled by $\alpha$ and $\delta$ respectively.
+Parameter values are obtained from the ATNF pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.}
+\label{tab:pulsar_values}
+\end{table*}
+
+\section{Dispersion of ${\theta}_{\rm gw}$ estimates}\label{sec:wasserstein}
+Every random realization of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$ leads to a different ${\boldsymbol{\theta}}_{\rm gw}$ posterior, when the synthetic data are analysed according to the procedure in Section \ref{sec:methodsummary}.
+This cosmic variance is unavoidable.
+It is a key source of uncertainty, when the analysis is performed on real astronomical data, where the specific random realizations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$ are unknown.
+To quantify the cosmic variance, one needs a metric to describe the distance between two posteriors (i.e.\ how similar they are).
+Many valid metrics of this sort exist, including those related to the Kolmogorov-Smirnoff test \citep{corder2014nonparametric}.
+In this paper, we use the WD \citep{Wasserstein,Villani2009} which is popular in machine learning \citep[e.g.][]{2017arXiv170107875A} and other domains.
+In Section \ref{sec:appendix_overview_WD} we define the WD and summarize its main properties.
+In Section \ref{sec:appendix_WD_results} we present for reproducibility the WD calculated between every pair of posteriors for each static parameter in $\boldsymbol{\theta}_{\rm gw}$ across the $10^3$ noise realisations of Section  \ref{sec:multiple_noise}.
+The results are summarised via the median values reported in Table \ref{tab:Wasserstein}.
+
+\subsection{Overview of the WD}\label{sec:appendix_overview_WD}
+The WD is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
+It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$.
+For this reason it is sometimes known as the ``Earth mover's distance''.
+The $p$-th order Wasserstein distance between two distributions is
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left[ \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \lambda (x,y)\right]^{1/p} \label{eq:wasserstein} \ ,
+\end{eqnarray}
+for $p \in [1,\infty)$, where $\lambda(x,y)$ is the transport plan, and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
+$\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
+\newline
+
+The cost function can be freely chosen to suit the nature of the problem.
+Often, as in this paper, it is taken to be the absolute value function, $c(x,y) = |x-y|$.
+In general, $W_p(\mu,\nu)$ with $c(x,y)=|x-y|$ can be computed from $n$ samples by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
+However, for $\mu$ and $\nu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that $E[W_p(\mu,\nu)]$ converges slowly $\propto n^{-1/d}$.
+In this paper, we calculate the WD exclusively between the one-dimensional marginalized posteriors, setting $d=1$ and obtaining
+\begin{eqnarray}
+	W_p(\mu,\nu)= \left\{\int_0^1 dz \left[ F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) \right] \right\}^{1/p} \ , \label{eq:wasserstein}
+\end{eqnarray}
+where $F_{\mu}^{-1}(z)$ is the inverse cumulative distribution function of $\mu$. \newline
+
+The WD holds certain advantages over valid alternatives like the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test \citep{gelman2013bayesian,corder2014nonparametric}.
+sIt is intuitive, being the minimum cost required to transform one distribution into another.
+It satisfies the Monge-Kantorovich duality \citep{villani2003topics,Villani2009},
+\begin{eqnarray}
+	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
+\end{eqnarray}
+where $X_{\mu}$ and $Y_{\nu}$ are random variates drawn from the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
+It also satisfies desirable properties of a measure of distance, such as symmetry and the triangle inequality.
+For $p=1$, the WD inherits the units of the underlying distributions.
+Finally, the WD is versatile; any two distributions can be compared, irrespective of whether they are continuous, discrete, or singular.
+\newline
+
+\section{pulsar terms stuff}
+
+\subsection{Challenges of pulsar term}\label{sec:pulsar_terms_challenge}
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =0.5\textwidth ]{example-image-a}
+	\caption{Logarithm of the one-dimensional likelihood $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
+All other static parameters of $\boldsymbol{\theta}$ are held constant at their true injected values.
+In the left-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}(\alpha)$, i.e.
+inclusive of the pulsar terms.
+In the right-hand panel the Kalman filter uses Equation \ref{eq:g_func_trig_earth}, i.e.
+just the Earth terms.
+The vertical dashed lines indicate the true injected value of $\alpha$ ($=1.0$ rad) used to generate $\boldsymbol{Y}$.
+The inset plot covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
+In the left hand panel $\mathcal{L}(\alpha)$ is very noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
+This noise hampers the ability of sampling algorithms to converge to a posterior distribution.
+Conversely, in the right hand panel $\mathcal{L}(\alpha)$ is globally smooth.
+}
+	\label{fig:likelihood_comparison}
+\end{figure*}
+
+To elucidate the complications introduced by including the pulsar term, we generate some noisy synthetic data $\boldsymbol{Y}$ using the methods outlined in K23.
+$\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
+The choice of $\boldsymbol{\theta}_{\rm psr}$ corresponds to the 47 pulsars in the 12.5 year NANOGrav data set  \citep{2020ApJ...905L..34A}.
+We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig} on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad.
+We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve).
+The dashed grey line indicates the injected value of $\alpha$ (= 1.0 rad).
+The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
+\newline
+
+There are two important features to discuss in the left-hand panel of Figure \ref{fig:likelihood_comparison}.
+The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
+Similar noisy curves are obtained for $\mathcal{L}(\Omega)$ and $\mathcal{L}(\delta)$.
+For all other static parameters, the $\mathcal{L}(\Theta)$ curves are smooth and globally concave.
+The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.
+This means that whilst it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$, in practice the computation is intractable since the noise in the likelihood curve inhibits sampling algorithms to reliably converge to a definitive posterior.
+Moreover, $\mathcal{L}(\alpha)$ is only a one-dimensional function.
+The challenge for sampling algorithms is further exacerbated when we consider the three-dimensional parameter space in $\alpha - \delta - \Omega$, or the full $7 + 5N$ dimensions of Section \ref{sec:ss_params}.
+\newline
+
+Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\Theta)$ curves whereas other static parameters do not?
+These parameters appear in the pulsar term as a product in the pulsar-dependent phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, where we remind the reader that $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
+Over the typical pulsar distances $d^{(n)} \sim 1$ kpc, a 10 nHz GW can accumulate $10^3$ cycles.
+This creates a confusion problem when taking the cosine of  $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ since we do not know how many cycles the wave has gone through, modulo $2 \pi$.
+\textcolor{red}{TK: this explanation is rubbish!
+How can we say this better?}.
+Conversely, if we drop the pulsar term,  the $\mathcal{L}(\alpha)$ curve (as well as $\mathcal{L}(\Omega)$ , $\mathcal{L}(\delta)$   )
+This is shown by the orange curve in the right-hand panel of Figure \ref{fig:likelihood_comparison} where we plot $\mathcal{L}(\alpha)$, analogous to the left-hand panel, but now use Equation \ref{eq:g_func_trig_earth} in our Kalman filter (i.e.
+just the Earth terms) rather than the full Equation \ref{eq:g_func_trig}.
+In this case there is no confusion regarding the number of cycles the wave has gone through over a distance $d^{(n)}$ and the curve is smooth.
+
+\section{KF and NS review}
 \section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
-
-In K1
-
-The workflow presented in K1, hich we also follow here proceeds as follows
+The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters controlling the intrinsic frequency evolution of the pulsars in the array and the GW source, i.e.
+\begin{equation}
+	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
+\end{equation}
+with
+\begin{equation}
+	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
+\end{equation}
+and
+\begin{equation}
+	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
+\end{equation}
+where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively \footnote{$\iota$ is the angle between the unit normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
+$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.}.
+In Equation \eqref{eq:params3} we reparameterize the two GW polarisation amplitudes, $h_{+}$ and $h_{\times}$, in terms of $\iota$ and the characteristic wave strain $h_0$ through
+\begin{align}
+	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
+	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ .
+\end{align}
+We use the parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
+A PTA containing $N$ pulsars comprises $7 + 5N$ parameters to estimate.
+Typically the pulsar parameters are constrained better \textit{a priori} by electromagnetic observations than the GW parameters.
+For example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information about $\delta$ and $\alpha$.
+\newline
 
@@
 A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
-The method complements traditional PTA analyses because it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations.
+The method complements traditional PTA analyses because (i) it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations; and (ii) it infers ${\boldsymbol{\theta}}$ conditional on the specific, time-ordered, random realization of the noise corresponding to the observed sequence of TOAs rather than on ensemble-averaged quantities like the power spectral density.
 
 \subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
-%\begin{figure*}
-%	%\centering % Not needed
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/Kalman_example_true_params_single}
-%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
-%		\label{fig:6MB_BFS}
-%	\end{subfigure}  
-%    \hfill
-%	\begin{subfigure}[b]{0.49\textwidth}
-%		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
-%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
-%		\label{fig:25MB_bfs}
-%	\end{subfigure}
-%	\caption{
-%		}
-%	\label{fig:four figures}
-%\end{figure*}
-
-The Kalman filter \citep{Kalman1} is
-
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
+	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column).
+The top panels show the true pulsar state $f_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$  (green curve).
+We subtract $f_{\rm em}^{(1)}(t)$ to better illustrate the stochastic wandering of the pulsar frequency.
+In the left column, the blue/green solutions overlap almost perfectly; in the right column, they do not.
+The middle panels show the true measured frequency $f_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve).
+Again we subtract $f_{\rm em}^{(1)}(t)$.
+In the left column the red/magenta solutions overlap almost perfectly; in the right column, they do not.
+The bottom panels show the residual or innovation $\epsilon(t) =f_{\rm m}^{(1)}(t) - \hat{f}_{\rm m}^{(1)}(t)$.
+In the right column $\Omega$ is displaced from its true value by 20\%.
+Results are shown for a single pulsar.
+}
+	\label{fig:kalman_example}
+\end{figure*}
+
+The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$.
+It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
 In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
@@
 The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel).
-By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters to the Kalman filter, where $\Omega$ has been perturbed from its true value by $20 \%$.
+By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$) to the Kalman filter, where $\Omega$ is displaced from its true value by $20 \%$.
 In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates.
 It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel).
-
-%n Figure ; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity.
-The residuals correspond to random noise and are normally distributed.
-Conversely, if $\boldsymbol{\hat{\theta}}$ is not close to the true parameters then the filter is unable to recover the state evolution.
-This is demonstrated in Figure \ref{fig:25MB_bfs} where the Kalman filter was run with $\boldsymbol{\hat{\theta}}$ slightly perturbed away from the true values.
-In this case the filter cannot track the state variable accurately and the residuals are no longer Gaussian.
+In Section \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$.
 
 \subsection{Nested sampling}\label{sec:nested_sampling}
-We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of
-
+We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
+\begin{equation}
+	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
+\end{equation}
+where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
+\begin{equation}
+	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
+\end{equation}
 We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper.
-Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, which also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}.
+It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
 It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
@@
 	\begin{enumerate}[leftmargin=2em]
-	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
+		\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
 		\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
@@
 	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
-	\item Constrained sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood $\log \mathcal{L}_{\rm new}$ > $\log \mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (a)--(b).
-	\item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ 
-	\item Repeat steps (vii) - (ix) until convergence criteria are satisfied.
+	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a)--(vi)(b).
+	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
+	\item Repeat steps (vii)--(ix) until convergence criteria are satisfied.
 \end{enumerate}
@@
 The resulting $\mathcal{Z}$ values can then be compared.
-We remind the reader that the above workflow differs from a realistic PTA analysis in on important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
+We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
 The generalization to TOAs is subtle and will be tackled in a forthcoming paper.
 
-\section{Validation procedure with synthetic data} \label{sec:testing}
-In this section we outline how synthetic data is generated in order to validate the analysis scheme in Section \ref{sec:detect}.
-In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
-In Section \ref{sec:gendata} we demonstrate how to solve the equations of Section \ref{sec:model} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.
-
-\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
-
-\section{Representative example: single SMBHB source}\label{sec:rep_example}
-In this Section we apply the analysis scheme of Section \ref{sec:detect} and the validation procedure of Section \ref{sec:testing} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-We remind the reader that the analysis in this paper targets a single SMBHB source.
-The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-In Section \ref{sec:priors} we outline the priors chosen on $\boldsymbol{\theta}$.
-In Section \ref{sec:parameter_estim} we apply the workflow of Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
-We go on in Section \ref{sec:multiple_noise} to extend this parameter estimation procedure to multiple realisations of the noise processes.
-In Section \ref{sec:detection} we calculate the the detectability of the source as a function of $h_0$.
-\newline
-
-\subsection{Prior distributions}\label{sec:priors}
-
-%\begin{figure*}
-%	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
-%	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
-The vertical orange lines indicate the true injected values.
-The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-We are able to accurately estimate each parameter of interest.
-We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
-%	\label{fig:corner_plot_1}
-%\end{figure*}
-
-For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate $W_1$ between every pair of posteriors across our $10^3$ noise realisations.
-The results are presented in Figure \ref{fig:pairwise_wasserstein}.
-The figure shows a lower triangular heat-map where each point denotes the $W_1$ value between a pair of posteriors.
-Small values of $W_1$ are magenta moving to red for large values of $W_1$.
-Note that the heat-map colour scaling is not the same for all subplots of the figure.
-It is evident that $W_1$ is generally small across all pairs of posteriors.
-We define $W_{1, \rm median}$ as the median value of $W_1$ across all $10^3 \choose 2$ pairs of noise realisations.
-The values of $W_{1, \rm median}$ for each parameter are quoted in Table \ref{tab:Wasserstein}.
-As a percentage of the width of the prior space (e.g.
-for $\alpha$ the width of the prior is $2 \pi$), the median $W_1$ for each parameter is $\lesssim 3 \%$.
-Generally speaking, the method converges to consistent one-dimensional posteriors for each parameter across different noise realisations.
-\newline
-
-There are a small fraction of the total number of noise realisations which have a large Wasserstein distance from all the other posteriors of all the other noise realisations.
-These are the strong yellow lines in Figure \ref{fig:pairwise_wasserstein}.
-These correspond exclusively to posteriors which have converged poorly and not found the likelihood global maximum.
-For instance, the top strong horizontal yellow line in Figure \ref{figWD_psi}, which is identified by an asterisk in the figure, corresponds to a posterior for $\psi$ which has converged sub-optimally, with an estimated value of $\psi = 1.01 \pm 0.06$ rad, highly diverged from the true injected value of $2.50$ rad.
-The $W_1$ value between this posterior and all the posteriors inferred for all other noise realisations is $\sim 1.7$ radians.
-We also observe a strong correlation across all noise pairs (Pearson correlation coefficient $=0.96$) between the $W_1$ inferred for $\alpha$ and $\psi$, indicating that noise realisations which corrupt the inference of $\psi$ similarly affect $\alpha$.
-This is due to multimodal likelihood surface for $\psi-\alpha$, shown in Figure \ref{fig:likelihood_surface_alpha_psi}.
-We can see that the likelihood surface has a global optima at the approximate location of the injected values ($\alpha \sim 1.0$, $\psi\sim2.5$) and local optima that corresponds to the location of the example poorly converged posterior discussed previously, with $\psi = 1.01$ rad.
-This problem of poor global convergence can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
-
-\subsection{Detection} \label{sec:detection}
-We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-$\mathcal{M}_1$ is the Earth-terms only model, i.e.
-the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
-The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
-The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
-We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
-We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
-For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
-This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
-$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
-\newline
-
-There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
-The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
-The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
-For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
-That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
-For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
-For the results presented in this subsection we set $n_{\rm live} =2000$.
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
-	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}.
-The horizontal grey dashed line labels the detection threshold of $\beta = 10$.
-The minimal detectable strain, below which $\beta < 10$, is $\sim 3.9 \times 10^{-15}$.
-Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$.}
-	\label{fig:bayes}
-\end{figure}
-
-\section{Exploring a broader parameter space} \label{sec:parameter_space}
+\section{Bias from dropping the pulsar terms}\label{sec:bias_appendix}
+Why does dropping the pulsar terms in Equation \ref{eq:g_func_trig} bias some components of $\boldsymbol{\theta}_{\rm gw}$, whilst leaving others nearly unaffected?
+For instance, why is the bias in $\iota$ strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
+In this appendix, we examine the above questions in the context of the Kalman analysis in this paper, complementing work on similar questions in the context of traditional PTA analyses \citep{Zhupulsarterms,Chen2022}.
+\newline
+
+ When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is modulated by the interference of two equal-amplitude, phase-shifted oscillations (c.f.
+Equation \eqref{eq:measurement}).
+Specifically we have an
+``Earth-term oscillation'' of form $y_{\rm Earth} = A \cos \xi$, with amplitude $A$ and phase $\xi$, and a ``pulsar-terms oscillation'' of form $y_{\rm pulsar} = A \cos (\xi + \kappa)$, with phase offset $\kappa$.
+The two terms are related to the measurement equation, Equation \eqref{eq:g_func_trig}, through
+\begin{align}
+	1 - g^{(n)}(t) &=  y_{\rm Earth}  - y_{\rm pulsar} \  \nonumber \\
+		 &= 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ . \label{eq:psr_interference}
+\end{align}
+The interference in Equation \ref{eq:psr_interference} multiplies the amplitude by $ 2 \sin (\kappa/2)$, but leaves the frequency unchanged.
+When we model the data using the Kalman filter without the pulsar terms, $y_{\rm Earth}$ must try to compensate for the modified amplitude on the right-hand side of Equation \ref{eq:psr_interference} by adjusting the parameters which govern the amplitude, such as $h_0$ and $\iota$.
+Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
+The wave amplitude is independent of $\Omega$, and so $\Omega$ exhibits no bias.
+A shift in phase is covariant with the amplitude, so $\Phi_0$ also exhibits a small bias.
+The same hierarchy of biases is seen also in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are over-constrained whilst $\Omega$ is well-estimated.
+\newline
+
+Quantifying the bias for practical, astronomical PTA continuous wave searches requires a thorough exploration of the SMBHB parameter space which falls outside the scope of this introductory paper.
+The bias also depends on the PTA configuration.
+If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate.
+This is what we observe for the synthetic data, for example in Figure \ref{fig:corner_plot_2}; the marginalised posterior broadens as the SNR decreases, until it overwhelms the shift due to the bias.
+Conversely if the bias is sufficiently strong, or the source is sufficiently loud, then the bias may dominate.
+\newline
+
+%%%%%%%%%%%%%%%
+
+\begin{figure*}
+	\setkeys{Gin}{width=\linewidth}
+
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/WD_0}
+		\caption{$\Omega$}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/WD_1}
+		\caption{$\Phi_0$}
+	\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/WD_2}
+		\caption{$\psi$} \label{figWD_psi}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{images/WD_3}
+		\caption{$\iota$}
+	\end{subfigure}
+	\medskip
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/WD_4}
+		\caption{$\delta$}
+	\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/WD_5}
+		\caption{$\alpha$}
+	\end{subfigure}
+	\hfill	
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{images/WD_6}
+		\caption{$h$}
+	\end{subfigure}
+	\caption{First moment of the WD, $W_1$ (colour scale), calculated between each pair of one-dimensional posteriors for the representative system in Table \ref{tab:parameters_and_priors} across $10^3$ realisations as discussed in Section \ref{sec:appendix_WD_results}, for each component of $\boldsymbol{\theta}_{\rm gw}$.
+$W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
+$W_1$ is generally low across all parameters and posteriors, but some posteriors converge poorly and return high values of $W_1$.
+The $W_1$ values inferred for $\psi$ and $\alpha$ are correlated with Pearson correlation coefficient = 0.96.
+An example of a noise realisation with particularly high $W_1$ is labelled with an asterisk in (c) and is discussed in the text.} \label{fig:pairwise_wasserstein}
+\end{figure*}
 \begin{figure}
 	\centering
-	\includegraphics[width=\columnwidth]{example-image-a}
-	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}.
-Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$.
-The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
-The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals.
-Well estimated posteriors should fall along the diagonal $y=x$.
-$\Omega$ is generally well-estimated but many other parameters show evidence of being over-constrained due to a modelling bias.
-}
-	\label{fig:parameter_space}
+	\includegraphics[width=\columnwidth]{images/likelihood_surface_alpha_psi}
+	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\alpha$-$\psi$ plane for a single noise realisation.
+The red point labels the maximum, and the magenta point labels the median values of the inferred one-dimensional posteriors for $\alpha$ and $\psi$.
+The likelihood is normalised with respect to the absolute value of the maximum; note the compressed colour scale.
+A multi-modal structure is clearly visible for this noise realisation.
+The nested sampler finds a local maximum (magenta point) instead of the global maximum (red point)}
+	\label{fig:likelihood_surface_alpha_psi}
 \end{figure}
-We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
-In this section we test the method in different regions of the parameter space, varying the source parameters through astrophysically relevant ranges.
-\newline
-
-We consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
-We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space.
-For each simulated injection we can then attempt to recover the posterior distributions for each parameter.
-To summarise the results across the parameter space we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
-A PP plot describes the fraction of the total number of injected parameters which are included within a given credible interval of the estimated posterior with respect to the credible interval itself.
-In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage credible interval.
-The results are shown in Figure \ref{fig:parameter_space}.
-The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections.
-We can see that only $\Omega$ falls within the $3\sigma$ shaded region.
-The other parameters deviate from the $y=x$ diagonal.
-The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$.
-The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and there are more injections contained within the low value credible intervals.
-This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally confident, but are systematically biased away from the true injection values.
-This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval.
-An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.
-
-\section{Identifiability and bias} \label{sec:bias_and_identifiability}
-There are two underlying issues in the preceding tests that we now explore in more detail.
-The first relates to parameter identifiability, i.e.
-can we uniquely identify the parameters of the model given the data and the model structure?
-This question is explored in Section \ref{sec:identif}.
-The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
-This is the focus of Section \ref{sec:bias}.
-In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f.
-Figure \ref{fig:bayes}) for our representative example system (Table \ref{tab:parameters_and_priors}).
-\footnote{\tiny \textcolor{red}{TK: This discussion on bias and identifiability is perhaps less relevant than it was when our representative example had a 'high' strain ($10^{-12}$).
-However it IS still a feature/bug of our approach and I think it is worth discussing in the main body rather than relegating to an appendix.
-Opinions welcome...} \normalsize}
-
-\subsection{Bias by dropping pulsar terms}\label{sec:bias}
-
-Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g.
-the median of the one-dimensional marginalised posterior) is biased away from the true injection value.
-In Figure \ref{fig:bias_for_large_h} we show the resulting posteriors for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but now for the high-SNR system with $h_0 = 10^{-12}$.
-We can see that for all parameters, with the exception of $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it is not visible on the scales over which we have plotted the posteriors.
-This effect is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
-Similar results are obtained for other noise realisations.
-Whilst $h_0$ does not exhibit any noticeable bias for this particular noise realisation, in general across multiple noise realisations it does exhibit a correlated bias with $\iota$.
-In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
-\newline
-
-In Figure \ref{fig:likelihood_surface_iota} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, as we vary $\iota$ across the prior space, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
-We do this for 10 separate realisations of the system noise.
-We consider three separate situations: \textit{(i)} The solid blue line is the likelihood curve with just the Earth-terms included (i.e.
-using Equation \ref{eq:measuremen_earth} in the Kalman filter), \textit{(ii)} the solid orange line is the likelihood curve with the additional pulsar terms included (i.e.
-using Equation \ref{eq:measurement}), \textit{(iii)} the solid green line is the solution where we drop the pulsar terms from both the synthetic data and the measurement equation.
-We have not used nested sampling for situations \textit{(ii)} or \textit{(iii)} in this work.
-However, we are able to run the Kalman filter either using the pulsar terms in the model or else dropping them from the synthetic data consider how the resulting likelihood varies for different parameter values.
-The dashed lines show the location of the likelihood maxima for each of the likelihood solutions; finding the location of the maxima across all static parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
-\newline
-
-The key observation from Figure \ref{fig:likelihood_surface_iota} is that for the Earth-terms only solution the location of the likelihood optima and the location of the injected value are not the same.
-Consequently Bayesian likelihood inference techniques settle in optima that do not necessarily align with the true injection value.
-This is the origin of the bias in the one-dimensional marginalised posteriors that we observe.
-For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ radians, whereas the injected value was $\iota = 1.0$ radians.
-This optima is comparable to the median of the posterior inferred for $\iota$ in Figure \ref{fig:bias_for_large_h}.
-\newline
-
-In contrast to the Earth-terms solution, for the pulsar-terms solution the optima of the likelihood curve now aligns with the true injected value.
-The corollary is also true; if we drop the pulsar terms from the synthetic data and use the Earth terms equation for inference then the optima is also correctly located.
-This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.
-Similar results are observed for the other parameters which exhibit a bias in the nested sampling inference.
-\newline
-
-Why does dropping the pulsar terms bias these parameters in particular, whilst leaving others unaffected?
-For instance, why is the bias induced in $\iota$ so strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
-When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is effective modulated by the interference of two equal-amplitude, phase shifted waves (c.f.
-Equation \ref{eq:measurement}).
-Explicitly, we have the ``Earth-terms wave"	 with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
-The net interference wave that modulates $f_{\rm p}$ is
-\begin{eqnarray}
-	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ .
-\end{eqnarray}
-The additional modulations from $y_{\rm pulsar}$ influence the amplitude of the resulting interference wave, but do not affect its frequency.
-When we model this data using the Kalman filter without the pulsar terms, the additional amplitude modulations from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
-As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude of the single wave.
-Consequently the parameters which are most affected are those which determine the amplitude of the GW.
-Terms which influence the amplitude strongly, such as $h_0$ and $\iota$, are strongly affected and exhibit a strong bias.
-Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
-The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
-Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
-This hierarchy of biases for each parameter is also what we observe in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
-\newline
-
-In order to ascertain how important the bias is for PTA continuous wave searches requires a thorough exploration of the astrophysical parameter space which we do not carry out in this work.
-The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
-If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate and the bias may not be important for searches.
-This is what we observe for our synthetic data (e.g.
-Figure \ref{fig:corner_plot_2}); the decreased SNR manifests as a broader marginalised posterior which generally overwhelms the shift due to the bias.
-Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may come to systematically affect the inferred parameters.
-\newline
-
-We have shown in this section how the dropping of the pulsar terms results in a bias in the inferred parameters, not just in the sky localisation parameters  \citep[e.g.][]{Zhupulsarterms,Chen2022} but additionally all parameters which influence the wave amplitude.
-Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for our example system, it is present and may need to be accounted for for analysis in other parts of the parameter space.
-This bias is not particular to our method but a shared feature across likelihood based methods that do not include the pulsar terms in the inference model.
-The inclusion of the pulsar terms in the methods presented in this work in order to alleviate the bias will be a key goal of a future work.
-
-\section{Discussion}\label{sec:discussion}
-
-Whilst computational methods for approximate inference from noisy likelihoods do exist \citep[e.g.][]{doi:10.1137/21M1410853,2021arXiv210403942J,2019arXiv190501252J},
-
-We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.
-There are a number of potential extensions of this work which we now discuss.
-\newline
-
-The natural first extension is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.
-Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred one-dimensional marginalized posteriors for the model parameters.
-Biases in sky localisation for synthetic datasets as a result of dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}.
-Alternatively, if the method continues to be used with solely the Earth terms then it would be highly desirable to make systematic, quantitative estimates of the incurred biases in the model parameters across an astrophysically representative parameter space.
-\newline
-
-We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
-Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA.
-Indeed, since adding pulsars to PTAs increases the computational demands of the data analysis it may be advantageous to optimally select a subset of the total available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.
-\newline
-
-Throughout this work we have assumed that all pulsars are observed at the same time, with a constant time sampling.
-In practice this assumption does not hold and different pulsars will be observed with different cadences at different times.
-Extending this method to non-constant time sampling is straightforward and the performance of the method should be evaluated in this regime.
-\newline
-
-We have taken our GW source to be monochromatic, i.e.
-$\Omega$ is a fixed parameter with no time dependence.
-Whilst this is an astronomically well-justified assumption (see discussion in Section \ref{sec:plane_gw}), and appropriate for this initial methods work, it would also be of interest to extend the state-space construction to enable the GW frequency to evolve in time.
-The GW source is monochromatic over the decadal PTA observation period, but the timescale set by the Earth-pulsar distance is comparable to the SMBHB evolution timescale \citep{Sesana2010}.
-For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) the GW source frequency evolution manifests as an incoherent source of noise in the pulsar terms, whilst for  $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms can affect the phase coherency of the Earth terms \citep{Perrodin2018}.
-Careful consideration of the evolution of the GW frequency will be important as we look to extend our method to include the pulsar terms in the inference model.
-\newline
-
-We have also assumed that there is only a single GW source that influences the received pulsar frequencies.
-However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
-Our method naturally extends to detecting multiple GW sources simultaneously; Equation \eqref{eq:measurement} is straightforwardly modified to become a linear superposition of multiple GWs.
-The stochastic background itself is the incoherent summation of many individual GW sources.
-It therefore stands to reason that it may be possible to extend this method to also search for a stochastic GW background, providing a complementary approach to detect the stochastic background without requiring a cross-correlation of the pulsar residuals to uncover the Hellings-Downs curve.
-\newline
-
-\textcolor{red}{TK: Lets also pre-empt reviewer and make it explicit why we are using frequencies rather than TOAs. Need to chat with group to get this clear in my head.}
-
-%https://arxiv.org/pdf/2107.03047.pdf
-\section{Conclusion}\label{sec:conclusion}
-In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
-We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
-That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW induced modulations.
-This enables the inference of GW parameters conditioned on the specific observed realisation of the noisy data.
-We introduce the Kalman filter in order to track the intrinsic state evolution of the pulsar and combine this with a Bayesian sampling framework to estimate the posterior distributions of each parameter of the system, as well as the associated evidence (marginalised likelihood) of the model.
-We test our method on synthetic data.
-We initially focus on a single, astrophysically representative, SMBHB GW source detected with the 12.5-year NANOGrav pulsars observed over a 10 year timespan.
-The parameters of the model are accurately recovered and the minimal detectable strain is estimated to be $h_0 \sim 4 \times 10^{-15}$ for $\iota=1.0$ rad.
-The method is then further tested over 1000 noise realisations.
-Consistent posterior estimates of the parameters are obtained for the majority of noise realisations, with the median Wasserstein distance across all noise realisations $W_{1, \rm median}$ generally small compared to the prior space ($\lesssim 3 \%$).
-For a minority of noise realisations the nested sampler converges poorly and settles into a local likelihood optima for the posteriors estimates of $\psi$ and $\alpha$.
-This problem can be solved in general by increasing the number of live points $n_{\rm live}$ used by the sampler.
-Exploration of a broader parameter space via 200 randomly sampled parameter vectors highlights biases in some parameters as a result of dropping the pulsar terms from the inference model.
-The cause of these biases is examined for the case of high SNR, along with the structural weak-identifiability between $\iota$ and $h_0$.
-The extension of this method to include the pulsar term is a key challenge for a future work.
-
-\appendix
-
-\section{Kalman recursion equations} \label{sec:kalman}
-The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
-\begin{equation}
-	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
-\end{equation}
-where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance, and the subscript $k$ labels the time-step.
-The Kalman filter takes the underlying states to evolve according to
-\begin{equation}
-	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
-\end{equation}
-where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix.
-$\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$.
-\newline
-
-The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage.
-The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
-Specifically, the predict step proceeds as,
-\begin{equation}
-	\hat{\boldsymbol{X}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k
-\end{equation}
-\begin{equation}
-	\hat{\boldsymbol{P}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k 
-\end{equation}
-Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
-\begin{equation}
-	\boldsymbol{\epsilon}_{k} = \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1}
-\end{equation} 
-\begin{equation}
-	\boldsymbol{S}_k = \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k
-\end{equation}
-\begin{equation}
-	\boldsymbol{K}_k = \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \label{eq:kalman gain}
-\end{equation}
-\begin{equation}
-	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
-\end{equation}
-\begin{equation}
-		\hat{\boldsymbol{P}}_{k|k} = \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1}
-\end{equation}
-Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
-$\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$.
-For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
-\newline
-
-We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
-\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
-\begin{equation}
-	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
-\end{equation}
-Similarly,  
-\begin{equation}
-\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
-\end{equation}
-The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
-\begin{equation}
-	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
-\end{equation}
-where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix
-\begin{equation}
-\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
-\end{equation}
-and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
-\begin{equation}
- C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
-\end{equation}
-The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
-\begin{equation}
-\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)
-\end{equation}
-\newline
-
-The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
-\begin{equation}
-	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \label{eq:gardenier}
-\end{equation} 
-From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
-\begin{equation}
-\boldsymbol{F}_k = e^{\boldsymbol{A}\left( t_{k+1} - t_k \right)} 
-\end{equation}
-\begin{equation}
-	\boldsymbol{G}_k = \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' 
-\end{equation}
-\begin{equation}
-	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t')
-	\end{equation}
-For our system, the state, control and process noise covariance matrices then take the form:
-\begin{equation}
-	\boldsymbol{F}_k = 	\text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right)
-\end{equation}
-\begin{equation}
-\boldsymbol{G}_k	= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right)
-\end{equation}
-with
-\begin{align}
-	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
-	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
-\end{align}
-and
-\begin{equation}
-	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) 
-\end{equation}
-for 
-\begin{equation}
-	Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right)
-\end{equation}
-where $\Delta t$ is the constant time sampling interval $=t_{k+1} -t_{k}$. \newline
-
-The remaining component matrices of the Kalman filter that we have not yet specified are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
-These are defined straightforwardly from Equation \eqref{eq:measurement}.
-Specifically,
-$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
-The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
-
-\section{Fiducial pulsar values}\label{appendix_fiducial}
-\textcolor{red}{TK: Do we really need this?
-The argument is for reproducibility, but we have a nice github repo that holds all the data we actually used, jupyter notebooks that show how the calculations are done.
-Perhaps it is sufficient to point the reader to the repo...
-}
-\section{Identifiability}\label{appendix_identifiability}
-\textcolor{red}{TK TBD. An example calculation for a simple system using the methods of e.g. } \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
-
-\section{Wasserstein distance}\label{sec:wasserstein}
-The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
-It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$.
-For this reason it is sometimes known as the "Earth mover's distance".
-The $p$-th order Wasserstein distance between two distributions is
-\begin{eqnarray}
-	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
-\end{eqnarray}
-where $p \in [1,\infty)$ and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
-$\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
-\newline
-
-The cost function can be freely chosen for the nature of the problem, but is commonly taken to be the absolute value function, $c(x,y) = |x-y|$, and we adopt the same cost function in this work.
-Given this choice of $c(x,y)$, for an empirical distribution $\hat{\mu}_n$ composed of $n$ i.i.d samples drawn from the underlying probability distribution $\mu$, the solution to the discrete form of Equation \eqref{eq:wasserstein} can be computed generally by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
-However, for $\mu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
-\begin{eqnarray}
-	\boldsymbol{E} [W_1(\hat{\mu}_n,\mu)] \asymp n^{-\frac{1}{d}} \ .
-\end{eqnarray}
-In order to circumvent this issue, in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors, i.e.
-setting $d=1$.
-In this case Equation \eqref{eq:wasserstein} has a convenient, easily computed, closed form:
-\begin{eqnarray}
-	W_p(\mu,\nu)= \left(\int_0^1  F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) dz \right)^{1/p} \, \label{eq:wasserstein}
-\end{eqnarray}
-where $F_{\mu}(z)$ is the cumulative distribution function of $\mu$. \newline
-
-The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages.
-Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another.
-It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g.
-radians then we can interpret this as the cost to transform between the two distributions is 1 radian.
-Further interpretability is provided by the Monge-Kantarovih duality, Equation \eqref{eq:WDdefn}; the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian.
-Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g.
-Gaussianity.
-Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces.
-Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality.
-It also respects the geometry of the underlying metric space over which the distributions are defined.
-
-%\section{References}
-%\label{sec:ref_list}
-
-\bibliographystyle{mnras}
-\bibliography{example} % if your bibtex file is called example.bib
 

### 2023-11-15 manuscript/manuscript_2/paper.tex
attempt tp reproduce the earth/psr run 1251

@@
 Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-However, dropping the pulsar terms leads to biases in the inferred parameters and reduced detection probabilities \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
+However, dropping the pulsar terms leads to biases in the inferred parameters, and reduces the detection probability \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
 \newline
@@
 Secondly we quantify how the inclusion of the pulsar term affects the accuracy with which the static GW parameters can be estimated, and the minimal detectable GW strain.
-
-This paper is organised as follows. In Section \ref{sec:2} we briefly review the state-space formulation of PTA data analysis introduced in \cite{KimpsonPTA}.
-
-Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline
-
-%In Section 3 we introduce how to include the pulsar term in the state-space model.
-In Section 4 we test the updated method on the synthetic data for a single representative example GW system, and compare the results to the original Earth-terms model.
-In Section 9 we extend the tests to cover an astrophysically relevant parameter space.
-Discussion on future extensions and conclusions are made in Sections 11 and 12 respectively.
-Throughout the paper we adopt the natural units, with 𝑐 = 𝐺 = ħ = 1, and metric signature (−, +, +, +).
-
-%
-%
-%Intrinsic pulsar timing noise -- i.e.
-random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}.
-This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Myers2021MNRAS.502.3113M,Meyers2021,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}.
-In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators.
-However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.
-In modern Bayesian PTA searches, the power spectral density of the red intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the noise induced by a stochastic GW background (whose spectrum is also red).
-In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}.
+\newline
+
+The paper is organised as follows.
+In Section \ref{sec:2} we briefly review the state-space model of PTA data analysis introduced in \cite{KimpsonPTA}.
+In Section \ref{sec:pulsar_term} we show how to include the pulsar term in the model via a convenient reparametrisation.
+In Section \ref{sec:rep_example1} we test the updated model which includes the pulsar term on synthetic data for a single representative GW source.
+In Section \ref{sec:earth_vs_psr} we compare the parameter estimation accuracy and the minimal detectable GW strain when including the pulsar terms to the case where the pulsar terms are dropped.
+In Section \ref{sec:parameter_space} we extend the tests of Section \ref{sec:rep_example1} to cover an astrophysically relevant domain of SMBHB source parameters.
+Conclusions are drawn in Section \ref{sec:discussion}.
+Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$.
 \newline
@@
 In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates.
-Conclusions are drawn in Section \ref{sec:discussion}.
 
@@
 The model is formulated in terms of pulse frequency time series as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
-It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data.
-We defer this subtle generalization to future work.
+It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization we defer to future work.
 
@@
 \begin{equation}
-	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
+	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ , 
 	\label{eq:frequency_evolution}
@@
 K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection.
-That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
+That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
 \begin{equation}
@@
 and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}.
+For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
 Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
-We refer the reader to Appendix X for a review of Kalman filtering and nested sampling in the context of PTA searches for GWs.
+We refer the reader to Appendix X for a review of Kalman filtering, nested sampling and the combined workflow in the context of PTA searches for GWs.
 \newline
@@
 \begin{equation}
-	\chi^{(n)} = \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} 
+	\chi^{(n)} = \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \mod 2 \pi 
 \end{equation}
@@
 
-\section{Representative example: single SMBHB source and Comparison with Earth terms}\label{sec:rep_example1}
-
-In this section we apply the analysis scheme of K23 and the new model parametrisation of Section \ref{sec:pulsar_term} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-In Section \ref{sec:synthetic_data} we briefly review how synthetic data is generated in order to validate the analysis scheme.
-In Section \ref{sec:param_estimation} we apply the analysis scheme in conjunction with the model of Section \ref{sec:pulsar_term} in order to estimate $\boldsymbol{\theta}$ and compare the results with those using Earth-terms model, Equation \eqref{eq:g_func_trig_earth}.
-In Section \ref{sec:detection_new} we calculate the detectability of the GW source as a function of $h_0$ for both the pulsar terms and the Earth terms model.
-\newline
-
-This section considers only a single realisation of the system (i.e.
-process and measurement) noise and a particular, arbitrary set of injection parameters $\boldsymbol{\theta}$.
-Extension to multiple noise realisations and broader parameter spaces is considered in Sections X and Y respectively.
+\section{Representative example: single SMBHB source including the pulsar terms}\label{sec:rep_example1}
+In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}.
+As outlined in Section \ref{sec:pulsar_term}, the analysis scheme consists of a Kalman filter to track the state evolution of the intrinsic pulsar spin frequency which provides a likelihood for nested sampling.
+Nested sampling then returns estimates of the static parameters of the model and $\mathcal{Z}$.
+We refer the reader to Appendix X and K24 for details of the Kalman filter - nested sampling scheme.
+We apply the analysis scheme to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
+The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
+The goal of this section is to validate the analysis scheme when the pulsar terms are included.
+\newline
+
+In Section \ref{sec:synthetic_data} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme.
+In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ and $\mathcal{Z}$.
+The scheme is validated on multiple noise realisations of the pulsar process noise and the measurement noise in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
+This section considers only a particular, arbitrary set of injection parameters $\boldsymbol{\theta'}$.
+Extension to broader parameter spaces is considered in Section X.
 
 \subsection{Generating synthetic data}\label{sec:synthetic_data}
-\begin{table*}
-	\centering
-	%	\resizebox{\textwidth}{!}{%
-		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
-		\begin{tabular}{lccll}
-			\toprule
-			&Parameter & Injected Values & Units & Prior  \\
-			\hline
-			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
-			& $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
-			& $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
-			& $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
-			& $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
-			& $h_0$            & $5 \times 10^{-15}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
-			& $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
-			\hline
-			\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left( f_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{f} \right)$ \\
-			& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left( \dot{f}_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{\dot{f}} \right)$ \\
-			&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
-			& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
-			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
-			\bottomrule
-		\end{tabular}
-		\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter.
-The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}.
-The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
-		\label{tab:parameters_and_priors}
-	\end{table*}
-
-	Following K23, we consider a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
-The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
-Fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, are used with the latter two quantities evaluated at the Solar System barycenter.
-These values are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
-No direct measurements exist for $\gamma^{(n)}$, though the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
-In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
-Similarly, no direct measurements exist for $\sigma^{(n)}$.
-Following K23, we use
-	the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} in order to  set physically reasonable values.
-Within this model , the standard deviation in the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, is
-	\begin{align}
-		\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
-		&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
-	\end{align}
-	where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits.
-The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value.
-Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
-In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
-	\begin{eqnarray}
-		\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
-	\end{eqnarray}
-	For the synthetic 12.5 year NANOGrav dataset, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline
-
-	This completely specifies the static injection parameters for $\theta_{\rm psr}$.
-The static injection GW source parameters $\boldsymbol{\theta}_{\rm gw}$ are selected arbitrarily, subject to being astrophysically reasonable and representative.
-We consider a "high-SNR" source with $h_0 = 10^{-12}$ and a "low-SNR" source with $h_0 =5 \times 10^{-15}$.
-The complete set of injected parameters $\boldsymbol{\theta}$ is summarised in Table \ref{tab:parameters_and_priors} \newline
-
-	Given the injection parameters $\boldsymbol{\theta}$, we generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
-	\begin{enumerate}[leftmargin=2em]
-		\item Integrate Equation \eqref{eq:frequency_evolution} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
-This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
-		\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func_trig}
-	\end{enumerate}
-	We consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
-The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
-	\begin{equation}
-		\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
-	\end{equation}
-	Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
-The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
-Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
-Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar.
-\newline
-
-	\subsection{Parameter estimation}\label{sec:param_estimation}
-	\begin{figure*}
-		\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_highSNR_compare}
-		\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
-The vertical orange lines indicate the true injected values.
-The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
-We are able to accurately estimate each parameter of interest.
-We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
-Note that the $x$-axis scaling does not span the total prior space.}
-		\label{fig:corner_plot_1}
-	\end{figure*}
-
-	\begin{figure*}
-		\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
-		\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
-The vertical orange lines indicate the true injected values.
-The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
-We are able to accurately estimate each parameter of interest.
-We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
-Note that the $x$-axis scaling does not span the total prior space.}
-		\label{fig:corner_plot_2}
-	\end{figure*}
-
-	\begin{figure*}
-		\includegraphics[width=0.45\textwidth, height =0.45\textwidth ]{images/corner_highSNR_chi} 	
-		\includegraphics[width=0.45\textwidth, height =0.45\textwidth ]{images/corner_lowSNR_chi}
-		\caption{hellow}
-		\label{fig:corner_plot_3}
-	\end{figure*}
-
-	\begin{table}
-		\centering
-		%	\resizebox{\textwidth}{!}{%
-			%	\renewcommand{\arraystretch}{1.0} % Default value: 1
-			\begin{tabular}{lcclc}
-				\toprule
-				&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm pulsar}(\theta) -\Delta_{\rm earth}(\theta) $   \\
-				\hline
-				\multirow{7}{2mm}{High SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
-				& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
-				& $\psi$ &0.0386& 0.000254& -0.0384 \\
-				& $\iota$ & 0.374 &0.126 &-0.248 \\
-				& $\delta$ & 0.00223 &0.000242& -0.00198 \\
-				&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
-				&$h_0$ & 0.168 &0.0846 &-0.0839 \\
-				\hline
-				\multirow{7}{2mm}{Low SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
-				& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
-				& $\psi$ &0.0386& 0.000254& -0.0384 \\
-				& $\iota$ & 0.374 &0.126 &-0.248 \\
-				& $\delta$ & 0.00223 &0.000242& -0.00198 \\
-				&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
-				&$h_0$ & 0.168 &0.0846 &-0.0839 \\
-
-				\bottomrule
-			\end{tabular}
-			\caption{dfdfddd}
-			\label{tab:posterior_errors}
-		\end{table}
-
-		%\begin{table}
-		%	\centering
-		%	%	\resizebox{\textwidth}{!}{%
-			%		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
-			%		\begin{tabular}{cccc}
-				%			\toprule
-				%			$\theta$ & \multicolumn{2}{c}{RMSE$(\theta)$} & $\Delta$ \\
-				%			& Earth & Pulsar & \\
-				%			\hline
-				%			$\Omega$ & 1 & 2 & 3 \\ 
-				%			
-				%			\bottomrule
-				%		\end{tabular}
-			%		\caption{}
-			%		\label{tab:posterior_errors}
-			%	\end{table}
-
-		%
-		%\begin{table}
-		%	\centering
-		%	%	\resizebox{\textwidth}{!}{%
-			%		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
-			%		\begin{tabular}{lccll}
-				%			\toprule
-				%			$\theta$& \multirow{2}{*}{RMSE$(\theta)$} & Injected Values & Units & Prior  \\
-				%			& 
-				%			\hline
-				%			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
-				%			& $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
-				%			& $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
-				%			& $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
-				%
-				%
-				%			\bottomrule
-				%		\end{tabular}
-			%		\caption{}
-			%		\label{tab:posterior_errors}
-			%	\end{table}
-
-		Initially we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$ for the representative example system described in Table \ref{tab:parameters_and_priors}.
-We consider a "high-SNR" system with $h_0 = 10^{-12}$ and a ``low-SNR" system with $h_0 = 5 \times 10^{-15}$.
-Priors on $\boldsymbol{\theta}$ are selected as follows.
-For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
-For $\pi(\boldsymbol{\theta}_{\rm psr})$ we use the same priors discussed in K23.
-Additionally, we set a prior on $\chi^{(n)}$ as $\text{Uniform}(0,2 \pi)$.
-All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
-\newline
-
-		We apply the Kalman filter using Equation \eqref{eq:measuremen_earth} in conjunction with nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ for both the high/low SNR systems.
-The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_1} for the high-SNR system and in Figure \ref{fig:corner_plot_2} for the low-SNR system.
-The histograms shown in the figures are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the solid orange line indicates the true injection value.
-The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours.
-The green curves are the results derived using the Earth-terms measurement equation, Equation \ref{eq:g_func_trig_earth}.
-The blue curves are the results derived using the full measurement equation, including the pulsar terms, Equation \eqref{eq:g_func_trig}.
-\newline
-
-		For the high-SNR case, Figure \ref{fig:corner_plot_1}, the one-dimensional posteriors for the Earth-terms solution are evidently biased, as was observed in K23.
-Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
-For example, for $\iota$, whilst the injected value is contained within the 90 \% credible interval, the median of the posterior of the Earth-terms solution is shifted from the injected value by 0.35 radians.
-The inclusion of the pulsar terms corrects for this bias.
-The green one-dimensional posteriors exhibit no bias and a generally symmetric about the injected value.
-Moreover, for every pair of parameter the injection values are contained within the 2-sigma equivalent contours.
-This is not true in the Earth-terms case.
-More quantitatively, we define a relative error metric to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
-		\begin{equation}
-			\Delta_{\rm M}(\Theta) = \frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}} 
-		\end{equation}
-		where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ the total number of samples, $\hat{\theta}$ the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimates using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
-$\Delta_{\rm M}(\theta)$ then represents the average relative error in the one dimensional posterior.
-The values for each parameter in Figure \ref{fig:corner_plot_1} are summarised it Table \ref{tab:posterior_errors}, along with the relative difference between the two models, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$.
-For every parameter except $\Omega$, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is negative, indicating that the parameter estimates are more accurate using the pulsar terms.
-$\Omega$ is recovered with high accuracy in both cases, and the relative error is small, $\Delta_{\rm M}(\Omega) \sim 10^{-5}$.
-The improvement in $\iota$ is largest, with $\Delta_{\rm pulsar}(\iota) - \Delta_{\rm earth}(\iota) = 0.25$.
-\newline
-
-		For the low-SNR case,  Figure \ref{fig:corner_plot_2}, there is less of a notable improvement in the parameter estimates.
-Qualitatively, the green and blue one-dimensional posteriors are generally overlaid.
-The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest.
-The relative errors 	$\Delta_{\rm M}(\Theta)$ for the low SNR case are also reported in Table \ref{tab:posterior_errors}.
-We see that....
-\textcolor{red}{TK: numbers to be inserted here}.
-We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case.
-This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates.
-Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:detection_new} that they are important for increasing the detection probability for a given signal strength.
-\newline
-
-		In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
-In both the high and low SNR cases $\chi^{(n)}$ is unambiguously recovered.
-		This is shown in Figure \ref{fig:corner_plot_3} where we plot the results for five of the $\chi^{(n)}$ values for both high SNR (left-hand panel) and low SNR (right hand panel).
-The estimates are naturally more constrained in the high SNR case, but in both cases the injection is contained within the 90 \% credible interval.
-Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
-
-		In summary, the main results of this section are:
-		\begin{itemize}
-			\item We are able to estimate all the static parameters of the system using the full pulsar-terms model, for this representative system with characteristic strain $h_0 = 5 \times 10^{-15}$.
-			\item The inclusion of pulsar terms improves the parameter estimation accuracy, especially when the GW signal is loud with respect to the noise.	
-		\end{itemize}
-
-		\subsection{Detection}\label{sec:detection_new}
-
-		\begin{figure}
-			\includegraphics[width=\columnwidth, height = \columnwidth ]{images/CanonicalBayesPlot2000} 	
-			\caption{hellow}
-			\label{fig:bayes1}
-		\end{figure}
-
-		The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
-		\begin{equation}
-			\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
-		\end{equation}
-		Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
-$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
-This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
-The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
-
-		We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-$\mathcal{M}_1$ is the Earth-terms only model, i.e.
-the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
-The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
-The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
-We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
-We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
-For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
-This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
-$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
-\newline
-
-		There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
-The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
-The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
-For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
-That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
-For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
-For the results presented in this subsection we set $n_{\rm live} =2000$.
-
-\section{Validation with synthetic data} \label{sec:testing}
-In this section we outline how synthetic data are generated in order to validate the analysis scheme in Section \ref{sec:detect}.
-Synthetic data enable validation to occur systematically and under controlled conditions.
-In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$.
-In Section \ref{sec:gendata} we demonstrate how to solve Equations \eqref{eq:spinevol}--\eqref{eq:frequency_evolution}, and \eqref{eq:measurement}--\eqref{eq:g_func} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) $.
-
-\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
-	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work.
-The pulsar distances relative to the observer are also indicated, with distance $\leq 2 \, {\rm kpc}$ for $38$ pulsars.
-The grey dashed curve denotes the Galactic plane.}
-	\label{fig:pulsar_distrib}
-\end{figure}
-We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
-The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA.
-\newline
-
-For each pulsar we adopt fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter.
-A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility.
-The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}.
-The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
-\newline
-
-The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist.
-The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency}, and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
-In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
-We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$.
-The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
-\begin{align}
-	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
-	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
-\end{align}
-where $T_{\rm cad}$ is the cadence of the timing observations.
-For MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$ ($\pm 2\sigma$ confidence limits).
-The uncertainties are broad; for the purpose of generating astrophysically representative synthetic data we straightforwardly adopt the central values in this paper.
-Throughout this work we assume for simplicity that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$.
-In order to relate Equation \eqref{eq:sigmap} to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
-\begin{eqnarray}
-	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
-\end{eqnarray}
-For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, with $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline
-
-As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
-\begin{eqnarray}
-	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
-\end{eqnarray}
-and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach.
-We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
-A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$.
-In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$.
-In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}.
-We see that the synthetic and empirical residuals are qualitatively similar.
-For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is much less than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ s$^{-3/2}$ for PSR J1939$+$2134.
-
-%This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars.
-If we set $\ln \alpha_1 = -38$ in Equation \eqref{eq:sigmap}, close to its lower limit, then the estimated $\sigma^{(n)}$ for  PSR J1939+2134 is $\sigma^{(n)} =3.90 \times 10^{-27}$ which is in much better agreement with the value inferred from the residuals comparison method.
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
-	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J1939+2134.
-The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}.
-The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$.
-The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}.
-We set the TOA uncertainty to be a constant, viz.
-$0.1 \mu$s.
-The blue and orange residuals are qualitatively similar by inspection.
-Similar results are obtained for the other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} by tuning the value of $\sigma^{(n)}$.}
-	\label{fig:qualitative_compare}
-\end{figure}
-\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
-We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
-\begin{enumerate}[leftmargin=2em]
-	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
-	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}.
-\end{enumerate}
-Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
-The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \ref{sec:synt_pta}.
-For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence.
-\newline
-
-The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, through
-\begin{equation}
-	\sigma_{\rm m} \approx \sigma_{\rm TOA} f_{\rm p}^{(n)}  \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
-\end{equation}
-Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct.
-Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise.
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 10^2$ Hz observed with weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s, Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
-The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz.
-Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred for the sake of simplicity; this assumption can be relaxed easily when analysing real data.
-Whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}(t)$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar.
-\newline
-
-Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} dB(t) | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $dB(t)$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}).
-To fix the problem, we subtract the deterministic frequency evolution and track the new variable
-\begin{equation}
-	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ , \label{eq:hetero1}
-\end{equation}  
-equivalent to a change of variables. We similarly modify the measurement variable to be
-\begin{equation}
-	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ , \label{eq:hetero2}
-\end{equation}
-where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic evolution based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice.
-For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately.
-Equation \eqref{eq:measurement} is then updated to read
-\begin{equation}
-	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
-	\label{eq:measurement_cov}
-\end{equation}
-We emphasise that the change of variables in Equations \eqref{eq:hetero1} and \eqref{eq:hetero2} is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g.
-long double, quadruple).
-It does not remove any degrees of freedom nor does it involve an approximation.
-In particular $f_{\rm em}^{(n)}(t_1)$
-and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, i.e.
-Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.
-
-\section{Representative end-to-end analysis} \label{sec:rep_example}
-In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space} and \ref{sec:bias_and_identifiability}.
-To assist with reproducibility, we present intermediate outputs from each step.
-In Section \ref{sec:priors} we define the priors on $\boldsymbol{\theta}$.
-In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter.
-In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
-In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates.
-In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$.
-\newline
-
-The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
-The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
-
-\subsection{Prior distribution}\label{sec:priors}
-The first step is to select a reasonable Bayesian prior, $\pi(\boldsymbol{\theta})$, for the static parameters.
-For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}.
-The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires additional discussion below.
-\newline
-
-The parameters that govern the deterministic evolution of $f_{\rm p}^{(n)}(t)$, namely $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, are well-determined by radio timing observations.
-We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues \citep{Manchester2005}.
-For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} $ per cent and $\pm 2.31 \times 10^{-3}$ per cent respectively.
-In this paper we adopt uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database \citep{Manchester2005}.
-Wider-than-necessary priors, such as those above, test the method more stringently than narrow priors.
-The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are wide or narrow.
-\newline
-
-The pulsar distances $d^{(n)}$ are less constrained than $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with typical uncertainties $\sim 10 \%$ \citep{Yao2017,Arzoumanian2018ApJS..235...37A}.
-In this paper, however, we omit the term involving $d^{(n)}$ from Equation \eqref{eq:g_func}, as justified in Section \ref{sec:parameter_estim}, so there is no need to set a prior on $d^{(n)}$.
-Furthermore we do not set a prior on $\gamma^{(n)}$, because we have $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and $\gamma^{(n)}$ is effectively ``unobservable'' over a decade; that is, for $T_{\rm obs}=10$ years the solution of Equation \eqref{eq:spinevol} is approximately independent of $\gamma^{(n)}$, as long as $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
-It is therefore sufficient for validation purposes to carry $\gamma^{(n)}$ through the analysis at its injected value.
-\footnote{We check this empirically in two independent ways for an informal selection of test cases: (i) we set an uninformative prior $\pi\left [\gamma^{(n)} / 1 \, {\rm s^{-1}} \right] \sim$ LogUniform($10^{-15}, 10^{-10}$); and (ii) we deliberately displace $\gamma^{(n)}$ in the inference analysis from its true, injected value in the synthetic data, e.g.
-$\gamma^{(n)} = 10^{-14} \text{ s}^{-1}$ versus $10^{-13} \text{ s}^{-1}$, respectively.
-The results from (i) and (ii) are found to be the same as those reported in Section \ref{sec:rep_example}.} \newline
-
-Most pulsars in the synthetic PTA have $10^{-22} \leq \sigma^{(n)} / (1 \, {\rm s^{-3/2}}) \leq 10^{-20}$ as calculated from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}.
-For these pulsars we set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$).
-The sole exception is PSR J1939+2134 which has $\sigma^{(n)} \sim 10^{-16} \text{s}^{-3/2}$.
-For validation purposes, we artificially set $\sigma^{(n)} = 10^{-20} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within the aforementioned log-uniform prior.
-This is done purely for testing convenience; it is straightforward to expand the prior when analysing real, astronomical data.
-\newline
-
-By not setting priors on $\gamma^{(n)}$ and $d^{(n)}$ we reduce the number of parameters to $7 + 3N$.
-We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter set, c.f.
-Equation \eqref{eq:psrparams}.
-Explicitly, we write
-\begin{equation}
-	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ . \label{eq:psrreduced}
-\end{equation}
-The injected static parameters and their corresponding priors for the representative analysis in Section \ref{sec:rep_example} are summarised in Table \ref{tab:parameters_and_priors}.
 \begin{table*}
@@
 	\end{table*}
-
-\subsection{Earth and pulsar terms}\label{sec:earth_psr_terms}
-
-	\subsection{Posterior distribution}\label{sec:parameter_estim}
+In order to generate synthetic frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$ Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
+The measured $f_{\rm m}^{(n)}(t)$ are then produced via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
+Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and \eqref{eq:measurement}--\eqref{eq:g_func_trig} depend on the choice of static parameters $\boldsymbol{\theta'}$, as well as the choice of pulsar directions ${\boldsymbol{q}}^{(n)}$ and detector measurement noise covariance $\sigma_{\rm m}$ (Equation \eqref{eq:sigma_m_eqn}).
+We now describe how these values are chosen so as to generate astrophysically representative synthetic data.
+\newline
+
+In this paper we use the same values for $\boldsymbol{\theta}_{\rm psr}$ as in K24.
+That is, we adopt the 47 millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}, i.e.
+$N = 47$.
+We consider all pulsars to be observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period.
+For each pulsar fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$ are obtained via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues.
+No direct measurements exist for $\gamma^{(n)}$, though the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
+In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
+Similarly, no direct measurements exist for $\sigma^{(n)}$.
+We relate the process noise in the frequency $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week such that
+\begin{eqnarray}
+	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
+\end{eqnarray}
+where, following K24, we use the empirical timing noise model for MSPs from \cite{Shannon2010ApJ...725.1607S} in order to set physically reasonable values on $\sigma_{\rm TOA}^{(n)}$.
+The median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134.
+Compared to the other 46 pulsars of the array, PSR J1939+2134 is a notable outlier with an estimated $\sigma^{(n)}$ much larger than the median.
+In practice, in order to avoid setting an overly broad prior on $\sigma^{(n)}$ (see Section \ref{sec:set_priors}) for validation purposes, we artificially set $\sigma^{(n)} = 10^{-20} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within an reasonable width log-uniform prior.
+This is done purely for testing convenience; it is straightforward to expand the prior when analysing real, astronomical data.
+\newline
+
+The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
+\begin{equation}
+	\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
+\end{equation}
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
+Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
+Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar.
+\newline
+
+The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
+The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
+Different values of the static GW source parameters are explored in Section XX.
+\newline
+
+The synthetic data generated by the above procedure is passed to the analysis scheme in Section \ref{sec:pe_and_ms}.
+
+\subsection{Prior distribution}\label{sec:set_priors}
+To deploy the analysis scheme of K24 it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters. \newline
+
+For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}.
+For $\pi(\boldsymbol{\theta}_{\rm psr})$ \textit{a priori} information from electromagnetic observations does exist.
+We adopt constrained uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database.
+By using wider-than-necessary priors we expose the analysis scheme to a more stringent test.
+We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$) which covers the range of values inferred from Equation \ref{eq:sigmap_f}.
+We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.
+We do not set a prior on $\gamma^{(n)}$, because $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and consequently $\gamma^{(n)}$ is effectively ``unobservable'' over $T_{\rm obs}$.
+For validation purposes it is therefore sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value.
+This reduces the total dimension of the parameter space to $7 + 4N$.
+The chosen prior $\boldsymbol{\theta}$ is the same as used in K24, with the addition of a prior on $\chi^{(n)}$ due to the inclusion of the pulsar term.
+\newline
+
+The chosen priors for the representative analysis in Section \ref{sec:rep_example1} are summarised in \ref{Table tab:parameters_and_priors}
+
+\subsection{Parameter estimation} \label{sec:pe_and_ms} 
 In this section, we calculate the posterior probability distribution for the static parameters $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$ and compare it to the known, injected values.
-The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior), and (ii) to give a preliminary sense of its accuracy.
-Initially we consider a single noise realisation when generating the synthetic data for the representative example in Table \ref{tab:parameters_and_priors}.
+The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
+That is, given the injected static parameters $\boldsymbol{\theta}$ summarised in Table \ref{Table tab:parameters_and_priors}, multiple realisations of synthetic data are generated using the method described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta}$.The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) demonstrate that quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
 We apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$.
@@
 
+In Figure \ref{fig:corner_plot_2} we plot the estimates of $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary realisations.
+The seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$ are recovered unambiguously.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, which stems from one realisation.
+We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
+As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm psr}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
+\newline
+
+Figure \ref{fig:corner_plot_2} confirms two main points: (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
+Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 (i.e.
+$92 \%$) out of the 63 possible combinations of the seven parameters and nine realizations.
+Figure \ref{fig:corner_plot_2}, like Figure \ref{fig:corner_plot_1}, displays tentative signs of bias, where the one-dimensional posteriors are not symmetric about the injected value.
+For example, the maximum a posteriori probability estimates of $\iota$ appear left-skewed, consistently underestimating the injected value in all nine realisations.
+A similar but messier trend is seen in $\delta$.
+However, the width of the posteriors is comparable to the putative bias, so it is difficult to draw strong conclusions.
+Biases are discussed in detail in Section \ref{sec:bias_and_identifiability}.
+There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.
+\newline
+
 The posterior is approximately symmetric about the known injected value for some parameters such as $\Omega$, where the posterior median and the injected value coincide approximately.
@@
 
+In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
+In both the high and low SNR cases $\chi^{(n)}$ is unambiguously recovered.
+This is shown in Figure \ref{fig:corner_plot_3} where we plot the results for five of the $\chi^{(n)}$ values for both high SNR (left-hand panel) and low SNR (right hand panel).
+The estimates are naturally more constrained in the high SNR case, but in both cases the injection is contained within the 90 \% credible interval.
+Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
+
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
-	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/stacked_GW_plot_small_h_definitive2}
+	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for nine realisations of the noise processes, with curves coloured uniquely.
 The horizontal and vertical orange lines indicate the true injected values.
@@
 The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter.
-The supertitles of the marginalized one-dimensional histograms specify the posterior median and the 0.16 and 0.84 quantiles.
+The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation.
 We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$.
 The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
-The horizontal axes span a subset of the prior domain for all seven parameters.}
+The horizontal axes span a subset of the prior domain for all seven parameters.
+The known, injected value lies within the 90\% credible interval for \textcolor{red}{X} out of the 63 combinations of seven parameters and nine noise realizations.
+There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation \textcolor{red}{X} across the 63 combinations.}
 	\label{fig:corner_plot_1}
@@
 
-%\begin{figure*}
-%	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
-%	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
-The vertical orange lines indicate the true injected values.
-The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-We are able to accurately estimate each parameter of interest.
-We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
-%	\label{fig:corner_plot_1}
-%\end{figure*}
-
-\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
-	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely.
-The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation.
-The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations.
-There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation $\sim10 \%$ across the 63 combinations.
-A slight skew-bias is apparent in some of the parameters, e.g.
-$\iota$.
-Weak correlations can be seen between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.}
-	\label{fig:corner_plot_2}
-\end{figure*}
+	\begin{figure}
+	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi}
+	\caption{hellow}
+	\label{fig:corner_plot_3}
+\end{figure}
+
 \begin{table}
@@
 
-The results in Section \ref{sec:parameter_estim} are obtained for a single realisation of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
-It is important to confirm that the analysis scheme returns accurate answers for arbitrary noise realisations and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous by accident.
-It is also important to quantify, albeit approximately, the natural random dispersion in the one-dimensional posterior medians from one noise realization to the next, as the dispersion is a practical measure of the accuracy of the parameter estimation scheme, when it is applied to real astronomical data, where the true parameter values and specific noise realisation are unknown.
-\newline
-
 To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
 For each realisation we independently estimate the static parameters, $\boldsymbol{\theta}$.
-In Figure \ref{fig:corner_plot_2} we plot the estimates of $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary realisations.
-The seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$ are recovered unambiguously.
-The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, which stems from one realisation.
-We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
-As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm psr}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
-\newline
-
-Figure \ref{fig:corner_plot_2} confirms two main points: (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
-Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 (i.e.
-$92 \%$) out of the 63 possible combinations of the seven parameters and nine realizations.
-Figure \ref{fig:corner_plot_2}, like Figure \ref{fig:corner_plot_1}, displays tentative signs of bias, where the one-dimensional posteriors are not symmetric about the injected value.
-For example, the maximum a posteriori probability estimates of $\iota$ appear left-skewed, consistently underestimating the injected value in all nine realisations.
-A similar but messier trend is seen in $\delta$.
-However, the width of the posteriors is comparable to the putative bias, so it is difficult to draw strong conclusions.
-Biases are discussed in detail in Section \ref{sec:bias_and_identifiability}.
-There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.
-\newline
 
@@
 
-\subsection{Detectability versus $h_0$} \label{sec:detection}
+\section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}
+In this section we compare the results obtained using the pulsar-term model (i.e.
+Equation \eqref{eq:g_func_trig_chi}) to those obtained using the Earth-term model (i.e.
+Equation \eqref{eq:g_func_trig_earth}) for the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
+In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$ returned by the two models.
+We consider two separate cases; a ``high-SNR" case with $h_0 = 10^{-12}$ and a ``low-SNR" case with $h_0 = 5 \times 10^{-15}$.
+In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable strain for the two models.
+In this Section, the procedure for generating synthetic data and the specification of $\pi(\boldsymbol{\theta})$ is as described in Sections \ref{sec:synthetic_data} and \ref{sec:set_priors} respectively.
+Both the pulsar-term model and the Earth-terms model are applied to identical realisations of the data.
+The priors in both cases are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for the pulsar terms model, as described in Section \ref{sec:set_priors}.
+
+\subsection{Accuracy of parameter estimation}\label{sec:psr_v_earth_pe}
+
+In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$.
+For the Earth-term model we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$.
+For the pulsar-term model we we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi}.
+The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+The settings of the nested sampler (for example the number of live points, see Appendix X) are identical for both models.
+\newline
+
+We consider two representative systems.
+The first is a ``low-SNR" system with $h_0 = 5 \times 10^{-15}$, i.e.
+the exact system described in Table \ref{tab:parameters_and_priors}.
+The second is a ``high-SNR" system which has the same static injection parameters, but with an increased $h_0 = 10^{-12}$.
+The ``high-SNR" system is considered in order to evaluate the parameter estimation accuracy, and in particular the bias, without confusion from the measurement noise.
+This section considers a single noise realisation of the synthetic data $\boldsymbol{Y}$ for the representative example system described in Table \ref{tab:parameters_and_priors}.
+\newline
+
+The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_compare_high} for the high-SNR system and in Figure \ref{fig:corner_plot_compare_low} for the low-SNR system.
+The corner plot is arranged identically to \ref{fig:corner_plot_1}.
+The green curves are the results derived using the Earth-terms measurement equation, Equation \ref{eq:g_func_trig_earth}.
+The blue curves are the results derived using the full measurement equation, including the pulsar terms, Equation \eqref{eq:g_func_trig}.
+\newline
+
+For the high-SNR case, Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for the Earth-term model are evidently biased, as was observed in K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
+Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
+For example, for $\iota$, whilst the injected value is contained within the 90 \% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians.
+The inclusion of the pulsar terms corrects for this bias.
+The green one-dimensional posteriors exhibit no bias and a generally symmetric about the injected value.
+Moreover, for every pair of parameters the injection values are contained within the 2-sigma equivalent contours for the pulsar-term model, which is not true in the Earth-terms case.
+\newline
+
+We define a relative error metric to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
+\begin{equation}
+\Delta_{\rm M}(\theta) = \frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}} \ , 
+\end{equation}
+where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimates using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+The dependence of $p_k(\theta)$ on $M$ is implied.
+$\Delta_{\rm M}(\theta)$ represents the mean relative error in the one-dimensional posterior.
+The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_1} are summarised in Table \ref{tab:posterior_errors}.
+The difference in the mean relative error between the two models, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is also presented.
+For every parameter except $\Omega$, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is negative, indicating that the parameter estimates are more accurate using the pulsar terms (i.e.
+the mean relative error when using the pulsar model is smaller).
+The exception for $\Omega$ is due to the fact that $\Omega$ is recovered with high accuracy in both cases; the the mean relative error is small $\Delta_{\rm M}(\Omega) \sim 10^{-5}$.
+The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm pulsar}(\iota) - \Delta_{\rm earth}(\iota) = 0.25$.
+\newline
+
+For the low-SNR case,  Figure \ref{fig:corner_plot_compare_low}, there is less of a notable improvement in the parameter estimates.
+Qualitatively, the green and blue one-dimensional posteriors are generally overlaid.
+The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest.
+The relative errors 	$\Delta_{\rm M}(\Theta)$ for the low SNR case are also reported in Table \ref{tab:posterior_errors}.
+We see that....
+\textcolor{red}{TK: numbers to be inserted here}.
+We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case.
+This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates.
+Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:detection_new} that they are important for increasing the detection probability for a given signal strength.
+\newline
+
+	\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_highSNR_compare}
+	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The vertical orange lines indicate the true injected values.
+The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
+We are able to accurately estimate each parameter of interest.
+We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
+Note that the $x$-axis scaling does not span the total prior space.}
+	\label{fig:corner_plot_compare_high}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
+	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+The vertical orange lines indicate the true injected values.
+The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
+We are able to accurately estimate each parameter of interest.
+We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
+Note that the $x$-axis scaling does not span the total prior space.}
+	\label{fig:corner_plot_compare_low}
+\end{figure*}
+
+\begin{table}
+	\centering
+	%	\resizebox{\textwidth}{!}{%
+		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+		\begin{tabular}{lcclc}
+			\toprule
+			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm pulsar}(\theta) -\Delta_{\rm earth}(\theta) $   \\
+			\hline
+			\multirow{7}{2mm}{High SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
+			& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
+			& $\psi$ &0.0386& 0.000254& -0.0384 \\
+			& $\iota$ & 0.374 &0.126 &-0.248 \\
+			& $\delta$ & 0.00223 &0.000242& -0.00198 \\
+			&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
+			&$h_0$ & 0.168 &0.0846 &-0.0839 \\
+			\hline
+			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
+			& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
+			& $\psi$ &0.0386& 0.000254& -0.0384 \\
+			& $\iota$ & 0.374 &0.126 &-0.248 \\
+			& $\delta$ & 0.00223 &0.000242& -0.00198 \\
+			&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
+			&$h_0$ & 0.168 &0.0846 &-0.0839 \\
+
+			\bottomrule
+		\end{tabular}
+		\caption{dfdfddd}
+		\label{tab:posterior_errors}
+	\end{table}
+
+\subsection{Detectability vs $h_0$}\label{sec:psr_v_earth_bayes}
+
+		\begin{figure}
+			\includegraphics[width=\columnwidth, height = \columnwidth ]{images/CanonicalBayesPlot2000} 	
+			\caption{hellow}
+			\label{fig:bayes1}
+		\end{figure}
+
+		The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
+		\begin{equation}
+			\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
+		\end{equation}
+		Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
+$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
+This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
+The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
+
+		In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$. \newline
+
 		We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}.
@@
 In this section we set $n_{\rm live} =2000$.
+
+		We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
+$\mathcal{M}_1$ is the Earth-terms only model, i.e.
+the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
+The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
+The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
+We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
+We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
+For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
+This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
+$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
+\newline
+
+		There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
+The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
+The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
+For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
+That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
+For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
+For the results presented in this subsection we set $n_{\rm live} =2000$.
+
+\section{Representative end-to-end analysis} \label{sec:rep_example}
+In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
+The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
+The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space} and \ref{sec:bias_and_identifiability}.
+To assist with reproducibility, we present intermediate outputs from each step.
+In Section \ref{sec:priors} we define the priors on $\boldsymbol{\theta}$.
+In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter.
+In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
+In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates.
+In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$.
+\newline
+
+The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
+The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
+
+\subsection{Detectability versus $h_0$} \label{sec:detection}
+
 \begin{figure}
@@
 
-\section{Systematic biases} \label{sec:bias_and_identifiability}
- Two systematic aspects of the tests in Sections \ref{sec:rep_example} and \ref{sec:parameter_space} merit further study.
-The first relates to parameter identifiability, i.e.
-can we uniquely identify the parameters of the model given the data and the model structure?
-This question is explored in Section \ref{sec:identif}.
-The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
-It is the focus of Section \ref{sec:pulsar_terms_drop}.
-In order to elucidate these matters without confusion from the measurement noise, we switch to the high-SNR regime and set $h_0 = 10^{-12}$ (c.f.
-Figure \ref{fig:bayes}) for the representative system in Table \ref{tab:parameters_and_priors}.
-
-\subsection{Identifiability}\label{sec:identif}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
-	\caption{Corner plot of the joint posterior distribution of the GW source parameters $h_0$ and $\iota$ for the representative system in Table 1 in the high-SNR regime ($h_0 = 10^{-12}$).
-Nine coloured curves are plotted for nine random noise realizations.
-The posteriors overlap less than for the other five parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
-A correlation between $h_0$ and $\iota$ is evident in the bottom left panel.}
-	\label{fig:just_iota_h}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_surface}
-	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$-$h_0$ plane for a single noise realisation.
-The red dot marks the maximum of $\log {\cal L}$.
-The colour scale is normalised with respect to the absolute value of the maximum.
-An S-shaped, flat-topped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
-	\label{fig:likelihood_surface}
-\end{figure}
-In Figure \ref{fig:just_iota_h} we plot the one- and two-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations.
-The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$.
-One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section \ref{sec:pulsar_terms_drop}; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high.
-One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$.
-Specifically, across $10^3$ noise realisations, the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
-Both $W_{1,{\rm median}}$ values are higher (as a fraction of the true, injected value) than in the lower-SNR tests conducted in Sections \ref{sec:rep_example} and \ref{sec:parameter_space}.
-In contrast, $W_{1,\rm median}$ for the other five parameters in $\boldsymbol{\theta}_{\rm gw}$ decreases, as the SNR increases.
-\newline
-
-We infer different $h_0$ and $\iota$ posteriors for different noise realizations, because $h_0$ and $\iota$ are weakly identifiable.
-Identifiability is a formal concept from the theory of signal processing, which is applied routinely to engineering problems \citep{e5be7c83a0d24500826f6e1b414d1733}.
-It refers to whether it is theoretically possible to infer unique and accurate parameter values, given the measured data and the model structure.
-Sources of non-identifiability are generally categorised as either ``structural'' , arising from the structure of the model (e.g.
-its invertibility), or ``practical'', arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
-Regarding $\iota$ and $h_0$, the source is structural.
-Equations \eqref{eq:hphx} and \eqref{eq:hphx2} contain a weak degeneracy between $\iota$ and $h_0$.
-For example, the cross polarisation strain $h_{\times}$ involves a trade-off between $\iota$ and $h_0$, which is resolved by knowing $h_+$ via Equation \eqref{eq:hphx}.
-However, the measured strain is a linear combination of $h_\times$ and $h_+$; the two polarizations are not measured independently, which makes it challenging to separate $\iota$ and $h_0$.
-In fact $\iota$ and $h_0$ are structurally identifiable --- but only weakly.
-That is, the $\iota$-$h_0$ likelihood contours plateau to a flat-topped ridge with a small, non-zero gradient close to the maximum likelihood solution.
-The likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
-Whilst a single, unique maximum does exist (labelled by the red dot in the figure), it lies along an S-shaped ridge where the likelihood values are all similar.
-Consequently it is difficult for the nested sampling algorithm to settle on the red dot; it trades off accuracy in one parameter against the other, depending on the particular realisation of the noise.
-By way of verification, if we fix one parameter at its true value (i.e.
-set a delta function prior) during the inference analysis, the posteriors for the other parameter overlap indistinguishably  for different noise realisations, i.e.
-$W_1 \to 0$.
-
-\subsection{Dropping the pulsar terms}\label{sec:pulsar_terms_drop}
-
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
-	\caption{Same as Figure \ref{fig:corner_plot_1} but for a high-SNR system with $h_0 = 10^{-12}$.
-For five out of seven parameters the true, injected value does not fall within the 90\% credible interval and falls outside the plotted domain.
-The estimated posteriors are biased away from the injected value due to dropping the pulsar terms.
-Note that the plotted domain is narrower than in Figure \ref{fig:corner_plot_1}.}
-	\label{fig:bias_for_large_h}
-\end{figure*}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
-	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth term in the Kalman filter (Equation \ref{eq:measuremen_earth}, blue curve), the pulsar terms in the Kalman filter (Equation \ref{eq:measurement}, orange curve), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter (green curve).
-All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
-The coloured dashed vertical lines mark the maximum for the corresponding model; the green and orange dashed lines overlap at this resolution.
-For the Earth-term model the likelihood maximum ($\iota \approx 0.7$ rad) does not coincide with the injected value ($\iota = 1.0$ rad).
-For the pulsar-terms model the maximum and injected value do coincide.
-10 noise realisations are plotted for each situation but overlap at this resolution; they are indistinguishable to the eye.}
-	\label{fig:likelihood_surface_iota}
-\end{figure}
-
-Sections \ref{sec:rep_example} and \ref{sec:parameter_space} present preliminary evidence that the inferred value of ${\boldsymbol{\theta}}_{\rm gw}$ (e.g.
-the medians of the one-dimensional marginalised posteriors) is biased away from the true injected value.
-Figure \ref{fig:bias_for_large_h} displays a corner plot for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but for high SNR $(h_0 = 10^{-12})$.
-Except for the elements $\Omega$ and $h_0$, the injected values lie outside the 90\% credible interval, and indeed fall outside the plotted domain.
-The deviation is most severe for $\iota$ with a bias of $\approx 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$.
-Similar results are obtained for other noise realisations.
-Whilst $h_0$ does not exhibit any noticeable bias for the particular noise realisation generating Figure \ref{fig:bias_for_large_h}, in general it does exhibit a correlated bias with $\iota$ across multiple noise realisations, as noted in the bottom left panel of Figure \ref{fig:just_iota_h}.
-In this section we demonstrate how the bias stems from dropping the pulsar terms described in Section \ref{sec:parameter_estim}.
-\newline
-
-In Figure \ref{fig:likelihood_surface_iota} we plot the log-likelihood returned by the Kalman filter, viz.
-Equation \eqref{eq:likelihood}, as a function of $\iota$ (i.e.
-$\log \mathcal{L}(\iota)$) across the prior domain, holding all other parameters constant at the true injected value of the representative system in Table \ref{tab:parameters_and_priors}.
-We generate 10 plots of $\log \mathcal{L}(\iota)$ corresponding to 10 separate realisations of the system noise.
-The solid blue curves only includes the Earth term in the measurement equation (i.e.
-using Equation \eqref{eq:measuremen_earth} in the Kalman filter).
-The solid orange curves include the Earth and pulsar terms in the measurement equation (i.e.
-using Equation \eqref{eq:measurement}).
-The solid green curve corresponds to a control experiment, where we drop the pulsar terms from both the synthetic data and the measurement equation.
-The vertical dashed lines mark the corresponding maxima, which nested sampling seeks.
-\newline
-
-The key observation from Figure \ref{fig:likelihood_surface_iota} is that the location of the maximum ($\iota \approx 0.7 \, {\rm rad}$) does not coincide with the true, injected value ($\iota = 1.0 \, {\rm rad}$), when only the Earth terms are included.
-This is the origin of the bias in the one-dimensional marginalised posteriors in Figure \ref{fig:bias_for_large_h}.
-In contrast, the maximum aligns with the true, injected value, when the pulsar terms are included.
-As a cross-check, agreement is also achieved, if we drop the pulsar terms from the synthetic data (which is not the situation in reality, of course) and use the Earth term only for inference.
-Similar results are obtained for other parameters, e.g.
-$\delta$.
-A discussion about why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches, are presented in Appendix \ref{sec:bias_appendix}.
-Dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}.
-The bias is not particular to our method; it is shared by all likelihood-based methods that do not include the pulsar terms in the inference model.
-Generalizing the Kalman tracker and nested sampler to correct for the bias will be a key goal of a forthcoming paper.
-
 \section{Conclusion}\label{sec:discussion}
@@
 \appendix
+\newpage
+\newpage
+\clearpage
 
@@
 The mapping onto the specific continuous-time state-space model in Section \ref{sec:model} is written down in Section \ref{sec_kalman_specific}.
-
-\subsection{Recursion equations}\label{sec_kalman_general}
-The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation
-\begin{equation}
-	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
-\end{equation}
-where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
-The Kalman filter evolves the underlying states according to
-\begin{equation}
-	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
-\end{equation}
-where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix.
-$\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline
-
-The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage.
-The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
-Specifically, the predict step proceeds as
-\begin{align}
-	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
-		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ ,
-\end{align}
-where $\hat{\boldsymbol{P}}_{k|k-1}$ is the covariance of the prediction.
-Note that the predict stage is independent of the measurements.
-The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
-\begin{align}
-	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \\
-	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \\
-	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
-		\hat{\boldsymbol{X}}_{k|k} &=\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \ , \label{eq:kalmangainupdate} \\
-			\hat{\boldsymbol{P}}_{k|k} &= \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1} \ .
-\end{align}
-Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
-$\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$.
-For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
-\newline
-
-To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$.
-In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.
-
-\subsection{State space representation of a PTA analysis}\label{sec_kalman_specific}
-We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:model} as follows. \newline
-
-We identify $\boldsymbol{X}(t)$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e. 
-\begin{equation}
-	\boldsymbol{X}(t) = \left(f_{\rm p}^{(1)}(t), f_{\rm p}^{(2)}(t), ..., f_{\rm p}^{(N)}(t)\right) \ .
-\end{equation}
-Analogously,  we package the measured pulsar frequencies as
-\begin{equation}
-	\boldsymbol{Y}(t) = \left(f_{\rm m}^{(1)}(t), f_{\rm m}^{(2)}(t), ..., f_{\rm m}^{(N)}(t) \right) \ .
-\end{equation}
-The states evolve according to the continuous stochastic differential equation (c.f. Equation \eqref{eq:frequency_evolution})
-\begin{equation}
-	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \ , \label{eq:kalmn2}
-\end{equation}
-where $\boldsymbol{A}$ is a diagonal $N \times N$ matrix,
-\begin{equation}
-	\boldsymbol{A} = \text{diag} \left(-\gamma^{(1)}, -\gamma^{(2)}, ..., -\gamma^{(N)}\right) \ ,
-\end{equation}
-and $\boldsymbol{C}(t)$ is a time-dependent vector with $n$-th component
-\begin{equation}
-	C^{(n)} =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
-\end{equation}
-The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, with
-\begin{equation}
-	\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right) \ .
-\end{equation}
-
-In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic},
-\begin{equation}
-	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ .
-\label{eq:gardenier}
-\end{equation} 
-From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2}, with
-\begin{align}
-	\boldsymbol{F}_k &= e^{\boldsymbol{A} \Delta t } \  \\
-	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
-\end{align}
-\begin{align}
-	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
-	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ,
-\end{align}
-\begin{equation}
-	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \ ,  \label{eq:appendix_noise}
-\end{equation}
-\begin{align}
-	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left(\Delta t + t_k \right) \nonumber \\ 
-	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ ,
-\end{align}
-and $\Delta t = t_{k+1} - t_k$. From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
-\begin{equation}
-	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) \ ,
-\end{equation}
-with 
-\begin{equation}
-Q^{(n)} = \frac{[\sigma^{n}]^2}{2 \gamma^{(n)}} \left[ 1 - e^{-2 \gamma^{(n)} \Delta t}\right] \ .
-\end{equation}
-
-The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
-These are defined straightforwardly from Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
-Specifically,
-$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
-The measurement covariance satisfies $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$ for all $k$.
-
-\section{Synthetic PTA specifications}\label{appendix_fiducial}
-The synthetic PTA deployed for testing in this paper is constructed to mimic the $N=47$ pulsars from the 12.5-year NANOGrav data release, whose sky positions are plotted in Figure \ref{fig:pulsar_distrib}.
-The construction recipe is described in full in Section \ref{sec:synt_pta}.
-In this appendix, we record a complete list of the pulsar parameters in Table \ref{tab:pulsar_values}, to assist the interested reader in implementing the analysis scheme and reproducing the test results in Sections \ref{sec:rep_example}--\ref{sec:bias_and_identifiability}.
-
-\begin{table*}
-\begin{tabular}{lccccc}
-	\toprule
-	JNAME &    $f_{\rm em} (t_1)$ (Hz) &  $\dot{f}^{(n)}_{\rm em}(t_1)\times 10^{16}$ (s$^{-2}$) &  $\alpha$ (deg) &  $\delta$ (deg) &  $d$ (kpc) \\
-	\midrule
-	J0023+0923 & 327.8 &      -12.3 &   5.8 &    9.4 &   1.8 \\
-	J0030+0451 & 205.5 &       -4.3 &   7.6 &    4.9 &   0.3 \\
-	J0340+4130 & 303.1 &       -6.5 &  55.1 &   41.5 &   1.6 \\
-	J0613-0200 & 326.6 &      -10.2 &  93.4 &   -2.0 &   0.9 \\
-	J0636+5128 & 348.6 &       -4.2 &  99.0 &   51.5 &   0.7 \\
-	J0645+5158 & 112.9 &       -0.6 & 101.5 &   52.0 &   1.2 \\
-	J0740+6620 & 346.5 &      -14.6 & 115.2 &   66.3 &   1.1 \\
-	J0931-1902 & 215.6 &       -1.7 & 142.8 &  -19.0 &   3.7 \\
-	J1012+5307 & 190.3 &       -6.2 & 153.1 &   53.1 &   0.7 \\
-	J1024-0719 & 193.7 &       -7.0 & 156.2 &   -7.3 &   1.2 \\
-	J1125+7819 & 238.0 &       -3.9 & 171.5 &   78.3 &   0.9 \\
-	J1453+1902 & 172.6 &       -3.5 & 223.4 &   19.0 &   1.3 \\
-	J1455-3330 & 125.2 &       -3.8 & 223.9 &  -33.5 &   0.7 \\
-	J1600-3053 & 277.9 &       -7.3 & 240.2 &  -30.9 &   1.9 \\
-	J1614-2230 & 317.4 &       -9.7 & 243.7 &  -22.5 &   0.7 \\
-	J1640+2224 & 316.1 &       -2.8 & 250.1 &   22.4 &   1.5 \\
-	J1643-1224 & 216.4 &       -8.6 & 250.9 &  -12.4 &   0.7 \\
-	J1713+0747 & 218.8 &       -4.1 & 258.5 &    7.8 &   1.3 \\
-	J1738+0333 & 170.9 &       -7.0 & 264.7 &    3.6 &   1.5 \\
-	J1741+1351 & 266.9 &      -21.5 & 265.4 &   13.9 &   1.7 \\
-	J1744-1134 & 245.4 &       -5.4 & 266.1 &  -11.6 &   0.4 \\
-	J1747-4036 & 607.7 &      -48.5 & 267.0 &  -40.6 &   7.1 \\
-	J1832-0836 & 367.8 &      -11.2 & 278.1 &   -8.6 &   2.1 \\
-	J1853+1303 & 244.4 &       -5.2 & 283.5 &   13.1 &   2.1 \\
-	J1857+0943 & 186.5 &       -6.2 & 284.4 &    9.7 &   1.2 \\
-	J1903+0327 & 465.1 &      -40.7 & 285.8 &    3.5 &   7.0 \\
-	J1909-3744 & 339.3 &      -16.1 & 287.4 &  -37.7 &   1.1 \\
-	J1910+1256 & 200.7 &       -3.9 & 287.5 &   12.9 &   1.5 \\
-	J1911+1347 & 216.2 &       -7.9 & 288.0 &   13.8 &   1.4 \\
-	J1918-0642 & 130.8 &       -4.4 & 289.7 &   -6.7 &   1.1 \\
-	J1923+2515 & 264.0 &       -6.7 & 290.8 &   25.3 &   1.2 \\
-	J1939+2134 & 641.9 &     -433.1 & 294.9 &   21.6 &   3.5 \\
-	J1944+0907 & 192.9 &       -6.4 & 296.0 &    9.1 &   1.2 \\
-	J1946+3417 & 315.4 &       -3.1 & 296.6 &   34.3 &   6.9 \\
-	J1955+2908 & 163.0 &       -7.9 & 298.9 &   29.1 &   6.3 \\
-	J2010-1323 & 191.5 &       -1.8 & 302.7 &  -13.4 &   2.4 \\
-	J2017+0603 & 345.3 &       -9.5 & 304.3 &    6.1 &   1.4 \\
-	J2033+1734 & 168.1 &       -3.1 & 308.4 &   17.6 &   1.7 \\
-	J2043+1711 & 420.2 &       -9.3 & 310.8 &   17.2 &   1.4 \\
-	J2145-0750 &  62.3 &       -1.2 & 326.5 &   -7.8 &   0.7 \\
-	J2214+3000 & 320.6 &      -15.1 & 333.7 &   30.0 &   0.6 \\
-	J2229+2643 & 335.8 &       -1.7 & 337.5 &   26.7 &   1.8 \\
-	J2234+0611 & 279.6 &       -9.4 & 338.6 &    6.2 &   1.0 \\
-	J2234+0944 & 275.7 &      -15.3 & 338.7 &    9.7 &   1.6 \\
-	J2302+4442 & 192.6 &       -5.1 & 345.7 &   44.7 &   0.9 \\
-	J2317+1439 & 290.3 &       -2.0 & 349.3 &   14.7 &   1.7 \\
-	J2322+2057 & 208.0 &       -4.2 & 350.6 &   21.0 &   1.0 \\
-	\bottomrule
-\end{tabular}
-\caption{List of the fiducial pulsar parameters for the 47 pulsars used to construct the synthetic PTA in Section \ref{sec:synt_pta}.
-The right ascension and declination of an individual pulsar in J2000 coordinates are labelled by $\alpha$ and $\delta$ respectively.
-Parameter values are obtained from the ATNF pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.}
-\label{tab:pulsar_values}
-\end{table*}
-
-\section{Dispersion of ${\theta}_{\rm gw}$ estimates}\label{sec:wasserstein}
-Every random realization of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$ leads to a different ${\boldsymbol{\theta}}_{\rm gw}$ posterior, when the synthetic data are analysed according to the procedure in Section \ref{sec:methodsummary}.
-This cosmic variance is unavoidable.
-It is a key source of uncertainty, when the analysis is performed on real astronomical data, where the specific random realizations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$ are unknown.
-To quantify the cosmic variance, one needs a metric to describe the distance between two posteriors (i.e.\ how similar they are).
-Many valid metrics of this sort exist, including those related to the Kolmogorov-Smirnoff test \citep{corder2014nonparametric}.
-In this paper, we use the WD \citep{Wasserstein,Villani2009} which is popular in machine learning \citep[e.g.][]{2017arXiv170107875A} and other domains.
-In Section \ref{sec:appendix_overview_WD} we define the WD and summarize its main properties.
-In Section \ref{sec:appendix_WD_results} we present for reproducibility the WD calculated between every pair of posteriors for each static parameter in $\boldsymbol{\theta}_{\rm gw}$ across the $10^3$ noise realisations of Section  \ref{sec:multiple_noise}.
-The results are summarised via the median values reported in Table \ref{tab:Wasserstein}.
-
-\subsection{Overview of the WD}\label{sec:appendix_overview_WD}
-The WD is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$.
-It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$.
-For this reason it is sometimes known as the ``Earth mover's distance''.
-The $p$-th order Wasserstein distance between two distributions is
-\begin{eqnarray}
-	W_p(\mu,\nu)= \left[ \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \lambda (x,y)\right]^{1/p} \label{eq:wasserstein} \ ,
-\end{eqnarray}
-for $p \in [1,\infty)$, where $\lambda(x,y)$ is the transport plan, and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e.
-$\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
-\newline
-
-The cost function can be freely chosen to suit the nature of the problem.
-Often, as in this paper, it is taken to be the absolute value function, $c(x,y) = |x-y|$.
-In general, $W_p(\mu,\nu)$ with $c(x,y)=|x-y|$ can be computed from $n$ samples by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}.
-However, for $\mu$ and $\nu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that $E[W_p(\mu,\nu)]$ converges slowly $\propto n^{-1/d}$.
-In this paper, we calculate the WD exclusively between the one-dimensional marginalized posteriors, setting $d=1$ and obtaining
-\begin{eqnarray}
-	W_p(\mu,\nu)= \left\{\int_0^1 dz \left[ F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) \right] \right\}^{1/p} \ , \label{eq:wasserstein}
-\end{eqnarray}
-where $F_{\mu}^{-1}(z)$ is the inverse cumulative distribution function of $\mu$. \newline
-
-The WD holds certain advantages over valid alternatives like the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test \citep{gelman2013bayesian,corder2014nonparametric}.
-sIt is intuitive, being the minimum cost required to transform one distribution into another.
-It satisfies the Monge-Kantorovich duality \citep{villani2003topics,Villani2009},
-\begin{eqnarray}
-	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
-\end{eqnarray}
-where $X_{\mu}$ and $Y_{\nu}$ are random variates drawn from the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value.
-It also satisfies desirable properties of a measure of distance, such as symmetry and the triangle inequality.
-For $p=1$, the WD inherits the units of the underlying distributions.
-Finally, the WD is versatile; any two distributions can be compared, irrespective of whether they are continuous, discrete, or singular.
-\newline
-
-\section{pulsar terms stuff}
-
-\subsection{Challenges of pulsar term}\label{sec:pulsar_terms_challenge}
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =0.5\textwidth ]{example-image-a}
-	\caption{Logarithm of the one-dimensional likelihood $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
-All other static parameters of $\boldsymbol{\theta}$ are held constant at their true injected values.
-In the left-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}(\alpha)$, i.e.
-inclusive of the pulsar terms.
-In the right-hand panel the Kalman filter uses Equation \ref{eq:g_func_trig_earth}, i.e.
-just the Earth terms.
-The vertical dashed lines indicate the true injected value of $\alpha$ ($=1.0$ rad) used to generate $\boldsymbol{Y}$.
-The inset plot covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
-In the left hand panel $\mathcal{L}(\alpha)$ is very noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
-This noise hampers the ability of sampling algorithms to converge to a posterior distribution.
-Conversely, in the right hand panel $\mathcal{L}(\alpha)$ is globally smooth.
-}
-	\label{fig:likelihood_comparison}
-\end{figure*}
-
-To elucidate the complications introduced by including the pulsar term, we generate some noisy synthetic data $\boldsymbol{Y}$ using the methods outlined in K23.
-$\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
-The choice of $\boldsymbol{\theta}_{\rm psr}$ corresponds to the 47 pulsars in the 12.5 year NANOGrav data set  \citep{2020ApJ...905L..34A}.
-We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig} on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad.
-We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve).
-The dashed grey line indicates the injected value of $\alpha$ (= 1.0 rad).
-The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
-\newline
-
-There are two important features to discuss in the left-hand panel of Figure \ref{fig:likelihood_comparison}.
-The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
-Similar noisy curves are obtained for $\mathcal{L}(\Omega)$ and $\mathcal{L}(\delta)$.
-For all other static parameters, the $\mathcal{L}(\Theta)$ curves are smooth and globally concave.
-The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.
-This means that whilst it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$, in practice the computation is intractable since the noise in the likelihood curve inhibits sampling algorithms to reliably converge to a definitive posterior.
-Moreover, $\mathcal{L}(\alpha)$ is only a one-dimensional function.
-The challenge for sampling algorithms is further exacerbated when we consider the three-dimensional parameter space in $\alpha - \delta - \Omega$, or the full $7 + 5N$ dimensions of Section \ref{sec:ss_params}.
-\newline
-
-Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\Theta)$ curves whereas other static parameters do not?
-These parameters appear in the pulsar term as a product in the pulsar-dependent phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, where we remind the reader that $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
-Over the typical pulsar distances $d^{(n)} \sim 1$ kpc, a 10 nHz GW can accumulate $10^3$ cycles.
-This creates a confusion problem when taking the cosine of  $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ since we do not know how many cycles the wave has gone through, modulo $2 \pi$.
-\textcolor{red}{TK: this explanation is rubbish!
-How can we say this better?}.
-Conversely, if we drop the pulsar term,  the $\mathcal{L}(\alpha)$ curve (as well as $\mathcal{L}(\Omega)$ , $\mathcal{L}(\delta)$   )
-This is shown by the orange curve in the right-hand panel of Figure \ref{fig:likelihood_comparison} where we plot $\mathcal{L}(\alpha)$, analogous to the left-hand panel, but now use Equation \ref{eq:g_func_trig_earth} in our Kalman filter (i.e.
-just the Earth terms) rather than the full Equation \ref{eq:g_func_trig}.
-In this case there is no confusion regarding the number of cycles the wave has gone through over a distance $d^{(n)}$ and the curve is smooth.
-
-\section{KF and NS review}
-\section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
-The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters controlling the intrinsic frequency evolution of the pulsars in the array and the GW source, i.e.
-\begin{equation}
-	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
-\end{equation}
-with
-\begin{equation}
-	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
-\end{equation}
-and
-\begin{equation}
-	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
-\end{equation}
-where $\delta$, $\alpha$, and $\iota$ are the declination, right ascension and inclination of the GW source respectively \footnote{$\iota$ is the angle between the unit normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer's line of sight, i.e.
-$\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$.}.
-In Equation \eqref{eq:params3} we reparameterize the two GW polarisation amplitudes, $h_{+}$ and $h_{\times}$, in terms of $\iota$ and the characteristic wave strain $h_0$ through
-\begin{align}
-	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
-	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ .
-\end{align}
-We use the parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work.
-A PTA containing $N$ pulsars comprises $7 + 5N$ parameters to estimate.
-Typically the pulsar parameters are constrained better \textit{a priori} by electromagnetic observations than the GW parameters.
-For example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information about $\delta$ and $\alpha$.
-\newline
-
-In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e.
-the model evidence).
-In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter.
-In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference.
-Model selection and the specification of the null model are described in Section \ref{sec:model_selection}.
-A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
-The method complements traditional PTA analyses because (i) it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations; and (ii) it infers ${\boldsymbol{\theta}}$ conditional on the specific, time-ordered, random realization of the noise corresponding to the observed sequence of TOAs rather than on ensemble-averaged quantities like the power spectral density.
-
-\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
-
-\begin{figure*}
-	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
-	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column).
-The top panels show the true pulsar state $f_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$  (green curve).
-We subtract $f_{\rm em}^{(1)}(t)$ to better illustrate the stochastic wandering of the pulsar frequency.
-In the left column, the blue/green solutions overlap almost perfectly; in the right column, they do not.
-The middle panels show the true measured frequency $f_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve).
-Again we subtract $f_{\rm em}^{(1)}(t)$.
-In the left column the red/magenta solutions overlap almost perfectly; in the right column, they do not.
-The bottom panels show the residual or innovation $\epsilon(t) =f_{\rm m}^{(1)}(t) - \hat{f}_{\rm m}^{(1)}(t)$.
-In the right column $\Omega$ is displaced from its true value by 20\%.
-Results are shown for a single pulsar.
-}
-	\label{fig:kalman_example}
-\end{figure*}
 
@@
 
-\subsection{Nested sampling}\label{sec:nested_sampling}
+\subsection{Recursion equations}\label{sec_kalman_general}
+The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation
+\begin{equation}
+	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
+\end{equation}
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
+The Kalman filter evolves the underlying states according to
+\begin{equation}
+	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
+\end{equation}
+where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix.
+$\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline
+
+The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage.
+The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
+Specifically, the predict step proceeds as
+\begin{align}
+	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
+		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ ,
+\end{align}
+where $\hat{\boldsymbol{P}}_{k|k-1}$ is the covariance of the prediction.
+Note that the predict stage is independent of the measurements.
+The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
+\begin{align}
+	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \\
+	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \\
+	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
+		\hat{\boldsymbol{X}}_{k|k} &=\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \ , \label{eq:kalmangainupdate} \\
+			\hat{\boldsymbol{P}}_{k|k} &= \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1} \ .
+\end{align}
+Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
+$\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$.
+For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
+\newline
+
+To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$.
+In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.
+
+\subsection{State space representation of a PTA analysis}\label{sec_kalman_specific}
+We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:model} as follows. \newline
+
+We identify $\boldsymbol{X}(t)$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e. 
+\begin{equation}
+	\boldsymbol{X}(t) = \left(f_{\rm p}^{(1)}(t), f_{\rm p}^{(2)}(t), ..., f_{\rm p}^{(N)}(t)\right) \ .
+\end{equation}
+Analogously,  we package the measured pulsar frequencies as
+\begin{equation}
+	\boldsymbol{Y}(t) = \left(f_{\rm m}^{(1)}(t), f_{\rm m}^{(2)}(t), ..., f_{\rm m}^{(N)}(t) \right) \ .
+\end{equation}
+The states evolve according to the continuous stochastic differential equation (c.f. Equation \eqref{eq:frequency_evolution})
+\begin{equation}
+	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \ , \label{eq:kalmn2}
+\end{equation}
+where $\boldsymbol{A}$ is a diagonal $N \times N$ matrix,
+\begin{equation}
+	\boldsymbol{A} = \text{diag} \left(-\gamma^{(1)}, -\gamma^{(2)}, ..., -\gamma^{(N)}\right) \ ,
+\end{equation}
+and $\boldsymbol{C}(t)$ is a time-dependent vector with $n$-th component
+\begin{equation}
+	C^{(n)} =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
+\end{equation}
+The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, with
+\begin{equation}
+	\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right) \ .
+\end{equation}
+
+In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic},
+\begin{equation}
+	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ .
+\label{eq:gardenier}
+\end{equation} 
+From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2}, with
+\begin{align}
+	\boldsymbol{F}_k &= e^{\boldsymbol{A} \Delta t } \  \\
+	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
+\end{align}
+\begin{align}
+	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
+	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ,
+\end{align}
+\begin{equation}
+	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \ ,  \label{eq:appendix_noise}
+\end{equation}
+\begin{align}
+	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left(\Delta t + t_k \right) \nonumber \\ 
+	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ ,
+\end{align}
+and $\Delta t = t_{k+1} - t_k$. From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
+\begin{equation}
+	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) \ ,
+\end{equation}
+with 
+\begin{equation}
+Q^{(n)} = \frac{[\sigma^{n}]^2}{2 \gamma^{(n)}} \left[ 1 - e^{-2 \gamma^{(n)} \Delta t}\right] \ .
+\end{equation}
+
+The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
+These are defined straightforwardly from Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
+Specifically,
+$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
+The measurement covariance satisfies $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$ for all $k$.
+
+\section{Nested sampling}
+
 We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
@@
 
-%See https://arxiv.org/pdf/2102.12478.pdf for a good NS explanation
-
-\subsection{Model selection}\label{sec:model_selection}
-The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
-\begin{equation}
-	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
-\end{equation}
-Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
-$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data, and is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
-The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW signal compared to there being no GW signal present.
-
-\subsection{Summary of workflow}\label{sec:methodsummary}
+\section{Complete workflow summary}
+
 For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
@@
 
-\section{Bias from dropping the pulsar terms}\label{sec:bias_appendix}
-Why does dropping the pulsar terms in Equation \ref{eq:g_func_trig} bias some components of $\boldsymbol{\theta}_{\rm gw}$, whilst leaving others nearly unaffected?
-For instance, why is the bias in $\iota$ strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
-In this appendix, we examine the above questions in the context of the Kalman analysis in this paper, complementing work on similar questions in the context of traditional PTA analyses \citep{Zhupulsarterms,Chen2022}.
-\newline
-
- When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is modulated by the interference of two equal-amplitude, phase-shifted oscillations (c.f.
-Equation \eqref{eq:measurement}).
-Specifically we have an
-``Earth-term oscillation'' of form $y_{\rm Earth} = A \cos \xi$, with amplitude $A$ and phase $\xi$, and a ``pulsar-terms oscillation'' of form $y_{\rm pulsar} = A \cos (\xi + \kappa)$, with phase offset $\kappa$.
-The two terms are related to the measurement equation, Equation \eqref{eq:g_func_trig}, through
-\begin{align}
-	1 - g^{(n)}(t) &=  y_{\rm Earth}  - y_{\rm pulsar} \  \nonumber \\
-		 &= 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ . \label{eq:psr_interference}
-\end{align}
-The interference in Equation \ref{eq:psr_interference} multiplies the amplitude by $ 2 \sin (\kappa/2)$, but leaves the frequency unchanged.
-When we model the data using the Kalman filter without the pulsar terms, $y_{\rm Earth}$ must try to compensate for the modified amplitude on the right-hand side of Equation \ref{eq:psr_interference} by adjusting the parameters which govern the amplitude, such as $h_0$ and $\iota$.
-Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
-The wave amplitude is independent of $\Omega$, and so $\Omega$ exhibits no bias.
-A shift in phase is covariant with the amplitude, so $\Phi_0$ also exhibits a small bias.
-The same hierarchy of biases is seen also in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are over-constrained whilst $\Omega$ is well-estimated.
-\newline
-
-Quantifying the bias for practical, astronomical PTA continuous wave searches requires a thorough exploration of the SMBHB parameter space which falls outside the scope of this introductory paper.
-The bias also depends on the PTA configuration.
-If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate.
-This is what we observe for the synthetic data, for example in Figure \ref{fig:corner_plot_2}; the marginalised posterior broadens as the SNR decreases, until it overwhelms the shift due to the bias.
-Conversely if the bias is sufficiently strong, or the source is sufficiently loud, then the bias may dominate.
-\newline
-
-%%%%%%%%%%%%%%%
+\section{}
+
+\section{Challenges of the pulsar term}
+In this section....
 
 \begin{figure*}
-	\setkeys{Gin}{width=\linewidth}
-
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_0}
-		\caption{$\Omega$}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_1}
-		\caption{$\Phi_0$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_2}
-		\caption{$\psi$} \label{figWD_psi}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_3}
-		\caption{$\iota$}
-	\end{subfigure}
-	\medskip
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_4}
-		\caption{$\delta$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_5}
-		\caption{$\alpha$}
-	\end{subfigure}
-	\hfill	
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\textwidth]{images/WD_6}
-		\caption{$h$}
-	\end{subfigure}
-	\caption{First moment of the WD, $W_1$ (colour scale), calculated between each pair of one-dimensional posteriors for the representative system in Table \ref{tab:parameters_and_priors} across $10^3$ realisations as discussed in Section \ref{sec:appendix_WD_results}, for each component of $\boldsymbol{\theta}_{\rm gw}$.
-$W_1$ provides an upper bound on the difference in expected values between any two probability distributions.
-$W_1$ is generally low across all parameters and posteriors, but some posteriors converge poorly and return high values of $W_1$.
-The $W_1$ values inferred for $\psi$ and $\alpha$ are correlated with Pearson correlation coefficient = 0.96.
-An example of a noise realisation with particularly high $W_1$ is labelled with an asterisk in (c) and is discussed in the text.} \label{fig:pairwise_wasserstein}
+	\includegraphics[width=\textwidth, height =0.5\textwidth ]{example-image-a}
+	\caption{Logarithm of the one-dimensional likelihood $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
+All other static parameters of $\boldsymbol{\theta}$ are held constant at their true injected values.
+In the left-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}(\alpha)$, i.e.
+inclusive of the pulsar terms.
+In the right-hand panel the Kalman filter uses Equation \ref{eq:g_func_trig_earth}, i.e.
+just the Earth terms.
+The vertical dashed lines indicate the true injected value of $\alpha$ ($=1.0$ rad) used to generate $\boldsymbol{Y}$.
+The inset plot covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
+In the left hand panel $\mathcal{L}(\alpha)$ is very noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
+This noise hampers the ability of sampling algorithms to converge to a posterior distribution.
+Conversely, in the right hand panel $\mathcal{L}(\alpha)$ is globally smooth.
+}
+	\label{fig:likelihood_comparison}
 \end{figure*}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/likelihood_surface_alpha_psi}
-	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\alpha$-$\psi$ plane for a single noise realisation.
-The red point labels the maximum, and the magenta point labels the median values of the inferred one-dimensional posteriors for $\alpha$ and $\psi$.
-The likelihood is normalised with respect to the absolute value of the maximum; note the compressed colour scale.
-A multi-modal structure is clearly visible for this noise realisation.
-The nested sampler finds a local maximum (magenta point) instead of the global maximum (red point)}
-	\label{fig:likelihood_surface_alpha_psi}
-\end{figure}
-
-%%%%%%%%%%%%%%%%Scratch space
-
-%Taking a tangible example, for pulsar J0023+0923 ATNF returns a frequency $f_{\rm em}^{(n)} (t_1) \sim 327.8$ Hz with an error $\epsilon \sim 4 \times 10^{13}$ Hz.
-	The prior on $f_{\rm em}^{(n)} (t_1) \sim 327.8$ for this pulsar is then Uniform$(327.8 - 4 \times 10^{10},327.8 + 4 \times 10^{10} )$
-
-%Alternatively, one can consider instead the alternative parametrisation in terms of $h_{+}$ and $h_{\times}$. 
-%
-%from Equation \ref{eq:hij} the amplitude of the plane GW perturbation at a particular sky location $\boldsymbol{n}$ is a linear combination of $h_{+}$ and $h_{\times}$. 
-%
-%
-%
-%Considering a single pulsar where the only unknown parameters of the model are $h_{+}$ and $h_{\times}$, the system is clearly under-determined - we have one equation with two unknowns - and the parameters are not identifiable
-%
-%
-
-%
-%
-%
-%
-%For instance, in Figure \ref{fig:omega_likelihood} we pass different values of $\Omega$ in the range $10^{-9}$ to $10^{-5}$ Hz into the Kalman filter, whilst holding the remaining parameters constant, and plot the retuned log-likelihood as a function of $\Omega$.
-%
-
-%
-%The second is that the log-likelihood curves for $\Omega, \delta$ and $\alpha$ are much more noisy than in the Earth-terms case.
-Whilst clear optima exist above the noise, and whilst local to the true value the curve becomes much more smooth, this noise presents an additional challenge to typical likelihood inference techniques which perform best on smooth likelihoods with a single clear global optima.
-It affects these parameters in particular since Equation \eqref{eq:g_func_trig} contains a phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, which is essentially a combination of these 3 parameters ($ \boldsymbol{n}$ is defined by $\delta$ and $\alpha$).
-When calculating a likelihood we are asking "How likely is this data, given these parameters".
-The inclusion of the pulsar terms results in effectively 2 continuous waves with different frequencies and phase.
-Perturbing e.g.
-$\alpha$ influences wave 1 in some non-liner way and wave 2 in some different non linear way.
-Whether this new set of parameters makes the data more likely depends on the particular combination of those two waves.
-\textcolor{red}{TK: this explanation of why these parameters is clunky because it is not properly clear to myself.
-Think in terms of $g(\theta)$ and $g(\hat{\theta})$}.
-
-%
-%.
-In turn, each posterior distribution has a median value.
-We plot the distribution of these medians over the 1000 noise realisations for each parameter of $\boldsymbol{\theta}_{\rm gw}$ in Figure \ref{fig:median_distriubutins}.
-For each of these distributions we can also calculate the median value (i.e.
-the median of the medians) and compare it with the true injection value.
-This is shown by the red and orange dashed lines respectively in the Figure.
-Analogous to the results we saw in Figure \ref{fig:corner_plot_2}, the distributions of the median values of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are very narrow, with a generally small discrepancy between the inferred value and the injected value.
-Conversely, the distributions for $\iota$ and $h_0$ are much broader and exhibit a bias of $\sim 0.18$ radians and $0.17 \times 10^{-12}$ respectively.
-This agrees with the results we discussed for the 9 noise realisations in Figure \ref{fig:corner_plot_2} where a bias in $\iota$ and $h_0$ was also observed.
-In addition to  $\iota$ and $h_0$  exhibiting a strong bias, the parameters $\psi$ and $\alpha$ also exhibit a small bias of magnitude $\sim 0.1$ and $\sim 0.04$ radians respectively.
-This bias is again a result of dropping the pulsar terms from the measurement equation and will be discussed in .
+
+To elucidate the complications introduced by including the pulsar term, we generate some noisy synthetic data $\boldsymbol{Y}$ using the methods outlined in K23.
+$\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
+The choice of $\boldsymbol{\theta}_{\rm psr}$ corresponds to the 47 pulsars in the 12.5 year NANOGrav data set  \citep{2020ApJ...905L..34A}.
+We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig} on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad.
+We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve).
+The dashed grey line indicates the injected value of $\alpha$ (= 1.0 rad).
+The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
+\newline
+
+There are two important features to discuss in the left-hand panel of Figure \ref{fig:likelihood_comparison}.
+The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
+Similar noisy curves are obtained for $\mathcal{L}(\Omega)$ and $\mathcal{L}(\delta)$.
+For all other static parameters, the $\mathcal{L}(\Theta)$ curves are smooth and globally concave.
+The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.
+This means that whilst it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$, in practice the computation is intractable since the noise in the likelihood curve inhibits sampling algorithms to reliably converge to a definitive posterior.
+Moreover, $\mathcal{L}(\alpha)$ is only a one-dimensional function.
+The challenge for sampling algorithms is further exacerbated when we consider the three-dimensional parameter space in $\alpha - \delta - \Omega$, or the full $7 + 5N$ dimensions of Section \ref{sec:ss_params}.
+\newline
+
+Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\Theta)$ curves whereas other static parameters do not?
+These parameters appear in the pulsar term as a product in the pulsar-dependent phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, where we remind the reader that $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
+Over the typical pulsar distances $d^{(n)} \sim 1$ kpc, a 10 nHz GW can accumulate $10^3$ cycles.
+This creates a confusion problem when taking the cosine of  $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ since we do not know how many cycles the wave has gone through, modulo $2 \pi$.
+\textcolor{red}{TK: this explanation is rubbish!
+How can we say this better?}.
+Conversely, if we drop the pulsar term,  the $\mathcal{L}(\alpha)$ curve (as well as $\mathcal{L}(\Omega)$ , $\mathcal{L}(\delta)$   )
+This is shown by the orange curve in the right-hand panel of Figure \ref{fig:likelihood_comparison} where we plot $\mathcal{L}(\alpha)$, analogous to the left-hand panel, but now use Equation \ref{eq:g_func_trig_earth} in our Kalman filter (i.e.
+just the Earth terms) rather than the full Equation \ref{eq:g_func_trig}.
+In this case there is no confusion regarding the number of cycles the wave has gone through over a distance $d^{(n)}$ and the curve is smooth.
+
+In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e.
+the model evidence).
+In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter.
+In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference.
+Model selection and the specification of the null model are described in Section \ref{sec:model_selection}.
+A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
+The method complements traditional PTA analyses because (i) it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations; and (ii) it infers ${\boldsymbol{\theta}}$ conditional on the specific, time-ordered, random realization of the noise corresponding to the observed sequence of TOAs rather than on ensemble-averaged quantities like the power spectral density.
+
+\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
+
+%See https://arxiv.org/pdf/2102.12478.pdf for a good NS explanation
+
+\subsection{Model selection}\label{sec:model_selection}
+The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
+\begin{equation}
+	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
+\end{equation}
+Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
+$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data, and is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
+The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW signal compared to there being no GW signal present.
 

### 2023-11-15 py_src/configs/config_batch.py

@@ -24,18 +24,42 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
     
     
 
-# MULTIPLE NOISE REALISATIONS
-N = 100
-seeds = np.arange(1235+10,1235+10+N,1)
-strain = 5e-15
-model = "pulsar"
+# # MULTIPLE NOISE REALISATIONS
+# N = 100
+# seeds = np.arange(1235+10,1235+10+N,1)
+# strain = 5e-15
+# model = "pulsar"
+
+# with open('batch.sh','w') as b: 
+
+#     for s in seeds:
+#         arg_name = f"multiple_noise_canonical_{model}_{strain}_{s}"
+#         create_slurm_job(arg_name,strain,model,s)
+#         b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+
+
+
+
+
+N = 10
+#seeds = np.arange(1235+10,1235+10+N,1)
+seeds = [1251]
+
+strains = [1e-12,5e-15]
+models = ["earth", "pulsar"]
 
 with open('batch.sh','w') as b: 
 
     for s in seeds:
-        arg_name = f"multiple_noise_canonical_{model}_{strain}_{s}"
-        create_slurm_job(arg_name,strain,model,s)
-        b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+        for h in strains:
+            for m in models:
+                arg_name = f"reproduce_eg_canonical_{m}_{h}_{s}"
+                create_slurm_job(arg_name,h,m,s)
+                b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+       
+
+
+
        
 
 #BAYES PLOT

### 2023-11-20 manuscript/manuscript_2/paper.tex
test a different prior on sigma p

@@
 \begin{abstract}	
-
 	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.
-
-	 The gravitational wave signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
-
+The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
 The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-
 We show how to extend previous state-space analyses to include the pulsar terms.
-
-It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy relative to the Earth terms approach by X and the sensitivity by Y
-
+It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the Earth-terms-only approach.
 \end{abstract}
@@
 Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
-SMBHBs which are sufficiently close and massive may be individually resolvable, though conclusive evidence is currently lacking \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
+SMBHBs which are sufficiently close and massive may be individually resolvable, though no definitive detection has yet been made \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
 \newline
@@
 Firstly we extend the methods presented in K24 to include the pulsar term.
-We show how to reparametrise the model in terms of an additional pulsars-specific phase term.
+We show how to reparametrise the state-space model in terms of an additional pulsars-specific phase term.
 We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}.
-Secondly we quantify how the inclusion of the pulsar term affects the accuracy with which the static GW parameters can be estimated, and the minimal detectable GW strain.
+Secondly we quantify how the inclusion of the pulsar term affects (i) the accuracy with which the static GW parameters can be estimated, and (ii) the minimal detectable GW strain, compared to the case where the pulsar terms are omitted.
 \newline
@@
 The paper is organised as follows.
-In Section \ref{sec:2} we briefly review the state-space model of PTA data analysis introduced in \cite{KimpsonPTA}.
+In Section \ref{sec:2} we briefly review the state-space model of PTA data analysis introduced in K24.
 In Section \ref{sec:pulsar_term} we show how to include the pulsar term in the model via a convenient reparametrisation.
-In Section \ref{sec:rep_example1} we test the updated model which includes the pulsar term on synthetic data for a single representative GW source.
-In Section \ref{sec:earth_vs_psr} we compare the parameter estimation accuracy and the minimal detectable GW strain when including the pulsar terms to the case where the pulsar terms are dropped.
-In Section \ref{sec:parameter_space} we extend the tests of Section \ref{sec:rep_example1} to cover an astrophysically relevant domain of SMBHB source parameters.
+In Section \ref{sec:rep_example1} we test the updated model, inclusive of the pulsar term, on synthetic data for both a single representative GW source and across an astrophysically relevant domain of SMBHB source parameters.
+In Section \ref{sec:earth_vs_psr} we compare the parameter estimation accuracy and the minimal detectable GW strain when including the pulsar terms to the case where the pulsar terms are omitted.
 Conclusions are drawn in Section \ref{sec:discussion}.
@@
 
-%In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations.
-The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density.
-Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations summarized by the power spectral density, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization.
-The new and existing approaches are therefore complementary.
-In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}.
-We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the static GW parameters and calculate the marginal likelihood (i.e.
-the model evidence) for model selection.
-\newline
-
-%The paper is organised as follows.
-In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW.
-In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence.
-In Section \ref{sec:testing} we describe how we create synthetic validation data with which to test the method.
-We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source.
-In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant domain of SMBHB source parameters.
-In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates.
-
 \section{State-Space Formulation}\label{sec:2}
@@
 The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$.
-The intrinsic spin frequency of the $n$-th pulsar as measured in the local, freely-falling rest frame of the pulsar's centre of mass, $f_{\rm p}^{(n)}(t)$, evolves according to a stochastic differential equation.
+The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, as measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation.
 The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW.
@@
 In Section \ref{sec:ss_params} we summarise the parameters of the model.
-The model is formulated in terms of pulse frequency time series as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
+The model is formulated in terms of pulse frequency time series as a proof-of-principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
 It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization we defer to future work.
@@
 \end{equation}
-where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
+where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
 \begin{align}
@@
 \end{align}
-In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
 The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
-PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and can be approximated accurately by
+PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$.
+Consequently, 	$f_{\rm em}^{(n)}(t)$ can be approximated accurately by
 \begin{equation}
@@
 However, Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} do contain certain modelling omissions and idealisations which much be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
-These include the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and the formal interpretation of $d^2 f_{\rm p} / dt^2$ as $\xi(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.
+These include the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and the formal interpretation of $d^2 f_{\rm p} / dt^2$ due to $\xi(t)$ in Equation \eqref{eq:frequency_evolution} not being differentiable.
 
@@
 \end{align}
-where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor and $\Phi_0$ the phase offset of the GW with respect to some reference time.
+where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor, and $\Phi_0$ the phase offset of the GW with respect to some reference time.
 \newline
@@
 These parameters enter the model through Equation \eqref{eq:g_func_trig} where $H_{ij} = H_{ij}(h_0, \iota, \psi, \delta, \alpha)$ and $\boldsymbol{n}=\boldsymbol{n}(\delta,\alpha)$.
-The complete set of $7 + 5N$ static parameters is denoted $\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw}$.
+The complete set of $7 + 5N$ static parameters is denoted $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$.
 While we have no prior information about the values of the members of $\boldsymbol{\theta}_{\rm gw}$, the members of $\boldsymbol{\theta}_{\rm psr}$
-do have constraints from electromagnetic observations; for example
-for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
+do have constraints from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
 
@@
 \begin{equation}
-	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
+	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ , \label{eq:model_evidence}
 \end{equation}
@@
 For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
+Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e.
+$\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$.
 Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
-We refer the reader to Appendix X for a review of Kalman filtering, nested sampling and the combined workflow in the context of PTA searches for GWs.
+We refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, nested sampling and the combined workflow in the context of PTA searches for GWs.
 \newline
@@
 In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar term for likelihood-based inference methods like nested sampling.
-This new parametrisation is validated with synthetic data in Section \ref{sec:testing}.
+This new parametrisation is validated with synthetic data in Section \ref{sec:rep_example1}.
 
@@
 The GW signal is composed of an ``Earth term", corresponding to the $\cos(-\Omega t + \Phi_0)$ component, and a ``pulsar term", corresponding to the $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \}$ component.
-The Earth term describes the GW phase at the observer on Earth, while the pulsar term describes the GW phase at each the individual pulsar.
+The Earth term describes the GW phase at the observer on Earth, while the pulsar term describes the GW phase at the individual pulsar.
 The Earth term depends only on the GW source parameters and so is shared across all pulsars.
@@
 
-Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases which cannot be coherently summed and are instead generally considered as a source of self noise.
+Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases which cannot be coherently summed and are instead generally considered as a source of self-noise.
 The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties much greater than the typical target GW wavelength.
 It is therefore not possible to set the phase of the pulsar term.
-Consequently the pulsar terms is typically dropped from in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space, making the parameter estimation computationally cheaper.
-Within the state-space model described in Section \ref{sec:2}, dropping the pulsar terms equates to using a modified measurement equation
-\begin{equation}
-	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) \ , 
+Consequently the pulsar term is typically dropped from in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space, since the $N$ values of $d^{(n)}$ do not have to be inferred, making the parameter estimation computationally cheaper.
+Within the state-space model described in Section \ref{sec:2}, dropping the pulsar terms equates to using a modified measurement equation, c.f.
+Equations \eqref{eq:measurement} and \eqref{eq:g_func_trig},
+\begin{equation}
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) + \varepsilon^{(n)}(t) \ , 
 	\label{eq:measuremen_earth}
@@
 \begin{equation}
-	\chi^{(n)} = \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \mod 2 \pi 
-\end{equation}
-Such that Equation \eqref{eq:g_func_trig} becomes
+	\chi^{(n)} \equiv \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \mod 2 \pi \ ,
+\end{equation}
+such that Equation \eqref{eq:g_func_trig} becomes
 \begin{align}
@@
 \end{align}
-and the static pulsar parameters to be inferred become
-\begin{equation}
-	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
-\end{equation}
-whilst $\boldsymbol{\theta}_{\rm gw}$  remain unchanged. We define $\boldsymbol{\theta'} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta'}_{\rm gw}$. \newline
-
 The reparametrisation in terms of $\chi^{(n)}$ treats the pulsar-dependent phase correction $\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ in the argument of the cosine of the pulsar term as an explicit parameter to be inferred for each pulsar.
-The advantage of this approach, compared to trying to infer $d^{(n)}$ directly, is that the resulting likelihood surfaces are smoother enables sampling algorithms to efficiently find the likelihood maxima and estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
-This point is discussed in more detail in Appendix YY.
+The advantage of this approach, compared to trying to infer $d^{(n)}$ directly, is that the resulting likelihood surface across the parameter space (i.e.
+the shape of the function $\mathcal{L}(\boldsymbol{\theta})$) is smoother, with smaller derivatives.
+This enables sampling algorithms to efficiently find the likelihood maxima and estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+This point is discussed in more detail in Appendix \ref{sec:psr_term_challenges}.
 The new parametrisation also does not increase the dimension of the parameter space, because we have traded $d^{(n)}$ for $\chi^{(n)}$.
-
-\section{Representative example: single SMBHB source including the pulsar terms}\label{sec:rep_example1}
+The static parameters specific to the pulsars in the array (c.f.
+Equation \eqref{eq:psrparams}) are now
+\begin{equation}
+	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1) \right\}_{1\leq n \leq N} \ ,  \label{eq:psrparamsnew}
+\end{equation}
+whilst $\boldsymbol{\theta}_{\rm gw}$  remains unchanged.
+We exclude $\chi^{(n)}$ from membership of $\boldsymbol{\theta'}_{\rm psr}$ as the other elements of $\boldsymbol{\theta'}_{\rm psr}$ depend solely on the properties of the pulsar, with no dependence on the properties of the GW source.
+In contrast $\chi^{(n)}$ is a cross term that depends on both the individual pulsar and the GW.
+We define the complete set of $7 +4N + N$ static parameters to be inferred as $\boldsymbol{\theta'} = \boldsymbol{\theta'}_{\rm gw} \cup \boldsymbol{\theta'}_{\rm psr} \cup \chi^{(n)}$ .
+\newline
+
+\section{Validation of pulsar-terms inclusion with synthetic data}\label{sec:rep_example1}
 In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}.
-As outlined in Section \ref{sec:pulsar_term}, the analysis scheme consists of a Kalman filter to track the state evolution of the intrinsic pulsar spin frequency which provides a likelihood for nested sampling.
-Nested sampling then returns estimates of the static parameters of the model and $\mathcal{Z}$.
-We refer the reader to Appendix X and K24 for details of the Kalman filter - nested sampling scheme.
+As outlined in Section \ref{sec:pulsar_term}, the analysis scheme consists of a Kalman filter to track the state evolution of the intrinsic pulsar spin frequency $f_{\rm p}^{(n)}(t)$ which provides a likelihood for nested sampling.
+Nested sampling then returns estimates of $\boldsymbol{\theta'}$ and $\mathcal{Z}$.
+We refer the reader to Appendix \ref{sec:kalman} and K24 for details of the Kalman filter - nested sampling scheme.
 We apply the analysis scheme to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
 The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-The goal of this section is to validate the analysis scheme when the pulsar terms are included.
+The goal of this section is to validate the analysis scheme when the pulsar terms are included in the inference model.
 \newline
@@
 In Section \ref{sec:synthetic_data} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme.
-In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ and $\mathcal{Z}$.
+In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ and $\mathcal{Z}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$.
 The scheme is validated on multiple noise realisations of the pulsar process noise and the measurement noise in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
-This section considers only a particular, arbitrary set of injection parameters $\boldsymbol{\theta'}$.
-Extension to broader parameter spaces is considered in Section X.
+In Section \ref{sec:parameter_space} we extend the tests of Section  \ref{sec:pe_and_ms}
+to cover a broader parameter space and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.
 
@@
 			\hline
-			%		\vspace{1mm}\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
 			& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
-			& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
-			\vspace{1mm} $\boldsymbol{\theta}_{\rm psr}$ &  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
-			\vspace{1mm} & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
+			\multirow{2}{2mm}{$\boldsymbol{\theta'}_{\rm psr}$} & $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
+		     & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
 			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
+			\hline
+			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
 			\bottomrule
 		\end{tabular}
-		\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the prior used for Bayesian inference on each parameter (rightmost column).
-The top and bottom halves of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ respectively.The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}.
-The subscript ``SC'' indicates that the injected value has been calculated using Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} in \protect \cite{Shannon2010}.
+		\caption{Summary of injected static parameters used for generating synthetic data for validation of the analysis scheme when including the pulsar-term in Section \ref{sec:rep_example}.
+The prior used for Bayesian inference (Section \ref{sec:pe_and_ms}) on each parameter is also displayed (rightmost column).
+The top, middle and bottom sections of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$, $\boldsymbol{\theta}_{\rm psr}$ and $\chi^{(n)}$ respectively.
+The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synthetic_data}.
+The subscript ``SC'' on the pulsar noise amplitude $\sigma^{(n)}$ indicates that the injected value has been calculated using Equation and \eqref{eq:sigmap_f} the empirical timing noise model for MSPs from \protect \cite{Shannon2010}.
 The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
-No priors are set for $d^{(n)}$ and $\gamma^{(n)}$ since those parameters do not enter the inference model,  as discussed in Section \ref{sec:priors}, c.f.
-Equation \eqref{eq:psrreduced}.
+The choice of priors on $\boldsymbol{\theta}_{\rm psr}$ is discussed in Appendix \ref{sec:set_priors}.
 		}
@@
 	\end{table*}
-In order to generate synthetic frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$ Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+In order to generate synthetic frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
 This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
@@
 Compared to the other 46 pulsars of the array, PSR J1939+2134 is a notable outlier with an estimated $\sigma^{(n)}$ much larger than the median.
-In practice, in order to avoid setting an overly broad prior on $\sigma^{(n)}$ (see Section \ref{sec:set_priors}) for validation purposes, we artificially set $\sigma^{(n)} = 10^{-20} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within an reasonable width log-uniform prior.
+In practice, in order to avoid setting an overly broad prior on $\sigma^{(n)}$ (see Appendix \ref{sec:set_priors}) for validation purposes, we artificially set $\sigma^{(n)} = 10^{-24} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within an reasonable width log-uniform prior.
 This is done purely for testing convenience; it is straightforward to expand the prior when analysing real, astronomical data.
@@
 
+\subsection{Parameter estimation} \label{sec:pe_and_ms}
+
 The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
@@
 
-\subsection{Prior distribution}\label{sec:set_priors}
-To deploy the analysis scheme of K24 it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters. \newline
-
-For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}.
-For $\pi(\boldsymbol{\theta}_{\rm psr})$ \textit{a priori} information from electromagnetic observations does exist.
-We adopt constrained uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database.
-By using wider-than-necessary priors we expose the analysis scheme to a more stringent test.
-We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$) which covers the range of values inferred from Equation \ref{eq:sigmap_f}.
-We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.
-We do not set a prior on $\gamma^{(n)}$, because $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and consequently $\gamma^{(n)}$ is effectively ``unobservable'' over $T_{\rm obs}$.
-For validation purposes it is therefore sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value.
-This reduces the total dimension of the parameter space to $7 + 4N$.
-The chosen prior $\boldsymbol{\theta}$ is the same as used in K24, with the addition of a prior on $\chi^{(n)}$ due to the inclusion of the pulsar term.
-\newline
-
-The chosen priors for the representative analysis in Section \ref{sec:rep_example1} are summarised in \ref{Table tab:parameters_and_priors}
-
-\subsection{Parameter estimation} \label{sec:pe_and_ms} 
 In this section, we calculate the posterior probability distribution for the static parameters $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$ and compare it to the known, injected values.
@@
 We apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$.
-\newline
+The specification of the priors on $\boldsymbol{\theta}$ is described in section \ref{sec:set_priors} and summarised in \ref{Table tab:parameters_and_priors}
 
@@
 Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
+
+\subsection{Extension to broader parameter space} \label{sec:parameter_space}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/pp_plot_new1}
+	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
+PP plot).
+The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}.
+Each coloured curve corresponds to a different parameter (see legend).
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
+The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals.
+For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope.
+$\Omega$ is generally well-estimated (i.e.
+it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e.
+the curves lie above the unit diagonal for low credible intervals, and below the unit diagonal for high credible intervals).
+This is due to a modelling bias whose origin is discussed in Sections \ref{sec:parameter_estim} and \ref{sec:pulsar_terms_drop}.}
+	\label{fig:parameter_space}
+\end{figure}
+Section \ref{sec:rep_example} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}.
+In this section we test the method for various systems, varying the SMBHB source parameters through astrophysically relevant ranges.
+\newline
+
+We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
+For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
+To summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself.
+In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
+\newline
+
+Figure \ref{fig:parameter_space} displays the results of the numerical experiment described in the previous paragraph.
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections.
+We see that only $\Omega$ falls wholly within the $3\sigma$ shaded region.
+The PP curves for the other parameters deviate from the diagonal of unit slope.
+The deviation is more pronounced for $\alpha$ and $\delta$ and less for $\psi$ and $\Phi_0$.
+The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within higher-value credible intervals than would be expected statistically, and there are more injections contained within lower-value credible intervals.
+This stems from the bias noted in Section \ref{sec:multiple_noise}.
+The origin of the bias is discussed in Section \ref{sec:pulsar_terms_drop}.
 
@@
 
-\section{Representative end-to-end analysis} \label{sec:rep_example}
+\section{Validation of pulsar-terms inclusion with synthetic data} \label{sec:rep_example}
 In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
@@
 \end{figure}
-
-\section{SMBHB source parameters} \label{sec:parameter_space}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/pp_plot_new1}
-	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
-PP plot).
-The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}.
-Each coloured curve corresponds to a different parameter (see legend).
-The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
-The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals.
-For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope.
-$\Omega$ is generally well-estimated (i.e.
-it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e.
-the curves lie above the unit diagonal for low credible intervals, and below the unit diagonal for high credible intervals).
-This is due to a modelling bias whose origin is discussed in Sections \ref{sec:parameter_estim} and \ref{sec:pulsar_terms_drop}.}
-	\label{fig:parameter_space}
-\end{figure}
-Section \ref{sec:rep_example} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}.
-In this section we test the method for various systems, varying the SMBHB source parameters through astrophysically relevant ranges.
-\newline
-
-We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
-We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
-For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
-To summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
-A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself.
-In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
-\newline
-
- Figure \ref{fig:parameter_space} displays the results of the numerical experiment described in the previous paragraph.
-The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections.
-We see that only $\Omega$ falls wholly within the $3\sigma$ shaded region.
-The PP curves for the other parameters deviate from the diagonal of unit slope.
-The deviation is more pronounced for $\alpha$ and $\delta$ and less for $\psi$ and $\Phi_0$.
-The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within higher-value credible intervals than would be expected statistically, and there are more injections contained within lower-value credible intervals.
-This stems from the bias noted in Section \ref{sec:multiple_noise}.
-The origin of the bias is discussed in Section \ref{sec:pulsar_terms_drop}.
 
@@
 
+\section{Choice of priors}\label{sec:set_priors}
+To deploy the analysis scheme of K24 it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters. \newline
+
+For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}.
+For $\pi(\boldsymbol{\theta}_{\rm psr})$ \textit{a priori} information from electromagnetic observations does exist.
+We adopt constrained uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database.
+By using wider-than-necessary priors we expose the analysis scheme to a more stringent test.
+We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$) which covers the range of values inferred from Equation \ref{eq:sigmap_f}.
+We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.
+We do not set a prior on $\gamma^{(n)}$, because $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and consequently $\gamma^{(n)}$ is effectively ``unobservable'' over $T_{\rm obs}$.
+For validation purposes it is therefore sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value.
+This reduces the total dimension of the parameter space to $7 + 4N$.
+The chosen prior $\boldsymbol{\theta}$ is the same as used in K24, with the addition of a prior on $\chi^{(n)}$ due to the inclusion of the pulsar term.
+\newline
+
+The chosen priors for the representative analysis in Section \ref{sec:rep_example1} are summarised in \ref{Table tab:parameters_and_priors}
+
 \section{Complete workflow summary}
@@
 
-\section{Challenges of the pulsar term}
+\section{Challenges of the pulsar term}\label{sec:psr_term_challenges}
 In this section....

### 2023-11-20 manuscript/manuscript_2/paper.tex
restore standard priors on sigma_p

@@
 			\hline
-			& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
+			\vspace{1mm}& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
       \multirow{2}{2mm}{$\boldsymbol{\theta'}_{\rm psr}$} & $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
-		     & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
+		     & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$ \\
 			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
@@
 The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
+We do not try to infer $\gamma^{(n)}$ as $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$ and so no prior is set.
 The choice of priors on $\boldsymbol{\theta}_{\rm psr}$ is discussed in Appendix \ref{sec:set_priors}.
@@
 The median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134.
-Compared to the other 46 pulsars of the array, PSR J1939+2134 is a notable outlier with an estimated $\sigma^{(n)}$ much larger than the median.
-In practice, in order to avoid setting an overly broad prior on $\sigma^{(n)}$ (see Appendix \ref{sec:set_priors}) for validation purposes, we artificially set $\sigma^{(n)} = 10^{-24} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within an reasonable width log-uniform prior.
-This is done purely for testing convenience; it is straightforward to expand the prior when analysing real, astronomical data.
-\newline
 
@@
 Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
-Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar.
+When analysing real data this assumption can easily be relaxed.
+Whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is constructed by a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar.
 \newline

### 2023-11-20 py_src/run.py

@@ -15,8 +15,8 @@ def bilby_inference_run(arg_name,h,measurement_model,seed):
 
     logger = logging.getLogger().setLevel(logging.INFO)
     #Setup the system
-    #P   = SystemParameters(h=h,σp=None,σm=1e-11,use_psr_terms_in_data=True,measurement_model=measurement_model,seed=seed) # define the system parameters as a dict. Todo: make this a class
-    P   = SystemParameters(h=h,σp=True,σm=1e-11,use_psr_terms_in_data=True,measurement_model=measurement_model,seed=seed) # define the system parameters as a dict. Todo: make this a class
+    P   = SystemParameters(h=h,σp=None,σm=1e-11,use_psr_terms_in_data=True,measurement_model=measurement_model,seed=seed) # define the system parameters as a dict. Todo: make this a class
+    #P   = SystemParameters(h=h,σp=True,σm=1e-11,use_psr_terms_in_data=True,measurement_model=measurement_model,seed=seed) # define the system parameters as a dict. Todo: make this a class
 
     PTA = Pulsars(P)                                       # setup the PTA
     data = SyntheticData(PTA,P)                            # generate some synthetic data

### 2023-11-21 manuscript/manuscript_2/paper.tex
repro original plot, c.f. iota issues

@@
 	\end{table*}
-In order to generate synthetic frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+In order to generate synthetic frequency time series $ \boldsymbol{Y}(t) = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
 This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
@@
 Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and \eqref{eq:measurement}--\eqref{eq:g_func_trig} depend on the choice of static parameters $\boldsymbol{\theta'}$, as well as the choice of pulsar directions ${\boldsymbol{q}}^{(n)}$ and detector measurement noise covariance $\sigma_{\rm m}$ (Equation \eqref{eq:sigma_m_eqn}).
-We now describe how these values are chosen so as to generate astrophysically representative synthetic data.
+We now describe how $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ are chosen.
+This is equivalent to specifying the configuration of a synthetic PTA.
+In Section  \ref{sec:pe_and_ms} and Section \ref{sec:parameter_space} we describe how $\boldsymbol{\theta}_{\rm gw}$ chosen.
+This is equivalent to specifying the synthetic SMBHB source.
 \newline
@@
 \subsection{Parameter estimation} \label{sec:pe_and_ms} 
-
-The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
+In this section, we calculate the joint posterior probability distribution $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$
+
+for synthetic data $\boldsymbol{Y}$
+
+The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative.
+
+The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}
+
 The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
-Different values of the static GW source parameters are explored in Section XX.
-\newline
-
-The synthetic data generated by the above procedure is passed to the analysis scheme in Section \ref{sec:pe_and_ms}.
-
-In this section, we calculate the posterior probability distribution for the static parameters $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$ and compare it to the known, injected values.
+The specification of the priors on $\boldsymbol{\theta}$ is described in Appendix \ref{sec:set_priors} and summarised in Table \ref{tab:parameters_and_priors}.
+Different values of the static GW source parameters are explored in Section \ref{sec:parameter_space}.
+
+posterior probability distribution for the static parameters $\boldsymbol{\theta}$ and compare it to the known, injected values.
+
  The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
-That is, given the injected static parameters $\boldsymbol{\theta}$ summarised in Table \ref{Table tab:parameters_and_priors}, multiple realisations of synthetic data are generated using the method described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta}$.The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) demonstrate that quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
-We apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$.
-The specification of the priors on $\boldsymbol{\theta}$ is described in section \ref{sec:set_priors} and summarised in \ref{Table tab:parameters_and_priors}
+That is, given the injected static parameters $\boldsymbol{\theta}$ multiple realisations of synthetic data are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta}$.
+The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) demonstrate that quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
+
+We apply the Kalman filter in conjunction with nested sampling in order to infer the
 

### 2023-11-21 py_src/configs/config_batch.py

@@ -24,31 +24,11 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
     
     
 
-# MULTIPLE NOISE REALISATIONS
-N = 100
-seeds = np.arange(1235,1235+N,1)
-#strains = [1e-12,5e-15]
-strains = [1e-14]
-models = ["earth", "pulsar"]
-
-with open('batch.sh','w') as b: 
-
-    for s in seeds:
-        for h in strains:
-            for m in models:
-                arg_name = f"bigrun_{m}_{h}_{s}"
-                create_slurm_job(arg_name,h,m,s)
-                b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
-
-
-
-
-
-# N = 10
-# #seeds = np.arange(1235+10,1235+10+N,1)
-# seeds = [1251]
-
-# strains = [1e-12,5e-15]
+# # MULTIPLE NOISE REALISATIONS
+# N = 100
+# seeds = np.arange(1235,1235+N,1)
+# #strains = [1e-12,5e-15]
+# strains = [1e-14]
 # models = ["earth", "pulsar"]
 
 # with open('batch.sh','w') as b: 
@@ -56,11 +36,30 @@ with open('batch.sh','w') as b:
 #     for s in seeds:
 #         for h in strains:
 #             for m in models:
-#                 arg_name = f"reproduce_eg_canonical_{m}_{h}_{s}"
+#                 arg_name = f"bigrun_{m}_{h}_{s}"
 #                 create_slurm_job(arg_name,h,m,s)
 #                 b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
-       
 
+
+
+
+
+N = 10
+#seeds = np.arange(1235+10,1235+10+N,1)
+seeds = [1251]
+
+strains = [1e-12,5e-15]
+models = ["earth", "pulsar"]
+
+with open('batch.sh','w') as b: 
+
+    for s in seeds:
+        for h in strains:
+            for m in models:
+                arg_name = f"V2reproduce_eg_canonical_{m}_{h}_{s}"
+                create_slurm_job(arg_name,h,m,s)
+                b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+       
 
 
        

### 2023-11-21 manuscript/manuscript_2/paper.tex
some clean up to resemble canonical

@@
 
-We apply the Kalman filter in conjunction with nested sampling in order to infer the
+We apply the Kalman filter in conjunction with nested sampling in order to infer $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$ .  \newline
 
@@
 
+% stacked_GW_plot_small_h_definitive2
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{images/stacked_GW_plot_small_h_definitive2}
+	\includegraphics[width=\textwidth, height =\textwidth ]{example-image-a}
 	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for nine realisations of the noise processes, with curves coloured uniquely.
@@
 
+%images/corner_lowSNR_chi
 	\begin{figure}
-	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi}
+	\includegraphics[width=\columnwidth, height =\columnwidth ]{example-image-b}
 	\caption{hellow}

### 2023-11-21 py_src/kalman_filter.py

@@ -14,7 +14,6 @@ def log_likelihood(S,innovation):
     value = -0.5*(slogdet+innovation @ x + N*np.log(2*np.pi))
     #value = x @ x
     #return np.log(np.abs(value))
-    #value = -np.sum(innovation)
     return value
 
 
@@ -285,7 +284,6 @@ class KalmanFilter:
         y_results[0,:] = (1.0 - X_factor[0,:])*x - X_factor[0,:]*f_EM[0,:] 
 
 
-
         for i in np.arange(1,self.Nsteps):
 
             
@@ -299,7 +297,7 @@ class KalmanFilter:
              y_results[i,:] = (1.0 - X_factor[i,:])*x - X_factor[i,:]*f_EM[i,:] 
 
    
-        #print(iota_gw,likelihood)
+
         return likelihood,x_results,y_results
 
 

### 2023-11-21 manuscript/manuscript_2/paper.tex
test an increased P initialisation

@@
 The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-We show how to extend previous state-space analyses to include the pulsar terms.
+We show how to extend previous state-space analyses to include the pulsar terms and the method is verified on synthetic data.
 It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the Earth-terms-only approach.
@@
 In Section \ref{sec:synthetic_data} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme.
-In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ and $\mathcal{Z}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$.
+In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$.
 The scheme is validated on multiple noise realisations of the pulsar process noise and the measurement noise in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
@@
 \subsection{Parameter estimation} \label{sec:pe_and_ms} 
-In this section, we calculate the joint posterior probability distribution $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$
-
-for synthetic data $\boldsymbol{Y}$
-
+In this section, we apply the Kalman filter in conjunction with nested sampling to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ for synthetic data $\boldsymbol{Y}$, and compare it to the known, injected values.
+The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
+That is, given the injected static parameters $\boldsymbol{\theta'}$, multiple realisations of $\boldsymbol{Y}$ are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta'}$.
+The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) demonstrate that quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
 The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative.
-
-The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}
-
-The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
-The specification of the priors on $\boldsymbol{\theta}$ is described in Appendix \ref{sec:set_priors} and summarised in Table \ref{tab:parameters_and_priors}.
 Different values of the static GW source parameters are explored in Section \ref{sec:parameter_space}.
-
-posterior probability distribution for the static parameters $\boldsymbol{\theta}$ and compare it to the known, injected values.
-
- The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
-That is, given the injected static parameters $\boldsymbol{\theta}$ multiple realisations of synthetic data are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta}$.
-The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) demonstrate that quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
-
-We apply the Kalman filter in conjunction with nested sampling in order to infer $p({\boldsymbol{\theta}} | {\boldsymbol{Y}})$ .  \newline
-
-Figure \ref{fig:corner_plot_1} displays results for the seven parameters in  $\boldsymbol{\theta}_{\rm gw}$ in the form of a traditional corner plot.
+The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}.
+All injected static parameters are summarised in the second column of Table \ref{tab:parameters_and_priors}.
+The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}.
+\newline
+
+Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for \textcolor{red}{YY} arbitrary noise realisations in the form of a traditional corner plot.
 The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
-The dashed vertical blue lines mark the 0.16 and 0.84 quantiles; the solid orange line marks the known injected value.
+The solid orange line marks the known injected value.
 The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces.
@@
 There is scant evidence of railing against the prior bounds.
-For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$, the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
-The injected values are contained within the 90\% credible interval for all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ .
-\newline
-
-In Figure \ref{fig:corner_plot_2} we plot the estimates of $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary realisations.
-The seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$ are recovered unambiguously.
-The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, which stems from one realisation.
-We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
-As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm psr}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
-\newline
-
-Figure \ref{fig:corner_plot_2} confirms two main points: (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
-Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 (i.e.
-$92 \%$) out of the 63 possible combinations of the seven parameters and nine realizations.
-Figure \ref{fig:corner_plot_2}, like Figure \ref{fig:corner_plot_1}, displays tentative signs of bias, where the one-dimensional posteriors are not symmetric about the injected value.
-For example, the maximum a posteriori probability estimates of $\iota$ appear left-skewed, consistently underestimating the injected value in all nine realisations.
-A similar but messier trend is seen in $\delta$.
-However, the width of the posteriors is comparable to the putative bias, so it is difficult to draw strong conclusions.
-Biases are discussed in detail in Section \ref{sec:bias_and_identifiability}.
-There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.
-\newline
-
-The posterior is approximately symmetric about the known injected value for some parameters such as $\Omega$, where the posterior median and the injected value coincide approximately.
-For other parameters (e.g.
-$\iota$) the distribution is not symmetric about the injected value and the posterior median and the injected value do not coincide.
-The median value of the posterior for $\iota$ is shifted by $\approx 0.35$ rad relative to the injected value, although the injected value does still remain inside the 90\% credible interval.
-Similar effects are seen, albeit with a smaller shift, in other parameters such as $\delta$ and $\alpha$.
-For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
-a bias, or a random outcome specific to this particular noise realisation.
-We explore this further in Sections \ref{sec:multiple_noise} and \ref{sec:pulsar_terms_drop} using more noise realizations and show that indeed there is a systematic bias from dropping the pulsar terms, similar to that reported by \cite{Zhupulsarterms}.
-\newline
-
-Similar results are obtained for the 3$N$ parameters in $\boldsymbol{\theta}_{\rm psr, reduced}$.
-Again, the parameters are recovered unambiguously, in the sense that the nested sampler converges smoothly to a unimodal joint posterior near the known, injected parameter values, with all $3N$ injected parameters lying within the 90\% credible interval of the $3N$ one-dimensional, marginalized posteriors.
-We do not display the resulting corner plot because it is too big ($3N = 141$), and because inferring ${\boldsymbol{\theta}}_{\rm gw}$ is the focus of this paper and contemporary PTA analyses more generally.
+\textcolor{red}{There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.} The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e.
+\textcolor{red}{YY} $\%$) out of the \textcolor{red}{YY} possible combinations of the seven parameters and \textcolor{red}{YY} realizations.
+For this representative example, with characteristic strain \textcolor{red}{XX} the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
+\newline
 
 In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
-In both the high and low SNR cases $\chi^{(n)}$ is unambiguously recovered.
-This is shown in Figure \ref{fig:corner_plot_3} where we plot the results for five of the $\chi^{(n)}$ values for both high SNR (left-hand panel) and low SNR (right hand panel).
-The estimates are naturally more constrained in the high SNR case, but in both cases the injection is contained within the 90 \% credible interval.
-Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work.
+Figure \ref{fig:corner_plot_3} displays the results for a subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over \textcolor{red}{YY} noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
+We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big (total number of panels is $N(N+1)/2 = 1128$).
+All elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations, with the injected value contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e.
+\textcolor{red}{YY} $\%$) out of the \textcolor{red}{YY} possible combinations of the seven parameters and \textcolor{red}{YY} realizations.
+Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$; the parameters are recovered unambiguously across the \textcolor{red}{XX} noise realisations.
+For the sake of brevity we do not display the calculated posterior distributions here because a) the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses, and b) the resulting corner plot is too big.
+\newline
+
+Figure \ref{fig:corner_plot_1} confirms two points: (i) the analysis scheme returns accurate answers for arbitrary noise realisations and (ii) there is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
+Regarding the second point, quantifying the dispersion is important since it is a practical measure of the accuracy of the parameter estimation scheme when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
+\newline
+
+\textcolor{red}{TK: do we want an extended discussion re Wasserstein distance over many noise realisations analogous to what we did in paper 1?
+I think it is a bit distracting to main structure of paper which is: (i) show that including pulsar terms works generally (Section 4) and (ii) show the improvement over the Earth terms (Section 5).
+Relegating WD stuff to an appendix for now.}
 
@@
 	\includegraphics[width=\columnwidth, height =\columnwidth ]{example-image-b}
-	\caption{hellow}
+	\caption{Same as Figure \ref{fig:corner_plot_1}	but for 5 static parameters of $\chi^{(n)}$.
+The known, injected value lies within the 90\% credible interval for \textcolor{red}{X} out of the \textcolor{red}{X} combinations of five parameters and v\textcolor{red}{X} noise realizations.
+}
 	\label{fig:corner_plot_3}
@@
 
-\begin{table}
-	\centering
-	\begin{tabular}{ccll}
-		\toprule
-		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
-		\hline
-		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
-		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
-		$\psi$              & $2.50$ & rad & $0.16$ \\
-		$\iota$             & $1.0$ & rad & $0.14$ \\ 
-		$\delta$              & $1.0$  & rad & $0.09$ \\
-		$\alpha$          & $1.0$  & rad & $0.17$\\
-		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
-		\bottomrule
-	\end{tabular}
-	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
-$W_{1, \rm median}$ is generally smaller than the prior domain.}
-	\label{tab:Wasserstein}
-\end{table}
-
-To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
-For each realisation we independently estimate the static parameters, $\boldsymbol{\theta}$.
-
-We now consider the complete set of $10^3$ noise realisations, going beyond the subset of nine realisations in the preceding discussion.
-We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$?
-There is no unique way to answer this question.
-In this paper we appeal to the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory, which defines an intuitive notion of similarity between probability distributions.
-The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}; a short review is presented in Appendix \ref{sec:wasserstein}.
-The WD measures the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
-In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}.
-Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$ with respect to the PDFs $\mu$ and $\nu$.
-Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad.
-Then we can infer $| \langle \iota \rangle_\mu - \langle \iota \rangle_\nu | \leq 0.5 \, {\rm rad}$.
-\newline
-
-Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
-The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$.
-As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
-These values quantify approximately the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
-Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$.
-These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for $10^3$ noise realizations.
-Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
-
 \subsection{Extension to broader parameter space} \label{sec:parameter_space}
+%images/pp_plot_new1
 \begin{figure}
 	\centering
-	\includegraphics[width=\columnwidth]{images/pp_plot_new1}
+	\includegraphics[width=\columnwidth]{example-image-c}
 	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
@@
 \end{figure}
-Section \ref{sec:rep_example} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}.
-In this section we test the method for various systems, varying the SMBHB source parameters through astrophysically relevant ranges.
-\newline
-
-We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+Section \ref{sec:pe_and_ms} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}.
+In this section we test the method for various systems, varying the SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ through astrophysically relevant ranges.
+The aim of this section is to confirm that the analysis scheme can be applied across an astrophysically relevant parameter space and that the arbitrary choice of $\boldsymbol{\theta}_{\rm gw}$ in Section \ref{sec:pe_and_ms} is not particularly advantageous by accident.
+\newline
+
+We analyse \textcolor{red}{X} injections constructed by fixing $h_0 =$ \textcolor{red}{X}  and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
 We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
 For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
-To summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+\newline
+
+The results are displayed in Figure \ref{fig:parameter_space}.
+Since it is prohibitive to produce corner plots analogous to Figure \ref{fig:corner_plot_1} for \textcolor{red}{X} injections, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
 A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself.
 In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
-\newline
-
-Figure \ref{fig:parameter_space} displays the results of the numerical experiment described in the previous paragraph.
-The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections.
-We see that only $\Omega$ falls wholly within the $3\sigma$ shaded region.
-The PP curves for the other parameters deviate from the diagonal of unit slope.
-The deviation is more pronounced for $\alpha$ and $\delta$ and less for $\psi$ and $\Phi_0$.
-The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within higher-value credible intervals than would be expected statistically, and there are more injections contained within lower-value credible intervals.
-This stems from the bias noted in Section \ref{sec:multiple_noise}.
-The origin of the bias is discussed in Section \ref{sec:pulsar_terms_drop}.
+In Figure \ref{fig:parameter_space} the shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for \textcolor{red}{X} injections.
+We see that...
+\textcolor{red}{TBD once simulations complete}
 
 \section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}
+In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model.
 In this section we compare the results obtained using the pulsar-term model (i.e.
 Equation \eqref{eq:g_func_trig_chi}) to those obtained using the Earth-term model (i.e.
-Equation \eqref{eq:g_func_trig_earth}) for the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
+Equation \eqref{eq:g_func_trig_earth}) for the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
+The pulsar-term model and the Earth-terms model are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}.
+The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for the pulsar terms model, as described in Appendix \ref{sec:set_priors}.
+\newline
+
 In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$ returned by the two models.
-We consider two separate cases; a ``high-SNR" case with $h_0 = 10^{-12}$ and a ``low-SNR" case with $h_0 = 5 \times 10^{-15}$.
+We consider two separate cases: a ``high-SNR" case with $h_0 = 10^{-12}$ and a ``low-SNR" case with $h_0 = 5 \times 10^{-15}$.
 In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable strain for the two models.
-In this Section, the procedure for generating synthetic data and the specification of $\pi(\boldsymbol{\theta})$ is as described in Sections \ref{sec:synthetic_data} and \ref{sec:set_priors} respectively.
-Both the pulsar-term model and the Earth-terms model are applied to identical realisations of the data.
-The priors in both cases are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for the pulsar terms model, as described in Section \ref{sec:set_priors}.
 
@@
 The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
-The settings of the nested sampler (for example the number of live points, see Appendix X) are identical for both models.
+The settings of the nested sampler (for example the number of live points, see Appendix \ref{sec:nested_sampling}) are identical for both models.
 \newline
@@
 The ``high-SNR" system is considered in order to evaluate the parameter estimation accuracy, and in particular the bias, without confusion from the measurement noise.
-This section considers a single noise realisation of the synthetic data $\boldsymbol{Y}$ for the representative example system described in Table \ref{tab:parameters_and_priors}.
-\newline
+
+In order to enable a clear comparison
+
+This section considers a single noise realisation of the synthetic data $\boldsymbol{Y}$ for the representative example system described in Table \ref{tab:parameters_and_priors}. \newline
 
@@
 			\includegraphics[width=\columnwidth, height = \columnwidth ]{images/CanonicalBayesPlot2000} 	
-			\caption{hellow}
+			\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strains, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
+The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
+The minimum detectable strain, for $\beta < 10$, equals $4 \times 10^{-15}$.
+Missing points for $\beta \lesssim 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
 			\label{fig:bayes1}
@@
 For the results presented in this subsection we set $n_{\rm live} =2000$.
-
-\section{Validation of pulsar-terms inclusion with synthetic data} \label{sec:rep_example}
-In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space} and \ref{sec:bias_and_identifiability}.
-To assist with reproducibility, we present intermediate outputs from each step.
-In Section \ref{sec:priors} we define the priors on $\boldsymbol{\theta}$.
-In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter.
-In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise.
-In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates.
-In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$.
-\newline
-
-The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative.
-The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
-
-\subsection{Detectability versus $h_0$} \label{sec:detection}
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
-	\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strains, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
-The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
-The minimum detectable strain, for $\beta < 10$, equals $4 \times 10^{-15}$.
-Missing points for $\beta \lesssim 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
-	\label{fig:bayes}
-\end{figure}
 
@@
 
-\section{Nested sampling}
+\section{Nested sampling}\label{sec:nested_sampling}
 
@@
 
-\section{}
+\section{Multiple noise realisaitons and the Wasserstein distance}
+
+We can extend the analysis of these two points and consider 1000 noise realisations of $\boldsymbol{Y}$, beyond the \textcolor{red}{X} plotted in Figure \ref{fig:corner_plot_1}.
+As in Figure \ref{fig:corner_plot_1} we independently estimate the static parameters, $\boldsymbol{\theta}$ for each realisation and calculate $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$.
+
+To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
+For each realisation we
+
+We now consider the complete set of $10^3$ noise realisations, going beyond the subset of nine realisations in the preceding discussion.
+We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$?
+There is no unique way to answer this question.
+In this paper we appeal to the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory, which defines an intuitive notion of similarity between probability distributions.
+The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}; a short review is presented in Appendix \ref{sec:wasserstein}.
+The WD measures the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
+In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}.
+Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$ with respect to the PDFs $\mu$ and $\nu$.
+Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad.
+Then we can infer $| \langle \iota \rangle_\mu - \langle \iota \rangle_\nu | \leq 0.5 \, {\rm rad}$.
+\newline
+
+Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$.
+As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
+These values quantify approximately the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
+Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$.
+These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for $10^3$ noise realizations.
+Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
+
+\begin{table}
+	\centering
+	\begin{tabular}{ccll}
+		\toprule
+		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
+		\hline
+		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
+		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
+		$\psi$              & $2.50$ & rad & $0.16$ \\
+		$\iota$             & $1.0$ & rad & $0.14$ \\ 
+		$\delta$              & $1.0$  & rad & $0.09$ \\
+		$\alpha$          & $1.0$  & rad & $0.17$\\
+		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
+		\bottomrule
+	\end{tabular}
+	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
+$W_{1, \rm median}$ is generally smaller than the prior domain.}
+	\label{tab:Wasserstein}
+\end{table}
 

### 2023-11-22 manuscript/manuscript_2/paper.tex
version of manuscript passed to internal comments

@@
 The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-We show how to extend previous state-space analyses to include the pulsar terms and the method is verified on synthetic data.
+We show how to extend previous state-space analyses to include the pulsar terms.
 It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the Earth-terms-only approach.
@@
 
-We analyse \textcolor{red}{X} injections constructed by fixing $h_0 =$ \textcolor{red}{X}  and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
-We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
+We analyse \textcolor{red}{X} injections constructed by fixing $h_0 =$ \textcolor{red}{Y}  and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the  \textcolor{red}{X} injections.
 For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
@@
 In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$.
-For the Earth-term model we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$.
+For the Earth-term model we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta'})$.
 For the pulsar-term model we we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi}.
-The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$.
 The settings of the nested sampler (for example the number of live points, see Appendix \ref{sec:nested_sampling}) are identical for both models.
@@
 The second is a ``high-SNR" system which has the same static injection parameters, but with an increased $h_0 = 10^{-12}$.
-The ``high-SNR" system is considered in order to evaluate the parameter estimation accuracy, and in particular the bias, without confusion from the measurement noise.
-
-In order to enable a clear comparison
-
-This section considers a single noise realisation of the synthetic data $\boldsymbol{Y}$ for the representative example system described in Table \ref{tab:parameters_and_priors}. \newline
+The ``high-SNR" system is considered in order to evaluate the parameter estimation accuracy, and in particular any potential bias, without confusion from the measurement noise.
+In order to enable a clear comparison between the posterior probability distributions calculated using the two models, in this section we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$.
+The results quoted in the below paragraph are generally consistent across different realisations, but plotting multiple realisations simultaneously overcrowds the figure and makes it harder to draw conclusions.
+\newline
 
 The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_compare_high} for the high-SNR system and in Figure \ref{fig:corner_plot_compare_low} for the low-SNR system.
-The corner plot is arranged identically to \ref{fig:corner_plot_1}.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
 The green curves are the results derived using the Earth-terms measurement equation, Equation \ref{eq:g_func_trig_earth}.
@@
 Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
-For example, for $\iota$, whilst the injected value is contained within the 90 \% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians.
-The inclusion of the pulsar terms corrects for this bias.
-The green one-dimensional posteriors exhibit no bias and a generally symmetric about the injected value.
+For example, for $\iota$, whilst the injected value is contained within the 90\% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians.
+The inclusion of the pulsar terms corrects for this bias; the green one-dimensional exhibit no bias and are generally symmetric about the injected value.
 Moreover, for every pair of parameters the injection values are contained within the 2-sigma equivalent contours for the pulsar-term model, which is not true in the Earth-terms case.
@@
 \begin{equation}
-\Delta_{\rm M}(\theta) = \frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}} \ , 
-\end{equation}
-where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimates using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+\Delta_{\rm M}(\theta) = \frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}} \ , \label{eq:mean_rel_error}
+\end{equation}
+where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
 The dependence of $p_k(\theta)$ on $M$ is implied.
 $\Delta_{\rm M}(\theta)$ represents the mean relative error in the one-dimensional posterior.
-The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_1} are summarised in Table \ref{tab:posterior_errors}.
+The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}.
 The difference in the mean relative error between the two models, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is also presented.
@@
 The exception for $\Omega$ is due to the fact that $\Omega$ is recovered with high accuracy in both cases; the the mean relative error is small $\Delta_{\rm M}(\Omega) \sim 10^{-5}$.
-The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm pulsar}(\iota) - \Delta_{\rm earth}(\iota) = 0.25$.
+The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm pulsar}(\iota) - \Delta_{\rm earth}(\iota) = -0.25$.
 \newline
@@
 The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest.
-The relative errors 	$\Delta_{\rm M}(\Theta)$ for the low SNR case are also reported in Table \ref{tab:posterior_errors}.
+The mean relative errors $\Delta_{\rm M}(\theta)$ for the low SNR case are also reported in Table \ref{tab:posterior_errors}.
 We see that....
@@
 This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates.
-Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:detection_new} that they are important for increasing the detection probability for a given signal strength.
+Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:psr_v_earth_bayes} that they are important for increasing the detection probability for a given signal strength.
 \newline
@@
 	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_highSNR_compare}
-	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
+	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, with $h_0 = 10^{-12}$, for a single realisation of the system noise.
+The blue curves are the posteriors derived using the Earth-term model, Equation \eqref{eq:g_func_trig_earth}, the green curves using the pulsar-term model, Equation \eqref{eq:g_func_trig}.
 The vertical orange lines indicate the true injected values.
 The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
-We are able to accurately estimate each parameter of interest.
-We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
-Note that the $x$-axis scaling does not span the total prior space.}
+We plot the scaled variables $10^9 \Omega$ (units: rad s$^{-1}$) and $10^{15} h_0$.
+Some parameters (e.g.
+$\psi, \iota$) exhibit a bias when using the Earth-term model which is absent when using the pulsar-term model.
+\textcolor{red}{TK: need to remove these subtitles and correct the labels for omega and h.
+The actual results here may also be updated (same for Figure 5)}}
 	\label{fig:corner_plot_compare_high}
@@
 	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
-	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise.
-The vertical orange lines indicate the true injected values.
-The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles.
-We are able to accurately estimate each parameter of interest.
-We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$.
-Note that the $x$-axis scaling does not span the total prior space.}
+	\caption{Same as Figure \ref{fig:corner_plot_compare_high}, but for a low-SNR system with $h_0 = 5 \times 10^{-15}$.
+The posterior distributions from the Earth-term and pulsar-term models are generally overlaid }
 	\label{fig:corner_plot_compare_low}
@@
 		\end{tabular}
-		\caption{dfdfddd}
+		\caption{Mean relative error $\Delta_{\rm M}(\Theta)$ (Equation \eqref{eq:mean_rel_error}) in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+The injection values are as summarised in Table \ref{tab:parameters_and_priors}.
+The top and bottom halves of the table contain the high-SNR ($h_0 = 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.
+The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.\textcolor{red}{TK numbers to be revised and double checked, but probably representative}}
 		\label{tab:posterior_errors}

### 2023-11-24 manuscript/manuscript_2/paper.tex
generate pp plot for smaller h

@@
 The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-We show how to extend previous state-space analyses to include the pulsar terms.
-It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the Earth-terms-only approach.
+It is shown how to extend previous state-space analyses to include the pulsar terms.
+Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the previous approach.
 \end{abstract}
@@
 Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
-SMBHBs which are sufficiently close and massive may be individually resolvable, though no definitive detection has yet been made \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
+SMBHBs which are sufficiently close and massive are perhaps individually resolvable, though no definitive detection is yet claimed \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
 \newline
@@
 Firstly we extend the methods presented in K24 to include the pulsar term.
-We show how to reparametrise the state-space model in terms of an additional pulsars-specific phase term.
+We show how to reparametrise the state-space model in terms of an additional pulsar-specific phase term.
 We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}.
@@
 There are $N$ pulsars in the array.
+The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, as measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation.
 The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$.
-The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, as measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation.
 The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW.
@@
 The model is formulated in terms of pulse frequency time series as a proof-of-principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
-It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization we defer to future work.
+It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, an extension we defer to future work.
 
@@
 \end{align}
-In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
 The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
@@
 Such a phenomenological model is necessary as a predictive, first-principles theory of timing noise does not currently exist c.f.
-the multiple theorized mechanisms of timing noise (see discussion in Section 1 of \cite{KimpsonPTA}).
+the multiple theorized mechanisms of timing noise (see discussion in Section 1 of K24).
 Similar approaches, analogous to Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
-However, Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} do contain certain modelling omissions and idealisations which much be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
-These include the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and the formal interpretation of $d^2 f_{\rm p} / dt^2$ due to $\xi(t)$ in Equation \eqref{eq:frequency_evolution} not being differentiable.
+However, Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} do contain certain modelling omissions and idealisations which must be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
+These include (i) the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$ due to $\xi(t)$ in Equation \eqref{eq:frequency_evolution} not being differentiable.
 
 \subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
-In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a a non-linear measurement equation:
+In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation:
 \begin{equation}
@@
 
-Equation \eqref{eq:g_func_trig} contains two simplifying assumptions which must be recognised:  
-\begin{enumerate}
+\noindent Equation \eqref{eq:g_func_trig} contains two simplifying assumptions: 
+\begin{enumerate}[leftmargin=2em]
 	\item The GW is a monochromatic plane wave.
-	\item The pulsar locations are constant with respect to the observer 
+	\item The pulsar locations are constant with respect to the observer.
 \end{enumerate}
@@
 where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years.
-A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
+A source can be considered monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
 Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \gtrsim 20$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$.
 The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
-We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
+We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
 \newline
@@
 The complete set of $7 + 5N$ static parameters is denoted $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$.
-While we have no prior information about the values of the members of $\boldsymbol{\theta}_{\rm gw}$, the members of $\boldsymbol{\theta}_{\rm psr}$
+While we have no prior information about the values of the elements of $\boldsymbol{\theta}_{\rm gw}$, the elements of $\boldsymbol{\theta}_{\rm psr}$
 do have constraints from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
@@
 Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
-We refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, nested sampling and the combined workflow in the context of PTA searches for GWs.
+We refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, nested sampling, and the combined workflow in the context of PTA searches for GWs.
 \newline
@@
 In Section \ref{sec:earth_term} we review how the pulsar term is dropped from typical PTA analyses and the modification to Equation \eqref{eq:g_func_trig} used in K24.
-In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar term for likelihood-based inference methods like nested sampling.
-This new parametrisation is validated with synthetic data in Section \ref{sec:rep_example1}.
+In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar term for likelihood-based inference methods such as nested sampling.
+The new parametrisation is validated with synthetic data in Section \ref{sec:rep_example1}.
 
 \subsection{Earth term}\label{sec:earth_term}
-The influence of the GW on the pulsar radio pulses is completely described by Equation \eqref{eq:g_func_trig}.
+The influence of the GW on the pulsar radio pulses is described by Equation \eqref{eq:g_func_trig}.
 The GW signal is composed of an ``Earth term", corresponding to the $\cos(-\Omega t + \Phi_0)$ component, and a ``pulsar term", corresponding to the $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \}$ component.
@@
 Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases which cannot be coherently summed and are instead generally considered as a source of self-noise.
-The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties much greater than the typical target GW wavelength.
+The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties much greater than the target GW wavelength.
 It is therefore not possible to set the phase of the pulsar term.
-Consequently the pulsar term is typically dropped from in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space, since the $N$ values of $d^{(n)}$ do not have to be inferred, making the parameter estimation computationally cheaper.
+Consequently the pulsar term is often dropped in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space; the $N$ values of $d^{(n)}$ are not inferred, making the parameter estimation computationally cheaper.
 Within the state-space model described in Section \ref{sec:2}, dropping the pulsar terms equates to using a modified measurement equation, c.f.
@@
 \begin{equation}
-	\chi^{(n)} \equiv \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \mod 2 \pi \ ,
+	\chi^{(n)} \equiv \left[ \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right ] \mod 2 \pi \ ,
 \end{equation}
@@
 The reparametrisation in terms of $\chi^{(n)}$ treats the pulsar-dependent phase correction $\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ in the argument of the cosine of the pulsar term as an explicit parameter to be inferred for each pulsar.
-The advantage of this approach, compared to trying to infer $d^{(n)}$ directly, is that the resulting likelihood surface across the parameter space (i.e.
-the shape of the function $\mathcal{L}(\boldsymbol{\theta})$) is smoother, with smaller derivatives.
+The advantage of this approach, compared to trying to infer $d^{(n)}$ directly, is that the resulting likelihood surface across the parameter space is smoother, with smaller derivatives.
 This enables sampling algorithms to efficiently find the likelihood maxima and estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
@@
 In contrast $\chi^{(n)}$ is a cross term that depends on both the individual pulsar and the GW.
-We define the complete set of $7 +4N + N$ static parameters to be inferred as $\boldsymbol{\theta'} = \boldsymbol{\theta'}_{\rm gw} \cup \boldsymbol{\theta'}_{\rm psr} \cup \chi^{(n)}$ .
-\newline
-
-\section{Validation of pulsar-terms inclusion with synthetic data}\label{sec:rep_example1}
+We define the complete set of $7 +4N + N$ static parameters to be inferred as $\boldsymbol{\theta'} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta'}_{\rm psr} \cup \chi^{(n)}$ .
+\newline
+
+\section{Validation with synthetic data}\label{sec:rep_example1}
 In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}.
@@
 This is equivalent to specifying the configuration of a synthetic PTA.
-In Section  \ref{sec:pe_and_ms} and Section \ref{sec:parameter_space} we describe how $\boldsymbol{\theta}_{\rm gw}$ chosen.
+In Section  \ref{sec:pe_and_ms} and Section \ref{sec:parameter_space} we describe how the remaining elements of $\boldsymbol{\theta}'$, namely $\boldsymbol{\theta}_{\rm gw}$, are chosen.
 This is equivalent to specifying the synthetic SMBHB source.
@@
 
-In this paper we use the same values for $\boldsymbol{\theta}_{\rm psr}$ as in K24.
-That is, we adopt the 47 millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}, i.e.
+In this paper we use the same values for $\boldsymbol{\theta}_{\rm psr}$ as in K24, i.e.
+we adopt the $N=47$ millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}, i.e.
 $N = 47$.
-We consider all pulsars to be observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period.
+We assume all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period.
 For each pulsar fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$ are obtained via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
@@
 The median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134.
-
-The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
+\newline
+
+The measurement noise covariance, as defined in Equation \eqref{eq:vareps}, can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
 \begin{equation}
@@
 \end{equation}
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s,  Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
 The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
@@
 That is, given the injected static parameters $\boldsymbol{\theta'}$, multiple realisations of $\boldsymbol{Y}$ are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta'}$.
-The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) demonstrate that quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
+The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
 The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative.
@@
 In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
-Figure \ref{fig:corner_plot_3} displays the results for a subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over \textcolor{red}{YY} noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
-We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big (total number of panels is $N(N+1)/2 = 1128$).
+Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over \textcolor{red}{YY} noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
+We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big.
 All elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations, with the injected value contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e.
@@
 Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$; the parameters are recovered unambiguously across the \textcolor{red}{XX} noise realisations.
-For the sake of brevity we do not display the calculated posterior distributions here because a) the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses, and b) the resulting corner plot is too big.
+For the sake of brevity we do not display the calculated posterior distributions here because the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses.
 \newline
@@
 \newline
-
-\textcolor{red}{TK: do we want an extended discussion re Wasserstein distance over many noise realisations analogous to what we did in paper 1?
-I think it is a bit distracting to main structure of paper which is: (i) show that including pulsar terms works generally (Section 4) and (ii) show the improvement over the Earth terms (Section 5).
-Relegating WD stuff to an appendix for now.}
 
@@
 Since it is prohibitive to produce corner plots analogous to Figure \ref{fig:corner_plot_1} for \textcolor{red}{X} injections, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
-A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself.
+A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself.
 In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
@@
 In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model.
-In this section we compare the results obtained using the pulsar-term model (i.e.
-Equation \eqref{eq:g_func_trig_chi}) to those obtained using the Earth-term model (i.e.
-Equation \eqref{eq:g_func_trig_earth}) for the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
-The pulsar-term model and the Earth-terms model are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}.
-The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for the pulsar terms model, as described in Appendix \ref{sec:set_priors}.
+In this section, critically, we compare the affect of including the pulsar-term in the model (i.e.
+using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e.
+Equation \eqref{eq:g_func_trig_earth}).
+We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
+We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm P \& E}$.
+We refer to the model where the pulsar-term is omitted as
+$\mathcal{M}_{\rm E}$.
+Model $\mathcal{M}_{\rm P \& E}$ is parameterised by $\boldsymbol{\theta}'$.
+Model $\mathcal{M}_{\rm E}$ is parameterised by $\boldsymbol{\theta}_{\rm E} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$.
+Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}.
+The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm P \& E}$, as described in Appendix \ref{sec:set_priors}.
 \newline
@@
 In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$ returned by the two models.
-We consider two separate cases: a ``high-SNR" case with $h_0 = 10^{-12}$ and a ``low-SNR" case with $h_0 = 5 \times 10^{-15}$.
 In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable strain for the two models.
@@
 
-In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$.
-For the Earth-term model we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta'})$.
-For the pulsar-term model we we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi}.
+In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model.
+For $\mathcal{M}_{\rm E}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm E})$.
+The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta}_{\rm E} | \boldsymbol{Y})$.
+For $\mathcal{M}_{\rm P \& E}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$.
 The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$.
@@
 In order to enable a clear comparison between the posterior probability distributions calculated using the two models, in this section we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$.
-The results quoted in the below paragraph are generally consistent across different realisations, but plotting multiple realisations simultaneously overcrowds the figure and makes it harder to draw conclusions.
+The results quoted are consistent across different noise realisations.
 \newline
@@
 The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
-The green curves are the results derived using the Earth-terms measurement equation, Equation \ref{eq:g_func_trig_earth}.
-The blue curves are the results derived using the full measurement equation, including the pulsar terms, Equation \eqref{eq:g_func_trig}.
+The blue curves are the results derived using $\mathcal{M}_{E}$.
+The green curves are the results derived using $\mathcal{M}_{\rm P \& E}$.
 \newline
@@
 The top and bottom halves of the table contain the high-SNR ($h_0 = 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.
-The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.\textcolor{red}{TK numbers to be revised and double checked, but probably representative}}
+The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.\textcolor{red}{TK numbers to be revised and double checked, but probably representative.
+Julian notes: swap so positive is good, normalise omega, add units, align}}
 		\label{tab:posterior_errors}
@@
 For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
+Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Appendix \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
 For the results presented in this subsection we set $n_{\rm live} =2000$.
@@
 First, one must still generalize the Kalman tracker and nested sampler to ingest TOAs directly, as happens traditionally, instead of ingesting a frequency time series, as in this paper.
-Generalizing the algorithm is a subtle task and is postponed to a forthcoming paper.
+Generalizing the algorithm is a subtle task and is postponed to a future work.
 Second, the Kalman tracker and nested sampler are conditional on a noise model that is related but different to the noise model in traditional approaches.

### 2023-11-24 py_src/configs/config_batch_pp.py

@@ -29,7 +29,7 @@ def create_slurm_job(arg_name,h,measurement_model,seed,omega,phi0,psi,delta,alph
 
 N = 500
 seed = 1250 
-h = 1e-14 
+h = 1e-15 
 model = "pulsar"
         
 #Create a bilby prior object to sample from
@@ -57,7 +57,7 @@ with open('batch.sh','w') as b:
     for i in range(N):
         p = priors.sample()
 
-        arg_name = f"november_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
+        arg_name = f"november15_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
         print(arg_name)
         create_slurm_job(arg_name,h,model,seed,p['omega_gw'],p['phi0_gw'],p['psi_gw'],p['delta_gw'],p['alpha_gw'])
         b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")

### 2023-11-24 manuscript/manuscript_2/paper.tex
some manuscript updates

@@
 \usepackage{multirow}
+
+\usepackage{stfloats}
 
@@
 		\end{figure}
-
-		The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
-		\begin{equation}
-			\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
-		\end{equation}
-		Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
-$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data.
-This is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
-The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW compared to there being no GW signal present.
-
-		In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$. \newline
-
-		We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}.
-In Equation \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-term-only model with a GW present, i.e.
-the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}.
-The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
+The evidence integral $\mathcal{Z}$ returned by nested sampling, Equation \eqref{eq:model_evidence}, is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$.
+We define $\mathcal{M}_{\rm null}$ as the null model that assumes no GW exists in the data.
+This is equivalent to setting $g^{(n)}(t)=1$ in Equation \eqref{eq:g_func_trig}.
+The support in the data for the presence of a GW signal, described by model $\mathcal{M}_i$, over the absence of a GW signal is quantified via a Bayes factor
+		\begin{equation}
+			\beta_i = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_i)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_{\rm null})} \ . \label{eq:bayes}
+		\end{equation}
+In this paper we consider $i \in \left \{ \rm P\& E, E\right \}$ \newline
+
+The Bayes factor, $\beta_i$, for each of the two models is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
 To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
-\footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.} \newline
+\footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.}
+
+	Multiple noise...
 
@@
 
-		The odds ratio $\beta$ drops off for $h_0 \lesssim 4 \times10^{-15}$.
+%		The odds ratio $\beta$ drops off for $h_0 \lesssim 4 \times10^{-15}$.
 This happens because the two competing models becoming increasingly indistinguishable once the measurement noise dominates the GW signal.
@@
 In this section we set $n_{\rm live} =2000$.
-
-		We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
+%		
+%		
+%
+
+%		
+%		We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
 $\mathcal{M}_1$ is the Earth-terms only model, i.e.
@@
 
-		There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
+%		There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
 The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
@@
 For the results presented in this subsection we set $n_{\rm live} =2000$.
+%
 
 \section{Conclusion}\label{sec:discussion}
-
-In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs in PTA data.
+In this paper we demonstrate how to extend state-space methods for PTA data analysis to include the pulsar term.
+We compare the results when the pulsar terms are includes to those when they are omitted.
+\newline
+
+The intrinsic rotational state evolution of the pulsar, described by a mean-reverting Ornstein-Uhlenbeck process (Equations \eqref{eq:xieqn}--\eqref{eq:spinevol}) is tracked using a Kalman filter.
+The Kalman filter is combined with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model with and without a GW.
+\newline
+
+The updated state-space model including the pulsar term is tested on synthetic data.
+We consider $X$ noise realisations of synthetic data for a single, astrophysically representative, SMBHB GW source observed by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$.
+We find that the updated state-space model successfully detects injected signals and estimates their static parameters accurately for all noise realisations with relatively low computational cost.
+There is a non-negligible natural random dispersion in the median of the marginalised one-dimensional posteriors which \textcolor{red}{...means something}.
+The updated model is further tested across a broader parameter domain of SMBHB sources.
+500 randomly sampled parameter vectors at fixed $\iota$ and $h_0$ are explored; \textcolor{red}{the results are...} \newline
+
+We compare the accuracy of the parameter estimates  from the updated model to those when the pulsar term is omitted.
+We find that the inclision of the pulsar terms increases the accuaray by X percent in teh case of high SNR and Y per cent in the case of low SNR.
+The inclusion of pulsar terms also lowers the floor of the minimal detectable strain by YY percent.
+\newline
+
+\textcolor{red}{Caceats, extensions, other things}
+
 The new method is complementary to traditional approaches.
@@
 That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data.
-We implement a Kalman filter in order to track the intrinsic rotational state evolution of the pulsar and combine it with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model with and without a GW.
-\newline
-
- We test the new method on synthetic data and find that it detects injected signals successfully and estimates their static parameters accurately with relatively low computational cost.
-We initially focus on a single, astrophysically representative, SMBHB GW source observed synthetically by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$.
-The minimum detectable strain is estimated to be $\min(h_0) \approx 4 \times 10^{-15}$ for $\iota=1.0$ rad.
-We then repeat the parameter estimation exercise for 1000 noise realisations to compute the natural dispersion (``cosmic variance'') in the recovered values and hence quantify the statistical accuracy of the method in a real astrophysical application, when the true parameter values are unknown.
-Consistent posteriors are obtained for most realisations.
-The median WD is limited to $\lesssim 3\%$ of the width of the prior domain (i.e.\ negligible railing).
-The median WD divided by the injected value, an approximate measure of the natural dispersion, ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
-\newline
-
-Exploration of a broader SMBHB parameter domain at fixed $\iota$ and $h_0$ via 200 randomly sampled parameter vectors reveals a bias in the estimates of the static parameters $\delta$ and $\alpha$, and to a lesser extent $\psi$ and $\Phi_0$.
-The bias is examined for a specific SMBHB GW source in the limit of high-SNR.
-It is greatest for $\iota$, amounting to $\approx 0.3$ rad.
-Smaller biases of $\lesssim 0.1$ rad are also observed in $\Phi_0, \psi, \delta$ and $\alpha$.
-The bias is shown to result from dropping the pulsar terms from the measurement equation in the Kalman filter, consistent with the work of previous authors \citep{Zhupulsarterms,Chen2022}.
-The method also exhibits a trade-off between $h_0$ and $\iota$, which follows formally from the weak identifiability of $h_0$ and $\iota$.
-\newline
 
@@
 Second, the Kalman tracker and nested sampler are conditional on a noise model that is related but different to the noise model in traditional approaches.
-The analysis in this paper assumes a specific, time-ordered realization of a mean-reverting Ornstein-Uhlenbeck process satisfying Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law power spectral density, whose amplitude and exponent are adjustable.
+
+The analysis in this paper assumes a specific, time-ordered realization of a ,
+
+whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law power spectral density, whose amplitude and exponent are adjustable.
 Hence the sensitivities cannot be compared directly.
@@
 
-(i) A natural first extension is to retain the pulsar terms in Equation \eqref{eq:measurement}.
+%(i) A natural first extension is to retain the pulsar terms in Equation \eqref{eq:measurement}.
 Alternatively, if the method continues to be used with solely the Earth term, in line with standard practice in published PTA analyses, it would be desirable to evaluate systematically the incurred biases in the model parameters across an astrophysically representative parameter domain, supplementing the results obtained by other authors \citep{Zhupulsarterms,Chen2022}.\newline

### 2023-11-27 manuscript/manuscript_2/paper.tex
Whats with the shift in iota at low SNR?

@@
 % Keep the title short and informative.
-	\title[Kalman PTA]{Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array: inclusion of pulsar terms}
+	\title[Kalman Pulsar-terms]{State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms}
 
@@
 The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
-The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
+The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
 It is shown how to extend previous state-space analyses to include the pulsar terms.
@@
 The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}.
-All injected static parameters are summarised in the second column of Table \ref{tab:parameters_and_priors}.
+All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
 The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}.
@@
 
-Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for \textcolor{red}{YY} arbitrary noise realisations in the form of a traditional corner plot.
+Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for 10 arbitrary noise realisations in the form of a traditional corner plot.
 The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
+Each coloured curve is the result for different noise realisation.
 The solid orange line marks the known injected value.
@@
 There is scant evidence of railing against the prior bounds.
-\textcolor{red}{There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.} The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e.
+There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.
+The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e.
 \textcolor{red}{YY} $\%$) out of the \textcolor{red}{YY} possible combinations of the seven parameters and \textcolor{red}{YY} realizations.
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{example-image-a}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
 	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for nine realisations of the noise processes, with curves coloured uniquely.
@@
 
-%images/corner_lowSNR_chi
+%images/corner_lowSNR_chiF
 	\begin{figure}
-	\includegraphics[width=\columnwidth, height =\columnwidth ]{example-image-b}
+	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi}
 	\caption{Same as Figure \ref{fig:corner_plot_1}	but for 5 static parameters of $\chi^{(n)}$.

### 2023-11-27 manuscript/manuscript_2/paper.tex
bayes for multiple strains

@@
 
+\textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection.
+I think this is likely a sampler setting.
+I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.}
+
 \section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}

### 2023-11-27 py_src/configs/config_batch.py

@@ -30,21 +30,41 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
 
 
 
-# MULTIPLE NOISE REALISATIONS, low snr
+
+
+# Bayes vs h0, multiple noise
 N = 10
 seeds = np.arange(1235+10,1235+10+N,1)
-
-strains = [5e-15]
-models = ["pulsar"]
-
+strains = [5e-15,1e-14,5e-14,1e-13,5e-13,1e-12]
+models = ["pulsar","earth"]
 with open('batch.sh','w') as b: 
-
     for s in seeds:
         for h in strains:
             for m in models:
-                arg_name = f"eg_canonical_bias_exploration_single_bounding_{m}_{h}_{s}"
+                arg_name = f"eg_canonical_bayes_{m}_{h}_{s}"
                 create_slurm_job(arg_name,h,m,s)
                 b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+
+
+
+
+
+
+# MULTIPLE NOISE REALISATIONS, low snr
+# N = 10
+# seeds = np.arange(1235+10,1235+10+N,1)
+
+# strains = [5e-15]
+# models = ["pulsar"]
+
+# with open('batch.sh','w') as b: 
+
+#     for s in seeds:
+#         for h in strains:
+#             for m in models:
+#                 arg_name = f"eg_canonical_bias_exploration_single_bounding_{m}_{h}_{s}"
+#                 create_slurm_job(arg_name,h,m,s)
+#                 b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
 
 
 

### 2023-11-27 manuscript/manuscript_2/paper.tex
PP plot for large npoints

@@
 			\hline
-			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
+			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)_{\rm ATNF}} \right)  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
 			\bottomrule
@@
 The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
+Regarding point (iii), quantifying the dispersion is important since it is a practical measure of the accuracy of the parameter estimation scheme when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
 The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative.
@@
 
-Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for 10 arbitrary noise realisations in the form of a traditional corner plot.
+Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary noise realisations in the form of a traditional corner plot.
 The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
@@
 There is scant evidence of railing against the prior bounds.
-There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.
-The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e.
-\textcolor{red}{YY} $\%$) out of the \textcolor{red}{YY} possible combinations of the seven parameters and \textcolor{red}{YY} realizations.
-For this representative example, with characteristic strain \textcolor{red}{XX} the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
+There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
+The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e.
+70 $\%$).
+There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
+The minimum coefficient of variation is obtained for $\Omega$ at  0.2 \%, the maximum value is obtained for $\Phi_0$ at 43 \%.
+The remaining parameters have typical values $\lessapprox 10 \%$.
+For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
+Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (Figure \ref{fig:corner_high_snr_appendix}).
+At higher strains the dispersion between noise realisations is reduced.
 \newline
@@
 In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
-Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over \textcolor{red}{YY} noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
+Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the nine noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
 We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big.
-All elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations, with the injected value contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e.
-\textcolor{red}{YY} $\%$) out of the \textcolor{red}{YY} possible combinations of the seven parameters and \textcolor{red}{YY} realizations.
-Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$; the parameters are recovered unambiguously across the \textcolor{red}{XX} noise realisations.
+With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values.
+There is no evidence for correlations between parameter pairs.
+The majority of the elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations; for the 47 elements of $\chi^{n}$ (i.e.
+not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 325 out of the 423 possible combinations of the 47 parameters and nine realizations (i.e.
+$\sim$ 77 $\%$).
+There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$, (third row, third column of Figure \ref{fig:corner_plot_3}).
+This corresponds to the phase-correction term for PSR J0340+4130.
+This pulsar lies in the closest direction to our synthetic gw source, such that $| \boldsymbol{n} \cdot \boldsymbol{q}^{(2)}| = 0.96$.
+As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the pulsar term is minimal.
+The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately, see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
+\newline
+
+Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$; the injected values are recovered unambiguously across the nine noise realisations.
 For the sake of brevity we do not display the calculated posterior distributions here because the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses.
@@
 
-Figure \ref{fig:corner_plot_1} confirms two points: (i) the analysis scheme returns accurate answers for arbitrary noise realisations and (ii) there is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
-Regarding the second point, quantifying the dispersion is important since it is a practical measure of the accuracy of the parameter estimation scheme when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
-\newline
-
-% stacked_GW_plot_small_h_definitive2
+ \textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection.
+The systematic-ness makes me think this is likely a sampler setting.
+I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.}
+
 \begin{figure*}
@@
 The horizontal axes span a subset of the prior domain for all seven parameters.
-The known, injected value lies within the 90\% credible interval for \textcolor{red}{X} out of the 63 combinations of seven parameters and nine noise realizations.
-There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation \textcolor{red}{X} across the 63 combinations.}
+The known, injected value lies within the 90\% credible interval for 45 out of the 63 combinations of seven parameters and nine noise realizations.
+There is an appreciable dispersion among the peaks of the one-dimensional posteriors; the maximum coefficient of variation is $43 \%$ for $\Phi_0$, the minimum is $0.02 \%$ for $\Omega$.}
 	\label{fig:corner_plot_1}
@@
 	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi}
-	\caption{Same as Figure \ref{fig:corner_plot_1}	but for 5 static parameters of $\chi^{(n)}$.
-The known, injected value lies within the 90\% credible interval for \textcolor{red}{X} out of the \textcolor{red}{X} combinations of five parameters and v\textcolor{red}{X} noise realizations.
+	\caption{Same as Figure \ref{fig:corner_plot_1}	but for five static parameters of $\chi^{(n)}$.
+Consistent unimodal posteriors are obtained across the nine noise realisations for four out of the five displayed parameters.
+No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar has a similar sky position to that of the synthetic GW source.
+The known, injected value lies within the 90\% credible interval for 37 out of the 45 combinations of five parameters and nine noise realizations (i.e.
+$82 \%$).
 }
@@
 	\centering
-	\includegraphics[width=\columnwidth]{example-image-c}
+	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
 	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
@@
 For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope.
-$\Omega$ is generally well-estimated (i.e.
-it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e.
-the curves lie above the unit diagonal for low credible intervals, and below the unit diagonal for high credible intervals).
-This is due to a modelling bias whose origin is discussed in Sections \ref{sec:parameter_estim} and \ref{sec:pulsar_terms_drop}.}
+$\Omega$ and $\alpha$ are generally well-estimated (i.e.
+the curves lie close to the unit diagonal).
+The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
 	\label{fig:parameter_space}
 \end{figure}
-Section \ref{sec:pe_and_ms} focuses on a single representative system, parameterised in Table \ref{tab:parameters_and_priors}.
+Section \ref{sec:pe_and_ms} focuses on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
 In this section we test the method for various systems, varying the SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ through astrophysically relevant ranges.
@@
 
-We analyse \textcolor{red}{X} injections constructed by fixing $h_0 =$ \textcolor{red}{Y}  and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
-We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the  \textcolor{red}{X} injections.
+We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
 For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
-\newline
-
-The results are displayed in Figure \ref{fig:parameter_space}.
-Since it is prohibitive to produce corner plots analogous to Figure \ref{fig:corner_plot_1} for \textcolor{red}{X} injections, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:synthetic_data} and Table  \ref{tab:parameters_and_priors}.
+\newline
+
+It is prohibitive to produce corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$.
+Therefore, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
 A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself.
 In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
-In Figure \ref{fig:parameter_space} the shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for \textcolor{red}{X} injections.
-We see that...
-\textcolor{red}{TBD once simulations complete}
-
-\textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection.
-I think this is likely a sampler setting.
-I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.}
+
+The results are displayed in Figure \ref{fig:parameter_space}.  
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters generally exhibit the discussed linear behaviour.
+
+Some parameters are consistently well estimated across the parameter space; the blue curve for $\Omega$ is consistently contained within the 2-$\sigma$ shaded region.
+
+Conversely some parameters e.g.
+$\psi$ stray outside the $1\sigma$ bounded region and show evidence of being over constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and more injections contained within higher-value credible intervals.
+
+\textcolor{red}{TK: can we do better than this?}
+
+The
+
+At this signal-to-noise ratio
+
+the polarisation angle does stray above /below
+
+At lowe SNR the ability to resolve the depends upon the noise realisation and the source location.
+The antenna pattern function formed by the NANOGrav 12.5 year pulsars is sensitive to different parts of the sky.
+Areas with lower sensitivity may need to be
+
+Conversely the ability to infer $\Omega$ depends less on sky position or values.
 
@@
 The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
-The blue curves are the results derived using $\mathcal{M}_{E}$.
+The blue curves are the results derived using $\mathcal{M}_{\rm E}$.
 The green curves are the results derived using $\mathcal{M}_{\rm P \& E}$.
@@
 The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-We plot the scaled variables $10^9 \Omega$ (units: rad s$^{-1}$) and $10^{15} h_0$.
+The supertitles of the one-dimensional histograms record the posterior medians and the 0.16 and 0.84 quantiles of the green curves, i.e.
+the pulsar-term model.
+We plot the scaled variables $10^9 \Omega$ (units: rad s$^{-1}$) and $10^{12} h_0$.
 Some parameters (e.g.
-$\psi, \iota$) exhibit a bias when using the Earth-term model which is absent when using the pulsar-term model.
-\textcolor{red}{TK: need to remove these subtitles and correct the labels for omega and h.
-The actual results here may also be updated (same for Figure 5)}}
+$\psi, \iota$) exhibit a bias when using the Earth-term model which is absent when using the pulsar-term model.}
 	\label{fig:corner_plot_compare_high}
@@
 
+\section{Canonical example, high SNR} \label{sec:app_high_SNR}
+
+		\begin{figure}
+	\includegraphics[width=\columnwidth, height = \columnwidth ]{example-image-a} 	
+	\caption{}
+	\label{fig:corner_high_snr_appendix}
+\end{figure}
+
+	\begin{figure}
+	\includegraphics[width=\columnwidth, height = \columnwidth ]{example-image-b} 	
+	\caption{}
+	\label{fig:corner_high_snr_appendix_chi}
+\end{figure}
+
 \section{Multiple noise realisaitons and the Wasserstein distance}

### 2023-11-27 py_src/configs/config_batch_pp.py

@@ -1,0 +1,66 @@
+
+
+#As config_batch.py but now using main_pp.py for generating PP plots
+
+
+import os
+import sys 
+import numpy as np 
+import bilby
+
+
+def create_slurm_job(arg_name,h,measurement_model,seed,omega,phi0,psi,delta,alpha):
+
+    with open(f'slurm_jobs/slurm_{arg_name}.sh','w') as g:
+
+
+        g.write("#!/bin/bash \n \n")  
+        g.write("#SBATCH --ntasks=1 \n")  
+        g.write("#SBATCH --mem=8000MB \n")  
+        g.write("#SBATCH --time=48:00:00 \n")  
+        g.write(f"#SBATCH --job-name={arg_name} \n")  
+        g.write(f"#SBATCH --output=outputs/{arg_name}_out.txt \n \n")
+
+        g.write("source ~/.bashrc \n")
+        g.write("conda activate OzStar \n")
+        g.write(f"time python main_pp.py {arg_name} {h} {measurement_model} {seed} {omega} {phi0} {psi} {delta} {alpha}")
+    
+    
+
+N = 200
+seed = 1251 
+h = 5e-15 
+model = "pulsar"
+        
+#Create a bilby prior object to sample from
+init_parameters = {}
+priors = bilby.core.prior.PriorDict()
+
+init_parameters["omega_gw"] = None
+priors["omega_gw"] = bilby.core.prior.LogUniform(1e-8, 1e-6, 'omega_gw')
+
+init_parameters["phi0_gw"] = None
+priors["phi0_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'phi0_gw')
+
+init_parameters["psi_gw"] = None
+priors["psi_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'psi_gw')
+
+init_parameters["delta_gw"] = None
+priors["delta_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2, 'delta_gw')
+
+init_parameters["alpha_gw"] = None
+priors["alpha_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'alpha_gw')
+
+
+with open('batch.sh','w') as b: 
+
+    for i in range(N):
+        p = priors.sample()
+
[…]

### 2023-11-27 py_src/plotting.py

@@ -243,7 +243,7 @@ def _extract_value_from_title(title_string):
 def _argmedian(x):
     return np.argpartition(x, len(x) // 2)[len(x) // 2]
 
-def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,labels,injection_parameters,ranges,axes_scales,scalings=[1.0,1.0],savefig=None,logscale=False,title=None,smooth=True,smooth1d=True,seed=1):
+def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,labels,injection_parameters,ranges,axes_scales,scalings=[1.0,1.0],savefig=None,logscale=False,title=None,smooth=True,smooth1d=True,seed=1,no_titles=False):
 
     #Some arrays to hold the title value returned by corner.corner
     num_params = len(variables_to_plot)
@@ -376,8 +376,19 @@ def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,label
 
 
             new_title_string = rf'{labels[kk]} $= {title_values[kk,selected_idx]:.2f}_{{{title_lower[kk,selected_idx]:.2f}}}^{{+{title_upper[kk,selected_idx]:.2f}}}$'
-            ax.set_title(new_title_string, fontsize=18)
+            
+            
+            if no_titles:
+                selected_idx=1
+                new_title_string = rf'{labels[kk]} $= {title_values[kk,selected_idx]:.2f}_{{{title_lower[kk,selected_idx]:.2f}}}^{{+{title_upper[kk,selected_idx]:.2f}}}$'
+
+                #ax.set_title('')
+                ax.set_title(new_title_string, fontsize=18)
+            else:
+                ax.set_title(new_title_string, fontsize=18)
+            
             kk += 1
+
 
 
 

### 2023-11-27 manuscript/manuscript_2/paper.tex
Eod
n4000, low SNR

@@
 The systematic-ness makes me think this is likely a sampler setting.
-I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.}
+I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.
+This is not an issue in the high SNR case: can noise drive results away from the injection systematically?
+Naively I expect a scatter, symmetric about the injection...
+}
 
@@
 
+With a strain of X, we are close to the nosie floor (see e.g. figure X), so small variations in
+
 \textcolor{red}{TK: can we do better than this?}
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR_compare}
 	\caption{Same as Figure \ref{fig:corner_plot_compare_high}, but for a low-SNR system with $h_0 = 5 \times 10^{-15}$.
@@
 \section{Kalman filter} \label{sec:kalman}
-In this appendix, we describe the Kalman filter algorithm used in this paper.
-General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}.
-The mapping onto the specific continuous-time state-space model in Section \ref{sec:model} is written down in Section \ref{sec_kalman_specific}.
-
 The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$.
 It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
+In this appendix, we describe the Kalman filter algorithm used in this paper.
+General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the calculation of the likelihood.
+The application of the Kalman filter algorithm to the specific continuous-time state-space model in Section \ref{sec:2} is outlined in Section \ref{sec_kalman_specific}.
+
+%Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
+Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
+ %Implementation of the linear Kalman filter for the PTA state-space model in Section \ref{sec:model}, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
+%
+%At each discrete timestep indexed by , the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.
+
+\subsection{Recursion equations and likelihoood }\label{sec_kalman_general}
 In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
-Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
-Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
-Hence inferring the static parameters is a non-linear exercise, to be tackled separately after the linear Kalman filter operates on the data for a fixed set of static parameters.
-Inference of the static parameters in Equations \eqref{eq:measurement} and \eqref{eq:g_func} by nested sampling is discussed in Section \ref{sec:nested_sampling}.
-\newline
-
-Implementation of the linear Kalman filter for the PTA state-space model in Section \ref{sec:model}, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
-At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.
-The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by converting ${\boldsymbol{\hat X}}_i$ into predicted measurements ${\boldsymbol{\hat Y}}_i$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func} and comparing with the actual, noisy measurements ${\boldsymbol{\hat Y}}_i$.
-This defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation.
-The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$.
-The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}.
-For a fixed set of static parameters, the Kalman filter returns an estimate of the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error.
-We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline
-
-The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ can  then be calculated at each timestep from the Kalman filter output according to
-\begin{eqnarray}
-	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ .
-\end{eqnarray}
-The total log-likelihood for the entire sequence is
-\begin{eqnarray}
-	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
-\end{eqnarray}
-Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
-$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$.
-Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$.
-If $\boldsymbol{\hat{\theta}}$ is close to the true underlying parameters $\boldsymbol{\theta}$, then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are minimized and $\mathcal{L}$ is maximised.
-This is illustrated with synthetic data in Figure \ref{fig:kalman_example}.
-In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations \eqref{eq:spinevol}-- \eqref{eq:xieqn}, \eqref{eq:measurement}, and \eqref{eq:g_func} for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$.
-The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve).
-The predicted state $\hat{f}_{\rm p}^{(1)}(t)$ is converted  into a predicted measurement $\hat{f}_{\rm m}^{(1)}(t)$ (middle panel, magenta curve), which again overlaps almost perfectly with the true measurement.
-The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel).
-By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$) to the Kalman filter, where $\Omega$ is displaced from its true value by $20 \%$.
-In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates.
-It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel).
-In Section \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$.
-
-\subsection{Recursion equations}\label{sec_kalman_general}
-The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation
+The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k = \boldsymbol{Y}(t_k)$, which are related via a linear transformation to a set of unobservable discrete system states $\boldsymbol{X}_k= \boldsymbol{Y}(t_k)$.
+Each discrete timestep is indexed by $ 1 \leq k  \leq K$.
+The measurements are related to the states via
 \begin{equation}
@@
 \begin{align}
-	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \\
-	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \\
+	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \label{eq:residual} \\
+	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \label{eq:innv_covar}\\
 	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
@@
 \end{align}
+Equation \eqref{eq:residual} defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation.
+The uncertainty in $\boldsymbol{\epsilon}_i$ is quantified via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$, Equation \eqref{q:innv_covar}.
 Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
@@
 
+The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ is calculated at each timestep from the Kalman filter output according to
+\begin{eqnarray}
+	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ ,
+\end{eqnarray}
+where $N$ is the number of data points at timestep $i$. The total log-likelihood for the entire sequence is
+\begin{eqnarray}
+	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
+\end{eqnarray}
+Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
+$\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$.
+Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$.
+In Section \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$.
+\newline
+
 To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$.
-In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.
+In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:2}.
 
 \subsection{State space representation of a PTA analysis}\label{sec_kalman_specific}
-We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:model} as follows. \newline
+We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:2} as follows. \newline
 
@@
 \end{equation}
-
 The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$.
@@
 Specifically,
-$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}.
+$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func_trig}.
 The measurement covariance satisfies $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$ for all $k$.
@@
 \section{Nested sampling}\label{sec:nested_sampling}
-
 We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
@@
 \section{Choice of priors}\label{sec:set_priors}
-To deploy the analysis scheme of K24 it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters. \newline
+To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters.
+In this section we outline how the priors are chosen.
+The priors on each static parameter are summarised in Table \ref{tab:parameters_and_priors}.
+\newline
 
@@
 By using wider-than-necessary priors we expose the analysis scheme to a more stringent test.
-We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$) which covers the range of values inferred from Equation \ref{eq:sigmap_f}.
+We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$) which covers the range of values inferred from Equation \eqref{eq:sigmap_f}.
 We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.

### 2023-11-28 manuscript/manuscript_2/paper.tex
also do the null model for Bayes

@@
 In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model.
-In this section, critically, we compare the affect of including the pulsar-term in the model (i.e.
+In this section, critically, we compare the effect of including the pulsar-term in the model (i.e.
 using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e.
@@
 
-In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$ returned by the two models.
+In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm E}$ returned by the respective models.
 In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable strain for the two models.
@@
 
-We define a relative error metric to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
-\begin{equation}
-\Delta_{\rm M}(\theta) = \frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}} \ , \label{eq:mean_rel_error}
+We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
+\begin{equation}
+\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}}} \ , \label{eq:mean_rel_error}
 \end{equation}
@@
 The dependence of $p_k(\theta)$ on $M$ is implied.
-$\Delta_{\rm M}(\theta)$ represents the mean relative error in the one-dimensional posterior.
 The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}.
-The difference in the mean relative error between the two models, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is also presented.
-For every parameter except $\Omega$, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is negative, indicating that the parameter estimates are more accurate using the pulsar terms (i.e.
-the mean relative error when using the pulsar model is smaller).
-The exception for $\Omega$ is due to the fact that $\Omega$ is recovered with high accuracy in both cases; the the mean relative error is small $\Delta_{\rm M}(\Omega) \sim 10^{-5}$.
-The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm pulsar}(\iota) - \Delta_{\rm earth}(\iota) = -0.25$.
+The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is also presented.
+For every parameter $\Omega$, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms.
+In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm pulsar}(\Omega) = 6.7 \times 10^{-3}$.
+The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm pulsar}(\iota)  = 0.24$.
 \newline
@@
 The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest.
-The mean relative errors $\Delta_{\rm M}(\theta)$ for the low SNR case are also reported in Table \ref{tab:posterior_errors}.
-We see that....
-\textcolor{red}{TK: numbers to be inserted here}.
+The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR case is also reported in Table \ref{tab:posterior_errors}.
+We see that the error is larger for every element in $\theta_{\rm gw}$ than in the high SNR case, as expected.
+
+The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive for all parameters except $\psi$ and $h_0$.
+However it is hard to draw strong conclusions; for different realisations of the noise different parameters exhibit either improvement or deterioration in $Delta_{\rm M}(\theta)$.
 We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case.
@@
 		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
-		\begin{tabular}{lcclc}
+		\begin{tabular}{lcllc}
 			\toprule
-			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm pulsar}(\theta) -\Delta_{\rm earth}(\theta) $   \\
+			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm pulsar}(\theta)$   \\
 			\hline
-			\multirow{7}{2mm}{High SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
-			& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
-			& $\psi$ &0.0386& 0.000254& -0.0384 \\
-			& $\iota$ & 0.374 &0.126 &-0.248 \\
-			& $\delta$ & 0.00223 &0.000242& -0.00198 \\
-			&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
-			&$h_0$ & 0.168 &0.0846 &-0.0839 \\
+			\multirow{7}{2mm}{High SNR} & $\Omega$       & 0.011& 0.0043& 0.00671\\
+			& $\Phi_0$ &0.00332 &0.00115 &0.00217\\
+			& $\psi$ &0.0968 &0.000653 &0.0962 \\
+			& $\iota$ & 0.343 &0.106 &0.237 \\
+			& $\delta$ & 0.00187 &0.000285 &0.00158 \\
+			&$\alpha$ &0.0403 &0.000435 &0.0399 \\
+			&$h_0$ & 0.227 &0.075 &0.152 \\
 			\hline
-			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
-			& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
-			& $\psi$ &0.0386& 0.000254& -0.0384 \\
-			& $\iota$ & 0.374 &0.126 &-0.248 \\
-			& $\delta$ & 0.00223 &0.000242& -0.00198 \\
-			&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
-			&$h_0$ & 0.168 &0.0846 &-0.0839 \\
+			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& 1.09 \\
+			& $\Phi_0$ &0.335 &0.229 &0.106 \\
+			& $\psi$ &0.112 &0.123 &-0.0119 \\
+			& $\iota$ & 0.443 &0.37 &0.0724 \\
+			& $\delta$ & 0.181 &0.0782 &0.103 \\
+			&$\alpha$ &0.145 &0.139 &0.00637 \\
+			&$h_0$ & 1.12 &1.41 &-0.291 \\
 
@@
 		\end{tabular}
-		\caption{Mean relative error $\Delta_{\rm M}(\Theta)$ (Equation \eqref{eq:mean_rel_error}) in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$ ,Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
 The injection values are as summarised in Table \ref{tab:parameters_and_priors}.
 The top and bottom halves of the table contain the high-SNR ($h_0 = 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.
-The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.\textcolor{red}{TK numbers to be revised and double checked, but probably representative.
-Julian notes: swap so positive is good, normalise omega, add units, align}}
+For all parameters the units are in rad, except for $h_0$ which is dimensionless.
+The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.}
 		\label{tab:posterior_errors}
@@
 
-The Bayes factor, $\beta_i$, for each of the two models is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
+The Bayes factor, $\beta_i$, for each of the two models is presented in Figure \ref{fig:bayes1} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
+
+For a given $h_0$, we generate 10 realisations of synthetic data
+
 To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
@@
 \begin{equation}
-	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
+	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence2}
 \end{equation}
 We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper.
-Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}.
+Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence2}.
 It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
@@
 
-For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}.
+For $\pi(\boldsymbol{\theta}_{\rm gw})$ we have no \textit{a priori} information on the parameters.
+We therefore choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}.
 For $\pi(\boldsymbol{\theta}_{\rm psr})$ \textit{a priori} information from electromagnetic observations does exist.
@@
 By using wider-than-necessary priors we expose the analysis scheme to a more stringent test.
-We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$) which covers the range of values inferred from Equation \eqref{eq:sigmap_f}.
+We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$
+where $\sigma_{\rm SC}^{(n)}$ is the value for pulsar $n$ inferred from
+Equation \eqref{eq:sigmap_f}.
 We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.
@@
 This reduces the total dimension of the parameter space to $7 + 4N$.
-The chosen prior $\boldsymbol{\theta}$ is the same as used in K24, with the addition of a prior on $\chi^{(n)}$ due to the inclusion of the pulsar term.
-\newline
-
-The chosen priors for the representative analysis in Section \ref{sec:rep_example1} are summarised in \ref{Table tab:parameters_and_priors}
+\newline
 
 \section{Complete workflow summary}
-
 For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
@@
 
-\section{Canonical example, high SNR} \label{sec:app_high_SNR}
-
-		\begin{figure}
-	\includegraphics[width=\columnwidth, height = \columnwidth ]{example-image-a} 	
+\section{Parameter estimation of representative example, high-SNR} \label{sec:app_high_SNR}
+
+In this Section we repeat the analysis of Section \ref{sec:pe_and_ms}
+
+In this section, we apply the Kalman filter in conjunction with nested sampling to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ for synthetic data $\boldsymbol{Y}$, and compare it to the known, injected values.
+The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
+That is, given the injected static parameters $\boldsymbol{\theta'}$, multiple realisations of $\boldsymbol{Y}$ are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta'}$.
+The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
+Regarding point (iii), quantifying the dispersion is important since it is a practical measure of the accuracy of the parameter estimation scheme when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
+The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative.
+Different values of the static GW source parameters are explored in Section \ref{sec:parameter_space}.
+The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}.
+All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
+The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}.
+\newline
+
+Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary noise realisations in the form of a traditional corner plot.
+The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
+Each coloured curve is the result for different noise realisation.
+The solid orange line marks the known injected value.
+The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces.
+All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values.
+There is scant evidence of railing against the prior bounds.
+There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
+The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e.
+70 $\%$).
+There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
+The minimum coefficient of variation is obtained for $\Omega$ at  0.2 \%, the maximum value is obtained for $\Phi_0$ at 43 \%.
+The remaining parameters have typical values $\lessapprox 10 \%$.
+For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
+Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (Figure \ref{fig:corner_high_snr_appendix}).
+At higher strains the dispersion between noise realisations is reduced.
+\newline
+
+In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
+Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the nine noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
+We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big.
+With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values.
+There is no evidence for correlations between parameter pairs.
+The majority of the elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations; for the 47 elements of $\chi^{n}$ (i.e.
+not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 325 out of the 423 possible combinations of the 47 parameters and nine realizations (i.e.
+$\sim$ 77 $\%$).
+There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$, (third row, third column of Figure \ref{fig:corner_plot_3}).
+This corresponds to the phase-correction term for PSR J0340+4130.
+This pulsar lies in the closest direction to our synthetic gw source, such that $| \boldsymbol{n} \cdot \boldsymbol{q}^{(2)}| = 0.96$.
+As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the pulsar term is minimal.
+The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately, see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
+\newline
+
+Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$; the injected values are recovered unambiguously across the nine noise realisations.
+For the sake of brevity we do not display the calculated posterior distributions here because the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses.
+\newline
+
+		\begin{figure*}
+	\includegraphics[width=\textwidth, height = \textwidth ]{images/corner_highSNR} 	
 	\caption{}
 	\label{fig:corner_high_snr_appendix}
-\end{figure}
+\end{figure*}
 
 	\begin{figure}
-	\includegraphics[width=\columnwidth, height = \columnwidth ]{example-image-b} 	
+	\includegraphics[width=\columnwidth, height = \columnwidth ]{images/corner_highSNR_chi} 	
 	\caption{}
@@
 
-\section{Multiple noise realisaitons and the Wasserstein distance}
-
-We can extend the analysis of these two points and consider 1000 noise realisations of $\boldsymbol{Y}$, beyond the \textcolor{red}{X} plotted in Figure \ref{fig:corner_plot_1}.
+%
+%\section{Multiple noise realisaitons and the Wasserstein distance}
+%
+%
+%We can extend the analysis of these two points and consider 1000 noise realisations of $\boldsymbol{Y}$, beyond the \textcolor{red}{X} plotted in Figure \ref{fig:corner_plot_1}.
 As in Figure \ref{fig:corner_plot_1} we independently estimate the static parameters, $\boldsymbol{\theta}$ for each realisation and calculate $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$.
-
-To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
+%
+%
+%
+%
+%To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
 For each realisation we
-
-We now consider the complete set of $10^3$ noise realisations, going beyond the subset of nine realisations in the preceding discussion.
+%
+%
+%We now consider the complete set of $10^3$ noise realisations, going beyond the subset of nine realisations in the preceding discussion.
 We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$?
@@
 \newline
-
-Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
+%
+%
+%Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
 The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$.
@@
 Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
-
-\begin{table}
-	\centering
-	\begin{tabular}{ccll}
-		\toprule
-		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
-		\hline
-		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
-		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
-		$\psi$              & $2.50$ & rad & $0.16$ \\
-		$\iota$             & $1.0$ & rad & $0.14$ \\ 
-		$\delta$              & $1.0$  & rad & $0.09$ \\
-		$\alpha$          & $1.0$  & rad & $0.17$\\
-		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
-		\bottomrule
-	\end{tabular}
-	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
+%
+%\begin{table}
+%	\centering
+%	\begin{tabular}{ccll}
+%		\toprule
+%		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
+%		\hline
+%		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
+%		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
+%		$\psi$              & $2.50$ & rad & $0.16$ \\
+%		$\iota$             & $1.0$ & rad & $0.14$ \\ 
+%		$\delta$              & $1.0$  & rad & $0.09$ \\
+%		$\alpha$          & $1.0$  & rad & $0.17$\\
+%		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
+%		\bottomrule
+%	\end{tabular}
+%	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
 $W_{1, \rm median}$ is generally smaller than the prior domain.}
-	\label{tab:Wasserstein}
-\end{table}
+%	\label{tab:Wasserstein}
+%\end{table}
+%
+%
 
@@
 
-\subsection{Kalman filter and likelihood}\label{sec:kalman_filter}
-
 %See https://arxiv.org/pdf/2102.12478.pdf for a good NS explanation
 
-\subsection{Model selection}\label{sec:model_selection}
-The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
-\begin{equation}
-	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
-\end{equation}
-Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
+%
+%\section{Model selection}\label{sec:model_selection}
+%The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
+%\begin{equation}
+%	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
+%\end{equation}
+%Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
 $\mathcal{M}_0$ is the null model, which assumes no GW exists in the data, and is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
 The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW signal compared to there being no GW signal present.
+%
+%
 

### 2023-11-28 py_src/bilby_wrapper.py

@@ -29,7 +29,7 @@ def BilbySampler(KalmanFilter,init_parameters,priors,label,outdir):
                               bound = 'single',
                               sample='rwalk_dynesty',
                               check_point_plot=False,
-                              npoints=4000,
+                              npoints=2000,
                               dlogz=0.1,
                               npool=1,
                               plot=False,resume=False)

### 2023-11-28 py_src/configs/config_batch.py

@@ -28,22 +28,19 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
 
 
 
-
-
-
-
-# Bayes vs h0, multiple noise
-# N = 10
-# seeds = np.arange(1235+10,1235+10+N,1)
-# strains = [5e-15,1e-14,5e-14,1e-13,5e-13,1e-12]
-# models = ["pulsar","earth"]
-# with open('batch.sh','w') as b: 
-#     for s in seeds:
-#         for h in strains:
-#             for m in models:
-#                 arg_name = f"eg_canonical_bayes_{m}_{h}_{s}"
-#                 create_slurm_job(arg_name,h,m,s)
-#                 b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+#Bayes vs h0, multiple noise
+N = 10
+seeds = np.arange(1235+10,1235+10+N,1)
+strains = [5e-15,1e-14,5e-14,1e-13,5e-13,1e-12]
+#models = ["pulsar","earth"]
+models = ["null"]
+with open('batch.sh','w') as b: 
+    for s in seeds:
+        for h in strains:
+            for m in models:
+                arg_name = f"eg_canonical_bayes_{m}_{h}_{s}"
+                create_slurm_job(arg_name,h,m,s)
+                b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
 
 
 
@@ -51,20 +48,20 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
 
 
 # MULTIPLE NOISE REALISATIONS, low snr
-N = 10
-seeds = np.arange(1235+10,1235+10+N,1)
+# N = 10
+# seeds = np.arange(1235+10,1235+10+N,1)
 
-strains = [5e-15]
-models = ["pulsar"]
+# strains = [5e-15]
+# models = ["pulsar"]
 
-with open('batch.sh','w') as b: 
+# with open('batch.sh','w') as b: 
 
-    for s in seeds:
-        for h in strains:
-            for m in models:
-                arg_name = f"eg_canonical_bias_exploration_n4000_{m}_{h}_{s}"
-                create_slurm_job(arg_name,h,m,s)
-                b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+#     for s in seeds:
+#         for h in strains:
+#             for m in models:
[…]

### 2023-11-28 py_src/plotting.py

@@ -400,37 +400,38 @@ def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,label
 
 
     #Surface some numbers
-    print("Surfacing some numbers for comparing two posteriors")
+    print("Surfacing numbers for comparing two posteriors")
+    print("The RMSRE is:")
     if len(selected_files) ==2:
-        errors1 = error_files[0]
-        errors2 = error_files[1]
+        earth_term_errors = error_files[0]
+        pulsar_term_errors = error_files[1]
 
         #print(errors1)
         #print(errors2)
         #relative_error = (errors2 - errors1) / errors1
-        relative_error = (errors2 - errors1) #/ errors1
+        difference = (earth_term_errors - pulsar_term_errors) #/ errors1
 
         #print(relative_error)
-        for i in range(len(relative_error)):
-            print("%.3g" % errors1[i],"%.3g" %errors2[i],"%.3g" %relative_error[i]) #printing to 3 sig fig
+        for i in range(len(earth_term_errors)):
+            print(variables_to_plot[i], "%.3g" % earth_term_errors[i],"%.3g" %pulsar_term_errors[i],"%.3g" %difference[i]) #printing to 3 sig fig
 
 
 
 def get_posterior_accuracy(posterior,injection,labels):
 
     print("The error in the 1D posteriors is as follows:")
-    rmse_errors =np.zeros(posterior.shape[-1])
+    rmse_errors =np.zeros(posterior.shape[-1]) # vector of length n parameters
     for i in range(posterior.shape[-1]):
         y = posterior[:,i]
         inj = injection[i]
-        error = np.mean(np.abs(inj - y) / inj) #julian error
+        
+        #error = np.mean(np.abs(inj - y) / inj) #i.e. average absolute error
 
-        rmse = np.sqrt(np.sum((y - inj)**2) / len(y))
-        #rmse_errors[i] = rmse
-        rmse_errors[i] = error
+        rmse = np.sqrt(np.sum((y - inj)**2) / len(y)) # RMSRE https://stats.stackexchange.com/questions/413209/is-there-something-like-a-root-mean-square-relative-error-rmsre-or-what-is-t
+        rmse_errors[i] = rmse
 
-        print(labels[i], error,rmse)
-    print('*****************************')
+        #print(labels[i], error,rmse)
+    #print('*****************************')
 
 
     return rmse_errors

### 2023-11-28 manuscript/manuscript_2/paper.tex
PP plot for a known chi

@@
 % Keep the title short and informative.
-	\title[Kalman Pulsar-terms]{State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms}
+	\title[State-space PTA]{State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms}
 
@@
 In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
-
 The results are displayed in Figure \ref{fig:parameter_space}.
-The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters generally exhibit the discussed linear behaviour.
-
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections.
+The curves for the five parameters generally exhibit the discussed linear behaviour.
 Some parameters are consistently well estimated across the parameter space; the blue curve for $\Omega$ is consistently contained within the 2-$\sigma$ shaded region.
@@
 Conversely some parameters e.g.
-$\psi$ stray outside the $1\sigma$ bounded region and show evidence of being over constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and more injections contained within higher-value credible intervals.
+$\psi$ stray outside the $1\sigma$ bounded region and show evidence of being over constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and less injections contained within higher-value credible intervals.
 
@@
 
-\textcolor{red}{TK: can we do better than this?}
+Inspecting the individual resulting corner plots for each of the 200 injections, it is evident that for a minority of SMBHB sources the method either does not converge to a consistent unimodal one-dimensional posterior, or else a posterior is calculated but is slightly offset, not symmetric about the injected value.
+
+due to challenge of etimateng phi
 
@@
 
-At lowe SNR the ability to resolve the depends upon the noise realisation and the source location.
+At low SNR the ability to resolve the depends upon the noise realisation and the source location.
 The antenna pattern function formed by the NANOGrav 12.5 year pulsars is sensitive to different parts of the sky.
@@
 To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
+
 \footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.}

### 2023-11-28 manuscript/manuscript_2/paper.tex
try a tighter h prior

@@
 There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
+
+There is an evident offset in the peaks of the posteriors of $\iota$ and $h_0$ relative to the injection value, across all noise realisations.
+
+iota and h0 are offset/ not symmetric about the injection value
+
+The
+
 The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e.
 70 $\%$).
+
+Of the 18 posteriors which do not contain the injected value within the  90\% credible interval, 12
+
+This offset is due to the weak-identifiability of $\iota$-$h$
+
 There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
-The minimum coefficient of variation is obtained for $\Omega$ at  0.2 \%, the maximum value is obtained for $\Phi_0$ at 43 \%.
-The remaining parameters have typical values $\lessapprox 10 \%$.
+The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 43\% for $\Phi_0$.
+The remaining parameters have typical coefficients $\lessapprox 10 \%$.
 For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
@@
 \newline
+
+ Moreover the systematic offset in $\iota$ in $h_0$ is reduced and becomes symmetric about the injection value
+
+  \textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection.
+The systematic-ness makes me think this is likely a sampler setting.
+I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.
+This is not an issue in the high SNR case: can noise drive results away from the injection systematically?
+Naively I expect a scatter, symmetric about the injection...
+How is this related to the chi vals}
 
@@
 $\sim$ 77 $\%$).
-There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$, (third row, third column of Figure \ref{fig:corner_plot_3}).
+There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}).
 This corresponds to the phase-correction term for PSR J0340+4130.
@@
 \newline
-
- \textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection.
-The systematic-ness makes me think this is likely a sampler setting.
-I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.
-This is not an issue in the high SNR case: can noise drive results away from the injection systematically?
-Naively I expect a scatter, symmetric about the injection...
-}
 
@@
 For every parameter $\Omega$, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms.
-In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm pulsar}(\Omega) = 6.7 \times 10^{-3}$.
-The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm pulsar}(\iota)  = 0.24$.
+In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm pulsar}(\Omega) = 6.7 \times 10^{-3}$ rad.
+The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm pulsar}(\iota)  = 0.24$ rad.
 \newline
@@
 The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR case is also reported in Table \ref{tab:posterior_errors}.
-We see that the error is larger for every element in $\theta_{\rm gw}$ than in the high SNR case, as expected.
-
+We see that the error is larger for every element in $\boldsymbol{\theta}_{\rm gw}$ than in the high SNR case, as expected.
 The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive for all parameters except $\psi$ and $h_0$.
-However it is hard to draw strong conclusions; for different realisations of the noise different parameters exhibit either improvement or deterioration in $Delta_{\rm M}(\theta)$.
+However it is hard to draw strong conclusions; for different realisations of the noise different parameters exhibit either improvement or deterioration in $\Delta_{\rm M}(\theta)$.
 We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case.
@@
 		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
-		\begin{tabular}{lcllc}
+		\begin{tabular}{lclll}
 			\toprule
@@
 			\hline
-			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& 1.09 \\
+			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& \hspace{5mm} 1.09 \\
 			& $\Phi_0$ &0.335 &0.229 &0.106 \\
@@
 		\end{tabular}
-		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$ ,Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
 The injection values are as summarised in Table \ref{tab:parameters_and_priors}.

### 2023-11-28 py_src/configs/config_batch.py

@@ -29,39 +29,39 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
 
 
 #Bayes vs h0, multiple noise
-N = 10
-seeds = np.arange(1235+10,1235+10+N,1)
-strains = [5e-15,1e-14,5e-14,1e-13,5e-13,1e-12]
-#models = ["pulsar","earth"]
-models = ["null"]
-with open('batch.sh','w') as b: 
-    for s in seeds:
-        for h in strains:
-            for m in models:
-                arg_name = f"eg_canonical_bayes_{m}_{h}_{s}"
-                create_slurm_job(arg_name,h,m,s)
-                b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
-
-
-
-
-
-
-# MULTIPLE NOISE REALISATIONS, low snr
 # N = 10
 # seeds = np.arange(1235+10,1235+10+N,1)
-
-# strains = [5e-15]
-# models = ["pulsar"]
-
+# strains = [5e-15,1e-14,5e-14,1e-13,5e-13,1e-12]
+# #models = ["pulsar","earth"]
+# models = ["null"]
 # with open('batch.sh','w') as b: 
-
 #     for s in seeds:
 #         for h in strains:
 #             for m in models:
-#                 arg_name = f"eg_canonical_bias_exploration_n4000_{m}_{h}_{s}"
+#                 arg_name = f"eg_canonical_bayes_{m}_{h}_{s}"
 #                 create_slurm_job(arg_name,h,m,s)
 #                 b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
+
+
+
+
+
+
+#MULTIPLE NOISE REALISATIONS, low snr
+N = 10
+seeds = np.arange(1235+10,1235+10+N,1)
+
+strains = [5e-15]
+models = ["pulsar"]
+
+with open('batch.sh','w') as b: 
+
+    for s in seeds:
+        for h in strains:
+            for m in models:
+                arg_name = f"eg_canonical_bias_exploration_hprior_{m}_{h}_{s}"
+                create_slurm_job(arg_name,h,m,s)
+                b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")
 
 
 

### 2023-11-28 manuscript/manuscript_2/paper.tex
define null model properly

@@
 It is shown how to extend previous state-space analyses to include the pulsar terms.
-Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the previous approach.
+
+	 Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increases the parameter accuracy, and decrease the minimal detectable strain relative to the previous approach.
+
+	 For a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the root-mean-square relative error in the estimated one-dimensional posteriors decreases by up to 99\% by including the pulsar terms.
+
+	 For a quiet GW source with $h_0 = 5 \times 10^{-15}$ since the system is heavily noise dominated.
+
+	 However, including the pulsar terms decreases the minimal detectable strain
+
+	by \textcolor{red}{Y percent} .
 \end{abstract}
@@
 There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
-
-There is an evident offset in the peaks of the posteriors of $\iota$ and $h_0$ relative to the injection value, across all noise realisations.
-
-iota and h0 are offset/ not symmetric about the injection value
-
-The
-
+There is an evident offset in the peaks of the posteriors of $\iota$ and $h_0$ relative to the injection value, across all noise realisations \textcolor{red}{TK: to discuss}.
 The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e.
 70 $\%$).
-
-Of the 18 posteriors which do not contain the injected value within the  90\% credible interval, 12
-
-This offset is due to the weak-identifiability of $\iota$-$h$
-
 There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
@@
 Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (Figure \ref{fig:corner_high_snr_appendix}).
-At higher strains the dispersion between noise realisations is reduced.
-\newline
-
- Moreover the systematic offset in $\iota$ in $h_0$ is reduced and becomes symmetric about the injection value
-
-  \textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection.
-The systematic-ness makes me think this is likely a sampler setting.
-I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.
-This is not an issue in the high SNR case: can noise drive results away from the injection systematically?
-Naively I expect a scatter, symmetric about the injection...
-How is this related to the chi vals}
+At higher strains the dispersion between noise realisations is reduced and the one dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injection value.
+\newline
 
@@
 \begin{equation}
-\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}}} \ , \label{eq:mean_rel_error}
+\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{ \left[p_k(\theta) - \hat{\theta}\right]^2}{\hat{\theta}}} \ , \label{eq:mean_rel_error}
 \end{equation}
@@
 The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm pulsar}(\iota)  = 0.24$ rad.
+As a percentage change, i.e.
+$1 - \Delta_{\rm earth}(\theta)/\Delta_{\rm pulsar}(\theta)$, the improvement from using $\mathcal{M}_{\rm P \& E}$ over $\mathcal{M}_{E}$ ranges from 61\% for $\Omega$ to 99\% for $\psi$ and $\alpha$.
+These improvements are consistent across noise realisations.
 \newline
@@
 		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
-		\begin{tabular}{lclll}
+		\begin{tabular}{lcllc}
 			\toprule
@@
 			\hline
-			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& \hspace{5mm} 1.09 \\
+			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& 1.09 \\
 			& $\Phi_0$ &0.335 &0.229 &0.106 \\
@@
 			&$h_0$ & 1.12 &1.41 &-0.291 \\
-
 			\bottomrule

### 2023-11-28 manuscript/manuscript_2/paper.tex
handle null dict parsing

@@
 It is shown how to extend previous state-space analyses to include the pulsar terms.
-
 Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increases the parameter accuracy, and decrease the minimal detectable strain relative to the previous approach.
-
 For a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the root-mean-square relative error in the estimated one-dimensional posteriors decreases by up to 99\% by including the pulsar terms.
-
-	 For a quiet GW source with $h_0 = 5 \times 10^{-15}$ since the system is heavily noise dominated.
-
-	 However, including the pulsar terms decreases the minimal detectable strain
-
-	by \textcolor{red}{Y percent} .
+For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated.
+However, including the pulsar terms decreases the minimal detectable strain by \textcolor{red}{Y percent} .
 \end{abstract}
@@
 The intrinsic rotational state evolution of the pulsar, described by a mean-reverting Ornstein-Uhlenbeck process (Equations \eqref{eq:xieqn}--\eqref{eq:spinevol}) is tracked using a Kalman filter.
-The Kalman filter is combined with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model with and without a GW.
+The Kalman filter is combined with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model.
 \newline
@@
 The updated state-space model including the pulsar term is tested on synthetic data.
-We consider $X$ noise realisations of synthetic data for a single, astrophysically representative, SMBHB GW source observed by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$.
+We consider 10 noise realisations of synthetic data for a single, astrophysically representative, SMBHB GW source observed by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$ and $h_0 = 5 \times 10^{-15}$.
 We find that the updated state-space model successfully detects injected signals and estimates their static parameters accurately for all noise realisations with relatively low computational cost.
-There is a non-negligible natural random dispersion in the median of the marginalised one-dimensional posteriors which \textcolor{red}{...means something}.
+There is a non-negligible natural random dispersion in the median of the marginalised one-dimensional posteriors, with a maximum coefficient of variation of 43\% for $\Phi_0$, and a median value across the seven elements of $\boldsymbol{\theta}_{\rm gw}$ 7\%.
+
 The updated model is further tested across a broader parameter domain of SMBHB sources.
-500 randomly sampled parameter vectors at fixed $\iota$ and $h_0$ are explored; \textcolor{red}{the results are...} \newline
+200 randomly sampled parameter vectors of $\boldsymbol{\theta}_{\rm gw}$ are explored at fixed $\iota$ and $h_0$.
+
+Extensions to the state-space analysis scheme are discussed at length in K24.
+
+We note two relevant extensions here
 

### 2023-11-28 manuscript/manuscript_2/paper.tex
restart high strain bayes

@@
 It is shown how to extend previous state-space analyses to include the pulsar terms.
-Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increases the parameter accuracy, and decrease the minimal detectable strain relative to the previous approach.
-For a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the root-mean-square relative error in the estimated one-dimensional posteriors decreases by up to 99\% by including the pulsar terms.
+Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%.
 For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated.

### 2023-11-28 py_src/configs/config_batch.py

@@ -31,9 +31,11 @@ def create_slurm_job(arg_name,h,measurement_model,seed):
 #Bayes vs h0, multiple noise
 N = 10
 seeds = np.arange(1235+10,1235+10+N,1)
-strains = [5e-15,1e-14,5e-14,1e-13,5e-13,1e-12]
+#trains = [5e-15,1e-14,5e-14,1e-13,5e-13,1e-12]
+strains = [5e-13,1e-12]
+models = ["pulsar"]
 #models = ["pulsar","earth"]
-models = ["null"]
+#models = ["null"]
 with open('batch.sh','w') as b: 
     for s in seeds:
         for h in strains:

### 2023-11-28 manuscript/manuscript_2/paper.tex
eod

@@
 We compare the results when the pulsar terms are includes to those when they are omitted.
-\newline
+The state-space method is  \newline
 
@@
 There is a non-negligible natural random dispersion in the median of the marginalised one-dimensional posteriors, with a maximum coefficient of variation of 43\% for $\Phi_0$, and a median value across the seven elements of $\boldsymbol{\theta}_{\rm gw}$ 7\%.
-
 The updated model is further tested across a broader parameter domain of SMBHB sources.
 200 randomly sampled parameter vectors of $\boldsymbol{\theta}_{\rm gw}$ are explored at fixed $\iota$ and $h_0$.
-
-Extensions to the state-space analysis scheme are discussed at length in K24.
-
-We note two relevant extensions here
+Consistent parameter estimates are obtained for the majority of cases; \textcolor{red}{TBD once the main section updates}.
+\newline
 
 We compare the accuracy of the parameter estimates from the updated model to those when the pulsar term is omitted.
-We find that the inclision of the pulsar terms increases the accuaray by X percent in teh case of high SNR and Y per cent in the case of low SNR.
-The inclusion of pulsar terms also lowers the floor of the minimal detectable strain by YY percent.
-\newline
-
-\textcolor{red}{Caceats, extensions, other things}
-
-The new method is complementary to traditional approaches.
-We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than fitting for the ensemble-averaged power spectral density of the noise.
-That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data.
-
-We emphasize that the Kalman tracker and nested sampler in this paper do not supplant traditional PTA analysis approaches; they complement traditional approaches and are most powerful when used in tandem.
-Relatedly, it is misleading to ask whether the Kalman tracker and nested sampler are more or less sensitive than traditional approaches for two reasons.
-First, one must still generalize the Kalman tracker and nested sampler to ingest TOAs directly, as happens traditionally, instead of ingesting a frequency time series, as in this paper.
-Generalizing the algorithm is a subtle task and is postponed to a future work.
-Second, the Kalman tracker and nested sampler are conditional on a noise model that is related but different to the noise model in traditional approaches.
-
-The analysis in this paper assumes a specific, time-ordered realization of a ,
-
-whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law power spectral density, whose amplitude and exponent are adjustable.
-Hence the sensitivities cannot be compared directly.
+We find that the inclusion of the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) by 61\% for $\Omega$, up to 99\% for $\psi$ and $\alpha$ for a representative SMBHB system with a high SNR ($h_0 = 10^{-12}$).
+For low SNR ($h_0 = 5 \times 10^{-15}$) there is no significant improvement because the data is noise dominated.
+The inclusion of pulsar terms lowers the floor of the minimal detectable strain by \textcolor{red}{X}.
+\newline
+
+State-space methods for PTA data analysis are complementary to traditional approaches, and enable the evolution of the intrinsic pulsar timing noise to be tracked explicitly rather than fitting for the ensemble-averaged power spectral density of the noise.
+That is, it is possible to disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data.
+State-space models are conditional on a noise model that is related but different to the noise model in traditional approaches.
+The analysis in this paper assumes a specific, time-ordered realization of a mean-reverting Ornstein-Uhlenbeck process, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law power spectral density, whose amplitude and exponent are adjustable.
 Clarifying the similarities and differences between various approaches promises to be a fruitful avenue of future work.
@@
 
- The approach in this paper can be extended in at least five ways, enumerated as (i)--(v) below. \newline
-
-%(i) A natural first extension is to retain the pulsar terms in Equation \eqref{eq:measurement}.
-Alternatively, if the method continues to be used with solely the Earth term, in line with standard practice in published PTA analyses, it would be desirable to evaluate systematically the incurred biases in the model parameters across an astrophysically representative parameter domain, supplementing the results obtained by other authors \citep{Zhupulsarterms,Chen2022}.\newline
-
-(ii) In this paper, we consider one specific configuration of the synthetic PTA, namely the same pulsars that make up the 12.5-year NANOGrav (Section \ref{sec:synt_pta}).
-It is interesting to compare the performance of the method using different pulsar configurations, e.g.
-those in the PPTA and EPTA.
-Adding pulsars to PTAs increases the computational cost so it may be advantageous to select a subset of pulsars by exploiting formal optimization techniques from electrical engineering \citep{2023MNRAS.518.1802S}, although the Kalman filter and nested sampler are already cheaper computationally than some other methods.
-\newline
-
-(iii) In practice different PTA pulsars are observed with different cadences at different times.
-Extending the Kalman filter to non-uniform time sampling is straightforward \citep{zarchan2000fundamentals}.
-\newline
-
-(iv) The assumption of a monochromatic source is well-justified astrophysically in various regimes (see Section \ref{sec:plane_gw}) and is an appropriate starting point for this introductory paper.
+Extensions to the state-space methods for PTA data analysis are discussed at length in K24. We note three relevant extensions here:
+
+(i) The assumption of a monochromatic source is well-justified astrophysically in various regimes (see Section \ref{sec:plane_gw}).
 Nevertheless, SMBHBs are not strictly monochromatic.
-It is interesting to extend the state-space framework such that $f_{\rm gw}$ evolves in time.
-For $\Delta f_{\rm gw} < 1 / T_{\rm obs}$ (see Equation \eqref{eq:f_evolution}) evolution adds noise incoherently to the pulsar terms, whilst for $\Delta f_{\rm gw} > 1 / T_{\rm obs}$ the pulsar terms induce phase shifts that affect the overall phase coherence \citep{Sesana2010,Perrodin2018}.
-Careful consideration of the evolution of $f_{\rm gw}$ will be needed when including the pulsar terms in the inference model, as in point (i).
-\newline
-
-(v) Finally, we assume in this paper that there is only one GW source.
+In principle it is straightforward, to extend the analysis to account for the evolution of $\Omega(t)$, namely by including $\Omega(t)$ in the differential state equations.
+However, issues related to identifiability \citep{e5be7c83a0d24500826f6e1b414d1733}, the form of the measurement equation Equation \eqref{eq:g_func_trig}, and the linear structure the Kalman filter must be considered carefully and are postponed to a future paper.
+Regarding the last point, the Kalman filter can be applied to non-linear problems using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
+
+(ii) The state-space model presented here ingests a frequency time series.
+It is necessary to generalize this approach to ingest TOAs directly, as happens traditionally.
+Generalizing the algorithm is a subtle task and will be presented in a forthcoming work.
+
+(iii) We assume in this paper that there is only one GW source.
 However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
@@
 %Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
-Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
+
+% Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
  %Implementation of the linear Kalman filter for the PTA state-space model in Section \ref{sec:model}, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.

### 2023-11-28 manuscript/manuscript_2/paper.tex
another Bayes run

@@
 The model is formulated in terms of pulse frequency time series as a proof-of-principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
-It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, an extension we defer to future work.
+It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, an extension we defer to a forthcoming work.
 

### 2023-11-29 manuscript/manuscript_2/paper.tex
update manuscript for n4000 results

@@
 	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.
+The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
 The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
-The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-It is shown how to extend previous state-space analyses to include the pulsar terms.
+It is shown how to extend previous state-space analyses to include the pulsar term.
 Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%.
@@
 \end{equation}
-where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
-\begin{equation}
-	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ , \label{eq:model_evidence}
-\end{equation}
-and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}.
+where $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}, $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$, and $\mathcal{Z}$ is the marginalised likelihood, or evidence
+\begin{equation}
+	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
+\end{equation}
 For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
@@
 Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
-We refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, nested sampling, and the combined workflow in the context of PTA searches for GWs.
+In the context of PTA searches for GWs, we refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, Appendix \ref{sec:nested_sampling} for a review of nested sampling, Appendix \ref{sec:workflow} for an overview of the combined Kalman filter - nested sampling workflow.
 \newline
@@
 			\hline
-			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)_{\rm ATNF}} \right)  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
+			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}_{\rm ATNF} \right)  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
 			\bottomrule
 		\end{tabular}
-		\caption{Summary of injected static parameters used for generating synthetic data for validation of the analysis scheme when including the pulsar-term in Section \ref{sec:rep_example}.
+		\caption{Summary of the injected static parameters used for generating synthetic data for validation of the analysis scheme when including the pulsar-term in Section \ref{sec:pe_and_ms}.
 The prior used for Bayesian inference (Section \ref{sec:pe_and_ms}) on each parameter is also displayed (rightmost column).
@@
 The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synthetic_data}.
-The subscript ``SC'' on the pulsar noise amplitude $\sigma^{(n)}$ indicates that the injected value has been calculated using Equation and \eqref{eq:sigmap_f} the empirical timing noise model for MSPs from \protect \cite{Shannon2010}.
+The subscript ``SC'' on the pulsar noise amplitude $\sigma^{(n)}$ indicates that the injected value has been calculated using Equation \eqref{eq:sigmap_f} the empirical timing noise model for MSPs from \protect \cite{Shannon2010}.
 The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
@@
 The measured $f_{\rm m}^{(n)}(t)$ are then produced via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
-Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and \eqref{eq:measurement}--\eqref{eq:g_func_trig} depend on the choice of static parameters $\boldsymbol{\theta'}$, as well as the choice of pulsar directions ${\boldsymbol{q}}^{(n)}$ and detector measurement noise covariance $\sigma_{\rm m}$ (Equation \eqref{eq:sigma_m_eqn}).
+Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and \eqref{eq:measurement}--\eqref{eq:g_func_trig} depend on the choice of static parameters $\boldsymbol{\theta'}$, as well as the choice of pulsar directions ${\boldsymbol{q}}^{(n)}$ and detector measurement noise covariance $\sigma_{\rm m}$ (Equation \eqref{eq:vareps}).
 We now describe how $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ are chosen.
@@
 
-Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary noise realisations in the form of a traditional corner plot.
+Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for ten arbitrary noise realisations in the form of a traditional corner plot.
 The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
@@
 There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
-There is an evident offset in the peaks of the posteriors of $\iota$ and $h_0$ relative to the injection value, across all noise realisations \textcolor{red}{TK: to discuss}.
-The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e.
-70 $\%$).
+The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations (i.e.
+86$\%$).
+The injected value is contained within the 90\% credible interval in 100\% of cases for $\Omega, \Phi_0, \psi$, and  $\iota$.
+The injected value is contained within the 5\% and 95\% quantiles for all of the one-dimensional marginalized posteriors.
+
 There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
-The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 43\% for $\Phi_0$.
+The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 52\% for $\Phi_0$.
 The remaining parameters have typical coefficients $\lessapprox 10 \%$.
 For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
-Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (Figure \ref{fig:corner_high_snr_appendix}).
+Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR}, Figure \ref{fig:corner_high_snr_appendix}.
 At higher strains the dispersion between noise realisations is reduced and the one dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injection value.
@@
 
-In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
-Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the nine noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
+In addition to $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
+Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the ten noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
 We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big.
@@
 The majority of the elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations; for the 47 elements of $\chi^{n}$ (i.e.
-not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 325 out of the 423 possible combinations of the 47 parameters and nine realizations (i.e.
-$\sim$ 77 $\%$).
+not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations (i.e.
+$\sim$ 81 $\%$).
 There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}).
 This corresponds to the phase-correction term for PSR J0340+4130.
-This pulsar lies in the closest direction to our synthetic gw source, such that $| \boldsymbol{n} \cdot \boldsymbol{q}^{(2)}| = 0.96$.
-As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the pulsar term is minimal.
-The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately, see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
+This pulsar lies in the closest direction to our synthetic GW source, such that $| \boldsymbol{n} \cdot \boldsymbol{q}^{(2)}| = 0.96$.
+As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the GW is minimal ($g^{(2)}(t) \to 1$) The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately, see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
 \newline
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
-	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for nine realisations of the noise processes, with curves coloured uniquely.
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR_n4000}
+	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for ten realisations of the noise processes, with curves coloured uniquely.
 The horizontal and vertical orange lines indicate the true injected values.
@@
 The horizontal axes span a subset of the prior domain for all seven parameters.
-The known, injected value lies within the 90\% credible interval for 45 out of the 63 combinations of seven parameters and nine noise realizations.
-There is an appreciable dispersion among the peaks of the one-dimensional posteriors; the maximum coefficient of variation is $43 \%$ for $\Phi_0$, the minimum is $0.02 \%$ for $\Omega$.}
+The known, injected value lies within the 90\% credible interval for 60 out of the 70 combinations of seven parameters and ten noise realizations.
+There is an appreciable dispersion among the peaks of the one-dimensional posteriors; the maximum coefficient of variation is $52 \%$ for $\Phi_0$, the minimum is $0.02 \%$ for $\Omega$.}
 	\label{fig:corner_plot_1}
@@
 	\begin{figure}
-	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi}
+	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi_n4000}
 	\caption{Same as Figure \ref{fig:corner_plot_1}	but for five static parameters of $\chi^{(n)}$.
@@
 No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar has a similar sky position to that of the synthetic GW source.
-The known, injected value lies within the 90\% credible interval for 37 out of the 45 combinations of five parameters and nine noise realizations (i.e.
-$82 \%$).
+The known, injected value lies within the 90\% credible interval for 38 out of the 50 combinations of five parameters and nine noise realizations (i.e.
+$76 \%$).
 }
@@
 Section \ref{sec:pe_and_ms} focuses on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
-In this section we test the method for various systems, varying the SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ through astrophysically relevant ranges.
+In this section we test the method for a range of systems, varying the SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ through astrophysically relevant ranges.
 The aim of this section is to confirm that the analysis scheme can be applied across an astrophysically relevant parameter space and that the arbitrary choice of $\boldsymbol{\theta}_{\rm gw}$ in Section \ref{sec:pe_and_ms} is not particularly advantageous by accident.
@@
 
-It is prohibitive to produce corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$.
+It is prohibitive to display corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$.
 Therefore, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
@@
 We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
-We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm P \& E}$.
+We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$.
 We refer to the model where the pulsar-term is omitted as
-$\mathcal{M}_{\rm E}$.
-Model $\mathcal{M}_{\rm P \& E}$ is parameterised by $\boldsymbol{\theta}'$.
-Model $\mathcal{M}_{\rm E}$ is parameterised by $\boldsymbol{\theta}_{\rm E} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$.
+$\mathcal{M}_{\rm earth}$.
+Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$.
+Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$.
 Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}.
-The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm P \& E}$, as described in Appendix \ref{sec:set_priors}.
-\newline
-
-In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm E}$ returned by the respective models.
-In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable strain for the two models.
+The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}.
+\newline
+
+In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm earth}$ returned by the respective models.
+In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable GW strain for the two models.
 
@@
 In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model.
-For $\mathcal{M}_{\rm E}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm E})$.
-The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta}_{\rm E} | \boldsymbol{Y})$.
-For $\mathcal{M}_{\rm P \& E}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$.
+For $\mathcal{M}_{\rm earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm earth})$.
+The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta}_{\rm earth} | \boldsymbol{Y})$.
+For $\mathcal{M}_{\rm psr \& earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$.
 The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$.
@@
 The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
-The blue curves are the results derived using $\mathcal{M}_{\rm E}$.
-The green curves are the results derived using $\mathcal{M}_{\rm P \& E}$.
-\newline
-
-For the high-SNR case, Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for the Earth-term model are evidently biased, as was observed in K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
+The blue curves are the results derived using $\mathcal{M}_{\rm earth}$.
+The green curves are the results derived using $\mathcal{M}_{\rm psr \& earth}$.
+\newline
+
+For the high-SNR case, Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm earth}$ are evidently biased, as was observed in K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
 Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
 For example, for $\iota$, whilst the injected value is contained within the 90\% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians.
-The inclusion of the pulsar terms corrects for this bias; the green one-dimensional exhibit no bias and are generally symmetric about the injected value.
-Moreover, for every pair of parameters the injection values are contained within the 2-sigma equivalent contours for the pulsar-term model, which is not true in the Earth-terms case.
+The inclusion of the pulsar terms corrects for this bias; the green one-dimensional posteriors exhibit no bias and are generally symmetric about the injected value.
+Moreover, for every pair of parameters the injection values are contained within the 2-sigma equivalent contours for the $\mathcal{M}_{\rm psr \& earth}$, which is not true for $\mathcal{M}_{\rm earth}$.
 \newline
@@
 The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is also presented.
-For every parameter $\Omega$, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms.
+For every parameter, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms.
 In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm pulsar}(\Omega) = 6.7 \times 10^{-3}$ rad.
@@
 Qualitatively, the green and blue one-dimensional posteriors are generally overlaid.
-The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest.
+The two-dimensional contours suggest a slight improvement through using $\mathcal{M}_{\rm psr \& earth}$, but the difference is modest.
 The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR case is also reported in Table \ref{tab:posterior_errors}.
@@
 For all parameters the units are in rad, except for $h_0$ which is dimensionless.
-The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.}
+The rightmost column quotes the difference in the mean relative error between the models for each parameter; a positive value indicates that the pulsar-term model is more accurate.}
 		\label{tab:posterior_errors}
@@
 
-Extensions to the state-space methods for PTA data analysis are discussed at length in K24. We note three relevant extensions here:
-
-(i) The assumption of a monochromatic source is well-justified astrophysically in various regimes (see Section \ref{sec:plane_gw}).
+Extensions to the state-space methods for PTA data analysis are discussed at length in K24. We note three relevant extensions here: \newline
+
+(0) there is no real improvement in PE for this particular representative source. Exploration of a broader parameter space.... \textcolor{red}{TK: TBD}
+
+(i) The assumption of a monochromatic source is well-justified astrophysically in various regimes (see Section \ref{sec:psr_measured}).
 Nevertheless, SMBHBs are not strictly monochromatic.
@@
 However, issues related to identifiability \citep{e5be7c83a0d24500826f6e1b414d1733}, the form of the measurement equation Equation \eqref{eq:g_func_trig}, and the linear structure the Kalman filter must be considered carefully and are postponed to a future paper.
-Regarding the last point, the Kalman filter can be applied to non-linear problems using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
+Regarding the last point, the Kalman filter can be applied to non-linear problems if needed using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
 
@@
 
-\section{Complete workflow summary}
+\section{Complete workflow summary}\label{sec:workflow}
 For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
@@
 \section{Parameter estimation of representative example,  high-SNR} \label{sec:app_high_SNR}
-
-In this Section we repeat the analysis of Section \ref{sec:pe_and_ms}
-
-In this section, we apply the Kalman filter in conjunction with nested sampling to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ for synthetic data $\boldsymbol{Y}$, and compare it to the known, injected values.
-The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
-That is, given the injected static parameters $\boldsymbol{\theta'}$, multiple realisations of $\boldsymbol{Y}$ are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta'}$.
-The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
-Regarding point (iii), quantifying the dispersion is important since it is a practical measure of the accuracy of the parameter estimation scheme when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
-The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative.
-Different values of the static GW source parameters are explored in Section \ref{sec:parameter_space}.
-The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}.
-All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
-The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}.
-\newline
-
-Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary noise realisations in the form of a traditional corner plot.
-The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
-Each coloured curve is the result for different noise realisation.
-The solid orange line marks the known injected value.
-The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces.
+In this Section we repeat the analysis of Section \ref{sec:pe_and_ms}, but in the high-SNR limit.
+We do this as a sanity check to confirm that the state-space analysis scheme with the pulsar term included returns reasonable results in the ``easy" case with a strong GW signal.
+The high-SNR limit allows us to inspect the results without without confusion from a measurement noise which is large compared to the GW signal.
+\newline
+
+We apply the Kalman filter - nested sampling scheme as in Section \ref{sec:pe_and_ms}.
+All injected static parameters $\boldsymbol{\theta}'$, and the associated priors are as specified in Table \ref{tab:parameters_and_priors}, which the exception of $h_0$ which now takes a value $h_0 = 10^{-12}$.
+The procedure is undertaken for ten realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
+\newline
+
+Figure \ref{fig:corner_high_snr_appendix} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for ten arbitrary noise realisations in the form of a traditional corner plot.
+The figure is exactly analogous to Figure \ref{fig:corner_plot_1}.
 All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values.
-There is scant evidence of railing against the prior bounds.
-There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
-The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e.
-70 $\%$).
-There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
-The minimum coefficient of variation is obtained for $\Omega$ at  0.2 \%, the maximum value is obtained for $\Phi_0$ at 43 \%.
-The remaining parameters have typical values $\lessapprox 10 \%$.
-For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
-Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (Figure \ref{fig:corner_high_snr_appendix}).
-At higher strains the dispersion between noise realisations is reduced.
-\newline
-
-In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
-Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the nine noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
-We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big.
-With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values.
-There is no evidence for correlations between parameter pairs.
-The majority of the elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations; for the 47 elements of $\chi^{n}$ (i.e.
-not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 325 out of the 423 possible combinations of the 47 parameters and nine realizations (i.e.
-$\sim$ 77 $\%$).
-There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$, (third row, third column of Figure \ref{fig:corner_plot_3}).
-This corresponds to the phase-correction term for PSR J0340+4130.
-This pulsar lies in the closest direction to our synthetic gw source, such that $| \boldsymbol{n} \cdot \boldsymbol{q}^{(2)}| = 0.96$.
-As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the pulsar term is minimal.
-The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately, see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
-\newline
-
-Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$; the injected values are recovered unambiguously across the nine noise realisations.
-For the sake of brevity we do not display the calculated posterior distributions here because the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses.
-\newline
+There is no evidence of railing against the prior bounds.
+Correlations between $\iota$ and $h_0$ are evident over multiple noise realisations due to the weak-identifiability of these two parameters.
+For five of the seven parameters, there is effectively zero dispersion in the one-dimensional posterior medians between noise realisations.
+For the remaining two parameters, $\iota$ and $h_0$, an appreciable is an appreciable dispersion, again due to weak-identifiability.
+\newline
+
+Figure \ref{fig:corner_high_snr_appendix_chi} displays results a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the ten noise realisations.
+The figure is exactly analogous to Figure \ref{fig:corner_plot_3}, in the high-SNR limit.
+All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values, with no evidence for correlations between parameter pairs.
+Unlike in the low-SNR case, it is now possible to infer $\chi^{(2)}$ (the phase correction for PSR J0340+4130) consistently (third row, third column of Figure \ref{fig:corner_high_snr_appendix_chi}).
+Despite the proximity on the sky of PSR J0340+4130 to our synthetic gw source, the loundness of the GW signal is sufficient to enable $\chi^{(2)}$ to be estimated.
 
@@
 	\includegraphics[width=\textwidth, height = \textwidth ]{images/corner_highSNR} 	
-	\caption{}
+	\caption{Same as Figure \ref{fig:corner_plot_1}, but for synthetic data with $h_0 = 10^{-12}$.
+The Kalman filter and nested sampler estimate all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ accurately.
+There is a reduced dispersion in the  one-dimensional posterior medians between noise realisations compared to the low-SNR case (c.f.
+Figure \ref{fig:corner_plot_1}).
+An appreciable dispersion remains for $\iota$ and $h_0$, along with a correlation across noise realisations for these two parameters, due to their weak-identifiability.}
 	\label{fig:corner_high_snr_appendix}
@@
 	\includegraphics[width=\columnwidth, height = \columnwidth ]{images/corner_highSNR_chi} 	
-	\caption{}
+	\caption{Same as Figure \ref{fig:corner_plot_3}, but for synthetic data with $h_0 = 10^{-12}$.
+Consistent unimodal posteriors are obtained across all noise realisations and all five displayed parameters.
+It is now possible to infer $\chi^{(2)}$ due to the loundness of the GW signal, unlike in the low-SNR case where consistent posteriors for $\chi^{(2)}$ could not be estimated.}
 	\label{fig:corner_high_snr_appendix_chi}

### 2023-11-29 manuscript/manuscript_2/paper.tex
anohter pp plot run, n2000

@@
 It is shown how to extend previous state-space analyses to include the pulsar term.
-Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%.
+Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms corrects for biases in the parameter estimates and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%.
 For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated.
@@
 
-due to challenge of etimateng phi
+due to challenge of etimateng chi
 
@@
 Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:psr_v_earth_bayes} that they are important for increasing the detection probability for a given signal strength.
-\newline
+Moreover, we have considered only a single \newline
 

### 2023-11-29 py_src/configs/config_batch_pp.py

@@ -28,7 +28,7 @@ def create_slurm_job(arg_name,h,measurement_model,seed,omega,phi0,psi,delta,alph
     
 
 N = 200
-seed = 1250 
+seed = 1251 
 h = 5e-15 
 model = "pulsar"
         
@@ -57,7 +57,7 @@ with open('batch.sh','w') as b:
     for i in range(N):
         p = priors.sample()
 
-        arg_name = f"N200chi_single_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
+        arg_name = f"finalcanonical_single_pp_plot_h_{h}_model_{model}_seed_{seed}_omega_{p['omega_gw']}_phi0_{p['phi0_gw']}_psi_{p['psi_gw']}_delta_{p['delta_gw']}_alpha_{p['alpha_gw']}"
         print(arg_name)
         create_slurm_job(arg_name,h,model,seed,p['omega_gw'],p['phi0_gw'],p['psi_gw'],p['delta_gw'],p['alpha_gw'])
         b.write(f"sbatch slurm_jobs/slurm_{arg_name}.sh & \n")

### 2023-11-29 py_src/plotting.py

@@ -243,7 +243,7 @@ def _extract_value_from_title(title_string):
 def _argmedian(x):
     return np.argpartition(x, len(x) // 2)[len(x) // 2]
 
-def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,labels,injection_parameters,ranges,axes_scales,scalings=[1.0,1.0],savefig=None,logscale=False,title=None,smooth=True,smooth1d=True,seed=1,no_titles=False):
+def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,labels,injection_parameters,ranges,axes_scales,scalings=[1.0,1.0],savefig=None,logscale=False,title=None,smooth=True,smooth1d=True,seed=1,no_titles=False,plot_datapoints=True):
 
     #Some arrays to hold the title value returned by corner.corner
     num_params = len(variables_to_plot)
@@ -299,7 +299,7 @@ def stacked_corner(list_of_files,number_of_files_to_plot,variables_to_plot,label
                             label_kwargs=dict(fontsize=fs),
                             axes_scales = axes_scales,
                             weights = np.ones(nsamples)/nsamples,
-                            plot_datapoints=True,fig=fig)
+                            plot_datapoints=plot_datapoints,fig=fig)
 
 
         #Extract the axis titles 

### 2023-11-30 manuscript/manuscript_2/paper.tex
version passed to AM

@@
 	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.
-The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
+The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
 The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
@@
 For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated.
-However, including the pulsar terms decreases the minimal detectable strain by \textcolor{red}{Y percent} .
+However, including the pulsar terms decreases the minimal detectable strain by $8\%$ .
 \end{abstract}
@@
 Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}.
-A nHz GW influences the time of arrival (TOA) at the Earth of radio pulses from pulsars.
+A nHz GW can influence the time of arrival (TOA) at the Earth of radio pulses from pulsars.
 A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars.
@@
 We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}.
-Secondly we quantify how the inclusion of the pulsar term affects (i) the accuracy with which the static GW parameters can be estimated, and (ii) the minimal detectable GW strain, compared to the case where the pulsar terms are omitted.
+Secondly, we quantify how the inclusion of the pulsar term affects (i) the accuracy with which the static GW parameters can be estimated, and (ii) the minimal detectable GW strain, compared to the case where the pulsar terms are omitted.
 \newline
@@
 \end{equation}
-where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
+where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative where an overdot denotes a derivative with respect to $t$, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
 \begin{align}
@@
 \end{align}
-In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+In Equation \eqref{eq:xieqn}, $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
 The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
@@
 \end{equation} 
-where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline
+where $t_1$ labels the time of the first TOA. \newline
 
@@
 \subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
-In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation:
+In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation
 \begin{equation}
@@
 	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
+	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:vareps}
 \end{align}
-where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
+In Equation \eqref{eq:vareps}, $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars.
+In Equation \eqref{eq:measurement} the measurement function $g^{(n)}(t)$ is
 \begin{align}
@@
 	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
-	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
+	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ ,
 	\label{eq:g_func_trig}
@@
 
-Regarding point (ii) whilst in practice the pulsar locations vary with respect to the Earth, pulsar TOAs are defined relative to the Solar System barycentre (SSB).The barycentering correction is typically applied when generating TOAs, e.g.
+Regarding point (ii) whilst in practice the pulsar locations vary with respect to the Earth, pulsar TOAs are defined relative to the Solar System barycentre (SSB).
+The barycentering correction is typically applied when generating TOAs, e.g.
 with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
@@
 Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
-In the context of PTA searches for GWs, we refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, Appendix \ref{sec:nested_sampling} for a review of nested sampling, Appendix \ref{sec:workflow} for an overview of the combined Kalman filter - nested sampling workflow.
+Within the context of PTA searches for GWs, we refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, Appendix \ref{sec:nested_sampling} for a review of nested sampling and Appendix \ref{sec:workflow} for an overview of the combined Kalman filter - nested sampling workflow.
 \newline
@@
 The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties much greater than the target GW wavelength.
-It is therefore not possible to set the phase of the pulsar term.
 Consequently the pulsar term is often dropped in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
@@
 Nested sampling then returns estimates of $\boldsymbol{\theta'}$ and $\mathcal{Z}$.
-We refer the reader to Appendix \ref{sec:kalman} and K24 for details of the Kalman filter - nested sampling scheme.
+We refer the reader to Appendices \ref{sec:kalman} -- \ref{sec:workflow} and K24 for details of the Kalman filter - nested sampling scheme.
 We apply the analysis scheme to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
 The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-The goal of this section is to validate the analysis scheme when the pulsar terms are included in the inference model.
+The goal of this section is to validate the state-space analysis scheme when the pulsar terms are included in the inference model.
 \newline
@@
 In Section \ref{sec:synthetic_data} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme.
-In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$.
-The scheme is validated on multiple noise realisations of the pulsar process noise and the measurement noise in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
+In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e.
+the SMBHB system.
+The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
 In Section \ref{sec:parameter_space} we extend the tests of Section  \ref{sec:pe_and_ms}
@@
 In this paper we use the same values for $\boldsymbol{\theta}_{\rm psr}$ as in K24, i.e.
-we adopt the $N=47$ millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}, i.e.
-$N = 47$.
+we adopt the $N=47$ millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
 We assume all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period.
@@
 86$\%$).
-The injected value is contained within the 90\% credible interval in 100\% of cases for $\Omega, \Phi_0, \psi$, and  $\iota$.
+The injected value is contained within the 90\% credible interval for all posterior realisations of $\Omega, \Phi_0, \psi$, and  $\iota$.
 The injected value is contained within the 5\% and 95\% quantiles for all of the one-dimensional marginalized posteriors.
-
 There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
@@
 $\sim$ 81 $\%$).
-There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}).
+There are some elements of $\chi^{(n)}$ for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}).
 This corresponds to the phase-correction term for PSR J0340+4130.
-This pulsar lies in the closest direction to our synthetic GW source, such that $| \boldsymbol{n} \cdot \boldsymbol{q}^{(2)}| = 0.96$.
-As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the GW is minimal ($g^{(2)}(t) \to 1$) The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately, see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
+Out of all the pulsars in the array, this pulsar lies in the closest direction to our synthetic GW source, such that $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$.
+As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the GW is minimal ($g^{(2)}(t) \to 1$) The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with a larger strain $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
 \newline
@@
 	\caption{Same as Figure \ref{fig:corner_plot_1}	but for five static parameters of $\chi^{(n)}$.
-Consistent unimodal posteriors are obtained across the nine noise realisations for four out of the five displayed parameters.
+Consistent unimodal posteriors are obtained across the ten noise realisations for four out of the five displayed parameters.
 No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar has a similar sky position to that of the synthetic GW source.
-The known, injected value lies within the 90\% credible interval for 38 out of the 50 combinations of five parameters and nine noise realizations (i.e.
+The known, injected value lies within the 90\% credible interval for 38 out of the 50 combinations of five parameters and ten noise realizations (i.e.
 $76 \%$).
@@
 \subsection{Extension to broader parameter space} \label{sec:parameter_space}
-%images/pp_plot_new1
+\textcolor{red}{TK: Figures 3 and 4 may be updated.
+Some additional simulations are currently running to double check some things.
+I expect the current figures to be representative of any updated results.} \newline
 \begin{figure}
@@
 \end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi}
+	\caption{Same as Figure \ref{fig:parameter_space} but now assuming that $\chi^{(n)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior for the nested sampling inference).
+The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, are contained within the shaded region.
+}
+	\label{fig:parameter_space2}
+\end{figure}
+
 Section \ref{sec:pe_and_ms} focuses on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
@@
 In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
+\newline
+
 The results are displayed in Figure \ref{fig:parameter_space}.
@@
 The curves for the five parameters generally exhibit the discussed linear behaviour.
-Some parameters are consistently well estimated across the parameter space; the blue curve for $\Omega$ is consistently contained within the 2-$\sigma$ shaded region.
-
-Conversely some parameters e.g.
-$\psi$ stray outside the $1\sigma$ bounded region and show evidence of being over constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and less injections contained within higher-value credible intervals.
-
-With a strain of X, we are close to the nosie floor (see e.g. figure X), so small variations in
-
-Inspecting the individual resulting corner plots for each of the 200 injections, it is evident that for a minority of SMBHB sources the method either does not converge to a consistent unimodal one-dimensional posterior, or else a posterior is calculated but is slightly offset, not symmetric about the injected value.
-
-due to challenge of etimateng chi
-
-The
-
-At this signal-to-noise ratio
-
-the polarisation angle does stray above /below
-
-At low SNR the ability to resolve the depends upon the noise realisation and the source location.
-The antenna pattern function formed by the NANOGrav 12.5 year pulsars is sensitive to different parts of the sky.
-Areas with lower sensitivity may need to be
-
-Conversely the ability to infer $\Omega$ depends less on sky position or values.
+Some parameters are consistently well estimated across the parameter space; the blue curve for $\Omega$ is consistently contained within the 2$\sigma$ shaded region.
+Conversely some parameters, e.g.
+$\psi$, stray outside the $1\sigma$ bounded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and less injections contained within higher-value credible intervals.
+This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky.
+Manually inspecting the individual corner plots for each of the 200 injections, it is evident that for a minority of SMBHB sources which lie in an unfavourable direction ($\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} = -1$) the method either does not converge to a consistent unimodal one-dimensional posterior, or else a posterior is converged to but the estimate is suboptimal.
+For these sources their location with respect to the PTA makes it difficult to infer some elements of $\chi^{(n)}$ accurately, which impacts the inference of the other elements of $\boldsymbol{\theta}_{\rm gw}$.
+All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$, $n_{\rm live}$ etc.
+Some sources in unfavourable sky locations  may only be resolvable with different observation settings; we defer an in-depth exploration of sensitivity across the parameter space to a future work.
+In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but now we assume that the injected values of $\chi^{(n)}$ are known exactly, i.e.
+setting a delta-function prior on $\chi^{(n)}$ during the nested sampling inference.
+Now the excursions out of the shaded region are almost completely reduced.
+This indicates the importance of accurate estimates of $\chi^{(n)}$ and the sensitivity of a specific PTA configuration to different parts of the sky.
+In spite of the small excursions of some parameters in Figure \ref{fig:parameter_space}, it is reassuring that in this low-SNR regime all parameters generally follow the desired linear trend, and that key parameters such as $\Omega$ lie close to the unit diagonal.
 
@@
 In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm earth}$ returned by the respective models.
-In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable GW strain for the two models.
+In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable GW strain for the two models by calculating the model evidence and comparing it to the evidence for a null model that does not contain a GW.
 
@@
 \end{equation}
-where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, psr} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
 The dependence of $p_k(\theta)$ on $M$ is implied.
 The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}.
-The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is also presented.
-For every parameter, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms.
-In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm pulsar}(\Omega) = 6.7 \times 10^{-3}$ rad.
-The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm pulsar}(\iota)  = 0.24$ rad.
+The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is also presented.
+For every parameter, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms, i.e.
+$\Delta_{\rm earth}(\theta) > \Delta_{\rm psr}$.
+In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm psr}(\Omega) = 6.7 \times 10^{-3}$ rad.
+The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm psr}(\iota)  = 0.24$ rad.
 As a percentage change, i.e.
-$1 - \Delta_{\rm earth}(\theta)/\Delta_{\rm pulsar}(\theta)$, the improvement from using $\mathcal{M}_{\rm P \& E}$ over $\mathcal{M}_{E}$ ranges from 61\% for $\Omega$ to 99\% for $\psi$ and $\alpha$.
+$1 - \Delta_{\rm earth}(\theta)/\Delta_{\rm psr}(\theta)$, the improvement from using $\mathcal{M}_{\rm psr \& earth}$ over $\mathcal{M}_{\rm earth}$ ranges from 61\% for $\Omega$ to 99\% for $\psi$ and $\alpha$.
 These improvements are consistent across noise realisations.
@@
 We see that the error is larger for every element in $\boldsymbol{\theta}_{\rm gw}$ than in the high SNR case, as expected.
-The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive for all parameters except $\psi$ and $h_0$.
+The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is positive for all parameters except $\psi$ and $h_0$.
 However it is hard to draw strong conclusions; for different realisations of the noise different parameters exhibit either improvement or deterioration in $\Delta_{\rm M}(\theta)$.
@@
 Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:psr_v_earth_bayes} that they are important for increasing the detection probability for a given signal strength.
-Moreover, we have considered only a single \newline
+\newline
 
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR_compare}
+	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR_compare_n4000}
 	\caption{Same as Figure \ref{fig:corner_plot_compare_high}, but for a low-SNR system with $h_0 = 5 \times 10^{-15}$.
-The posterior distributions from the Earth-term and pulsar-term models are generally overlaid }
+The posterior distributions from the Earth-term and pulsar-term models are generally overlaid.
+The two-dimensional contours suggest a slight improvement through using $\mathcal{M}_{\rm psr \& earth}$; qualitatively the injected value generally lies closer to the centre of the green contours than the blue contours.
+However the difference is modest and contingent on the specific realisation of the data.}
 	\label{fig:corner_plot_compare_low}
@@
 			\toprule
-			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm pulsar}(\theta)$   \\
+			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm psr}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm psr}(\theta)$   \\
 			\hline
@@
 		\end{tabular}
-		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=psr, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
 The injection values are as summarised in Table \ref{tab:parameters_and_priors}.
@@
 \subsection{Detectability vs $h_0$}\label{sec:psr_v_earth_bayes}
-
+		\textcolor{red}{TK: Figure 7 may be updated with the results of ongoing simulations.
+However, this won't affect any general conclusions we draw; just a smoother graph and a slight change in minimal detectable strain.} \newline
 		\begin{figure}
 			\includegraphics[width=\columnwidth, height = \columnwidth ]{images/CanonicalBayesPlot2000} 	
-			\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strains, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
+			\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_i$, $i \in \left \{ \rm psr \& earth, earth \right \}$ (GW present in data) and $\mathcal{M}_{\rm null}$ (GW not present in data) at different GW strains, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
 The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
-The minimum detectable strain, for $\beta < 10$, equals $4 \times 10^{-15}$.
-Missing points for $\beta \lesssim 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
+The minimum detectable strain for which $\beta < 10$ equals $3.19 \times 10^{-15}$ for $\mathcal{M}_{\rm earth}$ and $2.93 \times 10^{-15}$ for $\mathcal{M}_{\rm psr \& earth}$.
+\textcolor{red}{TK: we should also mention scatter at lower h, but this should be eliminated when new simulations complete.}}
 			\label{fig:bayes1}
 		\end{figure}
-The evidence integral $\mathcal{Z}$ returned by nested sampling, Equation \eqref{eq:model_evidence}, is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$.
+
+In this section we compute the minimal detectable strain for the representative source, Table \ref{tab:parameters_and_priors}, using $\mathcal{M}_{\rm earth}$ and $\mathcal{M}_{\rm psr \& earth}$.
+We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure.
 We define $\mathcal{M}_{\rm null}$ as the null model that assumes no GW exists in the data.
 This is equivalent to setting $g^{(n)}(t)=1$ in Equation \eqref{eq:g_func_trig}.
+The evidence integral $\mathcal{Z}$ returned by nested sampling, Equation \eqref{eq:model_evidence}, is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$.
 The support in the data for the presence of a GW signal, described by model $\mathcal{M}_i$, over the absence of a GW signal is quantified via a Bayes factor
@@
 		\end{equation}
-In this paper we consider $i \in \left \{ \rm P\& E, E\right \}$ \newline
+In this paper we consider $i \in \left \{ \rm psr \& earth, earth \right \}$ \newline
 
 The Bayes factor, $\beta_i$, for each of the two models is presented in Figure \ref{fig:bayes1} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
-
-For a given $h_0$, we generate 10 realisations of synthetic data
-
-To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
-
-\footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.}
-
-	Multiple noise...
-
-		We see in Figure \ref{fig:bayes} an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$.
-The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 4 \times 10^{-15}$.
-Of course, the minimum detectable strain is particular to the system in Table \ref{tab:parameters_and_priors}.
-It is influenced in general by $T_{\rm obs}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$, as discussed in Section \ref{sec:parameter_space}.
-Adjusting $\sigma_{\rm m}$ we find that an approximate quadratic relationship $\beta \propto h_0^2 / \sigma_{\rm m}^2$ also exists, where $h_0 / \sigma_{\rm m}$ is the effective signal-to-noise ratio.
-\newline
-
-%		The odds ratio $\beta$ drops off for $h_0 \lesssim 4 \times10^{-15}$.
-This happens because the two competing models becoming increasingly indistinguishable once the measurement noise dominates the GW signal.
-Moreover, the points in Figure \ref{fig:bayes} become sparser for $h_0 \lesssim 4 \times 10^{-15}$.
-This happens because we obtain $\beta <0$ for the missing, intermediate points.
-This is a noise artefact of the nested sampler.
-When the sampler converges sub-optimally, the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails, and $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-There is nothing special about the particular missing points; if one recreates the $\beta(h_0)$ curve by rerunning the nested sampler with another random seed, a different set of points are missing.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$, because the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(n_{\rm live}^{-1/2})$ (see Section \ref{sec:nested_sampling}).
-In this section we set $n_{\rm live} =2000$.
-%		
-%		
-%
-
-%		
-%		We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}.
-$\mathcal{M}_1$ is the Earth-terms only model, i.e.
-the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}.
-The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$.
-The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$.
-We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$.
-We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW.
-For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$.
-This minimal detectable strain is of course particular to this system and will be influenced in general by e.g.
-$T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky.
-\newline
-
-%		There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$.
-The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate.
-The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method.
-For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails.
-That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
-There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$.
-For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds.
-Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Appendix \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$.
-For the results presented in this subsection we set $n_{\rm live} =2000$.
-%
+As in Section \ref{sec:psr_v_earth_pe} in order to enable a clear comparison between the results from $\mathcal{M}_{\rm psr \& earth}$ and $\mathcal{M}_{\rm psr}$ we consider only a single noise realisation of the synthetic data $\boldsymbol{Y}$.
+Moreover, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
+The conclusions drawn below are consistent across noise realisations.
+\newline
+
+We see in Figure \ref{fig:bayes1} an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$ for both $\mathcal{M}_{\rm earth}$ and $\mathcal{M}_{\rm psr \& earth}$.
+Moreover, $\beta_{\rm psr \& earth} > \beta_{\rm earth}$ for all $h_0$.
+That is, for a given $h_0$, $\mathcal{M}_{\rm psr \& earth}$ provides greater evidence for a GW signal in the noisy data than $\mathcal{M}_{\rm psr \& earth}$.
+The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.19 \times 10^{-15}$ for  $\mathcal{M}_{\rm earth}$ and for $h_0 \gtrsim 2.93 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& earth}$, a relative improvement in the minimal detectable strain of $8\%$.
+The minimum detectable strain and the relative improvement through using $\mathcal{M}_{\rm psr \& earth}$ is of course particular to the system in Table \ref{tab:parameters_and_priors} and the realisation of $\boldsymbol{Y}$.
+It will be influenced in general by $T_{\rm obs}$, $T_{\rm cad}$ as well as ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$.
 
@@
 We compare the results when the pulsar terms are includes to those when they are omitted.
-The state-space method is  \newline
+\newline
 
 The intrinsic rotational state evolution of the pulsar, described by a mean-reverting Ornstein-Uhlenbeck process (Equations \eqref{eq:xieqn}--\eqref{eq:spinevol}) is tracked using a Kalman filter.
+An updated state-space model for the influence of the GW is introduced that reparameterises the pulsar term as a separate pulsar-specific static parameter $\chi^{(n)}$ to be inferred for each pulsar.
 The Kalman filter is combined with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model.
@@
 We find that the updated state-space model successfully detects injected signals and estimates their static parameters accurately for all noise realisations with relatively low computational cost.
-There is a non-negligible natural random dispersion in the median of the marginalised one-dimensional posteriors, with a maximum coefficient of variation of 43\% for $\Phi_0$, and a median value across the seven elements of $\boldsymbol{\theta}_{\rm gw}$ 7\%.
-The updated model is further tested across a broader parameter domain of SMBHB sources.
-200 randomly sampled parameter vectors of $\boldsymbol{\theta}_{\rm gw}$ are explored at fixed $\iota$ and $h_0$.
-Consistent parameter estimates are obtained for the majority of cases; \textcolor{red}{TBD once the main section updates}.
-\newline
-
-We compare the accuracy of the parameter estimates from the updated model to those when the pulsar term is omitted.
+There is a non-negligible natural random dispersion in the median of the marginalised one-dimensional posteriors, with a maximum coefficient of variation of 52\% for $\Phi_0$, and a median value across the seven elements of $\boldsymbol{\theta}_{\rm gw}$ of 10\%.
+The updated model is further tested across a broader parameter domain of SMBHB sources; 200 randomly sampled parameter vectors of $\boldsymbol{\theta}_{\rm gw}$ are explored at fixed $\iota$ and $h_0$.
+Consistent parameter estimates are obtained for the majority of cases, although for some SMBHB source positions on the sky the sensitivity of the specific PTA configuration used in this paper is reduced leading to sub-optimally converged posteriors.
+\newline
+
+We compare the accuracy of the parameter estimates from the updated model to those when the pulsar term is omitted for a representative SMBHB source.
 We find that the inclusion of the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) by 61\% for $\Omega$, up to 99\% for $\psi$ and $\alpha$ for a representative SMBHB system with a high SNR ($h_0 = 10^{-12}$).
 For low SNR ($h_0 = 5 \times 10^{-15}$) there is no significant improvement because the data is noise dominated.
-The inclusion of pulsar terms lowers the floor of the minimal detectable strain by \textcolor{red}{X}.
-\newline
+The inclusion of pulsar terms lowers the floor of the minimal detectable strain by $8\%$ \newline
 
@@
 
-Extensions to the state-space methods for PTA data analysis are discussed at length in K24. We note three relevant extensions here: \newline
-
-(0) there is no real improvement in PE for this particular representative source. Exploration of a broader parameter space.... \textcolor{red}{TK: TBD}
-
-(i) The assumption of a monochromatic source is well-justified astrophysically in various regimes (see Section \ref{sec:psr_measured}).
+There are at least four potential extensions to this work, as follows:
+
+\begin{enumerate}[leftmargin=2em]
+
+	\item For the single representative SMBHB source in Section \ref{sec:psr_v_earth_pe} there was no obvious improvement in the parameter estimation accuracy in the low-SNR case.
+It would be of interest to explore a broader parameter space of $\boldsymbol{\theta}'$ (i.e.
+varying both the SMBHB source and the PTA configuration) to quantify the accuracy gained through including the pulsar terms across an astrophysically representative regime.
+A similar exercise should be undertaken for Section \ref{sec:psr_v_earth_bayes} to quantify the improvement in the minimal detectable strain.
+\newline
+
+	\item The assumption of a monochromatic source is well-justified astrophysically in various regimes (see Section \ref{sec:psr_measured}).
 Nevertheless, SMBHBs are not strictly monochromatic.
@@
 Regarding the last point, the Kalman filter can be applied to non-linear problems if needed using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
-
-(ii) The state-space model presented here ingests a frequency time series.
+\newline
+
+	\item The state-space model presented here ingests a frequency time series.
 It is necessary to generalize this approach to ingest TOAs directly, as happens traditionally.
 Generalizing the algorithm is a subtle task and will be presented in a forthcoming work.
-
-(iii) We assume in this paper that there is only one GW source.
+\newline
+
+	\item We assume in this paper that there is only one GW source.
 However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
@@
 Taking the logic further, the stochastic background itself is arguably an incoherent sum of many individual GW sources.
-As long as a way can be found to summarize economically the many static parameters associated with the background sources, it should be possible for a Kalman filter and nested sampler to operate together to detect the stochastic background by generalizing the model selection procedure in Sections \ref{sec:model_selection} and \ref{sec:detection}.
+As long as a way can be found to summarize economically the many static parameters associated with the background sources, it should be possible for a Kalman filter and nested sampler to operate together to detect the stochastic background.
 Summarizing the parameter set economically, while respecting the mathematical structure of the Kalman filter, is a subtle challenge, which we postpone to a forthcoming paper.
 If successful, it will complement the traditional approach of cross-correlating pulsar residuals to uncover the Hellings-Downs curve \citep{Hellings,2023ApJ...951L...8A}.
-\newline
+\end{enumerate}
 
@@
 \subsection{Recursion equations and likelihoood }\label{sec_kalman_general}
-In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
+In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement).
 The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k = \boldsymbol{Y}(t_k)$, which are related via a linear transformation to a set of unobservable discrete system states $\boldsymbol{X}_k= \boldsymbol{Y}(t_k)$.
@@
 Equation \eqref{eq:residual} defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation.
-The uncertainty in $\boldsymbol{\epsilon}_i$ is quantified via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$, Equation \eqref{q:innv_covar}.
+The uncertainty in $\boldsymbol{\epsilon}_i$ is quantified via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$, Equation \eqref{eq:innv_covar}.
 Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
@@
 Informal empirical tests conducted as part of this paper support the trade-off suggested by \cite{Ashton2022}; we find typically that the true ${\boldsymbol{\theta}}$ is contained within the 90\% credible interval of the one-dimensional marginalised posteriors of ${\boldsymbol{\hat{\theta}}}$ for $n_{\rm live} > 7 + 5N$ with $N \leq 50$.
-Unless stated otherwise we take $n_{\rm live} = 1000$ for all results presented in this work.
-\newline
-
-\section{Choice of priors}\label{sec:set_priors}
-To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters.
-In this section we outline how the priors are chosen.
-The priors on each static parameter are summarised in Table \ref{tab:parameters_and_priors}.
-\newline
-
-For $\pi(\boldsymbol{\theta}_{\rm gw})$ we have no \textit{a priori} information on the parameters.
-We therefore choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}.
-For $\pi(\boldsymbol{\theta}_{\rm psr})$ \textit{a priori} information from electromagnetic observations does exist.
-We adopt constrained uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database.
-By using wider-than-necessary priors we expose the analysis scheme to a more stringent test.
-We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$
-where $\sigma_{\rm SC}^{(n)}$ is the value for pulsar $n$ inferred from
-Equation \eqref{eq:sigmap_f}.
-We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.
-We do not set a prior on $\gamma^{(n)}$, because $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and consequently $\gamma^{(n)}$ is effectively ``unobservable'' over $T_{\rm obs}$.
-For validation purposes it is therefore sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value.
-This reduces the total dimension of the parameter space to $7 + 4N$.
+Unless stated otherwise we take $n_{\rm live} = 2000$ for all results presented in this work.
 \newline
@@
 
+\section{Choice of priors}\label{sec:set_priors}
+To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} and the workflow of Appendix \ref{sec:workflow} it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters.
+In this section we outline how the priors are chosen.
+The priors on each static parameter are summarised in Table \ref{tab:parameters_and_priors}.
+\newline
+
+For $\pi(\boldsymbol{\theta}_{\rm gw})$ we have no \textit{a priori} information on the parameters.
+We therefore choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}.
+For $\pi(\boldsymbol{\theta}_{\rm psr})$ \textit{a priori} information from electromagnetic observations does exist.
+We adopt constrained uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database.
+By using wider-than-necessary priors we expose the analysis scheme to a more stringent test.
+We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$
+where $\sigma_{\rm SC}^{(n)}$ is the value for pulsar $n$ inferred from
+Equation \eqref{eq:sigmap_f}.
+We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.
+We do not set a prior on $\gamma^{(n)}$, because $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and consequently $\gamma^{(n)}$ is effectively ``unobservable'' over $T_{\rm obs}$.
+For validation purposes it is therefore sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value.
+This reduces the total dimension of the parameter space to $7 + 4N$.
+\newline
+
 \section{Parameter estimation of representative example,  high-SNR} \label{sec:app_high_SNR}
-In this Section we repeat the analysis of Section \ref{sec:pe_and_ms}, but in the high-SNR limit.
+In this section we repeat the analysis of Section \ref{sec:pe_and_ms}, but in the high-SNR limit.
 We do this as a sanity check to confirm that the state-space analysis scheme with the pulsar term included returns reasonable results in the ``easy" case with a strong GW signal.
@@
 For five of the seven parameters, there is effectively zero dispersion in the one-dimensional posterior medians between noise realisations.
-For the remaining two parameters, $\iota$ and $h_0$, an appreciable is an appreciable dispersion, again due to weak-identifiability.
+For the remaining two parameters, $\iota$ and $h_0$, an appreciable dispersion is observed, again due to weak-identifiability.
 \newline
@@
 	\includegraphics[width=\textwidth, height = \textwidth ]{images/corner_highSNR} 	
-	\caption{Same as Figure \ref{fig:corner_plot_1}, but for synthetic data with $h_0 = 10^{-12}$.
+	\caption{Same as Figure \ref{fig:corner_plot_1}, but for a SMBHB with $h_0 = 10^{-12}$.
 The Kalman filter and nested sampler estimate all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ accurately.
@@
 	\includegraphics[width=\columnwidth, height = \columnwidth ]{images/corner_highSNR_chi} 	
-	\caption{Same as Figure \ref{fig:corner_plot_3}, but for synthetic data with $h_0 = 10^{-12}$.
+	\caption{Same as Figure \ref{fig:corner_plot_3}, but for a SMBHB with $h_0 = 10^{-12}$.
 Consistent unimodal posteriors are obtained across all noise realisations and all five displayed parameters.
@@
 \section{Challenges of the pulsar term}\label{sec:psr_term_challenges}
-In this section....
 
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =0.5\textwidth ]{example-image-a}
+	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
 	\caption{Logarithm of the one-dimensional likelihood $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
-All other static parameters of $\boldsymbol{\theta}$ are held constant at their true injected values.
-In the left-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}(\alpha)$, i.e.
+All other static elements of $\boldsymbol{\theta}$ are held constant at their true injected values.
+In the left panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}(\alpha)$, i.e.
 inclusive of the pulsar terms.
@@
 The inset plot covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
-In the left hand panel $\mathcal{L}(\alpha)$ is very noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
+In the left panel $\mathcal{L}(\alpha)$ is noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
 This noise hampers the ability of sampling algorithms to converge to a posterior distribution.
@@
 \end{figure*}
-
-To elucidate the complications introduced by including the pulsar term, we generate some noisy synthetic data $\boldsymbol{Y}$ using the methods outlined in K23.
-$\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
-The choice of $\boldsymbol{\theta}_{\rm psr}$ corresponds to the 47 pulsars in the 12.5 year NANOGrav data set  \citep{2020ApJ...905L..34A}.
-We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig} on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad.
-We obtain a curve $\mathcal{L}(\alpha)$, which is plotted in the left hand panel of Figure \ref{fig:likelihood_comparison} (blue curve).
+In this section we elucidate the complications introduced by including the pulsar term into the state-space analysis and justify the reparametrisation in terms of $\chi^{(n)}$ introduced in Section \ref{sec:pulsar_term2}.
+\newline
+
+We generate some noisy synthetic data $\boldsymbol{Y}$ as outlined in Section \ref{sec:synthetic_data}.
+The SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
+We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig}, on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad passed to the  Kalman filter.
+All other elements of $\boldsymbol{\theta}_{\rm gw}$ are fixed at their true injection values.
+At this point we do not use the $\chi^{(n)}$ parametrisation.
+Therefore varying $\alpha$ changes the value of $\boldsymbol{n}$ everywhere it appears in Equation \eqref{eq:g_func_trig}.
+For each value of $\alpha$, the Kalman filter returns a likelihood $\mathcal{L}(\boldsymbol{Y} | \alpha) = \mathcal{L}(\alpha)$.
+\newline
+
+The function $\mathcal{L}(\alpha)$ is displayed in the left panel of Figure \ref{fig:likelihood_comparison} (orange curve).
 The dashed grey line indicates the injected value of $\alpha$ (= 1.0 rad).
 The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
-\newline
-
-There are two important features to discuss in the left-hand panel of Figure \ref{fig:likelihood_comparison}.
+There are two important features to discuss in the left panel.
 The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
 Similar noisy curves are obtained for $\mathcal{L}(\Omega)$ and $\mathcal{L}(\delta)$.
-For all other static parameters, the $\mathcal{L}(\Theta)$ curves are smooth and globally concave.
+For all other static parameters, the $\mathcal{L}(\theta)$ curves are smooth and globally concave.
 The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.
@@
 
-Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\Theta)$ curves whereas other static parameters do not?
+Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\theta)$ curves whereas other static parameters do not?
 These parameters appear in the pulsar term as a product in the pulsar-dependent phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, where we remind the reader that $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
@@
 This creates a confusion problem when taking the cosine of  $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ since we do not know how many cycles the wave has gone through, modulo $2 \pi$.
-\textcolor{red}{TK: this explanation is rubbish!
-How can we say this better?}.
-Conversely, if we drop the pulsar term,  the $\mathcal{L}(\alpha)$ curve (as well as $\mathcal{L}(\Omega)$ , $\mathcal{L}(\delta)$   )
-This is shown by the orange curve in the right-hand panel of Figure \ref{fig:likelihood_comparison} where we plot $\mathcal{L}(\alpha)$, analogous to the left-hand panel, but now use Equation \ref{eq:g_func_trig_earth} in our Kalman filter (i.e.
-just the Earth terms) rather than the full Equation \ref{eq:g_func_trig}.
-In this case there is no confusion regarding the number of cycles the wave has gone through over a distance $d^{(n)}$ and the curve is smooth.
-
-In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e.
-the model evidence).
-In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter.
-In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference.
-Model selection and the specification of the null model are described in Section \ref{sec:model_selection}.
-A complete summary of the workflow is presented in Section \ref{sec:methodsummary}.
-The method complements traditional PTA analyses because (i) it does not assume a specific functional form (e.g.\ power law) for the power spectral density of the TOA fluctuations; and (ii) it infers ${\boldsymbol{\theta}}$ conditional on the specific, time-ordered, random realization of the noise corresponding to the observed sequence of TOAs rather than on ensemble-averaged quantities like the power spectral density.
-
-%See https://arxiv.org/pdf/2102.12478.pdf for a good NS explanation
-
-%
-%\section{Model selection}\label{sec:model_selection}
-%The evidence integral $\mathcal{Z}$ returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$. We compare competing models via a Bayes factor,
-%\begin{equation}
-%	\beta = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)} \ . \label{eq:bayes}
-%\end{equation}
-%Throughout this work we take $\mathcal{M}_1$ to be the state-space model that includes the presence of a GW.
-$\mathcal{M}_0$ is the null model, which assumes no GW exists in the data, and is equivalent to setting $g^{(n)}(t)=1$ in Equations \eqref{eq:g_func} and \eqref{eq:g_func_trig}.
-The Bayes factors we quote in this work therefore quantify whether the data supports evidence for a GW signal compared to there being no GW signal present.
-%
-%
+\newline
+
+In order to smooth the likelihood curve there are two options.
+The first is to drop the pulsar term completely, i.e.
+using $\mathcal{M}_{\rm earth}$.
+This is the approach taken in K24 and other standard PTA analysis.
+The second is to reparametrise the pulsar term such that the phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ is no longer a function of $\alpha$ (or $\Omega$ or $\delta$), but an independent parameter to be estimated.
+This is the approach presented in this paper, i.e.
+using $\mathcal{M}_{\rm psr \& earth}$.
+The right panel of Figure \ref{fig:likelihood_comparison} displays $\mathcal{L}(\alpha)$ calculated using $\mathcal{M}_{\rm earth}$ (blue curve) and $\mathcal{M}_{\rm psr \& earth}$ (green curve).
+In both of these cases there is no confusion regarding the number of cycles the wave has gone through over a distance $d^{(n)}$ and the resulting likelihood curve is smooth.
 

### 2024-01-07 manuscript/manuscript_2/paper.tex
quickl typo fixes
misc fixes
comments on abstract and intro

@@
 % Keep the title short and informative.
-	\title[State-space PTA]{State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms}
+	\title[State-space PTA]{State-space analysis of a continuous gravitational wave source with a pulsar timing array: inclusion of the pulsar terms}
 
@@
 $^{2}$OzGrav, University of Melbourne, Parkville, VIC 3010, Australia \\
-$^{3}$Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia }
+$^{3}$Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, VIC 3010, Australia }
 
@@
 % Enter the current year, for the copyright statements etc.
-\pubyear{2023}
+\pubyear{2024}
 
@@
 \begin{abstract}	
-	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.
-The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival.
-The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
-It is shown how to extend previous state-space analyses to include the pulsar term.
-Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms corrects for biases in the parameter estimates and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%.
-For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated.
-However, including the pulsar terms decreases the minimal detectable strain by $8\%$ .
+	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.
+The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source.
+The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array.
+Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms.
+The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise.
+It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$.
+For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $8\%$.
 \end{abstract}
@@
 Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}.
-A nHz GW can influence the time of arrival (TOA) at the Earth of radio pulses from pulsars.
-A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars.
-In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers.
-Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
-SMBHBs which are sufficiently close and massive are perhaps individually resolvable, though no definitive detection is yet claimed \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
-\newline
-
-The modulation in the pulsar TOAs due to the presence of a GW has two contributions: an ``Earth term" and a ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.
+They modulate sinusoidally the times of arrival (TOAs) at the Earth of radio pulses from pulsars.
+A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple to search for a coincident GW signature.
+Congruent evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
+SMBHBs which are sufficiently close and massive may be individually resolvable, though no definitive detection is yet claimed \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
+\newline
+
+The modulation in a pulsar's TOAs due to a GW has two contributions: an ``Earth term" and a ``pulsar term" which are proportional to the GW strain at the Earth and the pulsar respectively.
 Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases.
 Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-However, dropping the pulsar terms leads to biases in the inferred parameters, and reduces the detection probability \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
-\newline
-
-\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations.
-The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term.
-In this paper we undertake two key tasks.
-Firstly we extend the methods presented in K24 to include the pulsar term.
-We show how to reparametrise the state-space model in terms of an additional pulsar-specific phase term.
-We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}.
-Secondly, we quantify how the inclusion of the pulsar term affects (i) the accuracy with which the static GW parameters can be estimated, and (ii) the minimal detectable GW strain, compared to the case where the pulsar terms are omitted.
+It is acknowledged that, dropping the pulsar terms leads to biases in the inferred parameters and reduces the detection probability, but both impacts are modest \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
+\newline
+
+\cite{KimpsonPTA} (K24 hereafter) introduced a state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, which self-consistently tracks the intrinsic, achromatic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations.
+The method described in K24 follows the example of many standard PTA analyses and drops the pulsar term.
+In this paper we achieve two goals.
+First, we extend K24 to include the pulsar terms.
+We demonstrate how the static GW parameters and the GW phase at each pulsar can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}.
+Second, we quantify how including the pulsar term improves (i) the accuracy with which the static GW parameters are estimated, and (ii) the minimum detectable GW strain, compared to when the pulsar terms are omitted.
 \newline
@@
 In Section \ref{sec:2} we briefly review the state-space model of PTA data analysis introduced in K24.
-In Section \ref{sec:pulsar_term} we show how to include the pulsar term in the model via a convenient reparametrisation.
-In Section \ref{sec:rep_example1} we test the updated model, inclusive of the pulsar term, on synthetic data for both a single representative GW source and across an astrophysically relevant domain of SMBHB source parameters.
-In Section \ref{sec:earth_vs_psr} we compare the parameter estimation accuracy and the minimal detectable GW strain when including the pulsar terms to the case where the pulsar terms are omitted.
+In Section \ref{sec:pulsar_term} we show how to include the pulsar term via a convenient reparametrisation.
+In Section \ref{sec:rep_example1} we test the updated model, inclusive of the pulsar term, on synthetic data for a single representative GW source across an astrophysically relevant domain of SMBHB source parameters.
+In Section \ref{sec:earth_vs_psr} we evaluate the parameter estimation accuracy and the minimum detectable GW strain.
 Conclusions are drawn in Section \ref{sec:discussion}.
-Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$.
+Throughout the paper we adopt natural units, with $c = G = \hbar = 1$, and the metric signature $(-,+,+,+)$.
 \newline

### 2024-01-08 manuscript/manuscript_2/paper.tex
initial edits

@@
 \section{State-Space Formulation}\label{sec:2}
-In this section we briefly review the state-space formulation of PTA analysis presented in K24. \newline
-
-There are $N$ pulsars in the array.
-The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, as measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation.
-The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$.
-The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW.
-In Section \ref{sec:psr_frequency} we outline the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$.
-In Section \ref{sec:psr_measured} we outline the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$.
-In Section \ref{sec:ss_params} we summarise the parameters of the model.
-The model is formulated in terms of pulse frequency time series as a proof-of-principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
-It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, an extension we defer to a forthcoming work.
+In this section we briefly review the state-space formulation of a PTA experiment, as presented in K24.
+There are $N$ pulsars in the array, labelled $1\leq n\leq N$.
+The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation,  which describes secular braking combined with intrinsic, achromatic spin wandering (``timing noise'').
+The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation, which describes the TOA modulation induced by the GW.
+In Section \ref{sec:psr_frequency} we define and justify the phenomenological equation of motion for $f_{\rm p}^{(n)}(t)$.
+In Section \ref{sec:psr_measured} we outline the measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$.
+In Section \ref{sec:ss_params} we summarise the static parameters of the model, including the GW source parameters, which are estimated ultimately from the data by nested sampling.
+The analysis method accepts as an input a sequence of pulse frequencies, instead of a sequence of pulse TOAs, in order to validate the approach in its simplest formulation and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
+It will be necessary to modify the method to accept pulse TOAs when implementing it to analyze real data, a subtle extension which we defer to a forthcoming paper.
 
 \subsection{Spin evolution} \label{sec:psr_frequency}
-The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
+We assume that the rest frame spin frequency of the $n$-th pulsar evolves according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
 \begin{equation}
@@
 \end{equation}
-where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative where an overdot denotes a derivative with respect to $t$, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
+where $f_{\rm em}^{(n)}$ is the deterministic evolution, an overdot denotes a derivative with respect to $t$, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
 \begin{align}
 	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
+	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:xieqn}
 \end{align}
-In Equation \eqref{eq:xieqn}, $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
-The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
-PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$.
+In Equation \eqref{eq:xieqn}, $[\sigma^{(n)}]^2$ parametrizes the noise amplitude and produces characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
+The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking for the sake of definiteness, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
+PTAs are typically composed of millisecond pulsars (MSPs), for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$.
 Consequently, 	$f_{\rm em}^{(n)}(t)$ can be approximated accurately by
@@
 
-Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} are inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
-Such a phenomenological model is necessary as a predictive, first-principles theory of timing noise does not currently exist c.f.
-the multiple theorized mechanisms of timing noise (see discussion in Section 1 of K24).
-Similar approaches, analogous to Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
-However, Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} do contain certain modelling omissions and idealisations which must be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
-These include (i) the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$ due to $\xi(t)$ in Equation \eqref{eq:frequency_evolution} not being differentiable.
+Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are not unique.
+Rather, they offer one possible phenomenological description consistent with the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}.
+A phenomenological approach is obligatory, because a predictive, first-principles theory of timing noise does not currently exist, c.f.
+the multiple theorized mechanisms referenced in Section 1 of K24.
+Langevin equations like \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} have been applied successfully to analyse anomalous braking indices \citep{Vargas} and in hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
+However, they are highly idealised \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
+Idealisations include (i) the exclusion of elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the whitenes of $\xi^{(n)}(t)$ for MSPs in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$, noting that $\xi^{(n)}(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.
 
 \subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
-In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation
+In the presence of a GW, the rest-frame spin frequency of the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation,
 \begin{equation}
@@
 \end{equation}
-where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
+where $d^{(n)}$ labels the distance to the $n$-th pulsar, $f_{\rm p}$ is evaluated at the retarded time $t-d^{(n)}$, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
 \begin{align}
@@
 In Equation \eqref{eq:vareps}, $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars.
-In Equation \eqref{eq:measurement} the measurement function $g^{(n)}(t)$ is
+In Equation \eqref{eq:measurement} the measurement function $g^{(n)}(t)$ is given by \citep[e.g.][]{Maggiore}
 \begin{align}
@@
 \end{align}
-where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor, and $\Phi_0$ the phase offset of the GW with respect to some reference time.
-\newline
-
-\noindent Equation \eqref{eq:g_func_trig} contains two simplifying assumptions: 
-\begin{enumerate}[leftmargin=2em]
-	\item The GW is a monochromatic plane wave.
-	\item The pulsar locations are constant with respect to the observer.
-\end{enumerate}
-Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Sesana2010},
+where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the GW amplitude tensor, and $\Phi_0$ is the phase offset of the GW with respect to some reference time.
+\newline
+
+\noindent Equation \eqref{eq:g_func_trig} assumes that (i) the GW is a monochromatic plane wave; and (ii) $\boldsymbol{q}^{(n)}$ is constant.
+Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves by an amount \citep[e.g.][]{Sesana2010}
 \begin{equation}
@@
 A source can be considered monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$.
-Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \gtrsim 20$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$.
+Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \lesssim 20$ for $\Delta f_{\rm gw} \lesssim 1/T_{\rm obs}$.
 The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
@@
 
-Regarding point (ii) whilst in practice the pulsar locations vary with respect to the Earth, pulsar TOAs are defined relative to the Solar System barycentre (SSB).
+Regarding point (ii) a pulsar's sky position varies due to its own orbital motion (if it is located in a binary) as well as the rotation and revolution of the Earth.
+However, pulsar TOAs are defined relative to the Solar System barycentre, after correcting for the pulsar's binary motion (if any).
 The barycentering correction is typically applied when generating TOAs, e.g.
-with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series.
-Some pulsars, including some PTA pulsars, do have non-negligible proper motions of order $10^2$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.
-
-\subsection{Static parameters of the model}\label{sec:ss_params}
-For the model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured}, the set of static parameters which are specific to the pulsars in the array are
+with {\sc tempo2} \citep{tempo2,edwardstempo} and related timing software, and is inherited by $f_{\rm m}^{(n)}(t)$.
+Some PTA pulsars do have non-negligible proper motions of order $10^2$ km s$^{-1}$ after barycentering \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this introductory paper.
+
+\subsection{Static parameters}\label{sec:ss_params}
+The model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured}, compromises $5N$ static parameters, that are specific to the pulsars in the array,
 \begin{equation}
@@
 \end{equation}
-The parameters which are specific to the GW source are 
+It also compromises seven parameters, that are specific to the GW source, viz. 
 \begin{equation}
@@
 \end{equation}
-where $h_0$ is the characteristic wave strain, $\iota$ the orbital inclination, $\psi$ the polarisation angle of the wave, $\delta$ the declination of the source and $\alpha$ its right ascension.
-These parameters enter the model through Equation \eqref{eq:g_func_trig} where $H_{ij} = H_{ij}(h_0, \iota, \psi, \delta, \alpha)$ and $\boldsymbol{n}=\boldsymbol{n}(\delta,\alpha)$.
-The complete set of $7 + 5N$ static parameters is denoted $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$.
-While we have no prior information about the values of the elements of $\boldsymbol{\theta}_{\rm gw}$, the elements of $\boldsymbol{\theta}_{\rm psr}$
-do have constraints from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
-
-\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
-K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection.
-That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
+where $h_0$ is the characteristic wave strain, $\iota$ is the orbital inclination, $\psi$ is the polarisation angle, $\delta$ is the declination and $\alpha$ is the right ascension.
+These parameters enter the model through Equation \eqref{eq:g_func_trig}, with $H_{ij} = H_{ij}(h_0, \iota, \psi, \delta, \alpha)$ and $\boldsymbol{n}=\boldsymbol{n}(\delta,\alpha)$.
+The complete set of $7 + 5N$ static parameters is denoted by $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$.
+While we have no prior information about $\boldsymbol{\theta}_{\rm gw}$, there are constraints on $\boldsymbol{\theta}_{\rm psr}$ from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ typically \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
+
+\section{Inference with the pulsar terms}\label{sec:pulsar_term}
+K24 developed a likelihood-based Bayesian framework to infer the static parameters in Section \ref{sec:ss_params} and select between models with and without a GW.
+Given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
 \begin{equation}
@@
 \end{equation}
-where $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}, $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$, and $\mathcal{Z}$ is the marginalised likelihood, or evidence
+where $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated with a Kalman filter \citep{Kalman1}, as discussed in Appendix \ref{sec:kalman}, $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$, and $\mathcal{Z}$ is the marginalised likelihood or evidence,
 \begin{equation}
@@
 \end{equation}
-For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
-Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e.
-$\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$.
-Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$.
-Within the context of PTA searches for GWs, we refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, Appendix \ref{sec:nested_sampling} for a review of nested sampling and Appendix \ref{sec:workflow} for an overview of the combined Kalman filter - nested sampling workflow.
-\newline
-
-The inference model used by the Kalman filter in K24 drops the pulsar term, consistent with other PTA analyses.
-In this section we show how to remove this limitation and include the pulsar term in the state-space inference model passed to the Kalman filter.
-In Section \ref{sec:earth_term} we review how the pulsar term is dropped from typical PTA analyses and the modification to Equation \eqref{eq:g_func_trig} used in K24.
+In the paper, the measurements are ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
+Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is a function of $\boldsymbol{\theta}$.
+Appendix \ref{sec:kalman} explains how to compute ${\mathcal{L}}({\boldsymbol{Y}}|{\boldsymbol{\theta}})$ from $f_{\rm m}^{(n)}(t)$ by discretizing the dynamical equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and the measurement equations \eqref{eq:measurement}--\eqref{eq:g_func_trig} and solving them recursively using a Kalman filter.
+Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$ given ${\mathcal{L}}({\boldsymbol{Y}}|{\boldsymbol{\theta}})$.
+Appendix \ref{sec:nested_sampling} reviews nested sampling briefly as applied in this paper.
+Appendix \ref{sec:workflow} summarizes the workflow of the combined Kalman filter and nested sampler.
+\newline
+
+The inference procedure in K24 drops the pulsar term on the last line of Equation \eqref{eq:g_func_trig}, in keeping with other PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+In this section we show how to remove this limitation, following previous authors \cite{Zhupulsarterms,Chen2022}.
+In Section \ref{sec:earth_term} we review briefly how the pulsar term is dropped from typical PTA analyses, as well as K24.
 In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar term for likelihood-based inference methods such as nested sampling.
@@
 \subsection{Earth term}\label{sec:earth_term}
-The influence of the GW on the pulsar radio pulses is described by Equation \eqref{eq:g_func_trig}.
-The GW signal is composed of an ``Earth term", corresponding to the $\cos(-\Omega t + \Phi_0)$ component, and a ``pulsar term", corresponding to the $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \}$ component.
-The Earth term describes the GW phase at the observer on Earth, while the pulsar term describes the GW phase at the individual pulsar.
-The Earth term depends only on the GW source parameters and so is shared across all pulsars.
-In contrast the pulsar term is a function of $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars.
-\newline
-
-Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases which cannot be coherently summed and are instead generally considered as a source of self-noise.
-The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties much greater than the target GW wavelength.
-Consequently the pulsar term is often dropped in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space; the $N$ values of $d^{(n)}$ are not inferred, making the parameter estimation computationally cheaper.
-Within the state-space model described in Section \ref{sec:2}, dropping the pulsar terms equates to using a modified measurement equation, c.f.
-Equations \eqref{eq:measurement} and \eqref{eq:g_func_trig},
-\begin{equation}
-	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) + \varepsilon^{(n)}(t) \ , 
+The GW modulates radio pulses according to Equation \eqref{eq:g_func_trig}.
+The modulation is composed of an Earth term, proportional to $\cos(-\Omega t + \Phi_0)$, and a pulsar term, proportional to $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \}$.
+The Earth term describes the GW phase at the observer on Earth, while the pulsar term describes the GW phase at the $n$-th pulsar.
+The Earth term depends only on the GW source parameters and is common across all pulsars.
+In contrast the pulsar term is a function of $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and varies between pulsars.
+\newline
+
+The phases of the pulsar terms are related unpredictably, which is why some published PTA analyses approximate the pulsar terms collectively as a source of self-noise.
+The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties greater than the typical GW wavelength.
+Consequently the pulsar term is often dropped in standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space; the $N$ values of $d^{(n)}$ are not inferred.
+Within the state-space model described in Section \ref{sec:2}, dropping the pulsar terms equates to using a modified measurement equation,
+\begin{equation}
+	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}_{\rm Earth}(t) + \varepsilon^{(n)}(t) \ , 
 	\label{eq:measuremen_earth}
@@
 \end{equation}
-However, dropping the pulsar term leads to biases in the parameter estimation, in particular the source localisation, and a reduced detection probability \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
-It is therefore important to remove this limitation and include the pulsar term in PTA data analysis pipelines.
-
-\subsection{Reparametrisation of the pulsar term} \label{sec:pulsar_term2}
-We define a new parameter 
-\begin{equation}
-	\chi^{(n)} \equiv \left[ \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right ] \mod 2 \pi \ ,
+However, dropping the pulsar term leads to biases in the parameter estimation, in $\alpha$ and $\delta$, and reduces the detection probability by $\sim 5 \%$ \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
+
+\subsection{Reparametrisation of the pulsar terms} \label{sec:pulsar_term2}
+In this paper, we generalize the Kalman filter analysis in K24 to include the pulsar terms. To do so, we define a new parameter 
+\begin{equation}
+	\chi^{(n)} = \left\{ \Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \mod 2 \pi \ , \label{eq:chi_param}
 \end{equation}
@@
 	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
-	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
-	&- \cos \left \{-\Omega t +\Phi_0 + \chi^{(n)} \right \} \Big ] \ .
+	& \times \Big \{\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
+	&- \cos \left [-\Omega t +\Phi_0 + \chi^{(n)} \right ] \Big \} \ .
 	\label{eq:g_func_trig_chi}
 \end{align}
-The reparametrisation in terms of $\chi^{(n)}$ treats the pulsar-dependent phase correction $\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ in the argument of the cosine of the pulsar term as an explicit parameter to be inferred for each pulsar.
-The advantage of this approach, compared to trying to infer $d^{(n)}$ directly, is that the resulting likelihood surface across the parameter space is smoother, with smaller derivatives.
-This enables sampling algorithms to efficiently find the likelihood maxima and estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$.
+The reparametrisation in terms of $\chi^{(n)}$ treats the pulsar-dependent phase correction $\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}$ in the argument of the cosine of the pulsar term as a composite parameter to be inferred for each pulsar.
+
+In principle, it is possible to disentangle $\Omega$, $\boldsymbol{n}$, and $d^{(n)}$
+
+and infer them individually, if $N$ is large enough; for example, $\boldsymbol{n}$ appears independently in the phase of the pulsar term and in the denominator of the first line of Equation \eqref{eq:g_func_trig}.
+In practice, however, the likelihood surface is highly corrugated along the $d^{(n)}$-axis, and the prior on $d^{(n)}$ is broad compared to the wavelength $2\pi/\Omega$, so $d^{(n)}$ and hence $\Omega$ are not identifiable for reasonable $N$.
+By replacing $d^{(n)}$ with $\chi^{(n)}$, we obtain a smooth likelihood function, whose maximum is located efficiently and accurately by the nested sampler.
 This point is discussed in more detail in Appendix \ref{sec:psr_term_challenges}.
-The new parametrisation also does not increase the dimension of the parameter space, because we have traded $d^{(n)}$ for $\chi^{(n)}$.
+The new parametrisation does not increase the dimension of the parameter space, because we trade $d^{(n)}$ for $\chi^{(n)}$.
+Once $\Omega$, ${\boldsymbol{n}}$, and $\chi^{(n)}$ are inferred, it is possible to solve for $d^{(n)}$, albeit not uniquely; $d^{(n)}$ can be inferred up to an integer multiple of the Doppler-shifted wavelength, due to the ${\rm mod} \, 2 \pi$ operation in Equation \eqref{eq:chi_param}.
+However, this ambiguity is unavoidable, whether we replace $d^{(n)}$ with $\chi^{(n)}$ or not, and occurs in every PTA analysis.
 The static parameters specific to the pulsars in the array (c.f.
@@
 \begin{equation}
-	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1) \right\}_{1\leq n \leq N} \ ,  \label{eq:psrparamsnew}
+	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)} \right\}_{1\leq n \leq N} \ ,  \label{eq:psrparamsnew}
 \end{equation}
 whilst $\boldsymbol{\theta}_{\rm gw}$  remains unchanged.
-We exclude $\chi^{(n)}$ from membership of $\boldsymbol{\theta'}_{\rm psr}$ as the other elements of $\boldsymbol{\theta'}_{\rm psr}$ depend solely on the properties of the pulsar, with no dependence on the properties of the GW source.
+We define the complete set of $7 +5N$ static parameters to be inferred as $\boldsymbol{\theta'} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta'}_{\rm psr}$.
+%We exclude $\chi^{(n)}$ from membership of $\boldsymbol{\theta'}_{\rm psr}$ as the other elements of $\boldsymbol{\theta'}_{\rm psr}$ depend solely on the properties of the pulsar, with no dependence on the properties of the GW source.
 In contrast $\chi^{(n)}$ is a cross term that depends on both the individual pulsar and the GW.
-We define the complete set of $7 +4N + N$ static parameters to be inferred as $\boldsymbol{\theta'} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta'}_{\rm psr} \cup \chi^{(n)}$ .
-\newline
+%
 
 \section{Validation with synthetic data}\label{sec:rep_example1}
-In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}.
+In the rest of this paper, we test the performance of the PTA analysis scheme in K24, once it is generalized to include the pulsar terms in the inference model as described in Section 3.
+We refer the reader to the appendices for detailed instructions about how to implement the Kalman filter (Appendix \ref{sec:kalman}) and nested sampler (Appendix \ref{sec:nested_sampling}) and integrate them in a unified workflow (Appendix \ref{sec:workflow}).
+In this section, by way of preparation, we explain how to generate the synthetic data employed in the tests, namely noisy frequency time series $f_{\rm m}^{(n)}(t)$ for $1 \leq n \leq N$.
+Tests are performed for multiple random realizations of $\xi^{(n)}(t)$ in order to quantify the irreducible ``cosmic'' variance in the inference output (e.g.\ estimates of ${\boldsymbol{\theta}}_{\rm gw}$).
+Every real PTA analysis witnesses a unique realization of $\xi^{(n)}(t)$ --- the actual, astronomical one --- but there is no way to determine where this realization lies within the admissible statistical ensemble.
+\newline
+%[end of first para of sec 4]
+
+%In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}.
 As outlined in Section \ref{sec:pulsar_term}, the analysis scheme consists of a Kalman filter to track the state evolution of the intrinsic pulsar spin frequency $f_{\rm p}^{(n)}(t)$ which provides a likelihood for nested sampling.
@@
 \newline
-
-In Section \ref{sec:synthetic_data} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme.
-In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e.
-the SMBHB system.
-The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
-In Section \ref{sec:parameter_space} we extend the tests of Section  \ref{sec:pe_and_ms}
-to cover a broader parameter space and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.
-
-\subsection{Generating synthetic data}\label{sec:synthetic_data}
+%
+
 \begin{table*}
@@
 			\toprule
-			&Parameter & Injected value & Units & Prior  \\
+			Set&Parameter & Injected value & Units & Prior  \\
 			\hline
@@
 			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
-			\hline
-			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}_{\rm ATNF} \right)  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
+			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}_{\rm ATNF} \right]  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
 			\bottomrule
 		\end{tabular}
-		\caption{Summary of the injected static parameters used for generating synthetic data for validation of the analysis scheme when including the pulsar-term in Section \ref{sec:pe_and_ms}.
-The prior used for Bayesian inference (Section \ref{sec:pe_and_ms}) on each parameter is also displayed (rightmost column).
-The top, middle and bottom sections of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$, $\boldsymbol{\theta}_{\rm psr}$ and $\chi^{(n)}$ respectively.
-The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synthetic_data}.
-The subscript ``SC'' on the pulsar noise amplitude $\sigma^{(n)}$ indicates that the injected value has been calculated using Equation \eqref{eq:sigmap_f} the empirical timing noise model for MSPs from \protect \cite{Shannon2010}.
-The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
-We do not try to infer $\gamma^{(n)}$ as $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$ and so no prior is set.
-The choice of priors on $\boldsymbol{\theta}_{\rm psr}$ is discussed in Appendix \ref{sec:set_priors}.
+		\caption{Injected static parameters used to generate synthetic data to validate the analysis scheme including the pulsar-terms.
+The prior used for Bayesian inference is also displayed (rightmost column).
+The top and bottom sections of the table contain $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta'}_{\rm psr}$ respectively.
+The subscript ``ATNF'' denotes values obtained from the ATNF pulsar catalogue as described in Section \ref{sec:rep_example1}.
+The subscript ``SC'' on $\sigma^{(n)}$ indicates that the injected value is calculated from Equation \eqref{eq:sigmap_f} and the empirical timing noise model for MSPs in \protect \cite{Shannon2010}.
+The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the uncertainties in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
+We do not infer infer $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$, so no prior is set.
+The priors on $\boldsymbol{\theta}_{\rm psr}$ are justified in Appendix \ref{sec:set_priors}.
 		}
@@
 	\end{table*}
-In order to generate synthetic frequency time series $ \boldsymbol{Y}(t) = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
-This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
-The measured $f_{\rm m}^{(n)}(t)$ are then produced via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
-Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and \eqref{eq:measurement}--\eqref{eq:g_func_trig} depend on the choice of static parameters $\boldsymbol{\theta'}$, as well as the choice of pulsar directions ${\boldsymbol{q}}^{(n)}$ and detector measurement noise covariance $\sigma_{\rm m}$ (Equation \eqref{eq:vareps}).
-We now describe how $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ are chosen.
-This is equivalent to specifying the configuration of a synthetic PTA.
-In Section  \ref{sec:pe_and_ms} and Section \ref{sec:parameter_space} we describe how the remaining elements of $\boldsymbol{\theta}'$, namely $\boldsymbol{\theta}_{\rm gw}$, are chosen.
-This is equivalent to specifying the synthetic SMBHB source.
-\newline
-
-In this paper we use the same values for $\boldsymbol{\theta}_{\rm psr}$ as in K24, i.e.
-we adopt the $N=47$ millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
-We assume all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period.
-For each pulsar fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$ are obtained via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
-We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues.
-No direct measurements exist for $\gamma^{(n)}$, though the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}.
-In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
-Similarly, no direct measurements exist for $\sigma^{(n)}$.
-We relate the process noise in the frequency $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week such that
+In order to synthesize $\boldsymbol{Y}(t) = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, we integrate Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$, which we convert to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
+The numerical solutions depend on how we choose $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ or, equivalently, how we specify the configuration of a synthetic PTA.
+In Section  \ref{sec:pe_and_ms} we describe how we choose the remaining elements of $\boldsymbol{\theta}'$, namely $\boldsymbol{\theta}_{\rm gw}$.
+This latter step is equivalent to specifying the synthetic SMBHB source and differs from one test to the next according to the goal of the test.
+\newline
+
+In this paper we adopt for consistency the same $\boldsymbol{\theta}_{\rm psr}$ values as in K24, i.e.
+the $N=47$ MSPs in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
+We assume all pulsars are observed with cadence $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period.
+Fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$ are read from the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.
+No direct measurements exist for $\gamma^{(n)}$.
+The mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}; in this paper, for the sake of simplicity, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$.
+No direct measurements exist for $\sigma^{(n)}$ either.
+We relate $\sigma^{(n)}$ to the root mean square TOA noise $\sigma^{(n)}_{\rm TOA}$ accumulated over an interval of length $T_{\rm cad}$ by
 \begin{eqnarray}
@@
 \end{eqnarray}
-where, following K24, we use the empirical timing noise model for MSPs from \cite{Shannon2010ApJ...725.1607S} in order to set physically reasonable values on $\sigma_{\rm TOA}^{(n)}$.
-The median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134.
-\newline
-
-The measurement noise covariance, as defined in Equation \eqref{eq:vareps}, can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
+As in K24, the empirical timing noise model for MSPs from \cite{Shannon2010ApJ...725.1607S} implies $\text{median} [\sigma^{(n)}] = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134.
+Likewise, $\sigma_{\rm m}$ can be related to the TOA uncertainty, $\sigma_{\rm TOA}$, by
 \begin{equation}
@@
 \end{equation}
-For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s,  Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
-The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz.
-Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
-When analysing real data this assumption can easily be relaxed.
-Whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is constructed by a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar.
-\newline
-
-\subsection{Parameter estimation} \label{sec:pe_and_ms} 
-In this section, we apply the Kalman filter in conjunction with nested sampling to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ for synthetic data $\boldsymbol{Y}$, and compare it to the known, injected values.
-The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
-That is, given the injected static parameters $\boldsymbol{\theta'}$, multiple realisations of $\boldsymbol{Y}$ are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta'}$.
-The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations.
-Regarding point (iii), quantifying the dispersion is important since it is a practical measure of the accuracy of the parameter estimation scheme when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
-The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative.
-Different values of the static GW source parameters are explored in Section \ref{sec:parameter_space}.
-The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}.
+For an MSP with $f_{\rm p}^{(n)} \sim 100$ Hz, $T_{\rm cad} = 1 \, {\rm week}$, and $\sigma_{\rm TOA} \sim 1 \mu$s,  Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz.
+Here, for the sake of definiteness, we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
+When analysing real data this assumption is easily relaxed.
+Although $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar.
+\newline
+
+\section{Parameter estimation} \label{sec:pe_and_ms} 
+In this section, we apply the Kalman filter and nested sampler to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ and compare it to the known, injected values.
+The procedure is undertaken for multiple realisations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$, and we estimate $\boldsymbol{\theta'}$ independently for each realisation.
+The aims are (i) to demonstrate that the analysis scheme works, i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations; (ii) to give a preliminary sense of its accuracy, and (iii) to quantify the natural random dispersion in the one-dimensional posterior medians.
+Quantifying the dispersion is important since it is a practical measure of the scheme's accuracy when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
+The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:rep_example1}.
 All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
 The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}.
-\newline
+
+In Section \ref{sec:rep_smbh_source} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme.
+In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e.
+the SMBHB system.
+The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
+In Section \ref{sec:parameter_space} we extend the tests of Section  \ref{sec:pe_and_ms}
+to cover a broader parameter space and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.
+
+\subsection{Representative SMBHB source}\label{sec:rep_smbh_source}
 

### 2024-01-08 manuscript/manuscript_2/paper.tex
all minor comments addressed

@@
 
-Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for ten arbitrary noise realisations in the form of a traditional corner plot.
-The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters.
-Each coloured curve is the result for different noise realisation.
+Figure \ref{fig:corner_plot_1} displays the posterior distribution of $\boldsymbol{\theta}_{\rm gw}$ for ten noise realisations in the form of a traditional corner plot for the representative SMBHB source parametrized as per the top portion of Table \ref{tab:parameters_and_priors}.
+The histograms are one-dimensional posteriors, marginalized over the six other parameters.
+Each coloured curve corresponds to a different noise realisation.
 The solid orange line marks the known injected value.
@@
 There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
-The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations (i.e.
-86$\%$).
-The injected value is contained within the 90\% credible interval for all posterior realisations of $\Omega, \Phi_0, \psi$, and  $\iota$.
-The injected value is contained within the 5\% and 95\% quantiles for all of the one-dimensional marginalized posteriors.
-There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations.
-The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 52\% for $\Phi_0$.
+The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations.
+The injected value falls within the 90\% credible interval for $\Omega, \Phi_0, \psi$, and  $\iota$ 40 out of 40 times.
+There is appreciable natural dispersion in the one-dimensional posterior medians between realisations.
+The minimum and maximum coefficients of variation are 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively.
 The remaining parameters have typical coefficients $\lessapprox 10 \%$.
 For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
-Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR}, Figure \ref{fig:corner_high_snr_appendix}.
-At higher strains the dispersion between noise realisations is reduced and the one dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injection value.
-\newline
-
-In addition to $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach.
-Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the ten noise realisations, analogously to Figure \ref{fig:corner_plot_1}.
-We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big.
+Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}).
+The dispersion between noise realisations decreases with increasing $h_0$ is reduced and the one-dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injected value.
+\newline
+
+\subsection{GW phase at each pulsar} \label{sec:chi_estim}
+
+ It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach.
+
+ approach, as explained in Section \ref{sec:pulsar_term2}. In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$.
+
+ Figure \ref{fig:corner_plot_3} displays the results as a corner plot for the representative subset $\chi^{(1)}, \dots, \chi^{(5)}$ over the ten noise realisations, in the same style as Figure \ref{fig:corner_plot_1}.
+We do not display the corner plot for $\chi^{(6)}$, $\dots$, $\chi^{(47)}$ because it is too big.
 With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values.
 There is no evidence for correlations between parameter pairs.
-The majority of the elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations; for the 47 elements of $\chi^{n}$ (i.e.
-not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations (i.e.
-$\sim$ 81 $\%$).
-There are some elements of $\chi^{(n)}$ for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}).
-This corresponds to the phase-correction term for PSR J0340+4130.
-Out of all the pulsars in the array, this pulsar lies in the closest direction to our synthetic GW source, such that $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$.
-As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the GW is minimal ($g^{(2)}(t) \to 1$) The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with a larger strain $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
-\newline
-
-Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$; the injected values are recovered unambiguously across the nine noise realisations.
-For the sake of brevity we do not display the calculated posterior distributions here because the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses.
+Most phases are recovered unambiguously across all noise realisations; for the 47 $\chi^{(n)}$ parameters (i.e.
+not just the five plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations.
+
+Rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}), corresponding to PSR J0340+4130.
+Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$,  $\cos \chi^{(2)} \approx 0.07$, and $g^{(2)}(t) \approx XYZ$ The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR); for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered accurately; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
+\newline
+
+\subsection{Timing parameters} \label{sec:timing_parameters}
+Similar results are obtained for the 4$N$ parameters in $\boldsymbol{\theta}_{\rm psr}'$ besides $\chi^{(n)}$.
+The injected values are recovered unambiguously across the 10 noise realisations.
+For the sake of brevity we do not display the calculated posterior distributions, because inferring $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses.
+In short, the estimates of $f_{\rm em}(t_1)$ and $\dot{f}_{\rm em}(t_1)$ are guided into narrow ranges by the narrow priors.
+[then add 1-2 sentences about sigma including giving some numbers on percentage accuracy (min and max)] We remind the reader that $\gamma^{(n)}$ is not estimated in this paper; typically it satisfies $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ [refs from before], so its influence is muted.
 \newline
@@
 
-\subsection{Extension to broader parameter space} \label{sec:parameter_space}
-\textcolor{red}{TK: Figures 3 and 4 may be updated.
-Some additional simulations are currently running to double check some things.
-I expect the current figures to be representative of any updated results.} \newline
+\subsection{Exploring the SMBHB parameter domain} \label{sec:parameter_space} 
 \begin{figure}
@@
 
-Section \ref{sec:pe_and_ms} focuses on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
-In this section we test the method for a range of systems, varying the SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ through astrophysically relevant ranges.
-The aim of this section is to confirm that the analysis scheme can be applied across an astrophysically relevant parameter space and that the arbitrary choice of $\boldsymbol{\theta}_{\rm gw}$ in Section \ref{sec:pe_and_ms} is not particularly advantageous by accident.
-\newline
-
-We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
-We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
+Section \ref{sec:rep_smbh_source} focuses on a single representative source, summarised in Table \ref{tab:parameters_and_priors}.
+In this section we test the method for a range of sources, varying $\boldsymbol{\theta}_{\rm gw}$.
+The aim is to verify that the analysis scheme works across an astrophysically relevant domain and that the arbitrary choice of $\boldsymbol{\theta}_{\rm gw}$ in Table \ref{tab:parameters_and_priors} is not advantageous by accident.
+\newline
+
+We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the 200 injections.
 For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
-The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:synthetic_data} and Table  \ref{tab:parameters_and_priors}.
+The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1} and Table  \ref{tab:parameters_and_priors}.
 \newline
@@
 It is prohibitive to display corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$.
-Therefore, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+Therefore, we summarise the results with the aid of a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
 A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself.
@@
 
-The results are displayed in Figure \ref{fig:parameter_space}.
+The PP results are displayed in Figure \ref{fig:parameter_space}.
 The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections.
-The curves for the five parameters generally exhibit the discussed linear behaviour.
-Some parameters are consistently well estimated across the parameter space; the blue curve for $\Omega$ is consistently contained within the 2$\sigma$ shaded region.
+The curves for the five parameters, colour-coded in the legend, are approximately linear.
+Some parameters are consistently well estimated across the parameter domain, e.g.
+the blue curve for $\Omega$ lies everywhere within the 2$\sigma$ shaded region.
 Conversely some parameters, e.g.
-$\psi$, stray outside the $1\sigma$ bounded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and less injections contained within higher-value credible intervals.
+$\psi$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals.
 This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky.
-Manually inspecting the individual corner plots for each of the 200 injections, it is evident that for a minority of SMBHB sources which lie in an unfavourable direction ($\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} = -1$) the method either does not converge to a consistent unimodal one-dimensional posterior, or else a posterior is converged to but the estimate is suboptimal.
-For these sources their location with respect to the PTA makes it difficult to infer some elements of $\chi^{(n)}$ accurately, which impacts the inference of the other elements of $\boldsymbol{\theta}_{\rm gw}$.
-All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$, $n_{\rm live}$ etc.
-Some sources in unfavourable sky locations  may only be resolvable with different observation settings; we defer an in-depth exploration of sensitivity across the parameter space to a future work.
-In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but now we assume that the injected values of $\chi^{(n)}$ are known exactly, i.e.
-setting a delta-function prior on $\chi^{(n)}$ during the nested sampling inference.
-Now the excursions out of the shaded region are almost completely reduced.
-This indicates the importance of accurate estimates of $\chi^{(n)}$ and the sensitivity of a specific PTA configuration to different parts of the sky.
+\newline
+
+Manually inspecting the individual corner plots for each of the 200 injections, we see that the nested sampler does not return a unimodal posterior in the rare event that the source is located unfavourably, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} \approx -1$, such that one cannot infer $\chi^{(n)}$ accurately (see Section \ref{sec:chi_estim}).
+All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$ and $n_{\rm live}$.
+In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but assuming that the injected values of $\chi^{(n)}$ are known exactly for the sake of testing, i.e.
+setting a delta-function prior on $\chi^{(n)}$.
+The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$.
+
+%that for a minority of SMBHB sources which lie in an unfavourable direction ($\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} = -1$) the method either does not converge to a consistent unimodal one-dimensional posterior, or else a posterior is converged to but the estimate is suboptimal.
+
+%For these sources their location with respect to the PTA makes it difficult to infer some elements of $\chi^{(n)}$ accurately, which impacts the inference of the other elements of $\boldsymbol{\theta}_{\rm gw}$.
+%
+
+%Some sources in unfavourable sky locations  may only be resolvable with different observation settings; we defer an in-depth exploration of sensitivity across the parameter space to a future work.
+
+%
+%and the sensitivity of a specific PTA configuration to different parts of the sky.
 In spite of the small excursions of some parameters in Figure \ref{fig:parameter_space}, it is reassuring that in this low-SNR regime all parameters generally follow the desired linear trend, and that key parameters such as $\Omega$ lie close to the unit diagonal.
 
-\section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}
+\section{Bias when neglecting the pulsar terms}\label{sec:earth_vs_psr}
 In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model.

### 2024-01-09 manuscript/manuscript_2/paper.tex
EoD

@@
 It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$.
-For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $8\%$.
+For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
 \end{abstract}
@@
 The reparametrisation in terms of $\chi^{(n)}$ treats the pulsar-dependent phase correction $\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}$ in the argument of the cosine of the pulsar term as a composite parameter to be inferred for each pulsar.
-
-In principle, it is possible to disentangle $\Omega$, $\boldsymbol{n}$, and $d^{(n)}$
-
-and infer them individually, if $N$ is large enough; for example, $\boldsymbol{n}$ appears independently in the phase of the pulsar term and in the denominator of the first line of Equation \eqref{eq:g_func_trig}.
+In principle, it is possible to disentangle $\Omega$, $\boldsymbol{n}$, and $d^{(n)}$ and infer them individually, if $N$ is large enough; for example, $\boldsymbol{n}$ appears independently in the phase of the pulsar term and in the denominator of the first line of Equation \eqref{eq:g_func_trig}.
 In practice, however, the likelihood surface is highly corrugated along the $d^{(n)}$-axis, and the prior on $d^{(n)}$ is broad compared to the wavelength $2\pi/\Omega$, so $d^{(n)}$ and hence $\Omega$ are not identifiable for reasonable $N$.
@@
 The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}.
-
-In Section \ref{sec:rep_smbh_source} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme.
-In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e.
+\newline
+
+In Section \ref{sec:rep_smbh_source} we apply the analysis scheme to synthetic data and estimate $\boldsymbol{\theta}_{\rm gw}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e.
 the SMBHB system.
 The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
-In Section \ref{sec:parameter_space} we extend the tests of Section  \ref{sec:pe_and_ms}
-to cover a broader parameter space and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.
+In Section \ref{sec:chi_estim} we inspect the estimates of  $\chi^{(n)}$.
+In Section \ref{sec:timing_parameters} we briefly discuss the estimates of the remaining 4N parameters in $\boldsymbol{\theta}_{\rm psr}$.
+In Section \ref{sec:parameter_space} we extend the tests across a broader parameter domain and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.
 
@@
 
- It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach.
-
- approach, as explained in Section \ref{sec:pulsar_term2}. In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$.
-
+ It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach, as explained in Section \ref{sec:pulsar_term2}.
+In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$.
 Figure \ref{fig:corner_plot_3} displays the results as a corner plot for the representative subset $\chi^{(1)}, \dots, \chi^{(5)}$ over the ten noise realisations, in the same style as Figure \ref{fig:corner_plot_1}.
@@
 not just the five plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations.
+\newline
 
 Rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}), corresponding to PSR J0340+4130.
-Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$,  $\cos \chi^{(2)} \approx 0.07$, and $g^{(2)}(t) \approx XYZ$ The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR); for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered accurately; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
+Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$ and $\cos \chi^{(2)} \approx 0.07$.
+The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR); for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered unambiguously; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
 \newline
@@
 In short, the estimates of $f_{\rm em}(t_1)$ and $\dot{f}_{\rm em}(t_1)$ are guided into narrow ranges by the narrow priors.
-[then add 1-2 sentences about sigma including giving some numbers on percentage accuracy (min and max)] We remind the reader that $\gamma^{(n)}$ is not estimated in this paper; typically it satisfies $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ [refs from before], so its influence is muted.
+The one-dimensional posteriors inferred for $\sigma^{(n)}$ are generally broader due to the broader prior, but contain the injection within the  90\% credible interval in a majority of cases.
+We remind the reader that $\gamma^{(n)}$ is not estimated in this paper; typically it satisfies $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}, so its influence is muted.
 \newline
@@
 \subsection{Exploring the SMBHB parameter domain} \label{sec:parameter_space} 
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
-	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
+%	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
 PP plot).
@@
 The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
-	\label{fig:parameter_space}
-\end{figure}
+%	\label{fig:parameter_space}
+%\end{figure}
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi}
+%	\caption{Same as Figure \ref{fig:parameter_space} but }
+%	\label{fig:parameter_space2}
+%\end{figure}
+
 \begin{figure}
 	\centering
-	\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi}
-	\caption{Same as Figure \ref{fig:parameter_space} but now assuming that $\chi^{(n)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior for the nested sampling inference).
+	\subfloat[a][]{\includegraphics[width=\columnwidth]{images/pp_plot_canonical500} \label{fig:parameter_space}} \\
+	\subfloat[b][]{\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi} \label{fig:parameter_space2}}
+	\caption{(a) Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
+PP plot).
+The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}.
+Each coloured curve corresponds to a different parameter (see legend).
+The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
+The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals.
+For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope.
+$\Omega$ and $\alpha$ are generally well-estimated (i.e.
+the curves lie close to the unit diagonal).
+The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.
+(b) Same as (a) but now assuming that $\chi^{(n)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior for the nested sampling inference).
 The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, are contained within the shaded region.
-}
-	\label{fig:parameter_space2}
+} \label{fig:AB}
 \end{figure}
@@
 Conversely some parameters, e.g.
-$\psi$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals.
+$\Phi_0$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals.
 This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky.
+More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases.
+For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases.
 \newline
@@
 The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$.
+For the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}.
+For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 89\% of cases, an improvement over the results of Figure Figure \ref{fig:parameter_space}.
+\newline
 
@@
 Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$.
-Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}.
+Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:rep_example1}.
 The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}.
@@
 
-In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model.
+In this section we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model.
 For $\mathcal{M}_{\rm earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm earth})$.
-The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta}_{\rm earth} | \boldsymbol{Y})$.
+The likelihood returned by the Kalman filter is then used by the nested sampler in order to estimate $p(\boldsymbol{\theta}_{\rm earth} | \boldsymbol{Y})$.
 For $\mathcal{M}_{\rm psr \& earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$.
-The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$.
+The likelihood returned by the Kalman filter is then used by the nested sampler in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$.
 The settings of the nested sampler (for example the number of live points, see Appendix \ref{sec:nested_sampling}) are identical for both models.
@@
 the exact system described in Table \ref{tab:parameters_and_priors}.
-The second is a ``high-SNR" system which has the same static injection parameters, but with an increased $h_0 = 10^{-12}$.
+The second is a ``high-SNR" system which has the same static injection parameters, but with an increased strain amplitude $h_0 = 10^{-12}$.
 The ``high-SNR" system is considered in order to evaluate the parameter estimation accuracy, and in particular any potential bias, without confusion from the measurement noise.
@@
 The vertical orange lines indicate the true injected values.
-The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
+The contours in the two-dimensional histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
 The supertitles of the one-dimensional histograms record the posterior medians and the 0.16 and 0.84 quantiles of the green curves, i.e.
@@
 \subsection{Detectability vs $h_0$}\label{sec:psr_v_earth_bayes}
-		\textcolor{red}{TK: Figure 7 may be updated with the results of ongoing simulations.
-However, this won't affect any general conclusions we draw; just a smoother graph and a slight change in minimal detectable strain.} \newline
 		\begin{figure}
@@
 The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
-The minimum detectable strain for which $\beta < 10$ equals $3.19 \times 10^{-15}$ for $\mathcal{M}_{\rm earth}$ and $2.93 \times 10^{-15}$ for $\mathcal{M}_{\rm psr \& earth}$.
-\textcolor{red}{TK: we should also mention scatter at lower h, but this should be eliminated when new simulations complete.}}
+The minimum detectable strain for which $\beta < 10$ equals $3.19 \times 10^{-15}$ for $\mathcal{M}_{\rm earth}$ and $2.78 \times 10^{-15}$ for $\mathcal{M}_{\rm psr \& earth}$.
+The axes are plotted on logarithmic scales.}
 			\label{fig:bayes1}
@@
 		\end{equation}
-In this paper we consider $i \in \left \{ \rm psr \& earth, earth \right \}$ \newline
+In this paper we consider $i \in \left \{ \rm psr \& earth, earth \right \}$. \newline
 
@@
 That is, for a given $h_0$, $\mathcal{M}_{\rm psr \& earth}$ provides greater evidence for a GW signal in the noisy data than $\mathcal{M}_{\rm psr \& earth}$.
-The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.19 \times 10^{-15}$ for  $\mathcal{M}_{\rm earth}$ and for $h_0 \gtrsim 2.93 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& earth}$, a relative improvement in the minimal detectable strain of $8\%$.
+The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.19 \times 10^{-15}$ for  $\mathcal{M}_{\rm earth}$ and for $h_0 \gtrsim 2.78 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& earth}$, a relative improvement in the minimal detectable strain of $13\%$.
 The minimum detectable strain and the relative improvement through using $\mathcal{M}_{\rm psr \& earth}$ is of course particular to the system in Table \ref{tab:parameters_and_priors} and the realisation of $\boldsymbol{Y}$.
@@
 
-The intrinsic rotational state evolution of the pulsar, described by a mean-reverting Ornstein-Uhlenbeck process (Equations \eqref{eq:xieqn}--\eqref{eq:spinevol}) is tracked using a Kalman filter.
+The intrinsic rotational state evolution of the pulsar, described by a mean-reverting Ornstein-Uhlenbeck process, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol}, is tracked using a Kalman filter.
 An updated state-space model for the influence of the GW is introduced that reparameterises the pulsar term as a separate pulsar-specific static parameter $\chi^{(n)}$ to be inferred for each pulsar.
@@
 In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement).
-The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k = \boldsymbol{Y}(t_k)$, which are related via a linear transformation to a set of unobservable discrete system states $\boldsymbol{X}_k= \boldsymbol{Y}(t_k)$.
+The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k = \boldsymbol{Y}(t_k)$, which are related via a linear transformation to a set of unobservable discrete system states $\boldsymbol{X}_k= \boldsymbol{X}(t_k)$.
 Each discrete timestep is indexed by $ 1 \leq k  \leq K$.
@@
 \begin{equation}
-	C^{(n)} =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
+	C^{(n)} =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}^{(n)} _{\rm em}(t_1)\ .
 \end{equation}
@@
 \section{Choice of priors}\label{sec:set_priors}
-To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} and the workflow of Appendix \ref{sec:workflow} it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters.
+To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} and the workflow of Appendix \ref{sec:workflow} it is necessary to specify a Bayesian prior $\pi(\boldsymbol{\theta})$ on the static parameters.
 In this section we outline how the priors are chosen.
@@
 We do this as a sanity check to confirm that the state-space analysis scheme with the pulsar term included returns reasonable results in the ``easy" case with a strong GW signal.
-The high-SNR limit allows us to inspect the results without without confusion from a measurement noise which is large compared to the GW signal.
-\newline
-
-We apply the Kalman filter - nested sampling scheme as in Section \ref{sec:pe_and_ms}.
-All injected static parameters $\boldsymbol{\theta}'$, and the associated priors are as specified in Table \ref{tab:parameters_and_priors}, which the exception of $h_0$ which now takes a value $h_0 = 10^{-12}$.
+The high-SNR limit allows us to inspect the results without confusion from a measurement noise which is large compared to the GW signal.
+\newline
+
+We apply the parameter estimation framework as in Section \ref{sec:pe_and_ms}.
+All injected static parameters $\boldsymbol{\theta}'$, and the associated priors are as specified in Table \ref{tab:parameters_and_priors}, with the exception of $h_0$ which now takes a value $h_0 = 10^{-12}$.
 The procedure is undertaken for ten realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
@@
 	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
-	\caption{Logarithm of the one-dimensional likelihood $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
+	\caption{Logarithm of the likelihood curve $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
 All other static elements of $\boldsymbol{\theta}$ are held constant at their true injected values.
@@
 inclusive of the pulsar terms.
-In the right-hand panel the Kalman filter uses Equation \ref{eq:g_func_trig_earth}, i.e.
+In the right-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig_earth}, i.e.
 just the Earth terms.
@@
 There are two important features to discuss in the left panel.
-The first is that on scales of order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
+The first is that on scales of the order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
 Similar noisy curves are obtained for $\mathcal{L}(\Omega)$ and $\mathcal{L}(\delta)$.

### 2024-01-09 py_src/plotting.py

@@ -178,7 +178,7 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
 
 
 
-    
+    print(ranges)
     #Now plot it using corner.corner
     fs = 20
     newfig = corner.corner(y_post, 
@@ -217,6 +217,16 @@ def plot_custom_corner(path,variables_to_plot,labels,injection_parameters,ranges
     if title is not None:
         newfig.suptitle(title, fontsize=20)  
     #plt.show()
+
+    
+
+
+
+
+
+
+
+
 
     
 

### 2024-01-09 manuscript/manuscript_2/paper.tex
complete draft

@@
 It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$.
-For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
+For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain decreases by $13\%$.
 \end{abstract}
@@
 \section{Validation with synthetic data}\label{sec:rep_example1}
-In the rest of this paper, we test the performance of the PTA analysis scheme in K24, once it is generalized to include the pulsar terms in the inference model as described in Section 3.
+In the rest of this paper, we test the performance of the PTA analysis scheme in K24, once it is generalized to include the pulsar terms in the inference model as described in Section \ref{sec:pulsar_term}.
 We refer the reader to the appendices for detailed instructions about how to implement the Kalman filter (Appendix \ref{sec:kalman}) and nested sampler (Appendix \ref{sec:nested_sampling}) and integrate them in a unified workflow (Appendix \ref{sec:workflow}).
@@
 			\hline
-			\vspace{1mm}& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
-      \multirow{2}{2mm}{$\boldsymbol{\theta'}_{\rm psr}$} & $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
-		     & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$ \\
+			& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
+       & $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
+		     $\boldsymbol{\theta'}_{\rm psr}$ & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$ \\
 			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
@@
 Quantifying the dispersion is important since it is a practical measure of the scheme's accuracy when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
-The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1}.
+The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.
+The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1}.
 All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
@@
 For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
-Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}).
-The dispersion between noise realisations decreases with increasing $h_0$ is reduced and the one-dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injected value.
+Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}); the dispersion between noise realisations decreases with increasing $h_0$ and the one-dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injected value.
 \newline
@@
 This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky.
-More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases.
+More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases.
 For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases.
@@
 The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$.
-For the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}.
-For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 89\% of cases, an improvement over the results of Figure Figure \ref{fig:parameter_space}.
+For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}.
+For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 89\% of cases, an improvement over the results of Figure \ref{fig:parameter_space}.
 \newline
@@
 \section{Bias when neglecting the pulsar terms}\label{sec:earth_vs_psr}
-In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model.
-In this section, critically, we compare the effect of including the pulsar-term in the model (i.e.
-using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e.
-Equation \eqref{eq:g_func_trig_earth}).
+In Section \ref{sec:pe_and_ms} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model.
+In this section, critically, we compare the effect of including the pulsar-term in the model, i.e.
+using Equation \eqref{eq:g_func_trig_chi}, as opposed to the omitting the pulsar-term, i.e.
+using Equation \eqref{eq:g_func_trig_earth}.
 We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
@@
 
-In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm earth}$ returned by the respective models.
+In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$ and $\boldsymbol{\theta}_{\rm earth}$ returned by the respective models.
 In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable GW strain for the two models by calculating the model evidence and comparing it to the evidence for a null model that does not contain a GW.
@@
 The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_compare_high} for the high-SNR system and in Figure \ref{fig:corner_plot_compare_low} for the low-SNR system.
-The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curves and contours now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
 The blue curves are the results derived using $\mathcal{M}_{\rm earth}$.
@@
 Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
-For example, for $\iota$, whilst the injected value is contained within the 90\% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians.
+For example, considering $\iota$, whilst the injected value is contained within the 90\% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians.
 The inclusion of the pulsar terms corrects for this bias; the green one-dimensional posteriors exhibit no bias and are generally symmetric about the injected value.
@@
 \end{equation}
-where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, psr} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
-The dependence of $p_k(\theta)$ on $M$ is implied.
+where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript ${\rm M} \in \{ \text{earth, psr \& earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+The dependence of $p_k(\theta)$ on ${\rm M}$ is implied.
 The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}.
-The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is also presented.
-For every parameter, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms, i.e.
-$\Delta_{\rm earth}(\theta) > \Delta_{\rm psr}$.
-In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm psr}(\Omega) = 6.7 \times 10^{-3}$ rad.
-The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm psr}(\iota)  = 0.24$ rad.
+The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr \& earth}(\theta)$ is also presented.
+For every parameter, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr \& earth}(\theta)$ is positive, indicating that the parameter estimates are more accurate when including the pulsar terms, i.e.
+$\Delta_{\rm earth}(\theta) > \Delta_{\rm psr \& earth}$.
+In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm psr \& earth}(\Omega) = 6.7 \times 10^{-3}$ rad.
+The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm psr \& earth}(\iota)  = 0.24$ rad.
 As a percentage change, i.e.
-$1 - \Delta_{\rm earth}(\theta)/\Delta_{\rm psr}(\theta)$, the improvement from using $\mathcal{M}_{\rm psr \& earth}$ over $\mathcal{M}_{\rm earth}$ ranges from 61\% for $\Omega$ to 99\% for $\psi$ and $\alpha$.
+$1 - \Delta_{\rm earth}(\theta)/\Delta_{\rm psr \& earth}(\theta)$, the improvement from using $\mathcal{M}_{\rm psr \& earth}$ over $\mathcal{M}_{\rm earth}$ ranges from 61\% for $\Omega$ to 99\% for $\psi$ and $\alpha$.
 These improvements are consistent across noise realisations.
@@
 We see that the error is larger for every element in $\boldsymbol{\theta}_{\rm gw}$ than in the high SNR case, as expected.
-The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is positive for all parameters except $\psi$ and $h_0$.
+The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr \& earth}(\theta)$ is positive for all parameters except $\psi$ and $h_0$.
 However it is hard to draw strong conclusions; for different realisations of the noise different parameters exhibit either improvement or deterioration in $\Delta_{\rm M}(\theta)$.
@@
 			\toprule
-			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm psr}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm psr}(\theta)$   \\
+			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm psr \& earth}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm psr \& earth}(\theta)$   \\
 			\hline
@@
 			\hline
-			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& 1.09 \\
+			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.69 &1.60& 1.09 \\
 			& $\Phi_0$ &0.335 &0.229 &0.106 \\
@@
 		\end{tabular}
-		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=psr, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=psr \& earth, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
 The injection values are as summarised in Table \ref{tab:parameters_and_priors}.
@@
 		\end{equation}
-In this paper we consider $i \in \left \{ \rm psr \& earth, earth \right \}$. \newline
+In this paper we consider $i \in \left \{ \rm earth,psr \& earth \right \}$. \newline
 
@@
 
-We see in Figure \ref{fig:bayes1} an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$ for both $\mathcal{M}_{\rm earth}$ and $\mathcal{M}_{\rm psr \& earth}$.
+We see in Figure \ref{fig:bayes1} an approximate quadratic relationship $\beta \propto h_0^2$ for both $\mathcal{M}_{\rm earth}$ and $\mathcal{M}_{\rm psr \& earth}$.
 Moreover, $\beta_{\rm psr \& earth} > \beta_{\rm earth}$ for all $h_0$.
 That is, for a given $h_0$, $\mathcal{M}_{\rm psr \& earth}$ provides greater evidence for a GW signal in the noisy data than $\mathcal{M}_{\rm psr \& earth}$.
-The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.19 \times 10^{-15}$ for  $\mathcal{M}_{\rm earth}$ and for $h_0 \gtrsim 2.78 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& earth}$, a relative improvement in the minimal detectable strain of $13\%$.
+The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.19 \times 10^{-15}$ for  $\mathcal{M}_{\rm earth}$ and for $h_0 \gtrsim 2.78 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& earth}$, a relative improvement in the minimum detectable strain of $13\%$.
 The minimum detectable strain and the relative improvement through using $\mathcal{M}_{\rm psr \& earth}$ is of course particular to the system in Table \ref{tab:parameters_and_priors} and the realisation of $\boldsymbol{Y}$.
@@
 In this paper we demonstrate how to extend state-space methods for PTA data analysis to include the pulsar term.
-We compare the results when the pulsar terms are includes to those when they are omitted.
+By testing the extended methods on astrophysical representative synthetic data we compare the results when the pulsar terms are includes to those when they are omitted.
 \newline
@@
 We find that the inclusion of the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) by 61\% for $\Omega$, up to 99\% for $\psi$ and $\alpha$ for a representative SMBHB system with a high SNR ($h_0 = 10^{-12}$).
-For low SNR ($h_0 = 5 \times 10^{-15}$) there is no significant improvement because the data is noise dominated.
-The inclusion of pulsar terms lowers the floor of the minimal detectable strain by $8\%$ \newline
+For low SNR ($h_0 = 5 \times 10^{-15}$) there is no significant improvement because the data is noise dominated, however the inclusion of pulsar terms lowers the floor of the minimal detectable strain by $13\%$ \newline
 
@@
 However, issues related to identifiability \citep{e5be7c83a0d24500826f6e1b414d1733}, the form of the measurement equation Equation \eqref{eq:g_func_trig}, and the linear structure the Kalman filter must be considered carefully and are postponed to a future paper.
-Regarding the last point, the Kalman filter can be applied to non-linear problems if needed using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
+The Kalman filter can be applied to non-linear problems if needed using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
 \newline
@@
 
-Figure \ref{fig:corner_high_snr_appendix_chi} displays results a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the ten noise realisations.
+Figure \ref{fig:corner_high_snr_appendix_chi} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $\{\chi^{(1)} \dots \chi^{(5)} \}$ over the ten noise realisations.
 The figure is exactly analogous to Figure \ref{fig:corner_plot_3}, in the high-SNR limit.
@@
 \end{figure*}
-In this section we elucidate the complications introduced by including the pulsar term into the state-space analysis and justify the reparametrisation in terms of $\chi^{(n)}$ introduced in Section \ref{sec:pulsar_term2}.
-\newline
-
-We generate some noisy synthetic data $\boldsymbol{Y}$ as outlined in Section \ref{sec:synthetic_data}.
+In this section we elucidate the complications introduced by including the pulsar term in the state-space analysis and justify the reparametrisation in terms of $\chi^{(n)}$ introduced in Section \ref{sec:pulsar_term2}.
+\newline
+
+We generate some noisy synthetic data $\boldsymbol{Y}$ as outlined in Section \ref{sec:rep_example1}.
 The SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
-We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig}, on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad passed to the  Kalman filter.
-All other elements of $\boldsymbol{\theta}_{\rm gw}$ are fixed at their true injection values.
+We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig}, on $\boldsymbol{Y}$.
+The Kalman filter is fed different values $ 0.0 \le \alpha \le \pi$ rad.
+All other components of $\boldsymbol{\theta}_{\rm gw}$ are fixed at their true injection values.
 At this point we do not use the $\chi^{(n)}$ parametrisation.
@@
 Moreover, $\mathcal{L}(\alpha)$ is only a one-dimensional function.
-The challenge for sampling algorithms is further exacerbated when we consider the three-dimensional parameter space in $\alpha - \delta - \Omega$, or the full $7 + 5N$ dimensions of Section \ref{sec:ss_params}.
-\newline
-
-Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\theta)$ curves whereas other static parameters do not?
+The challenge for sampling algorithms is further exacerbated when we consider the three-dimensional parameter domain e.g.
+$\alpha - \delta - \Omega$, or the full $7 + 5N$ dimensions of Section \ref{sec:ss_params}.
+\newline
+
+Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}\boldsymbol{\theta})$ curves whereas other static parameters do not?
 These parameters appear in the pulsar term as a product in the pulsar-dependent phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, where we remind the reader that $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.

### 2024-01-23 manuscript/manuscript_2/paper.tex
majority of Am comments

--- 2024-01-09 manuscript/manuscript_2/paper.tex
@@
 	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.
-The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source.
+The data analysis procedure can be formulated within a state-space framework, wherein achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source.
 The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array.
@@
 The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise.
-It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$.
-For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
+It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source)
+
+	 which also arise in traditional matched-filter analyses that exclude the pulsar terms.
+
+ Including the pulsar terms also decreases the root-mean-square relative error in the estimated one-dimensional posteriors \textcolor{red}{by up to 99\%}, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$.
+For a quieter source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
 \end{abstract}
@@
 They modulate sinusoidally the times of arrival (TOAs) at the Earth of radio pulses from pulsars.
-A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple to search for a coincident GW signature.
+A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars to search for a coincident GW signature.
 Congruent evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}.
-SMBHBs which are sufficiently close and massive may be individually resolvable, though no definitive detection is yet claimed \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
-\newline
-
-The modulation in a pulsar's TOAs due to a GW has two contributions: an ``Earth term" and a ``pulsar term" which are proportional to the GW strain at the Earth and the pulsar respectively.
+SMBHBs that are sufficiently close and massive may be resolvable individually but no definitive detection has been claimed so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}.
+\newline
+
+The modulation in a pulsar's TOAs due to a GW has two contributions: an ``Earth term" and ``pulsar terms" which are proportional to the GW strain at the Earth and every pulsar respectively.
 Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases.
 Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-It is acknowledged that, dropping the pulsar terms leads to biases in the inferred parameters and reduces the detection probability, but both impacts are modest \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
+It is acknowledged in standard analyses that dropping the pulsar terms leads to biases in the inferred parameters and reduces the detection probability, but both impacts are modest \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
 \newline
@@
 \cite{KimpsonPTA} (K24 hereafter) introduced a state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, which self-consistently tracks the intrinsic, achromatic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations.
-The method described in K24 follows the example of many standard PTA analyses and drops the pulsar term.
+The method described in K24 complements standard PTA analyses by tracking the specific, measured, time-ordered realization of the TOA fluctuations instead of fitting for their ensemble-averaged statistics (such as their power spectral density).
+The method described in K24 also follows the example of many standard PTA analyses and drops the pulsar terms.
 In this paper we achieve two goals.
 First, we extend K24 to include the pulsar terms.
-We demonstrate how the static GW parameters and the GW phase at each pulsar can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}.
-Second, we quantify how including the pulsar term improves (i) the accuracy with which the static GW parameters are estimated, and (ii) the minimum detectable GW strain, compared to when the pulsar terms are omitted.
+We demonstrate how the static GW parameters and the GW phase at each pulsar can be estimated successfully by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}, repeating the successful demonstration in K24 but with the pulsar terms now included.
+Second, we quantify how including the pulsar terms improves (i) the accuracy with which the static GW parameters are estimated, and (ii) the minimum detectable GW strain, compared to when the pulsar terms are omitted.
+We emphasize that the biases incurred by omitting the pulsar terms have been studied thoroughly in the context of standard PTA analyses \citep{Zhupulsarterms,Chen2022}; they are not new effects discovered here for the first time.
+Rather, the goal of this paper is to study them again in the context of the promising state-space formulation of the PTA analysis problem introduced by K24.
+The state-space formulation complements standard analysis techniques, e.g.\ based on matched filters \citep{PhysRevD.79.084030,PhysRevD.91.044048}.
+It does not supplant them and is likely to be most informative when run in tandem.
 \newline
@@
 The paper is organised as follows.
-In Section \ref{sec:2} we briefly review the state-space model of PTA data analysis introduced in K24.
-In Section \ref{sec:pulsar_term} we show how to include the pulsar term via a convenient reparametrisation.
-In Section \ref{sec:rep_example1} we test the updated model, inclusive of the pulsar term, on synthetic data for a single representative GW source across an astrophysically relevant domain of SMBHB source parameters.
-In Section \ref{sec:earth_vs_psr} we evaluate the parameter estimation accuracy and the minimum detectable GW strain.
+In Section \ref{sec:2} we briefly review the state-space formulation of PTA data analysis introduced by K24.
+In Section \ref{sec:pulsar_term} we show how to include the pulsar terms via a convenient reparametrisation.
+In Sections \ref{sec:rep_example1} and \ref{sec:pe_and_ms} we explain how to test the updated model, inclusive of the pulsar terms, by employing synthetic data for a single representative GW source across an astrophysically relevant domain of SMBHB source parameters.
+In Section \ref{sec:earth_vs_psr} we quantify the parameter estimation accuracy and the minimum detectable GW strain, comparing the performance when the pulsar terms are included and excluded..
 Conclusions are drawn in Section \ref{sec:discussion}.
@@
 In Section \ref{sec:ss_params} we summarise the static parameters of the model, including the GW source parameters, which are estimated ultimately from the data by nested sampling.
-The analysis method accepts as an input a sequence of pulse frequencies, instead of a sequence of pulse TOAs, in order to validate the approach in its simplest formulation and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
-It will be necessary to modify the method to accept pulse TOAs when implementing it to analyze real data, a subtle extension which we defer to a forthcoming paper.
+The analysis method accepts as an input a sequence of pulse frequencies, instead of a sequence of pulse TOAs, in order to validate the approach in its simplest incarnation and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}.
+It will be necessary to modify the method to accept pulse TOAs when analyzing real data, a subtle extension which we defer to a forthcoming paper.
 
@@
 \end{equation} 
-where $t_1$ labels the time of the first TOA. \newline
+where $t_1$ labels the first TOA. \newline
 
@@
 Langevin equations like \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} have been applied successfully to analyse anomalous braking indices \citep{Vargas} and in hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}.
-However, they are highly idealised \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}.
-Idealisations include (i) the exclusion of elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the whitenes of $\xi^{(n)}(t)$ for MSPs in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$, noting that $\xi^{(n)}(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.
+However, they are highly idealised \citep{Meyers2021,Myers2021MNRAS.502.3113M,2023MNRAS.520.2813A,Vargas}.
+Idealisations include (i) the exclusion of physics, which is likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the whitenes of $\xi^{(n)}(t)$ for MSPs in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$, noting that $\xi^{(n)}(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.
 
@@
 \end{equation}
-where $d^{(n)}$ labels the distance to the $n$-th pulsar, $f_{\rm p}$ is evaluated at the retarded time $t-d^{(n)}$, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
+where $d^{(n)}$ labels the distance to the $n$-th pulsar, $f_{\rm p}^{(n)}$ is evaluated at the retarded time $t-d^{(n)}$, and $\varepsilon^{(n)}(t)$ is a Gaussian measurement noise which satisfies 
 \begin{align}
@@
 \end{align}
-In Equation \eqref{eq:vareps}, $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars.
+In Equation \eqref{eq:vareps}, $\sigma_{\rm m}^2$ is the variance of the measurement noise at the telescope and is shared between all pulsars.
 In Equation \eqref{eq:measurement} the measurement function $g^{(n)}(t)$ is given by \citep[e.g.][]{Maggiore}
@@
 \noindent Equation \eqref{eq:g_func_trig} assumes that (i) the GW is a monochromatic plane wave; and (ii) $\boldsymbol{q}^{(n)}$ is constant.
-Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves by an amount \citep[e.g.][]{Sesana2010}
+Regarding point (i), studies of SMBHB inspirals in the PTA context show that the GW frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves by an amount \citep[e.g.][]{Sesana2010}
 \begin{equation}
@@
 \subsection{Static parameters}\label{sec:ss_params}
-The model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured}, compromises $5N$ static parameters, that are specific to the pulsars in the array,
+The model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured} comprises $5N$ static parameters, that are specific to the pulsars in the array, viz.
 \begin{equation}
@@
 \end{equation}
-It also compromises seven parameters, that are specific to the GW source, viz. 
+It also comprises seven parameters, that are specific to the GW source, viz. 
 \begin{equation}
@@
 \end{equation}
-where $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated with a Kalman filter \citep{Kalman1}, as discussed in Appendix \ref{sec:kalman}, $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$, and $\mathcal{Z}$ is the marginalised likelihood or evidence,
+where $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated with a Kalman filter \citep{Kalman1}, as discussed in Appendix \ref{sec:kalman}, $\pi(\boldsymbol{\theta})$ is the prior distribution of $\boldsymbol{\theta}$, and $\mathcal{Z}$ is the marginalised likelihood or evidence,
 \begin{equation}
@@
 \end{equation}
-In the paper, the measurements are ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$.
+In the paper, the measurements are ${\boldsymbol{Y}} = \{ f^{(n)}_{\rm m}(t_1), \dots, f^{(n)}_{\rm m}(T_{\rm obs})  \}_{1 \leq n \leq N}$.
 Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is a function of $\boldsymbol{\theta}$.
@@
 Appendix \ref{sec:workflow} summarizes the workflow of the combined Kalman filter and nested sampler.
+Appendices \ref{sec:kalman}--\ref{sec:workflow} are designed to equip the interested reader with the requisite tools to reproduce the key results in Sections \ref{sec:pe_and_ms} and \ref{sec:earth_vs_psr}.
 \newline
@@
 The inference procedure in K24 drops the pulsar term on the last line of Equation \eqref{eq:g_func_trig}, in keeping with other PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-In this section we show how to remove this limitation, following previous authors \cite{Zhupulsarterms,Chen2022}.
-In Section \ref{sec:earth_term} we review briefly how the pulsar term is dropped from typical PTA analyses, as well as K24.
-In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar term for likelihood-based inference methods such as nested sampling.
+In this section we show how to remove this limitation, following previous authors \citep{Zhupulsarterms,Chen2022}.
+In Section \ref{sec:earth_term} we review briefly how the pulsar terms are dropped from typical PTA analyses as well as K24.
+In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar terms for likelihood-based inference methods such as nested sampling.
 The new parametrisation is validated with synthetic data in Section \ref{sec:rep_example1}.
@@
 \end{equation}
-However, dropping the pulsar term leads to biases in the parameter estimation, in $\alpha$ and $\delta$, and reduces the detection probability by $\sim 5 \%$ \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
+However, dropping the pulsar term leads to well-known parameter estimation biases, especially in $\alpha$ and $\delta$, and reduces the detection probability by $\sim 5 \%$ \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
 
@@
 In principle, it is possible to disentangle $\Omega$, $\boldsymbol{n}$, and $d^{(n)}$ and infer them individually, if $N$ is large enough; for example, $\boldsymbol{n}$ appears independently in the phase of the pulsar term and in the denominator of the first line of Equation \eqref{eq:g_func_trig}.
-In practice, however, the likelihood surface is highly corrugated along the $d^{(n)}$-axis, and the prior on $d^{(n)}$ is broad compared to the wavelength $2\pi/\Omega$, so $d^{(n)}$ and hence $\Omega$ are not identifiable for reasonable $N$.
+In practice, however, the likelihood surface is highly corrugated along the $d^{(n)}$-axis, and the prior on $d^{(n)}$ is broad compared to the wavelength $2\pi/\Omega$, so $d^{(n)}$ and hence $\Omega$ are not identifiable for reasonable $N \lesssim 10^2$.
 By replacing $d^{(n)}$ with $\chi^{(n)}$, we obtain a smooth likelihood function, whose maximum is located efficiently and accurately by the nested sampler.
-This point is discussed in more detail in Appendix \ref{sec:psr_term_challenges}.
+This point is discussed in more detail in Appendix \ref{sec:psr_term_challenges}.\newline
+
 The new parametrisation does not increase the dimension of the parameter space, because we trade $d^{(n)}$ for $\chi^{(n)}$.
@@
 We define the complete set of $7 +5N$ static parameters to be inferred as $\boldsymbol{\theta'} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta'}_{\rm psr}$.
-%We exclude $\chi^{(n)}$ from membership of $\boldsymbol{\theta'}_{\rm psr}$ as the other elements of $\boldsymbol{\theta'}_{\rm psr}$ depend solely on the properties of the pulsar, with no dependence on the properties of the GW source.
-In contrast $\chi^{(n)}$ is a cross term that depends on both the individual pulsar and the GW.
-%
 
@@
 \newline
-%[end of first para of sec 4]
-
-%In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}.
-As outlined in Section \ref{sec:pulsar_term}, the analysis scheme consists of a Kalman filter to track the state evolution of the intrinsic pulsar spin frequency $f_{\rm p}^{(n)}(t)$ which provides a likelihood for nested sampling.
-Nested sampling then returns estimates of $\boldsymbol{\theta'}$ and $\mathcal{Z}$.
-We refer the reader to Appendices \ref{sec:kalman} -- \ref{sec:workflow} and K24 for details of the Kalman filter - nested sampling scheme.
-We apply the analysis scheme to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source.
-The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper.
-The goal of this section is to validate the state-space analysis scheme when the pulsar terms are included in the inference model.
-\newline
-%
 
@@
 		\end{tabular}
-		\caption{Injected static parameters used to generate synthetic data to validate the analysis scheme including the pulsar-terms.
+		\caption{Injected static parameters used to generate synthetic data to validate the analysis scheme including the pulsar terms in Equation \eqref{eq:g_func_trig_chi}.
 The prior used for Bayesian inference is also displayed (rightmost column).
@@
 The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the uncertainties in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.
-We do not infer infer $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$, so no prior is set.
-The priors on $\boldsymbol{\theta}_{\rm psr}$ are justified in Appendix \ref{sec:set_priors}.
+We do not infer $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ for simplicity, so no prior is set.
+The priors on $\boldsymbol{\theta'}_{\rm psr}$ are justified in Appendix \ref{sec:set_priors}.
 		}
@@
 	\end{table*}
-In order to synthesize $\boldsymbol{Y}(t) = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, we integrate Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
+In order to synthesize $\boldsymbol{Y} = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, we integrate Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}.
 This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$, which we convert to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}.
 The numerical solutions depend on how we choose $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ or, equivalently, how we specify the configuration of a synthetic PTA.
-In Section  \ref{sec:pe_and_ms} we describe how we choose the remaining elements of $\boldsymbol{\theta}'$, namely $\boldsymbol{\theta}_{\rm gw}$.
+In Section  \ref{sec:pe_and_ms} we describe how we choose the remaining elements of $\boldsymbol{\theta}$, namely $\boldsymbol{\theta}_{\rm gw}$.
 This latter step is equivalent to specifying the synthetic SMBHB source and differs from one test to the next according to the goal of the test.
@@
 In this paper we adopt for consistency the same $\boldsymbol{\theta}_{\rm psr}$ values as in K24, i.e.
-the $N=47$ MSPs in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}.
+the $N=47$ MSPs in the 12.5-year NANOGrav dataset \citep{2020ApJ...905L..34A}.
 We assume all pulsars are observed with cadence $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period.
@@
 \begin{eqnarray}
-	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
+	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)}(t_1) {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
 \end{eqnarray}
-As in K24, the empirical timing noise model for MSPs from \cite{Shannon2010ApJ...725.1607S} implies $\text{median} [\sigma^{(n)}] = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134.
+As in K24, the empirical timing noise model for MSPs from \cite{Shannon2010ApJ...725.1607S}, applied to the 12.5-year NANOGrav dataset, implies $\text{median} [\sigma^{(n)}] = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134.
+\newline
+
 Likewise, $\sigma_{\rm m}$ can be related to the TOA uncertainty, $\sigma_{\rm TOA}$, by 
 \begin{equation}
-	\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
-\end{equation}
-For an MSP with $f_{\rm p}^{(n)} \sim 100$ Hz, $T_{\rm cad} = 1 \, {\rm week}$, and $\sigma_{\rm TOA} \sim 1 \mu$s,  Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
+	\sigma_{\rm m} \approx f_{\rm p}^{(n)}(t_1) \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
+\end{equation}
+For an MSP with $f_{\rm p}^{(n)} \sim 0.1$ kHz, $T_{\rm cad} = 1 \, {\rm week}$, and $\sigma_{\rm TOA} \sim 1 \mu$s,  Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
 The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz.
@@
 The procedure is undertaken for multiple realisations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$, and we estimate $\boldsymbol{\theta'}$ independently for each realisation.
-The aims are (i) to demonstrate that the analysis scheme works, i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations; (ii) to give a preliminary sense of its accuracy, and (iii) to quantify the natural random dispersion in the one-dimensional posterior medians.
-Quantifying the dispersion is important since it is a practical measure of the scheme's accuracy when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown.
+The aims are (i) to demonstrate that the analysis scheme works, i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations; (ii) to give a preliminary sense of its accuracy; and (iii) to quantify the natural random dispersion in the one-dimensional posterior medians.
+Quantifying the dispersion is important since it is a practical measure of the scheme's accuracy when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown (i.e.
+cosmic variance).
 The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1}.
@@
 the SMBHB system.
-The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates.
-In Section \ref{sec:chi_estim} we inspect the estimates of  $\chi^{(n)}$.
-In Section \ref{sec:timing_parameters} we briefly discuss the estimates of the remaining 4N parameters in $\boldsymbol{\theta}_{\rm psr}$.
+The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible cosmic variance in the parameter estimates.
+In Section \ref{sec:chi_estim} we inspect and verify the estimates of  $\chi^{(n)}$.
+In Section \ref{sec:timing_parameters} we briefly discuss the estimates of the remaining $4N$ parameters in $\boldsymbol{\theta'}_{\rm psr}$.
 In Section \ref{sec:parameter_space} we extend the tests across a broader parameter domain and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.
@@
 There is scant evidence of railing against the prior bounds.
-There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
-The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations.
-The injected value falls within the 90\% credible interval for $\Omega, \Phi_0, \psi$, and  $\iota$ 40 out of 40 times.
+There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.
+The injected value falls within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations.
+Indeed, the injected value falls within the 90\% credible interval for $\Omega, \Phi_0, \psi$, and  $\iota$ 40 out of 40 times.
 There is appreciable natural dispersion in the one-dimensional posterior medians between realisations.
-The minimum and maximum coefficients of variation are 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively.
-The remaining parameters have typical coefficients $\lessapprox 10 \%$.
-For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately.
-Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}).
-The dispersion between noise realisations decreases with increasing $h_0$ is reduced and the one-dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injected value.
+We quantify the degree of dispersion using the coefficient of variation,
+\begin{eqnarray}
+	CV = \frac{\mu}{\sigma}
+\end{eqnarray}
+ where $\mu$ is the mean value of the 10 posterior medians and $\sigma$ is the standard deviation.
+The minimum and maximum $CV$ is 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively.
+The remaining parameters have typical $CV$ $\lesssim 10 \%$.
+Analogous results for synthetic data with higher strain ($h_0 = 1 \times 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}) for completeness.
+The dispersion between noise realisations decreases as $h_0$ increases, and the one-dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injected value.
 \newline
@@
 \subsection{GW phase at each pulsar} \label{sec:chi_estim}
-
 It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach, as explained in Section \ref{sec:pulsar_term2}.
@@
 not just the five plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations.
-\newline
-
-Rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}), corresponding to PSR J0340+4130.
+The next paragraph explains why 383 out of 470 is fewer than 90\%.
+\newline
+
+Sometimes, albeit rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}), corresponding to PSR J0340+4130.
 Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$ and $\cos \chi^{(2)} \approx 0.07$.
-The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR); for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered unambiguously; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
+The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR).
+For example, for for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered unambiguously; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
 \newline
@@
 The injected values are recovered unambiguously across the 10 noise realisations.
-For the sake of brevity we do not display the calculated posterior distributions, because inferring $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses.
+For the sake of brevity we do not display the calculated posterior distributions, because inferring $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and most published PTA analyses.
 In short, the estimates of $f_{\rm em}(t_1)$ and $\dot{f}_{\rm em}(t_1)$ are guided into narrow ranges by the narrow priors.
@@
 	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR_n4000}
-	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for ten realisations of the noise processes, with curves coloured uniquely.
+	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for ten realisations of the noise processes, with curves coloured uniquely per realisation.
 The horizontal and vertical orange lines indicate the true injected values.
@@
 The known, injected value lies within the 90\% credible interval for 60 out of the 70 combinations of seven parameters and ten noise realizations.
-There is an appreciable dispersion among the peaks of the one-dimensional posteriors; the maximum coefficient of variation is $52 \%$ for $\Phi_0$, the minimum is $0.02 \%$ for $\Omega$.}
+There is appreciable dispersion among the peaks of the one-dimensional posteriors: the maximum coefficient of variation is $52 \%$ for $\Phi_0$, and the minimum is $0.02 \%$ for $\Omega$.}
 	\label{fig:corner_plot_1}
@@
 	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi_n4000}
-	\caption{Same as Figure \ref{fig:corner_plot_1}	but for five static parameters of $\chi^{(n)}$.
+	\caption{Same as Figure \ref{fig:corner_plot_1}	but for the static parameters of $\chi^{(1)}, \dots, \chi^{(5)}$ (the remaining pulsar phases $\chi^{(6)}, \dots, \chi^{(N)}$ are omitted for readability).
 Consistent unimodal posteriors are obtained across the ten noise realisations for four out of the five displayed parameters.
-No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar has a similar sky position to that of the synthetic GW source.
-The known, injected value lies within the 90\% credible interval for 38 out of the 50 combinations of five parameters and ten noise realizations (i.e.
-$76 \%$).
-}
+No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar's sky position is close to the synthetic GW source.
+The known, injected value lies within the 90\% credible interval for 38 out of the 50 combinations of five parameters and ten noise realizations.}
 	\label{fig:corner_plot_3}
@@
 \subsection{Exploring the SMBHB parameter domain} \label{sec:parameter_space}
-%\begin{figure}
-%	\centering
-%	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
-%	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
-PP plot).
-The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}.
-Each coloured curve corresponds to a different parameter (see legend).
-The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
-The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals.
-For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope.
-$\Omega$ and $\alpha$ are generally well-estimated (i.e.
-the curves lie close to the unit diagonal).
-The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
-%	\label{fig:parameter_space}
-%\end{figure}
-%\begin{figure}
-%	\centering
-%	\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi}
-%	\caption{Same as Figure \ref{fig:parameter_space} but }
-%	\label{fig:parameter_space2}
-%\end{figure}
 
@@
 	\subfloat[b][]{\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi} \label{fig:parameter_space2}}
-	\caption{(a) Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
+	\caption{Accuracy of SMBHB parameter estimation across an astrophysically plausible domain.
+(a) Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
 PP plot).
@@
 The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.
-(b) Same as (a) but now assuming that $\chi^{(n)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior for the nested sampling inference).
-The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, are contained within the shaded region.
+(b) Same as (a) but now assuming that $\chi^{(1)}, \dots, \chi^{(47)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior).
+The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, lie within the shaded region.
 } \label{fig:AB}
@@
 $\Phi_0$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals.
-This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky.
-More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases.
-For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases.
+This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration as a function of sky position.
+More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases.
+For the worst estimated parameter $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases.
 \newline
@@
 The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$.
-For the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}.
-For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 89\% of cases, an improvement over the results of Figure Figure \ref{fig:parameter_space}.
-\newline
-
-%that for a minority of SMBHB sources which lie in an unfavourable direction ($\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} = -1$) the method either does not converge to a consistent unimodal one-dimensional posterior, or else a posterior is converged to but the estimate is suboptimal.
-
-%For these sources their location with respect to the PTA makes it difficult to infer some elements of $\chi^{(n)}$ accurately, which impacts the inference of the other elements of $\boldsymbol{\theta}_{\rm gw}$.
-%
-
-%Some sources in unfavourable sky locations  may only be resolvable with different observation settings; we defer an in-depth exploration of sensitivity across the parameter space to a future work.
-
-%
-%and the sensitivity of a specific PTA configuration to different parts of the sky.
-In spite of the small excursions of some parameters in Figure \ref{fig:parameter_space}, it is reassuring that in this low-SNR regime all parameters generally follow the desired linear trend, and that key parameters such as $\Omega$ lie close to the unit diagonal.
+For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}.
+For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90\% credible interval in 89\% of cases, an improvement over the results of Figure \ref{fig:parameter_space}.
+\newline
 
 \section{Bias when neglecting the pulsar terms}\label{sec:earth_vs_psr}
-In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model.
-In this section, critically, we compare the effect of including the pulsar-term in the model (i.e.
-using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e.
-Equation \eqref{eq:g_func_trig_earth}).
-We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
-We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$.
-We refer to the model where the pulsar-term is omitted as
-$\mathcal{M}_{\rm earth}$.
-Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$.
-Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$.
-Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:rep_example1}.
-The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}.
-\newline
-
-In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm earth}$ returned by the respective models.
-In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable GW strain for the two models by calculating the model evidence and comparing it to the evidence for a null model that does not contain a GW.
+In Section \ref{sec:rep_example1} we validate the state-space analysis scheme when the pulsar terms are included in the inference model, i.e.
+Equation \eqref{eq:g_func_trig_chi}.
+In this section, we compare what happens when we omit the pulsar terms intentionally by using Equation \eqref{eq:g_func_trig_earth}.
+We do this for the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}.
+We refer to the model where the pulsar terms are included alongside the Earth term as $\mathcal{M}_{\rm psr \& Earth}$.
+We refer to the model where the pulsar terms are omitted as
+$\mathcal{M}_{\rm Earth}$.
+Model $\mathcal{M}_{\rm psr \& Earth}$ is parameterised by $\boldsymbol{\theta}'$.
+Model $\mathcal{M}_{\rm Earth}$ is parameterised by $\boldsymbol{\theta}_{\rm Earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$.
+Both models are applied to identical realisations of the data, generated using the procedure in Section  \ref{sec:rep_example1}.
+The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& Earth}$, as described in Appendix \ref{sec:set_priors}.
+\newline
+
+In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm Earth}$ returned by the respective models.
+In Section \ref{sec:psr_v_earth_bayes} we compare the minimum detectable GW strain for the two models by calculating the model evidence and comparing it to the evidence for a null model that does not contain a GW.
 
@@
 
-In this section we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model.
-For $\mathcal{M}_{\rm earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm earth})$.
-The likelihood returned by the Kalman filter is then used by the nested sampler in order to estimate $p(\boldsymbol{\theta}_{\rm earth} | \boldsymbol{Y})$.
-For $\mathcal{M}_{\rm psr \& earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$.
-The likelihood returned by the Kalman filter is then used by the nested sampler in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$.
-The settings of the nested sampler (for example the number of live points, see Appendix \ref{sec:nested_sampling}) are identical for both models.
+In this section we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters with and without the pulsar terms.
+For $\mathcal{M}_{\rm Earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm Earth})$ and the nested sampler to estimate $p(\boldsymbol{\theta}_{\rm Earth} | \boldsymbol{Y})$.
+For $\mathcal{M}_{\rm psr \& Earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$ and the nested sampler to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$.
+The settings of the nested sampler (for example the number of live points; see Appendix \ref{sec:nested_sampling}) are identical for both models.
 \newline
@@
 We consider two representative systems.
-The first is a ``low-SNR" system with $h_0 = 5 \times 10^{-15}$, i.e.
-the exact system described in Table \ref{tab:parameters_and_priors}.
-The second is a ``high-SNR" system which has the same static injection parameters, but with an increased strain amplitude $h_0 = 10^{-12}$.
-The ``high-SNR" system is considered in order to evaluate the parameter estimation accuracy, and in particular any potential bias, without confusion from the measurement noise.
-In order to enable a clear comparison between the posterior probability distributions calculated using the two models, in this section we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$.
+The first is a ``low-SNR'' system with $h_0 = 5 \times 10^{-15}$, i.e.
+the system described in Table \ref{tab:parameters_and_priors}.
+The second is a ``high-SNR'' system which has the same static injection as in Table \ref{tab:parameters_and_priors} except for $h_0 = 1 \times 10^{-12}$.
+The ``high-SNR'' system is considered in order to evaluate any systematic biases, as distinct from random errors caused by the measurement noise.
+In order to enable a clear comparison between the posterior probability distributions calculated using the two models, in this section we present a single noise realisation of the synthetic data $\boldsymbol{Y}$.
 The results quoted are consistent across different noise realisations.
@@
 
-The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_compare_high} for the high-SNR system and in Figure \ref{fig:corner_plot_compare_low} for the low-SNR system.
-The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
-The blue curves are the results derived using $\mathcal{M}_{\rm earth}$.
-The green curves are the results derived using $\mathcal{M}_{\rm psr \& earth}$.
-\newline
-
-For the high-SNR case, Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm earth}$ are evidently biased, as was observed in K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
+The results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_compare_high} for the high-SNR system and in Figure \ref{fig:corner_plot_compare_low} for the low-SNR system.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, except that the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
+The blue curves are the results derived using $\mathcal{M}_{\rm Earth}$.
+The green curves are the results derived using $\mathcal{M}_{\rm psr \& Earth}$.
+\newline
+
+For the high-SNR case results in Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm Earth}$ are evidently biased, as was observed by K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
 Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
-For example, for $\iota$, whilst the injected value is contained within the 90\% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians.
-The inclusion of the pulsar terms corrects for this bias; the green one-dimensional posteriors exhibit no bias and are generally symmetric about the injected value.
-Moreover, for every pair of parameters the injection values are contained within the 2-sigma equivalent contours for the $\mathcal{M}_{\rm psr \& earth}$, which is not true for $\mathcal{M}_{\rm earth}$.
-\newline
-
-We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
-\begin{equation}
-\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{ \left[p_k(\theta) - \hat{\theta}\right]^2}{\hat{\theta}}} \ , \label{eq:mean_rel_error}
-\end{equation}
-where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, psr} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
-The dependence of $p_k(\theta)$ on $M$ is implied.
-The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}.
-The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is also presented.
-For every parameter, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms, i.e.
-$\Delta_{\rm earth}(\theta) > \Delta_{\rm psr}$.
-In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm psr}(\Omega) = 6.7 \times 10^{-3}$ rad.
-The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm psr}(\iota)  = 0.24$ rad.
-As a percentage change, i.e.
-$1 - \Delta_{\rm earth}(\theta)/\Delta_{\rm psr}(\theta)$, the improvement from using $\mathcal{M}_{\rm psr \& earth}$ over $\mathcal{M}_{\rm earth}$ ranges from 61\% for $\Omega$ to 99\% for $\psi$ and $\alpha$.
-These improvements are consistent across noise realisations.
-\newline
-
-For the low-SNR case,  Figure \ref{fig:corner_plot_compare_low}, there is less of a notable improvement in the parameter estimates.
-Qualitatively, the green and blue one-dimensional posteriors are generally overlaid.
-The two-dimensional contours suggest a slight improvement through using $\mathcal{M}_{\rm psr \& earth}$, but the difference is modest.
-The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR case is also reported in Table \ref{tab:posterior_errors}.
-We see that the error is larger for every element in $\boldsymbol{\theta}_{\rm gw}$ than in the high SNR case, as expected.
-The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr}(\theta)$ is positive for all parameters except $\psi$ and $h_0$.
-However it is hard to draw strong conclusions; for different realisations of the noise different parameters exhibit either improvement or deterioration in $\Delta_{\rm M}(\theta)$.
-We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case.
-This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates.
-Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:psr_v_earth_bayes} that they are important for increasing the detection probability for a given signal strength.
-\newline
-
+For example, for $\iota$, the injected value is contained within the 90\% credible interval, but the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians.
+The inclusion of the pulsar terms corrects for this bias.
+The green one-dimensional posteriors exhibit no bias and are generally symmetric about the injected value.
+Moreover, for every pair of parameters in ${\boldsymbol{\theta}}_{\rm gw}$ the injected values are contained within the 2-sigma contours for the $\mathcal{M}_{\rm psr \& Earth}$.
+This is not true for $\mathcal{M}_{\rm Earth}$ where the injected values fall outside the 2-sigma contour for 15 out of the 21 parameter pairs.\newline
+
+We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value, viz.
+\begin{equation}
+\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{ \left[\theta_{k,\rm M} - \theta_{\rm inj}\right]^2}{\theta_{\rm inj}}} \ . \label{eq:mean_rel_error}
+\end{equation}
+In Equation \eqref{eq:mean_rel_error} $\theta_{k,\rm M}$ is the $k$-th sample of the one-dimensional posterior for parameter $\theta$ returned by nested sampling, and the subscript $\rm M \in \{ \text{Earth, psr \& Earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+The total number of samples is $K$ and $\theta_{\rm inj}$ is the true injection value.
+The error $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} is summarised in Table \ref{tab:posterior_errors}.
+We find that the estimates are more accurate using the pulsar terms, i.e.
+$\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$.
+In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models, with $\Delta_{\rm Earth}(\Omega)-\Delta_{\rm psr}(\Omega) = 6.7 \times 10^{-3}$.
+The improvement from including the pulsar terms is largest for $\iota$, with $\Delta_{\rm Earth}(\iota) - \Delta_{\rm psr \& Earth}(\iota)  = 0.24$.
+Whilst we present only a single noise realisation in this section, the improvements from including the pulsar terms are generally of a comparable magnitude across different realisations.
+\newline
+
+For the low-SNR results in Figure \ref{fig:corner_plot_compare_low}, there is less improvement in the parameter estimates.
+Qualitatively, the green and blue contours and histograms mostly overlap, although the green contours are centred slightly better on the injected values.
+The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR case is reported in the lower half of Table \ref{tab:posterior_errors}.
+We see that the error is larger for every element in $\boldsymbol{\theta}_{\rm gw}$ than in the high-SNR case, as expected.
+The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; we find $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for all parameters except $\psi$ and $h_0$.
+However it is hard to draw strong conclusions; the improvements from including the pulsar terms vary across different noise realisations.
+Whilst the inclusion of the pulsar terms improves the accuracy of the parameter estimates at low SNR only modestly, we show in Section \ref{sec:psr_v_earth_bayes} that it increases the detection probability.
+\newline
 	\begin{figure*}
 	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_highSNR_compare}
-	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, with $h_0 = 10^{-12}$, for a single realisation of the system noise.
-The blue curves are the posteriors derived using the Earth-term model, Equation \eqref{eq:g_func_trig_earth}, the green curves using the pulsar-term model, Equation \eqref{eq:g_func_trig}.
-The vertical orange lines indicate the true injected values.
+	\caption{Posterior distribution in the form of a standard corner plot of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, with $h_0 = 1 \times 10^{-12}$, for a single realisation of the system noise.
+The blue curves are the posteriors calculated using the Earth-term model, Equation \eqref{eq:g_func_trig_earth}.
+The green curves are calculated by including the Earth term and pulsar terms, Equation \eqref{eq:g_func_trig}.
+The vertical and horizontal orange lines indicate the true injected values.
 The contours in the two-dimensional histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
@@
 Some parameters (e.g.
-$\psi, \iota$) exhibit a bias when using the Earth-term model which is absent when using the pulsar-term model.}
+$\psi, \iota$) exhibit a bias when using the Earth-term model which disappears when the pulsar terms are included.}
 	\label{fig:corner_plot_compare_high}
@@
 	\caption{Same as Figure \ref{fig:corner_plot_compare_high}, but for a low-SNR system with $h_0 = 5 \times 10^{-15}$.
-The posterior distributions from the Earth-term and pulsar-term models are generally overlaid.
-The two-dimensional contours suggest a slight improvement through using $\mathcal{M}_{\rm psr \& earth}$; qualitatively the injected value generally lies closer to the centre of the green contours than the blue contours.
-However the difference is modest and contingent on the specific realisation of the data.}
+The posterior distributions with (green curves) and without (blue curves) the pulsar terms overlap more closely than in Figure \ref{fig:corner_plot_compare_high}, but the green curves are still centered slightly more accurately on the injected values (horizontal and vertical orange lines).}
 	\label{fig:corner_plot_compare_low}
@@
 		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
-		\begin{tabular}{lcllc}
+		\begin{tabular}{lcll}
 			\toprule
-			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm psr}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm psr}(\theta)$   \\
+			&$\theta$ & $\Delta_{\rm Earth}(\theta)$ & $\Delta_{\rm psr \& earth}(\theta)$  \\
 			\hline
-			\multirow{7}{2mm}{High SNR} & $\Omega$       & 0.011& 0.0043& 0.00671\\
-			& $\Phi_0$ &0.00332 &0.00115 &0.00217\\
-			& $\psi$ &0.0968 &0.000653 &0.0962 \\
-			& $\iota$ & 0.343 &0.106 &0.237 \\
-			& $\delta$ & 0.00187 &0.000285 &0.00158 \\
-			&$\alpha$ &0.0403 &0.000435 &0.0399 \\
-			&$h_0$ & 0.227 &0.075 &0.152 \\
+			\multirow{7}{2mm}{High SNR} & $\Omega$       & $1.1 \times 10^{-2}$& $4.3 \times 10^{-3}$\\
+			& $\Phi_0$ &$3.3 \times 10^{-3}$ &$1.2 \times 10^{-3}$ \\
+			& $\psi$ &$9.7 \times 10^{-2}$ &$6.5 \times 10^{-4}$  \\
+			& $\iota$ & $3.4 \times 10^{-1}$ & $1.0 \times 10^{-1}$  \\
+			& $\delta$ &$1.9 \times10^{-3}$ &$2.9 \times 10^{-4}$  \\
+			&$\alpha$ &$4.0 \times 10^{-2}$ &$4.4 \times10^{-4}$  \\
+			&$h_0$ & $2.2 \times 10^{-1}$ &$7.5 \times 10^{-2}$  \\
 			\hline
-			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& 1.09 \\
-			& $\Phi_0$ &0.335 &0.229 &0.106 \\
-			& $\psi$ &0.112 &0.123 &-0.0119 \\
-			& $\iota$ & 0.443 &0.37 &0.0724 \\
-			& $\delta$ & 0.181 &0.0782 &0.103 \\
-			&$\alpha$ &0.145 &0.139 &0.00637 \\
-			&$h_0$ & 1.12 &1.41 &-0.291 \\
+			\multirow{7}{2mm}{Low SNR} & $\Omega$       &  $2/7 \times 10^0$ &$1.6 \times 10^0$ \\
+			& $\Phi_0$ &$3.4 \times 10^{-1}$ & $2.3 \times 10^{-1}$  \\
+			& $\psi$ &$1.1 \times 10^{-1}$ &$1.2 \times 10^{-1}$  \\
+			& $\iota$ & $4.4 \times 10^{-1}$ & $3.7 \times 10^{-1}$  \\
+			& $\delta$ & $1.8 \times 10^{-1}$ & $7.8 \times 10^{-2}$  \\
+			&$\alpha$ & $1.5 \times 10^{-1}$ & $1.4 \times 10^{-1}$ \\
+			&$h_0$ & $1.1 \times 10^0$ &$1.4 \times 10^0$ \\
 			\bottomrule
 		\end{tabular}
-		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=psr, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$.
-The injection values are as summarised in Table \ref{tab:parameters_and_priors}.
-The top and bottom halves of the table contain the high-SNR ($h_0 = 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.
-For all parameters the units are in rad, except for $h_0$ which is dimensionless.
-The rightmost column quotes the difference in the mean relative error between the models for each parameter; a positive value indicates that the pulsar-term model is more accurate.}
+		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = Earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=$\rm psr \& Earth$, Equation \eqref{eq:g_func_trig}) for $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+The injected values are as summarised in Table \ref{tab:parameters_and_priors}.
+The top and bottom halves of the table contain the high-SNR ($h_0 =1 \times 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.
+The psr\&Earth model is more accurate than the Earth model for all $\theta \in \boldsymbol{\theta}_{\rm gw}$ at high SNR and five out of seven parameters at low SNR.}
 		\label{tab:posterior_errors}
@@
 			\includegraphics[width=\columnwidth, height = \columnwidth ]{images/CanonicalBayesPlot2000} 	
-			\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_i$, $i \in \left \{ \rm psr \& earth, earth \right \}$ (GW present in data) and $\mathcal{M}_{\rm null}$ (GW not present in data) at different GW strains, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
-The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
-The minimum detectable strain for which $\beta < 10$ equals $3.19 \times 10^{-15}$ for $\mathcal{M}_{\rm earth}$ and $2.78 \times 10^{-15}$ for $\mathcal{M}_{\rm psr \& earth}$.
+			\caption{Bayes factor (odds ratio) $\beta_{\rm M}$ between the competing models $\mathcal{M}_{\rm M}$, with $\rm M \in \left \{ \rm psr \& Earth, Earth \right \}$ (GW present in data) and $\mathcal{M}_{\rm null}$ (GW not present in data) as a function of the signal amplitude, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}.
+The horizontal grey dashed line labels an arbitrary detection threshold, $\beta_{\rm M} = 10$.
+The minimum detectable strain for which $\beta < 10$ equals $3.2 \times 10^{-15}$ for $\mathcal{M}_{\rm Earth}$ (blue points) and $2.8 \times 10^{-15}$ for $\mathcal{M}_{\rm psr \& Earth}$ (green points).
 The axes are plotted on logarithmic scales.}
@@
 
-In this section we compute the minimal detectable strain for the representative source, Table \ref{tab:parameters_and_priors}, using $\mathcal{M}_{\rm earth}$ and $\mathcal{M}_{\rm psr \& earth}$.
-We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure.
+In this section we compute the minimum detectable strain for the representative source in Table \ref{tab:parameters_and_priors}, using $\mathcal{M}_{\rm Earth}$ and $\mathcal{M}_{\rm psr \& Earth}$.
+We frame the problem of claiming a GW detection in terms of a Bayesian model selection procedure, following the lead of other PTA analyses \citep[e.g.][]{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X}.
 We define $\mathcal{M}_{\rm null}$ as the null model that assumes no GW exists in the data.
 This is equivalent to setting $g^{(n)}(t)=1$ in Equation \eqref{eq:g_func_trig}.
-The evidence integral $\mathcal{Z}$ returned by nested sampling, Equation \eqref{eq:model_evidence}, is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_i$.
-The support in the data for the presence of a GW signal, described by model $\mathcal{M}_i$, over the absence of a GW signal is quantified via a Bayes factor
-		\begin{equation}
-			\beta_i = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_i)}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_{\rm null})} \ . \label{eq:bayes}
-		\end{equation}
-In this paper we consider $i \in \left \{ \rm psr \& earth, earth \right \}$. \newline
-
-The Bayes factor, $\beta_i$, for each of the two models is presented in Figure \ref{fig:bayes1} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).
-As in Section \ref{sec:psr_v_earth_pe} in order to enable a clear comparison between the results from $\mathcal{M}_{\rm psr \& earth}$ and $\mathcal{M}_{\rm psr}$ we consider only a single noise realisation of the synthetic data $\boldsymbol{Y}$.
-Moreover, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself.
-The conclusions drawn below are consistent across noise realisations.
-\newline
-
-We see in Figure \ref{fig:bayes1} an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$ for both $\mathcal{M}_{\rm earth}$ and $\mathcal{M}_{\rm psr \& earth}$.
-Moreover, $\beta_{\rm psr \& earth} > \beta_{\rm earth}$ for all $h_0$.
-That is, for a given $h_0$, $\mathcal{M}_{\rm psr \& earth}$ provides greater evidence for a GW signal in the noisy data than $\mathcal{M}_{\rm psr \& earth}$.
-The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.19 \times 10^{-15}$ for  $\mathcal{M}_{\rm earth}$ and for $h_0 \gtrsim 2.78 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& earth}$, a relative improvement in the minimal detectable strain of $13\%$.
-The minimum detectable strain and the relative improvement through using $\mathcal{M}_{\rm psr \& earth}$ is of course particular to the system in Table \ref{tab:parameters_and_priors} and the realisation of $\boldsymbol{Y}$.
-It will be influenced in general by $T_{\rm obs}$, $T_{\rm cad}$ as well as ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$.
+The evidence integral $\mathcal{Z}$ returned by nested sampling, Equation \eqref{eq:model_evidence}, is the probability of the data $\boldsymbol{Y}$ given a model $\mathcal{M}_{\rm M}$.
+The support in the data for the presence of a GW signal, described by model $\mathcal{M}_{\rm M}$, over the absence of a GW signal is quantified via the Bayes factor
+		\begin{equation}
+			\beta_{\rm M} = \frac{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_{\rm M})}{\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_{\rm null})} \ . \label{eq:bayes}
+		\end{equation}
+In this paper we consider $\rm M \in \left \{ \rm Earth, psr \& Earth \right \}$. \newline
+
+The Bayes factors $\beta_{\rm Earth}$ and $\beta_{\rm psr{\&}Earth}$ are plotted as functions of $h_0$ in Figure \ref{fig:bayes1} for the representative source in Table \ref{tab:parameters_and_priors}.
+We vary the source amplitude from $h_0 = 10^{-15}$ (undetectable) to $h_0 = 10^{-12}$ (easily detectable).
+To sharpen the comparison between $\mathcal{M}_{\rm psr \& Earth}$ and $\mathcal{M}_{\rm Earth}$ we present only a single noise realisation of the synthetic data $\boldsymbol{Y}$, as in Section \ref{sec:psr_v_earth_pe}; the conclusions drawn below are consistent across different noise realisations.
+Moreover, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself, to smooth the curves in Figure \ref{sec:psr_v_earth_pe}.
+\newline
+
+Figure \ref{fig:bayes1} reveals an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$ for both $\mathcal{M}_{\rm Earth}$ and $\mathcal{M}_{\rm psr \& Earth}$.
+Moreover, we obtain $\beta_{\rm psr \& Earth} > \beta_{\rm Earth}$ for all $h_0$.
+That is, for a given $h_0$, $\mathcal{M}_{\rm psr \& Earth}$ provides greater evidence for a GW signal in the noisy data than $\mathcal{M}_{\rm Earth}$.
+The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.2 \times 10^{-15}$ for  $\mathcal{M}_{\rm Earth}$ and $h_0 \gtrsim 2.8 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& Earth}$, a relative improvement in the minimal detectable strain of $14\%$.
+The minimum detectable strain and the relative improvement through using $\mathcal{M}_{\rm psr \& Earth}$ are particular to the system in Table \ref{tab:parameters_and_priors} and the realisation of $\boldsymbol{Y}$.
+They are influenced in general by $T_{\rm obs}$, $T_{\rm cad}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$.
+A full parameter sweep is postponed to future work, after the analysis scheme in this paper is upgraded from ingesting pulse frequencies to pulse TOAs to enable a like-for-like comparison with standard PTA analyses.
+In the interim, we note that a $13\%$ improvement in sensitivity when including the pulsar terms is comparable to the improvements of $5\%$ achieved by \cite{Zhupulsarterms}.
 
 \section{Conclusion}\label{sec:discussion}
-In this paper we demonstrate how to extend state-space methods for PTA data analysis to include the pulsar term.
-We compare the results when the pulsar terms are includes to those when they are omitted.
-\newline
-
-The intrinsic rotational state evolution of the pulsar, described by a mean-reverting Ornstein-Uhlenbeck process, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol}, is tracked using a Kalman filter.
-An updated state-space model for the influence of the GW is introduced that reparameterises the pulsar term as a separate pulsar-specific static parameter $\chi^{(n)}$ to be inferred for each pulsar.
-The Kalman filter is combined with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model.
-\newline
-
-The updated state-space model including the pulsar term is tested on synthetic data.
-We consider 10 noise realisations of synthetic data for a single, astrophysically representative, SMBHB GW source observed by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$ and $h_0 = 5 \times 10^{-15}$.
+In this paper we demonstrate how to extend state-space methods for PTA data analysis to include the pulsar terms as well as the Earth term.
+We emphasize that including the pulsar terms is not a new idea; the advantages of doing so are well known in standard PTA analyses
+\citep[e.g.][]{Zhupulsarterms,Chen2022,2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X,Arzoumanian2023,2023arXiv230616226A}.
+The goal of this paper is to verify whether the advantages apply equally to state-space methods, which complement standard analyses.
+In the state-space formulation, the rotational state of each pulsar evolves according to a mean-reverting Ornstein-Uhlenbeck process, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol}, and is tracked using a Kalman filter.
+The measurement equation in the Kalman filter is reparameterized in terms of a pulsar-specific static phase $\chi^{(n)}$ ($1 \leq n \leq N$) to be inferred for each pulsar.
+The Kalman filter is combined with a nested sampler to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence of the signals models with and without the pulsar terms included.
+\newline
+
+The updated state-space model including the pulsar terms is tested on synthetic data.
+We consider 10 noise realisations for a single, astrophysically representative, SMBHB GW source with $h_0 = 5 \times 10^{-15}$, observed by the 12.5-year NANOGrav pulsars ($N=47$) with $T_{\rm obs} = 10 \, {\rm years}$ and $T_{\rm cad} = 1$ week.
 We find that the updated state-space model successfully detects injected signals and estimates their static parameters accurately for all noise realisations with relatively low computational cost.
-There is a non-negligible natural random dispersion in the median of the marginalised one-dimensional posteriors, with a maximum coefficient of variation of 52\% for $\Phi_0$, and a median value across the seven elements of $\boldsymbol{\theta}_{\rm gw}$ of 10\%.
-The updated model is further tested across a broader parameter domain of SMBHB sources; 200 randomly sampled parameter vectors of $\boldsymbol{\theta}_{\rm gw}$ are explored at fixed $\iota$ and $h_0$.
-Consistent parameter estimates are obtained for the majority of cases, although for some SMBHB source positions on the sky the sensitivity of the specific PTA configuration used in this paper is reduced leading to sub-optimally converged posteriors.
-\newline
-
-We compare the accuracy of the parameter estimates from the updated model to those when the pulsar term is omitted for a representative SMBHB source.
-We find that the inclusion of the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) by 61\% for $\Omega$, up to 99\% for $\psi$ and $\alpha$ for a representative SMBHB system with a high SNR ($h_0 = 10^{-12}$).
-For low SNR ($h_0 = 5 \times 10^{-15}$) there is no significant improvement because the data is noise dominated.
-The inclusion of pulsar terms lowers the floor of the minimal detectable strain by $8\%$ \newline
-
-State-space methods for PTA data analysis are complementary to traditional approaches, and enable the evolution of the intrinsic pulsar timing noise to be tracked explicitly rather than fitting for the ensemble-averaged power spectral density of the noise.
-That is, it is possible to disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data.
+The irreducible random dispersion (cosmic variance) in the median of the marginalised one-dimensional posteriors, with a maximum $CV$ of 52\% for $\Phi_0$, a minimum of 0.2 \% for $\Omega$ and a median of $10 \%$ for $h_0$.
+The updated model is further tested across a broad and astrophysically plausible parameter domain of SMBHB sources, exploring 200 randomly sampled $\boldsymbol{\theta}_{\rm gw}$ at fixed $\iota$ and $h_0$.
+Consistent parameter estimates are obtained, although the accuracy is lower, when one or more PTA pulsars coincide approximately on the sky with the SMBHB source.For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases.
+For the worst estimated parameter $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases.
+Including the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) by 61\% for $\Omega$, up to 99\% for $\psi$ and $\alpha$ for a representative SMBHB system with a high SNR ($h_0 = 10^{-12}$).
+For low SNR ($h_0 = 5 \times 10^{-15}$), including the pulsar terms lowers the minimum detectable strain by $14\%$, comparable with standard PTA analyses \citep[e.g.][]]{Zhupulsarterms} \newline
+
+State-space methods for PTA data analysis are complementary to traditional approaches.
+They track the actual, astrophysical, time-ordered realization of the timing noise in every PTA pulsar instead of fitting for the noise power spectral density (PSD), which averages over the ensemble of possible timing noise realizations (not just the actual one) and discards time-ordered information when taking the modulus of the complex Fourier phase in each PSD frequency bin.
+That is, it is possible to disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters.
 State-space models are conditional on a noise model that is related but different to the noise model in traditional approaches.
-The analysis in this paper assumes a specific, time-ordered realization of a mean-reverting Ornstein-Uhlenbeck process, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law power spectral density, whose amplitude and exponent are adjustable.
+The analysis in this paper assumes a mean-reverting Ornstein-Uhlenbeck process, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law PSD, whose amplitude and exponent are adjustable.
+The Ornstein-Uhlenbeck process also maps onto a stationary Gaussian process, whose PSD is a power law at high frequencies and rolls over at low frequencies, and whose amplitude and exponent can be adjusted by modifying the form of the damping term and Langevin driver in Equation \eqref{eq:frequency_evolution}.
+However, the Ornstein-Uhlenbeck process in the time domain in Equation \eqref{eq:frequency_evolution} contains more information than its associated PSD in the frequency domain for the two reasons stated above: the Kalman filter ``fits'' the actual noise realization rather than an ensemble average, and it preserves time ordering by implicitly preserving the Fourier phases, which the PSD discards.
 Clarifying the similarities and differences between various approaches promises to be a fruitful avenue of future work.
@@
 
-There are at least four potential extensions to this work, as follows:
+There are at least four useful extensions to this work.
 
@@
 
-	\item For the single representative SMBHB source in Section \ref{sec:psr_v_earth_pe} there was no obvious improvement in the parameter estimation accuracy in the low-SNR case.
-It would be of interest to explore a broader parameter space of $\boldsymbol{\theta}'$ (i.e.
-varying both the SMBHB source and the PTA configuration) to quantify the accuracy gained through including the pulsar terms across an astrophysically representative regime.
-A similar exercise should be undertaken for Section \ref{sec:psr_v_earth_bayes} to quantify the improvement in the minimal detectable strain.
+	\item For the single representative SMBHB source in Section \ref{sec:psr_v_earth_pe} there is no appreciable improvement in the parameter estimation accuracy for $h_0 = 5\times 10^{-15}$ (i.e.\ low SNR).
+It would be of interest to explore a broader $\boldsymbol{\theta}'$ domain (i.e.
+varying both the SMBHB source and the PTA configuration) to check whether including the pulsar terms improves the estimation accuracy at low SNR in some overlooked pockets of the astrophysically plausible ${\boldsymbol{\theta}}'$ domain.
+A similar exercise should be undertaken with respect to the minimum detectable GW strain (see Section \ref{sec:psr_v_earth_bayes}  ).
 \newline
@@
 Nevertheless, SMBHBs are not strictly monochromatic.
-In principle it is straightforward, to extend the analysis to account for the evolution of $\Omega(t)$, namely by including $\Omega(t)$ in the differential state equations.
-However, issues related to identifiability \citep{e5be7c83a0d24500826f6e1b414d1733}, the form of the measurement equation Equation \eqref{eq:g_func_trig}, and the linear structure the Kalman filter must be considered carefully and are postponed to a future paper.
-Regarding the last point, the Kalman filter can be applied to non-linear problems if needed using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
-\newline
-
-	\item The state-space model presented here ingests a frequency time series.
-It is necessary to generalize this approach to ingest TOAs directly, as happens traditionally.
-Generalizing the algorithm is a subtle task and will be presented in a forthcoming work.
+In principle it is straightforward to include the evolution of $\Omega(t)$ in the differential state equations.
+However, issues related to identifiability \citep{e5be7c83a0d24500826f6e1b414d1733}, the form of Equation \eqref{eq:g_func_trig}, and the linear structure of the Kalman filter must be considered carefully and are postponed to a future paper.
+The Kalman framework can be applied to non-linear problems if needed using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
+\newline
+
+	\item The state-space analysis presented here ingests a frequency time series $f_{\rm m}^{(n)}(t)$.
+It is necessary to generalize this approach to ingest TOAs directly, as happens in standard PTA analyses \citep[e.g.][]{Zhupulsarterms,Chen2022,2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X,Arzoumanian2023,2023arXiv230616226A}.
+Generalizing the algorithm is a surprisingly subtle task and will be presented in a forthcoming paper.
 \newline
@@
 	\item We assume in this paper that there is only one GW source.
-However, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
-The Kalman filter extends naturally to multiple sources; one can modify Equation \eqref{eq:measurement} easily to accommodate a linear superposition of GWs.
-Taking the logic further, the stochastic background itself is arguably an incoherent sum of many individual GW sources.
-As long as a way can be found to summarize economically the many static parameters associated with the background sources, it should be possible for a Kalman filter and nested sampler to operate together to detect the stochastic background.
-Summarizing the parameter set economically, while respecting the mathematical structure of the Kalman filter, is a subtle challenge, which we postpone to a forthcoming paper.
+The Kalman framework extends naturally to multiple sources.
+It is straightforward to modify Equation \eqref{eq:measurement} to describe a linear superposition of GWs.
+This is useful for two reasons.
+Firstly, it may be possible to resolve multiple continuous GW sources concurrently \citep{PhysRevD.85.044034}.
+Secondly, the stochastic background itself is an incoherent sum of many individual GW sources.
+It should be possible for a Kalman filter and nested sampler to operate together to detect the stochastic background.
+For the first case, it is straightforward to apply the Kalman filter and nested sampler with an extended range of static parameters associated with the GW sources.
+For the second case it is necessary to summarize economically the additional static parameters, whilst respecting the mathematical structure of the Kalman filter.
+This is a subtle challenge which we postpone to a forthcoming paper.
 If successful, it will complement the traditional approach of cross-correlating pulsar residuals to uncover the Hellings-Downs curve \citep{Hellings,2023ApJ...951L...8A}.
@@
 In this appendix, we describe the Kalman filter algorithm used in this paper.
-General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the calculation of the likelihood.
+General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the formula for the likelihood.
 The application of the Kalman filter algorithm to the specific continuous-time state-space model in Section \ref{sec:2} is outlined in Section \ref{sec_kalman_specific}.
-
-%Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}.
-
-% Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
- %Implementation of the linear Kalman filter for the PTA state-space model in Section \ref{sec:model}, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
-%
-%At each discrete timestep indexed by , the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.
 
@@
 \end{equation}
-where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix.
-$\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline
+where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix. $\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process 
+noise, with $\boldsymbol{w}_k \sim {\cal N}(0, \boldsymbol{Q}_k)$ and covariance $\boldsymbol{Q}_k$. \newline
 
 The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage.
-The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimates from step $k-1$.
+The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimate from step $k-1$.
 Specifically, the predict step proceeds as
@@
 Note that the predict stage is independent of the measurements.
-The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
+The measurement $\boldsymbol{Y}_k$ updates the prediction during the update stage as follows:
 \begin{align}
@@
 \end{align}
-Equation \eqref{eq:residual} defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation.
-The uncertainty in $\boldsymbol{\epsilon}_i$ is quantified via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$, Equation \eqref{eq:innv_covar}.
-Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.
-$\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$.
-For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
-\newline
-
-The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ is calculated at each timestep from the Kalman filter output according to
+Equation \eqref{eq:residual} defines a residual $\boldsymbol{\epsilon}_k = \boldsymbol{Y}_k  - \hat{\boldsymbol{Y}}_k$, which is sometimes termed the innovation.
+The uncertainty in $\boldsymbol{\epsilon}_k$ is quantified via the innovation covariance $\boldsymbol{S}_k = \langle \boldsymbol{\epsilon}_k \boldsymbol{\epsilon}_k^{T} \rangle$, viz.
+Equation \eqref{eq:innv_covar} (noting that the Einstein summation convention is suppressed temporarily in the latter definition).
+Equation \eqref{eq:kalman gain} defines the Kalman gain $\boldsymbol{K}_k$ For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}.
+\newline
+%i.e.
+$\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$ which is defined so as to minimise the mean squared error in the state estimate.
+
+The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_k$ given ${\boldsymbol{\hat X}}_k$ is calculated at each timestep from the Kalman filter output according to \citep{zarchan2000fundamentals}
 \begin{eqnarray}
-	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ ,
+	\log \mathcal{L}_k =  -\frac{1}{2} \left (D_k \log 2 \pi + \log  \left | \boldsymbol{S}_k \right | + \boldsymbol{\epsilon}_k^{\intercal} \boldsymbol{S}_k^{-1}  \boldsymbol{\epsilon}_k \right ) \ ,
 \end{eqnarray}
-where $N$ is the number of data points at timestep $i$. The total log-likelihood for the entire sequence is
+where $D_k$ is the number of data points at timestep $k$. The total log-likelihood for the entire sequence is
 \begin{eqnarray}
-	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
+	\log \mathcal{L} =  \sum_{k=1}^{M} \log \mathcal{L}_k \ . \label{eq:likelihood}
 \end{eqnarray}
-Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
+Given ${\boldsymbol{Y}}_k$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e.
 $\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$.
 Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$.
-In Section \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$.
+In Appendix \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$.
 \newline
@@
 To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$.
-In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:2}.
+In Appendix \ref{sec_kalman_specific} we define the machinery for the state-space model in Section \ref{sec:2}.
 
@@
 \end{equation}
-and $\boldsymbol{C}(t)$ is a time-dependent vector with $n$-th component
-\begin{equation}
-	C^{(n)} =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}^{(n)} _{\rm em}(t_1)\ .
-\end{equation}
-The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, with
+and $\boldsymbol{C}(t)$ is a time-dependent $N \times 1$ vector with $n$-th component
+\begin{equation}
+	C^{(n)}(t) =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}^{(n)} _{\rm em}(t_1)\ .
+\end{equation}
+The $N \times N$ matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, with
 \begin{equation}
@@
 
-In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic},
+In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic}
 \begin{equation}
@@
 \begin{align}
-	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
+	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}} dt' e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t')  \ , \\
 	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ,
@@
 \begin{equation}
-	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \ ,  \label{eq:appendix_noise}
+	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} d \boldsymbol{B}(t') e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma}  \ ,  \label{eq:appendix_noise}
 \end{equation}
@@
 It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point.
-The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies.
+The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies, as in Sections \ref{sec:pulsar_term} and \ref{sec:psr_v_earth_bayes}.
 Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}.
@@
 For reviews of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}.
-Multiple nested sampling algorithms and computational libraries exist.
-\citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}.
+Multiple nested sampling algorithms and computational libraries exist \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}.
 In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library.
@@
 The primary tunable parameter in nested sampling is $n_{\rm live}$.
-More live points assists with large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
+More live points addresses larger parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
 However the computational runtime scales as $\mathcal{O}(n_{\rm live})$.
@@
 
-\section{Complete workflow summary}\label{sec:workflow}
+\section{Workflow summary}\label{sec:workflow}
 For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
@@
 
-\section{Choice of priors}\label{sec:set_priors}
-To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} and the workflow of Appendix \ref{sec:workflow} it is necessary to specify a Bayesian prior $\pi(\boldsymbol{\theta})$ on the static parameters.
+\section{Priors}\label{sec:set_priors}
+To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} and the workflow in Appendix \ref{sec:workflow} it is necessary to specify a Bayesian prior $\pi(\boldsymbol{\theta})$ on the static parameters.
 In this section we outline how the priors are chosen.
@@
 We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$
-where $\sigma_{\rm SC}^{(n)}$ is the value for pulsar $n$ inferred from
+where $\sigma_{\rm SC}^{(n)}$ is the noise amplitude for pulsar $n$ inferred from
 Equation \eqref{eq:sigmap_f}.
-We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.
-We do not set a prior on $\gamma^{(n)}$, because $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and consequently $\gamma^{(n)}$ is effectively ``unobservable'' over $T_{\rm obs}$.
-For validation purposes it is therefore sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value.
+We set a non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$.
+We do not set a prior on $\gamma^{(n)}$, because one typically has $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$ astrophysically, and $\gamma^{(n)}$ is effectively ``unobservable'' for $T_{\rm obs} \sim 10 \, {\rm years}$.
+For validation purposes it is sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value.
 This reduces the total dimension of the parameter space to $7 + 4N$.
@@
 
-\section{Parameter estimation of representative example,  high-SNR} \label{sec:app_high_SNR}
-In this section we repeat the analysis of Section \ref{sec:pe_and_ms}, but in the high-SNR limit.
-We do this as a sanity check to confirm that the state-space analysis scheme with the pulsar term included returns reasonable results in the ``easy" case with a strong GW signal.
-The high-SNR limit allows us to inspect the results without confusion from a measurement noise which is large compared to the GW signal.
+\section{High-SNR example} \label{sec:app_high_SNR}
+In this appendix we repeat the analysis of Section \ref{sec:pe_and_ms} in the high-SNR regime.
+We do so as a sanity check to confirm that the state-space analysis scheme with the pulsar terms included returns reasonable results in the ``easy'' case with a strong GW signal.
+The high-SNR limit allows us to check for estimation biases unobscured by measurement noise which is large compared to the GW signal.
 \newline
@@
 We apply the parameter estimation framework as in Section \ref{sec:pe_and_ms}.
-All injected static parameters $\boldsymbol{\theta}'$, and the associated priors are as specified in Table \ref{tab:parameters_and_priors}, with the exception of $h_0$ which now takes a value $h_0 = 10^{-12}$.
+All injected static parameters $\boldsymbol{\theta}'$, and the associated priors are as specified in Table \ref{tab:parameters_and_priors}, with the exception of $h_0$ which now takes the value $h_0 = 1 \times 10^{-12}$.
 The procedure is undertaken for ten realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
@@
 There is no evidence of railing against the prior bounds.
-Correlations between $\iota$ and $h_0$ are evident over multiple noise realisations due to the weak-identifiability of these two parameters.
+Correlations between $\iota$ and $h_0$ are evident over multiple noise realisations due to the weak identifiability \footnote{By weak identifiability in this context we mean that the $\iota-h_0$ likelihood contours plateau to a flat-topped ridge with a small, non-zero gradient close to the maximum likelihood solution.} of these two parameters \citep{e5be7c83a0d24500826f6e1b414d1733}.
 For five of the seven parameters, there is effectively zero dispersion in the one-dimensional posterior medians between noise realisations.
-For the remaining two parameters, $\iota$ and $h_0$, an appreciable dispersion is observed, again due to weak-identifiability.
-\newline
-
-Figure \ref{fig:corner_high_snr_appendix_chi} displays results a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the ten noise realisations.
-The figure is exactly analogous to Figure \ref{fig:corner_plot_3}, in the high-SNR limit.
+For the remaining two parameters, $\iota$ and $h_0$, appreciable dispersion is observed, again due to weak-identifiability.
+\newline
+
+Figure \ref{fig:corner_high_snr_appendix_chi} displays results a representative subset of five out of 47 pulsar phases $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ for the ten noise realisations.
+The figure is exactly analogous to Figure \ref{fig:corner_plot_3}.
 All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values, with no evidence for correlations between parameter pairs.
-Unlike in the low-SNR case, it is now possible to infer $\chi^{(2)}$ (the phase correction for PSR J0340+4130) consistently (third row, third column of Figure \ref{fig:corner_high_snr_appendix_chi}).
-Despite the proximity on the sky of PSR J0340+4130 to our synthetic gw source, the loundness of the GW signal is sufficient to enable $\chi^{(2)}$ to be estimated.
+Unlike in the low-SNR regime, it is now possible to infer $\chi^{(2)}$ (the phase correction for PSR J0340+4130) consistently (third row, third column of Figure \ref{fig:corner_high_snr_appendix_chi}).
+Even though PSR J0340+4130 nearly coincides on the sky with the synthetic GW source, the GW signal is loud enough for $\chi^{(2)}$ to be estimated.
 
@@
 	\includegraphics[width=\textwidth, height = \textwidth ]{images/corner_highSNR} 	
-	\caption{Same as Figure \ref{fig:corner_plot_1}, but for a SMBHB with $h_0 = 10^{-12}$.
+	\caption{Same as Figure \ref{fig:corner_plot_1}, but for a loud SMBHB with $h_0 = 1 \times 10^{-12}$.
 The Kalman filter and nested sampler estimate all seven parameters in $\boldsymbol{\theta}_{\rm gw}$ accurately.
-There is a reduced dispersion in the  one-dimensional posterior medians between noise realisations compared to the low-SNR case (c.f.
+There is less dispersion in the  one-dimensional posterior medians among noise realisations compared to the low-SNR case (c.f.
 Figure \ref{fig:corner_plot_1}).
-An appreciable dispersion remains for $\iota$ and $h_0$, along with a correlation across noise realisations for these two parameters, due to their weak-identifiability.}
+An appreciable dispersion remains for $\iota$ and $h_0$, which are correlated and weakly identifiable.}
 	\label{fig:corner_high_snr_appendix}
@@
 	\includegraphics[width=\columnwidth, height = \columnwidth ]{images/corner_highSNR_chi} 	
-	\caption{Same as Figure \ref{fig:corner_plot_3}, but for a SMBHB with $h_0 = 10^{-12}$.
+	\caption{Same as Figure \ref{fig:corner_plot_3}, but for a loud SMBHB with $1 \times h_0 = 10^{-12}$.
 Consistent unimodal posteriors are obtained across all noise realisations and all five displayed parameters.
-It is now possible to infer $\chi^{(2)}$ due to the loundness of the GW signal, unlike in the low-SNR case where consistent posteriors for $\chi^{(2)}$ could not be estimated.}
+It is now possible to infer $\chi^{(2)}$ due to the loundness of the GW signal, unlike in the low-SNR case in Figure \ref{fig:corner_plot_3}.
+The colour scheme is the same as in Figure \ref{fig:corner_plot_3}, but the curves are all overlapping and so are obscured by the orange curves.}
 	\label{fig:corner_high_snr_appendix_chi}
@@
 
-%
-%\section{Multiple noise realisaitons and the Wasserstein distance}
-%
-%
-%We can extend the analysis of these two points and consider 1000 noise realisations of $\boldsymbol{Y}$, beyond the \textcolor{red}{X} plotted in Figure \ref{fig:corner_plot_1}.
-As in Figure \ref{fig:corner_plot_1} we independently estimate the static parameters, $\boldsymbol{\theta}$ for each realisation and calculate $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$.
-%
-%
-%
-%
-%To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$.
-For each realisation we
-%
-%
-%We now consider the complete set of $10^3$ noise realisations, going beyond the subset of nine realisations in the preceding discussion.
-We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$?
-There is no unique way to answer this question.
-In this paper we appeal to the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory, which defines an intuitive notion of similarity between probability distributions.
-The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}; a short review is presented in Appendix \ref{sec:wasserstein}.
-The WD measures the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$.
-In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}.
-Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$ with respect to the PDFs $\mu$ and $\nu$.
-Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad.
-Then we can infer $| \langle \iota \rangle_\mu - \langle \iota \rangle_\nu | \leq 0.5 \, {\rm rad}$.
-\newline
-%
-%
-%Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$.
-The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$.
-As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$.
-These values quantify approximately the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
-Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$.
-These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for $10^3$ noise realizations.
-Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
-%
-%\begin{table}
-%	\centering
-%	\begin{tabular}{ccll}
-%		\toprule
-%		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
-%		\hline
-%		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
-%		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
-%		$\psi$              & $2.50$ & rad & $0.16$ \\
-%		$\iota$             & $1.0$ & rad & $0.14$ \\ 
-%		$\delta$              & $1.0$  & rad & $0.09$ \\
-%		$\alpha$          & $1.0$  & rad & $0.17$\\
-%		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
-%		\bottomrule
-%	\end{tabular}
-%	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}.
-$W_{1, \rm median}$ is generally smaller than the prior domain.}
-%	\label{tab:Wasserstein}
-%\end{table}
-%
-%
-
 \section{Challenges of the pulsar term}\label{sec:psr_term_challenges}
@@
 	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
-	\caption{Logarithm of the likelihood curve $\mathcal{L}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
-All other static elements of $\boldsymbol{\theta}$ are held constant at their true injected values.
-In the left panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}(\alpha)$, i.e.
+	\caption{Logarithm of the cross-section along the $\alpha$-axis of the likelihood function $\mathcal{L}_{\rm slice}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
+All static elements of $\boldsymbol{\theta}$ other than $\alpha$ are held fixed at their true injected values for the purpose of the tests in Appendix \ref{sec:psr_term_challenges}.
+In the left panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}_{\rm slice}(\alpha)$, i.e.
 inclusive of the pulsar terms.
@@
 just the Earth terms.
-The vertical dashed lines indicate the true injected value of $\alpha$ ($=1.0$ rad) used to generate $\boldsymbol{Y}$.
-The inset plot covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
-In the left panel $\mathcal{L}(\alpha)$ is noisy, though a smooth concave optima exists on scales $\mathcal{O}(10^{-4})$ rad.
-This noise hampers the ability of sampling algorithms to converge to a posterior distribution.
-Conversely, in the right hand panel $\mathcal{L}(\alpha)$ is globally smooth.
+The vertical dashed lines indicate the true injected value $\alpha = 1.0$ rad used to generate $\boldsymbol{Y}$.
+The inset in the left panel covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
+In the left panel $\mathcal{L}_{\rm slice}(\alpha)$ is jagged, although a unique peak exists in principle at $\alpha=1.0 \, {\rm rad}$, and $\mathcal{L}_{\rm slice}(\alpha)$ is smooth within the neighbourhood plotted in the inset.
+The jagged form of $\mathcal{L}_{\rm slice}(\alpha)$ hampers the convergence of the nested sampler.
+In the right panel $\mathcal{L}_{\rm slice}(\alpha)$ is smooth for $0\leq\alpha\leq \pi$.
 }
@@
 \end{figure*}
-In this section we elucidate the complications introduced by including the pulsar term into the state-space analysis and justify the reparametrisation in terms of $\chi^{(n)}$ introduced in Section \ref{sec:pulsar_term2}.
-\newline
-
-We generate some noisy synthetic data $\boldsymbol{Y}$ as outlined in Section \ref{sec:synthetic_data}.
+In this appendix we elucidate the complications introduced by the pulsar terms into the state-space analysis and justify the reparametrisation in terms of $\chi^{(n)}$ introduced in Section \ref{sec:pulsar_term2}.
+\newline
+
+As a first step, we demonstrate empirically that sticking with the original parameterization of the pulsar terms in Equation \eqref{eq:g_func_trig}, i.e.\ writing them in terms of $d^{(n)}$ instead of $\chi^{(n)}$, produces a highly oscillatory likelihood function, which the nested sampler fails to navigate successfully.
+Physically, this happens because the Doppler factor depends on the GW wavelength, and there are $\gtrsim 10^3$ GW wavelengths across the Galactic volume spanned by the PTA.
+To enact the demonstration, we generate noisy synthetic data $\boldsymbol{Y}$ as outlined in Section \ref{sec:rep_example1}.
 The SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
-We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig}, on $\boldsymbol{Y}$ for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ rad passed to the  Kalman filter.
-All other elements of $\boldsymbol{\theta}_{\rm gw}$ are fixed at their true injection values.
+We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig}, for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ passed to the  Kalman filter.
+All other elements of $\boldsymbol{\theta}_{\rm gw}$ are fixed at their true injected values.
 At this point we do not use the $\chi^{(n)}$ parametrisation.
-Therefore varying $\alpha$ changes the value of $\boldsymbol{n}$ everywhere it appears in Equation \eqref{eq:g_func_trig}.
-For each value of $\alpha$, the Kalman filter returns a likelihood $\mathcal{L}(\boldsymbol{Y} | \alpha) = \mathcal{L}(\alpha)$.
-\newline
-
-The function $\mathcal{L}(\alpha)$ is displayed in the left panel of Figure \ref{fig:likelihood_comparison} (orange curve).
-The dashed grey line indicates the injected value of $\alpha$ (= 1.0 rad).
-The inset plot shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
-There are two important features to discuss in the left panel.
-The first is that on scales of the order the width of the prior $\pi(\alpha)$ (i.e radians), the curve of $\mathcal{L}(\alpha)$ is very noisy.
-Similar noisy curves are obtained for $\mathcal{L}(\Omega)$ and $\mathcal{L}(\delta)$.
-For all other static parameters, the $\mathcal{L}(\theta)$ curves are smooth and globally concave.
-The second is that despite being noisy, a true likelihood maximum coincident with the injected value does exist and is locally concave on scales $\mathcal{O} \left( 10^{-4}\right)$ radians.
-This means that whilst it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$, in practice the computation is intractable since the noise in the likelihood curve inhibits sampling algorithms to reliably converge to a definitive posterior.
-Moreover, $\mathcal{L}(\alpha)$ is only a one-dimensional function.
-The challenge for sampling algorithms is further exacerbated when we consider the three-dimensional parameter space in $\alpha - \delta - \Omega$, or the full $7 + 5N$ dimensions of Section \ref{sec:ss_params}.
-\newline
-
-Why do $\alpha$, $\delta$ and $\Omega$ exhibit noisy $\mathcal{L}(\theta)$ curves whereas other static parameters do not?
-These parameters appear in the pulsar term as a product in the pulsar-dependent phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, where we remind the reader that $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
-Over the typical pulsar distances $d^{(n)} \sim 1$ kpc, a 10 nHz GW can accumulate $10^3$ cycles.
-This creates a confusion problem when taking the cosine of  $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ since we do not know how many cycles the wave has gone through, modulo $2 \pi$.
-\newline
-
-In order to smooth the likelihood curve there are two options.
+Therefore changing $\alpha$ also changes $\boldsymbol{n}$ everywhere it appears in Equation \eqref{eq:g_func_trig}.
+For each value of $\alpha$, the Kalman filter returns a likelihood $\mathcal{L}(\boldsymbol{Y} | \alpha) = \mathcal{L}_{\rm slice}(\alpha)$, i.e.\ a cross-section of the $\mathcal {L}(\boldsymbol{\theta}_{\rm gw})$ along the $\alpha$-axis for the purpose of testing.
+\newline
+
+The function $\mathcal{L}_{\rm slice}(\alpha)$ is displayed in the left panel of Figure \ref{fig:likelihood_comparison} (orange curve).
+The dashed grey line indicates the injected value $\alpha= 1.0$ rad.
+The inset shows a zoomed-in section of $\mathcal{L}(\alpha)$ within the region $ 1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad.
+There are two important features in the left panel.
+First, on scales of the order the width of the prior $\pi(\alpha)$ (i.e radians), the curve $\mathcal{L}_{\rm slice}(\alpha)$ is jagged.
+Similar jagged curves are obtained for $\mathcal{L}_{\rm slice}(\Omega)$ and $\mathcal{L}_{\rm slice}(\delta)$ (not plotted for brevity).
+For the other static parameters, the curves are smooth and globally concave.
+Second, despite being jagged, a true likelihood maximum coincident with the injected value does exist and is locally concave in a neighbourhood spanning $\sim 10^{-4} \, {\rm rad}$.
+Hence it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$ without reparameterizing in terms of $\chi^{(n)}$, but in practice the computation is intractable and does not converge for reasonable choices of $n_{\rm live}$.
+The challenge for sampling algorithms is exacerbated when we extend the above test from $\mathcal{L}(\alpha)$ tp the $\alpha$-$\delta$-$\Omega$ subspace or the full $7 + 5N$ domain in Section \ref{sec:ss_params}.
+\newline
+
+Why do $\alpha$, $\delta$ and $\Omega$ exhibit jagged likelihood curves whereas other static parameters do not?
+The culprit is the Doppler factor flagged above: $\alpha$, $\delta$, and $\Omega$ appear in the phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, with $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
+For $d^{(n)} \sim 1$ kpc, a 10 nHz GW accumulates $ \gtrsim 10^3$ cycles.
+The inference problem becomes multiply degenerate, once the cosine of the phase is calculated (modulo $2\pi$).
+\newline
+
+In order to smooth the likelihood function, one can try two things.
 The first is to drop the pulsar term completely, i.e.
-using $\mathcal{M}_{\rm earth}$.
-This is the approach taken in K24 and other standard PTA analysis.
-The second is to reparametrise the pulsar term such that the phase correction $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ is no longer a function of $\alpha$ (or $\Omega$ or $\delta$), but an independent parameter to be estimated.
+using $\mathcal{M}_{\rm Earth}$.
+This is the approach taken in K24 and some (not all) other standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+The second is to reparameterize the pulsar terms such that the phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ is no longer a function of $\alpha$ (or $\Omega$ or $\delta$), but an independent parameter $\chi^{(n)}$ to be estimated.
 This is the approach presented in this paper, i.e.
-using $\mathcal{M}_{\rm psr \& earth}$.
-The right panel of Figure \ref{fig:likelihood_comparison} displays $\mathcal{L}(\alpha)$ calculated using $\mathcal{M}_{\rm earth}$ (blue curve) and $\mathcal{M}_{\rm psr \& earth}$ (green curve).
-In both of these cases there is no confusion regarding the number of cycles the wave has gone through over a distance $d^{(n)}$ and the resulting likelihood curve is smooth.
+using $\mathcal{M}_{\rm psr \& Earth}$, as in Section \ref{sec:pulsar_term2}.
+The right panel of Figure \ref{fig:likelihood_comparison} displays $\mathcal{L}_{\rm slice}(\alpha)$ calculated using $\mathcal{M}_{\rm Earth}$ (blue curve) and $\mathcal{M}_{\rm psr \& Earth}$ (green curve).
+In both instances there is no ambiguity about the number of cycles the wave has gone through over a distance $d^{(n)}$, and the likelihood function is smooth.
 

### 2024-01-23 docs/manuscript/all_internal_comments/internal_comments_5/difference_file/new.tex

--- 2023-10-23 docs/manuscript/internal_comments_4/difference_file/new.tex
@@
 We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues \citep{Manchester2005}.
-For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} \%$ and $\pm 2.31 \times 10^{-3} \%$ respectively.
+For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} $ per cent and $\pm 2.31 \times 10^{-3}$ per cent respectively.
 In this paper we adopt uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database \citep{Manchester2005}.
-Adopting more relaxed priors tests the method stringently.
-The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are relaxed or narrow.
-\newline
-
-The pulsar distances $d^{(n)}$ are not as well constrained as $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with typical uncertainties $\sim 10 \%$ \citep{Yao2017,Arzoumanian2018ApJS..235...37A}.
+Wider-than-necessary priors test the method more stringently than narrow priors.
+The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are wide or narrow.
+\newline
+
+The pulsar distances $d^{(n)}$ are less constrained than $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, with typical uncertainties $\sim 10 \%$ \citep{Yao2017,Arzoumanian2018ApJS..235...37A}.
 In this paper, however, we omit the term involving $d^{(n)}$ from Equation \eqref{eq:g_func}, as justified in Section \ref{sec:parameter_estim}, so there is no need to set a prior on $d^{(n)}$.
-Furthermore we do not set a prior on $\gamma^{(n)}$, because we have $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and $\gamma^{(n)}$ is effectively ``unobservable" over a decade; that is,for $T_{\rm obs}$ =10 years the solution of Equation \eqref{eq:spinevol} is approximately independent of $\gamma^{(n)}$, as long as $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
+Furthermore we do not set a prior on $\gamma^{(n)}$, because we have $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and $\gamma^{(n)}$ is effectively ``unobservable" over a decade; that is, for $T_{\rm obs}=10$ years the solution of Equation \eqref{eq:spinevol} is approximately independent of $\gamma^{(n)}$, as long as $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ is satisfied.
 It is therefore sufficient for validation purposes to carry $\gamma^{(n)}$ through the analysis at its injected value.
@@
 $\gamma^{(n)} = 10^{-14} \text{ s}^{-1}$ versus $10^{-13} \text{ s}^{-1}$, respectively.
-The results in (i) and (ii) are the same as those reported in Section \ref{sec:rep_example}.} \newline
+The results in (i) and (ii) are found to be the same as those reported in Section \ref{sec:rep_example}.} \newline
 
@@
 They are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}.
-In this work we follow the standard approach and drop the pulsar terms from the inference model (but not the source model).
+In this work we follow the standard approach and drop the pulsar terms from the inference calculation (but not the source model).
 Explicitly the measurement equation used in the Kalman filter reduces to
@@
 	\end{equation}
-	We defer the inclusion of the pulsar terms to a future paper.
-As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem since the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
+	We defer the inclusion of the pulsar terms in the inference calculation to a future paper.
+As discussed in Section \ref{sec:priors} this choice reduces the dimensionality of the parameter estimation problem because the measurement equation, Equation \eqref{eq:measurement}, is no longer a function of the pulsar distance.
 We stress that the pulsar terms are dropped only when doing Bayesian inference, i.e.
@@
 For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e.
-a bias or a random outcome specific to this particular noise realisation.
-We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} using more noise realizations and show that indeed there is a bias from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
+a bias, or a random outcome specific to this particular noise realisation.
+We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} using more noise realizations and show that indeed there is a systematic bias from dropping the pulsar terms, similar to that reported in \cite{Zhupulsarterms}.
 \newline
@@
 	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
-	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely.
+	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely.
 The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation.
 The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations.
-There is an appreciable dispersion among the peaks of the one-dimensional posteriors is appreciable, with coefficient of variation $\sim10 \%$ across the 63 combinations.
+There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation $\sim10 \%$ across the 63 combinations.
 A slight skew-bias is apparent in some of the parameters, e.g.
 $\iota$.
-Weak correlations can be seen between $\Omega - \Phi_0$, $\psi-\alpha$ and $\iota - h_0$.}
+Weak correlations can be seen between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.}
 	\label{fig:corner_plot_2}
@@
 We plot nine realisations rather than the full set of 1000 to avoid overcrowding.
-As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm gw}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
-\newline
-
-Figure \ref{fig:corner_plot_2} confirms two main points (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
-Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 out of the 63 (i.e.
-$92 \%$) possible combinations of the seven parameters and nine realizations.
+As in Section \ref{sec:parameter_estim}, $\boldsymbol{\theta}_{\rm psr}$ is also recovered unambiguously across the 1000 noise realisations, but for the sake of brevity we do not show the results here.
+\newline
+
+Figure \ref{fig:corner_plot_2} confirms two main points: (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations.
+Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 (i.e.
+$92 \%$) out of the 63 possible combinations of the seven parameters and nine realizations.
 Figure \ref{fig:corner_plot_2}, like Figure \ref{fig:corner_plot_1}, displays tentative signs of bias, where the one-dimensional posteriors are not symmetric about the injected value.
@@
 In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}.
-Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$" with respect to the PDFs $\mu$ and $\nu$.
+Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$ with respect to the PDFs $\mu$ and $\nu$.
 Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad.
@@
 These values are an approximate measure of the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown.
-Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.14 \times 10^{-4}$ \% for $h_0$ to 2.69\% for $\psi$.
-These values serve as confirmation that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for a wide variety of noise realizations.
+Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$.
+These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for a wide variety of noise realizations.
 Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.
@@
 We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}.
-In Equatin \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-terms-only model, i.e.
+In Equation \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-terms-only model with a GW present, i.e.
 the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}.
@@
 
-We see in Figure \ref{fig:bayes} an approximate quadratic relationship $\beta \propto h_0^2$ exists.
+We see in Figure \ref{fig:bayes} an approximate quadratic relationship $\beta \propto h_0^2$.
 The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 4 \times 10^{-15}$.
 Of course, the minimal detectable strain is particular to the system in Table \ref{tab:parameters_and_priors}.
-It is influenced in general by e.g.
-$T_{\rm obs}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$, as discussed in Section \ref{sec:parameter_space}.
-Adjusting $\sigma_{\rm m}$ linearly scales the effective signal-to-noise ratio $h_0 / \sigma_{\rm m}$, such that an approximate quadratic relationship $\beta \propto h_0^2 / \sigma_{\rm m}^2$ also exists.
+It is influenced in general by $T_{\rm obs}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$, as discussed in Section \ref{sec:parameter_space}.
+Adjusting $\sigma_{\rm m}$ we find that an approximate quadratic relationship $\beta \propto h_0^2 / \sigma_{\rm m}^2$ also exists, where $h_0 / \sigma_{\rm m}$ is the effective signal-to-noise ratio.
 \newline
@@
 Moreover, the points in Figure \ref{fig:bayes} become sparser for $h_0 \lesssim 4 \times 10^{-15}$.
-This happens because we find $\beta <0$ for the missing, intermediate points.
-This is a noise artefact of the nested sampler.When the sampler converges sub-optimally, the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails, and $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
+This happens because we obtain $\beta <0$ for the missing, intermediate points.
+This is a noise artefact of the nested sampler.
+When the sampler converges sub-optimally, the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails, and $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds.
 There is nothing special about the particular missing points; if one recreates the $\beta(h_0)$ curve by rerunning the nested sampler with another random seed, a different set of points are missing.
@@
 The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$.
-The minimum detectable strain for $\beta < 10$, is $ 4 \times 10^{-15}$.
-Missing points for $\beta\leq 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
+The minimum detectable strain, for $\beta < 10$, equals $4 \times 10^{-15}$.
+Missing points for $\beta \lesssim 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
 	\label{fig:bayes}
@@
 
-\section{Exploring a broader parameter space} \label{sec:parameter_space}
+\section{SMBHB source parameters} \label{sec:parameter_space}
 \begin{figure}
@@
 	\includegraphics[width=\columnwidth]{images/pp_plot_new}
-	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}.
-Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$.
+	\caption{The fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e.
+PP plot).
+The injections are 200 simulated GW sources for a subset of five parameters in $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions in Table \ref{tab:parameters_and_priors}.
+Each coloured curve corresponds to a different parameter (see legend).
 The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR.
-The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals.
-Well estimated posteriors should fall along the diagonal $y=x$.
-$\Omega$ is generally well-estimated but many other parameters show evidence of being over-constrained due to a modelling bias.
-}
+The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals given our statistical sample.
+For parameters with well estimated posteriors, the PP plot curve should fall along the diagonal of unit slope.
+$\Omega$ is generally well-estimated (i.e.
+it lies close to the unit diagonal) but the four other parameters show evidence of being over-constrained (i.e.
+the curves lie above the unit diagonal for low credible intervals, below the unit diagonal for high credible intervals).
+This is due to a modelling bias.}
 	\label{fig:parameter_space}
 \end{figure}
-We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
-In this section we test the method in different regions of the parameter space, varying the source parameters through astrophysically relevant ranges.
-\newline
-
-We consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}.
-We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space.
-For each simulated injection we can then attempt to recover the posterior distributions for each parameter.
-To summarise the results across the parameter space we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
-A PP plot describes the fraction of the total number of injected parameters which are included within a given credible interval of the estimated posterior with respect to the credible interval itself.
-In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage credible interval.
-The results are shown in Figure \ref{fig:parameter_space}.
-The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections.
-We can see that only $\Omega$ falls within the $3\sigma$ shaded region.
-The other parameters deviate from the $y=x$ diagonal.
-The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$.
-The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and there are more injections contained within the low value credible intervals.
-This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally confident, but are systematically biased away from the true injection values.
+Section \ref{sec:rep_example} focuses on a single representative system, summarised in Table \ref{tab:parameters_and_priors}.
+In this section we test the method for various systems, varying the SMBHB source parameters through astrophysically relevant ranges.
+\newline
+
+We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}.
+We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections.
+For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
+To summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}.
+A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself.
+In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope.
+\newline
+
+ Figure \ref{fig:parameter_space} displays the results of the numerical experiment described in the previous paragraph.
+The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections.
+We see that only $\Omega$ falls wholly within the $3\sigma$ shaded region.
+The PP curves for the other parameters deviate from the diagonal of unit slope.
+The deviation is more pronounced for $\alpha$ and $\delta$ and less for $\psi$ and $\Phi_0$.
+The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high-value credible intervals than would be expected statistically, and there are more injections contained within the low-value credible intervals.
+This stems from the bias that we noted in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally confident, but are systematically biased away from the true injection values.
 This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval.
-An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.
-
-\section{Identifiability and bias} \label{sec:bias_and_identifiability}
-There are two underlying issues in the preceding tests that we now explore in more detail.
+The origin of the bias is discussed in Section \ref{sec:bias}.
+
+\section{Systematic biases} \label{sec:bias_and_identifiability}
+ Two systematic aspects of the tests in Sections \ref{sec:rep_example} and \ref{sec:parameter_space} merit further study.
 The first relates to parameter identifiability, i.e.
@@
 The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}.
-This is the focus of Section \ref{sec:bias}.
-In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f.
-Figure \ref{fig:bayes}) for our representative example system (Table \ref{tab:parameters_and_priors}).
+It is the focus of Section \ref{sec:bias}.
+In order to elucidate these matters without confusion from the measurement noise, we switch to the high-SNR regime and set $h_0 = 10^{-12}$ (c.f.
+Figure \ref{fig:bayes}) for the representative system in Table \ref{tab:parameters_and_priors}.
 
@@
 	\includegraphics[width=\columnwidth]{images/likelihood_surface}
-	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation.
-The red point labels the location of the likelihood maxima.
-The likelihood surface has been normalised with respect to the absolute value of the maxima.
-A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
+	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota - h_0$ plane for a single noise realisation.
+The red dot marks the maximum of $\log {\cal L}$.
+The colour scale is normalised with respect to the absolute value of the maximum.
+An S-shaped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
 	\label{fig:likelihood_surface}
 \end{figure}
-In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations. 
-The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but just for $\iota$ and $h_0$ rather than all members of $\boldsymbol{\theta}_{\rm gw}$.
-In the case of high-SNR we might expect \textit{a priori} that our estimates would be increasingly accurate over the low-SNR case, and that the method would converge robustly to similar posteriors irrespective of the noise realisation (i.e.
-$W_1 \to 0$).
-This is generally true for the majority of the component parameters of $\boldsymbol{\theta}_{\rm gw}$; the estimated posteriors across noise realisations are largely indistinguishable.
-However for $\iota$ and $h_0$ this is not the case.
-Instead the inferred one-dimensional posteriors do not broadly overlap and are not consistent across different noise realisations.
-Specifically across $10^3$ noise realisations the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
-For $\iota$, $W_{1,\rm median}$ is greater than the value inferred in the low SNR case ($=0.14$, Table \ref{tab:Wasserstein}), whilst for $h_0$ $W_{1,\rm median}$ is also greater as a proportion of the scale set by the strain amplitude.
-This is in contrast to $W_{1,\rm median}$ inferred for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ in the high-SNR regime where as SNR increases, $W_1$ decreases; e.g.
-for $\Phi_0$, $W_{1, \rm median} = 4.5 \times 10^{-3}$ rad.
-\newline
-
-This issue of inferring very different posteriors depending on the particular realisation of the noise is due to these parameters being only weakly-identifiable.
-Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}.
-Sources of non-identifiability are generally categorised as either ``structural", arising from the structure of the model, or ``practical", arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
-Regarding $\iota$ and $h_0$, we are concerned with structural identifiability.
-Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx} and \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.
-For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$.
-This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}.
-It can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable.
-That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a small, non-zero gradient close to the maximum likelihood solution.
-This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
-Whilst a single, unique maxima exists (labelled by the red point in the figure), along the ridge in the parameter space the likelihood values are all very similar.
-Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise.
-If during the inference we fix one of the parameters at its true value (i.e.
-set a delta function prior), taking a slice through the likelihood surface it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably, i.e.
+In Figure \ref{fig:just_iota_h} we plot the one-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations. 
+The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$.
+One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, which is related to the bias discussed in Section 7.2; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high.
+One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$.
+Specifically, across $10^3$ noise realisations, the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$.
+Both $W_{1,{\rm median}}$ values are higher (as a fraction of the true, injected value) than in the lower-SNR tests conducted in Sections \ref{sec:rep_example} and \ref{sec:parameter_space}.
+In contrast, $W_{1,\rm median}$ for the other five parameters in $\boldsymbol{\theta}_{\rm gw}$ decreases, as the SNR increases.
+\newline
+
+We infer different $h_0$ and $\iota$ posteriors for different noise realizations, because $h_0$ and $\iota$ are weakly identifiable.
+Identifiability is a formal concept from the theory of signal processing, which is applied routinely to engineering problems \citep{e5be7c83a0d24500826f6e1b414d1733}.
+It refers to whether it is theoretically possible to infer unique and accurate parameter values, given the measured data and the model structure .
+Sources of non-identifiability are generally categorised as either ``structural'' , arising from the structure of the model (e.g.
+its invertibility), or ``practical'', arising from insufficient data or measurement errors \citep{GUILLAUME2019418}.
+Regarding $\iota$ and $h_0$, the source is structural.
+Equations \eqref{eq:hphx} and \eqref{eq:hphx2} contain a weak degeneracy between $\iota$ and $h_0$.
+For example, the cross polarisation strain $h_{\times}$ involves a trade-off between $\iota$ and $h_0$, which is broken by knowing $h_+$ via Equation \eqref{eq:hphx}.
+However, the measured strain is a linear combination of $h_\times$ and $h_+$; the two polarizations are not measured independently, which makes it challenging to separate $\iota$ and $h_0$.
+In fact $\iota$ and $h_0$ are structurally identifiable --- but only weakly.
+That is, the $\iota$-$h_0$ likelihood contours plateau to a ridge with a small, non-zero gradient close to the maximum likelihood solution.
+The likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation.
+Whilst a single, unique maximum does exists (labelled by the red dot in the figure), it lies along an S-shaped ridge where the likelihood values are all similar.
+Consequently it is difficult for the nested sampling algorithm to settle on the red dot; it trades off accuracy in one parameter against the other, depending on the particular realisation of the noise.
+By way of verification, if we fix one parameter at its true value (i.e.
+set a delta function prior) during the inference analysis, the posteriors for the other parameter overlap indistinguishably  for different noise realisations, i.e.
 $W_1 \to 0$.
-In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" \footnote{\tiny \textcolor{red}{TK: "washes out" may be too colloquial / unspecific, but I think it is intuitive...}\normalsize}the likelihood ridge, leading to a more concave likelihood surface.
-This enables consistent posteriors to be estimated for $\iota$ and $h$.
-In the case where we fix e.g $h_0$ at its true value to infer indistinguishable posteriors for $\iota$, all of the posteriors for $\iota$ are all strongly biased away from the injected value; the maximum a posteriori probability estimates underestimate the injected value by $\sim 0.34$ rad.
-This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be now be discussed further in Section \ref{sec:bias}.
-\subsection{Bias by dropping pulsar terms}\label{sec:bias}
+
+\subsection{Dropping the pulsar terms}\label{sec:bias}
 
@@
 	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
-	\caption{As Figure \ref{fig:corner_plot_1} but for the high-SNR system with $h_0 = 10^{-12}$.
-For the majority of parameters the injection value does not fall within the 90\% credible interval and is not visible over these scales.
-The estimated posteriors are biased away from the injected value due to the dropping of the pulsar terms.
-Note that the scale of the $x$-axes are generally highly constrained compared to Figure \ref{fig:corner_plot_1}.}
+	\caption{Same as Figure \ref{fig:corner_plot_1} but for a high-SNR system with $h_0 = 10^{-12}$.
+For five out of seven parameters the true, injected value does not fall within the 90\% credible interval and falls outside the plotted domain.
+The estimated posteriors are biased away from the injected value due to dropping the pulsar terms.
+Note that the plotted domain is narrower than in Figure \ref{fig:corner_plot_1}.}
 	\label{fig:bias_for_large_h}
@@
 	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
-	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and the Earth terms model for Earth terms only synthetic data (green line), with respect to the inclination $\iota$, over the prior space.
+	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter model (green line).
 All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}.
-The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model; the green/orange dashed lines are overlaid on this scale.
-For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$).
-For the pulsar-terms model the location of the maxima and injected value do coincide.
-10 noise realisations are plotted for each situation but the solutions are completely overlaid on this scale.}
+The coloured dashed vertical lines mark the maximum for the corresponding model; the green and orange dashed lines overlap at this resolution.
+For the Earth-terms model the likelihood maximum ($\iota \approx 0.7$) does not coincide with the injected value ($\iota = 1.0$).
+For the pulsar-terms model the maximum and injected value do coincide.
+10 noise realisations are plotted for each situation but overlap at this resolution; they are indistinguishable to the eye.}
 	\label{fig:likelihood_surface_iota}
 \end{figure}
-Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g.
-the median of the one-dimensional marginalised posterior) is biased away from the true injection value.
-In Figure \ref{fig:bias_for_large_h} we show the resulting posteriors for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but now for the high-SNR system with $h_0 = 10^{-12}$.
-We can see that for all parameters, with the exception of $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it is not visible on the scales over which we have plotted the posteriors.
-This effect is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$.
+
+Sections \ref{sec:rep_example} and \ref{sec:parameter_space} present preliminary evidence that the inferred value of ${\boldsymbol{\theta}}_{\rm gw}$ (e.g.
+the medians of the one-dimensional marginalised posteriors) is biased away from the true injected value.
+Figure \ref{fig:bias_for_large_h} displays a corner plot for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but for high SNR $(h_0 = 10^{-12})$.
+Except for $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it falls outside the plotted domain.
+The deviation is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$.
 Similar results are obtained for other noise realisations.
-Whilst $h_0$ does not exhibit any noticeable bias for this particular noise realisation, in general across multiple noise realisations it does exhibit a correlated bias with $\iota$.
-In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}.
-\newline
-
-In Figure \ref{fig:likelihood_surface_iota} we plot the variation in the log-likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, as we vary $\iota$ across the prior space, holding all other parameters constant at the true injected value of the representative example system, Table \ref{tab:parameters_and_priors}.
-We do this for 10 separate realisations of the system noise.
-We consider three separate situations: \textit{(i)} The solid blue line is the likelihood curve with just the Earth-terms included (i.e.
-using Equation \ref{eq:measuremen_earth} in the Kalman filter), \textit{(ii)} the solid orange line is the likelihood curve with the additional pulsar terms included (i.e.
-using Equation \ref{eq:measurement}), \textit{(iii)} the solid green line is the solution where we drop the pulsar terms from both the synthetic data and the measurement equation.
-We have not used nested sampling for situations \textit{(ii)} or \textit{(iii)} in this work.
-However, we are able to run the Kalman filter either using the pulsar terms in the model or else dropping them from the synthetic data consider how the resulting likelihood varies for different parameter values.
-The dashed lines show the location of the likelihood maxima for each of the likelihood solutions; finding the location of the maxima across all static parameters is effectively the goal of any likelihood based inference method, such as nested sampling.
-\newline
-
-The key observation from Figure \ref{fig:likelihood_surface_iota} is that for the Earth-terms only solution the location of the likelihood optima and the location of the injected value are not the same.
-Consequently Bayesian likelihood inference techniques settle in optima that do not necessarily align with the true injection value.
-This is the origin of the bias in the one-dimensional marginalised posteriors that we observe.
-For the Earth-terms solution the optima of the likelihood curve lies around $\iota \sim 0.7$ rad, whereas the injected value was $\iota = 1.0$ rad.
-This optima is comparable to the median of the posterior inferred for $\iota$ in Figure \ref{fig:bias_for_large_h}.
-\newline
-
-In contrast to the Earth-terms solution, for the pulsar-terms solution the optima of the likelihood curve now aligns with the true injected value.
-The corollary is also true; if we drop the pulsar terms from the synthetic data and use the Earth terms equation for inference then the optima is also correctly located.
-This evidences that it is the effect of dropping the pulsar terms from the measurement equation which causes this bias in the inferred posteriors.
-Similar results are observed for the other parameters which exhibit a bias in the nested sampling inference.
-\newline
-
-Why does dropping the pulsar terms bias these parameters in particular, whilst leaving others unaffected?
-For instance, why is the bias induced in $\iota$ so strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
-When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is effective modulated by the interference of two equal-amplitude, phase shifted waves (c.f.
-Equation \ref{eq:measurement}).
-Explicitly, we have the ``Earth-terms wave"	 with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$ and the ``pulsar-terms wave" with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
-The net interference wave that modulates $f_{\rm p}$ is
-\begin{eqnarray}
-	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ .
-\end{eqnarray}
-The additional modulations from $y_{\rm pulsar}$ influence the amplitude of the resulting interference wave, but do not affect its frequency.
-When we model this data using the Kalman filter without the pulsar terms, the additional amplitude modulations from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
-As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude of the single wave.
-Consequently the parameters which are most affected are those which determine the amplitude of the GW.
-Terms which influence the amplitude strongly, such as $h_0$ and $\iota$, are strongly affected and exhibit a strong bias.
-Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
-The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
-Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
-This hierarchy of biases for each parameter is also what we observe in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
-\newline
-
-In order to ascertain how important the bias is for PTA continuous wave searches requires a thorough exploration of the astrophysical parameter space which we do not carry out in this work.
-The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
-If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate and the bias may not be important for searches.
-This is what we observe for our synthetic data (e.g.
-Figure \ref{fig:corner_plot_2}); the decreased SNR manifests as a broader marginalised posterior which generally overwhelms the shift due to the bias.
-Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may come to systematically affect the inferred parameters.
-\newline
-
-We have shown in this section how the dropping of the pulsar terms results in a bias in the inferred parameters, not just in the sky localisation parameters  \citep[e.g.][]{Zhupulsarterms,Chen2022} but additionally all parameters which influence the wave amplitude.
-Whilst the bias is generally small and does not prevent the recovery of accurate posteriors for our example system, it is present and may need to be accounted for for analysis in other parts of the parameter space.
-This bias is not particular to our method but a shared feature across likelihood based methods that do not include the pulsar terms in the inference model.
-The inclusion of the pulsar terms in the methods presented in this work in order to alleviate the bias will be a key goal of a future work.
+Whilst $h_0$ does not exhibit any noticeable bias for the particular noise realisation generating Figure \ref{fig:bias_for_large_h}, in general it does exhibit a correlated bias with $\iota$ across multiple noise realisations, as noted in the bottom left panel of Figure \ref{fig:just_iota_h}.
+In this section we demonstrate how the bias stems from dropping the pulsar terms described in Section \ref{sec:parameter_estim}.
+\newline
+
+In Figure \ref{fig:likelihood_surface_iota} we plot the log-likelihood returned by the Kalman filter, viz.
+Equation \eqref{eq:likelihood}, as a function of $\iota$ (i.e.
+$\log \mathcal{L}(\iota)$) across the prior domain, holding all other parameters constant at the true injected value of the representative system in Table \ref{tab:parameters_and_priors}.
+We generate 10 $\log \mathcal{L}(\iota)$ solutions corresponding to 10 separate realisations of the system noise.
+The solid blue curves only includes the Earth terms (i.e.
+using Equation \eqref{eq:measuremen_earth} in the Kalman filter).
+The solid orange curves include the Earth and pulsar terms (i.e.
+using Equation \eqref{eq:measurement}).
+The solid green curve corresponds to a control experiment, where we drop the pulsar terms from both the synthetic data and the measurement equation.
+The vertical dashed lines mark the maxima, which nested sampling seeks.
+\newline
+
+The key observation from Figure \ref{fig:likelihood_surface_iota} is that the location of the maximum ($\iota \approx 0.7 \, {\rm rad}$) does not coincide with the true, injected value ($\iota = 1.0 \, {\rm rad}$), when only the Earth terms are included.
+This is the origin of the bias in the one-dimensional marginalised posteriors.
+In contrast, the maximum aligns with the true, injected value, when the pulsar terms are included.
+As a cross-check, agreement is also achieved, if we drop the pulsar terms from the synthetic data (which is not the situation in reality, of course) and use the Earth terms only for inference.
+Similar results are obtained for other parameters, e.g.
+$\delta$.
+A discussion on why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches is presented in Appendix \ref{sec:bias_appendix}.
+\newline
+
+We show in this section that dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just in the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}.
+The bias is not particular to our method; it is shared by likelihood based methods that do not include the pulsar terms in the inference model.
+Generalizing the Kalman tracker and nested sampler to correct for the bias will be a key goal of a future paper.
 
@@
 
-\textcolor{red}{TK: Lets also pre-empt reviewer and make it explicit why we are using frequencies rather than TOAs. Need to chat with group to get this clear in my head.}
-
-%https://arxiv.org/pdf/2107.03047.pdf
-\section{Conclusion}\label{sec:conclusion}
-In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
+To conclude, in this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs.
 We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics.
@@
 }
-\section{Identifiability}\label{appendix_identifiability}
-\textcolor{red}{TK TBD. An example calculation for a simple system using the methods of e.g. } \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
 
@@
 This problem of poor global convergence can be solved in principle by increasing the number of live points $n_{\rm live}$ in the nested sampler so as to better cover the likelihood surface and prevent falling into local minima.
+
+\section{Bias from dropping the pulsar terms}\label{sec:bias_appendix}
+Why does dropping the pulsar terms bias the above parameters in particular, whilst leaving others nearly unaffected?
+For instance, why is the bias in $\iota$ so strong, whereas $\psi$ shows only a weak bias and  $\Omega$ shows no bias at all?
+When we generate synthetic data including the pulsar term, the state frequency of an individual pulsar $f_{\rm p}$ is effective modulated by the interference of two equal-amplitude, phase shifted waves (c.f.
+Equation \eqref{eq:measurement}).
+Explicitly we have an
+``Earth-terms wave'' with a general form $y_{\rm Earth} = A \cos \xi$ for amplitude $A$ and phase $\xi$, and the ``pulsar-terms wave'' with a general form $y_{\rm pulsar} = A \cos (\xi + \kappa)$ for phase offset $\kappa$.
+These two terms are related to the measurement equation, Equation \eqref{eq:g_func_trig} as,
+\begin{eqnarray}
+	1 - g^{(n)}(t) =  y_{\rm Earth}  - y_{\rm pulsar} \ .
+\end{eqnarray}
+The net interference wave that modulates $f_{\rm p}$ is 
+\begin{eqnarray}
+	y_{\rm Earth} - y_{\rm pulsar} = 2 A \sin \left(\frac{\kappa}{2}\right) \sin \left(\xi + \frac{\kappa}{2}\right) \ .
+\end{eqnarray}
+The inclusion of  $y_{\rm pulsar}$ influences the amplitude of the resulting interference wave by a factor $ 2 \sin (\kappa/2)$, but does but do not affect its frequency.
+When we model this data using the Kalman filter without the pulsar terms, the additional factor from $y_{\rm pulsar}$ must be described by the single wave model, $y_{\rm Earth}$.
+As a result the $y_{\rm Earth}$ model must try to compensate for these amplitude modulations by adjusting the parameters which govern the amplitude.
+Terms which influence the amplitude strongly, such as $h_0$ and $\iota$, are strongly affected and exhibit a strong bias.
+Terms which influence the amplitude weakly such as $\psi$ exhibit a weaker bias.
+The wave amplitude is independent of the frequency and so $\Omega$ exhibits no bias.
+Since a shift in phase affects the instantaneous amplitude of the wave, $\Phi_0$ also exhibits a small bias, despite $A$ not being a function of $\Phi_0$.
+The same hierarchy of biases is seen also in Figure \ref{fig:parameter_space}, where $\alpha$ and $\delta$ are strongly over-constrained whilst $\Omega$ is well-estimated.
+\newline
+
+In order to ascertain how important the bias is for PTA continuous wave searches requires a thorough exploration of the astrophysical parameter space which we do not carry out in this work.
+The magnitude of the bias will vary for different points in the parameter space, as well as for different PTA configurations.
+Quantifying the bias in preparation for a practical PTA analysis with real data lies outside the scope of this introductory paper.
+If the bias is sufficiently modest then for quiet continuous wave sources with low SNR the uncertainty in the one-dimensional marginalised posterior may dominate.
+This is what we observe for our synthetic data (e.g.
+Figure \ref{fig:corner_plot_2}); the marginalised posterior broadens as the SNR decreases, until it overwhelms the shift due to the bias.
+Conversely if the bias is sufficiently strong, or the source sufficiently loud with low uncertainties, then the bias may systematically affect the inferred parameters.
+\newline
 

### 2024-01-24 manuscript/manuscript_2/paper.tex
Final manuscript version

@@
 The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise.
-It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source)
-
-	 which also arise in traditional matched-filter analyses that exclude the pulsar terms.
-
- Including the pulsar terms also decreases the root-mean-square relative error in the estimated one-dimensional posteriors \textcolor{red}{by up to 99\%}, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$.
-For a quieter source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
+It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) which also arise in traditional matched-filter analyses that exclude the pulsar terms.
+Additionally, including the pulsar terms decreases the minimum detectable strain by $14\%$.
 \end{abstract}
@@
 			& $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
-			& $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
+			& $\delta$              & $1.0$  & rad & Cosine($-\pi/2, \pi/2$) \\
 			& $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
@@
 			& $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
-			& $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
+			& $\iota$             & $1.0$ & rad & Sin($0, \pi$) \\ 
 			\hline
@@
 			\toprule
-			&$\theta$ & $\Delta_{\rm Earth}(\theta)$ & $\Delta_{\rm psr \& earth}(\theta)$  \\
+			&$\theta$ & $\Delta_{\rm Earth}(\theta)$ & $\Delta_{\rm psr \& Earth}(\theta)$  \\
 			\hline
@@
 For the worst estimated parameter $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases.
-Including the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) by 61\% for $\Omega$, up to 99\% for $\psi$ and $\alpha$ for a representative SMBHB system with a high SNR ($h_0 = 10^{-12}$).
-For low SNR ($h_0 = 5 \times 10^{-15}$), including the pulsar terms lowers the minimum detectable strain by $14\%$, comparable with standard PTA analyses \citep[e.g.][]]{Zhupulsarterms} \newline
+Including the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) for five out of the seven static parameters in $\boldsymbol{\theta}_{\rm gw}$; we find $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for all parameters except $\psi$ and $h_0$.
+In the high-SNR case ($h_0 = 1 \times 10^{-12}$) all of the seven static parameters are estimated more accurately; $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$.
+Including the pulsar terms lowers the minimum detectable strain by $14\%$, comparable with standard PTA analyses \citep[e.g.][]{Zhupulsarterms}.
+\newline
 
@@
 \section{Kalman filter} \label{sec:kalman}
-The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$.
-It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
-In this appendix, we describe the Kalman filter algorithm used in this paper.
-General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the formula for the likelihood.
+The Kalman filter \citep{Kalman1} is an algorithmic optimal solver for Gauss-Markov processes.
+Given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$ the Kalman filter recovers a temporal sequence of stochastically evolving system state variables, $\boldsymbol{X}(t)$, which are hidden from the observer.
+It is a standard method in control theory \citep{zarchan2000fundamentals,rob_KF_book,10.5555/3103280} and finds common use in engineering applications \citep[e.g.][]{KFexample1,KFexample2,KFexample3,KFexample4} as well as successful application to neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
+In this appendix, we outline the Kalman filter used in this paper.
+The general recursion relations for the discrete-time Kalman filter are presented for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the formula for the likelihood.
 The application of the Kalman filter algorithm to the specific continuous-time state-space model in Section \ref{sec:2} is outlined in Section \ref{sec_kalman_specific}.
 
-\subsection{Recursion equations and likelihoood }\label{sec_kalman_general}
+\subsection{Recursion equations and likelihood }\label{sec_kalman_general}
 In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement).
@@
 		\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
-		\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
+		\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value through Equation \eqref{eq:likelihood}
 	\end{enumerate}
@@
 \end{enumerate}
-In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$.
+In order to compute the odds ratio $\beta$ the above workflow is repeated for a different $\mathcal{M}$.
 The resulting $\mathcal{Z}$ values can then be compared.
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
+	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha_v3}
 	\caption{Logarithm of the cross-section along the $\alpha$-axis of the likelihood function $\mathcal{L}_{\rm slice}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$.
@@
 In the right panel $\mathcal{L}_{\rm slice}(\alpha)$ is smooth for $0\leq\alpha\leq \pi$.
-}
+The green (blue) curve is the likelihood returned by $\mathcal{M}_{\rm psr\&Earth}$ ($\mathcal{M}_{\rm Earth}$).}
 	\label{fig:likelihood_comparison}

### 2024-01-24 py_src/priors.py

@@ -204,16 +204,21 @@ def set_prior_on_measurement_parameters(init_parameters,priors,measurement_model
 
 
         init_parameters["phi0_gw"] = None
-        priors["phi0_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'phi0_gw')
+        #priors["phi0_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'phi0_gw')
+        priors["phi0_gw"] = bilby.core.prior.Uniform(0.0, 2.0*np.pi, 'phi0_gw')
+
 
         init_parameters["psi_gw"] = None
         priors["psi_gw"] = bilby.core.prior.Uniform(0.0, np.pi, 'psi_gw')
 
         init_parameters["iota_gw"] = None
-        priors["iota_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'iota_gw')
+        #priors["iota_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2.0, 'iota_gw')
+        priors["iota_gw"] = bilby.core.prior.Sin(0.0, np.pi, 'iota_gw')
+
 
         init_parameters["delta_gw"] = None
-        priors["delta_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2, 'delta_gw')
+        #priors["delta_gw"] = bilby.core.prior.Uniform(0.0, np.pi/2, 'delta_gw')
+        priors["delta_gw"] = bilby.core.prior.Cosine(-np.pi/2, np.pi/2, 'delta_gw')
 
 
         init_parameters["alpha_gw"] = None

### 2024-01-24 manuscript/manuscript_2/paper.tex
actual version passed to AM

@@
 \end{equation}
-In Equation \eqref{eq:mean_rel_error} $\theta_{k,\rm M}$ is the $k$-th sample of the one-dimensional posterior for parameter $\theta$ returned by nested sampling, and the subscript $\rm M \in \{ \text{Earth, psr \& Earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+In Equation \eqref{eq:mean_rel_error} $\theta_{k,\rm M}$ is the $k$-th sample of the one-dimensional posterior for parameter $\theta$ returned by nested sampling, and the subscript $\rm M \in \{ \text{Earth, psr\&Earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
 The total number of samples is $K$ and $\theta_{\rm inj}$ is the true injection value.
@@
 $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$.
-In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models, with $\Delta_{\rm Earth}(\Omega)-\Delta_{\rm psr}(\Omega) = 6.7 \times 10^{-3}$.
+In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models, with $\Delta_{\rm Earth}(\Omega)-\Delta_{\rm psr \& Earth}(\Omega) = 6.7 \times 10^{-3}$.
 The improvement from including the pulsar terms is largest for $\iota$, with $\Delta_{\rm Earth}(\iota) - \Delta_{\rm psr \& Earth}(\iota)  = 0.24$.
@@
 A full parameter sweep is postponed to future work, after the analysis scheme in this paper is upgraded from ingesting pulse frequencies to pulse TOAs to enable a like-for-like comparison with standard PTA analyses.
-In the interim, we note that a $13\%$ improvement in sensitivity when including the pulsar terms is comparable to the improvements of $5\%$ achieved by \cite{Zhupulsarterms}.
+In the interim, we note that a $14\%$ improvement in sensitivity when including the pulsar terms is comparable to the improvements of $5\%$ achieved by \cite{Zhupulsarterms}.
 

### 2024-02-09 manuscript/manuscript_2/paper.tex
small edits before sub

@@
 \newcommand{\bibtex}{\textsc{Bib}\!\TeX} % bibtex. Not quite the correct typesetting, but close enough
+
+\DeclareMathOperator*{\argmax}{argmax} % thin space, limits underneath in displays
 
@@
 	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.
-The data analysis procedure can be formulated within a state-space framework, wherein achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source.
+The data analysis procedure can be formulated within a time-domain, state-space framework, wherein achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source.
 The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array.
-Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms.
+Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms; that is, we copy the pulsar terms from traditional, non-state-space analyses over to the state-space framework.
 The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise.
-It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) which also arise in traditional matched-filter analyses that exclude the pulsar terms.
+It is shown that including the pulsar terms corrects for previously identified biases in the parameter estimates (especially the sky position of the source) which also arise in traditional matched-filter analyses that exclude the pulsar terms.
 Additionally, including the pulsar terms decreases the minimum detectable strain by $14\%$.
+Overall, the study verifies that the pulsar terms do not raise any special extra impediments for the state-space framework, beyond those studied in traditional analyses.
 \end{abstract}
@@
 
-The modulation in a pulsar's TOAs due to a GW has two contributions: an ``Earth term" and ``pulsar terms" which are proportional to the GW strain at the Earth and every pulsar respectively.
+The modulation in a pulsar's TOAs due to a GW has two contributions: an ``Earth term'' and ``pulsar terms'' which are proportional to the GW strain at the Earth and every pulsar respectively.
 Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases.
-Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many --- although not all --- standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
 It is acknowledged in standard analyses that dropping the pulsar terms leads to biases in the inferred parameters and reduces the detection probability, but both impacts are modest \citep{Zhupulsarterms,Chen2022,KimpsonPTA}.
@@
 \cite{KimpsonPTA} (K24 hereafter) introduced a state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, which self-consistently tracks the intrinsic, achromatic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations.
-The method described in K24 complements standard PTA analyses by tracking the specific, measured, time-ordered realization of the TOA fluctuations instead of fitting for their ensemble-averaged statistics (such as their power spectral density).
+The method described in K24 complements standard PTA analyses by tracking the specific, measured, time-ordered realization of the TOA fluctuations instead of fitting for their ensemble-averaged statistics (such as their power spectral density), following the customary signal processing approach in many industrial and scientific electrical engineering applications.
 The method described in K24 also follows the example of many standard PTA analyses and drops the pulsar terms.
@@
 We emphasize that the biases incurred by omitting the pulsar terms have been studied thoroughly in the context of standard PTA analyses \citep{Zhupulsarterms,Chen2022}; they are not new effects discovered here for the first time.
-Rather, the goal of this paper is to study them again in the context of the promising state-space formulation of the PTA analysis problem introduced by K24.
+Rather, the goal of this paper is to study them again in the context of the promising state-space formulation of the PTA analysis problem introduced by K24, to check whether they raise any special extra impediments beyond those studied in standard PTA analyses.
 The state-space formulation complements standard analysis techniques, e.g.\ based on matched filters \citep{PhysRevD.79.084030,PhysRevD.91.044048}.
@@
 \end{equation}
-where $f_{\rm em}^{(n)}$ is the deterministic evolution, an overdot denotes a derivative with respect to $t$, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
+where $f_{\rm em}^{(n)}$ is the deterministic component of the evolution, an overdot denotes a derivative with respect to $t$, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
 \begin{align}
@@
 In Equation \eqref{eq:xieqn}, $[\sigma^{(n)}]^2$ parametrizes the noise amplitude and produces characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}.
-The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking for the sake of definiteness, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
+The deterministic evolution $f_{\rm em}^{(n)}(t)$ is attributed to magnetic dipole braking for the sake of definiteness, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}.
 PTAs are typically composed of millisecond pulsars (MSPs), for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$.
@@
 \begin{equation}
-	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
+	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1) (t - t_1) \ , \label{eq:spinevol}
 \end{equation} 
@@
 However, they are highly idealised \citep{Meyers2021,Myers2021MNRAS.502.3113M,2023MNRAS.520.2813A,Vargas}.
-Idealisations include (i) the exclusion of physics, which is likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the whitenes of $\xi^{(n)}(t)$ for MSPs in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$, noting that $\xi^{(n)}(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.
+Idealisations include (i) the exclusion of physics, which is likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the whiteness of $\xi^{(n)}(t)$ for MSPs in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$, noting that $\xi^{(n)}(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.
 
@@
 	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
-	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:vareps}
+	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \left[\sigma_{\rm m}^{(n)}\right]^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:vareps}
 \end{align}
-In Equation \eqref{eq:vareps}, $\sigma_{\rm m}^2$ is the variance of the measurement noise at the telescope and is shared between all pulsars.
+In Equation \eqref{eq:vareps}, $[\sigma_{\rm m}^{(n)}]^2$ is the variance of the measurement noise at the telescope.
 In Equation \eqref{eq:measurement} the measurement function $g^{(n)}(t)$ is given by \citep[e.g.][]{Maggiore}
@@
 The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}.
-We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
+We are therefore justified in treating the GW source as monochromatic as a first approximation \citep{Sesana10,Sesana2010,Ellis2012ApJ}.
 \newline
@@
 The complete set of $7 + 5N$ static parameters is denoted by $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$.
-While we have no prior information about $\boldsymbol{\theta}_{\rm gw}$, there are constraints on $\boldsymbol{\theta}_{\rm psr}$ from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ typically \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
+While we assume no prior information about $\boldsymbol{\theta}_{\rm gw}$, there are constraints on $\boldsymbol{\theta}_{\rm psr}$ from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ typically \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
 
@@
 \begin{equation}
-	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
+	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ , \label{eq:posterior_distrib}
 \end{equation}
@@
 \end{equation}
-In the paper, the measurements are ${\boldsymbol{Y}} = \{ f^{(n)}_{\rm m}(t_1), \dots, f^{(n)}_{\rm m}(T_{\rm obs})  \}_{1 \leq n \leq N}$.
+The measurements are ${\boldsymbol{Y}} = \{ f^{(n)}_{\rm m}(t_1), \dots, f^{(n)}_{\rm m}(T_{\rm obs})  \}_{1 \leq n \leq N}$.
 Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is a function of $\boldsymbol{\theta}$.
@@
 Appendix \ref{sec:workflow} summarizes the workflow of the combined Kalman filter and nested sampler.
-Appendices \ref{sec:kalman}--\ref{sec:workflow} are designed to equip the interested reader with the requisite tools to reproduce the key results in Sections \ref{sec:pe_and_ms} and \ref{sec:earth_vs_psr}.
-\newline
-
-The inference procedure in K24 drops the pulsar term on the last line of Equation \eqref{eq:g_func_trig}, in keeping with other PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+Appendices \ref{sec:kalman}--\ref{sec:workflow} are designed to equip the interested reader to reproduce the key results in Sections \ref{sec:pe_and_ms} and \ref{sec:earth_vs_psr}.
+\newline
+
+The inference procedure in K24 drops the pulsar term on the last line of Equation \eqref{eq:g_func_trig}, in keeping with several other PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
 In this section we show how to remove this limitation, following previous authors \citep{Zhupulsarterms,Chen2022}.
@@
 The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties greater than the typical GW wavelength.
-Consequently the pulsar term is often dropped in standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
+Consequently the pulsar term is often --- although not always --- dropped in standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
 Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space; the $N$ values of $d^{(n)}$ are not inferred.
@@
 
-Likewise, $\sigma_{\rm m}$ can be related to the TOA uncertainty, $\sigma_{\rm TOA}$, by 
-\begin{equation}
-	\sigma_{\rm m} \approx f_{\rm p}^{(n)}(t_1) \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
-\end{equation}
-For an MSP with $f_{\rm p}^{(n)} \sim 0.1$ kHz, $T_{\rm cad} = 1 \, {\rm week}$, and $\sigma_{\rm TOA} \sim 1 \mu$s,  Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz.
-The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz.
-Here, for the sake of definiteness, we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred.
+In a similar vein, $\sigma_{\rm m}^{(n)}$ can be related to, $\sigma_{\rm TOA}^{(n)}$, by 
+\begin{equation}
+	\sigma_{\rm m}^{(n)} \approx f_{\rm p}^{(n)}(t_1) \sigma_{\rm TOA}^{(n)} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
+\end{equation}
+For an MSP with $f_{\rm p}^{(n)} \sim 0.1$ kHz, $T_{\rm cad} = 1 \, {\rm week}$, and $\sigma_{\rm TOA}^{(n)} \sim 1 \mu$s,  Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m}^{(n)} \sim 10^{-10}$ Hz.
+The most accurately timed pulsars have $\sigma_{\rm TOA}^{(n)} \sim 10 $ ns and $\sigma_{\rm m}^{(n)} \sim 10^{-12}$ Hz.
+In this paper, for simplicity and the sake of definiteness, we fix $\sigma_{\rm m}^{(n)} = 10^{-11}$ Hz for all $n$, and take it as known \textit{a priori} rather than a parameter to be inferred.
 When analysing real data this assumption is easily relaxed.
-Although $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar.
+Although $\sigma_{\rm m}^{(n)}$ is assumed to be the same for every pulsar, $f_{\rm m} ^{(n)}$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar.
 \newline
@@
 cosmic variance).
-The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1}.
+The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.
+The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are specified in Section \ref{sec:rep_example1}.
 All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.
@@
 
-In Section \ref{sec:rep_smbh_source} we apply the analysis scheme to synthetic data and estimate $\boldsymbol{\theta}_{\rm gw}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e.
+In Section \ref{sec:rep_smbh_source} we apply the analysis scheme to synthetic data and start by estimating $\boldsymbol{\theta}_{\rm gw}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e.
 the SMBHB system.
-The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible cosmic variance in the parameter estimates.
+The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme multiple times and quantify the  irreducible cosmic variance in the parameter estimates.
 In Section \ref{sec:chi_estim} we inspect and verify the estimates of  $\chi^{(n)}$.
@@
 
-Figure \ref{fig:corner_plot_1} displays the posterior distribution of $\boldsymbol{\theta}_{\rm gw}$ for ten noise realisations in the form of a traditional corner plot for the representative SMBHB source parametrized as per the top portion of Table \ref{tab:parameters_and_priors}.
+Figure \ref{fig:corner_plot_1} displays the posterior distribution of $\boldsymbol{\theta}_{\rm gw}$ for ten noise realisations in the form of a traditional corner plot for the representative SMBHB source parametrized in the top portion of Table \ref{tab:parameters_and_priors}.
 The histograms are one-dimensional posteriors, marginalized over the six other parameters.
@@
 Indeed, the injected value falls within the 90\% credible interval for $\Omega, \Phi_0, \psi$, and  $\iota$ 40 out of 40 times.
-There is appreciable natural dispersion in the one-dimensional posterior medians between realisations.
-We quantify the degree of dispersion using the coefficient of variation,
+\newline
+
+There is appreciable natural dispersion in the one-dimensional posterior medians between realisations. We quantify the degree of dispersion using the coefficient of variation,
 \begin{eqnarray}
-	CV = \frac{\mu}{\sigma}
+	CV = \mu_{\rm med}/\sigma_{\rm med}
 \end{eqnarray}
- where $\mu$ is the mean value of the 10 posterior medians and $\sigma$ is the standard deviation.
-The minimum and maximum $CV$ is 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively.
-The remaining parameters have typical $CV$ $\lesssim 10 \%$.
+ where $\mu_{\rm med}$ is the mean value of the 10 posterior medians and $\sigma_{\rm med}$ is their standard deviation.
+The minimum and maximum $CV$ are 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively.
+The remaining parameters typically satisfy $CV$ $\lesssim 10 \%$.
 Analogous results for synthetic data with higher strain ($h_0 = 1 \times 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}) for completeness.
@@
 It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach, as explained in Section \ref{sec:pulsar_term2}.
-In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$.
+In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$., except under special circumstances.
 Figure \ref{fig:corner_plot_3} displays the results as a corner plot for the representative subset $\chi^{(1)}, \dots, \chi^{(5)}$ over the ten noise realisations, in the same style as Figure \ref{fig:corner_plot_1}.
@@
 Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$ and $\cos \chi^{(2)} \approx 0.07$.
-The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR).
-For example, for for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered unambiguously; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
+The ability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR).
+For example, for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered unambiguously; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}.
 \newline
@@
 	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR_n4000}
-	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for ten realisations of the noise processes, with curves coloured uniquely per realisation.
+	\caption[Caption of LOF]{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for 10 realisations of the noise processes, with curves coloured uniquely per realisation.
 The horizontal and vertical orange lines indicate the true injected values.
@@
 The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter.
-The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation.
+The supertitles of the one-dimensional histograms record the median and the 0.16 and 0.84 quantiles of the posterior median realisation (in this context the median realisation is defined as follows: a posterior is generated for all 10 realisations; the medians of the posteriors are ranked in ascending order; the median of the ranked list is defined to correspond to the median realisation).
 We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$.
@@
 The known, injected value lies within the 90\% credible interval for 60 out of the 70 combinations of seven parameters and ten noise realizations.
-There is appreciable dispersion among the peaks of the one-dimensional posteriors: the maximum coefficient of variation is $52 \%$ for $\Phi_0$, and the minimum is $0.02 \%$ for $\Omega$.}
+There is appreciable dispersion among the peaks of the one-dimensional posteriors: the maximum coefficient of variation is $52 \%$ for $\Phi_0$, and the minimum is $0.2 \%$ for $\Omega$.}
 	\label{fig:corner_plot_1}
@@
 	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi_n4000}
-	\caption{Same as Figure \ref{fig:corner_plot_1}	but for the static parameters of $\chi^{(1)}, \dots, \chi^{(5)}$ (the remaining pulsar phases $\chi^{(6)}, \dots, \chi^{(N)}$ are omitted for readability).
+	\caption{Same as Figure \ref{fig:corner_plot_1}	but for the static parameters $\chi^{(1)}, \dots, \chi^{(5)}$ (the remaining pulsar phases $\chi^{(6)}, \dots, \chi^{(N)}$ are omitted for readability).
 Consistent unimodal posteriors are obtained across the ten noise realisations for four out of the five displayed parameters.
@@
 The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.
-(b) Same as (a) but now assuming that $\chi^{(1)}, \dots, \chi^{(47)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior).
-The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, lie within the shaded region.
+(b) Same as (a) but now assuming that $\chi^{(1)}, \dots, \chi^{(47)}$ are known exactly \textit{a priori} (i.e.
+with a delta-function prior).
+The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, lie wholly within the shaded region.
 } \label{fig:AB}
@@
 For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$.
-The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1} and Table  \ref{tab:parameters_and_priors}.
+The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are specified in Section \ref{sec:rep_example1} and Table \ref{tab:parameters_and_priors}.
 \newline
@@
 More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases.
-For the worst estimated parameter $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases.
+For the worst estimated parameter $\Phi_0$, the injection is contained within the 90\% credible interval in 81\% of cases; the next paragraph explains why 81\% is less than the theoretical ideal 90\%.
 \newline
@@
 Manually inspecting the individual corner plots for each of the 200 injections, we see that the nested sampler does not return a unimodal posterior in the rare event that the source is located unfavourably, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} \approx -1$, such that one cannot infer $\chi^{(n)}$ accurately (see Section \ref{sec:chi_estim}).
-All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$ and $n_{\rm live}$.
+All injected sources are ``observed'' with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$ and $n_{\rm live}$.
 In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but assuming that the injected values of $\chi^{(n)}$ are known exactly for the sake of testing, i.e.
 setting a delta-function prior on $\chi^{(n)}$.
-The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$.
+The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of estimating $\chi^{(n)}$ accurately.
 For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}.
@@
 Model $\mathcal{M}_{\rm psr \& Earth}$ is parameterised by $\boldsymbol{\theta}'$.
-Model $\mathcal{M}_{\rm Earth}$ is parameterised by $\boldsymbol{\theta}_{\rm Earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$.
+Model $\mathcal{M}_{\rm Earth}$ is parameterised by $\boldsymbol{\theta}_{\rm Earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}''_{\rm psr}$ where $\boldsymbol{\theta}''_{\rm psr}$ equals $\boldsymbol{\theta}'_{\rm psr}$ with $\chi^{(n)}$ removed.
 Both models are applied to identical realisations of the data, generated using the procedure in Section  \ref{sec:rep_example1}.
@@
 
-In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm Earth}$ returned by the respective models.
+In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$ and $\boldsymbol{\theta}_{\rm Earth}$ returned by the respective models.
 In Section \ref{sec:psr_v_earth_bayes} we compare the minimum detectable GW strain for the two models by calculating the model evidence and comparing it to the evidence for a null model that does not contain a GW.
@@
 the system described in Table \ref{tab:parameters_and_priors}.
-The second is a ``high-SNR'' system which has the same static injection as in Table \ref{tab:parameters_and_priors} except for $h_0 = 1 \times 10^{-12}$.
-The ``high-SNR'' system is considered in order to evaluate any systematic biases, as distinct from random errors caused by the measurement noise.
+The second is a ``high-SNR'' system which has the same static parameters as in Table \ref{tab:parameters_and_priors} except with $h_0 = 1 \times 10^{-12}$.
+The ``high-SNR'' system is considered in order to quantify any systematic biases, as distinct from random errors caused by the measurement noise.
 In order to enable a clear comparison between the posterior probability distributions calculated using the two models, in this section we present a single noise realisation of the synthetic data $\boldsymbol{Y}$.
@@
 The results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_compare_high} for the high-SNR system and in Figure \ref{fig:corner_plot_compare_low} for the low-SNR system.
-The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, except that the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
+The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, except that the different coloured curves now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$.
 The blue curves are the results derived using $\mathcal{M}_{\rm Earth}$.
@@
 
-For the high-SNR case results in Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm Earth}$ are evidently biased, as was observed by K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
+For the high-SNR results in Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm Earth}$ are evidently biased, as was observed by K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
 Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
@@
 The green one-dimensional posteriors exhibit no bias and are generally symmetric about the injected value.
-Moreover, for every pair of parameters in ${\boldsymbol{\theta}}_{\rm gw}$ the injected values are contained within the 2-sigma contours for the $\mathcal{M}_{\rm psr \& Earth}$.
+Moreover, for every pair of parameters in ${\boldsymbol{\theta}}_{\rm gw}$ the injected values are contained within the 2-sigma contours for $\mathcal{M}_{\rm psr \& Earth}$.
 This is not true for $\mathcal{M}_{\rm Earth}$ where the injected values fall outside the 2-sigma contour for 15 out of the 21 parameter pairs.\newline
 
-We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value, viz.
-\begin{equation}
-\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{ \left[\theta_{k,\rm M} - \theta_{\rm inj}\right]^2}{\theta_{\rm inj}}} \ . \label{eq:mean_rel_error}
-\end{equation}
-In Equation \eqref{eq:mean_rel_error} $\theta_{k,\rm M}$ is the $k$-th sample of the one-dimensional posterior for parameter $\theta$ returned by nested sampling, and the subscript $\rm M \in \{ \text{Earth, psr\&Earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
-The total number of samples is $K$ and $\theta_{\rm inj}$ is the true injection value.
+We define a relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injected value, viz.
+\begin{eqnarray}
+\Delta_{\rm M}(\theta) = \frac{ \left| \underset{\theta}{\text{argmax }} p_{\rm M}\left(\theta| \boldsymbol{Y}\right) - \theta_{\rm inj}\right|}{\theta_{\rm inj}} \ . \label{eq:mean_rel_error}
+\end{eqnarray}
+In Equation \eqref{eq:mean_rel_error}, $p_{\rm M}\left(\theta| \boldsymbol{Y}\right)$ is the one-dimensional posterior for parameter $\theta$ returned by nested sampling (c.f.
+Equation \eqref{eq:posterior_distrib}).
+The subscript $\rm M \in \{ \text{Earth, psr\&Earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively.
+$\theta_{\rm inj}$ denotes the true injection value.
+We note that $\Delta_{\rm M}(\theta)$ quantifies the accuracy of the estimates returned by the two models, i.e.
+the closeness of the most probable estimate of $\theta$ relative to $\theta_{\rm inj}$.
+In contrast, $\Delta_{\rm M}(\theta)$ does not quantify the uncertainty in the estimates.
 The error $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} is summarised in Table \ref{tab:posterior_errors}.
@@
 $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$.
-In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models, with $\Delta_{\rm Earth}(\Omega)-\Delta_{\rm psr \& Earth}(\Omega) = 6.7 \times 10^{-3}$.
-The improvement from including the pulsar terms is largest for $\iota$, with $\Delta_{\rm Earth}(\iota) - \Delta_{\rm psr \& Earth}(\iota)  = 0.24$.
-Whilst we present only a single noise realisation in this section, the improvements from including the pulsar terms are generally of a comparable magnitude across different realisations.
+In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models, with $\Delta_{\rm Earth}(\Omega)-\Delta_{\rm psr \& Earth}(\Omega) = 1.5 \times 10^{-5}$.
+The improvement from including the pulsar terms is largest for $\iota$, with $\Delta_{\rm Earth}(\iota) - \Delta_{\rm psr \& Earth}(\iota)  = 0.14$.
+Whilst we present only a single noise realisation in this section, the improvements from including the pulsar terms are found to be comparable across different realisations.
 \newline
@@
 Qualitatively, the green and blue contours and histograms mostly overlap, although the green contours are centred slightly better on the injected values.
-The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR case is reported in the lower half of Table \ref{tab:posterior_errors}.
+The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR results is reported in the lower half of Table \ref{tab:posterior_errors}.
 We see that the error is larger for every element in $\boldsymbol{\theta}_{\rm gw}$ than in the high-SNR case, as expected.
 The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; we find $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for all parameters except $\psi$ and $h_0$.
-However it is hard to draw strong conclusions; the improvements from including the pulsar terms vary across different noise realisations.
-Whilst the inclusion of the pulsar terms improves the accuracy of the parameter estimates at low SNR only modestly, we show in Section \ref{sec:psr_v_earth_bayes} that it increases the detection probability.
+However it is hard to draw strong conclusions; the improvements from including the pulsar terms vary randomly across different noise realisations.
+Whilst including the pulsar terms improves the accuracy of the parameter estimates at low SNR modestly, we show in Section \ref{sec:psr_v_earth_bayes} that it increases the detection probability.
 \newline
@@
 
+%\begin{table}
+%	\centering
+%	%	\resizebox{\textwidth}{!}{%
+%		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
+%		\begin{tabular}{lcll}
+%			\toprule
+%			&$\theta$ & $\Delta_{\rm Earth}(\theta)$ & $\Delta_{\rm psr \& Earth}(\theta)$  \\
+%			\hline
+%			\multirow{7}{2mm}{High SNR} & $\Omega$       & $1.1 \times 10^{-2}$& $4.3 \times 10^{-3}$\\
+%			& $\Phi_0$ &$3.3 \times 10^{-3}$ &$1.2 \times 10^{-3}$ \\
+%			& $\psi$ &$9.7 \times 10^{-2}$ &$6.5 \times 10^{-4}$  \\
+%			& $\iota$ & $3.4 \times 10^{-1}$ & $1.0 \times 10^{-1}$  \\
+%			& $\delta$ &$1.9 \times10^{-3}$ &$2.9 \times 10^{-4}$  \\
+%			&$\alpha$ &$4.0 \times 10^{-2}$ &$4.4 \times10^{-4}$  \\
+%			&$h_0$ & $2.2 \times 10^{-1}$ &$7.5 \times 10^{-2}$  \\
+%			\hline
+%			\multirow{7}{2mm}{Low SNR} & $\Omega$       &  $2.7 \times 10^0$ &$1.6 \times 10^0$ \\
+%			& $\Phi_0$ &$3.4 \times 10^{-1}$ & $2.3 \times 10^{-1}$  \\
+%			& $\psi$ &$1.1 \times 10^{-1}$ &$1.2 \times 10^{-1}$  \\
+%			& $\iota$ & $4.4 \times 10^{-1}$ & $3.7 \times 10^{-1}$  \\
+%			& $\delta$ & $1.8 \times 10^{-1}$ & $7.8 \times 10^{-2}$  \\
+%			&$\alpha$ & $1.5 \times 10^{-1}$ & $1.4 \times 10^{-1}$ \\
+%			&$h_0$ & $1.1 \times 10^0$ &$1.4 \times 10^0$ \\
+%			\bottomrule
+%		\end{tabular}
+%		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = Earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M = $\rm psr \& Earth$, Equation \eqref{eq:g_func_trig}) for $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+The injected values are as summarised in Table \ref{tab:parameters_and_priors}.
+The top and bottom halves of the table contain the high-SNR ($h_0 =1 \times 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.
+The psr\&Earth model is more accurate than the Earth model for all $\theta \in \boldsymbol{\theta}_{\rm gw}$ at high SNR and five out of seven $\theta \in \boldsymbol{\theta}_{\rm gw}$ at low SNR.}
+%		\label{tab:posterior_errors}
+%	\end{table}
+
 \begin{table}
@@
 			\hline
-			\multirow{7}{2mm}{High SNR} & $\Omega$       & $1.1 \times 10^{-2}$& $4.3 \times 10^{-3}$\\
-			& $\Phi_0$ &$3.3 \times 10^{-3}$ &$1.2 \times 10^{-3}$ \\
-			& $\psi$ &$9.7 \times 10^{-2}$ &$6.5 \times 10^{-4}$  \\
-			& $\iota$ & $3.4 \times 10^{-1}$ & $1.0 \times 10^{-1}$  \\
-			& $\delta$ &$1.9 \times10^{-3}$ &$2.9 \times 10^{-4}$  \\
-			&$\alpha$ &$4.0 \times 10^{-2}$ &$4.4 \times10^{-4}$  \\
-			&$h_0$ & $2.2 \times 10^{-1}$ &$7.5 \times 10^{-2}$  \\
+			\multirow{7}{2mm}{High SNR} & $\Omega$       & $1.9 \times 10^{-5}$& $3.4 \times 10^{-6}$\\
+			& $\Phi_0$ &$1.6 \times 10^{-2}$ &$2.2 \times 10^{-3}$ \\
+			& $\psi$ &$3.9 \times 10^{-2}$ &$2.0 \times 10^{-4}$  \\
+			& $\iota$ & $2.0 \times 10^{-1}$ & $6.1 \times 10^{-2}$  \\
+			& $\delta$ &$1.9 \times10^{-3}$ &$2.0 \times 10^{-4}$  \\
+			&$\alpha$ &$4.0 \times 10^{-2}$ &$2.8 \times10^{-4}$  \\
+			&$h_0$ & $9.3 \times 10^{-2}$ &$4.9 \times 10^{-2}$  \\
 			\hline
-			\multirow{7}{2mm}{Low SNR} & $\Omega$       &  $2/7 \times 10^0$ &$1.6 \times 10^0$ \\
-			& $\Phi_0$ &$3.4 \times 10^{-1}$ & $2.3 \times 10^{-1}$  \\
-			& $\psi$ &$1.1 \times 10^{-1}$ &$1.2 \times 10^{-1}$  \\
-			& $\iota$ & $4.4 \times 10^{-1}$ & $3.7 \times 10^{-1}$  \\
-			& $\delta$ & $1.8 \times 10^{-1}$ & $7.8 \times 10^{-2}$  \\
-			&$\alpha$ & $1.5 \times 10^{-1}$ & $1.4 \times 10^{-1}$ \\
-			&$h_0$ & $1.1 \times 10^0$ &$1.4 \times 10^0$ \\
+			\multirow{7}{2mm}{Low SNR} & $\Omega$       &  $4.8 \times 10^{-3}$ &$1.5 \times 10^{-3}$ \\
+			& $\Phi_0$ &$1.4 \times 10^{0}$ & $3.3 \times 10^{-1}$  \\
+			& $\psi$ &$5.3 \times 10^{-3}$ &$1.1 \times 10^{-2}$  \\
+			& $\iota$ & $3.3 \times 10^{-1}$ & $2.1 \times 10^{-1}$  \\
+			& $\delta$ & $1.3 \times 10^{-1}$ & $4.2 \times 10^{-2}$  \\
+			&$\alpha$ & $2.5 \times 10^{-2}$ & $3.3 \times 10^{-3}$ \\
+			&$h_0$ & $1.1 \times 10^{-1}$ &$1.6 \times 10^{-1}$ \\
 			\bottomrule
 		\end{tabular}
-		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = Earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=$\rm psr \& Earth$, Equation \eqref{eq:g_func_trig}) for $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+		\caption{Relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = Earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M = $\rm psr \& Earth$, Equation \eqref{eq:g_func_trig}) for $\theta \in \boldsymbol{\theta}_{\rm gw}$.
 The injected values are as summarised in Table \ref{tab:parameters_and_priors}.
 The top and bottom halves of the table contain the high-SNR ($h_0 =1 \times 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.
-The psr\&Earth model is more accurate than the Earth model for all $\theta \in \boldsymbol{\theta}_{\rm gw}$ at high SNR and five out of seven parameters at low SNR.}
+The psr\&Earth model is more accurate than the Earth model for all $\theta \in \boldsymbol{\theta}_{\rm gw}$ at high SNR, and five out of seven $\theta \in \boldsymbol{\theta}_{\rm gw}$ at low SNR.}
 		\label{tab:posterior_errors}
@@
 The horizontal grey dashed line labels an arbitrary detection threshold, $\beta_{\rm M} = 10$.
-The minimum detectable strain for which $\beta < 10$ equals $3.2 \times 10^{-15}$ for $\mathcal{M}_{\rm Earth}$ (blue points) and $2.8 \times 10^{-15}$ for $\mathcal{M}_{\rm psr \& Earth}$ (green points).
+The minimum detectable strain at $\beta_{\rm M} = 10$ equals $3.2 \times 10^{-15}$ for $\mathcal{M}_{\rm Earth}$ (blue points) and $2.8 \times 10^{-15}$ for $\mathcal{M}_{\rm psr \& Earth}$ (green points).
 The axes are plotted on logarithmic scales.}
@@
 The measurement equation in the Kalman filter is reparameterized in terms of a pulsar-specific static phase $\chi^{(n)}$ ($1 \leq n \leq N$) to be inferred for each pulsar.
-The Kalman filter is combined with a nested sampler to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence of the signals models with and without the pulsar terms included.
+The Kalman filter is combined with a nested sampler to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence of the signal models with and without the pulsar terms included.
 \newline
@@
 The updated state-space model including the pulsar terms is tested on synthetic data.
-We consider 10 noise realisations for a single, astrophysically representative, SMBHB GW source with $h_0 = 5 \times 10^{-15}$, observed by the 12.5-year NANOGrav pulsars ($N=47$) with $T_{\rm obs} = 10 \, {\rm years}$ and $T_{\rm cad} = 1$ week.
+We start by considering 10 noise realisations for a single, astrophysically representative, SMBHB GW source with $h_0 = 5 \times 10^{-15}$, observed by the 12.5-year NANOGrav pulsars ($N=47$) with $T_{\rm obs} = 10 \, {\rm years}$ and $T_{\rm cad} = 1$ week.
 We find that the updated state-space model successfully detects injected signals and estimates their static parameters accurately for all noise realisations with relatively low computational cost.
-The irreducible random dispersion (cosmic variance) in the median of the marginalised one-dimensional posteriors, with a maximum $CV$ of 52\% for $\Phi_0$, a minimum of 0.2 \% for $\Omega$ and a median of $10 \%$ for $h_0$.
+The irreducible random dispersion (cosmic variance) in the median of the marginalised one-dimensional posteriors, with a maximum $CV$ of 52\% for $\Phi_0$, a minimum of 0.2\% for $\Omega$ and a median of $10 \%$ for $h_0$.
+\newline
+
 The updated model is further tested across a broad and astrophysically plausible parameter domain of SMBHB sources, exploring 200 randomly sampled $\boldsymbol{\theta}_{\rm gw}$ at fixed $\iota$ and $h_0$.
-Consistent parameter estimates are obtained, although the accuracy is lower, when one or more PTA pulsars coincide approximately on the sky with the SMBHB source.For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases.
-For the worst estimated parameter $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases.
+Consistent parameter estimates are obtained, although the accuracy is lower, when one or more PTA pulsars coincide approximately on the sky with the SMBHB source.
+For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases.
+For the worst estimated parameter $\Phi_0$, the injection is contained within the 90\% credible interval in 81\% of cases.
 Including the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) for five out of the seven static parameters in $\boldsymbol{\theta}_{\rm gw}$; we find $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for all parameters except $\psi$ and $h_0$.
-In the high-SNR case ($h_0 = 1 \times 10^{-12}$) all of the seven static parameters are estimated more accurately; $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$.
+In the high-SNR case ($h_0 = 1 \times 10^{-12}$) all of the seven static parameters are estimated more accurately; we find $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$.
 Including the pulsar terms lowers the minimum detectable strain by $14\%$, comparable with standard PTA analyses \citep[e.g.][]{Zhupulsarterms}.
@@
 It should be possible for a Kalman filter and nested sampler to operate together to detect the stochastic background.
-For the first case, it is straightforward to apply the Kalman filter and nested sampler with an extended range of static parameters associated with the GW sources.
-For the second case it is necessary to summarize economically the additional static parameters, whilst respecting the mathematical structure of the Kalman filter.
+With respect to the first reason, it is straightforward to apply the Kalman filter and nested sampler with an extended range of static parameters associated with the GW sources.
+With respect to the second reason, it is necessary to summarize economically the additional static parameters, whilst respecting the mathematical structure of the Kalman filter.
 This is a subtle challenge which we postpone to a forthcoming paper.
-If successful, it will complement the traditional approach of cross-correlating pulsar residuals to uncover the Hellings-Downs curve \citep{Hellings,2023ApJ...951L...8A}.
+If successful, it will complement the traditional approach of cross-correlating TOA residuals to uncover the Hellings-Downs curve \citep{Hellings,2023ApJ...951L...8A}.
 \end{enumerate}
@@
 \section{Kalman filter} \label{sec:kalman}
-The Kalman filter \citep{Kalman1} is an algorithmic optimal solver for Gauss-Markov processes.
-Given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$ the Kalman filter recovers a temporal sequence of stochastically evolving system state variables, $\boldsymbol{X}(t)$, which are hidden from the observer.
+The Kalman filter \citep{Kalman1} is an optimal solver for Gauss-Markov processes.
+Given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$, the Kalman filter recovers a temporal sequence of stochastically evolving system state variables, $\boldsymbol{X}(t)$, which are hidden from the observer.
 It is a standard method in control theory \citep{zarchan2000fundamentals,rob_KF_book,10.5555/3103280} and finds common use in engineering applications \citep[e.g.][]{KFexample1,KFexample2,KFexample3,KFexample4} as well as successful application to neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}.
 In this appendix, we outline the Kalman filter used in this paper.
-The general recursion relations for the discrete-time Kalman filter are presented for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the formula for the likelihood.
-The application of the Kalman filter algorithm to the specific continuous-time state-space model in Section \ref{sec:2} is outlined in Section \ref{sec_kalman_specific}.
+The general recursion relations for the discrete-time Kalman filter are presented for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the formula for the Bayesian likelihood.
+The application to the specific continuous-time state-space model in Section \ref{sec:2} is outlined in Section \ref{sec_kalman_specific}.
 
 \subsection{Recursion equations and likelihood }\label{sec_kalman_general}
-In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement).
+In this work we use the linear Kalman filter, which assumes linear relations between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement).
 The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k = \boldsymbol{Y}(t_k)$, which are related via a linear transformation to a set of unobservable discrete system states $\boldsymbol{X}_k= \boldsymbol{X}(t_k)$.
@@
 \end{equation}
-where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
+where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, one has $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step.
 The Kalman filter evolves the underlying states according to
@@
 
-The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage.
+The Kalman filter is a recursive estimator with two distinct stages: a ``predict'' stage and an ``update'' stage.
 The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimate from step $k-1$.
@@
 \end{eqnarray}
-where $D_k$ is the number of data points at timestep $k$. The total log-likelihood for the entire sequence is
+where $D_k$ is the dimension of $\boldsymbol{\epsilon}_k$ at timestep $k$. The total log-likelihood for the entire sequence is
 \begin{eqnarray}
-	\log \mathcal{L} =  \sum_{k=1}^{M} \log \mathcal{L}_k \ . \label{eq:likelihood}
+	\log \mathcal{L} =  \sum_{k=1}^{K} \log \mathcal{L}_k \ . \label{eq:likelihood}
 \end{eqnarray}
@@
 \begin{equation}
-	C^{(n)}(t) =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}^{(n)} _{\rm em}(t_1)\ .
+	C^{(n)}(t) =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, (t-t_1) \right] +  \dot{f}^{(n)} _{\rm em}(t_1)\ .
 \end{equation}
@@
 
-In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic}
-\begin{equation}
-	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ .
+In the idealized model in this paper, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic}
+\begin{equation}
+	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t dt'  e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') + \int_0^t d\boldsymbol{B}(t') e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma}  \ .
 \label{eq:gardenier}
@@
 \end{equation}
-where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
+where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence,
 \begin{equation}
@@
 The primary tunable parameter in nested sampling is $n_{\rm live}$.
-More live points addresses larger parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
+More live points address larger parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}\left(n_{\rm live}^{-1/2}\right)$.
 However the computational runtime scales as $\mathcal{O}(n_{\rm live})$.
@@
 Informal empirical tests conducted as part of this paper support the trade-off suggested by \cite{Ashton2022}; we find typically that the true ${\boldsymbol{\theta}}$ is contained within the 90\% credible interval of the one-dimensional marginalised posteriors of ${\boldsymbol{\hat{\theta}}}$ for $n_{\rm live} > 7 + 5N$ with $N \leq 50$.
-Unless stated otherwise we take $n_{\rm live} = 2000$ for all results presented in this work.
+Unless stated otherwise we take $n_{\rm live} = 2000$ conservatively for all results presented in this work.
 \newline
@@
 In order to compute the odds ratio $\beta$ the above workflow is repeated for a different $\mathcal{M}$.
-The resulting $\mathcal{Z}$ values can then be compared.
+The resulting $\mathcal{Z}$ values can then be divided.
 We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs.
@@
 
-For $\pi(\boldsymbol{\theta}_{\rm gw})$ we have no \textit{a priori} information on the parameters.
+For $\pi(\boldsymbol{\theta}_{\rm gw})$ we assume no \textit{a priori} information about the parameters.
 We therefore choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}.
@@
 We do so as a sanity check to confirm that the state-space analysis scheme with the pulsar terms included returns reasonable results in the ``easy'' case with a strong GW signal.
-The high-SNR limit allows us to check for estimation biases unobscured by measurement noise which is large compared to the GW signal.
+The high-SNR limit allows us to check for estimation biases unobscured by measurement noise.
 \newline
@@
 We apply the parameter estimation framework as in Section \ref{sec:pe_and_ms}.
-All injected static parameters $\boldsymbol{\theta}'$, and the associated priors are as specified in Table \ref{tab:parameters_and_priors}, with the exception of $h_0$ which now takes the value $h_0 = 1 \times 10^{-12}$.
+All injected static parameters $\boldsymbol{\theta}'$, and the associated priors are specified in Table \ref{tab:parameters_and_priors}, with the exception of $h_0$ which now takes the value $h_0 = 1 \times 10^{-12}$.
 The procedure is undertaken for ten realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$.
@@
 There is no evidence of railing against the prior bounds.
-Correlations between $\iota$ and $h_0$ are evident over multiple noise realisations due to the weak identifiability \footnote{By weak identifiability in this context we mean that the $\iota-h_0$ likelihood contours plateau to a flat-topped ridge with a small, non-zero gradient close to the maximum likelihood solution.} of these two parameters \citep{e5be7c83a0d24500826f6e1b414d1733}.
+Correlations between $\iota$ and $h_0$ are evident over multiple noise realisations due to the weak identifiability \footnote{By weak identifiability in this context we mean that the $\iota$-$h_0$ likelihood contours plateau to a flat-topped ridge with a small, non-zero gradient close to the maximum likelihood solution.} of these two parameters \citep{e5be7c83a0d24500826f6e1b414d1733}.
 For five of the seven parameters, there is effectively zero dispersion in the one-dimensional posterior medians between noise realisations.
@@
 
-Figure \ref{fig:corner_high_snr_appendix_chi} displays results a representative subset of five out of 47 pulsar phases $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ for the ten noise realisations.
+Figure \ref{fig:corner_high_snr_appendix_chi} displays results for a representative subset of five out of 47 pulsar phases $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ for the ten noise realisations.
 The figure is exactly analogous to Figure \ref{fig:corner_plot_3}.
@@
 It is now possible to infer $\chi^{(2)}$ due to the loundness of the GW signal, unlike in the low-SNR case in Figure \ref{fig:corner_plot_3}.
-The colour scheme is the same as in Figure \ref{fig:corner_plot_3}, but the curves are all overlapping and so are obscured by the orange curves.}
+The colour scheme is the same as in Figure \ref{fig:corner_plot_3}, but all the curves overlap and so are obscured by the orange curves.}
 	\label{fig:corner_high_snr_appendix_chi}
@@
 The SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ are chosen arbitrarily, with $\alpha = 1.0 $ rad.
-We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig}, for different values of $\alpha$, $ 0.0 \le \alpha \le \pi$ passed to the  Kalman filter.
+We run the Kalman filter using the measurement equation inclusive of the pulsar terms, Equation \eqref{eq:g_func_trig}, for $ 0 \le \alpha \le \pi$.
 All other elements of $\boldsymbol{\theta}_{\rm gw}$ are fixed at their true injected values.
@@
 Therefore changing $\alpha$ also changes $\boldsymbol{n}$ everywhere it appears in Equation \eqref{eq:g_func_trig}.
-For each value of $\alpha$, the Kalman filter returns a likelihood $\mathcal{L}(\boldsymbol{Y} | \alpha) = \mathcal{L}_{\rm slice}(\alpha)$, i.e.\ a cross-section of the $\mathcal {L}(\boldsymbol{\theta}_{\rm gw})$ along the $\alpha$-axis for the purpose of testing.
+For each value of $\alpha$, the Kalman filter returns a likelihood $\mathcal{L}(\boldsymbol{Y} | \alpha) = \mathcal{L}_{\rm slice}(\alpha)$, i.e.\ a cross-section of $\mathcal {L}(\boldsymbol{\theta}_{\rm gw})$ along the $\alpha$-axis for the purpose of testing.
 \newline
@@
 Second, despite being jagged, a true likelihood maximum coincident with the injected value does exist and is locally concave in a neighbourhood spanning $\sim 10^{-4} \, {\rm rad}$.
-Hence it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$ without reparameterizing in terms of $\chi^{(n)}$, but in practice the computation is intractable and does not converge for reasonable choices of $n_{\rm live}$.
-The challenge for sampling algorithms is exacerbated when we extend the above test from $\mathcal{L}(\alpha)$ tp the $\alpha$-$\delta$-$\Omega$ subspace or the full $7 + 5N$ domain in Section \ref{sec:ss_params}.
+Hence it is theoretically possible to use likelihood-based methods for Bayesian estimation of $\alpha$ without reparameterizing in terms of $\chi^{(n)}$.
+In practice, however, the computation is intractable and does not converge for reasonable choices of $n_{\rm live}$.
+The challenge for sampling algorithms is exacerbated when we extend the above test from $\mathcal{L}_{\rm slice}(\alpha)$ to the $\alpha$-$\delta$-$\Omega$ subspace or the full $7 + 5N$ domain in Section \ref{sec:ss_params}.
 \newline
@@
 Why do $\alpha$, $\delta$ and $\Omega$ exhibit jagged likelihood curves whereas other static parameters do not?
-The culprit is the Doppler factor flagged above: $\alpha$, $\delta$, and $\Omega$ appear in the phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$, with $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
+The culprit is the Doppler factor flagged above: $\alpha$, $\delta$, and $\Omega$ appear in the phase term $\Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}$, with $\boldsymbol{n} = \boldsymbol{n}(\delta, \alpha)$.
 For $d^{(n)} \sim 1$ kpc, a 10 nHz GW accumulates $ \gtrsim 10^3$ cycles.
@@
 This is the approach taken in K24 and some (not all) other standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}.
-The second is to reparameterize the pulsar terms such that the phase term $\Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}$ is no longer a function of $\alpha$ (or $\Omega$ or $\delta$), but an independent parameter $\chi^{(n)}$ to be estimated.
+The second is to reparameterize the pulsar terms such that the phase term $\Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}$ is no longer a function of $\alpha$ (or $\Omega$ or $\delta$), but an independent parameter $\chi^{(n)}$ to be estimated.
 This is the approach presented in this paper, i.e.

### 2024-02-20 manuscript/manuscript_2/paper.tex
version submitted to MNRAS

@@
 \end{eqnarray}
- where $\mu_{\rm med}$ is the mean value of the 10 posterior medians and $\sigma_{\rm med}$ is their standard deviation.
+ where $\mu_{\rm med}$ is the mean of the 10 posterior medians and $\sigma_{\rm med}$ is their standard deviation.
 The minimum and maximum $CV$ are 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively.
@@
 It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach, as explained in Section \ref{sec:pulsar_term2}.
-In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$., except under special circumstances.
+In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$, except under special circumstances.
 Figure \ref{fig:corner_plot_3} displays the results as a corner plot for the representative subset $\chi^{(1)}, \dots, \chi^{(5)}$ over the ten noise realisations, in the same style as Figure \ref{fig:corner_plot_1}.
@@
 
-Sometimes, albeit rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}), corresponding to PSR J0340+4130.
+Sometimes, albeit rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}) corresponding to PSR J0340+4130.
 Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$ and $\cos \chi^{(2)} \approx 0.07$.
@@
 The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter.
-The supertitles of the one-dimensional histograms record the median and the 0.16 and 0.84 quantiles of the posterior median realisation (in this context the median realisation is defined as follows: a posterior is generated for all 10 realisations; the medians of the posteriors are ranked in ascending order; the median of the ranked list is defined to correspond to the median realisation).
+The supertitles of the one-dimensional histograms record the median and the 0.16 and 0.84 quantiles of the median realisation.
+In this context the median realisation is defined as follows: a posterior is generated for all 10 realisations; the medians of the posteriors are ranked in ascending order; the median of the ranked list is associated with the median realisation.
 We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$.
@@
 The known, injected value lies within the 90\% credible interval for 60 out of the 70 combinations of seven parameters and ten noise realizations.
-There is appreciable dispersion among the peaks of the one-dimensional posteriors: the maximum coefficient of variation is $52 \%$ for $\Phi_0$, and the minimum is $0.2 \%$ for $\Omega$.}
+There is appreciable dispersion among the peaks of the one-dimensional posteriors: the maximum coefficient of variation is $CV = 52 \%$ for $\Phi_0$, and the minimum is $CV = 0.2 \%$ for $\Omega$.}
 	\label{fig:corner_plot_1}
@@
 The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are specified in Section \ref{sec:rep_example1} and Table \ref{tab:parameters_and_priors}.
-\newline
-
 It is prohibitive to display corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$.
@@
 More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases.
-For the worst estimated parameter $\Phi_0$, the injection is contained within the 90\% credible interval in 81\% of cases; the next paragraph explains why 81\% is less than the theoretical ideal 90\%.
+For the worst estimated parameter $\Phi_0$, the injection is contained within the 90\% credible interval in 81\% of cases.
+The next paragraph explains why 81\% is less than the theoretical ideal 90\%.
 \newline
@@
 
-For the high-SNR results in Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm Earth}$ are evidently biased, as was observed by K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
+For the high-SNR results in Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm Earth}$ are biased, as was observed by K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}.
 Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line.
@@
 The green one-dimensional posteriors exhibit no bias and are generally symmetric about the injected value.
-Moreover, for every pair of parameters in ${\boldsymbol{\theta}}_{\rm gw}$ the injected values are contained within the 2-sigma contours for $\mathcal{M}_{\rm psr \& Earth}$.
-This is not true for $\mathcal{M}_{\rm Earth}$ where the injected values fall outside the 2-sigma contour for 15 out of the 21 parameter pairs.\newline
-
-We define a relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injected value, viz.
+Moreover, for every pair of parameters in ${\boldsymbol{\theta}}_{\rm gw}$, the injected values are contained within the 2-sigma contours for $\mathcal{M}_{\rm psr \& Earth}$.
+This is not true for $\mathcal{M}_{\rm Earth}$, where the injected values fall outside the 2-sigma contour for 15 out of the 21 parameter pairs.\newline
+
+We define a relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injected value.
+The relative error is defined to be the relative unsigned displacement of the mode of $p_{\rm M}\left(\theta| \boldsymbol{Y}\right)$ from the injection, viz.
 \begin{eqnarray}
@@
 We find that the estimates are more accurate using the pulsar terms, i.e.
-$\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$.
+$\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta \in \boldsymbol{\theta}_{\rm gw}$.
 In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models, with $\Delta_{\rm Earth}(\Omega)-\Delta_{\rm psr \& Earth}(\Omega) = 1.5 \times 10^{-5}$.
@@
 Qualitatively, the green and blue contours and histograms mostly overlap, although the green contours are centred slightly better on the injected values.
-The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR results is reported in the lower half of Table \ref{tab:posterior_errors}.
+The relative error $\Delta_{\rm M}(\theta)$ for the low-SNR results is reported in the lower half of Table \ref{tab:posterior_errors}.
 We see that the error is larger for every element in $\boldsymbol{\theta}_{\rm gw}$ than in the high-SNR case, as expected.
@@
 However it is hard to draw strong conclusions; the improvements from including the pulsar terms vary randomly across different noise realisations.
-Whilst including the pulsar terms improves the accuracy of the parameter estimates at low SNR modestly, we show in Section \ref{sec:psr_v_earth_bayes} that it increases the detection probability.
-\newline
+We show in Section \ref{sec:psr_v_earth_bayes} that including the pulsar terms increases the detection probability.
+\newline
+
 \begin{figure*}
@@
 The contours in the two-dimensional histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels.
-The supertitles of the one-dimensional histograms record the posterior medians and the 0.16 and 0.84 quantiles of the green curves, i.e.
+The supertitles of the one-dimensional histograms record the medians and the 0.16 and 0.84 quantiles of the green curves, i.e.
 the pulsar-term model.
@@
 Some parameters (e.g.
-$\psi, \iota$) exhibit a bias when using the Earth-term model which disappears when the pulsar terms are included.}
+$\psi, \iota$) exhibit a bias when using the Earth-term model, which disappears when the pulsar terms are included.}
 	\label{fig:corner_plot_compare_high}
@@
 		\end{tabular}
-		\caption{Relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = Earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M = $\rm psr \& Earth$, Equation \eqref{eq:g_func_trig}) for $\theta \in \boldsymbol{\theta}_{\rm gw}$.
-The injected values are as summarised in Table \ref{tab:parameters_and_priors}.
+		\caption{Relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the mode of the one-dimensional posteriors calculated using the Earth-term model (M = Earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M = $\rm psr \& Earth$, Equation \eqref{eq:g_func_trig}) for $\theta \in \boldsymbol{\theta}_{\rm gw}$.
+The injected values are summarised in Table \ref{tab:parameters_and_priors}.
 The top and bottom halves of the table contain the high-SNR ($h_0 =1 \times 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.
@@
 In this section we compute the minimum detectable strain for the representative source in Table \ref{tab:parameters_and_priors}, using $\mathcal{M}_{\rm Earth}$ and $\mathcal{M}_{\rm psr \& Earth}$.
-We frame the problem of claiming a GW detection in terms of a Bayesian model selection procedure, following the lead of other PTA analyses \citep[e.g.][]{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X}.
+We frame the detection problem in terms of a Bayesian model selection procedure, following the lead of other PTA analyses \citep[e.g.][]{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X}.
 We define $\mathcal{M}_{\rm null}$ as the null model that assumes no GW exists in the data.
